### 2012-09-27 docs/roadmap/problems.tex
Initial problem definition

% LaTeX Template for a short article
\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym}
\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
 \pagestyle{empty} % remove page numbers

%\oddsidemargin   0.0cm
%\evensidemargin  0.0cm
%\topmargin       0.0cm
%\headheight      0.0cm
%\headsep         1.0cm
%\textheight     21.0cm
%\textwidth      16.0cm
\parskip         0.1cm
\parindent       0.0cm
%\footskip        1.0cm

\usepackage[compact]{titlesec}
\titlespacing{\section}{0pt}{8pt}{0pt}
\titlespacing{\subsection}{0pt}{8pt}{0pt}
\titlespacing{\subsubsection}{0pt}{8pt}{0pt}
\usepackage{enumitem}
\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}
\setenumerate{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}
\def\ul{\begin{itemize}}
\def\ule{\end{itemize}}
\def\ol{\begin{enumerate}}
\def\ole{\end{enumerate}}

\title{DP problems of interest}
\author{Manohar, Thierry}
\date{}
\begin{document}
\maketitle
\pagestyle{headings}

\section{Introduction}
\subsection{Smith-Waterman}
Problem: matching two strings $S$, $T$.

Alphabets:\ul
\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
\ule

Initialization:\ul
\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
\ule

Recurrence:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	M_{(i-1,j)}-d & N\\
	M_{(i,j-1)}-d & W
\end{array}\right\}=B_{(i,j)} \]

\newpage
\subsection{Nussinov algorithm}
Problem: folding a RNA string $S$ over itself.

Alphabets:\ul
\item Input: $\Sigma(S)=\{A,C,G,U\}$.
\item Cost matrix: $\Sigma(M)=\{0..n\}, n={\rm length}(S)/2$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,S,SW, 1..n\}$
\ule

Initialization: \ul
\item Cost matrix: $\left\{\begin{array}{l} M_{(i,i)}=0 \\ M_{(i,i-1)}=0 \end{array}\right. \forall i \in 1..{\rm length}(S)$
\item Backtrack matrix: $\left\{\begin{array}{l} B_{(i,i)}=stop \\ B_{(i,i-1)}=stop \end{array}\right.  \forall i \in 1..{\rm length}(S)$
\ule

Recurrence:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	M_{(i+1,j-1)}+\omega(i,j) & SW\\
	M_{(i+1,j)} & S\\
	M_{(i,j-1)} & W\\
	\max_{i<k<j}M_{(i,k)}+W_{(k+1,j)} & k
\end{array}\right\} = B_{(i,j)} \]
With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.

% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
%\newpage
%\subsection{Zuker folding}
%Problem: folding a RNA string $S$ over itself.
%
%Alphabets:\ul
%\item Input: $\Sigma(S)=\{A,C,G,U\}$.
%\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
%\item Backtrack matrix: $\Sigma(B)=\{\}$
%\ule
%
%Initialization:\ul
%\item Cost matrices:\ul
%	\item $\Sigma(W) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
%	\item x
%	\item x
%\ule
%\item Backtrack matrix
%\ule
%
%Recurrence:
%\[\begin{array}{rcl}
%W_{(i,j)}&=&\min\left\{\begin{array}{l}
%	W_{(i+1,j)}+b \\
%	W_{(i,j-1)}+b \\
%	V_{(i,j)}+\delta(S_i,S_j) \\
%	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
%\end{array}\right.\\\\
%V_{(i,j)}&=&\min\left\{\begin{array}{ll}
%	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
%	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
%	V_{(i+1,j-1)}+es(i,j) \\
%	VBI(i,j) \\
%	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
%\end{array}\right.\\\\
% F_{(j)}&=&\min\left\{\begin{array}{ll}
%F_{(j-i)} \\
%\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
%\end{array}\right. \text{(Free Energy)}\\\\
%VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
%\end{array}\]
%
%XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...

\end{document}

### 2012-09-28 docs/roadmap/problems.tex
Nits.
ideas and problems
more problems, more ideas
Some specs for problems.

@@
 
-%\oddsidemargin   0.0cm
-%\evensidemargin  0.0cm
-%\topmargin       0.0cm
-%\headheight      0.0cm
-%\headsep         1.0cm
-%\textheight     21.0cm
-%\textwidth      16.0cm
+\oddsidemargin   0.0cm
+\evensidemargin  0.0cm
+\topmargin       0.0cm
+\headheight      0.0cm
+\headsep         1.0cm
+\textheight     21.0cm
+\textwidth      16.0cm
 \parskip         0.1cm
 \parindent       0.0cm
-%\footskip        1.0cm
+\footskip        1.0cm
 
@@
 
+% ----------------------------------------------------------------------------------------------------------------------------------------
 \section{Introduction}
+\subsection{Definitions}\ul
+\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
+\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
+\ule
+
+\subsection{Problems classification}
+In the literature dynamic programming problems (DP) are classified according to two criteria:
+
+\textbf{Monadic / polyadic}\ul
+\item Monadic: on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
+	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
+\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
+\ule
+
+\textbf{Serial/non-serial} \ul
+\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial.
+\item \textbf{Non-serial:} when the solution of a level depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
+	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
+\ule
+
+\subsection{Simplifications}
+In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example, we can transform
+	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
+	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
+	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
+Where $C$ is a matrix that stores the maximum along the column and $A$ is a matrix that stores the sum of the array of the previous elements.
+They can be easily computed with the additional recurrence.
+	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
+	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]
+
+This simplification avoids the non-serial dependencies at the cost of an additional storage at the wavefront.
+
+% ----------------------------------------------------------------------------------------------------------------------------------------
+\section{Problems of interest}
+We describe problems structures: inputs, cost matrices, backtracking matrix.
+They all have an alphabet (that hopefully corresponds to a maximal bit-size) and matrixes dimensions are specified by number of indices and their number of elements is usually the same as the input.
+Unless otherwise specified, number are all unsigned integers.
+We describe the problem processing in terms of both initialization and recurrences.
+
+% ----------------------------------------------
 \subsection{Smith-Waterman}
@@
 
+% ----------------------------------------------
+\subsection{Smith-Waterman with gap extension at different cost}
+Problem: matching two strings $S$, $T$.
+
+Alphabets:\ul
+\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
+\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
+\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
+\ule
+
+Initialization:\ul
+\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
+\item Gap opening matrix: $E_{(i,0)}= 0, 0 \le i \le {\rm length}(S)$
+\item Gap extending matrix: $E_{(0,j)}= 0, 0 \le j \le {\rm length}(T)$
+\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
+\ule
+
+Recurrence for the cost matrix:
+\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+	0 & stop\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	E_{(i,j)} & N\\
+	F_{(i,j)} & W
+\end{array}\right\}=B_{(i,j)} \]
+
+Recurrence for the gap opening/extending matrices:
+\[E_{(i,j)}=\max\left\{\begin{array}{l|l}
+	M_{(i, j-1)} - \alpha & NW\\
+	E_{(i,j-1)} - \beta & N\\
+\end{array}\right\}=B_{(i,j)} \]
+
+\[F_{(i,j)}=\max\left\{\begin{array}{l|l}
+	M_{(i-1, j)} - \alpha & NW\\
+	E_{(i-1,j)} - \beta & N\\
+\end{array}\right\}=B_{(i,j)} \]
+
+Otherwise written as:
+\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+	0 & stop\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
+	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
+\end{array}\right\}=B_{(i,j)} \]
+
+% ----------------------------------------------
 \newpage
@@
 
-% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
+% ------------------------------------------------------------------------------------------------
 %\newpage
 %\subsection{Zuker folding}
+% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
 %Problem: folding a RNA string $S$ over itself.
@@
 %\item Cost matrices:\ul
-%	\item $\Sigma(W) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
-%	\item x
-%	\item x
+%	\item xxx
 %\ule
-%\item Backtrack matrix
+%\item Backtrack matrix: $B$
 %\ule

### 2012-09-28 docs/roadmap/problems.tex
Nits in problems.

@@
 \title{DP problems of interest}
-\author{Manohar, Thierry}
+\author{Manohar Jonnalagedda, Thierry Coppey}
 \date{}
@@
 \subsection{Simplifications}
+\subsubsection{Calculus}
 In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example, we can transform
@@
 
+\subsubsection{Precomputations}
+When a calculus computation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
+For instance, for Nussinov's $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$ we can precompute it for all rows and columns of the block, and for all elements that are not part of the block, and pass these partial results together at the block launch.
+
+On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
+
 % ----------------------------------------------------------------------------------------------------------------------------------------
@@
 % ----------------------------------------------
-\subsection{Smith-Waterman}
+\subsection{Smith-Waterman (simple)}
 Problem: matching two strings $S$, $T$.
@@
 % ----------------------------------------------
-\subsection{Smith-Waterman with gap extension at different cost}
+\subsection{Smith-Waterman (with gap extension at different cost)}
 Problem: matching two strings $S$, $T$.

### 2012-09-28 docs/roadmap/problems.tex
more problem instances

@@
 % ----------------------------------------------
+\subsection{Convex polygon triangulation}
+Problem: triangulating a polygon with least cost overall for added edges
+
+Alphabets:\ul
+\item Input: $\Sigma(S)= $ a matrix of costs for every possible edge in the polygon. The 
+size of the input is $\frac{n^2}{2}$, where $n$ is the number of vertices in the polygon. 
+\item Cost matrix: $\Sigma(M)=\mathbb{N}$. The sum of cost of edges is not "bounded".
+\item Backtrack matrix: $\Sigma(B)=\{n\}$
+\ule
+
+Recurrence:
+\[M_{(i,j)}= \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + c(i,k) | k = B_{(i,j)} \]
+
+It is interesting to note that even in the sequential world, this problem is best solved 
+by filling the diagonals, ie. computing sub-solutions for all polygons on size $k$ before
+those of size $k+1$.
+
+% ----------------------------------------------
+
+% ----------------------------------------------
 \newpage
@@
 	M_{(i,j-1)} & W\\
-	\max_{i<k<j}M_{(i,k)}+W_{(k+1,j)} & k
+	\max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} & k
 \end{array}\right\} = B_{(i,j)} \]
@@
 % ------------------------------------------------------------------------------------------------
-%\newpage
-%\subsection{Zuker folding}
-% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
-%Problem: folding a RNA string $S$ over itself.
-%
-%Alphabets:\ul
-%\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-%\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
-%\item Backtrack matrix: $\Sigma(B)=\{\}$
-%\ule
-%
-%Initialization:\ul
-%\item Cost matrices:\ul
-%	\item xxx
-%\ule
-%\item Backtrack matrix: $B$
-%\ule
-%
-%Recurrence:
-%\[\begin{array}{rcl}
-%W_{(i,j)}&=&\min\left\{\begin{array}{l}
-%	W_{(i+1,j)}+b \\
-%	W_{(i,j-1)}+b \\
-%	V_{(i,j)}+\delta(S_i,S_j) \\
-%	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
-%\end{array}\right.\\\\
-%V_{(i,j)}&=&\min\left\{\begin{array}{ll}
-%	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
-%	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
-%	V_{(i+1,j-1)}+es(i,j) \\
-%	VBI(i,j) \\
-%	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
-%\end{array}\right.\\\\
-% F_{(j)}&=&\min\left\{\begin{array}{ll}
-%F_{(j-i)} \\
-%\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
-%\end{array}\right. \text{(Free Energy)}\\\\
-%VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
-%\end{array}\]
-%
+\newpage
+\subsection{Zuker folding}
+XXX: we let this aside for the moment as the recurrence has too many unknown parameters
+Problem: folding a RNA string $S$ over itself.
+
+Alphabets:\ul
+\item Input: $\Sigma(S)=\{A,C,G,U\}$.
+\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
+\item Backtrack matrix: $\Sigma(B)=\{\}$
+\ule
+
+Initialization:\ul
+\item Cost matrices:\ul
+	\item xxx
+\ule
+\item Backtrack matrix: $B$
+\ule
+
+Recurrence:
+\[\begin{array}{rcl}
+W_{(i,j)}&=&\min\left\{\begin{array}{l}
+	W_{(i+1,j)}+b \\
+	W_{(i,j-1)}+b \\
+	V_{(i,j)}+\delta(S_i,S_j) \\
+	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
+\end{array}\right.\\\\
+
+V_{(i,j)}&=&\min\left\{\begin{array}{ll}
+	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
+	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
+	V_{(i+1,j-1)}+es(i,j) \\
+	VBI(i,j) \\
+	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c
+\end{array}\right.\\\\
+
+VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c \\\\
+
+F_{(j)}&=&\min\left\{\begin{array}{ll}
+F_{(j-i)} \\
+\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
+\end{array}\right. \text{(Free Energy)}\\\\
+
+\end{array}\]
+
+\subsection{Comments}
+
+VBI stores scores for internal loops and bulges.
+Notice in VBI, two moving indices $i', j'$.
+In general this gives a quadratic complexity for that particular recurrence, making the overall complexity of the algorithm $O(n^4)$.
+In practice (as read in Arpith's work), we have that $i' - i + j'- j < 30$, so the complexity is reduced to $O(n^3)$, albeit with a large constant.
+
 %XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...

### 2012-09-28 docs/roadmap/problems.tex
Added matrix mul, init plan.

@@
 % ----------------------------------------------
-\subsection{Convex polygon triangulation}
-Problem: triangulating a polygon with least cost overall for added edges
+\newpage
+\subsection{Matrix chain multiplication}
+Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$.
 
 Alphabets:\ul
-\item Input: $\Sigma(S)= $ a matrix of costs for every possible edge in the polygon. The 
-size of the input is $\frac{n^2}{2}$, where $n$ is the number of vertices in the polygon. 
-\item Cost matrix: $\Sigma(M)=\mathbb{N}$. The sum of cost of edges is not "bounded".
-\item Backtrack matrix: $\Sigma(B)=\{n\}$
+\item Input: matrix size is defined as pairs of integers $(r_i,c_i)$.
+\item Cost matrix: the cost is an integer (might blow up, use a float or a double?).
+\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$ with $n$ the input length.
 \ule
 
-Recurrence:
-\[M_{(i,j)}= \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + c(i,k) | k = B_{(i,j)} \]
+Initialization (line=length of sequence, column=start):\ul
+\item Cost matrix: $M_{(0,j)}=0$.
+\item Backtrack matrix: $B_{(0,j)}=stop$.
+\ule
 
-It is interesting to note that even in the sequential world, this problem is best solved 
-by filling the diagonals, ie. computing sub-solutions for all polygons on size $k$ before
-those of size $k+1$.
-
-% ----------------------------------------------
+Recurrence for the cost matrix:
+\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
+	M_{(i,k)}+M_{(k,j)}+r_i \cdot r_k \cdot col_j & k
+\end{array}\right\}=B_{(i,j)} \]
 
@@
 	M_{(i,j-1)} & W\\
-	\max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} & k
+	\max_{i<k<j}M_{(i,k)}+W_{(k+1,j)} & k
 \end{array}\right\} = B_{(i,j)} \]
@@
 % ------------------------------------------------------------------------------------------------
-\newpage
+%\newpage
 \subsection{Zuker folding}
@@
 \end{array}\right.\\\\
-
 V_{(i,j)}&=&\min\left\{\begin{array}{ll}
@@
 	VBI(i,j) \\
-	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c
+	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
 \end{array}\right.\\\\
-
-VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c \\\\
-
  F_{(j)}&=&\min\left\{\begin{array}{ll}
@@
 \end{array}\right. \text{(Free Energy)}\\\\
-
+VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
 \end{array}\]
-
-\subsection{Comments}
-
-VBI stores scores for internal loops and bulges.
-Notice in VBI, two moving indices $i', j'$.
-In general this gives a quadratic complexity for that particular recurrence, making the overall complexity of the algorithm $O(n^4)$.
-In practice (as read in Arpith's work), we have that $i' - i + j'- j < 30$, so the complexity is reduced to $O(n^3)$, albeit with a large constant.
-
+%
 %XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...

### 2012-09-29 docs/roadmap/problems.tex
Nits.

@@
 % ----------------------------------------------
-\newpage
 \subsection{Nussinov algorithm}
@@
 % ------------------------------------------------------------------------------------------------
-%\newpage
+\newpage
 \subsection{Zuker folding}
- XXX: we let this aside for the moment as the recurrence has too many unknown parameters
 Problem: folding a RNA string $S$ over itself.
@@
 \item Cost matrices:\ul
-	\item xxx
+	\item XXX: $W$,$V$
+	\item $F_{(0)}=0$
 \ule
@@
 \[\begin{array}{rcl}
-W_{(i,j)}&=&\min\left\{\begin{array}{l}
-	W_{(i+1,j)}+b \\
-	W_{(i,j-1)}+b \\
-	V_{(i,j)}+\delta(S_i,S_j) \\
-	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
-\end{array}\right.\\\\
+W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
+	W_{(i+1,j)}+b & S\\
+	W_{(i,j-1)}+b & W\\
+	V_{(i,j)}+\delta(S_i,S_j) & ?V? \\
+	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} & ?k?
+\end{array}\right.\\
+\\
 V_{(i,j)}&=&\min\left\{\begin{array}{ll}
@@
 	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
-\end{array}\right.\\\\
- F_{(j)}&=&\min\left\{\begin{array}{ll}
-F_{(j-i)} \\
-\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
+\end{array}\right.\\
+\\
+ F_{(j)}&=&\min\left\{\begin{array}{ll} F_{(j-1)} \\ \min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
+%
 \end{array}\right. \text{(Free Energy)}\\\\
@@
 %
-%XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...
 

### 2012-09-29 docs/roadmap/problems.tex
More ideas, reworking of first problem (need to do the same for the 3 last).

@@
 \documentclass[11pt]{article}
-\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym}
+\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs}
 \hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
+\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
 \usepackage[utf8]{inputenc}
@@
 
+% Recurrence visualisation
+\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
+\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
+\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
+\def\Cm{\put(6.5,1.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
+
 \title{DP problems of interest}
@@
 \subsection{Problems classification}
-In the literature dynamic programming problems (DP) are classified according to two criteria:
+In the literature, dynamic programming problems (DP) are classified according to two criteria:
 
-\textbf{Monadic / polyadic}\ul
-\item Monadic: on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
+\textbf{Monadic/polyadic}\ul
+\item \textbf{Monadic:} on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
 	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
@@
 \textbf{Serial/non-serial} \ul
-\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial.
-\item \textbf{Non-serial:} when the solution of a level depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
+\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial (it accesses only 2 cells backward).
+\item \textbf{Non-serial:} when the solution depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
 	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
@@
 \subsubsection{Calculus}
-In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example, we can transform
+In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
 	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
@@
 	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
-Where $C$ is a matrix that stores the maximum along the column and $A$ is a matrix that stores the sum of the array of the previous elements.
-They can be easily computed with the additional recurrence.
+Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
 	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
@@
 
-This simplification avoids the non-serial dependencies at the cost of an additional storage at the wavefront.
+This simplification avoids the non-serial dependencies at the cost of extra storage in the wavefront, unfortunately, it might be applicable only for some special cases.
 
 \subsubsection{Precomputations}
-When a calculus computation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
-For instance, for Nussinov's $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$ we can precompute it for all rows and columns of the block, and for all elements that are not part of the block, and pass these partial results together at the block launch.
+When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
+For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.
 
 On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
+We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).
 
@@
 \section{Problems of interest}
-We describe problems structures: inputs, cost matrices, backtracking matrix.
-They all have an alphabet (that hopefully corresponds to a maximal bit-size) and matrixes dimensions are specified by number of indices and their number of elements is usually the same as the input.
-Unless otherwise specified, number are all unsigned integers.
+We describe problems structures: inputs, cost matrices and backtracking matrix.
+These all have an alphabet (that must be bounded in terms of bit-size).
+Unless otherwise specified, we adopt the following conventions:\ul
+\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length.
+\item Number are all unsigned integers
+\item Problem dimension is $m,n$ (or $n$) indices $i,j$ ranges are respectively $0\le i<m$, $0\le j<n$.
+\item Unless otherwise specified, the recurrence applies to all non-initialized matrix elements.
+\ule
 We describe the problem processing in terms of both initialization and recurrences.
@@
 % ----------------------------------------------
+\newpage
 \subsection{Smith-Waterman (simple)}
-Problem: matching two strings $S$, $T$.
+Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+
+Matrices: $M_{m \times n}, B_{m \times n}$
 
@@
 \item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
+\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
 \item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
@@
 
+Backtracking: starts from the cell $M_{(m,j)} \cup M_{(i,n)} $
+
+Visualisation:
+\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
+	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
+	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
+	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
+	\Cd{6,2}{0.8}
+	\Cd[1,0]{5,1}{0.8}
+	\Cd[1,-1]{5,2}{0.8}
+\Cm\end{picture}\end{center}
+
+Optimizations\ul
+\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built).
+This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
+\ule
+
 % ----------------------------------------------
+\newpage
 \subsection{Smith-Waterman (with gap extension at different cost)}
-Problem: matching two strings $S$, $T$.
+Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+
+Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
 
@@
 \item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
+\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
 \item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
@@
 Initialization:\ul
-\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-\item Gap opening matrix: $E_{(i,0)}= 0, 0 \le i \le {\rm length}(S)$
-\item Gap extending matrix: $E_{(0,j)}= 0, 0 \le j \le {\rm length}(T)$
+\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
+\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
+\item S-gap extension cost matrix: $F_{(0,j)}= 0$
 \item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
@@
 
-Recurrence for the cost matrix:
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+Recurrence for the cost matrices:
+\[\begin{array}{rcl}
+M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
 	0 & stop\\
@@
 	F_{(i,j)} & W
-\end{array}\right\}=B_{(i,j)} \]
-
-Recurrence for the gap opening/extending matrices:
-\[E_{(i,j)}=\max\left\{\begin{array}{l|l}
+\end{array}\right\}=B_{(i,j)}\\
+\\
+E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
 	M_{(i, j-1)} - \alpha & NW\\
 	E_{(i,j-1)} - \beta & N\\
-\end{array}\right\}=B_{(i,j)} \]
+\end{array}\right\}=B_{(i,j)}\\
+\\
+F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
+	M_{(i-1,j)} - \alpha & NW\\
+	F_{(i-1,j)} - \beta & W\\
+\end{array}\right\}=B_{(i,j)}
+\end{array}\]
 
-\[F_{(i,j)}=\max\left\{\begin{array}{l|l}
-	M_{(i-1, j)} - \alpha & NW\\
-	E_{(i-1,j)} - \beta & N\\
-\end{array}\right\}=B_{(i,j)} \]
+{\color{red} XXX: Are we sure we do not need to maintain separate backtrack tables for the gap extensions?}
 
@@
 
+{\color{red} XXX: how to simplify this relation, can we move the max in the wavefront?}
+
+Recurrence visualisation:
+
+{\color{red} XXX: todo}
+
+Optimizations:
+
+{\color{red} XXX: todo}
+
 % ----------------------------------------------
+\newpage
 \subsection{Convex polygon triangulation}
@@
 \subsection{Matrix chain multiplication}
-Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$.
+Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$ \textit{(r=rows, c=columns)}.
 
@@
 
+%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
+%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
+%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
+%$\begin{array}{|C|C|C|C|C|C|} \hline
+%0 & 0 & 0 & 0 & 0 & 0 \mnl
+%0 &  &  &  &  & \mnl
+%0 & M_{23}  &  &  &  & \mnl
+%0 &  & \sum  &  &  & \mnl
+%0 &  &  &  &  & \mnl
+%0 &  &  &  &  & \mnl
+%\end{array}$
+
 \end{document}

### 2012-09-30 docs/roadmap/problems.tex
More cleanup of problems.

@@
 \documentclass[11pt]{article}
-\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs}
+\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
 \hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
@@
 \def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
-\def\Cm{\put(6.5,1.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
+\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
 
@@
 \newpage
-\subsection{Smith-Waterman (simple)}
-Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
-
-Matrices: $M_{m \times n}, B_{m \times n}$
-
-Alphabets:\ul
+\subsection{Smith-Waterman (simple)}\label{sswat}\ol
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Matrices: $M_{m \times n}, B_{m \times n}$
+\item Alphabets:\ul
 	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
@@
 	\ule
-
-Initialization:\ul
+\item Initialization:\ul
 	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
@@
 	\ule
-
-Recurrence:
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
 		0 & stop\\
@@
 
-Backtracking: starts from the cell $M_{(m,j)} \cup M_{(i,n)} $
-
-Visualisation:
+\item Backtracking: starts from the cell $M_{(m,j)} \cup M_{(i,n)} $
+\item Visualisation: by convention, we put the longest string vertically ($m\ge n$):
 \begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-	\Cd{6,2}{0.8}
-	\Cd[1,0]{5,1}{0.8}
-	\Cd[1,-1]{5,2}{0.8}
+	\Cd{6,5}{0.8}
+	\Cd[1,0]{5,4}{0.8}
+	\Cd[1,-1]{5,5}{0.8}
 \Cm\end{picture}\end{center}
 
-Optimizations\ul
+\item Optimizations:\ul
 	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built).
@@
 	\ule
+\ole
 
@@
 \newpage
-\subsection{Smith-Waterman (with gap extension at different cost)}
-Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
-
-Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
-
-Alphabets:\ul
+\subsection{Smith-Waterman (with gap extension at different cost)}\ol
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
+\item Alphabets:\ul
 	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
@@
 	\ule
-
-Initialization:\ul
+\item Initialization:\ul
 	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
@@
 	\ule
-
-Recurrence for the cost matrices:
+\item Recurrence for the cost matrices:
 \[\begin{array}{rcl}
@@
 
-{\color{red} XXX: Are we sure we do not need to maintain separate backtrack tables for the gap extensions?}
-
-Otherwise written as:
+That can be written alternatively as:
 \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
@@
 
-{\color{red} XXX: how to simplify this relation, can we move the max in the wavefront?}
+Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is 
+\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
+Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
 
-Recurrence visualisation:
-
-{\color{red} XXX: todo}
-
-Optimizations:
-
-{\color{red} XXX: todo}
+\item Backtracking: same as \nameref{sswat}
+\item Visualisation: same as \nameref{sswat}
+\item Optimizations: same as \nameref{sswat}
+\ole
 
@@
 \newpage
-\subsection{Convex polygon triangulation}
-Problem: triangulating a polygon with least cost overall for added edges
+\subsection{Convex polygon triangulation}\ol
+\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
+We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is a lower triangular matrix compacted in memory (rows are contiguous) with a 0 diagonal that is omitted \footnote{Arbitrary convention for both architectural implementation and code generator.
+Rationale: in lower triangular matrix, element address is independent of the matrix size.}, hence $|S|=\tfrac{n^2}{2}=N$.
 
-Alphabets:\ul
-\item Input: $\Sigma(S)= $ a matrix of costs for every possible edge in the polygon. The 
-size of the input is $\frac{n^2}{2}$, where $n$ is the number of vertices in the polygon. 
-\item Cost matrix: $\Sigma(M)=\mathbb{N}$. The sum of cost of edges is not "bounded".
-\item Backtrack matrix: $\Sigma(B)=\{n\}$
+% ------------------------------------------------------------------------------------------
+% ------------------------------------------------------------------------------------------
+% ------------------------------------------------------------------------------------------
+{\color{red}
+\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
+\item Alphabets:\ul
+	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
+	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
+	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
 	\ule
-
-Recurrence:
-\[M_{(i,j)}= \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + c(i,k) | k = B_{(i,j)} \]
-
+\item Initialization: $M_{(i,i)}=0, B_{(i,i)}=stop \quad\forall i$
+\item Recurrence: \[M_{(i,j)}=\left\{ \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + S(i,k) \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
 	It is interesting to note that even in the sequential world, this problem is best solved 
@@
 	those of size $k+1$.
+\item Backtracking: Starting at $B_{(0,n)}$ and let current point you backtrack drawing an edge from the current position to the backtracking position ??
 
-% ----------------------------------------------
+\item Visualisation: in its traditional form the matrix is upper triangular with diagonal excluded.
+\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
+	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
+	\Cg{0,6}
+	\Cg{0,5} \Cg{1,5}
+	\Cg{0,4} \Cg{1,4} \Cg{2,4}
+	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
+	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
+	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
+	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}
+
+	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
+	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
+	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
+	\put(3.5,4.5){\line(3,-1){3}}
+	\put(4.5,4.5){\line(2,-2){2}}
+	\put(5.5,4.5){\line(1,-3){1}}
+
+	%\Cd[1,-1]{5,2}{0.8}
+
+\Cm\end{picture}\end{center}
+
+\item Optimizations: XXX we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
+}
+\ole
 
@@
 \newpage
-\subsection{Matrix chain multiplication}
-Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$ \textit{(r=rows, c=columns)}.
-
-Alphabets:\ul
-\item Input: matrix size is defined as pairs of integers $(r_i,c_i)$.
-\item Cost matrix: the cost is an integer (might blow up, use a float or a double?).
-\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$ with $n$ the input length.
+\subsection{Matrix chain multiplication}\ol
+\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ \textit{<<r=rows, c=columns>>}. 
+\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
+\item Alphabets:\ul
+	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
+	\item Cost matrix: $\Sigma(M):=$ huge integer\footnote{Integer multiplication might blow up, use float or doubles and addition of logarithms instead.}.
+	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
+	\ule
+\item Initialization:\ul
+	\item Cost matrix: $M_{(i,i)}=0$.
+	\item Backtrack matrix: $B_{(i,i)}=stop$.
 	\ule
 
-Initialization (line=length of sequence, column=start):\ul
-\item Cost matrix: $M_{(0,j)}=0$.
-\item Backtrack matrix: $B_{(0,j)}=stop$.
-\ule
-
-Recurrence for the cost matrix:
+{\color{red}
+\item Recurrence for the cost matrix:
 	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
@@
 	\end{array}\right\}=B_{(i,j)} \]
+\item Backtracking: XXX
+\item Visualisation: XXX
+
+\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
+	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
+	\Cg{0,6}
+	\Cg{0,5} \Cg{1,5}
+	\Cg{0,4} \Cg{1,4} \Cg{2,4}
+	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
+	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
+	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
+	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}
+
+	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
+	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
+	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
+	\put(3.5,4.5){\line(3,-1){3}}
+	\put(4.5,4.5){\line(2,-2){2}}
+	\put(5.5,4.5){\line(1,-3){1}}
+
+	%\Cd[1,-1]{5,2}{0.8}
+
+\Cm\end{picture}\end{center}
+
+\item Optimizations: XXX
+}
+\ole
 
 % ----------------------------------------------
-\subsection{Nussinov algorithm}
-Problem: folding a RNA string $S$ over itself.
-
-Alphabets:\ul
+\newpage
+\subsection{Nussinov algorithm}\ol
+\item Problem: folding a RNA string $S$ over itself.
+{\color{red}
+\item Matrices: XXX
+\item Alphabets:\ul
 	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
@@
 
-Initialization: \ul
+\item Initialization: \ul
 	\item Cost matrix: $\left\{\begin{array}{l} M_{(i,i)}=0 \\ M_{(i,i-1)}=0 \end{array}\right. \forall i \in 1..{\rm length}(S)$
@@
 
-Recurrence:
+\item Recurrences:
 	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
@@
 	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
+\item Backtracking: XXX
+\item Visualisation: XXX
+\item Optimizations: XXX
+}
+\ole
 
@@
 \newpage
-\subsection{Zuker folding}
-Problem: folding a RNA string $S$ over itself.
+\subsection{Zuker folding}\ol
+\item Problem: folding a RNA string $S$ over itself.
+{\color{red}
+\item Matrices: XXX
 
-Alphabets:\ul
+\item Alphabets:\ul
 \item Input: $\Sigma(S)=\{A,C,G,U\}$.
@@
 
-Initialization:\ul
+\item Initialization:\ul
 	\item Cost matrices:\ul
@@
 
-Recurrence:
+\item Recurrence:
 \[\begin{array}{rcl}
@@
 \end{array}\]
-%
+
+\item Backtracking: XXX
+\item Visualisation: XXX
+\item Optimizations: XXX
+}
+\ole
 
@@
 %\end{array}$
-
 \end{document}

### 2012-09-30 docs/roadmap/problems.tex
More problems.

@@
-% LaTeX Template for a short article
 \documentclass[11pt]{article}
@@
 
-% Recurrence visualisation
+% Recurrence visualisation helpers
 \newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
@@
 \def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
+\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
+	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
+\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
 
-\title{DP problems of interest}
+\title{DP Problems of Interest}
 \author{Manohar Jonnalagedda, Thierry Coppey}
@@
 \newpage
+\subsection{Smith-Waterman with arbitrary gap cost (3D)}\ol
+
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function.
+Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.
+Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
+
+\item Matrices: $M_{m \times n \times m}, B_{m \times n \times m}$
+\item Alphabets:\ul
+	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
+	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
+	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
+	\ule
+
+{\color{red}
+\item Initialization: XXX: we build this, because we go 3D.
+\item Recurrences:
+\item Backtracking:
+\item Visualisation:
+\item Optimizations:
+}
+\ole
+
+% ----------------------------------------------
+\newpage
 \subsection{Convex polygon triangulation}\ol
@@
 
-% ------------------------------------------------------------------------------------------
-% ------------------------------------------------------------------------------------------
-% ------------------------------------------------------------------------------------------
 {\color{red}
@@
 	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cg{0,6}
-	\Cg{0,5} \Cg{1,5}
-	\Cg{0,4} \Cg{1,4} \Cg{2,4}
-	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
-	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
-	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
-	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}
-
-	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
+	\Cfl{\Cg}\Cfd{\Cz}
 	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 	\put(5.5,4.5){\line(1,-3){1}}
-
 	%\Cd[1,-1]{5,2}{0.8}
-
 \Cm\end{picture}\end{center}
@@
 \subsection{Matrix chain multiplication}\ol
-\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$. Each matrix $A_i$ is of dimension $r_i \times c_i$ \textit{<<r=rows, c=columns>>}. 
+\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$.
+Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$.
+\textit{<<r=rows, c=columns>>}
 \item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
@@
 	\ule
+\item Recurrence: $c_k=r_{k+1}$
+	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
+		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
+	\end{array}\right\}=B_{(i,j)} \]
+\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
+	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
+		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
 
-{\color{red}
-\item Recurrence for the cost matrix:
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
-		M_{(i,k)}+M_{(k,j)}+r_i \cdot r_k \cdot col_j & k
-	\end{array}\right\}=B_{(i,j)} \]
-\item Backtracking: XXX
-\item Visualisation: XXX
-
+\item Visualisation:
 	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cg{0,6}
-	\Cg{0,5} \Cg{1,5}
-	\Cg{0,4} \Cg{1,4} \Cg{2,4}
-	\Cg{0,3} \Cg{1,3} \Cg{2,3} \Cg{3,3}
-	\Cg{0,2} \Cg{1,2} \Cg{2,2} \Cg{3,2} \Cg{4,2}
-	\Cg{0,1} \Cg{1,1} \Cg{2,1} \Cg{3,1} \Cg{4,1} \Cg{5,1}
-	\Cg{0,0} \Cg{1,0} \Cg{2,0} \Cg{3,0} \Cg{4,0} \Cg{5,0} \Cg{6,0}
-
-	\Cz{0,7}\Cz{1,6}\Cz{2,5}\Cz{3,4}\Cz{4,3}\Cz{5,2}\Cz{6,1}\Cz{7,0}
+		\Cfl{\Cg}\Cfd{\Cz}
 		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
 		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-	\put(3.5,4.5){\line(3,-1){3}}
-	\put(4.5,4.5){\line(2,-2){2}}
-	\put(5.5,4.5){\line(1,-3){1}}
-
-	%\Cd[1,-1]{5,2}{0.8}
-
+		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
 	\Cm\end{picture}\end{center}
 
-\item Optimizations: XXX
-}
+\item Optimizations:\ul
+	\item {\color{red}We need to swap vertically the matrix to have a normalized progression towards bottom right.
+To do that, we need to map all indices $i \mapsto n-1-i$, but since we want to store the matrix sparsely, we might want to transform it into a lower triangular matrix as we can provide a size-independent mapping of element indices.
+	XXX: what's the best trade off ? progress towards bottom right VS map indices more efficiently ?}
+	\ule
 \ole
@@
 \subsection{Nussinov algorithm}\ol
-\item Problem: folding a RNA string $S$ over itself.
-{\color{red}
-\item Matrices: XXX
+\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
+\item Matrices: $M_{n\times n}, B_{n \times n}$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrix: $\Sigma(M)=\{0..n\}, n={\rm length}(S)/2$
+	\item Cost matrix: $\Sigma(M)=\{0..n\}$
 	\item Backtrack matrix: $\Sigma(B)=\{stop,W,S,SW, 1..n\}$
 	\ule
-
 \item Initialization: \ul
-	\item Cost matrix: $\left\{\begin{array}{l} M_{(i,i)}=0 \\ M_{(i,i-1)}=0 \end{array}\right. \forall i \in 1..{\rm length}(S)$
-	\item Backtrack matrix: $\left\{\begin{array}{l} B_{(i,i)}=stop \\ B_{(i,i-1)}=stop \end{array}\right.  \forall i \in 1..{\rm length}(S)$
+	\item Cost matrix: $ M_{(i,i)}=M_{(i,i-1)}=0$
+	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i-1)}=stop$
 	\ule
 
-\item Recurrences:
+{\color{red}
+\item Recurrences (XXX: isn't it $\max_{i\le k < j}$ ?):
 	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
@@
 \subsection{Zuker folding}\ol
-\item Problem: folding a RNA string $S$ over itself.
-{\color{red}
-\item Matrices: XXX
-
+\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
+\item Matrices: $M_{n\times n}, V_{n\times n}, B_{n \times n}, {\color{red} F???}$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
-\item Backtrack matrix: $\Sigma(B)=\{\}$
+	\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\}$
+	\item Backtrack matrix: $\Sigma(B)=\{stop, {\color{red} S,W, XXX here} \}$
 	\ule
-
+{\color{red}
 \item Initialization:\ul

### 2012-10-01 docs/roadmap/problems.tex
Nits.

@@
 \newpage
-\subsection{Smith-Waterman with arbitrary gap cost (3D)}\ol
-
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function.
+\subsection{Smith-Waterman with arbitrary gap cost}\ol
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
 Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.
 Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
-
-\item Matrices: $M_{m \times n \times m}, B_{m \times n \times m}$
+\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
 \item Alphabets:\ul
@@
 	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
+	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
+	\ule
+
+\item Initialization:\ul
+	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
+	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
 	\ule
@@
 {\color{red}
-\item Initialization: XXX: we build this, because we go 3D.
-\item Recurrences:
+\item Initialization: 
+\item Recurrence:
+
+\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+	0 & stop\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	\max_{1 \le k \le j-1} M_{(i,j-k)} - g(k) & N_k\\
+	\max_{1 \le k \le i-1} M_{(i-k,j)} - g(k) & W_k\\
+\end{array}\right\}=B_{(i,j)} \]
+
 \item Backtracking:

### 2012-10-01 docs/roadmap/problems.tex
Stub plan.

@@
 
-% ------------------------------------------------------------------------------------------------
+% ----------------------------------------------
 \newpage
@@
 
+% ------------------------------------------------------------------------------------------------
+\newpage
+\section{Plan} \ol
+\item Define/agree some problems to take into consideration, agree on them
+\item Concerns separation: common architecture enables flexibility (exchange components) \ul
+	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
+	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
+	\item Scheduler (CPU): interleave block computation and memory statistics
+	\ule
+\item Common description \ul
+	\item Block kernel processor
+	\item Full block
+	\ule
+\item Discussion: wavefront, design similarities, polyhedral theory
+\ole
+
 %%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}

### 2012-10-01 docs/roadmap/problems.tex
Nits.

@@
 \usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
-\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
+%\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
 \usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
@@
 \section{Plan} \ol
-\item Define/agree some problems to take into consideration, agree on them
+\item Define/agree the maoin problems/classes to take into consideration: dimension, serial/non-serial, monadic/polyadic, ... add some other problems that belong to this class or simpler classes
 \item Concerns separation: common architecture enables flexibility (exchange components) \ul
@@
 \item Discussion: wavefront, design similarities, polyhedral theory
+\ole
+
+TODO:\ol
+\item See the feasibility for FPGA (possibly helped by CPU)
+\item Implement this algorithm in both CUDA and FPGA $\to$ specify all parameters so that we can compare both implementations
+\item Find a common IR for the problems
+\item Write a code generator for both architectures from IR $\to$ make sure all the test-DP are compiled well from IR
+\item Find an expressive and compact language for the user-facing representation
 \ole

### 2012-10-01 docs/roadmap/problems.tex
local changes
Nussinov.

@@
 \usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
-%\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
+\hypersetup{
+	%colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black
+	urlbordercolor=1 0 0,linkbordercolor=0 1 1, citebordercolor=0 1 1,
+}
 \usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
@@
 	\item Cost matrix: $\Sigma(M)=\{0..n\}$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,S,SW, 1..n\}$
+	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
 	\ule
@@
 	\ule
-
-{\color{red}
-\item Recurrences (XXX: isn't it $\max_{i\le k < j}$ ?):
+\item Recurrences:
 	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-		M_{(i+1,j-1)}+\omega(i,j) & SW\\
-		M_{(i+1,j)} & S\\
-		M_{(i,j-1)} & W\\
-		\max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} & k
+		M_{(i+1,j-1)}+\omega(i,j) & D\\
+		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
 	\end{array}\right\} = B_{(i,j)} \]
 	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
-\item Backtracking: XXX
-\item Visualisation: XXX
-\item Optimizations: XXX
-}
+\item Backtracking: Start the backtracking in $B_{(1,n)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
+\item Visualisation:
+	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
+		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
+		\Cfl{\Cg}\Cfd{\Cz}
+		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
+		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
+		\Cd[1,1]{5,3}{0.8}
+		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
+	\Cm\end{picture}\end{center}
+
+\item Optimizations: note that this is very similar to the matrix multiplication except that we also need the diagonal one step backward, so the same optimization can apply.
 \ole

### 2012-10-01 docs/roadmap/problems.tex
Nussinov.

@@
 \end{array}\]
-
 \item Backtracking: XXX
@@
 \section{Plan} \ol
-\item Define/agree the maoin problems/classes to take into consideration: dimension, serial/non-serial, monadic/polyadic, ... add some other problems that belong to this class or simpler classes
+\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
+\item Let the user tune the window size if he wants to reduce non-serial to serial.
+
 \item Concerns separation: common architecture enables flexibility (exchange components) \ul

### 2012-10-01 docs/roadmap/problems.tex
better definitions of the problem
Plan for future.

--- 2012-10-01 docs/roadmap/problems.tex
@@
 \title{DP Problems of Interest}
-\author{Manohar Jonnalagedda, Thierry Coppey}
+\author{Manohar Jonnalagedda, Thierry Coppey, Nithin George}
 \date{}
@@
 
-\item Backtracking: starts from the cell $M_{(m,j)} \cup M_{(i,n)} $
+\item Backtracking: starts from the cell $\max \{M_{(m,j)} \cup M_{(i,n)}\}$, stops at
+the first cell containing a $0$.
 \item Visualisation: by convention, we put the longest string vertically ($m\ge n$):
@@
 \newpage
-\subsection{Smith-Waterman with arbitrary gap cost}\ol
+\subsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
 \item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
@@
 \item Initialization:\ul
-	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
+	\item Match cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
 	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
@@
 
-{\color{red}
-\item Initialization: 
 \item Recurrence:
@@
 
-\item Backtracking:
+\item Backtracking: same as in \nameref{sswat}
 \item Visualisation:
-\item Optimizations:
-}
+
+	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+		\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
+		\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
+%		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
+%		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
+	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
+	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
+		\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
+		\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
+		\Cd[1,-1]{5,5}{0.8}
+	\Cm\end{picture}\end{center}
+
+\item Optimizations: The dependencies here are non-serial, there is no optimization that we can 
+apply out of the box here.
 \ole
@@
 \newpage
-\section{Plan} \ol
-\item Define/agree the maoin problems/classes to take into consideration: dimension, serial/non-serial, monadic/polyadic, ... add some other problems that belong to this class or simpler classes
-\item Concerns separation: common architecture enables flexibility (exchange components) \ul
-	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
-	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
-	\item Scheduler (CPU): interleave block computation and memory statistics
+\section{Plan}
+\ol
+\item \textbf{Problems:} finish the problem description.
+\\ We might look towards parallel tree-raking, but it does not share much with the above algorithms (sparse version of above computations, might not scale that well).
+Most of the common patterns are already enclosed by the above problems.
+Real input size is around 300'000 (we may want to target the million, using disk).
+We might want to also look at an $O(n^3)$-space complexity problem (like matching 3 strings $S,T,U$).
+\item \textbf{User facing language:} goals are flexibility and compactness.\ul
+	\item Design the user-facing language that we want to support, which should be similar to related papers <<Algebraic Dynamic Programming>> and <<ADP fusion>>.
+We want to reuse their transformation to map (problem description) $\mapsto$ (kernel implementation) for a single element.
+	\item We also may want to try to make implicit transformation for code like \\
+		{\tt @DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
+	\item Windowing: the user should be able to force a windowing (i.e. force a non-serial problem to be a $k$-polyadic serial problem).	
+	\item Consider 3 different cases: we care about backtrack, the costs or both.
+	\item Backtracking: we might want to create an operator/class that given an item produces the previous in the backtrack, or the whole sequence in correct order, or only indices.
 	\ule
-\item Common description \ul
-	\item Block kernel processor
-	\item Full block
-	\ule
-\item Discussion: wavefront, design similarities, polyhedral theory
+\item \textbf{Prototyping:} get a prototype to understand difficulties and share common base.
+\\ Implement a working prototype of \nameref{aswat} on CPU (for correctness), and specific platform (CUDA/FPGA).
+This will give us an idea of how to implement the general case.
+We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
+\item \textbf{Formalize IR:} According to experiences, describe the intermediate representation, also formalize the framework that will be provided to the code generators (i.e.
+memory management, ...).
+\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
+\item \textbf{Test and benchmark:} make sure our implementations are correct and compare them with previous papers implementations.
 \ole
 
-TODO:\ol
-\item See the feasibility for FPGA (possibly helped by CPU)
-\item Implement this algorithm in both CUDA and FPGA $\to$ specify all parameters so that we can compare both implementations
-\item Find a common IR for the problems
-\item Write a code generator for both architectures from IR $\to$ make sure all the test-DP are compiled well from IR
-\item Find an expressive and compact language for the user-facing representation
-\ole
+\newpage
+\subsection{LMS compiler stack}
+\textbf{User-language}: Windowing, input sizes, alphabet, recurrence functions, devices available, what to keep in memory (cost, backtrack or both). \\
+$\Downarrow$ Conversion (using an existing technique)
+
+\textbf{Intermediate representation}
+
+$\Downarrow$ Optimizations\ul
+\item Transform non-serial into serial \ul
+	\item Use aggregation functions/transformations
+	\item Use windowing from user (if no other technique succeed)
+	\ule
+\item Define the wavefront depth
+\item Avoiding the cost matrix by moving it into the wavefront
+\ule
+
+\textbf{Code specification}\ul
+\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
+\item Device-level interface => setup the block sizes(w/h), input and memory sizes
+\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
+\item Define the co-processor memory aggregation function
+\item Define the scheduling of the blocks and aggregation (software pipelining)
+\item Define the data movement back and forth to disk
+\ule
+
+$\Downarrow$ Generation\ul
+\item Generate the kernel for specific device
+\item Generate the scheduling and barriers
+\ule
+
+\textbf{Binary program}
+
+%\ol
+%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
+%\item Let the user tune the window size if he wants to reduce non-serial to serial.
+%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
+%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
+%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
+%	\item Scheduler (CPU): interleave block computation and memory statistics
+%	\ule
+%\item Common description \ul
+%	\item Block kernel processor
+%	\item Full block
+%	\ule
+%\item Discussion: wavefront, design similarities, polyhedral theory
+%\ole
 

### 2012-10-02 docs/roadmap/problems.tex
problem description precisions

@@
 \def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
+
+\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
+\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
 
@@
 
-{\color{red}
-\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
+\item Matrices: $M_{n\times 2n}, B_{n \times 2n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
 \item Alphabets:\ul
@@
 	\ule
-\item Initialization: $M_{(i,i)}=0, B_{(i,i)}=stop \quad\forall i$
+\item Initialization: $M_{(i,i)}=0, M_{(i,i+1)}=0, B_{(i,i)}=stop \quad\forall i$
+
 \item Recurrence: \[M_{(i,j)}=\left\{ \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + S(i,k) \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-	It is interesting to note that even in the sequential world, this problem is best solved 
-	by filling the diagonals, ie. computing sub-solutions for all polygons on size $k$ before
+	It is interesting to note that even in the sequential world, this problem is solved 
+	by filling the diagonals, ie. computing sub-solutions for all polygons of size $k$ before
 	those of size $k+1$.
-\item Backtracking: Starting at $B_{(0,n)}$ and let current point you backtrack drawing an edge from the current position to the backtracking position ??
 
-\item Visualisation: in its traditional form the matrix is upper triangular with diagonal excluded.
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for the smaller polygons:
+	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
+		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
+
+\item Visualisation: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
+"cyclical" in nature.
+\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,9)
 	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
 	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cfl{\Cg}\Cfd{\Cz}
+	\Cfl{\Cg}\Cfd{\Cz}\Cfu{\Cg}
 	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 	%\Cd[1,-1]{5,2}{0.8}
-\Cm\end{picture}\end{center}
+\Cmlong\end{picture}\end{center}
 
-\item Optimizations: XXX we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
-}
+\item Optimizations: we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
+{\color{red}}
 \ole

### 2012-10-02 docs/roadmap/problems.tex
More Zucker.

--- 2012-10-01 docs/roadmap/problems.tex
@@
 
-\item Backtracking: same as in \nameref{sswat}
+\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells.
 \item Visualisation:
@@
 \item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
-\item Matrices: $M_{n\times n}, V_{n\times n}, B_{n \times n}, {\color{red} F???}$
+\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\}$
-	\item Backtrack matrix: $\Sigma(B)=\{stop, {\color{red} S,W, XXX here} \}$
+	\item Cost matrices:\ul
+		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n*b+c$
+		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
 		\ule
-{\color{red}
+	\item Backtrack matrices: \ul
+		\item $\Sigma(BW)=\{stop, S,W,V,k\}$
+		\item $\Sigma(BV)=\{stop, EH, ES, SW, i' \times j', k\}$
+		\item $\Sigma(BF)=\{stop, L, k\}$
+		\ule
+	\ule
 \item Initialization:\ul
-	\item Cost matrices:\ul
-		\item XXX: $W$,$V$
-		\item $F_{(0)}=0$
+	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
+	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
 	\ule
-	\item Backtrack matrix: $B$
-	\ule
-
 \item Recurrence:
@@
 	W_{(i,j-1)}+b & W\\
-	V_{(i,j)}+\delta(S_i,S_j) & ?V? \\
-	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} & ?k?
-\end{array}\right.\\
+	V_{(i,j)}+\delta(S_i,S_j) & V \\
+	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
+\end{array}\right\} = BW_{(i,j)}\\
 \\
-V_{(i,j)}&=&\min\left\{\begin{array}{ll}
+V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
 	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
-	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
-	V_{(i+1,j-1)}+es(i,j) \\
-	VBI(i,j) \\
-	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
-\end{array}\right.\\
+	eh(i,j)+b \qquad\qquad\, \text{otherwise} & EH\\
+	V_{(i+1,j-1)}+es(i,j) & ES \\
+	VBI(i,j) & i',j' \\
+	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
+\end{array}\right\} = BV_{(i,j)}\\
 \\
- F_{(j)}&=&\min\left\{\begin{array}{ll} F_{(j-1)} \\ \min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
-%
-\end{array}\right. \text{(Free Energy)}\\\\
+F_{(j)}&=&\min\left\{\begin{array}{l|l}
+	F_{(j-1)} & L \\ 
+	\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)}) & k
+\end{array}\right\} = BF_{(j)}\\
+\\
 VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
 \end{array}\]
+
+In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $W$,$V$, $VBI$ and $F$.
+
+{\color{red}
 \item Backtracking: XXX
@@
 \subsection{LMS compiler stack}
-\textbf{User-language}: Windowing, input sizes, alphabet, recurrence functions, devices available, what to keep in memory (cost, backtrack or both). \\
+\textbf{User-language}: define additional parameters for the recurrence\ul
+\item Windowing (to convert non-serial into serial problems)
+\item Input sizes, and alphabets (backtrack, input, cost)
+\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
+\item Recurrence functions, devices available
+\item What to keep in memory (cost, backtrack or both).
+\ule
 $\Downarrow$ Conversion (using an existing technique)

### 2012-10-02 docs/roadmap/problems.tex
More Zucker.

@@
 		\item $\Sigma(BW)=\{stop, S,W,V,k\}$
-		\item $\Sigma(BV)=\{stop, EH, ES, SW, i' \times j', k\}$
-		\item $\Sigma(BF)=\{stop, L, k\}$
+		\item $\Sigma(BV)=\{stop, HL, IL, SW, (i,j) , k\}$ with $0\le i,j,k < n$ \\
+		$HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
+
+		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
 		\ule
@@
 V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
-	eh(i,j)+b \qquad\qquad\, \text{otherwise} & EH\\
-	V_{(i+1,j-1)}+es(i,j) & ES \\
-	VBI(i,j) & i',j' \\
+	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
+	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
+	V_{(i+1,j-1)}+es(i,j) & IL \\
+	VBI(i,j) & (i',j') \\
 	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
@@
 	F_{(j-1)} & L \\ 
-	\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)}) & k
+	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
 \end{array}\right\} = BF_{(j)}\\
 \\
-VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
+VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c \quad|\quad (i',j')=BV_{(i,j)}
 \end{array}\]
@@
 {\color{red}
-\item Backtracking: XXX
-\item Visualisation: XXX
-\item Optimizations: XXX
+\item Backtracking: Start at $BF_{(n)}$ using the recurrences
+ \[\begin{array}{rcl}
+	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
+	\\
+	BV(i,j) &=& \left\{\begin{array}{rcl}
+		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
+		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
+		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
+		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
+	\end{array}\right.\\
+	\\
+	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
+	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
+	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
+	V &\implies& BV_{(i,j)} \\
+	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
+	\end{array}\right.
+\end{array}\]
 }
+\item Visualisation: \begin{center}\includegraphics[width=7cm]{problems_zucker.pdf}\end{center}
+	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
+\item Optimizations: {\color{red} XXX}
 \ole

### 2012-10-02 docs/roadmap/problems.tex
dates

--- 2012-10-02 docs/roadmap/problems.tex
@@
 	\ule
+Get an implementation by \emph{end of October}.
+
 \item \textbf{Prototyping:} get a prototype to understand difficulties and share common base.
@@
 We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
+Here we aim to do as good an implementaion for the specific platform (CPU/GPA) as possible.
+
+\item \textbf{Baseline:} get some implementations working on the different platforms and use them as baselines. Goal to get something working by \emph{end of October}.
+
 \item \textbf{Formalize IR:} According to experiences, describe the intermediate representation, also formalize the framework that will be provided to the code generators (i.e.

### 2012-10-08 docs/roadmap/problems.tex
dates
Nits.

--- 2012-10-02 docs/roadmap/problems.tex
@@
 \ol
-\item \textbf{Problems:} finish the problem description.
+\item {\color{gray}
+\textbf{Problems:} finish the problem description.
 \\ We might look towards parallel tree-raking, but it does not share much with the above algorithms (sparse version of above computations, might not scale that well).
@@
 We might want to also look at an $O(n^3)$-space complexity problem (like matching 3 strings $S,T,U$).
+}
+
+All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
+\item Serial dependencies
+\item Non-serial horizontal or vertical dependencies (1D non-serial)
+\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
+\item We have not found other type of dependencies in the litterature
+\ule
+
 \item \textbf{User facing language:} goals are flexibility and compactness.\ul
@@
 	\ule
+
+$\implies$ \emph{end of October}.
+
 \item \textbf{Prototyping:} get a prototype to understand difficulties and share common base.
@@
 We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
+Here we aim to do as good an implementaion for the specific platform (CPU/GPA) as possible.\\
+$\implies$ \emph{end of October}.
+
+\item \textbf{Baseline:} Also use benchmarks provided by existing implementations as baselines.\\
+$\implies$ \emph{end of October}.
+
 \item \textbf{Formalize IR:} According to experiences, describe the intermediate representation, also formalize the framework that will be provided to the code generators (i.e.

### 2012-10-15 docs/roadmap/problems.tex
CudaAlign 2.0.1 source, added build on paper.

@@
 	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
-	\item Cost matrix: $\Sigma(M):=$ huge integer\footnote{Integer multiplication might blow up, use float or doubles and addition of logarithms instead.}.
+	\item Cost matrix: $\Sigma(M)= 1..z$ with $z\le n\cdot \big[ \max_i(r_i,c_i) \big]^3 $.
 	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
@@
 To do that, we need to map all indices $i \mapsto n-1-i$, but since we want to store the matrix sparsely, we might want to transform it into a lower triangular matrix as we can provide a size-independent mapping of element indices.
-	XXX: what's the best trade off ? progress towards bottom right VS map indices more efficiently ?}
+	XXX: what's the best trade off ? progress towards bottom right VS map indices more efficiently ? Also consider efficient memory usage$\to$ larger problem VS standardized layout}
 	\ule
@@
 \item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
-\item We have not found other type of dependencies in the litterature
+\item We have not found other type of dependencies in the literature
 \ule
@@
 We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
-Here we aim to do as good an implementaion for the specific platform (CPU/GPA) as possible.\\
+Here we aim to do as good an implementation for the specific platform (CPU/GPA) as possible.\\
 $\implies$ \emph{end of October}.

### 2012-10-15 docs/roadmap/problems.tex
Initial, although probably bugged matrix multiplication CPU algorithm.

@@
 \item Optimizations:\ul
-	\item {\color{red}We need to swap vertically the matrix to have a normalized progression towards bottom right.
-To do that, we need to map all indices $i \mapsto n-1-i$, but since we want to store the matrix sparsely, we might want to transform it into a lower triangular matrix as we can provide a size-independent mapping of element indices.
-	XXX: what's the best trade off ? progress towards bottom right VS map indices more efficiently ? Also consider efficient memory usage$\to$ larger problem VS standardized layout}
+	\item We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$ and we obtain a new recurrence relation:
+	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l} M_{(i,k)}+M_{(2i-1 -k,j)}+r_i \cdot c_k \cdot c_j \end{array}\right. \]
+	With the initialization at $M_{(i,n-i-1)}$
 	\ule

### 2012-10-17 docs/roadmap/problems.tex
Finished CPU version, should be correct.

@@
 
-\item Recurrence: \[M_{(i,j)}=\left\{ \max_{i<k<j}M_{(i,k)}+M_{(k+1,j)} + S(i,k) \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
+\item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
 	It is interesting to note that even in the sequential world, this problem is solved 

### 2012-10-18 docs/roadmap/problems.tex
Fixed SWat strings by introducing padding.

@@
 \subsection{Smith-Waterman (simple)}\label{sswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
 \item Matrices: $M_{m \times n}, B_{m \times n}$
@@
 This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
+	\item Padding: since first line and column of the matrix are zeroes, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
+Also since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
+Hence valid input indices are $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
+We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
 	\ule
@@
 \subsection{Smith-Waterman (with gap extension at different cost)}\ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
 \item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
@@
 \subsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
+\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
 Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.

### 2012-10-18 docs/roadmap/benchmarks.tex
Extracted some benchmarks.

\input{style.sty}

\title{Benchmarks}
\begin{document}
\maketitle
\pagestyle{headings}

The goal of this document is to see how we are comparing with other papers in terms of performance and to note performance progress of the improvements.

\subsection*{Graphic cards}
\def\unt#1{& \footnotesize #1}
\begin{center}\begin{tabular}{lrrr} \toprule
\bf Paper		&				& \bf -- 		& \bf  ATLP\cite{gpu_atlp} \\ \midrule
Serie		&				& GeForce	& Tesla \\
Model		&				& GT 650M	& C1060 \\
Architecture	&				& Kepler		& GT200 \\
Capability		&				& 3.0		& 1.3 \\
Memory \unt{Mb}				& 1024		& 4096 \\
CUDA cores &					& 384		& 240 \\
Core clock \unt{MHz}			& 756		& 1300 \\
Memory clock \unt{MHz}			& 1953		& 1600 \\
Memory bus \unt{bit}				& 128		& 512 \\
Memory bandwidth \unt{GB/s}		& 28.8		& 102.4 \\
Processing power \unt{GFLOPS}	&564.5		& 622.08 \\ \midrule
\bf Processing speedup & 		& 1			& 1.07 \\
\bf Memory speedup & 			& 1			& 3.55 \\ \bottomrule
\end{tabular}\end{center}

\subsection*{Results}
\subsubsection*{ATLP\cite{gpu_atlp}}
\begin{center}\begin{tabular}{lrrrrrrrrrr} \toprule
\bf Matrix size & 128 & 256 & 512 & 1024 & 1536 & 2048 & 2560 & 3072 & 3584 & 4096 \\ \midrule
\bf No split & 0.07 & 0.09 & 0.19 & 0.59 & 1.27 & 2.25 & 3.51 & 5.07 & 6.92 & 9.06 \\
\bf Split at 1 & 0.06 & 0.07 & 0.08 & 0.14 & 0.26 & 0.47 & 0.77 & 1.21 & 1.80 & 2.57 \\ \bottomrule
\end{tabular} \\[4pt]
Matrix multiplication timing in seconds
\end{center}

\bibliographystyle{plain}
\bibliography{bibliography.bib}

\end{document}

### 2012-10-19 docs/roadmap/benchmarks.tex
Improvement of non-serial version.

@@
 
+% --------------------------------------------------------------------------------
 \subsection*{Graphic cards}
@@
 
+% --------------------------------------------------------------------------------
 \subsection*{Results}
@@
 
+% --------------------------------------------------------------------------------
+\subsubsection*{Preliminary results}
+\begin{center}\begin{tabular}{cclrrr} \toprule
+\bf Matrix & \bf Block & \bf Comment & \bf R(ms) & \bf T(ms) & \bf P(ms) \\ \midrule
+1024 & 32 & CPU baseline & 1965.63 & 1191.13 & 6069.66 \\
+1024 & 32 & GPU baseline & 838.67 & 499.87 & 516.82 \\
+1024 & 32 & GPU sync improved & 642.37 & 316.40 & 343.29 \\
+
+\bottomrule \end{tabular} \\[4pt]
+Best timings for R=rectangle, T=triangle, P=parallelogram
+\end{center}
+
+% --------------------------------------------------------------------------------
+\subsection*{Micro-benchmarks \footnote{Micro-benchmark do not provide extensive results as they stand only to validate implementation strategies.}}
+Problem are described by the triplet <matrix size/block size/shape>.
+\ul
+\item It has been put in evidence in a previous work\footnote{Performance Evaluation 2012 miniproject, \it Performance Evaluation of Mersenne arithmetic on GPUs} that with Mac OS operating system, using the GPU exclusively for CUDA or combining with UI display may affect the performance (GeForce 330M, architecture 1.2).
+With the new architecture (3.0, GeForce 650M), this difference has been reduced to less than 3.5\% with the decoupling of UI and CUDA performing best.
+So in micro-benchmarks, we can safely ignore the graphic card usage.
+
+\item Synchronization  between blocks\ul
+	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but still remains correct, this validates the observation made by \cite{gpu_barrier}.
+Speedup for <1024/32/*> are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
+	\item In the parallelogram case, using all threads to monitor other blocks status instead of 1 thread results in a 6.4x speedup (22.72$\to$3.52ms) on <1024/32/para>.
+	\ule
+\ule
+
+% --------------------------------------------------------------------------------
 \bibliographystyle{plain}
 \bibliography{bibliography.bib}
-
 \end{document}

### 2012-10-19 docs/roadmap/benchmarks.tex
Nits.

@@
 % --------------------------------------------------------------------------------
-\subsection*{Graphic cards}
+\subsection*{Graphic cards\footnote{Source: \url{http://en.wikipedia.org/wiki/Comparison_of_Nvidia_graphics_processing_units}}}
 \def\unt#1{& \footnotesize #1}
-\begin{center}\begin{tabular}{lrrr} \toprule
-\bf Paper		&				& \bf -- 		& \bf  ATLP\cite{gpu_atlp} \\ \midrule
-Serie		&				& GeForce	& Tesla \\
-Model		&				& GT 650M	& C1060 \\
-Architecture	&				& Kepler		& GT200 \\
-Capability		&				& 3.0		& 1.3 \\
-Memory \unt{Mb}				& 1024		& 4096 \\
-CUDA cores &					& 384		& 240 \\
-Core clock \unt{MHz}			& 756		& 1300 \\
-Memory clock \unt{MHz}			& 1953		& 1600 \\
-Memory bus \unt{bit}				& 128		& 512 \\
-Memory bandwidth \unt{GB/s}		& 28.8		& 102.4 \\
-Processing power \unt{GFLOPS}	&564.5		& 622.08 \\ \midrule
-\bf Processing speedup & 		& 1			& 1.07 \\
-\bf Memory speedup & 			& 1			& 3.55 \\ \bottomrule
+\begin{center}\begin{tabular}{lrrrr} \toprule
+\bf Paper		&				& \bf -- 		& \bf  ATLP\cite{gpu_atlp} & \bf SWMB\cite{swat_mega} \\ \midrule
+Serie		&				& GeForce	& Tesla	& GeForce  \\
+Model		&				& GT 650M	& C1060	& GTX 560 \\
+Architecture	&				& Kepler		& GT200	& GF114 \\
+Capability		&				& 3.0		& 1.3	& 2.1 \\
+Memory \unt{Mb}				& 1024		& 4096	& 1024 \\
+CUDA cores &					& 384		& 240	& 384 \\
+Core clock \unt{MHz}			& 756		& 1300	& 822 \\
+Memory clock \unt{MHz}			& 1953		& 1600	& 4008 \\
+Memory bus \unt{bit}				& 128		& 512	& 256 \\
+Memory bandwidth \unt{GB/s}		& 28.8		& 102.4	& 128.26 \\
+Processing power \unt{GFLOPS}	&564.5		& 622.08	& 1263.4 \\ \midrule
+\bf Processing speedup & 		& 1			& 1.07	& \\
+\bf Memory speedup & 			& 1			& 3.55	& \\ \bottomrule
 \end{tabular}\end{center}
@@
 \bf Split at 1 & 0.06 & 0.07 & 0.08 & 0.14 & 0.26 & 0.47 & 0.77 & 1.21 & 1.80 & 2.57 \\ \bottomrule
-\end{tabular} \\[4pt]
-Matrix multiplication timing in seconds
-\end{center}
+\end{tabular} \\[4pt] Matrix multiplication timing in seconds \end{center}
+
+\subsubsection*{SWMB\cite{swat_mega}}
+\begin{center}\begin{tabular}{rlrr} \toprule
+\bf Matrix size & \bf Sequences & \bf No pruning & \bf Pruning \\ \midrule
+162K $\times$ 172K		& NC\_000898.1, NC\_007605.1	& 1.2 & 1.2 \\
+543K $\times$ 536K		& NC\_003064.2, NC\_000914.1	& 10.8 & 10.8 \\
+1044K $\times$ 1073K	& CP000051.1, AE002160.2		& 40.3 & 36.2 \\
+3147K $\times$ 3283K	& BA000035.2, BX927147.1		& 363.6 & 363.2 \\ 
+5227K $\times$ 5229K	& AE016879.1, AE017225.1		& 962.4 & 469.5 \\
+7146K $\times$ 5227K	& NC\_005027.1, NC\_003997.3	& 1309 & 1309 \\
+23012K $\times$ 24544K	& NT\_033779.4, NT\_037436.3	& 19701 & 19694 \\
+59374K $\times$ 23953K	& NC\_000024.9, NC\_006492.2	& 49634 & 46869 \\
+32799K $\times$ 46944K	& BA000046.3, NC\_000021.7		& 53869 & 29133 \\ \bottomrule
+\end{tabular} \\[4pt] Smith-Waterman, timing in seconds \end{center}
 
 % --------------------------------------------------------------------------------
-\subsubsection*{Preliminary results}
+\subsubsection*{Intermediate results}
 \begin{center}\begin{tabular}{cclrrr} \toprule

### 2012-10-19 src/test/ns_test.cu

@@ -17,13 +17,6 @@
 #include "include/ns_gpu.h"  // gpu implementation
 // -----------------------------------------------------------------------------
 
-/*
-PRELIMINARY RESULTS (i7-3720QM-1core vs GeForce GT 650M-attached)
------------------------------------------------------------------
-
-
-*/
-
 __global__ void gpu_solve(const TI* in0, const TI* in1, TC* cost, TB* back, volatile unsigned* lock) {
 	const unsigned tI = threadIdx.x + blockIdx.x * blockDim.x; // * (  + blockIdx.y*gridDim.x );
 	const unsigned tN = blockDim.x * gridDim.x;

### 2012-10-19 docs/roadmap/problems.tex

@@
-\documentclass[11pt]{article}
-\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs,nameref}
-\hypersetup{
-	%colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black
-	urlbordercolor=1 0 0,linkbordercolor=0 1 1, citebordercolor=0 1 1,
-}
-\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage[english]{babel}
-\pagestyle{empty} % remove page numbers
-
-\oddsidemargin 0.0cm
-\evensidemargin 0.0cm
-\topmargin 0.0cm
-\headheight 0.0cm
-\headsep 1.0cm
-\textheight 21.0cm
-\textwidth 16.0cm
-\parskip 0.1cm
-\parindent 0.0cm
-\footskip 1.0cm
-
-\usepackage[compact]{titlesec}
-\titlespacing{\section}{0pt}{8pt}{0pt}
-\titlespacing{\subsection}{0pt}{8pt}{0pt}
-\titlespacing{\subsubsection}{0pt}{8pt}{0pt}
-\usepackage{enumitem}
-\setitemize{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}
-\setenumerate{noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt}
-\def\ul{\begin{itemize}}
-\def\ule{\end{itemize}}
-\def\ol{\begin{enumerate}}
-\def\ole{\end{enumerate}}
+\input{style.sty}
 
@@
 \title{DP Problems of Interest}
-\author{Manohar Jonnalagedda, Thierry Coppey, Nithin George}
-\date{}
 \begin{document}

### 2012-10-23 docs/roadmap/benchmarks.tex
Pre-documentation-rework.

@@
 1024 & 32 & GPU sync improved & 642.37 & 316.40 & 343.29 \\
-
+2048 & 32 & CPU baseline & 27229.28 & 15296.09 & 57323.04 \\
+2048 & 32 & GPU P 32 blocks & 2864.27 & 1427.21 & 2096.34 \\
 \bottomrule \end{tabular} \\[4pt]
-Best timings for R=rectangle, T=triangle, P=parallelogram
+Best timings for R=rectangle, T=triangle, P=parallelogram\footnote{$\le 32$ blocks on my GPU to prevent deadlock}
 \end{center}

### 2012-10-23 docs/roadmap/problems.tex
Pre-documentation-rework.

@@
 % ----------------------------------------------------------------------------------------------------------------------------------------
-\section{Introduction}
+\section{Problems classification}
 \subsection{Definitions}\ul
@@
 
-\subsection{Problems classification}
+\subsection{Classification}
 In the literature, dynamic programming problems (DP) are classified according to two criteria:
@@
 \newpage
-\subsection{Smith-Waterman (with gap extension at different cost)}\ol
+\subsection{Smith-Waterman with affine gap extension cost}\ol
 \item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
@@
 	V_{(i+1,j-1)}+es(i,j) & IL \\
-	VBI(i,j) & (i',j') \\
+	VBI_{(i,j)} & (i',j') \\
 	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
@@
 \\
-VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c \quad|\quad (i',j')=BV_{(i,j)}
+VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}
 \end{array}\]
 
-In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $W$,$V$, $VBI$ and $F$.
+In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$ {\color{red} $W$,$V$ and $F$}.
 
-{\color{red}
 \item Backtracking: Start at $BF_{(n)}$ using the recurrences
@@
 	\\
-	BV(i,j) &=& \left\{\begin{array}{rcl}
+	BV_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
 		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
@@
 	\\
-	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
+	BW_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
 	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
@@
 \end{array}\]
-}
-\item Visualisation: \begin{center}\includegraphics[width=7cm]{problems_zucker.pdf}\end{center}
+
+\item Visualisation: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
 	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
-\item Optimizations: {\color{red} XXX}
+\item Optimizations: {\color{red} XXX: we need to find how to encode $F_{(j)}$ such that we can normalize our computation into an existing recurrence shape.}
 \ole
+\newpage
+
+% ------------------------------------------------------------------------------------------------
+\section{Related problems}
 

### 2012-10-23 docs/roadmap/problems.tex
Ideas stub.

@@
 
-% Recurrence visualisation helpers
+% Recurrence visualization helpers
 \newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
@@
 \subsection{Definitions}\ul
-\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
-\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
+\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
+\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest.
+Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
+\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute.
+A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
+\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix.
+It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
+\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
 \ule
@@
 \subsection{Classification}
-In the literature, dynamic programming problems (DP) are classified according to two criteria:
+\subsubsection{Litterature classification}
+In the literature, dynamic programming problems (DP) are classified according to two criteria:\ul
+\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman).
+When two or more terms appear, the problem is polyadic (ex: Fibonacci, $F_n = F_{n-1} + F_{n-2}$).
+When a problem is polyadic with index $p$, it also means that its backtracking forms a $p$-ary tree (where each node has at most $p$ children).
 
-\textbf{Monadic/polyadic}\ul
-\item \textbf{Monadic:} on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
-\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
-\ule
-
-\textbf{Serial/non-serial} \ul
-\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial (it accesses only 2 cells backward).
-\item \textbf{Non-serial:} when the solution depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
+\item \textbf{Serial/non-serial:} a problem is serial ($s=0$) when the solutions depends on a fixed number of previous solutions (ex: Fibonacci), otherwise it is said to be non-serial ($s\ge 1$), as the number of dependencies grows with the size of the subproblem.
+That is computing an element of the matrix would require $O(n^s)$.
+(ex: Smith-Waterman with arbitrary gap is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula)
 	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
@@
 
-\subsection{Simplifications}
-\subsubsection{Calculus}
+Note that the algorithmic complexity of a problem is exactly $O\big(n^{d+s}\big)$.
+
+\subsubsection{Calculus simplifications}
 In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
@@
 
-This simplification avoids the non-serial dependencies at the cost of extra storage in the wavefront, unfortunately, it might be applicable only for some special cases.
-
-\subsubsection{Precomputations}
-When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
-For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.
-
-On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
-We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).
+Although this simplification removes some non-serial dependencies at the cost of extra storage in the wavefront, it is not sufficient to transform all non-serial monadic problems into serial problems (ex: this does not apply to Smith-Waterman with arbitrary gap cost).
 
@@
 \section{Problems of interest}
+We usually focus on problem that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$).
+Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability.
+Also we tend to limit algorithmic complexity of the problems as from $O(n^4)$ on, running time becomes a severely limiting factor.
+
 We describe problems structures: inputs, cost matrices and backtracking matrix.
@@
 the first cell containing a $0$.
-\item Visualisation: by convention, we put the longest string vertically ($m\ge n$):
+\item Visualization: by convention, we put the longest string vertically ($m\ge n$):
 \begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 \item Backtracking: same as \nameref{sswat}
-\item Visualisation: same as \nameref{sswat}
+\item Visualization: same as \nameref{sswat}
 \item Optimizations: same as \nameref{sswat}
@@
 \item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells.
-\item Visualisation:
-
+\item Visualization:
 	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 		\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
-%		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-%		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
 	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
@@
 
-\item Visualisation: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
+\item Visualization: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
 "cyclical" in nature.
@@
 
-\item Visualisation:
+\item Visualization:
 	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 \item Backtracking: Start the backtracking in $B_{(1,n)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
-\item Visualisation:
+\item Visualization:
 	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 
-\item Visualisation: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
+\item Visualization: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
 	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
@@
 \ole
-\newpage
 
 % ------------------------------------------------------------------------------------------------
+\newpage
 \section{Related problems}
+
+{\color{red}
+Add additional problems into their related category
+\ul
+\item \url{http://en.wikipedia.org/wiki/Dynamic_programming}
+\item \url{http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf}
+\item \url{http://www.algorithmist.com/index.php/Dynamic_Programming}
+\item \url{http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf}
+\item \url{http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf}
+\ule
+}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\section{Implementation}
+We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done.
+This can be the GPU or the FPGA internal memory.
+Usually the main memory is larger and can ultimately be extended by disk or network storage.
+\subsection{Memory requirements}
+We propose to evaluate the device memory requirements to solve the above problem classes. First we need to define additional problem properties related to implementation:\ul
+\item
+\ule
+
+\subsection{Performance considerations}
+
+%\subsubsection{Precomputations}
+%When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
+For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.
+%
+%On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
+We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).
+
+Zucker: il y a une matrice 1d additionnelle
+la question n'est pas "comment on fait", j'ai relu la recurrence, mais d'avantage "comment rendre a de maniere generique sans utiliser une grosse hash map" ?
+
+XXX: do we want to support larger non-serial problem that do not fit in memory
+
+%\subsubsection{Towards a complexity metric}
+Considering the serial polyadic case, the number of dependencies of each option in the right hand-side of the formula (polyadic count) needs to be weighted with the number of achievable options (lines in the right hand-side of the formula) to provide some usable metric (number of dependencies, or memory accesses).
+Hence unless all options are evaluated in parallel, polyadic count itself can not be used as a problem complexity metric.
+
+The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
+Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
+
+These considerations obviously do not change the problem's algorithmic complexity but might provide some estimate of the constant factor that is omitted in the big $O$ notation.
+
+%\subsubsection{New classification proposal}
+We propose 5 criteria to categorize dynamic programming problems:\ol
+\item \textbf{Dimensions of the matrix:} the dimension in which the intermediate computation matrix lies.
+For the majority of the problems, the matrix is 2D, but it might be 1D (ex: Fibonacci), 3D (aligning 3 sequences) or more.
+\item \textbf{Complexity of dependencies:} the number of dependencies can be a constant (serial) or arbitrarily growing (non-serial).
+Also in the non-serial case, we allow to differentiate between dependences along one or more dimensions of the matrix.
+\item \textbf{Number of matrices:} (and element size) counts the number of cost and associated backtrack matrices to be maintained.
+This factor (combined with respective data encoding) allows to give a precise memory consumption and defines whether the problem fits into one device memory or needs to be split into subproblems.
+\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
+This amount is usually proportional to the delay of dependencies.
+\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell of the matrix (ex: maximum along one line).
+Hence such information can be maintained in a single place (wavefront) and progress with the computation.
+\ole
+
+The first 2 criteria provide the theoretical complexity of the dynamic program.
+Indeed, DP proceeds by filling every element of the matrix, and for each element of the matrix, compute the aggregate over all dependencies, hence the overall complexity is equal to the product of the complexities.
+
+The three last elements, combined with the above one, provide a precise estimation of the memory consumption, and the implementation difficulty
+
+all the problem are subject to the input dimensions $n$.
+
+the two latter one gives an estimation of the constant factor.
+
+delay
+polyadic
+
+We might need to introduce new parameters for serial
+- delay (access 1 or up to k elements backwards)
+- number of cost matrices to maintain
+
+- size of wavefront
+
+Implementation: number of matrices, size of data, size of wavefront
+
+Since in some special case, transforming it might be possible to transform a non-serial 
+We propose a slightly different classification as
+
+non-serial monadic -> serial polyadic
+
+XXX: transformation to polyadic serial into monadic serial
+
+-> aggregation
+-> delay
+
+monadic / polyadic defines the complexity of the backtrack
 

### 2012-10-23 docs/roadmap/problems.tex
Problems tighter estimates.

@@
 
-\item Recurrence:
-
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
+\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
 	0 & stop\\
@@
 	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
-\item Optimizations: {\color{red} XXX: we need to find how to encode $F_{(j)}$ such that we can normalize our computation into an existing recurrence shape.}
+\item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?).
+We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?)}
 \ole
@@
 {\color{red}
-Add additional problems into their related category
+XXX: describe common parts between problems (sub categories)
+
+XXX: Add additional problems into their related category
 \ul
@@
 \section{Implementation}
-We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done.
+We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
 This can be the GPU or the FPGA internal memory.
-Usually the main memory is larger and can ultimately be extended by disk or network storage.
+Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
 \subsection{Memory requirements}
-We propose to evaluate the device memory requirements to solve the above problem classes. First we need to define additional problem properties related to implementation:\ul
-\item
+We propose to evaluate the device memory requirements to solve the above problem classes. We need first to define additional problem properties related to implementation:\ul
+\item \textbf{Number of matrices:} multiple matrices can be encoded as 1 matrix with multiple values per cell.
+Hence the implementation differentiates only between cost and backtrack matrices with respective element sizes $S_C$ and $S_B$.
+\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
+Such data amount is usually proportional to the delay of dependencies.
+If this delay is small, it might be worth to duplicate matrix data in the wavefront, otherwise it might be more efficient to allow access to the previous computational blocks of the matrix.
+\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
+Hence such information can be maintained in a single place (in the wavefront) and progress together with the computation.
+We denote by $S_W$ the size of wavefront elements.
+\item \textbf{Input size:} the size of an input letter (from input alphabet) is denoted by $S_I$.
 \ule
 
-\subsection{Performance considerations}
+\subsection{Small problems (in-memory)}
+Problem that can fit in memory can be solved in a single pass on the device. Such problem must satisfy the equation:
+	\[(S_I+S_W) \cdot (m+n) + (S_C+S_B) \cdot (m\cdot n) \le M_D\]
 
-%\subsubsection{Precomputations}
-%When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
-For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.
-%
-%On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
-We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).
+For instance, assuming that $m=n$, $M_D=1024{\rm Mb}$, that backtrack is 2b (<16384, 3 directions) and that the cost can be represented on 4 b (int or float), that input is 1b (char) and that there is no wavefront, we can treat problems of size $n$ such that $2n+5n^2 \le 2^{30} \implies  n\le 14650$.
+We might also possibly need to take into account extra padding memory used for coalesced accesses.
+But it is reasonable to estimate that problems up to 14K fit in memory.
 
-Zucker: il y a une matrice 1d additionnelle
-la question n'est pas "comment on fait", j'ai relu la recurrence, mais d'avantage "comment rendre a de maniere generique sans utiliser une grosse hash map" ?
+\subsection{Large problems}
+To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$.
 
-XXX: do we want to support larger non-serial problem that do not fit in memory
+\subsubsection{Non-serial problems}
+Non-serial problems need to potentially access all elements that have been previously computed.
+Assuming the restriction that dependencies are only along the rows and columns and that the serial dependencies delay is $\le$ than one block size, we need all the block on the same line and column and one in diagonal.
 
-%\subsubsection{Towards a complexity metric}
-Considering the serial polyadic case, the number of dependencies of each option in the right hand-side of the formula (polyadic count) needs to be weighted with the number of achievable options (lines in the right hand-side of the formula) to provide some usable metric (number of dependencies, or memory accesses).
-Hence unless all options are evaluated in parallel, polyadic count itself can not be used as a problem complexity metric.
+For simplification, let $m=n$, assume that we have $b$ rectangular block per row and per column. We need to transfer forth and back (in terms of blocks):
+	\[b \cdot b +% computed line (once)
+		(b-1) \cdot (b-1) + % diagonal block (1 per block)
+		\sum_{i=0}^{b-1} i \cdot b % column blocks (for line i)
+		= \tfrac{1}{2}b^2(b+5)-2b+1\]
 
-The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
-Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
+With $b=8$ we have 401 block transfer (forth and back), which corresponds to transferring $12.5\times$ the whole matrix, and the problem size can now be (following the same formula and assertion as for small problems) $2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = 2n+10n^2/8 \le 2^{30} \implies n\le 29307$, so we roughly doubled matrix size at the cost of $12.5\times$ matrix transfer time plus indices overhead.
 
-These considerations obviously do not change the problem's algorithmic complexity but might provide some estimate of the constant factor that is omitted in the big $O$ notation.
+{\color{red} XXX: Do we want to support larger non-serial problem that do not fit in memory? Is it worth implementing ? Argue.}
 
-%\subsubsection{New classification proposal}
-We propose 5 criteria to categorize dynamic programming problems:\ol
-\item \textbf{Dimensions of the matrix:} the dimension in which the intermediate computation matrix lies.
-For the majority of the problems, the matrix is 2D, but it might be 1D (ex: Fibonacci), 3D (aligning 3 sequences) or more.
-\item \textbf{Complexity of dependencies:} the number of dependencies can be a constant (serial) or arbitrarily growing (non-serial).
-Also in the non-serial case, we allow to differentiate between dependences along one or more dimensions of the matrix.
-\item \textbf{Number of matrices:} (and element size) counts the number of cost and associated backtrack matrices to be maintained.
-This factor (combined with respective data encoding) allows to give a precise memory consumption and defines whether the problem fits into one device memory or needs to be split into subproblems.
-\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
-This amount is usually proportional to the delay of dependencies.
-\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell of the matrix (ex: maximum along one line).
-Hence such information can be maintained in a single place (wavefront) and progress with the computation.
-\ole
+\subsubsection{Serial problems}
 
-The first 2 criteria provide the theoretical complexity of the dynamic program.
-Indeed, DP proceeds by filling every element of the matrix, and for each element of the matrix, compute the aggregate over all dependencies, hence the overall complexity is equal to the product of the complexities.
+{\color{red}
+Split into blocks:\ul
+\item Decide the shape of the blocks
+\item Decide the size of the blocks
+\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
+\ule
+
+XXX: make it work up to 14K for all 3 problems using multiple kernels
+XXX: example of non-symmetric serial problem
 
@@
 
-delay
-polyadic
-
-We might need to introduce new parameters for serial
-- delay (access 1 or up to k elements backwards)
-- number of cost matrices to maintain
-
-- size of wavefront
-
-Implementation: number of matrices, size of data, size of wavefront
-
-Since in some special case, transforming it might be possible to transform a non-serial 
-We propose a slightly different classification as
-
-non-serial monadic -> serial polyadic
-
-XXX: transformation to polyadic serial into monadic serial
-
--> aggregation
--> delay
-
-monadic / polyadic defines the complexity of the backtrack
+The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
+Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
+}
 

### 2012-10-24 docs/roadmap/benchmarks.tex
Pre splitting rework.

@@
 2048 & 32 & CPU baseline & 27229.28 & 15296.09 & 57323.04 \\
-2048 & 32 & GPU P 32 blocks & 2864.27 & 1427.21 & 2096.34 \\
+2048 & 32 & GPU P $\le32$ blocks & 2864.27 & 1427.21 & 2096.34 \\
+4096 & 32 & GPU 8 splits & & & 16767.58 \\
+8192 & 32 & GPU 64 splits & & & 135793.22 \\
+12288 & 32 & GPU 256 splits & & & 460912.44 \\
+
 \bottomrule \end{tabular} \\[4pt]

### 2012-10-24 docs/roadmap/benchmarks.tex
Allowed splits for the 3 shapes.

--- 2012-10-23 docs/roadmap/benchmarks.tex
@@
-\input{style.sty}
+%\input{style.sty}
 
-\title{Benchmarks}
-\begin{document}
-\maketitle
-\pagestyle{headings}
+%\title{Benchmarks}
+%\begin{document}
+%\maketitle
+%\pagestyle{headings}
+
+\newpage
+\section{Benchmarks}
 
@@
 \begin{center}\begin{tabular}{cclrrr} \toprule
-\bf Matrix & \bf Block & \bf Comment & \bf R(ms) & \bf T(ms) & \bf P(ms) \\ \midrule
-1024 & 32 & CPU baseline & 1965.63 & 1191.13 & 6069.66 \\
-1024 & 32 & GPU baseline & 838.67 & 499.87 & 516.82 \\
-1024 & 32 & GPU sync improved & 642.37 & 316.40 & 343.29 \\
-2048 & 32 & CPU baseline & 27229.28 & 15296.09 & 57323.04 \\
-2048 & 32 & GPU P 32 blocks & 2864.27 & 1427.21 & 2096.34 \\
+\bf Matrix & \bf Block & \bf Comment & \bf R(s) & \bf T(s) & \bf P(s) \\ \midrule
+1024 & 1 & CPU					& 1.965		& 1.191		& 6.069 \\
+2048 & 1 & CPU					& 27.229		& 15.296		& 57.323 \\
+4096 & 1 & CPU					& 			& 177.608	&  \\
+1024 & 32 & GPU baseline			& 0.838		& 0.500		& 0.516 \\
+1024 & 32 & GPU sync improved		& 0.642		& 0.316		& 0.343 \\
+2048 & 32 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
+4096 & 32 & GPU 8 splits				& timeout		& 9.285		& 16.767 \\
+8192 & 32 & GPU 64 splits			&			& 62.064		& 135.793 \\
+12288 & 32 & GPU 256 splits			&			& 196.971	& 460.912 \\
 \bottomrule \end{tabular} \\[4pt]
@@
 
-% --------------------------------------------------------------------------------
-\bibliographystyle{plain}
-\bibliography{bibliography.bib}
-\end{document}
+%\end{document}

### 2012-10-24 docs/roadmap/draft.tex
Allowed splits for the 3 shapes.

--- 2012-10-23 docs/roadmap/problems.tex
@@
 
-% Recurrence visualisation helpers
-\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
-\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
-\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
-\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
-\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
-	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
-\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
-
-\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
-\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
-
-\title{DP Problems of Interest}
+\title{Draft}
 \begin{document}
@@
 \pagestyle{headings}
-
-% ----------------------------------------------------------------------------------------------------------------------------------------
-\section{Problems classification}
-\subsection{Definitions}\ul
-\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
-\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
-\ule
-
-\subsection{Classification}
-In the literature, dynamic programming problems (DP) are classified according to two criteria:
-
-\textbf{Monadic/polyadic}\ul
-\item \textbf{Monadic:} on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
-\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
-\ule
-
-\textbf{Serial/non-serial} \ul
-\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial (it accesses only 2 cells backward).
-\item \textbf{Non-serial:} when the solution depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
-\ule
-
-\subsection{Simplifications}
-\subsubsection{Calculus}
-In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
-	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
-	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
-Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
-	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
-	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]
-
-This simplification avoids the non-serial dependencies at the cost of extra storage in the wavefront, unfortunately, it might be applicable only for some special cases.
-
-\subsubsection{Precomputations}
-When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block.
-For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.
-
-On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.
-We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).
-
-% ----------------------------------------------------------------------------------------------------------------------------------------
-\section{Problems of interest}
-We describe problems structures: inputs, cost matrices and backtracking matrix.
-These all have an alphabet (that must be bounded in terms of bit-size).
-Unless otherwise specified, we adopt the following conventions:\ul
-\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length.
-\item Number are all unsigned integers
-\item Problem dimension is $m,n$ (or $n$) indices $i,j$ ranges are respectively $0\le i<m$, $0\le j<n$.
-\item Unless otherwise specified, the recurrence applies to all non-initialized matrix elements.
-\ule
-We describe the problem processing in terms of both initialization and recurrences.
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman (simple)}\label{sswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
-\item Matrices: $M_{m \times n}, B_{m \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
-	\ule
-\item Initialization:\ul
-	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-		0 & stop\\
-		M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-		M_{(i-1,j)}-d & N\\
-		M_{(i,j-1)}-d & W
-	\end{array}\right\}=B_{(i,j)} \]
-
-\item Backtracking: starts from the cell $\max \{M_{(m,j)} \cup M_{(i,n)}\}$, stops at
-the first cell containing a $0$.
-\item Visualisation: by convention, we put the longest string vertically ($m\ge n$):
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
-	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
-	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-	\Cd{6,5}{0.8}
-	\Cd[1,0]{5,4}{0.8}
-	\Cd[1,-1]{5,5}{0.8}
-\Cm\end{picture}\end{center}
-
-\item Optimizations:\ul
-	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built).
-This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
-	\item Padding: since first line and column of the matrix are zeroes, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
-Also since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
-Hence valid input indices are $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
-We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
-	\ule
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman with affine gap extension cost}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
-\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
-	\ule
-\item Initialization:\ul
-	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
-	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-\item Recurrence for the cost matrices:
-\[\begin{array}{rcl}
-M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	E_{(i,j)} & N\\
-	F_{(i,j)} & W
-\end{array}\right\}=B_{(i,j)}\\
-\\
-E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	M_{(i, j-1)} - \alpha & NW\\
-	E_{(i,j-1)} - \beta & N\\
-\end{array}\right\}=B_{(i,j)}\\
-\\
-F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	M_{(i-1,j)} - \alpha & NW\\
-	F_{(i-1,j)} - \beta & W\\
-\end{array}\right\}=B_{(i,j)}
-\end{array}\]
-
-That can be written alternatively as:
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
-	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
-\end{array}\right\}=B_{(i,j)} \]
-
-Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is 
-\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
-Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
-
-\item Backtracking: same as \nameref{sswat}
-\item Visualisation: same as \nameref{sswat}
-\item Optimizations: same as \nameref{sswat}
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
-Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.
-Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
-\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
-	\ule
-
-\item Initialization:\ul
-	\item Match cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-
-\item Recurrence:
-
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	\max_{1 \le k \le j-1} M_{(i,j-k)} - g(k) & N_k\\
-	\max_{1 \le k \le i-1} M_{(i-k,j)} - g(k) & W_k\\
-\end{array}\right\}=B_{(i,j)} \]
-
-\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells.
-\item Visualisation:
-
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
-%		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-%		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
-	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-		\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
-		\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\Cd[1,-1]{5,5}{0.8}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations: The dependencies here are non-serial, there is no optimization that we can 
-apply out of the box here.
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Convex polygon triangulation}\ol
-\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
-We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is a lower triangular matrix compacted in memory (rows are contiguous) with a 0 diagonal that is omitted \footnote{Arbitrary convention for both architectural implementation and code generator.
-Rationale: in lower triangular matrix, element address is independent of the matrix size.}, hence $|S|=\tfrac{n^2}{2}=N$.
-
-\item Matrices: $M_{n\times 2n}, B_{n \times 2n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
-\item Alphabets:\ul
-	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
-	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
-	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
-	\ule
-\item Initialization: $M_{(i,i)}=0, M_{(i,i+1)}=0, B_{(i,i)}=stop \quad\forall i$
-
-\item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-	It is interesting to note that even in the sequential world, this problem is solved 
-	by filling the diagonals, ie. computing sub-solutions for all polygons of size $k$ before
-	those of size $k+1$.
-
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for the smaller polygons:
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
-		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
-
-\item Visualisation: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
-"cyclical" in nature.
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,9)
-	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cfl{\Cg}\Cfd{\Cz}\Cfu{\Cg}
-	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-	\put(3.5,4.5){\line(3,-1){3}}
-	\put(4.5,4.5){\line(2,-2){2}}
-	\put(5.5,4.5){\line(1,-3){1}}
-	%\Cd[1,-1]{5,2}{0.8}
-\Cmlong\end{picture}\end{center}
-
-\item Optimizations: we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
-{\color{red}}
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Matrix chain multiplication}\ol
-\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$.
-Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$.
-\textit{<<r=rows, c=columns>>}
-\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
-\item Alphabets:\ul
-	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
-	\item Cost matrix: $\Sigma(M)= 1..z$ with $z\le n\cdot \big[ \max_i(r_i,c_i) \big]^3 $.
-	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
-	\ule
-\item Initialization:\ul
-	\item Cost matrix: $M_{(i,i)}=0$.
-	\item Backtrack matrix: $B_{(i,i)}=stop$.
-	\ule
-\item Recurrence: $c_k=r_{k+1}$
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
-		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
-	\end{array}\right\}=B_{(i,j)} \]
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
-		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
-
-\item Visualisation:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-		\Cfl{\Cg}\Cfd{\Cz}
-		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations:\ul
-	\item We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$ and we obtain a new recurrence relation:
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l} M_{(i,k)}+M_{(2i-1 -k,j)}+r_i \cdot c_k \cdot c_j \end{array}\right. \]
-	With the initialization at $M_{(i,n-i-1)}$
-	\ule
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Nussinov algorithm}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
-\item Matrices: $M_{n\times n}, B_{n \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrix: $\Sigma(M)=\{0..n\}$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
-	\ule
-\item Initialization: \ul
-	\item Cost matrix: $ M_{(i,i)}=M_{(i,i-1)}=0$
-	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i-1)}=stop$
-	\ule
-\item Recurrences:
-	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-		M_{(i+1,j-1)}+\omega(i,j) & D\\
-		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
-	\end{array}\right\} = B_{(i,j)} \]
-	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
-\item Backtracking: Start the backtracking in $B_{(1,n)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
-\item Visualisation:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-		\Cfl{\Cg}\Cfd{\Cz}
-		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\Cd[1,1]{5,3}{0.8}
-		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations: note that this is very similar to the matrix multiplication except that we also need the diagonal one step backward, so the same optimization can apply.
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Zuker folding}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
-\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrices:\ul
-		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n*b+c$
-		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
-		\ule
-	\item Backtrack matrices: \ul
-		\item $\Sigma(BW)=\{stop, S,W,V,k\}$
-		\item $\Sigma(BV)=\{stop, HL, IL, SW, (i,j) , k\}$ with $0\le i,j,k < n$ \\
-		$HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
-
-		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
-		\ule
-	\ule
-\item Initialization:\ul
-	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
-	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
-	\ule
-\item Recurrence:
-\[\begin{array}{rcl}
-W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	W_{(i+1,j)}+b & S\\
-	W_{(i,j-1)}+b & W\\
-	V_{(i,j)}+\delta(S_i,S_j) & V \\
-	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
-\end{array}\right\} = BW_{(i,j)}\\
-\\
-V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
-	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
-	V_{(i+1,j-1)}+es(i,j) & IL \\
-	VBI_{(i,j)} & (i',j') \\
-	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
-\end{array}\right\} = BV_{(i,j)}\\
-\\
-F_{(j)}&=&\min\left\{\begin{array}{l|l}
-	F_{(j-1)} & L \\ 
-	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
-\end{array}\right\} = BF_{(j)}\\
-\\
-VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}
-\end{array}\]
-
-In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$ {\color{red} $W$,$V$ and $F$}.
-
-\item Backtracking: Start at $BF_{(n)}$ using the recurrences
- \[\begin{array}{rcl}
-	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
-	\\
-	BV_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
-		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
-		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
-		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
-	\end{array}\right.\\
-	\\
-	BW_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
-	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
-	V &\implies& BV_{(i,j)} \\
-	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
-	\end{array}\right.
-\end{array}\]
-
-\item Visualisation: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
-	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
-\item Optimizations: {\color{red} XXX: we need to find how to encode $F_{(j)}$ such that we can normalize our computation into an existing recurrence shape.}
-\ole
-\newpage
-
-% ------------------------------------------------------------------------------------------------
-\section{Related problems}
+\setcounter{tocdepth}{2} \tableofcontents
 
@@
 \newpage
-\section{Plan}
-\ol
+\section{Planning/draft}
+\subsubsection*{Contributions}\ul
+\item Identify DP classes and give a generalization of their shape/dependency graph
+\item Provide state of the art/on par parallel implementations for all these classes
+\item Generalize/simplify expression by using a DSL to implement these problems
+\item Provide multi-platform support (CPU/GPU/FPGA)
+\ule
+
+\subsubsection*{Todo @TCK}
+\begin{verbatim}
+- backtrack on GPU for non-serial
+- timeout workaround for non-serial / alleviate timeout on in-memory => up to 14K
+- deadlock check/limit for non-serial (i.e. make sure no deadlock)
+- serial implementation for large memory
+\end{verbatim}
+
+\subsubsection*{Todo @Manohar}
+\begin{verbatim}
+
+\end{verbatim}
+
+Steps/timeline
+\begin{verbatim}
+- Theorie (?|?)
+- Implementation CPU+GPU pour P/NS (3 problemes: rect, tri, parall)
+- Implementation CPU+GPU pour _/S
+- Lien avec le DSL
+- Benchmarks
+- Optimisations
+\end{verbatim}
+
+\subsection*{Plan} \ol
 \item {\color{gray}
@@
 
-\newpage
-\subsection{LMS compiler stack}
-\textbf{User-language}: define additional parameters for the recurrence\ul
-\item Windowing (to convert non-serial into serial problems)
-\item Input sizes, and alphabets (backtrack, input, cost)
-\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
-\item Recurrence functions, devices available
-\item What to keep in memory (cost, backtrack or both).
-\ule
-$\Downarrow$ Conversion (using an existing technique)
+% --------------------------------------------------------------------------------
+\input{problems}
+\input{benchmarks}
 
-\textbf{Intermediate representation}
-
-$\Downarrow$ Optimizations\ul
-\item Transform non-serial into serial \ul
-	\item Use aggregation functions/transformations
-	\item Use windowing from user (if no other technique succeed)
-	\ule
-\item Define the wavefront depth
-\item Avoiding the cost matrix by moving it into the wavefront
-\ule
-
-\textbf{Code specification}\ul
-\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
-\item Device-level interface => setup the block sizes(w/h), input and memory sizes
-\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
-\item Define the co-processor memory aggregation function
-\item Define the scheduling of the blocks and aggregation (software pipelining)
-\item Define the data movement back and forth to disk
-\ule
-
-$\Downarrow$ Generation\ul
-\item Generate the kernel for specific device
-\item Generate the scheduling and barriers
-\ule
-
-\textbf{Binary program}
-
-%\ol
-%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
-%\item Let the user tune the window size if he wants to reduce non-serial to serial.
-%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
-%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
-%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
-%	\item Scheduler (CPU): interleave block computation and memory statistics
-%	\ule
-%\item Common description \ul
-%	\item Block kernel processor
-%	\item Full block
-%	\ule
-%\item Discussion: wavefront, design similarities, polyhedral theory
-%\ole
-
-%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
-%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
-%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
-%$\begin{array}{|C|C|C|C|C|C|} \hline
-%0 & 0 & 0 & 0 & 0 & 0 \mnl
-%0 &  &  &  &  & \mnl
-%0 & M_{23}  &  &  &  & \mnl
-%0 &  & \sum  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%\end{array}$
+% --------------------------------------------------------------------------------
+\bibliographystyle{plain}
+\bibliography{bibliography.bib}
 \end{document}

### 2012-10-24 docs/roadmap/problems.tex
Allowed splits for the 3 shapes.

@@
-\input{style.sty}
+%\input{style.sty}
 
@@
 
-\title{DP Problems of Interest}
-\begin{document}
-\maketitle
-\pagestyle{headings}
+%\title{DP Problems of Interest}
+%\begin{document}
+%\maketitle
+%\pagestyle{headings}
 
 % ----------------------------------------------------------------------------------------------------------------------------------------
+\newpage
 \section{Problems classification}
@@
 Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
+Also notice that it is possible to propagate $E$ and $F$ inside a resp.
+horizontal and vertical wavefront.
 
@@
 \section{Related problems}
+The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered thoroughly at the design time.
+
+\subsection{Serial problems}
+\begin{tabular}{llll} \toprule
+\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
+Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
+Smith-Waterman \footnotesize affine gap extension & rectangle & 1 & 2 \\
+\end{tabular}
+
+\subsection{Non-serial problems}
+\begin{tabular}{llll} \toprule
+\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
+Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
+Convex polygon triangulation & parallelogram & 1 & -- \\
+Matrix chain multiplication & triangle & 1 & -- \\
+Nussinov & triangle & 1 & -- \\
+Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
+\end{tabular}
 
 {\color{red}
-XXX: describe common parts between problems (sub categories)
+I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems. We need to describe common parts between problems (sub categories).
 
-XXX: Add additional problems into their related category
+Add additional problems into their related category:
 \ul
@@
 \ule
+
+XXX: see comments
 }
+
+%
+%Dijkstra shortest path, Fibonacci, Balanced 0-1 matrix, Checkerboard, Tower of Hanoi
+%Recurrent solutions to lattice models for protein-DNA binding
+%Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
+%Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
+%Many string algorithms including longest common subsequence, longest increasing subsequence, longest common substring, Levenshtein distance (edit distance).
+%Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
+%The Cocke-Younger-Kasami (CYK) algorithm which determines whether and how a given string can be generated by a given context-free grammar
+%Knuth's word wrapping algorithm that minimizes raggedness when word wrapping text
+%The use of transposition tables and refutation tables in computer chess
+%The Viterbi algorithm (used for hidden Markov models)
+%The Earley algorithm (a type of chart parser)
+%The Needleman-Wunsch and other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.
+%Floyd's all-pairs shortest path algorithm
+%Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
+%The dynamic time warping algorithm for computing the global distance between two time series
+%The Selinger (a.k.a. System R) algorithm for relational database query optimization
+%De Boor algorithm for evaluating B-spline curves
+%Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
+%The Value Iteration method for solving Markov decision processes
+%Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
+%Some methods for solving interval scheduling problems
+%Some methods for solving word wrap problems
+%Some methods for solving the travelling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
+%Recursive least squares method
+%Beat tracking in music information retrieval.
+%Adaptive-critic training strategy for artificial neural networks
+%Stereo algorithms for solving the correspondence problem used in stereo vision.
+%Seam carving (content aware image resizing)
+%The Bellman-Ford algorithm for finding the shortest distance in a graph.
+%Some approximate solution methods for the linear search problem.
+%Kadane's algorithm for the maximum subarray problem.
+%	http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
+%
+%Shortest path in DAGs
+%Longest increasing subsequences
+%Knapsack
+%Shortest path
+%All pair shortest paths
+%Independent sets in trees
+%=> also see exercises for more problems
+%	http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
+%
+%Longest Increasing Subsequence
+%Longest Common Subsequence
+%Subset Sum
+%Coin Change
+%Family Graph
+%	http://www.algorithmist.com/index.php/Dynamic_Programming
+%
+%Longest Increasing Subsequence
+%Edit/Levenshtein distance
+%Optimal Binary Search Trees
+%	http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
+%
+%Independent set on a tree
+%0-1 Knapsack
+%Floyd-Warshall
+%	http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
 
@@
 \newpage
-\section{Plan}
-\ol
-\item {\color{gray}
-\textbf{Problems:} finish the problem description.
-\\ We might look towards parallel tree-raking, but it does not share much with the above algorithms (sparse version of above computations, might not scale that well).
-Most of the common patterns are already enclosed by the above problems.
-Real input size is around 300'000 (we may want to target the million, using disk).
-We might want to also look at an $O(n^3)$-space complexity problem (like matching 3 strings $S,T,U$).
-}
-
-All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
-\item Serial dependencies
-\item Non-serial horizontal or vertical dependencies (1D non-serial)
-\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
-\item We have not found other type of dependencies in the literature
-\ule
-
-\item \textbf{User facing language:} goals are flexibility and compactness.\ul
-	\item Design the user-facing language that we want to support, which should be similar to related papers <<Algebraic Dynamic Programming>> and <<ADP fusion>>.
-We want to reuse their transformation to map (problem description) $\mapsto$ (kernel implementation) for a single element.
-	\item We also may want to try to make implicit transformation for code like \\
-		{\tt @DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
-	\item Windowing: the user should be able to force a windowing (i.e. force a non-serial problem to be a $k$-polyadic serial problem).	
-	\item Consider 3 different cases: we care about backtrack, the costs or both.
-	\item Backtracking: we might want to create an operator/class that given an item produces the previous in the backtrack, or the whole sequence in correct order, or only indices.
-	\ule
-
-$\implies$ \emph{end of October}.
-
-\item \textbf{Prototyping:} get a prototype to understand difficulties and share common base.
-\\ Implement a working prototype of \nameref{aswat} on CPU (for correctness), and specific platform (CUDA/FPGA).
-This will give us an idea of how to implement the general case.
-We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
-Here we aim to do as good an implementation for the specific platform (CPU/GPA) as possible.\\
-$\implies$ \emph{end of October}.
-
-\item \textbf{Baseline:} Also use benchmarks provided by existing implementations as baselines.\\
-$\implies$ \emph{end of October}.
-
-\item \textbf{Formalize IR:} According to experiences, describe the intermediate representation, also formalize the framework that will be provided to the code generators (i.e.
-memory management, ...).
-\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
-\item \textbf{Test and benchmark:} make sure our implementations are correct and compare them with previous papers implementations.
-\ole
-
-\newpage
 \subsection{LMS compiler stack}
@@
 %\end{array}$
-\end{document}
+%\end{document}

### 2012-10-24 src/test/include/ns.h

@@ -244,7 +244,7 @@ void dbg_track(bool gpu, FILE* f) {
 
 // Checks that the CPU and GPU implementation return the same costs
 // Note that backtrack may vary depending the ordering of paths selection
-void dbg_compare(bool full=false) {
+void dbg_compare() {
 	double ttc=0,ttg=0; cuTimer t;
 	TC* tc=(TC*)malloc(sizeof(TC)*MEM_MATRIX);
 	TB* tb=(TB*)malloc(sizeof(TB)*MEM_MATRIX);
@@ -263,10 +263,11 @@ void dbg_compare(bool full=false) {
 	int err=0;
 	#ifdef SH_RECT // some extra memory is needed, we want to ignore it
 	for (int i=0;i<M_H;++i) for (int j=0;j<M_W;++j) { if (tc[idx(i,j)]!=c_cost[idx(i,j)]) ++err; }
+	// for (int j=0;j<M_W;++j) { for (int i=0;i<M_H;++i) printf("%c",tc[idx(i,j)]!=c_cost[idx(i,j)]?'!':' '); printf("\n"); }
 	#else
 	for (unsigned i=0;i<MEM_MATRIX;++i) { if (tc[i]!=c_cost[i]) ++err; }
 	#endif
-	fprintf(stderr,"- Matrix cpu(%.3f)/gpu(%.3f) : ",ttc,ttg);
+	fprintf(stderr,"- Matrix cpu(%.3f)/gpu(%.3f) : ",ttc/1000,ttg/1000);
 	if (err==0) fprintf(stderr,"identical.\n"); else fprintf(stderr,"WARNING %d ERRORS !!\n",err);
 	free(tc);
 	free(tb);
@@ -274,10 +275,10 @@ void dbg_compare(bool full=false) {
 	#ifdef __CUDACC__
 
 	// XXX: implement backtrack comparison
-	
+
 	#else
 	fprintf(stderr,"NO GPU implementation available\n");
-	#endif	
+	#endif
 }
 
 // Initialize some structures

### 2012-10-24 docs/roadmap/draft.tex
Init C code.

@@
 - backtrack on GPU for non-serial
-- timeout workaround for non-serial / alleviate timeout on in-memory => up to 14K
-- deadlock check/limit for non-serial (i.e. make sure no deadlock)
+- (timeout workaround / splitting up to 14K->12K)
 - serial implementation for large memory
@@
 \begin{verbatim}
-
+- generate chain matrix recurrences to generate implementation
 \end{verbatim}

### 2012-10-24 docs/roadmap/problems.tex
Init C code.

@@
 \\
+VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
+\\
 F_{(j)}&=&\min\left\{\begin{array}{l|l}
@@
 	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
-\end{array}\right\} = BF_{(j)}\\
-\\
-VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}
+\end{array}\right\} = BF_{(j)}
 \end{array}\]
 
-In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$ {\color{red} $W$,$V$ and $F$}.
+In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$.
 

### 2012-10-24 src/main/c/matmul.cu

@@ -1,0 +1,54 @@
+#include "include/common.h" // XXX: fix path
+#define SH_TRI
+#define B_W 32LU
+#define B_H 32LU
+#define M_W 128LU
+#define M_H 128LU
+
+////////////////////////////////////////////////////////////////////////////////
+// Data types
+typedef struct { unsigned rows,cols; char print() { return 'X'; } } mat_t;
+#define TI mat_t         // input data type
+#define TI_CHR(X) ('0'+(X).rows) // conversion to char (debug)
+#define TC unsigned long // cost type
+#define TB short         // backtrack type (2 bits for direction + 14 for value)
+
+// Initialization
+#define INIT(i,j) (j<=i) // matrix initialization at [stop]
+
+// Matrix multiplication parenthesizing (triangular matrix)
+//
+//   M[i,j]= min {i<=k<j} M[i,k] + M [k+1,j] + r_i * c_k * c_j
+//
+#define p_kernel \
+	_infinity \
+	_min_loop(k,i,j,  _cost(i,k) + _cost(k+1,j) + _in(i).rows * _in(k).cols * _in(j).cols, k )
+
+// Once generated: execute
+// nvcc -arch=sm_21 -O2 <FILE>.cu -DSH_RECT -o <OUT>
+// optirun <OUT>
+////////////////////////////////////////////////////////////////////////////////
+
+// Input
+TI* p_input() {
+	static unsigned s=time(NULL); mseed(s); // keep consistent
+	TI* in = (TI*)malloc(M_H*sizeof(TI));
+	#define RNZ ({ unsigned x; do { x=mrand()%10; } while (!x); x; })
+	in[0].rows=RNZ;
+	for (unsigned i=1;i<M_H;++i) { in[i-1].cols=in[i].rows=RNZ; }
+	in[M_H-1].cols=RNZ;
+	return in;
+}
+
+// XXX: fix path
+#include "include/small.h"      // common functions
+#include "include/small_cpu.h"  // cpu implementation
+#include "include/small_gpu.h"  // gpu implementation
+
+int main(int argc, char** argv) {
+	dbg_init();
+	dbg_compare();
+	dbg_cleanup();
+	return 0;
+}
+

### 2012-10-24 src/main/c/run.sh

@@ -1,0 +1,12 @@
+#!/bin/sh
+if [ "$OSTYPE" = "darwin" ]; then
+	CCARGS="-arch=sm_30 -O2 --linker-options -rpath,/usr/local/cuda/lib"
+	EXEC="./test"
+else
+	CCARG="-arch=sm_20 -O2"
+	EXEC="optirun ./test"
+fi
+
+nvcc $CCARGS test.cu -DSH_RECT -o test && $EXEC
+nvcc $CCARGS test.cu -DSH_TRI -o test && $EXEC
+nvcc $CCARGS test.cu -DSH_PARA -o test && $EXEC

### 2012-10-25 docs/roadmap/benchmarks.tex
Scalability fixes.

@@
 2048 & 32 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
-4096 & 32 & GPU 8 splits				& timeout		& 9.285		& 16.767 \\
-8192 & 32 & GPU 64 splits			&			& 62.064		& 135.793 \\
-12288 & 32 & GPU 256 splits			&			& 196.971	& 460.912 \\
+4096 & 32 & GPU 8 splits				& 21.902		& 9.285		& 16.767 \\
+8192 & 32 & GPU 64 splits			& 159.058	& 62.064		& 135.793 \\
+12288 & 32 & GPU 256 splits			& 419.030	& 196.971	& 460.912 \\
 \bottomrule \end{tabular} \\[4pt]
@@
 \item Synchronization  between blocks\ul
-	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but still remains correct, this validates the observation made by \cite{gpu_barrier}.
+	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}.
 Speedup for <1024/32/*> are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).

### 2012-10-25 src/main/c/test.cu

@@ -7,7 +7,7 @@
 // Problem dimensions
 #define B_W 32LU    // block width
 #define B_H 32LU    // block height
-#define M_W 128LU   // matrix dimension (at most 14336LU for in-memory, 12288LU OK)
+#define M_W 128LU   // matrix dimension (<=14336LU theo., <=12288LU actually)
 #define M_H 128LU   // matrix dimension
 //#define SPLITS 8  // number of kernels to be successively launched
 
@@ -18,25 +18,35 @@
 #include "include/small_gpu.h"  // gpu implementation
 // -----------------------------------------------------------------------------
 
+//#define FAST
+
 int main(int argc, char** argv) {
 	cuTimer t;
 	dbg_init();
+	#ifndef FAST
 	// CPU solving
 	for (int i=0;i<1;++i) { t.start(); c_solve(); t.stop(); }
 	fprintf(stderr,"- CPU: "); t.print(); fprintf(stderr,"\n");
 	fflush(stderr);
+	#endif
 
 	#ifdef __CUDACC__
 	// GPU solving
-	for (int i=0;i<4;++i) { t.start(); g_solve(); t.stop(); }
+		#ifdef FAST
+		const unsigned loops=1;
+		#else
+		const unsigned loops=4;
+		#endif
+	for (int i=0;i<loops;++i) { t.start(); g_solve(); t.stop(); }
 	fprintf(stderr,"- GPU: "); t.print(); fprintf(stderr,"\n");
 	#endif
 
-	dbg_compare(); // XXX: also compare backtrack
-
+	#ifndef FAST
+	dbg_compare();
 	//dbg_print(false,stdout);
 	dbg_track(false,stdout);
 	dbg_track(true,stdout);
+	#endif
 
 	dbg_cleanup();
 	return 0;

### 2012-10-25 docs/roadmap/draft.tex
Scalability fixes.
Docs nits.

@@
 \begin{verbatim}
-- backtrack on GPU for non-serial
-- (timeout workaround / splitting up to 14K->12K)
 - serial implementation for large memory
+- optimizations for in-memory computations
+- CPU+GPU implementation for S (NS?) problems larger than device memory
 \end{verbatim}
@@
 - generate chain matrix recurrences to generate implementation
-\end{verbatim}
-
-Steps/timeline
-\begin{verbatim}
-- Theorie (?|?)
-- Implementation CPU+GPU pour P/NS (3 problemes: rect, tri, parall)
-- Implementation CPU+GPU pour _/S
-- Lien avec le DSL
-- Benchmarks
-- Optimisations
+- more theory behind what we want to support, how we encompass all cases, ...
 \end{verbatim}
@@
 \item {\color{gray}
-\textbf{Problems:} finish the problem description.
-\\ We might look towards parallel tree-raking, but it does not share much with the above algorithms (sparse version of above computations, might not scale that well).
-Most of the common patterns are already enclosed by the above problems.
-Real input size is around 300'000 (we may want to target the million, using disk).
-We might want to also look at an $O(n^3)$-space complexity problem (like matching 3 strings $S,T,U$).
-}
+\textbf{Problems description:} parallel tree-raking, does not share much with other algorithms (sparse version of computations, might not scale efficiently).
+Most common patterns are already enclosed by the above problems.
+Real input size is around 300K.
+We might want to also look at an $O(n^3)$-space-complex problem (like matching 3 strings $S,T,U$).}
 
@@
 \item \textbf{User facing language:} goals are flexibility and compactness.\ul
-	\item Design the user-facing language that we want to support, which should be similar to related papers <<Algebraic Dynamic Programming>> and <<ADP fusion>>.
-We want to reuse their transformation to map (problem description) $\mapsto$ (kernel implementation) for a single element.
+	\item User-facing language should be similar to related paper \cite{adp_gpu} or \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}.
+We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
 	\item We also may want to try to make implicit transformation for code like \\
@@
 	\item Windowing: the user should be able to force a windowing (i.e. force a non-serial problem to be a $k$-polyadic serial problem).	
-	\item Consider 3 different cases: we care about backtrack, the costs or both.
-	\item Backtracking: we might want to create an operator/class that given an item produces the previous in the backtrack, or the whole sequence in correct order, or only indices.
+	\item 3 different cases: we care about backtrack cost or both.
+	\item Backtracking: create an operator that produces the whole backtrack sequence indices.
 	\ule
@@
 
-\item \textbf{Formalize IR:} According to experiences, describe the intermediate representation, also formalize the framework that will be provided to the code generators (i.e.
-memory management, ...).
+\item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
 \item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
-\item \textbf{Test and benchmark:} make sure our implementations are correct and compare them with previous papers implementations.
+\item \textbf{Benchmark:} make sure implementations are correct, compare them other papers.
+\item \textbf{Optimizations:} improve as much as possible / as long as time permits
 \ole

### 2012-10-25 docs/roadmap/problems.tex
Computations.

@@
 \subsection{Large problems}
-To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$.
+To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$. For simplification, we assume a square matrix made of square blocks with $b$ blocks per row/column.
 
@@
 Non-serial problems need to potentially access all elements that have been previously computed.
-Assuming the restriction that dependencies are only along the rows and columns and that the serial dependencies delay is $\le$ than one block size, we need all the block on the same line and column and one in diagonal.
+We restrict\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.} ourselves to the following dependencies: \ul
+\item Non-serial dependencies along row and column
+\item Serial dependencies along diagonal, with delay smaller or equal to one block size
+\ule
+Such restriction implies that all the block of the line and the row, and one additional block to cover diagonal dependencies must be held in memory (independently of the matrix shape).
 
-For simplification, let $m=n$, assume that we have $b$ rectangular block per row and per column. We need to transfer forth and back (in terms of blocks):
-	\[b \cdot b +% computed line (once)
-		(b-1) \cdot (b-1) + % diagonal block (1 per block)
-		\sum_{i=0}^{b-1} i \cdot b % column blocks (for line i)
-		= \tfrac{1}{2}b^2(b+5)-2b+1\]
+For simplification, let $m=n$ and assume that we have $b$ square blocks per row and per column. Hence we have the following memory restriction:
+\[ 2\frac{n}{b}(S_I+S_W) + 2 \cdot \frac{n^2}{b}S_C + \frac{n^2}{b^2} S_B \le M_D\]
 
-With $b=8$ we have 401 block transfer (forth and back), which corresponds to transferring $12.5\times$ the whole matrix, and the problem size can now be (following the same formula and assertion as for small problems) $2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = 2n+10n^2/8 \le 2^{30} \implies n\le 29307$, so we roughly doubled matrix size at the cost of $12.5\times$ matrix transfer time plus indices overhead.
+We also need to take into account the transfer between main memory (or disk) and device memory.
+Dependency blocks only need to be read, computed blocks need to be written back.
+Focusing on the cost blocks only:
+\[\begin{array}{rclll}
+		b^2 +% computed line writeback (once)
+		(b-1)^2 + % diagonal block loads (1 per block)
+		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
+		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
+	\\
+		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
+		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
+	\\
+		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
+		&=& b^3 &\qquad& \rm (Parallelogram)
+\end{array}\]
+
+Backtrack blocks need only to be write back and optionally reloaded during backtrack, hence we need to transfer at most:
+\[\begin{array}{cccll}
+	b^2&+&2b &\qquad& \rm (Rectangle)\\
+	\tfrac{1}{2}b^2 + \tfrac{1}{2}b &+& 2b-1 % binary tree with b leafs, hence b-1 interior nodes.
+		&\qquad& \rm (Triangle)\\
+	b^2 &+& 2b-1 &\qquad& \rm (Parallelogram)
+
+\end{array}\]
+Since matrix can grow larger, we store backtrack on 4 bytes. Reusing other assertions from small problems we have $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$. From first constraint we have
+\[
+\frac{n}{b} (1 + 4n + 2\tfrac{n}{b}) \le 2^{29}
+\quad\Longleftarrow\quad 5n^2 \le b \cdot 2^{29} \]
+Since blocks can not divide cells, we must have $b\le n$ hence $n\le 2^{29}/5 \approx 1.07\cdot 10^8$.
+
+{\color{red} XXX: Continue rewriting here
+
+With $b=8$ we have 337 block transfer, which corresponds to transferring $5.3\times$ the whole matrix, and the problem size can now be (same assertions as for small problems) $2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = 2n+10n^2/8 \le 2^{30} \implies n\le 29307$, so we roughly doubled matrix size at the cost of $12.5\times$ matrix transfer time plus indices overhead.
+
+Assume a different implementation might be worth to support problems order of magnitude larger, hence $n \approx 120K$ thus we need $2^{30} > 2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = $
+}
 

### 2012-10-25 docs/roadmap/problems.tex
More reports.

@@
 Dependency blocks only need to be read, computed blocks need to be written back.
-Focusing on the cost blocks only:
+Ignoring the backtrack and focusing only on the cost blocks, the transfers (in blocks) are:
 \[\begin{array}{rclll}
@@
 
-Backtrack blocks need only to be write back and optionally reloaded during backtrack, hence we need to transfer at most:
-\[\begin{array}{cccll}
-	b^2&+&2b &\qquad& \rm (Rectangle)\\
-	\tfrac{1}{2}b^2 + \tfrac{1}{2}b &+& 2b-1 % binary tree with b leafs, hence b-1 interior nodes.
-		&\qquad& \rm (Triangle)\\
-	b^2 &+& 2b-1 &\qquad& \rm (Parallelogram)
+Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
+\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}
 
-\end{array}\]
-Since matrix can grow larger, we store backtrack on 4 bytes. Reusing other assertions from small problems we have $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$. From first constraint we have
-\[
-\frac{n}{b} (1 + 4n + 2\tfrac{n}{b}) \le 2^{29}
-\quad\Longleftarrow\quad 5n^2 \le b \cdot 2^{29} \]
-Since blocks can not divide cells, we must have $b\le n$ hence $n\le 2^{29}/5 \approx 1.07\cdot 10^8$.
+Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in between 10 and 25 minutes of delay due to transfers.
+{\color{blue} Using our preliminary benchmarks, such a computation would require at least 7.6 days, assuming there is no other scalability issues.}
+
+%Backtrack blocks need only to be write back and optionally reloaded during backtrack, hence we need to transfer at most:
+%\[\begin{array}{cccll}
+%	b^2&+&2b &\qquad& \rm (Rectangle)\\
+%	\tfrac{1}{2}b^2 + \tfrac{1}{2}b &+& 2b-1 % binary tree with b leafs, hence b-1 interior nodes.
+%		&\qquad& \rm (Triangle)\\
+%	b^2 &+& 2b-1 &\qquad& \rm (Parallelogram)
+%	
+%\end{array}\]
+%Since matrix can grow larger, we store backtrack on 4 bytes. Reusing other assertions from small problems we have $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$. From first constraint we have
+%\[
+%\frac{n}{b} (1 + 4n + 2\tfrac{n}{b}) \le 2^{29}
+%\quad\Longleftarrow\quad 5n^2 \le b \cdot 2^{29} \]
+%Since blocks can not divide cells, we must have $b\le n$ hence $n\le 2^{29}/5 \approx 1.07\cdot 10^8$.
+%
+%\textbf{Transfer time is the limiting factor to scalability}\\
+%Since we are interested in an efficient implementation of DP, we want to have an acceptable ratio
+%\[R= \frac{\text{transfer time}}{\text{computation time}}\]
+
+%Let's count the minimal transfer time for various sizes using all the available memory.
 

### 2012-10-25 src/test/ship.cu

@@ -13,131 +13,60 @@ __global__ void fun1(int* a, int* b, unsigned size) {
 
 }
 
+/*
+ * Processing 992 Mb of data
+ * CPU: 162.10 ms
+ * GPU: 483.20 ms (including transfers)
+ * GPU: 122.69 ms (without transfers)
+ * => 5.3743 GB/s for transfer
+ */
+
 int main(int argc,char** argv) {
 	cudaDeviceReset();
-	const unsigned size=3<<21; // 3<<24 = max can be reserved for 512MB GPU
-	// 1<<22 with backtrack
+	// We transfer 2 pairs of int, hence 8 bytes, my card mem is 2^30 bytes
+	const unsigned size=31<<22; // we need some memory for the program itself
+	fprintf(stderr,"Processing %ld Mb of data:\n",size*sizeof(int)*2 /1024/1024);
+
 	const unsigned int ms=size*sizeof(int);
-
-	// 3<<24 = 48M pairs of (4+4) chars => 384Mb data to and back from GPU
-	// Note that it goes faster with -d64 on my CPU (!)
-
 	int* ca = (int*)malloc(ms);
 	int* cb = (int*)malloc(ms);
-	int* ga; cuErr(cudaMalloc(&ga,ms));
-	int* gb; cuErr(cudaMalloc(&gb,ms));
-
-for (int i=0;i<10;++i) {
-	struct timeval ts,te;
-	gettimeofday(&ts, NULL);
-	for (unsigned i=0;i<size;++i) { OP(ca,cb,i) }
-
-	gettimeofday(&te, NULL);
-	printf("CPU time: %.3f ms\n", (float)( 1000.0*(te.tv_sec-ts.tv_sec) + 0.001*(te.tv_usec-ts.tv_usec) ) );
+	int* ga; cuMalloc(ga,ms);
+	int* gb; cuMalloc(gb,ms);
 
 	cudaStream_t stream;
 	cuErr(cudaStreamCreate(&stream));
+	cuTimer tc,tg,tp;
 
-	// Test 1: forth and back memory transfer
-	cuSync(stream); gettimeofday(&ts, NULL);
-	cuPut(ca,ga,ms,stream);
-	cuPut(cb,gb,ms,stream);
-	fun<<<size/32, 32, 0, stream>>>(ga,gb,size);
[…]

### 2012-10-26 docs/roadmap/draft.tex
Feeding some problems.

@@
 \begin{verbatim}
-- serial implementation for large memory
+- continue performance evaluation
+- class problem into categories
+- think of hashmap implementation. is it useful ?
+- CPU+GPU implementation for serial (NS?) problems larger than device memory
 - optimizations for in-memory computations
-- CPU+GPU implementation for S (NS?) problems larger than device memory
+
 \end{verbatim}

### 2012-10-26 docs/roadmap/problems.tex
Feeding some problems.

--- 2012-10-25 docs/roadmap/problems.tex
@@
 \section{Related problems}
-The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered thoroughly at the design time.
+The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.
 
 \subsection{Serial problems}
-\begin{tabular}{llll} \toprule
+\begin{tabular}{llcc} \toprule
 \bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
 Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
-Smith-Waterman \footnotesize affine gap extension & rectangle & 1 & 2 \\
+Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & -- \\
+\href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
 \end{tabular}
@@
 \subsection{Non-serial problems}
-\begin{tabular}{llll} \toprule
+\begin{tabular}{llcc} \toprule
 \bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
@@
 Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
+\href{http://en.wikipedia.org/wiki/CYK_algorithm}{Cocke-Younger-Kasami (CYK)} & triangle & \#rules & -- \\
+
 \end{tabular}
 
+\subsection{Other problems}\ul
+\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
+\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
+\item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
+\item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
+
+\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
+
 {\color{red}
-I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems. We need to describe common parts between problems (sub categories).
-
-Add additional problems into their related category:
-\ul
-\item \url{http://en.wikipedia.org/wiki/Dynamic_programming}
-\item \url{http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf}
-\item \url{http://www.algorithmist.com/index.php/Dynamic_Programming}
-\item \url{http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf}
-\item \url{http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf}
+\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
+\item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
+}
 \ule
 
-XXX: see comments
+{\color{red}
+Other algorithms to consider/process:\ul
+\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
+\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
+\item \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} (sequences alignement)
+\ule
+See more in comments
 }
 
-%
-%Dijkstra shortest path, Fibonacci, Balanced 0-1 matrix, Checkerboard, Tower of Hanoi
+Other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.
+
+% PROBLEM LISTS
+% I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems.
+% http://en.wikipedia.org/wiki/Dynamic_programming
+% http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
+% http://www.algorithmist.com/index.php/Dynamic_Programming
+% http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
+% http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
+% ------------------------------------------
+% Balanced 0-1 matrix
 %Recurrent solutions to lattice models for protein-DNA binding
@@
 %Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
-%Many string algorithms including longest common subsequence, longest increasing subsequence, longest common substring, Levenshtein distance (edit distance).
 %Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
-%The Cocke-Younger-Kasami (CYK) algorithm which determines whether and how a given string can be generated by a given context-free grammar
-%Knuth's word wrapping algorithm that minimizes raggedness when word wrapping text
 %The use of transposition tables and refutation tables in computer chess
-%The Viterbi algorithm (used for hidden Markov models)
-%The Earley algorithm (a type of chart parser)
-%The Needleman-Wunsch and other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.
-%Floyd's all-pairs shortest path algorithm
 %Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
@@
 %Seam carving (content aware image resizing)
-%The Bellman-Ford algorithm for finding the shortest distance in a graph.
 %Some approximate solution methods for the linear search problem.
 %Kadane's algorithm for the maximum subarray problem.
-%	http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
+% =====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
 %
 %Shortest path in DAGs
-%Longest increasing subsequences
 %Knapsack
@@
 %=> also see exercises for more problems
-%	http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
+% =====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
 %
-%Longest Increasing Subsequence
-%Longest Common Subsequence
 %Subset Sum
@@
 %Family Graph
-%	http://www.algorithmist.com/index.php/Dynamic_Programming
+% =====> http://www.algorithmist.com/index.php/Dynamic_Programming
 %
-%Longest Increasing Subsequence
-%Edit/Levenshtein distance
 %Optimal Binary Search Trees
@@
 %0-1 Knapsack
-%Floyd-Warshall
-%	http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
+% =====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
 
@@
 
-Backtrack blocks need only to be write back and optionally reloaded during backtrack, hence we need to transfer at most:
-\[\begin{array}{cccll}
-	b^2&+&2b &\qquad& \rm (Rectangle)\\
-	\tfrac{1}{2}b^2 + \tfrac{1}{2}b &+& 2b-1 % binary tree with b leafs, hence b-1 interior nodes.
-		&\qquad& \rm (Triangle)\\
-	b^2 &+& 2b-1 &\qquad& \rm (Parallelogram)
+Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
+\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}
 
-\end{array}\]
-Since matrix can grow larger, we store backtrack on 4 bytes. Reusing other assertions from small problems we have $S_C=4, S_B=4, S_I=1, S_W=0$ and $M_D=2^{30}$. From first constraint we have
-\[
-\frac{n}{b} (1 + 4n + 2\tfrac{n}{b}) \le 2^{29}
-\quad\Longleftarrow\quad 5n^2 \le b \cdot 2^{29} \]
-Since blocks can not divide cells, we must have $b\le n$ hence $n\le 2^{29}/5 \approx 1.07\cdot 10^8$.
-
-{\color{red} XXX: Continue rewriting here
-
-With $b=8$ we have 337 block transfer, which corresponds to transferring $5.3\times$ the whole matrix, and the problem size can now be (same assertions as for small problems) $2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = 2n+10n^2/8 \le 2^{30} \implies n\le 29307$, so we roughly doubled matrix size at the cost of $12.5\times$ matrix transfer time plus indices overhead.
-
-Assume a different implementation might be worth to support problems order of magnitude larger, hence $n \approx 120K$ thus we need $2^{30} > 2n + 5 \cdot \big(  (n/b)^2 \cdot 2b \big) = $
-}
-
-{\color{red} XXX: Do we want to support larger non-serial problem that do not fit in memory? Is it worth implementing ? Argue.}
+Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay.
+Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h\up{(T)} and 6 days 10h\up{(P)}, assuming there is no other scalability issues.
+Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
+Given that real problems -- like RNA folding -- operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
 
 \subsubsection{Serial problems}
+The serial problem have the interesting property to access to a fixed number of previous elements. These elements
+
+propose a memory consumption scheme
+
+try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront
 
 {\color{red}
+
+XXX: what's the maximal size of the wavefront ??
+
+XXX: can we avoid to store some matrices and put them in the wavefront ??
+
 Split into blocks:\ul

### 2012-10-26 docs/roadmap/problems.tex
Nits.

@@
 \section{Related problems}
-The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.
+The goal of this section is to demonstrate that our framework can accommodate with many {\color{red} (20-50)} problems that we have not considered at the design time.
 
@@
 Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & -- \\
+\href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} & rectangle & 1 & --\\
+
 \href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
@@
 \href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \\
 \end{tabular}
@@
 Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
-\href{http://en.wikipedia.org/wiki/CYK_algorithm}{Cocke-Younger-Kasami (CYK)} & triangle & \#rules & -- \\
+\href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
 
@@
 \item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
-
 \item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
+\item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial
 
@@
 \item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
+\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
+\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
+\item \href{http://en.wikipedia.org/wiki/Maximum_subarray_problem}{Kadane maximum subarray} 1D serial, look at 
+\href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
+\item Structural alignment (MAMMOTH, SSAP), \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}.
+\item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
+\item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
 }
@@
 
-{\color{red}
-Other algorithms to consider/process:\ul
-\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
-\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
-\item \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} (sequences alignement)
-\ule
-See more in comments
-}
+\begin{verbatim}
+- Balanced 0-1 matrix
+- Recurrent solutions to lattice models for protein-DNA binding
+- Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
+- Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
+- Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
+- Transposition tables and refutation tables in computer chess
+- Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
+- The dynamic time warping algorithm for computing the global distance between two time series
+- The Selinger (a.k.a. System R) algorithm for relational database query optimization
+- Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
+- The Value Iteration method for solving Markov decision processes
+- Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
+- Some methods for solving interval scheduling problems
+- Some methods for solving word wrap problems
+- Some methods for solving the traveling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
+- Beat tracking in music information retrieval.
+- Adaptive-critic training strategy for artificial neural networks
+- Stereo algorithms for solving the correspondence problem used in stereo vision.
+- Seam carving (content aware image resizing)
+- Some approximate solution methods for the linear search problem.
+=====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
 
-Other algorithms used in bioinformatics, including sequence alignment, structural alignment, RNA structure prediction.
+- Shortest path in DAGs
+- Shortest path
+- All pair shortest paths
+- Independent sets in trees
+=> also see exercises for more problems
+=====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
+- 
+- Subset Sum, Coin Change, Family Graph
+=====> http://www.algorithmist.com/index.php/Dynamic_Programming
 
-% PROBLEM LISTS
-% I guess we need between 20-50 DP problems to demonstrate that we identified almost all types of problems.
-% http://en.wikipedia.org/wiki/Dynamic_programming
-% http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-% http://www.algorithmist.com/index.php/Dynamic_Programming
-% http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
-% http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
-% ------------------------------------------
-% Balanced 0-1 matrix
-%Recurrent solutions to lattice models for protein-DNA binding
-%Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
-%Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
-%Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
-%The use of transposition tables and refutation tables in computer chess
-%Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
-%The dynamic time warping algorithm for computing the global distance between two time series
-%The Selinger (a.k.a. System R) algorithm for relational database query optimization
-%De Boor algorithm for evaluating B-spline curves
-%Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
-%The Value Iteration method for solving Markov decision processes
-%Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
-%Some methods for solving interval scheduling problems
-%Some methods for solving word wrap problems
-%Some methods for solving the travelling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
-%Recursive least squares method
-%Beat tracking in music information retrieval.
-%Adaptive-critic training strategy for artificial neural networks
-%Stereo algorithms for solving the correspondence problem used in stereo vision.
-%Seam carving (content aware image resizing)
-%Some approximate solution methods for the linear search problem.
-%Kadane's algorithm for the maximum subarray problem.
-% =====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
-%
-%Shortest path in DAGs
-%Knapsack
-%Shortest path
-%All pair shortest paths
-%Independent sets in trees
-%=> also see exercises for more problems
-% =====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-%
-%Subset Sum
-%Coin Change
-%Family Graph
-% =====> http://www.algorithmist.com/index.php/Dynamic_Programming
-%
-%Optimal Binary Search Trees
-%	http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
-%
-%Independent set on a tree
-%0-1 Knapsack
-% =====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
+- Optimal Binary Search Trees
+=====> http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
+
+- Independent set on a tree
+- 0-1 Knapsack
+=====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
+\end{verbatim}
 

### 2012-10-29 docs/roadmap/draft.tex
Plan for upcoming work.

@@
 \begin{verbatim}
-- continue performance evaluation
-- class problem into categories
-- think of hashmap implementation. is it useful ?
+1. check the vanilla implementation
+2. read ADP legacy parsers
+3. presentation for Thursday
+4. Propose an IR and code generators related to
+Note. We only need to generate code directly for CUDA, no interaction. Look at string templating .
+"""xxx $var xxx"""
+- Fork lms on github and cleanup all the CUDA generator
+
+% class problem into categories
+% think of hashmap implementation. is it useful ?
 - CPU+GPU implementation for serial (NS?) problems larger than device memory
 - optimizations for in-memory computations
-
 \end{verbatim}

### 2012-10-29 docs/roadmap/problems.tex
Plan for upcoming work.

@@
 \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} & rectangle & 1 & --\\
 
@@
 \href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & rectangle & 1 & -- \\
+\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & triangle & 1 & -- \\
 \href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
@@
 \href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
+\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} \footnotesize (input sensitive) & rectangle & 1 & --\\
 

### 2012-10-30 docs/roadmap/draft.tex
Nits.

@@
 """xxx $var xxx"""
-- Fork lms on github and cleanup all the CUDA generator
+- Fork LMS on GitHub and cleanup all the CUDA generator
 
@@
 - optimizations for in-memory computations
+
+1. restriction vs ADP: we give only the best answer, do we care about semi-optimal ones?
+2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
+3. use sub-cell parallelism (seen in 2 papers already)
+
+Plan:
+We need to extend the language to support all 3 cases
+- Parallelogram is triangular + cyclic => add a cyclic keyword that applies to the whole problem
+- Square: what is the transform being done in ADP extension?
+
+1. optimize: terminals of bounded yield + binary decomposition if possible
+2. transform into: axioms (init/fixed) + rules (iterations)
+3. transform to code
+     axiom(i := k(i)) -> M[i,i]=k(i)
+     rule(x~y := h(x,y)) -> M[i,j]=forall_i<k<j h(M[i,k],M[k,j]) for both unbounded yield
+     rule(x~y_l := h(x,y)) -> M[i,j]=forall_j-l<k<j h(M[i,k],M[k,j]) for y bounded by l
+     rule(x~y_l := h(x,y)) -> M[i,j]=h(M[i,j-l],M[j-l,j]) for y of exactly l
+
+XXX: how to encode multi-dimensional matrices
+1. assume they have the same type put one after another => different dimensions ok
+2. assume of same size => put into a struct
+=> but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)
 \end{verbatim}

### 2012-10-31 docs/roadmap/draft.tex
Roadmap.

@@
 
+\begin{verbatim}
+restrict to only 1 max/min (more restrictive than ADP)
+
+for optimization we want to split grammar into binary productions, to do that we need
+to have equality on the following formula. We can solve it more easily if functions are linear
+\end{verbatim}
+\[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]
+
+\begin{verbatim}
+Roadmap:
+2.11 Come with an IR prototype: data structures and nodes
+9.11 GenScala on LMS, JNI working for arbitrary types
+16.11 GenCuda + LMS CudaCompiler
+23.11 Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization
+30.11-- Gap due to LMS missing knowledge
+7.12 Benchmarking, grammar analysis
+14.12
+21.12
+28.12 -------
+4.01 -------
+11.01 writing report = implementation description + plan for future work
+18.01 writing report
+\end{verbatim}
+
 \subsubsection*{Todo @TCK}
 \begin{verbatim}
-1. check the vanilla implementation
-2. read ADP legacy parsers
-3. presentation for Thursday
+cleanup ADP code
 4. Propose an IR and code generators related to

### 2012-11-12 docs/roadmap/problems.tex
Backtrack solution.

@@
 \newpage
+\section{User facing language}
+\subsection{Parsing grammar (ADP)}
+Borrowing from ADP XXX
+
+XXX: explain ADP concepts
+
+XXX: normalization, transforms applied
+
+\subsection{Backtracking}
+In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
+To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack reads are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
+To deal with this problem, we are facing two options:\ul
+\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack and is not implemented in vanilla ADP (unless the whole backtrack is made part of the scoring: big performance impact and non-constant memory requirement issues make its GPU implementation hard and not desirable).
+Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
+
+\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem.
+To do that we need to apply the following transformations\ol
+	\item Normalize rules and identify them uniquely by \ul
+		\item Removing all the <<or>> operators by breaking a parser into multiple subrules
+		\item Forwarding names: replace a rule that is a single invocation by its content (i.e. name a terminal parser with its own name instead of its containing parser's)
+		\item Assigning each subrule an unique identifier $r$ and create a mapping table $T: r \to$ user-defined name of the parser.
+	\ule
+	\item The data elements corresponding to a rule are (scoring, backtracking) and named after the tabulation.
+The scoring is a user-defined primary type (int, double, ...) and the backtracking is a tuple $({\rm id}_{\rm subrule}, k_1, k_2, \ldots , k_m)$ where $m$ is the maximal number of splits that can occur in the subrules.
+	\item During the matrix computation of cell $(i,j)$, if the subrule $r$ applies, the backtrack will be set as $(r,k_1,k_2,\ldots,k_{m_r})$ (obviously with $i\le k_1\le k_2\le \ldots k_{m_r}\le j$ and $m_r\le m$).
+	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r,k_1,k_2,\ldots,k_m)$, given $r$, we recover the subrule hence the user-defined name of the rule that applies and $k_{m_r}$, which allows us to enqueue the subwords $(i,k_1), (k_1,k_2), ..., (k_{m_r})$ for further backtracking.
+If $r$ refers to a terminal, we stop the backtracking.
+	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $(r,i,j)$ where $r$ is the index of the sub-rule that can be easily mapped into the user name (such that $T: r_u \to$ user-defined rule name) and $(i,j)$ is the subword on which the rules applies.
+\ole
+In short, we break parsers into normalized rules, the backtracking information is the rule id (which rule to unfold) and a list of indices (how to unfold it).
+\ule
+Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed by the user program in-order, applying for each user-defined rule the underlying transformation, and storing intermediate results (i.e.
+in a hash map) until they are processed by another rule.
+Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.
+
+% ------------------------------------------------------------------------------------------------
+\newpage
 \section{Implementation}
@@
 \subsubsection{Serial problems}
-The serial problem have the interesting property to access to a fixed number of previous elements. These elements
+The serial problem have the interesting property to access to a fixed number of previous elements.
+These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into a wavefront.
+Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems) or line-wise or column-wiser.
+This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
+horizontally) along the matrix.
+
+Since serial problems are of complexity $O(n^2)$
+
+Different strategies
+XXX: Do we need/want to use divide and conquer?\ul
+\item Divide and co
+\item
+\ule
 

### 2012-11-12 docs/roadmap/benchmarks.tex

@@
 2048 & 32 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
-4096 & 32 & GPU 8 splits				& 21.902		& 9.285		& 16.767 \\
+4096 & 32 & GPU 8 splits				& 21.902		& 8.841		& 16.767 \\
 8192 & 32 & GPU 64 splits			& 159.058	& 62.064		& 135.793 \\
@@
 	\ule
+\item Using multiple threads per matrix cell:
+In the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increase (there is one more dependency).
+According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity.
+However, in our setup (memory layout+algorithm+hardware), we did not found any improvement by doing so.
+We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
+\item
+
+First we need to prove that we use almost all the available memory bandwidth: for matrix multiplication, in a triangular matrix, we have
+\[\text{Total transfer}=\frac{n(n+1)}{2} \text{ writes} + \sum_{i=0}^{n-1} 2 i \cdot (n-i) \text{ reads}\]
+where each write is 10 bytes (long+short), and each read is 8 bytes (long). For $n=4096$ we transfer
+% n*(n + 1)/2*12 + Sum[2*i*(n - i), {i, 0, n - 1}]*8
+183'352'614'912 bytes which corresponds to 183.35GB.
+In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254 \rm GB$.
+Hence we can say that 72\% of the algorithm running time is spent into memory accesses.
+
+\item On a 4096 matrix, if we assume that ATLP card would have the same bandwidth as our card, their running time would be
+\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 9.43\rm s_{\text{ ATLP}} > 8.84\rm s_{\text{ our}}\]
+Which shows that our algorithm is comparable to theirs.
+However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have ATLP source code).
+\ule
+As a conclusion, (1) we must remain away to invalidate their result as previous hardware generations might be subject to more constraint to our hardware and (2) we are on par, if not better with one of the best current implementation.
+
+\item Reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
+Even worse, it slows down slightly the computation.
+We might attribute this to a better constant transformation by the compiler.
+Hence, having many idle threads does not impede performance.
 \ule

### 2012-11-12 docs/roadmap/problems.tex
Nits.

@@
 \subsection{Parsing grammar (ADP)}
+{\color{red}
 Borrowing from ADP XXX
@@
 XXX: normalization, transforms applied
+}
 
@@
 
-Since serial problems are of complexity $O(n^2)$
+Since serial problems are of complexity $O(n^2)$ (due to the matrix dimension and the finite number of dependencies), it is possible to tackle much larger problem than non-serial during the same running time.
+Hence, it seems obvious to let serial problems grow larger than the memory.
 
-Different strategies
-XXX: Do we need/want to use divide and conquer?\ul
-\item Divide and co
-\item
+Mixing the dependency property and size requirements, we can split the matrix into sub-matrices, store special lines (and/or columns) into memory (or hard disk), and repeat computations to solve the backtrack (similarly as in \cite{swat_gpu},\cite{swat_mega}, but this implementation use problem-specific knowledge that might not generalize).
+
+To store intermediate lines and columns, we are facing two different strategies to explore:\ul
+\item \textbf{Fixed subproblem size:} we decompose the algorithm as follows\ol
+\item Define a grid of <<major column and rows>>, where each cell's data (input, output, cost and backtrack matrices) fits into the device memory.
+\item Compute the values of the grid's major columns and rows in one pass.
+\item Second (on-demand) computation to process backtracking inside relevant cells.
+\ole
+Let $b$ the number of cells that we have on each row/column, the total computation running time would be $(b^2 + 2b) \cdot t_b$ where $t_b$ is the time to compute one cell's matrix.
+This division has the advantage of providing the minimal computation time at the expense of external storage proportional to $O(n)$ (if we store only lines or columns) or $O(n^2)$ (if we store both).
+
+\item \textbf{Myers and Miller’s algorithm:} (divide and conquer)
+This algorithm break the DP problem into 2 (or 4) subproblems such that once the middle line/column is computed, the problem can be solved for 1 submatrix and backtracking among up to 2 of the 3 other.
+This breaking is applied recursively until the submatrix data fits into memory.
+The storage requirements are $4 \cdot O(n)$ (we store along both dimension $1+\tfrac{1}{2}+\tfrac{1}{4}+...$ lines/columns).
+
+The algorithm proceeds as follows: first it solves the problem to obtain the first backtracking element, then it breaks the matrix in 4 submatrices, and refine it until backtrack is tractable.
+Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
+\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used.
 \ule
 
-propose a memory consumption scheme
-
+{\color{red}
 try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront
-
-{\color{red}
 

### 2012-11-12 docs/roadmap/draft.tex
Merged all the documents into the draft. Now we need to do heavy cleanup.

--- 2012-10-30 docs/roadmap/draft.tex
@@
-\input{style.sty}
+\documentclass[11pt]{article}
+\input{inc/style.sty}
 
-\title{Draft}
+\title{DynaProg for Scala}
+\subtitle{A Scala DSL for Dynamic Programming on GPU and FPGA}
 \begin{document}
-\maketitle
+%\maketitle
+\shorttitle
+
+\subsection*{Abstract}
+Dynamic programming (solving a problem by leveraging subproblems solutions) is a common pattern used in various domains.
+Yet underlying matrix recurrences are non-trivial and error prone.
+Additionally, domain experts might not have the skills to make an efficient parallel implementation.
+In this paper, we leverage the Scala / LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs.
+Our contributions are: \ul
+\item Classification of dynamic programming problems, and generalization of their characteristics % shape, dependency graph
+\item State of the art parallel implementation of these classes on GPUs
+\item Grammar embedded in Scala (DSL) to express dynamic programming problems efficiently
+%\item Normalization of the grammar into efficient productions
+\item Code generator to transform a grammar into efficient code for CPU, GPU, and FPGA
+\item Systematic approach to generate and process backtracking information
+\ule
+
 \pagestyle{headings}
@@
 \newpage
+\section{Introduction}
+\subsection{Dynamic programming}
+\subsection{Scala and LMS}
+
+% ------------------------------------------------------------------------------------------------
+\input{problems}
+\input{benchmarks}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
 \section{Planning/draft}
-\subsubsection*{Contributions}\ul
-\item Identify DP classes and give a generalization of their shape/dependency graph
-\item Provide state of the art/on par parallel implementations for all these classes
-\item Generalize/simplify expression by using a DSL to implement these problems
-\item Provide multi-platform support (CPU/GPU/FPGA)
-\ule
+\subsubsection*{Roadmap}
+\begin{verbatim}
+2.11 Come with an IR prototype: data structures and nodes
+9.11 GenScala on LMS, JNI working for arbitrary types
+16.11 GenCuda + LMS CudaCompiler
+23.11 Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization
+30.11-- Gap due to LMS missing knowledge
+ 7.12 Benchmarking, grammar analysis
+14.12
+21.12
+28.12 --- holiday ---
+ 4.01 --- holiday ---
+11.01 Writing report: implementation description and plan for future work
+18.01 Writing report
+\end{verbatim}
 
@@
 \begin{verbatim}
-1. check the vanilla implementation
-2. read ADP legacy parsers
-3. presentation for Thursday
 4. Propose an IR and code generators related to
@@
 """xxx $var xxx"""
-- Fork LMS on GitHub and cleanup all the CUDA generator
-
-% class problem into categories
-% think of hashmap implementation. is it useful ?
+- Fork LMS on GitHub and cleanup the CUDA generator
 - CPU+GPU implementation for serial (NS?) problems larger than device memory
@@
 2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
-3. use sub-cell parallelism (seen in 2 papers already)
+3. use sub-cell parallelism (seen in 2 papers already) => does not improve
 
@@
 
-% --------------------------------------------------------------------------------
-\input{problems}
-\input{benchmarks}
+\begin{verbatim}
+ideas.txt
+---------------------------------------------------
+Plan:
+1. Define the main problem in focus and its class
+   - Dimension
+   - Serial/non-serial
+   - Monadic/polyadic
+2. Define some other problems that belong to this class or simpler classes
+3. See the feasibility for FPGA (possibly helped by CPU)
+4. Implement this algorithm in both CUDA and FPGA
+   -> specify all parameters so that we can compare both implementations
+5. Find a common IR for the problems
+6. Write a code generator for both architectures from IR
+   -> make sure all the test-DP are compiled well from IR
+7. Find an expressive and compact language for the user-facing representation
 
-% --------------------------------------------------------------------------------
-\bibliographystyle{plain}
+Translation into C++:
+http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
+
+----------------------------------
+
+Ideas for the language:
+- we want maximum flexibility
+- backtrack can be custom, we may want just to give back indices
+- two problems: we care about cost and we care about whole backtrack.
+	@DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)
+  should be translated into the more complete syntax that allows more flexibility
+  => get inspiration from papers to write an internal language
+
+- we may want to keep track of indices but we do not want to give access to the backtrack matrix / cost matrix
+
+Language side:
+- define a simple way to represent some DP programs
+- define a comprehensive way to define them => micro-syntax
+- define the mapping into IR (actually the IR might BE the micro-syntax)
+- define optimizations to IR
+  - making serializable by using aggregation functions/transformations
+  - avoiding the cost matrix by moving it into the wavefront
+- scheduling (with memory loads for non-serializable)
+- code generation for both platforms
+  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)
+
+----------------------------------
+
+Core function F:
+- in: s,t strings
+- in: neighbor costs: top, left, top+left
+- in: neighbor stats: top, left, top+left
+- out: backtrack information
+- out: cost(i,j)
+- out: stats
+
+To be most efficient, we need to get exact bit-sizes of
+- strings' char
+- costs: min/max values
+- stats: min/max values
+- backtrack
+
+=> we might want to parallelize processing within a thread that operate on 32 or 64 bit inputs (to reduce memory accesses)
+
+Some ideas:
+- define a way to pack the characters => less memory transfer (i.e. GATC=>4 letters in 1 char)
+- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
+
+- write a kernel that takes stats from left and r
+
+		stats (x)
+		 ||
+		 vv
+stats -> KK -> new_stats (y')
+(y)		 || \
+		 vv  --+ backtrack info (Bxy)
+		 new_stats (x')
+
+and compute the backtracking for its cell (optional as template boolean)
+
+- multi-grain
+  + within the 64b-words : multiple letters at once (thread level)
+  + group threads into blocks that operate on (block level)
+  + kernel that operates one after another (GPU/CPU level)
+    => can we catch a stream's event at CPU level to alloc memory to get back results ?
+- keep track of
+  + stats O(m+n)
+  + backtrack O(m*n)
+
+- are we sure we need squares ?
+  - rectangles can increase the full usage of all threads during (x-y) runs
+  - assuming we put longest word vertically we play hot potato for stats
+    left->right => block.x exchanges within block, last thread writes to
+    global exchange memory (if short, so that it does not penalize running time)
+  - we may go up to running in stripes
+
+- if we put in diagonal-major data in the memory, why not stream all chunks like
+  that using a cyclic buffer (?) => no empty slot
+  ==> this might be useful for the backtrack information
+
+- use extensively profiling : CUDA profiler
+
+- Robust DP paper provides 2 insights: coalesced access + GPU synchronization
+  => can we do a producer-consumer scenario at the block border so that we can
+     execute a large diagonal horizontal/vertical swipe made of multiple blocks?
+  => multiple swipes at different delays (so that we avoid syncing at every step
+
+          s-->
+  	  +----------------/------------+
+	t | B1            /             |
+	| +------/-------/              | KERNEL1 (keep benefit of t in shared mem)
+	| | B2  /                       | delay between b1 and b2 can be 1% of line length
+	v +----/------------------------+
+	  |                             |
+	  |                             | KERNEL2
+	Memory organized as
+	b1 -  -
+	b1 b1 b1
+	b1 b1 b1
+	.. .. ..
+	b1 b1 b1
+	b2 b1 b1
+	b2 b2 b1
+	b2 b2 b2
+	.. .. ..
+
+For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
+For backtracking we need to maintain the whole table
+     .. | . |
+	..A | B | min/max: Min(D)=Min(Min(C),Min(B))
+	----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
+	..C | D | avg    : sum both #cells and values then divide at appropriate cell
+	----+---+
+
+we many not need to store what's the previous cell, however, the previous
+cell information (usually limited range) is much more compact than the score(32-64 bits)
+
+Memory usage of CPU:
+
+we can potentially extend the GPU's memory by storing some information in the CPU RAM, which in general is larger.
+For large sequences, this technique may apply to problems where one needs to access elements "all the way to the top" and aggregate them in a non trivial (max,min,sum,avg) way.
+
+Problem parameters:
+- bit size (alphabet,stats,cost,backtrack)
+- dimensionality (2D,3D for the matrix)
+  - 3D problem example is Smith Waterman with arbitrary gap weight function.
+- Summarizability of the set: for computing the stats of an element, we can limit ourselves to computing values up to a certain value
+  -in SW with gap extending penalty, we can summarize easily.
+  -in Nussinov, we _need_ to access all the previous row-column elements for the pairwise sum (without a window limit).
+
+user-interface.txt
+--------------------------------------
+This is how we would like to encode the DP programs into Scala DSL.
+
+I think the best would be to reuse most of the syntactic elements and write it
+as an uneducated programmer would write, apply annotations/additions on top of
+it or detect that it is DP and then converge to our system.
+
+We need to keep in mind that we need both to express the backtracking and the scoring at the same time
+
+Let's build some examples
+
+Smith-Waterman (from problems of interest)
+--------------------------------------------
+
+trait SWOps {
+	type T
+	type Cost = Int
+	type Backtrack = (i,j) => Direction
+	def rightGap(_ :_, c:Char) : [Cost, Backtrack]
+}
+
+def myRec = {
+	def M[i,j] : Int = max {
+		M[i,j-1] + gapRight,
+		M[i-1, j] + gapLeft,
+		M[i-1,j-1] + delta(S(i),T(j))
+	}
+}
+
+def SWat(S0: List[Char], T0:List[Char]) : List[Pair[Char,Char]] = {
+	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function
+
+	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S,T) match {
+		case (_, Nil) => (0, stop)
+		case (Nil, _) => (0, stop)
+		case Si::Sr, Tj::Tr) =>
+			val nw = SWat(Sr,Tr);
+			val n  = SWat(Sr,T);
+			val w  = SWat(S,Tr);
+			val c = cost(Si,Tj);
+
+			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
+			else if (n>=w) then (n._1-d, "N",[Si,_])
+			else (w._1-d, "W",[_,Tj])
+	}
+	backtrack(DP(S0,T0))
+}
+
+def Nussinov(S0: List[Char]) : List[Pair[Char,Char]] = {
+	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function
+
+	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S.zipWithIndex,T.zipWithIndex) match {
+		case (_, Nil) => (0, stop)
+		case (Nil, _) => (0, stop)
+		case (Si::Sr, Tj::Tr) =>
+			val nw = SWat(Sr,Tr);
+			val n  = SWat(Sr,T);
+			val w  = SWat(S,Tr);
+			val cols_rows = 
+			val c = cost(Si,Tj);
+
+			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
+			else if (n>=w) then (n._1-d, "N",[Si,_])
+			else (w._1-d, "W",[_,Tj])
+	}
+
+	backtrack(DP(S0,S0))
+}
+
+- Char, Score, Backtrack, Output are LMS Rep objects so that we can easily detect
+  when we need to use dynamic programming.
+- backtrack() is a new function that take the Backtrack information and reconstruct
+  the desired output backwards
+
+We would unfold this program into two:
+1. The dynamic programming that constructs all the solution and that is the core of the algorithm
+2. The backtrack function that reconstruct the solution and can be run anywhere (Java, CPU, ...)
+
+URLs.txt
+------------------------------------
+Some GPU algorithms:
+http://hgpu.org/?cat=11
+
+CUDPP libraries (but awfully big resulting binary):
+http://code.google.com/p/cudpp/
+
+13 dwarfs:
+http://developer.amd.com/afds/assets/presentations/2155_final.pdf
+
+http://tutorials.jenkov.com/java-reflection/fields.html
+http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
+
+XXX: use TypeClass to put a predicate on types
+--------------------------------------------
+
+def fun[T: CanTranslateToC](...)
+def fun[T](implicit ev:CanTranslateToC[T])
+
+class CanTranslateToC[T] {
+  def translate:String
+}
+
+implicit def canTranslateInt = new CanTranslateToC[Int] = {
+  def translate = "Int"
+}
+
+\end{verbatim}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\bibliographystyle{finplain}
 \bibliography{bibliography.bib}

### 2012-11-12 docs/roadmap/problems.tex
Merged all the documents into the draft. Now we need to do heavy cleanup.

@@
 
-\item Visualization: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
+\item Visualization: \begin{center}\includegraphics[width=8cm]{inc/zucker.pdf}\end{center}
 	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
@@
 XXX: normalization, transforms applied
+
+XXX: Think how to implement {\tt cyclic} problems and {\tt windowing} properties within the language
+
+\begin{verbatim}
+restrict to only 1 max/min (more restrictive than ADP)
+for optimization we want to split grammar into binary productions, to do that we need
+to have equality on the following formula. We can solve it more easily if functions are linear
+\end{verbatim}
+\[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]
+
 }
@@
 Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
-\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}
+\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}
 

### 2012-11-13 docs/roadmap/benchmarks.tex
Rework of the draft.

@@
-%\input{style.sty}
-
-%\title{Benchmarks}
-%\begin{document}
-%\maketitle
-%\pagestyle{headings}
-
 \newpage
 \section{Benchmarks}
-
-The goal of this document is to see how we are comparing with other papers in terms of performance and to note performance progress of the improvements.
+% --------------------------------------------------------------------------------
+\subsection{Related work}
+The goal of this section is to see how we are comparing with other papers in terms of performance and to note performance progress of the improvements.
 
 % --------------------------------------------------------------------------------
-\subsection*{Graphic cards\footnote{Source: \url{http://en.wikipedia.org/wiki/Comparison_of_Nvidia_graphics_processing_units}}}
+\subsubsection*{Graphic cards\footnote{Source: \url{http://en.wikipedia.org/wiki/Comparison_of_Nvidia_graphics_processing_units}}}
 \def\unt#1{& \footnotesize #1}
@@
 % --------------------------------------------------------------------------------
-\subsection*{Results}
-\subsubsection*{ATLP\cite{gpu_atlp}}
+\subsubsection*{Results}
+\textbf{ATLP\cite{gpu_atlp}}
 \begin{center}\begin{tabular}{lrrrrrrrrrr} \toprule
@@
 
-\subsubsection*{SWMB\cite{swat_mega}}
+\textbf{SWMB\cite{swat_mega}}
 \begin{center}\begin{tabular}{rlrr} \toprule
@@
 % --------------------------------------------------------------------------------
+\subsection{Results}
 \subsubsection*{Intermediate results}
@@
 \ule
-
-%\end{document}

### 2012-11-13 docs/roadmap/design.tex
Rework of the draft.

--- 2012-11-12 docs/roadmap/problems.tex
@@
-%\input{style.sty}
-
-% Recurrence visualization helpers
-\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
-\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
-\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
-\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
-\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
-	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
-\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
-
-\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
-\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
-
-%\title{DP Problems of Interest}
-%\begin{document}
-%\maketitle
-%\pagestyle{headings}
-
-% ----------------------------------------------------------------------------------------------------------------------------------------
 \newpage
-\section{Problems classification}
-\subsection{Definitions}\ul
-\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
-\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest.
-Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
-\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute.
-A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
-\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix.
-It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
-\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
-\ule
-
-\subsection{Classification}
-\subsubsection{Litterature classification}
-In the literature, dynamic programming problems (DP) are classified according to two criteria:\ul
-\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman).
-When two or more terms appear, the problem is polyadic (ex: Fibonacci, $F_n = F_{n-1} + F_{n-2}$).
-When a problem is polyadic with index $p$, it also means that its backtracking forms a $p$-ary tree (where each node has at most $p$ children).
-
-\item \textbf{Serial/non-serial:} a problem is serial ($s=0$) when the solutions depends on a fixed number of previous solutions (ex: Fibonacci), otherwise it is said to be non-serial ($s\ge 1$), as the number of dependencies grows with the size of the subproblem.
-That is computing an element of the matrix would require $O(n^s)$.
-(ex: Smith-Waterman with arbitrary gap is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula)
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
-\ule
-
-Note that the algorithmic complexity of a problem is exactly $O\big(n^{d+s}\big)$.
-
-\subsubsection{Calculus simplifications}
-In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
-	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
-	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
-Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
-	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
-	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]
-
-Although this simplification removes some non-serial dependencies at the cost of extra storage in the wavefront, it is not sufficient to transform all non-serial monadic problems into serial problems (ex: this does not apply to Smith-Waterman with arbitrary gap cost).
-
-% ----------------------------------------------------------------------------------------------------------------------------------------
-\section{Problems of interest}
-We usually focus on problem that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$).
-Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability.
-Also we tend to limit algorithmic complexity of the problems as from $O(n^4)$ on, running time becomes a severely limiting factor.
-
-We describe problems structures: inputs, cost matrices and backtracking matrix.
-These all have an alphabet (that must be bounded in terms of bit-size).
-Unless otherwise specified, we adopt the following conventions:\ul
-\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length.
-\item Number are all unsigned integers
-\item Problem dimension is $m,n$ (or $n$) indices $i,j$ ranges are respectively $0\le i<m$, $0\le j<n$.
-\item Unless otherwise specified, the recurrence applies to all non-initialized matrix elements.
-\ule
-We describe the problem processing in terms of both initialization and recurrences.
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman (simple)}\label{sswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
-\item Matrices: $M_{m \times n}, B_{m \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
-	\ule
-\item Initialization:\ul
-	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-		0 & stop\\
-		M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-		M_{(i-1,j)}-d & N\\
-		M_{(i,j-1)}-d & W
-	\end{array}\right\}=B_{(i,j)} \]
-
-\item Backtracking: starts from the cell $\max \{M_{(m,j)} \cup M_{(i,n)}\}$, stops at
-the first cell containing a $0$.
-\item Visualization: by convention, we put the longest string vertically ($m\ge n$):
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
-	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
-	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-	\Cd{6,5}{0.8}
-	\Cd[1,0]{5,4}{0.8}
-	\Cd[1,-1]{5,5}{0.8}
-\Cm\end{picture}\end{center}
-
-\item Optimizations:\ul
-	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built).
-This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
-	\item Padding: since first line and column of the matrix are zeroes, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
-Also since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
-Hence valid input indices are $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
-We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
-	\ule
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman with affine gap extension cost}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
-\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
-	\ule
-\item Initialization:\ul
-	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
-	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-\item Recurrence for the cost matrices:
-\[\begin{array}{rcl}
-M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	E_{(i,j)} & N\\
-	F_{(i,j)} & W
-\end{array}\right\}=B_{(i,j)}\\
-\\
-E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	M_{(i, j-1)} - \alpha & NW\\
-	E_{(i,j-1)} - \beta & N\\
-\end{array}\right\}=B_{(i,j)}\\
-\\
-F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
-	M_{(i-1,j)} - \alpha & NW\\
-	F_{(i-1,j)} - \beta & W\\
-\end{array}\right\}=B_{(i,j)}
-\end{array}\]
-
-That can be written alternatively as:
-\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
-	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
-\end{array}\right\}=B_{(i,j)} \]
-
-Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is 
-\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
-Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
-Also notice that it is possible to propagate $E$ and $F$ inside a resp.
-horizontal and vertical wavefront.
-
-\item Backtracking: same as \nameref{sswat}
-\item Visualization: same as \nameref{sswat}
-\item Optimizations: same as \nameref{sswat}
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
-Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.
-Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
-\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
-	\ule
-
-\item Initialization:\ul
-	\item Match cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
-	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
-	\ule
-
-\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
-	\max_{1 \le k \le j-1} M_{(i,j-k)} - g(k) & N_k\\
-	\max_{1 \le k \le i-1} M_{(i-k,j)} - g(k) & W_k\\
-\end{array}\right\}=B_{(i,j)} \]
-
-\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells.
-\item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
-	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-		\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
-		\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\Cd[1,-1]{5,5}{0.8}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations: The dependencies here are non-serial, there is no optimization that we can 
-apply out of the box here.
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Convex polygon triangulation}\ol
-\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
-We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is a lower triangular matrix compacted in memory (rows are contiguous) with a 0 diagonal that is omitted \footnote{Arbitrary convention for both architectural implementation and code generator.
-Rationale: in lower triangular matrix, element address is independent of the matrix size.}, hence $|S|=\tfrac{n^2}{2}=N$.
-
-\item Matrices: $M_{n\times 2n}, B_{n \times 2n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
-\item Alphabets:\ul
-	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
-	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
-	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
-	\ule
-\item Initialization: $M_{(i,i)}=0, M_{(i,i+1)}=0, B_{(i,i)}=stop \quad\forall i$
-
-\item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-	It is interesting to note that even in the sequential world, this problem is solved 
-	by filling the diagonals, ie. computing sub-solutions for all polygons of size $k$ before
-	those of size $k+1$.
-
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for the smaller polygons:
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
-		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
-
-\item Visualization: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
-"cyclical" in nature.
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,9)
-	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cfl{\Cg}\Cfd{\Cz}\Cfu{\Cg}
-	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-	\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-	\put(3.5,4.5){\line(3,-1){3}}
-	\put(4.5,4.5){\line(2,-2){2}}
-	\put(5.5,4.5){\line(1,-3){1}}
-	%\Cd[1,-1]{5,2}{0.8}
-\Cmlong\end{picture}\end{center}
-
-\item Optimizations: we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
-{\color{red}}
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Matrix chain multiplication}\ol
-\item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$.
-Each matrix $A_i$ is of dimension $r_i \times c_i$ and $c_i=r_{i+1} \forall i$.
-\textit{<<r=rows, c=columns>>}
-\item Matrices: $M_{n \times n}, B_{n \times n}$ \textit{(first, last matrix)}
-\item Alphabets:\ul
-	\item Input: matrix $A_i$ size is defined as pairs of integers $(r_i,c_i)$.
-	\item Cost matrix: $\Sigma(M)= 1..z$ with $z\le n\cdot \big[ \max_i(r_i,c_i) \big]^3 $.
-	\item Backtrack matrix: $\Sigma(B)=\{stop\} \cup \{0..n\}$.
-	\ule
-\item Initialization:\ul
-	\item Cost matrix: $M_{(i,i)}=0$.
-	\item Backtrack matrix: $B_{(i,i)}=stop$.
-	\ule
-\item Recurrence: $c_k=r_{k+1}$
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l|l}
-		M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j & k
-	\end{array}\right\}=B_{(i,j)} \]
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
-		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
-
-\item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-		\Cfl{\Cg}\Cfd{\Cz}
-		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations:\ul
-	\item We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$ and we obtain a new recurrence relation:
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l} M_{(i,k)}+M_{(2i-1 -k,j)}+r_i \cdot c_k \cdot c_j \end{array}\right. \]
-	With the initialization at $M_{(i,n-i-1)}$
-	\ule
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Nussinov algorithm}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
-\item Matrices: $M_{n\times n}, B_{n \times n}$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrix: $\Sigma(M)=\{0..n\}$
-	\item Backtrack matrix: $\Sigma(B)=\{stop,D,1..n\}$
-	\ule
-\item Initialization: \ul
-	\item Cost matrix: $ M_{(i,i)}=M_{(i,i-1)}=0$
-	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i-1)}=stop$
-	\ule
-\item Recurrences:
-	\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
-		M_{(i+1,j-1)}+\omega(i,j) & D\\
-		\max_{i\le k<j}M_{(i,k)}+M_{(k+1,j)} & k
-	\end{array}\right\} = B_{(i,j)} \]
-	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
-\item Backtracking: Start the backtracking in $B_{(1,n)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
-\item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
-		\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
-		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-		\Cfl{\Cg}\Cfd{\Cz}
-		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
-		\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
-		\Cd[1,1]{5,3}{0.8}
-		\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
-	\Cm\end{picture}\end{center}
-
-\item Optimizations: note that this is very similar to the matrix multiplication except that we also need the diagonal one step backward, so the same optimization can apply.
-\ole
-
-% ----------------------------------------------
-\newpage
-\subsection{Zuker folding}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
-\item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
-\item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
-	\item Cost matrices:\ul
-		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n*b+c$
-		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
-		\ule
-	\item Backtrack matrices: \ul
-		\item $\Sigma(BW)=\{stop, S,W,V,k\}$
-		\item $\Sigma(BV)=\{stop, HL, IL, SW, (i,j) , k\}$ with $0\le i,j,k < n$ \\
-		$HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
-
-		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
-		\ule
-	\ule
-\item Initialization:\ul
-	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
-	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
-	\ule
-\item Recurrence:
-\[\begin{array}{rcl}
-W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	W_{(i+1,j)}+b & S\\
-	W_{(i,j-1)}+b & W\\
-	V_{(i,j)}+\delta(S_i,S_j) & V \\
-	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
-\end{array}\right\} = BW_{(i,j)}\\
-\\
-V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
-	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
-	V_{(i+1,j-1)}+es(i,j) & IL \\
-	VBI_{(i,j)} & (i',j') \\
-	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
-\end{array}\right\} = BV_{(i,j)}\\
-\\
-VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}\\
-\\
-F_{(j)}&=&\min\left\{\begin{array}{l|l}
-	F_{(j-1)} & L \\ 
-	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
-\end{array}\right\} = BF_{(j)}
-\end{array}\]
-
-In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$.
-
-\item Backtracking: Start at $BF_{(n)}$ using the recurrences
- \[\begin{array}{rcl}
-	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
-	\\
-	BV_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
-		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
-		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
-		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
-	\end{array}\right.\\
-	\\
-	BW_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
-	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
-	V &\implies& BV_{(i,j)} \\
-	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
-	\end{array}\right.
-\end{array}\]
-
-\item Visualization: \begin{center}\includegraphics[width=8cm]{inc/zucker.pdf}\end{center}
-	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
-\item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?).
-We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?)}
-\ole
-
+\section{Architecture design and technical decisions}
 % ------------------------------------------------------------------------------------------------
-\newpage
-\section{Related problems}
-The goal of this section is to demonstrate that our framework can accommodate with many {\color{red} (20-50)} problems that we have not considered at the design time.
-
-\subsection{Serial problems}
-\begin{tabular}{llcc} \toprule
-\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
-Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
-Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & -- \\
-\href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
-
-\href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/Longest_common_subsequence_problem\#Code_for_the_dynamic_programming_solution}{Longest common subsequence} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/Longest_common_substring_problem\#Pseudocode}{Longest common substring} & triangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \\
-\end{tabular}
-
-\subsection{Non-serial problems}
-\begin{tabular}{llcc} \toprule
-\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
-Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
-Convex polygon triangulation & parallelogram & 1 & -- \\
-Matrix chain multiplication & triangle & 1 & -- \\
-Nussinov & triangle & 1 & -- \\
-Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
-\href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
-\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} \footnotesize (input sensitive) & rectangle & 1 & --\\
-
-\end{tabular}
-
-\subsection{Other problems}\ul
-\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
-\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
-\item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
-\item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
-\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
-\item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial
-
-{\color{red}
-\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
-\item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
-\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
-\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
-\item \href{http://en.wikipedia.org/wiki/Maximum_subarray_problem}{Kadane maximum subarray} 1D serial, look at 
-\href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
-\item Structural alignment (MAMMOTH, SSAP), \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}.
-\item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
-\item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
-}
-\ule
-
-\begin{verbatim}
-- Balanced 0-1 matrix
-- Recurrent solutions to lattice models for protein-DNA binding
-- Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
-- Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
-- Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
-- Transposition tables and refutation tables in computer chess
-- Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
-- The dynamic time warping algorithm for computing the global distance between two time series
-- The Selinger (a.k.a. System R) algorithm for relational database query optimization
-- Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
-- The Value Iteration method for solving Markov decision processes
-- Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
-- Some methods for solving interval scheduling problems
-- Some methods for solving word wrap problems
-- Some methods for solving the traveling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
-- Beat tracking in music information retrieval.
-- Adaptive-critic training strategy for artificial neural networks
-- Stereo algorithms for solving the correspondence problem used in stereo vision.
-- Seam carving (content aware image resizing)
-- Some approximate solution methods for the linear search problem.
-=====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
-
-- Shortest path in DAGs
-- Shortest path
-- All pair shortest paths
-- Independent sets in trees
-=> also see exercises for more problems
-=====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-- 
-- Subset Sum, Coin Change, Family Graph
-=====> http://www.algorithmist.com/index.php/Dynamic_Programming
-
-- Optimal Binary Search Trees
-=====> http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
-
-- Independent set on a tree
-- 0-1 Knapsack
-=====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
-\end{verbatim}
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section{User facing language}
 \subsection{Parsing grammar (ADP)}
@@
 XXX: explain ADP concepts
+
+restriction from ADP: we give only the best answer, do we care about semi-optimal ones? no
 
@@
 % ------------------------------------------------------------------------------------------------
-\newpage
-\section{Implementation}
+\subsection{Memory management}
 We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
@@
 Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
-\subsection{Memory requirements}
+
 We propose to evaluate the device memory requirements to solve the above problem classes. We need first to define additional problem properties related to implementation:\ul
@@
 
-\subsection{Small problems (in-memory)}
+% ----------------------------------------------
+\subsubsection{Small problems (in-memory)}
 Problem that can fit in memory can be solved in a single pass on the device. Such problem must satisfy the equation:
@@
 
-\subsection{Large problems}
+% ----------------------------------------------
+\subsubsection{Large problems}
 To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$. For simplification, we assume a square matrix made of square blocks with $b$ blocks per row/column.
 
+% ----------------------------------------------
 \subsubsection{Non-serial problems}
@@
 
+% ----------------------------------------------
 \subsubsection{Serial problems}
@@
 }
+
+% ------------------------------------------------------------------------------------------------
+\subsection{Memory layout}
 

### 2012-11-13 docs/roadmap/draft.tex
Rework of the draft.

@@
 \subsection*{Abstract}
-Dynamic programming (solving a problem by leveraging subproblems solutions) is a common pattern used in various domains.
-Yet underlying matrix recurrences are non-trivial and error prone.
+Dynamic programming is a common pattern of Computer Science used in various domains.
+Yet underlying matrix recurrences might be difficult to express and error prone.
 Additionally, domain experts might not have the skills to make an efficient parallel implementation.
@@
 Our contributions are: \ul
-\item Classification of dynamic programming problems, and generalization of their characteristics % shape, dependency graph
+\item Classification of DP problems characteristics (matrix shape, dependency graph, ...)
 \item State of the art parallel implementation of these classes on GPUs
-\item Grammar embedded in Scala (DSL) to express dynamic programming problems efficiently
+\item Grammar embedded in Scala (DSL) to express DP problems efficiently
 %\item Normalization of the grammar into efficient productions
-\item Code generator to transform a grammar into efficient code for CPU, GPU, and FPGA
 \item Systematic approach to generate and process backtracking information
+\item Code generator to transform a grammar into efficient code for CPU, GPU and FPGA
 \ule
@@
 \subsection{Dynamic programming}
+Dynamic programming consists of solving a problem by reusing subproblems solutions. The most famous example of dynamic programming is the Fibonacci series that is defined as 
+\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
+
+A typical characteristic of dynamic programming is that a subproblem solution is often reused multiple time, so it is often efficient to store intermediate results in a table to speed-up computation of larger problems.
+
+There exist various categories of dynamic programming:\ul
+\item Series, knapsack : that operates usually on a single dimension
+\item Sequences alignment (matching two sequences at best), top-down grammar analysis, ...
+\item Tree-related algorithms: phylogenetic trees raking, maximum tree independent set, ... 
+\ule
+
+Since the first category is inherently sequential and the third category is hard to parallelize efficiently, we focus on the second type of problems, which is also the most common.
+
+{\color{red}
+parallel tree-raking, does not share much with other algorithms (sparse version of computations, might not scale efficiently).
+Most common patterns are already enclosed by the above problems.
+Real input size is around 300K.
+We might want to also look at an $O(n^3)$-space-complex problem (like matching 3 strings $S,T,U$).
+
+All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
+\item Serial dependencies
+\item Non-serial horizontal or vertical dependencies (1D non-serial)
+\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
+\item We have not found other type of dependencies in the literature
+\ule}
+
 \subsection{Scala and LMS}
@@
 \input{problems}
+\input{design}
+\input{implementation}
 \input{benchmarks}
+\section{Future work}
+\section{Conclusion}
 
@@
 \subsubsection*{Roadmap}
-\begin{verbatim}
-2.11 Come with an IR prototype: data structures and nodes
-9.11 GenScala on LMS, JNI working for arbitrary types
-16.11 GenCuda + LMS CudaCompiler
-23.11 Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization
-30.11-- Gap due to LMS missing knowledge
- 7.12 Benchmarking, grammar analysis
-14.12
-21.12
-28.12 --- holiday ---
- 4.01 --- holiday ---
-11.01 Writing report: implementation description and plan for future work
-18.01 Writing report
-\end{verbatim}
+\begin{tabular}{rl}
+16.11 & Rules normalization and automatic backtracking as in 5.2 \\
+	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
+23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
+30.11 &--- Gap due to LMS missing knowledge \\
+7.12 & Benchmarking, grammar analysis \\
+14.12 & \\
+21.12 & \\
+28.12 & --- holiday --- \\
+ 4.01 & --- holiday --- \\
+11.01 & Writing report: implementation description and plan for future work \\
+18.01 & Writing report
+\end{tabular}
 
@@
 \begin{verbatim}
-4. Propose an IR and code generators related to
-Note. We only need to generate code directly for CUDA, no interaction. Look at string templating .
-"""xxx $var xxx"""
-- Fork LMS on GitHub and cleanup the CUDA generator
+Look at string templating. """xxx $var xxx"""
+- Cleanup LMS CUDA generator
 - CPU+GPU implementation for serial (NS?) problems larger than device memory
-- optimizations for in-memory computations
 
-1. restriction vs ADP: we give only the best answer, do we care about semi-optimal ones?
 2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
-3. use sub-cell parallelism (seen in 2 papers already) => does not improve
 
@@
 \subsection*{Plan} \ol
-\item {\color{gray}
-\textbf{Problems description:} parallel tree-raking, does not share much with other algorithms (sparse version of computations, might not scale efficiently).
-Most common patterns are already enclosed by the above problems.
-Real input size is around 300K.
-We might want to also look at an $O(n^3)$-space-complex problem (like matching 3 strings $S,T,U$).}
-
-All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
-\item Serial dependencies
-\item Non-serial horizontal or vertical dependencies (1D non-serial)
-\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
-\item We have not found other type of dependencies in the literature
-\ule
 
@@
 - operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
-
 - write a kernel that takes stats from left and r
@@
   - we may go up to running in stripes
-
 - if we put in diagonal-major data in the memory, why not stream all chunks like
@@
 Memory organized as
-	b1 -  -
+b1 -- --
+b1 b1 --
 b1 b1 b1
@@
 	----+---+
-
 we many not need to store what's the previous cell, however, the previous
 cell information (usually limited range) is much more compact than the score(32-64 bits)
-
-Memory usage of CPU:
-
-we can potentially extend the GPU's memory by storing some information in the CPU RAM, which in general is larger.
-For large sequences, this technique may apply to problems where one needs to access elements "all the way to the top" and aggregate them in a non trivial (max,min,sum,avg) way.
 
@@
 --------------------------------------
-This is how we would like to encode the DP programs into Scala DSL.
-
-I think the best would be to reuse most of the syntactic elements and write it
-as an uneducated programmer would write, apply annotations/additions on top of
-it or detect that it is DP and then converge to our system.
-
-We need to keep in mind that we need both to express the backtracking and the scoring at the same time
-
-Let's build some examples
+Reuse most of the syntactic elements and write it as uneducated programmer, apply annotations/additions on top of it or detect that it is DP and then converge to our system.
 
@@
 
-def Nussinov(S0: List[Char]) : List[Pair[Char,Char]] = {
-	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function
-
-	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S.zipWithIndex,T.zipWithIndex) match {
-		case (_, Nil) => (0, stop)
-		case (Nil, _) => (0, stop)
-		case (Si::Sr, Tj::Tr) =>
-			val nw = SWat(Sr,Tr);
-			val n  = SWat(Sr,T);
-			val w  = SWat(S,Tr);
-			val cols_rows = 
-			val c = cost(Si,Tj);
-
-			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
-			else if (n>=w) then (n._1-d, "N",[Si,_])
-			else (w._1-d, "W",[_,Tj])
-	}
-
-	backtrack(DP(S0,S0))
-}
-
 - Char, Score, Backtrack, Output are LMS Rep objects so that we can easily detect
@@
 --------------------------------------------
-
 def fun[T: CanTranslateToC](...)
@@
 
-class CanTranslateToC[T] {
-  def translate:String
-}
-
-implicit def canTranslateInt = new CanTranslateToC[Int] = {
-  def translate = "Int"
-}
-
+class CanTranslateToC[T] { def translate:String }
+implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
 \end{verbatim}

### 2012-11-13 docs/roadmap/implementation.tex
Rework of the draft.

--- 2012-11-13 docs/roadmap/design.tex
@@
 \newpage
-\section{Architecture design and technical decisions}
+\section{Implementation}
 % ------------------------------------------------------------------------------------------------
-\subsection{Parsing grammar (ADP)}
-{\color{red}
-Borrowing from ADP XXX
-
-XXX: explain ADP concepts
-
-restriction from ADP: we give only the best answer, do we care about semi-optimal ones? no
-
-XXX: normalization, transforms applied
-
-XXX: Think how to implement {\tt cyclic} problems and {\tt windowing} properties within the language
-
-\begin{verbatim}
-restrict to only 1 max/min (more restrictive than ADP)
-for optimization we want to split grammar into binary productions, to do that we need
-to have equality on the following formula. We can solve it more easily if functions are linear
-\end{verbatim}
-\[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]
-
-}
-
-\subsection{Backtracking}
-In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
-To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack reads are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
-To deal with this problem, we are facing two options:\ul
-\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack and is not implemented in vanilla ADP (unless the whole backtrack is made part of the scoring: big performance impact and non-constant memory requirement issues make its GPU implementation hard and not desirable).
-Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
-
-\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem.
-To do that we need to apply the following transformations\ol
-	\item Normalize rules and identify them uniquely by \ul
-		\item Removing all the <<or>> operators by breaking a parser into multiple subrules
-		\item Forwarding names: replace a rule that is a single invocation by its content (i.e. name a terminal parser with its own name instead of its containing parser's)
-		\item Assigning each subrule an unique identifier $r$ and create a mapping table $T: r \to$ user-defined name of the parser.
-	\ule
-	\item The data elements corresponding to a rule are (scoring, backtracking) and named after the tabulation.
-The scoring is a user-defined primary type (int, double, ...) and the backtracking is a tuple $({\rm id}_{\rm subrule}, k_1, k_2, \ldots , k_m)$ where $m$ is the maximal number of splits that can occur in the subrules.
-	\item During the matrix computation of cell $(i,j)$, if the subrule $r$ applies, the backtrack will be set as $(r,k_1,k_2,\ldots,k_{m_r})$ (obviously with $i\le k_1\le k_2\le \ldots k_{m_r}\le j$ and $m_r\le m$).
-	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r,k_1,k_2,\ldots,k_m)$, given $r$, we recover the subrule hence the user-defined name of the rule that applies and $k_{m_r}$, which allows us to enqueue the subwords $(i,k_1), (k_1,k_2), ..., (k_{m_r})$ for further backtracking.
-If $r$ refers to a terminal, we stop the backtracking.
-	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $(r,i,j)$ where $r$ is the index of the sub-rule that can be easily mapped into the user name (such that $T: r_u \to$ user-defined rule name) and $(i,j)$ is the subword on which the rules applies.
-\ole
-In short, we break parsers into normalized rules, the backtracking information is the rule id (which rule to unfold) and a list of indices (how to unfold it).
-\ule
-Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed by the user program in-order, applying for each user-defined rule the underlying transformation, and storing intermediate results (i.e.
-in a hash map) until they are processed by another rule.
-Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.
-
-% ------------------------------------------------------------------------------------------------
-\subsection{Memory management}
-We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
-This can be the GPU or the FPGA internal memory.
-Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
-
-We propose to evaluate the device memory requirements to solve the above problem classes. We need first to define additional problem properties related to implementation:\ul
-\item \textbf{Number of matrices:} multiple matrices can be encoded as 1 matrix with multiple values per cell.
-Hence the implementation differentiates only between cost and backtrack matrices with respective element sizes $S_C$ and $S_B$.
-\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
-Such data amount is usually proportional to the delay of dependencies.
-If this delay is small, it might be worth to duplicate matrix data in the wavefront, otherwise it might be more efficient to allow access to the previous computational blocks of the matrix.
-\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
-Hence such information can be maintained in a single place (in the wavefront) and progress together with the computation.
-We denote by $S_W$ the size of wavefront elements.
-\item \textbf{Input size:} the size of an input letter (from input alphabet) is denoted by $S_I$.
-\ule
-
-% ----------------------------------------------
-\subsubsection{Small problems (in-memory)}
-Problem that can fit in memory can be solved in a single pass on the device. Such problem must satisfy the equation:
-	\[(S_I+S_W) \cdot (m+n) + (S_C+S_B) \cdot (m\cdot n) \le M_D\]
-
-For instance, assuming that $m=n$, $M_D=1024{\rm Mb}$, that backtrack is 2b (<16384, 3 directions) and that the cost can be represented on 4 b (int or float), that input is 1b (char) and that there is no wavefront, we can treat problems of size $n$ such that $2n+5n^2 \le 2^{30} \implies  n\le 14650$.
-We might also possibly need to take into account extra padding memory used for coalesced accesses.
-But it is reasonable to estimate that problems up to 14K fit in memory.
-
-% ----------------------------------------------
-\subsubsection{Large problems}
-To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$. For simplification, we assume a square matrix made of square blocks with $b$ blocks per row/column.
-
-% ----------------------------------------------
-\subsubsection{Non-serial problems}
-Non-serial problems need to potentially access all elements that have been previously computed.
-We restrict\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.} ourselves to the following dependencies: \ul
-\item Non-serial dependencies along row and column
-\item Serial dependencies along diagonal, with delay smaller or equal to one block size
-\ule
-Such restriction implies that all the block of the line and the row, and one additional block to cover diagonal dependencies must be held in memory (independently of the matrix shape).
-
-For simplification, let $m=n$ and assume that we have $b$ square blocks per row and per column. Hence we have the following memory restriction:
-\[ 2\frac{n}{b}(S_I+S_W) + 2 \cdot \frac{n^2}{b}S_C + \frac{n^2}{b^2} S_B \le M_D\]
-
-We also need to take into account the transfer between main memory (or disk) and device memory.
-Dependency blocks only need to be read, computed blocks need to be written back.
-Ignoring the backtrack and focusing only on the cost blocks, the transfers (in blocks) are:
-\[\begin{array}{rclll}
-		b^2 +% computed line writeback (once)
-		(b-1)^2 + % diagonal block loads (1 per block)
-		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
-		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
-		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
-		&=& b^3 &\qquad& \rm (Parallelogram)
-\end{array}\]
-
-Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
-\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}
-
-Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay.
-Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h\up{(T)} and 6 days 10h\up{(P)}, assuming there is no other scalability issues.
-Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
-Given that real problems -- like RNA folding -- operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
-
-% ----------------------------------------------
-\subsubsection{Serial problems}
-The serial problem have the interesting property to access to a fixed number of previous elements.
-These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into a wavefront.
-Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems) or line-wise or column-wiser.
-This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
-horizontally) along the matrix.
-
-Since serial problems are of complexity $O(n^2)$ (due to the matrix dimension and the finite number of dependencies), it is possible to tackle much larger problem than non-serial during the same running time.
-Hence, it seems obvious to let serial problems grow larger than the memory.
-
-Mixing the dependency property and size requirements, we can split the matrix into sub-matrices, store special lines (and/or columns) into memory (or hard disk), and repeat computations to solve the backtrack (similarly as in \cite{swat_gpu},\cite{swat_mega}, but this implementation use problem-specific knowledge that might not generalize).
-
-To store intermediate lines and columns, we are facing two different strategies to explore:\ul
-\item \textbf{Fixed subproblem size:} we decompose the algorithm as follows\ol
-\item Define a grid of <<major column and rows>>, where each cell's data (input, output, cost and backtrack matrices) fits into the device memory.
-\item Compute the values of the grid's major columns and rows in one pass.
-\item Second (on-demand) computation to process backtracking inside relevant cells.
-\ole
-Let $b$ the number of cells that we have on each row/column, the total computation running time would be $(b^2 + 2b) \cdot t_b$ where $t_b$ is the time to compute one cell's matrix.
-This division has the advantage of providing the minimal computation time at the expense of external storage proportional to $O(n)$ (if we store only lines or columns) or $O(n^2)$ (if we store both).
-
-\item \textbf{Myers and Miller’s algorithm:} (divide and conquer)
-This algorithm break the DP problem into 2 (or 4) subproblems such that once the middle line/column is computed, the problem can be solved for 1 submatrix and backtracking among up to 2 of the 3 other.
-This breaking is applied recursively until the submatrix data fits into memory.
-The storage requirements are $4 \cdot O(n)$ (we store along both dimension $1+\tfrac{1}{2}+\tfrac{1}{4}+...$ lines/columns).
-
-The algorithm proceeds as follows: first it solves the problem to obtain the first backtracking element, then it breaks the matrix in 4 submatrices, and refine it until backtrack is tractable.
-Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
-\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used.
-\ule
-
-{\color{red}
-try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront
-
-XXX: what's the maximal size of the wavefront ??
-
-XXX: can we avoid to store some matrices and put them in the wavefront ??
-
-Split into blocks:\ul
-\item Decide the shape of the blocks
-\item Decide the size of the blocks
-\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
-\ule
-
-XXX: make it work up to 14K for all 3 problems using multiple kernels
-XXX: example of non-symmetric serial problem
-
-The three last elements, combined with the above one, provide a precise estimation of the memory consumption, and the implementation difficulty
-
-all the problem are subject to the input dimensions $n$.
-
-the two latter one gives an estimation of the constant factor.
-
-The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
-Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
-}
-
-% ------------------------------------------------------------------------------------------------
-\subsection{Memory layout}
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{LMS compiler stack}
-\textbf{User-language}: define additional parameters for the recurrence\ul
-\item Windowing (to convert non-serial into serial problems)
-\item Input sizes, and alphabets (backtrack, input, cost)
-\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
-\item Recurrence functions, devices available
-\item What to keep in memory (cost, backtrack or both).
-\ule
-$\Downarrow$ Conversion (using an existing technique)
-
-\textbf{Intermediate representation}
-
-$\Downarrow$ Optimizations\ul
-\item Transform non-serial into serial \ul
-	\item Use aggregation functions/transformations
-	\item Use windowing from user (if no other technique succeed)
-	\ule
-\item Define the wavefront depth
-\item Avoiding the cost matrix by moving it into the wavefront
-\ule
-
-\textbf{Code specification}\ul
-\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
-\item Device-level interface => setup the block sizes(w/h), input and memory sizes
-\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
-\item Define the co-processor memory aggregation function
-\item Define the scheduling of the blocks and aggregation (software pipelining)
-\item Define the data movement back and forth to disk
-\ule
-
-$\Downarrow$ Generation\ul
-\item Generate the kernel for specific device
-\item Generate the scheduling and barriers
-\ule
-
-\textbf{Binary program}
-
-%\ol
-%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
-%\item Let the user tune the window size if he wants to reduce non-serial to serial.
-%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
-%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
-%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
-%	\item Scheduler (CPU): interleave block computation and memory statistics
-%	\ule
-%\item Common description \ul
-%	\item Block kernel processor
-%	\item Full block
-%	\ule
-%\item Discussion: wavefront, design similarities, polyhedral theory
-%\ole
-
-%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
-%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
-%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
-%$\begin{array}{|C|C|C|C|C|C|} \hline
-%0 & 0 & 0 & 0 & 0 & 0 \mnl
-%0 &  &  &  &  & \mnl
-%0 & M_{23}  &  &  &  & \mnl
-%0 &  & \sum  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%\end{array}$
-%\end{document}
+\subsection{ADP parsers}
+\subsection{Transformations and simplifications}
+\subsection{LMS integration}
+\subsection{CUDA implementation}

### 2012-11-13 docs/roadmap/problems.tex
Rework of the draft.

--- 2012-11-12 docs/roadmap/problems.tex
@@
-%\input{style.sty}
-
-% Recurrence visualization helpers
-\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
-\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
-\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
-\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
-\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
-	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
-\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
-
-\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
-\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
-
-%\title{DP Problems of Interest}
-%\begin{document}
-%\maketitle
-%\pagestyle{headings}
-
-% ----------------------------------------------------------------------------------------------------------------------------------------
 \newpage
-\section{Problems classification}
-\subsection{Definitions}\ul
+\section{Dynamic programming problems}
+% ------------------------------------------------------------------------------------------------
+\subsection{Problems classification}
+\subsubsection{Definitions}\ul
 \item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
@@
 
-\subsection{Classification}
 \subsubsection{Litterature classification}
@@
 
-% ----------------------------------------------------------------------------------------------------------------------------------------
-\section{Problems of interest}
+% ------------------------------------------------------------------------------------------------
+\subsection{Problems of interest}
 We usually focus on problem that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$).
@@
 
+% ------------------------------------------------------------------------------------------------
+% Recurrence visualization helpers
+\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
+\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
+\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
+\def\Cm{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base
+\def\Cfl#1{#1{0,6}#1{0,5}#1{1,5}#1{0,4}#1{1,4}#1{2,4}#1{0,3}#1{1,3}#1{2,3}#1{3,3}#1{0,2}#1{1,2}#1{2,2}#1{3,2}#1{4,2}
+	#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{4,1}#1{5,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}#1{5,0}#1{6,0}} % triangular lower (function)
+\def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
+
+\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
+\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
+
 % ----------------------------------------------
 \newpage
-\subsection{Smith-Waterman (simple)}\label{sswat}\ol
+\subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
 \item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
@@
 \newpage
-\subsection{Smith-Waterman with affine gap extension cost}\ol
+\subsubsection{Smith-Waterman with affine gap extension cost}\ol
 \item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
@@
 \newpage
-\subsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
+\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
 \item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
@@
 \newpage
-\subsection{Convex polygon triangulation}\ol
+\subsubsection{Convex polygon triangulation}\ol
 \item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
@@
 \newpage
-\subsection{Matrix chain multiplication}\ol
+\subsubsection{Matrix chain multiplication}\ol
 \item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$.
@@
 \newpage
-\subsection{Nussinov algorithm}\ol
+\subsubsection{Nussinov algorithm}\ol
 \item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
@@
 \newpage
-\subsection{Zuker folding}\ol
+\subsubsection{Zuker folding}\ol
 \item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
@@
 \newpage
-\section{Related problems}
+\subsection{Related problems}
 The goal of this section is to demonstrate that our framework can accommodate with many {\color{red} (20-50)} problems that we have not considered at the design time.
 
-\subsection{Serial problems}
+\subsubsection{Serial problems}
 \begin{tabular}{llcc} \toprule
@@
 
-\subsection{Non-serial problems}
+\subsubsection{Non-serial problems}
 \begin{tabular}{llcc} \toprule
@@
 \href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} \footnotesize (input sensitive) & rectangle & 1 & --\\
-
 \end{tabular}
 
-\subsection{Other problems}\ul
+\subsubsection{Other problems}\ul
 \item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
@@
 \end{verbatim}
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section{User facing language}
-\subsection{Parsing grammar (ADP)}
-Borrowing from ADP XXX
-
-XXX: explain ADP concepts
-
-XXX: normalization, transforms applied
-
-\subsection{Backtracking}
-In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
-To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack reads are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
-To deal with this problem, we are facing two options:\ul
-\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack and is not implemented in vanilla ADP (unless the whole backtrack is made part of the scoring: big performance impact and non-constant memory requirement issues make its GPU implementation hard and not desirable).
-Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
-
-\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem.
-To do that we need to apply the following transformations\ol
-	\item Normalize rules and identify them uniquely by \ul
-		\item Removing all the <<or>> operators by breaking a parser into multiple subrules
-		\item Forwarding names: replace a rule that is a single invocation by its content (i.e. name a terminal parser with its own name instead of its containing parser's)
-		\item Assigning each subrule an unique identifier $r$ and create a mapping table $T: r \to$ user-defined name of the parser.
-	\ule
-	\item The data elements corresponding to a rule are (scoring, backtracking) and named after the tabulation.
-The scoring is a user-defined primary type (int, double, ...) and the backtracking is a tuple $({\rm id}_{\rm subrule}, k_1, k_2, \ldots , k_m)$ where $m$ is the maximal number of splits that can occur in the subrules.
-	\item During the matrix computation of cell $(i,j)$, if the subrule $r$ applies, the backtrack will be set as $(r,k_1,k_2,\ldots,k_{m_r})$ (obviously with $i\le k_1\le k_2\le \ldots k_{m_r}\le j$ and $m_r\le m$).
-	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r,k_1,k_2,\ldots,k_m)$, given $r$, we recover the subrule hence the user-defined name of the rule that applies and $k_{m_r}$, which allows us to enqueue the subwords $(i,k_1), (k_1,k_2), ..., (k_{m_r})$ for further backtracking.
-If $r$ refers to a terminal, we stop the backtracking.
-	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $(r,i,j)$ where $r$ is the index of the sub-rule that can be easily mapped into the user name (such that $T: r_u \to$ user-defined rule name) and $(i,j)$ is the subword on which the rules applies.
-\ole
-In short, we break parsers into normalized rules, the backtracking information is the rule id (which rule to unfold) and a list of indices (how to unfold it).
-\ule
-Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed by the user program in-order, applying for each user-defined rule the underlying transformation, and storing intermediate results (i.e.
-in a hash map) until they are processed by another rule.
-Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section{Implementation}
-We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
-This can be the GPU or the FPGA internal memory.
-Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
-\subsection{Memory requirements}
-We propose to evaluate the device memory requirements to solve the above problem classes. We need first to define additional problem properties related to implementation:\ul
-\item \textbf{Number of matrices:} multiple matrices can be encoded as 1 matrix with multiple values per cell.
-Hence the implementation differentiates only between cost and backtrack matrices with respective element sizes $S_C$ and $S_B$.
-\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
-Such data amount is usually proportional to the delay of dependencies.
-If this delay is small, it might be worth to duplicate matrix data in the wavefront, otherwise it might be more efficient to allow access to the previous computational blocks of the matrix.
-\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
-Hence such information can be maintained in a single place (in the wavefront) and progress together with the computation.
-We denote by $S_W$ the size of wavefront elements.
-\item \textbf{Input size:} the size of an input letter (from input alphabet) is denoted by $S_I$.
-\ule
-
-\subsection{Small problems (in-memory)}
-Problem that can fit in memory can be solved in a single pass on the device. Such problem must satisfy the equation:
-	\[(S_I+S_W) \cdot (m+n) + (S_C+S_B) \cdot (m\cdot n) \le M_D\]
-
-For instance, assuming that $m=n$, $M_D=1024{\rm Mb}$, that backtrack is 2b (<16384, 3 directions) and that the cost can be represented on 4 b (int or float), that input is 1b (char) and that there is no wavefront, we can treat problems of size $n$ such that $2n+5n^2 \le 2^{30} \implies  n\le 14650$.
-We might also possibly need to take into account extra padding memory used for coalesced accesses.
-But it is reasonable to estimate that problems up to 14K fit in memory.
-
-\subsection{Large problems}
-To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$. For simplification, we assume a square matrix made of square blocks with $b$ blocks per row/column.
-
-\subsubsection{Non-serial problems}
-Non-serial problems need to potentially access all elements that have been previously computed.
-We restrict\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.} ourselves to the following dependencies: \ul
-\item Non-serial dependencies along row and column
-\item Serial dependencies along diagonal, with delay smaller or equal to one block size
-\ule
-Such restriction implies that all the block of the line and the row, and one additional block to cover diagonal dependencies must be held in memory (independently of the matrix shape).
-
-For simplification, let $m=n$ and assume that we have $b$ square blocks per row and per column. Hence we have the following memory restriction:
-\[ 2\frac{n}{b}(S_I+S_W) + 2 \cdot \frac{n^2}{b}S_C + \frac{n^2}{b^2} S_B \le M_D\]
-
-We also need to take into account the transfer between main memory (or disk) and device memory.
-Dependency blocks only need to be read, computed blocks need to be written back.
-Ignoring the backtrack and focusing only on the cost blocks, the transfers (in blocks) are:
-\[\begin{array}{rclll}
-		b^2 +% computed line writeback (once)
-		(b-1)^2 + % diagonal block loads (1 per block)
-		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
-		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
-		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
-		&=& b^3 &\qquad& \rm (Parallelogram)
-\end{array}\]
-
-Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
-\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}
-
-Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay.
-Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h\up{(T)} and 6 days 10h\up{(P)}, assuming there is no other scalability issues.
-Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
-Given that real problems -- like RNA folding -- operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
-
-\subsubsection{Serial problems}
-The serial problem have the interesting property to access to a fixed number of previous elements.
-These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into a wavefront.
-Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems) or line-wise or column-wiser.
-This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
-horizontally) along the matrix.
-
-Since serial problems are of complexity $O(n^2)$
-
-Different strategies
-XXX: Do we need/want to use divide and conquer?\ul
-\item Divide and co
-\item
-\ule
-
-propose a memory consumption scheme
-
-try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront
-
-{\color{red}
-
-XXX: what's the maximal size of the wavefront ??
-
-XXX: can we avoid to store some matrices and put them in the wavefront ??
-
-Split into blocks:\ul
-\item Decide the shape of the blocks
-\item Decide the size of the blocks
-\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
-\ule
-
-XXX: make it work up to 14K for all 3 problems using multiple kernels
-XXX: example of non-symmetric serial problem
-
-The three last elements, combined with the above one, provide a precise estimation of the memory consumption, and the implementation difficulty
-
-all the problem are subject to the input dimensions $n$.
-
-the two latter one gives an estimation of the constant factor.
-
-The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
-Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
-}
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{LMS compiler stack}
-\textbf{User-language}: define additional parameters for the recurrence\ul
-\item Windowing (to convert non-serial into serial problems)
-\item Input sizes, and alphabets (backtrack, input, cost)
-\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
-\item Recurrence functions, devices available
-\item What to keep in memory (cost, backtrack or both).
-\ule
-$\Downarrow$ Conversion (using an existing technique)
-
-\textbf{Intermediate representation}
-
-$\Downarrow$ Optimizations\ul
-\item Transform non-serial into serial \ul
-	\item Use aggregation functions/transformations
-	\item Use windowing from user (if no other technique succeed)
-	\ule
-\item Define the wavefront depth
-\item Avoiding the cost matrix by moving it into the wavefront
-\ule
-
-\textbf{Code specification}\ul
-\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
-\item Device-level interface => setup the block sizes(w/h), input and memory sizes
-\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
-\item Define the co-processor memory aggregation function
-\item Define the scheduling of the blocks and aggregation (software pipelining)
-\item Define the data movement back and forth to disk
-\ule
-
-$\Downarrow$ Generation\ul
-\item Generate the kernel for specific device
-\item Generate the scheduling and barriers
-\ule
-
-\textbf{Binary program}
-
-%\ol
-%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
-%\item Let the user tune the window size if he wants to reduce non-serial to serial.
-%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
-%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
-%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
-%	\item Scheduler (CPU): interleave block computation and memory statistics
-%	\ule
-%\item Common description \ul
-%	\item Block kernel processor
-%	\item Full block
-%	\ule
-%\item Discussion: wavefront, design similarities, polyhedral theory
-%\ole
-
-%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
-%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
-%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
-%$\begin{array}{|C|C|C|C|C|C|} \hline
-%0 & 0 & 0 & 0 & 0 & 0 \mnl
-%0 &  &  &  &  & \mnl
-%0 & M_{23}  &  &  &  & \mnl
-%0 &  & \sum  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%\end{array}$
-%\end{document}

### 2012-11-13 docs/roadmap/design.tex
NIts.

@@
 \section{Architecture design and technical decisions}
+% ------------------------------------------------------------------------------------------------
+\subsection{User facing language requirements}
+\quote{The Algebraic Dynamic Programming approach (ADP) introduces a conceptual splitting of a DP algorithm into a recognition and an evaluation phase.
+The evaluation phase is specified by an evaluation algebra, the recognition phase by a yield grammar.
+Each grammar can be combined with a variety of algebras to solve different but related problems, for which heretofore DP recurrences had to be developed independently.
+Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.
+A notation for yield grammars is provided that serves as a domain-specific language.}
+
+Given such an existing formalization \cite{adp} of dynamic programming on sequences, it seems natural to reuse it and possibly extend it to  other types of DP problems.
+In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only desired candidates for further combination.
+
+The benefits of ADP framework is that it does not constraint the result of the evaluation to be a single value, but could extend parsers to backtracking parsers or pretty-printers.
+Such functionality is easy to support in plain Scala, however, it could hampers the performance of the GPU implementation if we leave to the user the duty of representing and processing all the information within the parsers.
+Its major shortcoming in expressivity is that it only operates on strings, making only triangular-matrix problem easily representable.
+
+The additional features we want to support are:\ol
+\item Cyclic problems: are inherently very similar to string problems, except that the input is cyclic.
+To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
+\item Input pair algebra: the original ADP framework only support single 
+\item Device restrictions and automatic backtracking: since efficient code has to be devised for C/CUDA implementation, we need to impose restrictions on the output that could be generated by the parsers.
+In particular, we would like to constraint the result to primary types (int, float, ...).
+However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we need to automate the backtracking to both fulfill the goals of efficiency and usefulness of the device specific implementation.
+
+\ole
+
+\textbf{=> have a pair of subwords for multiple sequences}
+
+XXX: explain ADP
+
+Explain shortcomings
+
+We want to offer the following features to the user:\ul
+\item Support of cyclic problems (optionally with number of iterations)
+\item Full features of ADP available in Scala, restriction on the CUDA available language
+\item Windowing
+\item Pair of inputs algebra (similar to SWat)
+\ule
+
+We define the aggregate function h as $List[T]\to List[T]$ and a special string emitter for C code that produces the initial element and a comparator between two (cost) elements.
+
 % ------------------------------------------------------------------------------------------------

### 2012-11-14 docs/roadmap/design.tex
More design.

@@
 
-Given such an existing formalization \cite{adp} of dynamic programming on sequences, it seems natural to reuse it and possibly extend it to  other types of DP problems.
-In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only desired candidates for further combination.
+Given such formalization \cite{adp} of dynamic programming on sequences, it seems natural to borrow from it and extend it to other types of DP problems.
+In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only some candidates for further combination.
 
@@
 The additional features we want to support are:\ol
-\item Cyclic problems: are inherently very similar to string problems, except that the input is cyclic.
+\item \textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
 To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
-\item Input pair algebra: the original ADP framework only support single 
-\item Device restrictions and automatic backtracking: since efficient code has to be devised for C/CUDA implementation, we need to impose restrictions on the output that could be generated by the parsers.
-In particular, we would like to constraint the result to primary types (int, float, ...).
+\item \textbf{Input pair algebra:} the original ADP framework only support single input, we want to support a pair of inputs such that we can treat problem such as Smith Waterman or Needleman-Wunsch.
+However, it does not make sense to treat more than two sequences because of the $O(n^3)$ storage requirements that limits the problems size more dramatically.
+\item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$ best results.
+\item \textbf{Device restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ...
+and structures of them).
+However, we want to preserve the expressiveness available for Scala and impose restrictions on the input and answers available to CUDA.
+A typical restriction we want to make is that processed structures are of fixed size so that we avoid memory management issues and thread divergence.
+
+The general ADP framework supports multiple solutions for intermediate results.
+Similarly, we want to impose the restriction that only the best result would be stored on device while preserving the support for multiple partial results in Scala.
+
+\item \textbf{Automatic backtracking:} Since efficient code has to be devised we impose restrictions on the output that could be generated by the parsers on devices.
 However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we need to automate the backtracking to both fulfill the goals of efficiency and usefulness of the device specific implementation.
 
+Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement while memoizing the backtracking with the results would have the drawback of growing towards final result and duplicated unnecessary information, hence requiring both $O(n^3)$ space and memory management features on devices.
+Collecting the backtracking list can be easily done in $O(n)$ and then inverted whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
 \ole
+Since we construct the dependency tree bottom up, and since we need to work with a fixed-size memory, we need to do the additional steps in our pre-processing:\ul
+\item \textbf{Normalization:} in order to automate the backtracking, we need the rule to present a certain shape (in particular we want to remove alternatives) so that we can define uniquely the backtracking information.
+Another step we can make in the process is to break down complex rules into simpler ones if they are expressible so, hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
+Also we can strip away unnecessary rules.
+\item \textbf{Dependency analysis:} a precise order between the rules application must be given.
+This is required as we use a bottom-up approach where all previous results must be computed versus a top-down approach with a table where a fallback to computation always exist if the desired element is not present.
+\ule
+
+Size analysis to know what storage size we require
 
@@
 
-XXX: explain ADP
+We define the aggregate function h as $List[T]\to List[T]$ and a special string emitter for C code that produces the initial element and a comparator between two (cost) elements.
 
-Explain shortcomings
+What about the other way round for aggregation: instead of specifying List[U]->List[T] specify comparator of T and a k to keep the k bests(?)
 
-We want to offer the following features to the user:\ul
-\item Support of cyclic problems (optionally with number of iterations)
-\item Full features of ADP available in Scala, restriction on the CUDA available language
-\item Windowing
-\item Pair of inputs algebra (similar to SWat)
-\ule
-
-We define the aggregate function h as $List[T]\to List[T]$ and a special string emitter for C code that produces the initial element and a comparator between two (cost) elements.
+XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?
 
@@
 XXX: explain ADP concepts
-
-restriction from ADP: we give only the best answer, do we care about semi-optimal ones? no
 

### 2012-11-14 docs/roadmap/design.tex
Nits.

@@
 Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
-\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used.
+\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using twice more storage space would only result in a $2\times$ speedup (measuring only the computation time).
+Hence an hybrid approach would be to decide a $k$ such that at each step we partition the desired submatix into a intermediate grid of $k$ rows/columns.
+The space usage would be in $2 k \log_k (n/b)$ and the running time complexity would be $O(n^2 \cdot \log_k n)$.
+Then the user would be able to fix a storage space $S \ge 4 \log_2 (n/b)$ and obtain the corresponding $k$ for a given $n$.
 \ule

### 2012-11-14 docs/roadmap/draft.tex
More design.
Nits.

@@
 \section{Planning/draft}
+\subsubsection*{Steps}\ul
+\item Re-implement vanilla combinators with lists
+\item Add following problems and features:\ul
+	\item Polygon triangulation (cyclic problem)
+	\item SWat with arbitrary gap (multi-track grammar)
+	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
+	\item Rules normalization
+	\item Rules dependency
+	\item Automatic backtracking
+	\ule
+\item Test / proof that parsers are correct
+\item Automate test to compare against implementation
+\item Build bottom-up vanilla version\ul
+	\item Separate initialization (terminals) and non-terminals (speed up?)
+	\item In parallel implement the code generator for CUDA, hardcode in text function bodies for CUDA
+	\ule
+\item LMS implementation: from normal to Rep/Exp\ul
+	\item Re-implement bottom-up vanilla into LMS
+	\item Re-implement CUDA version and use functions body to infer CUDA code
+	\ule
+\item Write report
+\item Further work\ul
+	\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
+	\item 
+	\ule
+\ule
+
 \subsubsection*{Roadmap}

### 2012-11-14 docs/roadmap/design.tex

@@
 
+{\color{red}
 Size analysis to know what storage size we require
@@
 XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?
+}
 
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Memory management}
+\subsection{Memory constraints}
 We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
@@
 \subsection{Memory layout}
+{\color{red} XXX}
 
@@
 \newpage
+{\color{red} 
 \subsection{LMS compiler stack}
@@
 \textbf{Binary program}
-
+}
 %\ol

### 2012-11-14 docs/roadmap/draft.tex

@@
 
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section{Introduction}
-\subsection{Dynamic programming}
-Dynamic programming consists of solving a problem by reusing subproblems solutions. The most famous example of dynamic programming is the Fibonacci series that is defined as 
-\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
-
-A typical characteristic of dynamic programming is that a subproblem solution is often reused multiple time, so it is often efficient to store intermediate results in a table to speed-up computation of larger problems.
-
-There exist various categories of dynamic programming:\ul
-\item Series, knapsack : that operates usually on a single dimension
-\item Sequences alignment (matching two sequences at best), top-down grammar analysis, ...
-\item Tree-related algorithms: phylogenetic trees raking, maximum tree independent set, ... 
-\ule
-
-Since the first category is inherently sequential and the third category is hard to parallelize efficiently, we focus on the second type of problems, which is also the most common.
-
-{\color{red}
-parallel tree-raking, does not share much with other algorithms (sparse version of computations, might not scale efficiently).
-Most common patterns are already enclosed by the above problems.
-Real input size is around 300K.
-We might want to also look at an $O(n^3)$-space-complex problem (like matching 3 strings $S,T,U$).
-
-All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
-\item Serial dependencies
-\item Non-serial horizontal or vertical dependencies (1D non-serial)
-\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
-\item We have not found other type of dependencies in the literature
-\ule}
-
-\subsection{Scala and LMS}
-
-% ------------------------------------------------------------------------------------------------
+\input{introduction}
 \input{problems}
@@
 \input{benchmarks}
-\section{Future work}
+\section{Future work} \ul
+\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
+\item Annotation on recursive functions to use dynamic programming like \\
+	{\tt @DynaProg def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
+\ule
+
 \section{Conclusion}
@@
 \newpage
-\section{Planning/draft}
-\subsubsection*{Steps}\ul
+\section{Planning --- Work in progress}
+\subsubsection*{Steps}\ol
 \item Re-implement vanilla combinators with lists
 \item Add following problems and features:\ul
-	\item Polygon triangulation (cyclic problem)
-	\item SWat with arbitrary gap (multi-track grammar)
+	\item Polygon triangulation (cyclic problem) (parallelogram matrix)
+	\item SWat with arbitrary gap (multi-track grammar) (rectangular matrix)
 	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
@@
 	\ule
-\item Test / proof that parsers are correct
+\item Test/proof parsers are correct --- make sure implementation is correct
 \item Automate test to compare against implementation
@@
 	\ule
+\item Benchmarks --- compare also versus other papers.
 \item Write report
-\item Further work\ul
-	\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
-	\item 
-	\ule
-\ule
+\ole
 
@@
 \subsubsection*{Todo @TCK}
+\subsubsection*{Todo @Manohar}
 \begin{verbatim}
@@
 - Cleanup LMS CUDA generator
-- CPU+GPU implementation for serial (NS?) problems larger than device memory
 
-2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
-
-Plan:
-We need to extend the language to support all 3 cases
-- Parallelogram is triangular + cyclic => add a cyclic keyword that applies to the whole problem
-- Square: what is the transform being done in ADP extension?
-
-1. optimize: terminals of bounded yield + binary decomposition if possible
-2. transform into: axioms (init/fixed) + rules (iterations)
-3. transform to code
-     axiom(i := k(i)) -> M[i,i]=k(i)
-     rule(x~y := h(x,y)) -> M[i,j]=forall_i<k<j h(M[i,k],M[k,j]) for both unbounded yield
-     rule(x~y_l := h(x,y)) -> M[i,j]=forall_j-l<k<j h(M[i,k],M[k,j]) for y bounded by l
-     rule(x~y_l := h(x,y)) -> M[i,j]=h(M[i,j-l],M[j-l,j]) for y of exactly l
+Separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
 
@@
 
-\subsubsection*{Todo @Manohar}
-\begin{verbatim}
-- generate chain matrix recurrences to generate implementation
-- more theory behind what we want to support, how we encompass all cases, ...
-\end{verbatim}
-
-\subsection*{Plan} \ol
-
-\item \textbf{User facing language:} goals are flexibility and compactness.\ul
-	\item User-facing language should be similar to related paper \cite{adp_gpu} or \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}.
+\subsection*{OldPlan} \ol
+\item \textbf{User facing language:} similar to \cite{adp_gpu} or \cite{adp_fusion} \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}.
 We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
-	\item We also may want to try to make implicit transformation for code like \\
-		{\tt @DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
-	\item Windowing: the user should be able to force a windowing (i.e. force a non-serial problem to be a $k$-polyadic serial problem).	
-	\item 3 different cases: we care about backtrack cost or both.
-	\item Backtracking: create an operator that produces the whole backtrack sequence indices.
-	\ule
-
-$\implies$ \emph{end of October}.
-
-\item \textbf{Prototyping:} get a prototype to understand difficulties and share common base.
-\\ Implement a working prototype of \nameref{aswat} on CPU (for correctness), and specific platform (CUDA/FPGA).
-This will give us an idea of how to implement the general case.
-We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much).
-Here we aim to do as good an implementation for the specific platform (CPU/GPA) as possible.\\
-$\implies$ \emph{end of October}.
-
-\item \textbf{Baseline:} Also use benchmarks provided by existing implementations as baselines.\\
-$\implies$ \emph{end of October}.
-
+\item \textbf{Prototyping:} understand difficulties and share common base.
+Prototype of \nameref{aswat} on CUDA/FPGA.
+Give an idea of how to implement the general case.
+Benchmark and compare both implementations.
+Aim to do as good an implementation for the specific platform (CPU/GPA) as possible.
 \item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
 \item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
-\item \textbf{Benchmark:} make sure implementations are correct, compare them other papers.
-\item \textbf{Optimizations:} improve as much as possible / as long as time permits
 \ole
@@
 \begin{verbatim}
-ideas.txt
----------------------------------------------------
-Plan:
-1. Define the main problem in focus and its class
-   - Dimension
-   - Serial/non-serial
-   - Monadic/polyadic
-2. Define some other problems that belong to this class or simpler classes
-3. See the feasibility for FPGA (possibly helped by CPU)
-4. Implement this algorithm in both CUDA and FPGA
-   -> specify all parameters so that we can compare both implementations
-5. Find a common IR for the problems
-6. Write a code generator for both architectures from IR
-   -> make sure all the test-DP are compiled well from IR
-7. Find an expressive and compact language for the user-facing representation
-
-Translation into C++:
-http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
-
-----------------------------------
-
-Ideas for the language:
-- we want maximum flexibility
-- backtrack can be custom, we may want just to give back indices
-- two problems: we care about cost and we care about whole backtrack.
-	@DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)
-  should be translated into the more complete syntax that allows more flexibility
-  => get inspiration from papers to write an internal language
-
-- we may want to keep track of indices but we do not want to give access to the backtrack matrix / cost matrix
-
-Language side:
-- define a simple way to represent some DP programs
-- define a comprehensive way to define them => micro-syntax
-- define the mapping into IR (actually the IR might BE the micro-syntax)
+1. optimize: terminals of bounded yield + binary decomposition if possible
+2. transform into: axioms (init/fixed) + rules (iterations)
 - define optimizations to IR
@@
 ----------------------------------
-
 Core function F:
@@
 
-To be most efficient, we need to get exact bit-sizes of
-- strings' char
-- costs: min/max values
-- stats: min/max values
-- backtrack
-
-=> we might want to parallelize processing within a thread that operate on 32 or 64 bit inputs (to reduce memory accesses)
+we might want to pack into memory the matrix data and make threads operate on more than one cell.
 
@@
 For backtracking we need to maintain the whole table
-     .. | . |
+... | . |
 ..A | B | min/max: Min(D)=Min(Min(C),Min(B))
@@
 
-Problem parameters:
-- bit size (alphabet,stats,cost,backtrack)
-- dimensionality (2D,3D for the matrix)
-  - 3D problem example is Smith Waterman with arbitrary gap weight function.
-- Summarizability of the set: for computing the stats of an element, we can limit ourselves to computing values up to a certain value
-  -in SW with gap extending penalty, we can summarize easily.
-  -in Nussinov, we _need_ to access all the previous row-column elements for the pairwise sum (without a window limit).
-
-user-interface.txt
---------------------------------------
-Reuse most of the syntactic elements and write it as uneducated programmer, apply annotations/additions on top of it or detect that it is DP and then converge to our system.
-
-Smith-Waterman (from problems of interest)
---------------------------------------------
-
-trait SWOps {
-	type T
-	type Cost = Int
-	type Backtrack = (i,j) => Direction
-	def rightGap(_ :_, c:Char) : [Cost, Backtrack]
-}
-
-def myRec = {
-	def M[i,j] : Int = max {
-		M[i,j-1] + gapRight,
-		M[i-1, j] + gapLeft,
-		M[i-1,j-1] + delta(S(i),T(j))
-	}
-}
-
-def SWat(S0: List[Char], T0:List[Char]) : List[Pair[Char,Char]] = {
-	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function
-
-	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S,T) match {
-		case (_, Nil) => (0, stop)
-		case (Nil, _) => (0, stop)
-		case Si::Sr, Tj::Tr) =>
-			val nw = SWat(Sr,Tr);
-			val n  = SWat(Sr,T);
-			val w  = SWat(S,Tr);
-			val c = cost(Si,Tj);
-
-			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
-			else if (n>=w) then (n._1-d, "N",[Si,_])
-			else (w._1-d, "W",[_,Tj])
-	}
-	backtrack(DP(S0,T0))
-}
-
-- Char, Score, Backtrack, Output are LMS Rep objects so that we can easily detect
-  when we need to use dynamic programming.
-- backtrack() is a new function that take the Backtrack information and reconstruct
-  the desired output backwards
-
-We would unfold this program into two:
-1. The dynamic programming that constructs all the solution and that is the core of the algorithm
-2. The backtrack function that reconstruct the solution and can be run anywhere (Java, CPU, ...)
-
-URLs.txt
-------------------------------------
-Some GPU algorithms:
-http://hgpu.org/?cat=11
-
-CUDPP libraries (but awfully big resulting binary):
-http://code.google.com/p/cudpp/
-
-13 dwarfs:
-http://developer.amd.com/afds/assets/presentations/2155_final.pdf
-
+----------------------------------------------------------------------------
+Some GPU algorithms: http://hgpu.org/?cat=11
+Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
+CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
+13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
 http://tutorials.jenkov.com/java-reflection/fields.html

### 2012-11-14 docs/roadmap/introduction.tex

\newpage
\section{Introduction}
% ------------------------------------------------------------------------------------------------
\subsection{Dynamic programming}
Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
which expands to (first 21 numbers)
\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]

A typical characteristic is that an intermediate solution is reused multiple times to construct larger solutions (here $F(3)$ helps constructing $F(4)$ and $F(5)$). Reusing an existing solution avoid redoing expensive computations: with memoization (memorizing all the intermediate results we obtain) the solution of $F(n)$ would be obtained after only $n$ additions whereas without memoization it would require $F(n)-1$ additions !

Given that characteristic, there exist various categories of dynamic programming:\ul
\item Series that operates usually on a single dimension
\item Sequences alignment (matching two sequences at best), top-down grammar analysis, ...
\item Tree-related algorithms: phylogenetic, trees raking, maximum tree independent set, ...
\ule

Since the first category is inherently sequential and the third category is both hard to parallelize efficiently (similar to a sparse version of the second category) and does not share much with other classes, we focus on the second type of problems, which is also the most common.

Taking real-world examples, the average input size for sequence alignment is around 300K whereas problems like RNA folding input are usually around few thousands. Multiple input problems also require more memory: for instance matching 3 is $O(n^3)$-space complex. Since we target a single computer with one or more attached devices (GPUs, FPGAs), the available storage is relatively small if we compare it to a cluster of machines. Hence in general, we would focus on problems that are around $O(n^2)$-space and $O(n^3)$-time complexity. We encourage you to refer to the section~\ref{problems} for further classification and examples.

% ------------------------------------------------------------------------------------------------
\subsection{Scala and LMS}
{\color{red} XXX}

### 2012-11-14 docs/roadmap/problems.tex

@@
 \newpage
-\section{Dynamic programming problems}
+\section{Dynamic programming problems} \label{problems}
 % ------------------------------------------------------------------------------------------------
@@
 \subsubsection{Definitions}\ul
+\item \textbf{Alphabets:} the alphabet represent an enumeration of the possible values.
+Their size helps determining how many bits are required in the implementation to store the values.
+Alphabets must be defined for input, wavefront, cost and backtrack.
 \item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
@@
 \subsection{Related problems}
-The goal of this section is to demonstrate that our framework can accommodate with many {\color{red} (20-50)} problems that we have not considered at the design time.
+The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.
 
@@
 \item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial
+\ule
 
  {\color{red}
+XXX: Fill up to 20-50 related problems to see how general we are
+\ul
 \item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
@@
 \item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
-}
 \ule
@@
 \end{verbatim}
+}

### 2012-11-14 docs/roadmap/draft.tex
Polygon triangulation example working.

@@
 7.12 & Benchmarking, grammar analysis \\
-14.12 & \\
-21.12 & \\
+14.12 & First though for larger than mem \\
+21.12 & Writing report \\
 28.12 & --- holiday --- \\

### 2012-11-14 src/main/scala/v2/examples/Triangulation.scala

@@ -1,0 +1,50 @@
+package v2.examples
+
+import v2._
+
+// Convex polygon triangulation problem
+trait TriangulationSignature extends Signature {
+  def edgeCost(v1:Char, v2:Char):Int
+}
+
+trait TriangulationAlgebra extends TriangulationSignature {
+  override val cyclic = true
+  override type Alphabet = Char // name of the vertex
+
+  type Answer = (Int,Int,Int,String) // (start, end, cost, pretty-print)
+  def h(l:List[Answer]) = if(l.isEmpty) List() else List(l.minBy(_._3))
+}
+
+object TriangulationApp extends LexicalParsers with TriangulationAlgebra {
+
+  def edgeCost(v1:Char, v2:Char):Int = (v1,v2) match {
+    case (a,b) if (a==b) => -100 // should not be used
+    case ('a','d') | ('a','e') | ('b','d') => 1
+    case ('d','a') | ('e','a') | ('d','b') => 1
+    case _ => 2
+  }
+
+  def triangle = new Parser[Answer] {
+    def apply(sw:Subword) = sw match { // we need to use in() instead of input for cyclic
+      case (i,j) if(j == i+2) => List((i,j, edgeCost(in(i),in(j)), in(i)+""+in(j) ))
+      case (i,j) => List()
+    }
+  }
+
+  val size = input.size
+  def triangulation: Parser[Answer] = tabulate("M",(
+    chari ^^ { i => (i,i+1,0,"") }
+  | triangulation +~+ triangulation ^^ { case((a1,b1,c1,s1),(a2,b2,c2,s2)) => // assert((a2-b1)%size==0);
+      val (s,c) = if (Math.abs(a1%size-b2%size)<=1) ("",0) else (in(a1)+""+in(b2), edgeCost(in(a1),in(b2)))
+      (a1,b2,c1+c2+c,s1+s2+s)
+    }
+  ) aggregate h)
+
+  // XXX: this should be moved as argument of the parse function
+  def input = "abcdef".toArray
+
+  def main(args: Array[String]) = {
+    println(parse(triangulation))
+    //println(gen)
+  }
+}

### 2012-11-15 docs/roadmap/design.tex
Polygon triangulation example working.
Nits.

@@
 However, it does not make sense to treat more than two sequences because of the $O(n^3)$ storage requirements that limits the problems size more dramatically.
-\item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$ best results.
+\item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$-best results.
 \item \textbf{Device restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ...
@@
 XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?
+
+IS CYCLIC EQUIVALENT TO WINDOWING ??
+No: in cyclic we compute 2 triangles, in windowing we compute 3 (no reuse) => make graph to make this explicit
 }

### 2012-11-15 docs/roadmap/problems.tex
Nits.

@@
 \href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
-\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Unbounded Knapsack} \footnotesize (input sensitive) & rectangle & 1 & --\\
+\href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Knapsack} \footnotesize (pseudo-polynomial) & rectangle & 1 & --\\
 \end{tabular}
@@
 - Transposition tables and refutation tables in computer chess
-- Pseudo-polynomial time algorithms for the subset sum and knapsack and partition problems
+- Pseudo-polynomial time algorithms for the subset sum and partition problems
 - The dynamic time warping algorithm for computing the global distance between two time series
@@
 =====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-- 
+
 - Subset Sum, Coin Change, Family Graph

### 2012-11-15 src/main/scala/v2/examples/SmithWaterman.scala

@@ -13,34 +13,124 @@ trait SWatAlgebra extends SWatSignature {
   type Answer = (Int,String,String) // score, string1, string2
   def h(l:List[Answer]) = if(l.isEmpty) List() else List(l.maxBy(_._1))
 
-  private val open = -2
-  private val extend = -1
+  private val open = 2
+  private val extend = 1
   def gap(size:Int) = open + (size-1) * extend
   def pair(a:Char,b:Char) = if (a==b) 1 else 0
 }
 
-object SWatApp extends App with LexicalParsers with SWatAlgebra {
+// ------------------------------------------------------------------------------------------
+
+// Two-tracks combinators
+trait TTParsers { this:Signature =>
+  type Subword = (Int, Int) // (end_in1, end_in2)
+  type Input = Array[Alphabet]
+
+  private var input1: Input = null
+  private var input2: Input = null
+  def in1(k:Int):Alphabet = input1(k)
+  def in2(k:Int):Alphabet = input2(k)
+  def size1:Int = input1.size
+  def size2:Int = input2.size
+
+  import scala.collection.mutable.HashMap
+  val tabs = new HashMap[String,HashMap[Subword,List[Answer]]]
+  val rules = new HashMap[String,Parser[Answer]]
+  def tabulate(name:String, inner:Parser[Answer]) = new Parser[Answer] {
+    val map = tabs.getOrElseUpdate(name,new HashMap[Subword,List[Answer]])
+    rules += ((name,this))
+
+    def apply(sw: Subword) = sw match {
+      case (i,j) => map.getOrElseUpdate(sw, inner(sw))
+    }
+  }
+
+  abstract class Parser[T] { inner =>
+    // duplicate
+    def apply(sw: Subword): List[T]
+    def ^^[U](f: T => U) = this.map(f)
+    private def map[U](f: T => U) = new Parser[U] { def apply(sw:Subword) = inner(sw) map f }
+    def |(that: => Parser[T]) = or(that)
+    private def or (that: => Parser[T]) = new Parser[T] { def apply(sw: Subword) = inner(sw)++that(sw) }
+    def aggregate[U](h: List[T] => List[U]) = new Parser[U] { def apply(sw: Subword) = h(inner(sw)) }
[…]

### 2012-11-16 docs/roadmap/problems.tex
Nits.

@@
 W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
-	W_{(i+1,j)}+b & S\\
-	W_{(i,j-1)}+b & W\\
+	W_{(i+1,j)}+b & L\\
+	W_{(i,j-1)}+b & R\\
 	V_{(i,j)}+\delta(S_i,S_j) & V \\
@@
 	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
-	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
-	V_{(i+1,j-1)}+es(i,j) & IL \\
+	eh(i,j)+b \qquad\qquad\, \text{otherwise} & hairpin\\
+	V_{(i+1,j-1)}+es(i,j) & stack \\
 	VBI_{(i,j)} & (i',j') \\
@@
 F_{(j)}&=&\min\left\{\begin{array}{l|l}
-	F_{(j-1)} & L \\ 
+	F_{(j-1)} & P \\ 
 	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
@@
  \[\begin{array}{rcl}
-	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
+	BF_{(j)} &=& \left\{\begin{array}{rcl} P&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
 	\\
-	BV_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
-		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
-		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
-		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
+	BV_{(i,j)} &=& \left\{\begin{array}{rcl}
+		hairpin &\implies&\big< {\rm hairpin}(i,j) \big> \\
+		stack &\implies& \big< {\rm stack}(i,j) \big> \oplus BV_{(i+1,j-1)} \\
+		(i',j') &\implies& \big< \text{bulge from $(i,j)$ to }(i',j') \big> \oplus BV(i',j')\\
+		k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
 	\end{array}\right.\\
 	\\
-	BW_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
-	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
-	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
+	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
+	L & \implies & \big< multi\_loop(i) \big> \oplus BW_{(i+1,j)} \\
+	R & \implies & \big< multi\_loop(j) \big> \oplus BW_{(i,j+1)} \\
 	V &\implies& BV_{(i,j)} \\
-	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
+	k &\implies& BW_{(i+1,k)} \oplus BW_{(k+1,j-1)}
 	\end{array}\right.
@@
 
-\item Visualization: \begin{center}\includegraphics[width=8cm]{inc/zucker.pdf}\end{center}
-	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
+\item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
+The recurrence consists of two non-serial dependencies as in <<\nameref{aswat}>> plus a bounded 2-dimensional dependency for bulges.
+\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}
+
+Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
+\begin{center}
+\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
+\small An example of an RNA folded into a secondary structure.
+Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).
+\end{center}
+
 \item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?).
-We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?)}
+We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?) Also notice that the $k$ of $BV$ and $BW$ describe almost the same backtrack, but there is an additional cost $c$ in $BV$.}
 \ole
@@
 =====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
-
 - Shortest path in DAGs
@@
 =====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-
 - Subset Sum, Coin Change, Family Graph
 =====> http://www.algorithmist.com/index.php/Dynamic_Programming
-
 - Optimal Binary Search Trees
 =====> http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
-
 - Independent set on a tree
-- 0-1 Knapsack
 =====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf

### 2012-11-28 docs/roadmap/implementation.tex
Nits.

@@
 \subsection{Transformations and simplifications}
+
+\begin{verbatim}
+ * Parser construction:
+ * 1. Assign a "OR_id" (sub_rule_id) and "CONCAT_id" to all parsers so that we know which rule applies
+ *    How to skip some indices ?
+ *
+ * Scala running:
+ * 2. Provide meaningful backtrack: rule_id and list of concat indices
+ *
+ * Code generation:
+ * 1. Normalize rules (at hash-map insertion (?))
+ * 2. Compute dependency analysis between rules => order them into a list/queue
+ *    - If rule R contains another rule S unconcatenated (or concatenated with empty)
+ *      then we have S -> T (S before T)
+ * 3. Compute the maximal number of concatenation among each rule (field in Treeable(?))
+ * 4. Break rules into subrules (at each Or, which must be at top of the rule)
+\end{verbatim}
+
 \subsection{LMS integration}

### 2012-11-28 src/main/scala/v3/CodeGen.scala

@@ -4,7 +4,7 @@ trait CodeGen { this:Signature =>
   val twotracks = false
   trait Treeable {
     def makeTree = tree
-    def tree:PTree
+    val tree:PTree
   }
 
   import scala.collection.mutable.HashMap
@@ -12,8 +12,8 @@ trait CodeGen { this:Signature =>
 
   // Tree structure for recurrences generation
   sealed abstract class PTree {
-    // numAlt = number of alternatives in the subtree
-    // numCC = number of varying-index-concatenations in the subtree
+    var idxAlt = -1; // alternative(subrule) unique identifier
+    var idxCC = -1;  // concatenation backtrack index
     lazy val (numAlt:Int,numCC:Int) = this match {
       case PAggr(_,p) => (p.numAlt,p.numCC)
       case PMap(_,p) => (p.numAlt,p.numCC)
@@ -61,11 +61,45 @@ trait CodeGen { this:Signature =>
   case class CLeq(a:Char,b:Char,delta:Int) extends Cond // 'a'+delta<='b'
   case class CEq(a:Char,b:Char,delta:Int) extends Cond // 'a'+delta=='b'
 
+  // ------------------------------------------------------------------------------
+  // Recurrence analysis, done once when grammar is complete, before the computation.
+  // 1) Unique identifiers for subrules
+  // 2) Maximal backtrack size (cuda: #define TB)
+  // 3) Compute cost storage (cuda: #define TC)
+  // 4) Dependency analysis (order within the pass, possibly if multiple pass are needed)
+
+  private var analyzed=false;
+  def analyze { if (!analyzed) { analyzed=true
+    println("Proceed with tree anlysis to define the backtrack")
+    for((n,p) <- rules) {
+      println(n+" => "+p+" ("+p.tree.numAlt+","+p.tree.numCC+")")
+    }
+  }}
+
+/*
+    val tc = scala.reflect.classTag[TC].runtimeClass.toString
+    val ti:(String,String) = scala.reflect.classTag[TI].runtimeClass match {
+      case cl if (type_jni.contains(cl.toString)) => val cn=cl.toString;
+        ("#define TI "+type_c(cn),"  for (i=0;i<size;++i) (*in)[i] = (TI)env->Get"+up1(cn)+"ArrayElement(input, i);\n")
+      case cl =>
+        //println(manifest[T].erasure.getConstructors.mkString(", "))
[…]

### 2012-11-28 src/main/scala/v3/examples/Brackets.scala

@@ -20,7 +20,7 @@ object Brackets extends LexicalParsers with BracketsAlgebra {
     case(i,j) => j > i+1 && in(i) == '(' && in(j-1) == ')'
   }
 
-  val myParser: Parser[Int] = tabulate("M",(
+  val myParser:Parser[Answer] = tabulate("M",(
       digit
     | (char -~~ myParser ~~- char).filter(areBrackets _).^^{ case (c1,(i,c2)) => i}
     | myParser +~+ myParser ^^ {case (x,y) => x+y}
@@ -32,7 +32,7 @@ object Brackets extends LexicalParsers with BracketsAlgebra {
   // Score of best valid subproblem of size exactely 8 := "(1)((6))"
   override val window = 8
   def main(args: Array[String]) = {
-    println(parse(myParser)("(((3)))((2))(1)((6))((((8))))".toArray))
+    println(parse(myParser)("(((3)))((2))(1)((6))((((8))))"))
     println(gen)
   }
 }

### 2012-11-28 src/main/scala/v3/examples/MatrixMult.scala

@@ -43,7 +43,7 @@ trait PrettyMatrixAlgebra extends MatrixSig {
 }
 
 trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val matrixGrammar: Parser[Answer] = tabulate("M",(
+  val matrixGrammar:Parser[Answer] = tabulate("M",(
     el ^^ single
   | (matrixGrammar +~+ matrixGrammar) ^^ { case (a1,a2) => mult(a1, a2) }
   ) aggregate h)
@@ -51,7 +51,6 @@ trait MatrixGrammar extends ADPParsers with MatrixSig {
 
 object MatrixMult extends MatrixGrammar with PrettyMatrixAlgebra with App {
   val input = List((10,100),(100,5),(5,50)).toArray
-
   println(parse(matrixGrammar)(input))
   println(gen)
 }

### 2012-11-28 src/main/scala/v3/examples/Nussinov.scala

@@ -75,7 +75,6 @@ trait NussinovGrammar extends LexicalParsers with PrettyPairingAlgebra {
 }
 
 object Nussinov extends NussinovGrammar with App {
-  val input = "guaugu".toArray
-  println(parse(s)(input))
+  println(parse(s)("guaugu"))
   println(gen)
 }

### 2012-11-29 docs/roadmap/design.tex
More backtracking theory to help implementation.

@@
 
-\textbf{=> have a pair of subwords for multiple sequences}
-
 We define the aggregate function h as $List[T]\to List[T]$ and a special string emitter for C code that produces the initial element and a comparator between two (cost) elements.
-
-What about the other way round for aggregation: instead of specifying List[U]->List[T] specify comparator of T and a k to keep the k bests(?)
-
-XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?
-
-IS CYCLIC EQUIVALENT TO WINDOWING ??
-No: in cyclic we compute 2 triangles, in windowing we compute 3 (no reuse) => make graph to make this explicit
 }
@@
 In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
-To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack reads are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
+To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
 To deal with this problem, we are facing two options:\ul
@@
 Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.
+
+\subsubsection{Backtracking with multiple backtrack elements}
+The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems).
+However, in the generalization introduced by ADP, it is possible that a matrix cell stores multiple results.
+In such case, we need to select a correct result to avoid backtracking inconsistencies.
+
+Additionally, we need to keep track of the multiplicities of the solutions, that is if we want to obtain the $k$ best solutions, we need to make sure that we return $k$ different backtracks.
+To do that, we maintain a multiplicity counter at every step: \ul
+\item While there is a single solution for all possible incoming paths, we continue in this direction with the same multiplicity (we have no choice).
+\item When there is $r$ different solutions available, and the path multiplicity at this point is $k$ we have the following cases: \ol
+	\item If $k\ge r$: we explore all paths with multiplicity $k-r+1$.
+This is because each branch may produce only one solution and we don't know ahead of time which path will provide multiple solutions.
+Finally, we retain only the $k$ best solutions.
+	\item If $k<r$ (there is more paths than needed): we can only explore the $k$ first paths with multiplicity 1 and safely ignore the other (as we only need $k$ results).
+	\ole
+\ule
+
+Now it remains the problem to generate all possible results and check whether they are correct correct.
+To do that we can simply re-apply the parsers while maintaining the source elements of all production and then retain only those with desired score and backtrack.
+Since we know the backtrack for one element, we can do the following optimization at backtrack parser computation:\ol
+\item Defuse or's: since we know exactly (by the subrule id, maintained in the backtrack) which alternative has been taken to obtain the result, we can skip undesired branches of or parsers.
+\item Feed concatenation indices: since the backtrack stores the concatenation indices, we can reuse in the concatenation parsers.
+This removes the $f(n)$ factor in the backtrack complexity (as concatenation backtrack parsers <<know>> where to split).
+\item Skip filters: since filter are applied before their inner solution is computed, their are only position-dependent.
+Hence if a backtrack involves a filter, since its position is fixed, the filter must have been passed at matrix construction time.
+\ole
+
+\subsubsection{Backtracking complexity}
+We want to have an estimation of the complexity of the two operations to see what is the overhead of multiple backtracking.
+For single-element backtrack, we only need to revert the parser to find involved subwords, which is linear in the parser size (because the backtracking identifies uniquely the alternative in or's and indices in concatenations).
+At every backtrack step, either:\ul
+\item The word is removed at least one character, which leads to maximal backtrack length of $n$.
+\item The word is split in $s$ subwords, with recurrence $f(n)=2f(n/2)+1$ by solving this recurrence we see that there can be at most $n$ final nodes and $n$ intermediate nodes (when $s=2$).
+Hence the backtrack length is at most $2n$.
+\ule
+
+Hence for single backtrack we would have at most $O(2n) \cdot O({\rm parser})$ complexity.
+For the $k$-elements backtrack since we regenerate all possible solutions, that is $O(k^c)$ candidates (with $c$ the maximal number of concatenation in the parser), the overall complexity is $O(2n \cdot k^c) \cdot O({\rm parser})$.
+Hence there is a $k^c$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same factor lies in the matrix computation complexity.}.
+
+%Algo: Read matrix, reapply the parser but by maintaining list of origins. Filter results to match multiplicities
+% à priori ça a une complexité en k^3*2n pour les k meilleurs backtracks
 

### 2012-11-29 docs/roadmap/design.tex
Doc nits.

@@
 
-{\color{red}
-Size analysis to know what storage size we require
-
-We define the aggregate function h as $List[T]\to List[T]$ and a special string emitter for C code that produces the initial element and a comparator between two (cost) elements.
-}
+{\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
 
@@
 \subsection{Parsing grammar (ADP)}
+Borrowing from the ADP grammar, parsers operate on a subsequence of the input (sequence) and return a list of parse results on these subsequences.
+These results are obtained by the internal transformation that is defined within a parser. The parsers are classified by types:\ul
+\item \textbf{Terminal:} operates on a subsequence and returns its content or position (or nothing if the sequence does not fit the terminal).
+\item \textbf{Filter:} this parser accepts only subsequences matching a certain predicate
+\item \textbf{Union:} this parser expresses alternative between two different parsers
+\item \textbf{Concatenation:} construct a larger sequence from two subsequences.
+The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
+\item \textbf{Transformation:} this parser transform its input using an arbitrary function. It is typically used to transform a parse list into a score that can later be aggregated.
+\item \textbf{Aggregation:} the aggregation reduces the list of results, typically using minimum or maximum but can be arbitrarily defined.
+\ule
+
 {\color{red}
-Borrowing from ADP XXX
-
 XXX: explain ADP concepts
@@
 
-XXX: Think how to implement {\tt cyclic} problems and {\tt windowing} properties within the language
-
 \begin{verbatim}
-restrict to only 1 max/min (more restrictive than ADP)
-for optimization we want to split grammar into binary productions, to do that we need
+Optimizations: we want to split grammar into binary productions, to do that we need
 to have equality on the following formula. We can solve it more easily if functions are linear
@@
 \[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]
-
 }

### 2012-11-29 docs/roadmap/introduction.tex
Doc nits.

@@
 \subsection{Scala and LMS}
-{\color{red} XXX}
+\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
+It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive.
+Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org}}
+
+Lightweight Modular Staging (LMS) is a runtime code generation framework built on top of Scala that uses only types to distinguish between code being transformed at compilation and at runtime.
+Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.

### 2012-11-30 docs/roadmap/design.tex
Nits but not backtrack working (empty), find out why !

@@
 Additionally, we need to keep track of the multiplicities of the solutions, that is if we want to obtain the $k$ best solutions, we need to make sure that we return $k$ different backtracks.
-To do that, we maintain a multiplicity counter at every step: \ul
+To do that, we maintain a multiplicity counter in each backtrack process: \ul
 \item While there is a single solution for all possible incoming paths, we continue in this direction with the same multiplicity (we have no choice).
@@
 
-Hence for single backtrack we would have at most $O(2n) \cdot O({\rm parser})$ complexity.
-For the $k$-elements backtrack since we regenerate all possible solutions, that is $O(k^c)$ candidates (with $c$ the maximal number of concatenation in the parser), the overall complexity is $O(2n \cdot k^c) \cdot O({\rm parser})$.
-Hence there is a $k^c$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same factor lies in the matrix computation complexity.}.
+Let one parser reversal complexity be $O(p)$, single backtrack has $O(2n\cdot p)$ complexity.
+For the $k$-elements backtrack, since we regenerate all possible solutions, that is $O(k^{c+1})$ candidates (with $c$ the maximal number of concatenation in the parser), the overall complexity is $O(2n \cdot k^{c+1} \cdot p)$.
+Hence there is a $k^{c+1}$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same factor lies in the matrix computation complexity.}.
 
 %Algo: Read matrix, reapply the parser but by maintaining list of origins. Filter results to match multiplicities
-% à priori ça a une complexité en k^3*2n pour les k meilleurs backtracks
 

### 2012-11-30 src/main/scala/v3/ADPParsers.scala

@@ -10,6 +10,13 @@ class ADPParsers extends BaseParsers { this:Signature =>
     val res = if (cyclic) aggr( ((0 until size).flatMap{ x => p(x,size+x) }).toList, h )
               else if (window>0) aggr( ((0 to size-window).flatMap{ x => p(x,window+x) }).toList, h)
               else p(0,size)
+
+    // XXX: make it clean
+    // XXX: does not work for the moment, find out why
+    println("Backtrack = {")
+    println(backtrack(p(0,size).map{case(x,y)=>((0,size),x,y)}))
+    println("}")
+
     input = null; reset(); res
   }
 

### 2012-12-01 docs/roadmap/design.tex
Some more docs.

@@
 XXX: explain ADP concepts
+
+explain split into signature and grammar
 
@@
 For the $k$-elements backtrack, since we regenerate all possible solutions, that is $O(k^{c+1})$ candidates (with $c$ the maximal number of concatenation in the parser), the overall complexity is $O(2n \cdot k^{c+1} \cdot p)$.
-Hence there is a $k^{c+1}$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same factor lies in the matrix computation complexity.}.
+Hence there is a $k^{c+1}$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same $k^{c+1}$ factor lies in the forward matrix computation complexity.}.
 
-%Algo: Read matrix, reapply the parser but by maintaining list of origins. Filter results to match multiplicities
+\subsubsection{Backtrack utilization}
+Since the dynamic programming input may be a subset of the larger problem that we want to resolve\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation on values in a different domain.
+The easiest way to do that, is to reuse the same grammar on a different domain, and only compute the resulting trace obtained from the DP solver.
+
+This step is pretty straightforward: since ADP parsers emphasize on the split between signature and grammar and decouples them, we only need to modify the signature to operate on another domain, and reuse the same grammar.
+The key point here is to notice that a backtrack trace on one parser can be reused on another, providing that they have the same grammar.
+For instance, to compute optimally a matrix chain multiplication, we solve the DP problem in a domain where matrices are represented by their dimensions, we obtain an optimal trace and feed it to a parser operating on the <<real matrices>> domain that will do the actual computation.
 

### 2012-12-01 docs/roadmap/problems.tex
Algebrae systematic product.
Some more docs.

@@
 \item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
-We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is a lower triangular matrix compacted in memory (rows are contiguous) with a 0 diagonal that is omitted \footnote{Arbitrary convention for both architectural implementation and code generator.
-Rationale: in lower triangular matrix, element address is independent of the matrix size.}, hence $|S|=\tfrac{n^2}{2}=N$.
-
-\item Matrices: $M_{n\times 2n}, B_{n \times 2n}$ \textit{<<first edge, last edge>>} upper triangular including main diagonal 
+We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is an upper triangular matrix with a 0 diagonal that is omitted, hence $|S|=\tfrac{n^2}{2}=N$.
+\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first vertex, last vertex>>} upper triangular including main diagonal. Note that if the vertex $n$ is the same as the vertex $0$.
 \item Alphabets:\ul
@@
 	\ule
+{\color{red}
 \item Initialization: $M_{(i,i)}=0, M_{(i,i+1)}=0, B_{(i,i)}=stop \quad\forall i$
@@
 \item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-	It is interesting to note that even in the sequential world, this problem is solved 
-	by filling the diagonals, ie. computing sub-solutions for all polygons of size $k$ before
-	those of size $k+1$.
 
@@
 "cyclical" in nature.
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,9)
+\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
 	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
 	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cfl{\Cg}\Cfd{\Cz}\Cfu{\Cg}
+	\Cfl{\Cg}
+	\Cfd{\Cz}
+	%\Cfu{\Cg}
 	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 	%\Cd[1,-1]{5,2}{0.8}
-\Cmlong\end{picture}\end{center}
-
-\item Optimizations: we need to rotate that matrix to progress in the same direction as usual, that is towards bottom right.
-{\color{red}}
+	\Cm %\Cmlong
+\end{picture}\end{center}
+}
+\item Optimizations: we need the cost of edges between contiguous polygon vertices to have a cost of 0, so that we do not need to have special cases in the DP program (i.e.
+edges are <<already present>>, hence we cannot add them).
 \ole
@@
 
-In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$.
+With $\delta$ a lookup table. In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$.
 

### 2012-12-02 docs/roadmap/draft.tex
Nits.
Fixed nasty long-lived bugs, created the reflection version of parsers.
Some more docs.

@@
 Our contributions are: \ul
-\item Classification of DP problems characteristics (matrix shape, dependency graph, ...)
-\item State of the art parallel implementation of these classes on GPUs
+\item {\color{red} Reuse of existing compiler technology for a specific purpose}
+%\item Classification of DP problems characteristics (matrix shape, dependency graph, ...)
 \item Grammar embedded in Scala (DSL) to express DP problems efficiently
+%\item State of the art parallel implementation of these classes on GPUs
 %\item Normalization of the grammar into efficient productions
-\item Systematic approach to generate and process backtracking information
+\item Systematic approach to generate and process backtracking information (better than gapc)
 \item Code generator to transform a grammar into efficient code for CPU, GPU and FPGA
@@
 \section{Planning --- Work in progress}
+\subsection{Write introduction}
+\begin{verbatim}
+what exists
+problem
+how we solve it compared to other
+how to evaluate
+contributions (3): 3 tensed sentences
+
+related work
+benchmark => prove by evaluation intro statements
+\end{verbatim}
+
 \subsubsection*{Steps}\ol
@@
 \newpage
-\bibliographystyle{finplain}
+\bibliographystyle{plain}
 \bibliography{bibliography.bib}

### 2012-12-02 docs/roadmap/draft.tex
Nits.

@@
 Our contributions are: \ul
-\item {\color{red} Reuse of existing compiler technology for a specific purpose}
+\item Reuse of existing compiler technology (fusion) for a specific purpose
 %\item Classification of DP problems characteristics (matrix shape, dependency graph, ...)
-\item Grammar embedded in Scala (DSL) to express DP problems efficiently
+\item Grammar embedded in Scala (DSL) to express DP problems concisely
 %\item State of the art parallel implementation of these classes on GPUs
 %\item Normalization of the grammar into efficient productions
-\item Systematic approach to generate and process backtracking information (better than gapc)
-\item Code generator to transform a grammar into efficient code for CPU, GPU and FPGA
+\item Systematic approach to generate and process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
+\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
 \ule

### 2012-12-12 docs/roadmap/design.tex
Some recurrence analysis.

@@
 }
+
+\subsection{Recurrences analysis}
+Since the original ADP introduces many concatenation combinators to differentiate between empty/non-empty, and floating/fixed concatenations, it might become quite involved for the programmer to make sure that the concatenation operator exactly matches the size of each sub-word involved.
+Additionally the priority of operators might vary in Scala, depending the operator's first character whereas it is possible to define arbitrary priorities in Haskell.
+To overcome these problems, we propose to automate the computation of the length of the left and right handside subwords of concat operators.
+
+Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
+A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ...
+\to A$).
+Such case only happen when there is no concatenation or a concatenation with an empty word.
+
+Knowing that fact we can now easily track the dependencies of tabulations and infer an order of computation between them.
+Although unnecessary in a top-down approach (Scala), this order becomes essential in a bottom-up approach (CUDA) to make sure we always have all dependencies computed.
+Once we have such an order we can also trigger a min/max analysis on parsers length.
+Since a parser is made of terminals, concat, operations (aggregate, map, filter) and tabulations and since the min/max length of terminals is fixed, we only need to assign appropriate size to tabulations,  other operation simply need to propagate the information.
+
+It is possible to obtain the yield size using the following algorithm (assuming all parser loops contain at least one tabulation):\ol
+\item Set the yield minimum size of all tabulations to a large
+\item Repeat $k$ times ($k$ is the number of tabulations): for each tabulation, compute its minimum yield size and update its value (without recursion at tabulations).
+This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all tabulations.
+\item Set the maximal yield of all the tabulations to the minimal value.
+For each tabulation, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size.
+If the depth reach the maximum, return infinity.
+\ole
+
+The last part of this algorithm seems to have an exponential complexity, however, if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$.
+Obtaining the yield size of tabulations provides the following benefits:\ul
+\item Minimum size: prevents self-loops (computing one matrix cell using its result) and avoid considering subsequences that would yield empty result (hence reducing time complexity).
+\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(k\cdot n)$ instead of $O(n^2)$ (where $k$ is the maximum yield size), possibly providing substantial space savings.
+\ule
 

### 2012-12-20 docs/roadmap/design.tex
Nits about yield.
Fixed emptiness issue, need to document.

@@
 
+% ------------------------------------------------------------------------------------------------
 \subsection{Recurrences analysis}
@@
 For each tabulation, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size.
-If the depth reach the maximum, return infinity.
+If the depth reaches the maximum $k$, there is a loop between tabulations, hence return infinity.
 \ole
@@
 
+% ------------------------------------------------------------------------------------------------
 \subsection{Backtracking}
@@
 Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
-
 \item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem.
@@
 
+% ----------------------------------------------
 \subsubsection{Backtracking with multiple backtrack elements}
@@
 
+% ----------------------------------------------
 \subsubsection{Backtracking complexity}
@@
 
+% ----------------------------------------------
 \subsubsection{Backtrack utilization}
@@
 For instance, to compute optimally a matrix chain multiplication, we solve the DP problem in a domain where matrices are represented by their dimensions, we obtain an optimal trace and feed it to a parser operating on the <<real matrices>> domain that will do the actual computation.
+
+% ------------------------------------------------------------------------------------------------
+\subsection{CUDA storage: from list to single-value}
+Since lists are one of the primary type of the Scala language, it is natural to gather parsers results into lists in Scala implementation.
+However, when it comes to efficient CUDA implementation, lists must be avoided because memory allocation (and management) is not very efficient.
+A workaround might be to use fixed-length lists but we assume here that the programmer is most often interested in a single optimal solution (this also alleviates the complexity of constructing multiple distinct backtrack traces).
+Even if this design choice apparently greatly simplifies the design issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
+
+\item \textbf{Minimizing memory consumption:} Under the restriction that we only store the best result, we first need to transform aggregation such that they return at most one element.
+Useful aggregator belonging to this class are quite limited: minimum/maximum (optionally with respect to an object's property), count and sum\footnote{Notice that all these operations can be implemented with a folding operation on a single variable.}, hence it is possible to provide them to the user.
+To benefit of this fixed memory aggregation, we need to do some structural transformation of inner parsers.
+In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order.
+\[T < \textbf{Aggregate} < \textbf{Or} < \textbf{Filter} < \textbf{Map} < \textbf{Concat} < \text{(Tabulate | Terminal)}\]
+
+For obvious performance reasons, we want to maintain aggregations wherever they are present. However, we can partially normalize the rest of the evaluation tree:\ul
+	\item We must ensure that all parsers potentially generating multiple possibilities are aggregated.
+To do that, we simply wrap the original parser in an $h$-aggregation (where $h$ is the default aggregation function that must be specified by the user)
+	\item Since the aggregation now operates on a single element, we want to push it as close to the leaves as possible, as long as we do not change operational domain\footnote{We cannot push an aggregation through a mapping or a concatenation operation}.
+	\item Filters can be hoisted within the same concatenation / alternative
+	\item Alternatives must be hoisted outside of maps and concatenations, the reason being that we need to avoid maintaining list of candidates (that will be later aggregated).
+	\ule
+
+\begin{center}
+\begin{tabular}{l|lllll}
+Outer $\diagdown$ inner	& Aggregate	& Or			& Map	& Filter		& Concat \\ \hline
+Aggregate			& merge?		& swap$_R$	& ---		& swap$_P$	& ---\\
+Or					& ---			& simplify?	& ---		& swap$_P$?	& ---\\
+Map					& ---			& swap$_R$	& fuse	& swap$_P$	& ---\\
+Filter					& ---			& ---			& ---		& merge?		& ---\\
+Concat				& ---			& swap$_R$	& ---		& ---			& ---\\
+\end{tabular} \\[10pt]
+\footnotesize{$R=\ $required, $P=\ $performance optimization}
+\end{center}
+
+Note that there can be no swap with Map and Concat internal parsers due to domain change. Fusing is done by the C compiler (declaring mapping functions inline).
+
+\item \textbf{Handling nested aggregations:} since the parser might contain nested aggregation, they must be preserved in order not to increase its time complexity.
+To do that, each internal aggregation has to define its own intermediate variables and backtrack, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
+\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result, {\color{red} XXX: also for terminals!!
+hence we need to encode the failure either in the result or in the backtrack.
+We discuss advantages of each strategy:\ul
+	\item Special 'empty' value:
+	\item Backtrack encoding:
+	\ule
+	XXX: this is a more serious problem
+}
+\ule
 

### 2012-12-20 src/main/scala/v4/CodeGen.scala

@@ -3,8 +3,8 @@ package v4
 trait CodeGen extends BaseParsers { this:Signature =>
   private val head = new CodeHeader(this) // User code and helpers store
   private var order:List[String]=Nil // Order of parsers evaluation
-  private lazy val tpAlphabet = head.getType(tags._1.tpe.toString)
-  private lazy val tpAnswer = head.getType(tags._2.tpe.toString)
+  private lazy val tpAlphabet = if (tags._1==null) "XXX" else head.getType(tags._1.tpe.toString) // XXX: bad fix to avoid initialization issues
+  private lazy val tpAnswer = if (tags._1==null) "XXX" else head.getType(tags._2.tpe.toString)
   
   // Additional typing informations required for codegen
   import scala.reflect.runtime.universe.TypeTag
@@ -98,31 +98,25 @@ trait CodeGen extends BaseParsers { this:Signature =>
 // XXX: rework below
 // XXX: handle user filters and empty results properly
 // XXX: merge different ands together
-
+// XXX: have one global 'valid' switch per alternative ?? does this make sense ??
 
   // Code indentation helper
   def ind(s:String,n:Int=1) = { val i="  "*n; i+s.replaceAll(" +$","").replace("\n","\n"+i)+"\n" }
 
   // Tabulation code generation
   def genTab(t:Tabulate):String = {
+    var aggid=0; // id for intermediate results in nested aggregation
     def scs(min:Int,max:Int,i:Var,j:Var) = if (min==max) List(i.eq(j,min)) else (if (max==maxN) Nil else List(j.leq(i,-max))):::(if (min>0) List(i.leq(j,min)) else Nil)
     def genFun[T,U](f0:T=>U):String = { val f1 = f0 match { case d:DeTuple => d.f case f => f }
       f1 match { case f:CFun => head.add(f) case _ => "UnsupportedFunction" } // TODO: error
     }
-    var aggid=0; // id for intermediate results in nested aggregation
-    var depid=0; // hoisted tabulations (for non-emptiness check)
-   
[…]

### 2012-12-20 docs/roadmap/design.tex
Optimization for full (never empty) tabulations.

@@
 To do that, each internal aggregation has to define its own intermediate variables and backtrack, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
-\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result, {\color{red} XXX: also for terminals!!
-hence we need to encode the failure either in the result or in the backtrack.
-We discuss advantages of each strategy:\ul
-	\item Special 'empty' value:
-	\item Backtrack encoding:
+\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result; failure can happen in terminals, (source) tabulations and filters.
+These 3 cases can be reduced to one by wrapping terminals and tabulations into a filter that checks validity conditions.
+It remains to discuss failure encoding strategies:\ul
+	\item \textbf{Special <<empty>> value: }The benefit of such encoding is a reduced number of memory accesses, indeed since we need to access anyway the value to make computations, we do not generate additional memory accesses to check the validity of the value.
+The drawback of such approach is that it becomes necessary to specify a special <<empty>> value that cannot be used, except to denote the absence of result.
+Since types can be arbitrary at every step of the parser, it becomes cumbersome to ask the DSL user to provide a special value for every intermediate result.
+	\item \textbf{Backtrack encoding:} Reusing the backtrack to encode the validity of a result is a more general approach, and allow greater flexibility for the user.
+Indeed, since we maintain subrules in the backtrack, it suffice to use a special subrule number to denote that related value is invalid.
+	This approach also work with nested aggregations by storing an <<intermediate subrule>> temporary that would only grant validity of the related value.
 	\ule
-	XXX: this is a more serious problem
-}
+Since backtrack encoding comes at the price of an additional memory access to test the validity (memory accesses usually accounts for most of the time on CUDA devices), it is also relevant to allow the user to completely disable this test to speed-up computations.
 \ule

### 2012-12-20 src/main/scala/v4/examples/MatrixMultGen.scala

@@ -37,7 +37,7 @@ trait MatrixGrammarGen extends ADPParsers with MatrixSig {
   val matrixGrammar:Tabulate = tabulate("m1",(
     el ^^ single
   | (matrixGrammar ~ matrixGrammar) ^^ mult
-  ) aggregate h)
+  ) aggregate h,true)
 
   // Let us be more fancy and define testing
   val fooBar:Tabulate = tabulate("m2",matrixGrammar aggregate h)
@@ -49,12 +49,11 @@ trait MatrixGrammarGen extends ADPParsers with MatrixSig {
     val tpe ="Boolean"
   }
 
+  val ffail = new Terminal[Answer](0,0,(i:Var,j:Var)=>(List(CUser("1==0")),"(T3iii){}") ) { def apply(sw:Subword) = Nil }
   val nestedAggr = tabulate("aggr",( // complexity = O(n^2) / element
-    (matrixGrammar ~ matrixGrammar ^^ mult aggregate h) ~
+    (ffail /*matrixGrammar*/ ~ matrixGrammar ^^ mult aggregate h) ~
     ((matrixGrammar filter ps filter ps) ~ matrixGrammar ^^ mult aggregate h) ^^ mult
   ) aggregate h)
-
-  // XXX: use a filter
 
   val axiom=matrixGrammar
 }

### 2012-12-28 docs/roadmap/draft.tex
Nits. Issues with classes as JNI I/O.

--- 2012-12-02 docs/roadmap/draft.tex
@@
 \title{DynaProg for Scala}
-\subtitle{A Scala DSL for Dynamic Programming on GPU and FPGA}
+\subtitle{A Scala DSL for Dynamic Programming on CPU and GPU}
 \begin{document}
-%\maketitle
-\shorttitle
+\maketitle
+%\shorttitle
 
@@
 Additionally, domain experts might not have the skills to make an efficient parallel implementation.
-In this paper, we leverage the Scala / LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs.
+In this project, we leverage the Scala and LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs.
 Our contributions are: \ul
-\item {\color{red} Reuse of existing compiler technology for a specific purpose}
-%\item Classification of DP problems characteristics (matrix shape, dependency graph, ...)
-\item Grammar embedded in Scala (DSL) to express DP problems efficiently
+\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
+\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
+\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
+\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
+%\item Reuse of existing compiler technology (fusion) for a specific purpose
 %\item State of the art parallel implementation of these classes on GPUs
 %\item Normalization of the grammar into efficient productions
-\item Systematic approach to generate and process backtracking information (better than gapc)
-\item Code generator to transform a grammar into efficient code for CPU, GPU and FPGA
+%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
 \ule
+\vfill
+This project has been achieved in collaboration with Manohar Jonnalagedda.
+I would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I hope you will enjoy your reading.
+\vspace{.3cm}\\
+\textit{Thierry Coppey}
 
-\pagestyle{headings}
+\newpage
 \setcounter{tocdepth}{2} \tableofcontents

### 2012-12-28 docs/roadmap/problems.tex
Nits. Issues with classes as JNI I/O.

@@
 \subsubsection{Litterature classification}
-In the literature, dynamic programming problems (DP) are classified according to two criteria:\ul
+In the literature, dynamic programming problems are classified according to two criteria:\ul
 \item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman).
@@
 Unless otherwise specified, we adopt the following conventions:\ul
-\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length.
+\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length (possibly with 1 extra row/column).
 \item Number are all unsigned integers
@@
 \newpage
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 \subsubsection{Smith-Waterman (simple)}\label{sswat}\ol

### 2012-12-28 src/main/scala/v4/examples/MatrixMultGen.scala

@@ -64,8 +64,7 @@ object MatrixMultGen extends MatrixGrammarGen with MatrixAlgebraGen with CodeGen
   val input = List((1,2),(2,20),(20,2),(2,4),(4,2),(2,1),(1,7),(7,3)).toArray // -> 1x3, 122 multiplications
 
   // ------- Extra codegen initialization
-  import scala.reflect.runtime.universe.typeTag;
-  override val tags=(typeTag[Alphabet],typeTag[Answer])
+  override val tps=(manifest[Alphabet],manifest[Answer])
   // ------- Extra codegen initialization
   //println(gen)
   println(backtrackCU(input))

### 2012-12-30 docs/roadmap/draft.tex
Nits. Missing: Zucker, proper user function codegen, report.

@@
 This project has been achieved in collaboration with Manohar Jonnalagedda.
-I would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I also would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
 I hope you will enjoy your reading.

### 2012-12-30 docs/roadmap/problems.tex
Nits. Missing: Zucker, proper user function codegen, report.

@@
 
+Although not necessary to understand the project, the description of some of the most common dynamic programming problem is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems.
+Would the reader be familiar with dynamic programming, he could immediately jump to the next section.
+
 % ------------------------------------------------------------------------------------------------
@@
 \newpage
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 \subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
+\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
 \item Matrices: $M_{m \times n}, B_{m \times n}$
@@
 		0 & stop\\
-		M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+		M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
 		M_{(i-1,j)}-d & N\\
@@
 This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
-	\item Padding: since first line and column of the matrix are zeroes, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
-Also since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
-Hence valid input indices are $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
+	\item Since first line and column of the matrix (corresponding to a matching with one string being empty) are zeroed, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
+	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
+Hence valid input indices would become $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
 We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
@@
 \subsubsection{Smith-Waterman with affine gap extension cost}\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
+
+\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
 \item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
@@
 	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
 	E_{(i,j)} & N\\
@@
 	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
 	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
@@
 \newpage
-\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
+\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+\ol
 \item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.

### 2012-12-30 docs/roadmap/problems.tex
Fixed some problems.

@@
 Unless otherwise specified, we adopt the following conventions:\ul
+\item Vectors of size $n$ are indexed from 0 to $n-1$, matrices follow the same convention
 \item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length (possibly with 1 extra row/column).
@@
 Would the reader be familiar with dynamic programming, he could immediately jump to the next section.
+
+A tighter analysis on the alphabet and intermediate results size is done because FPGA was also considered as a possible execution platform.\footnote{Initially, the project envisioned a collaboration with \href{mailto:nithin.george@epfl.ch}{Nithin George}, a PhD student at Processor Laboratory (LAP), EPFL.
+Unfortunately, at the time of writing, progress in this direction is not sufficient to be covered in this report.
+However, FPGA is considered as a future deployment platform for a particular class of problems.}
 
@@
 \subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
-\item Matrices: $M_{m \times n}, B_{m \times n}$
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
+	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
 	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
@@
 
-\item Backtracking: starts from the cell $\max \{M_{(m,j)} \cup M_{(i,n)}\}$, stops at
-the first cell containing a $0$.
+\item Backtracking: starts from the cell $M_{(m,n)}$ and stops at the first cell containing a $0$.
 \item Visualization: by convention, we put the longest string vertically ($m\ge n$):
@@
 	\item Since first line and column of the matrix (corresponding to a matching with one string being empty) are zeroed, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
-	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
-Hence valid input indices would become $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$.
-We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
+	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$, we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
+Hence valid input indices would become $S[1 \cdots m]$ and $T[1 \cdots n]$.
 	\ule
@@
 
-\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
-\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Matrices: $M_{(m+1) \times (n+1)}, E_{(m+1) \times (n+1)}, F_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
+	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
 	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
@@
 
-\item Backtracking: same as \nameref{sswat}
-\item Visualization: same as \nameref{sswat}
-\item Optimizations: same as \nameref{sswat}
+\item Backtracking, visualization and optimizations are the same as in \nameref{sswat}
 \ole
@@
 \newpage
-\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}
-
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-\ol
-\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
-Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}.
-Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
-\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
+\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
+Without loss of generality, let $m\ge n$ (swap inputs if necessary).
+Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching with smaller gaps.} $g(x)=m-x$.
+\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
 \item Alphabets:\ul
 	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
-	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
+	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
 	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
@@
 	0 & stop\\
-	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
+	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
 	\max_{1 \le k \le j-1} M_{(i,j-k)} - g(k) & N_k\\
@@
 
-\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells.
+\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells along the rows or along the columns.
 \item Visualization:
@@
 
-\item Optimizations: The dependencies here are non-serial, there is no optimization that we can 
-apply out of the box here.
+\item Optimizations: The dependencies here are non-serial, there is no optimization that we can apply out of the box here.
+In general, this problem has an $O(n^3)$ complexity (whereas simple and affine gap variants are $O(n^2)$).
 \ole
@@
 \item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
-We denote the cost of adding an edge between the  pair of edges $i,j$ by $S(i,j)$, Where $S_{n \times n}$ is an upper triangular matrix with a 0 diagonal that is omitted, hence $|S|=\tfrac{n^2}{2}=N$.
-\item Matrices: $M_{n\times n}, B_{n \times n}$ \textit{<<first vertex, last vertex>>} upper triangular including main diagonal. Note that if the vertex $n$ is the same as the vertex $0$.
+We denote the cost of adding an edge between the  pair of edges $i,j$ by $S_{(i,j)}$, Where $S_{n \times n}$ is a symmetric matrix (can be stored as a triangular matrix with 0 diagonal that can be omitted), hence $|S|=\tfrac{n(n-1)}{2}=N$.
+\item Matrices: $M_{(n+1)\times (n+1)}, B_{(n+1)\times (n+1)}$ \textit{indices denote <<first vertex, last vertex>>}.\\ Upper triangular matrices including main diagonal.
+Note that the vertex $n$ is the same as the vertex $0$ due to the cyclic nature of the problem.
 \item Alphabets:\ul
-	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_S(i,j) \forall i,j$ determined at runtime\footnote{We need to scan/have stats about $S$ and that's where LMS plays a role}.
-	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (we add at most $n-2$ edges).
-	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (the index of the edge we add)
+	\item Input: $\Sigma(S_{(i,j)})=\{0..m\}$ with $m=\max_{i,j} S_{(i,j)}$ determined at runtime\footnote{We need to have statistics about $S$, this is where dynamic compilation might play a role}.
+	\item Cost matrix: $\Sigma(M)=\{0..z\}$ with $z = m \cdot (n-2)$ (a triangulation of a polygon of $n$ edges adds at most $n-2$ edges).
+	\item Backtrack matrix: $\Sigma(B)=\{stop, 0..n\}$ (index of intermediate edge)
 	\ule
-{\color{red}
-\item Initialization: $M_{(i,i)}=0, M_{(i,i+1)}=0, B_{(i,i)}=stop \quad\forall i$
-
+\item Initialization: $M_{(i,i)}=M_{(i,i+1)}=0, B_{(i,i)}=B_{(i,i+1)}=stop \quad\forall i$
 \item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for the smaller polygons:
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
-		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
-
-\item Visualization: the layout is the a matrix of size $n \times (2n-2)$, because of polygons being 
-"cyclical" in nature.
+	Intuition: triangulate the partial polygon $(i,..j)$ recursively. For the last triangle of the polygon consider 3 cases:\ul
+	\item Given 2 triangulations $(1..k)$ and $(k..n)$, we close the polygon with $\bigtriangleup(1,k,n)$
+	\item Given a triangulation $(1..n-1)$, we close the polygon with $\bigtriangleup(1,n-1,n)$
+	\item Given a triangulation $(2..n)$, we close the polygon with $\bigtriangleup(1,2,n)$
+	\ule
+	Since the edge to close the last triangle is already part of the polygon, its cost is 0.
+\item Backtracking: Add the edges in the set given by the set ${\rm BT}(B_{(0,n)})$ where
+	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} \{\} & \text{if } k=stop \\
+		 \{(i,j)\} \cup {\rm BT}(B_{(i,k)}) \cup {\rm BT}(B_{(k,j)}) & \text{otherwise} \end{array}\right.\]
+\item Visualization:
 \begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
@@
 \end{picture}\end{center}
-}
-\item Optimizations: we need the cost of edges between contiguous polygon vertices to have a cost of 0, so that we do not need to have special cases in the DP program (i.e.
-edges are <<already present>>, hence we cannot add them).
+
+\item Optimizations: we need the cost of edges between contiguous polygon vertices to have a cost of 0, so that we do not need to handle special cases in the DP program (i.e.
+existing edges cannot be added).
 \ole
@@
 \item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
-	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=0 \lor k=j \\
+	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=stop \\
 		\Big( {\rm BT}(B_{(i,k)}) \Big) \cdot \Big( {\rm BT}(B_{(k+1,j)}) \Big) & \text{otherwise} \end{array}\right.\]
@@
 
-\item Optimizations:\ul
-	\item We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$ and we obtain a new recurrence relation:
-	\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l} M_{(i,k)}+M_{(2i-1 -k,j)}+r_i \cdot c_k \cdot c_j \end{array}\right. \]
-	With the initialization at $M_{(i,n-i-1)}$
-	\ule
+\item Optimizations: we might want to normalize the semantics of indices and use $(n+1) \times (n+1)$ matrices where the meaning of cell $(i,j)$ would be $\underset{i\le k<j}{\rm chain}(A_k)$.
+	%\ul
+	%\item We need to swap vertically the matrix to have a normalized progression towards bottom right. To do that, we need to map all indices $i \mapsto n-1-i$ and we obtain a new recurrence relation:
+	%\[M_{(i,j)}=\min_{i\le k<j}\left\{\begin{array}{l} M_{(i,k)}+M_{(2i-1 -k,j)}+r_i \cdot c_k \cdot c_j \end{array}\right. \]
+	%With the initialization at $M_{(i,n-i-1)}$
+	%\ule
 \ole
@@
 \newpage
-\subsubsection{Nussinov algorithm}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
+\subsubsection{Nussinov algorithm}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+\ol
+\item Problem: folding a RNA string $S$ over itself $|S|=n$.
 \item Matrices: $M_{n\times n}, B_{n \times n}$

### 2012-12-31 docs/roadmap/design.tex
Finished problems rework.

@@
 \section{Architecture design and technical decisions}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 % ------------------------------------------------------------------------------------------------
@@
 The additional features we want to support are:\ol
-\item \textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
-To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
+\item {\color{red}\textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
+To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.}
 \item \textbf{Input pair algebra:} the original ADP framework only support single input, we want to support a pair of inputs such that we can treat problem such as Smith Waterman or Needleman-Wunsch.

### 2012-12-31 docs/roadmap/problems.tex
Finished problems rework.

@@
 Unless otherwise specified, we adopt the following conventions:\ul
-\item Vectors of size $n$ are indexed from 0 to $n-1$, matrices follow the same convention
+\item Vectors of size $n$ are indexed from 0 to $n-1$, matrices follow the same convention ($M_{(m,n)}$ is indexed from $(0,0)$ to $(m-1,n-1)$)
 \item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length (possibly with 1 extra row/column).
@@
 \def\Cfd#1{#1{0,7}#1{1,6}#1{2,5}#1{3,4}#1{4,3}#1{5,2}#1{6,1}#1{7,0}} % main diagonal
+\def\Cfdu#1{#1{7,1}#1{6,2}#1{5,3}#1{4,4}#1{3,5}#1{2,6}#1{1,7}} % upper diagonal
 
-\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
-\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
+%\def\Cmlong{\put(6.5,4.5){\circle*{.4}}\multiput(0,0)(1,0){16}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){15}}} % matrix base
+%\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
 
@@
 	\Cfl{\Cg}
-	\Cfd{\Cz}
-	%\Cfu{\Cg}
+	\Cfd{\Cz}\Cfdu{\Cz}
 	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 	\end{array}\right\}=B_{(i,j)} \]
-\item Backtracking: Start at $B_{(1,n)}$. Use the following recursive function for parenthesizing
+\item Backtracking: Start at $B_{(0,n-1)}$. Use the following recursive function for parenthesizing
 	\[{\rm BT}(B_{(i,j)}=k) \mapsto \left\{\begin{array}{ll} A_i & \text{if } k=stop \\
@@
 
-\item Optimizations: we might want to normalize the semantics of indices and use $(n+1) \times (n+1)$ matrices where the meaning of cell $(i,j)$ would be $\underset{i\le k<j}{\rm chain}(A_k)$.
+\item Optimizations: we could normalize the semantics of indices and use $(n+1) \times (n+1)$ matrices where the meaning of cell $(i,j)$ would be $\underset{i\le k<j}{\rm chain}(A_k)$.
 	%\ul
@@
 \newpage
-\subsubsection{Nussinov algorithm}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-\ol
+\subsubsection{Nussinov algorithm}\ol
 \item Problem: folding a RNA string $S$ over itself $|S|=n$.
@@
 \item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
+	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
 	\item Cost matrix: $\Sigma(M)=\{0..n\}$
@@
 \item Initialization: \ul
-	\item Cost matrix: $ M_{(i,i)}=M_{(i,i-1)}=0$
-	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i-1)}=stop$
+	\item Cost matrix: $ M_{(i,i)}=M_{(i,i+1)}=0$
+	\item Backtrack matrix: $B_{(i,i)}=B_{(i,i+1)}=stop$
 	\ule
@@
 	With $\omega(i,j)=1$ if $i,j$ are complementary. 0 otherwise.
-\item Backtracking: Start the backtracking in $B_{(1,n)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
+\item Backtracking: Start the backtracking in $B_{(0,n-1)}$ and go backward. The backtracking is very similar to that of the matrix multiplication, except that we also introduce the diagonal matching.
 \item Visualization:
@@
 		\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-		\Cfl{\Cg}\Cfd{\Cz}
+		\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
 		\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 \subsubsection{Zuker folding}\ol
-\item Problem: folding a RNA string $S$ over itself $\left\lfloor |S| / 2 \right\rfloor = n$.
+\item Problem: folding a RNA string $S$ over itself $|S|=n$.
 \item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
 \item Alphabets:\ul
-	\item Input: $\Sigma(S)=\{A,C,G,U\}$.
+	\item Input: $\Sigma(S)=\{a,c,g,u\}$.
 	\item Cost matrices:\ul
-		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n*b+c$
+		\item $\Sigma(W)=\Sigma(V)=\{0..z\}$ with $z \le n \cdot b+c$
 		\item $\Sigma(F)=\{0..y\}$ with $y\le \min(F_0, z\cdot n)$
@@
 	\item Backtrack matrices: \ul
-		\item $\Sigma(BW)=\{stop, S,W,V,k\}$
-		\item $\Sigma(BV)=\{stop, HL, IL, SW, (i,j) , k\}$ with $0\le i,j,k < n$ \\
-		$HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
-
-		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
+		\item $\Sigma(BW)=\{stop, L,R,V,k\}$
+		\item $\Sigma(BV)=\{stop, hairpin, stack, (i,j) , k\}$ with $0\le i,j,k < n$ % \\ $HL$=HairpinLoop, $IL$=InteriorLoop, $(i,j)$=MultiLoop
+		\item $\Sigma(BF)=\{stop, P, k\}$ with $0\le k < n$
 		\ule
@@
 \item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
-The recurrence consists of two non-serial dependencies as in <<\nameref{aswat}>> plus a bounded 2-dimensional dependency for bulges.
 \begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}
-
+The recurrence consists of two non-serial dependencies as in \textit{\nameref{aswat}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
 Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
@@
 
-\item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?).
-We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?) Also notice that the $k$ of $BV$ and $BW$ describe almost the same backtrack, but there is an additional cost $c$ in $BV$.}
+\item Optimizations: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction.
+Also notice that the $k$ of $BV$ and $BW$ describe almost the same backtrack, but there is an additional cost $c$ in $BV$.
+%{\color{red} XXX: We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?) }
 \ole
@@
 \subsection{Related problems}
-The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.
+The goal of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems.
+(There is an hyperlink on the problems name to their detailed description).
 
@@
 Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
-Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & -- \\
+Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & (can replace 2 matrices) \\
 \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
@@
 Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
-Convex polygon triangulation & parallelogram & 1 & -- \\
+Convex polygon triangulation & triangle & 1 & -- \\
 Matrix chain multiplication & triangle & 1 & -- \\
 Nussinov & triangle & 1 & -- \\
-Zuker folding & triangle & \color{red} 3? & \color{red} 0? \\
+Zuker folding & triangle & 3 & -- \\
 \href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
@@
 \item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
-\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
+\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial (binary search is more efficient).
 \item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial
@@
 
- {\color{red}
-XXX: Fill up to 20-50 related problems to see how general we are
-\ul
-\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
+These algorithm also involve dynamic programming. However, we do not evaluate thoroughly their shape and number of matrices as this is not relevant in the scope of this project.\ul
+\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}
 \item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
@@
 \href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
-\item Structural alignment (MAMMOTH, SSAP), \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}.
+\item \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}
 \item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
 \item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
+\item \href{http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf}{Shortest path, Shortest path in DAGs, All pair shortest paths, Independent sets in trees}
+\item \href{http://www.algorithmist.com/index.php/Dynamic_Programming}{Subset Sum, Family Graph}
+\item \href{http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf}{Optimal Binary Search Trees}
+\item \href{http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf}{Independent set on a tree}
+\item \href{http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix}{More dynamic programming problems from Wikipedia}
 \ule
 
-\begin{verbatim}
-- Balanced 0-1 matrix
-- Recurrent solutions to lattice models for protein-DNA binding
-- Backward induction as a solution method for finite-horizon discrete-time dynamic optimization problems
-- Method of undetermined coefficients can be used to solve the Bellman equation in infinite-horizon, discrete-time, discounted, time-invariant dynamic optimization problems
-- Many algorithmic problems on graphs can be solved efficiently for graphs of bounded treewidth or bounded clique-width by using dynamic programming on a tree decomposition of the graph.
-- Transposition tables and refutation tables in computer chess
-- Pseudo-polynomial time algorithms for the subset sum and partition problems
-- The dynamic time warping algorithm for computing the global distance between two time series
-- The Selinger (a.k.a. System R) algorithm for relational database query optimization
-- Duckworth-Lewis method for resolving the problem when games of cricket are interrupted
-- The Value Iteration method for solving Markov decision processes
-- Some graphic image edge following selection methods such as the "magnet" selection tool in Photoshop
-- Some methods for solving interval scheduling problems
-- Some methods for solving word wrap problems
-- Some methods for solving the traveling salesman problem, either exactly (in exponential time) or approximately (e.g. via the bitonic tour)
-- Beat tracking in music information retrieval.
-- Adaptive-critic training strategy for artificial neural networks
-- Stereo algorithms for solving the correspondence problem used in stereo vision.
-- Seam carving (content aware image resizing)
-- Some approximate solution methods for the linear search problem.
-=====> http://en.wikipedia.org/wiki/Dynamic_programming#A_type_of_balanced_0.E2.80.931_matrix
-- Shortest path in DAGs
-- Shortest path
-- All pair shortest paths
-- Independent sets in trees
-=> also see exercises for more problems
-=====> http://www.cs.berkeley.edu/~vazirani/algorithms/chap6.pdf
-- Subset Sum, Coin Change, Family Graph
-=====> http://www.algorithmist.com/index.php/Dynamic_Programming
-- Optimal Binary Search Trees
-=====> http://www.cs.uiuc.edu/~jeffe/teaching/algorithms/notes/05-dynprog.pdf
-- Independent set on a tree
-=====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
-\end{verbatim}
-}
+\subsubsection{Conclusion}
+In the rest of the project, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar).
+Although not of immediate use, the description of the above problem and ad-hoc CUDA implementation of three of them (Smith-Waterman with arbitrary gap cost, Matrix chain multiplication and Convex polygon triangulation) helped us to understand:\ol
+\item There is a difference between dynamic programming as seen in algorithmic classes and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
+\item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences). Most of the common dynamic programming problems fall in these two categories.
+\item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
+\ole

### 2012-12-31 docs/roadmap/design.tex
Dead rule elimination, 3.1 reworked.

--- 2012-12-20 docs/roadmap/design.tex
@@
 \subsection{User facing language requirements}
+The field of dynamic programming has been influenced in the recent years by a methodology known as Algebraic Dynamic Programming which uses a grammar and an algebra to separate between the parsing and the score computation:
 \quote{The Algebraic Dynamic Programming approach (ADP) introduces a conceptual splitting of a DP algorithm into a recognition and an evaluation phase.
@@
 Each grammar can be combined with a variety of algebras to solve different but related problems, for which heretofore DP recurrences had to be developed independently.
-Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.
-A notation for yield grammars is provided that serves as a domain-specific language.}
+Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.}
 
@@
 
-The benefits of ADP framework is that it does not constraint the result of the evaluation to be a single value, but could extend parsers to backtracking parsers or pretty-printers.
-Such functionality is easy to support in plain Scala, however, it could hampers the performance of the GPU implementation if we leave to the user the duty of representing and processing all the information within the parsers.
-Its major shortcoming in expressivity is that it only operates on strings, making only triangular-matrix problem easily representable.
-
-The additional features we want to support are:\ol
-\item \textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
-To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
+The benefits of ADP framework is that it does not constraint the result of the evaluation to be a single value, but can extend parsers to backtracking parsers or pretty-printers.
+Additionally, we want to support the following features:\ol
+%\item {\color{red}\textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
+To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.}
 \item \textbf{Input pair algebra:} the original ADP framework only support single input, we want to support a pair of inputs such that we can treat problem such as Smith Waterman or Needleman-Wunsch.
-However, it does not make sense to treat more than two sequences because of the $O(n^3)$ storage requirements that limits the problems size more dramatically.
+However, it does not make sense to treat more than two sequences because of the $\Omega(n^3)$ storage requirements that limits the problems size more dramatically.
 \item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$-best results.
-\item \textbf{Device restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ...
+\item \textbf{Input restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ...
 and structures of them).
@@
 A typical restriction we want to make is that processed structures are of fixed size so that we avoid memory management issues and thread divergence.
+\item \textbf{Single-result on devices:} The general ADP framework supports multiple solutions for intermediate results.
+Such functionality is easy to support in Scala; however, memory management hampers the performance of the GPU implementation.
+To overcome this issue, the user could manually manage the memory, but this would defeat most of the benefits of automatic code generation.
+Hence the trade-off solution we propose is to restrict ADP to only one optimal result on CUDA, while leaving the freedom to obtain co-optimal (or even all possible solutions) with the Scala version.
 
-The general ADP framework supports multiple solutions for intermediate results.
-Similarly, we want to impose the restriction that only the best result would be stored on device while preserving the support for multiple partial results in Scala.
+\item \textbf{Automatic backtracking:} Since efficient code has to be devised we imposed restrictions on the output that could be generated by the parsers on devices.
+However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we would like to to automate the backtracking to fulfill goals of usefulness, efficiency and ease-of-use in device-specific implementation:\ul
+	\item Leaving the backtrack implementation to the user would force him to memoize the backtracking information together with the results (backtrack would  grow towards final result and duplicate unnecessarily information), hence requiring both $O(n^3)$ space and memory management features on devices.
+	\item Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement.
+Collecting the backtracking list can be easily done in $O(n)$ and then inverted whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
+	\ule
+\ole
 
-\item \textbf{Automatic backtracking:} Since efficient code has to be devised we impose restrictions on the output that could be generated by the parsers on devices.
-However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we need to automate the backtracking to both fulfill the goals of efficiency and usefulness of the device specific implementation.
+The support of these features has the following implications:\ul
+\item \textbf{Dependency analysis:} Since we target GPUs (and FPGAs) which are massively parallel architecture, a top-down execution using hash tables is impractical (fallback computation if element is not present is hard to parallelize), hence we need to construct the result tree bottom-up, therefore ensure that the (partial) evaluation order between rules is respected.
+\item \textbf{Normalization:} in order to automate the backtracking, we need the rule to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives).
+Also we need to maintain coherency between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (for instance pretty-printing or effectively multiplying matrices).
 
-Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement while memoizing the backtracking with the results would have the drawback of growing towards final result and duplicated unnecessary information, hence requiring both $O(n^3)$ space and memory management features on devices.
-Collecting the backtracking list can be easily done in $O(n)$ and then inverted whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
-\ole
-Since we construct the dependency tree bottom up, and since we need to work with a fixed-size memory, we need to do the additional steps in our pre-processing:\ul
-\item \textbf{Normalization:} in order to automate the backtracking, we need the rule to present a certain shape (in particular we want to remove alternatives) so that we can define uniquely the backtracking information.
-Another step we can make in the process is to break down complex rules into simpler ones if they are expressible so, hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
-Also we can strip away unnecessary rules.
-\item \textbf{Dependency analysis:} a precise order between the rules application must be given.
-This is required as we use a bottom-up approach where all previous results must be computed versus a top-down approach with a table where a fallback to computation always exist if the desired element is not present.
+\item \textbf{Optimizations:} Another step we could make is to break down complex rules into simpler ones if they are expressible so; hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
+Unfortunately this analysis is very involved: we need to analyze a combination functions $f(a,b,c)$ and find a pair $(f_2,f_3)$ or $(f_3,f_4)$ such that $f(a,b,c)=f_2(f_1(a,b),c)$ or $f(a,b,c)=f_3(a,f_4(b,c))$.
+Since this is overly complicated for the scope of the project, we leave this optimization to the responsibility of the user.
+
+Another optimization that is easier to implement is dead rule elimination.
+Traditional dead code elimination only reduces the size of the generated code.
+In the GPU implementation, this optimization will not only reduce the binary size but also the CUDA memory consumption as well as speed-up computation (since all rules are executed together).
 \ule
-
-{\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
+% {\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
 
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{Parsing grammar (ADP)}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
 Borrowing from the ADP grammar, parsers operate on a subsequence of the input (sequence) and return a list of parse results on these subsequences.
@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{Recurrences analysis}

### 2012-12-31 docs/roadmap/problems.tex
Dead rule elimination, 3.1 reworked.

@@
 \item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
+\item Incidentally, we proposed a cyclic variant of the convex polygon triangulation, which uses a parallelogram matrix instead of a triangular one (upper right triangle extended by a lower triangle on the right).
+Unfortunately, this proved to be based on an erroneous recurrence relation analysis, which could be reduced to the more natural triangular case.
+It is not impossible that some problems do require such special structure, but we did not find any problem that could justify such structure.
 \ole

### 2013-01-01 docs/roadmap/design.tex
More report.

@@
 	\ule
+\item \textbf{Yield analysis:} in the vanilla ADP, the user has to define for each concatenation the minimal and maximal length of the subsequence on each side.
+Although non-emptiness information is necessary to avoid infinite recursion in the parsers, forcing an explicit definition can become cumbersome for the DSL user.
+We want to provide an automatic computation of concatenation boundaries, while at the same time leaving the possibility to manually specify it for maximum flexibility.
 \ole
@@
 
-\item \textbf{Optimizations:} Another step we could make is to break down complex rules into simpler ones if they are expressible so; hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
-Unfortunately this analysis is very involved: we need to analyze a combination functions $f(a,b,c)$ and find a pair $(f_2,f_3)$ or $(f_3,f_4)$ such that $f(a,b,c)=f_2(f_1(a,b),c)$ or $f(a,b,c)=f_3(a,f_4(b,c))$.
-Since this is overly complicated for the scope of the project, we leave this optimization to the responsibility of the user.
+\item \textbf{Optimizations:} An possible optimization is to break down complex rules into simpler ones if they are expressible so; hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
+Unfortunately this analysis is very involved: we need to solve the following problem:
+
+Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
+\[\begin{array}{rcll}
+f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
+	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
+f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
+	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
+\end{array}\]
+% We can solve this more easily if functions are linear
+Since this require complex mathematical analysis that are out of the scope of the project, we leave this optimization to the responsibility of the user.
 
@@
 \subsection{Parsing grammar (ADP)}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+In this section, we provide a concise description of the Algebraic Dynamic Programming.
+Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.
 
-Borrowing from the ADP grammar, parsers operate on a subsequence of the input (sequence) and return a list of parse results on these subsequences.
-These results are obtained by the internal transformation that is defined within a parser. The parsers are classified by types:\ul
-\item \textbf{Terminal:} operates on a subsequence and returns its content or position (or nothing if the sequence does not fit the terminal).
-\item \textbf{Filter:} this parser accepts only subsequences matching a certain predicate
-\item \textbf{Union:} this parser expresses alternative between two different parsers
-\item \textbf{Concatenation:} construct a larger sequence from two subsequences.
+ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
+Such separation makes possible to define multiple algebra for the same grammar.
+This has two main applications:\ol
+\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Watermann share the same grammar but have a different evaluation algebra
+\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data.
+For example in matrix chain multiplication, the first step solves the underlying dynamic problem by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
+\ole
+
+Practically, an ADP program is constituted of 3 components: a \textbf{signature} containing functions signatures, which are implemented by \textbf{algebrae} and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
+The concrete program instance mixes-in the algebra with the grammar.
+The grammar parsers intermediate results are memoized in an array (tabulation parser).
+A parser usually consist of a tree of:\ul
+\item \textbf{Terminal:} operates on a subsequence of input elements and returns either its content or position (or a failure if the sequence does not fit the terminal).
+\item \textbf{Filter:} accepts only subsequences matching a certain predicate. The condition is evaluated ahead of its actual content evaluation.
+\item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
+\item \textbf{Concatenation:} constructs a larger sequence from two subsequences.
 The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
-\item \textbf{Transformation:} this parser transform its input using an arbitrary function. It is typically used to transform a parse list into a score that can later be aggregated.
-\item \textbf{Aggregation:} the aggregation reduces the list of results, typically using minimum or maximum but can be arbitrarily defined.
+\item \textbf{Map:} this parser transform its input using a user-defined function. It is typically used to transform a subword into a score that can later be aggregated.
+\item \textbf{Aggregation:} the aggregation applies a functions that reduces the list of results, typically minimum or maximum, but the function can be arbitrarily defined.
+\item \textbf{Tabulation:} the tabulation's primary function is to store intermediate results and possibly serve as connection point between different parsers.
 \ule
 
-{\color{red}
-XXX: explain ADP concepts
+Additionally, the signature must define input alphabet ({\tt Alphabet}), and output alphabet ({\tt Answer}) can be defined either in the signature or in the algebra.
+Finally, the grammar needs to have a starting point, denoted as axiom.
+Finally, the default aggregation function $h$ must be defined\footnote{Although aggregation usage is not mandatory in the framework, we force the existence of an aggregation function over the ouput type so that we can use it to aggregate windowed results.}.
+To make it more clear, we propose an example of the matrix chain multiplication problem\footnote{The original ADP framework was embedded in Haskell, however, we assume that the reader is more familiar with Scala notation and immediately present the syntax of our implementation.}.
 
-explain split into signature and grammar
+\newpage
+\begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
+trait MatrixSig extends Signature {
+  type Alphabet = (Int,Int) // Matrix(rows, columns)
+  val single:Alphabet=>Answer
+  val mult:(Answer,Answer)=>Answer
+}
 
-XXX: normalization, transforms applied
+trait MatrixAlgebra extends MatrixSig {
+  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
+  override val h = minBy((a:Answer) => a._1)
+  val single = (a: Alphabet) => (0, a)
+  val mult = (a:Answer,b:Answer) =>
+    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
+}
 
-\begin{verbatim}
-Optimizations: we want to split grammar into binary productions, to do that we need
-to have equality on the following formula. We can solve it more easily if functions are linear
-\end{verbatim}
-\[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]
+trait MatrixGrammar extends ADPParsers with MatrixSig {
+  val axiom:Tabulate = tabulate("M",
+    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
 }
+
+object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
+  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
+}
+\end{lstlisting}
+\begin{center}
+\vspace{-24pt}
+with or: $|\,\,$ map: $\land\land\,\,$ concatenation: $\sim$
+\end{center}
+
+\vspace{3cm}
+This program grammar can also be expressed in BNF:
+\[\begin{array}{lrl}
+axiom &::=& \text{matrix} \\
+ &|& axiom \,\,\, axiom
+\end{array}\]
+
+and it encodes the following recurrence (cost only):
+	\[M_{(i,j)}=\left\{\begin{array}{ll}
+		0 & \text{if } i+1= j\\
+		\min_{i<k<j} M_{(i,k)}+M_{(k,j)}+r_i \cdot c_k \cdot c_j & \text{otherwise}
+	\end{array}\right. \]
+
+Notice that the semantics of indices differ slightly from the problem presented in \ref{mat_mult_plain}; this is because empty chain are made expressible (denoted $M_{(i,i)}$, single matrices are denoted $M_{(i,i+1)}$).
 
@@
 \subsection{Recurrences analysis}
-Since the original ADP introduces many concatenation combinators to differentiate between empty/non-empty, and floating/fixed concatenations, it might become quite involved for the programmer to make sure that the concatenation operator exactly matches the size of each sub-word involved.
+Since the original ADP introduces many concatenation combinators to differentiate between empty/non-empty, and floating/fixed-length concatenations, it might become quite involved for the programmer to make sure that the concatenation operator exactly matches the size of each subsequence involved.
 Additionally the priority of operators might vary in Scala, depending the operator's first character whereas it is possible to define arbitrary priorities in Haskell.
@@
 Such case only happen when there is no concatenation or a concatenation with an empty word.
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 

### 2013-01-01 docs/roadmap/design.tex
More report, fixed related nits.

@@
 \subsection{Recurrences analysis}
-Since the original ADP introduces many concatenation combinators to differentiate between empty/non-empty, and floating/fixed-length concatenations, it might become quite involved for the programmer to make sure that the concatenation operator exactly matches the size of each subsequence involved.
-Additionally the priority of operators might vary in Scala, depending the operator's first character whereas it is possible to define arbitrary priorities in Haskell.
-To overcome these problems, we propose to automate the computation of the length of the left and right handside subwords of concat operators.
+\subsubsection{Dead rules elimination}
+Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are mandatory in the grammar).
+The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules.
+Although seemingly useless for the Scala implementation, this step is necessary to maintain coherency between Scala and CUDA rules numbering (that happen in a later stage on the valid rules).
+In CUDA, this analysis not only provides dead code elimination, but it also prevents useless computation execution, since all rules are computed sequentially for a particular subsequence before the next subsequence is processed.
 
-Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
-A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ...
-\to A$).
-Such case only happen when there is no concatenation or a concatenation with an empty word.
+\subsubsection{Yield analysis}
+Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved.
+Additionally the priority of operators varies in Scala, depending the operator's first character whereas it is possible to specify arbitrary priorities in Haskell.
+To overcome these issues, we propose to automate the computation of the minimum/maximum length of (subsequences) parsers.
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-Knowing that fact we can now easily track the dependencies of tabulations and infer an order of computation between them.
-Although unnecessary in a top-down approach (Scala), this order becomes essential in a bottom-up approach (CUDA) to make sure we always have all dependencies computed.
-Once we have such an order we can also trigger a min/max analysis on parsers length.
-Since a parser is made of terminals, concat, operations (aggregate, map, filter) and tabulations and since the min/max length of terminals is fixed, we only need to assign appropriate size to tabulations,  other operation simply need to propagate the information.
-
-It is possible to obtain the yield size using the following algorithm (assuming all parser loops contain at least one tabulation):\ol
-\item Set the yield minimum size of all tabulations to a large
-\item Repeat $k$ times ($k$ is the number of tabulations): for each tabulation, compute its minimum yield size and update its value (without recursion at tabulations).
-This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all tabulations.
-\item Set the maximal yield of all the tabulations to the minimal value.
-For each tabulation, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size.
+Parsers are made of terminals, concat, operations (aggregate, map, filter) and tabulations; minimum/maximum yield of terminals is set, hence it only remains to assign appropriate yield sizes to tabulations; other operations simply propagate that information.
+It is possible to obtain the yield size of tabulations using the following algorithm (assuming recursive parsers contain at least one tabulation that is part of the loop):\ol
+\item Set the yield minimum size of all tabulations to a large number $M_0$ (such that all tabulation would reasonably have a minimum yield size smaller than $M_0$)
+\item Repeat $k$ times ($k$ is the number of rules of the grammar): for each rule, compute its minimum yield size and update its value (without recursion at tabulations).
+This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all rules.
+\item Set the maximal yield of all the rules to the minimal value.
+For each rule, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size.
 If the depth reaches the maximum $k$, there is a loop between tabulations, hence return infinity.
@@
 Obtaining the yield size of tabulations provides the following benefits:\ul
-\item Minimum size: prevents self-loops (computing one matrix cell using its result) and avoid considering subsequences that would yield empty result (hence reducing time complexity).
-\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(k\cdot n)$ instead of $O(n^2)$ (where $k$ is the maximum yield size), possibly providing substantial space savings.
+\item Minimum size: prevents self-loops (computing one matrix cell using its own result) and avoid considering subsequences that would yield empty result (hence slightly reducing time complexity).
+\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(m\cdot n)$ instead of $O(n^2)$ (where $m$ is the maximum yield size), possibly providing substantial space savings.
+As the rules with bounded maximum yield are very rare, we did not implement this optimization, although we might consider it for future work.
+\ule
+
+\subsection{Dependency analysis}
+Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
+A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ...
+\to A$).
+Such case only happen when there is no concatenation or a concatenation with an empty word.
+Being able to track the dependencies of tabulations and infer a computation order between them has two benefits:\ul
+\item Although seemingly unnecessary in a top-down approach (Scala), this analysis detects dependency loops which would result in infinite call loops (stack overflow) at execution.
+\item Ordering tabulations is critical in a bottom-up approach (CUDA) to make sure that all dependencies are valid before an element computation is triggered.
 \ule
@@
 \subsection{Backtracking}
-In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
-To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
+In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
+To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
 To deal with this problem, we are facing two options:\ul
-\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack and is not implemented in vanilla ADP (unless the whole backtrack is made part of the scoring: big performance impact and non-constant memory requirement issues make its GPU implementation hard and not desirable).
-Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
-\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem.
+\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack which is not implemented in ADP, unless we consider the whole backtrack being part of the scoring (which has a big performance impact and non-constant memory requirement issues that make such GPU implementation hard and not desirable).
+Additionally, since the backtracking data is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
+\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the sub-solutions combination that contribute to the problem solution.
 To do that we need to apply the following transformations\ol
-	\item Normalize rules and identify them uniquely by \ul
-		\item Removing all the <<or>> operators by breaking a parser into multiple subrules
-		\item Forwarding names: replace a rule that is a single invocation by its content (i.e. name a terminal parser with its own name instead of its containing parser's)
-		\item Assigning each subrule an unique identifier $r$ and create a mapping table $T: r \to$ user-defined name of the parser.
+	\item Normalize rules and identify them uniquely by exploding alternatives: each rule is decomposed into the union of multiple sub-rules uniquely identified by an index, where sub-rules do not contain alternatives (Or parsers).
+Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
+	\item Let ${\rm cc}(r_s)$ the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
+		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
+		\item The backtrack part is a tuple $(r_s,(k_1, k_2, \ldots , k_m))$ where $m$ is the maximal number of concatenations occurring in the enclosing rule of $r_s$; more formally $m=\max_z\big[{\rm cc}(r_z) | r_z \in {\rm rule}(r_s)\big]$, and let $m_s = {\rm cc}(r_s) \le m$.
 	\ule
-	\item The data elements corresponding to a rule are (scoring, backtracking) and named after the tabulation.
-The scoring is a user-defined primary type (int, double, ...) and the backtracking is a tuple $({\rm id}_{\rm subrule}, k_1, k_2, \ldots , k_m)$ where $m$ is the maximal number of splits that can occur in the subrules.
-	\item During the matrix computation of cell $(i,j)$, if the subrule $r$ applies, the backtrack will be set as $(r,k_1,k_2,\ldots,k_{m_r})$ (obviously with $i\le k_1\le k_2\le \ldots k_{m_r}\le j$ and $m_r\le m$).
-	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r,k_1,k_2,\ldots,k_m)$, given $r$, we recover the subrule hence the user-defined name of the rule that applies and $k_{m_r}$, which allows us to enqueue the subwords $(i,k_1), (k_1,k_2), ..., (k_{m_r})$ for further backtracking.
-If $r$ refers to a terminal, we stop the backtracking.
-	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $(r,i,j)$ where $r$ is the index of the sub-rule that can be easily mapped into the user name (such that $T: r_u \to$ user-defined rule name) and $(i,j)$ is the subword on which the rules applies.
+	\item During the matrix computation of cell $(i,j)$, if the sub-rule $r_s$ applies, the backtrack will be set as $(r_s,(k_1,k_2,\ldots,k_{m_s}))$; with $i\le k_1\le k_2\le \ldots k_{m_s}\le j$.
+Note that if the backtrack occupies a fixed-length memory, the backtrack will contain exactly $m$ indices, hence $k_i\,|\,m_s<i\le m$ will be unspecified.
+	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r_s,(k_1,k_2,\ldots,k_m))$, given $r_s$, we recover $s=T(r_s)$, the sub-rule that applies.
+Hence we can determine $m_s$, which allows us to enqueue the subsequences $(i,k_1), (k_1,k_2), ..., (k_{m_s},j)$ for recursive backtracking.
+If $s$ refers to a terminal, we stop the backtracking.
+	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $((i,j),r_s,(k_1,k_2,\ldots,k_{m_s}))$ where $(i,j)$ denotes the subsequence on which the sub-rule $T(r_s)$ has to be applied with concatenation indices $(k_1,k_2,\ldots,k_{m_s})$.
 \ole
-In short, we break parsers into normalized rules, the backtracking information is the rule id (which rule to unfold) and a list of indices (how to unfold it).
+In short, we break parsers into normalized rules, the backtracking information is the subword, the sub-rule id (which rule to unfold) and a list of indices (how to unfold it).
 \ule
-Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed by the user program in-order, applying for each user-defined rule the underlying transformation, and storing intermediate results (i.e.
+In order to reduce the storage required by the backtracking indices, we can avoid storing fixed indices (where at least one of the two subsequences involved in a concatenation has a fixed size) and leverage the knowledge contained in $s$ to reconstruct the appropriate backtrack.
+
+Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed in-order, applying for each rule the underlying transformation, and storing intermediate results (i.e.
 in a hash map) until they are processed by another rule.
-Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.
+Since there is only one consumer for each intermediate result, every read in the hash map can remove the corresponding value, hence reducing the memory consumption.
+Ultimately, only the problem solution will be stored in the hash map.
 
@@
 \subsubsection{Backtracking with multiple backtrack elements}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
 The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems).

### 2013-01-01 src/main/scala/v4/CodeGen.scala

@@ -2,7 +2,6 @@ package v4
 
 trait CodeGen extends BaseParsers { this:Signature =>
   private val head = new CodeHeader(this) // User code and helpers store
-  private var order:List[String]=Nil // Order of parsers evaluation
   private lazy val tpAlphabet = if (tps._1==null) "XXX" else head.getType(tps._1.toString) // XXX: bad fix to avoid initialization issues
   private lazy val tpAnswer = if (tps._2==null) "XXX" else head.getType(tps._2.toString)
   
@@ -15,22 +14,10 @@ trait CodeGen extends BaseParsers { this:Signature =>
   // Backtracking: back_t = parser_name -> (rule, positions)
   def btTpe(n:Int) = head.addType("short rule"+(if(n>0)"; short pos["+n+"]"else""),"bt"+n)
   override def analyze:Boolean = { if (!super.analyze) return false;
-    // Dependency analysis: computation order between tabulations
-    def deps[T](q:Parser[T]): List[String] = q match { // (A->B) <=> B(i,j) = ... | A(i,j)
-      case Aggregate(p,_) => deps(p) case Filter(p,_) => deps(p) case Map(p,_) => deps(p) case Or(l,r) => deps(l)++deps(r)
-      case cc@Concat(l,r,_) => val(lm,_,rm,_)=cc.indices; (if (lm==0) deps(r) else Nil) ::: (if (rm==0) deps(l) else Nil)
-      case t:Tabulate => List(t.name) case _ => Nil
-    }
-    val cs = rules.map{case (n,p)=>(n,deps(p.inner)) }
-    while (!cs.isEmpty) { var rem=false;
-      cs.foreach { case (n,ds) if (ds.isEmpty||(true/:ds.map{d=>order.contains(d)}){case(a,b)=>a&&b}) => rem=true; order=n::order; cs.remove(n); case _ => }
-      if (rem==false) sys.error("Loop between tabulations, error in grammar")
-    }
-    order.reverse
     // Prepare major structures header
     head.add("#define input_t "+tpAlphabet)
-    head.add("typedef struct { "+order.map{n=>tpAnswer+" "+n+";"}.mkString(" ")+" } cost_t;")
[…]

### 2013-01-01 docs/roadmap/design.tex
Nits.

@@
 \end{lstlisting}
-\begin{center}
-\vspace{-24pt}
-with or: $|\,\,$ map: $\land\land\,\,$ concatenation: $\sim$
-\end{center}
+\begin{center}\vspace{-18pt} with or: $|\,\,$ map: $\land\land\,\,$ concatenation: $\sim$ \end{center}
 
-\vspace{3cm}
+%\vspace{3cm}
 This program grammar can also be expressed in BNF:
@@
 % ------------------------------------------------------------------------------------------------
-\newpage
+%\newpage
 \subsection{Recurrences analysis}
@@
 Obtaining the yield size of tabulations provides the following benefits:\ul
-\item Minimum size: prevents self-loops (computing one matrix cell using its own result) and avoid considering subsequences that would yield empty result (hence slightly reducing time complexity).
+\item Minimum size: prevents self-reference parsers (on the same subsequence) and avoid considering subsequences yielding empty results (hence slightly reducing time complexity).
 \item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(m\cdot n)$ instead of $O(n^2)$ (where $m$ is the maximum yield size), possibly providing substantial space savings.
@@
 
-\subsection{Dependency analysis}
+\subsubsection{Dependency analysis}
 Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
@@
 \subsubsection{Backtracking with multiple backtrack elements}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
 The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems).
 However, in the generalization introduced by ADP, it is possible that a matrix cell stores multiple results.
-In such case, we need to select a correct result to avoid backtracking inconsistencies.
+In such case, we need to select a correct intermediate result to avoid backtracking inconsistencies.
 
-Additionally, we need to keep track of the multiplicities of the solutions, that is if we want to obtain the $k$ best solutions, we need to make sure that we return $k$ different backtracks.
-To do that, we maintain a multiplicity counter in each backtrack process: \ul
-\item While there is a single solution for all possible incoming paths, we continue in this direction with the same multiplicity (we have no choice).
+Additionally, we need to keep track of the multiplicities of the solutions, that is if we want to obtain the $k$ best solutions, we need to make sure that we return $k$ different traces.
+To do that, we maintain a multiplicity counter in each backtrack path: \ul
+\item While there is an unique solution for all possible incoming paths, we continue in this direction with the same multiplicity (we have no choice).
 \item When there is $r$ different solutions available, and the path multiplicity at this point is $k$ we have the following cases: \ol
@@
 Finally, we retain only the $k$ best solutions.
-	\item If $k<r$ (there is more paths than needed): we can only explore the $k$ first paths with multiplicity 1 and safely ignore the other (as we only need $k$ results).
+	\item If $k<r$ (there is more paths than needed): we explore the $k$ first paths with multiplicity 1 and safely ignore the other (as we only need $k$ distinct results).
 	\ole
@@
 
-Now it remains the problem to generate all possible results and check whether they are correct correct.
-To do that we can simply re-apply the parsers while maintaining the source elements of all production and then retain only those with desired score and backtrack.
+Now it remains the problem to generate all possible results and check whether they are valid candidates.
+To do that we simply re-apply the parsers while maintaining the source elements of all production and then retain only those with desired score and backtrack.
 Since we know the backtrack for one element, we can do the following optimization at backtrack parser computation:\ol
-\item Defuse or's: since we know exactly (by the subrule id, maintained in the backtrack) which alternative has been taken to obtain the result, we can skip undesired branches of or parsers.
+\item Defuse alternatives: since we know exactly (by the subrule id $r_s$, maintained in the backtrack) which alternative has been taken to obtain the result, we can skip undesired branches of or parsers.
 \item Feed concatenation indices: since the backtrack stores the concatenation indices, we can reuse in the concatenation parsers.
-This removes the $f(n)$ factor in the backtrack complexity (as concatenation backtrack parsers <<know>> where to split).
+This removes the $O(f(n))$ factor in the backtrack complexity (as concatenation backtrack parsers <<know>> where to split).
 \item Skip filters: since filter are applied before their inner solution is computed, their are only position-dependent.
-Hence if a backtrack involves a filter, since its position is fixed, the filter must have been passed at matrix construction time.
+Hence if a backtrack involves a filter, since its position is set by the backtrack, the filter must have been passed at matrix construction time.
 \ole
@@
 \subsubsection{Backtracking complexity}
-We want to have an estimation of the complexity of the two operations to see what is the overhead of multiple backtracking.
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+We want to have an estimation of the complexity of the two ??XXX??
+operations to see what is the overhead of multiple backtracking.
 For single-element backtrack, we only need to revert the parser to find involved subwords, which is linear in the parser size (because the backtracking identifies uniquely the alternative in or's and indices in concatenations).

### 2013-01-01 docs/roadmap/design.tex
Nits.

@@
 \subsubsection{Backtracking complexity}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+Since the ADP parsers can store multiple results, we are interested in measuring the overhead of $k$-best backtracking (compared to single-element backtracking).
 
-We want to have an estimation of the complexity of the two ??XXX??
-operations to see what is the overhead of multiple backtracking.
-For single-element backtrack, we only need to revert the parser to find involved subwords, which is linear in the parser size (because the backtracking identifies uniquely the alternative in or's and indices in concatenations).
+For single-element backtrack, we only need to <<revert>> the parser to find involved subsequences, which is linear in the parser size (because the backtracking identifies uniquely the alternative and concatenations indices).
+
 At every backtrack step, either:\ul
-\item The word is removed at least one character, which leads to maximal backtrack length of $n$.
-\item The word is split in $s$ subwords, with recurrence $f(n)=2f(n/2)+1$ by solving this recurrence we see that there can be at most $n$ final nodes and $n$ intermediate nodes (when $s=2$).
+\item The sequence is removed at one element, which leads to maximal backtrack length of $n$.
+\item The sequence is split in $k$ subsequences, with recurrence $f(n)=k\cdot f(n/k)+1$ by solving this recurrence we see that there can be at most $n$ final nodes and $n$ intermediate nodes (when $k=2$).
 Hence the backtrack length is at most $2n$.
@@
 
+Another special case we might be interested in is to obtain all co-optimal solutions (all the solutions that have an optimal score).
+We need here to notice that in the parser reversal, no sub-solution is discarded, because either it is not co-optimal (and would have been discarded at an earlier stage) or it is co-optimal, hence contributes to create a co-optimal result.
+It follows that the complexity of co-optimal solutions backtrack is proportional to the number of solutions.
+
 % ----------------------------------------------
 \subsubsection{Backtrack utilization}
-Since the dynamic programming input may be a subset of the larger problem that we want to resolve\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation on values in a different domain.
-The easiest way to do that, is to reuse the same grammar on a different domain, and only compute the resulting trace obtained from the DP solver.
+Since the dynamic programming problem may help solving a larger problem\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation in a different domain.
+The easiest way to do that, is to reuse the same input and grammar, but use a different output domain, and only compute the result of the trace obtained from the DP backtrack.
 
-This step is pretty straightforward: since ADP parsers emphasize on the split between signature and grammar and decouples them, we only need to modify the signature to operate on another domain, and reuse the same grammar.
-The key point here is to notice that a backtrack trace on one parser can be reused on another, providing that they have the same grammar.
-For instance, to compute optimally a matrix chain multiplication, we solve the DP problem in a domain where matrices are represented by their dimensions, we obtain an optimal trace and feed it to a parser operating on the <<real matrices>> domain that will do the actual computation.
+This step is pretty straightforward: since ADP parsers emphasize on the split between signature and grammar and decouples them, we only need to modify the algebra to operate on another domain, and reuse the same grammar.
+The key point here is to notice that a backtrack trace of a parser can be reused by another, providing that they share the same grammar.
+For instance, to compute optimally a matrix chain multiplication, we solve the DP problem in a domain where matrices are represented by their dimensions, we obtain an optimal trace and feed it to another parser operating on <<concrete matrices>> domain that will do the actual matrix multiplication (instead of the cost estimation).
 
 % ------------------------------------------------------------------------------------------------
-\subsection{CUDA storage: from list to single-value}
-Since lists are one of the primary type of the Scala language, it is natural to gather parsers results into lists in Scala implementation.
-However, when it comes to efficient CUDA implementation, lists must be avoided because memory allocation (and management) is not very efficient.
+\subsection{CUDA storage: from list to optional value}
+Since lists are natively supported, it is natural to gather parsers results into lists in Scala implementation.
+However, when it comes to efficient CUDA implementation, lists must be avoided because memory allocation (and management) is not very efficient \cite{scatter_alloc}.
 A workaround might be to use fixed-length lists but we assume here that the programmer is most often interested in a single optimal solution (this also alleviates the complexity of constructing multiple distinct backtrack traces).
-Even if this design choice apparently greatly simplifies the design issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
+Even if this restriction apparently greatly simplifies the design, issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
 
 \item \textbf{Minimizing memory consumption:} Under the restriction that we only store the best result, we first need to transform aggregation such that they return at most one element.
-Useful aggregator belonging to this class are quite limited: minimum/maximum (optionally with respect to an object's property), count and sum\footnote{Notice that all these operations can be implemented with a folding operation on a single variable.}, hence it is possible to provide them to the user.
+Useful aggregator belonging to this class are quite limited: minimum/maximum (optionally with respect to an object's property), count and sum\footnote{Notice that all these operations can be implemented with a folding operation on a single variable.}, hence it is possible to provide the user with a tailored implementation.
 To benefit of this fixed memory aggregation, we need to do some structural transformation of inner parsers.
-In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order.
+In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order:
 \[T < \textbf{Aggregate} < \textbf{Or} < \textbf{Filter} < \textbf{Map} < \textbf{Concat} < \text{(Tabulate | Terminal)}\]
@@
 To do that, we simply wrap the original parser in an $h$-aggregation (where $h$ is the default aggregation function that must be specified by the user)
-	\item Since the aggregation now operates on a single element, we want to push it as close to the leaves as possible, as long as we do not change operational domain\footnote{We cannot push an aggregation through a mapping or a concatenation operation}.
+	\item Since the aggregation now operates on a single element, we want to push it as close to the leaves as possible, as long as we do not change operational domain\footnote{We cannot push an aggregation through a mapping or a concatenation operation.}.
 	\item Filters can be hoisted within the same concatenation / alternative
-	\item Alternatives must be hoisted outside of maps and concatenations, the reason being that we need to avoid maintaining list of candidates (that will be later aggregated).
+	\item Alternatives must be hoisted outside of maps and concatenations, the reason being that we need to avoid maintaining lists of candidates (that will be later aggregated).
 	\ule
-
+	We summarize the required transformations in the following table:
+\begin{table}[H]
 \begin{center}
@@
 \footnotesize{$R=\ $required, $P=\ $performance optimization}
+\caption{Parsers normalization towards CUDA code generation}
 \end{center}
+\end{table}
 
@@
 
-\item \textbf{Handling nested aggregations:} since the parser might contain nested aggregation, they must be preserved in order not to increase its time complexity.
-To do that, each internal aggregation has to define its own intermediate variables and backtrack, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
-\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result; failure can happen in terminals, (source) tabulations and filters.
+\item \textbf{Handling nested aggregations:} since a tabulation might contain nested aggregation, they must be preserved in order not to increase its time complexity.
+To do that, each internal aggregation has to define its own intermediate score and backtrack variables, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
+\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result; failure can happen in terminals, (input) tabulations and filters.
 These 3 cases can be reduced to one by wrapping terminals and tabulations into a filter that checks validity conditions.
 It remains to discuss failure encoding strategies:\ul
-	\item \textbf{Special <<empty>> value: }The benefit of such encoding is a reduced number of memory accesses, indeed since we need to access anyway the value to make computations, we do not generate additional memory accesses to check the validity of the value.
-The drawback of such approach is that it becomes necessary to specify a special <<empty>> value that cannot be used, except to denote the absence of result.
+	\item \textbf{Special <<empty>> value: }The benefit of such encoding is a reduced number of memory accesses; indeed since we anyway need to access the value to make computations, we do not generate additional memory accesses to check the validity of the value.
+The drawback of such approach is that it becomes necessary to specify a special {\tt empty} value that cannot be used, except to denote the absence of result.
 Since types can be arbitrary at every step of the parser, it becomes cumbersome to ask the DSL user to provide a special value for every intermediate result.
 	\item \textbf{Backtrack encoding:} Reusing the backtrack to encode the validity of a result is a more general approach, and allow greater flexibility for the user.
-Indeed, since we maintain subrules in the backtrack, it suffice to use a special subrule number to denote that related value is invalid.
-	This approach also work with nested aggregations by storing an <<intermediate subrule>> temporary that would only grant validity of the related value.
+Indeed, since we maintain sub-rules identifiers in the backtrack, it suffice to use a special identifier to denote that related value is invalid.
+This approach also work with nested aggregations by storing intermediate sub-rule identifiers that would only grant the validity of the related value.
 	\ule
+
 Since backtrack encoding comes at the price of an additional memory access to test the validity (memory accesses usually accounts for most of the time on CUDA devices), it is also relevant to allow the user to completely disable this test to speed-up computations.
@@
 \subsection{Memory constraints}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
 We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.

### 2013-01-01 docs/report/design.tex
Renamed roadmap into report.

--- 2013-01-01 docs/roadmap/design.tex
@@
 	\ule
-
 Since backtrack encoding comes at the price of an additional memory access to test the validity (memory accesses usually accounts for most of the time on CUDA devices), it is also relevant to allow the user to completely disable this test to speed-up computations.
@@
 \subsection{Memory constraints}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
 We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
@@
 If this delay is small, it might be worth to duplicate matrix data in the wavefront, otherwise it might be more efficient to allow access to the previous computational blocks of the matrix.
-\item \textbf{Wavefront size:} Finally some aggregation that is made along some dimension of the matrix does not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
+\item \textbf{Wavefront size:} Finally aggregations that are made along the dimensions of the matrix do not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
 Hence such information can be maintained in a single place (in the wavefront) and progress together with the computation.
 We denote by $S_W$ the size of wavefront elements.
-\item \textbf{Input size:} the size of an input letter (from input alphabet) is denoted by $S_I$.
+\item \textbf{Input size:} the size of an input symbol (from input alphabet) is denoted by $S_I$.
 \ule
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 

### 2013-01-01 docs/report/introduction.tex
Renamed roadmap into report.

\newpage
\section{Introduction}
% ------------------------------------------------------------------------------------------------
\subsection{Dynamic programming}
Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
which expands to (first 21 numbers)
\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]

A typical characteristic is that an intermediate solution is reused multiple times to construct larger solutions (here $F(3)$ helps constructing $F(4)$ and $F(5)$). Reusing an existing solution avoid redoing expensive computations: with memoization (memorizing intermediate results), the solution of $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !

Formally, dynamic programming problems respect the Bellman's principle of optimality: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}. This means that every intermediate result is computed only once, although it might be reused as basis for multiple larger problems, hence our first observation.

There exist various categories of dynamic programming:\ul
\item Series that operates usually on a single dimension (like Fibonacci)
\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ...
\item Tree-related algorithms: phylogenetic, trees raking, maximum tree independent set, ...
\ule

Since the first category is inherently sequential (progress cannot be faster than one element at a time) and the third category is both hard to parallelize efficiently (similar to a sparse version of the second category) and does not share much with the previous category, we focus on the second type of problems, which is also the most common.

Taking real-world examples, the average input size for sequence alignment is around 300K whereas for problems like RNA folding, input are usually around few thousands. Multiple input problems also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex. Since we target a single computer with one or more attached devices (GPUs, FPGAs), and since we plan to maintain data in memory (due to the multiple reuse of intermediate solutions) the storage complexity must be relatively limited, compared to other problem that could leverage the disk storage. Hence in general, we focus on problems that have $O(n^2)$-space complexity whereas time complexity is usually $O(n^3)$ or larger. We encourage you to refer to the section~\ref{problems} for further classification and examples.

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Scala and LMS}
\textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way. It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive. Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org}}

As the Scala programming language is initially developed by the LAMP, it seems natural to use it as host language for our project, however, we would list its features that makes it an interesting development language for this project:\ul
\item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.
\item Scala is largely adopted in the industry, which makes both the adoption of related project easier and offer a steeper learning curve to their potential users.
\item Finally, through the Java VM and JNI interface, Scala offers the possibility to load dynamically external libraries, to leverage best underlying hardware by mixing with CUDA kernels to obtain optimal performance.
\ule

Lightweight Modular Staging (LMS) is a runtime code generation framework built on top of Scala that uses only types to distinguish between code being transformed at compilation and at runtime. Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.

LMS can be leveraged to transform Scala code into its C-like equivalent. However, the concern in this project is that the code for the GPU would be sensibly different from the original CPU code as both implementation serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment. Hence the use of LMS would be restricted to user-specific function, over which our DSL has no control.

### 2013-01-02 docs/report/design.tex
Nits.

@@
 In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order:
-\[T < \textbf{Aggregate} < \textbf{Or} < \textbf{Filter} < \textbf{Map} < \textbf{Concat} < \text{(Tabulate | Terminal)}\]
+\[T < \textbf{Aggregate} < \textbf{Or} < \textbf{Filter} < \textbf{Map} < \textbf{Concat} < \text{(Tabulation | Terminal)}\]
 
@@
 Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
+Without loss of generality, let the underlying dynamic programming matrices be of dimension $m \times n$.
 
@@
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
 % ----------------------------------------------
+\newpage
 \subsubsection{Small problems (in-memory)}
@@
 \subsubsection{Large problems}
-To handle large problems, we need to split the matrix into blocks of size $B_H \times B_W$. For simplification, we assume a square matrix made of square blocks with $b$ blocks per row/column.
+To handle large problems, we need to split the $(m \times n)$ matrix into blocks of size $B_H \times B_W$.
+For simplification in our estimations, we assume a square matrix ($m=n$) made of square blocks with $b$ blocks per row/column ($B_H=B_W=n/b$).
 
@@
 Non-serial problems need to potentially access all elements that have been previously computed.
-We restrict\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.} ourselves to the following dependencies: \ul
+We restrict ourselves to the following dependencies\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.}: \ul
 \item Non-serial dependencies along row and column
@@
 Such restriction implies that all the block of the line and the row, and one additional block to cover diagonal dependencies must be held in memory (independently of the matrix shape).
-
-For simplification, let $m=n$ and assume that we have $b$ square blocks per row and per column. Hence we have the following memory restriction:
+Hence we have the following memory restriction:
 \[ 2\frac{n}{b}(S_I+S_W) + 2 \cdot \frac{n^2}{b}S_C + \frac{n^2}{b^2} S_B \le M_D\]
@@
 Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
-\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}
+\begin{figure}[H]\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}\caption{Transfer overhead for non-serial problems larger than device memory}\end{figure}
 
@@
 Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
-Given that real problems -- like RNA folding -- operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
+Given that real problems (like RNA folding) operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
 
@@
 \subsubsection{Serial problems}
-The serial problem have the interesting property to access to a fixed number of previous elements.
-These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into a wavefront.
-Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems) or line-wise or column-wiser.
+The serial problems have the interesting property to access a fixed number of previous elements.
+These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into the wavefront.
+Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems), line-wise or column-wise.
 This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
@@
 
-Since serial problems are of complexity $O(n^2)$ (due to the matrix dimension and the finite number of dependencies), it is possible to tackle much larger problem than non-serial during the same running time.
-Hence, it seems obvious to let serial problems grow larger than the memory.
+Since serial problems are of complexity $O(n^2)$ (due to the matrix dimension and the finite number of dependencies), it is possible to tackle much larger problem than non-serial given the same running time.
+Hence, it seems natural to let serial problems grow larger than the memory.
 
@@
 
+\newpage
 To store intermediate lines and columns, we are facing two different strategies to explore:\ul
@@
 This division has the advantage of providing the minimal computation time at the expense of external storage proportional to $O(n)$ (if we store only lines or columns) or $O(n^2)$ (if we store both).
-
 \item \textbf{Myers and Miller’s algorithm:} (divide and conquer)
@@
 Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
-\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using twice more storage space would only result in a $2\times$ speedup (measuring only the computation time).
-Hence an hybrid approach would be to decide a $k$ such that at each step we partition the desired submatix into a intermediate grid of $k$ rows/columns.
+\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using $4 \times$ more storage space would only result in a $2\times$ speedup (measuring only the computation time).
+Hence a hybrid approach would be to decide a $k$ such that at each step, we partition the desired sub-matrix into a intermediate grid of $k$ rows/columns.
 The space usage would be in $2 k \log_k (n/b)$ and the running time complexity would be $O(n^2 \cdot \log_k n)$.
@@
 \ule
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 
@@
 \subsection{Memory layout}
+A major bottleneck on massively parallel architecture with shared memory (like CUDA) is the global memory access.
+To do that, and according to the manufacturer documentation, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
+This is justified by the memory hardware architecture, where additional latency (precharge rows and columns of the memory chip) is mandatory to access data at different positions.
+Since all thread share the same global memory (and most of the time the same scheduler), accessing non-contiguous memory cumulates latencies before progress can be made.
+
+% challenge in programming CUDA devices is to make efficient 
 {\color{red} XXX}

### 2013-01-02 docs/report/implementation.tex
Nits.

\newpage
\section{Implementation}

% ------------------------------------------------------------------------------------------------
\subsection{CUDA ad-hoc}
During the first phase of project, and after the decision to focus on non-serial problems implementation on GPU, we developed an ad-hoc version of 3 non-serial problems. Since we erroneously complicated the matrix of the polygon triangulation problem, we devised 3 different matrix shape, that is upper triangular for single-track problems, rectangular for two-track problems and parallelogram for the polygon triangulation.

Although the polygon triangulation problem can be solved with a regular triangular matrix and we have not found real problem requiring a parallelogram matrix, we still present this version, which could be used for cyclic problem that require dynamic programming for each unfolding position (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting sequence).

%\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
%\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store

\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,2.5)
	\put(1,0){\moveto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath\put(1.1,1.1){\linethickness{1.5pt}\vector(1,1){.75}}
		\multiput(0,0)(0,.1){20}{\circle*{.03}}\multiput(0,0)(.1,0){20}{\circle*{.03}}}
	\put(0,-.8){\makebox[4\unitlength]{Triangle}}

	\put(7,0){\moveto(0,0)\lineto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
		\put(.5,1.5){\linethickness{1.5pt}\vector(1,-1){1}}}
	\put(6,-.8){\makebox[4\unitlength]{Rectangle}}

	\put(12,0){\moveto(2,0)\lineto(4,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
		\put(1.6,0.6){\linethickness{1.5pt}\vector(1,1){.8}}
		\multiput(0,0)(2,0){3}{\multiput(0,0)(0,.1){20}{\circle*{.03}}}
		\multiput(0,0)(.1,0){20}{\circle*{.03}}\multiput(4,2)(-.1,0){20}{\circle*{.03}}}
	\put(12,-.8){\makebox[4\unitlength]{Parallelogram}}
\end{picture}\end{center}\caption{Matrix shapes, the arrow indicates the computation wavefront progress direction}
\end{figure}

\subsection{LibRNA (?)}

\begin{verbatim}
1 week on hash maps
2 weeks to define and analyze the problems (also with FPGA in mind)
3 weeks restriction to non-serial and ad-hoc implementation on GPU (rectangle, triangle, parallelogram); translation of ADP parsers in Scala (Manohar)
1 week run-time engine for Scala/JNI/CUDA
05.11 - rework of the ADP parser to aim at generating C-like code
12.11 - rework of ADP parsers to extend usages (cyclic, two-track) and aim at automatic backtracking
19.11 - explorations in LMS / macros
26.11 - full backtracking: apply/unapply/reapply, rework of the classes
03.12 - Zuker/JNI
10.12 - Yield analysis, code generation
17.12 - code generation: detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support
24.12 - code generation, sick
31.12 - report
\end{verbatim}

\subsection{ADP parsers}
\subsection{Transformations and simplifications}

\begin{verbatim}
 * Parser construction:
 * 1. Assign a "OR_id" (sub_rule_id) and "CONCAT_id" to all parsers so that we know which rule applies
 *    How to skip some indices ?
 *
 * Scala running:
 * 2. Provide meaningful backtrack: rule_id and list of concat indices
 *
 * Code generation:
 * 1. Normalize rules (at hash-map insertion (?))
 * 2. Compute dependency analysis between rules => order them into a list/queue
 *    - If rule R contains another rule S unconcatenated (or concatenated with empty)
 *      then we have S -> T (S before T)
 * 3. Compute the maximal number of concatenation among each rule (field in Treeable(?))
 * 4. Break rules into subrules (at each Or, which must be at top of the rule)
\end{verbatim}

\subsection{LMS integration}
Why cannot use LMS for everything
Where and how is LMS useful

\subsection{Code generation}

\subsection{Runtime execution engine}

\subsection{CUDA implementation}

### 2013-01-02 docs/report/introduction.tex
Nits.

@@
 \subsection{Scala and LMS}
+\subsubsection{Scala}
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
@@
 
+\subsubsection{LMS}
 Lightweight Modular Staging (LMS) is a runtime code generation framework built on top of Scala that uses only types to distinguish between code being transformed at compilation and at runtime.

### 2013-01-02 docs/report/problems.tex
Renamed roadmap into report.
Nits.

--- 2012-12-31 docs/roadmap/problems.tex
@@
 % Recurrence visualization helpers
+
 \newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
@@
 %\def\Cfu#1{#1{8,7}#1{9,7}#1{10,7}#1{11,7}#1{12,7}#1{13,7}#1{14,7}#1{9,6}#1{10,6}#1{11,6}#1{12,6}#1{13,6}#1{14,6}#1{10,5}#1{11,5}#1{12,5}#1{13,5}#1{14,5}#1{11,4}#1{12,4}#1{13,4}#1{14,4}#1{12,3}#1{13,3}#1{14,3}#1{13,2}#1{14,2}#1{14,1}} % triangular upper (function)
+
+\def\CFig#1#2{\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9) #2 \Cm\end{picture}\end{center}\caption{#1}\end{figure}}
 
@@
 \item Visualization: by convention, we put the longest string vertically ($m\ge n$):
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\CFig{Smith-Waterman (affine gap cost) dependencies (serial)}{
 	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
@@
 	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
-	\Cd{6,5}{0.8}
-	\Cd[1,0]{5,4}{0.8}
-	\Cd[1,-1]{5,5}{0.8}
-\Cm\end{picture}\end{center}
+	\Cd{6,5}{0.8}\Cd[1,0]{5,4}{0.8}\Cd[1,-1]{5,5}{0.8}
+}
 
@@
 \[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
-Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
-Also notice that it is possible to propagate $E$ and $F$ inside a resp.
-horizontal and vertical wavefront.
 
-\item Backtracking, visualization and optimizations are the same as in \nameref{sswat}
+\item Backtracking: similar to \nameref{sswat}
+\item Visualization: same as \nameref{sswat}
+\item Optimizations: Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
+Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of two additional matrices.
 \ole
@@
 \item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\CFig{Smith-Waterman (arbitrary gap cost) dependencies}{
 	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
@@
 	\Cd[1,-1]{5,5}{0.8}
-	\Cm\end{picture}\end{center}
+}
 
@@
 \item Recurrence: \[M_{(i,j)}=\left\{ S(i,j) + \max_{i<k<j}M_{(i,k)}+M_{(k,j)} \,\,\rule[-.75em]{.5pt}{2em}\,\,  k \right\} = B_{(i,j)} \]
-	Intuition: triangulate the partial polygon $(i,..j)$ recursively. For the last triangle of the polygon consider 3 cases:\ul
+	Intuition: triangulate the partial polygon $(i,..j)$ recursively. 3 cases for the last triangle:\ul
 	\item Given 2 triangulations $(1..k)$ and $(k..n)$, we close the polygon with $\bigtriangleup(1,k,n)$
@@
 \item Visualization:
-\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\CFig{Convex polygon triangulation dependencies}{
 	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
 	\put(.2,8.2){Last}\put(1.5,8.4){\linethickness{1pt}\vector(1,0){2}}
-	\Cfl{\Cg}
-	\Cfd{\Cz}\Cfdu{\Cz}
+	\Cfl{\Cg}\Cfd{\Cz}\Cfdu{\Cz}
 	\Cd[0,1]{6,1}{2.8}\Cd[0,1]{6,2}{1.8}\Cd[0,1]{6,3}{0.8}
@@
 	\put(5.5,4.5){\line(1,-3){1}}
-	%\Cd[1,-1]{5,2}{0.8}
-	\Cm %\Cmlong
-\end{picture}\end{center}
+}
 
-\item Optimizations: we need the cost of edges between contiguous polygon vertices to have a cost of 0, so that we do not need to handle special cases in the DP program (i.e.
-existing edges cannot be added).
+\item Optimizations: if the cost of edges between contiguous vertices is 0, we do not need to handle special cases in the DP program (i.e. existing edges cannot be added).
 \ole
@@
 \newpage
-\subsubsection{Matrix chain multiplication}\ol
+\subsubsection{Matrix chain multiplication}\label{mat_mult_plain}\ol
 \item Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$.
@@
 \item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\CFig{Matrix chain multiplication dependencies}{
 	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
@@
 	\put(3.5,4.5){\line(3,-1){3}}\put(4.5,4.5){\line(2,-2){2}}\put(5.5,4.5){\line(1,-3){1}}
-	\Cm\end{picture}\end{center}
+}
 
@@
 \item Visualization:
-	\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
+\CFig{Nussinov dependencies}{
 	\put(-.7,6.5){\rotatebox{90}{First}}\put(-.4,6.4){\linethickness{1pt}\vector(0,-1){2}}
@@
 \item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
-\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}
+\begin{figure}[H]\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}\caption{Zuker folding dependencies}\end{figure}
+
 The recurrence consists of two non-serial dependencies as in \textit{\nameref{aswat}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
 Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
-\begin{center}
+\begin{figure}[H]\begin{center}
 \includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
-\small An example of an RNA folded into a secondary structure.
-Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).
-\end{center}
+\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}
 
@@
 \subsection{Related problems}
-The goal of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems.
+The aim of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems.
 (There is an hyperlink on the problems name to their detailed description).
 
-\subsubsection{Serial problems}
-\begin{tabular}{llcc} \toprule
-\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
+\begin{table}[H]\begin{center}\begin{tabular}{llcc} \toprule
+\bf Serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
 Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
@@
 \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
-
 \href{http://en.wikipedia.org/wiki/Dynamic_programming#Checkerboard}{Checkerboard} & rectangle & 1 & -- \\
@@
 \href{http://en.wikipedia.org/wiki/Levenshtein_distance\#Computing_Levenshtein_distance}{Levenshtein distance} & rectangle & 1 & -- \\
-\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \\
-\end{tabular}
-
-\subsubsection{Non-serial problems}
-\begin{tabular}{llcc} \toprule
-\bf Problem & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
+\href{http://en.wikipedia.org/wiki/De_Boor's_algorithm}{De Boor} \footnotesize evaluating B-spline curves & rectangle & 1 & -- \vspace{.4cm} \\
+\toprule
+\bf Non-serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
 Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
@@
 \href{http://en.wikipedia.org/wiki/Knapsack_problem#Dynamic_programming}{Knapsack} \footnotesize (pseudo-polynomial) & rectangle & 1 & --\\
-\end{tabular}
+\end{tabular}\end{center}
+\caption{Classification of related problems}
+\end{table}
 
@@
 Unfortunately, this proved to be based on an erroneous recurrence relation analysis, which could be reduced to the more natural triangular case.
-It is not impossible that some problems do require such special structure, but we did not find any problem that could justify such structure.
+
+Although the polygon triangulation problem can be solved with a regular triangular matrix and we have not found real problem requiring a parallelogram matrix, we still present this version in the analysis part of the project, which could be justified for cyclic problems that require dynamic programming for each unfolding position (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting flattened sequence).
+For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.
+
 \ole

### 2013-01-02 docs/report/design.tex
Nits.

@@
 
-% challenge in programming CUDA devices is to make efficient 
-{\color{red} XXX}
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,2.5)
+	\put(1,0){\moveto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath\put(1.1,1.1){\linethickness{1.5pt}\vector(1,1){.75}}
+		\multiput(0,0)(0,.1){20}{\circle*{.03}}\multiput(0,0)(.1,0){20}{\circle*{.03}}}
+	\put(0,-.8){\makebox[4\unitlength]{Triangle}}
+
+	\put(7,0){\moveto(0,0)\lineto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
+		\put(.5,1.5){\linethickness{1.5pt}\vector(1,-1){1}}}
+	\put(6,-.8){\makebox[4\unitlength]{Rectangle}}
+
+	\put(12,0){\moveto(2,0)\lineto(4,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
+		\put(1.6,0.6){\linethickness{1.5pt}\vector(1,1){.8}}
+		\multiput(0,0)(2,0){3}{\multiput(0,0)(0,.1){20}{\circle*{.03}}}
+		\multiput(0,0)(.1,0){20}{\circle*{.03}}\multiput(4,2)(-.1,0){20}{\circle*{.03}}}
+	\put(12,-.8){\makebox[4\unitlength]{Parallelogram}}
+\end{picture}\end{center}\caption{Matrix shapes, the arrow indicates the computation progress direction}\end{figure}
+
+Since progress is made along the diagonal of the matrices, a naive row/column addressing of the matrix elements would result in no coalesced access.
+Hence instead of addressing by row the matrix elements, we address them by diagonal.
+Let $M_W$ the width of the matrix and $M_H$ its height.
+Coalesced addressing is most easy in the parallelogram matrix, as we just need to map parallelogram indices to addresses into a square matrix (with diagonal elements being contiguous in memory) as follows:
+\[(i,j) \to i + (j-i)*M_H\]
+
+Noticing that the triangle is just half the parallelogram, we could use the same addressing, but we would use twice the memory necessary.
+Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl.
+diagonal) is $\frac{k\cdot(k+1)}{2}$ knowing that fact, the index of the $i^{\rm th}$ element on the $j^{\rm th}$ diagonal of the triangle (starting with the main diagonal when $j=0$), we obtain the following formula (with $M_H=M_W=n$):
+\[(i,j) \to M-T+i \text{ with }\left\{\begin{array}{rcll}
+	M&=&\dfrac{(n\cdot (n+1)}{2} \vspace{.15cm} & \textit{total }\bigtriangleup\\
+	d&=&n+1+i-j \vspace{.15cm} & \textit{diagonal of }(i,j)\\
+	T &=&\dfrac{(d\cdot (d+1)}{2} & \bigtriangleup \textit{ of current diagonal}
+\end{array}\right.\]
+
+Finally, for the rectangular matrix, since the parallelogram indexing looks efficient, we want to reuse the same idea.
+However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle).
+A solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, hereby dramatically reducing the size of the triangles needed.
+Finally, the stripes can be stitched together to form a single parallelogram (beyond a certain number of lines, coalescing access does not improve as memory is anyway not stored contiguously).
+
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(4,3.5)
+\put(0,-.5) {
+	\put(0,0){\color{lightgray}\moveto(4,0)\lineto(4.5,.5)\lineto(4,.5)\closepath\fillpath\moveto(0,4)\lineto(-.5,3.5)\lineto(0,3.5)\closepath\fillpath}
+	\put(1,3){\linethickness{1.5pt}\vector(1,-1){2}}
+	\multiput(0,.5)(0,.5){7}{\line(1,0){4}}
+	\moveto(0,0)\lineto(4,0)\lineto(4.5,.5)\lineto(4,.5)\lineto(4,4)\lineto(0,4)\lineto(-.5,3.5)\lineto(0,3.5)\closepath\strokepath
+}
+\end{picture}\end{center}\caption{Parallelogram memory representation of a rectangular matrix}\end{figure}
+
+It follows that the total memory required to store the matrix is:
+\[M_W \cdot \left\lceil\dfrac{M_H}{B_H}\right\rceil \cdot B_H  + (B_H)^2\]
+
+and the mapping of indices is given by
+\[ (i,j) \to (B_H \cdot \underset{\rm diagonal}{(j + m)} + m + \underset{\text{stripe}}{\left\lfloor \dfrac{i}{B_H} \right\rfloor} \cdot \underset{\text{stripe size}}{M_W \cdot B_H}) \quad \text{with } m=i \mod B_H \]
 

### 2013-01-02 docs/report/implementation.tex
Nits.

@@
 Since we erroneously complicated the matrix of the polygon triangulation problem, we devised 3 different matrix shape, that is upper triangular for single-track problems, rectangular for two-track problems and parallelogram for the polygon triangulation.
-
-Although the polygon triangulation problem can be solved with a regular triangular matrix and we have not found real problem requiring a parallelogram matrix, we still present this version, which could be used for cyclic problem that require dynamic programming for each unfolding position (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting sequence).
-
-%\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
-%\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
-
-\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,2.5)
-	\put(1,0){\moveto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath\put(1.1,1.1){\linethickness{1.5pt}\vector(1,1){.75}}
-		\multiput(0,0)(0,.1){20}{\circle*{.03}}\multiput(0,0)(.1,0){20}{\circle*{.03}}}
-	\put(0,-.8){\makebox[4\unitlength]{Triangle}}
-
-	\put(7,0){\moveto(0,0)\lineto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
-		\put(.5,1.5){\linethickness{1.5pt}\vector(1,-1){1}}}
-	\put(6,-.8){\makebox[4\unitlength]{Rectangle}}
-
-	\put(12,0){\moveto(2,0)\lineto(4,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
-		\put(1.6,0.6){\linethickness{1.5pt}\vector(1,1){.8}}
-		\multiput(0,0)(2,0){3}{\multiput(0,0)(0,.1){20}{\circle*{.03}}}
-		\multiput(0,0)(.1,0){20}{\circle*{.03}}\multiput(4,2)(-.1,0){20}{\circle*{.03}}}
-	\put(12,-.8){\makebox[4\unitlength]{Parallelogram}}
-\end{picture}\end{center}\caption{Matrix shapes, the arrow indicates the computation wavefront progress direction}
-\end{figure}
 

### 2013-01-02 docs/report/design.tex
More report.

@@
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-{\color{red}
-try to setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront
-
-XXX: what's the maximal size of the wavefront ??
-
-XXX: can we avoid to store some matrices and put them in the wavefront ??
-
-Split into blocks:\ul
-\item Decide the shape of the blocks
-\item Decide the size of the blocks
-\item Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
+Finally, although such problem is interesting because targeted platforms could include FPGA, where efficient implementations exist \cite{swat_fpga}, several reasons made us considering this class of problem as a future work: \ul
+\item The most prominent problem in this category is Smith-Waterman, for which an ad-hoc efficient implementation already exists\cite{swat_gpu},\cite{swat_mega}.
+Additionally, the authors are planning to write extensions to support some variants of this problem like Needleman-Wunsch.
+\item The implementation sensibly differs from the class of small problems, as the solving strategy is completely different from non-serial small problems, hereby requiring larger development time that would be out of the scope of this project.
+\item Finally, such implementation would be only valuable for problems that are larger than the memory device, whereas smaller problem could perfectly use the existing implementation.
 \ule
 
-XXX: make it work up to 14K for all 3 problems using multiple kernels
-XXX: example of non-symmetric serial problem
-
-The three last elements, combined with the above one, provide a precise estimation of the memory consumption, and the implementation difficulty
-
-all the problem are subject to the input dimensions $n$.
-
-the two latter one gives an estimation of the constant factor.
-
-The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
+% Setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront. What's the maximal size of the wavefront ?
+% Can we avoid to store some matrices and put them in the wavefront ??
+% Split into blocks:
+% - Decide the shape of the blocks
+% - Decide the size of the blocks
+% - Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
+% The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
 Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
-}
 
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{Memory layout}
 A major bottleneck on massively parallel architecture with shared memory (like CUDA) is the global memory access.
-To do that, and according to the manufacturer documentation, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
-This is justified by the memory hardware architecture, where additional latency (precharge rows and columns of the memory chip) is mandatory to access data at different positions.
+To address it, according to the manufacturer documentation, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
+This is justified by the memory hardware architecture, where additional latency (precharge rows and columns of the memory chip) is mandatory to access data at very different positions.
 Since all thread share the same global memory (and most of the time the same scheduler), accessing non-contiguous memory cumulates latencies before progress can be made.
@@
 Since progress is made along the diagonal of the matrices, a naive row/column addressing of the matrix elements would result in no coalesced access.
-Hence instead of addressing by row the matrix elements, we address them by diagonal.
+Hence instead of addressing by row (or column) the matrix elements, we address them by diagonal.
 Let $M_W$ the width of the matrix and $M_H$ its height.
-Coalesced addressing is most easy in the parallelogram matrix, as we just need to map parallelogram indices to addresses into a square matrix (with diagonal elements being contiguous in memory) as follows:
+Coalesced addressing is most easy in the parallelogram matrix, as we could pretend that the parallelogram is simply a tilted rectangle  (with diagonal elements being contiguous in memory) as follows:
 \[(i,j) \to i + (j-i)*M_H\]
 
-Noticing that the triangle is just half the parallelogram, we could use the same addressing, but we would use twice the memory necessary.
+Noticing that the triangle is just half the parallelogram, we could use the same addressing, but we would use twice the necessary memory.
 Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl.
-diagonal) is $\frac{k\cdot(k+1)}{2}$ knowing that fact, the index of the $i^{\rm th}$ element on the $j^{\rm th}$ diagonal of the triangle (starting with the main diagonal when $j=0$), we obtain the following formula (with $M_H=M_W=n$):
+diagonal) is $\frac{k\cdot(k+1)}{2}$; knowing that fact, the index of the element $(i,j)$ of the triangle (starting with the main diagonal when $i=j$), we obtain the following formula (with $M_H=M_W=n$):
 \[(i,j) \to M-T+i \text{ with }\left\{\begin{array}{rcll}
@@
 However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle).
-A solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, hereby dramatically reducing the size of the triangles needed.
-Finally, the stripes can be stitched together to form a single parallelogram (beyond a certain number of lines, coalescing access does not improve as memory is anyway not stored contiguously).
+The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, hereby dramatically reducing the size of the additional triangles.
+Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe.
+Noticing that beyond a certain number of lines, coalescing access does not improve latency as memory is anyway not stored contiguously, we can make $B_H$ constant .
 
@@
 
-and the mapping of indices is given by
+and the mapping of indices is given by:
 \[ (i,j) \to (B_H \cdot \underset{\rm diagonal}{(j + m)} + m + \underset{\text{stripe}}{\left\lfloor \dfrac{i}{B_H} \right\rfloor} \cdot \underset{\text{stripe size}}{M_W \cdot B_H}) \quad \text{with } m=i \mod B_H \]
@@
 % ------------------------------------------------------------------------------------------------
-\newpage
-{\color{red} 
-\subsection{LMS compiler stack}
-\textbf{User-language}: define additional parameters for the recurrence\ul
-\item Windowing (to convert non-serial into serial problems)
-\item Input sizes, and alphabets (backtrack, input, cost)
-\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
-\item Recurrence functions, devices available
-\item What to keep in memory (cost, backtrack or both).
+\subsection{Compilation stack}
+Since the generated code sensibly differs from the Scala version, due to the reasons previously discussed, we cannot reuse LMS, although we borrow some of its ideas.
+Having the full control on the compilation stack also provides us the following benefits:\ul
+\item Since CUDA target and runtime compilation/execution is not supported in LMS, but in Delite \cite{delite}: an additional framework that runs on top of LMS.
+Although conceptually very similar, implementing our own stack reduces the number of dependencies, hence possible sources of misconfiguration errors for the final user of our framework.
+Additionally, since we do not share much of the functionality of Delite, an ad-hoc stack helps keeping the project featherweight.
+\item LMS can generate code for monadic functions that operate on arrays.
+However dynamic programming problems might require multiple inputs (for multi-track grammars) and special scheduling (to respect dependencies in the matrix), hence we need to issue specific C/CUDA code to handle problems correctly.
+\item LMS works on array of primitive types, possibly array of structures broken into array of simple types.
+Since in DP problems, composite types are representing a single logical element, and since we want to benefit from coalesced accesses (and possibly storing structures into efficient on-chip shared memory), we do not want to break structures.
+Also we want to offer support for tuples, which are a convenient way to write data containers (also we want to support case classes and composites types).
+\item Some information is only known at run-time (for instance input and matrix dimensions), hence we want to benefit from this knowledge as it helps computing the matrix indices more efficiently.
+Since such information could possibly be reused, we want to make the process as transparent as possible for the DSL user.
 \ule
-$\Downarrow$ Conversion (using an existing technique)
+Finally, we reap most of the LMS benefits in the generation of user-defined functions, as they are completely independent of the rest of the generated program thus could be generated without additional processing.
 
-\textbf{Intermediate representation}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 
-$\Downarrow$ Optimizations\ul
-\item Transform non-serial into serial \ul
-	\item Use aggregation functions/transformations
-	\item Use windowing from user (if no other technique succeed)
+Although our compilation stack might look quite similar to that of Delite, we do not share any component but LMS for the user functions. The compilation and execution process works as follows:\ul
+% XXX: Macros ?
+\item At compilation time:\ul
+	\item LMS generates the C code corresponding to the user-defined function and embeds it into the program
 	\ule
-\item Define the wavefront depth
-\item Avoiding the cost matrix by moving it into the wavefront
+\item At run-time, per problem description instance\ul
+	\item Recurrence analysis 
+	\item Code generation
+	\ule
+\item At run-time, per problem instance function call\ul
+	\item Adaptation of code to problem dimensions
+	\item Compilation of CUDA and C/JNI code as an external library (jnilib) then loaded
+	\item Compilation of a Scala wrapper to invoke the generated code
+	\item Execution of the compiled code:
+		\[ \text{Scala wrapper} \iff \text{JNI interface} \iff \text{CPU C code} \iff \text{CUDA kernel}\]
+	\ule
 \ule
 
-\textbf{Code specification}\ul
-\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
-\item Device-level interface => setup the block sizes(w/h), input and memory sizes
-\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
-\item Define the co-processor memory aggregation function
-\item Define the scheduling of the blocks and aggregation (software pipelining)
-\item Define the data movement back and forth to disk
-\ule
-
-$\Downarrow$ Generation\ul
-\item Generate the kernel for specific device
-\item Generate the scheduling and barriers
-\ule
-
-\textbf{Binary program}
-}
+%{\color{red} 
+%\textbf{User-language}: define additional parameters for the recurrence\ul
+%\item Windowing (to convert non-serial into serial problems)
+%\item Input sizes, and alphabets (backtrack, input, cost)
+%\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
+%\item Recurrence functions, devices available
+%\item What to keep in memory (cost, backtrack or both).
+%\ule
+%$\Downarrow$ Conversion (using an existing technique)
+%
+%\textbf{Intermediate representation}
+%
+%$\Downarrow$ Optimizations\ul
+%\item Transform non-serial into serial \ul
+%	\item Use aggregation functions/transformations
+%	\item Use windowing from user (if no other technique succeed)
+%	\ule
+%\item Define the wavefront depth
+%\item Avoiding the cost matrix by moving it into the wavefront
+%\ule
+%
+%\textbf{Code specification}\ul
+%\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
+%\item Device-level interface => setup the block sizes(w/h), input and memory sizes
+%\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
+%\item Define the co-processor memory aggregation function
+%\item Define the scheduling of the blocks and aggregation (software pipelining)
+%\item Define the data movement back and forth to disk
+%\ule
+%
+%$\Downarrow$ Generation\ul
+%\item Generate the kernel for specific device
+%\item Generate the scheduling and barriers
+%\ule
+%
+%\textbf{Binary program}
+%}
 %\ol

### 2013-01-03 docs/report/design.tex
More report.

@@
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+Although our compilation stack might look quite similar to that of Delite, we do not share any component but LMS for the user functions generation.
+The compilation and execution process works as follows:\ul
+\item Compilation: LMS generates the C code corresponding to the user-defined function and embeds them into the program bytecode (note that we present them separately for clarity).
+\item Runtime, for each parser (grammar+algebra): recurrences analysis is done in order to generate code (with placeholders for constants)
+\item Runtime, at every parse function call: the input size is known, hence replaced into the generic problem code, which is then processed by CUDA, C++ and Scala compilers.
+Then the JNI library resulting of the compilation is loaded and its corresponding Scala wrapper is invoked on the data to be processed.
+\ule
 
-Although our compilation stack might look quite similar to that of Delite, we do not share any component but LMS for the user functions. The compilation and execution process works as follows:\ul
-% XXX: Macros ?
-\item At compilation time:\ul
-	\item LMS generates the C code corresponding to the user-defined function and embeds it into the program
-	\ule
-\item At run-time, per problem description instance\ul
-	\item Recurrence analysis 
-	\item Code generation
-	\ule
-\item At run-time, per problem instance function call\ul
-	\item Adaptation of code to problem dimensions
-	\item Compilation of CUDA and C/JNI code as an external library (jnilib) then loaded
-	\item Compilation of a Scala wrapper to invoke the generated code
-	\item Execution of the compiled code:
-		\[ \text{Scala wrapper} \iff \text{JNI interface} \iff \text{CPU C code} \iff \text{CUDA kernel}\]
-	\ule
-\ule
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(15,19)
+\put(0,17){\tbox{5}{1}{DSL libraries}}\put(6,17){\tbox{5}{1}{User program}}
+{\linethickness{1.5pt}\moveto(8,17)\lineto(8,16.25)\lineto(2.5,16.25)\strokepath\put(2.5,17){\vector(0,-1){2}}}\put(-1.5,16){\small Scala compiler}
+\put(8.5,17){\linethickness{1.5pt}\vector(0,-1){2}}\put(8.75,16){\small LMS}
+% -----
+\put(0,14){\tbox{5}{1}{JVM bytecode}}\put(6,14){\tbox{5}{1}{C functions}}
+\put(2.5,14){\vector(0,-1){2.75}}\put(8.5,14){\vector(0,-1){1.75}}
+\multiput(-2,13)(.15,0){130}{\circle*{.05}}\put(14,13.25){\small compilation}\put(14,12.5){\small run-time}
+% -----
+{\linethickness{1.5pt}\moveto(2.5,7)\lineto(2.5,11)\lineto(5,11)\strokepath\put(5,11){\vector(1,0){1}}}
+\put(2.75,9){\small recurrences analysis}\put(2.75,8.4){\small code generation}
+% -----
+\put(6,10){\tbox{5}{2}{C/CUDA/JNI code}}
+\put(8.5,10){\linethickness{1.5pt}\vector(0,-1){3}}\put(8.75,9){\small constants evaluation}\put(8.75,8.4){\small compilation}
+{\linethickness{1.5pt}\moveto(12,1.5)\lineto(12,8)\lineto(10,8)\strokepath\put(10,8){\vector(-1,0){1.5}}}\put(12.25,6.5){\rotatebox{-90}{\small input size}}
+% -----
+\put(0,2.5){\tbox{5}{4.5}{Scala parsers}}
+\put(6,2.5){\put(0,0){\tbox{5}{4.5}{}}
+	\put(.25,3.25){\tbox{4.5}{1}{CUDA kernels}}\put(.25,1.75){\tbox{4.5}{1}{JNI / host C}}\put(.25,.25){\tbox{4.5}{1}{Scala wrapper}}
+	\put(3.5,3.25){\vector(0,-1){.5}}\put(3.5,1.75){\vector(0,-1){.5}}\put(1.5,1.25){\vector(0,1){.5}}\put(1.5,2.75){\vector(0,1){.5}}}
+\put(13,1){\tbox[l0]{4}{1}{Input data}}
+\put(13,0){\tbox[l0]{4}{1}{Output data}}
+\put(1.5,1.5){\vector(0,1){1}}\put(7.5,1.5){\vector(0,1){1}}\moveto(1.5,2)\lineto(1.5,1.5)\lineto(13,1.5)\strokepath % input
+\put(9.5,.5){\line(0,1){2}}\moveto(3.5,2.5)\lineto(3.5,.5)\lineto(12,.5)\strokepath\put(12,.5){\vector(1,0){1}} % output
+
+\end{picture}\end{center}\caption{Compilation and execution scheme of a parser}\end{figure}
 

### 2013-01-03 docs/report/implementation.tex
More report.

@@
 \section{Implementation}
-
 % ------------------------------------------------------------------------------------------------
 \subsection{CUDA ad-hoc}
+In the project planning, an ad-hoc implementation phase immediately followed the problem analysis. The goals of this phase is threefold:\ol
+\item Better understand the challenges in CUDA implementation of dynamic programming problems and get on par with state-of-art implementations.
+\item Have a baseline implementation that is independent of the hardware and that could be benchmarked.
+We also tried to contact the authors of \cite{swat_mega} and \cite{gpu_atlp} to obtain their implementation.
+The former provided us with their implementation, which turned out to address large serial problems whereas our focus was on smaller non-serial problems, the latter did not respond to our solicitations.
+\item Have an optimal implementation that can serve as target to be imitated and generalized by the code generation.
+\ole
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+The project development was scheduled in \ul
+\item Problems analysis
+\item Ad-hoc implementation (CUDA)
+\item Porting ADP parsers
+\item Improving parsers / adding functionalities
+\item Integrating CUDA generation into parsers
+\ule
+
 During the first phase of project, and after the decision to focus on non-serial problems implementation on GPU, we developed an ad-hoc version of 3 non-serial problems.

### 2013-01-03 docs/report/implementation.tex
Nits in report.

@@
 \subsection{CUDA ad-hoc}
-In the project planning, an ad-hoc implementation phase immediately followed the problem analysis. The goals of this phase is threefold:\ol
+In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
 \item Better understand the challenges in CUDA implementation of dynamic programming problems and get on par with state-of-art implementations.
@@
 
+Leveraging the insights provided by \cite{gpu_atlp} and \cite{gpu_barrier}, we started with a basic implementation (where each CUDA thread processes one matrix line) with three additional optimizations:\ul
+\item Memory accesses must be coalesced (memory accesses account for a significant part of the total running time, according to both manufacturer documentation and experiments \cite{perfeval})
+\item Synchronization between threads can be done according to \cite{gpu_barrier}, additionally, we can slightly loosen the synchronization restrictions, as the paper describes a thread barrier whereas we only require a condition on previous thread progress (except for the parallelogram case, where we still require a barrier).
+\item Computation progresses element-wise along the diagonal (maximizes the parallelism level)
+\ule
+
+\subsubsection{Related work}
+Since \cite{swat_mega} focuses on a different class of problem, we compare our implementation against \cite{gpu_atlp}, which provides an efficient matrix multiplication implementation.
+However, since we have neither the source code (or binary) nor the same evaluation hardware, we need to normalize the results.
+To do that, we present hardware differences and their result:
+
+\def\unt#1{& \footnotesize #1}
+\begin{table}[H]\begin{center}\begin{tabular}{lrrr} \toprule
+\bf Graphic card &				& \bf Our 		& \bf  ATLP\cite{gpu_atlp} \\ \midrule
+Model &						& GeForce GT 650M	& Tesla C1060 \\
+Architecture, capability  &			& Kepler (3.0)	& GT200 (1.3) \\
+Memory \unt{Mb}				& 1024		& 4096 \\
+CUDA cores &					& 384		& 240 \\
+Clock (core, memory) \unt{MHz}	& 756, 1953	& 1300, 1600 \\
+Memory bus \unt{bit}				& 128		& 512 \\
+Memory bandwidth \unt{GB/s}		& 28.8		& 102.4 \\
+Processing power \unt{GFLOPS}	&564.5		& 622.08 \\ \midrule
+\bf Processing speedup & 		& 1			& 1.07 \\
+\bf Memory speedup & 			& 1			& 3.55 \\ \bottomrule
+\end{tabular}\end{center}\caption{Graphic cards technical specifications (source: \href{http://en.wikipedia.org/wiki/Comparison_of_Nvidia_graphics_processing_units}{Wikipedia})}\end{table}
+
+\begin{table}[H]\begin{center}\begin{tabular}{lrrrrrrrrrr} \toprule
+\bf Matrix size & 128 & 256 & 512 & 1024 & 1536 & 2048 & 2560 & 3072 & 3584 & 4096 \\ \midrule
+\bf No split & 0.07 & 0.09 & 0.19 & 0.59 & 1.27 & 2.25 & 3.51 & 5.07 & 6.92 & 9.06 \\
+\bf Split at 1 & 0.06 & 0.07 & 0.08 & 0.14 & 0.26 & 0.47 & 0.77 & 1.21 & 1.80 & 2.57 \\ \bottomrule
+\end{tabular}\end{center}\caption{ATLP\cite{gpu_atlp} results summary: matrix multiplication timing in seconds}\end{table}
+
+\subsubsection{Results}
+We here present the timings of our implementation.
+For correctness, we first implemented a CPU version that we used to compare CUDA results against.
+Input is made of random numbers.
+The implemented dynamic programming problems are:\ul
+\item Rectangle: Smith-Waterman with arbitrary cost
+\item Triangle: matrix chain multiplication
+\item Parallelogram: polygon triangulation using a matrix larger than necessary.
+Note that this implementation uses at most 32 blocks to prevent dead locks on our hardware (restriction due to the number of concurrent threads on the device).
+\ule
+
+\begin{table}[H]
+\begin{center}\begin{tabular}{cclrrr} \toprule
+\bf Matrix & \bf Block & \bf Comment & \bf R\footnotesize (s) & \bf T\footnotesize (s) & \bf P\footnotesize (s) \\ \midrule
+1024 & 1 & CPU					& 1.965		& 1.191		& 6.069 \\
+2048 & 1 & CPU					& 27.229		& 15.296		& 57.323 \\
+4096 & 1 & CPU					& 			& 177.608	&  \\
+1024 & 32 & GPU baseline			& 0.838		& 0.500		& 0.516 \\
+1024 & 32 & GPU sync improved		& 0.642		& 0.316		& 0.343 \\
+2048 & 32 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
+4096 & 32 & GPU 8 splits				& 21.902		& 8.841		& 16.767 \\
+8192 & 32 & GPU 64 splits			& 159.058	& 62.064		& 135.793 \\
+12288 & 32 & GPU 256 splits			& 419.030	& 196.971	& 460.912 \\
+\bottomrule \end{tabular}\end{center}
+\caption{Best timings for R=rectangle, T=triangle, P=parallelogram}
+\end{table}
+
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+\subsubsection{Discussion}
+Problem are described by the triplet <matrix size/block size/shape>.\ul
+\item It has been put in evidence in a previous work\footnote{Performance Evaluation 2012 miniproject, \it Performance Evaluation of Mersenne arithmetic on GPUs} that with Mac OS operating system, using the GPU exclusively for CUDA or combining with UI display may affect the performance (GeForce 330M, architecture 1.2).
+With the new architecture (3.0, GeForce 650M), this difference has been reduced to less than 3.5\% with the decoupling of UI and CUDA performing best.
+So in micro-benchmarks, we can safely ignore the graphic card usage.
+
+\item Synchronization  between blocks\ul
+	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}.
+Speedup for <1024/32/*> are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
+	\item In the parallelogram case, using all threads to monitor other blocks status instead of 1 thread results in a 6.4x speedup (22.72$\to$3.52ms) on <1024/32/para>.
+	\ule
+\item Using multiple threads per matrix cell:
+In the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increase (there is one more dependency).
+According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity.
+However, in our setup (memory layout+algorithm+hardware), we did not found any improvement by doing so.
+We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
+\item
+
+First we need to prove that we use almost all the available memory bandwidth: for matrix multiplication, in a triangular matrix, we have
+\[\text{Total transfer}=\frac{n(n+1)}{2} \text{ writes} + \sum_{i=0}^{n-1} 2 i \cdot (n-i) \text{ reads}\]
+where each write is 10 bytes (long+short), and each read is 8 bytes (long). For $n=4096$ we transfer
+% n*(n + 1)/2*12 + Sum[2*i*(n - i), {i, 0, n - 1}]*8
+183'352'614'912 bytes which corresponds to 183.35GB.
+In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254 \rm GB$.
+Hence we can say that 72\% of the algorithm running time is spent into memory accesses.
+
+\item On a 4096 matrix, if we assume that ATLP card would have the same bandwidth as our card, their running time would be
+\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 9.43\rm s_{\text{ ATLP}} > 8.84\rm s_{\text{ our}}\]
+Which shows that our algorithm is comparable to theirs.
+However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have ATLP source code).
+\ule
+As a conclusion, (1) we must remain away to invalidate their result as previous hardware generations might be subject to more constraint to our hardware and (2) we are on par, if not better with one of the best current implementation.
+
+\item Reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
+Even worse, it slows down slightly the computation.
+We might attribute this to a better constant transformation by the compiler.
+Hence, having many idle threads does not impede performance.
+
+\item Unrolling the inner loops (non-serial dependencies) a small number of time provide some speedup, for a 2048-matrix respectively 10.9\% (rectangle, $2.765\to 2.464$), 14.1\% (triangle, $1.427\to 1.225$) and 9.7\% (parallelogram $1.539\to 1.389$).
+
+\ule
+
+Future work: multi-splits but proved not relevant
 

### 2013-01-03 docs/report/implementation.tex
More report.

@@
 \item Computation progresses element-wise along the diagonal (maximizes the parallelism level)
+\item Thread block size = warp size (32) to benefit from implicit synchronization within warps
 \ule
 
+% ----------------------------------------------
 \subsubsection{Related work}
@@
 \bf Split at 1 & 0.06 & 0.07 & 0.08 & 0.14 & 0.26 & 0.47 & 0.77 & 1.21 & 1.80 & 2.57 \\ \bottomrule
-\end{tabular}\end{center}\caption{ATLP\cite{gpu_atlp} results summary: matrix multiplication timing in seconds}\end{table}
+\end{tabular}\end{center}\caption{ATLP\cite{gpu_atlp} results: matrix chain multiplication, execution time (in seconds)}\end{table}
 
+% ----------------------------------------------
 \subsubsection{Results}
-We here present the timings of our implementation.
+We present here the timings of our implementation.
 For correctness, we first implemented a CPU version that we used to compare CUDA results against.
-Input is made of random numbers.
+Input data is made of random numbers.
 The implemented dynamic programming problems are:\ul
@@
 \begin{table}[H]
-\begin{center}\begin{tabular}{cclrrr} \toprule
-\bf Matrix & \bf Block & \bf Comment & \bf R\footnotesize (s) & \bf T\footnotesize (s) & \bf P\footnotesize (s) \\ \midrule
-1024 & 1 & CPU					& 1.965		& 1.191		& 6.069 \\
-2048 & 1 & CPU					& 27.229		& 15.296		& 57.323 \\
-4096 & 1 & CPU					& 			& 177.608	&  \\
-1024 & 32 & GPU baseline			& 0.838		& 0.500		& 0.516 \\
-1024 & 32 & GPU sync improved		& 0.642		& 0.316		& 0.343 \\
-2048 & 32 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
-4096 & 32 & GPU 8 splits				& 21.902		& 8.841		& 16.767 \\
-8192 & 32 & GPU 64 splits			& 159.058	& 62.064		& 135.793 \\
-12288 & 32 & GPU 256 splits			& 419.030	& 196.971	& 460.912 \\
+\begin{center}\begin{tabular}{rlrrr} \toprule
+\bf Matrix size & \bf Comment & \bf R & \bf T & \bf P \\ \midrule
+1024 & CPU					& 1.965		& 1.191		& 6.069 \\
+2048 & CPU					& 27.229		& 15.296		& 57.323 \\
+4096 & CPU					& 			& 177.608	&  \\
+1024 & GPU baseline			& 0.838		& 0.500		& 0.516 \\
+1024 & GPU sync improved		& 0.642		& 0.316		& 0.343 \\
+2048 & GPU P $\le32$ blocks		& 2.864		& 1.427		& 2.096 \\
+4096 & GPU 8 splits				& 21.902		& 8.841		& 16.767 \\
+8192 & GPU 64 splits			& 159.058	& 62.064		& 135.793 \\
+12288 & GPU 256 splits			& 419.030	& 196.971	& 460.912 \\
 \bottomrule \end{tabular}\end{center}
-\caption{Best timings for R=rectangle, T=triangle, P=parallelogram}
+\caption{Execution time (in seconds) for R=rectangle, T=triangle, P=parallelogram}
 \end{table}
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-\subsubsection{Discussion}
-Problem are described by the triplet <matrix size/block size/shape>.\ul
-\item It has been put in evidence in a previous work\footnote{Performance Evaluation 2012 miniproject, \it Performance Evaluation of Mersenne arithmetic on GPUs} that with Mac OS operating system, using the GPU exclusively for CUDA or combining with UI display may affect the performance (GeForce 330M, architecture 1.2).
-With the new architecture (3.0, GeForce 650M), this difference has been reduced to less than 3.5\% with the decoupling of UI and CUDA performing best.
-So in micro-benchmarks, we can safely ignore the graphic card usage.
-
-\item Synchronization  between blocks\ul
+% ----------------------------------------------
+\subsubsection{Results discussion}\ul
+\item \textbf{User interface:} It has been put in evidence in \cite{perfeval} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M).
+With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best.
+So we can safely ignore this issue.
+\item \textbf{Blocks synchronization:}\ul
 	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}.
-Speedup for <1024/32/*> are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
-	\item In the parallelogram case, using all threads to monitor other blocks status instead of 1 thread results in a 6.4x speedup (22.72$\to$3.52ms) on <1024/32/para>.
+Speedup for matrix size of 1024 are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
+	\item In the parallelogram case, using all threads to monitor other blocks status instead of the first one only results in a 6.4x speedup (22.72$\to$3.52ms) for the parallelogram.
 	\ule
-\item Using multiple threads per matrix cell:
-In the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increase (there is one more dependency).
+\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increase (there is one more dependency).
 According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity.
@@
 In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254 \rm GB$.
-Hence we can say that 72\% of the algorithm running time is spent into memory accesses.
+Hence  72\% of the algorithm running time is spent into memory accesses.
 
@@
 
-\item Reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
+\item \textbf{Threads number:} reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
 Even worse, it slows down slightly the computation.
@@
 
-\item Unrolling the inner loops (non-serial dependencies) a small number of time provide some speedup, for a 2048-matrix respectively 10.9\% (rectangle, $2.765\to 2.464$), 14.1\% (triangle, $1.427\to 1.225$) and 9.7\% (parallelogram $1.539\to 1.389$).
-
+\item \textbf{Unrolling:} unrolling the inner loops (non-serial dependencies) a small number of time provide some speedup, for a 2048-matrix respectively 10.9\% (rectangle, $2.765\to 2.464$), 14.1\% (triangle, $1.427\to 1.225$) and 9.7\% (parallelogram $1.539\to 1.389$).
+The best experimental number of unrolling is 5.
 \ule
 
-Future work: multi-splits but proved not relevant
-
-The project development was scheduled in \ul
-\item Problems analysis
-\item Ad-hoc implementation (CUDA)
-\item Porting ADP parsers
-\item Improving parsers / adding functionalities
-\item Integrating CUDA generation into parsers
+% ------------------------------------------------------------------------------------------------
+\subsection{Scala parsers}
+The Scala parsers consist in 4 traits that are used to construct a DSL program:\ul
+\item \textbf{Signature:} abstraction to define input and output types, and the aggregation function. The signature is implemented by all other traits (in particular algebras and grammars).
+\item \textbf{BaseParsers:} serves as basis for the two other traits and defines common features.
+It implements the {\tt Parser} abstraction and all its inheriting classes: {\tt Tabulate}, (abstract) {\tt Terminal}, {\tt Aggregate}, {\tt Filter}, {\tt Map}, {\tt Or}, {\tt Concat}.
+Terminals are further specialized in the two other traits (ADPParsers and TTParsers).
+The parser abstraction specifies 3 methods:\ul
+	\item {\tt apply(subword)} computes the parser result; it is used to obtain the corresponding results.
+	\item {\tt unapply(subword,backtrack)} computes the previous step of the backtrack by returning subsequences at the origin of the result; it is invoked recursively to obtain the full backtrack trace.
+	\item {\tt reapply(subword,backtrack)} is very similar to apply, except that it  computes only the results matching the backtrack.
+It is used to construct the result corresponding to a backtrack trace (possibly in a different domain, pretty printing, ...).
+	\ule
+	Additionally, the BaseParser implements the analysis that are shared by both the Scala and the CUDA version: dead rules elimination, yield analysis and dependencies ordering.
+Finally, it provides some implicit functions to flatten nested tuples (that are constructed by multiple concatenations).
+\item \textbf{ADPParsers:} used as basis for a single track DP grammar (using one input sequence).
+It defines the concatenation operator $\sim$ ({\tt Concat} wrapper), and the terminals (empty, element and sequence).
+Additionally, it defines the interface functions {\tt parse(input)}, {\tt backtrack(input)} and {\tt build(in,backtrack)} that respectively compute the result, the backtrack and the result corresponding to a trace.
+\item \textbf{TTParsers:} used to define two-track DP grammar (using a pair of sequences as input).
+Similarly, this class defines concatenations $-\!\!\sim$ and $\sim\!\!-$, terminals (for each track) and the {\tt parse(in1,in2)}, {\tt backtrack(in1,in2)} and {\tt build(in1,in2,backtrack)} functions.
 \ule
 
-During the first phase of project, and after the decision to focus on non-serial problems implementation on GPU, we developed an ad-hoc version of 3 non-serial problems.
-Since we erroneously complicated the matrix of the polygon triangulation problem, we devised 3 different matrix shape, that is upper triangular for single-track problems, rectangular for two-track problems and parallelogram for the polygon triangulation.
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 
-\subsection{LibRNA (?)}
+XXX: make a class diagram ? (classes hierarchies)
 
-\begin{verbatim}
-1 week on hash maps
-2 weeks to define and analyze the problems (also with FPGA in mind)
-3 weeks restriction to non-serial and ad-hoc implementation on GPU (rectangle, triangle, parallelogram); translation of ADP parsers in Scala (Manohar)
-1 week run-time engine for Scala/JNI/CUDA
-05.11 - rework of the ADP parser to aim at generating C-like code
-12.11 - rework of ADP parsers to extend usages (cyclic, two-track) and aim at automatic backtracking
-19.11 - explorations in LMS / macros
-26.11 - full backtracking: apply/unapply/reapply, rework of the classes
-03.12 - Zuker/JNI
-10.12 - Yield analysis, code generation
-17.12 - code generation: detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support
-24.12 - code generation, sick
-31.12 - report
-\end{verbatim}
+XXX: explain thoroughly all the functionalities and dig into the code ?
 
-\subsection{ADP parsers}
 \subsection{Transformations and simplifications}
@@
 
+\subsection{LibRNA (?)}
+The project development was scheduled in \ul
+\item Problems analysis
+\item Ad-hoc implementation (CUDA)
+\item Porting ADP parsers
+\item Improving parsers / adding functionalities
+\item Integrating CUDA generation into parsers
+\ule
+
+\begin{verbatim}
+1 week on hash maps
+2 weeks to define and analyze the problems (also with FPGA in mind)
+3 weeks restriction to non-serial and ad-hoc implementation on GPU (rectangle, triangle, parallelogram); translation of ADP parsers in Scala (Manohar)
+1 week run-time engine for Scala/JNI/CUDA
+05.11 - rework of the ADP parser to aim at generating C-like code
+12.11 - rework of ADP parsers to extend usages (cyclic, two-track) and aim at automatic backtracking
+19.11 - explorations in LMS / macros
+26.11 - full backtracking: apply/unapply/reapply, rework of the classes
+03.12 - Zuker/JNI
+10.12 - Yield analysis, code generation
+17.12 - code generation: detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support
+24.12 - code generation, sick
+31.12 - report
+\end{verbatim}
+
 \subsection{LMS integration}

### 2013-01-03 docs/report/report.tex
Renamed roadmap into report.
Nits in report.
More report.

--- 2012-12-30 docs/roadmap/draft.tex
@@
 \input{implementation}
-\input{benchmarks}
+\section{Benchmarks}
+XXX
+
 \section{Future work} \ul
@@
 % ------------------------------------------------------------------------------------------------
-\newpage
-\section{Planning --- Work in progress}
-\subsection{Write introduction}
+\section*{Planning --- Work in progress}
+\textbf{Paper introduction:} 
 \begin{verbatim}
-what exists
-problem
-how we solve it compared to other
-how to evaluate
+problem to solve, what exists (related work), how we solve it compared to other, how to evaluate
 contributions (3): 3 tensed sentences
-
-related work
 benchmark => prove by evaluation intro statements
@@
 	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
-	\item Rules normalization
-	\item Rules dependency
-	\item Automatic backtracking
 	\ule
@@
 Core function F:
-- in: s,t strings
-- in: neighbor costs: top, left, top+left
-- in: neighbor stats: top, left, top+left
-- out: backtrack information
-- out: cost(i,j)
-- out: stats
+- in: s,t strings, neighbor costs: top, left, top+left, neighbor stats: top, left, top+left
+- out: backtrack information, cost(i,j), stats
 
@@
 
-          s-->
-  +----------------/------------+
-t | B1            /             |
-| +------/-------/              | KERNEL1 (keep benefit of t in shared mem)
-| | B2  /                       | delay between b1 and b2 can be 1% of line length
-v +----/------------------------+
-  |                             |
-  |                             | KERNEL2
-Memory organized as
-b1 -- --
-b1 b1 --
-b1 b1 b1
-b1 b1 b1
-.. .. ..
-b1 b1 b1
-b2 b1 b1
-b2 b2 b1
-b2 b2 b2
-.. .. ..
-
 For stats (aggregation of the table): we need to maintain both an horizontal and vertical front

### 2013-01-03 src/main/scala/v4/BaseParsers.scala

@@ -7,6 +7,7 @@ trait Signature {
   val window = 0 // windowing size, 0=disabled
   val h:List[Answer]=>List[Answer] = l=>l // optimization function
 
+  // Predefined CUDA-compatible aggregation functions
   def max[T:Numeric]: List[T]=>List[T]
   def min[T:Numeric]: List[T]=>List[T]
   def sum[T:Numeric]: List[T]=>List[T]

### 2013-01-04 docs/report/implementation.tex
More report.

@@
 % ------------------------------------------------------------------------------------------------
-\subsection{CUDA ad-hoc}
+\subsection{Ad-hoc CUDA}
 In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
@@
 The Scala parsers consist in 4 traits that are used to construct a DSL program:\ul
-\item \textbf{Signature:} abstraction to define input and output types, and the aggregation function. The signature is implemented by all other traits (in particular algebras and grammars).
+\item \textbf{Signature:} abstraction to define input ({\tt Alphabet}) and output ({\tt Answer}) types, and the aggregation function.
+The signature is implemented by all other traits (in particular algebras and grammars).
 \item \textbf{BaseParsers:} serves as basis for the two other traits and defines common features.
@@
 It is used to construct the result corresponding to a backtrack trace (possibly in a different domain, pretty printing, ...).
+	\ule
+	To support analysis, the parsers carry additional values:\ul
+	\item Minimum and maximum yield size: functions evaluated recursively except for tabulations where value is attributed in the yield analysis phase.
+	\item Number of inner alternatives: helps counting alternatives, hereby guaranteeing an unique number for each (provided that parsers obtain non-overlapping ranges).
+	\item Number of inner moving concatenations: helps determining required storage for the backtrack as well as retrieving the appropriate index in the backtrack phase
 	\ule
@@
 
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(14,11)
+\put(3,8){\tbox{8}{2.5}{{\bf Signature} \footnotesize\\ Types: Alphabet, Answer \\ $h$ (aggregation function)}}
+\put(0,4){\tbox{14}{2.5}{{\bf BaseParsers} \footnotesize\\ Tabulate, Terminal, Aggregate, Filter, Map, Or, Concat \\ Analysis: dead rules, yield analysis, dependencies}}
+\put(0,0){\tbox{6}{2.5}{{\bf ADPParsers} \footnotesize\\ $\sim$, $\sim(a,b,c,d)\sim$ \\ Single track terminals}}
+\put(8,0){\tbox{6}{2.5}{{\bf TTParsers} \footnotesize\\ $-\!\!\sim$, $\sim\!\!-$ \\ Two-tracks terminals}}
+{\linethickness{1.5pt}\put(3,2.5){\vector(1,1){1.5}}\put(11,2.5){\vector(-1,1){1.5}}\put(7,6.5){\vector(0,1){1.5}}}
+\end{picture}\end{center}\caption{Class diagram (simplified)}\end{figure}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\subsection{Code generation}
+
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
-
-XXX: make a class diagram ? (classes hierarchies)
-
-XXX: explain thoroughly all the functionalities and dig into the code ?
-
-\subsection{Transformations and simplifications}
 
@@
 \end{verbatim}
+
+\subsection{JNI wrappers and runtime execution engine}
 
@@
 Where and how is LMS useful
-
-\subsection{Code generation}
-
-\subsection{Runtime execution engine}
-
-\subsection{CUDA implementation}

### 2013-01-04 docs/report/report.tex
More report.

@@
 Additionally, domain experts might not have the skills to make an efficient parallel implementation.
-In this project, we leverage the Scala and LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs.
+In this project, we present \textit{DynaProg}, a Scala DSL for dynamic programming on heterogeneous platforms which allow to write concise programs and execute them efficiently on GPUs.
+
+Existing work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
+% XXX: How we compare to other, how to evaluate
+% Benchmark => prove by evaluation intro statements
+
 Our contributions are: \ul
-\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
+%\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
 \item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
@@
 \ule
+
 \vfill
@@
 \section{Benchmarks}
-XXX
+{\color{red} XXX}
 
@@
 
-\section{Conclusion}
-
-% ------------------------------------------------------------------------------------------------
-\section*{Planning --- Work in progress}
-\textbf{Paper introduction:} 
 \begin{verbatim}
-problem to solve, what exists (related work), how we solve it compared to other, how to evaluate
-contributions (3): 3 tensed sentences
-benchmark => prove by evaluation intro statements
-\end{verbatim}
-
-\subsubsection*{Steps}\ol
-\item Re-implement vanilla combinators with lists
-\item Add following problems and features:\ul
-	\item Polygon triangulation (cyclic problem) (parallelogram matrix)
-	\item SWat with arbitrary gap (multi-track grammar) (rectangular matrix)
-	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
-	\ule
-\item Test/proof parsers are correct --- make sure implementation is correct
-\item Automate test to compare against implementation
-\item Build bottom-up vanilla version\ul
-	\item Separate initialization (terminals) and non-terminals (speed up?)
-	\item In parallel implement the code generator for CUDA, hardcode in text function bodies for CUDA
-	\ule
-\item LMS implementation: from normal to Rep/Exp\ul
-	\item Re-implement bottom-up vanilla into LMS
-	\item Re-implement CUDA version and use functions body to infer CUDA code
-	\ule
-\item Benchmarks --- compare also versus other papers.
-\item Write report
-\ole
-
-\subsubsection*{Roadmap}
-\begin{tabular}{rl}
-16.11 & Rules normalization and automatic backtracking as in 5.2 \\
-	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
-23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
-30.11 &--- Gap due to LMS missing knowledge \\
-7.12 & Benchmarking, grammar analysis \\
-14.12 & First though for larger than mem \\
-21.12 & Writing report \\
-28.12 & --- holiday --- \\
- 4.01 & --- holiday --- \\
-11.01 & Writing report: implementation description and plan for future work \\
-18.01 & Writing report
-\end{tabular}
-
-\subsubsection*{Todo @TCK}
-\subsubsection*{Todo @Manohar}
-\begin{verbatim}
-Look at string templating. """xxx $var xxx"""
-- Cleanup LMS CUDA generator
-
-Separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
-
-XXX: how to encode multi-dimensional matrices
+How to Encode multi-dimensional matrices efficiently
 1. assume they have the same type put one after another => different dimensions ok
@@
 => but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)
-\end{verbatim}
 
-\subsection*{OldPlan} \ol
-\item \textbf{User facing language:} similar to \cite{adp_gpu} or \cite{adp_fusion} \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}.
-We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
-\item \textbf{Prototyping:} understand difficulties and share common base.
-Prototype of \nameref{aswat} on CUDA/FPGA.
-Give an idea of how to implement the general case.
-Benchmark and compare both implementations.
-Aim to do as good an implementation for the specific platform (CPU/GPA) as possible.
-\item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
-\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
-\ole
+Possible extensions / optimizations:
+- making serializable by using aggregation functions/transformations
+- avoiding the cost matrix by moving data in the wavefront
+- Scheduling for large problems (with memory loads for non-serializable)
+- add FPGA target platform
+- pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char)
+- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
 
-\begin{verbatim}
-1. optimize: terminals of bounded yield + binary decomposition if possible
-2. transform into: axioms (init/fixed) + rules (iterations)
-- define optimizations to IR
-  - making serializable by using aggregation functions/transformations
-  - avoiding the cost matrix by moving it into the wavefront
-- scheduling (with memory loads for non-serializable)
-- code generation for both platforms
-  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)
-
-----------------------------------
 Core function F:
@@
 - out: backtrack information, cost(i,j), stats
-
-we might want to pack into memory the matrix data and make threads operate on more than one cell.
-
-Some ideas:
-- define a way to pack the characters => less memory transfer (i.e. GATC=>4 letters in 1 char)
-- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
-- write a kernel that takes stats from left and r
-
         stats (x)
@@
      new_stats (x')
-
-and compute the backtracking for its cell (optional as template boolean)
-
-- multi-grain
-  + within the 64b-words : multiple letters at once (thread level)
-  + group threads into blocks that operate on (block level)
-  + kernel that operates one after another (GPU/CPU level)
-    => can we catch a stream's event at CPU level to alloc memory to get back results ?
-- keep track of
-  + stats O(m+n)
-  + backtrack O(m*n)
-
-- are we sure we need squares ?
-  - rectangles can increase the full usage of all threads during (x-y) runs
-  - assuming we put longest word vertically we play hot potato for stats
-    left->right => block.x exchanges within block, last thread writes to
-    global exchange memory (if short, so that it does not penalize running time)
-  - we may go up to running in stripes
-- if we put in diagonal-major data in the memory, why not stream all chunks like
-  that using a cyclic buffer (?) => no empty slot
-  ==> this might be useful for the backtrack information
-
-- use extensively profiling : CUDA profiler
-
-- Robust DP paper provides 2 insights: coalesced access + GPU synchronization
-  => can we do a producer-consumer scenario at the block border so that we can
-     execute a large diagonal horizontal/vertical swipe made of multiple blocks?
-  => multiple swipes at different delays (so that we avoid syncing at every step
 
@@
 cell information (usually limited range) is much more compact than the score(32-64 bits)
+\end{verbatim}
 
-----------------------------------------------------------------------------
-Some GPU algorithms: http://hgpu.org/?cat=11
-Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
-CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
-13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
-http://tutorials.jenkov.com/java-reflection/fields.html
-http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
+\section{Conclusion}
+{\color{red} XXX}
 
-XXX: use TypeClass to put a predicate on types
---------------------------------------------
-def fun[T: CanTranslateToC](...)
-def fun[T](implicit ev:CanTranslateToC[T])
+% ------------------------------------------------------------------------------------------------
+\section*{Planning}
+\textbf{Paper introduction}\ul
+\item Problem to solve, what exists (related work), how we compare to other, how to evaluate
+\item Contributions (3): 3 tensed sentences
+\item benchmark => prove by evaluation intro statements
+\ule
 
-class CanTranslateToC[T] { def translate:String }
-implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
-\end{verbatim}
+\subsubsection*{Roadmap}
+\begin{tabular}{rl}
+16.11 & Rules normalization and automatic backtracking \\
+	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
+23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
+30.11 &--- Gap due to LMS missing knowledge \\
+7.12 & Benchmarking, grammar analysis \\
+14.12 & First thoughts for larger than device memory \\
+21.12 & Writing report \\
+28.12 & --- holiday --- \\
+ 4.01 & --- holiday --- \\
+11.01 & Writing report: implementation description and plan for future work \\
+18.01 & Writing report
+\end{tabular}
+
+\subsubsection*{Todo @TCK}\ul
+\item Test/proof parsers are correct --- make sure implementation is correct
+\item Automate test to compare against implementation
+\item Benchmarks -- use extensively CUDA profiler
+\item Write report
+\item Port LibRNA for CUDA?
+\ule
+
+\subsubsection*{Todo @Manohar}\ul
+\item Integrate LMS code generation into v4.
+\item Fix Zuker coefficients
+\ule
+
+%Some GPU algorithms: http://hgpu.org/?cat=11
+%Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
+%CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
+%13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
+%http://tutorials.jenkov.com/java-reflection/fields.html
+%http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
+%
+%Hint: use TypeClass to put a predicate on types
+%  def fun[T: CanTranslateToC](...)
+%  def fun[T](implicit ev:CanTranslateToC[T])
+%  class CanTranslateToC[T] { def translate:String }
+%  implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
 

### 2013-01-04 docs/report/implementation.tex
NIts.

@@
 \subsection{Code generation}
+The code generation step produces multiple outputs that are tightly bound to each other. Besides the Scala wrapper (a simple JNI interface), in the C/CUDA code generated we distinguish:\ol
+\item JNI input and output conversion functions
+\item Host memory management and scheduling of CUDA kernels
+\item CUDA matrix computation, which can be further decomposed into matrix scheduling (loops) and (matrix cell) computation. 
+\item CUDA backtrack collection kernel
+\ole
 
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+\subsubsection{Scala structures conversion}
+
+\subsubsection{Host wrappers}
+
+\subsubsection{Matrix computation scheduling}
+Similarly as in the ad-hoc implementation, progress is made along the diagonal (see \ref{mem_layout}) and each thread is responsible of one line.
+That is, the matrix is swept horizontally by a <<diagonal of threads>>, that are enabled only if they are within a valid matrix cell.
+
+Special care must be taken to handle computation dependencies: within a warp, all threads are executed at the same time, hence no synchronization is necessary, otherwise, thread may execute in arbitrary order.
+To enforce synchronization, we first set the block size being equal to the warp size to benefit of the implicit synchronization, it remains to provide inter-block synchronization.
+
+First notice that the DP matrix element has dependencies: along its line, column and possibly intermediate elements in diagonal.
+By induction, it suffice to have the last column and row element valid (this also guarantees the validity of all intermediate elements).
+Since line is computed by the current thread (and is valid), it only remains to guarantee that the element of the previous line is valid.
+
+Since the line is computed by the same thread, data validity is guaranteed.
+Noticing that column dependency is along one direction and that diagonal dependency is valid if line and column are valid, it suffice to have a valid column element.
+
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(6,6)
+	\def\Cfl2#1{#1{0,4}#1{0,3}#1{1,3}#1{0,2}#1{1,2}#1{2,2}#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}}
+	\Cfl2{\Cg}
+	\Cd[0,1]{4,1}{2.8}\Cd[0,1]{4,2}{1.8}\Cd[0,1]{4,3}{0.8}
+	\Cd[1,0]{1,4}{2.8}\Cd[1,0]{2,4}{1.8}\Cd[1,0]{3,4}{0.8}
+	\Cd[1,1]{3,3}{0.8}
+	\multiput(3.5,5.5)(1,-1){6}{\circle{.4}}
+	\multiput(0,0)(1,0){7}{\line(0,1){6}}\multiput(0,0)(0,1){7}{\line(1,0){6}} % matrix
+	\put(3.5,5.5){\color{red}\line(1,-1){5}}\put(7,2){\color{red}\linethickness{1.5pt}\vector(1,0){2}}
+\end{picture}\end{center}\caption{Computation diagonal and dependencies}\end{figure}
+
+\begin{verbatim}
+	// Sync between blocks, removing __threadfence() is incorrect but works
+	// __threadfence();
+	#ifdef SH_PARA // wait for all blocks
+	__syncthreads();
+	++tP; if (threadIdx.x==0) lock[tB]=tP;
+	for (unsigned b=threadIdx.x;b<gridDim.x;b+=blockDim.x) { // sync with all blocks
+		while(lock[(tB+b)%gridDim.x]<tP) {}
+	}
+	#else // wait for previous block only
+	if (threadIdx.x==0) { lock[tB]=++tP; if (tB) while(lock[tB-1]<tP) {} }
+	#endif
+	__syncthreads();
+\end{verbatim}
+
+\subsubsection{Parsers code generation}
+
+\subsubsection{Backtracking on the GPU}
+
+\subsection{Runtime execution engine}
+
+XXX: try to add const wherever possible in the CUDA code, maybe this allows the compiler to produce more efficient constant folding.
 
@@
 \end{verbatim}
-
-\subsection{JNI wrappers and runtime execution engine}
 

### 2013-01-04 src/main/c/include/small_gpu.h

@@ -28,25 +28,23 @@ __global__ void gpu_solve(const TI* in0, const TI* in1, TC* cost, TB* back, vola
 	const unsigned tB = blockIdx.x;
 	unsigned tP=s_start; // block progress
 
-#ifdef SH_RECT
 	for (unsigned jj=s_start; jj<s_stop; ++jj) {
+		#ifdef SH_RECT
 		for (unsigned i=tI; i<M_H; i+=tN) {
 			unsigned j = jj-tI;
 			if (j<M_W) {
-#endif
-#ifdef SH_TRI
-	for (unsigned jj=s_start; jj<s_stop; ++jj) {
+		#endif
+		#ifdef SH_TRI
 		for (unsigned ii=tI; ii<M_H; ii+=tN) {
 			unsigned i=M_H-1-ii;
 			unsigned j=i+jj;
 			if (j<M_W) {
-#endif
-#ifdef SH_PARA
-	for (unsigned jj=s_start; jj<s_stop; ++jj) {
+		#endif
+		#ifdef SH_PARA
 		for (unsigned i=tI; i<M_H; i+=tN) {
 			unsigned j=jj+i;
 			{
-#endif
+		#endif
 				TB b=BT_STOP; TC c=0,c2; // stop
 				if (!INIT(i,j)) { p_kernel }
 				cost[idx(i,j)] = c;

### 2013-01-05 docs/report/implementation.tex
More report.

--- 2013-01-04 docs/report/implementation.tex
@@
 \subsection{Code generation}
+The code generation step produces multiple outputs that are tightly bound to each other. Besides the Scala wrapper (a simple JNI interface), in the C/CUDA code generated we distinguish:\ol
+\item JNI input and output conversion functions
+\item Host memory management and scheduling of CUDA kernels
+\item CUDA matrix computation, which can be further decomposed into matrix scheduling (loops) and (matrix cell) computation. 
+\item CUDA backtrack collection kernel
+\ole
+
+\subsubsection{Scala structures conversion (JNI)}
+Since Scala general types can be extremely complex and might heavily depend of the JVM, we want to restrict the supported types; additionally types should be of fixed size for more efficient processing and easier memory allocation.
+We support the following types:\ul
+\item \textbf{Primitive types:} natively supported in both Java and C.
+Since there is some little semantics difference between these two languages types, we used C (signed) types as reference.
+Supported types are: boolean, byte (unsigned char), char, short, int (32bit), long (64bit), float and double.
+\item \textbf{Empty case classes:} user-defined types might be more complex, so we allow users to define case classes that serve as data container and would be translated into C {\tt struct}s.
+\item \textbf{Tuples:} if the user-defined type is fairly simple, a named case class might be cumbersome.
+Tuples are a syntactical lightweight alternative to case classes, although they translate very similarly.
+Since Tuple classes are generic and can carry different member types; need to name tuple types uniquely, according to their arity and inner types.
+\ule
+
+{\color{red} Currently we use {\tt Manifest}s to extract types, and convert their string representation into our restricted subset.
+This imposes the additional restriction that we can not nest tuples into case classes, because generic types are then erased.
+However, the same effect could be achieved with Scala 2.10 {\tt TypeTag}s although we then need to rely on macros expansion to convert them properly into concrete classes.
+[with help from Eugene, \url{https://gist.github.com/4407488}].}
 
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+XXX: how to get inner types
+XXX: explain hierarchy restrictions
+
+XXX: supported types: primitive, tuples, classes, discuss limitations.
+XXX: input conversion (native types for classes, wrapped types for tuples), output conversion (similar as input).
+
+\begin{verbatim}
+static unsigned jni_read(JNIEnv* env, jobjectArray input, input_t** in) {
+  if (*in) free(*in); *in=NULL; if (input==NULL) return 0;
+  jsize i,size = env->GetArrayLength(input); if (size==0) return 0;
+  *in=(input_t*)malloc(size*sizeof(input_t)); if (!*in) { fprintf(stderr,"Not enough memory.\n"); exit(1); }
+  // Loading method handles
+  jobject el = env->GetObjectArrayElement(input, 0);
+  jclass cls = env->GetObjectClass(el);
+  jmethodID m_1 = env->GetMethodID(cls, "_1", "()I");
+  jmethodID m_2 = env->GetMethodID(cls, "_2", "()I");
+  // Loading data elements
+  for (i=0;i<size;++i) {
+    el = env->GetObjectArrayElement(input, i);
+    (*in)[i]._1 = (int)env->CallIntMethod(el, m_1);
+    (*in)[i]._2 = (int)env->CallIntMethod(el, m_2);
+  }
+  return size;
+}
+
+static jobject jni_write(JNIEnv* env, T3iii score, trace_t* trace, size_t size) {
+  // Result object
+  jclass cls_1 = env->FindClass("java/lang/Integer");
+  jmethodID ctr_1 = env->GetMethodID(cls_1,"<init>","(I)V");
+  jobject res_1 = env->NewObject(cls_1,ctr_1,score._1);
+  jclass cls_2 = env->FindClass("java/lang/Integer");
+  jmethodID ctr_2 = env->GetMethodID(cls_2,"<init>","(I)V");
+  jobject res_2 = env->NewObject(cls_2,ctr_2,score._2);
+  jclass cls_3 = env->FindClass("java/lang/Integer");
+  jmethodID ctr_3 = env->GetMethodID(cls_3,"<init>","(I)V");
+  jobject res_3 = env->NewObject(cls_3,ctr_3,score._3);
+  jclass cls = env->FindClass("scala/Tuple3");
+  jmethodID ctr = env->GetMethodID(cls,"<init>","(Ljava/lang/Object;Ljava/lang/Object;Ljava/lang/Object;)V");
+  jobject res = env->NewObject(cls,ctr,res_1,res_2,res_3);
+  // Backtrack trace
+  if (trace==NULL) return res;
+  jclass cl2 = env->FindClass("scala/Tuple2");
+  jmethodID ct2 = env->GetMethodID(cl2, "<init>", "(Ljava/lang/Object;Ljava/lang/Object;)V");
+  #define J_PAIR(A,B) env->NewObject(cl2, ct2, A, B)
+  jclass cli = env->FindClass("java/lang/Integer");
+  jmethodID cti = env->GetMethodID(cli, "<init>", "(I)V");
+  #define J_INT(X) env->NewObject(cli, cti, X)
+  jclass cln = env->FindClass("scala/collection/immutable/List");
+  jmethodID ctn = env->GetStaticMethodID(cln, "empty", "()Lscala/collection/immutable/List;");
+  const jobject nil = env->CallStaticObjectMethod(cln, ctn);
+  jclass clc = env->FindClass("scala/collection/immutable/$colon$colon");
+  jmethodID ctc = env->GetMethodID(clc, "<init>", "(Ljava/lang/Object;Lscala/collection/immutable/List;)V");
+  #define J_CONS(H,T) env->NewObject(clc, ctc, H, T)
+  jobject jtrace = nil;
+  for (int i=0; i<size; ++i) {
+    jobject ixs = nil; trace_t* tr=&trace[i];
+    for (int l=trace_len[tr->rule]; l>0; --l) ixs = J_CONS(J_INT(tr->pos[l-1]), ixs);
+    jobject el = J_PAIR( J_PAIR(J_INT(tr->i),J_INT(tr->j)) , J_PAIR(J_INT(tr->rule),ixs) );
+    jtrace = J_CONS(el, jtrace);
+  }
+  return J_PAIR(res,jtrace);
+}
+
+jobject Java_{className}_parse(JNIEnv* env, jobject obj, jobjectArray input1) {
+  input_t *in1=NULL; jni_read(env,input1,&in1);
+  g_init(in1,NULL); free(in1); g_solve();
+  T3iii score=g_backtrack(NULL,NULL);
+  jobject result = jni_write(env, score, NULL, 0);
+  g_free(); return result;
+}
+
+jobject Java_{className}_backtrack(JNIEnv* env, jobject obj, jobjectArray input1) {
+  input_t *in1=NULL; jni_read(env,input1,&in1);
+  g_init(in1,NULL); free(in1); g_solve();
+  trace_t *trace=NULL; unsigned size=0;
+  T3iii score=g_backtrack(&trace,&size);
+  jobject result = jni_write(env, score, trace, size); free(trace);
+  g_free(); return result;
+}
+\end{verbatim}
+
+\subsubsection{Host wrappers}
+XXX: Explain the complexity analysis and split into multiple kernels.
+
+XXX: interface JNI and CUDA kernel launch, allocate/free host/CUDA memory, sets the number of slices for the computation (overcome the kernel timeout issue).
+
+\begin{verbatim}
+void g_init(input_t* in1, input_t* in2) {
+  cuMalloc(g_in1,sizeof(input_t)*(M_H-1));
+  cuPut(in1,g_in1,sizeof(input_t)*(M_H-1),NULL);
+  g_in2=NULL;
+  cuMalloc(g_cost,sizeof(cost_t)*MEM_MATRIX);
+  cuMalloc(g_back,sizeof(back_t)*MEM_MATRIX);
+}
+
+void g_free() { cuFree(g_in1); cuFree(g_cost); cuFree(g_back); cudaDeviceReset(); }
+
+void g_solve() {
+  #define WARP_SIZE 32 // constant over CUDA devices
+  unsigned blk_size = WARP_SIZE;
+  unsigned blk_num = (M_H+blk_size-1)/blk_size;
+  unsigned* lock; cuMalloc(lock,sizeof(unsigned)*blk_num);
+  cuErr(cudaMemset(lock,0,sizeof(unsigned)*blk_num));
+  gpu_solve<<<blk_num, blk_size, 0, NULL>>>(g_in1, g_in2, g_cost, g_back, lock, 0, (M_W));
+  cuFree(lock);
+}
+
+T3iii g_backtrack(trace_t** trace, unsigned* size) {
+  T3iii res; unsigned i0=0, j0=M_W-1;
+  cuGet(&res,&g_cost[idx(i0,j0)].M,sizeof(T3iii),NULL);
+  if (trace && size) {
+    unsigned mem=(M_W+M_H)*sizeof(trace_t);
+    trace_t *g_trace=NULL; cuMalloc(g_trace,mem);
+    unsigned *g_size=NULL; cuMalloc(g_size,sizeof(unsigned));
+// XXX: new size adaptation
+    gpu_backtrack<<<1,1,0,NULL>>>(g_trace, g_size, g_back, i0, j0);
+    cuGet(size,g_size,sizeof(unsigned),NULL); cuFree(g_size); mem=(*size)*sizeof(trace_t);
+    *trace=(trace_t*)malloc(mem); cuGet(*trace,g_trace,mem,NULL); cuFree(g_trace);
+  }
+  return res;
+}
+\end{verbatim}
+
+\subsubsection{Matrix computation scheduling}
+Similarly as in the ad-hoc implementation, progress is made along the diagonal (see \ref{mem_layout}) and each thread is responsible of one line.
+That is, the matrix is swept horizontally by a <<diagonal of threads>>, that are enabled only if they are within a valid matrix cell.
+
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(6,6)
+	\def\Cfl2#1{#1{0,4}#1{0,3}#1{1,3}#1{0,2}#1{1,2}#1{2,2}#1{0,1}#1{1,1}#1{2,1}#1{3,1}#1{0,0}#1{1,0}#1{2,0}#1{3,0}#1{4,0}}
+	\Cfl2{\Cg}
+	{\color{cyan}\Cd[0,1]{4,1}{2.8}\Cd[0,1]{4,2}{1.8}\Cd[1,0]{1,4}{2.8}\Cd[1,0]{2,4}{1.8}\Cd[1,1]{3,3}{0.8}\Cd[2,1]{2,3}{1.8}\Cd[1,2]{3,2}{.9}}
+	\Cd[0,1]{4,3}{0.8}\Cd[1,0]{3,4}{0.8}
+	\multiput(3.5,5.5)(1,-1){6}{\circle{.4}}
+	\multiput(0,0)(1,0){7}{\line(0,1){6}}\multiput(0,0)(0,1){7}{\line(1,0){6}} % matrix
+	\put(3.5,5.5){\color{lightgray}\line(1,-1){5}}
+	\multiput(3.7,5.5)(1,-1){6}{\color{red}\linethickness{1.5pt}\vector(1,0){2}}
+	\put(8.6,0.05){\tiny thread 0}
+	\put(3.6,5.75){\tiny thread 5}
+\end{picture}\end{center}\caption{<<Diagonal of threads>> and maximal dependencies}\label{fig:diag_deps}\end{figure}
+
+Special care must be taken to handle computation dependencies: within a warp, all threads are executed at the same time, hence no synchronization is necessary.
+To benefit from this implicit synchronization, we set block size being equal to wrap size.
+It remains to provide inter-block synchronization: dependencies are along line, column and possibly intermediate elements.
+By induction on rows and columns, it suffice to have the last column and row element valid.
+Since line is computed by the current thread (hereby valid), it only remains to guarantee that the column element of the previous line is valid (in figure \ref{fig:diag_deps}, previous refers to the line immediately below).
+To do that, each block writes last valid diagonal in a <<lock>> array, and next block need only to wait (polling) until desired element is marked valid.
+
+\begin{lstlisting}[language=C,caption=Synchronization with previous thread block (active waiting)]
+__global__ void gpu_solve(/*...*/ volatile unsigned* lock, unsigned d_start, unsigned d_stop) {
+	const unsigned tB = blockIdx.x;
+	unsigned tP=d_start; // block progress
+
+	for (unsigned diag=d_start; diag<d_stop; ++diag) {
+		/* ... compute diagonal values ... */
+
+		// __threadfence();
+		if (threadIdx.x == 0) {
+			lock[tB] = ++tP;
+			if (tB > 0) while(lock[tB-1]<tP) {}
+		}
+		__syncthreads();
+	}
+}
+\end{lstlisting}
+
+\subsubsection{Parsers code generation}
+XXX: structures generation(?)
+XXX: refer to architecture, explain concrete transformations for all parsers
+
+\begin{verbatim}
+__device__ inline T3iii fun0(T2ii i) { return (T3iii){i._1,0,i._2}; }
+__device__ inline int fun1(T3iii a) { return a._2; }
+__device__ inline T3iii fun2(T3iii l, T3iii r) { return (T3iii){l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3}; }
+
+__global__ void gpu_solve(const input_t* in1, const input_t* in2, cost_t* cost, back_t* back, /*...*/) {
+...
+        #define VALID(I,J,RULE) (back[idx(I,J)].RULE.rule!=-1)
+        /* --- M[i,j] --- */
+        if (i+1==j) {
+          T3iii _c=fun0(in1[i]); if (fun1(_c)<fun1(_cost.M) || _back.M.rule==-1) { _cost.M=_c; _back.M=(bt1){0}; }
+        }
+        _unroll for(int k=i+1; k<j; ++k) {
+          if (VALID(i,k,M) && VALID(k,j,M)) {
+            T3iii _c=fun2(cost[idx(i,k)].M,cost[idx(k,j)].M); if (fun1(_c)<fun1(_cost.M) || _back.M.rule==-1) { _cost.M=_c; _back.M=(bt1){1,{k}}; }
+          }
+        }
+        cost[idx(i,j)] = _cost;
+        back[idx(i,j)] = _back;
+...
+}
+\end{verbatim}
+
+\subsubsection{Backtracking on the GPU}
+XXX: explain how it is done, how to get list reversal for free, explain lattice of elements with construction order.
+
+\begin{verbatim}
+__global__ void gpu_backtrack(trace_t* trace, unsigned* size, back_t* back, int i0, int j0) {
+  const unsigned trace_len[2] = {1,1};
+  trace_t *rd=trace, *wr=trace; *size=0;
+  #define PUSH_BACK(I,J,RULE) { wr->i=I; wr->j=J; wr->rule=RULE; ++wr; ++(*size); }
+  PUSH_BACK(i0,j0,0);
+  for(;rd<wr;++rd) {
+    bt1* bt;
+    switch (rd->rule) {
+      case 0: bt=(bt1*)&back[idx(rd->i,rd->j)].M; break;
+      case 1: bt=(bt1*)&back[idx(rd->i,rd->j)].M; break;
+    }
+    rd->rule=bt->rule;
+    for (int i=0,l=trace_len[rd->rule]; i<l; ++i) rd->pos[i]=bt->pos[i];
+    switch (rd->rule) {
+      case 0: break;
+      case 1: PUSH_BACK(rd->i,rd->pos[0],0); PUSH_BACK(rd->pos[0],rd->j,0); break;
+    }
+  }
+}
+\end{verbatim}
+
+\subsection{Runtime execution engine}
 

### 2013-01-05 docs/report/implementation.tex
Nits.

@@
 
-{\color{red} Currently we use {\tt Manifest}s to extract types, and convert their string representation into our restricted subset.
+{\color{red} Currently we use {\tt Manifest}s and reflection to extract types, and convert their string representation into our restricted subset.
+Manifests expands tuple inner types and reflection can be used to find class member's types.
 This imposes the additional restriction that we can not nest tuples into case classes, because generic types are then erased.
 However, the same effect could be achieved with Scala 2.10 {\tt TypeTag}s although we then need to rely on macros expansion to convert them properly into concrete classes.
-[with help from Eugene, \url{https://gist.github.com/4407488}].}
+[see hint from Eugene, \url{https://gist.github.com/4407488}].}
+
+The JNI functions are involved at input to decode sequences arrays and at output, to encode the result and possibly its corresponding trace. Input method is constructed in two steps:\ul
+\item Recursively obtain the classes and accessor methods of the composite input type.
+A subtle variation is that case classes primitive types are immediately converted into native types whereas tuple members are boxed in their respective class (i.e.
+{\tt java.lang.Integer}, ...).
+\item For each element of the input array, retrieve the objects recursively and write their primitive values in the corresponding {\tt struct} array.
+\ule
+The output method consist of two different steps:\ul
+\item Converting the result into its JVM counterpart by using the opposite rule as for decoding input (but with JNI types specified in the constructor lookup instead of accessors).
+\item Optionally encoding the backtrack: this is pretty straightforward as the structure is more regular (and make uses of Lists); additional care should be taken to avoid bloating concatenation indices lists with unnecessary elements (as C uses fixed memory whereas Scala lists length might vary).
+\ule
 
@@
 
-XXX: how to get inner types
-XXX: explain hierarchy restrictions
-
-XXX: supported types: primitive, tuples, classes, discuss limitations.
-XXX: input conversion (native types for classes, wrapped types for tuples), output conversion (similar as input).
-
 \begin{verbatim}
-static unsigned jni_read(JNIEnv* env, jobjectArray input, input_t** in) {
-  if (*in) free(*in); *in=NULL; if (input==NULL) return 0;
-  jsize i,size = env->GetArrayLength(input); if (size==0) return 0;
-  *in=(input_t*)malloc(size*sizeof(input_t)); if (!*in) { fprintf(stderr,"Not enough memory.\n"); exit(1); }
-  // Loading method handles
-  jobject el = env->GetObjectArrayElement(input, 0);
-  jclass cls = env->GetObjectClass(el);
-  jmethodID m_1 = env->GetMethodID(cls, "_1", "()I");
-  jmethodID m_2 = env->GetMethodID(cls, "_2", "()I");
-  // Loading data elements
-  for (i=0;i<size;++i) {
-    el = env->GetObjectArrayElement(input, i);
-    (*in)[i]._1 = (int)env->CallIntMethod(el, m_1);
-    (*in)[i]._2 = (int)env->CallIntMethod(el, m_2);
-  }
-  return size;
-}
-
-static jobject jni_write(JNIEnv* env, T3iii score, trace_t* trace, size_t size) {
-  // Result object
-  jclass cls_1 = env->FindClass("java/lang/Integer");
-  jmethodID ctr_1 = env->GetMethodID(cls_1,"<init>","(I)V");
-  jobject res_1 = env->NewObject(cls_1,ctr_1,score._1);
-  jclass cls_2 = env->FindClass("java/lang/Integer");
-  jmethodID ctr_2 = env->GetMethodID(cls_2,"<init>","(I)V");
-  jobject res_2 = env->NewObject(cls_2,ctr_2,score._2);
-  jclass cls_3 = env->FindClass("java/lang/Integer");
-  jmethodID ctr_3 = env->GetMethodID(cls_3,"<init>","(I)V");
-  jobject res_3 = env->NewObject(cls_3,ctr_3,score._3);
-  jclass cls = env->FindClass("scala/Tuple3");
-  jmethodID ctr = env->GetMethodID(cls,"<init>","(Ljava/lang/Object;Ljava/lang/Object;Ljava/lang/Object;)V");
-  jobject res = env->NewObject(cls,ctr,res_1,res_2,res_3);
-  // Backtrack trace
-  if (trace==NULL) return res;
-  jclass cl2 = env->FindClass("scala/Tuple2");
-  jmethodID ct2 = env->GetMethodID(cl2, "<init>", "(Ljava/lang/Object;Ljava/lang/Object;)V");
-  #define J_PAIR(A,B) env->NewObject(cl2, ct2, A, B)
-  jclass cli = env->FindClass("java/lang/Integer");
-  jmethodID cti = env->GetMethodID(cli, "<init>", "(I)V");
-  #define J_INT(X) env->NewObject(cli, cti, X)
-  jclass cln = env->FindClass("scala/collection/immutable/List");
-  jmethodID ctn = env->GetStaticMethodID(cln, "empty", "()Lscala/collection/immutable/List;");
-  const jobject nil = env->CallStaticObjectMethod(cln, ctn);
-  jclass clc = env->FindClass("scala/collection/immutable/$colon$colon");
-  jmethodID ctc = env->GetMethodID(clc, "<init>", "(Ljava/lang/Object;Lscala/collection/immutable/List;)V");
-  #define J_CONS(H,T) env->NewObject(clc, ctc, H, T)
-  jobject jtrace = nil;
-  for (int i=0; i<size; ++i) {
-    jobject ixs = nil; trace_t* tr=&trace[i];
-    for (int l=trace_len[tr->rule]; l>0; --l) ixs = J_CONS(J_INT(tr->pos[l-1]), ixs);
-    jobject el = J_PAIR( J_PAIR(J_INT(tr->i),J_INT(tr->j)) , J_PAIR(J_INT(tr->rule),ixs) );
-    jtrace = J_CONS(el, jtrace);
-  }
-  return J_PAIR(res,jtrace);
-}
-
 jobject Java_{className}_parse(JNIEnv* env, jobject obj, jobjectArray input1) {

### 2013-01-05 docs/report/implementation.tex
Nits in complexity analysis.

@@
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+\subsubsection{Host wrappers}
+The host wrappers are the functions bridging between JNI and CUDA; their duties are:\ul
+\item Exposing JNI parsing and backtracking functions
+\item Calling appropriate conversion methods
+\item Allocating host and CUDA memory (and managing transfers between them)
+\item Launching CUDA kernels: matrix computation, backtrack, and possibly aggregation within window (additional aggregation among window results, would this option be set)
+\ule
 
-\begin{verbatim}
-jobject Java_{className}_parse(JNIEnv* env, jobject obj, jobjectArray input1) {
-  input_t *in1=NULL; jni_read(env,input1,&in1);
-  g_init(in1,NULL); free(in1); g_solve();
-  T3iii score=g_backtrack(NULL,NULL);
-  jobject result = jni_write(env, score, NULL, 0);
-  g_free(); return result;
-}
+These simple operations do not require a detailed description.
+The only peculiarity of our execution environment, is that the kernel execution duration is bound to approximately 10 seconds\footnote{Hard limit imposed by the operating system.
+Although workarounds exist for Linux and Windows (requiring a second graphic card to display the UI), none of them is compatible with Mac OS.
+Eventually, a hack has been devised to force the UI on CPU while keeping the dedicated CUDA card powered; unfortunately this does not alleviate the kernel execution timeout.}.
+To solve this issue, we estimate the overall complexity of matrix computation, which allows us to estimate running time, then break computation into multiple kernels sufficiently small to fit in the time limit.
 
-jobject Java_{className}_backtrack(JNIEnv* env, jobject obj, jobjectArray input1) {
-  input_t *in1=NULL; jni_read(env,input1,&in1);
-  g_init(in1,NULL); free(in1); g_solve();
-  trace_t *trace=NULL; unsigned size=0;
-  T3iii score=g_backtrack(&trace,&size);
-  jobject result = jni_write(env, score, trace, size); free(trace);
-  g_free(); return result;
-}
-\end{verbatim}
+Since computations are made diagonal-by-diagonal (see \ref{matrix_scheduling}), we can easily decompose the matrix computation by adapting the number of diagonals computed per kernel.
+The global complexity being the product of the number of elements and the complexity per element, the latter being equal to the number of unbounded concatenations (where maximal size is infinite).
 
-\subsubsection{Host wrappers}
-XXX: Explain the complexity analysis and split into multiple kernels.
-
-XXX: interface JNI and CUDA kernel launch, allocate/free host/CUDA memory, sets the number of slices for the computation (overcome the kernel timeout issue).
-
-\begin{verbatim}
-void g_init(input_t* in1, input_t* in2) {
-  cuMalloc(g_in1,sizeof(input_t)*(M_H-1));
-  cuPut(in1,g_in1,sizeof(input_t)*(M_H-1),NULL);
-  g_in2=NULL;
-  cuMalloc(g_cost,sizeof(cost_t)*MEM_MATRIX);
-  cuMalloc(g_back,sizeof(back_t)*MEM_MATRIX);
-}
-
-void g_free() { cuFree(g_in1); cuFree(g_cost); cuFree(g_back); cudaDeviceReset(); }
-
-void g_solve() {
-  #define WARP_SIZE 32 // constant over CUDA devices
-  unsigned blk_size = WARP_SIZE;
-  unsigned blk_num = (M_H+blk_size-1)/blk_size;
-  unsigned* lock; cuMalloc(lock,sizeof(unsigned)*blk_num);
-  cuErr(cudaMemset(lock,0,sizeof(unsigned)*blk_num));
-  gpu_solve<<<blk_num, blk_size, 0, NULL>>>(g_in1, g_in2, g_cost, g_back, lock, 0, (M_W));
-  cuFree(lock);
-}
-
-T3iii g_backtrack(trace_t** trace, unsigned* size) {
-  T3iii res; unsigned i0=0, j0=M_W-1;
-  cuGet(&res,&g_cost[idx(i0,j0)].M,sizeof(T3iii),NULL);
-  if (trace && size) {
-    unsigned mem=(M_W+M_H)*sizeof(trace_t);
-    trace_t *g_trace=NULL; cuMalloc(g_trace,mem);
-    unsigned *g_size=NULL; cuMalloc(g_size,sizeof(unsigned));
-// XXX: new size adaptation
-    gpu_backtrack<<<1,1,0,NULL>>>(g_trace, g_size, g_back, i0, j0);
-    cuGet(size,g_size,sizeof(unsigned),NULL); cuFree(g_size); mem=(*size)*sizeof(trace_t);
-    *trace=(trace_t*)malloc(mem); cuGet(*trace,g_trace,mem,NULL); cuFree(g_trace);
-  }
-  return res;
-}
-\end{verbatim}
-
-\subsubsection{Matrix computation scheduling}
+\subsubsection{Matrix computation scheduling} \label{matrix_scheduling}
 Similarly as in the ad-hoc implementation, progress is made along the diagonal (see \ref{mem_layout}) and each thread is responsible of one line.
@@
 \begin{lstlisting}[language=C,caption=Synchronization with previous thread block (active waiting)]
-__global__ void gpu_solve(/*...*/ volatile unsigned* lock, unsigned d_start, unsigned d_stop) {
+__global__ void gpu_solve(/*...*/ volatile unsigned* lock, // = {0}
+		unsigned d_start, unsigned d_stop) {
 	const unsigned tB = blockIdx.x;
@@
 \end{lstlisting}
+Notice that {\tt \_\_threadfence} is not necessary here, verifying the observation of \cite{gpu_barrier}.
 
 \subsubsection{Parsers code generation}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
 XXX: structures generation(?)

### 2013-01-05 src/main/scala/v4/CodeGen.scala

@@ -365,12 +365,12 @@ trait CodeGen extends BaseParsers { this:Signature =>
       head.add("#define input_t "+tpAlphabet)
       head.add("typedef struct { "+rulesOrder.map{n=>tpAnswer+" "+n+";"}.mkString(" ")+" } cost_t;")
       head.add("typedef struct { "+rulesOrder.map{n=>val c=rules(n).inner.cat; btTpe(c)+" "+n+";"}.mkString(" ")+" } back_t;")
-      // Matrix cell time complexity: O(n^ max(#unaggregated floating concatenations) )
-      var cpx:Int=0; rules.foreach{case(n,t) => ufc(t.inner)}
-      def ufc[T](p0:Parser[T]):Int = p0 match {
-        case Aggregate(p,_) => ufc(p); 0 case cc@Concat(l,r,_) => val c=(if (cc.hasBt) 1 else 0)+ufc(l)+ufc(r); if (c>cpx) cpx=c; c
-        case Filter(p,_) => ufc(p) case Map(p,_) => ufc(p) case Or(l,r) => Math.max(ufc(l),ufc(r)) case _ => 0
+      // Time complexity of computations: O(n^( max(#unbounded concatenations) + matrix_dimensions=2 ))
+      def ucc[T](p0:Parser[T]):Int = p0 match {
+        case cc@Concat(l,r,_) => ucc(l)+ucc(r)+(if (cc.max== -1) 1 else 0)
+        case Aggregate(p,_)=>ucc(p) case Filter(p,_)=>ucc(p) case Map(p,_)=>ucc(p) case Or(l,r)=>Math.max(ucc(l),ucc(r)) case _=>0
       }
+      val cpx = rules.map{case(_,t)=>ucc(t.inner)}.max + 2
       // Code generation informations
       val info="// Type: "+(if (twotracks) "sequence alignment" else "sequence parser" )+(if (window>0) ", window="+window else "")+", complexity/element=O(n^"+cpx+"), splits={SPLITS}\n"+
         rulesOrder.zipWithIndex.map { case(n,i)=> val p=rules(n); "// Rule #%-2d %-8s : id=%-2d alt=%-2d cat=%-2d min=%-2d max=%-2d\n".format(i,"'"+n+"'",p.id,p.inner.alt,p.inner.cat,p.min,p.max) }.mkString
@@ -383,9 +383,8 @@ trait CodeGen extends BaseParsers { this:Signature =>
   abstract class CodeCompiler extends CCompiler with ScalaCompiler {
[…]

### 2013-01-05 docs/report/design.tex
Nits in report.
NIts.
More report.

@@
 % ------------------------------------------------------------------------------------------------
-\subsection{CUDA storage: from list to optional value}
+\subsection{CUDA storage: from list to optional value} \label{normalization}
 Since lists are natively supported, it is natural to gather parsers results into lists in Scala implementation.
@@
 \newpage
-\subsection{Memory layout}
+\subsection{Memory layout}\label{mem_layout}
 A major bottleneck on massively parallel architecture with shared memory (like CUDA) is the global memory access.

### 2013-01-05 docs/report/implementation.tex
More report.

@@
 
+% ----------------------------------------------
 \subsubsection{Scala structures conversion (JNI)}
@@
 
+% ----------------------------------------------
 \subsubsection{Host wrappers}
@@
 
+% ----------------------------------------------
 \subsubsection{Matrix computation scheduling} \label{matrix_scheduling}
@@
 To do that, each block writes last valid diagonal in a <<lock>> array, and next block need only to wait (polling) until desired element is marked valid.
+Notice that {\tt \_\_threadfence} is not mandatory (hereby slightly improving performance), verifying the observation of \cite{gpu_barrier}.
 
@@
 \end{lstlisting}
-Notice that {\tt \_\_threadfence} is not necessary here, verifying the observation of \cite{gpu_barrier}.
 
+% ----------------------------------------------
 \subsubsection{Parsers code generation}
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+Parsers generation is independent of user-defined function generation (see \ref{user_fun_gen}).
+Tabulation inner parsers are first wrapped in additional aggregation (by $h$, hereby ensuring they produce at most one result) and normalized (according to \ref{normalization}); code generation then occurs recursively, producing a list of loops and conditions, and body (possibly with a hoisted part).
+Additionally, position variables are maintained and sub-rule index and concatenation indices are propagated.
+We give an overview of each parser transformation:\ul
+\item \textbf{Terminal:} provides its own C code, which correspond usually to the input element value, its position or the position of the matching range.
+\item \textbf{Tabulate:} is a simple value load, possibly wrapped into a validity check. Useless validity verification can be removed by marking the tabulation as <<always valid>>.
+\item \textbf{Aggregate:} corresponds to an intermediate (value,backtrack) pair where inner parsers write their result; outermost aggregation is written back to corresponding (cost, backtrack) matrices.
+Validity information, and concatenation indices are propagated within backtrack.
+To preserve a correct semantic, inner aggregations body is hoisted outside loops and condition checks of the enclosing parser.
+\item \textbf{Or:} since parsers are normalized and operate on a single aggregation result, it suffice to emit sequentially code of alternatives.
+\item \textbf{Map:} wraps its argument into a the user-defined function call
+\item \textbf{Filter:} wraps its body into user-defined condition check
+\item \textbf{Concat:} fixed size concatenation are wrapped in simple conditions; moving concatenations are wrapped in a {\tt for} loop.
+The loops and conditions are further simplified to reduce range and remove useless conditions before actual code is emitted.
+\ule
 
-XXX: structures generation(?)
-XXX: refer to architecture, explain concrete transformations for all parsers
+Intermediate types must be correctly declared.
+To do that each user-defined function provides its input and output types.
+Aggregation temporary values declaration is ensured by a \textit{exists-or-declare} header policy that is called for every type declaration.
 
-\begin{verbatim}
-__device__ inline T3iii fun0(T2ii i) { return (T3iii){i._1,0,i._2}; }
-__device__ inline int fun1(T3iii a) { return a._2; }
-__device__ inline T3iii fun2(T3iii l, T3iii r) { return (T3iii){l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3}; }
+% ----------------------------------------------
+\subsubsection{Backtracking on the GPU}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
+XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
 
-__global__ void gpu_solve(const input_t* in1, const input_t* in2, cost_t* cost, back_t* back, /*...*/) {
-...
-        #define VALID(I,J,RULE) (back[idx(I,J)].RULE.rule!=-1)
-        /* --- M[i,j] --- */
-        if (i+1==j) {
-          T3iii _c=fun0(in1[i]); if (fun1(_c)<fun1(_cost.M) || _back.M.rule==-1) { _cost.M=_c; _back.M=(bt1){0}; }
-        }
-        _unroll for(int k=i+1; k<j; ++k) {
-          if (VALID(i,k,M) && VALID(k,j,M)) {
-            T3iii _c=fun2(cost[idx(i,k)].M,cost[idx(k,j)].M); if (fun1(_c)<fun1(_cost.M) || _back.M.rule==-1) { _cost.M=_c; _back.M=(bt1){1,{k}}; }
-          }
-        }
-        cost[idx(i,j)] = _cost;
-        back[idx(i,j)] = _back;
-...
-}
-\end{verbatim}
-
-\subsubsection{Backtracking on the GPU}
 XXX: explain how it is done, how to get list reversal for free, explain lattice of elements with construction order.
@@
 
+% ----------------------------------------------
+\subsubsection{User functions generation} \label{user_fun_gen}
+How LMS is used to generate function
+Possibly write that we can do the same with macros
+Why cannot use LMS for everything
+Where and how is LMS useful
+
 \subsection{Runtime execution engine}
-
-\begin{verbatim}
- * Parser construction:
- * 1. Assign a "OR_id" (sub_rule_id) and "CONCAT_id" to all parsers so that we know which rule applies
- *    How to skip some indices ?
- *
- * Scala running:
- * 2. Provide meaningful backtrack: rule_id and list of concat indices
- *
- * Code generation:
- * 1. Normalize rules (at hash-map insertion (?))
- * 2. Compute dependency analysis between rules => order them into a list/queue
- *    - If rule R contains another rule S unconcatenated (or concatenated with empty)
- *      then we have S -> T (S before T)
- * 3. Compute the maximal number of concatenation among each rule (field in Treeable(?))
- * 4. Break rules into subrules (at each Or, which must be at top of the rule)
-\end{verbatim}
+XXX: compiler for C and for Scala
 
@@
 \end{verbatim}
-
-\subsection{LMS integration}
-Why cannot use LMS for everything
-Where and how is LMS useful

### 2013-01-05 docs/report/implementation.tex
Nits.

@@
 \newpage
-\subsection{Code generation}
+\subsection{Code generation} \label{codegen}
 The code generation step produces multiple outputs that are tightly bound to each other. Besides the Scala wrapper (a simple JNI interface), in the C/CUDA code generated we distinguish:\ol
@@
 Tabulation inner parsers are first wrapped in additional aggregation (by $h$, hereby ensuring they produce at most one result) and normalized (according to \ref{normalization}); code generation then occurs recursively, producing a list of loops and conditions, and body (possibly with a hoisted part).
-Additionally, position variables are maintained and sub-rule index and concatenation indices are propagated.
+Additionally, position variables are maintained and subrule index and concatenation indices are propagated.
 We give an overview of each parser transformation:\ul
@@
 \subsubsection{Backtracking on the GPU}
+The backtracking is processed similarly as with the Scala parser, the major difference being that since we are generating the C code, we can provide an immediate mapping from the subrule index to the backtrack elements to add to the trace.
+The backtrack is done in 3 steps:\ul
+\item If window is set, the windowing aggregation kernel is run to determine the position of the best result within the matrix.
+Otherwise the best position can be found at the last computed element of the matrix.
+\item For a $m\times n$ matrix, allocate a $m+n$ vector with two heads (reading, writing, initialized at the same position). Write the best element in the vector.
+\item While there is a vector element that has been written but not read\ul
+	\item From the parser id and its position retrieve the corresponding (subrule, concatenation indices) pair by reading in the corresponding matrix cell
+	\item Using this, write new backtrack items that are at the origin of the current element.
+	\ule
+\ule
+
+Since code is generated, it is possible to write the last step using a switch case, hereby flattening the writes in the vector (compared to recursive calls in Scala).
+Finally, since the trace has to be reversed, we can obtain this transformation for free by constructing the trace list from the end in the JNI conversion.
+Reversing the list present the advantage that trace is immediately usable to construct the desired element.
+It might be possible that Scala and CUDA parser provide different traces to construct the same result, because the trace verifies the dependency order, which is only a partial order.
+
+% ----------------------------------------------
+\subsubsection{User functions generation} \label{user_fun_gen}
+{\color{red}
+\begin{verbatim}
+XXX: How LMS is used to generate function
+Possibly write that we can do the same with macros
+Why cannot use LMS for everything
+Where and how is LMS useful
+Also user function code generation with macros ?
+\end{verbatim}
+}
+
+\subsection{Runtime execution engine}
+The runtime execution engine is made of two instrumented compilers:\ul
+\item A wrapper for {\tt g++} and {\tt nvcc} that can combine different file types ({\tt .h}, {\tt .c}, {\tt .cu}) into a JNI library which is then loaded into the current JVM instance.
+If necessary, paths can be customized to fit the user environment.
+\item A wrapper for the Scala compiler, which allow the creation of Scala interface to the freshly compiled JNI libraries.
+It should be noted there that using {\tt VirtualDirectory} as compilation target prevents the interaction with JNI, hence physical path has to be used.
+\ule
+
+These two compilers interfaces are then mixed in another class that transform the previously (see \ref{codegen}) generated code, fixing input sizes and splits (number of kernels to launch to respect the time limit) constants, and execute it.
+
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
@@
 
-XXX: explain how it is done, how to get list reversal for free, explain lattice of elements with construction order.
+\subsection{LibRNA (?)}
 
-\begin{verbatim}
-__global__ void gpu_backtrack(trace_t* trace, unsigned* size, back_t* back, int i0, int j0) {
-  const unsigned trace_len[2] = {1,1};
-  trace_t *rd=trace, *wr=trace; *size=0;
-  #define PUSH_BACK(I,J,RULE) { wr->i=I; wr->j=J; wr->rule=RULE; ++wr; ++(*size); }
-  PUSH_BACK(i0,j0,0);
-  for(;rd<wr;++rd) {
-    bt1* bt;
-    switch (rd->rule) {
-      case 0: bt=(bt1*)&back[idx(rd->i,rd->j)].M; break;
-      case 1: bt=(bt1*)&back[idx(rd->i,rd->j)].M; break;
-    }
-    rd->rule=bt->rule;
-    for (int i=0,l=trace_len[rd->rule]; i<l; ++i) rd->pos[i]=bt->pos[i];
-    switch (rd->rule) {
-      case 0: break;
-      case 1: PUSH_BACK(rd->i,rd->pos[0],0); PUSH_BACK(rd->pos[0],rd->j,0); break;
-    }
-  }
-}
-\end{verbatim}
-
-% ----------------------------------------------
-\subsubsection{User functions generation} \label{user_fun_gen}
-How LMS is used to generate function
-Possibly write that we can do the same with macros
-Why cannot use LMS for everything
-Where and how is LMS useful
-
-\subsection{Runtime execution engine}
-XXX: compiler for C and for Scala
-
-\subsection{LibRNA (?)}
 The project development was scheduled in \ul

### 2013-01-05 docs/report/report.tex
NIts.
Nits.

@@
 \section{Future work} \ul
-\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
-\item Annotation on recursive functions to use dynamic programming like \\
-	{\tt @DynaProg def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
+\item Implementation for serial problems larger than memory, using hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory)
+\item Transforming non-serial into serial using aggregation functions/transformations
+\item Avoid to maintain the cost matrix in memory by moving data in the wavefront
+\item Non-serial scheduling for large problems (with memory loads for non-serializable)
+\item Add FPGA target platform (George Nithin)
+\item Pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char)
+\item Operate on larger words (ex 64 bits) to increase thread locality and reduce memory accesses
 \ule
@@
 
-Possible extensions / optimizations:
-- making serializable by using aggregation functions/transformations
-- avoiding the cost matrix by moving data in the wavefront
-- Scheduling for large problems (with memory loads for non-serializable)
-- add FPGA target platform
-- pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char)
-- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
-
 Core function F:
@@
           vv
-stats -> KK -> new_stats (y')
+stats -> [  ] -> new_stats (y')
 (y)       || \
@@
 
-For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
+For stats, we need to maintain an horizontal and vertical front
 For backtracking we need to maintain the whole table
@@
 ----+---+
-we many not need to store what's the previous cell, however, the previous
+we may not need to store what's the previous cell, however, the previous
 cell information (usually limited range) is much more compact than the score(32-64 bits)
@@
 \subsubsection*{Todo @TCK}\ul
-\item Test/proof parsers are correct --- make sure implementation is correct
+\item Test/proof parsers are correct -- make sure implementation is correct
 \item Automate test to compare against implementation
-\item Benchmarks -- use extensively CUDA profiler
+\item Benchmarks -- use CUDA profiler(?)
 \item Write report
@@
 \newpage
+%\usepackage{multicol}
+%\usepackage{etoolbox}
+%\patchcmd{\thebibliography}{\section*{\refname}}{\begin{multicols}{2}[\section*{\refname}]}{}{}
+%\patchcmd{\endthebibliography}{\endlist}{\endlist\end{multicols}}{}{}
+%\bibliographystyle{acm}
 \bibliographystyle{plain}

### 2013-01-06 docs/report/implementation.tex
Report cleanup, automatic use of CUDA if CodeGen is mixed in.

@@
 {\linethickness{1.5pt}\put(3,2.5){\vector(1,1){1.5}}\put(11,2.5){\vector(-1,1){1.5}}\put(7,6.5){\vector(0,1){1.5}}}
-\end{picture}\end{center}\caption{Class diagram (simplified)}\end{figure}
+\end{picture}\end{center}\caption{Scala parsers class diagram (simplified)}\end{figure}
 
@@
 \item JNI input and output conversion functions
-\item Host memory management and scheduling of CUDA kernels
+\item Host helpers for memory management and scheduling of CUDA kernels
 \item CUDA matrix computation, which can be further decomposed into matrix scheduling (loops) and (matrix cell) computation. 
@@
 \ole
+
+\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,8)
+\put(0,-.5){
+{\color{lightgray}\put(0,7){\tbox{4}{1}{\footnotesize\bf BaseParsers}}
+	\put(0,2){\rotatebox{90}{\tbox{4}{1}{\footnotesize\bf ADPParsers}}}
+	\put(1.5,2){\rotatebox{90}{\tbox{4}{1}{\footnotesize\bf TTParsers}}}
+	{\linethickness{1.5pt}\put(.5,6){\vector(1,1){1}}\put(2,6){\vector(0,1){1}}}}
+{\linethickness{1.5pt}\put(6,6){\vector(-7,2){3.5}}\put(8,1.33){\vector(3,1){2}}\put(8,1.33){\vector(3,-1){2}}}
+\put(3,0){\tbox{6}{6}{{\bf CodeGen} \footnotesize\\ code generation: parsers, backtrack, helpers, JNI \\[12pt] CodeCompiler}}
+\put(10,3.5){\tbox{6}{2.5}{{\bf CodeHeader} \footnotesize\\ Types conversion, headers management}}
+\put(10,1.5){\tbox{6}{1}{\bf ScalaCompiler}}\put(10,0){\tbox{6}{1}{\bf CCompiler}}
+\moveto(2,2)\lineto(2,1.5)\lineto(2.5,1.5)\strokepath\put(2.5,1){\vector(1,0){.5}}
+\moveto(.5,2)\lineto(.5,1)\lineto(2.5,1)\strokepath\put(2.5,1.5){\vector(1,0){.5}}
+\put(-1.25,-.1){\tbox[0l]{4.5}{1}{\footnotesize automatic call \tiny\\ if CodeGen mixed-in}}
+}
+\end{picture}\end{center}\caption{Code generation and runtime engine class diagram (simplified)}\end{figure}
 
@@
 \subsection{LibRNA (?)}
-
-The project development was scheduled in \ul
-\item Problems analysis
-\item Ad-hoc implementation (CUDA)
-\item Porting ADP parsers
-\item Improving parsers / adding functionalities
-\item Integrating CUDA generation into parsers
-\ule
-
-\begin{verbatim}
-1 week on hash maps
-2 weeks to define and analyze the problems (also with FPGA in mind)
-3 weeks restriction to non-serial and ad-hoc implementation on GPU (rectangle, triangle, parallelogram); translation of ADP parsers in Scala (Manohar)
-1 week run-time engine for Scala/JNI/CUDA
-05.11 - rework of the ADP parser to aim at generating C-like code
-12.11 - rework of ADP parsers to extend usages (cyclic, two-track) and aim at automatic backtracking
-19.11 - explorations in LMS / macros
-26.11 - full backtracking: apply/unapply/reapply, rework of the classes
-03.12 - Zuker/JNI
-10.12 - Yield analysis, code generation
-17.12 - code generation: detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support
-24.12 - code generation, sick
-31.12 - report
-\end{verbatim}
+{\color{red} XXX}

### 2013-01-06 docs/report/report.tex
Report cleanup, automatic use of CUDA if CodeGen is mixed in.

@@
 
+% Paper introduction:
+% Problem to solve, what exists (related work) and how do we compare to other
+% Contributions (3): 3 tensed sentences
+% Benchmarks => evaluation metrics, prove introduction statements by evaluation
 \subsection*{Abstract}
@@
 \section*{Planning}
-\textbf{Paper introduction}\ul
-\item Problem to solve, what exists (related work), how we compare to other, how to evaluate
-\item Contributions (3): 3 tensed sentences
-\item benchmark => prove by evaluation intro statements
-\ule
-
-\subsubsection*{Roadmap}
-\begin{tabular}{rl}
-16.11 & Rules normalization and automatic backtracking \\
-	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
-23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
-30.11 &--- Gap due to LMS missing knowledge \\
-7.12 & Benchmarking, grammar analysis \\
-14.12 & First thoughts for larger than device memory \\
-21.12 & Writing report \\
-28.12 & --- holiday --- \\
- 4.01 & --- holiday --- \\
-11.01 & Writing report: implementation description and plan for future work \\
-18.01 & Writing report
-\end{tabular}
-
 \subsubsection*{Todo @TCK}\ul
@@
 \ule
+
+Legacy roadmap deadlines: \\
+\begin{tabular}{ll}
+Nov 16 & Rules normalization and automatic backtracking \\
+	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
+Nov 23 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
+Nov 30 &--- gap due to LMS missing knowledge --- \\
+Dec 7 & Benchmarking, grammar analysis \\
+Dec 14 & First thoughts for larger than device memory \\
+Dec 21 & Writing report \\
+Jan 4 & --- holiday --- \\
+Jan 18 & Writing report: implementation description and plan for future work
+\end{tabular}
+
+\subsubsection*{Journal}
+\begin{tabular}{ll}
+Sep. 17	& Getting started with the project, reading related work on hash maps. \\
+Sep. 24	& Parallel hashing paper solve the problem for 32bit key/value pair. Stripped CudPP. \\
+		& Devised (but not implemented) extension beyond 64bit using memory areas locks. \\
+---		& Change of project suggested by Vojin (supervisor): joint work with Manohar on DP \\
+Oct. 01	& Problems specifications: serial/non-serial, started CUDA implementation with blocks. \\
+Oct. 08	& CudAlign solves serial monadic, might adapt it (but complicated / ad-hoc / in progress). \\
+		& Focus on non-serial problems that fit in GPU memory. \\
+Oct. 15	& First working implementation for non-serial problems (rectangle, triangle, parallelogram) \\
+Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
+Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. \\
+Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
+Nov. 12	& Rework ADP parsers: cyclic, two-track grammars and automatic backtracking discussion. \\
+Nov. 19	& Explorations LMS and Macros for C code, implementation is quite different, hence ad-hoc. \\
+Nov. 26	& Full backtracking stack: apply / unapply / reapply, refactoring of the classes. \\
+Dec. 03	& JNI for LibRNA to get coefficients for Zuker, errors in algebra (quite hairy code). \\
+Dec. 10	& Rework concatenation operators, yield analysis, code generation. \\
+Dec. 17	& Detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support. \\
+Dec. 24	& (sick 4 days), code generation: full JNI conversion support, rework CodeCompiler. \\
+Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
+Jan. 7	& {\color{red} XXX} \\
+Jan. 14	& {\color{red} XXX} \\
+\end{tabular}
 

### 2013-01-06 src/main/scala/v4/ADPParsers.scala

@@ -5,11 +5,13 @@ trait ADPParsers extends BaseParsers { this:Signature =>
   private var input: Input = null
   def in(k:Int):Alphabet = input(k)
   def size:Int = input.size
-  def parse(in:Input):List[Answer] = {
-    run(in,()=>(if (window>0) aggr(((0 to size-window).flatMap{x=>axiom(x,window+x)}).toList, h) else axiom(0,size)).map{_._1})
+  def parse(in:Input):List[Answer] = this match {
+    case c:CodeGen => List(c.parseCU(in.asInstanceOf[c.Input]).asInstanceOf[Answer])
+    case _ => run(in,()=>(if (window>0) aggr(((0 to size-window).flatMap{x=>axiom(x,window+x)}).toList, h) else axiom(0,size)).map{_._1})
   }
-  def backtrack(in:Input):List[(Answer,List[(Subword,Backtrack)])] = {
-    run(in,()=>if (window>0) aggr(((0 to size-window).flatMap{x=>axiom.backtrack(x,window+x)}).toList, h) else axiom.backtrack(0,size))
+  def backtrack(in:Input):List[(Answer,List[(Subword,Backtrack)])] = this match {
+    case c:CodeGen => List(c.backtrackCU(in.asInstanceOf[c.Input]).asInstanceOf[(Answer,List[(Subword,Backtrack)])])
+    case _ => run(in,()=>if (window>0) aggr(((0 to size-window).flatMap{x=>axiom.backtrack(x,window+x)}).toList, h) else axiom.backtrack(0,size))
   }
   def build(in:Input,bt:List[(Subword,Backtrack)]):Answer = run(in, ()=>axiom.build(bt))
   private def run[T](in:Input, f:()=>T) = { input=in; analyze; val res=f(); input=null; reset(); res }

### 2013-01-06 src/main/scala/v4/examples/MatrixMultGen2.scala

@@ -69,8 +69,9 @@ object MatrixMultGen2 extends MatrixMultGen2Grammar with CodeGen with App {
   override val window = 3 // only 3 consecutive matrices (aka 2x4, 4x2, 2x1 => 16)
 
   // ------- Extra codegen initialization
+  override val benchmark = true
   override val tps = (manifest[Alphabet],manifest[Answer])
   // ------- Extra codegen initialization
   //println(gen)
-  println(backtrackCU(input))
+  println(backtrack(input).head)
 }

### 2013-01-06 docs/report/report.tex
Added host memory fallback, nits in code.

@@
 
-\section{Future work} \ul
+Compare current implementation versus ad-hoc implementation. Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).
+If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.
+
+\section{Future work}
+We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
+\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in a general context, with the extra penalties of memory transfers between main and device memory and a more complex indexing strategy (since we need to find the matrix block in which lies the data, then address the element within the block.
+
+Since recent CUDA devices implement the possibility to address host memory directly from the device, a simple change in the memory allocation strategy could allow larger problems to be solved within the main memory instead of the GPU memory.
+However, a more thorough analysis on the addressing cost and load penalties should be done to decide which of the strategy is the best.
+
+{\color{red} XXX: need to force 64 bits mode -m64}
+
+% http://www.opensource.apple.com/source/xnu/xnu-2050.18.24/osfmk/vm/vm_map.c : add_wire_counts
+
+ large problems (with memory loads for non-serializable)
+\ole
+
+ \ul
 \item Implementation for serial problems larger than memory, using hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory)
@@
 \item Avoid to maintain the cost matrix in memory by moving data in the wavefront
-\item Non-serial scheduling for large problems (with memory loads for non-serializable)
+\item 
 \item Add FPGA target platform (George Nithin)

### 2013-01-06 src/main/scala/v4/examples/MatrixMult.scala

@@ -47,7 +47,7 @@ trait PrettyMatrixAlgebra extends MatrixSig {
 trait MatrixGrammar extends ADPParsers with MatrixSig {
   val matrixGrammar:Tabulate = tabulate("M",(
     el ^^ single
-  | (matrixGrammar ~ matrixGrammar) ^^ mult // { case (a1,a2) => mult(a1, a2) }
+  | (matrixGrammar ~ matrixGrammar) ^^ mult
   ) aggregate h,true)
 
   val axiom=matrixGrammar

### 2013-01-06 docs/report/design.tex
Added host memory fallback, nits in code.
More report, fixed memory allocation.

@@
 % ------------------------------------------------------------------------------------------------
-\subsection{User facing language requirements}
+\subsection{User facing language requirements} \label{user_lang}
 The field of dynamic programming has been influenced in the recent years by a methodology known as Algebraic Dynamic Programming which uses a grammar and an algebra to separate between the parsing and the score computation:
@@
 % ----------------------------------------------
-\subsubsection{Non-serial problems}
+\subsubsection{Non-serial problems} \label{ns_mem_transfer}
 Non-serial problems need to potentially access all elements that have been previously computed.
@@
 Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay.
-Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h\up{(T)} and 6 days 10h\up{(P)}, assuming there is no other scalability issues.
-Although this overhead seems appealing compared to the computation time, the total time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
+Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h and 6 days 10h\up{(P)}, assuming there is no other scalability issues, hence transfer would account respectively for 0.1\%\up{(R)}, 0.2\%\up{(T)} and 0.3\%\up{(P)} of the running time.
+Although this overhead seems appealing compared to the computation time, the total running time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
 Given that real problems (like RNA folding) operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
@@
 % ----------------------------------------------
-\subsubsection{Serial problems}
+\subsubsection{Serial problems} \label{serial_memory}
 The serial problems have the interesting property to access a fixed number of previous elements.

### 2013-01-06 docs/report/implementation.tex
More report, fixed memory allocation.

@@
 
-These simple operations do not require a detailed description.
-The only peculiarity of our execution environment, is that the kernel execution duration is bound to approximately 10 seconds\footnote{Hard limit imposed by the operating system.
+One peculiarity of our execution environment, is that the kernel execution duration is bound to approximately 10 seconds\footnote{Hard limit imposed by the operating system.
 Although workarounds exist for Linux and Windows (requiring a second graphic card to display the UI), none of them is compatible with Mac OS.
@@
 The global complexity being the product of the number of elements and the complexity per element, the latter being equal to the number of unbounded concatenations (where maximal size is infinite).
+
+\textbf{Bonus: problems larger than device memory}\\
+Problems larger than the device memory can actually be processed on recent CUDA devices (architecture $\ge 2.0$) as these are able to address the main memory from the device.
+However, since the distance between CUDA processors and memory is increased, there is a $5\times$ slowdown penalty to be paid in this configuration (experimentally, on a $1024\times 1024$ triangular matrix).
+Nevertheless, this workaround implementation has 2 benefits:\ul
+\item It allows larger problem to be solved, with very little implementation effort, would the user be patient enough for the computation to terminate
+\item It provides a good estimation of the main memory usage penalty, and hereby a strong argument in favor of the implementation described in \ref{ns_mem_transfer} (with only up to $\le 0.3\%$ transfer overhead).
+However, since we have not found concrete applications with such matrix size, the benefit of supporting large matrices is unclear, hence we leave the optimal implementation for future work.
+\ule
 

### 2013-01-06 docs/report/problems.tex
Nits in report.
More report, fixed memory allocation.

@@
 
-\subsubsection{Calculus simplifications}
+\subsubsection{Calculus simplifications} \label{calc_simplifications}
 In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:

### 2013-01-06 docs/report/report.tex
More report, fixed memory allocation.

--- 2013-01-06 docs/report/report.tex
@@
 \section{Benchmarks}
-{\color{red} XXX}
+{\color{red} XXX: explain why slower: approx 3x more memory used per score (maintain matrix dimensions)
 
-\section{Future work} \ul
-\item Implementation for serial problems larger than memory, using hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory)
-\item Transforming non-serial into serial using aggregation functions/transformations
-\item Avoid to maintain the cost matrix in memory by moving data in the wavefront
-\item Non-serial scheduling for large problems (with memory loads for non-serializable)
-\item Add FPGA target platform (George Nithin)
-\item Pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char)
-\item Operate on larger words (ex 64 bits) to increase thread locality and reduce memory accesses
-\ule
+XXX: continue benchmarking here
+}
+\begin{table}[H]\begin{center}\begin{tabular}{lrrrrrr} \toprule
+Matrix size			& 1024	& 2048	& 4096	& 8192 \\
+Splits				& 1		& 8		& 64		& 512 \\ \midrule
+Analysis				& 0.007	& 0.007	& 0.007	& 0.007 \\
+Code generation		& 0.071	& 0.072	& 0.073	& 0.071 \\
+C/CUDA compilation	& 3.025	& 1.836	& 1.832	& 1.841 \\
+Scala compilation		& 1.837	& 1.853	& 1.830	& 1.767 \\ \midrule
+- JNI read				& 0.027	& 0.028	& 0.029	& 0.058 \\
+- CUDA compute		& 0.868	& 3.378	& 19.782	& 139.697 \\
+- CUDA backtrack		& 0.009	& 0.012	& 0.023	& 0.045 \\
+- JNI output			& 0.005	& 0.009	& 0.018	& 0.032 \\
+Total execution			& 0.915	& 3.432	& 19.860	& 139.846 \\ \bottomrule
+\end{tabular}\end{center}\caption{Preliminary results for MatrixMultGen}\end{table}
 
+{\color{red} XXX: Compare current implementation versus ad-hoc implementation.
+Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.}
+
+\section{Future work}
+We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
+\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, hereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
+However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
+\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program.
+Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}).
+These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar).
+\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation.
+Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
+\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
+The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
+\item \textbf{Multi-dimensional matrices:} in the current implementation, all the matrices are encoded such that they are of the same size.
+Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size).
+In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.
+
+{\color{red}
 \begin{verbatim}
@@
 => but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)
+\end{verbatim}
+}
 
-Core function F:
-- in: s,t strings, neighbor costs: top, left, top+left, neighbor stats: top, left, top+left
-- out: backtrack information, cost(i,j), stats
-         stats (x)
-          ||
-          vv
-stats -> [  ] -> new_stats (y')
-(y)       || \
-          vv  --+ backtrack info (Bxy)
-      new_stats (x')
+\item \textbf{Independent computations:} {\color{red}The attentive reader might have noticed a discrepancy between the problem analysis and the recurrence rules of the Zuker problem implementation.
+This is due to the usage of multiple source for the problem description.
+}
 
-For stats, we need to maintain an horizontal and vertical front
-For backtracking we need to maintain the whole table
-... | . |
-..A | B | min/max: Min(D)=Min(Min(C),Min(B))
-----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
-..C | D | avg    : sum both #cells and values then divide at appropriate cell
-----+---+
-we may not need to store what's the previous cell, however, the previous
-cell information (usually limited range) is much more compact than the score(32-64 bits)
-\end{verbatim}
+\item {\color{red} Pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char), but only applies to input; score is hardly compressible if appropriate type is initially chosen}
+\item {\color{red} Operate on larger words (ex 64 bits) to increase thread locality and reduce memory accesses}
+\ole
 
@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \section*{Planning}

### 2013-01-07 docs/report/report.tex
Nits.

@@
 This project has been achieved in collaboration with Manohar Jonnalagedda.
-I also would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
 I hope you will enjoy your reading.
@@
 The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
-\item \textbf{Multi-dimensional matrices:} in the current implementation, all the matrices are encoded such that they are of the same size.
+\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.
 Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size).
@@
 
-{\color{red}
-\begin{verbatim}
-How to Encode multi-dimensional matrices efficiently
-1. assume they have the same type put one after another => different dimensions ok
-2. assume of same size => put into a struct
-=> but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)
-\end{verbatim}
-}
+Matrix of different dimensions must be stored in their own array (versus being in a single array of struct enclosing corresponding element of all matrices).
+Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description, there are two $O(n^2)$ matrices and one $O(n)$ matrix.
+This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once.
 
-\item \textbf{Independent computations:} {\color{red}The attentive reader might have noticed a discrepancy between the problem analysis and the recurrence rules of the Zuker problem implementation.
-This is due to the usage of multiple source for the problem description.
-}
-
-\item {\color{red} Pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char), but only applies to input; score is hardly compressible if appropriate type is initially chosen}
-\item {\color{red} Operate on larger words (ex 64 bits) to increase thread locality and reduce memory accesses}
+\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently.
+For example, RNA is represented with only 4 letters (g,a,t,c), thus 4 symbols could be encoded in a single byte.
+Unfortunately, this optimizations seems to only apply for the input data.
+Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, hereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
+\item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded.
+Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, hereby making it possible to prune entire computation tiles.
 \ole

### 2013-01-07 docs/report/tmp_gpu_rnafold.tex
RNAFold, but not working.

\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\pagestyle{empty} % remove page numbers

\begin{document}

$Q'_{i,j}$ is the minimum energy of folding of a subsequence $i,j$ given that bases $i$ and $j$ form a base pair; $Q_{i,j}$ and $QM_{i,j}$ are the minimum energy of folding of the subsequence $i,j$ assuming that this subsequence is inside a multiloop and that it contains respectively at least one and two base pairs. A simplified model of the recursion relations can be written as:

\[\begin{array}{rcl}
Q'_{i,j}&=&\left\{\begin{array}{ll}
	\min\left\{\begin{array}{l}
	Eh(i,j) \\
	Es(i,j)+Q'_{i+1,j-1} \\
	\min\limits_{i<k<l<j}Ei(i,j,k,l)+Q'_{k,l} \\
	QM_{i+1,j-1}
	\end{array}\right. & \text{if } (i,j) \text{ is a basepair}\\
	\infty & \text{otherwise} \\
	\end{array}\right. \\
QM_{i,j} &=& \min\limits_{i<k<j}(Q_{i,k}+Q_{k+1,j})\\
Q_{i,j} &=& \min\{ QM_{i,j}, Q_{i+1,j},Q_{i,j-1},Q'_{i,j} \}
\end{array}\]
\begin{itemize}
\item $Eh(i,j)$ energy of hairpin loop closed by the pair $i \cdot j$.
\item $Ei(i,j,k,l)$ energy of interior loop formed by two base pairs $i \cdot j$, $k \cdot l$.
\item $Es(i,j)$ energy of two stacked base pairs $i\cdot j$ and $(i+1)\cdot(j-1)$.
\end{itemize}

This converts to the grammar
\begin{verbatim}

Qp=tabulate("Qp",(
  seq            ^^ x => Eh(x._1,x._2-1)
| eli ~ Qp ~ eli ^^ (a,b,c) => Es(a,c) + b
| seq ~ Qp ~ seq ^^ (a,b,c) => Ei(a._1,b._2-1,a._2-1,b._1) + b
| el ~ QM  ~ el  ^^ (a,b,c) => b
) filter base aggregate h)

QM=tabulate("QM",(
  Q ~ Q ^^ (a,b)=>a+b
) aggregate h) // with size >= 4

Q=tabulate("Q",(
  QM
| el ~ Q ^^ (a,b) => b
| Q ~ el ^^ (a,b) => a
| Qp
) aggregate h) // with size >= 2

\end{verbatim}

\end{document}

### 2013-01-08 docs/report/report.tex
Nits.
Tabulations uses array in Scala.

@@
 \section{Benchmarks}
+\subsection{Java parsers}
+{\color{red}
+MatrixMult-512, Mac+JDK7\ul
+\item Original: 24.937 sec / 24.658 sec
+\item Optimized concatenation: 19.329 sec / 18.924 sec
+\item Tabulation as arrays+inline: 15.068 sec / 14.900 sec
+\ule}
+
+\subsection{CUDA parsers}
 {\color{red} XXX: explain why slower: approx 3x more memory used per score (maintain matrix dimensions)
@@
 Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, hereby making it possible to prune entire computation tiles.
+\item \textbf{Fusion:} {\color{red} pass the Scala grammar code through LMS to generate a more efficient version by using the fusion optimization}
 \ole

### 2013-01-10 docs/report/introduction.tex
Nits in report.
More report.

@@
 
-As the Scala programming language is initially developed by the LAMP, it seems natural to use it as host language for our project, however, we would list its features that makes it an interesting development language for this project:\ul
+As the Scala \cite{scala} programming language is initially developed by the LAMP, it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
 \item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.

### 2013-01-10 docs/report/problems.tex
More report.

@@
 For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.
+\ole
 
-\ole
+\subsection{Related work}
+Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.
+Grammar-based dynamic programming was initiated sparked by \cite{adp_discipline} and implemented as a Haskell DSL \cite{adp}.
+To overcome performance issues, multiple solutions were devised: converting the grammar to C or CUDA code \cite{adp_gpu} or using loop fusion to improve Haskell performance \cite{adp_fusion}.
+Ultimately, the dynamic programming algebra and grammar were formalized into an external DSL
+and ad-hoc compiler
+
+Related work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.

### 2013-01-10 docs/report/report.tex
Nits.
More report.

@@
 
-Existing work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
+Related work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
 % XXX: How we compare to other, how to evaluate
@@
 \item Optimized concatenation: 19.329 sec / 18.924 sec
-\item Tabulation as arrays+inline: 15.068 sec / 14.900 sec
+\item Tabulation as arrays+inline: 14.976 sec / 14.849 sec
 \ule}

### 2013-01-10 docs/report/introduction.tex
More related papers.

@@
 
-\subsubsection{LMS}
-Lightweight Modular Staging (LMS) is a runtime code generation framework built on top of Scala that uses only types to distinguish between code being transformed at compilation and at runtime.
+\subsubsection{Lightweight Modular Staging}
+Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.
+This framework can be extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime \cite{lms2}, \cite{lms3}, \cite{delite}.
+
  Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.

### 2013-01-10 docs/report/problems.tex
More related papers.

--- 2013-01-06 docs/report/problems.tex
@@
 For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.
+\ole
 
-\ole
+% ------------------------------------------------------------------------------------------------
+\newpage
+\subsection{Related work}
+Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.
+Grammar-based dynamic programming was initiated inseminated sparked by ADP \cite{adp_discipline} and first implemented as a Haskell DSL \cite{adp}.
+To overcome performance issues, multiple solutions were devised: converting the grammar to C or CUDA code \cite{adp_gpu} or using loop fusion to improve Haskell performance \cite{adp_fusion}, \cite{adp_fusion_pkg}.
+Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} and ad-hoc compiler \cite{gapc_thesis}, allowing more advanced analysis \cite{gapc_yield} of the grammar.
+
+The research on ad-hoc implementation have also been prolific by going from general DP techniques
+\cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl} to biology-specific applications \cite{nussinov_gpu}, \cite{gpu_rnafold}, \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.

### 2013-01-10 docs/report/report.tex
More related papers.

@@
 
-Related work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
+Related work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming-specific language written in C \cite{gapc_thesis} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
 % XXX: How we compare to other, how to evaluate
@@
 XXX: continue benchmarking here
+
+XXX: relate to staging operation in LMS
+
+XXX: include "RNAfold" grammar description into problems
+
+XXX: talk about memory management and hash tables
+XXX: relate to staging operation in LMS
+
+XXX: fix layout in codegen (newline missing)
+
 }

### 2013-01-10 src/librna/vienna/vienna.c

@@ -186,9 +186,9 @@ void read_parameter_file(const char *filename) {
   char        *line, ident[256];
   int         r;
 
-  if (!(fp=fopen(filename,"r"))) { fprintf(stderr,"\nread_parameter_file: can't open file %s. Using default parameters instead.\n", filename); return; }
-  if (!(line = get_line(fp))) { fprintf(stderr," File %s is inproper.\n", filename); fclose(fp); return; }
-  if (strncmp(line,"## RNAfold parameter file v2.0",30)!=0) { fprintf(stderr, "Missing header line in file. May be this file has not v2.0 format.\n"); }
+  if (!(fp=fopen(filename,"r"))) { fprintf(stderr,"Can't open '%s'. Using default parameters instead.\n", filename); return; }
+  if (!(line = get_line(fp))) { fprintf(stderr,"Parameter file '%s' is corrupted.\n", filename); fclose(fp); return; }
+  if (strncmp(line,"## RNAfold parameter file v2.0",30)!=0) { fprintf(stderr, "Missing v2.0 format header missing in '%s'.\n", filename); }
   free(line);
 
   while((line=get_line(fp))) {
@@ -234,7 +234,7 @@ void read_parameter_file(const char *filename) {
     else if IDENT("Hexaloops")                       rd_Hexaloop37();
     else if IDENT("Misc")                           { int values[4]; rd_1dim(&values[0],4,0); DuplexInit37=values[0]; DuplexInitdH=values[1]; TerminalAU37=values[2]; TerminalAUdH=values[3]; }
     else if IDENT("END")                            { free(line); break; }
-    else fprintf(stderr,"read_epars: Unknown field identifier in `%s'\n",line); /* do nothing but complain */
+    else fprintf(stderr,"Unknown field identifier in '%s'\n",line); /* do nothing but complain */
     free(line);
   }
   fclose(fp);

### 2013-01-10 docs/report/introduction.tex
Nits in build.sbt.

@@
 Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.
-This framework can be extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime \cite{lms2}, \cite{lms3}, \cite{delite}.
+This framework has been extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime (Delite) \cite{lms2}, \cite{lms3}, \cite{delite}.
 

### 2013-01-11 docs/report/implementation.tex
More report.

@@
 
-{\color{red} Currently we use {\tt Manifest}s and reflection to extract types, and convert their string representation into our restricted subset.
+Currently we use {\tt Manifest}s and reflection to extract types, and convert their string representation into our restricted subset.
 Manifests expands tuple inner types and reflection can be used to find class member's types.
 This imposes the additional restriction that we can not nest tuples into case classes, because generic types are then erased.
-However, the same effect could be achieved with Scala 2.10 {\tt TypeTag}s although we then need to rely on macros expansion to convert them properly into concrete classes.
-[see hint from Eugene, \url{https://gist.github.com/4407488}].}
+However, the same effect could be achieved with Scala 2.10 {\tt TypeTag}s, converting immediately to concrete type tree representation using macros expansion\footnote{Hint provided by Eugene Burmako, \url{https://gist.github.com/4407488}}.
 
@@
 \subsubsection{User functions generation} \label{user_fun_gen}
-{\color{red}
-\begin{verbatim}
-XXX: How LMS is used to generate function
-Possibly write that we can do the same with macros
-Why cannot use LMS for everything
-Where and how is LMS useful
-Also user function code generation with macros ?
-\end{verbatim}
-}
+The user generation function needs to be tightly integrated with the rest of the code generation.
+To do that, we need to establish a relation between the Scala function and its C counterpart.
+This is done by modifying the Scala function such that it embeds its C code and related types (input, output and possibly internal structures).
+To do that, LMS is used to generate both Scala and C code (as the user would want to write only once his function, using the corresponding LMS {\tt Rep} types).
+The two implementations are then mixed to provide the augmented Scala function that can then be used at appropriate places by either the Scala parsers or the code generator.
+
+Actually, the idea of mixing the two implementations into a single function emerged from experiments with the Scala macros \cite{scala_macros}, where it is possible to modify the AST of the Scala program before actually compiling it.
+Macros could also be an alternative to LMS in the sense that they have the same power in this particular case (because the code is just converted from Scala to C and does benefit of additional run-time information); however, relying only on macros would imply rewriting significant portions of code conversion, which might end up being a duplicated effort with LMS.
+The most interesting use of the macros would actually be to convert Scala types to LMS types in the <<context>> of user functions, hereby unleashing the power of LMS without forcing the DSL user to explicitly specify {\tt Rep} types\footnote{Since this is an ongoing project at LAMP with different schedule as this project, we do not want to duplicate effort currently but might integrate it at a later stage.}.
+
+Another advantage of using LMS only for user-specific function, is that it does not impose any restriction on the types manipulated by Scala, hereby providing the opportunity to solve the DP problem (possibly on CUDA using restricted types) and apply the solution (in Scala) on complex types that would have no representation in LMS.
 
@@
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
-XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+\subsection{LibRNA}
+Since the energy computation for RNA folding involve complex coefficient and computations (that are also standardized through libraries and coefficient files), we might want to provide the user with it.
+To do that, we based our library on \cite{gapc_thesis} which itself is based on \href{http://www.tbi.univie.ac.at/~ivo/RNA/}{ViennaRNA}.\cite{vienna_rna}.
+Since the library is provided in C, we rely on JNI to reuse the code without modifying it; this allows Scala to immediately benefit from it, but also makes possible to write a GPU version, provided that the related functions are simple enough to be expressible in CUDA.
 
-\subsection{LibRNA (?)}
-{\color{red} XXX}
+{\color{red}
+XXX: attempt to strip it to make CUDA version
+}

### 2013-01-11 docs/report/problems.tex
More report.

--- 2013-01-06 docs/report/problems.tex
@@
 \newpage
-\subsubsection{Zuker folding}\ol
-\item Problem: folding a RNA string $S$ over itself $|S|=n$.
+\subsubsection{Zuker RNA folding}\ol
+\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy.
 \item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
@@
 
-\item Visualizations\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
+\item Visualization\footnote{Reproductions of the illustrations from \cite{para_dprec} pp.148,149}:
 \begin{figure}[H]\begin{center}\includegraphics[width=8cm]{inc/zuker_rec.pdf}\end{center}\caption{Zuker folding dependencies}\end{figure}
@@
 Also notice that the $k$ of $BV$ and $BW$ describe almost the same backtrack, but there is an additional cost $c$ in $BV$.
-%{\color{red} XXX: We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?) }
 \ole
+
+\textbf{Alternative:} Since the recurrence matrices described in \cite{para_dprec} are of different dimensions ($F$ matrix is $O(n)$), we might want to use another description \cite{gpu_rnafold} where all matrices are of the same dimension, such that we can have a more uniform description across DP problems:
+
+Let $Q'_{(i,j)}$ the minimum energy of folding of a subsequence $i,j$ given that bases $i$ and $j$ form a base pair.
+$Q_{(i,j)}$ and $QM_{(i,j)}$ are the minimum energy of folding of the subsequence $i,j$ assuming that this subsequence is inside a multi-loop and that it contains respectively at least one and two base pairs.
+
+A simplified model of the recursion relations can be written as:
+\[\begin{array}{rcl}
+Q'_{(i,j)}&=&\left\{\begin{array}{ll}
+	\min\left\{\begin{array}{l}
+	Eh(i,j) \\
+	Es(i,j)+Q'_{i+1,j-1} \\
+	\min\limits_{i<k<l<j}Ei(i,j,k,l)+Q'_{k,l} \\
+	QM_{i+1,j-1}
+	\end{array}\right. & \text{if } (i,j) \text{ is a basepair}\\
+	\infty & \text{otherwise} \\
+	\end{array}\right. \\
+QM_{i,j} &=& \min\limits_{i<k<j}(Q_{i,k}+Q_{k+1,j})\\
+Q_{i,j} &=& \min\{ QM_{i,j}, Q_{i+1,j},Q_{i,j-1},Q'_{i,j} \}
+\end{array}\]
+
+\newpage
+The corresponding energy functions are:\ul
+\item $Eh(i,j)$ energy of hairpin loop closed by the pair $i \cdot j$.
+\item $Ei(i,j,k,l)$ energy of interior loop formed by two base pairs $i \cdot j$, $k \cdot l$.
+\item $Es(i,j)$ energy of two stacked base pairs $i\cdot j$ and $(i+1)\cdot(j-1)$.
+\ule
+
+This recurrence is more amenable to be converted into a grammar (in pseudo-code)
+\begin{lstlisting}[language=Scala,caption=Pseudocode of the alternative Zuker grammar]
+Qp : matrix(
+  ( seq            ^^ x => Eh(x)
+  | el  ~ Qp ~ el  ^^ (a,b,c) => Es(a,c) + b
+  | seq ~ Qp ~ seq ^^ (a,b,c) => Ei(a.start, a.end, c.start, c.end) + b
+  | el  ~ QM  ~ el ^^ (a,b,c) => b
+  ).filter(basepair).getMinimum)
+
+QM : matrix(
+  ( Q ~ Q ^^ (a,b) => a+b
+  ).filter(size>=4).getMinimum)
+
+Q : matrix(
+  ( QM
+  | el ~ Q ^^ (a,b) => b
+  | Q ~ el ^^ (a,b) => a
+  | Qp
+  ).filter(size>=2).getMinimum)
+
+\end{lstlisting}
+\begin{center}
+\vspace{-24pt}
+\footnotesize ({\tt seq} represent a nonempty sequence of elements and {\tt el} a single element)
+\end{center}
 
@@
 For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.
+\ole
 
-\ole
+% ------------------------------------------------------------------------------------------------
+\newpage
+\subsection{Related work}
+Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.
+The former focus on the performance for a specific problem whereas the latter generalize and formalize the dynamic programming problem description into a parsing grammar paired with a costing algebra.
+
+Grammar-based dynamic programming was inseminated by ADP \cite{adp_discipline} and first implemented as a Haskell DSL \cite{adp}. To overcome performance issues, multiple solutions were devised:\ul
+\item Converting Haskell parsers in their C or CUDA equivalent \cite{adp_gpu}
+\item Modifying Haskell execution environment to provide loop fusion to improve ADP parsers performance \cite{adp_fusion}, \cite{adp_fusion_pkg}.
+\item Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} provided with an ad-hoc compiler \cite{gapc_thesis}, hereby allowing more advanced analysis of the grammar \cite{gapc_yield}.
+\ule
+
+The research on ad-hoc implementation has focused on three kind of problems:\ul
+\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
+\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
+\item Biological sequence alignment (Smith-Watermann) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
+\ule

### 2013-01-11 docs/report/report.tex
More report.

@@
 \section{Benchmarks}
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
+XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
 \subsection{Java parsers}
@@
 This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once.
-
 \item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently.
@@
 Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, hereby making it possible to prune entire computation tiles.
-\item \textbf{Fusion:} {\color{red} pass the Scala grammar code through LMS to generate a more efficient version by using the fusion optimization}
+\item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
+Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser function calls into a single function making equivalent calls.
+The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
+\item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
+We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
 \ole
@@
 Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
-Jan. 7	& {\color{red} XXX} \\
+Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA {\color{red} XXX} \\
 Jan. 14	& {\color{red} XXX} \\

### 2013-01-11 docs/report/implementation.tex
More report.

@@
 Actually, the idea of mixing the two implementations into a single function emerged from experiments with the Scala macros \cite{scala_macros}, where it is possible to modify the AST of the Scala program before actually compiling it.
-Macros could also be an alternative to LMS in the sense that they have the same power in this particular case (because the code is just converted from Scala to C and does benefit of additional run-time information); however, relying only on macros would imply rewriting significant portions of code conversion, which might end up being a duplicated effort with LMS.
-The most interesting use of the macros would actually be to convert Scala types to LMS types in the <<context>> of user functions, hereby unleashing the power of LMS without forcing the DSL user to explicitly specify {\tt Rep} types\footnote{Since this is an ongoing project at LAMP with different schedule as this project, we do not want to duplicate effort currently but might integrate it at a later stage.}.
+Macros could also be an alternative to LMS in the sense that they have the same power in this particular case (because the code is just converted from Scala to C and does not benefit of additional run-time information); however, relying only on macros would imply rewriting significant portions of code conversion, which might end up being a duplicated effort with LMS.
+The most interesting use of the macros would actually be to stage plain Scala to its LMS representation in the <<context>> of user functions, hereby unleashing the power of LMS without forcing the DSL user to explicitly specify {\tt Rep} types\footnote{Since this is an ongoing project at LAMP with different schedule as this project, we do not want to duplicate effort currently but might integrate it at a later stage.}.
 
@@
 
+% ------------------------------------------------------------------------------------------------
 \subsection{Runtime execution engine}
@@
 
+% ------------------------------------------------------------------------------------------------
 \subsection{LibRNA}

### 2013-01-11 docs/report/problems.tex
More report.

@@
 \ule
+
+Considered as possible solutions to store a varying number of results per matrix cell were dynamic memory allocation \cite{nvidia} available on recent graphic cards, ad-hoc memory allocation \cite{scatter_alloc} or hash tables \cite{parallel_hashing}, however the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, also the use of cuckoo hash tables adds a constant factor penalty to every memory access.
+Finally both solution introduce the possibility of algorithm failure in the middle of the computation process.

### 2013-01-11 docs/report/report.tex
More report.

@@
 \section{Benchmarks}
+In an attempt to provide realistic benchmarks, we tried to gather related implementations.
+The authors of \cite{gpu_atlp} did not respond to our multiple solicitations.
+The authors of \cite{swat_mega} were very friendly and provided us their source code.
+Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
+
+Test 1 : Ad-hoc VS CUDA parsers (implies ATLP rescaled comparison)
+
+Test 2 : CUDAlign vs CUDA parsers for small sequences
+
+Test 3 : GAPC vs Java parsers vs CUDA parsers
+
+Test 4 ? : GPU RNAfolding vs CUDA parsers
+
 {\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
@@
 
-XXX: relate to staging operation in LMS
-
-XXX: include "RNAfold" grammar description into problems
-
-XXX: talk about memory management and hash tables
-XXX: relate to staging operation in LMS
-
-XXX: fix layout in codegen (newline missing)
-
+XXX: talk about memory management and hash tables (?)
 }
@@
 We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
+\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
+Hence cost and backtrack matrix would contain $k$ elements per cell.
+Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
 \ole

### 2013-01-11 docs/report/report.tex
Stripped circular out of RNAfold.

@@
 Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
+Finally, we asked lately the authors of \cite{gpu_rnafold} {\color{red} who did not respond so far}.
+
+We organize the benchmarks as follow: if we have at our disposal a working implementation that could be run on our evaluation platform, we use it, otherwise, we refer to the related paper and rescale the result according to the memory throughput and computation throughput of the related device so that we can have a good approximation of how they could compare.
+
+\subsection{Metrics}
+As the device memory is quite limited, it seems interesting to take into account the space usage.
+The space requirement limits the maximal size of addressable problems  on a particular hardware.
+This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently.
+However, simple solutions like using a device with larger memory or using main memory (if $5\times$ slowdown is still acceptable) could solve this issue.
+
+memory
+
+running time
+
+\subsection{Benchmarking platform}
 

### 2013-01-11 src/librna/rnafold/part_func.c

@@ -37,11 +37,9 @@ PRIVATE FLT_OR_DBL  *q=NULL, *qb=NULL, *qm=NULL, *qm1=NULL, *qqm=NULL, *qqm1=NUL
 PRIVATE FLT_OR_DBL  *probs=NULL, *prml=NULL, *prm_l=NULL, *prm_l1=NULL, *q1k=NULL, *qln=NULL;
 PRIVATE FLT_OR_DBL  *scale=NULL;
 PRIVATE FLT_OR_DBL  *expMLbase=NULL;
-PRIVATE FLT_OR_DBL  qo=0., qho=0., qio=0., qmo=0., *qm2=NULL;
 PRIVATE int         *jindx=NULL;
 PRIVATE int         *my_iindx=NULL;
 PRIVATE int         init_length = -1;   /* length in last call to init_pf_fold() */
-PRIVATE int         circular=0;
 PRIVATE int         do_bppm = 1;             /* do backtracking per default */
 PRIVATE int         struct_constrained = 0;
 PRIVATE char        *pstruc=NULL;
@@ -59,13 +57,11 @@ PRIVATE void  init_partfunc(int length, pf_paramT *parameters);
 PRIVATE void  scale_pf_params(unsigned int length, pf_paramT *parameters);
 PRIVATE void  get_arrays(unsigned int length);
 PRIVATE void  make_ptypes(const short *S, const char *structure);
-PRIVATE void  pf_circ(const char *sequence, char *structure);
 PRIVATE void  pf_linear(const char *sequence, char *structure);
 PRIVATE void  pf_create_bppm(const char *sequence, char *structure);
 PRIVATE void  backtrack(int i, int j);
 PRIVATE void  backtrack_qm(int i, int j);
 PRIVATE void  backtrack_qm1(int i,int j);
-PRIVATE void  backtrack_qm2(int u, int n);
 
 /*
 #################################
@@ -103,8 +99,7 @@ PRIVATE void get_arrays(unsigned int length){
   q     = (FLT_OR_DBL *) space(size);
   qb    = (FLT_OR_DBL *) space(size);
   qm    = (FLT_OR_DBL *) space(size);
-  qm1   = (st_back || circular) ? (FLT_OR_DBL *) space(size) : NULL;
-  qm2   = (circular) ? (FLT_OR_DBL *) space(sizeof(FLT_OR_DBL)*(length+2)) : NULL;
[…]

### 2013-01-11 src/main/scala/v4/examples/RNAFold.scala

@@ -11,7 +11,7 @@ trait RNAFoldSig extends Signature {
   type Alphabet = Char
 
   def Eh(i:Int,j:Int) = LibRNA.hl_energy(i,j) // hairpin loop
-  def Ei(i:Int,k:Int,l:Int,j:Int) = LibRNA.il_energy(i,k,l,j) // internal loop XXX isn't it i-j __enclosing__ basepair and k-l inner basepair
+  def Ei(i:Int,k:Int,l:Int,j:Int) = LibRNA.il_energy(i,k,l,j) // internal loop
   def Es(i:Int,j:Int) = LibRNA.sr_energy(i,j) // 2 stacked base pairs
 
   def hairpin(ij:(Int,Int)):Answer
@@ -40,7 +40,7 @@ trait RNAFoldPrettyPrint extends RNAFoldSig {
   private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
   def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
   def stack(i:Int,s:String,j:Int) = "("+s+")"
-  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-ik._1-1)+s+dots(lj._2-lj._1-1)+")"
+  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
   def mloop(i:Int,s:String,j:Int) = "("+s+")"
   def left(l:String,r:Int) = l+"."
   def right(l:Int,r:String) = "."+r

### 2013-01-11 docs/report/report.tex
Nits, fixed cudalign.

@@
 \input{inc/style.sty}
+\usepackage{multicol}
+\usepackage{etoolbox}
 
@@
 \subsection{Metrics}
-As the device memory is quite limited, it seems interesting to take into account the space usage.
+{\color{red} The main metric we focus on is the running time.}
+
+As the device memory is quite limited, it seems interesting to also take into account the space usage.
 The space requirement limits the maximal size of addressable problems  on a particular hardware.
 This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently.
-However, simple solutions like using a device with larger memory or using main memory (if $5\times$ slowdown is still acceptable) could solve this issue.
-
-memory
-
-running time
+However, simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue.
 
@@
 
-Test 1 : Ad-hoc VS CUDA parsers (implies ATLP rescaled comparison)
-
-Test 2 : CUDAlign vs CUDA parsers for small sequences
-
-Test 3 : GAPC vs Java parsers vs CUDA parsers
-
-Test 4 ? : GPU RNAfolding vs CUDA parsers
+Candidates : 
+- CUDASW++
+- SeqNFind
 
@@
 XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+\subsubsection{Test 1 : Ad-hoc VS CUDA parsers (implies ATLP rescaled comparison)}
+
+\subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}
+
+\subsubsection{Test 3 : GAPC vs Java parsers vs CUDA parsers}
+
+\subsubsection{Test 4 ? : GPU RNAfolding vs CUDA parsers}
 
@@
 XXX: continue benchmarking here
-
-XXX: talk about memory management and hash tables (?)
 }
@@
 Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
-Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. \\
+Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. ADP (Manohar) \\
 Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
@@
 Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
-Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA {\color{red} XXX} \\
+Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
 Jan. 14	& {\color{red} XXX} \\
@@
 \newpage
-%\usepackage{multicol}
-%\usepackage{etoolbox}
-%\patchcmd{\thebibliography}{\section*{\refname}}{\begin{multicols}{2}[\section*{\refname}]}{}{}
-%\patchcmd{\endthebibliography}{\endlist}{\endlist\end{multicols}}{}{}
-%\bibliographystyle{acm}
-\bibliographystyle{plain}
-\bibliography{bibliography.bib}
+\patchcmd{\thebibliography}{\section*{\refname}}{\begin{multicols}{2}[\section*{\refname}]}{}{}
+\patchcmd{\endthebibliography}{\endlist}{\endlist\end{multicols}}{}{}
+\bibliographystyle{plain} % acm
+{\small \bibliography{bibliography.bib}}
 \end{document}

### 2013-01-11 src/librna/rnafold/libs/data_structures.h

@@ -1,114 +1,0 @@
-#ifndef __VIENNA_RNA_PACKAGE_DATA_STRUCTURES_H__
-#define __VIENNA_RNA_PACKAGE_DATA_STRUCTURES_H__
-
-#include "vienna.h"
-#include "energy_const.h"
-
-/**
- *  \file data_structures.h
- *  \brief All datastructures and typedefs shared among the Vienna RNA Package can be found here
- */
-
-/* to use floats instead of doubles in pf_fold() comment next line */
-#define LARGE_PF
-#ifdef  LARGE_PF
-#define FLT_OR_DBL double
-#else
-#define FLT_OR_DBL float
-#endif
-
-#ifndef MAXALPHA
-/**
- *  \brief Maximal length of alphabet
- */
-#define MAXALPHA              20
-#endif
-
-/**
- *  \brief Maximum density of states discretization for subopt
- */
-#define MAXDOS                1000
-
-/*
-* ############################################################
-* Here are the type definitions of various datastructures
-* shared among the Vienna RNA Package
-* ############################################################
-*/
-
-/**
- *  \brief this datastructure is used as input parameter in functions of PS_dot.h and others
- */
-typedef struct plist {
-  int i;
-  int j;
-  float p;
-} plist;
-
-/**
- *  stack of partial structures for backtracking
- */
-typedef struct sect {
-  int  i;
-  int  j;
-  int ml;
-} sect;
-
-/**
- *  base pair
- */
-typedef struct bondT {
-   unsigned int i;
-   unsigned int j;
-} bondT;
-
-/**
- *  The datastructure that contains temperature scaled Boltzmann weights of the energy parameters.
- */
-typedef struct{
-  int     id;
-  double  expstack[NBPAIRS+1][NBPAIRS+1];
-  double  exphairpin[31];
-  double  expbulge[MAXLOOP+1];
-  double  expinternal[MAXLOOP+1];
-  double  expmismatchExt[NBPAIRS+1][5][5];
-  double  expmismatchI[NBPAIRS+1][5][5];
-  double  expmismatch23I[NBPAIRS+1][5][5];
-  double  expmismatch1nI[NBPAIRS+1][5][5];
-  double  expmismatchH[NBPAIRS+1][5][5];
-  double  expmismatchM[NBPAIRS+1][5][5];
-  double  expdangle5[NBPAIRS+1][5];
-  double  expdangle3[NBPAIRS+1][5];
[…]

### 2013-01-11 src/librna/rnafold/libs/fold_vars.h

@@ -1,193 +1,0 @@
-#ifndef __VIENNA_RNA_PACKAGE_FOLD_VARS_H__
-#define __VIENNA_RNA_PACKAGE_FOLD_VARS_H__
-
-#include "data_structures.h"
-
-/**
- *  \file fold_vars.h
- *  \brief Here all all declarations of the global variables used throughout RNAlib
- */
-
-
-#define PUBLIC
-#define PRIVATE static
-
-/**
- *  \brief Global switch to avoid/allow helices of length 1
- *
- *  Disallow all pairs which can only occur as lonely pairs (i.e. as helix
- *  of length 1). This avoids lonely base pairs in the predicted structures in
- *  most cases.
- */
-extern int    noLonelyPairs;
-
-/**
- *  \brief Switch the energy model for dangling end contributions (0, 1, 2, 3)
- *
- *  If set to 0 no stabilizing energies are assigned to bases adjacent to
- *  helices in free ends and multiloops (so called dangling ends). Normally
- *  (dangles = 1) dangling end energies are assigned only to unpaired
- *  bases and a base cannot participate simultaneously in two dangling ends. In
- *  the partition function algorithm pf_fold() these checks are neglected.
- *  If #dangles is set to 2, all folding routines will follow this convention.
- *  This treatment of dangling ends gives more favorable energies to helices
- *  directly adjacent to one another, which can be beneficial since such
- *  helices often do engage in stabilizing interactions through co-axial
- *  stacking.\n
- *  If dangles = 3 co-axial stacking is explicitly included for
- *  adjacent helices in mutli-loops. The option affects only mfe folding
- *  and energy evaluation (fold() and energy_of_structure()), as
- *  well as suboptimal folding (subopt()) via re-evaluation of energies.
- *  Co-axial stacking with one intervening mismatch is not considered so far.
- *
- *  Default is 2 in most algorithms, partition function algorithms can only handle 0 and 2
- */
-extern int  dangles;
-
-/**
- *  \brief Global switch to forbid/allow GU base pairs at all
- */
-extern int  noGU;
-
-/**
- *  \brief GU allowed only inside stacks if set to 1
- */
-extern int  no_closingGU;
-
-/**
- *  \brief Include special stabilizing energies for some tri-, tetra- and hexa-loops;
- *
- *  default is 1.
- */
-extern int  tetra_loop;
-
-/**
- *  \brief 0 = BP; 1=any mit GC; 2=any mit AU-parameter
- *
[…]

### 2013-01-11 src/librna/rnafold/part_func.c

@@ -1,907 +1,0 @@
-/*
-                  partiton function for RNA secondary structures
-
-                  Ivo L Hofacker
-                  Vienna RNA package
-*/
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-#include <math.h>
-#include <float.h>    /* #defines FLT_MAX ... */
-#include <limits.h>
-
-#include "utils.h"
-#include "energy_par.h"
-#include "fold_vars.h"
-#include "pair_mat.h"
-#include "params.h"
-#include "loop_energies.h"
-#include "part_func.h"
-
-#define ISOLATED  256.0
-
-/*
-#################################
-# GLOBAL VARIABLES              #
-#################################
-*/
-PUBLIC  int         st_back = 0;
-
-/*
-#################################
-# PRIVATE VARIABLES             #
-#################################
-*/
-PRIVATE FLT_OR_DBL  *q=NULL, *qb=NULL, *qm=NULL, *qm1=NULL, *qqm=NULL, *qqm1=NULL, *qq=NULL, *qq1=NULL;
-PRIVATE FLT_OR_DBL  *probs=NULL, *prml=NULL, *prm_l=NULL, *prm_l1=NULL, *q1k=NULL, *qln=NULL;
-PRIVATE FLT_OR_DBL  *scale=NULL;
-PRIVATE FLT_OR_DBL  *expMLbase=NULL;
-PRIVATE int         *jindx=NULL;
-PRIVATE int         *my_iindx=NULL;
-PRIVATE int         init_length = -1;   /* length in last call to init_pf_fold() */
-PRIVATE int         do_bppm = 1;             /* do backtracking per default */
-PRIVATE char        *pstruc=NULL;
-PRIVATE char        *sequence=NULL;
-PRIVATE char        *ptype=NULL;        /* precomputed array of pair types */
-PRIVATE pf_paramT   *pf_params=NULL;    /* the precomputed Boltzmann weights */
-PRIVATE short       *S=NULL, *S1=NULL;
-
-/*
-#################################
-# PRIVATE FUNCTION DECLARATIONS #
-#################################
-*/
-PRIVATE void  init_partfunc(int length, pf_paramT *parameters);
-PRIVATE void  scale_pf_params(unsigned int length, pf_paramT *parameters);
[…]

### 2013-01-11 src/librna/rnafold/utils.c

@@ -1,0 +1,249 @@
+/*
+                               utils.c
+
+                 c  Ivo L Hofacker and Walter Fontana
+                          Vienna RNA package
+*/
+/* Last changed Time-stamp: <2008-11-25 16:34:36 ivo> */
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <ctype.h>
+#include <errno.h>
+#include <time.h>
+#include <string.h>
+#include "utils.h"
+
+#define PRIVATE  static
+#define PUBLIC
+
+/*@notnull@ @only@*/
+PUBLIC unsigned short xsubi[3];
+
+PRIVATE char  *inbuf = NULL;
+PRIVATE char  *inbuf2 = NULL;
+PRIVATE unsigned int typebuf2 = 0;
+
+/*-------------------------------------------------------------------------*/
+
+void *xrealloc (void *p, unsigned size) {
+  if (p == 0) return space(size);
+  p = (void *) realloc(p, size);
+  if (p == NULL) {
+    fprintf(stderr,"xrealloc: requested size: %d\n", size);
+    nrerror("xrealloc allocation failure -> no memory");
+  }
+  return p;
+}
+
+/*------------------------------------------------------------------------*/
+
+PUBLIC void nrerror(const char message[])       /* output message upon error */
+{
+  fprintf(stderr, "ERROR: %s\n", message);
+  exit(EXIT_FAILURE);
+}
+
+/*------------------------------------------------------------------------*/
+
+PRIVATE char *get_line(FILE *fp) /* reads lines of arbitrary length from fp */
+{
+  char s[512], *line, *cp;
+  int len=0, size=0, l;
+  line=NULL;
+  do {
+    if (fgets(s, 512, fp)==NULL) break;
+    cp = strchr(s, '\n');
+    if (cp != NULL) *cp = '\0';
+    l = len + (int)strlen(s);
+    if (l+1>size) {
+      size = (int)((l+1)*1.2);
+      line = (char *) xrealloc(line, size*sizeof(char));
+    }
+    strcat(line+len, s);
+    len=l;
+  } while(cp==NULL);
+
+  return line;
+}
+
+PRIVATE  unsigned int get_multi_input_line(char **string, unsigned int option){
+  char  *line;
+  int   i, l;
+  int   state = 0;
+  int   str_length = 0;
+
+  line = (inbuf) ? inbuf : get_line(stdin);
+  inbuf = NULL;
+  do{
+
+    /*
[…]

### 2013-01-12 docs/report/introduction.tex
Beginning of intro rework.

@@
+% abstract = resume the paper
+% introduction
+%
+
+% Paper introduction:
+% Problem to solve, what exists (related work) and how do we compare to other
+% Contributions (3): 3 tensed sentences
+% Benchmarks => evaluation metrics, prove introduction statements by evaluation
+\subsection*{Abstract}
+Dynamic programming is a common pattern of Computer Science; yet underlying matrix recurrences might be difficult to express, error prone and efficient parallel implementation might be difficult.
+In this project, we present \textit{DynaProg}, a Scala DSL for dynamic programming on heterogeneous platforms which allow to write concise programs based on a pair of parsing grammar and algebra and execute them either on CPU or on GPU.
+
+We then evaluate the performance of both CPU and GPU implementations against existing work and ad-hoc implementations of our own.
+Experimental results show a average speedup of {\color{red} XXX} if run on GPU (compared to CPU), {\color{red} is comparable to GAPC?} on CPU and has a slowdown of {\color{red} XXX} versus ad-hoc problem implementations on GPU {\color{red} (ours, cudalign, rnafold?)}.
+
+\vfill
+This project has been achieved in collaboration with Manohar Jonnalagedda.
+I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I hope you will enjoy your reading.
+\vspace{.3cm}\\
+\textit{Thierry Coppey}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\setcounter{tocdepth}{2} \tableofcontents
 \newpage
 \section{Introduction}
+\textbf{Difficulties with dynamic programming:} Dynamic programming (DP) is a technique used to solve combinatorial optimization problems.
+Often, it allows to evaluate a search space of exponential size in polynomial time.
+Variations of the basic algorithm not only return an optimal solution, but may also report co- or near-optimal solutions, or compute synthetic properties of the search space such as its size or the sum of all scores.
+
+In several areas of applied Computer Science, such as operational research, natural language processing, or biosequence analysis, dynamic programming problems arise in many variations and with a considerable degree of sophistication.
+Hence there is a mismatch between the simple textbook algorithms and the real-world implementation of their variants.
+Writing correct recurrences is difficult and error prone because it needs a lot of attention to get correct indices, and implementation debugging is tedious.
+Additionally, little changes in the computational process might imply large rewrites of the matrices and recurrences.
+
+For efficiency reasons, the computation is split in two: a matrix elements computation followed by a backtracking stage to retrieve the solution associated with the optimal score.
+The backtrack information gathering depends on the matrix design and the final result needs also to be presented in a form that makes sense for the user.
+This might add an additional source of errors and tedious change, would the initial recurrence be changed.
+
+Finally, once the implementation is correct, it is possible to turn it into efficient code for many core architectures such as multi-CPU, GPU or implementation on programmable hardware (FPGA).
+However, the domain specialist who write the recurrences might not be very familiar with these platforms, whereas parallelization experts might not deeply understand the domain of the dynamic programming recurrences.
+
+\textbf{Separating concerns:} Algebraic Dynamic Programming (ADP) \cite{adp} is a language-independent declarative approach that separate the concerns of dynamic programming algorithms over data sequences:\ul
+\item Search space definition (using a tree parsing grammar)
+\item Candidates scoring and optimization objective (with mapping and aggregation functions)
+\item Tabulation of the result in matrices (determines running time and space efficiency)
+\ule
+
+By the use of a parsing grammar, ADP makes the invisible candidate structure explicit.
+An additional signature serves as interface between the grammar and the scoring algebra and aggregation function, which makes possible that different grammars share different algebra or vice versa.
+Tabulation issues are hidden from the programmer, thus there are no subscripts any more hereby removing this source of errors.
+
+Finally, since the expression of the dynamic program is formalized, it makes it possible to convert it to efficient recurrences to target many-core platforms such as GPUs.
+
+\textbf{DynaProg system:} DynaProg implements ADP with a syntax matching that of the Scala parsers.
+
+{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
+XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+
+% XXX: How we compare to other, how to evaluate
+% Benchmark => prove by evaluation intro statements
+
+Our contributions are: \ul
+\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
+\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
+\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
+\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
+%\item Reuse of existing compiler technology (fusion) for a specific purpose
+%\item State of the art parallel implementation of these classes on GPUs
+%\item Normalization of the grammar into efficient productions
+%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
+\ule
+
 % ------------------------------------------------------------------------------------------------
+\section{Background}
 \subsection{Dynamic programming}
@@
 \newpage
-\subsection{Scala and LMS}
-\subsubsection{Scala}
+\subsection{Scala}
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
@@
 
-\subsubsection{Lightweight Modular Staging}
+% ------------------------------------------------------------------------------------------------
+\subsection{Lightweight Modular Staging}
 Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.

### 2013-01-12 docs/report/report.tex
Beginning of intro rework.

@@
 \maketitle
-%\shorttitle
-
-% Paper introduction:
-% Problem to solve, what exists (related work) and how do we compare to other
-% Contributions (3): 3 tensed sentences
-% Benchmarks => evaluation metrics, prove introduction statements by evaluation
-\subsection*{Abstract}
-Dynamic programming is a common pattern of Computer Science used in various domains.
-Yet underlying matrix recurrences might be difficult to express and error prone.
-Additionally, domain experts might not have the skills to make an efficient parallel implementation.
-In this project, we present \textit{DynaProg}, a Scala DSL for dynamic programming on heterogeneous platforms which allow to write concise programs and execute them efficiently on GPUs.
-
-Related work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming-specific language written in C \cite{gapc_thesis} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
-% XXX: How we compare to other, how to evaluate
-% Benchmark => prove by evaluation intro statements
-
-Our contributions are: \ul
-%\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
-\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
-\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
-\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
-%\item Reuse of existing compiler technology (fusion) for a specific purpose
-%\item State of the art parallel implementation of these classes on GPUs
-%\item Normalization of the grammar into efficient productions
-%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
-\ule
-
-\vfill
-This project has been achieved in collaboration with Manohar Jonnalagedda.
-I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
-I hope you will enjoy your reading.
-\vspace{.3cm}\\
-\textit{Thierry Coppey}
-
-\newpage
-\setcounter{tocdepth}{2} \tableofcontents
-
 \input{introduction}

### 2013-01-13 docs/report/introduction.tex
Reworked introduction.

@@
 
-\textbf{DynaProg system:} DynaProg implements ADP with a syntax matching that of the Scala parsers.
+\textbf{DynaProg framework:} DynaProg implements ADP in Scala as an embedded DSL (domain-specific language) with a syntax matching that of the existing parsers.
+It extends over the original ADP parsers by providing multi-track grammars, simplifies the combinators by providing yield analysis, and can translates programs into CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} parsers, that are competitive to handwritten code, and arguably more reliable.
+It performs additional optimization for correct parallel scheduling and removal of unused tabulations (dead code elimination).
 
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
-XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+DynaProg provides a generic way of backtracking the parsers, such that the same backtrack can be used with multiple algebras, provided that they use the same grammar.
+This allows to construct a generic pattern <<solve-compute>> of solving problem: for example, assume that the problem consists of multiplying a chain of matrices efficiently.
+In the first step, the used would want to find the optimal execution schedule of the multiplication using dynamic programming.
+In the second step, the computed optimal solution will be used to execute the actual computation on actual data, hereby producing effortlessly the desired result: it suffice to provide an additional computation algebra.
 
-% XXX: How we compare to other, how to evaluate
-% Benchmark => prove by evaluation intro statements
+Offloading dynamic programming computations to CUDA devices is achieved effortlessly, providing that the used has correctly configured few variables to match its system configuration: it suffices to add two lines of code (mixing a trait and providing two data types) to schedule the dynamic compilation and execution of the CUDA optimized program and obtain the result, as if it was executed in plain Scala.
 
-Our contributions are: \ul
-\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
-\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
-\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
-\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
-%\item Reuse of existing compiler technology (fusion) for a specific purpose
-%\item State of the art parallel implementation of these classes on GPUs
-%\item Normalization of the grammar into efficient productions
-%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
+In the rest of the document we\ul
+\item Briefly describe dynamic programming background and some interesting features of the Scala language
+\item Provide a classification of DP problems characteristics (matrix shape, dependency graph) and provide a detailed analysis of some related problems that we want to implement.
+\item Propose a systematic approach to encode backtracking information such that the backtracking process can be made linear to the size of the problem
+\item Describe how to convert parsers efficiently in parallel code from a top-down description to a bottom-up implementation (focusing on running time and memory efficiency)
+\item Provide an concrete implementation in the form of a language embedded DSL in Scala, leveraging the grammar and algebra concepts of ADP
+\item Describe two implementations: Scala for CPU (focusing on multiple backtracking) and an CUDA for GPU (focusing on efficiency)
+\item Evaluate the performance of our work by providing appropriate benchmarks against existing implementations
 \ule
 
+%Our contributions are: \ul
+%\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
+%\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
+%\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
+%\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
+%%\item Reuse of existing compiler technology (fusion) for a specific purpose
+%%\item State of the art parallel implementation of these classes on GPUs
+%%\item Normalization of the grammar into efficient productions
+%%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
+%\ule
+
 % ------------------------------------------------------------------------------------------------
-\section{Background}
+\newpage
+%\section{Background}
 \subsection{Dynamic programming}
@@
 
-As the Scala \cite{scala} programming language is initially developed by the LAMP, it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
+As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
 \item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.
@@
 Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.
-This framework has been extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime (Delite) \cite{lms2}, \cite{lms3}, \cite{delite}.
-
+This makes possible to annotate parts of the code with special types, such that their compilation is delayed until the program is executed.
+At run time, these parts are represented as multiple nodes that serve as the basis for another compilation phase where all the code executed until this point can provide additional information to produce a more efficient compilation.
+The process of delaying the compilation is known as \textit{lifting} whereas \textit{lowering} corresponds to transforming this intermediate representation into executable code.
 Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.
 
-LMS can be leveraged to transform Scala code into its C-like equivalent.
-However, the concern in this project is that the code for the GPU would be sensibly different from the original CPU code as both implementation serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
-Hence the use of LMS would be restricted to user-specific function, over which our DSL has no control.
+Delite \cite{lms2}, \cite{lms3}, \cite{delite} leverages the benefits of delayed compilation provided by LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
+For instance, this can be used to transform embarrassingly parallel operations on collections into their parallel counterpart that can be executed on GPU.
+
+At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
+However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
+Ad-hoc C translation seems more appropriate than writing Scala code and convert it with LMS because LMS only understand a subset of both languages.
+
+LMS would still be helpful for generating user-specific functions, as they are independent of the rest of the program and user wants to write his functions only once for both Scala and CUDA.

### 2013-01-13 src/main/scala/v4/examples/Zuker.scala

@@ -28,7 +28,6 @@ trait ZukerSig extends Signature {
   def ssadd(e:SSeq, x:Answer) : Answer
   def nil(d:Unit) : Answer
 
-  def size:Int
   def in(x:Int):Alphabet
   def basepairing(i:Int, j:Int):Boolean = if (i+2>j) false else (in(i),in(j-1)) match {
     case ('a','u') | ('u','a') | ('u','g') | ('g','u') | ('g','c') | ('c','g') => true
@@ -41,7 +40,7 @@ trait ZukerMFE extends ZukerSig {
   type Answer = Int
   def sadd(lb:Int, e:Answer) = e
   def cadd(x:Answer, e:Answer) = x + e
-  def dlr(lb:Int, e:Answer, rb:Int) = e + LibRNA.ext_mismatch_energy(lb, rb-1, size) + LibRNA.termau_energy(lb, rb-1) // OK
+  def dlr(lb:Int, e:Answer, rb:Int) = e + LibRNA.ext_mismatch_energy(lb, rb-1) + LibRNA.termau_energy(lb, rb-1) // OK
   def sr(lb:Int, e:Answer, rb:Int) = e + LibRNA.sr_energy(lb,rb) // OK
   def hl(lb:Int, f1:Int, x:SSeq, f2:Int, rb:Int) = LibRNA.hl_energy(f1,f2) + LibRNA.sr_energy(lb, rb) // OK
   def bl(lb:Int, f1:Int, b:SSeq, x:Answer, f2:Int, rb:Int) = x + LibRNA.bl_energy(f1,b._1,b._2-1,f2,f2-1) + LibRNA.sr_energy(lb, rb) // OK

### 2013-01-13 docs/report/implementation.tex
Nits in librna.

@@
 \textbf{Bonus: problems larger than device memory}\\
-Problems larger than the device memory can actually be processed on recent CUDA devices (architecture $\ge 2.0$) as these are able to address the main memory from the device.
-However, since the distance between CUDA processors and memory is increased, there is a $5\times$ slowdown penalty to be paid in this configuration (experimentally, on a $1024\times 1024$ triangular matrix).
+Problems larger than the device memory can actually be processed on recent CUDA devices (with CUDA architecture $\ge 2.0$) as these are able to address the main memory from the device.
+However, since the distance between CUDA processors and memory is increased, there is an approximate $5\times$ slowdown penalty to be paid in this configuration (experimentally, on a $1024\times 1024$ triangular matrix).
 Nevertheless, this workaround implementation has 2 benefits:\ul
 \item It allows larger problem to be solved, with very little implementation effort, would the user be patient enough for the computation to terminate
-\item It provides a good estimation of the main memory usage penalty, and hereby a strong argument in favor of the implementation described in \ref{ns_mem_transfer} (with only up to $\le 0.3\%$ transfer overhead).
+\item It provides a good estimation of the main memory usage penalty, and hereby a strong argument in favor of the implementation described in \ref{ns_mem_transfer} (with lest than 1\% overhead due to transfers).
 However, since we have not found concrete applications with such matrix size, the benefit of supporting large matrices is unclear, hence we leave the optimal implementation for future work.

### 2013-01-13 docs/report/problems.tex
Reworked introduction.
Nits in librna.

@@
 \subsection{Problems classification}
+Since <<dynamic programming>> defines a very general technique, we already focused on grammar and alignment problems in \ref{intro_dp}.
+Before exploring some particular problem instances, we want to define some characteristics that will be used through the rest of the document to describe dynamic programming problems.
+
 \subsubsection{Definitions}\ul
+\item \textbf{Cost or score:} refers to the result of the dynamic programming recurrence formula.
+\item \textbf{Backtrack:} the backtrack is the information related to a score that describe how it has been obtained by referring to immediately previous elements.
+By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
 \item \textbf{Alphabets:} the alphabet represent an enumeration of the possible values.
@@
 \subsection{Related problems}
-The aim of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems.
-(There is an hyperlink on the problems name to their detailed description).
+The aim of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems\footnote{There are hyperlinks on the problems name to their detailed description.}.
 
@@
 \subsubsection{Other problems}\ul
-\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
-\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
+\item Dijkstra shortest path: can be expressed in DP and requires a $E\times V$ matrix.
+Informally: along $E$, forall $V$, reduce the distance.
+The problem is serial along the $E$ dimension and non-serial along $V$, hence its complexity is $O(|E|\cdot |V^2|)$ which is worse than both $O(|V|^2)$ (using a minimum priority queue) and $O(|E|+|V|\log |V|)$ (with Fibonacci heap).
+\item Fibonacci numbers: this problem is serial 1D (in 1 dimension). $F(n)$ could be implemented with ADP using a sequence of $n$ placeholder elements, but this is inefficient.
 \item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
@@
 \item There is a difference between dynamic programming as seen in algorithmic classes and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
-\item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences). Most of the common dynamic programming problems fall in these two categories.
+\item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences).
+Most of the interesting dynamic programming problems that could be parallelized fall in these two categories.
 \item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
@@
 Considered as possible solutions to store a varying number of results per matrix cell were dynamic memory allocation \cite{nvidia} available on recent graphic cards, ad-hoc memory allocation \cite{scatter_alloc} or hash tables \cite{parallel_hashing}, however the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, also the use of cuckoo hash tables adds a constant factor penalty to every memory access.
-Finally both solution introduce the possibility of algorithm failure in the middle of the computation process.
+Finally both solution introduce undesirable possibility of failure (respectively out of memory or unrecoverable collision) in the middle of the algorithm computation process.

### 2013-01-14 docs/report/design.tex
Rework of intro.

@@
 \newpage
-\section{Architecture design and technical decisions}
+\section{Architecture design and technical decisions} \label{architecture}
 % ------------------------------------------------------------------------------------------------
@@
 \newpage
-\subsection{Parsing grammar (ADP)}
+\subsection{Parsing grammar (ADP)} \label{adp_grammar}
 In this section, we provide a concise description of the Algebraic Dynamic Programming.
@@
 %\newpage
-\subsection{Recurrences analysis}
+\subsection{Recurrences analysis} \label{recurrences}
 \subsubsection{Dead rules elimination}
@@
 
-\subsubsection{Yield analysis}
+\subsubsection{Yield analysis} \label{yield_analysis}
 Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved.
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Backtracking}
+\subsection{Backtracking} \label{backtracking}
 In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Memory constraints}
+\subsection{Memory constraints} \label{memory_constr}
 We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.

### 2013-01-14 docs/report/introduction.tex
Nits in librna.
Rework of intro.

--- 2013-01-10 docs/report/introduction.tex
@@
+% abstract = resume the paper
+% introduction
+%
+
+% Paper introduction:
+% Problem to solve, what exists (related work) and how do we compare to other
+% Contributions (3): 3 tensed sentences
+% Benchmarks => evaluation metrics, prove introduction statements by evaluation
+\subsection*{Abstract}
+Dynamic programming is a common pattern of Computer Science to solve problem that follow the Bellman's principle: optimal solutions depends on optimal sub-problem solutions.
+Solving a dynamic programming problem decomposes in two phases: filling one or more matrices with sub-problems intermediate results and decomposing how the final result was constructed (backtracking).
+Yet matrices computation formulae and recurrences might be difficult to express, indexing element is often error prone and writing an efficient parallel implementation can take a significant amount of time.
+In this project, we present \textit{DynaProg}, a language embedded in Scala (DSL) to address dynamic programming problems on heterogeneous platforms.
+DynaProg allows the programmer to write concise programs based on a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.
+
+We evaluate the performance of our implementation against existing work and own ad-hoc implementations for both the CPU and GPU versions of DynaProg.
+
+Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU).
+Compared to ad-hoc GPU implementations, the generated parsers have an average slowdown of {\color{red} XXX}.
+
+The CPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,GAPC?}.
+The GPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,CudAlign, RNAFold}.
+
+\vfill
+This project has been achieved in collaboration with Manohar Jonnalagedda.
+I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I hope you will enjoy your reading.
+\vspace{.3cm}\\
+\textit{Thierry Coppey}
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\setcounter{tocdepth}{2} \tableofcontents
 \newpage
 \section{Introduction}
+Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle: optimal solutions depends on optimal solutions of sub-problem.
+Dynamic programming often allows to find solutions in an exponential search space (number of solutions) in polynomial running time.
+The user is usually interested in one optimal solution of the problem, but he might also desire to have \textit{all} optimal solutions (co-optimal), a fixed number of near-optimal solutions, or some synthetic properties of the search space such as its size or the sum of all scores.
+
+Dynamic programming problems arise in several areas of applied Computer Science such as biosequence analysis, natural language processing and operational research.
+Unfortunately, they appear in multiple variations, and often with a considerable degree of sophistication such that there is a mismatch between textbook solutions and the real-world implementation.
+Additionally, implementation debugging is tedious and require a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences.
+\cite{gapc_yield}
+
+For efficiency reasons, the process of dynamic programming computation is split in two: first intermediate scores are memoized in a matrix (tabulation), and reused to construct scores for larger problems; then a backtrack stage retrieve the solution associated with the optimal score for the problem.
+This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
+The obtained information then needs to be presented in a form that makes sense for the user (or possibly might be used to drive further computations).
+Each of these two phases must be tightly correlated as a divergence would be a source of errors; would the recurrence formulae be even slightly changed, the programmer must ensure that both phase are still synchronized, which might be tedious.
+
+Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA).
+However, the domain specialist who write the recurrences might not be very familiar with these platforms, whereas parallelization and hardware experts might not deeply understand the domain of the dynamic programming recurrences.
+
+To solve these concerns, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming algorithms into four distinct components that are tightly connected:\ol
+\item The search space is described using a context-free \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
+\item Constructed candidates are then evaluated by a \textbf{scoring function} (where all these functions form an \textbf{algebra}), so that they can be compared appropriately.
+\item The optimization objective (which candidates to retain as possible solutions) is described using an \textbf{aggregation function} that operates on the scores previously obtained.
+\item Finally, results are \textbf{tabulated} (memoized in an array) in the corresponding matrices.
+The tabulation regulates the trade-off between running time and space efficiency by memoizing appropriate results that are reused multiple times.
+\ole
+
+By the use of a parsing grammar, ADP makes the candidate structure explicit.
+An additional signature serves as interface between the grammar, the scoring algebra and the aggregation function which makes possible that different grammars share different algebra or vice versa.
+Tabulation indices issues are hidden from the programmer, hereby removing one source of errors.
+
+Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and algebra, it becomes possible to convert it to efficient recurrences for many-core platforms such as GPUs.
+\cite{adp_gpu}
+
+% -----------------------------
+DynaProg implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax matching similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
+It extends over original ADP by allowing grammars for pairing two sequences (multi-track grammars), simplifies the process of writing program by inferring additional informations (see \ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart.
+Since the program is formalized, it can be analyzed to remove unused grammar productions (dead code elimination) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, hereby producing an arguably more reliable program.
+
+DynaProg provides a generic way of backtracking the results, such that the same backtrack trace can be used with multiple algebras if they share the same grammar.
+This allows to construct a two steps pattern for solving problems: first DP problem is solved, then from optimal solution backtrack, the desired result is computed.
+For example, assume the problem of multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently.
+In the first step, optimal execution scheduling (or parenthesization) is found using dynamic programming.
+At the second step, the backtrack trace of optimal solution is used to multiply the actual matrices (given the corresponding algebra).
+
+Finally, offloading dynamic programming computations to CUDA devices has been made effortless for the programmer: it suffices to enable code generation to schedule dynamic compilation and execution of the GPU-optimized program, as if it was executed in plain Scala.
+
+This project resulted is an open-source\footnote{\url{https://github.com/manojo/lamp-dp-mt}} implementation of dynamic programming parsers for Scala optimized for both on CPU and GPU and featuring several analysis (see section~\ref{architecture}) to ease the writing of dynamic programs.
+Its contribution is an automated approach to encode and process backtracking information such that reconstruction complexity is reduced and backtrack trace be exchanged among different algebras sharing the same grammar.
+
+The rest of the document consists of:\ul
+\item A brief description of dynamic programming background (\ref{intro_dp}), followed by an explanation of some of the Scala programming language (\ref{intro_scala}) and LMS framework (\ref{intro_lms}) features.
+\item Section~\ref{problems} proposes a classification of DP problems in terms of matrix shape and dependencies.
+A detailed analysis of some specific problems is provided.
+Related work addressing the dynamic programming challenges is then presented (\ref{related_work}).
+\item In section~\ref{architecture} we describe the whole parser stack in an abstract manner, going from the user facing language (\ref{user_lang}, \ref{adp_grammar}) to optimizations (\ref{recurrences}, \ref{backtracking}) and implementation constraints (\ref{normalization}, \ref{memory_constr}), describing all the architectural decisions we made.
+\item Section~\ref{implementation} describes how these idea are concretely implemented in the form of a DSL for Scala (\ref{scala_parsers}) and in particular how is efficient CUDA code generated (\ref{codegen})
+\item Finally, in section~\ref{benchmarks} we evaluate the performance of our work by providing appropriate benchmarks against existing implementations
+\ule
+
+%\item Propose a systematic approach to encode backtracking information such that the backtracking process can be made linear to the size of the problem
+%\item Provide an concrete implementation in the form of a language embedded DSL in Scala, leveraging the grammar and algebra concepts of ADP
+%\item Describe two implementations: Scala for CPU (focusing on multiple backtracking) and an CUDA for GPU (focusing on efficiency)
+
+%The contributions of this project are: \ul
+%\item A classification of dynamic programming problems characteristics in terms of matrix shape and recurrence formulae dependencies.
+%\item A systematic approach to convert a top-down recurrence description (grammar) into efficient bottom-up 
+%\item A systematic approach to process backtracking information (focus on running time and memory efficiency)
+%\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
+%\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
+%%\item Reuse of existing compiler technology (fusion) for a specific purpose
+%%\item State of the art parallel implementation of these classes on GPUs
+%%\item Normalization of the grammar into efficient productions
+%%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
+%\ule
+
 % ------------------------------------------------------------------------------------------------
-\subsection{Dynamic programming}
+\newpage
+%\section{Background}
+\subsection{Dynamic programming} \label{intro_dp}
 Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
@@
 \newpage
-\subsection{Scala and LMS}
-\subsubsection{Scala}
+\subsection{Scala} \label{intro_scala}
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
@@
 
-\subsubsection{Lightweight Modular Staging}
+{\color{red} talk of features: traits, ... }
+
+% ------------------------------------------------------------------------------------------------
+\subsection{Lightweight Modular Staging} \label{intro_lms}
 Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.
-This framework has been extended to generate from the same source code efficient implementation for heterogeneous platforms at runtime (Delite) \cite{lms2}, \cite{lms3}, \cite{delite}.
-
+This makes possible to annotate parts of the code with special types, such that their compilation is delayed until the program is executed.
+At run time, these parts are represented as multiple nodes that serve as the basis for another compilation phase where all the code executed until this point can provide additional information to produce a more efficient compilation.
+The process of delaying the compilation is known as \textit{lifting} whereas \textit{lowering} corresponds to transforming this intermediate representation into executable code.
 Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.
 
-LMS can be leveraged to transform Scala code into its C-like equivalent.
-However, the concern in this project is that the code for the GPU would be sensibly different from the original CPU code as both implementation serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
-Hence the use of LMS would be restricted to user-specific function, over which our DSL has no control.
+{\color{red} move delite to related work ... }
+LMS code generation is not limited to Scala, it can also target other languages.
+To automate code generation and execution flow, Delite \cite{lms2}, \cite{lms3}, \cite{delite} leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
+This can be used to transform operations on collections into efficient parallel GPU implementation.
+
+At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
+However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
+Ad-hoc C translation seems more appropriate than writing Scala code and convert it with LMS because LMS only understand a subset of both languages.
+
+LMS would still be helpful for generating user-specific functions, as they are independent of the rest of the program and user wants to write his functions only once for both Scala and CUDA.

### 2013-01-14 docs/report/problems.tex
Rework of intro.

@@
 \newpage
-\subsection{Related work}
+\subsection{Related work} \label{related_work}
 Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.

### 2013-01-14 docs/report/report.tex
Nits in librna.
Rework of intro.

@@
 \input{implementation}
-\section{Benchmarks}
+\section{Benchmarks} \label{benchmarks}
 In an attempt to provide realistic benchmarks, we tried to gather related implementations.
@@
 
-Candidates : 
-- CUDASW++
-- SeqNFind
-
-{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ 
-XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
+% ------------------------------------------------------------------------------------------------
+{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+% ------------------------------------------------------------------------------------------------
 
@@
 \subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}
+Also CUDASW++
 

### 2013-01-14 docs/report/design.tex
Nits.

@@
 
-\item \textbf{Optimizations:} An possible optimization is to break down complex rules into simpler ones if they are expressible so; hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
+\item \textbf{Optimizations:} An possible optimization is to break down complex rules into simpler ones if they are expressible so; thereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
 Unfortunately this analysis is very involved: we need to solve the following problem:
@@
 Additionally, the authors are planning to write extensions to support some variants of this problem like Needleman-Wunsch.
-\item The implementation sensibly differs from the class of small problems, as the solving strategy is completely different from non-serial small problems, hereby requiring larger development time that would be out of the scope of this project.
+\item The implementation sensibly differs from the class of small problems, as the solving strategy is completely different from non-serial small problems, thereby requiring larger development time that would be out of the scope of this project.
 \item Finally, such implementation would be only valuable for problems that are larger than the memory device, whereas smaller problem could perfectly use the existing implementation.
@@
 However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle).
-The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, hereby dramatically reducing the size of the additional triangles.
+The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles.
 Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe.

### 2013-01-14 docs/report/implementation.tex
Rework of intro.
Nits.

@@
 \newpage
-\section{Implementation}
+\section{Implementation} \label{implementation}
 % ------------------------------------------------------------------------------------------------
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Scala parsers}
+\subsection{Scala parsers} \label{scala_parsers}
 The Scala parsers consist in 4 traits that are used to construct a DSL program:\ul
@@
 	\item Minimum and maximum yield size: functions evaluated recursively except for tabulations where value is attributed in the yield analysis phase.
-	\item Number of inner alternatives: helps counting alternatives, hereby guaranteeing an unique number for each (provided that parsers obtain non-overlapping ranges).
+	\item Number of inner alternatives: helps counting alternatives, thereby guaranteeing an unique number for each (provided that parsers obtain non-overlapping ranges).
 	\item Number of inner moving concatenations: helps determining required storage for the backtrack as well as retrieving the appropriate index in the backtrack phase
@@
 \item It allows larger problem to be solved, with very little implementation effort, would the user be patient enough for the computation to terminate
-\item It provides a good estimation of the main memory usage penalty, and hereby a strong argument in favor of the implementation described in \ref{ns_mem_transfer} (with lest than 1\% overhead due to transfers).
+\item It provides a good estimation of the main memory usage penalty, and thereby a strong argument in favor of the implementation described in \ref{ns_mem_transfer} (with lest than 1\% overhead due to transfers).
 However, since we have not found concrete applications with such matrix size, the benefit of supporting large matrices is unclear, hence we leave the optimal implementation for future work.
@@
 By induction on rows and columns, it suffice to have the last column and row element valid.
-Since line is computed by the current thread (hereby valid), it only remains to guarantee that the column element of the previous line is valid (in figure \ref{fig:diag_deps}, previous refers to the line immediately below).
+Since line is computed by the current thread (thereby valid), it only remains to guarantee that the column element of the previous line is valid (in figure \ref{fig:diag_deps}, previous refers to the line immediately below).
 To do that, each block writes last valid diagonal in a <<lock>> array, and next block need only to wait (polling) until desired element is marked valid.
-Notice that {\tt \_\_threadfence} is not mandatory (hereby slightly improving performance), verifying the observation of \cite{gpu_barrier}.
+Notice that {\tt \_\_threadfence} is not mandatory (thereby slightly improving performance), verifying the observation of \cite{gpu_barrier}.
 
@@
 Parsers generation is independent of user-defined function generation (see \ref{user_fun_gen}).
-Tabulation inner parsers are first wrapped in additional aggregation (by $h$, hereby ensuring they produce at most one result) and normalized (according to \ref{normalization}); code generation then occurs recursively, producing a list of loops and conditions, and body (possibly with a hoisted part).
+Tabulation inner parsers are first wrapped in additional aggregation (by $h$, thereby ensuring they produce at most one result) and normalized (according to \ref{normalization}); code generation then occurs recursively, producing a list of loops and conditions, and body (possibly with a hoisted part).
 Additionally, position variables are maintained and subrule index and concatenation indices are propagated.
@@
 
-Since code is generated, it is possible to write the last step using a switch case, hereby flattening the writes in the vector (compared to recursive calls in Scala).
+Since code is generated, it is possible to write the last step using a switch case, thereby flattening the writes in the vector (compared to recursive calls in Scala).
 Finally, since the trace has to be reversed, we can obtain this transformation for free by constructing the trace list from the end in the JNI conversion.
@@
 Macros could also be an alternative to LMS in the sense that they have the same power in this particular case (because the code is just converted from Scala to C and does not benefit of additional run-time information); however, relying only on macros would imply rewriting significant portions of code conversion, which might end up being a duplicated effort with LMS.
-The most interesting use of the macros would actually be to stage plain Scala to its LMS representation in the <<context>> of user functions, hereby unleashing the power of LMS without forcing the DSL user to explicitly specify {\tt Rep} types\footnote{Since this is an ongoing project at LAMP with different schedule as this project, we do not want to duplicate effort currently but might integrate it at a later stage.}.
+The most interesting use of the macros would actually be to stage plain Scala to its LMS representation in the <<context>> of user functions, thereby unleashing the power of LMS without forcing the DSL user to explicitly specify {\tt Rep} types\footnote{Since this is an ongoing project at LAMP with different schedule as this project, we do not want to duplicate effort currently but might integrate it at a later stage.}.
 
-Another advantage of using LMS only for user-specific function, is that it does not impose any restriction on the types manipulated by Scala, hereby providing the opportunity to solve the DP problem (possibly on CUDA using restricted types) and apply the solution (in Scala) on complex types that would have no representation in LMS.
+Another advantage of using LMS only for user-specific function, is that it does not impose any restriction on the types manipulated by Scala, thereby providing the opportunity to solve the DP problem (possibly on CUDA using restricted types) and apply the solution (in Scala) on complex types that would have no representation in LMS.
 

### 2013-01-14 docs/report/introduction.tex
Nits.

@@
-% abstract = resume the paper
-% introduction
-%
+\subsection*{Abstract}
+Dynamic programming is an algorithmic technique to solve problems that follow the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal sub-problem solutions.
+The core idea behind dynamic programming is to memoize intermediate results into matrices to avoid multiple computations.
+Solving a dynamic programming problem consists of two phases: filling one or more matrices with intermediate solutions for sub-problems and recomposing how the final result was constructed (backtracking).
+In the textbooks, problems are usually described in terms of recurrence relations between matrices elements.
+However, matrices computation formulae and recurrences might be difficult to express, indexing element is often error prone and writing an efficient parallel implementation can take a significant amount of time.
+
+In this project, we present \textit{DynaProg}, a language embedded in Scala (DSL) to address dynamic programming problems on heterogeneous platforms.
+DynaProg allows the programmer to write concise programs based on ADP \cite{adp}, using a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.
+We evaluate the performance of our implementation against existing work and own ad-hoc implementations for both the CPU and GPU versions.
+Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU).
+Compared to ad-hoc GPU implementations, the generated parsers have an average slowdown of {\color{red} XXX}.
+The CPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,GAPC?}.
+The GPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,CudAlign, RNAFold}.
 
@@
 % Benchmarks => evaluation metrics, prove introduction statements by evaluation
-\subsection*{Abstract}
-Dynamic programming is a common pattern of Computer Science to solve problem that follow the Bellman's principle: optimal solutions depends on optimal sub-problem solutions.
-Solving a dynamic programming problem decomposes in two phases: filling one or more matrices with sub-problems intermediate results and decomposing how the final result was constructed (backtracking).
-Yet matrices computation formulae and recurrences might be difficult to express, indexing element is often error prone and writing an efficient parallel implementation can take a significant amount of time.
-In this project, we present \textit{DynaProg}, a language embedded in Scala (DSL) to address dynamic programming problems on heterogeneous platforms.
-DynaProg allows the programmer to write concise programs based on a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.
-
-We evaluate the performance of our implementation against existing work and own ad-hoc implementations for both the CPU and GPU versions of DynaProg.
-
-Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU).
-Compared to ad-hoc GPU implementations, the generated parsers have an average slowdown of {\color{red} XXX}.
-
-The CPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,GAPC?}.
-The GPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,CudAlign, RNAFold}.
 
@@
 This project has been achieved in collaboration with Manohar Jonnalagedda.
-I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advice and suggestions.
 I hope you will enjoy your reading.
@@
 \section{Introduction}
-Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle: optimal solutions depends on optimal solutions of sub-problem.
+Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
 Dynamic programming often allows to find solutions in an exponential search space (number of solutions) in polynomial running time.
@@
 
-Dynamic programming problems arise in several areas of applied Computer Science such as biosequence analysis, natural language processing and operational research.
-Unfortunately, they appear in multiple variations, and often with a considerable degree of sophistication such that there is a mismatch between textbook solutions and the real-world implementation.
-Additionally, implementation debugging is tedious and require a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences.
-\cite{gapc_yield}
+Dynamic programming problems arise in several disciplines of applied Computer Science such as biosequence analysis, natural language processing and operational research: sequence alignment, RNA sequence folding or expression parenthesisation are example of such problems which are usually described by matrix recurrence relations.
+Unfortunately, these problems often appear in multiple variations and with a considerable degree of sophistication such that there is a mismatch between the textbook solution and its concrete implementation.
+Additionally, debugging is tedious and requires a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences\cite{gapc_yield}.
 
-For efficiency reasons, the process of dynamic programming computation is split in two: first intermediate scores are memoized in a matrix (tabulation), and reused to construct scores for larger problems; then a backtrack stage retrieve the solution associated with the optimal score for the problem.
+Dynamic programming computation is usually split in two phases: first intermediate scores are memoized in a matrix (tabulation), and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
 This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
-The obtained information then needs to be presented in a form that makes sense for the user (or possibly might be used to drive further computations).
-Each of these two phases must be tightly correlated as a divergence would be a source of errors; would the recurrence formulae be even slightly changed, the programmer must ensure that both phase are still synchronized, which might be tedious.
+The obtained information then needs to be presented in a form that makes sense for the user (or  might drive further computations).
+These two phases need to be kept consistent with each other, thereby introducing a new potential source of errors.
 
 Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA).
-However, the domain specialist who write the recurrences might not be very familiar with these platforms, whereas parallelization and hardware experts might not deeply understand the domain of the dynamic programming recurrences.
+However, a domain specialist who writes the recurrences might not be very familiar with these platforms, whereas parallelization and hardware experts might not deeply understand the domain of the dynamic programming recurrences.
 
+% http://en.wikipedia.org/wiki/Regular_tree_grammar: language class between regular languages and the deterministic context-free languages.
 To solve these concerns, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming algorithms into four distinct components that are tightly connected:\ol
-\item The search space is described using a context-free \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
+\item The search space is described using a tree \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
 \item Constructed candidates are then evaluated by a \textbf{scoring function} (where all these functions form an \textbf{algebra}), so that they can be compared appropriately.
-\item The optimization objective (which candidates to retain as possible solutions) is described using an \textbf{aggregation function} that operates on the scores previously obtained.
-\item Finally, results are \textbf{tabulated} (memoized in an array) in the corresponding matrices.
-The tabulation regulates the trade-off between running time and space efficiency by memoizing appropriate results that are reused multiple times.
+\item The \textbf{objective function} (or aggregation function) operates on the scores previously obtained to retain valid candidates.
+\item Finally, results are \textbf{tabulated} (memoized in an array) in corresponding matrices.
+Tabulation process regulates the trade-off between running time and space efficiency by memoizing appropriate results that are reused multiple times.
 \ole
 
-By the use of a parsing grammar, ADP makes the candidate structure explicit.
+By using a parsing grammar, ADP makes the candidate structure explicit.
 An additional signature serves as interface between the grammar, the scoring algebra and the aggregation function which makes possible that different grammars share different algebra or vice versa.
-Tabulation indices issues are hidden from the programmer, hereby removing one source of errors.
-
+Tabulation indices issues are hidden from the programmer, thereby removing one source of errors.
 Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and algebra, it becomes possible to convert it to efficient recurrences for many-core platforms such as GPUs.
@@
 % -----------------------------
-DynaProg implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax matching similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
-It extends over original ADP by allowing grammars for pairing two sequences (multi-track grammars), simplifies the process of writing program by inferring additional informations (see \ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart.
-Since the program is formalized, it can be analyzed to remove unused grammar productions (dead code elimination) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, hereby producing an arguably more reliable program.
+DynaProg implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
+It extends ADP by allowing grammars for pairing two sequences (multi-track grammars) similarly as GAPC\cite{gapc_thesis}, simplifies the process of writing programs by inferring additional information (\S\ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart (\S\ref{benchmarks}).
+Since the program structure is formalized in ADP framework, it can be analyzed to remove unused grammar productions (dead code elimination) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, thereby producing an arguably more reliable program.
 
 DynaProg provides a generic way of backtracking the results, such that the same backtrack trace can be used with multiple algebras if they share the same grammar.
-This allows to construct a two steps pattern for solving problems: first DP problem is solved, then from optimal solution backtrack, the desired result is computed.
-For example, assume the problem of multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently.
-In the first step, optimal execution scheduling (or parenthesization) is found using dynamic programming.
-At the second step, the backtrack trace of optimal solution is used to multiply the actual matrices (given the corresponding algebra).
+This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost function; then from the backtrack of its optimal, the desired result is computed.
+As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently.
+In the first step, optimal execution scheduling (or parenthesization) is found using dynamic programming (\S\ref{mat_mult_plain}).
+In the second step, the backtrack trace of optimal solution is used to multiply the actual matrices (given the corresponding algebra).
 
@@
 
-This project resulted is an open-source\footnote{\url{https://github.com/manojo/lamp-dp-mt}} implementation of dynamic programming parsers for Scala optimized for both on CPU and GPU and featuring several analysis (see section~\ref{architecture}) to ease the writing of dynamic programs.
-Its contribution is an automated approach to encode and process backtracking information such that reconstruction complexity is reduced and backtrack trace be exchanged among different algebras sharing the same grammar.
+This project resulted is an open-source\footnote{\url{https://github.com/manojo/lamp-dp-mt}} implementation of dynamic programming parsers for Scala optimized for both on CPU and GPU and featuring several analyses (\S\ref{architecture}) to ease the writing of dynamic programs.
+Its contribution is an automated approach to encode and process backtracking information such that the reconstruction complexity is reduced compared to \cite{gapc_thesis} and backtrack trace can be exchanged among different algebras sharing the same grammar.
 
 The rest of the document consists of:\ul
-\item A brief description of dynamic programming background (\ref{intro_dp}), followed by an explanation of some of the Scala programming language (\ref{intro_scala}) and LMS framework (\ref{intro_lms}) features.
-\item Section~\ref{problems} proposes a classification of DP problems in terms of matrix shape and dependencies.
-A detailed analysis of some specific problems is provided.
-Related work addressing the dynamic programming challenges is then presented (\ref{related_work}).
-\item In section~\ref{architecture} we describe the whole parser stack in an abstract manner, going from the user facing language (\ref{user_lang}, \ref{adp_grammar}) to optimizations (\ref{recurrences}, \ref{backtracking}) and implementation constraints (\ref{normalization}, \ref{memory_constr}), describing all the architectural decisions we made.
-\item Section~\ref{implementation} describes how these idea are concretely implemented in the form of a DSL for Scala (\ref{scala_parsers}) and in particular how is efficient CUDA code generated (\ref{codegen})
-\item Finally, in section~\ref{benchmarks} we evaluate the performance of our work by providing appropriate benchmarks against existing implementations
+\item A brief background on dynamic programming (\S\ref{intro_dp}), followed by an explanation of some of the Scala programming language (\S\ref{intro_scala}) and LMS framework (\S\ref{intro_lms}) features.
+\item A classification of DP problems in terms of matrix shape and dependencies in section~\ref{problems}, followed by a detailed analysis of some specific problems is provided.
+Related work addressing the dynamic programming challenges is presented in \ref{related_work}.
+\item A description of the whole parser stack (\S\ref{architecture}), going from the user facing language (\S\ref{user_lang}, \S\ref{adp_grammar}) to optimizations (\S\ref{recurrences}, \S\ref{backtracking}) and implementation constraints (\S\ref{normalization}, \S\ref{memory_constr}), describing all the architectural decisions we made.
+\item The concrete implementation of these ideas (\S\ref{implementation}) in the form of a DSL for Scala (\S\ref{scala_parsers}) and in efficient CUDA code generation (\S\ref{codegen}).
+\item An evaluation of the performance of our work (\S\ref{benchmarks}) by providing appropriate benchmarks against existing implementations.
 \ule
@@
 \item Scala is largely adopted in the industry, which makes both the adoption of related project easier and offer a steeper learning curve to their potential users.
-\item Finally, through the Java VM and JNI interface, Scala offers the possibility to load dynamically external libraries, to leverage best underlying hardware by mixing with CUDA kernels to obtain optimal performance.
+\item Since Scala programs execute in the JVM, they can benefit of the JNI, that is the possibility to dynamically load C libraries and interact with them, possibly leveraging  underlying hardware by launching CUDA kernels.
+\item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and mixed together, thereby allowing multiple inheritance.
 \ule
-
-{\color{red} talk of features: traits, ... }
 
@@
 This makes possible to annotate parts of the code with special types, such that their compilation is delayed until the program is executed.
-At run time, these parts are represented as multiple nodes that serve as the basis for another compilation phase where all the code executed until this point can provide additional information to produce a more efficient compilation.
+At run time, these parts are represented as a \textit{sea of nodes} that serve as the basis for another compilation phase where all the code executed until this point provides additional information to produce a more efficient compilation.
 The process of delaying the compilation is known as \textit{lifting} whereas \textit{lowering} corresponds to transforming this intermediate representation into executable code.
-Through extensive use of component technology, lightweight modular staging makes an optimizing compiler framework available at the library level, allowing programmers to tightly integrate domain-specific abstractions and optimizations into the generation process.
-
-{\color{red} move delite to related work ... }
-LMS code generation is not limited to Scala, it can also target other languages.
-To automate code generation and execution flow, Delite \cite{lms2}, \cite{lms3}, \cite{delite} leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
-This can be used to transform operations on collections into efficient parallel GPU implementation.
+LMS code generation is not limited to Scala, it can also target other languages like C.
+In short, LMS is an optimizing compiler framework that allows integration of domain-specific abstractions and optimizations into the generation process.
 

### 2013-01-14 docs/report/problems.tex
Nits.

@@
 \item Modifying Haskell execution environment to provide loop fusion to improve ADP parsers performance \cite{adp_fusion}, \cite{adp_fusion_pkg}.
-\item Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} provided with an ad-hoc compiler \cite{gapc_thesis}, hereby allowing more advanced analysis of the grammar \cite{gapc_yield}.
+\item Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} provided with an ad-hoc compiler \cite{gapc_thesis}, thereby allowing more advanced analysis of the grammar \cite{gapc_yield}.
 \ule
@@
 Finally both solution introduce undesirable possibility of failure (respectively out of memory or unrecoverable collision) in the middle of the algorithm computation process.
+
+Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
+Although our project shares many patterns with this project we can not reuse it because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{intro_lms}) whereas this framework focuses on parallelizing operations on independent elements.

### 2013-01-14 docs/report/report.tex
Nits.

@@
 \input{implementation}
+
+\section{Usage}
+{\color{red} usage, make reference from problems to usage
+
+XXX: do not introduce pseudocode but refere to usage in RNAfold problem
+
+}
+
 \section{Benchmarks} \label{benchmarks}
@@
 \subsection{Benchmarking platform}
+
+RUN 5-10x until running time stabilizes on Scala
 
@@
 We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
-\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, hereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
+\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
 However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
@@
 Unfortunately, this optimizations seems to only apply for the input data.
-Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, hereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
+Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
 \item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded.
-Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, hereby making it possible to prune entire computation tiles.
+Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, thereby making it possible to prune entire computation tiles.
 \item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.

### 2013-01-15 docs/report/design.tex
More report.

@@
 % ------------------------------------------------------------------------------------------------
+\subsection{LMS integration} \label{lms_use}
+At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
+However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
+
+Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
+\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}.
+We tried to use LMS to generate both the Scala and CUDA program.
+This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}), and this approach also limits the functionality of Scala parsers (as only a subset of the Scala language can be lifted into LMS).
+
+\item A second possibility would be to analyze the (user) Scala program and produce another Scala program (simplified) in the restricted subset that is compatible with LMS, hereby providing some ad-hoc transformations either during the conversion (user to simplified), or at a later stage during the optional optimization phase of LMS.
+This approach seems quite complicated and error prone as it introduces multiple sources of possible errors: during the ad-hoc conversion (user to simplified), during the optimization phase, and during the conversion to C code.
+Additionally, since dynamic programming require special scheduling to respect the dependencies between intermediate solutions, we would need at a later stage to combine ad-hoc scheduling code with the result of the LMS code generation.
+\item Since ADP normalizes the structure of dynamic programming, the concerns between grammar and algebra are well separated.
+Since the grammar has a limited number of parsers, it is possible to provide an ad-hoc implementation for each of them thereby allowing better integration.
+Given that the algebra can contain arbitrary used-defined functions, we avoid duplicating the efforts and rely on LMS to generate the corresponding C code.
+\ule
+For all the previous reasons it is more simple to provide an ad-hoc code generation for the grammar, whereas the algebra can be generated by LMS, hereby avoiding to duplicate code generation efforts while at the same time allowing us to integrate tightly the grammar parsers into the surrounding scheduling C code.
+
+% ------------------------------------------------------------------------------------------------
 \subsection{Compilation stack}

### 2013-01-15 docs/report/introduction.tex
More report.

@@
 By using a parsing grammar, ADP makes the candidate structure explicit.
-An additional signature serves as interface between the grammar, the scoring algebra and the aggregation function which makes possible that different grammars share different algebra or vice versa.
-Tabulation indices issues are hidden from the programmer, thereby removing one source of errors.
+A signature serves as interface between the grammar, the scoring algebra and the aggregation function making possible that different grammars share different algebra or vice versa.
+Tabulation indices issues are hidden from the programmer, thereby removing potential errors.
 Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and algebra, it becomes possible to convert it to efficient recurrences for many-core platforms such as GPUs.
@@
 This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost function; then from the backtrack of its optimal, the desired result is computed.
-As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently.
-In the first step, optimal execution scheduling (or parenthesization) is found using dynamic programming (\S\ref{mat_mult_plain}).
-In the second step, the backtrack trace of optimal solution is used to multiply the actual matrices (given the corresponding algebra).
+As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently: first, optimal execution scheduling (or parenthesization) trace is found using dynamic programming and cost algebra (\S\ref{mat_mult_plain}).
+The backtrack trace is then used (with a multiplication algebra) to multiply the actual matrices.
 
@@
 The rest of the document consists of:\ul
-\item A brief background on dynamic programming (\S\ref{intro_dp}), followed by an explanation of some of the Scala programming language (\S\ref{intro_scala}) and LMS framework (\S\ref{intro_lms}) features.
-\item A classification of DP problems in terms of matrix shape and dependencies in section~\ref{problems}, followed by a detailed analysis of some specific problems is provided.
-Related work addressing the dynamic programming challenges is presented in \ref{related_work}.
+\item A brief background on dynamic programming, followed by an explanation of some of the Scala programming language and LMS framework features (\S\ref{background}).
+\item A classification of DP problems in terms of matrix shape and dependencies, followed by a detailed analysis of some specific problems is provided (\S\ref{problems}).
+Related work addressing dynamic programming challenges is presented in (\S\ref{related_work}).
 \item A description of the whole parser stack (\S\ref{architecture}), going from the user facing language (\S\ref{user_lang}, \S\ref{adp_grammar}) to optimizations (\S\ref{recurrences}, \S\ref{backtracking}) and implementation constraints (\S\ref{normalization}, \S\ref{memory_constr}), describing all the architectural decisions we made.
 \item The concrete implementation of these ideas (\S\ref{implementation}) in the form of a DSL for Scala (\S\ref{scala_parsers}) and in efficient CUDA code generation (\S\ref{codegen}).
-\item An evaluation of the performance of our work (\S\ref{benchmarks}) by providing appropriate benchmarks against existing implementations.
+\item A brief usage explanation detailing the available features for the DSL user (\S\ref{usage}).
+\item An evaluation of the performance of our work by providing appropriate benchmarks against existing implementations (\S\ref{benchmarks}).
 \ule
@@
 \newpage
-%\section{Background}
-\subsection{Dynamic programming} \label{intro_dp}
-Dynamic programming consists of solving a problem by reusing subproblems solutions. A famous example of dynamic programming is the Fibonacci series that is defined by the recurrence
+\section{Background} \label{background}
+\subsection{Dynamic programming} \label{bg_dp}
+Dynamic programming consists of solving a problem by constructing its solution from solutions to subproblems.
+A famous example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
 \[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
@@
 A typical characteristic is that an intermediate solution is reused multiple times to construct larger solutions (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
-Reusing an existing solution avoid redoing expensive computations: with memoization (memorizing intermediate results), the solution of $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !
+Reusing an existing solution avoids redoing expensive computations: with memoization (memorizing intermediate results), the solution for $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !
 
-Formally, dynamic programming problems respect the Bellman's principle of optimality: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}.
-This means that every intermediate result is computed only once, although it might be reused as basis for multiple larger problems, hence our first observation.
+Formally, dynamic programming problems respect the Bellman's principle of optimality\cite{bellman_principle}: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}.
+This means that every intermediate result needs to be computed only once, although it might be reused as a basis for multiple larger problems, hence our first observation.
 
 There exist various categories of dynamic programming:\ul
-\item Series that operates usually on a single dimension (like Fibonacci)
-\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ...
-\item Tree-related algorithms: phylogenetic, trees raking, maximum tree independent set, ...
+\item Series that operate usually on a single dimension (like Fibonacci)
+\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
+\item Tree-related algorithms: phylogenetic, trees raking, maximum tree independent set, ... (can be viewed as a sparse version of the second category)
 \ule
 
-Since the first category is inherently sequential (progress cannot be faster than one element at a time) and the third category is both hard to parallelize efficiently (similar to a sparse version of the second category) and does not share much with the previous category, we focus on the second type of problems, which is also the most common.
+Since the first category operates on a single dimension, to benefit of the smaller solutions to compute larger ones, elements must be computed sequentially (one at a time), hence computations cannot be made parallel (unless duplicated, thereby hindering benefits of memoization).
+The third category suffers from limited parallelism \cite{philogeny} and its implementation does not share much with the previous category, hence we focus on the second type of problems.
 
-Taking real-world examples, the average input size for sequence alignment is around 300K whereas for problems like RNA folding, input are usually around few thousands.
-Multiple input problems also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex.
+Taking real-world examples in biology, the average input size for sequence alignment (\S\ref{swat_affine}) is around 300K whereas for problems like RNA folding (\S\ref{zuker}), input are usually around few thousands.
+Problems operating on multiple input sequences also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex (as intermediate results needs to be stored in a position representing the progress in each of the involved sequence).
 Since we target a single computer with one or more attached devices (GPUs, FPGAs), and since we plan to maintain data in memory (due to the multiple reuse of intermediate solutions) the storage complexity must be relatively limited, compared to other problem that could leverage the disk storage.
 Hence in general, we focus on problems that have $O(n^2)$-space complexity whereas time complexity is usually $O(n^3)$ or larger.
-We encourage you to refer to the section~\ref{problems} for further classification and examples.
+We encourage you to refer to \S\ref{problems} for further classification and examples.
 
@@
 \newpage
-\subsection{Scala} \label{intro_scala}
+\subsection{Scala} \label{bg_scala}
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
@@
 As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
-\item The functional programming style and syntactic sugar offered by Scala allows concise writing of implementation, analysis and transformations of our DSL, which would have been much more complex and tiresome in an imperative language like C.
-\item Scala is largely adopted in the industry, which makes both the adoption of related project easier and offer a steeper learning curve to their potential users.
-\item Since Scala programs execute in the JVM, they can benefit of the JNI, that is the possibility to dynamically load C libraries and interact with them, possibly leveraging  underlying hardware by launching CUDA kernels.
+\item The functional programming style and syntactic sugar offered by Scala allow concise writing of implementation, analysis and transformations of our DSL, allowing us to focus on \textit{what} we want to achieve instead of \textit{how}.
+\item Scala is largely adopted in the industry\footnote{\url{http://www.scala-lang.org/node/1658}}, which makes both the adoption of related project easier and reduces the learning time of its potential users.
+\item Since Scala programs execute in the Java Virtual Machine (JVM), they can benefit of the native interface (JNI) that offers the possibility to dynamically load libraries (usually written in C) and possibly interact with CUDA to leverage the GPU.
+\item Scala is equipped with a strong typing and type inference system that reduces the syntactical constraints while putting strong guarantees on type correctness at compilation.
+\item Implicit functions definition can help interfacing functions by automatically changing the type of passed arguments
+\item Manifests (or TypeTags and ClassTags) allow type extraction at runtime (we use this to convert a Scala type into a C/CUDA type)
+\item Macros\cite{scala_macros} and LMS (\S\ref{bg_lms}) could be used to modify the semantics of a specific part of the user program (we currently use LMS to produce the corresponding C function).
 \item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and mixed together, thereby allowing multiple inheritance.
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Lightweight Modular Staging} \label{intro_lms}
+\subsection{Lightweight Modular Staging} \label{bg_lms}
 Lightweight Modular Staging (LMS) \cite{lms}, \cite{lms_thesis} is a runtime code generation built on top of Scala virtualized \cite{scala_virtualized} that uses types to distinguish between binding time (compilation and runtime) for code compilation.
@@
 
-At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
-However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
-Ad-hoc C translation seems more appropriate than writing Scala code and convert it with LMS because LMS only understand a subset of both languages.
-
-LMS would still be helpful for generating user-specific functions, as they are independent of the rest of the program and user wants to write his functions only once for both Scala and CUDA.
+A discussion of the integration of LMS in our project can be found in \S\ref{lms_use}.

### 2013-01-15 docs/report/problems.tex
More report.

@@
 \subsection{Problems classification}
-Since <<dynamic programming>> defines a very general technique, we already focused on grammar and alignment problems in \ref{intro_dp}.
+Since <<dynamic programming>> defines a very general technique, we already focused on grammar and alignment problems in \ref{bg_dp}.
 Before exploring some particular problem instances, we want to define some characteristics that will be used through the rest of the document to describe dynamic programming problems.
@@
 \subsubsection{Litterature classification}
-In the literature, dynamic programming problems are classified according to two criteria:\ul
+In \cite{gpu_atlp}, dynamic programming problems are classified according to two criteria:\ul
 \item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman).
@@
 \newpage
-\subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
+\subsubsection{Smith-Waterman (simple)}\label{swat_simple}\ol
 \item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
@@
 \newpage
-\subsubsection{Smith-Waterman with affine gap extension cost}\ol
-
+\subsubsection{Smith-Waterman with affine gap extension cost} \label{swat_affine}
+\ol
 \item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
@@
 
-\item Backtracking: similar to \nameref{sswat}
-\item Visualization: same as \nameref{sswat}
-\item Optimizations: Notice that this recurrence is very similar to \nameref{sswat} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
+\item Backtracking: similar to \nameref{swat_simple}
+\item Visualization: same as \nameref{swat_simple}
+\item Optimizations: Notice that this recurrence is very similar to \nameref{swat_simple} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
 Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of two additional matrices.
@@
 \newpage
-\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
+\subsubsection{Smith-Waterman with arbitrary gap cost}\label{swat_arbitrary}\ol
 \item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
@@
 
-\item Backtracking: similar to \nameref{sswat} except that you can jump of $k$ cells along the rows or along the columns.
+\item Backtracking: similar to \nameref{swat_simple} except that you can jump of $k$ cells along the rows or along the columns.
 \item Visualization:
@@
 \newpage
-\subsubsection{Nussinov algorithm}\ol
+\subsubsection{Nussinov algorithm} \label{nussinov} \ol
 \item Problem: folding a RNA string $S$ over itself $|S|=n$.
@@
 \newpage
-\subsubsection{Zuker RNA folding}\ol
+\subsubsection{Zuker RNA folding} \label{zuker} \ol
 \item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy.
@@
 
-The recurrence consists of two non-serial dependencies as in \textit{\nameref{aswat}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
+The recurrence consists of two non-serial dependencies as in \textit{\nameref{swat_arbitrary}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
 Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
@@
 Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
-Although our project shares many patterns with this project we can not reuse it because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{intro_lms}) whereas this framework focuses on parallelizing operations on independent elements.
+Although our project shares many patterns with this project we can not reuse it because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{bg_lms}) whereas this framework focuses on parallelizing operations on independent elements.

### 2013-01-15 docs/report/design.tex
Reports: SWat problems.

@@
 This has two main applications:\ol
-\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Watermann share the same grammar but have a different evaluation algebra
+\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
 \item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data.

### 2013-01-15 docs/report/problems.tex
Reports: SWat problems.

@@
 
+% ----------------------------------------------
 \subsubsection{Definitions}\ul
@@
 By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
-\item \textbf{Alphabets:} the alphabet represent an enumeration of the possible values.
-Their size helps determining how many bits are required in the implementation to store the values.
-Alphabets must be defined for input, wavefront, cost and backtrack.
+\item \textbf{Alphabets:} an alphabet is an enumeration of the possible values.
+Its size helps determining how many bits are required in the implementation to store all values.
+Alphabets are defined for input, cost, backtrack and wavefront.
 \item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
-\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest.
-Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
-\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute.
-A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
-\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix.
+\item \textbf{Matrices:} we refer by \textit{matrix} or \textit{matrices} to all the memoized intermediate cost- and backtrack-related information that is necessary to solve the dynamic programming problem of interest.
+Matrix elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
+\item \textbf{Computation block:} this is a part of the DP matrix (cost and backtrack) that we want to compute.
+A block might be either a sub-matrix (rectangular) or a parallelogram (possibly reduced by taking the intersection with its enclosing matrix).
+\item \textbf{Wavefront:} the wavefront consists of the minimum data necessary to construct a computation block of the DP matrix.
 It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
-\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
+\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i+1,j)}, M_{(i+2,j-1)}\big)$ has delay 3).
 \ule
 
+% ----------------------------------------------
 \subsubsection{Litterature classification}
 In \cite{gpu_atlp}, dynamic programming problems are classified according to two criteria:\ul
-\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman).
+\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman \S\ref{swat_simple}).
 When two or more terms appear, the problem is polyadic (ex: Fibonacci, $F_n = F_{n-1} + F_{n-2}$).
@@
 That is computing an element of the matrix would require $O(n^s)$.
-(ex: Smith-Waterman with arbitrary gap is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula)
-	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
+For example, Smith-Waterman with arbitrary gap cost (\S\ref{swat_arbitrary}) is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula (see recurrence formulae in \S\ref{problem_examples}).
+%	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
 \ule
-
 Note that the algorithmic complexity of a problem is exactly $O\big(n^{d+s}\big)$.
 
-\subsubsection{Calculus simplifications} \label{calc_simplifications}
-In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
+% ----------------------------------------------
+\subsubsection{Recurrence formulae simplifications} \label{calc_simplifications}
+In some special cases, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
 	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Problems of interest}
-We usually focus on problem that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$).
+\subsection{Problems of interest} \label{problem_examples}
+We here focus on problems that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$).
 Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability.
-Also we tend to limit algorithmic complexity of the problems as from $O(n^4)$ on, running time becomes a severely limiting factor.
+Also it seems that most problems of interest have an algorithmic complexity of at most $O(n^4)$, probably because running time would otherwise becomes a severely limiting factor for the size of the problem.
 
@@
 
-Although not necessary to understand the project, the description of some of the most common dynamic programming problem is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems.
+Although not necessary to understand the project, the description of some of the most common dynamic programming problems is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems.
 Would the reader be familiar with dynamic programming, he could immediately jump to the next section.
 
-A tighter analysis on the alphabet and intermediate results size is done because FPGA was also considered as a possible execution platform.\footnote{Initially, the project envisioned a collaboration with \href{mailto:nithin.george@epfl.ch}{Nithin George}, a PhD student at Processor Laboratory (LAP), EPFL.
-Unfortunately, at the time of writing, progress in this direction is not sufficient to be covered in this report.
-However, FPGA is considered as a future deployment platform for a particular class of problems.}
+A tighter analysis on the alphabet and intermediate results size is done because FPGA was also considered as a possible execution platform.
 
@@
 \newpage
-\subsubsection{Smith-Waterman (simple)}\label{swat_simple}\ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\subsubsection{Smith-Waterman (simple)}\label{swat_simple}
+Smith-Waterman is a biological sequence alignment algorithm.
+It tries to find the maximum number of correspondences between two DNA sequences; variants of this algorithm include \href{http://en.wikipedia.org/wiki/Needleman–Wunsch_algorithm}{Needleman-Wunsch}, and minimum edit distance family that generalizes on strings (\href{http://en.wikipedia.org/wiki/Hamming_distance}{Hamming distance}, \href{http://en.wikipedia.org/wiki/Levenshtein_distance}{Levenshtein distance}, ...).
+We explore three variants of this algorithm: simple (\S\ref{swat_simple}), affine (\S\ref{swat_affine}) and arbitrary (\S\ref{swat_arbitrary}) gap cost models.
+We study this problem because it has the interesting properties of using multiple input sequences and being suitable for hardware generation \cite{swat_fpga}.\\[6pt]\ol
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$, with constant mismatch penalty ($d$) and arbitrary matching function (${\rm cost}(\_,\_)$).
 \item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
@@
 \item Optimizations:\ul
-	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built).
-This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
-	\item Since first line and column of the matrix (corresponding to a matching with one string being empty) are zeroed, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
+	\item In serial (monadic) problems we can avoid building the matrix $M$ by maintaining only the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being currently built).
+This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory, where $k$ is the maximum backward lookup.
+	\item We could eliminate the first line and column of the matrix as they are filled with zeroes (representing a match with empty string), however this implies more involved computations, which is cumbersome.
 	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$, we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned.
@@
 % ----------------------------------------------
-\newpage
 \subsubsection{Smith-Waterman with affine gap extension cost} \label{swat_affine}
 \ol
-\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
+\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$, where creating a gap in either sequence has an opening penalty ($\alpha$) and an extension penalty ($\beta$).
 \item Matrices: $M_{(m+1) \times (n+1)}, E_{(m+1) \times (n+1)}, F_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
@@
 
-Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is
-\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]
+Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules as
+\[ [M;E;F]_{(i,j)} = f \big( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} \big) \]
 
-\item Backtracking: similar to \nameref{swat_simple}
-\item Visualization: same as \nameref{swat_simple}
-\item Optimizations: Notice that this recurrence is very similar to \nameref{swat_simple} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
-Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of two additional matrices.
+\item Backtracking and visualization are similar to \S\ref{swat_simple}
+\item Optimizations: Notice that this recurrence is very similar to \S\ref{swat_simple} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$).
+Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of the two additional matrices.
 \ole
@@
 % ----------------------------------------------
-\newpage
 \subsubsection{Smith-Waterman with arbitrary gap cost}\label{swat_arbitrary}\ol
 \item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap.
-Without loss of generality, let $m\ge n$ (swap inputs if necessary).
-Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching with smaller gaps.} $g(x)=m-x$.
+For example\footnote{Intuition: long gaps should penalize less; one large gap might be better than matching with smaller gaps.}: $g(x)=\max(m,n)-x$.
+
 \item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
@@
 
-\item Backtracking: similar to \nameref{swat_simple} except that you can jump of $k$ cells along the rows or along the columns.
+\item Backtracking: similar to \S\ref{swat_simple} except that you can jump of $k$ cells along the rows or along the columns.
 \item Visualization:
@@
 
-The recurrence consists of two non-serial dependencies as in \textit{\nameref{swat_arbitrary}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
+The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
 Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
@@
 \item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
-\item Biological sequence alignment (Smith-Watermann) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
+\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
 \ule

### 2013-01-15 docs/report/implementation.tex
Report: problems reworked.

@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Ad-hoc CUDA}
+\subsection{Ad-hoc CUDA} \label{ad_hoc_impl}
 In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol

### 2013-01-15 docs/report/problems.tex
Report: problems reworked.

@@
 \newpage
-\subsubsection{Convex polygon triangulation}\ol
-\item Problem: triangulating a polygon of $n$ vertices with least total cost for added edges.
-We denote the cost of adding an edge between the  pair of edges $i,j$ by $S_{(i,j)}$, Where $S_{n \times n}$ is a symmetric matrix (can be stored as a triangular matrix with 0 diagonal that can be omitted), hence $|S|=\tfrac{n(n-1)}{2}=N$.
-\item Matrices: $M_{(n+1)\times (n+1)}, B_{(n+1)\times (n+1)}$ \textit{indices denote <<first vertex, last vertex>>}.\\ Upper triangular matrices including main diagonal.
-Note that the vertex $n$ is the same as the vertex $0$ due to the cyclic nature of the problem.
+\subsubsection{Convex polygon triangulation} \label{polygon_tri} \ol
+\item Problem: triangulating a convex polygon of $n$ vertices at minimal cost. Adding an edge $[i,j]$ has a cost $S_{(i,j)}$, where $S$ is a $(n\times n)$ matrix.
+\item Matrices: $M_{(n+1)\times (n+1)}, B_{(n+1)\times (n+1)}$, upper triangular matrices including main diagonal.
+Indices denote \textit{<<first vertex, last vertex>>}; the vertex $n$ is the same as the vertex $0$ due to the cyclic nature of the problem.
 \item Alphabets:\ul
@@
 
-\item Optimizations: if the cost of edges between contiguous vertices is 0, we do not need to handle special cases in the DP program (i.e. existing edges cannot be added).
+\item Optimizations:\ul
+	\item If the cost of edges between contiguous vertices is 0, we do not need to handle special cases in the DP program (i.e. existing edges cannot be added).
+	\item The matrix cost $S$ is a symmetric matrix and can be stored as a triangular matrix with 0 diagonal that can be omitted), hence $|S|=\tfrac{n(n-1)}{2}=N$.
+	\ule
 \ole
@@
 % ----------------------------------------------
-\newpage
 \subsubsection{Matrix chain multiplication}\label{mat_mult_plain}\ol
@@
 
-\item Optimizations: we could normalize the semantics of indices and use $(n+1) \times (n+1)$ matrices where the meaning of cell $(i,j)$ would be $\underset{i\le k<j}{\rm chain}(A_k)$.
+\item Optimizations:\ul
+	\item We could normalize the semantics of indices and use $(n+1) \times (n+1)$ matrices where the meaning of cell $(i,j)$ would be $\underset{i\le k<j}{\rm chain}(A_k)$.
 	%\ul
@@
 	%\ule
+	\item Alternatively, we could encode the dimension of the resulting matrix within the cost matrix by using a triplet (rows,columns,cost) and taking minimum appropriately.
+	\ule
 \ole
@@
 
-This recurrence is more amenable to be converted into a grammar (in pseudo-code)
-\begin{lstlisting}[language=Scala,caption=Pseudocode of the alternative Zuker grammar]
-Qp : matrix(
-  ( seq            ^^ x => Eh(x)
-  | el  ~ Qp ~ el  ^^ (a,b,c) => Es(a,c) + b
-  | seq ~ Qp ~ seq ^^ (a,b,c) => Ei(a.start, a.end, c.start, c.end) + b
-  | el  ~ QM  ~ el ^^ (a,b,c) => b
-  ).filter(basepair).getMinimum)
-
-QM : matrix(
-  ( Q ~ Q ^^ (a,b) => a+b
-  ).filter(size>=4).getMinimum)
-
-Q : matrix(
-  ( QM
-  | el ~ Q ^^ (a,b) => b
-  | Q ~ el ^^ (a,b) => a
-  | Qp
-  ).filter(size>=2).getMinimum)
-
-\end{lstlisting}
-\begin{center}
-\vspace{-24pt}
-\footnotesize ({\tt seq} represent a nonempty sequence of elements and {\tt el} a single element)
-\end{center}
+This latter recurrence is more amenable to be converted into a grammar as the matrix are all of the same dimension. See the example in \S\ref{ex_rnafold} for a detailed implementation of this problem.
 
 % ------------------------------------------------------------------------------------------------
-\newpage
 \subsection{Related problems}
-The aim of this section is to demonstrate that the problems described above are very similar or encompass a significant part of the common dynamic programming problems\footnote{There are hyperlinks on the problems name to their detailed description.}.
+The aim of this section is to demonstrate that the problems previously described are very similar or encompass a significant part of the common dynamic programming problems\footnote{There are hyperlinks on the problems name to their detailed description.}.
 
@@
 \bf Serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
-Smith-Waterman \footnotesize simple & rectangle & 1 & -- \\
-Smith-Waterman \footnotesize affine gap extension & rectangle & 3 & (can replace 2 matrices) \\
+Smith-Waterman \footnotesize simple (\S\ref{swat_simple}) & rectangle & 1 & -- \\
+Smith-Waterman \footnotesize affine gap extension (\S\ref{swat_affine}) & rectangle & 3 & (can replace 2 matrices) \\
 \href{http://en.wikipedia.org/wiki/Needleman-Wunsch_algorithm}{Needleman-Wunsch} & rectangle & 1 & -- \\
@@
 \bf Non-serial problems & \bf Shape & \bf Matrices & \bf Wavefront \\ \midrule
-Smith-Waterman \footnotesize arbitrary gap cost & rectangle & 1 & -- \\
-Convex polygon triangulation & triangle & 1 & -- \\
-Matrix chain multiplication & triangle & 1 & -- \\
-Nussinov & triangle & 1 & -- \\
-Zuker folding & triangle & 3 & -- \\
+Smith-Waterman \footnotesize arbitrary gap cost (\S\ref{swat_arbitrary}) & rectangle & 1 & -- \\
+Convex polygon triangulation \footnotesize (\S\ref{polygon_tri}) & triangle & 1 & -- \\
+Matrix chain multiplication \footnotesize (\S\ref{mat_mult_plain}) & triangle & 1 & -- \\
+Nussinov \footnotesize (\S\ref{nussinov}) & triangle & 1 & -- \\
+\href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{Zuker folding} \footnotesize (\S\ref{zuker}) & triangle & 3 & -- \\
 \href{http://en.wikipedia.org/wiki/CYK_algorithm}{CYK} \footnotesize Cocke-Younger-Kasami & triangle & \#rules & -- \\
@@
 
-These algorithm also involve dynamic programming. However, we do not evaluate thoroughly their shape and number of matrices as this is not relevant in the scope of this project.\ul
+These algorithms also involve dynamic programming. However, we do not thoroughly evaluate their shape and number of matrices as a detailed description is not the focus of this project.\ul
 \item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}
@@
 \href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
-\item \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}
 \item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
@@
 \subsubsection{Conclusion}
-In the rest of the project, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar).
+In the rest of the report, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar).
 Although not of immediate use, the description of the above problem and ad-hoc CUDA implementation of three of them (Smith-Waterman with arbitrary gap cost, Matrix chain multiplication and Convex polygon triangulation) helped us to understand:\ol
-\item There is a difference between dynamic programming as seen in algorithmic classes and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
+\item There is a difference between dynamic programming as seen in algorithmic schoolbooks and their concrete implementation, mainly because special care must be taken for correct indices and preventing off-by-one errors.
 \item Problems can be classified in two categories: single track (input) and two-tracks (2 input sequences).
@@
 \item Sometimes matrices are initially padded with zeroes (or initial value), although this might be ignored at algorithm design, care must be taken for these special values and their inclusion in the matrix should be decided according to the complexity of the recurrence formula.
-\item Incidentally, we proposed a cyclic variant of the convex polygon triangulation, which uses a parallelogram matrix instead of a triangular one (upper right triangle extended by a lower triangle on the right).
-Unfortunately, this proved to be based on an erroneous recurrence relation analysis, which could be reduced to the more natural triangular case.
+\item Incidentally, we proposed a cyclic variant of the convex polygon triangulation, which uses a parallelogram matrix (see \S\ref{mem_layout}).
+Unfortunately, this proved to be based on an erroneous recurrence relation analysis, and can only use a triangular matrix as described in \S\ref{polygon_tri}.
 
-Although the polygon triangulation problem can be solved with a regular triangular matrix and we have not found real problem requiring a parallelogram matrix, we still present this version in the analysis part of the project, which could be justified for cyclic problems that require dynamic programming for each unfolding position (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting flattened sequence).
+Although we have not found a real problem requiring a parallelogram matrix, we still present this version in \S\ref{mem_layout} and \S\ref{ad_hoc_impl}.
+Such matrix layout could be adapted for cyclic problems that could be broken into a linear sequence anywhere (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting flattened sequence).
 For example, one could be interested in finding the longest subsequence verifying some property in a cycle, such that the subsequence score changes if it is rotated.
@@
 
-Considered as possible solutions to store a varying number of results per matrix cell were dynamic memory allocation \cite{nvidia} available on recent graphic cards, ad-hoc memory allocation \cite{scatter_alloc} or hash tables \cite{parallel_hashing}, however the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, also the use of cuckoo hash tables adds a constant factor penalty to every memory access.
+Since this project involves various domains, we also investigated in the memory management on graphic cards and existing code generation frameworks.
+
+In an attempt to support a varying number of results per matrix cell, we considered dynamic memory allocation \cite{nvidia} (available on recent graphic cards), ad-hoc memory allocation \cite{scatter_alloc} and hash tables \cite{parallel_hashing}.
+However the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, and the use of cuckoo hash tables adds a constant factor penalty to every memory access.
 Finally both solution introduce undesirable possibility of failure (respectively out of memory or unrecoverable collision) in the middle of the algorithm computation process.
 
-Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS to generate from the same source code efficient implementation for heterogeneous platforms at runtime.
-Although our project shares many patterns with this project we can not reuse it because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{bg_lms}) whereas this framework focuses on parallelizing operations on independent elements.
+Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS\cite{lms_thesis} to generate from the same source code efficient implementation for heterogeneous platforms (including CUDA) at runtime.
+Although this shares many patterns with our project, we can not reuse this framework because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{bg_lms}) whereas Delite focuses on parallelizing operations on collections (array, lists, maps, ...) of independent elements.

### 2013-01-15 docs/report/report.tex
More report.
Report: problems reworked.

@@
 
-\section{Usage}
+\section{Usage} \label{usage}
 {\color{red} usage, make reference from problems to usage
@@
 }
+
+\subsection{Program examples}
+\subsubsection{Matrix multiplication} \label{ex_matmult}
+\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
+import v4._
+
+trait MatrixSig extends Signature {
+  type Alphabet = (Int,Int) // Matrix(rows, columns)
+  val single:Alphabet=>Answer
+  val mult:(Answer,Answer)=>Answer
+}
+
+trait MatrixAlgebra extends MatrixSig {
+  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
+
+  override val h = minBy((a:Answer) => a._1)
+  val single = (a: Alphabet) => (0, a)
+  val mult = (a:Answer,b:Answer) =>
+    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
+}
+
+trait MatrixGrammar extends ADPParsers with MatrixSig {
+  val axiom:Tabulate = tabulate("M",
+    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
+}
+
+object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
+  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
+}
+\end{lstlisting}
+
+\subsubsection{RNA folding} \label{ex_rnafold}
+\begin{lstlisting}[language=Scala,caption={RNA folding}]
+import v4._
+
+trait RNAFoldSig extends RNASignature {
+  def hairpin(ij:(Int,Int)):Answer
+  def stack(i:Int,s:Answer,j:Int):Answer
+  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
+  def mloop(i:Int,s:Answer,j:Int):Answer
+  def left(l:Answer,r:Int):Answer
+  def right(l:Int,r:Answer):Answer
+  def join(l:Answer,r:Answer):Answer
+}
+
+trait RNAFoldAlgebra extends RNAFoldSig {
+  type Answer = Int
+  import librna.LibRNA._ // indexing convention: first base,last base
+  private def Eh(i:Int,j:Int) = hl_energy(i,j) // hairpin loop
+  private def Ei(i:Int,k:Int,l:Int,j:Int) = il_energy(i,k,l,j) // internal loop
+  private def Es(i:Int,j:Int) = sr_energy(i,j) // stacked base pairs
+  def hairpin(ij:(Int,Int)) = Eh(ij._1,ij._2-1)
+  def stack(i:Int,s:Int,j:Int) = Es(i,j) + s
+  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) = Ei(ik._1,ik._2,lj._1-1,lj._2-1) + s
+  def mloop(i:Int,s:Int,j:Int) = s
+  def left(l:Int,r:Int) = l
+  def right(l:Int,r:Int) = r
+  def join(l:Int,r:Int) = l+r
+  override val h = min[Answer] _
+}
+
+trait RNAFoldPrettyPrint extends RNAFoldSig {
+  type Answer = String
+  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
+  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
+  def stack(i:Int,s:String,j:Int) = "("+s+")"
+  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
+  def mloop(i:Int,s:String,j:Int) = "("+s+")"
+  def left(l:String,r:Int) = l+"."
+  def right(l:Int,r:String) = "."+r
+  def join(l:String,r:String) = l+r
+}
+
+trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
+  lazy val Qp:Tabulate = tabulate("Qp",(
+    seq(3,maxN)        ^^ hairpin
+  | eli   ~ Qp ~ eli   ^^ stack
+  | seq() ~ Qp ~ seq() ^^ iloop
+  | eli   ~ QM ~ eli   ^^ mloop
+  ) filter basepairing aggregate h)
+
+  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
+
+  lazy val Q:Tabulate = tabulate("Q",(
+    QM
+  | Q ~ eli ^^ left
+  | eli ~ Q ^^ right
+  | Qp
+  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
+
+  override val axiom = Q
+}
+
+object RNAFold extends App {
+  object fold extends RNAFoldGrammar with RNAFoldAlgebra
+  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
+
+  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuu"
+
+  import librna.LibRNA._
+  setSequence(seq)
+  val (score,bt) = fold.backtrack(seq.toArray).head
+  val res = pretty.build(seq.toArray,bt)
+  clear
+  println("Folding : "+res+" (%5.2f)".format(score/100.0));
+}
+\end{lstlisting}
 

### 2013-01-15 docs/report/design.tex
Report first pass finished.

@@
 
-The benefits of ADP framework is that it does not constraint the result of the evaluation to be a single value, but can extend parsers to backtracking parsers or pretty-printers.
+The benefits of the ADP framework is that it does not constrains the result of the evaluation to be a single value, but can extend parsers to backtracking parsers or pretty-printers.
 Additionally, we want to support the following features:\ol
@@
 To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.}
-\item \textbf{Input pair algebra:} the original ADP framework only support single input, we want to support a pair of inputs such that we can treat problem such as Smith Waterman or Needleman-Wunsch.
-However, it does not make sense to treat more than two sequences because of the $\Omega(n^3)$ storage requirements that limits the problems size more dramatically.
+\item \textbf{Input pair algebra:} the original ADP framework \cite{adp} only support single input, we want to support pairs of inputs sequences similarly as \cite{gapl} such that we can treat problem such as Smith Waterman or Needleman-Wunsch.
+As discussed in \S\ref{bg_dp}, handling more than two sequences introduces an $\Omega(n^3)$ storage complexity that might limit more severely the size of problems that could be addressed.
+Since the problems seen in \S\ref{problems} use either one or two input sequences, we only need to support these two cases.
 \item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$-best results.
-\item \textbf{Input restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ...
+\item \textbf{Input restrictions:} since CUDA (and FPGA) cannot process arbitrary Scala objects, we need to restrict the language to primary types (int, float, ...
 and structures of them).
-However, we want to preserve the expressiveness available for Scala and impose restrictions on the input and answers available to CUDA.
-A typical restriction we want to make is that processed structures are of fixed size so that we avoid memory management issues and thread divergence.
+However, we want to preserve the expressivity available in Scala and impose restrictions on the data types processed by CUDA.
+A typical restriction we want to make is that data elements are of fixed size, to avoid memory management issues and thread divergence\footnote{Occurs when a single thread needs more processing than its peers, thereby delaying the whole computation.}.
 \item \textbf{Single-result on devices:} The general ADP framework supports multiple solutions for intermediate results.
-Such functionality is easy to support in Scala; however, memory management hampers the performance of the GPU implementation.
+Such functionality is easily supported in Scala; however, memory management hampers the performance of the GPU implementation (see \S\ref{related_work}).
 To overcome this issue, the user could manually manage the memory, but this would defeat most of the benefits of automatic code generation.
-Hence the trade-off solution we propose is to restrict ADP to only one optimal result on CUDA, while leaving the freedom to obtain co-optimal (or even all possible solutions) with the Scala version.
+Hence the trade-off solution we propose is to restrict ADP to only one optimal result on CUDA, while offering the possibility to obtain co-optimal (or even all possible solutions) with Scala.
 
-\item \textbf{Automatic backtracking:} Since efficient code has to be devised we imposed restrictions on the output that could be generated by the parsers on devices.
-However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we would like to to automate the backtracking to fulfill goals of usefulness, efficiency and ease-of-use in device-specific implementation:\ul
+\item \textbf{Automatic backtracking:} To produce efficient code, we imposed a fixed size on the output generated by the parsers on devices.
+However, on the other hand, the backtracking information (of varying size) is of primary interest for the DSL user, hence we would like to to automate the backtracking to fulfill goals of usefulness, efficiency and ease-of-use in device-specific implementation:\ul
 	\item Leaving the backtrack implementation to the user would force him to memoize the backtracking information together with the results (backtrack would  grow towards final result and duplicate unnecessarily information), hence requiring both $O(n^3)$ space and memory management features on devices.
 	\item Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement.
-Collecting the backtracking list can be easily done in $O(n)$ and then inverted whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
+Collecting the backtracking list can be easily done in $O(n)$ and then reversed depending on whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
 	\ule
-\item \textbf{Yield analysis:} in the vanilla ADP, the user has to define for each concatenation the minimal and maximal length of the subsequence on each side.
-Although non-emptiness information is necessary to avoid infinite recursion in the parsers, forcing an explicit definition can become cumbersome for the DSL user.
-We want to provide an automatic computation of concatenation boundaries, while at the same time leaving the possibility to manually specify it for maximum flexibility.
+\item \textbf{Yield analysis:} in vanilla ADP, the user has to define for each concatenation the minimal and maximal length of the subsequence on each side.
+Although non-emptiness information is necessary to avoid infinite recursions in the parsers, forcing an explicit definition can become cumbersome for the DSL user.
+Similarly as in \cite{gapc_yield}, we want to provide an automatic computation of concatenation boundaries, while at the same time leaving the possibility to manually specify it for maximum flexibility.
 \ole
@@
 \item \textbf{Dependency analysis:} Since we target GPUs (and FPGAs) which are massively parallel architecture, a top-down execution using hash tables is impractical (fallback computation if element is not present is hard to parallelize), hence we need to construct the result tree bottom-up, therefore ensure that the (partial) evaluation order between rules is respected.
-\item \textbf{Normalization:} in order to automate the backtracking, we need the rule to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives).
-Also we need to maintain coherency between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (for instance pretty-printing or effectively multiplying matrices).
-
-\item \textbf{Optimizations:} An possible optimization is to break down complex rules into simpler ones if they are expressible so; thereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal.
-Unfortunately this analysis is very involved: we need to solve the following problem:
-
-Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
-\[\begin{array}{rcll}
-f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
-	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
-f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
-	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
-\end{array}\]
-% We can solve this more easily if functions are linear
-Since this require complex mathematical analysis that are out of the scope of the project, we leave this optimization to the responsibility of the user.
-
-Another optimization that is easier to implement is dead rule elimination.
-Traditional dead code elimination only reduces the size of the generated code.
-In the GPU implementation, this optimization will not only reduce the binary size but also the CUDA memory consumption as well as speed-up computation (since all rules are executed together).
+\item \textbf{Normalization:} in order to automate the backtracking, we need the grammar rules to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives).
+Also we need to maintain coherence between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (pretty-printing or actual computation as described in \S\ref{intro}).
+\item \textbf{Optimizations:} Since ADP exposes a grammar, we have the possibility to do grammar optimizations (see \S\ref{grammar_opt}).
+Also since the grammar might define useless rules, we might want to eliminate them: dead rule elimination is very similar to traditional dead code elimination (it reduces the size of generated code) but also reduces the memory consumption (as storage matrix does not need to be reserved) and even speeds up the computation, (since all grammar rules for a particular element are computed at the same time in CUDA implementation, see \ref{dead_rules}).
 \ule
-% {\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
 
@@
 ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
-Such separation makes possible to define multiple algebra for the same grammar.
+Such separation makes it possible to define multiple algebra for the same grammar.
 This has two main applications:\ol
@@
 \item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data.
-For example in matrix chain multiplication, the first step solves the underlying dynamic problem by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
+For example in matrix chain multiplication, the first step solves the underlying dynamic program by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
 \ole
 
-Practically, an ADP program is constituted of 3 components: a \textbf{signature} containing functions signatures, which are implemented by \textbf{algebrae} and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
-The concrete program instance mixes-in the algebra with the grammar.
-The grammar parsers intermediate results are memoized in an array (tabulation parser).
+Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebrae} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
+The concrete program instance combines the algebra with the grammar.
+The grammar parsers' intermediate results are memoized in a matrix (tabulation parser).
 A parser usually consist of a tree of:\ul
@@
 \item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
-\item \textbf{Concatenation:} constructs a larger sequence from two subsequences.
+\item \textbf{Concatenation:} constructs all possible combinations from two subsequences.
 The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
@@
 
-Additionally, the signature must define input alphabet ({\tt Alphabet}), and output alphabet ({\tt Answer}) can be defined either in the signature or in the algebra.
+Additionally, the signature must define an input alphabet ({\tt Alphabet}), and an output alphabet ({\tt Answer}) can be defined either in the signature or in the algebra.
 Finally, the grammar needs to have a starting point, denoted as axiom.
@@
 %\vspace{3cm}
-This program grammar can also be expressed in BNF:
+This program grammar can also be expressed in BNF\footnote{\url{http://en.wikipedia.org/wiki/Backus-Naur\_Form}}:
 \[\begin{array}{lrl}
@@
 
+\subsubsection{Grammar optimization} \label{grammar_opt}
+Since ADP exposes a grammar, we might be able to break complex grammar rules into simpler ones (optimally binary production). For example, consider the following rule (in BNF):
+\[A := B\;\;C\;\;D \qquad \implies \qquad \left\{\begin{array}{l} A' := B\;\;X \\ X := C\;\;D \end{array} \right. \]
+If $B$, $C$ and $D$ are of varying length, evaluating the rule $A$ for a single subproblem is $O(n^2)$ (assuming B,C,D are stored in matrices hence $O(1)$).
+Adding a tabulation $X$ reduces the evaluation complexity because for each subproblem we will consider either rule $A'$ or $X$, each of evaluation complexity $O(n)$; hence we would have reduced the grammar evaluation complexity.
+
+Since grammar candidates are then evaluated with an algebra, we need to devise an equivalent transformation of the algebra.
+Unfortunately this analysis is very involved: we need to solve the following problem (to respect Bellman's optimality principle\cite{bellman_principle}):
+
+Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
+\[\begin{array}{rcll}
+f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
+	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
+f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
+	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
+\end{array}\]
+% Viktor Kuncak: We can solve this more easily if functions are linear
+Since this requires complex mathematical analysis that is out of the scope of the project, and since we have not found relevant literature on that particular subject, we leave this optimization to the responsibility of the user by assuming that the provided grammar is already optimal.
+
 % ------------------------------------------------------------------------------------------------
-%\newpage
 \subsection{Recurrences analysis} \label{recurrences}
-\subsubsection{Dead rules elimination}
-Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are mandatory in the grammar).
+In this section we use the following notation: let $A$ be a tabulation parser, we refer by $A_{(i,j)}$ to the element $i,j$ of the underlying matrix (in order to keep a lightweight notation).
+\subsubsection{Dead rules elimination} \label{dead_rules}
+Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are reachable from the axiom).
 The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules.
@@
 
-Parsers are made of terminals, concat, operations (aggregate, map, filter) and tabulations; minimum/maximum yield of terminals is set, hence it only remains to assign appropriate yield sizes to tabulations; other operations simply propagate that information.
-It is possible to obtain the yield size of tabulations using the following algorithm (assuming recursive parsers contain at least one tabulation that is part of the loop):\ol
+Parsers are made of terminals, concatenations, operations (aggregate, map, filter) and tabulations; minimum/maximum yield of terminals is set, hence it only remains to assign appropriate yield sizes to tabulations; other operations simply propagate that information.
+It is possible to obtain the yield size of tabulations using the following algorithm (assuming recursive parsers contain at least one tabulation that is part of the loop), similar to \cite{adp_c}:\ol
 \item Set the yield minimum size of all tabulations to a large number $M_0$ (such that all tabulation would reasonably have a minimum yield size smaller than $M_0$)
@@
 \ole
-
-The last part of this algorithm seems to have an exponential complexity, however, if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$.
+The last part of this algorithm has worst case exponential complexity, but if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$.
 Obtaining the yield size of tabulations provides the following benefits:\ul
-\item Minimum size: prevents self-reference parsers (on the same subsequence) and avoid considering subsequences yielding empty results (hence slightly reducing time complexity).
+\item Minimum size: prevents self-reference parsers (on the same subsequence) and avoids considering subsequences which yield empty results (hence slightly reducing time complexity).
 \item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(m\cdot n)$ instead of $O(n^2)$ (where $m$ is the maximum yield size), possibly providing substantial space savings.
@@
 \subsubsection{Dependency analysis}
-Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
-A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ...
+Let us introduce the concept of dependency (similar to $\to_{chain}$ in \cite{adp_c}): a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
+A grammar is unsolvable if there exists a dependency loop between parsers ($A\to ...
 \to A$).
@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{Backtracking} \label{backtracking}
-In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
+In order to produce an efficient transformation from an ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
 To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
@@
 Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
-	\item Let ${\rm cc}(r_s)$ the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
+	\item Let ${\rm cc}(r_s)$ be the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
 		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
@@
 
-Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed in-order, applying for each rule the underlying transformation, and storing intermediate results (i.e.
-in a hash map) until they are processed by another rule.
-Since there is only one consumer for each intermediate result, every read in the hash map can remove the corresponding value, hence reducing the memory consumption.
-Ultimately, only the problem solution will be stored in the hash map.
+Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed in-order, applying for each rule the underlying transformation, while intermediate results are stored (in a hash map) until they are processed by another rule.
+Since there is only one consumer for each intermediate result, every read value can be immediately discarded, thereby reducing the memory consumption.
+Ultimately, only the problem solution will be stored (in the hash map).
 
@@
 
-Now it remains the problem to generate all possible results and check whether they are valid candidates.
+Now remains the problem of generating all possible results and check whether they are valid candidates.
 To do that we simply re-apply the parsers while maintaining the source elements of all production and then retain only those with desired score and backtrack.
@@
 This removes the $O(f(n))$ factor in the backtrack complexity (as concatenation backtrack parsers <<know>> where to split).
-\item Skip filters: since filter are applied before their inner solution is computed, their are only position-dependent.
+\item Skip filters: since filters are applied before their inner solution is computed, they are only position-dependent.
 Hence if a backtrack involves a filter, since its position is set by the backtrack, the filter must have been passed at matrix construction time.
@@
 Another special case we might be interested in is to obtain all co-optimal solutions (all the solutions that have an optimal score).
-We need here to notice that in the parser reversal, no sub-solution is discarded, because either it is not co-optimal (and would have been discarded at an earlier stage) or it is co-optimal, hence contributes to create a co-optimal result.
+We can notice that in the parser reversal, no sub-solution is discarded, because either it is not co-optimal (and would have been discarded at an earlier stage) or it is co-optimal, hence contributes to create a co-optimal result.
 It follows that the complexity of co-optimal solutions backtrack is proportional to the number of solutions.
@@
 \subsubsection{Backtrack utilization}
-Since the dynamic programming problem may help solving a larger problem\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation in a different domain.
-The easiest way to do that, is to reuse the same input and grammar, but use a different output domain, and only compute the result of the trace obtained from the DP backtrack.
+Since the dynamic programming may help solving a larger problem\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation in a different domain.
+The easiest way to do that, is to reuse the same input and grammar, but use a different algebra with a different output domain, and only compute the result of the trace obtained from the DP backtrack.
 
@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{CUDA storage: from list to optional value} \label{normalization}
@@
 A workaround might be to use fixed-length lists but we assume here that the programmer is most often interested in a single optimal solution (this also alleviates the complexity of constructing multiple distinct backtrack traces).
-Even if this restriction apparently greatly simplifies the design, issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
+Even if this restriction simplifies the design, issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
 
@@
 Useful aggregator belonging to this class are quite limited: minimum/maximum (optionally with respect to an object's property), count and sum\footnote{Notice that all these operations can be implemented with a folding operation on a single variable.}, hence it is possible to provide the user with a tailored implementation.
-To benefit of this fixed memory aggregation, we need to do some structural transformation of inner parsers.
+To benefit from this fixed memory aggregation, we need to do some structural transformation of inner parsers.
 In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order:
@@
 \end{tabular} \\[10pt]
-\footnotesize{$R=\ $required, $P=\ $performance optimization}
+\footnotesize{$R=\ $required, $P=\ $performance optimization, $?=\ $if possible}
 \caption{Parsers normalization towards CUDA code generation}
@@
 
-\item \textbf{Handling nested aggregations:} since a tabulation might contain nested aggregation, they must be preserved in order not to increase its time complexity.
+\item \textbf{Handling nested aggregations:} since a tabulation might contain nested aggregations, they must be preserved in order not to increase its time complexity.
 To do that, each internal aggregation has to define its own intermediate score and backtrack variables, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
-\item \textbf{Failure handling:} a parser can either be successful and return a valid result or fail and return no result; failure can happen in terminals, (input) tabulations and filters.
+\item \textbf{Failure handling:} a parser can either be successful and return results or fail and return no result (encoded as an empty list); failure can happen in terminals, (input) tabulations and filters.
 These 3 cases can be reduced to one by wrapping terminals and tabulations into a filter that checks validity conditions.
@@
 % ----------------------------------------------
-\newpage
 \subsubsection{Small problems (in-memory)}
@@
 Although this overhead seems appealing compared to the computation time, the total running time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
-Given that real problems (like RNA folding) operate at input sizes up to 4096, it would not be of much relevancy to implement a version for larger cases, although perfectly feasible.
+Given that real problems (like RNA folding) operate on sequences length in the hundreds \cite{gpu_rnafold}, it would not be of much relevance to implement a version for larger cases, although perfectly feasible.
 
@@
 The serial problems have the interesting property to access a fixed number of previous elements.
-These elements can be either stored either explicitly in a scoring matrix or implicitly as moving aggregation into the wavefront.
-Since the dependencies are fixed, the computation direction gains an additional degree of freedom: matrix can be solved in diagonal (as non-serial problems), line-wise or column-wise.
+These elements can be stored either explicitly in a scoring matrix, or implicitly into the wavefront (as moving aggregations).
+Since the dependencies are fixed, the computation direction gains an additional degree of freedom: the matrix can be solved in diagonal (as non-serial problems), line-wise or column-wise.
 This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
@@
 This division has the advantage of providing the minimal computation time at the expense of external storage proportional to $O(n)$ (if we store only lines or columns) or $O(n^2)$ (if we store both).
-\item \textbf{Myers and Miller’s algorithm:} (divide and conquer)
-This algorithm break the DP problem into 2 (or 4) subproblems such that once the middle line/column is computed, the problem can be solved for 1 submatrix and backtracking among up to 2 of the 3 other.
+\item \textbf{Myers and Miller’s algorithm:} \cite{myers_miller} (divide and conquer) 
+This algorithm break the DP problem into 2 (or 4) subproblems such that once the middle line/column is computed, the problem can be solved for one submatrix while backtracking occurs in up to 3 other submatrices.
 This breaking is applied recursively until the submatrix data fits into memory.
@@
 
-The algorithm proceeds as follows: first it solves the problem to obtain the first backtracking element, then it breaks the matrix in 4 submatrices, and refine it until backtrack is tractable.
+The algorithm proceeds as follows: first it solves the problem of obtaining the first backtracking element, then it breaks the matrix in 4 submatrices, and refine it until backtrack is tractable.
 Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
-\item \textbf{Hybrid approach:} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using $4 \times$ more storage space would only result in a $2\times$ speedup (measuring only the computation time).
+\item \textbf{Hybrid approach:} \cite{swat_linear} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using $4 \times$ more storage space would only result in a $2\times$ speedup (measuring only the computation time).
 Hence a hybrid approach would be to decide a $k$ such that at each step, we partition the desired sub-matrix into a intermediate grid of $k$ rows/columns.
@@
 Finally, although such problem is interesting because targeted platforms could include FPGA, where efficient implementations exist \cite{swat_fpga}, several reasons made us considering this class of problem as a future work: \ul
-\item The most prominent problem in this category is Smith-Waterman, for which an ad-hoc efficient implementation already exists\cite{swat_gpu},\cite{swat_mega}.
+\item The most prominent problem in this category is Smith-Waterman, for which efficient implementation already exists\cite{swat_gpu},\cite{swat_mega}.
 Additionally, the authors are planning to write extensions to support some variants of this problem like Needleman-Wunsch.
 \item The implementation sensibly differs from the class of small problems, as the solving strategy is completely different from non-serial small problems, thereby requiring larger development time that would be out of the scope of this project.
-\item Finally, such implementation would be only valuable for problems that are larger than the memory device, whereas smaller problem could perfectly use the existing implementation.
+\item Finally, such an implementation would be only valuable for problems that are larger than the memory device, whereas smaller problems could perfectly use the existing implementation.
 \ule
@@
 A major bottleneck on massively parallel architecture with shared memory (like CUDA) is the global memory access.
-To address it, according to the manufacturer documentation, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
+To address it, according to the manufacturer documentation \cite{nvidia}, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
 This is justified by the memory hardware architecture, where additional latency (precharge rows and columns of the memory chip) is mandatory to access data at very different positions.
@@
 
-Since progress is made along the diagonal of the matrices, a naive row/column addressing of the matrix elements would result in no coalesced access.
-Hence instead of addressing by row (or column) the matrix elements, we address them by diagonal.
+Since progress is made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses.
+Hence we address them by diagonal.
 Let $M_W$ the width of the matrix and $M_H$ its height.
@@
 
-Noticing that the triangle is just half the parallelogram, we could use the same addressing, but we would use twice the necessary memory.
+Noticing that the triangle matrix shape is just half the parallelogram, we could use the same addressing scheme, but we would use twice more memory than necessary.
 Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl.
@@
 
+From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}.
+Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress in our project rather than helping improving and adding features to an existing project (but I would reconsider my position given different deadlines).
+Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros (that operate on the Scala abstract syntax tree), which represent a possible alternative to LMS for generating  simple functions for a costing algebra.
+
 % ------------------------------------------------------------------------------------------------
@@
 \item LMS works on array of primitive types, possibly array of structures broken into array of simple types.
-Since in DP problems, composite types are representing a single logical element, and since we want to benefit from coalesced accesses (and possibly storing structures into efficient on-chip shared memory), we do not want to break structures.
+Since in DP problems, composite types represent a single logical element, and since we want to benefit from coalesced accesses (and possibly storing structures into efficient on-chip shared memory), we do not want to break structures.
 Also we want to offer support for tuples, which are a convenient way to write data containers (also we want to support case classes and composites types).

### 2013-01-15 docs/report/implementation.tex
Report first pass finished.

@@
 The former provided us with their implementation, which turned out to address large serial problems whereas our focus was on smaller non-serial problems, the latter did not respond to our solicitations.
-\item Have an optimal implementation that can serve as target to be imitated and generalized by the code generation.
+\item Have an optimal implementation that can serve as a   to be imitated and generalized by the code generation.
 \ole
@@
 \subsubsection{Results}
-We present here the timings of our implementation.
-For correctness, we first implemented a CPU version that we used to compare CUDA results against.
+We present here the timings of our ad-hoc implementation.
+For correctness, we first implemented a CPU single thread version (in C) that we used to compare CUDA results against.
 Input data is made of random numbers.
 The implemented dynamic programming problems are:\ul
-\item Rectangle: Smith-Waterman with arbitrary cost
-\item Triangle: matrix chain multiplication
-\item Parallelogram: polygon triangulation using a matrix larger than necessary.
+\item Rectangle: Smith-Waterman with arbitrary cost (\S\ref{swat_arbitrary})
+\item Triangle: matrix chain multiplication (\S\ref{mat_mult_plain})
+\item Parallelogram: polygon triangulation (\S\ref{polygon_tri}) using a matrix larger than necessary (\S\ref{problems_end}).
 Note that this implementation uses at most 32 blocks to prevent dead locks on our hardware (restriction due to the number of concurrent threads on the device).
@@
 	\ule
-\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increase (there is one more dependency).
+\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increases (there is one more dependency).
 According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity.
-However, in our setup (memory layout+algorithm+hardware), we did not found any improvement by doing so.
+However, in our setup (memory layout+algorithm+hardware), we did not find any improvement by doing so.
 We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
@@
 
-\item On a 4096 matrix, if we assume that ATLP card would have the same bandwidth as our card, their running time would be
+\item On a 4096 matrix, if we assume that the \cite{gpu_atlp} card would have the same bandwidth as our card, their running time would be
 \[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 9.43\rm s_{\text{ ATLP}} > 8.84\rm s_{\text{ our}}\]
-Which shows that our algorithm is comparable to theirs.
-However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have ATLP source code).
+This shows that our algorithm is comparable to theirs.
+However, we must avoid a close comparison because the fundamental hardware differences would make a tight computation almost intractable (additionally, we do not have \cite{gpu_atlp} source code).
 \ule
-As a conclusion, (1) we must remain away to invalidate their result as previous hardware generations might be subject to more constraint to our hardware and (2) we are on par, if not better with one of the best current implementation.
+As a conclusion, (1) it seems that the technique used in \cite{gpu_atlp} only applies to legacy hardware, however this remains a supposition (as we can not compare) and (2) we are on par, if not better with one of the best current implementations.
 
-\item \textbf{Threads number:} reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
+\item \textbf{Number of threads:} reducing the number of threads launched at different splits of the algorithm (especially in latest splits in rectangular and triangular shapes) does not bring any speedup.
 Even worse, it slows down slightly the computation.
@@
 
-\item \textbf{Unrolling:} unrolling the inner loops (non-serial dependencies) a small number of time provide some speedup, for a 2048-matrix respectively 10.9\% (rectangle, $2.765\to 2.464$), 14.1\% (triangle, $1.427\to 1.225$) and 9.7\% (parallelogram $1.539\to 1.389$).
+\item \textbf{Unrolling:} unrolling the inner loops (non-serial dependencies) a small number of times provide some speedup, for a 2048-matrix respectively 10.9\% (rectangle, $2.765s\to 2.464s$), 14.1\% (triangle, $1.427s\to 1.225s$) and 9.7\% (parallelogram $1.539s\to 1.389s$).
 The best experimental number of unrolling is 5.
@@
 	\ule
-	Additionally, the BaseParser implements the analysis that are shared by both the Scala and the CUDA version: dead rules elimination, yield analysis and dependencies ordering.
+	Additionally, the BaseParser implements the analysis that is shared by both the Scala and the CUDA version: dead rules elimination, yield analysis and dependencies ordering.
 Finally, it provides some implicit functions to flatten nested tuples (that are constructed by multiple concatenations).
@@
 \subsubsection{Scala structures conversion (JNI)}
-Since Scala general types can be extremely complex and might heavily depend of the JVM, we want to restrict the supported types; additionally types should be of fixed size for more efficient processing and easier memory allocation.
+Since general Scala types can be extremely complex and might depend of the JVM context (file stream, closures, ...), we want to restrict the supported types; additionally types should be of fixed size for more efficient processing and easier memory allocation.
 We support the following types:\ul
@@
 \subsubsection{Host wrappers}
-The host wrappers are the functions bridging between JNI and CUDA; their duties are:\ul
+Host wrappers are functions bridging between JNI and CUDA; their duties are:\ul
 \item Exposing JNI parsing and backtracking functions
@@
 
-\textbf{Bonus: problems larger than device memory}\\
+\textbf{Problems larger than device memory}\\
 Problems larger than the device memory can actually be processed on recent CUDA devices (with CUDA architecture $\ge 2.0$) as these are able to address the main memory from the device.
@@
 \subsubsection{Backtracking on the GPU}
-The backtracking is processed similarly as with the Scala parser, the major difference being that since we are generating the C code, we can provide an immediate mapping from the subrule index to the backtrack elements to add to the trace.
+The backtracking is processed similarly to the Scala parser, the major difference being that since we are generating C code, we can provide an immediate mapping from the subrule index to the backtrack elements to add to the trace.
 The backtrack is done in 3 steps:\ul
-\item If window is set, the windowing aggregation kernel is run to determine the position of the best result within the matrix.
+\item If a window is set, the windowing aggregation kernel is run to determine the position of the best result within the matrix.
 Otherwise the best position can be found at the last computed element of the matrix.
@@
 Finally, since the trace has to be reversed, we can obtain this transformation for free by constructing the trace list from the end in the JNI conversion.
-Reversing the list present the advantage that trace is immediately usable to construct the desired element.
-It might be possible that Scala and CUDA parser provide different traces to construct the same result, because the trace verifies the dependency order, which is only a partial order.
+Reversing the list presents the advantage that the trace is immediately usable to construct the desired element.
+It might be possible that Scala and CUDA parsers provide different traces to construct the same result, because the trace verifies the dependency order, which is only a partial order.
 

### 2013-01-15 docs/report/introduction.tex
Report first pass finished.

@@
 \newpage
-\section{Introduction}
+\section{Introduction} \label{intro}
 Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
@@
 To solve these concerns, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming algorithms into four distinct components that are tightly connected:\ol
-\item The search space is described using a tree \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
+\item The search space is described using a context-free \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
 \item Constructed candidates are then evaluated by a \textbf{scoring function} (where all these functions form an \textbf{algebra}), so that they can be compared appropriately.
@@
 It smoothly integrates features of object-oriented and functional languages, enabling programmers to be more productive.
-Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org}}
+Many companies depending on Java for business critical applications are turning to Scala to boost their development productivity, applications scalability and overall reliability.>>}\footnote{\url{http://www.scala-lang.org/node/25}}
 

### 2013-01-15 docs/report/problems.tex
Report first pass finished.

@@
 
-\subsubsection{Conclusion}
+\subsubsection{Conclusion} \label{problems_end}
 In the rest of the report, we use a different description of the problems that is based on ADP \cite{adp}, which is more convenient but does not share much with the above description (even though ultimately the executed computations are very similar).

### 2013-01-15 docs/report/implementation.tex
Nits.

@@
 % ------------------------------------------------------------------------------------------------
-\subsection{Ad-hoc CUDA} \label{ad_hoc_impl}
-In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
+\subsection{CUDA baseline} \label{baseline_impl}
+In the project planning, an baseline implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
 \item Better understand the challenges in CUDA implementation of dynamic programming problems and get on par with state-of-art implementations.
@@
 \subsubsection{Results}
-We present here the timings of our ad-hoc implementation.
+We present here the timings of our baseline implementation.
 For correctness, we first implemented a CPU single thread version (in C) that we used to compare CUDA results against.
@@
 \subsubsection{Matrix computation scheduling} \label{matrix_scheduling}
-Similarly as in the ad-hoc implementation, progress is made along the diagonal (see \ref{mem_layout}) and each thread is responsible of one line.
+Similarly as in the baseline implementation, progress is made along the diagonal (see \ref{mem_layout}) and each thread is responsible of one line.
 That is, the matrix is swept horizontally by a <<diagonal of threads>>, that are enabled only if they are within a valid matrix cell.

### 2013-01-15 docs/report/introduction.tex
Nits.

@@
 In the textbooks, problems are usually described in terms of recurrence relations between matrices elements.
-However, matrices computation formulae and recurrences might be difficult to express, indexing element is often error prone and writing an efficient parallel implementation can take a significant amount of time.
+Expressing dynamic programming problems in terms of recursive formulae involving matrix indices might be difficult, if often error prone, and the notation does not capture the essence of the underlying problem (for example aligning two sequences).
+Moreover, writing correct and efficient parallel implementation require different competencies and potentially a significant amount of time.
 
@@
 DynaProg allows the programmer to write concise programs based on ADP \cite{adp}, using a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.
-We evaluate the performance of our implementation against existing work and own ad-hoc implementations for both the CPU and GPU versions.
+We evaluate the performance of our implementation against existing work and own baseline implementations for both the CPU and GPU versions.
 Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU).
@@
 I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advice and suggestions.
-I hope you will enjoy your reading.
+I hope you will enjoy reading this report.
 \vspace{.3cm}\\
@@
 \section{Introduction} \label{intro}
-Dynamic programming (DP) is a technique used to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
-Dynamic programming often allows to find solutions in an exponential search space (number of solutions) in polynomial running time.
-The user is usually interested in one optimal solution of the problem, but he might also desire to have \textit{all} optimal solutions (co-optimal), a fixed number of near-optimal solutions, or some synthetic properties of the search space such as its size or the sum of all scores.
+Dynamic programming (DP) is a technique to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
+This property allows to find solutions in an exponential search space in polynomial running time by memoizing intermediate optimal solutions and reusing them multiple times.
+Dynamic programming problems are usually described in the textbooks by matrices recurrence relations involving indices and optimality is defined in terms of an objective function (maximum/minimum cost, ...).
+The computation is generally split in two phases: first intermediate scores are memoized in a matrix (tabulation) and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
+This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
 
-Dynamic programming problems arise in several disciplines of applied Computer Science such as biosequence analysis, natural language processing and operational research: sequence alignment, RNA sequence folding or expression parenthesisation are example of such problems which are usually described by matrix recurrence relations.
-Unfortunately, these problems often appear in multiple variations and with a considerable degree of sophistication such that there is a mismatch between the textbook solution and its concrete implementation.
-Additionally, debugging is tedious and requires a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences\cite{gapc_yield}.
-
-Dynamic programming computation is usually split in two phases: first intermediate scores are memoized in a matrix (tabulation), and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
-This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
-The obtained information then needs to be presented in a form that makes sense for the user (or  might drive further computations).
-These two phases need to be kept consistent with each other, thereby introducing a new potential source of errors.
-
+Dynamic programming problems arise in several disciplines of applied Computer Science such as biosequence analysis, natural language processing and operational research: sequence alignment, RNA sequence folding or expression parenthesisation are examples of such problems.
+Unfortunately, these often appear in multiple variations and with a considerable degree of sophistication such that there is a mismatch between the textbook solution and its concrete implementation.
+The user is often interested in one optimal solution, but he might also request \textit{all} co-optimal solutions, a fixed number of near-optimal solutions, or some synthetic properties of the search space (size, sum of scores, ...).
+The backtracking is usually ad-hoc because it needs both to be kept consistent with matrix filling and present the information in a format suitable for the user (human readable or ready to drive further computations).
+Additionally, debugging matrix indices is tedious and requires a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences \cite{gapc_yield}.
 Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA).
@@
 
-% http://en.wikipedia.org/wiki/Regular_tree_grammar: language class between regular languages and the deterministic context-free languages.
-To solve these concerns, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming algorithms into four distinct components that are tightly connected:\ol
-\item The search space is described using a context-free \textbf{parsing grammar} that describes how to construct intermediate candidates whose score might be inserted in the matrix.
+To overcome these difficulties, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming on sequences into four distinct components that are tightly connected:\ol
+\item The search space is described by a context-free \textbf{parsing grammar} that produces intermediate solution candidates whose score might be inserted in the matrix.
 \item Constructed candidates are then evaluated by a \textbf{scoring function} (where all these functions form an \textbf{algebra}), so that they can be compared appropriately.
@@
 
-By using a parsing grammar, ADP makes the candidate structure explicit.
-A signature serves as interface between the grammar, the scoring algebra and the aggregation function making possible that different grammars share different algebra or vice versa.
-Tabulation indices issues are hidden from the programmer, thereby removing potential errors.
-Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and algebra, it becomes possible to convert it to efficient recurrences for many-core platforms such as GPUs.
-\cite{adp_gpu}
+A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, making possible that the same grammar share different algebrae or vice versa.
+Because recurrence relations are expressed by a parsing grammar, ADP makes the candidate structure explicit and hides tabulation indices, thereby preventing potential errors.
+Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and an algebra, it becomes possible to systematically convert dynamic programming descriptions into efficient recurrences for many-core platforms such as GPUs \cite{adp_gpu}.
 
 % -----------------------------
-DynaProg implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
+DynaProg, the DSL we present in this report, implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
 It extends ADP by allowing grammars for pairing two sequences (multi-track grammars) similarly as GAPC\cite{gapc_thesis}, simplifies the process of writing programs by inferring additional information (\S\ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart (\S\ref{benchmarks}).
-Since the program structure is formalized in ADP framework, it can be analyzed to remove unused grammar productions (dead code elimination) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, thereby producing an arguably more reliable program.
+Since the program structure is formalized in ADP framework, it can be analyzed to remove unused grammar rules (\S\ref{dead_rules}) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, thereby producing an arguably more reliable program.
 
-DynaProg provides a generic way of backtracking the results, such that the same backtrack trace can be used with multiple algebras if they share the same grammar.
-This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost function; then from the backtrack of its optimal, the desired result is computed.
+DynaProg provides a generic way of backtracking the results such that the same trace can be used with different algebras sharing the same grammar.
+This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost functions; then from the backtrack of its optimal, the desired result is computed.
 As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently: first, optimal execution scheduling (or parenthesization) trace is found using dynamic programming and cost algebra (\S\ref{mat_mult_plain}).
@@
 
-This project resulted is an open-source\footnote{\url{https://github.com/manojo/lamp-dp-mt}} implementation of dynamic programming parsers for Scala optimized for both on CPU and GPU and featuring several analyses (\S\ref{architecture}) to ease the writing of dynamic programs.
-Its contribution is an automated approach to encode and process backtracking information such that the reconstruction complexity is reduced compared to \cite{gapc_thesis} and backtrack trace can be exchanged among different algebras sharing the same grammar.
+This project is currently available online\footnote{\url{https://github.com/manojo/lamp-dp-mt}}; it implements dynamic programming parsers in Scala (CPU) and CUDA (GPU).
+Its contribution is an novel approach to systematically encode and process backtracking information such that the reconstruction complexity is reduced compared to \cite{gapc_thesis}, and backtrack trace can be exchanged among different algebras sharing the same grammar.
 
 The rest of the document consists of:\ul
-\item A brief background on dynamic programming, followed by an explanation of some of the Scala programming language and LMS framework features (\S\ref{background}).
-\item A classification of DP problems in terms of matrix shape and dependencies, followed by a detailed analysis of some specific problems is provided (\S\ref{problems}).
+\item A brief background on dynamic programming, followed by an introduction to some of the key features of the Scala programming language and LMS framework (\S\ref{background}).
+\item A classification of DP problems in terms of matrix shape and dependencies, followed by a detailed analysis of some specific problems (\S\ref{problems}).
 Related work addressing dynamic programming challenges is presented in (\S\ref{related_work}).
-\item A description of the whole parser stack (\S\ref{architecture}), going from the user facing language (\S\ref{user_lang}, \S\ref{adp_grammar}) to optimizations (\S\ref{recurrences}, \S\ref{backtracking}) and implementation constraints (\S\ref{normalization}, \S\ref{memory_constr}), describing all the architectural decisions we made.
-\item The concrete implementation of these ideas (\S\ref{implementation}) in the form of a DSL for Scala (\S\ref{scala_parsers}) and in efficient CUDA code generation (\S\ref{codegen}).
+\item A description of the parser stack (\S\ref{architecture}), going from the user facing language (\S\ref{user_lang}, \S\ref{adp_grammar}) to optimizations (\S\ref{recurrences}, \S\ref{backtracking}) and implementation constraints (\S\ref{normalization}, \S\ref{memory_constr}), describing all the architectural decisions we made.
+\item The concrete implementation of these ideas (\S\ref{implementation}) in the form of a DSL in Scala (\S\ref{scala_parsers}) and in efficient CUDA code generation (\S\ref{codegen}).
 \item A brief usage explanation detailing the available features for the DSL user (\S\ref{usage}).
@@
 \newpage
+% Is Fibonacci a good example wrt what we describe in the rest of the report? After all we only deal with problems on sequences ?
+% - A typical characteristic => As one can see, a Fibonacci number is described __recursively__ in terms of "smaller" Fibonacci numbers.
+The naive algorithm for computing F(n) would require recomputing subsolutions many times (here F(3) helps ...
+).
+Reusing ...
+F(n) - 1 additions!
+% - any links to tree-related algorithms?
+% - "input are usually few thousands" => input is usually few thousands of K long ?
+% - Scala
+%  - However, we would list some ... => ; this is of course not the only reason; Scala offers some interesting features which make it a nice language to develop a DSL in ?
+%  - large adoption in the industry is not really a feature is it?
+%  - Implicit functions definition => implicit functions and parameters offer a way to neatly integrate primitive/pre-existing constructs into a DSL, by converting their types in order to correspond to the DSL semantics, while at the same time guarantee-ing type safety.
+%  - Macros[6] and LMS are mechanisms to modify semantics of a program; in particular, these tools can be used to implement domain-specific optimizations of programs.
+LMS also contains a code generation platform which we use to generate some functions in C (link to section).
+%  - Mixin composition (traits) allow for multiple inheritance and composition, which makes it easy to represent, in our DSL, multiple algebrae, and also interface with different runtimes (link to section).
+
 \section{Background} \label{background}
 \subsection{Dynamic programming} \label{bg_dp}
-Dynamic programming consists of solving a problem by constructing its solution from solutions to subproblems.
-A famous example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
+A classic example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
 \[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]

### 2013-01-15 docs/report/problems.tex
Nits.

@@
 
-Although we have not found a real problem requiring a parallelogram matrix, we still present this version in \S\ref{mem_layout} and \S\ref{ad_hoc_impl}.
+Although we have not found a real problem requiring a parallelogram matrix, we still present this version in \S\ref{mem_layout} and \S\ref{baseline_impl}.
 Such matrix layout could be adapted for cyclic problems that could be broken into a linear sequence anywhere (that is for all position in the circular structure, break the cycle at this position, and solve the dynamic programming problem on the resulting flattened sequence).

### 2013-01-15 src/main/scala/v4/report/RNAFold.scala

@@ -1,0 +1,74 @@
+package v4.report
+import v4._
+
+trait RNAFoldSig extends RNASignature {
+  def hairpin(ij:(Int,Int)):Answer
+  def stack(i:Int,s:Answer,j:Int):Answer
+  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
+  def mloop(i:Int,s:Answer,j:Int):Answer
+  def left(l:Answer,r:Int):Answer
+  def right(l:Int,r:Answer):Answer
+  def join(l:Answer,r:Answer):Answer
+}
+
+trait RNAFoldAlgebra extends RNAFoldSig {
+  type Answer = Int
+  import librna.LibRNA._ // indexing convention: first base,last base
+  private def Eh(i:Int,j:Int) = hl_energy(i,j) // hairpin loop
+  private def Ei(i:Int,k:Int,l:Int,j:Int) = il_energy(i,k,l,j) // internal loop
+  private def Es(i:Int,j:Int) = sr_energy(i,j) // stacked base pairs
+  def hairpin(ij:(Int,Int)) = Eh(ij._1,ij._2-1)
+  def stack(i:Int,s:Int,j:Int) = Es(i,j) + s
+  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) = Ei(ik._1,ik._2,lj._1-1,lj._2-1) + s
+  def mloop(i:Int,s:Int,j:Int) = s
+  def left(l:Int,r:Int) = l
+  def right(l:Int,r:Int) = r
+  def join(l:Int,r:Int) = l+r
+  override val h = min[Answer] _
+}
+
+trait RNAFoldPrettyPrint extends RNAFoldSig {
+  type Answer = String
+  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
+  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
+  def stack(i:Int,s:String,j:Int) = "("+s+")"
+  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
+  def mloop(i:Int,s:String,j:Int) = "("+s+")"
+  def left(l:String,r:Int) = l+"."
+  def right(l:Int,r:String) = "."+r
+  def join(l:String,r:String) = l+r
+}
+
[…]

### 2013-01-15 docs/report/introduction.tex
Nits.

@@
 Dynamic programming (DP) is a technique to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
-This property allows to find solutions in an exponential search space in polynomial running time by memoizing intermediate optimal solutions and reusing them multiple times.
+This property allows to find solutions in an exponential search space (meaning an exponential running time) in polynomial running time by memoizing intermediate optimal solutions and reusing them multiple times.
 Dynamic programming problems are usually described in the textbooks by matrices recurrence relations involving indices and optimality is defined in terms of an objective function (maximum/minimum cost, ...).
-The computation is generally split in two phases: first intermediate scores are memoized in a matrix (tabulation) and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
+The computation is generally split in two phases: first intermediate optimal solutions are memoized in a matrix (tabulation) and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
 This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
@@
 \newpage
+\section{Background} \label{background}
+\subsection{Dynamic programming} \label{bg_dp}
 % Is Fibonacci a good example wrt what we describe in the rest of the report? After all we only deal with problems on sequences ?
@@
 Reusing ...
-F(n) - 1 additions!
-% - any links to tree-related algorithms?
-% - "input are usually few thousands" => input is usually few thousands of K long ?
-% - Scala
-%  - However, we would list some ... => ; this is of course not the only reason; Scala offers some interesting features which make it a nice language to develop a DSL in ?
-%  - large adoption in the industry is not really a feature is it?
-%  - Implicit functions definition => implicit functions and parameters offer a way to neatly integrate primitive/pre-existing constructs into a DSL, by converting their types in order to correspond to the DSL semantics, while at the same time guarantee-ing type safety.
-%  - Macros[6] and LMS are mechanisms to modify semantics of a program; in particular, these tools can be used to implement domain-specific optimizations of programs.
-LMS also contains a code generation platform which we use to generate some functions in C (link to section).
-%  - Mixin composition (traits) allow for multiple inheritance and composition, which makes it easy to represent, in our DSL, multiple algebrae, and also interface with different runtimes (link to section).
-
-\section{Background} \label{background}
-\subsection{Dynamic programming} \label{bg_dp}
+F(n)-1 additions!
 A classic example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
@@
 
-A typical characteristic is that an intermediate solution is reused multiple times to construct larger solutions (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
+As one can see, Fibonacci numbers are described recursively in terms of previous series elements, which are reused multiple times to construct the following elements (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
 Reusing an existing solution avoids redoing expensive computations: with memoization (memorizing intermediate results), the solution for $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !
@@
 \item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
-\item Tree-related algorithms: phylogenetic, trees raking, maximum tree independent set, ... (can be viewed as a sparse version of the second category)
+\item Tree-related algorithms: phylogeny \cite{philogeny}, trees raking \cite{tree_raking}, maximum tree independent set \cite{tree_independent_set}, ...
+(can be viewed as a sparse version of the second category)
 \ule
@@
 
-Taking real-world examples in biology, the average input size for sequence alignment (\S\ref{swat_affine}) is around 300K whereas for problems like RNA folding (\S\ref{zuker}), input are usually around few thousands.
+Taking real-world examples in biology, the average input size for sequence alignment (\S\ref{swat_affine}) is around 300'000 whereas for problems like RNA folding (\S\ref{zuker}), input length is usually below 1000.
 Problems operating on multiple input sequences also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex (as intermediate results needs to be stored in a position representing the progress in each of the involved sequence).
@@
 \subsection{Scala} \label{bg_scala}
+%  - However, we would list some ... => ; this is of course not the only reason; Scala offers some interesting features which make it a nice language to develop a DSL in ?
+%  - large adoption in the industry is not really a feature is it?
+%  - Implicit functions definition => implicit functions and parameters offer a way to neatly integrate primitive/pre-existing constructs into a DSL, by converting their types in order to correspond to the DSL semantics, while at the same time guarantee-ing type safety.
+%  - Macros[6] and LMS are mechanisms to modify semantics of a program; in particular, these tools can be used to implement domain-specific optimizations of programs.
+LMS also contains a code generation platform which we use to generate some functions in C (link to section).
+%  - Mixin composition (traits) allow for multiple inheritance and composition, which makes it easy to represent, in our DSL, multiple algebrae, and also interface with different runtimes (link to section).
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.

### 2013-01-16 docs/report/report.tex
Report first pass finished.
Nits.

@@
 
+\begin{verbatim}
+NEW:
+- CPU: Dynaprog vs ADPFusion vs RNAfold vs GAPC
+- GPU: (Zuker) Dynaprog vs GAPC (si on arrive a faire marcher)
+- GPU: (MatMult) Dynaprog vs ATLP
+- GPU: (SWat) Dynaprog vs  CUDAlign
+\begin{verbatim}
+
 RUN 5-10x until running time stabilizes on Scala
+
+ {\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
+
+XXX: future work: proper support of two-track grammars
+
+XXX: future work, resize tables appropriately as in GAPL.
 

### 2013-01-16 src/librna/misc/test.c

@@ -3,7 +3,7 @@
 #include <stdlib.h>
 #include <string.h>
 
-// FLAGS:= -std=c99 ../librna.c ../vienna/*.c
+// FLAGS:= ../vienna/*.c
 
 #define my_len c_len
 #define my_seq c_seq
@@ -19,55 +19,45 @@
 
 void freeSeq() { free(c_seq); free(c_P); }
 void initSeq(const char* param, const char* seq) {
-  read_parameter_file(param);
+  int i; read_parameter_file(param);
   c_P = get_scaled_parameters();
   c_len = strlen(seq);
   c_seq = malloc(c_len+1);
-  for (int i=0; i < c_len; ++i) switch (seq[i]) {
-    case 'a' : c_seq[i] = 1; break;
-    case 'c' : c_seq[i] = 2; break;
-    case 'g' : c_seq[i] = 3; break;
-    case 'u' : c_seq[i] = 4; break;
+  for (i=0; i<c_len; ++i) switch (seq[i]) {
+    case 'a': c_seq[i]=1; break;
+    case 'c': c_seq[i]=2; break;
+    case 'g': c_seq[i]=3; break;
+    case 'u': c_seq[i]=4; break;
     default: exit(1);
   }
 }
 
-
-int E_stack(int i, int j) {
+int E_stack(int i, int j) { // = sr_energy
   return c_P->stack[_bp(i,j)][_bp(j-1,i+1)];
 }
 
 
 
 int main(int argc, char **argv) {
-  const char* s=
-     "ugcucaggca";
-  // (((....)))   len=10
-  //  --    --    hairpin(1,8)
-  // -        -   sr(0,9)
+  const char* s= "agccccgguuaagaauaaaggagauuucuccgcccaaccccuguaaugcu";
 
   initSeq("../vienna/rna_turner2004.par",s);
 
+/*
   printf("%d -- /\n",ext_mismatch_energy(0, 9)); // dlr
   printf("%d -- /\n",termau_energy(0,9));
-
   printf("%d -- %d\n",sr_energy(0,9), E_stack(0,9)); // stack
   printf("%d -- %d\n",sr_energy(1,8), E_stack(1,8)); // hairpin
   printf("%d -- %d\n",hl_energy(2,7), E_Hairpin(4,_bp(2,7),3,6,"CUCAGG",c_P));
-/*
- *  \param  size  The size of the loop (number of unpaired nucleotides)
- *  \param  type  The pair type of the base pair closing the hairpin
[…]

### 2013-01-16 docs/report/report.tex
Nits.

@@
 \input{inc/style.sty}
-\usepackage{multicol}
+\usepackage{multicol,nameref}
 \usepackage{etoolbox}
@@
 \section{Usage} \label{usage}
+
+Example based
+- Matrix mult + parenthesizing
+- SWat / Needleman
+
+go check the examples/ folder
+
+List of options at the end
+
 {\color{red} usage, make reference from problems to usage
@@
 
-\begin{verbatim}
-NEW:
-- CPU: Dynaprog vs ADPFusion vs RNAfold vs GAPC
-- GPU: (Zuker) Dynaprog vs GAPC (si on arrive a faire marcher)
-- GPU: (MatMult) Dynaprog vs ATLP
-- GPU: (SWat) Dynaprog vs  CUDAlign
-\begin{verbatim}
+\subsubsection{Test Scala / C / CUDA}
+\subsubsection{Test 1 : CPU MatrixMult VS hand optimized VS GAPC VS ADPFusion)}
+\subsubsection{Test 2 : GPU MatirxMult VS hand optimized VS ATLP)}
+\subsubsection{Test 3 : CPU SWat VS hand optimized VS GAPC VS ADPFusion)}
+\subsubsection{Test 4 : GPU SWat VS hand optimized VS CUDAlign)}
+\subsubsection{Test 5 : CPU Zuker VS GAPC VS ADPFusion VS ViennaRNA)}
+\subsubsection{Test 6 : GPU Zuker VS RNAFold?/Lavenier VS ViennaRNA-OpenCL)}
 
@@
 
-\subsubsection{Test 1 : Ad-hoc VS CUDA parsers (implies ATLP rescaled comparison)}
-
-\subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}
-Also CUDASW++
-
-\subsubsection{Test 3 : GAPC vs Java parsers vs CUDA parsers}
-
-\subsubsection{Test 4 ? : GPU RNAfolding vs CUDA parsers}
-
-\subsection{Java parsers}
+\subsection{Scala parsers}
 {\color{red}
@@
 {\color{red} XXX}
+
+\begin{verbatim}
+
+recap of the report
+
+recap of the results
+
+big picture: with actual compiler technology we 
+Embedded DSL vs external DSL
+Larger picture how to go further
+
+\end{verbatim}
 

### 2013-01-16 docs/report/design.tex
Nits.
Report reordering.

@@
 \ule
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{Parsing grammar (ADP)} \label{adp_grammar}
-In this section, we provide a concise description of the Algebraic Dynamic Programming.
-Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.
-
-ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
-Such separation makes it possible to define multiple algebra for the same grammar.
-This has two main applications:\ol
-\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
-\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data.
-For example in matrix chain multiplication, the first step solves the underlying dynamic program by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
-\ole
-
-Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebrae} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
-The concrete program instance combines the algebra with the grammar.
-The grammar parsers' intermediate results are memoized in a matrix (tabulation parser).
-A parser usually consist of a tree of:\ul
-\item \textbf{Terminal:} operates on a subsequence of input elements and returns either its content or position (or a failure if the sequence does not fit the terminal).
-\item \textbf{Filter:} accepts only subsequences matching a certain predicate. The condition is evaluated ahead of its actual content evaluation.
-\item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
-\item \textbf{Concatenation:} constructs all possible combinations from two subsequences.
-The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
-\item \textbf{Map:} this parser transform its input using a user-defined function. It is typically used to transform a subword into a score that can later be aggregated.
-\item \textbf{Aggregation:} the aggregation applies a functions that reduces the list of results, typically minimum or maximum, but the function can be arbitrarily defined.
-\item \textbf{Tabulation:} the tabulation's primary function is to store intermediate results and possibly serve as connection point between different parsers.
-\ule
-
-Additionally, the signature must define an input alphabet ({\tt Alphabet}), and an output alphabet ({\tt Answer}) can be defined either in the signature or in the algebra.
-Finally, the grammar needs to have a starting point, denoted as axiom.
-Finally, the default aggregation function $h$ must be defined\footnote{Although aggregation usage is not mandatory in the framework, we force the existence of an aggregation function over the ouput type so that we can use it to aggregate windowed results.}.
-To make it more clear, we propose an example of the matrix chain multiplication problem\footnote{The original ADP framework was embedded in Haskell, however, we assume that the reader is more familiar with Scala notation and immediately present the syntax of our implementation.}.
-
-\newpage
-\begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
-trait MatrixSig extends Signature {
-  type Alphabet = (Int,Int) // Matrix(rows, columns)
-  val single:Alphabet=>Answer
-  val mult:(Answer,Answer)=>Answer
-}
-
-trait MatrixAlgebra extends MatrixSig {
-  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
-  override val h = minBy((a:Answer) => a._1)
-  val single = (a: Alphabet) => (0, a)
-  val mult = (a:Answer,b:Answer) =>
-    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
-}
-
-trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val axiom:Tabulate = tabulate("M",
-    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
-}
-
-object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
-  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
-}
-\end{lstlisting}
-\begin{center}\vspace{-18pt} with or: $|\,\,$ map: $\land\land\,\,$ concatenation: $\sim$ \end{center}
-
-%\vspace{3cm}
-This program grammar can also be expressed in BNF\footnote{\url{http://en.wikipedia.org/wiki/Backus-Naur\_Form}}:
-\[\begin{array}{lrl}
-axiom &::=& \text{matrix} \\
- &|& axiom \,\,\, axiom
-\end{array}\]
-
-and it encodes the following recurrence (cost only):
-	\[M_{(i,j)}=\left\{\begin{array}{ll}
-		0 & \text{if } i+1= j\\
-		\min_{i<k<j} M_{(i,k)}+M_{(k,j)}+r_i \cdot c_k \cdot c_j & \text{otherwise}
-	\end{array}\right. \]
-
-Notice that the semantics of indices differ slightly from the problem presented in \ref{mat_mult_plain}; this is because empty chain are made expressible (denoted $M_{(i,i)}$, single matrices are denoted $M_{(i,i+1)}$).
 

### 2013-01-16 docs/report/introduction.tex
Nits.
Report reordering.

@@
 \section{Introduction} \label{intro}
-Dynamic programming (DP) is a technique to solve combinatorial optimization problems that verify the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal solutions of sub-problems.
-This property allows to find solutions in an exponential search space (meaning an exponential running time) in polynomial running time by memoizing intermediate optimal solutions and reusing them multiple times.
-Dynamic programming problems are usually described in the textbooks by matrices recurrence relations involving indices and optimality is defined in terms of an objective function (maximum/minimum cost, ...).
-The computation is generally split in two phases: first intermediate optimal solutions are memoized in a matrix (tabulation) and reused to construct scores for larger problems; then a backtrack stage retrieves the solution associated with the optimal score for the problem.
-This solution describes how to obtain the optimal score and is called trace or backtrack trace, and heavily depends on the matrix design.
+Dynamic programming (DP) is an algorithmic technique to solve optimization problems.
+For example, we might want to multiply a chain of matrices\footnote{Such that matrices have appropriate dimensions to be multiplied with each other.} efficiently.
+The order in which matrices are combined changes the number of required scalar multiplications, we would want to find an optimal order that minimizes this number.
+Notice that if knew how to split of the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them.
+Such combinatorial problems verify the Bellman's principle\cite{bellman_principle}: \textit{<<optimal solutions depends on optimal solutions of sub-problems>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}.
+Because we don't know a priori where we could split the chain, we would need to explore an exponential number of possibilities.
+Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, hereby reducing the problem to a polynomial complexity.
+
+Dynamic programming problems are usually expressed in terms of recurrences on intermediate solutions that are stored in matrices, whereas optimality is defined in terms of an objective function (min/max cost, ...).
+In the case of a chain of $n$ matrices, the recurrence is:
+	\[M_{(i,i)}=0 \quad\land\quad M_{(i,j)}=\min_{1\le i\le k<j\le n}\{M_{(i,k)}+M_{(k+1,j)}+r_i \cdot c_k \cdot c_j\} \qquad \forall 1\le i,j\le n\]
+where $r_i$ and $c_i$ denotes respectively the row and column of the $i^{\rm th}$ matrix in the chain, $M$ is an $n \times n$ matrix and $M_{(i,j)}$ stores the number of multiplications to obtain the product of matrices in the chain $i,...,j$.
+The total number of required scalar multiplications is given by $M_{(1,n)}$ (refer to \S\ref{mat_mult_plain} for details).
+Once the optimal result is found, a second backtrack phase retrieves the construction trace associated with the optimal score for the problem.
+This trace (or backtrack trace) describes how to obtain the optimal score, and heavily depends on the matrix design.
 
@@
 The user is often interested in one optimal solution, but he might also request \textit{all} co-optimal solutions, a fixed number of near-optimal solutions, or some synthetic properties of the search space (size, sum of scores, ...).
+
 The backtracking is usually ad-hoc because it needs both to be kept consistent with matrix filling and present the information in a format suitable for the user (human readable or ready to drive further computations).
 Additionally, debugging matrix indices is tedious and requires a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences \cite{gapc_yield}.
+
 Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA).
@@
 
-To overcome these difficulties, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming on sequences into four distinct components that are tightly connected:\ol
+\newpage
+To simplify the expression of dynamic problems, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming on sequences into four distinct components that are tightly connected:\ol
 \item The search space is described by a context-free \textbf{parsing grammar} that produces intermediate solution candidates whose score might be inserted in the matrix.
@@
 This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost functions; then from the backtrack of its optimal, the desired result is computed.
-As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently: first, optimal execution scheduling (or parenthesization) trace is found using dynamic programming and cost algebra (\S\ref{mat_mult_plain}).
+As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently: first, optimal execution scheduling (or parenthesisation) trace is found using dynamic programming and cost algebra (\S\ref{mat_mult_plain}).
 The backtrack trace is then used (with a multiplication algebra) to multiply the actual matrices.
@@
 
+\newpage
 This project is currently available online\footnote{\url{https://github.com/manojo/lamp-dp-mt}}; it implements dynamic programming parsers in Scala (CPU) and CUDA (GPU).
@@
 
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section{Background} \label{background}
-\subsection{Dynamic programming} \label{bg_dp}
+%Optimization problem: 
+%1. cut / recursivity
+%2. bellman
+%classic = recurrence
+%B. complex problems
+%C. backtrack
+%D. execute fast
+
+% XXX: grammar for all the examples of problems
+
+% XXX: background sur graphic cards
+
 % Is Fibonacci a good example wrt what we describe in the rest of the report? After all we only deal with problems on sequences ?
@@
 F(n)-1 additions!
-A classic example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
-\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
-which expands to (first 21 numbers)
-\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]
-
-As one can see, Fibonacci numbers are described recursively in terms of previous series elements, which are reused multiple times to construct the following elements (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
+%A classic example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
+%\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
+%which expands to (first 21 numbers)
+%\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]
+%
+%As one can see, Fibonacci numbers are described recursively in terms of previous series elements, which are reused multiple times to construct the following elements (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
 Reusing an existing solution avoids redoing expensive computations: with memoization (memorizing intermediate results), the solution for $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !
-
-Formally, dynamic programming problems respect the Bellman's principle of optimality\cite{bellman_principle}: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}.
+%
+%Formally, dynamic programming problems respect the Bellman's principle of optimality\cite{bellman_principle}: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}.
 This means that every intermediate result needs to be computed only once, although it might be reused as a basis for multiple larger problems, hence our first observation.
 
-There exist various categories of dynamic programming:\ul
-\item Series that operate usually on a single dimension (like Fibonacci)
-\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
-\item Tree-related algorithms: phylogeny \cite{philogeny}, trees raking \cite{tree_raking}, maximum tree independent set \cite{tree_independent_set}, ...
-(can be viewed as a sparse version of the second category)
+% ------------------------------------------------------------------------------------------------
+\newpage
+\section{Background} \label{background}
+\subsection{Graphic cards}
+{\color{red} background on graphic cards}
+
+\subsection{ADP and parsing grammars} \label{adp_grammar}
+In this section, we provide a concise description of the Algebraic Dynamic Programming.
+Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.
+
+{\color{red} recall some ADP formal definitions}
+%    - terms like algebra, signature, etc.
+might need to be defined more formally later on, because these are highly mathematical terms which, for example, Viktor could ask you questions on during your presentation
+
+ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
+Such separation makes it possible to define multiple algebra for the same grammar.
+This has two main applications:\ol
+\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
+\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data.
+For example in matrix chain multiplication, the first step solves the underlying dynamic program by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
+\ole
+
+Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebrae} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
+The concrete program instance combines the algebra with the grammar.
+The grammar parsers' intermediate results are memoized in a matrix (tabulation parser).
+A parser usually consist of a tree of:\ul
+\item \textbf{Terminal:} operates on a subsequence of input elements and returns either its content or position (or a failure if the sequence does not fit the terminal).
+\item \textbf{Filter:} accepts only subsequences matching a certain predicate. The condition is evaluated ahead of its actual content evaluation.
+\item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
+\item \textbf{Concatenation:} constructs all possible combinations from two subsequences.
+The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
+\item \textbf{Map:} this parser transform its input using a user-defined function. It is typically used to transform a subword into a score that can later be aggregated.
+\item \textbf{Aggregation:} the aggregation applies a functions that reduces the list of results, typically minimum or maximum, but the function can be arbitrarily defined.
+\item \textbf{Tabulation:} the tabulation's primary function is to store intermediate results and possibly serve as connection point between different parsers.
 \ule
 
-Since the first category operates on a single dimension, to benefit of the smaller solutions to compute larger ones, elements must be computed sequentially (one at a time), hence computations cannot be made parallel (unless duplicated, thereby hindering benefits of memoization).
-The third category suffers from limited parallelism \cite{philogeny} and its implementation does not share much with the previous category, hence we focus on the second type of problems.
+Additionally, the signature must define an input alphabet ({\tt Alphabet}), and an output alphabet ({\tt Answer}) can be defined either in the signature or in the algebra.
+Finally, the grammar needs to have a starting point, denoted as axiom.
+Finally, the default aggregation function $h$ must be defined\footnote{Although aggregation usage is not mandatory in the framework, we force the existence of an aggregation function over the ouput type so that we can use it to aggregate windowed results.}.
+To make it more clear, we propose an example of the matrix chain multiplication problem\footnote{The original ADP framework was embedded in Haskell, however, we assume that the reader is more familiar with Scala notation and immediately present the syntax of our implementation.}.
 
-Taking real-world examples in biology, the average input size for sequence alignment (\S\ref{swat_affine}) is around 300'000 whereas for problems like RNA folding (\S\ref{zuker}), input length is usually below 1000.
-Problems operating on multiple input sequences also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex (as intermediate results needs to be stored in a position representing the progress in each of the involved sequence).
-Since we target a single computer with one or more attached devices (GPUs, FPGAs), and since we plan to maintain data in memory (due to the multiple reuse of intermediate solutions) the storage complexity must be relatively limited, compared to other problem that could leverage the disk storage.
-Hence in general, we focus on problems that have $O(n^2)$-space complexity whereas time complexity is usually $O(n^3)$ or larger.
-We encourage you to refer to \S\ref{problems} for further classification and examples.
+\newpage
+\begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
+trait MatrixSig extends Signature {
+  type Alphabet = (Int,Int) // Matrix(rows, columns)
+  val single:Alphabet=>Answer
+  val mult:(Answer,Answer)=>Answer
+}
+
+trait MatrixAlgebra extends MatrixSig {
+  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
+  override val h = minBy((a:Answer) => a._1)
+  val single = (a: Alphabet) => (0, a)
+  val mult = (a:Answer,b:Answer) =>
+    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
+}
+
+trait MatrixGrammar extends ADPParsers with MatrixSig {
+  val axiom:Tabulate = tabulate("M",
+    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
+}
+
+object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
+  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
+}
+\end{lstlisting}
+\begin{center}\vspace{-18pt} with or: $|\,\,$ map: $\land\land\,\,$ concatenation: $\sim$ \end{center}
+
+%\vspace{3cm}
+This program grammar can also be expressed in BNF\footnote{\url{http://en.wikipedia.org/wiki/Backus-Naur\_Form}}:
+\[\begin{array}{lrl}
+axiom &::=& \text{matrix} \\
+ &|& axiom \,\,\, axiom
+\end{array}\]
+
+and it encodes the following recurrence (cost only):
+	\[M_{(i,j)}=\left\{\begin{array}{ll}
+		0 & \text{if } i+1= j\\
+		\min_{i<k<j} M_{(i,k)}+M_{(k,j)}+r_i \cdot c_k \cdot c_j & \text{otherwise}
+	\end{array}\right. \]
+
+Notice that the semantics of indices differ slightly from the problem presented in \ref{mat_mult_plain}; this is because empty chain are made expressible (denoted $M_{(i,i)}$, single matrices are denoted $M_{(i,i+1)}$).
 
@@
 A discussion of the integration of LMS in our project can be found in \S\ref{lms_use}.
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\subsection{Related work} \label{related_work}
+Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.
+The former focus on the performance for a specific problem whereas the latter generalize and formalize the dynamic programming problem description into a parsing grammar paired with a costing algebra.
+
+Grammar-based dynamic programming was inseminated by ADP \cite{adp_discipline} and first implemented as a Haskell DSL \cite{adp}. To overcome performance issues, multiple solutions were devised:\ul
+\item Converting Haskell parsers in their C or CUDA equivalent \cite{adp_gpu}
+\item Modifying Haskell execution environment to provide loop fusion to improve ADP parsers performance \cite{adp_fusion}, \cite{adp_fusion_pkg}.
+\item Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} provided with an ad-hoc compiler \cite{gapc_thesis}, thereby allowing more advanced analysis of the grammar \cite{gapc_yield}.
+\ule
+
+The research on ad-hoc implementation has focused on three kind of problems:\ul
+\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
+\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
+\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
+\ule
+
+Since this project involves various domains, we also investigated in the memory management on graphic cards and existing code generation frameworks.
+
+In an attempt to support a varying number of results per matrix cell, we considered dynamic memory allocation \cite{nvidia} (available on recent graphic cards), ad-hoc memory allocation \cite{scatter_alloc} and hash tables \cite{parallel_hashing}.
+However the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, and the use of cuckoo hash tables adds a constant factor penalty to every memory access.
+Finally both solution introduce undesirable possibility of failure (respectively out of memory or unrecoverable collision) in the middle of the algorithm computation process.
+
+Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS\cite{lms_thesis} to generate from the same source code efficient implementation for heterogeneous platforms (including CUDA) at runtime.
+Although this shares many patterns with our project, we can not reuse this framework because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{bg_lms}) whereas Delite focuses on parallelizing operations on collections (array, lists, maps, ...) of independent elements.

### 2013-01-16 docs/report/problems.tex
Report reordering.

@@
 \section{Dynamic programming problems} \label{problems}
+There exist various categories of dynamic programming:\ul
+\item Series that operate usually on a single dimension (like Fibonacci\footnote{http://en.wikipedia.org/wiki/Fibonacci_number})
+\item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
+\item Tree-related algorithms: phylogeny \cite{philogeny}, trees raking \cite{tree_raking}, maximum tree independent set \cite{tree_independent_set}, ...
+(can be viewed as a sparse version of the second category)
+\ule
+
+Since the first category operates on a single dimension, to benefit of the smaller solutions to compute larger ones, elements must be computed sequentially (one at a time), hence computations cannot be made parallel (unless duplicated, thereby hindering benefits of memoization).
+The third category suffers from limited parallelism \cite{philogeny} and its implementation does not share much with the previous category, hence we focus on the second type of problems.
+
+Taking real-world examples in biology, the average input size for sequence alignment (\S\ref{swat_affine}) is around 300'000 whereas for problems like RNA folding (\S\ref{zuker}), input length is usually below 1000.
+Problems operating on multiple input sequences also require more memory: for instance matching 3 sequences is $O(n^3)$-space complex (as intermediate results needs to be stored in a position representing the progress in each of the involved sequence).
+Since we target a single computer with one or more attached devices (GPUs, FPGAs), and since we plan to maintain data in memory (due to the multiple reuse of intermediate solutions) the storage complexity must be relatively limited, compared to other problem that could leverage the disk storage.
+Hence in general, we focus on problems that have $O(n^2)$-space complexity whereas time complexity is usually $O(n^3)$ or larger.
+We encourage you to refer to \S\ref{problems} for further classification and examples.
+
 % ------------------------------------------------------------------------------------------------
 \subsection{Problems classification}
-Since <<dynamic programming>> defines a very general technique, we already focused on grammar and alignment problems in \ref{bg_dp}.
+Since <<dynamic programming>> defines a very general technique, we already focused on grammar and alignment problems.
 Before exploring some particular problem instances, we want to define some characteristics that will be used through the rest of the document to describe dynamic programming problems.
@@
 \ole
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{Related work} \label{related_work}
-Work related to dynamic programming can be separated in two categories: ad-hoc implementations and grammar-based implementations.
-The former focus on the performance for a specific problem whereas the latter generalize and formalize the dynamic programming problem description into a parsing grammar paired with a costing algebra.
-
-Grammar-based dynamic programming was inseminated by ADP \cite{adp_discipline} and first implemented as a Haskell DSL \cite{adp}. To overcome performance issues, multiple solutions were devised:\ul
-\item Converting Haskell parsers in their C or CUDA equivalent \cite{adp_gpu}
-\item Modifying Haskell execution environment to provide loop fusion to improve ADP parsers performance \cite{adp_fusion}, \cite{adp_fusion_pkg}.
-\item Ultimately, the dynamic programming algebra and grammar were formalized into a specific language \cite{gapl} provided with an ad-hoc compiler \cite{gapc_thesis}, thereby allowing more advanced analysis of the grammar \cite{gapc_yield}.
-\ule
-
-The research on ad-hoc implementation has focused on three kind of problems:\ul
-\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
-\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
-\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
-\ule
-
-Since this project involves various domains, we also investigated in the memory management on graphic cards and existing code generation frameworks.
-
-In an attempt to support a varying number of results per matrix cell, we considered dynamic memory allocation \cite{nvidia} (available on recent graphic cards), ad-hoc memory allocation \cite{scatter_alloc} and hash tables \cite{parallel_hashing}.
-However the costs associated with dynamic memory allocation makes it unattractive for this particular kind of problem, and the use of cuckoo hash tables adds a constant factor penalty to every memory access.
-Finally both solution introduce undesirable possibility of failure (respectively out of memory or unrecoverable collision) in the middle of the algorithm computation process.
-
-Automated code generation and execution flow is addressed by Delite \cite{lms2}, \cite{lms3}, \cite{delite}, that leverages LMS\cite{lms_thesis} to generate from the same source code efficient implementation for heterogeneous platforms (including CUDA) at runtime.
-Although this shares many patterns with our project, we can not reuse this framework because the scheduling and computation is tightly interleaved in dynamic programming (see \ref{bg_lms}) whereas Delite focuses on parallelizing operations on collections (array, lists, maps, ...) of independent elements.

### 2013-01-16 docs/report/introduction.tex
Cleanup.

@@
 
-%Optimization problem: 
-%1. cut / recursivity
-%2. bellman
-%classic = recurrence
-%B. complex problems
-%C. backtrack
-%D. execute fast
-
 % XXX: grammar for all the examples of problems
 
-% XXX: background sur graphic cards
-
-% Is Fibonacci a good example wrt what we describe in the rest of the report? After all we only deal with problems on sequences ?
-% - A typical characteristic => As one can see, a Fibonacci number is described __recursively__ in terms of "smaller" Fibonacci numbers.
-The naive algorithm for computing F(n) would require recomputing subsolutions many times (here F(3) helps ...
-).
-Reusing ...
-F(n)-1 additions!
-%A classic example of dynamic programming is the Fibonacci series that is defined by the following recurrence:
-%\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]
-%which expands to (first 21 numbers)
-%\[1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584, 4181, 6765, 10946, ...\]
-%
-%As one can see, Fibonacci numbers are described recursively in terms of previous series elements, which are reused multiple times to construct the following elements (here $F(3)$ helps constructing $F(4)$ and $F(5)$).
-Reusing an existing solution avoids redoing expensive computations: with memoization (memorizing intermediate results), the solution for $F(n)$ would be obtained after $n$ additions whereas without memoization it requires $F(n)-1$ additions !
-%
-%Formally, dynamic programming problems respect the Bellman's principle of optimality\cite{bellman_principle}: \textit{<<An optimal policy has the property that whatever the initial state and initial decision are, the remaining decisions must constitute an optimal policy with regard to the state resulting from the first decision>>}.
-This means that every intermediate result needs to be computed only once, although it might be reused as a basis for multiple larger problems, hence our first observation.
-
 % ------------------------------------------------------------------------------------------------
-\newpage
 \section{Background} \label{background}
 \subsection{Graphic cards}
-{\color{red} background on graphic cards}
+Modern graphic cards\footnote{We cover here interesting features of the CUDA devices and programming paradigm; however, the same concept should be applicable to graphic cards from other vendors.} are powered by massively parallel processor: they can typically run hundreds or thousands of cores, each able to schedule multiple threads.
+The threads are usually grouped in warps that are scheduled synchronously.
+This means that if there is a divergence in the execution path, both alternative are executed sequentially, hereby stalling other warp's threads.
+Threads are logically grouped in blocks by the programmer whereas warps correspond to a physical constraint.
+In a deliberate design decision to simplify the hardware, there exist no global synchronization.
 
+On graphic cards, there exist two levels of memory that are visible for the programmer: the global memory, which can be accessed by any thread, and the shared memory, that corresponds to an explicitly addressable cache memory, whose access is faster but restricted to threads in the same block.
+Finally, access to the main memory of the computer is possible on recent cards but suffers an additional penalty, which makes it not desirable.
+
+Since in such architecture the major bottleneck is often the access to the global memory, threads should access contiguous memory at the same time.
+This is called coalesced memory access and improving the memory layout in this direction can lead to significant speedup\footnote{\url{http://mc.stanford.edu/cgi-bin/images/5/5f/Darve_cme343_cuda_2.pdf}}.
+
+\newpage
 \subsection{ADP and parsing grammars} \label{adp_grammar}

### 2013-01-17 docs/report/introduction.tex
Nits.

@@
 \subsection{ADP and parsing grammars} \label{adp_grammar}
-In this section, we provide a concise description of the Algebraic Dynamic Programming.
-Would the reader be interested in the mathematical details of the notation, we encourage him to read the section 3 of \cite{adp}.
+\subsubsection{ADP formal specifications}
+\textit{This subsection is an excerpt of "Algebraic Dynamic Programming" \cite{adp}, section 3.
+Would the reader be interested in more details, we encourage him to read the corresponding paper.} \\[6pt]
+\textbf{Terminology}\\
+An \textbf{alphabet} $\mathcal{A}$ is a finite set of symbols.
+Sequence of symbols are called strings.
+$\epsilon$ denotes the empty string, $\mathcal{A}^1=\mathcal{A}$, $\mathcal{A}^{n+1}=\{ax|a\in\mathcal{A},x\in\mathcal{A}^n\}$, $\mathcal{A}^+=\bigcup_{n\ge 1}\mathcal{A}^n,\mathcal{A}^*=\mathcal{A}^+\cup\{\epsilon\}$.
 
-{\color{red} recall some ADP formal definitions}
-%    - terms like algebra, signature, etc.
-might need to be defined more formally later on, because these are highly mathematical terms which, for example, Viktor could ask you questions on during your presentation
+A \textbf{signature} $\Sigma$ over some alphabet $\mathcal{A}$ consists of a sort symbol $S$ with a family of operators.
+Each operator $\circ$ has fixed arity: $\circ:s_1,...,s_k\to S$ where each $s_i$ is either $S$ or $\mathcal{A}$.
 
+A $\Sigma$-\textbf{algebra} $\mathcal{I}$ over $\mathcal{A}$, also called an interpretation, is a set $S_\mathcal{I}$ of values together with a function $\circ_\mathcal{I}$ for each operator $\circ$.
+Each $\circ_\mathcal{I}$ has type $\circ_\mathcal{I}: (s_1)_\mathcal{I} ...(s_k)_\mathcal{I} \to S_\mathcal{I}$ where $\mathcal{A}_\mathcal{I} = \mathcal{A}$.
+
+A \textbf{term algebra} $T_\Sigma$ arises by interpreting the operators in $\Sigma$ as constructors, building bigger terms from smaller ones.
+When variables from a set $V$ can take the place of arguments to constructors, we speak of a term algebra with variables $T_\Sigma(V)$ with $V\subset T_\Sigma(V)$.
+
+Terms will be viewed as rooted, ordered, node-labeled trees in the obvious way.
+According to the special role of $\mathcal{A}$, only leaf nodes can carry symbols from $\mathcal{A}$.
+A term/tree with variables is called a tree pattern.
+A tree containing a designated occurrence of a subtree $t$ is denoted $C[...t...]$.
+A tree language over $\Sigma$ is a subset of $T_\Sigma$. Tree languages are described by tree grammars, which can be defined in analogy to the Chomsky hierarchy of string grammars.
+
+\textbf{Definition 1:} (Tree grammar $\mathcal{G}$ over $\Sigma$ and $\mathcal{A}$) \\
+A (regular) tree grammar $\mathcal{G}$ over $\Sigma$ and $\mathcal{A}$ is given by\ul
+\item A set $V$ of nonterminal symbols
+\item A designated nonterminal symbol $Ax$ called the axiom
+\item A set $P$ of productions of the form $v\to t$ where $v\in V$ and $t\in T_\Sigma(V)$
+\ule
+The derivation relation for tree grammars is $\to^*$, with $C[...v...] \to C[...t...]$ if $v\to t\in P$.
+The language of $v\in V$ is $\mathcal{L}(v)=\{t\in T_\Sigma | v\to^* t\}$.
+The language of $\mathcal{G}$ is $\mathcal{L}(\mathcal{G})=\mathcal{L}(Ax)$.
+
+\textbf{Definition 2:} (Evaluation algebra)\\ Let $\Sigma$ be a signature over $\mathcal{A}$ with sort symbol $Ans$.
+A $\Sigma$-evaluation algebra is a $\Sigma$-algebra augmented with an objective function $h: [Ans] \to [Ans]$, where $[Ans]$ denotes lists over $Ans$.
+
+\textbf{Definition 3:} (Yield grammars and yield languages)\\
+Let $\mathcal{G}$ be a tree grammar over $\Sigma$ and $\mathcal{A}$, and $y$ the yield function.
+The pair $(\mathcal{G}, y)$ is called a yield grammar.
+It defines the yield language $\mathcal{L}(\mathcal{G}, y) = y(\mathcal{L}(\mathcal{G}))$.
+
+\textbf{Definition 4:} (Yield parsing)\\
+Given a yield grammar $(\mathcal{G}, y)$ over $\mathcal{A}$ and $w\in\mathcal{A}^*$, the yield parsing problem is to find $P_\mathcal{G}(w) := \{t\in \mathcal{L}(\mathcal{G})|y(t) = w\}$..
+
+\textbf{Definition 5:} (Algebraic dynamic programming)\ul
+\item An ADP problem is specified by a signature $\Sigma$ over $\mathcal{A}$, a yield grammar $(\mathcal{G}, y)$ over $\Sigma$, and a $\Sigma$-evaluation algebra $I$ with objective function $h_I$ .
+\item An ADP problem instance is posed by a string $w\in\mathcal{A}^*$. The search space it spawns is the set of all its parses, $P_\mathcal{G}(w)$.
+\item Solving an ADP problem is computing $h_I\{t_I | t\in P_\mathcal{G}(w)\}$
+in polynomial time and space.
+\ule
+
+\textbf{Definition 6:} (Algebraic version of Bellman’s principle)\\
+For each $k$-ary operator $f$ in $\Sigma$, and all answer lists $z_1,\ldots, z_k$, the objective function $h$ satisfies
+\[\begin{array}{rl}
+& h( [ f(x_1,\ldots,x_k) | x_1\leftarrow z_1,\ldots ,x_k\leftarrow z_k ] ) \\[4pt]
+= & h( [ f(x_1,\ldots,x_k) | x_1\leftarrow h(z_1),\ldots,x_k\leftarrow h(z_k) ] )
+\end{array}\]
+Furthermore, the same property holds for the concatenation of answer lists: \[h(z_1 ::: z_2) = h(h(z_1) ::: h(z_2))\]
+
+% -----------------------------
+\subsubsection{ADP in practice}
 ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
@@
 Finally, the default aggregation function $h$ must be defined\footnote{Although aggregation usage is not mandatory in the framework, we force the existence of an aggregation function over the ouput type so that we can use it to aggregate windowed results.}.
-To make it more clear, we propose an example of the matrix chain multiplication problem\footnote{The original ADP framework was embedded in Haskell, however, we assume that the reader is more familiar with Scala notation and immediately present the syntax of our implementation.}.
+To make it more clear, we propose an example of the matrix chain multiplication problem\footnote{The original ADP framework is an embedded DSL of Haskell, however, we assume that the reader is more familiar with Scala notation and immediately present the syntax of our implementation.}.
 
-\newpage
 \begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
@@
 
-Notice that the semantics of indices differ slightly from the problem presented in \ref{mat_mult_plain}; this is because empty chain are made expressible (denoted $M_{(i,i)}$, single matrices are denoted $M_{(i,i+1)}$).
+Notice that the semantics of indices differ slightly from the problem presented in \S\ref{mat_mult_plain}; this is because empty chain are made expressible (denoted $M_{(i,i)}$, single matrices are denoted $M_{(i,i+1)}$).
 

### 2013-01-17 docs/report/design.tex
Nits.

@@
 Also we need to maintain coherence between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (pretty-printing or actual computation as described in \S\ref{intro}).
-\item \textbf{Optimizations:} Since ADP exposes a grammar, we have the possibility to do grammar optimizations (see \S\ref{grammar_opt}).
+\item \textbf{Optimizations:} Since ADP exposes a grammar, we might have the opportunity to do optimizations at the grammar level (see \S\ref{grammar_opt}).
 Also since the grammar might define useless rules, we might want to eliminate them: dead rule elimination is very similar to traditional dead code elimination (it reduces the size of generated code) but also reduces the memory consumption (as storage matrix does not need to be reserved) and even speeds up the computation, (since all grammar rules for a particular element are computed at the same time in CUDA implementation, see \ref{dead_rules}).
@@
 
-Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
+Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first and third equations denote breaking in two distinct functions, the second and fourth represent Bellman's optimality principle preservation.}
 \[\begin{array}{rcll}
@@
 \end{array}\]
+
 % Viktor Kuncak: We can solve this more easily if functions are linear
 Since this requires complex mathematical analysis that is out of the scope of the project, and since we have not found relevant literature on that particular subject, we leave this optimization to the responsibility of the user by assuming that the provided grammar is already optimal.
+
+Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebrae).
 
@@
 We tried to use LMS to generate both the Scala and CUDA program.
-This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}), and this approach also limits the functionality of Scala parsers (as only a subset of the Scala language can be lifted into LMS).
+This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}).
+This issue is particularly annoying in our particular case because concatenation construct pair of result (hence multiple concatenation would produce nested pairs) whereas we want to present its result as a flattened list of arguments suitable for immediate use by a function.
+
+This approach also complicates the usage of Scala parsers because the algebra can only contain a subset of the Scala language can be represented in LMS.
+Taking the example of matrix multiplication, the user would like not only to solve the dynamic problem on GPU, but also to leverage the graphic card to compute matrices products.
+Since associated resources description cannot be represented by LMS nodes, the user would need to extend LMS with ad-hoc DSL for every particular function that is not representable in LMS.
 
@@
 From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}.
-Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress in our project rather than helping improving and adding features to an existing project (but I would reconsider my position given different deadlines).
-Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros (that operate on the Scala abstract syntax tree), which represent a possible alternative to LMS for generating  simple functions for a costing algebra.
+Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress rather than improving an existing project (but I would reconsider my position given different deadlines).
+Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating simple functions for a costing algebra.
 

### 2013-01-17 docs/report/introduction.tex
Nits.

@@
 \subsection{Scala} \label{bg_scala}
-%  - However, we would list some ... => ; this is of course not the only reason; Scala offers some interesting features which make it a nice language to develop a DSL in ?
-%  - large adoption in the industry is not really a feature is it?
-%  - Implicit functions definition => implicit functions and parameters offer a way to neatly integrate primitive/pre-existing constructs into a DSL, by converting their types in order to correspond to the DSL semantics, while at the same time guarantee-ing type safety.
-%  - Macros[6] and LMS are mechanisms to modify semantics of a program; in particular, these tools can be used to implement domain-specific optimizations of programs.
-LMS also contains a code generation platform which we use to generate some functions in C (link to section).
-%  - Mixin composition (traits) allow for multiple inheritance and composition, which makes it easy to represent, in our DSL, multiple algebrae, and also interface with different runtimes (link to section).
 \textit{<<Scala is a general purpose programming language designed to express common programming patterns in a concise, elegant, and type-safe way.
@@
 
-As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural to use it as host language for our project, however, we would list some of its features \cite{scala_api} that makes it an interesting development language for this project:\ul
+As the Scala \cite{scala} programming language is developed by our laboratory (LAMP, EPFL), it seems natural host language for our project.
+Its large adoption\footnote{\url{http://www.scala-lang.org/node/1658}}, would make the adoption of our DSL easier while reducing the learning time of its potential users.
+Additionally, some features \cite{scala_api} of Scala makes it an interesting development language for this project:\ul
 \item The functional programming style and syntactic sugar offered by Scala allow concise writing of implementation, analysis and transformations of our DSL, allowing us to focus on \textit{what} we want to achieve instead of \textit{how}.
-\item Scala is largely adopted in the industry\footnote{\url{http://www.scala-lang.org/node/1658}}, which makes both the adoption of related project easier and reduces the learning time of its potential users.
 \item Since Scala programs execute in the Java Virtual Machine (JVM), they can benefit of the native interface (JNI) that offers the possibility to dynamically load libraries (usually written in C) and possibly interact with CUDA to leverage the GPU.
 \item Scala is equipped with a strong typing and type inference system that reduces the syntactical constraints while putting strong guarantees on type correctness at compilation.
-\item Implicit functions definition can help interfacing functions by automatically changing the type of passed arguments
+\item Implicit functions and parameters allow to simplify the syntactic usage of the DSL by implementing automatic conversions, while at the same time preserving type safety.
 \item Manifests (or TypeTags and ClassTags) allow type extraction at runtime (we use this to convert a Scala type into a C/CUDA type)
-\item Macros\cite{scala_macros} and LMS (\S\ref{bg_lms}) could be used to modify the semantics of a specific part of the user program (we currently use LMS to produce the corresponding C function).
-\item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and mixed together, thereby allowing multiple inheritance.
+\item Macros\cite{scala_macros} and LMS (\S\ref{bg_lms}) could be used to modify the semantics of specific parts, or implement domain-specific optimizations of the user program.
+LMS also contains a multi-language code generator that we leverage to produce C functions (see \S\ref{user_fun_gen}).
+\item One Scala concept that we heavily use is \textit{traits} that can be viewed as abstract classes and combined (mixin composition), thereby allowing multiple inheritance.
+We use this feature in particular to smoothly combine algebra, grammar and possibly code generation (\S\ref{codegen}) into a concrete program.
 \ule
@@
 
-A discussion of the integration of LMS in our project can be found in \S\ref{lms_use}.
+A discussion on the integration of LMS in our project can be found in \S\ref{lms_use}.
 

### 2013-01-17 docs/report/problems.tex
Nits.

@@
 There exist various categories of dynamic programming:\ul
-\item Series that operate usually on a single dimension (like Fibonacci\footnote{http://en.wikipedia.org/wiki/Fibonacci_number})
+\item Series that operate usually sequentially on a single dimension (like Fibonacci\footnote{\url{http://en.wikipedia.org/wiki/Fibonacci\_number}})
 \item Sequences alignment (matching two sequences at best), top-down grammar analysis (parenthesizing), sequence folding, ... (see \S\ref{problems} for more examples and detailed classification)
@@
 By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
-\item \textbf{Alphabets:} an alphabet is an enumeration of the possible values.
-Its size helps determining how many bits are required in the implementation to store all values.
+\item \textbf{Alphabets:} an alphabet is an set of possible values.
+Its size helps determining how many bits are required in the implementation to represent all its elements.
 Alphabets are defined for input, cost, backtrack and wavefront.
@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{Problems of interest} \label{problem_examples}
@@
 \subsubsection{Nussinov algorithm} \label{nussinov} \ol
-\item Problem: folding a RNA string $S$ over itself $|S|=n$.
+\item Problem: folding a RNA string $S$ over itself $|S|=n$, according to matching properties ($\omega$) of its elements (also called bases).
 \item Matrices: $M_{n\times n}, B_{n \times n}$
@@
 \subsubsection{Zuker RNA folding} \label{zuker} \ol
-\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy.
+\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hereby much more complicated than Nussinov \S\ref{nussinov}).
 \item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$
@@
 
-\item Backtracking: Start at $BF_{(n)}$ using the recurrences
+\item Backtracking: starts at $BF_{(n)}$ and uses the recurrences
  \[\begin{array}{rcl}
@@
 
-The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
+The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.
+
+\newpage
 Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.

### 2013-01-17 docs/report/benchmarks.tex
Nits. Bug in the backtrack of two track grammars.

--- 2013-01-16 docs/report/report.tex
@@
-\documentclass[11pt]{article}
-\input{inc/style.sty}
-\usepackage{multicol,nameref}
-\usepackage{etoolbox}
+\newpage
+\section{Usage} \label{usage}
+\subsection{Program examples}
+In this section, we explain how to use DynaProg, as a parser library.
+Use an example based approach, we focus on three additional examples: Smith-Waterman/Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} to describe RNA library usage and reconstruction from backtrack trace, and finally we extend Matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.
 
-\title{DynaProg for Scala}
-\subtitle{A Scala DSL for Dynamic Programming on CPU and GPU}
-\begin{document}
-\maketitle
-\input{introduction}
-\input{problems}
-\input{design}
-\input{implementation}
+\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
+We first define a signature that can fit both algebrae, then we specify for each algebra the functions.
+In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).
 
-\section{Usage} \label{usage}
+We also want to emphasize how bad this example is, with respect to the result construction, as every intermediate result would contain its whole backtrack, hereby duplicating the information multiple times.
+To overcome this issue {\color{red} XXX}
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait SeqAlignSignature extends Signature {
+  type Alphabet = Char
+  type Answer = (Int,String,String) // score, seq1, seq2
+  override val h = maxBy((x:Answer)=>x._1)
+
+  def gap(current:Int, g:(Int,Int)):Int
+  def pair(a:Alphabet,b:Alphabet):Int
+}
+
+trait SmithWatermanAlgebra extends SeqAlignSignature {
+  private val open = -3
+  private val extend = -1
+  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, score + ( open + (size-1) * extend )) }
+  def pair(a:Char,b:Char) = if (a==b) 10 else -3
+}
+
+trait NeedlemanWunschAlgebra extends SeqAlignSignature {
+  private val open = -15
+  private val extend = -1
+  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; score + ( open + (size-1) * extend ) }
+  def pair(a:Char,b:Char) = if (a==b) 4 else -3
+}
+\end{lstlisting}
+
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
+  private def prettyGap(sw:Subword,in:Function1[Int,Char]):(String,String) = {
+    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
+  }
+
+  val alignment:Tabulate = tabulate("M",(
+    empty                         ^^ { _ => (0,".",".") }
+  | seq1() -~ alignment           ^^ { case (g,(score,s1,s2)) => val (g1,g2)=prettyGap(g,in1); (gap(score,g),s1+g1,s2+g2) }
+  |           alignment ~- seq2() ^^ { case ((score,s1,s2),g) => val (g1,g2)=prettyGap(g,in2); (gap(score,g),s1+g2,s2+g1) }
+  | el1    -~ alignment ~- el2    ^^ { case (c1,((score,s1,s2),c2)) => (score+pair(c1,c2),s1+c1, s2+c2) }
+  ) aggregate h)
+
+  val axiom = alignment
+
+  def align(s1:String,s2:String) = parse(s1.toArray,s2.toArray).head
+}
+
+object SeqAlign extends App {
+  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra   // Smith-Waterman
+  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra // Needleman-Wunsch
+  val seq1 = "CGATTACA"
+  val seq2 = "CCCATTAGAG"
+
+  val (swScore,sw1,sw2) = SWat.align(seq1,seq2)
+  println("Smith-Waterman alignment\n- Score: "+swScore+"\n- Seq1: "+sw1+"\n- Seq2: "+sw2+"\n")
+  val (nwScore,nw1,nw2) = NWun.align(seq1,seq2)
+  println("Needleman-Wunsch alignment\n- Score: "+nwScore+"\n- Seq1: "+nw1+"\n- Seq2: "+nw2+"\n")
+}
+\end{lstlisting}
+
+\newpage
+\subsubsection{RNA folding} \label{ex_rnafold}
+We define a signature with two evaluation algebras: {\tt RNAFoldAlgebra} actually computes the folding whereas {\tt RNAFoldPrettyPrint} describes the folding in a string.
+The energy functions are provided by an external library ({\tt LibRNA}).
+This library encodes substring as (first character, last character) whereas our framework encodes them as (first character, first character + length), which explains the off-by-one corrections.
+{\tt energies} variable is set to false in {\tt RNAFoldPrettyPrint} because this algebra does not involve the LibRNA energies functions (that require encoding the input RNA sequence in a special format; this option is enabled by default in the RNASignature trait).
+
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait RNAFoldSig extends RNASignature {
+  def hairpin(ij:(Int,Int)):Answer
+  def stack(i:Int,s:Answer,j:Int):Answer
+  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
+  def mloop(i:Int,s:Answer,j:Int):Answer
+  def left(l:Answer,r:Int):Answer
+  def right(l:Int,r:Answer):Answer
+  def join(l:Answer,r:Answer):Answer
+}
+
+trait RNAFoldAlgebra extends RNAFoldSig {
+  type Answer = Int
+  import librna.LibRNA._ // indexing convention: first base,last base
+  def hairpin(ij:(Int,Int)) = hl_energy(ij._1,ij._2-1) // Eh
+  def stack(i:Int,s:Int,j:Int) = sr_energy(i,j) + s // Es
+  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) =
+      il_energy(ik._1,ik._2,lj._1-1,lj._2-1) + s // Ei
+  def mloop(i:Int,s:Int,j:Int) = s
+  def left(l:Int,r:Int) = l
+  def right(l:Int,r:Int) = r
+  def join(l:Int,r:Int) = l+r
+  override val h = min[Answer] _
+}
+
+trait RNAFoldPrettyPrint extends RNAFoldSig {
+  type Answer = String
+  override val energies=false
+  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
+  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
+  def stack(i:Int,s:String,j:Int) = "("+s+")"
+  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) =
+      "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
+  def mloop(i:Int,s:String,j:Int) = "("+s+")"
+  def left(l:String,r:Int) = l+"."
+  def right(l:Int,r:String) = "."+r
+  def join(l:String,r:String) = l+r
+}
+\end{lstlisting}
+
+\newpage
+We can then define the associated grammar
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
+  lazy val Qp:Tabulate = tabulate("Qp",(
+    seq(3,maxN)        ^^ hairpin
+  | eli   ~ Qp ~ eli   ^^ stack
+  | seq() ~ Qp ~ seq() ^^ iloop
+  | eli   ~ QM ~ eli   ^^ mloop
+  ) filter basepairing aggregate h)
+
+  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
+
+  lazy val Q:Tabulate = tabulate("Q",(
+    QM
+  | Q ~ eli ^^ left
+  | eli ~ Q ^^ right
+  | Qp
+  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
+
+  override val axiom = Q
+}
+\end{lstlisting}
+
+In the application, we create two objects, each combining the grammar with a particular algebra.
+We can optionally specify a coefficient parameter file with {\tt setParams(file:String)}, otherwise the \textit{Turner2004} coefficients are used.
+The library is automatically loaded and fed with the sequence to produce correct energy coefficients.
+We request both the score and the backtrack trace (in {\tt bt}) so that we can reconstruct the folding using the pretty printing grammar.
+\begin{lstlisting}[language=Scala,captionpos=none]
+object RNAFold extends App {
+  object fold extends RNAFoldGrammar with RNAFoldAlgebra
+  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
+
+  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuuuucaaaggaagagg"
+
+  val (score,bt) = fold.backtrack(seq.toArray).head
+  val res = pretty.build(seq.toArray,bt)
+  println("Folding : "+res+" (%5.2f)".format(score/100.0));
+}
+\end{lstlisting}
+
+\newpage
+\subsubsection{Matrix multiplication with CUDA code generation} \label{ex_matmult_cuda_lms}
+Leveraging the existing definitions of the signature and grammar
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait MatrixSig extends Signature {
+  type Alphabet = (Int,Int) // Matrix(rows, columns)
+  val single:Alphabet=>Answer
+  val mult:(Answer,Answer)=>Answer
+}
+
+trait MatrixGrammar extends ADPParsers with MatrixSig {
+  val axiom:Tabulate = tabulate("M",
+    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
+}
+\end{lstlisting}
+We need describe the algebra functions in the LMS syntax ({\tt RepWorld}) that we can later compile to use as regular functions, augmented with C code description (necessary for code generation).
+Finally, we need to mix the {\tt CodeGen} trait to enable code generation and provide the manifest for input and ouput types ({\tt Alphabet} and {\tt Answer}).
+
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait RepWorld extends NumericOps with TupleOps {
+  type Alphabet = (Int, Int)
+  type Answer = (Int, Int, Int)
+
+  def hf(a: Rep[Answer]) :Rep[Int] = a._2
+  def repSingle(a: Rep[Alphabet]): Rep[Answer] = (a._1, unit(0), a._2)
+  def repMult(l: Rep[Answer], r: Rep[Answer]): Rep[Answer] =
+    (l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3)
+}
+
+object MatrixMultLMS extends MatrixSig with MatrixGrammar
+    with CodeGen with App {
+  override val tps=(manifest[Alphabet],manifest[Answer])
+  override val benchmark = true // display timing measurements
+
+  // Algebra is defined immediately in the concrete program
+  type Answer = (Int, Int, Int)
+  val concreteProg = new RepWorld with RepPackage
+  override val h = minBy(concreteProg.gen(concreteProg.hf))
+  val single = concreteProg.gen(concreteProg.repSingle)
+  val mult = concreteProg.gen2(concreteProg.repMult)
+
+  val input = List((1,2),(2,20),(20,2),(2,4),(4,2),(2,1),(1,7),(7,3)).toArray
+  println(parse(input).head) // -> 1x3 matrix, 122 multiplications
+}
+\end{lstlisting}
+
+\newpage
+----------------------------------------
+
+ and focus on three different problem: our matrix multiplication example,
 
@@
 
-\subsection{Program examples}
 \subsubsection{Matrix multiplication} \label{ex_matmult}
@@
   println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
-}
-\end{lstlisting}
-
-\subsubsection{RNA folding} \label{ex_rnafold}
-\begin{lstlisting}[language=Scala,caption={RNA folding}]
-import v4._
-
-trait RNAFoldSig extends RNASignature {
-  def hairpin(ij:(Int,Int)):Answer
-  def stack(i:Int,s:Answer,j:Int):Answer
-  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
-  def mloop(i:Int,s:Answer,j:Int):Answer
-  def left(l:Answer,r:Int):Answer
-  def right(l:Int,r:Answer):Answer
-  def join(l:Answer,r:Answer):Answer
-}
-
-trait RNAFoldAlgebra extends RNAFoldSig {
-  type Answer = Int
-  import librna.LibRNA._ // indexing convention: first base,last base
-  private def Eh(i:Int,j:Int) = hl_energy(i,j) // hairpin loop
-  private def Ei(i:Int,k:Int,l:Int,j:Int) = il_energy(i,k,l,j) // internal loop
-  private def Es(i:Int,j:Int) = sr_energy(i,j) // stacked base pairs
-  def hairpin(ij:(Int,Int)) = Eh(ij._1,ij._2-1)
-  def stack(i:Int,s:Int,j:Int) = Es(i,j) + s
-  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) = Ei(ik._1,ik._2,lj._1-1,lj._2-1) + s
-  def mloop(i:Int,s:Int,j:Int) = s
-  def left(l:Int,r:Int) = l
-  def right(l:Int,r:Int) = r
-  def join(l:Int,r:Int) = l+r
-  override val h = min[Answer] _
-}
-
-trait RNAFoldPrettyPrint extends RNAFoldSig {
-  type Answer = String
-  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
-  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
-  def stack(i:Int,s:String,j:Int) = "("+s+")"
-  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
-  def mloop(i:Int,s:String,j:Int) = "("+s+")"
-  def left(l:String,r:Int) = l+"."
-  def right(l:Int,r:String) = "."+r
-  def join(l:String,r:String) = l+r
-}
-
-trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
-  lazy val Qp:Tabulate = tabulate("Qp",(
-    seq(3,maxN)        ^^ hairpin
-  | eli   ~ Qp ~ eli   ^^ stack
-  | seq() ~ Qp ~ seq() ^^ iloop
-  | eli   ~ QM ~ eli   ^^ mloop
-  ) filter basepairing aggregate h)
-
-  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
-
-  lazy val Q:Tabulate = tabulate("Q",(
-    QM
-  | Q ~ eli ^^ left
-  | eli ~ Q ^^ right
-  | Qp
-  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
-
-  override val axiom = Q
-}
-
-object RNAFold extends App {
-  object fold extends RNAFoldGrammar with RNAFoldAlgebra
-  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
-
-  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuu"
-
-  import librna.LibRNA._
-  setSequence(seq)
-  val (score,bt) = fold.backtrack(seq.toArray).head
-  val res = pretty.build(seq.toArray,bt)
-  clear
-  println("Folding : "+res+" (%5.2f)".format(score/100.0));
 }
@@
 \ole
-
-\section{Conclusion}
-{\color{red} XXX}
-
-\begin{verbatim}
-
-recap of the report
-
-recap of the results
-
-big picture: with actual compiler technology we 
-Embedded DSL vs external DSL
-Larger picture how to go further
-
-\end{verbatim}
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\section*{Planning}
-\subsubsection*{Todo @TCK}\ul
-\item Test/proof parsers are correct -- make sure implementation is correct
-\item Automate test to compare against implementation
-\item Benchmarks -- use CUDA profiler(?)
-\item Write report
-\item Port LibRNA for CUDA?
-\ule
-
-\subsubsection*{Todo @Manohar}\ul
-\item Integrate LMS code generation into v4.
-\item Fix Zuker coefficients
-\ule
-
-Legacy roadmap deadlines: \\
-\begin{tabular}{ll}
-Nov 16 & Rules normalization and automatic backtracking \\
-	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
-Nov 23 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
-Nov 30 &--- gap due to LMS missing knowledge --- \\
-Dec 7 & Benchmarking, grammar analysis \\
-Dec 14 & First thoughts for larger than device memory \\
-Dec 21 & Writing report \\
-Jan 4 & --- holiday --- \\
-Jan 18 & Writing report: implementation description and plan for future work
-\end{tabular}
-
-\subsubsection*{Journal}
-\begin{tabular}{ll}
-Sep. 17	& Getting started with the project, reading related work on hash maps. \\
-Sep. 24	& Parallel hashing paper solve the problem for 32bit key/value pair. Stripped CudPP. \\
-		& Devised (but not implemented) extension beyond 64bit using memory areas locks. \\
----		& Change of project suggested by Vojin (supervisor): joint work with Manohar on DP \\
-Oct. 01	& Problems specifications: serial/non-serial, started CUDA implementation with blocks. \\
-Oct. 08	& CudAlign solves serial monadic, might adapt it (but complicated / ad-hoc / in progress). \\
-		& Focus on non-serial problems that fit in GPU memory. \\
-Oct. 15	& First working implementation for non-serial problems (rectangle, triangle, parallelogram) \\
-Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
-Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. ADP (Manohar) \\
-Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
-Nov. 12	& Rework ADP parsers: cyclic, two-track grammars and automatic backtracking discussion. \\
-Nov. 19	& Explorations LMS and Macros for C code, implementation is quite different, hence ad-hoc. \\
-Nov. 26	& Full backtracking stack: apply / unapply / reapply, refactoring of the classes. \\
-Dec. 03	& JNI for LibRNA to get coefficients for Zuker, errors in algebra (quite hairy code). \\
-Dec. 10	& Rework concatenation operators, yield analysis, code generation. \\
-Dec. 17	& Detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support. \\
-Dec. 24	& (sick 4 days), code generation: full JNI conversion support, rework CodeCompiler. \\
-Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
-Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
-Jan. 14	& {\color{red} XXX} \\
-\end{tabular}
-
-%Some GPU algorithms: http://hgpu.org/?cat=11
-%Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
-%CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
-%13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
-%http://tutorials.jenkov.com/java-reflection/fields.html
-%http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
-%
-%Hint: use TypeClass to put a predicate on types
-%  def fun[T: CanTranslateToC](...)
-%  def fun[T](implicit ev:CanTranslateToC[T])
-%  class CanTranslateToC[T] { def translate:String }
-%  implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\patchcmd{\thebibliography}{\section*{\refname}}{\begin{multicols}{2}[\section*{\refname}]}{}{}
-\patchcmd{\endthebibliography}{\endlist}{\endlist\end{multicols}}{}{}
-\bibliographystyle{plain} % acm
-{\small \bibliography{bibliography.bib}}
-\end{document}

### 2013-01-17 docs/report/introduction.tex
Nits. Bug in the backtrack of two track grammars.

@@
 % -----------------------------
-\subsubsection{ADP in practice}
+\subsubsection{ADP in practice} \label{adp_practice}
 ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase).
@@
 
-\begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
+\begin{lstlisting}[language=Scala,caption={Matrix chain mulitiplication DSL implementation}]
 trait MatrixSig extends Signature {

### 2013-01-17 docs/report/report.tex
Report reordering.
Nits. Bug in the backtrack of two track grammars.

--- 2013-01-15 docs/report/report.tex
@@
 \input{implementation}
-
-\section{Usage} \label{usage}
-{\color{red} usage, make reference from problems to usage
-
-XXX: do not introduce pseudocode but refere to usage in RNAfold problem
-
-}
-
-\subsection{Program examples}
-\subsubsection{Matrix multiplication} \label{ex_matmult}
-\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
-import v4._
-
-trait MatrixSig extends Signature {
-  type Alphabet = (Int,Int) // Matrix(rows, columns)
-  val single:Alphabet=>Answer
-  val mult:(Answer,Answer)=>Answer
-}
-
-trait MatrixAlgebra extends MatrixSig {
-  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
-
-  override val h = minBy((a:Answer) => a._1)
-  val single = (a: Alphabet) => (0, a)
-  val mult = (a:Answer,b:Answer) =>
-    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
-}
-
-trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val axiom:Tabulate = tabulate("M",
-    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
-}
-
-object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
-  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
-}
-\end{lstlisting}
-
-\subsubsection{RNA folding} \label{ex_rnafold}
-\begin{lstlisting}[language=Scala,caption={RNA folding}]
-import v4._
-
-trait RNAFoldSig extends RNASignature {
-  def hairpin(ij:(Int,Int)):Answer
-  def stack(i:Int,s:Answer,j:Int):Answer
-  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
-  def mloop(i:Int,s:Answer,j:Int):Answer
-  def left(l:Answer,r:Int):Answer
-  def right(l:Int,r:Answer):Answer
-  def join(l:Answer,r:Answer):Answer
-}
-
-trait RNAFoldAlgebra extends RNAFoldSig {
-  type Answer = Int
-  import librna.LibRNA._ // indexing convention: first base,last base
-  private def Eh(i:Int,j:Int) = hl_energy(i,j) // hairpin loop
-  private def Ei(i:Int,k:Int,l:Int,j:Int) = il_energy(i,k,l,j) // internal loop
-  private def Es(i:Int,j:Int) = sr_energy(i,j) // stacked base pairs
-  def hairpin(ij:(Int,Int)) = Eh(ij._1,ij._2-1)
-  def stack(i:Int,s:Int,j:Int) = Es(i,j) + s
-  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) = Ei(ik._1,ik._2,lj._1-1,lj._2-1) + s
-  def mloop(i:Int,s:Int,j:Int) = s
-  def left(l:Int,r:Int) = l
-  def right(l:Int,r:Int) = r
-  def join(l:Int,r:Int) = l+r
-  override val h = min[Answer] _
-}
-
-trait RNAFoldPrettyPrint extends RNAFoldSig {
-  type Answer = String
-  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
-  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
-  def stack(i:Int,s:String,j:Int) = "("+s+")"
-  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
-  def mloop(i:Int,s:String,j:Int) = "("+s+")"
-  def left(l:String,r:Int) = l+"."
-  def right(l:Int,r:String) = "."+r
-  def join(l:String,r:String) = l+r
-}
-
-trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
-  lazy val Qp:Tabulate = tabulate("Qp",(
-    seq(3,maxN)        ^^ hairpin
-  | eli   ~ Qp ~ eli   ^^ stack
-  | seq() ~ Qp ~ seq() ^^ iloop
-  | eli   ~ QM ~ eli   ^^ mloop
-  ) filter basepairing aggregate h)
-
-  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
-
-  lazy val Q:Tabulate = tabulate("Q",(
-    QM
-  | Q ~ eli ^^ left
-  | eli ~ Q ^^ right
-  | Qp
-  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
-
-  override val axiom = Q
-}
-
-object RNAFold extends App {
-  object fold extends RNAFoldGrammar with RNAFoldAlgebra
-  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
-
-  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuu"
-
-  import librna.LibRNA._
-  setSequence(seq)
-  val (score,bt) = fold.backtrack(seq.toArray).head
-  val res = pretty.build(seq.toArray,bt)
-  clear
-  println("Folding : "+res+" (%5.2f)".format(score/100.0));
-}
-\end{lstlisting}
-
-\section{Benchmarks} \label{benchmarks}
-In an attempt to provide realistic benchmarks, we tried to gather related implementations.
-The authors of \cite{gpu_atlp} did not respond to our multiple solicitations.
-The authors of \cite{swat_mega} were very friendly and provided us their source code.
-Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
-Finally, we asked lately the authors of \cite{gpu_rnafold} {\color{red} who did not respond so far}.
-
-We organize the benchmarks as follow: if we have at our disposal a working implementation that could be run on our evaluation platform, we use it, otherwise, we refer to the related paper and rescale the result according to the memory throughput and computation throughput of the related device so that we can have a good approximation of how they could compare.
-
-\subsection{Metrics}
-{\color{red} The main metric we focus on is the running time.}
-
-As the device memory is quite limited, it seems interesting to also take into account the space usage.
-The space requirement limits the maximal size of addressable problems  on a particular hardware.
-This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently.
-However, simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue.
-
-\subsection{Benchmarking platform}
-
-RUN 5-10x until running time stabilizes on Scala
-
-% ------------------------------------------------------------------------------------------------
-{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
-% ------------------------------------------------------------------------------------------------
-
-\subsubsection{Test 1 : Ad-hoc VS CUDA parsers (implies ATLP rescaled comparison)}
-
-\subsubsection{Test 2 : CUDAlign vs CUDA parsers for small sequences}
-Also CUDASW++
-
-\subsubsection{Test 3 : GAPC vs Java parsers vs CUDA parsers}
-
-\subsubsection{Test 4 ? : GPU RNAfolding vs CUDA parsers}
-
-\subsection{Java parsers}
-{\color{red}
-MatrixMult-512, Mac+JDK7\ul
-\item Original: 24.937 sec / 24.658 sec
-\item Optimized concatenation: 19.329 sec / 18.924 sec
-\item Tabulation as arrays+inline: 14.976 sec / 14.849 sec
-\ule}
-
-\subsection{CUDA parsers}
-{\color{red} XXX: explain why slower: approx 3x more memory used per score (maintain matrix dimensions)
-
-XXX: continue benchmarking here
-}
-\begin{table}[H]\begin{center}\begin{tabular}{lrrrrrr} \toprule
-Matrix size			& 1024	& 2048	& 4096	& 8192 \\
-Splits				& 1		& 8		& 64		& 512 \\ \midrule
-Analysis				& 0.007	& 0.007	& 0.007	& 0.007 \\
-Code generation		& 0.071	& 0.072	& 0.073	& 0.071 \\
-C/CUDA compilation	& 3.025	& 1.836	& 1.832	& 1.841 \\
-Scala compilation		& 1.837	& 1.853	& 1.830	& 1.767 \\ \midrule
-- JNI read				& 0.027	& 0.028	& 0.029	& 0.058 \\
-- CUDA compute		& 0.868	& 3.378	& 19.782	& 139.697 \\
-- CUDA backtrack		& 0.009	& 0.012	& 0.023	& 0.045 \\
-- JNI output			& 0.005	& 0.009	& 0.018	& 0.032 \\
-Total execution			& 0.915	& 3.432	& 19.860	& 139.846 \\ \bottomrule
-\end{tabular}\end{center}\caption{Preliminary results for MatrixMultGen}\end{table}
-
-{\color{red} XXX: Compare current implementation versus ad-hoc implementation.
-Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.}
-
-\section{Future work}
-We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
-\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
-However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
-\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program.
-Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}).
-These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar).
-\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation.
-Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
-\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
-The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
-\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.
-Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size).
-In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.
-
-Matrix of different dimensions must be stored in their own array (versus being in a single array of struct enclosing corresponding element of all matrices).
-Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description, there are two $O(n^2)$ matrices and one $O(n)$ matrix.
-This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once.
-\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently.
-For example, RNA is represented with only 4 letters (g,a,t,c), thus 4 symbols could be encoded in a single byte.
-Unfortunately, this optimizations seems to only apply for the input data.
-Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
-\item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded.
-Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, thereby making it possible to prune entire computation tiles.
-\item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
-Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser function calls into a single function making equivalent calls.
-The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
-\item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
-We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
-\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
-Hence cost and backtrack matrix would contain $k$ elements per cell.
-Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
-\ole
+\input{benchmarks}
 
@@
 {\color{red} XXX}
+
+\begin{verbatim}
+
+recap of the report
+
+recap of the results
+
+big picture: with actual compiler technology we
+Embedded DSL vs external DSL
+Larger picture how to go further
+
+\end{verbatim}
 
@@
 Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
-Jan. 14	& {\color{red} XXX} \\
+Jan. 14	& Report improvements, fixed Zuker (coherent with GAPC), {\color{red} XXX} \\
 \end{tabular}

### 2013-01-17 src/main/scala/v4/report/SeqAlign.scala

@@ -1,0 +1,79 @@
+package v4.report
+import v4._
+
+trait SeqAlignSignature extends Signature {
+  type Alphabet = Char
+
+  def start(x:Unit):Answer
+  def gap1(g:(Int,Int),a:Answer):Answer
+  def gap2(a:Answer,g:(Int,Int)):Answer
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
+}
+
+trait SmithWatermanAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
+  private val open = -3
+  private val extend = -1
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
+  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, a + ( open + (size-1) * extend )) }
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 10 else -3)
+}
+
+trait NeedlemanWunschAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
+  private val open = -15
+  private val extend = -1
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
+  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; a + ( open + (size-1) * extend ) }
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 4 else -3)
+}
+
+trait SeqPrettyPrint extends SeqAlignSignature {
+  type Answer = (String,String)
+  private def gap(sw:(Int,Int)) = (sw._1 to sw._2).map{_=>"-"}.mkString
+  def start(x:Unit) = (".",".")
+  def gap1(g:(Int,Int),a:Answer) = (a._1+gap(g),a._2)
+  def gap2(a:Answer,g:(Int,Int)) = (a._1,a._2+gap(g))
+  def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
+}
+
+trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
+  val alignment:Tabulate = tabulate("M",(
+    empty                         ^^ start
+  | seq1() -~ alignment           ^^ gap1
[…]

### 2013-01-17 docs/report/benchmarks.tex
More report.

@@
 \subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
-We first define a signature that can fit both algebrae, then we specify for each algebra the functions.
+First define a signature that can fit both algebrae, then specify for each algebra the functions.
 In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).
-
-We also want to emphasize how bad this example is, with respect to the result construction, as every intermediate result would contain its whole backtrack, hereby duplicating the information multiple times.
-To overcome this issue {\color{red} XXX}
+A naive idea would be to construct the alignment string immediately in the forward phase (that is immediately in the {\tt Answer}).
+However, is not efficient at all, both in terms of running time and space complexity perspective because substring are duplicated for every cell.
+The correct way to solve this issue is to use backtracking and forward construct these strings, as described below:
 \begin{lstlisting}[language=Scala,captionpos=none]
@@
   type Alphabet = Char
-  type Answer = (Int,String,String) // score, seq1, seq2
-  override val h = maxBy((x:Answer)=>x._1)
-
-  def gap(current:Int, g:(Int,Int)):Int
-  def pair(a:Alphabet,b:Alphabet):Int
+  def start(x:Unit):Answer
+  def gap1(g:(Int,Int),a:Answer):Answer
+  def gap2(a:Answer,g:(Int,Int)):Answer
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
 }
@@
 trait SmithWatermanAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
   private val open = -3
   private val extend = -1
-  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, score + ( open + (size-1) * extend )) }
-  def pair(a:Char,b:Char) = if (a==b) 10 else -3
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
+  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, a + ( open + (size-1) * extend )) }
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 10 else -3)
 }
@@
 trait NeedlemanWunschAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
   private val open = -15
   private val extend = -1
-  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; score + ( open + (size-1) * extend ) }
-  def pair(a:Char,b:Char) = if (a==b) 4 else -3
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
+  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; a + ( open + (size-1) * extend ) }
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 4 else -3)
 }
@@
 \begin{lstlisting}[language=Scala,captionpos=none]
-trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
-  private def prettyGap(sw:Subword,in:Function1[Int,Char]):(String,String) = {
+trait SeqPrettyPrint extends SeqAlignSignature {
+  type Answer = (String,String)
+
+  def in1(k:Int):Alphabet
+  def in2(k:Int):Alphabet
+  private def prettyGap(sw:(Int,Int),in:Function1[Int,Char]):(String,String) = {
     val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
@@
 
+  def start(x:Unit) = (".",".")
+  def gap1(g:(Int,Int),a:Answer) = { val (g1,g2)=prettyGap(g,in1); (a._1+g1,a._2+g2) }
+  def gap2(a:Answer,g:(Int,Int)) = { val (g2,g1)=prettyGap(g,in2); (a._1+g1,a._2+g2) }
+  def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
+}
+
+trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
   val alignment:Tabulate = tabulate("M",(
-    empty                         ^^ { _ => (0,".",".") }
-  | seq1() -~ alignment           ^^ { case (g,(score,s1,s2)) => val (g1,g2)=prettyGap(g,in1); (gap(score,g),s1+g1,s2+g2) }
-  |           alignment ~- seq2() ^^ { case ((score,s1,s2),g) => val (g1,g2)=prettyGap(g,in2); (gap(score,g),s1+g2,s2+g1) }
-  | el1    -~ alignment ~- el2    ^^ { case (c1,((score,s1,s2),c2)) => (score+pair(c1,c2),s1+c1, s2+c2) }
+    empty                         ^^ start
+  | seq1() -~ alignment           ^^ gap1
+  |           alignment ~- seq2() ^^ gap2
+  | el1    -~ alignment ~- el2    ^^ pair
   ) aggregate h)
@@
   val axiom = alignment
-
-  def align(s1:String,s2:String) = parse(s1.toArray,s2.toArray).head
 }
@@
 object SeqAlign extends App {
-  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra   // Smith-Waterman
-  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra // Needleman-Wunsch
+  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra
+  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra
+  object pretty extends SeqAlignGrammar with SeqPrettyPrint
   val seq1 = "CGATTACA"
@@
 
-  val (swScore,sw1,sw2) = SWat.align(seq1,seq2)
-  println("Smith-Waterman alignment\n- Score: "+swScore+"\n- Seq1: "+sw1+"\n- Seq2: "+sw2+"\n")
-  val (nwScore,nw1,nw2) = NWun.align(seq1,seq2)
-  println("Needleman-Wunsch alignment\n- Score: "+nwScore+"\n- Seq1: "+nw1+"\n- Seq2: "+nw2+"\n")
+  def align(name:String,s1:String,s2:String,g:SeqAlignGrammar) = {
+    val (score,bt) = g.backtrack(s1.toArray,s2.toArray).head
+    val (a1,a2) = pretty.build(s1.toArray,s2.toArray,bt)
+    println(name+" alignment\n- Score: "+score+"\n- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
+  }
+
+  align("Smith-Waterman",seq1,seq2,SWat)
+  align("Needleman-Wunsch",seq1,seq2,SWat)
 }

### 2013-01-17 docs/report/benchmarks.tex
Nits.

@@
 In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).
-A naive idea would be to construct the alignment string immediately in the forward phase (that is immediately in the {\tt Answer}).
-However, is not efficient at all, both in terms of running time and space complexity perspective because substring are duplicated for every cell.
-The correct way to solve this issue is to use backtracking and forward construct these strings, as described below:
 \begin{lstlisting}[language=Scala,captionpos=none]
@@
   def start(x:Unit) = 0
-  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
-  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, a + ( open + (size-1) * extend )) }
-  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 10 else -3)
+  def gap1(g:(Int,Int),a:Int) = gap2(a,g) // by symmetry
+  def gap2(a:Int,g:(Int,Int)) =
+    { val size=g._2-g._1; Math.max(0, a + ( open + (size-1)*extend )) }
+  def pair(c1:Char,a:Int,c2:Char) = a + (if (c1==c2) 10 else -3)
 }
@@
   def start(x:Unit) = 0
-  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
-  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; a + ( open + (size-1) * extend ) }
-  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 4 else -3)
+  def gap1(g:(Int,Int),a:Int) = gap2(a,g) // by symmetry
+  def gap2(a:Int,g:(Int,Int)) =
+    { val size=g._2-g._1; a + ( open + (size-1)*extend ) }
+  def pair(c1:Char,a:Int,c2:Char) = a + (if (c1==c2) 4 else -3)
 }
@@
 
+To obtain a visual representation of the alignment, a naive idea would be to construct the two aligned strings immediately in the forward phase (in the {\tt Answer}).
+However, this approach must be avoided as it is not efficient at all, both in terms of running time and space complexity perspective because intermediate strings are created (and stored in memory) for every intermediate result.
+The correct way to solve this issue is to use backtracking and forward construct these strings with a pretty printing algebra:
 \begin{lstlisting}[language=Scala,captionpos=none]
@@
   type Answer = (String,String)
-
-  def in1(k:Int):Alphabet
-  def in2(k:Int):Alphabet
-  private def prettyGap(sw:(Int,Int),in:Function1[Int,Char]):(String,String) = {
-    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
+  def in1(k:Int):Alphabet; def in2(k:Int):Alphabet // make it visible
+  private def gap(sw:(Int,Int),in:Function1[Int,Char]) = {
+    val g=(sw._1 until sw._2).toList
+    (g.map{x=>in(x)}.mkString,g.map{x=>"-"}.mkString)
   }
-
   def start(x:Unit) = (".",".")
-  def gap1(g:(Int,Int),a:Answer) = { val (g1,g2)=prettyGap(g,in1); (a._1+g1,a._2+g2) }
-  def gap2(a:Answer,g:(Int,Int)) = { val (g2,g1)=prettyGap(g,in2); (a._1+g1,a._2+g2) }
+  def gap1(g:(Int,Int),a:Answer) =
+    { val (g1,g2)=gap(g,in1); (a._1+g1,a._2+g2) }
+  def gap2(a:Answer,g:(Int,Int)) =
+    { val (g2,g1)=gap(g,in2); (a._1+g1,a._2+g2) }
   def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
 }
+\end{lstlisting}
 
+Finally, we can describe the associated grammar and the programs that mixes algebra and grammar.
+Note that we need one instance of each pair of grammar and algebra.
+Once we have done that, we can request scores and backtracks associated with an evaluation algebra (Smith-Waterman or Needleman-Wunsch) and reuse the obtained backtrack to construct the matching aligned sequences:
+\begin{lstlisting}[language=Scala,captionpos=none]
 trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
-  val alignment:Tabulate = tabulate("M",(
+  val axiom:Tabulate = tabulate("M",(
     empty                     ^^ start
-  | seq1() -~ alignment           ^^ gap1
-  |           alignment ~- seq2() ^^ gap2
-  | el1    -~ alignment ~- el2    ^^ pair
+  | seq1() -~ axiom           ^^ gap1
+  |           axiom ~- seq2() ^^ gap2
+  | el1    -~ axiom ~- el2    ^^ pair
   ) aggregate h)
-
-  val axiom = alignment
 }
@@
     val (a1,a2) = pretty.build(s1.toArray,s2.toArray,bt)
-    println(name+" alignment\n- Score: "+score+"\n- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
+    println(name+" alignment\n- Score: "+score)
+    println("- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
   }
-
   align("Smith-Waterman",seq1,seq2,SWat)
@@
 
-\begin{lstlisting}[language=Scala,captionpos=none]
+\begin{lstlisting}[language=Scala]
 trait RepWorld extends NumericOps with TupleOps {
@@
     with CodeGen with App {
-  override val tps=(manifest[Alphabet],manifest[Answer])
+  val tps=(manifest[Alphabet],manifest[Answer])
   override val benchmark = true // display timing measurements
@@
 
+For further examples, we encourage you to have a look in the {\tt examples/} folder of the project.
+
 \newpage
-----------------------------------------
+\subsection{Other usage options}
+We here provide a list of relevant variables and traits that the programmer might be interested to use.
+This list only serves the purpose of documenting features that might otherwise be difficult to find within the code.
 
- and focus on three different problem: our matrix multiplication example,
+Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebrae easily.
+The signature needs to inherit either from {\tt Signature} or {\tt RNASignature}, would the RNA folding energies be needed.
+The grammar can be either single track or two-tracks by inheriting respectively from  {\tt ADPParsers} and {\tt TTParsers}.
+Note that RNA folding only works for single track grammars and enables the library setup (could be disabled by setting the RNASignature {\tt energies} flag to false).
 
-Example based
-- Matrix mult + parenthesizing
-- SWat / Needleman
+The code generator is used by simply mixing in the {\tt CodeGen} trait, which requires the idiom
+	\[\text{\tt val tps=(manifest[Alphabet],manifest[Answer])}\]
+to be placed anywhere at the intersection of the CodeGen inheritance and definition of these types (usually in the final program).
+Further configuration of the execution environment can be tuned by overriding the following variables: {\tt compiler} (for system paths and flags), {\tt cudaSplit} and {\tt cudaDevice}.
 
-go check the examples/ folder
+Finally, the user can look in the files {\tt ADPParsers.scala} and {\tt TTParsers.scala} for a list of the available terminals, and possibly create new ones, .
 
-List of options at the end
+---------------------------------------
 
-{\color{red} usage, make reference from problems to usage
+XXX: List of options at the end
 
-XXX: do not introduce pseudocode but refere to usage in RNAfold problem
-
-}
-
-\subsubsection{Matrix multiplication} \label{ex_matmult}
-\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
-import v4._
-
-trait MatrixSig extends Signature {
-  type Alphabet = (Int,Int) // Matrix(rows, columns)
-  val single:Alphabet=>Answer
-  val mult:(Answer,Answer)=>Answer
-}
-
-trait MatrixAlgebra extends MatrixSig {
-  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
-
-  override val h = minBy((a:Answer) => a._1)
-  val single = (a: Alphabet) => (0, a)
-  val mult = (a:Answer,b:Answer) =>
-    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
-}
-
-trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val axiom:Tabulate = tabulate("M",
-    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
-}
-
-object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
-  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
-}
-\end{lstlisting}
+{\color{red} usage, make reference from problems to usage}
 

### 2013-01-17 docs/report/benchmarks.tex
Nits.

@@
 \subsection{Program examples}
-In this section, we explain how to use DynaProg, as a parser library.
-Use an example based approach, we focus on three additional examples: Smith-Waterman/Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} to describe RNA library usage and reconstruction from backtrack trace, and finally we extend Matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.
+In this section, we explain how to use the DynaProg DSL using an example based approach.
+We focus on three additional examples: Smith-Waterman (\S\ref{swat_affine}) and Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} (\S\ref{zuker}, alternative) to describe RNA library usage and reconstruction from backtrack trace, and finally we extend matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.
 
 \subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
-First define a signature that can fit both algebrae, then specify for each algebra the functions.
-In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).
+First define a signature that can fit both algebrae, then specify for each algebra the related functions.
+In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general).
 \begin{lstlisting}[language=Scala,captionpos=none]
@@
 To obtain a visual representation of the alignment, a naive idea would be to construct the two aligned strings immediately in the forward phase (in the {\tt Answer}).
-However, this approach must be avoided as it is not efficient at all, both in terms of running time and space complexity perspective because intermediate strings are created (and stored in memory) for every intermediate result.
+However, this approach must be avoided as it is extremely inefficient, both in terms of running time and space complexity because intermediate strings are created (and stored in memory) for every intermediate result.
 The correct way to solve this issue is to use backtracking and forward construct these strings with a pretty printing algebra:
@@
 
-Finally, we can describe the associated grammar and the programs that mixes algebra and grammar.
+Finally, we describe the associated grammar and the programs that mixes the algebrae and the grammar.
 Note that we need one instance of each pair of grammar and algebra.
@@
 \subsubsection{Matrix multiplication with CUDA code generation} \label{ex_matmult_cuda_lms}
-Leveraging the existing definitions of the signature and grammar
+Leveraging the existing definitions of the signature and grammar (repeated here for convenience)
 \begin{lstlisting}[language=Scala,captionpos=none]
@@
 
-For further examples, we encourage you to have a look in the {\tt examples/} folder of the project.
+The complete source file of the presented problems can be found in the {\tt report/} folder. For further examples and variants, we encourage you to have a look in the {\tt examples/} folder.
 
@@
 The grammar can be either single track or two-tracks by inheriting respectively from  {\tt ADPParsers} and {\tt TTParsers}.
-Note that RNA folding only works for single track grammars and enables the library setup (could be disabled by setting the RNASignature {\tt energies} flag to false).
+Note that RNA folding only works for single track grammars and library setup is enabled by the usage of the trait {\tt RNASignature} (this could be changed by disabling the flag {\tt energies}).
 
-The code generator is used by simply mixing in the {\tt CodeGen} trait, which requires the idiom
+The code generator is used by simply mixing in the {\tt CodeGen} trait, using the following idiom
 	\[\text{\tt val tps=(manifest[Alphabet],manifest[Answer])}\]
-to be placed anywhere at the intersection of the CodeGen inheritance and definition of these types (usually in the final program).
+anywhere at the intersection of the CodeGen inheritance and definition of these types (usually in the final program).
 Further configuration of the execution environment can be tuned by overriding the following variables: {\tt compiler} (for system paths and flags), {\tt cudaSplit} and {\tt cudaDevice}.
+The {\tt benchmark} flag can be set to enable timing measurements.
 
-Finally, the user can look in the files {\tt ADPParsers.scala} and {\tt TTParsers.scala} for a list of the available terminals, and possibly create new ones, .
-
----------------------------------------
-
-XXX: List of options at the end
-
-{\color{red} usage, make reference from problems to usage}
+Finally, the user can look in the files {\tt ADPParsers.scala} and {\tt TTParsers.scala} for a list of the available terminals, and possibly create new ones.
 
@@
 \subsection{Metrics}
-{\color{red} The main metric we focus on is the running time.}
+The main metrix of interest is the running time.
+In an attempt to reduce the variance, we run 20 consecutive test and take the median running time (since median is less sensitive to outlier than the average).
 
@@
 \subsection{Benchmarking platform}
+Our benchmarking platform is a notebook with a Core i7-3720QM, 16Gb or main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8.
+Tests are made when the discrete GPU is not involved in rendering the user interface\footnote{This involves a workaround to keep the discrete GPU powered, while the CPU is involved in rendering graphics; source code available on demand.}.
+Due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.
 
-\subsubsection{Test Scala / C / CUDA}
-\subsubsection{Test 1 : CPU MatrixMult VS hand optimized VS GAPC VS ADPFusion)}
-\subsubsection{Test 2 : GPU MatirxMult VS hand optimized VS ATLP)}
-\subsubsection{Test 3 : CPU SWat VS hand optimized VS GAPC VS ADPFusion)}
-\subsubsection{Test 4 : GPU SWat VS hand optimized VS CUDAlign)}
-\subsubsection{Test 5 : CPU Zuker VS GAPC VS ADPFusion VS ViennaRNA)}
-\subsubsection{Test 6 : GPU Zuker VS RNAFold?/Lavenier VS ViennaRNA-OpenCL)}
+% ------------------------------------------------------------------------------------------------
+{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+% ------------------------------------------------------------------------------------------------
+
+\subsection{Matrix chain multiplication}
+\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
+\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?}
+
+\subsection{Smith-Watermann}
+\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
+\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign}
+
+\subsection{Zuker RNA folding}
+\subsubsection{CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA}
+\subsubsection{GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?}
+
+\subsection{Synthetical results}
+{\color{red} We want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
 
@@
 XXX: future work, resize tables appropriately as in GAPL.
-
-% ------------------------------------------------------------------------------------------------
-{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
-% ------------------------------------------------------------------------------------------------
 

### 2013-01-17 docs/report/benchmarks.tex
Nits.

--- 2013-01-17 docs/report/benchmarks.tex
@@
-\newpage
-\section{Usage} \label{usage}
-\subsection{Program examples}
-In this section, we explain how to use DynaProg, as a parser library.
-Use an example based approach, we focus on three additional examples: Smith-Waterman/Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} to describe RNA library usage and reconstruction from backtrack trace, and finally we extend Matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.
-
-\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
-First define a signature that can fit both algebrae, then specify for each algebra the functions.
-In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).
-A naive idea would be to construct the alignment string immediately in the forward phase (that is immediately in the {\tt Answer}).
-However, is not efficient at all, both in terms of running time and space complexity perspective because substring are duplicated for every cell.
-The correct way to solve this issue is to use backtracking and forward construct these strings, as described below:
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait SeqAlignSignature extends Signature {
-  type Alphabet = Char
-  def start(x:Unit):Answer
-  def gap1(g:(Int,Int),a:Answer):Answer
-  def gap2(a:Answer,g:(Int,Int)):Answer
-  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
-}
-
-trait SmithWatermanAlgebra extends SeqAlignSignature {
-  type Answer = Int
-  override val h = max[Int] _
-  private val open = -3
-  private val extend = -1
-  def start(x:Unit) = 0
-  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
-  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, a + ( open + (size-1) * extend )) }
-  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 10 else -3)
-}
-
-trait NeedlemanWunschAlgebra extends SeqAlignSignature {
-  type Answer = Int
-  override val h = max[Int] _
-  private val open = -15
-  private val extend = -1
-  def start(x:Unit) = 0
-  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
-  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; a + ( open + (size-1) * extend ) }
-  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 4 else -3)
-}
-\end{lstlisting}
-
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait SeqPrettyPrint extends SeqAlignSignature {
-  type Answer = (String,String)
-
-  def in1(k:Int):Alphabet
-  def in2(k:Int):Alphabet
-  private def prettyGap(sw:(Int,Int),in:Function1[Int,Char]):(String,String) = {
-    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
-  }
-
-  def start(x:Unit) = (".",".")
-  def gap1(g:(Int,Int),a:Answer) = { val (g1,g2)=prettyGap(g,in1); (a._1+g1,a._2+g2) }
-  def gap2(a:Answer,g:(Int,Int)) = { val (g2,g1)=prettyGap(g,in2); (a._1+g1,a._2+g2) }
-  def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
-}
-
-trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
-  val alignment:Tabulate = tabulate("M",(
-    empty                         ^^ start
-  | seq1() -~ alignment           ^^ gap1
-  |           alignment ~- seq2() ^^ gap2
-  | el1    -~ alignment ~- el2    ^^ pair
-  ) aggregate h)
-
-  val axiom = alignment
-}
-
-object SeqAlign extends App {
-  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra
-  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra
-  object pretty extends SeqAlignGrammar with SeqPrettyPrint
-  val seq1 = "CGATTACA"
-  val seq2 = "CCCATTAGAG"
-
-  def align(name:String,s1:String,s2:String,g:SeqAlignGrammar) = {
-    val (score,bt) = g.backtrack(s1.toArray,s2.toArray).head
-    val (a1,a2) = pretty.build(s1.toArray,s2.toArray,bt)
-    println(name+" alignment\n- Score: "+score+"\n- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
-  }
-
-  align("Smith-Waterman",seq1,seq2,SWat)
-  align("Needleman-Wunsch",seq1,seq2,SWat)
-}
-\end{lstlisting}
-
-\newpage
-\subsubsection{RNA folding} \label{ex_rnafold}
-We define a signature with two evaluation algebras: {\tt RNAFoldAlgebra} actually computes the folding whereas {\tt RNAFoldPrettyPrint} describes the folding in a string.
-The energy functions are provided by an external library ({\tt LibRNA}).
-This library encodes substring as (first character, last character) whereas our framework encodes them as (first character, first character + length), which explains the off-by-one corrections.
-{\tt energies} variable is set to false in {\tt RNAFoldPrettyPrint} because this algebra does not involve the LibRNA energies functions (that require encoding the input RNA sequence in a special format; this option is enabled by default in the RNASignature trait).
-
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait RNAFoldSig extends RNASignature {
-  def hairpin(ij:(Int,Int)):Answer
-  def stack(i:Int,s:Answer,j:Int):Answer
-  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
-  def mloop(i:Int,s:Answer,j:Int):Answer
-  def left(l:Answer,r:Int):Answer
-  def right(l:Int,r:Answer):Answer
-  def join(l:Answer,r:Answer):Answer
-}
-
-trait RNAFoldAlgebra extends RNAFoldSig {
-  type Answer = Int
-  import librna.LibRNA._ // indexing convention: first base,last base
-  def hairpin(ij:(Int,Int)) = hl_energy(ij._1,ij._2-1) // Eh
-  def stack(i:Int,s:Int,j:Int) = sr_energy(i,j) + s // Es
-  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) =
-      il_energy(ik._1,ik._2,lj._1-1,lj._2-1) + s // Ei
-  def mloop(i:Int,s:Int,j:Int) = s
-  def left(l:Int,r:Int) = l
-  def right(l:Int,r:Int) = r
-  def join(l:Int,r:Int) = l+r
-  override val h = min[Answer] _
-}
-
-trait RNAFoldPrettyPrint extends RNAFoldSig {
-  type Answer = String
-  override val energies=false
-  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
-  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
-  def stack(i:Int,s:String,j:Int) = "("+s+")"
-  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) =
-      "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
-  def mloop(i:Int,s:String,j:Int) = "("+s+")"
-  def left(l:String,r:Int) = l+"."
-  def right(l:Int,r:String) = "."+r
-  def join(l:String,r:String) = l+r
-}
-\end{lstlisting}
-
-\newpage
-We can then define the associated grammar
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
-  lazy val Qp:Tabulate = tabulate("Qp",(
-    seq(3,maxN)        ^^ hairpin
-  | eli   ~ Qp ~ eli   ^^ stack
-  | seq() ~ Qp ~ seq() ^^ iloop
-  | eli   ~ QM ~ eli   ^^ mloop
-  ) filter basepairing aggregate h)
-
-  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
-
-  lazy val Q:Tabulate = tabulate("Q",(
-    QM
-  | Q ~ eli ^^ left
-  | eli ~ Q ^^ right
-  | Qp
-  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
-
-  override val axiom = Q
-}
-\end{lstlisting}
-
-In the application, we create two objects, each combining the grammar with a particular algebra.
-We can optionally specify a coefficient parameter file with {\tt setParams(file:String)}, otherwise the \textit{Turner2004} coefficients are used.
-The library is automatically loaded and fed with the sequence to produce correct energy coefficients.
-We request both the score and the backtrack trace (in {\tt bt}) so that we can reconstruct the folding using the pretty printing grammar.
-\begin{lstlisting}[language=Scala,captionpos=none]
-object RNAFold extends App {
-  object fold extends RNAFoldGrammar with RNAFoldAlgebra
-  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
-
-  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuuuucaaaggaagagg"
-
-  val (score,bt) = fold.backtrack(seq.toArray).head
-  val res = pretty.build(seq.toArray,bt)
-  println("Folding : "+res+" (%5.2f)".format(score/100.0));
-}
-\end{lstlisting}
-
-\newpage
-\subsubsection{Matrix multiplication with CUDA code generation} \label{ex_matmult_cuda_lms}
-Leveraging the existing definitions of the signature and grammar
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait MatrixSig extends Signature {
-  type Alphabet = (Int,Int) // Matrix(rows, columns)
-  val single:Alphabet=>Answer
-  val mult:(Answer,Answer)=>Answer
-}
-
-trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val axiom:Tabulate = tabulate("M",
-    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
-}
-\end{lstlisting}
-We need describe the algebra functions in the LMS syntax ({\tt RepWorld}) that we can later compile to use as regular functions, augmented with C code description (necessary for code generation).
-Finally, we need to mix the {\tt CodeGen} trait to enable code generation and provide the manifest for input and ouput types ({\tt Alphabet} and {\tt Answer}).
-
-\begin{lstlisting}[language=Scala,captionpos=none]
-trait RepWorld extends NumericOps with TupleOps {
-  type Alphabet = (Int, Int)
-  type Answer = (Int, Int, Int)
-
-  def hf(a: Rep[Answer]) :Rep[Int] = a._2
-  def repSingle(a: Rep[Alphabet]): Rep[Answer] = (a._1, unit(0), a._2)
-  def repMult(l: Rep[Answer], r: Rep[Answer]): Rep[Answer] =
-    (l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3)
-}
-
-object MatrixMultLMS extends MatrixSig with MatrixGrammar
-    with CodeGen with App {
-  override val tps=(manifest[Alphabet],manifest[Answer])
-  override val benchmark = true // display timing measurements
-
-  // Algebra is defined immediately in the concrete program
-  type Answer = (Int, Int, Int)
-  val concreteProg = new RepWorld with RepPackage
-  override val h = minBy(concreteProg.gen(concreteProg.hf))
-  val single = concreteProg.gen(concreteProg.repSingle)
-  val mult = concreteProg.gen2(concreteProg.repMult)
-
-  val input = List((1,2),(2,20),(20,2),(2,4),(4,2),(2,1),(1,7),(7,3)).toArray
-  println(parse(input).head) // -> 1x3 matrix, 122 multiplications
-}
-\end{lstlisting}
-
-\newpage
-----------------------------------------
-
- and focus on three different problem: our matrix multiplication example,
-
-Example based
-- Matrix mult + parenthesizing
-- SWat / Needleman
-
-go check the examples/ folder
-
-List of options at the end
-
-{\color{red} usage, make reference from problems to usage
-
-XXX: do not introduce pseudocode but refere to usage in RNAfold problem
-
-}
-
-\subsubsection{Matrix multiplication} \label{ex_matmult}
-\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
-import v4._
-
-trait MatrixSig extends Signature {
-  type Alphabet = (Int,Int) // Matrix(rows, columns)
-  val single:Alphabet=>Answer
-  val mult:(Answer,Answer)=>Answer
-}
-
-trait MatrixAlgebra extends MatrixSig {
-  type Answer = (Int,(Int,Int)) // Answer(cost, Matrix(rows, columns))
-
-  override val h = minBy((a:Answer) => a._1)
-  val single = (a: Alphabet) => (0, a)
-  val mult = (a:Answer,b:Answer) =>
-    { val ((m1,(r1,c1)),(m2,(r2,c2)))=(a,b); (m1+m2+r1*c1*c2, (r1,c2)) }
-}
-
-trait MatrixGrammar extends ADPParsers with MatrixSig {
-  val axiom:Tabulate = tabulate("M",
-    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
-}
-
-object MatrixMult extends MatrixGrammar with MatrixAlgebra with App {
-  println(parse(Array((10,100),(100,5),(5,50)))) // List((7500,(10,50)))
-}
-\end{lstlisting}
-
 \section{Benchmarks} \label{benchmarks}
@@
 Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
-Finally, we asked lately the authors of \cite{gpu_rnafold} {\color{red} who did not respond so far}.
+Finally, we asked lately the authors of \cite{gpu_rnafold} who did not respond either to our sollicitations.
 
@@
 \subsection{Benchmarking platform}
+Our benchmarking platform is a notebook with a Core i7-3720QM, 16Gb or main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8.
+Tests are made when the discrete GPU is not involved in rendering the user interface\footnote{This involves a workaround to keep the discrete GPU powered, while the CPU is involved in rendering graphics; source code available on demand.}.
+Due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.
 
-\subsubsection{Test Scala / C / CUDA}
-\subsubsection{Test 1 : CPU MatrixMult VS hand optimized VS GAPC VS ADPFusion)}
-\subsubsection{Test 2 : GPU MatirxMult VS hand optimized VS ATLP)}
-\subsubsection{Test 3 : CPU SWat VS hand optimized VS GAPC VS ADPFusion)}
-\subsubsection{Test 4 : GPU SWat VS hand optimized VS CUDAlign)}
-\subsubsection{Test 5 : CPU Zuker VS GAPC VS ADPFusion VS ViennaRNA)}
-\subsubsection{Test 6 : GPU Zuker VS RNAFold?/Lavenier VS ViennaRNA-OpenCL)}
+% ------------------------------------------------------------------------------------------------
+{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+% ------------------------------------------------------------------------------------------------
+
+{\color{red}\ol
+\item Make LibRNA work with CUDA generated parsers
+\item Finish the LibRNA section, write some conclusion ideas
+\item Run benchmark and add results in report (automate, generate raw Matlab/TeX output)
+\item Stabilize times, statistics, use CUDA profiler(?)
+\ole}
+
+\subsection{Matrix chain multiplication}
+\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
+XXX
+
+\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?}
+XXX
+
+\subsection{Smith-Watermann}
+\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
+XXX
+
+\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign}
+XXX
+
+\subsection{Zuker RNA folding}
+\subsubsection{CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA}
+XXX
+
+\subsubsection{GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?}
+XXX
+
+\subsection{Synthetical results}
+XXX
+
+{\color{red} We want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
 
@@
 Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.}
-
-\section{Future work}
-We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
-\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
-However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
-\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program.
-Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}).
-These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar).
-\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation.
-Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
-\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
-The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
-\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.
-Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size).
-In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.
-
-Matrix of different dimensions must be stored in their own array (versus being in a single array of struct enclosing corresponding element of all matrices).
-Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description, there are two $O(n^2)$ matrices and one $O(n)$ matrix.
-This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once.
-\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently.
-For example, RNA is represented with only 4 letters (g,a,t,c), thus 4 symbols could be encoded in a single byte.
-Unfortunately, this optimizations seems to only apply for the input data.
-Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
-\item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded.
-Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, thereby making it possible to prune entire computation tiles.
-\item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
-Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser function calls into a single function making equivalent calls.
-The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
-\item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
-We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
-\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
-Hence cost and backtrack matrix would contain $k$ elements per cell.
-Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
-\ole

### 2013-01-17 docs/report/conclusion.tex
Nits.

% ------------------------------------------------------------------------------------------------
\section{Future work}
We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage. However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program. Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}). These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar). 
\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation. Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project. The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
\item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size. Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size). In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.

Matrix of different dimensions must be stored in their own array (versus being in a single array of struct enclosing corresponding element of all matrices). Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description, there are two $O(n^2)$ matrices and one $O(n)$ matrix. This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once. 
\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently. For example, RNA is represented with only 4 letters (g,a,t,c), thus 4 symbols could be encoded in a single byte. Unfortunately, this optimizations seems to only apply for the input data. Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
\item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded. Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, thereby making it possible to prune entire computation tiles.
\item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization. Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser function calls into a single function making equivalent calls. The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
\item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information. We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead. Hence cost and backtrack matrix would contain $k$ elements per cell. Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
\ole

% ------------------------------------------------------------------------------------------------
\section{Conclusion}
{\color{red} XXX}

\begin{verbatim}

recap of the report

recap of the results

big picture: with actual compiler technology we
Embedded DSL vs external DSL
Larger picture how to go further

\end{verbatim}

### 2013-01-17 docs/report/implementation.tex
Nits.

@@
 }
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\section{Usage} \label{usage}
+\subsection{Program examples}
+In this section, we explain how to use the DynaProg DSL using an example based approach.
+We focus on three additional examples: Smith-Waterman (\S\ref{swat_affine}) and Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} (\S\ref{zuker}, alternative) to describe RNA library usage and reconstruction from backtrack trace, and finally we extend matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.
+
+% ----------------------------------------------
+\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
+First define a signature that can fit both algebrae, then specify for each algebra the related functions.
+In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general).
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait SeqAlignSignature extends Signature {
+  type Alphabet = Char
+  def start(x:Unit):Answer
+  def gap1(g:(Int,Int),a:Answer):Answer
+  def gap2(a:Answer,g:(Int,Int)):Answer
+  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
+}
+
+trait SmithWatermanAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
+  private val open = -3
+  private val extend = -1
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Int) = gap2(a,g) // by symmetry
+  def gap2(a:Int,g:(Int,Int)) =
+    { val size=g._2-g._1; Math.max(0, a + ( open + (size-1)*extend )) }
+  def pair(c1:Char,a:Int,c2:Char) = a + (if (c1==c2) 10 else -3)
+}
+
+trait NeedlemanWunschAlgebra extends SeqAlignSignature {
+  type Answer = Int
+  override val h = max[Int] _
+  private val open = -15
+  private val extend = -1
+  def start(x:Unit) = 0
+  def gap1(g:(Int,Int),a:Int) = gap2(a,g) // by symmetry
+  def gap2(a:Int,g:(Int,Int)) =
+    { val size=g._2-g._1; a + ( open + (size-1)*extend ) }
+  def pair(c1:Char,a:Int,c2:Char) = a + (if (c1==c2) 4 else -3)
+}
+\end{lstlisting}
+
+To obtain a visual representation of the alignment, a naive idea would be to construct the two aligned strings immediately in the forward phase (in the {\tt Answer}).
+However, this approach must be avoided as it is extremely inefficient, both in terms of running time and space complexity because intermediate strings are created (and stored in memory) for every intermediate result.
+The correct way to solve this issue is to use backtracking and forward construct these strings with a pretty printing algebra:
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait SeqPrettyPrint extends SeqAlignSignature {
+  type Answer = (String,String)
+  def in1(k:Int):Alphabet; def in2(k:Int):Alphabet // make it visible
+  private def gap(sw:(Int,Int),in:Function1[Int,Char]) = {
+    val g=(sw._1 until sw._2).toList
+    (g.map{x=>in(x)}.mkString,g.map{x=>"-"}.mkString)
+  }
+  def start(x:Unit) = (".",".")
+  def gap1(g:(Int,Int),a:Answer) =
+    { val (g1,g2)=gap(g,in1); (a._1+g1,a._2+g2) }
+  def gap2(a:Answer,g:(Int,Int)) =
+    { val (g2,g1)=gap(g,in2); (a._1+g1,a._2+g2) }
+  def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
+}
+\end{lstlisting}
+
+Finally, we describe the associated grammar and the programs that mixes the algebrae and the grammar.
+Note that we need one instance of each pair of grammar and algebra.
+Once we have done that, we can request scores and backtracks associated with an evaluation algebra (Smith-Waterman or Needleman-Wunsch) and reuse the obtained backtrack to construct the matching aligned sequences:
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
+  val axiom:Tabulate = tabulate("M",(
+    empty                     ^^ start
+  | seq1() -~ axiom           ^^ gap1
+  |           axiom ~- seq2() ^^ gap2
+  | el1    -~ axiom ~- el2    ^^ pair
+  ) aggregate h)
+}
+
+object SeqAlign extends App {
+  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra
+  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra
+  object pretty extends SeqAlignGrammar with SeqPrettyPrint
+  val seq1 = "CGATTACA"
+  val seq2 = "CCCATTAGAG"
+
+  def align(name:String,s1:String,s2:String,g:SeqAlignGrammar) = {
+    val (score,bt) = g.backtrack(s1.toArray,s2.toArray).head
+    val (a1,a2) = pretty.build(s1.toArray,s2.toArray,bt)
+    println(name+" alignment\n- Score: "+score)
+    println("- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
+  }
+  align("Smith-Waterman",seq1,seq2,SWat)
+  align("Needleman-Wunsch",seq1,seq2,SWat)
+}
+\end{lstlisting}
+
+% ----------------------------------------------
+\newpage
+\subsubsection{RNA folding} \label{ex_rnafold}
+We define a signature with two evaluation algebras: {\tt RNAFoldAlgebra} actually computes the folding whereas {\tt RNAFoldPrettyPrint} describes the folding in a string.
+The energy functions are provided by an external library ({\tt LibRNA}).
+This library encodes substring as (first character, last character) whereas our framework encodes them as (first character, first character + length), which explains the off-by-one corrections.
+{\tt energies} variable is set to false in {\tt RNAFoldPrettyPrint} because this algebra does not involve the LibRNA energies functions (that require encoding the input RNA sequence in a special format; this option is enabled by default in the RNASignature trait).
+
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait RNAFoldSig extends RNASignature {
+  def hairpin(ij:(Int,Int)):Answer
+  def stack(i:Int,s:Answer,j:Int):Answer
+  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
+  def mloop(i:Int,s:Answer,j:Int):Answer
+  def left(l:Answer,r:Int):Answer
+  def right(l:Int,r:Answer):Answer
+  def join(l:Answer,r:Answer):Answer
+}
+
+trait RNAFoldAlgebra extends RNAFoldSig {
+  type Answer = Int
+  import librna.LibRNA._ // indexing convention: first base,last base
+  def hairpin(ij:(Int,Int)) = hl_energy(ij._1,ij._2-1) // Eh
+  def stack(i:Int,s:Int,j:Int) = sr_energy(i,j) + s // Es
+  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) =
+      il_energy(ik._1,ik._2,lj._1-1,lj._2-1) + s // Ei
+  def mloop(i:Int,s:Int,j:Int) = s
+  def left(l:Int,r:Int) = l
+  def right(l:Int,r:Int) = r
+  def join(l:Int,r:Int) = l+r
+  override val h = min[Answer] _
+}
+
+trait RNAFoldPrettyPrint extends RNAFoldSig {
+  type Answer = String
+  override val energies=false
+  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
+  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
+  def stack(i:Int,s:String,j:Int) = "("+s+")"
+  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) =
+      "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
+  def mloop(i:Int,s:String,j:Int) = "("+s+")"
+  def left(l:String,r:Int) = l+"."
+  def right(l:Int,r:String) = "."+r
+  def join(l:String,r:String) = l+r
+}
+\end{lstlisting}
+
+\newpage
+We can then define the associated grammar
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
+  lazy val Qp:Tabulate = tabulate("Qp",(
+    seq(3,maxN)        ^^ hairpin
+  | eli   ~ Qp ~ eli   ^^ stack
+  | seq() ~ Qp ~ seq() ^^ iloop
+  | eli   ~ QM ~ eli   ^^ mloop
+  ) filter basepairing aggregate h)
+
+  lazy val QM:Tabulate = tabulate("QM",(Q ~ Q ^^ join) filter((i:Int,j:Int)=>i<=j+4) aggregate h)
+
+  lazy val Q:Tabulate = tabulate("Q",(
+    QM
+  | Q ~ eli ^^ left
+  | eli ~ Q ^^ right
+  | Qp
+  ) filter((i:Int,j:Int)=>i<=j+2) aggregate h)
+
+  override val axiom = Q
+}
+\end{lstlisting}
+
+In the application, we create two objects, each combining the grammar with a particular algebra.
+We can optionally specify a coefficient parameter file with {\tt setParams(file:String)}, otherwise the \textit{Turner2004} coefficients are used.
+The library is automatically loaded and fed with the sequence to produce correct energy coefficients.
+We request both the score and the backtrack trace (in {\tt bt}) so that we can reconstruct the folding using the pretty printing grammar.
+\begin{lstlisting}[language=Scala,captionpos=none]
+object RNAFold extends App {
+  object fold extends RNAFoldGrammar with RNAFoldAlgebra
+  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint
+
+  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuuuucaaaggaagagg"
+
+  val (score,bt) = fold.backtrack(seq.toArray).head
+  val res = pretty.build(seq.toArray,bt)
+  println("Folding : "+res+" (%5.2f)".format(score/100.0));
+}
+\end{lstlisting}
+
+% ----------------------------------------------
+\newpage
+\subsubsection{Matrix multiplication with CUDA code generation} \label{ex_matmult_cuda_lms}
+Leveraging the existing definitions of the signature and grammar (repeated here for convenience)
+\begin{lstlisting}[language=Scala,captionpos=none]
+trait MatrixSig extends Signature {
+  type Alphabet = (Int,Int) // Matrix(rows, columns)
+  val single:Alphabet=>Answer
+  val mult:(Answer,Answer)=>Answer
+}
+
+trait MatrixGrammar extends ADPParsers with MatrixSig {
+  val axiom:Tabulate = tabulate("M",
+    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
+}
+\end{lstlisting}
+We need describe the algebra functions in the LMS syntax ({\tt RepWorld}) that we can later compile to use as regular functions, augmented with C code description (necessary for code generation).
+Finally, we need to mix the {\tt CodeGen} trait to enable code generation and provide the manifest for input and ouput types ({\tt Alphabet} and {\tt Answer}).
+
+\begin{lstlisting}[language=Scala]
+trait RepWorld extends NumericOps with TupleOps {
+  type Alphabet = (Int, Int)
+  type Answer = (Int, Int, Int)
+
+  def hf(a: Rep[Answer]) :Rep[Int] = a._2
+  def repSingle(a: Rep[Alphabet]): Rep[Answer] = (a._1, unit(0), a._2)
+  def repMult(l: Rep[Answer], r: Rep[Answer]): Rep[Answer] =
+    (l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3)
+}
+
+object MatrixMultLMS extends MatrixSig with MatrixGrammar
+    with CodeGen with App {
+  val tps=(manifest[Alphabet],manifest[Answer])
+  override val benchmark = true // display timing measurements
+
+  // Algebra is defined immediately in the concrete program
+  type Answer = (Int, Int, Int)
+  val concreteProg = new RepWorld with RepPackage
+  override val h = minBy(concreteProg.gen(concreteProg.hf))
+  val single = concreteProg.gen(concreteProg.repSingle)
+  val mult = concreteProg.gen2(concreteProg.repMult)
+
+  val input = List((1,2),(2,20),(20,2),(2,4),(4,2),(2,1),(1,7),(7,3)).toArray
+  println(parse(input).head) // -> 1x3 matrix, 122 multiplications
+}
+\end{lstlisting}
+
+The complete source file of the presented problems can be found in the {\tt report/} folder. For further examples and variants, we encourage you to have a look in the {\tt examples/} folder.
+
+% ------------------------------------------------------------------------------------------------
+\newpage
+\subsection{Other usage options}
+We here provide a list of relevant variables and traits that the programmer might be interested to use.
+This list only serves the purpose of documenting features that might otherwise be difficult to find within the code.
+
+Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebrae easily.
+The signature needs to inherit either from {\tt Signature} or {\tt RNASignature}, would the RNA folding energies be needed.
+The grammar can be either single track or two-tracks by inheriting respectively from  {\tt ADPParsers} and {\tt TTParsers}.
+Note that RNA folding only works for single track grammars and library setup is enabled by the usage of the trait {\tt RNASignature} (this could be changed by disabling the flag {\tt energies}).
+
+The code generator is used by simply mixing in the {\tt CodeGen} trait, using the following idiom
+	\[\text{\tt val tps=(manifest[Alphabet],manifest[Answer])}\]
+anywhere at the intersection of the CodeGen inheritance and definition of these types (usually in the final program).
+Further configuration of the execution environment can be tuned by overriding the following variables: {\tt compiler} (for system paths and flags), {\tt cudaSplit} and {\tt cudaDevice}.
+The {\tt benchmark} flag can be set to enable timing measurements.
+
+Finally, the user can look in the files {\tt ADPParsers.scala} and {\tt TTParsers.scala} for a list of the available terminals, and possibly create new ones.

### 2013-01-17 docs/report/report.tex
Nits.

@@
 \input{inc/style.sty}
-\usepackage{multicol}
-\usepackage{etoolbox}
+\usepackage{multicol,etoolbox}
 
@@
 \input{benchmarks}
-
-\section{Conclusion}
-{\color{red} XXX}
-
-\begin{verbatim}
-
-recap of the report
-
-recap of the results
-
-big picture: with actual compiler technology we
-Embedded DSL vs external DSL
-Larger picture how to go further
-
-\end{verbatim}
+\input{conclusion}
 
 % ------------------------------------------------------------------------------------------------
-\newpage
-\section*{Planning}
-\subsubsection*{Todo @TCK}\ul
-\item Test/proof parsers are correct -- make sure implementation is correct
-\item Automate test to compare against implementation
-\item Benchmarks -- use CUDA profiler(?)
-\item Write report
-\item Port LibRNA for CUDA?
-\ule
-
-\subsubsection*{Todo @Manohar}\ul
-\item Integrate LMS code generation into v4.
-\item Fix Zuker coefficients
-\ule
-
-Legacy roadmap deadlines: \\
-\begin{tabular}{ll}
-Nov 16 & Rules normalization and automatic backtracking \\
-	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
-Nov 23 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
-Nov 30 &--- gap due to LMS missing knowledge --- \\
-Dec 7 & Benchmarking, grammar analysis \\
-Dec 14 & First thoughts for larger than device memory \\
-Dec 21 & Writing report \\
-Jan 4 & --- holiday --- \\
-Jan 18 & Writing report: implementation description and plan for future work
-\end{tabular}
-
-\subsubsection*{Journal}
-\begin{tabular}{ll}
-Sep. 17	& Getting started with the project, reading related work on hash maps. \\
-Sep. 24	& Parallel hashing paper solve the problem for 32bit key/value pair. Stripped CudPP. \\
-		& Devised (but not implemented) extension beyond 64bit using memory areas locks. \\
----		& Change of project suggested by Vojin (supervisor): joint work with Manohar on DP \\
-Oct. 01	& Problems specifications: serial/non-serial, started CUDA implementation with blocks. \\
-Oct. 08	& CudAlign solves serial monadic, might adapt it (but complicated / ad-hoc / in progress). \\
-		& Focus on non-serial problems that fit in GPU memory. \\
-Oct. 15	& First working implementation for non-serial problems (rectangle, triangle, parallelogram) \\
-Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
-Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. ADP (Manohar) \\
-Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
-Nov. 12	& Rework ADP parsers: cyclic, two-track grammars and automatic backtracking discussion. \\
-Nov. 19	& Explorations LMS and Macros for C code, implementation is quite different, hence ad-hoc. \\
-Nov. 26	& Full backtracking stack: apply / unapply / reapply, refactoring of the classes. \\
-Dec. 03	& JNI for LibRNA to get coefficients for Zuker, errors in algebra (quite hairy code). \\
-Dec. 10	& Rework concatenation operators, yield analysis, code generation. \\
-Dec. 17	& Detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support. \\
-Dec. 24	& (sick 4 days), code generation: full JNI conversion support, rework CodeCompiler. \\
-Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
-Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
-Jan. 14	& Report improvements, fixed Zuker (coherent with GAPC), {\color{red} XXX} \\
-\end{tabular}
+%\section*{Planning}
+%Legacy roadmap deadlines: \\
+%\begin{tabular}{ll}
+%Nov 16 & Rules normalization and automatic backtracking \\
+%	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
+%Nov 23 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
+%Nov 30 &--- gap due to LMS missing knowledge --- \\
+%Dec 7 & Benchmarking, grammar analysis \\
+%Dec 14 & First thoughts for larger than device memory \\
+%Dec 21 & Writing report \\
+%Jan 4 & --- holiday --- \\
+%Jan 18 & Writing report: implementation description and plan for future work
+%\end{tabular}
+%
+%\subsubsection*{Journal}
+%\begin{tabular}{ll}
+%Sep. 17	& Getting started with the project, reading related work on hash maps. \\
+%Sep. 24	& Parallel hashing paper solve the problem for 32bit key/value pair. Stripped CudPP. \\
+%		& Devised (but not implemented) extension beyond 64bit using memory areas locks. \\
+%---		& Change of project suggested by Vojin (supervisor): joint work with Manohar on DP \\
+%Oct. 01	& Problems specifications: serial/non-serial, started CUDA implementation with blocks. \\
+%Oct. 08	& CudAlign solves serial monadic, might adapt it (but complicated / ad-hoc / in progress). \\
+%		& Focus on non-serial problems that fit in GPU memory. \\
+%Oct. 15	& First working implementation for non-serial problems (rectangle, triangle, parallelogram) \\
+%Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
+%Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. ADP (Manohar) \\
+%Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
+%Nov. 12	& Rework ADP parsers: cyclic, two-track grammars and automatic backtracking discussion. \\
+%Nov. 19	& Explorations LMS and Macros for C code, implementation is quite different, hence ad-hoc. \\
+%Nov. 26	& Full backtracking stack: apply / unapply / reapply, refactoring of the classes. \\
+%Dec. 03	& JNI for LibRNA to get coefficients for Zuker, errors in algebra (quite hairy code). \\
+%Dec. 10	& Rework concatenation operators, yield analysis, code generation. \\
+%Dec. 17	& Detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support. \\
+%Dec. 24	& (sick 4 days), code generation: full JNI conversion support, rework CodeCompiler. \\
+%Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
+%Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
+%Jan. 14	& Report improvements, fixed Zuker (coherent with GAPC), {\color{red} XXX} \\
+%\end{tabular}
 

### 2013-01-18 src/main/scala/v4/ADPParsers.scala
Fixed one bug, still incorrect.

@@ -81,11 +81,11 @@ trait RNASignature extends Signature {
       case ('\1','\4') | ('\4','\1') | ('\4','\3') | ('\3','\4') | ('\3','\2') | ('\2','\3') => true
       case _ => false
     }
-    val (args,body,tpe)=(List(("i","Int"),("j","Int")),"return bp_index(_in1[i],_in1[j])!=NO_BP;","Boolean")
+    val (args,body,tpe)=(List(("i","Int"),("j","Int")),"return bp_index(_in1[i],_in1[j-1])!=NO_BP;","Boolean")
   }
   val stackpairing = new ((Int,Int)=>Boolean) with CFun {
     def apply(i:Int,j:Int):Boolean = basepairing(i,j) && basepairing(i+1,j-1)
-    val (args,body,tpe)=(List(("i","Int"),("j","Int")),"return bp_index(_in1[i],_in1[j])!=NO_BP && bp_index(_in1[i+1],_in1[j-1])!=NO_BP;","Boolean")
+    val (args,body,tpe)=(List(("i","Int"),("j","Int")),"return bp_index(_in1[i],_in1[j-1])!=NO_BP && bp_index(_in1[i+1],_in1[j-2])!=NO_BP;","Boolean")
   }
   def convert(in:String):Array[Alphabet] = in.toArray.map { case 'A'|'a'=>'\1' case 'C'|'c'=>'\2' case 'G'|'g'=>'\3' case 'U'|'u'=>'\4' case _ => sys.error("Invalid sequence") }
 }

### 2013-01-18 docs/report/conclusion.tex
Nits.

@@
 Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}).
-These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (hereby knowing how to optimize manually the grammar).
+These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (thereby knowing how to optimize manually the grammar).
 \item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation.

### 2013-01-18 docs/report/design.tex
Nits.

@@
 
-Since progress is made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses.
+Because dependencies are along line, column or intermediate elements (\S\ref{problem_examples}),  parallel progress can be made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses.
 Hence we address them by diagonal.
@@
 Noticing that the triangle matrix shape is just half the parallelogram, we could use the same addressing scheme, but we would use twice more memory than necessary.
-Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl.
+Hence we need a tailored formula: the size of the triangle embedded in a square of side $k$ (incl.
 diagonal) is $\frac{k\cdot(k+1)}{2}$; knowing that fact, the index of the element $(i,j)$ of the triangle (starting with the main diagonal when $i=j$), we obtain the following formula (with $M_H=M_W=n$):
@@
 However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle).
-The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles.
+The solution consists of breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles.
 Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe.
@@
 At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
-However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
+However, the concern in this project is that the GPU code sensibly differs from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
+We want to discuss how LMS could best be used in our project:
 
-Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
-\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}.
-We tried to use LMS to generate both the Scala and CUDA program.
-This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}).
-This issue is particularly annoying in our particular case because concatenation construct pair of result (hence multiple concatenation would produce nested pairs) whereas we want to present its result as a flattened list of arguments suitable for immediate use by a function.
+LMS only supports a subset of Scala.
+Embedding both algebrae and grammars into the LMS framework would reduce the expressivity of algebrae.
+Taking the example of matrix multiplication, the user would like to not only solve the dynamic problem on GPU, but also leverage the graphic card, the disk or the network to compute the matrices products.
+Since these resources are outside for the LMS world, the user needs to write an ad-hoc DSL for every particular function.
+In order to avoid this constraint, we need to decouple the grammar and algebra generation.
 
-This approach also complicates the usage of Scala parsers because the algebra can only contain a subset of the Scala language can be represented in LMS.
-Taking the example of matrix multiplication, the user would like not only to solve the dynamic problem on GPU, but also to leverage the graphic card to compute matrices products.
-Since associated resources description cannot be represented by LMS nodes, the user would need to extend LMS with ad-hoc DSL for every particular function that is not representable in LMS.
+For the generation of algebra functions, since CUDA types are restricted, they are representable in LMS.
+Hence we can leverage LMS for CUDA-compatible algebrae whereas regular Scala can be used for complex algebrae being executed on CPU exclusively.
 
-\item A second possibility would be to analyze the (user) Scala program and produce another Scala program (simplified) in the restricted subset that is compatible with LMS, hereby providing some ad-hoc transformations either during the conversion (user to simplified), or at a later stage during the optional optimization phase of LMS.
-This approach seems quite complicated and error prone as it introduces multiple sources of possible errors: during the ad-hoc conversion (user to simplified), during the optimization phase, and during the conversion to C code.
-Additionally, since dynamic programming require special scheduling to respect the dependencies between intermediate solutions, we would need at a later stage to combine ad-hoc scheduling code with the result of the LMS code generation.
-\item Since ADP normalizes the structure of dynamic programming, the concerns between grammar and algebra are well separated.
-Since the grammar has a limited number of parsers, it is possible to provide an ad-hoc implementation for each of them thereby allowing better integration.
-Given that the algebra can contain arbitrary used-defined functions, we avoid duplicating the efforts and rely on LMS to generate the corresponding C code.
+It remains to generate the C code corresponding to the grammar; we argue in favor of a specialized conversion rather than generation through LMS for the following reasons:\ul
+\item We have only 6 parser classes to convert (tabulations, terminals, aggregations, filters, alternatives, concatenations), the rest of the user program is part of the algebra.
+These parsers are combined with very little modifications to create the grammar.
+\item The behavior of the CUDA parsers sensibly differs from that available in Scala (\S\ref{normalization}). This would imply writing optimization phases in LMS to convert them appropriately.
+\item LMS operations on collections (lists, hash maps, arrays, ...) do not require special scheduling as all elements are treated independently; nevertheless dynamic programming introduces dependencies between elements, thus requiring particular scheduling interleaved with the computation.
+Enriching LMS with primitives to solve these constraints might be very complex if we want to generalize.
+However, we could leverage the knowledge that the pattern belongs to dynamic programming, hereby reducing the complexity of the work and possibly being more efficient.
+Additionally, a specific memory pattern is required to increase efficiency (\S\ref{mem_layout}), we also need to reuse this knowledge to LMS, so that it can generate optimal code, so we might end up writing a specialized code generator.
+\item Finally, LMS is a rather complicated framework, with sparse documentation.
+Although many features are available (support for tuples, objects, matrices, ...), it is not easy to figure out where to find them (because they may reside on different branches or are not tested), hence the time taken to improve the framework to suit our needs far outweighs the time to build a tailored solution, which is the choice we are somehow forced to take, given the time constraints of this project.
 \ule
-For all the previous reasons it is more simple to provide an ad-hoc code generation for the grammar, whereas the algebra can be generated by LMS, hereby avoiding to duplicate code generation efforts while at the same time allowing us to integrate tightly the grammar parsers into the surrounding scheduling C code.
-
-From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}.
-Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress rather than improving an existing project (but I would reconsider my position given different deadlines).
-Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating simple functions for a costing algebra.
+Since we need to operate a lot of transformation, inside LMS or in specialized generation, the choice of specialized code generation was guided by the engineering principle of steering away from unnecessary complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}, hence reducing the number of potential sources of errors.
+During our experiments with LMS, we also investigated in the Scala 2.10 macros\footnote{(Macros operate directly on the internal representation of the Scala compiler, after the type analysis and and before the code generation phase.} \cite{scala_macros}, which represent a possible alternative to LMS for generating very simple functions for a costing algebra.
 

### 2013-01-18 docs/report/introduction.tex
Nits.

@@
 Because we don't know a priori where we could split the chain, we would need to explore an exponential number of possibilities.
-Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, hereby reducing the problem to a polynomial complexity.
+Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, thereby reducing the problem to a polynomial complexity.
 
@@
 \subsection{Graphic cards}
-Modern graphic cards\footnote{We cover here interesting features of the CUDA devices and programming paradigm; however, the same concept should be applicable to graphic cards from other vendors.} are powered by massively parallel processor: they can typically run hundreds or thousands of cores, each able to schedule multiple threads.
+Modern graphic cards\footnote{We cover here interesting features of the CUDA devices and programming paradigm; however, the same concept should be applicable to graphic cards from other vendors.} are powered by massively parallel processors: they can typically run hundreds or thousands of cores, each able to schedule multiple threads.
 The threads are usually grouped in warps that are scheduled synchronously.
-This means that if there is a divergence in the execution path, both alternative are executed sequentially, hereby stalling other warp's threads.
+This means that if there is a divergence in the execution path, both alternatives are executed sequentially, thereby stalling other warp's threads.
 Threads are logically grouped in blocks by the programmer whereas warps correspond to a physical constraint.

### 2013-01-18 docs/report/problems.tex
Nits.

@@
 \subsubsection{Zuker RNA folding} \label{zuker} \ol
-\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hereby much more complicated than Nussinov \S\ref{nussinov}).
+\item Problem: folding a RNA string $S$ over itself $|S|=n$ by minimizing the free energy (which is based on actual measurements, hence much more complicated than Nussinov \S\ref{nussinov}).
 \item Matrices: $V_{n\times n}, W_{n\times n}, F_n$ (Free Energy),  $BV_{n \times n}, BW_{n \times n}, BF_n$

### 2013-01-18 docs/report/introduction.tex
Bug with bounded sequences identified !

@@
 The order in which matrices are combined changes the number of required scalar multiplications, we would want to find an optimal order that minimizes this number.
-Notice that if knew how to split of the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them.
+Notice that if we knew how to split the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them.
 Such combinatorial problems verify the Bellman's principle\cite{bellman_principle}: \textit{<<optimal solutions depends on optimal solutions of sub-problems>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}.
-Because we don't know a priori where we could split the chain, we would need to explore an exponential number of possibilities.
-Using the Bellman's principle, we can memoize intermediate optimal solutions to save redundant computations, thereby reducing the problem to a polynomial complexity.
+In order to find the optimal way of splitting the chain, we would need to explore an exponential number of possibilities.
+Using the Bellman's principle, we can memorize intermediate optimal solutions to save redundant computations, thereby reducing the problem to a polynomial complexity.
 

### 2013-01-18 src/librna/librna_impl.h

@@ -20,10 +20,8 @@ enum bp_t { N_BP, CG_BP, GC_BP, GU_BP, UG_BP, AU_BP, UA_BP, NO_BP };
 typedef unsigned int rsize;
 
 my_dev static paramT *my_P = NULL;
-/*
-my_dev static char* my_seq = NULL;
-my_dev static int my_len = 0;
-*/
+// my_dev static char* my_seq = NULL;
+// my_dev static int my_len = 0;
 
 my_dev int termau_energy(rsize i, rsize j);
 my_dev int hl_energy(rsize i, rsize j);

### 2013-01-18 src/main/scala/v4/examples/Nussinov.scala

@@ -2,97 +2,90 @@ package v4.examples
 import v4._
 
 // Nussinov RNA folding
-trait PairingSig extends Signature {
-  def nil(a: Unit): Answer
-  def left(c: Alphabet, a: Answer): Answer
-  def right(a: Answer, c: Alphabet): Answer
-  def pair(l: Alphabet, a: Answer, r: Alphabet): Answer
-  def split(l: Answer, r: Answer): Answer
+trait NussinovSig extends Signature {
+  type Alphabet = Char
+  val nil:Unit=>Answer
+  val left:(Alphabet,Answer)=>Answer
+  val right:(Answer,Alphabet)=>Answer
+  val pair:(Alphabet,Answer,Alphabet)=>Answer
+  val split:(Answer,Answer)=>Answer
 }
 
-trait PairingAlgebra extends PairingSig {
+trait NussinovAlgebra extends NussinovSig {
   type Answer = Int
-
-  def nil(a: Unit) = 0
-  def left(c: Alphabet,a: Answer) = a
-  def right(a: Answer, c:Alphabet) = a
-  def pair(l: Alphabet, a: Answer, r: Alphabet) = a+1
-  def split(l: Answer, r: Answer) = l + r
-
+  /*
+  val nil = (a: Unit) => 0
+  val left = (c: Alphabet,a: Answer) => a
+  val right = (a: Answer, c:Alphabet) => a
+  val pair = (l: Alphabet, a: Answer, r: Alphabet) => a+1
+  val split = (l: Answer, r: Answer) => l + r
+  */
+  val nil   = cfun1((a: Unit) => 0, "", "return 0;")
+  val left  = cfun2((c: Alphabet,a: Answer) => a, "c,a", "return a;")
+  val right = cfun2((a: Answer,c: Alphabet) => a, "a,c", "return a;")
+  val pair  = cfun3((l: Alphabet, a: Answer, r: Alphabet) => a+1, "l,a,r", "return a+1;")
+  val split = cfun2((l: Answer, r: Answer) => l + r, "l,r", "return l+r;")
   override val h = max[Answer] _
 }
 
-/*
-trait PrettyPrintAlgebra extends PairingSig {
-  type Answer = (String,String)
-  type Alphabet = Char
+trait NussinovString extends NussinovSig {
+  type Answer = String
[…]

### 2013-01-19 docs/report/benchmarks.tex
Fixed one bug, still incorrect.
Nits.

@@
 \subsection{Metrics}
-The main metrix of interest is the running time.
-In an attempt to reduce the variance, we run 20 consecutive test and take the median running time (since median is less sensitive to outlier than the average).
+The main metrics of interest is the running time.
+In an attempt to reduce the variance, we would like to run multiple consecutive test and take the median running time (since median is less sensitive to outlier than the average).
+Unfortunately, several factors hampers these ideal conditions.
+First the variance in the running time of CUDA kernels might be quite significant, in particular for short running time.
+This is due to the fact that the GPU needs to be 'warmed-up' during few seconds before actual computation can happen
+
+  In case the variance is significant, we wou
+
+talk about input sensivity, backup with results on rnafold and matrix mult
 
@@
 
+XXX: future work: infer automatically "always non-emptiness"
+
 \subsection{Scala parsers}

### 2013-01-19 docs/report/implementation.tex
Nits.

@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \subsection{LibRNA}
-Since the energy computation for RNA folding involve complex coefficient and computations (that are also standardized through libraries and coefficient files), we might want to provide the user with it.
-To do that, we based our library on \cite{gapc_thesis} which itself is based on \href{http://www.tbi.univie.ac.at/~ivo/RNA/}{ViennaRNA}.\cite{vienna_rna}.
+Since the energy computation for RNA secondary structure prediction (folding the sequence in two dimensions) involve complex coefficients and computations (seemingly standardized in coefficient files), we might want to provide the user with a simple interface to benefit from it.
+To do that, we based our library on the work of GAPC\cite{gapc_thesis} which itself is based on \href{http://www.tbi.univie.ac.at/~ivo/RNA/}{ViennaRNA}\cite{vienna_rna}.
 Since the library is provided in C, we rely on JNI to reuse the code without modifying it; this allows Scala to immediately benefit from it, but also makes possible to write a GPU version, provided that the related functions are simple enough to be expressible in CUDA.
+Our work in this direction is mainly focused on integration, we do not want to discuss the implementation details here but simply give a overview of what we transformed and adapted to suit our needs.
 
-{\color{red}
-XXX: attempt to strip it to make CUDA version
-}
+In the first phase, we adapted the library embedded in GAPC to obtain coefficients.
+Since GAPC is written in C, we had to write a JNI interface to let the Scala code communicate with the libraries.
+Once this step has been achieved, we focused on obtaining correct results for RNA folding.
+This has been achieved by a thorough analysis of the GAPC related code, and took quite a long time due to bugs that were hard to find.
+
+In parallel with this work, we focused on making the code compatible with CUDA.
+We managed to significantly reduce its size by removing unused functions (actually, all the programs of the Vienna package reuse the library, each introducing its own functions).
+This also enabled us to have a better understanding of the involved computations, thereby helping us to clearly separate the coefficient file processing and the energies computations.
+We also provided small optimizations towards parallelization and simplified the sequence management (because it needs to be converted in a particular format for the library to efficiently process it).
+
+Once the library was ready to fit on CUDA devices, we integrated it into DynaProg.
+Unfortunately, since the library adds a significant volume of code (because some coefficients are embedded in C files), the compilation process length was increased (from approx.
+2 to 7 seconds).
+To reduce this penalty (towards benchmarking), we introduced memoization of the compilation results in our code generator, thereby avoiding duplicated compilations of the program for same length of sequences (generated programs are tailored for particular sequences lengths).
 

### 2013-01-19 docs/report/benchmarks.tex
More report.

--- 2013-01-17 docs/report/benchmarks.tex
@@
 \subsection{Metrics}
-The main metrix of interest is the running time.
-In an attempt to reduce the variance, we run 20 consecutive test and take the median running time (since median is less sensitive to outlier than the average).
+The main metrics of interest is the running time.
+In an attempt to reduce the variance, we would like to run multiple consecutive test and take the median running time, since the median is less sensitive to outlier than the average\cite{perfeval}.
+Unfortunately, several factors hampers these ideal conditions.
+First the variance in the running time of CUDA kernels might be significant, in particular for short running time.
+This is due to the fact that the GPU needs to be 'warmed-up' before actual computation can happen.
+Similarly, the JVM is also subject to running time variance that is mainly due to the garbage collection\footnote{\url{http://www.oracle.com/technetwork/java/javase/gc-tuning-6-140523.html\#cms.overhead}} and JIT optimizations \cite{java_jit}.
+
+Also the input and problem might introduce variance.
+As example, we can consider two extreme cases: matrix chain multiplication and Zuker RNA folding, with a test environment of 100 random inputs (of length respectively 512 and 80) and a GPU warmup of 10 computations.
+In this settings, matrix chain multiplication computations are executed in a perfectly constant time (0.127 seconds), which mean that we sufficiently reduced the noise.
+Oppositely, the Zuker RNA folding running times appear much more scattered as presented below:
+
+\begin{figure}[H]\begin{center}\includegraphics[width=16cm]{inc/var_zuker.pdf}\end{center}
+\caption{Zuker folding running time (seconds). Quartiles: 0.152, 0.183 (median), 0.214}\label{fig:var_zuker}\end{figure}
+
+Using the QQplot\footnote{Quantile-to-quantile plot, used to compare two distributions against each other.}, the distribution is heavily tailed (has more results towards the ends of the range) than a Gaussian distribution (fig.~\ref{fig:var_zuker} center) but fits better an uniform distribution (fig.~\ref{fig:var_zuker} right).
+If we run multiple time the program over the same input, we obtain the same behavior as with the matrix multiplication (strictly identical time); hence we can conclude that Zuker is an input sensitive problem whereas matrix chain multiplication is not.
+It follows that we need to be careful to test with exactly identical input set different implementations.
 
@@
 This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently.
-However, simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue.
+However, this metric heavily depends on the problem and simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue, hence we do not consider this metric hereafter (except as an upper bound on the dimension of the input).
 
 \subsection{Benchmarking platform}
-Our benchmarking platform is a notebook with a Core i7-3720QM, 16Gb or main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8.
-Tests are made when the discrete GPU is not involved in rendering the user interface\footnote{This involves a workaround to keep the discrete GPU powered, while the CPU is involved in rendering graphics; source code available on demand.}.
-Due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.
+Our benchmarking platform is an Apple notebook with a Core i7-3720QM with 16Gb of main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8 and Oracle JDK 1.7.0-10.
+A workaround (see listing~\ref{cpu_workaround}) allows us to use the CPU to render the user interface while leaving the graphic card available to execute CUDA kernels.
+Unfortunately, due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.
 
+% ----------------------------------------------
+% MatrixMult-512, Mac+JDK7\ul
+% Original: 24.937 sec / 24.658 sec
+% Optimized concatenation: 19.329 sec / 18.924 sec
+% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
+\newpage
 % ------------------------------------------------------------------------------------------------
@@
 % ------------------------------------------------------------------------------------------------
+\subsection{Matrix chain multiplication}
+
+\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
+XXX
+
+\begin{table}[H]\begin{center}\begin{tabular}{rllll}\toprule
+\bf Size	& \bf DynaProg & \bf Optimized C & \bf GAPC & \bf ADP Fusion \\ \midrule
+128		&		&		&		& -		\\
+256		&		&		&		& -		\\
+512		&		&		&		& -		\\
+1024	&		&		&		& -		\\
+2048	&		&		&		& -		\\
+4096	&		&		&		& -		\\
+8192	&		&		&		& -		\\ \bottomrule
+\end{tabular}
+\end{center}
+\caption{Matrix chain multiplication results (CPU)}
+\end{table}
 
 {\color{red}\ol
-\item Make LibRNA work with CUDA generated parsers
-\item Finish the LibRNA section, write some conclusion ideas
+\item Write some conclusion ideas
 \item Run benchmark and add results in report (automate, generate raw Matlab/TeX output)
@@
 \ole}
-
-\subsection{Matrix chain multiplication}
-\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
-XXX
 
@@
 
+XXX: future work: infer automatically "always non-empty" property
+
 \subsection{Scala parsers}
-{\color{red}
-MatrixMult-512, Mac+JDK7\ul
-\item Original: 24.937 sec / 24.658 sec
-\item Optimized concatenation: 19.329 sec / 18.924 sec
-\item Tabulation as arrays+inline: 14.976 sec / 14.849 sec
-\ule}
 

### 2013-01-19 docs/report/conclusion.tex
Bug with bounded sequences identified !
More report.

@@
 \begin{verbatim}
+Recap of the report:
+- Goal: generalize solving dynamic programming on sequences for GPUs
+- classification of the problems; in terms of recurrences = complicated, grammar = easier
+- general technique to describe backtracking information
+- implemented in Scala+CUDA
+- additional analysis to generate efficient code (1 result, memory alignment, scheduling)
 
-recap of the report
+Recap of the results:
+- XXX
 
-recap of the results
-
-big picture: with actual compiler technology we
-Embedded DSL vs external DSL
-Larger picture how to go further
+Big picture (Larger picture how to go further):
+- Code generation and actual compiler technology
+- Embedded DSL vs external DSL
+- Only a subset of the dynamic programming problems, easy to parallelize (similar memory accesses and scheduling pattern/progress)
+  - others are iterative or sparse, different scheduling => multiple scheduling patterns to be discovered and implemented efficiently, possibly in LMS (if time permits)
 

### 2013-01-19 docs/report/implementation.tex
More report.

@@
 Leveraging the insights provided by \cite{gpu_atlp} and \cite{gpu_barrier}, we started with a basic implementation (where each CUDA thread processes one matrix line) with three additional optimizations:\ul
-\item Memory accesses must be coalesced (memory accesses account for a significant part of the total running time, according to both manufacturer documentation and experiments \cite{perfeval})
+\item Memory accesses must be coalesced (memory accesses account for a significant part of the total running time, according to both manufacturer documentation and experiments \cite{perfeval_gpu})
 \item Synchronization between threads can be done according to \cite{gpu_barrier}, additionally, we can slightly loosen the synchronization restrictions, as the paper describes a thread barrier whereas we only require a condition on previous thread progress (except for the parallelogram case, where we still require a barrier).
@@
 \subsubsection{Results discussion}\ul
-\item \textbf{User interface:} It has been put in evidence in \cite{perfeval} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M).
+\item \textbf{User interface:} It has been put in evidence in \cite{perfeval_gpu} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M).
 With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best.

### 2013-01-19 docs/report/inc/gpu_power.m

@@ -1,0 +1,87 @@
+#import <Foundation/Foundation.h>
+#import <IOKit/IOKitLib.h>
+
+bool gpuOpen();  // Initialize driver
+void gpuClose(); // Close driver
+
+// User client method dispatch selectors.
+enum { kOpen, kClose, kmuxSet, kmuxGet };
+
+typedef enum {
+  muxFeatureInfo    = 0, // get: uint64_t with bits set as (1<<muxFeature)
+  muxForceSwitch    = 2, // set: force graphics switching
+  muxPowerGPU       = 3, // set: power down a gpu
+                         // get: graphics cards?, 0x8=Intel, 0x88=Nvidia
+  muxGpuSelect      = 4, // set/get: dynamic switching on=2/off=0
+  muxSwitchPolicy   = 5, // set: 0=immediate, 2=requires logout to switch
+  muxGraphicsCard   = 7, // get: returns active graphics card
+} muxState;
+
+typedef enum { Policy, Auto_PowerDown_GPU, Dynamic_Switching,
+  GPU_Powerpolling, // Inverted: 1=off, 0=on
+  Defer_Policy,
+  Synchronous_Launch,
+  Backlight_Control=8,
+  Recovery_Timeouts,
+  Power_Switch_Debounce,
+  Logging=16,
+} muxFeature;
+
+static io_connect_t conn = IO_OBJECT_NULL;
+
+#define muxCall(STATE,IN,IN_N,OUT,OUT_N) if (IOConnectCallScalarMethod(conn,STATE,IN,IN_N,OUT,OUT_N)!=KERN_SUCCESS) { perror("Mux error"); gpuClose(); exit(EXIT_FAILURE); }
+static uint64_t muxGet(muxState state) { uint32_t count=1; uint64_t out,in[2]={1, (uint64_t)state}; muxCall(kmuxGet, in, 2, &out, &count); return out; }
+static void muxSet(muxState state, uint64_t arg) { uint64_t in[3] = {1, (uint64_t) state, arg }; muxCall(kmuxSet, in, 3, NULL, NULL); }
+
+#define setFeature(feature,enabled) muxSet(enabled,1<<(feature))
+#define setDynamic(enabled) muxSet(muxGpuSelect,enabled)
+#define setSwitchPolicy(immediate) muxSet(muxSwitchPolicy,immediate?0:2)
+
+bool gpuOpen() {
+  kern_return_t res;
+  io_iterator_t iterator = IO_OBJECT_NULL;
[…]

### 2013-01-20 docs/report/conclusion.tex
Nits pre-rework.

@@
 Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
-\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform by George Nithin, a PhD student of the LAP (Laboratory of processors architecture), the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
-The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the memory can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
+\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform, the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
+The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the data can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
 \item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.

### 2013-01-20 src/main/scala/v4/ADPParsers.scala

@@ -32,31 +32,36 @@ trait ADPParsers extends BaseParsers { this:Signature =>
 
   // Terminal parsers
   val empty = new Terminal[Unit](0,0,(i:Var,j:Var) => (Nil,"")) {
-    def apply(sw:Subword) = sw match { case (i,j) => if (i==j) List(({},bt0)) else Nil }
+    def apply(sw:Subword) = { val (i,j)=sw; if (i==j) List(({},bt0)) else Nil }
   }
   val emptyi = new Terminal[Int](0,0,(i:Var,j:Var) => (Nil,"("+i+")")) {
-    def apply(sw:Subword) = sw match { case (i,j) => if (i==j) List((i,bt0)) else Nil }
+    def apply(sw:Subword) = { val (i,j)=sw; if (i==j) List((i,bt0)) else Nil }
   }
   val el = new Terminal[Alphabet](1,1,(i:Var,j:Var) => (Nil,"_in1["+i+"]")) {
-    def apply(sw:Subword) = sw match { case (i,j) => if(i+1==j) List((in(i),bt0)) else Nil }
+    def apply(sw:Subword) = { val (i,j)=sw; if(i+1==j) List((in(i),bt0)) else Nil }
   }
   val eli = new Terminal[Int](1,1,(i:Var,j:Var) => (Nil,"("+i+")")) {
-    def apply(sw:Subword) = sw match { case (i,j) => if(i+1==j) List((i,bt0)) else Nil }
+    def apply(sw:Subword) = { val (i,j)=sw; if(i+1==j) List((i,bt0)) else Nil }
   }
   def seq(min:Int=1, max:Int=maxN) = new Terminal[Subword](min,max,(i:Var,j:Var) => (Nil,"(T2ii){"+i+","+j+"}")) {
-    def apply(sw:Subword) = sw match { case (i,j) => if (i+min<=j && (max==maxN || i+max>=j)) List(((i,j),bt0)) else Nil }
+    def apply(sw:Subword) = { val (i,j)=sw; if (i+min<=j && (max==maxN || i+max>=j)) List(((i,j),bt0)) else Nil }
   }
 }
 
 trait LexicalParsers extends ADPParsers { this:Signature =>
   override type Alphabet = Char
-  def char = el
-  def chari = eli
-  def charf(f: Char => Boolean) = el filter { // TODO: CUDA version
-    case(i,j) if(i+1==j) => f(in(i))
-    case _ => false
-  }
[…]

### 2013-01-20 src/main/scala/v4/examples/ElMamun.scala

@@ -23,8 +23,8 @@ trait SellerAlgebra extends Bill {
 trait BillGrammar extends LexicalParsers with Bill {
   override type Alphabet = Char
 
-  def plus = charf(_ == '+')
-  def times = charf(_ == '*')
+  def plus = charf('+')
+  def times = charf('*')
 
   val billGrammar:Tabulate = tabulate("M",(
     digit

### 2013-01-20 src/main/scala/v4/examples/MatrixMultGen.scala

@@ -44,6 +44,7 @@ object MatrixMultGen extends MatrixGrammar with MatrixAlgebraGen with CodeGen wi
   // ------- Extra codegen initialization
   //println(gen)
 
+  /*
   println("------ SCALA -------------------")
   val (res1,bt1) = backtrack(input,true).head
   println("--> "+res1)
@@ -52,4 +53,15 @@ object MatrixMultGen extends MatrixGrammar with MatrixAlgebraGen with CodeGen wi
   val (res2,bt2) = backtrack(input).head
   println("--> "+res2)
   println("--> "+build(input,bt2))
+  */
+
+  import scala.util.Random
+  Random.setSeed(123456748299L)
+  def rnd:Int = Math.abs(Random.nextInt)%253+1
+
+  def genMats(n:Int) = {
+    val s = Seq.fill(n)(rnd)
+    (Seq(rnd) ++ s).zip(s).map{case(x,y)=>(x,y)/*unbox ints!!*/}.toArray
+  }
+  for (k<-0 until 100) backtrack(genMats(512))
 }

### 2013-01-20 src/main/scala/v4/report/SeqAlign.scala

@@ -50,10 +50,10 @@ trait SeqPrettyPrint extends SeqAlignSignature {
 
 trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
   val axiom:Tabulate = tabulate("M",(
-    empty                     ^^ start
-  | seq1() -~ axiom           ^^ gap1
-  |           axiom ~- seq2() ^^ gap2
-  | el1    -~ axiom ~- el2    ^^ pair
+    empty                   ^^ start
+  | seq() -~ axiom          ^^ gap1
+  |          axiom ~- seq() ^^ gap2
+  | el1   -~ axiom ~- el2   ^^ pair
   ) aggregate h)
 }
 

### 2013-01-20 docs/report/benchmarks.tex
Reworked SWat hand-optimized.

@@
 
-\subsection{Metrics}
+\subsection{Metrics} \label{metrics}
 The main metrics of interest is the running time.
@@
 As example, we can consider two extreme cases: matrix chain multiplication and Zuker RNA folding, with a test environment of 100 random inputs (of length respectively 512 and 80) and a GPU warmup of 10 computations.
-In this settings, matrix chain multiplication computations are executed in a perfectly constant time (0.127 seconds), which mean that we sufficiently reduced the noise.
+In this settings, matrix chain multiplication computations are executed in a perfectly constant time\footnote{With respect to truncation and measurment accurcy, has less than 1\% of variation (not observable).} (0.127 seconds), which mean that we sufficiently reduced the noise.
 Oppositely, the Zuker RNA folding running times appear much more scattered as presented below:
@@
 \subsection{Matrix chain multiplication}
+Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
 

### 2013-01-20 docs/report/benchmarks.tex
Report nits.

--- 2013-01-19 docs/report/benchmarks.tex
@@
 
-\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
-XXX
+For this problem, we needed to modify the hand optimized version because the memory access was slightly different from the DynaProg version (the original hand-optimized version was storing only the matrix multiplication score whereas the DynaProg version also stores the matrix dimension in intermediate results).
+Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled.
+Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing slightly the performance of CUDA operations.
+% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
+% GPU: GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?
 
-\begin{table}[H]\begin{center}\begin{tabular}{rllll}\toprule
-\bf Size	& \bf DynaProg & \bf Optimized C & \bf GAPC & \bf ADP Fusion \\ \midrule
-128		&		&		&		& -		\\
-256		&		&		&		& -		\\
-512		&		&		&		& -		\\
-1024	&		&		&		& -		\\
-2048	&		&		&		& -		\\
-4096	&		&		&		& -		\\
-8192	&		&		&		& -		\\ \bottomrule
-\end{tabular}
-\end{center}
-\caption{Matrix chain multiplication results (CPU)}
-\end{table}
+\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
+& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
+& \hdr{DynaProg}{Scala version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{C, single thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{ADP Fusion}{\cite{adp_fusion}}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
+& \hdr{DynaProg}{CUDA version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{CUDA, multi-thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+\\ \bottomrule\end{tabular}}\end{center}
+\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
 
@@
 
-\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?}
-XXX
+\subsection{Smith-Watermann}
+% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
+% GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign
+XXX: to compare appropriately either use 3 matrices or constant gap penalty
+XXX: move toward use of 3 matrices
 
-\subsection{Smith-Watermann}
-\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
-XXX
-
-\subsubsection{GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign}
-XXX
+\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
+& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
+& \hdr{DynaProg}{Scala version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{C, single thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{ADP Fusion}{\cite{adp_fusion}}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
+& \hdr{DynaProg}{CUDA version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{CUDA, multi-thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+\\ \bottomrule\end{tabular}}\end{center}
+\caption{Running time of Smith-Waterman (in seconds)}\end{table}
 
 \subsection{Zuker RNA folding}
-\subsubsection{CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA}
-XXX
-
-\subsubsection{GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?}
+% CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA
+% GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
 XXX

### 2013-01-20 docs/report/benchmarks.tex
Nits.

@@
 Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
+For this problem, we modified the hand optimized version because the memory access was slightly different from the generated version (the original hand-optimized version was storing only the matrix multiplication score whereas the generated version also stores the matrix dimension in intermediate results, which proved a better pattern).
 
-For this problem, we needed to modify the hand optimized version because the memory access was slightly different from the DynaProg version (the original hand-optimized version was storing only the matrix multiplication score whereas the DynaProg version also stores the matrix dimension in intermediate results).
 Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled.
-Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing slightly the performance of CUDA operations.
+Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing the performance of CUDA operations.
 % CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
@@
 	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{Optimized}{CUDA, multi-thread}
+& \hdr{Optimized 64 bit}{CUDA, multi-thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized 32 bit}{CUDA, multi-thread}
 	& 			& 			& 			& 			& 			& 			& 		 \\
@@
 \caption{Running time of matrix chain multiplication (in seconds)}\end{table}
+
+% 1024=0.316, 2048=1.427, 4096=8.841, 8192=62.064
 

### 2013-01-20 docs/report/benchmarks.tex
Nits.

@@
 Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
+% XXX: is this true ??
 For this problem, we modified the hand optimized version because the memory access was slightly different from the generated version (the original hand-optimized version was storing only the matrix multiplication score whereas the generated version also stores the matrix dimension in intermediate results, which proved a better pattern).
-
 Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled.
@@
 & \hdr{Optimized}{C, single thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.001		& 0.008		& 0.078		& 1.179		& 19.811		& 206.556	& 2010.485 \\
 & \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
@@
 & \hdr{ADP Fusion}{\cite{adp_fusion}}
-	& 			& 			& 			& 			& 			& 			& 		 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
+	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad\quad$}}
 & \hdr{DynaProg}{CUDA version}
 	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{Optimized 64 bit}{CUDA, multi-thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{Optimized 32 bit}{CUDA, multi-thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{CUDA, 64-bit}
+	& 0.011		& 0.020		& 0.074		& 0.315		& 1.651		& 10.354		& 72.376 \\
 & \hdr{ATLP}{\cite{gpu_atlp}, rescaled results}
@@
 	& 			& 			& 			& 			& 			& 			& 		 \\
-\\ \bottomrule\end{tabular}}\end{center}
+\\[-10pt] \bottomrule\end{tabular}}\end{center}
 \caption{Running time of matrix chain multiplication (in seconds)}\end{table}
@@
 & \hdr{Optimized}{C, single thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& <0.001		& <0.001		& 0.002		& 0.009		& 0.042		& 0.171		& 0.706 \\
 & \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
@@
 & \hdr{ADP Fusion}{\cite{adp_fusion}}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
 \midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
@@
 	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{Optimized 64 bit}{CUDA, multi-thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{Optimized 32 bit}{CUDA, multi-thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{Optimized}{CUDA, 64-bit}
+	& 0.002		& 0.004		& 0.004		& 0.007		& 0.024		& 0.070		& 0.270 \\
 & \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
@@
 	& 			& 			& 			& 			& 			& 			& 		 \\
-\\ \bottomrule\end{tabular}}\end{center}
+\\[-10pt] \bottomrule\end{tabular}}\end{center}
 \caption{Running time of Smith-Waterman (in seconds)}\end{table}

### 2013-01-21 docs/report/benchmarks.tex
Bug int SWat to be fixed, see SWatAffine.

@@
+\documentclass[11pt]{article}
+\input{inc/style.sty}
+\begin{document}
+
 \section{Benchmarks} \label{benchmarks}
@@
 Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
-Finally, we asked lately the authors of \cite{gpu_rnafold} who did not respond either to our sollicitations.
+Finally, we asked lately the authors of \cite{gpu_rnafold} who did not respond either to our solicitations.
 
@@
 Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
-% XXX: is this true ??
-For this problem, we modified the hand optimized version because the memory access was slightly different from the generated version (the original hand-optimized version was storing only the matrix multiplication score whereas the generated version also stores the matrix dimension in intermediate results, which proved a better pattern).
 Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled.
@@
 & \hdr{DynaProg}{Scala version}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.247		& 2.048		& 16.657		& 134.207	& 1101.134	&  			& 		 \\
 & \hdr{Optimized}{C, single thread}
@@
 & \hdr{DynaProg}{CUDA version}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.028		& 0.046		& 0.129		& 0.348		& 1.685		& 10.308		& 71.215 \\
 & \hdr{Optimized}{CUDA, 64-bit}
@@
 \caption{Running time of matrix chain multiplication (in seconds)}\end{table}
+
+The running time of DynaProg/CUDA includes the overhead of back and forth JNI conversion (scales linearly between 0.018 and 0.057 seconds) but does not include the overhead due to the code generation which decomposes in 0.068 seconds for analysis and code synthesis (once per algebra/grammar pair) and 0.086 + 1.753 seconds for respectively Scala and CUDA compilation (constant time, once per problem dimension).
+
+For DynaProg/Scala we use a variant of the problem description: the original version only stores the matrix multiplication score whereas the modified version also stores the matrix dimension.
+This allows a speedup of $2.9\times$ probably due to additional function calls to retrieve the values.
+Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit.
 
@@
 Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.}
+
+\end{document}

### 2013-01-21 docs/report/conclusion.tex
Nits.

@@
 
+This Master project is focused on how to solve efficiently dynamic programming.
+By restricting to the class of problem involving sequences as input, we were able to extract generic patterns and expose them to parallel architectures like CUDA.
+To do that, we first depicted the dynamic programming landscape and defined a set of problems we would like to solve (\S\ref{problems}).
+Then we discussed the requested functionalities to provide the user with a convenient embedded DSL that is based on the ADP formalization (\S\ref{architecture}).
+These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtain a single optimal result.
+We then provide some technical details explaining how these ideas are put in practice in \textit{DynaProg} (\S\ref{implementation}).
+Finally, we compare our implementation with existing implementations (\S\ref{benchmarks}).
+
+{\color{red} What's the conclusion on experimental results?}
+
+% http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html
+
 \begin{verbatim}

### 2013-01-21 docs/report/benchmarks.tex

@@
 \newpage
-% ------------------------------------------------------------------------------------------------
-{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
-% ------------------------------------------------------------------------------------------------
 \subsection{Matrix chain multiplication}
@@
 	& 0.011		& 0.020		& 0.074		& 0.315		& 1.651		& 10.354		& 72.376 \\
-& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results}
-	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+% 32-bit &&		&			&			& 0.316		& 1.427		& 8.841		& 62.064
+& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}
+	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& - \\
 \\[-10pt] \bottomrule\end{tabular}}\end{center}
 \caption{Running time of matrix chain multiplication (in seconds)}\end{table}
+
+$^{(1)}$ Assuming that 72\% of the running time is due to memory accesses, and considering a $3.55\times$ memory throughput slowdown (see \S\ref{results_discussion}).
+% 0.06 & 0.07 & 0.08 & 0.14 & 0.47 & 2.57 => f[x_] := (x*(1 - .72)) + (x*0.72)*(102.4/28.8)
+% 0.1704 & 0.1988 & 0.2272 & .3976 & 1.3348 & 7.2988
 
@@
 This allows a speedup of $2.9\times$ probably due to additional function calls to retrieve the values.
-Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit.
+Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit	.
 
-% 1024=0.316, 2048=1.427, 4096=8.841, 8192=62.064
+% ------------------------------------------------------------------------------------------------
+{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+% ------------------------------------------------------------------------------------------------
 
@@
 & \hdr{DynaProg}{Scala version}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.115		& 0.438		& 1.743		& 7.864		& 32.788		& 139.309	& $^{(1)}\infty$ \\
 & \hdr{Optimized}{C, single thread}
@@
 & \hdr{DynaProg}{CUDA version}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.031		& 0.037		& 0.047		& 0.068		& 0.116		& 0.202		& $^{(2)}$3.226 \\
 & \hdr{Optimized}{CUDA, 64-bit}
@@
 	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{GAPC}{\cite{gapc_thesis}, OpenCL}
-	& 			& 			& 			& 			& 			& 			& 		 \\
 \\[-10pt] \bottomrule\end{tabular}}\end{center}
 \caption{Running time of Smith-Waterman (in seconds)}\end{table}
+
+$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx8G -Xms8G -XX:MaxPermSize=8G}) this emphasizes that the top-down parsing approach might not be the best for such simple problems.
+
+$^{(2)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance ($4\times$ slowdown).
+This extra memory requirement is due to the use of 3 matrices to avoid non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
+
+X: java.lang.OutOfMemoryError: GC overhead limit exceeded
 

### 2013-01-21 docs/report/implementation.tex

@@
 % ----------------------------------------------
-\subsubsection{Results discussion}\ul
+\subsubsection{Results discussion} \label{results_discussion} \ul
 \item \textbf{User interface:} It has been put in evidence in \cite{perfeval_gpu} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M).
@@
 183'352'614'912 bytes which corresponds to 183.35GB.
-In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254 \rm GB$.
+In 8.841 seconds, we can transfer theoretically at most $8.841\cdot 28.8 = 254.6 \rm GB$.
 Hence  72\% of the algorithm running time is spent into memory accesses.
@@
 \item On a 4096 matrix, if we assume that the \cite{gpu_atlp} card would have the same bandwidth as our card, their running time would be
-\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 9.43\rm s_{\text{ ATLP}} > 8.84\rm s_{\text{ our}}\]
+\[2.57 \cdot (1-.72) + 2.57 \cdot 0.72 \cdot \tfrac{102.4_{GB/s}}{28.8_{GB/s}} = 7.30\rm s_{\text{ ATLP}} < 8.84\rm s_{\text{ our}}\]
 This shows that our algorithm is comparable to theirs.
@@
 \ule
-As a conclusion, (1) it seems that the technique used in \cite{gpu_atlp} only applies to legacy hardware, however this remains a supposition (as we can not compare) and (2) we are on par, if not better with one of the best current implementations.
+As a conclusion, (1) it seems that the technique used in \cite{gpu_atlp} brings more performance improvement with legacy hardware, however this remains a supposition (as we can not compare) and (2) we are slightly worse than one of the best current implementations.
 

### 2013-01-21 docs/report/benchmarks.tex
Nits, small improvement for Scala.

@@
-\documentclass[11pt]{article}
-\input{inc/style.sty}
-\begin{document}
+%\documentclass[11pt]{article}\input{inc/style.sty}\begin{document}
 
@@
 & \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.012		& 0.050		& 0.348		& 2.822		& 25.163		& 224.704	& 		 \\
 & \hdr{ADP Fusion}{\cite{adp_fusion}}
@@
 & \hdr{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}
-	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& - \\
+	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& --- \\
 \\[-10pt] \bottomrule\end{tabular}}\end{center}
@@
 % GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign
-XXX: to compare appropriately either use 3 matrices or constant gap penalty
-XXX: move toward use of 3 matrices
+To maintain a fair comparison with other implementation, we use affine gap penalty variant of the Smith-Waterman algorithm.
+For that, we need to encode it with 3 tabulations in our grammar, whereas domain-specific application can leverage the knowledge that two of these matrices can be removed by storing appropriate information in the computation wavefront (\S\ref{calc_simplifications}).
+%XXX: to compare appropriately either use 3 matrices or constant gap penalty
 
@@
 & \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.011		& 0.011		& 0.011		& 0.011		& 0.011		& 0.011		& 0.012 \\
 \\[-10pt] \bottomrule\end{tabular}}\end{center}
@@
 This extra memory requirement is due to the use of 3 matrices to avoid non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
-
-X: java.lang.OutOfMemoryError: GC overhead limit exceeded
 
@@
 }
-\begin{table}[H]\begin{center}\begin{tabular}{lrrrrrr} \toprule
-Matrix size			& 1024	& 2048	& 4096	& 8192 \\
-Splits				& 1		& 8		& 64		& 512 \\ \midrule
-Analysis				& 0.007	& 0.007	& 0.007	& 0.007 \\
-Code generation		& 0.071	& 0.072	& 0.073	& 0.071 \\
-C/CUDA compilation	& 3.025	& 1.836	& 1.832	& 1.841 \\
-Scala compilation		& 1.837	& 1.853	& 1.830	& 1.767 \\ \midrule
-- JNI read				& 0.027	& 0.028	& 0.029	& 0.058 \\
-- CUDA compute		& 0.868	& 3.378	& 19.782	& 139.697 \\
-- CUDA backtrack		& 0.009	& 0.012	& 0.023	& 0.045 \\
-- JNI output			& 0.005	& 0.009	& 0.018	& 0.032 \\
-Total execution			& 0.915	& 3.432	& 19.860	& 139.846 \\ \bottomrule
-\end{tabular}\end{center}\caption{Preliminary results for MatrixMultGen}\end{table}
 
@@
 
-\end{document}
+%\end{document}

### 2013-01-21 src/main/scala/v4/BaseParsers.scala

@@ -158,9 +158,10 @@ trait BaseParsers { this:Signature =>
     rules += ((name,this))
 
     var id:Int = -1 // subrules base index
-    @inline private def idx(sw:Subword):Int = if (twotracks) sw._1*mW+sw._2 else { val (i,j)=sw; val d=mH+1+i-j; ( mH*(mH+1) - d*(d-1) ) /2 + i }
+    @inline private def idx(sw:Subword):Int = if (twotracks) sw._1*mW+sw._2 else { val d=mH+1+sw._1-sw._2; ( mH*(mH+1) - d*(d-1) ) /2 + sw._1 }
     private def get(sw:Subword) = { val i=idx(sw); val v1=data(i); if (v1!=null) v1 else { val v=inner(sw).map{case(c,(r,b))=>(c,(id+r,b))}; data(i)=v; v } }
     private def put(sw:Subword,v:List[(Answer,Backtrack)]) { data(idx(sw))=v; }
+    
     def apply(sw: Subword) = get(sw) map {x=>(x._1,bt0)}
     def unapply(sw:Subword,bt:Backtrack) = get(sw) map { case (c,b) => (c,List((sw,c,b))) }
     def reapply(sw:Subword,bt:Backtrack) = { val v=data(idx(sw)); if (v!=null) v.head._1 else sys.error("Failed reapply"+sw) }
@@ -267,11 +268,23 @@ trait BaseParsers { this:Signature =>
       if (track==0) {
         val min_k = if (rU==maxN || i+lL > j-rU) i+lL else j-rU
         val max_k = if (lU==maxN || j-rL < i+lU) j-rL else i+lU
-        for(
-          k <- (min_k to max_k).toList;
-          (x,xb) <- left((i,k));
-          (y,yb) <- right((k,j))
-        ) yield(((x,y),bt(xb,yb,k)))
+        var k=min_k;
+        //for(
+        //  k <- (min_k to max_k).toList;
+        //  (x,xb) <- left((i,k));
+        //  (y,yb) <- right((k,j))
+        //) yield(((x,y),bt(xb,yb,k)))
+        //
+        // Optimization: 16.675 -> 11.370 (mm1_512)
+        var l=List[((T,U),Backtrack)]()
+        while (k<=max_k) { var ll = left((i,k)); val r = right((k,j))
+          if (r.size>0) while (ll.size>0) { var rr = r
+            while (rr.size>0) { val lh=ll.head; val rh=rr.head
[…]

### 2013-01-22 docs/report/benchmarks.tex
Improved slightly scala perf for concat, added bottom-up parsers.

@@
 & \hdr{DynaProg}{Scala version}
-	& 0.247		& 2.048		& 16.657		& 134.207	& 1101.134	&  			& 		 \\
+	& 0.167		& 1.377		& 10.760		& 86.966		& 712.688	&  			& 		 \\
 & \hdr{Optimized}{C, single thread}
@@
 & \hdr{DynaProg}{Scala version}
-	& 0.115		& 0.438		& 1.743		& 7.864		& 32.788		& 139.309	& $^{(1)}\infty$ \\
+	& 0.098		& 0.364		& 1.421		& 6.527		& 26.798		& 114.377	& $^{(1)}\infty$ \\
 & \hdr{Optimized}{C, single thread}
@@
 
-$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx8G -Xms8G -XX:MaxPermSize=8G}) this emphasizes that the top-down parsing approach might not be the best for such simple problems.
+$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx8G -Xms8G -XX:MaxPermSize=8G}), independently of whether top-down or bottom-up parsing approaches are taken.
+After some point, most of the time is spent in the garbage collection.
+The aggregation function contributes to approximately 30\% of the total running time.
+From these result, we see that the use of complex parsing techniques are less scalable than ad-hoc methods for simple problems.
 

### 2013-01-23 docs/report/benchmarks.tex
Nits.

@@
 % ----------------------------------------------
-% MatrixMult-512, Mac+JDK7\ul
-% Original: 24.937 sec / 24.658 sec
-% Optimized concatenation: 19.329 sec / 18.924 sec
-% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
+\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
+% ----------------------------------------------
+% MatrixMult-512, Mac+JDK7
+% -> Original: 24.658 sec
+% -> Optimized concatenation: 18.924 sec
+% -> Tabulation as arrays+inline: 14.849 sec
+% -> Re-optimized concatenation: 10.76 sec
 \newpage
@@
 
-\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
 \begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
 & \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
 & \hdr{DynaProg}{Scala version}
@@
 & \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
-	& 0.012		& 0.050		& 0.348		& 2.822		& 25.163		& 224.704	& 		 \\
-& \hdr{ADP Fusion}{\cite{adp_fusion}}
-	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad\quad$}}
+	& 0.012		& 0.050		& 0.348		& 2.822		& 25.163		& 224.704	& 		 \\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
 & \hdr{DynaProg}{CUDA version}
@@
 	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& --- \\
-\\[-10pt] \bottomrule\end{tabular}}\end{center}
-\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
+\bottomrule\end{tabular}}\end{center}\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
 
@@
 
-\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
 \begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
@@
 & \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
+	& 0.006		& 0.010		& 0.025		& 0.097		& 0.394		& 1.634		& 11.245 \\
 & \hdr{ADP Fusion}{\cite{adp_fusion}}
@@
 	& 0.002		& 0.004		& 0.004		& 0.007		& 0.024		& 0.070		& 0.270 \\
+& \hdr{ADP-C}{\cite{adp_gpu}}
+	& 			& 			& 			& 			& 			& 			& 		\\
 & \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
@@
 % GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
-XXX
+
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
+& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
+& \hdr{DynaProg}{Scala version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{ViennaRNA}{\cite{vienna_rna}}
+	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hdr{DynaProg}{CUDA version}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+& \hdr{RNAFold}{\cite{gpu_rnafold}}
+	& 			& 			& 			& 			& 			& 			& 		 \\
+\\[-10pt] \bottomrule\end{tabular}}\end{center}
+\caption{Running time of RNA folding (in seconds)}\end{table}
+
+% 0.041 & 0.053 & 0.124 & 0.115 & 0.150 & 0.180 & 0.698 & 1.641 & 3.055 & 3.519 & 9.077 & 19.594 & 67.316 & 163.139
 

### 2013-01-23 docs/report/implementation.tex
Nits.

@@
 The {\tt benchmark} flag can be set to enable timing measurements.
+Also it is possible to use bottom-up parsers with Scala to reduce the stack size by enabling the {\tt bottomUp} flag.
+If special concatenations are needed, it is possible to replace the $\sim$ concatenation by $\sim\.(l_{\rm min},l_{\rm max},r_{\rm min},r_{\rm max})\.\sim$ where $l,r$ design respectively the yield size of left and right operands.
 

### 2013-01-23 src/main/c/include/small_dbg.h

@@ -136,7 +136,7 @@ void dbg_init(bool info=false) {
 	#ifdef __CUDACC__
 	if (info) cuInfo();
 	#endif
-	fprintf(stderr,SH_SHAPE " %ldx%ld with block %ldx%ld\n",M_H,M_W,B_H,B_H);
+	fprintf(stderr,SH_SHAPE " %ldx%ld with block %ldx%ld and %d splits\n",M_H,M_W,B_H,B_H,SPLITS);
 	TI* in0=p_input();
 	TI* in1=NULL;
 	#ifdef SH_RECT

### 2013-01-23 docs/report/benchmarks.tex
Nits in report.

--- 2013-01-21 docs/report/benchmarks.tex
@@
 % ----------------------------------------------
-% MatrixMult-512, Mac+JDK7\ul
-% Original: 24.937 sec / 24.658 sec
-% Optimized concatenation: 19.329 sec / 18.924 sec
-% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
+\def\hh{\normalsize\bf}
+\def\hl#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
+\def\hdps{\hl{DynaProg}{Scala parsers}}
+\def\hdpc{\hl{DynaProg}{CUDA parsers}}
+\def\hhoc{\hl{Optimized}{C, single thread}}
+\def\hhog{\hl{Optimized}{CUDA, 64-bit}}
+\def\hgapc{\hl{GAPC}{\cite{gapc_thesis}, C, single thread}}
+\def\hatlp{\hl{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}}
+\def\hcua{\hl{CUDAlign}{\cite{swat_linear}, version 2.0}}
+
+\def\hvien{\hl{ViennaRNA}{\cite{vienna_rna}}}
+\def\hrna{\hl{RNAFold}{\cite{gpu_rnafold}}}
+
+% ----------------------------------------------
+% MatrixMult-512, Mac+JDK7
+% -> Original: 24.658 sec
+% -> Optimized concatenation: 18.924 sec
+% -> Tabulation as arrays+inline: 14.849 sec
+% -> Re-optimized concatenation: 10.76 sec
 \newpage
@@
 Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
-Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is because 64-bit mode is enabled.
-Because external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby reducing the performance of CUDA operations.
+Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is caused by enabling the 64-bit mode.
+Since external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby slightly reducing the performance of CUDA operations.
 % CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
@@
 
-\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
-\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
-& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
-& \hdr{DynaProg}{Scala version}
-	& 0.247		& 2.048		& 16.657		& 134.207	& 1101.134	&  			& 		 \\
-& \hdr{Optimized}{C, single thread}
-	& 0.001		& 0.008		& 0.078		& 1.179		& 19.811		& 206.556	& 2010.485 \\
-& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
-	& 0.012		& 0.050		& 0.348		& 2.822		& 25.163		& 224.704	& 		 \\
-& \hdr{ADP Fusion}{\cite{adp_fusion}}
-	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad\quad$}}
-& \hdr{DynaProg}{CUDA version}
-	& 0.028		& 0.046		& 0.129		& 0.348		& 1.685		& 10.308		& 71.215 \\
-& \hdr{Optimized}{CUDA, 64-bit}
-	& 0.011		& 0.020		& 0.074		& 0.315		& 1.651		& 10.354		& 72.376 \\
-% 32-bit &&		&			&			& 0.316		& 1.427		& 8.841		& 62.064
-& \hdr{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}
-	& 0.170		& 0.199		& 0.227		& 0.397		& 1.335		& 7.299 		& --- \\
-\\[-10pt] \bottomrule\end{tabular}}\end{center}
-\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 0.05	& 0.20	& 0.80	& 2.03	& 6.65	& 15.10	& 47.40	\\
+& \hhoc	& <0.01	& <0.01	& <0.01	& 0.01	& 0.03	& 0.08	& 0.28	\\
+& \hgapc	& 0.01	& 0.01	& 0.03	& 0.05	& 0.15	& 0.35	& 1.16	\\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hdpc	& 0.03	& 0.04	& 0.05	& 0.07	& 0.13	& 0.13	& 0.21	\\
+& \hhog	& <0.01	& 0.01	& 0.01	& 0.02	& 0.04	& 0.08	& 0.17	\\
+& \hatlp	& 0.17	&	 	&		& 0.20	& 		& 0.23	& 		\\
+\midrule
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 109.77	& 368.21	& 877.30	& 3059.42 & 		& 		& 		\\
+& \hhoc	& 1.18	& 7.06	& 19.81	& 78.90	& 206.56	& 799.53	& 2010.49 \\
+& \hgapc	& 2.82	& 10.02	& 25.16	& 91.69	& 224.70	& 		&  		\\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hdpc	& 0.35	& 0.85	& 1.69	& 4.79	& 10.32	& 31.60	& 71.22	\\
+& \hhog	& 0.32	& 0.82	& 1.65	& 4.74	& 10.35	& 31.94	& 72.38	\\
+& \hatlp	& 0.40	& 0.74	& 1.33	& 3.43	& 7.29	& 		& 		\\
+\bottomrule\end{tabular}}\end{center}\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
 
-$^{(1)}$ Assuming that 72\% of the running time is due to memory accesses, and considering a $3.55\times$ memory throughput slowdown (see \S\ref{results_discussion}).
+$^{(1)}$ Assuming that 72\% of the running time is due to memory accesses, and considering a $3.55\times$ memory throughput slowdown of the original results (see \S\ref{results_discussion}).
 % 0.06 & 0.07 & 0.08 & 0.14 & 0.47 & 2.57 => f[x_] := (x*(1 - .72)) + (x*0.72)*(102.4/28.8)
@@
 The running time of DynaProg/CUDA includes the overhead of back and forth JNI conversion (scales linearly between 0.018 and 0.057 seconds) but does not include the overhead due to the code generation which decomposes in 0.068 seconds for analysis and code synthesis (once per algebra/grammar pair) and 0.086 + 1.753 seconds for respectively Scala and CUDA compilation (constant time, once per problem dimension).
+The results of the following problems are presented similarly.
 
 For DynaProg/Scala we use a variant of the problem description: the original version only stores the matrix multiplication score whereas the modified version also stores the matrix dimension.
-This allows a speedup of $2.9\times$ probably due to additional function calls to retrieve the values.
-Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit	.
+This allows a speedup of $2.9\times$ probably due to the additional lookups overhead.
+Also with the default JVM parameters, the program cannot address sequences longer than $\sim 420$ elements due to a stack overflow, for these benchmarks, we increased this limit.
+
+\subsection{Smith-Watermann}
+% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
+% GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign
+To maintain a fair comparison with other implementations, our algorithm needs to have the same complexity.
+Hence we use a variant of the Smith-Waterman algorithm: we encode our grammar in 3 tabulations as described in \S\ref{swat_affine}, whereas application like CUDAlign can leverage the domain-specific information that information of two of these matrices can be stored in the computation wavefront (see \S\ref{calc_simplifications}).
+
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 0.04	& 0.13	& 0.27	& 0.48	& 1.07	& 1.92	& 4.33	\\
+& \hhoc	& <0.01	& <0.01	& <0.01	& <0.01	& <0.01	& <0.01	& 0.01	\\
+& \hgapc	& 0.01	& 0.01	& 0.01	& 0.01	& 0.02	& 0.03	& 0.06	\\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hdpc	& 0.03	& 0.03	& 0.03	& 0.04	& 0.05	& 0.05	& 0.06	\\
+& \hhog	& 0.00	& 0.00	& 0.00	& 0.00	& 0.00	& 0.00	& 0.00	\\
+& \hcua	& 0.11	& 0.12	& 0.07	& 0.07	& 0.07	& 0.07	& 0.13	\\
+\midrule
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 7.84	& 18.95	& 33.63	& 70.86	& 		& 		& $^{(1)}\infty$ \\
+& \hhoc	& 0.01	& 0.02	& 0.04	& 0.10	& 0.17	& 0.40	& 0.71	\\
+& \hgapc	& 0.10	& 0.22	& 0.39	& 0.91	& 1.62	& 4.41	& 11.20 	\\[-2pt]
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hdpc	& 0.07	& 0.11	& 0.15	& 0.13	& 0.20	& 0.32	& $^{(2)}$3.21 	\\
+& \hhog	& 0.01	& 0.02	& 0.02	& 0.04	& 0.07	& 0.14	& 0.27	\\
+& \hcua	& 0.13	& 0.15	& 0.14	& 0.14	& 0.15	& 0.17	& 0.20 	\\
+\bottomrule\end{tabular}}\end{center}\caption{Running time of Smith-Waterman (in seconds)}\end{table}
+
+$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx12G -Xms12G}), and independently of whether top-down or bottom-up parsing approaches are taken.
+After a moment, most of the time is spent in (full) garbage collection.
+The aggregation function application contributes to approximately 30\% of the total running time.
+
+$^{(2)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance.
+This extra memory requirement is due to the use of 3 matrices to avoid the non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
+
+\subsection{Zuker RNA folding}
+% CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA
+% GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
+
+\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 0.07	& 0.67	& 2.26	& 4.98	& 17.61	& 39.44	& 130.71	\\
+& \hvien	& 0.01	& 0.01	& 0.02	& 0.03	& 0.07	& 0.12	& 0.29	\\
+& \hgapc	& 0.01	& 0.03	& 0.07	& 0.13	& 0.41	& 0.93	& 2.89	\\[-2pt]
+\midrule \multirow{4}{*}{\vspace{12pt}\rotatebox{90}{\normalsize\bf GPU}}
+& \hdpc	& 0.15	& 0.53	& 0.98	& 1.55	& 3.13	& 5.04	& 9.43	\\
+& \hrna	& 0.09	& 0.14	& 0.20	& 0.44	& 0.80	& 1.89	& 3.52	\\
+\midrule
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
+\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+& \hdps	& 306.12	& 1036.78 &		& 		& 		& 		& 	 	\\
+& \hvien	& 0.57	& 1.53	& 3.19	& 9.37	& 20.18	& 59.65	& 133.65 	\\
+& \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 		& 		&  		\\[-2pt]
+\midrule \multirow{4}{*}{\vspace{12pt}\rotatebox{90}{\normalsize\bf GPU}}
+& \hdpc	& 15.10	& 33.62	& 61.76	& 166.84	& 393.76	& 		&  		\\
+& \hrna	& 9.08	& 19.59	& 67.32	& 163.14	& 		& 		& 		\\
+\bottomrule\end{tabular}}\end{center}\caption{Running time of Zuker RNA folding (in seconds)}\end{table}
+
+{\color{red} XXX: comment, explain energy coefficients, complexity, limit search space for bulges, algorithm variants and constraints RNAfold/Zuker}
 
@@
 \ole}
-
-\subsection{Smith-Watermann}
-% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
-% GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign
-To maintain a fair comparison with other implementation, we use affine gap penalty variant of the Smith-Waterman algorithm.
-For that, we need to encode it with 3 tabulations in our grammar, whereas domain-specific application can leverage the knowledge that two of these matrices can be removed by storing appropriate information in the computation wavefront (\S\ref{calc_simplifications}).
-%XXX: to compare appropriately either use 3 matrices or constant gap penalty
-
-\def\hdr#1#2{\begin{minipage}{3.5cm} {\bf #1} \\[-2pt] \footnotesize #2 \vspace{6pt} \end{minipage}}
-\begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrrr}\toprule
-& \normalsize\bf Matrix dimension & \normalsize\bf 128 & \normalsize\bf 256 & \normalsize\bf 512 & \normalsize\bf 1024 & \normalsize\bf 2048 & \normalsize\bf 4096 & \normalsize\bf 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\qquad$}}
-& \hdr{DynaProg}{Scala version}
-	& 0.115		& 0.438		& 1.743		& 7.864		& 32.788		& 139.309	& $^{(1)}\infty$ \\
-& \hdr{Optimized}{C, single thread}
-	& <0.001		& <0.001		& 0.002		& 0.009		& 0.042		& 0.171		& 0.706 \\
-& \hdr{GAPC}{\cite{gapc_thesis}, C, single thread}
-	& 			& 			& 			& 			& 			& 			& 		 \\
-& \hdr{ADP Fusion}{\cite{adp_fusion}}
-	& 			& 			& 			& 			& 			& 			& 		 \\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\qquad$}}
-& \hdr{DynaProg}{CUDA version}
-	& 0.031		& 0.037		& 0.047		& 0.068		& 0.116		& 0.202		& $^{(2)}$3.226 \\
-& \hdr{Optimized}{CUDA, 64-bit}
-	& 0.002		& 0.004		& 0.004		& 0.007		& 0.024		& 0.070		& 0.270 \\
-& \hdr{CUDAlign}{\cite{swat_linear}, version 2.0}
-	& 0.011		& 0.011		& 0.011		& 0.011		& 0.011		& 0.011		& 0.012 \\
-\\[-10pt] \bottomrule\end{tabular}}\end{center}
-\caption{Running time of Smith-Waterman (in seconds)}\end{table}
-
-$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx8G -Xms8G -XX:MaxPermSize=8G}) this emphasizes that the top-down parsing approach might not be the best for such simple problems.
-
-$^{(2)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance ($4\times$ slowdown).
-This extra memory requirement is due to the use of 3 matrices to avoid non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
-
-\subsection{Zuker RNA folding}
-% CPU: DynaProg(Scala), GAPC, ADPFusion and ViennaRNA
-% GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
-XXX
 
@@
 
-\subsection{Scala parsers}
-
-\subsection{CUDA parsers}
-{\color{red} XXX: explain why slower: approx 3x more memory used per score (maintain matrix dimensions)
-
-XXX: continue benchmarking here
-}
-
-{\color{red} XXX: Compare current implementation versus ad-hoc implementation.
-Compare CUDA vs Scala (we might need to ad-hoc fix stack overflows in Scala).If Zuker coefficients can be fixed, compare performance with \cite{adp_gpu} by rescaling numbers wrt to bandwidth and computation performance.}
+{\color{red} XXX: say that Haskell package did not install and also time limited to provide largest result (but that are also less relevant as users do not want to wait)}
 

### 2013-01-24 docs/report/benchmarks.tex
Off-by-one in the results...
Using constant memory for most of the RNA parameters.

@@
 \section{Benchmarks} \label{benchmarks}
-In an attempt to provide realistic benchmarks, we tried to gather related implementations.
+In an attempt to provide realistic benchmarks, we tried to gather state-of-art implementations.
 The authors of \cite{gpu_atlp} did not respond to our multiple solicitations.
 The authors of \cite{swat_mega} were very friendly and provided us their source code.
-Unfortunately, since they address a different category of problem (they focus on huge serial problems whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences that our implementation cannot address.
+Unfortunately, since they address a different category of problem (Smith-Waterman on huge sequences whereas we focus on smaller non-serial problems) their implementation might be biased towards large sequences, and leverage problem-specific information (wavefront) that our implementation cannot address.
 Finally, we asked lately the authors of \cite{gpu_rnafold} who did not respond either to our solicitations.
+The authors of \cite{gapc_thesis} kindly share their implementation on a dedicated website\footnote{\url{http://www.gapc.eu}}.
 
-We organize the benchmarks as follow: if we have at our disposal a working implementation that could be run on our evaluation platform, we use it, otherwise, we refer to the related paper and rescale the result according to the memory throughput and computation throughput of the related device so that we can have a good approximation of how they could compare.
+We organize the benchmarks as follow: if we have at our disposal a working implementation that could be run on our evaluation platform, we use it, otherwise, we refer to the related paper and rescale the part of the result corresponding to memory accesses according to the memory bandwidth of the related device so that we can have a good approximation of how they could compare.
 
@@
 % ----------------------------------------------
+% MatrixMult-512, Mac+JDK7
+% -> Original: 24.658 sec
+% -> Optimized concatenation: 18.924 sec
+% -> Tabulation as arrays+inline: 14.849 sec
+% -> Re-optimized concatenation: 10.76 sec
+% ----------------------------------------------
 \def\hh{\normalsize\bf}
@@
 \def\hcua{\hl{CUDAlign}{\cite{swat_linear}, version 2.0}}
-
 \def\hvien{\hl{ViennaRNA}{\cite{vienna_rna}}}
@@
 
-% ----------------------------------------------
-% MatrixMult-512, Mac+JDK7
-% -> Original: 24.658 sec
-% -> Optimized concatenation: 18.924 sec
-% -> Tabulation as arrays+inline: 14.849 sec
-% -> Re-optimized concatenation: 10.76 sec
-\newpage
 \subsection{Matrix chain multiplication}
@@
 The running time of DynaProg/CUDA includes the overhead of back and forth JNI conversion (scales linearly between 0.018 and 0.057 seconds) but does not include the overhead due to the code generation which decomposes in 0.068 seconds for analysis and code synthesis (once per algebra/grammar pair) and 0.086 + 1.753 seconds for respectively Scala and CUDA compilation (constant time, once per problem dimension).
-The results of the following problems are presented similarly.
+These execution time results are presented similarly for the following problems.
 
@@
 \midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
-& \hdps	& 7.84	& 18.95	& 33.63	& 70.86	& 		& 		& $^{(1)}\infty$ \\
+& \hdps	& 7.84	& 18.95	& 33.63	& 70.86	& 		& 		& $^{(2)}\infty$ \\
 & \hhoc	& 0.01	& 0.02	& 0.04	& 0.10	& 0.17	& 0.40	& 0.71	\\
@@
 \midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
-& \hdpc	& 0.07	& 0.11	& 0.15	& 0.13	& 0.20	& 0.32	& $^{(2)}$3.21 	\\
+& \hdpc	& 0.07	& 0.11	& 0.15	& 0.13	& 0.20	& 0.32	& $^{(1)}$3.21 	\\
 & \hhog	& 0.01	& 0.02	& 0.02	& 0.04	& 0.07	& 0.14	& 0.27	\\
@@
 
-$^{(1)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx12G -Xms12G}), and independently of whether top-down or bottom-up parsing approaches are taken.
-After a moment, most of the time is spent in (full) garbage collection.
-The aggregation function application contributes to approximately 30\% of the total running time.
+$^{(1)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance.
+This extra memory requirement is due to the use of 3 matrices to avoid the non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
 
-$^{(2)}$ Since the memory requirements are larger than the device capacity, the backtrack matrix overflows in the main memory, thereby significantly degrading the performance.
-This extra memory requirement is due to the use of 3 matrices to avoid the non-serial dependencies (hence requiring at least $3 \cdot 2$ bytes of memory per matrix element for backtrack).
+$^{(2)}$ Extremely little progress due to intensive JVM garbage collection after some delay, even by tuning the JVM parameters ({\tt -Xss512m -Xmx12G -Xms12G}), and independently of whether top-down or bottom-up parsing approaches are taken.
+After a minute, most of the time is spent in (full) garbage collections.
+In this algorithm, the aggregation function application contributes to approximately 30\% of the total running time.
 
@@
 % GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
+
+The Zuker RNA folding algorithm significantly differs from the two previous problems by the fact that it relies on energy functions.
+These function involve problem parameters that are based on actual observation.
+These constants are encoded in lookup tables that need to be accessed usually at multiple places in each energy function.
+Since the parameters size is approximately 2Mb, it is impossible to put them in the constant memory (which would improve lookup performance).
+In \cite{gpu_rnafold}, such optimization is done, hereby avoiding multiple accesses to the main memory.
+
+Additionally, energy functions involve conditions and possibly loops hence are "very complex" compared to simple regular patterns of matrix multiplication and Smith-Waterman, thereby introducing thread divergence that reduce the performance.
 
@@
 & \hdpc	& 0.15	& 0.53	& 0.98	& 1.55	& 3.13	& 5.04	& 9.43	\\
-& \hrna	& 0.09	& 0.14	& 0.20	& 0.44	& 0.80	& 1.89	& 3.52	\\
+& \hrna	& 0.06	& 0.11	& 0.14	& 0.20	& 0.44	& 0.80	& 1.89	\\
 \midrule
@@
 & \hvien	& 0.57	& 1.53	& 3.19	& 9.37	& 20.18	& 59.65	& 133.65 	\\
-& \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 		& 		&  		\\[-2pt]
+& \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 529.40	& 		&  		\\[-2pt]
 \midrule \multirow{4}{*}{\vspace{12pt}\rotatebox{90}{\normalsize\bf GPU}}
 & \hdpc	& 15.10	& 33.62	& 61.76	& 166.84	& 393.76	& 		&  		\\
-& \hrna	& 9.08	& 19.59	& 67.32	& 163.14	& 		& 		& 		\\
+& \hrna	& 3.52	& 9.08	& 19.59	& 67.32	& 163.14	& 		& 		\\
 \bottomrule\end{tabular}}\end{center}\caption{Running time of Zuker RNA folding (in seconds)}\end{table}
+
+XXX: break constants int two structures
 
@@
 \item Write some conclusion ideas
-\item Run benchmark and add results in report (automate, generate raw Matlab/TeX output)
-\item Stabilize times, statistics, use CUDA profiler(?)
+\item use CUDA profiler(?)
 \ole}

### 2013-01-24 src/librna/rnafold/fold.c

@@ -172,7 +172,6 @@ PUBLIC float fold_par(const char *string, char *structure, paramT *parameters) {
 
   if (parameters) init_fold(length, parameters);
   else if (length>init_length) init_fold(length, parameters);
-  else if (fabs(P->temperature - temperature)>1e-6) update_fold_params_par(NULL);
 
   S   = encode_sequence(string, 0);
   S1  = encode_sequence(string, 1);
@@ -194,12 +193,9 @@ PRIVATE int fill_arrays(const char *string) {
 
   int   i, j, k, length, energy, en;
   int   decomp, new_fML, max_separation;
-  int   no_close, type, type_2, tt;
+  int   type, type_2, tt;
   int   bonus=0;
 
-  int   noGUclosure;
-
-  noGUclosure   = P->model_details.noGUclosure;
   length = (int) strlen(string);
   max_separation = (int) ((1.-LOCALITY)*(double)(length-2)); /* not in use */
 
@@ -220,14 +216,12 @@ PRIVATE int fill_arrays(const char *string) {
       type = ptype[ij];
       energy = INF;
 
-      no_close = (((type==3)||(type==4))&&noGUclosure&&(bonus==0));
-
       if (j-i-1 > max_separation) type = 0;  /* forces locality degree */
 
       if (type) {   /* we have a pair */
         int new_c=0, stackEnergy=INF;
         /* hairpin ----------------------------------------------*/
-        new_c = (no_close) ? FORBIDDEN : E_Hairpin(j-i-1, type, S1[i+1], S1[j-1], string+i-1, P);
+        new_c = E_Hairpin(j-i-1, type, S1[i+1], S1[j-1], string+i-1, P);
         /* check for elementary structures involving more than one closing pair. */
         for (p = i+1; p <= MIN2(j-2-TURN,i+MAXLOOP+1) ; p++) {
           int minq = j-i+p-MAXLOOP-2;
@@ -238,10 +232,6 @@ PRIVATE int fill_arrays(const char *string) {
             if (type_2==0) continue;
             type_2 = rtype[type_2];
 
-            if (noGUclosure)
-              if (no_close||(type_2==3)||(type_2==4))
-                if ((p>i+1)||(q<j-1)) continue;  /* continue unless stack */
-
[…]

### 2013-01-24 src/librna/vienna/vienna.c

@@ -6,13 +6,9 @@
 #include "vienna.h"
 #include "energy_par.h"
 
-double temperature = 37.0; /* temperature */
-
 // -----------------------------------------
 
 static void nrerror(const char* msg) { fprintf(stderr,"ERROR: %s.\n",msg); exit(EXIT_FAILURE); }
-
-int id=-1;
 
 paramT *get_scaled_parameters() {
   unsigned int i,j,k,l,m,n;
@@ -20,73 +16,60 @@ paramT *get_scaled_parameters() {
   paramT *params = (paramT*)calloc(1,sizeof(paramT));
   if (params==NULL) nrerror("Out of memory");
 
-  // set_model_details:
-  model_detailsT md;
-  md.dangles     = 2; /* use dangling end energies */
-  md.special_hp  = 1; /* Fold with specially stable 4-loops */
-  md.noLP        = 1; /* avoid helices of length 1 */
-  md.noGU        = 0; /* GU not allowed at all */
-  md.noGUclosure = 0; /* GU allowed only inside stacks */
-  md.logML       = 0; /* use logarithmic scaling for multi loops */
-
   /* store the model details */
-  params->model_details = md;
-  params->temperature   = temperature;
-  tempf                 = ((temperature+K0)/Tmeasure);
+  double temperature = 37.0; /* temperature */
+  params->p0.temperature = temperature;
+  tempf                  = ((temperature+K0)/Tmeasure);
 
   #define RESCALE(X,Y) (X)-((X)-(Y))*tempf
 
-  for (i=0; i<31; i++) params->hairpin[i] = RESCALE(hairpindH[i], hairpin37[i]);
+  for (i=0; i<31; i++) params->p0.hairpin[i] = RESCALE(hairpindH[i], hairpin37[i]);
   #define MIN2(A,B) ((A) < (B) ? (A) : (B))
   for (i=0; i<=MIN2(30,MAXLOOP); i++) {
-    params->bulge[i]          = RESCALE(bulgedH[i], bulge37[i]);
-    params->internal_loop[i]  = RESCALE(internal_loopdH[i], internal_loop37[i]);
[…]

### 2013-01-24 src/main/scala/v4/CodeGen.scala

@@ -393,9 +393,11 @@ trait CodeGen extends BaseParsers { this:Signature =>
       def path ="../src/librna/"
       "// --------------------------------\n"+
       "#include \""+path+"vienna/vienna.h\"\n"+
+      "__constant__ paramT0 param0;\n"+
       "#define my_len (M_H-1)\n"+
       "#define my_seq _in1\n"+
       "#define my_P g_P\n"+
+      "#define my_P0 param0\n"+
       "#define my_dev __device__\n"+
       "#include \""+path+"librna_impl.h\"\n"+
       "#include \""+path+"vienna/vienna.c\"\n"+
@@ -405,6 +407,7 @@ trait CodeGen extends BaseParsers { this:Signature =>
       "static inline void rna_init() {\n"+
       "  read_parameter_file(\""+s.paramsFile+"\");\n"+
       "  paramT* P = get_scaled_parameters();\n"+
+      "  cudaMemcpyToSymbol(param0,&(P->p0),sizeof(paramT0));"+
       "  cuMalloc(cg_P,sizeof(paramT));\n"+
       "  cuPut(P,cg_P,sizeof(paramT),NULL);\n"+
       "  _initP<<<1,1>>>(cg_P); free(P);\n"+

### 2013-01-24 docs/report/benchmarks.tex
More report.

@@
 \def\hdpc{\hl{DynaProg}{CUDA parsers}}
+\def\hdpcz{\hl{DynaProg}{CUDA-Zuker}}
+\def\hdpcr{\hl{DynaProg}{CUDA-RNAfold}}
 \def\hhoc{\hl{Optimized}{C, single thread}}
@@
 
+\def\hcpu{\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}}
+\def\hgpu{\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}}
+
 \subsection{Matrix chain multiplication}
-Since we have seen that this problem is not input sensitive in (\S\ref{metrics}), we can safely use different random number generators among different implementations without compromising the validity of the results.
+% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
+% GPU: GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?
+We have seen previously that this problem is not input sensitive in (\S\ref{metrics}), hence we can safely use different random number generators among different implementations without compromising the validity of the results.
 Also note that the hand-optimized results are slightly worse than those presented in (\S\ref{baseline_impl}), this is caused by enabling the 64-bit mode.
 Since external libraries linked with the Java virtual machine must be in 64 bit, we also enabled this mode in hand-optimized version to maintain a fair comparison, thereby slightly reducing the performance of CUDA operations.
-% CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
-% GPU: GPU: DynaProg(CUDA), hand optimized CUDA, ATLP and GAPC-OpenCL?
 
 \begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
-&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\ \hcpu
 & \hdps	& 0.05	& 0.20	& 0.80	& 2.03	& 6.65	& 15.10	& 47.40	\\
 & \hhoc	& <0.01	& <0.01	& <0.01	& 0.01	& 0.03	& 0.08	& 0.28	\\
-& \hgapc	& 0.01	& 0.01	& 0.03	& 0.05	& 0.15	& 0.35	& 1.16	\\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hgapc	& 0.01	& 0.01	& 0.03	& 0.05	& 0.15	& 0.35	& 1.16	\\[-2pt] \hgpu
 & \hdpc	& 0.03	& 0.04	& 0.05	& 0.07	& 0.13	& 0.13	& 0.21	\\
@@
 \midrule
-&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\ \hcpu
 & \hdps	& 109.77	& 368.21	& 877.30	& 3059.42 & 		& 		& 		\\
 & \hhoc	& 1.18	& 7.06	& 19.81	& 78.90	& 206.56	& 799.53	& 2010.49 \\
-& \hgapc	& 2.82	& 10.02	& 25.16	& 91.69	& 224.70	& 		&  		\\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hgapc	& 2.82	& 10.02	& 25.16	& 91.69	& 224.70	& 		&  		\\[-2pt] \hgpu
 & \hdpc	& 0.35	& 0.85	& 1.69	& 4.79	& 10.32	& 31.60	& 71.22	\\
@@
 
-\subsection{Smith-Watermann}
+From the results we see that this problem is well suited for GPUs where the computation pattern is very regular across threads.
+In this case, the generated CUDA code produces performance that is comparable to hand-optimized CUDA code, and comparable to one of the state of the art implementation with less than $1.5\times$ performance degradation.
+ATLP leverages the regularity of the access pattern optimize the resource utilization and adaptively map subproblems\cite{gpu_atlp}.
+As we cannot benefit of such problem-specific knowledge, our approach is to compute elements independently and synchronize efficiently.
+
+\subsection{Smith-Waterman (affine gap cost)}
 % CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion
 % GPU: DynaProg(CUDA), hand optimized CUDA, CUDAlign
-To maintain a fair comparison with other implementations, our algorithm needs to have the same complexity.
-Hence we use a variant of the Smith-Waterman algorithm: we encode our grammar in 3 tabulations as described in \S\ref{swat_affine}, whereas application like CUDAlign can leverage the domain-specific information that information of two of these matrices can be stored in the computation wavefront (see \S\ref{calc_simplifications}).
+Smith-Waterman with affine gap cost is a serial problem, which is not directly the focus of our project but nevertheless can be addressed efficiently.
+The problem formal description (\S\ref{swat_affine}) uses 3 matrices.
+However, since two of these matrices only propagate information along line or column, it is possible to encode this information in a wavefront (see \S\ref{calc_simplifications}) instead of maintaining it in a memory-expensive matrix.
+This knowledge is leveraged by \cite{ATLP}, however, in our case, we cannot describe the wavefront in the grammar and need to maintain explicitly 3 matrices, thereby multiplying by 3 the memory usage\footnote{Actually slightly less than 3 because we replace 2 matrices in $O(n^2)$ by two vectors in $O(n)$}.
 
 \begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
-&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\ \hcpu
 & \hdps	& 0.04	& 0.13	& 0.27	& 0.48	& 1.07	& 1.92	& 4.33	\\
 & \hhoc	& <0.01	& <0.01	& <0.01	& <0.01	& <0.01	& <0.01	& 0.01	\\
-& \hgapc	& 0.01	& 0.01	& 0.01	& 0.01	& 0.02	& 0.03	& 0.06	\\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hgapc	& 0.01	& 0.01	& 0.01	& 0.01	& 0.02	& 0.03	& 0.06	\\[-2pt] \hgpu
 & \hdpc	& 0.03	& 0.03	& 0.03	& 0.04	& 0.05	& 0.05	& 0.06	\\
@@
 \midrule
-&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\ \hcpu
 & \hdps	& 7.84	& 18.95	& 33.63	& 70.86	& 		& 		& $^{(2)}\infty$ \\
 & \hhoc	& 0.01	& 0.02	& 0.04	& 0.10	& 0.17	& 0.40	& 0.71	\\
-& \hgapc	& 0.10	& 0.22	& 0.39	& 0.91	& 1.62	& 4.41	& 11.20 	\\[-2pt]
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf GPU $\quad$}}
+& \hgapc	& 0.10	& 0.22	& 0.39	& 0.91	& 1.62	& 4.41	& 11.20 	\\[-2pt] \hgpu
 & \hdpc	& 0.07	& 0.11	& 0.15	& 0.13	& 0.20	& 0.32	& $^{(1)}$3.21 	\\
@@
 
+From these result, it is possible to say that although not being the primary focus, the serial problem can be solved efficiently as well in our framework, assuming that the device memory is sufficiently large to store all the matrices.
+We might attribute the proportional overhead of our implementation (compared to the optimized version) to the additional verifications for result cell emptiness (these are hidden in matrix multiplication by the memory accesses).
+Removing them requires problem-specific knowledge (for example setting an infinite value), as described in \S\ref{normalization}.
+
 \subsection{Zuker RNA folding}
@@
 % GPU: DynaProg(CUDA), RNAFold?/Lavenier VS ViennaRNA-OpenCL?, GAPC-OpenCL?
+The Zuker RNA folding algorithm significantly differs from the two previous problems because it relies on energy functions involving experimental parameters (constants).
+These parameters are encoded in lookup tables that need to be accessed usually at multiple places in each energy function.
+Additionally, energy functions involve conditions and possibly loops hence are more complex than the simple regular patterns of matrix multiplication and Smith-Waterman, thereby introducing possible thread divergence that reduces the performance.
 
-The Zuker RNA folding algorithm significantly differs from the two previous problems by the fact that it relies on energy functions.
-These function involve problem parameters that are based on actual observation.
-These constants are encoded in lookup tables that need to be accessed usually at multiple places in each energy function.
-Since the parameters size is approximately 2Mb, it is impossible to put them in the constant memory (which would improve lookup performance).
-In \cite{gpu_rnafold}, such optimization is done, hereby avoiding multiple accesses to the main memory.
-
-Additionally, energy functions involve conditions and possibly loops hence are "very complex" compared to simple regular patterns of matrix multiplication and Smith-Waterman, thereby introducing thread divergence that reduce the performance.
+For this problem, we present two grammar variants.
+The Scala version and CUDA-Zuker share the same grammar as GAPC\cite{gapc_thesis}, which is more complex than the RNAfold grammar, that is shared by the two implementation with the same name.
+Although not stated explicitly in the paper \cite{gpu_rnafold}, this implementation limits the size of internal loops to 30 (following current practice in the domain \cite{rna_structure}).
+Limiting the size of the internal loops allows to reduce the running time complexity from $O(n^4)$ to $O(n^3)$, because large loops rarely occur in practice \cite{rna_structure}.
 
 \begin{table}[H]\begin{center}{\small\begin{tabular}{llrrrrrrr}\toprule
-&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh  Matrix dimension &\hh 64 &\hh 128 &\hh 192 &\hh 256 &\hh 384 &\hh 512 &\hh 768 \\ \hcpu
 & \hdps	& 0.07	& 0.67	& 2.26	& 4.98	& 17.61	& 39.44	& 130.71	\\
 & \hvien	& 0.01	& 0.01	& 0.02	& 0.03	& 0.07	& 0.12	& 0.29	\\
-& \hgapc	& 0.01	& 0.03	& 0.07	& 0.13	& 0.41	& 0.93	& 2.89	\\[-2pt]
-\midrule \multirow{4}{*}{\vspace{12pt}\rotatebox{90}{\normalsize\bf GPU}}
-& \hdpc	& 0.15	& 0.53	& 0.98	& 1.55	& 3.13	& 5.04	& 9.43	\\
+& \hgapc	& 0.01	& 0.03	& 0.07	& 0.13	& 0.41	& 0.93	& 2.89	\\[-2pt] \hgpu
+& \hdpcz	& 0.13	& 0.48	& 0.88	& 1.41	& 2.84	& 4.55	& 8.51	\\
+& \hdpcr	& 0.17	& 0.54	& 0.92	& 1.33	& 2.25	& 3.22	& 5.34	\\
 & \hrna	& 0.06	& 0.11	& 0.14	& 0.20	& 0.44	& 0.80	& 1.89	\\
 \midrule
-&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\
-\midrule \multirow{4}{*}{\rotatebox{90}{\normalsize\bf CPU $\quad$}}
+&\hh Matrix dimension &\hh 1024 &\hh 1536 &\hh 2048 &\hh 3072 &\hh 4096 &\hh 6144 &\hh 8192 \\ \hcpu
 & \hdps	& 306.12	& 1036.78 &		& 		& 		& 		& 	 	\\
 & \hvien	& 0.57	& 1.53	& 3.19	& 9.37	& 20.18	& 59.65	& 133.65 	\\
-& \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 529.40	& 		&  		\\[-2pt]
-\midrule \multirow{4}{*}{\vspace{12pt}\rotatebox{90}{\normalsize\bf GPU}}
-& \hdpc	& 15.10	& 33.62	& 61.76	& 166.84	& 393.76	& 		&  		\\
+& \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 529.40	& 		&  		\\[-2pt] \hgpu
+& \hdpcz	& 13.61	& 30.45	& 56.22	& 152.88	& 		& 		& 		\\
+& \hdpcr	& 7.68	& 15.89	& 26.43	& 66.30	& 153.90 & 		& 		\\
 & \hrna	& 3.52	& 9.08	& 19.59	& 67.32	& 163.14	& 		& 		\\
@@
 
-XXX: break constants int two structures
-
-{\color{red} XXX: comment, explain energy coefficients, complexity, limit search space for bulges, algorithm variants and constraints RNAfold/Zuker}
+\subsection{Synthetical results}
 
@@
 % ------------------------------------------------------------------------------------------------
+%
 
@@
 
-\subsection{Synthetical results}
 XXX

### 2013-01-24 docs/report/implementation.tex
More report.

@@
 
-In the first phase, we adapted the library embedded in GAPC to obtain coefficients.
+First, we adapted the library embedded in GAPC to obtain coefficients.
 Since GAPC is written in C, we had to write a JNI interface to let the Scala code communicate with the libraries.
@@
 To reduce this penalty (towards benchmarking), we introduced memoization of the compilation results in our code generator, thereby avoiding duplicated compilations of the program for same length of sequences (generated programs are tailored for particular sequences lengths).
+
+Finally, we made two small nevertheless important enhancements to the CUDA version: because the energy functions heavily rely on the sequence and the coefficients, we would like to have fast access to them.
+Since we address small sequences ($\le 16$KB), these can easily fit in the shared memory.
+We also would like to benefit from the constant memory to store the coefficients.
+Unfortunately, since this memory is too small to contain all of them, we need to distinguish two cases: the most frequently used coefficients are stored in the constant memory whereas the other have to be put in the global memory.
+These two modifications provided substantial speed improvement by moving the data frequently used closer to the processing units.
 

### 2013-01-24 docs/report/benchmarks.tex
More report.

@@
 \def\hvien{\hl{ViennaRNA}{\cite{vienna_rna}}}
-\def\hrna{\hl{RNAFold}{\cite{gpu_rnafold}}}
+\def\hrna{\hl{RNAfold}{\cite{gpu_rnafold}}}
 
@@
 & \hgapc	& 6.66	& 22.91	& 56.97	& 208.33	& 529.40	& 		&  		\\[-2pt] \hgpu
-& \hdpcz	& 13.61	& 30.45	& 56.22	& 152.88	& 		& 		& 		\\
+& \hdpcz	& 13.61	& 30.45	& 56.22	& 152.88	& 365.24	& 		& 		\\
 & \hdpcr	& 7.68	& 15.89	& 26.43	& 66.30	& 153.90	& 		& 		\\
@@
 
-\subsection{Synthetical results}
-
-% ------------------------------------------------------------------------------------------------
-{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
-% ------------------------------------------------------------------------------------------------
-%
-
-{\color{red}\ol
-\item Write some conclusion ideas
-\item use CUDA profiler(?)
-\ole}
-
-XXX
-
-{\color{red} We want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
-
-RUN 5-10x until running time stabilizes on Scala
-
- {\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
-
-XXX: future work: proper support of two-track grammars
-
-XXX: future work, resize tables appropriately as in GAPL.
-
-XXX: future work: infer automatically "always non-empty" property
-
-{\color{red} XXX: say that Haskell package did not install and also time limited to provide largest result (but that are also less relevant as users do not want to wait)}
+These results are interesting because they demonstrate that for complex problems, the GPU do not behave as good as we wished compared to the CPU (comparing ViennaRNA and RNAfold).
+Beside that, we can notice that for large sequences, our Zuker grammar GPU implementation is on par with GAPC, and the RNAfold grammar is on par with the RNAfold implementation, although we might argue that the purpose of RNAfold is slightly different (using synergistically CPU and GPU to fold multiple small sequences of RNA).
 

### 2013-01-24 docs/report/conclusion.tex
Nits.
More report.

@@
 \item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
-We could use them to either simplify the user-functions description (by implicitly bootstrapping LMS code translation) or even completely replace LMS types by either type conversion or by providing ad-hoc conversion from the Scala AST to C code.
+We could use them to either simplify the user-functions description (by converting types to bootstrap LMS code generation) or even provide ad-hoc conversion from the Scala AST to C code.
 \item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
@@
 \section{Conclusion}
-{\color{red} XXX}
-
 This Master project is focused on how to solve efficiently dynamic programming.
@@
 
+From the benchmarks, we see that our code generation is able to deal with problems from simple (Smith-Waterman) to complex (Zuker and RNAfold) and provide performance that is comparable to existing work.
+Since these problem are expressed with a grammar and algebra (following the ADP decomposition), the description of problems is simplified and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
+
+% ------------------------------------------------------------------------------------------------
+{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+% ------------------------------------------------------------------------------------------------
+
+{\color{red} XXX: fix the introduction}
+
+{\color{red} Benchmarks: we want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
+{\color{red} XXX: say that Haskell package did not install and also time limited to provide largest result (but that are also less relevant as users do not want to wait)}
 {\color{red} What's the conclusion on experimental results?}
+
+XXX: future work, resize tables appropriately as in GAPL.
+{\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
+
+XXX: future work: infer automatically "always non-empty" property
+
+{\color{red} Use CUDA profiler(?)}
 
@@
 
+explain better approach: LMS write intermediate nodes, we embed nodes in actual objects
+
 Big picture (Larger picture how to go further):
@@
   - others are iterative or sparse, different scheduling => multiple scheduling patterns to be discovered and implemented efficiently, possibly in LMS (if time permits)
-
 \end{verbatim}

### 2013-01-24 docs/report/benchmarks.tex
Nits.

@@
 \def\hgapc{\hl{GAPC}{\cite{gapc_thesis}, C, single thread}}
-\def\hatlp{\hl{ATLP}{\cite{gpu_atlp}, rescaled results$^{(1)}$}}
+\def\hatlp{\hl{ATLP}{\cite{gpu_atlp}, rescaled$^{(1)}$}}
 \def\hcua{\hl{CUDAlign}{\cite{swat_linear}, version 2.0}}
 \def\hvien{\hl{ViennaRNA}{\cite{vienna_rna}}}
-\def\hrna{\hl{RNAfold}{\cite{gpu_rnafold}}}
+\def\hrna{\hl{RNAfold}{\cite{gpu_rnafold}, leveraging \cite{adp_gpu}}}
 
@@
 & \hhog	& <0.01	& 0.01	& 0.01	& 0.02	& 0.04	& 0.08	& 0.17	\\
-& \hatlp	& 0.17	&	 	&		& 0.20	& 		& 0.23	& 		\\
+& \hatlp	& 0.17	& ---	 	& ---		& 0.20	& --- 		& 0.23	& --- 		\\
 \midrule
@@
 & \hhog	& 0.32	& 0.82	& 1.65	& 4.74	& 10.35	& 31.94	& 72.38	\\
-& \hatlp	& 0.40	& 0.74	& 1.33	& 3.43	& 7.29	& 		& 		\\
+& \hatlp	& 0.40	& 0.74	& 1.33	& 3.43	& 7.29	& ---		& ---		\\
 \bottomrule\end{tabular}}\end{center}\caption{Running time of matrix chain multiplication (in seconds)}\end{table}
@@
 
-These results are interesting because they demonstrate that for complex problems, the GPU do not behave as good as we wished compared to the CPU (comparing ViennaRNA and RNAfold).
+These results are interesting because they demonstrate that complex problems are hard to parallelize efficiently, and the GPU implementation might not exhibit significant speedups (comparing ViennaRNA and RNAfold).
 Beside that, we can notice that for large sequences, our Zuker grammar GPU implementation is on par with GAPC, and the RNAfold grammar is on par with the RNAfold implementation, although we might argue that the purpose of RNAfold is slightly different (using synergistically CPU and GPU to fold multiple small sequences of RNA).
 
+\subsection{Synthetic results}
+From the previous results, we can make the following observations:\ul
+\item Plain Scala parsers are not well suited for dynamic programming (even for simple problems like Smith-Waterman).
+A better approach could be to use LMS to rewrite the Scala parsers into more efficient code, possibly reducing the running time to be comparable with GAPC (modulo the penalty introduced by the JVM compared to native execution).
+\item Offloading the computations to the GPU might bring significant speedup ($\sim 20\times$ for matrix multiplication) but some problem do not perform as well as what is possible on the CPU (Zuker), hence the programmer decision should be driven by performance evaluation rather than assumptions.
+\item The CUDA code generated by DynaProg is comparable with hand-optimized code.
+\ule
+
 %\end{document}

### 2013-01-24 docs/report/conclusion.tex
Nits.

@@
 We consider several directions and possible extensions for our work. We briefly describe each of them and give an idea of how they could be implemented:\ol
-\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
+
+\item \textbf{Fusion and C with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
+Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser functions into a single one and providing loops fusion\footnote{For example if the aggregation function is simple enough as maximum or minimum, aggregation could be merged with element processing, simliarly as described in \S\ref{normalization}.}.
+The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
+Also constructing large lists of candidates that are later reduced to a single one  increase the work of the garbage collector.
+Generating the grammar through LMS has two major benefits:\ol
+\item Improve the performance of the parsers within the JVM.
+Since we previously argued in favor of a decoupling of the algebra and the grammar, and since we want to reach optimal performance, we would need to keep available both a generic version for result processing (independent of LMS restrictions so that we can use arbitrary functions within the algebra) and an optimized version for computations.
+The latter could be achieved similarly as the current CUDA code generation: the Scala parsers might produce an abstract syntax tree (AST) that could then be merged with the algebra nodes and passed to LMS for code generation.
+\item Since experimental results have shown that in some situations, the CPU outperforms the GPU, it might also be interesting to also target single thread C implementation to benefit from these situations (thereby removing the overhead of the JVM).
+Using LMS would provide us with the support for such code generation.
+Since the program can be single threaded, no complex synchronization mechanism is involved, hence it is an ideal candidate for LMS multi-architecture code generation.
+\ole
+\item \textbf{Macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
+We could use them to either simplify the user-functions description (by converting types to bootstrap LMS code generation) or even provide ad-hoc conversion from the Scala AST to C code.
+\item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \S\ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
 However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
-\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program.
-Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (see \ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (see \ref{user_lang}).
-These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (thereby knowing how to optimize manually the grammar).
-\item \textbf{Serial problems larger than memory:} As discussed in \ref{serial_memory}, this class of problems require a completely different implementation.
+\item \textbf{Serial problems larger than memory:} As discussed in \S\ref{serial_memory}, this class of problems require a completely different implementation.
 Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
-\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform, the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
-The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the data can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
+\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
+Hence cost and backtrack matrix would contain $k$ elements per cell.
+Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
 \item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.
@@
 Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
+\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform, the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
+The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the data can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
+\item \textbf{Algorithmic analysis:} so far, we considered that the DSL user would write an optimal program.
+Another direction in which we could improve the parsers is the recurrence analysis, either by removing serial dependencies when possible (\S\ref{calc_simplifications}) or reducing the algorithmic complexity by creating intermediate tabulations (\S\ref{user_lang}).
+These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (thereby knowing how to optimize manually the grammar).
+\item \textbf{Non-emptiness analysis:} At code generation level, when a tabulation is known to be \textit{non-empty} (every cell contain a valid result), it is possible to remove the validity checks, thereby reducing the parser complexity\footnote{The validity of a result is determined if its corresponding backtrack has a valid rule number.
+Skipping this test saves a memory load, and a condition testing.}.
+Such analysis needs to make sure that among all possible candidates, there exist at least one valid.
+Such analysis might be quite complex as it requires induction (for example in matrix multiplication, we inductively need to prove that every cell element is non-empty.
+Currently, this information needs to be explicitly provided by the programmer (by setting a flag on the non-empty tabulation).
 \item \textbf{Pruning:} described in \cite{swat_mega}, this optimization could lead to a reduction of the computation, provided that the algorithm final score can be bounded.
 Such optimization would only be relevant with a non-uniform computation strategy where the matrix is tiled, thereby making it possible to prune entire computation tiles.
-\item \textbf{Fusion with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
-Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser function calls into a single function making equivalent calls.
-The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
-\item \textbf{Using macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
-We could use them to either simplify the user-functions description (by converting types to bootstrap LMS code generation) or even provide ad-hoc conversion from the Scala AST to C code.
-\item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
-Hence cost and backtrack matrix would contain $k$ elements per cell.
-Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
 \ole
@@
 These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtain a single optimal result.
-We then provide some technical details explaining how these ideas are put in practice in \textit{DynaProg} (\S\ref{implementation}).
+We then provide some technical details explaining how these ideas are put in practice in DynaProg (\S\ref{implementation}).
 Finally, we compare our implementation with existing implementations (\S\ref{benchmarks}).
@@
 From the benchmarks, we see that our code generation is able to deal with problems from simple (Smith-Waterman) to complex (Zuker and RNAfold) and provide performance that is comparable to existing work.
-Since these problem are expressed with a grammar and algebra (following the ADP decomposition), the description of problems is simplified and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
+Since these problem are expressed with a grammar and algebra, their expression is simplified, and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
 
@@
 % ------------------------------------------------------------------------------------------------
-
 {\color{red} XXX: fix the introduction}
@@
 {\color{red} Benchmarks: we want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
+
 {\color{red} XXX: say that Haskell package did not install and also time limited to provide largest result (but that are also less relevant as users do not want to wait)}
+
 {\color{red} What's the conclusion on experimental results?}
-
-XXX: future work, resize tables appropriately as in GAPL.
-{\color{red} Size analysis to know what storage size we require: ex: Zucker requires $O(n^2)+O(n)$ storage...}
-
-XXX: future work: infer automatically "always non-empty" property
 

### 2013-01-24 docs/report/conclusion.tex
Nits.

@@
 % ------------------------------------------------------------------------------------------------
+\newpage
 \section{Conclusion}
@@
 Then we discussed the requested functionalities to provide the user with a convenient embedded DSL that is based on the ADP formalization (\S\ref{architecture}).
-These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtain a single optimal result.
+These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtain a single optimal result on graphic cards.
 We then provide some technical details explaining how these ideas are put in practice in DynaProg (\S\ref{implementation}).
-Finally, we compare our implementation with existing implementations (\S\ref{benchmarks}).
+Finally, we see in our benchmarks that generated code is comparable with existing implementations (\S\ref{benchmarks}).
 
@@
 
+From a larger perspective, our work only address the subset of dynamic programming problems on sequences, there exist different dynamic programming problems that iteratively refine an approximation to converge to the solution (for example Bellman-Ford\footnote{\url{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}}) these algorithm have completely different access patterns but might also be parallelized.
+From this initial observation emerges a pattern for addressing problems:\ol
+\item Understand the gist of the problem and generalize its specific characteristics (we might call it domain knowledge)
+\item Equipped with that information, it is possible to devise an efficient implementation, possibly parallel\footnote{Because the single core (processor) model has reached its limit and the industry is moving towards multi-core and many-core.}, and encompassing as many problems as possible, given the trade-off between generality and specific information required to maintaing high performance.
+With the current compiler technology and hardware architectures, it has become easier to express such solution with code generation: a specific problem is optimally addressed, while the generality of the technique is preserved within the code generator.
+\item Finally, once an efficient solver has been implemented, it should be provided to other people trying to solve the problems in the same category.
+To do that, everybody must agree on a <<standard>>.
+Instead of creating new languages, it would be best if everybody would speak the same.
+This is the very purpose of embedded DSL: people knowing how to use the host language will easily understand how to encode their problem.
+\ole
+
+From our perspective it seemed that Scala and LMS represents the ideal candidate as a host language.
+Our work only provide one small construction brick in the huge space of the problems\footnote{For example \url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.pdf}} that remain to be solved...
+
 % ------------------------------------------------------------------------------------------------
-{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
+%{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
 % ------------------------------------------------------------------------------------------------
-{\color{red} XXX: fix the introduction}
-
-{\color{red} Benchmarks: we want to see the benefits of moving to CUDA, also compare to how far Scala is from C.}
-
-{\color{red} XXX: say that Haskell package did not install and also time limited to provide largest result (but that are also less relevant as users do not want to wait)}
-
-{\color{red} What's the conclusion on experimental results?}
-
-{\color{red} Use CUDA profiler(?)}
-
+%{\color{red} XXX: fix the introduction}
+%{\color{red} Use CUDA profiler(?)}
 % http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html
-
-\begin{verbatim}
-Recap of the report:
-- Goal: generalize solving dynamic programming on sequences for GPUs
-- classification of the problems; in terms of recurrences = complicated, grammar = easier
-- general technique to describe backtracking information
-- implemented in Scala+CUDA
-- additional analysis to generate efficient code (1 result, memory alignment, scheduling)
-
-Recap of the results:
-- XXX
-
-explain better approach: LMS write intermediate nodes, we embed nodes in actual objects
-
-Big picture (Larger picture how to go further):
-- Code generation and actual compiler technology
-- Embedded DSL vs external DSL
-- Only a subset of the dynamic programming problems, easy to parallelize (similar memory accesses and scheduling pattern/progress)
-  - others are iterative or sparse, different scheduling => multiple scheduling patterns to be discovered and implemented efficiently, possibly in LMS (if time permits)
-\end{verbatim}
+%
+%\begin{verbatim}
+%Big picture (Larger picture how to go further):
+%- Code generation and actual compiler technology
+%- Embedded DSL vs external DSL
+%- Only a subset of the dynamic programming problems, easy to parallelize (similar memory accesses and scheduling pattern/progress)
+%  - others are iterative or sparse, different scheduling => multiple scheduling patterns to be discovered and implemented efficiently, possibly in LMS (if time permits)
+%\end{verbatim}

### 2013-01-24 docs/report/introduction.tex
More report.
Nits.

@@
 DynaProg allows the programmer to write concise programs based on ADP \cite{adp}, using a pair of parsing grammar and algebra; these program can then be executed either on CPU or on GPU.
-We evaluate the performance of our implementation against existing work and own baseline implementations for both the CPU and GPU versions.
-Experimental results show an average speedup of {\color{red} XXX} for large problems when they are run on the GPU (compared to CPU).
-Compared to ad-hoc GPU implementations, the generated parsers have an average slowdown of {\color{red} XXX}.
-The CPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,GAPC?}.
-The GPU implementation has a slowdown of {\color{red} XXX} for {\color{red} PROBLEM} against {\color{red} REF,CudAlign, RNAFold}.
+We evaluate the performance of our implementation against existing work and our own hand-optimized baseline implementations for both the CPU and GPU versions.
+Experimental results show that plain Scala has a large overhead and can only be used for small sequences ($\le 1024$) whereas the generated GPU version is comparable with existing implementations: matrix chain multiplication has the same performance as our hand-optimized version (142\% of the execution time of \cite{gpu_atlp}) for a sequence of 4096 matrices, Smith-Waterman is twice slower than \cite{swat_linear} on a pair of sequences of 6144 elements, and  RNA folding is on par with \cite{gpu_rnafold} (95\% running time) for sequences of 4096 elements.
 
@@
 On graphic cards, there exist two levels of memory that are visible for the programmer: the global memory, which can be accessed by any thread, and the shared memory, that corresponds to an explicitly addressable cache memory, whose access is faster but restricted to threads in the same block.
+A small amount of (global) memory can be marked as constant, so that its caching and reading strategy can be adapted consequently \cite{cuda_constant}.
 Finally, access to the main memory of the computer is possible on recent cards but suffers an additional penalty, which makes it not desirable.

### 2013-01-24 docs/report/benchmarks.tex
Nits.

@@
 \item Plain Scala parsers are not well suited for dynamic programming (even for simple problems like Smith-Waterman).
-A better approach could be to use LMS to rewrite the Scala parsers into more efficient code, possibly reducing the running time to be comparable with GAPC (modulo the penalty introduced by the JVM compared to native execution).
+A better approach would be to use LMS to rewrite the Scala parsers into more efficient code, possibly reducing the running time to be comparable with GAPC (modulo the penalty introduced by the JVM compared to native execution).
 \item Offloading the computations to the GPU might bring significant speedup ($\sim 20\times$ for matrix multiplication) but some problem do not perform as well as what is possible on the CPU (Zuker), hence the programmer decision should be driven by performance evaluation rather than assumptions.

### 2013-01-24 docs/report/conclusion.tex
Nits.

@@
 
-\item \textbf{Fusion and C with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimization.
-Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser functions into a single one and providing loops fusion\footnote{For example if the aggregation function is simple enough as maximum or minimum, aggregation could be merged with element processing, simliarly as described in \S\ref{normalization}.}.
+\item \textbf{Fusion and C with LMS:} Although we gained some speedup by optimizing manually the Scala parsers, we cannot benefit from grammar-specific optimizations.
+Passing the whole grammar to LMS could possibly lead to more efficient code, by folding multiple parser functions into a single one and providing loop fusion\footnote{For example if the aggregation function is simple enough as maximum or minimum, aggregation could be merged with element processing, simliarly as described in \S\ref{normalization}.}.
 The added cost of function lookup, even if minimal, might still account for a non-negligible part of the total running time as parser processing is very simple but run repeatedly a large number of time.
-Also constructing large lists of candidates that are later reduced to a single one  increase the work of the garbage collector.
+Also constructing large lists of candidates that are later reduced to single results increases the work of the garbage collector.
 Generating the grammar through LMS has two major benefits:\ol
 \item Improve the performance of the parsers within the JVM.
-Since we previously argued in favor of a decoupling of the algebra and the grammar, and since we want to reach optimal performance, we would need to keep available both a generic version for result processing (independent of LMS restrictions so that we can use arbitrary functions within the algebra) and an optimized version for computations.
-The latter could be achieved similarly as the current CUDA code generation: the Scala parsers might produce an abstract syntax tree (AST) that could then be merged with the algebra nodes and passed to LMS for code generation.
-\item Since experimental results have shown that in some situations, the CPU outperforms the GPU, it might also be interesting to also target single thread C implementation to benefit from these situations (thereby removing the overhead of the JVM).
-Using LMS would provide us with the support for such code generation.
-Since the program can be single threaded, no complex synchronization mechanism is involved, hence it is an ideal candidate for LMS multi-architecture code generation.
+Since we previously argued in favor of a decoupling of the algebra and the grammar, and since we want to reach optimal performance, we would need to provide both a plain version for result processing (independent of LMS restrictions so that we can use arbitrary functions within the algebra) and an optimized version for dynamic programming computations (possibly with restricted types).
+The latter could be achieved similarly as the current CUDA code generation: the Scala parsers could be converted in an abstract syntax tree (AST) that could then be merged with the algebra nodes and passed to LMS for code generation.
+\item Since experimental results have shown that in some situations, the CPU outperforms the GPU, it might be interesting to target single thread C implementation to benefit from these situations (thereby removing the overhead of the JVM).
+Using LMS would provide such code generation mechanism.
+Since the program can be single threaded, no complex synchronization mechanism needs to be involved, hence this is a good candidate for LMS multi-architecture code generation.
 \ole
-\item \textbf{Macros:} macros provide an interesting meta-programming opportunity as they are being run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
-We could use them to either simplify the user-functions description (by converting types to bootstrap LMS code generation) or even provide ad-hoc conversion from the Scala AST to C code.
+\item \textbf{Macros:} macros provide an interesting meta-programming opportunity as they are run after the typing phase of the Scala compiler and can leverage all the compile-time typing information.
+We could use them to either simplify the user-functions description (by converting their types into LMS) or even provide ad-hoc conversion from the Scala AST to C code.
 \item \textbf{Non-serial scheduling for problems larger than the device memory:} as described in \S\ref{ns_mem_transfer}, it could be possible to handle problems that are larger than the device memory in an efficient way, thereby dramatically reducing the memory transfer penalties introduced by the main memory usage.
-However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (since we first need to find the enclosing matrix block before addressing the element relatively to it).
+However, this comes at the price of a more involved kernel scheduling and a complex element indexing strategy (because the enclosing matrix block needs to be looked-up before addressing one of its element).
 \item \textbf{Serial problems larger than memory:} As discussed in \S\ref{serial_memory}, this class of problems require a completely different implementation.
-Since the authors of \cite{swat_mega} are planning to write extensions to their implementation, duplicating the effort might not be worth the price; however, would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
+Since the authors of \cite{swat_mega} are planning to publish an extension for their implementation, duplicating the effort might not be worth the price.
+Would their future implementation be sufficiently modular, we could integrate it in our framework and redirect compatible grammars to this state of art implementation.
 \item \textbf{CUDA $k$-best parsers:} since a $k$-best algorithm has constant memory requirements, an efficient algorithm for CUDA could be devised: instead of comparing with only one value to find the best value, it suffice to compare with $k$ values instead.
 Hence cost and backtrack matrix would contain $k$ elements per cell.
-Since the problem of result validity is already addressed, cells with fewer results would simply have fewer cells marked valid.
+Validity flag for one value can easily be extended to multiple values; cells with fewer results would simply have fewer elements marked valid.
 \item \textbf{Multi-dimensional matrices and independent computations:} In the current implementation, all the matrices are encoded such that they are of the same size.
-Leveraging the yield analysis, we could reduce the dimension of matrices that are of smaller dimension (for tabulations with bounded maximal size).
-In the problems we have analyzed, no such special case appeared, this is why we do not support this optimization at present.
-
-Matrix of different dimensions must be stored in their own array (versus being in a single array of struct enclosing corresponding element of all matrices).
-Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description, there are two $O(n^2)$ matrices and one $O(n)$ matrix.
-This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed one after another whereas in the current computation, the same cell in all matrices is computed at once.
-\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we focus on data representation; in the current project, data is stored in primary types but we could store them more efficiently.
+Leveraging the yield analysis, we could reduce the dimension of smaller matrices (for tabulations with bounded maximal size).
+Matrix of different dimensions must be stored in their own array (versus being in a single array of matrices of similar dimensions).
+Also matrices might possibly be of different storage complexity: looking back at the Zuker problem description (\S\ref{zuker}), there are two $O(n^2)$ matrices and one $O(n)$ matrix.
+This discrepancy in the sizes also leads to multiple indexing strategies (depending on the complexity) and a more complex scheduling where matrix must be computed sequentially whereas in the current computation, one position is computed in all matrices at once.
+\item \textbf{Data granularity:} Since the major bottleneck of CUDA architecture is the memory, we might focus on data representation: in the current project, data is stored in primary types but we could store them more efficiently.
 For example, RNA is represented with only 4 letters (g,a,t,c), thus 4 symbols could be encoded in a single byte.
 Unfortunately, this optimizations seems to only apply for the input data.
-Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on the shared axis.
-\item \textbf{Add FPGA as target platform:} Initially envisioned a second target platform, the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
+Another solution in this direction is to operate on multiple cells with one thread, the argument being that they could share a row or a column, thereby dividing the number of memory accesses for non-serial dependencies on that axis.
+\item \textbf{Adding FPGA as target platform:} Initially envisioned a second target platform, the underlying complexity of transforming DP recurrences into VHDL code made us leave this platform aside for the scope of this project.
 The reconfigurability possibilities of FPGA make them attractive whenever it comes to very simple and massively parallel computations where the data can be pipelined; this makes serial dynamic programming problems good candidates for such implementation.
@@
 These analysis would certainly involve a strong mathematical analysis and the ratio benefit over implementation complexity would be quite small under the initial assumption that the DSL users are experts in their field (thereby knowing how to optimize manually the grammar).
-\item \textbf{Non-emptiness analysis:} At code generation level, when a tabulation is known to be \textit{non-empty} (every cell contain a valid result), it is possible to remove the validity checks, thereby reducing the parser complexity\footnote{The validity of a result is determined if its corresponding backtrack has a valid rule number.
+\item \textbf{Non-emptiness analysis:} At code generation level, when a tabulation is known to be \textit{non-empty} (every cell contain a valid result), it is possible to remove the corresponding verifications, thereby reducing the parser complexity\footnote{The validity of a result is determined if its corresponding backtrack has a valid rule number.
 Skipping this test saves a memory load, and a condition testing.}.
-Such analysis needs to make sure that among all possible candidates, there exist at least one valid.
-Such analysis might be quite complex as it requires induction (for example in matrix multiplication, we inductively need to prove that every cell element is non-empty.
+Such analysis needs to guarantee that all paths would produce at least on candidate.
+Such analysis might be quite complex as it requires induction (for example in matrix multiplication, we need to prove inductively that every cell element is non-empty).
 Currently, this information needs to be explicitly provided by the programmer (by setting a flag on the non-empty tabulation).
@@
 \section{Conclusion}
-This Master project is focused on how to solve efficiently dynamic programming.
-By restricting to the class of problem involving sequences as input, we were able to extract generic patterns and expose them to parallel architectures like CUDA.
+This Master project report is focused on how to solve efficiently dynamic programming.
+By restricting to the class of problem involving sequences, we were able to extract generic patterns and expose them to parallel architectures like CUDA.
 To do that, we first depicted the dynamic programming landscape and defined a set of problems we would like to solve (\S\ref{problems}).
-Then we discussed the requested functionalities to provide the user with a convenient embedded DSL that is based on the ADP formalization (\S\ref{architecture}).
-These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtain a single optimal result on graphic cards.
+Then we discussed the required functionalities to provide the user with a convenient embedded DSL based on the ADP formalization (\S\ref{architecture}).
+These architectural decisions lead us to consider two different implementations: one in Scala that allows the same expressivity as ADP (in term of multiple solutions search), and one in CUDA that focuses on efficiently obtaining a single optimal result on GPU.
 We then provide some technical details explaining how these ideas are put in practice in DynaProg (\S\ref{implementation}).
-Finally, we see in our benchmarks that generated code is comparable with existing implementations (\S\ref{benchmarks}).
+Finally, we see in the benchmarks (\S\ref{benchmarks}) that generated code performance is comparable with existing implementations.
 
-From the benchmarks, we see that our code generation is able to deal with problems from simple (Smith-Waterman) to complex (Zuker and RNAfold) and provide performance that is comparable to existing work.
-Since these problem are expressed with a grammar and algebra, their expression is simplified, and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
+DynaProg is able to deal with problems from simple (Smith-Waterman) to complex (Zuker and RNAfold) that are expressed with a grammar and algebra.
+With this formalism, their expression is simplified, and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
 
-From a larger perspective, our work only address the subset of dynamic programming problems on sequences, there exist different dynamic programming problems that iteratively refine an approximation to converge to the solution (for example Bellman-Ford\footnote{\url{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}}) these algorithm have completely different access patterns but might also be parallelized.
+From a larger perspective, our work only addresses the subset of dynamic programming on sequences, but there exist different types of dynamic programming problems that for instance iteratively refine an approximation to converge to the solution (for example Bellman-Ford\footnote{\url{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}}) these algorithm have completely different access patterns but might also be parallelized.
 From this initial observation emerges a pattern for addressing problems:\ol
-\item Understand the gist of the problem and generalize its specific characteristics (we might call it domain knowledge)
-\item Equipped with that information, it is possible to devise an efficient implementation, possibly parallel\footnote{Because the single core (processor) model has reached its limit and the industry is moving towards multi-core and many-core.}, and encompassing as many problems as possible, given the trade-off between generality and specific information required to maintaing high performance.
+\item Understand the gist of the problem and generalize its specific characteristics (this can be called domain-specific knowledge)
+\item Equipped with that information, it is possible to devise an efficient implementation, possibly parallel\footnote{Because the single core (processor) model has reached its limit and the industry is moving towards multi-core and many-core.}, and encompassing as many problems as possible, given the trade-off between generality and specific information required to maintain high performance.
 With the current compiler technology and hardware architectures, it has become easier to express such solution with code generation: a specific problem is optimally addressed, while the generality of the technique is preserved within the code generator.
@@
 To do that, everybody must agree on a <<standard>>.
-Instead of creating new languages, it would be best if everybody would speak the same.
+Instead of creating a new dialect, it would be better that everybody speaks the same language.
 This is the very purpose of embedded DSL: people knowing how to use the host language will easily understand how to encode their problem.
@@
 From our perspective it seemed that Scala and LMS represents the ideal candidate as a host language.
-Our work only provide one small construction brick in the huge space of the problems\footnote{For example \url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.pdf}} that remain to be solved...
+Our work only provides one small construction brick in the huge space of the problems\footnote{For example \url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.pdf}} that remain to be solved...
 

### 2013-01-24 docs/report/conclusion.tex
Nits in paper.

@@
 \section{Conclusion}
-This Master project report is focused on how to solve efficiently dynamic programming.
+This Master project report focuses on how to solve efficiently dynamic programming.
 By restricting to the class of problem involving sequences, we were able to extract generic patterns and expose them to parallel architectures like CUDA.
@@
 We then provide some technical details explaining how these ideas are put in practice in DynaProg (\S\ref{implementation}).
-Finally, we see in the benchmarks (\S\ref{benchmarks}) that generated code performance is comparable with existing implementations.
+Finally, we see in the benchmarks (\S\ref{benchmarks}) that generated code performances are comparable with existing implementations.
 
-DynaProg is able to deal with problems from simple (Smith-Waterman) to complex (Zuker and RNAfold) that are expressed with a grammar and algebra.
-With this formalism, their expression is simplified, and the possibility to exploit the dynamic programming results is leveraged through the use of multiple grammars.
+From a larger perspective, we can describe our approach as a generic pattern to address domain-specific problems:\ol
+\item Understand the gist of the problem and generalize its specific characteristics to similar problems (this can be called domain-specific knowledge)
+\item Based on this information, it is possible to devise an efficient implementation, possibly parallel\footnote{Because the single core (processor) model has reached its limit and the industry is moving towards multi-core and many-core.} that encompasses as many problems as possible, balancing between generality and specific information required to maintain high performance.
+\item Once an efficient solver is implemented, it should be offered to other people who are trying to solve similar problems.
+To do that, a <<standard>> should be agreed on: instead of creating a new dialect, it would be better that everybody speaks the same language.
+This is the very purpose of embedded DSL: people who know the host language easily understand how to encode their problem.
+\ole
+With the recent compiler technology, expressing such solution with code generation has become easier.
+The code generation approach is efficient because it allows to optimally address one specific problem for a particular hardware architecture, while preserving the generality in the code generator.
+In this perspective, Scala and LMS are an interesting host language.
+Our work provide one additional construction brick to address the space of the problems\footnote{For example \url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.pdf}}.
 
-From a larger perspective, our work only addresses the subset of dynamic programming on sequences, but there exist different types of dynamic programming problems that for instance iteratively refine an approximation to converge to the solution (for example Bellman-Ford\footnote{\url{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}}) these algorithm have completely different access patterns but might also be parallelized.
-From this initial observation emerges a pattern for addressing problems:\ol
-\item Understand the gist of the problem and generalize its specific characteristics (this can be called domain-specific knowledge)
-\item Equipped with that information, it is possible to devise an efficient implementation, possibly parallel\footnote{Because the single core (processor) model has reached its limit and the industry is moving towards multi-core and many-core.}, and encompassing as many problems as possible, given the trade-off between generality and specific information required to maintain high performance.
-With the current compiler technology and hardware architectures, it has become easier to express such solution with code generation: a specific problem is optimally addressed, while the generality of the technique is preserved within the code generator.
-\item Finally, once an efficient solver has been implemented, it should be provided to other people trying to solve the problems in the same category.
-To do that, everybody must agree on a <<standard>>.
-Instead of creating a new dialect, it would be better that everybody speaks the same language.
-This is the very purpose of embedded DSL: people knowing how to use the host language will easily understand how to encode their problem.
-\ole
-
-From our perspective it seemed that Scala and LMS represents the ideal candidate as a host language.
-Our work only provides one small construction brick in the huge space of the problems\footnote{For example \url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.pdf}} that remain to be solved...
+We have demonstrated that our implementation, DynaProg, is able to deal with simple (Smith-Waterman) and complex (Zuker and RNAfold) dynamic programming problems on sequences that are expressed by a grammar and an algebra.
+With this formalism, their expression is simplified and exploiting the dynamic programming results becomes possible thanks to systematic backtrack encoding that allows traces to be exchanged between algebrae.
+This implementation is publicly available\footnote{\url{https://github.com/manojo/lamp-dp-mt/}} and can immediately be used in Scala projects.
 

### 2013-01-24 docs/report/introduction.tex
Nits in paper.

@@
-\subsection*{Abstract}
+\section*{Abstract}
 Dynamic programming is an algorithmic technique to solve problems that follow the Bellman's principle\cite{bellman_principle}: optimal solutions depends on optimal sub-problem solutions.
@@
 Solving a dynamic programming problem consists of two phases: filling one or more matrices with intermediate solutions for sub-problems and recomposing how the final result was constructed (backtracking).
-In the textbooks, problems are usually described in terms of recurrence relations between matrices elements.
+In textbooks, problems are usually described in terms of recurrence relations between matrices elements.
 Expressing dynamic programming problems in terms of recursive formulae involving matrix indices might be difficult, if often error prone, and the notation does not capture the essence of the underlying problem (for example aligning two sequences).
-Moreover, writing correct and efficient parallel implementation require different competencies and potentially a significant amount of time.
+Moreover, writing correct and efficient parallel implementation requires different competencies and often a significant amount of time.
 
@@
 We evaluate the performance of our implementation against existing work and our own hand-optimized baseline implementations for both the CPU and GPU versions.
-Experimental results show that plain Scala has a large overhead and can only be used for small sequences ($\le 1024$) whereas the generated GPU version is comparable with existing implementations: matrix chain multiplication has the same performance as our hand-optimized version (142\% of the execution time of \cite{gpu_atlp}) for a sequence of 4096 matrices, Smith-Waterman is twice slower than \cite{swat_linear} on a pair of sequences of 6144 elements, and  RNA folding is on par with \cite{gpu_rnafold} (95\% running time) for sequences of 4096 elements.
+Experimental results show that plain Scala has a large overhead and is recommended to be used with small sequences ($\le 1024$) whereas the generated GPU version is comparable with existing implementations: matrix chain multiplication has the same performance as our hand-optimized version (142\% of the execution time of \cite{gpu_atlp}) for a sequence of 4096 matrices, Smith-Waterman is twice slower than \cite{swat_linear} on a pair of sequences of 6144 elements, and  RNA folding is on par with \cite{gpu_rnafold} (95\% running time) for sequences of 4096 elements.
 
@@
 \vfill
+%\newpage
+\section*{Acknowledgement}
 This project has been achieved in collaboration with Manohar Jonnalagedda.
-I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advice and suggestions.
-I hope you will enjoy reading this report.
-\vspace{.3cm}\\
-\textit{Thierry Coppey}
+I also would like to thank the LAMP team, including Eugene Burmako, Sandro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions.
+% I hope you will enjoy reading this report.
+%\vspace{.3cm}\\
+%\textit{Thierry Coppey}
 
@@
 For example, we might want to multiply a chain of matrices\footnote{Such that matrices have appropriate dimensions to be multiplied with each other.} efficiently.
-The order in which matrices are combined changes the number of required scalar multiplications, we would want to find an optimal order that minimizes this number.
-Notice that if we knew how to split the chain into two subparts, we could recursively find an optimal order in these two parts and recombine them.
+The order in which matrices are combined changes the number of required scalar multiplications, therefore we would want to find an optimal order that minimizes this number.
+Notice that if we know how to split the chain into two subparts, we can recursively find an optimal order in these two parts and recombine them.
 Such combinatorial problems verify the Bellman's principle\cite{bellman_principle}: \textit{<<optimal solutions depends on optimal solutions of sub-problems>>}\footnote{\url{http://en.wikipedia.org/wiki/Bellman\_equation\#Bellman.27s\_Principle\_of\_Optimality}}.
@@
 
-Dynamic programming problems are usually expressed in terms of recurrences on intermediate solutions that are stored in matrices, whereas optimality is defined in terms of an objective function (min/max cost, ...).
+Dynamic programming problems are usually expressed in terms of recurrences on intermediate solutions that are stored in matrices, whereas optimality is defined in terms of an objective function (minimal or maximal cost, ...).
 In the case of a chain of $n$ matrices, the recurrence is:
@@
 
-Dynamic programming problems arise in several disciplines of applied Computer Science such as biosequence analysis, natural language processing and operational research: sequence alignment, RNA sequence folding or expression parenthesisation are examples of such problems.
-Unfortunately, these often appear in multiple variations and with a considerable degree of sophistication such that there is a mismatch between the textbook solution and its concrete implementation.
+In several disciplines of applied Computer Science, for example, biosequence analysis, natural language processing and operational research,  dynamic programming problems such as sequence alignment, RNA sequence folding or expression parenthesisation could arise.
+Unfortunately, these often appear in multiple variations with a considerable degree of sophistication such that there is a mismatch between the textbook solution and its concrete implementation.
 The user is often interested in one optimal solution, but he might also request \textit{all} co-optimal solutions, a fixed number of near-optimal solutions, or some synthetic properties of the search space (size, sum of scores, ...).
 
-The backtracking is usually ad-hoc because it needs both to be kept consistent with matrix filling and present the information in a format suitable for the user (human readable or ready to drive further computations).
-Additionally, debugging matrix indices is tedious and requires a lot of time, and little changes in the formulae might imply large rewrites of the matrices and recurrences \cite{gapc_yield}.
+The backtracking is usually ad-hoc because it needs to be kept consistent with matrix filling and presented in a format suitable for the user (readable by human or ready to drive further computations).
+Additionally, debugging matrix indices is tedious and requires a lot of time, and small changes in the formulae might imply large rewrites of the matrices and recurrences \cite{gapc_yield}.
 
-Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU or programmable hardware (FPGA).
-However, a domain specialist who writes the recurrences might not be very familiar with these platforms, whereas parallelization and hardware experts might not deeply understand the domain of the dynamic programming recurrences.
+Finally, once the implementation is correct, it is possible to turn it into an efficient implementation for specific architectures such as multi-CPU, GPU and programmable hardware (FPGA).
+However, domain specialists who write the recurrences might not be very familiar with these platforms, and parallelization or hardware experts might not deeply understand the domain of the dynamic programming recurrences.
 
 \newpage
-To simplify the expression of dynamic problems, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separate the concerns of dynamic programming on sequences into four distinct components that are tightly connected:\ol
+To simplify the expression of dynamic problems, Algebraic Dynamic Programming (ADP) \cite{adp} proposes a language-independent declarative approach that separates the concerns of dynamic programming on sequences into four distinct components that are tightly connected:\ol
 \item The search space is described by a context-free \textbf{parsing grammar} that produces intermediate solution candidates whose score might be inserted in the matrix.
@@
 
-A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, making possible that the same grammar share different algebrae or vice versa.
+A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, which allows the user to combine one grammar with different algebrae or vice versa.
 Because recurrence relations are expressed by a parsing grammar, ADP makes the candidate structure explicit and hides tabulation indices, thereby preventing potential errors.
@@
 DynaProg, the DSL we present in this report, implements the concepts of ADP in Scala as an embedded DSL (domain-specific language) with a syntax similar to the combinators parsers of Scala library\footnote{See \url{http://www.scala-lang.org/api/current/index.html\#scala.util.parsing.combinator.Parsers}}.
-It extends ADP by allowing grammars for pairing two sequences (multi-track grammars) similarly as GAPC\cite{gapc_thesis}, simplifies the process of writing programs by inferring additional information (\S\ref{yield_analysis}) and can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart (\S\ref{benchmarks}).
-Since the program structure is formalized in ADP framework, it can be analyzed to remove unused grammar rules (\S\ref{dead_rules}) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, thereby producing an arguably more reliable program.
+It extends ADP by allowing grammars for pairing two sequences (multi-track grammars) similarly as GAPC\cite{gapc_thesis}.
+It also simplifies the process of writing programs by inferring additional information (\S\ref{yield_analysis}).
+Moreover, it can translate them into efficient CUDA\footnote{Compute Unified Device Architecture: a parallel computing platform and programming model created by NVIDIA, supported by their graphics processing units (GPUs).} program that are competitive to their handwritten counterpart (\S\ref{benchmarks}).
+Since the program structure is formalized in ADP framework, it can be analyzed to remove unused grammar rules (\S\ref{dead_rules}) and avoid some non-termination issues; since it is generated, correct scheduling is guaranteed and indices errors are avoided, thereby it produces an arguably more reliable program.
 
 DynaProg provides a generic way of backtracking the results such that the same trace can be used with different algebras sharing the same grammar.
-This allows to construct a two step pattern for solving problems: first the DP problem is solved using the appropriate cost functions; then from the backtrack of its optimal, the desired result is computed.
+This allows constructing a two step pattern to solve problems: first the DP problem is solved using the appropriate cost functions; then from the backtrack of its optimum, the desired result is computed.
 As example, consider multiplying a chain\footnote{Assuming matrices are of appropriated dimension to be multiplied with each other} of matrices efficiently: first, optimal execution scheduling (or parenthesisation) trace is found using dynamic programming and cost algebra (\S\ref{mat_mult_plain}).
@@
 This project is currently available online\footnote{\url{https://github.com/manojo/lamp-dp-mt}}; it implements dynamic programming parsers in Scala (CPU) and CUDA (GPU).
-Its contribution is an novel approach to systematically encode and process backtracking information such that the reconstruction complexity is reduced compared to \cite{gapc_thesis}, and backtrack trace can be exchanged among different algebras sharing the same grammar.
+Its contribution is an novel approach to systematically encode and process backtracking information such that the reconstruction complexity is reduced compared to GAPC\cite{gapc_thesis}, and backtrack trace can be exchanged among different algebras sharing the same grammar.
 

### 2013-01-25 docs/report/benchmarks.tex
Final report.

@@
 However, since two of these matrices only propagate information along line or column, it is possible to encode this information in a wavefront (see \S\ref{calc_simplifications}) instead of maintaining it in a memory-expensive matrix.
-This knowledge is leveraged by \cite{ATLP}, however, in our case, we cannot describe the wavefront in the grammar and need to maintain explicitly 3 matrices, thereby multiplying by 3 the memory usage\footnote{Actually slightly less than 3 because we replace 2 matrices in $O(n^2)$ by two vectors in $O(n)$}.
+This knowledge is leveraged by \cite{gpu_atlp}, however, in our case, we cannot describe the wavefront in the grammar and need to maintain explicitly 3 matrices, thereby multiplying by 3 the memory usage\footnote{Actually slightly less than 3 because we replace 2 matrices in $O(n^2)$ by two vectors in $O(n)$}.
 

### 2013-01-28 bench/cudalign/src/common/Job.hpp
Extenral benchmarks + slides.

@@ -1,0 +1,101 @@
+/*
+ * File:   Job.hpp
+ * Author: edans
+ *
+ * Created on April 8, 2010, 11:00 PM
+ */
+class Job;
+
+#ifndef _JOB_HPP
+#define	_JOB_HPP
+
+#include <string>
+#include <vector>
+using namespace std;
+
+#include "Sequence.hpp"
+#include "SpecialRowReader.hpp"
+#include "SpecialRowWriter.hpp"
+#include "MidpointsFile.hpp"
+#include "Alignment.hpp"
+#include "buffer/Buffer.hpp"
+
+#define STAGE_1   (1)
+#define STAGE_2   (2)
+#define STAGE_3   (3)
+#define STAGE_4   (4)
+#define STAGE_5   (5)
+#define STAGE_6   (6)
+
+class Job {
+    public:
+        int blocks;
+        
+        Sequence seq[2];
+
+        vector<SpecialRowReader*>* special_lines1;
+        vector<SpecialRowReader*>* special_lines2;
+		
+        vector<midpoint_t> midpoints;
+
+        long long flush_limit;
+        int flush_interval;
+		
+		string flush_column_url;
+		string load_column_url;
+		
+		int gpu;
+		
+		long long cells_updates;
+
+        int stage4_maximum_partition_size;
+		bool stage4_orthogonal_execution;
+
+		int stage6_output_format;
+		Alignment* alignment;
+		
+        string alignment_binary_filename;
+		string alignment_text_filename;
+		string alignment_svg_filename;
+		
+        Job();
+        Job ( const Job& orig );
+        virtual ~Job();
+
+		void createPath(string path);
+        void setWorkPath ( string workPath );
+        string getWorkPath ( );
+        int initialize();
+
+		Buffer* loadBufferFromURL(string url);
+		
+        void loadSpecialRows ( int stage_id );
+        void loadMidpoints ( int id );
+        void writeMidpoints ( int id = -1 );
+        int getLargestMidpointSize ( int* max_i = NULL, int* max_j = NULL );
+		int getLargestSpecialRowInterval ( int stage_id, int* max_i = NULL, int* max_j = NULL );
+
+		SpecialRowWriter* fopenNewSpecialRow  ( int stage_id, int row, int col );
+		SpecialRowReader* fopenNextSpecialRow ( int stage_id, int row, int col, int min_dist, int *id );
+		FILE* fopenSpecialRow ( int stage_id, int row, int col, const char* mode );
+        FILE* fopenFirstMidpoint();
[…]

### 2013-01-28 bench/cudalign/src/common/SpecialRowWriter.hpp

@@ -1,0 +1,38 @@
+/* 
+ * File:   SpecialRowWriter.hpp
+ * Author: edans
+ *
+ * Created on April 9, 2010, 11:59 PM
+ */
+
+#ifndef _SPECIALROWWRITER_HPP
+#define	_SPECIALROWWRITER_HPP
+
+#include <stdio.h>
+#include <string>
+using namespace std;
+
+class SpecialRowWriter {
+public:
+    SpecialRowWriter(string directory, int row, int col);
+    SpecialRowWriter(const SpecialRowWriter& orig);
+    virtual ~SpecialRowWriter();
+
+    int getCol() const;
+    int getRow() const;
+
+    void open();
+    void close();
+    void cancel();
+
+    int write(void* buf, int len);
+private:
+    int row;
+    int col;
+    string directory;
+    FILE* file;
+    int element_size;
+};
+
+#endif	/* _SPECIALROWWRITER_HPP */
+

### 2013-01-28 bench/cudalign/src/common/blosum/blosum50.h

@@ -1,0 +1,40 @@
+/******************************************************************************/
+/* Código gerado automaticamente pelo blosum2c.pl                             */
+/* Arquivos BLOSUMxxx: ftp://ftp.ncbi.nih.gov/blast/matrices/                 */
+/******************************************************************************/
+/* Matrix made by matblas from blosum50.iij                                   */
+/* * column uses minimum score                                                */
+/* BLOSUM Clustered Scoring Matrix in 1/3 Bit Units                           */
+/* Blocks Database = /data/blocks_5.0/blocks.dat                              */
+/* Cluster Percentage: >= 50                                                  */
+/* Entropy =   0.4808, Expected =  -0.3573                                    */
+/******************************************************************************/
+static char *order_blosum50 = "ARNDCQEGHILKMFPSTWYVBZX*";
+static const char score_blosum50[24][24] = {
+    // A  R  N  D  C  Q  E  G  H  I  L  K  M  F  P  S  T  W  Y  V  B  Z  X  * 
+/*A*/{ 5,-2,-1,-2,-1,-1,-1, 0,-2,-1,-2,-1,-1,-3,-1, 1, 0,-3,-2, 0,-2,-1,-1,-5},
+/*R*/{-2, 7,-1,-2,-4, 1, 0,-3, 0,-4,-3, 3,-2,-3,-3,-1,-1,-3,-1,-3,-1, 0,-1,-5},
[…]

### 2013-01-28 bench/cudalign/src/common/blosum/blosum62.h

@@ -1,0 +1,40 @@
+/******************************************************************************/
+/* Código gerado automaticamente pelo blosum2c.pl                             */
+/* Arquivos BLOSUMxxx: ftp://ftp.ncbi.nih.gov/blast/matrices/                 */
+/******************************************************************************/
+/* Matrix made by matblas from blosum62.iij                                   */
+/* * column uses minimum score                                                */
+/* BLOSUM Clustered Scoring Matrix in 1/2 Bit Units                           */
+/* Blocks Database = /data/blocks_5.0/blocks.dat                              */
+/* Cluster Percentage: >= 62                                                  */
+/* Entropy =   0.6979, Expected =  -0.5209                                    */
+/******************************************************************************/
+static char *order_blosum62 = "ARNDCQEGHILKMFPSTWYVBZX*";
+static const char score_blosum62[24][24] = {
+    // A  R  N  D  C  Q  E  G  H  I  L  K  M  F  P  S  T  W  Y  V  B  Z  X  * 
+/*A*/{ 4,-1,-2,-2, 0,-1,-1, 0,-2,-1,-1,-1,-1,-2,-1, 1, 0,-3,-2, 0,-2,-1, 0,-4},
+/*R*/{-1, 5, 0,-2,-3, 1, 0,-2, 0,-3,-2, 2,-1,-3,-2,-1,-1,-3,-2,-3,-1, 0,-1,-4},
[…]

### 2013-01-28 bench/cudalign/src/common/buffer/SocketBuffer.hpp

@@ -1,0 +1,26 @@
+#ifndef SOCKETBUFFER_H
+#define SOCKETBUFFER_H
+
+#include "Buffer.hpp"
+#include <string>
+using namespace std;
+
+class SocketBuffer : public Buffer
+{
+    public:
+        SocketBuffer(string hostname, int port);
+        virtual ~SocketBuffer();
+        
+        virtual void initAutoFlush();
+        virtual void initAutoLoad();
+        
+        virtual void autoFlushThread();
+        virtual void autoLoadThread();
+        
+    private:
+        string hostname;
+        int port;
+        int socketfd;
+};
+
+#endif // SOCKETBUFFER_H

### 2013-01-28 bench/cudalign/src/common/buffer/teste.cpp

@@ -1,0 +1,104 @@
+#include <pthread.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+
+
+#include "Buffer.h"
+#include "PipedBuffer.h"
+#include "CircularBuffer.h"
+
+Buffer* b = NULL;
+
+
+pthread_mutex_t fakeMutex = PTHREAD_MUTEX_INITIALIZER;
+pthread_cond_t fakeCond = PTHREAD_COND_INITIALIZER;
+
+void mywait(int timeInSec)
+{
+	struct timespec timeToWait;
+	struct timeval now;
+	int rt;
+	
+	gettimeofday(&now,NULL);
+	
+	timeToWait.tv_sec = now.tv_sec + timeInSec;
+	timeToWait.tv_nsec = now.tv_usec*1000;
+	
+	pthread_mutex_lock(&fakeMutex);
+	rt = pthread_cond_timedwait(&fakeCond, &fakeMutex, &timeToWait);
+	pthread_mutex_unlock(&fakeMutex);
+	//printf("Done\n");
+}
+
+
+
+void *writeThread(void *arg)
+{
+	Buffer* b = (Buffer*)arg;
+	const int c = 16;
+	char buf[c];
+	for (int i=0; i<c; i++) {
+		buf[i] = i%100;
+	}
+	
+	for (int j=0; j<1000; j++) {
+		int rc = b->writeBuffer(buf, c);
+		printf ("WRITE(%d): %d\n", j*c, rc);
+	}
+	
+	pthread_exit(NULL);
+}
+
+void *readThread(void *arg)
+{
+	Buffer* b = (Buffer*)arg;
+	const int c = 11;
+	char buf[c];
+	
+	int k=0;
+	
+	for (int j=0; j<1000; j++) {
+		int rc = b->readBuffer(buf, c);
+		printf ("READ(%d): %d\n", j*c, rc);
+		for (int i=0; i<c; i++) {
+			if (buf[i]!=(k%16)%100) {
+				printf("%d %d = %s\n", buf[i], (k%16)%100, "ERRO");
+			} else {
+				//printf("%d %d = %s\n", buf[i], (k%16)%100, ".");
+			}
+			k++;
+		}
+		printf("\n");
+		mywait(1);
+	}
+	
+	pthread_exit(NULL);
+}
+
+int main (int argc, char *argv[])
+{
+	Buffer* b = new CircularBuffer();
+	pthread_t threads[2];
+	int rc;
+	long t;
+	
[…]

### 2013-01-28 bench/cudalign/src/stage3/sw_stage3.h

@@ -1,0 +1,5 @@
+#pragma once
+
+#include "../common/cudalign.hpp"
+
+void stage3(Job* job);

### 2013-01-28 bench/cudalign/src/stage4/sw_stage4.cu

@@ -1,0 +1,680 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <unistd.h>
+#include <string.h>
+
+#include <pthread.h>
+#include <sys/stat.h>
+#include <sys/types.h>
+
+#include "../common/cudalign.hpp"
+
+#include <cutil_inline.h>
+
+#define H_MAX (2*64*1024)
+
+#define NUM_THREADS 2
+
+#define PRINT (0)
+
+
+typedef struct {
+    int partition0;
+    int partition1;
+
+    int h0[H_MAX];
+    int e0[H_MAX];
+
+    int h1[H_MAX];
+    int e1[H_MAX];
+
+    Job* job;
+    /* OUT */
+    midpoint_t *out_pos;
+} split_args_t;
+
+static midpoint_t split(Sequence *seq0, Sequence *seq1, int i0, int j0, int i1, int j1, 
+						int type_s, int type_e, int score_s, int score_e,
+						int *h0, int *h1, int *e0, int *e1);
+static midpoint_t ort_split(Sequence *seq0, Sequence *seq1, int i0, int j0, int i1, int j1, 
+						int type_s, int type_e, int score_s, int score_e,
+						int *h0, int *h1, int *e0, int *e1);
+												
+static void *split_thread(void *thread_arg) {
+    static int inv_type[] = {0,2,1};
+
+    split_args_t* args = (split_args_t*)thread_arg;
+
+    Job* job = args->job;
+
+    int i0, j0, i1, j1, start_type, prev_type, start_score, prev_score;
+
+    int partition_id = args->partition0;
+    i1 = job->midpoints[partition_id].i;
+    j1 = job->midpoints[partition_id].j;
+    prev_type = job->midpoints[partition_id].type;
+	prev_score = job->midpoints[partition_id].score;
+
+    float last_percent = 0;
+    for (int k=args->partition0+1; k<args->partition1; k++) {
+        float percent = 100.0f*(k-args->partition0)/(args->partition1-args->partition0);
+        if (percent > last_percent+1) {
[…]

### 2013-01-28 bench/cudalign/src/stage4/sw_stage4.h

@@ -1,0 +1,1 @@
+void stage4(Job* job);

### 2013-01-28 bench/rnafold/cpu/adplib.h

@@ -1,0 +1,218 @@
+#ifndef adplib_h
+#define adplib_h
+
+/* abstraction for poor people - after including, undef and redef them
+ * if you use tupels */
+
+#define get_result_score(a) a
+
+#define is_suboptimal(a, b, c) abs(a - b) <= c
+
+/* Always needed
+  ====================================================================== */
+
+#define min(A, B) ((A) < (B) ? (A) : (B))
+#define max(A, B) ((A) > (B) ? (A) : (B))
+
+/* Input handling
+  ====================================================================== */
+
+void convert_input(int start, char *z, int n);
+
+/* Correct incomplete phases in adpc
+  ====================================================================== */
+
+#define decode(X) ((X)-'0')
+
+
+/* Memory handling
+  ====================================================================== */
+
+
+/* wrappers for standard C functions
+   ------------------------------------------- */
+
+void memerr_exit(char *f);
+void *mcalloc(size_t nobj, size_t size);
+void *mmalloc(size_t size);
+void *mrealloc(void *q, size_t size);
+
+
+/* Memory management
+   ------------------------------------------- */
+
+typedef struct {
+  char **address;
+  int currentBlock;
+  int currentPos;
+  int blockSize;
+  int numberOfBlocks;
+} tmemory;
+
+tmemory *adp_statmem;
+tmemory *adp_dynmem;
+
+void      set_adplib_debug(int debug);
+void     *myalloc(tmemory *mem, int size);
+tmemory  *memory_new();
+void      memory_clear(tmemory *mem);
+void      memory_free(tmemory *mem);
+
+/* Preprocessing tools
+  ====================================================================== */
+
+char arr_iupac_base[128][5];
+#define iupac_base(A,B) arr_iupac_base[A][B]
+char *calc_contains_region(char *z, int n, int *offset, char *pat1);
+
+/* String tools
+  ====================================================================== */
+
+char *mkstr(char *s);
+
+#define dots(i,j) libPP_repeat(i,j,'.')
+
+char *libPP_repeat(int i, int j, char c);
+
+/* File input
+  ====================================================================== */
+
+/* A single sequence
+   ------------------------------------------- */
+
+typedef struct {
+  char success;
+  char *descr;
+  char *seq;
+  int  length;
+  char *original_seq;  /* backup for window mode */
+  int   original_length;
[…]

### 2013-01-28 bench/rnafold/cuda/RNAfold-man.c

@@ -1,0 +1,62 @@
+if (manoptmode == '-')
+switch (manopt) {
+  case 'h':
+    printf("-h Display this information\n\nOption -h displays a command option overview.\n");
+    break;
+  case 'H':
+    printf("-H <option> Display detailed information on <option>\n\nThis displays the corresponding section of the RNAfold manual for the\ngiven command line option.\n");
+    break;
+  case 'v':
+    printf("-v Show version\n\nThis shows the version number of RNAfold.\n");
+    break;
+  case 'b':
+    printf("-b <value> Set number of blocks\n\nThis switch sets the number of CUDA-blocks.\n");
+    break;
+  case 'T':
+    printf("-T <value> Set number of threads per blocks\n\nThis switch sets the number of threads per CUDA-block.\n");
+    break;
+  case 'e':
+    printf("-e <value> Set energy range (kcal/mol)\n\nThis sets the energy range for suboptimal results. value is the\ndifference to the minimum free energy for the sequence.\n");
+    break;
+  case 'c':
+    printf("-c <value> Set energy range (%%%%)\n\nThis sets the energy range as percentage value of the minimum free\nenergy. For example, when -c 10 is specified, and the minimum free\nenergy is -10.0 kcal/mol, the energy range is set to -9.0 to -10.0\nkcal/mol.\n");
+    break;
+  case 'M':
+    printf("-M <value> Set maximal loop length\n\nThis option sets the maximum lengths of the considered internal and\nbulge loops. The default value here is 30. For unrestricted loop\nlengths, use option -M n. This will increase calculation times and\nmemory requirements.\n");
+    break;
+  case 'w':
+    printf("-w <value> Set window size\n\nBeginning with position 1 of the input sequence, the analysis is\nrepeatedly processed on subsequences of the specified size. After each\ncalculation, the results are printed out and the window is moved by the\nwindow position increment (-W), until the end of the input sequence is\nreached.\n");
+    break;
+  case 'W':
+    printf("-W <value> Set window position increment\n\nThis option specifies the increment for the window analysis mode (-w).\n");
+    break;
+  case 'B':
+    printf("-B Show best results in window mode\n\nThis option specifies, that only the best results are shown in window\nmode.\n");
+    break;
+  case 'S':
[…]

### 2013-01-28 bench/rnafold/cuda/cudalib.cu

@@ -1,0 +1,2327 @@
+
+#include <unistd.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <math.h>
+#include <errno.h>
+
+#include "config.h"
+#include "options.h"
+
+
+#include <cutil.h>
+
+#define is_suboptimal(a, b, c) abs(a - b) <= c
+
+FILE *energyFile;
+
+#define ENERGYFILE "energies.dat"
+
+// #define DIFF3
+// #define SHARED_Z // 2000
+
+// #define SHARED_ENERGY // 8000
+// #define SHARED_OFFSET  // 8000
+
+// #define SPARSE_CLOSED
+// #define MAX_DIFF 200
+
+#define SHMEM 1000
+
+/* Always needed
+  ====================================================================== */
+
+#define min(A, B) ((A) < (B) ? (A) : (B))
+#define max(A, B) ((A) > (B) ? (A) : (B))
+
+/* Input handling
+  ====================================================================== */
+
+void convert_input(int start, char *z, int n);
+
+
+/* Correct incomplete phases in adpc
+  ====================================================================== */
+
+#define decode(X) ((X)-'0')
+
+
+/* Memory handling
+  ====================================================================== */
+
+
+/* wrappers for standard C functions
+   ------------------------------------------- */
+
+void memerr_exit(char *f);
+void *mcalloc(size_t nobj, size_t size);
+void *mmalloc(size_t size);
+void *mrealloc(void *q, size_t size);
+
+
+/* Memory management
+   ------------------------------------------- */
+
+typedef struct {
+  char **address;
+  int currentBlock;
+  int currentPos;
+  int blockSize;
+  int numberOfBlocks;
+} tmemory;
+
+tmemory *adp_statmem;
+tmemory *adp_dynmem;
+
+void      set_adplib_debug(int debug);
+
+void     *myalloc(tmemory *mem, int size);
+
+tmemory  *memory_new();
+void      memory_clear(tmemory *mem);
+void      memory_free(tmemory *mem);
+
+/* Preprocessing tools
+  ====================================================================== */
+
+char arr_iupac_base[128][5];
+#define iupac_base(A,B) arr_iupac_base[A][B]
+char *calc_contains_region(char *z, int n, int *offset, char *pat1);
+
+/* String tools
+  ====================================================================== */
+
[…]

### 2013-01-29 docs/slides/res/trans.tex
Up to 2x faster than RNAfold-RL (1.77x on average) with optimization in LibRNA.

--- 2013-01-16 docs/report/design.tex
@@
-\newpage
-\section{Architecture design and technical decisions} \label{architecture}
-% ------------------------------------------------------------------------------------------------
-\subsection{User facing language requirements} \label{user_lang}
-The field of dynamic programming has been influenced in the recent years by a methodology known as Algebraic Dynamic Programming which uses a grammar and an algebra to separate between the parsing and the score computation:
-\quote{The Algebraic Dynamic Programming approach (ADP) introduces a conceptual splitting of a DP algorithm into a recognition and an evaluation phase.
-The evaluation phase is specified by an evaluation algebra, the recognition phase by a yield grammar.
-Each grammar can be combined with a variety of algebras to solve different but related problems, for which heretofore DP recurrences had to be developed independently.
-Grammar and algebra together describe a DP algorithm on a high level of abstraction, supporting the development of ideas and the comparison of algorithms.}
+\documentclass[11pt]{article}
+\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs}
+\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage[english]{babel}
+\pagestyle{empty} % remove page numbers
+\oddsidemargin -2.5cm
+\evensidemargin -2.5cm
+\topmargin -4cm
+\headsep 1cm
+\parskip 0cm
+\parindent 0cm
+\setlength{\paperwidth}{14cm}
+\setlength{\paperheight}{6cm}
+\begin{document}
 
-Given such formalization \cite{adp} of dynamic programming on sequences, it seems natural to borrow from it and extend it to other types of DP problems.
-In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only some candidates for further combination.
-
-The benefits of the ADP framework is that it does not constrains the result of the evaluation to be a single value, but can extend parsers to backtracking parsers or pretty-printers.
-Additionally, we want to support the following features:\ol
-%\item {\color{red}\textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic.
-To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.}
-\item \textbf{Input pair algebra:} the original ADP framework \cite{adp} only support single input, we want to support pairs of inputs sequences similarly as \cite{gapl} such that we can treat problem such as Smith Waterman or Needleman-Wunsch.
-As discussed in \S\ref{bg_dp}, handling more than two sequences introduces an $\Omega(n^3)$ storage complexity that might limit more severely the size of problems that could be addressed.
-Since the problems seen in \S\ref{problems} use either one or two input sequences, we only need to support these two cases.
-\item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$-best results.
-\item \textbf{Input restrictions:} since CUDA (and FPGA) cannot process arbitrary Scala objects, we need to restrict the language to primary types (int, float, ...
-and structures of them).
-However, we want to preserve the expressivity available in Scala and impose restrictions on the data types processed by CUDA.
-A typical restriction we want to make is that data elements are of fixed size, to avoid memory management issues and thread divergence\footnote{Occurs when a single thread needs more processing than its peers, thereby delaying the whole computation.}.
-\item \textbf{Single-result on devices:} The general ADP framework supports multiple solutions for intermediate results.
-Such functionality is easily supported in Scala; however, memory management hampers the performance of the GPU implementation (see \S\ref{related_work}).
-To overcome this issue, the user could manually manage the memory, but this would defeat most of the benefits of automatic code generation.
-Hence the trade-off solution we propose is to restrict ADP to only one optimal result on CUDA, while offering the possibility to obtain co-optimal (or even all possible solutions) with Scala.
-
-\item \textbf{Automatic backtracking:} To produce efficient code, we imposed a fixed size on the output generated by the parsers on devices.
-However, on the other hand, the backtracking information (of varying size) is of primary interest for the DSL user, hence we would like to to automate the backtracking to fulfill goals of usefulness, efficiency and ease-of-use in device-specific implementation:\ul
-	\item Leaving the backtrack implementation to the user would force him to memoize the backtracking information together with the results (backtrack would  grow towards final result and duplicate unnecessarily information), hence requiring both $O(n^3)$ space and memory management features on devices.
-	\item Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement.
-Collecting the backtracking list can be easily done in $O(n)$ and then reversed depending on whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
-	\ule
-\item \textbf{Yield analysis:} in vanilla ADP, the user has to define for each concatenation the minimal and maximal length of the subsequence on each side.
-Although non-emptiness information is necessary to avoid infinite recursions in the parsers, forcing an explicit definition can become cumbersome for the DSL user.
-Similarly as in \cite{gapc_yield}, we want to provide an automatic computation of concatenation boundaries, while at the same time leaving the possibility to manually specify it for maximum flexibility.
-\ole
-
-The support of these features has the following implications:\ul
-\item \textbf{Dependency analysis:} Since we target GPUs (and FPGAs) which are massively parallel architecture, a top-down execution using hash tables is impractical (fallback computation if element is not present is hard to parallelize), hence we need to construct the result tree bottom-up, therefore ensure that the (partial) evaluation order between rules is respected.
-\item \textbf{Normalization:} in order to automate the backtracking, we need the grammar rules to present a certain shape so that we can define uniquely the backtracking information (in particular we want to distinguish between alternatives).
-Also we need to maintain coherence between the Scala and the CUDA version so that they can inter-operate: we would like to reuse the backtracking information (from CUDA) to do actual processing in Scala (pretty-printing or actual computation as described in \S\ref{intro}).
-\item \textbf{Optimizations:} Since ADP exposes a grammar, we have the possibility to do grammar optimizations (see \S\ref{grammar_opt}).
-Also since the grammar might define useless rules, we might want to eliminate them: dead rule elimination is very similar to traditional dead code elimination (it reduces the size of generated code) but also reduces the memory consumption (as storage matrix does not need to be reserved) and even speeds up the computation, (since all grammar rules for a particular element are computed at the same time in CUDA implementation, see \ref{dead_rules}).
-\ule
-
-\subsubsection{Grammar optimization} \label{grammar_opt}
-Since ADP exposes a grammar, we might be able to break complex grammar rules into simpler ones (optimally binary production). For example, consider the following rule (in BNF):
-\[A := B\;\;C\;\;D \qquad \implies \qquad \left\{\begin{array}{l} A' := B\;\;X \\ X := C\;\;D \end{array} \right. \]
-If $B$, $C$ and $D$ are of varying length, evaluating the rule $A$ for a single subproblem is $O(n^2)$ (assuming B,C,D are stored in matrices hence $O(1)$).
-Adding a tabulation $X$ reduces the evaluation complexity because for each subproblem we will consider either rule $A'$ or $X$, each of evaluation complexity $O(n)$; hence we would have reduced the grammar evaluation complexity.
-
-Since grammar candidates are then evaluated with an algebra, we need to devise an equivalent transformation of the algebra.
-Unfortunately this analysis is very involved: we need to solve the following problem (to respect Bellman's optimality principle\cite{bellman_principle}):
-
-Given $f$ , find a pair of functions $(f_1,f_2)$ or $(f_3,f_4)$ such that\footnote{The first equation denotes breaking in two functions, the second is Bellman's optimality principle.}
-\[\begin{array}{rcll}
-f(i,k_1,k_2,j) &=& f_1(i,f_2(i,k_1,k_2),k_2,j) & \land \\
-	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_1(i,\min\limits_{i<k_1<k_2} \big[ (f_2(i,k_1,k_2  ) \big],k_2,j) \big] & \lor \vspace{12pt} \\
-f(i,k_1,k_2,j) &=& f_3(i,k_1,f_4(k_1,k_2,j),j) & \land \\
-	\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] &=& \min\limits_{i<k_2<j} \big[ f_3(i,k_1,\min\limits_{k_1<k_2<j} \big[ (f_4(k_1,k_2,j) \big],j) \big]
-\end{array}\]
-% Viktor Kuncak: We can solve this more easily if functions are linear
-Since this requires complex mathematical analysis that is out of the scope of the project, and since we have not found relevant literature on that particular subject, we leave this optimization to the responsibility of the user by assuming that the provided grammar is already optimal.
-
-% ------------------------------------------------------------------------------------------------
-\subsection{Recurrences analysis} \label{recurrences}
-In this section we use the following notation: let $A$ be a tabulation parser, we refer by $A_{(i,j)}$ to the element $i,j$ of the underlying matrix (in order to keep a lightweight notation).
-\subsubsection{Dead rules elimination} \label{dead_rules}
-Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are reachable from the axiom).
-The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules.
-Although seemingly useless for the Scala implementation, this step is necessary to maintain coherency between Scala and CUDA rules numbering (that happen in a later stage on the valid rules).
-In CUDA, this analysis not only provides dead code elimination, but it also prevents useless computation execution, since all rules are computed sequentially for a particular subsequence before the next subsequence is processed.
-
-\subsubsection{Yield analysis} \label{yield_analysis}
-Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved.
-Additionally the priority of operators varies in Scala, depending the operator's first character whereas it is possible to specify arbitrary priorities in Haskell.
-To overcome these issues, we propose to automate the computation of the minimum/maximum length of (subsequences) parsers.
-
-Parsers are made of terminals, concatenations, operations (aggregate, map, filter) and tabulations; minimum/maximum yield of terminals is set, hence it only remains to assign appropriate yield sizes to tabulations; other operations simply propagate that information.
-It is possible to obtain the yield size of tabulations using the following algorithm (assuming recursive parsers contain at least one tabulation that is part of the loop), similar to \cite{adp_c}:\ol
-\item Set the yield minimum size of all tabulations to a large number $M_0$ (such that all tabulation would reasonably have a minimum yield size smaller than $M_0$)
-\item Repeat $k$ times ($k$ is the number of rules of the grammar): for each rule, compute its minimum yield size and update its value (without recursion at tabulations).
-This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all rules.
-\item Set the maximal yield of all the rules to the minimal value.
-For each rule, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size.
-If the depth reaches the maximum $k$, there is a loop between tabulations, hence return infinity.
-\ole
-The last part of this algorithm has worst case exponential complexity, but if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$.
-Obtaining the yield size of tabulations provides the following benefits:\ul
-\item Minimum size: prevents self-reference parsers (on the same subsequence) and avoids considering subsequences which yield empty results (hence slightly reducing time complexity).
-\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(m\cdot n)$ instead of $O(n^2)$ (where $m$ is the maximum yield size), possibly providing substantial space savings.
-As the rules with bounded maximum yield are very rare, we did not implement this optimization, although we might consider it for future work.
-\ule
-
-\subsubsection{Dependency analysis}
-Let us introduce the concept of dependency (similar to $\to_{chain}$ in \cite{adp_c}): a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$.
-A grammar is unsolvable if there exists a dependency loop between parsers ($A\to ...
-\to A$).
-Such case only happen when there is no concatenation or a concatenation with an empty word.
-Being able to track the dependencies of tabulations and infer a computation order between them has two benefits:\ul
-\item Although seemingly unnecessary in a top-down approach (Scala), this analysis detects dependency loops which would result in infinite call loops (stack overflow) at execution.
-\item Ordering tabulations is critical in a bottom-up approach (CUDA) to make sure that all dependencies are valid before an element computation is triggered.
-\ule
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{Backtracking} \label{backtracking}
-In order to produce an efficient transformation from an ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules.
-To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial).
-To deal with this problem, we are facing two options:\ul
-\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack which is not implemented in ADP, unless we consider the whole backtrack being part of the scoring (which has a big performance impact and non-constant memory requirement issues that make such GPU implementation hard and not desirable).
-Additionally, since the backtracking data is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
-\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the sub-solutions combination that contribute to the problem solution.
-To do that we need to apply the following transformations\ol
-	\item Normalize rules and identify them uniquely by exploding alternatives: each rule is decomposed into the union of multiple sub-rules uniquely identified by an index, where sub-rules do not contain alternatives (Or parsers).
-Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
-	\item Let ${\rm cc}(r_s)$ be the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
-		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
-		\item The backtrack part is a tuple $(r_s,(k_1, k_2, \ldots , k_m))$ where $m$ is the maximal number of concatenations occurring in the enclosing rule of $r_s$; more formally $m=\max_z\big[{\rm cc}(r_z) | r_z \in {\rm rule}(r_s)\big]$, and let $m_s = {\rm cc}(r_s) \le m$.
-	\ule
-	\item During the matrix computation of cell $(i,j)$, if the sub-rule $r_s$ applies, the backtrack will be set as $(r_s,(k_1,k_2,\ldots,k_{m_s}))$; with $i\le k_1\le k_2\le \ldots k_{m_s}\le j$.
-Note that if the backtrack occupies a fixed-length memory, the backtrack will contain exactly $m$ indices, hence $k_i\,|\,m_s<i\le m$ will be unspecified.
-	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r_s,(k_1,k_2,\ldots,k_m))$, given $r_s$, we recover $s=T(r_s)$, the sub-rule that applies.
-Hence we can determine $m_s$, which allows us to enqueue the subsequences $(i,k_1), (k_1,k_2), ..., (k_{m_s},j)$ for recursive backtracking.
-If $s$ refers to a terminal, we stop the backtracking.
-	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $((i,j),r_s,(k_1,k_2,\ldots,k_{m_s}))$ where $(i,j)$ denotes the subsequence on which the sub-rule $T(r_s)$ has to be applied with concatenation indices $(k_1,k_2,\ldots,k_{m_s})$.
-\ole
-In short, we break parsers into normalized rules, the backtracking information is the subword, the sub-rule id (which rule to unfold) and a list of indices (how to unfold it).
-\ule
-In order to reduce the storage required by the backtracking indices, we can avoid storing fixed indices (where at least one of the two subsequences involved in a concatenation has a fixed size) and leverage the knowledge contained in $s$ to reconstruct the appropriate backtrack.
-
-Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed in-order, applying for each rule the underlying transformation, while intermediate results are stored (in a hash map) until they are processed by another rule.
-Since there is only one consumer for each intermediate result, every read value can be immediately discarded, thereby reducing the memory consumption.
-Ultimately, only the problem solution will be stored (in the hash map).
-
-% ----------------------------------------------
-\subsubsection{Backtracking with multiple backtrack elements}
-The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems).
-However, in the generalization introduced by ADP, it is possible that a matrix cell stores multiple results.
-In such case, we need to select a correct intermediate result to avoid backtracking inconsistencies.
-
-Additionally, we need to keep track of the multiplicities of the solutions, that is if we want to obtain the $k$ best solutions, we need to make sure that we return $k$ different traces.
-To do that, we maintain a multiplicity counter in each backtrack path: \ul
-\item While there is an unique solution for all possible incoming paths, we continue in this direction with the same multiplicity (we have no choice).
-\item When there is $r$ different solutions available, and the path multiplicity at this point is $k$ we have the following cases: \ol
-	\item If $k\ge r$: we explore all paths with multiplicity $k-r+1$.
-This is because each branch may produce only one solution and we don't know ahead of time which path will provide multiple solutions.
-Finally, we retain only the $k$ best solutions.
-	\item If $k<r$ (there is more paths than needed): we explore the $k$ first paths with multiplicity 1 and safely ignore the other (as we only need $k$ distinct results).
-	\ole
-\ule
-
-Now remains the problem of generating all possible results and check whether they are valid candidates.
-To do that we simply re-apply the parsers while maintaining the source elements of all production and then retain only those with desired score and backtrack.
-Since we know the backtrack for one element, we can do the following optimization at backtrack parser computation:\ol
-\item Defuse alternatives: since we know exactly (by the subrule id $r_s$, maintained in the backtrack) which alternative has been taken to obtain the result, we can skip undesired branches of or parsers.
-\item Feed concatenation indices: since the backtrack stores the concatenation indices, we can reuse in the concatenation parsers.
-This removes the $O(f(n))$ factor in the backtrack complexity (as concatenation backtrack parsers <<know>> where to split).
-\item Skip filters: since filters are applied before their inner solution is computed, they are only position-dependent.
-Hence if a backtrack involves a filter, since its position is set by the backtrack, the filter must have been passed at matrix construction time.
-\ole
-
-% ----------------------------------------------
-\subsubsection{Backtracking complexity}
-Since the ADP parsers can store multiple results, we are interested in measuring the overhead of $k$-best backtracking (compared to single-element backtracking).
-
-For single-element backtrack, we only need to <<revert>> the parser to find involved subsequences, which is linear in the parser size (because the backtracking identifies uniquely the alternative and concatenations indices).
-
-At every backtrack step, either:\ul
-\item The sequence is removed at one element, which leads to maximal backtrack length of $n$.
-\item The sequence is split in $k$ subsequences, with recurrence $f(n)=k\cdot f(n/k)+1$ by solving this recurrence we see that there can be at most $n$ final nodes and $n$ intermediate nodes (when $k=2$).
-Hence the backtrack length is at most $2n$.
-\ule
-
-Let one parser reversal complexity be $O(p)$, single backtrack has $O(2n\cdot p)$ complexity.
-For the $k$-elements backtrack, since we regenerate all possible solutions, that is $O(k^{c+1})$ candidates (with $c$ the maximal number of concatenation in the parser), the overall complexity is $O(2n \cdot k^{c+1} \cdot p)$.
-Hence there is a $k^{c+1}$ factor to pay if we want to backtrack the $k$ best solutions\footnote{Note that the same $k^{c+1}$ factor lies in the forward matrix computation complexity.}.
-
-Another special case we might be interested in is to obtain all co-optimal solutions (all the solutions that have an optimal score).
-We can notice that in the parser reversal, no sub-solution is discarded, because either it is not co-optimal (and would have been discarded at an earlier stage) or it is co-optimal, hence contributes to create a co-optimal result.
-It follows that the complexity of co-optimal solutions backtrack is proportional to the number of solutions.
-
-% ----------------------------------------------
-\subsubsection{Backtrack utilization}
-Since the dynamic programming may help solving a larger problem\footnote{For instance in matrix chain multiplication, we only care about matrix dimensions for dynamic programming, however, we ultimately want to multiply the real matrices and obtain a result.}, we need to be able to apply the result of the dynamic programming computation in a different domain.
-The easiest way to do that, is to reuse the same input and grammar, but use a different algebra with a different output domain, and only compute the result of the trace obtained from the DP backtrack.
-
-This step is pretty straightforward: since ADP parsers emphasize on the split between signature and grammar and decouples them, we only need to modify the algebra to operate on another domain, and reuse the same grammar.
-The key point here is to notice that a backtrack trace of a parser can be reused by another, providing that they share the same grammar.
-For instance, to compute optimally a matrix chain multiplication, we solve the DP problem in a domain where matrices are represented by their dimensions, we obtain an optimal trace and feed it to another parser operating on <<concrete matrices>> domain that will do the actual matrix multiplication (instead of the cost estimation).
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{CUDA storage: from list to optional value} \label{normalization}
-Since lists are natively supported, it is natural to gather parsers results into lists in Scala implementation.
-However, when it comes to efficient CUDA implementation, lists must be avoided because memory allocation (and management) is not very efficient \cite{scatter_alloc}.
-A workaround might be to use fixed-length lists but we assume here that the programmer is most often interested in a single optimal solution (this also alleviates the complexity of constructing multiple distinct backtrack traces).
-Even if this restriction simplifies the design, issues might arise for how to represent and deal with empty lists and how to minimize the amount of used memory:\ul
-
-\item \textbf{Minimizing memory consumption:} Under the restriction that we only store the best result, we first need to transform aggregation such that they return at most one element.
-Useful aggregator belonging to this class are quite limited: minimum/maximum (optionally with respect to an object's property), count and sum\footnote{Notice that all these operations can be implemented with a folding operation on a single variable.}, hence it is possible to provide the user with a tailored implementation.
-To benefit from this fixed memory aggregation, we need to do some structural transformation of inner parsers.
-In general, a tabulation $T$ is the root of its evaluation tree, with leaves being either other tabulations or terminals, note that any of the 5 intermediate element can appear multiple times (or not being present) and in any order:
+%\color{white}
 \[T < \textbf{Aggregate} < \textbf{Or} < \textbf{Filter} < \textbf{Map} < \textbf{Concat} < \text{(Tabulation | Terminal)}\]
 
-For obvious performance reasons, we want to maintain aggregations wherever they are present. However, we can partially normalize the rest of the evaluation tree:\ul
-	\item We must ensure that all parsers potentially generating multiple possibilities are aggregated.
-To do that, we simply wrap the original parser in an $h$-aggregation (where $h$ is the default aggregation function that must be specified by the user)
-	\item Since the aggregation now operates on a single element, we want to push it as close to the leaves as possible, as long as we do not change operational domain\footnote{We cannot push an aggregation through a mapping or a concatenation operation.}.
-	\item Filters can be hoisted within the same concatenation / alternative
-	\item Alternatives must be hoisted outside of maps and concatenations, the reason being that we need to avoid maintaining lists of candidates (that will be later aggregated).
-	\ule
-	We summarize the required transformations in the following table:
-\begin{table}[H]
 \begin{center}
@@
 \footnotesize{$R=\ $required, $P=\ $performance optimization, $?=\ $if possible}
-\caption{Parsers normalization towards CUDA code generation}
 \end{center}
-\end{table}
 
-Note that there can be no swap with Map and Concat internal parsers due to domain change. Fusing is done by the C compiler (declaring mapping functions inline).
-
-\item \textbf{Handling nested aggregations:} since a tabulation might contain nested aggregations, they must be preserved in order not to increase its time complexity.
-To do that, each internal aggregation has to define its own intermediate score and backtrack variables, whereas outermost aggregation can directly write in the cell of the cost/backtrack matrix.
-\item \textbf{Failure handling:} a parser can either be successful and return results or fail and return no result (encoded as an empty list); failure can happen in terminals, (input) tabulations and filters.
-These 3 cases can be reduced to one by wrapping terminals and tabulations into a filter that checks validity conditions.
-It remains to discuss failure encoding strategies:\ul
-	\item \textbf{Special <<empty>> value: }The benefit of such encoding is a reduced number of memory accesses; indeed since we anyway need to access the value to make computations, we do not generate additional memory accesses to check the validity of the value.
-The drawback of such approach is that it becomes necessary to specify a special {\tt empty} value that cannot be used, except to denote the absence of result.
-Since types can be arbitrary at every step of the parser, it becomes cumbersome to ask the DSL user to provide a special value for every intermediate result.
-	\item \textbf{Backtrack encoding:} Reusing the backtrack to encode the validity of a result is a more general approach, and allow greater flexibility for the user.
-Indeed, since we maintain sub-rules identifiers in the backtrack, it suffice to use a special identifier to denote that related value is invalid.
-This approach also work with nested aggregations by storing intermediate sub-rule identifiers that would only grant the validity of the related value.
-	\ule
-Since backtrack encoding comes at the price of an additional memory access to test the validity (memory accesses usually accounts for most of the time on CUDA devices), it is also relevant to allow the user to completely disable this test to speed-up computations.
-\ule
-
-% ------------------------------------------------------------------------------------------------
-\subsection{Memory constraints} \label{memory_constr}
-We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory.
-This can be the GPU or the FPGA internal memory.
-Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.
-Without loss of generality, let the underlying dynamic programming matrices be of dimension $m \times n$.
-
-We propose to evaluate the device memory requirements to solve the above problem classes. We need first to define additional problem properties related to implementation:\ul
-\item \textbf{Number of matrices:} multiple matrices can be encoded as 1 matrix with multiple values per cell.
-Hence the implementation differentiates only between cost and backtrack matrices with respective element sizes $S_C$ and $S_B$.
-\item \textbf{Delay of dependencies:} In case the problem does not fit into memory, partial matrix content needs to be transferred across sub-problems.
-Such data amount is usually proportional to the delay of dependencies.
-If this delay is small, it might be worth to duplicate matrix data in the wavefront, otherwise it might be more efficient to allow access to the previous computational blocks of the matrix.
-\item \textbf{Wavefront size:} Finally aggregations that are made along the dimensions of the matrix do not need to be written at every cell but can be propagated and aggregated along with computation (ex: maximum along one row or column).
-Hence such information can be maintained in a single place (in the wavefront) and progress together with the computation.
-We denote by $S_W$ the size of wavefront elements.
-\item \textbf{Input size:} the size of an input symbol (from input alphabet) is denoted by $S_I$.
-\ule
-
-% ----------------------------------------------
-\subsubsection{Small problems (in-memory)}
-Problem that can fit in memory can be solved in a single pass on the device. Such problem must satisfy the equation:
-	\[(S_I+S_W) \cdot (m+n) + (S_C+S_B) \cdot (m\cdot n) \le M_D\]
-
-For instance, assuming that $m=n$, $M_D=1024{\rm Mb}$, that backtrack is 2b (<16384, 3 directions) and that the cost can be represented on 4 b (int or float), that input is 1b (char) and that there is no wavefront, we can treat problems of size $n$ such that $2n+5n^2 \le 2^{30} \implies  n\le 14650$.
-We might also possibly need to take into account extra padding memory used for coalesced accesses.
-But it is reasonable to estimate that problems up to 14K fit in memory.
-
-% ----------------------------------------------
-\subsubsection{Large problems}
-To handle large problems, we need to split the $(m \times n)$ matrix into blocks of size $B_H \times B_W$.
-For simplification in our estimations, we assume a square matrix ($m=n$) made of square blocks with $b$ blocks per row/column ($B_H=B_W=n/b$).
-
-% ----------------------------------------------
-\subsubsection{Non-serial problems} \label{ns_mem_transfer}
-Non-serial problems need to potentially access all elements that have been previously computed.
-We restrict ourselves to the following dependencies\footnote{As we have not encountered a problem with non-serial dependencies along the diagonal.}: \ul
-\item Non-serial dependencies along row and column
-\item Serial dependencies along diagonal, with delay smaller or equal to one block size
-\ule
-Such restriction implies that all the block of the line and the row, and one additional block to cover diagonal dependencies must be held in memory (independently of the matrix shape).
-Hence we have the following memory restriction:
-\[ 2\frac{n}{b}(S_I+S_W) + 2 \cdot \frac{n^2}{b}S_C + \frac{n^2}{b^2} S_B \le M_D\]
-
-We also need to take into account the transfer between main memory (or disk) and device memory.
-Dependency blocks only need to be read, computed blocks need to be written back.
-Ignoring the backtrack and focusing only on the cost blocks, the transfers (in blocks) are:
-\[\begin{array}{rclll}
-		b^2 +% computed line writeback (once)
-		(b-1)^2 + % diagonal block loads (1 per block)
-		\sum\limits_{i=0}^{b-1} i \cdot b % column dependencies loads (for line i)
-		&=&\tfrac{1}{2}b^3+\tfrac{3}{2}b^2-2b+1 &\qquad& \rm (Rectangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot (b+1-i) % on i_th diagonal
-		&=& \tfrac{1}{3} b^3 + \tfrac{1}{2}b^2 + \tfrac{1}{6}b &\qquad& \rm (Triangle)
-	\\
-		\sum\limits_{i=1}^{b} \Big(1+2\cdot(i-1)\Big) \cdot b % on i_th diagonal
-		&=& b^3 &\qquad& \rm (Parallelogram)
-\end{array}\]
-
-Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
-\begin{figure}[H]\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}\caption{Transfer overhead for non-serial problems larger than device memory}\end{figure}
-
-Given an experimental bandwidth of 5.3743 Gb/s between CPU and GPU, processing matrices one order of magnitude larger (128K) would result in respectively 13\up{(R)}, 8.5\up{(T)} and 25.4\up{(P)} minutes of transfer delay.
-Extrapolating the preliminary results of small problems, a computation on input of size 128K would require respectively 7 days 13h\up{(R)}, 2 days 22h and 6 days 10h\up{(P)}, assuming there is no other scalability issues, hence transfer would account respectively for 0.1\%\up{(R)}, 0.2\%\up{(T)} and 0.3\%\up{(P)} of the running time.
-Although this overhead seems appealing compared to the computation time, the total running time blows up (because of the  $O(n^3)$ complexity) and make the processing of such large problem less relevant.
-Given that real problems (like RNA folding) operate on sequences length in the hundreds \cite{gpu_rnafold}, it would not be of much relevance to implement a version for larger cases, although perfectly feasible.
-
-% ----------------------------------------------
-\subsubsection{Serial problems} \label{serial_memory}
-The serial problems have the interesting property to access a fixed number of previous elements.
-These elements can be stored either explicitly in a scoring matrix, or implicitly into the wavefront (as moving aggregations).
-Since the dependencies are fixed, the computation direction gains an additional degree of freedom: the matrix can be solved in diagonal (as non-serial problems), line-wise or column-wise.
-This allows to store the whole necessary state to make progress into a limited number of lines (or columns), and sweep vertically (resp.
-horizontally) along the matrix.
-
-Since serial problems are of complexity $O(n^2)$ (due to the matrix dimension and the finite number of dependencies), it is possible to tackle much larger problem than non-serial given the same running time.
-Hence, it seems natural to let serial problems grow larger than the memory.
-
-Mixing the dependency property and size requirements, we can split the matrix into sub-matrices, store special lines (and/or columns) into memory (or hard disk), and repeat computations to solve the backtrack (similarly as in \cite{swat_gpu},\cite{swat_mega}, but this implementation use problem-specific knowledge that might not generalize).
-
-\newpage
-To store intermediate lines and columns, we are facing two different strategies to explore:\ul
-\item \textbf{Fixed subproblem size:} we decompose the algorithm as follows\ol
-\item Define a grid of <<major column and rows>>, where each cell's data (input, output, cost and backtrack matrices) fits into the device memory.
-\item Compute the values of the grid's major columns and rows in one pass.
-\item Second (on-demand) computation to process backtracking inside relevant cells.
-\ole
-Let $b$ the number of cells that we have on each row/column, the total computation running time would be $(b^2 + 2b) \cdot t_b$ where $t_b$ is the time to compute one cell's matrix.
-This division has the advantage of providing the minimal computation time at the expense of external storage proportional to $O(n)$ (if we store only lines or columns) or $O(n^2)$ (if we store both).
-\item \textbf{Myers and Miller’s algorithm:} \cite{myers_miller} (divide and conquer) 
-This algorithm break the DP problem into 2 (or 4) subproblems such that once the middle line/column is computed, the problem can be solved for one submatrix while backtracking occurs in up to 3 other submatrices.
-This breaking is applied recursively until the submatrix data fits into memory.
-The storage requirements are $4 \cdot O(n)$ (we store along both dimension $1+\tfrac{1}{2}+\tfrac{1}{4}+...$ lines/columns).
-
-The algorithm proceeds as follows: first it solves the problem of obtaining the first backtracking element, then it breaks the matrix in 4 submatrices, and refine it until backtrack is tractable.
-Since there is at most $\log n/b$ refinements and since every part of the matrix may be involved in backtrack, running time is $O(n^2 \log_2 n)$.
-\item \textbf{Hybrid approach:} \cite{swat_linear} a hybrid approach might be created to take advantage of additional available memory, however, the running time decreases logarithmically to the space used, this means that using $4 \times$ more storage space would only result in a $2\times$ speedup (measuring only the computation time).
-Hence a hybrid approach would be to decide a $k$ such that at each step, we partition the desired sub-matrix into a intermediate grid of $k$ rows/columns.
-The space usage would be in $2 k \log_k (n/b)$ and the running time complexity would be $O(n^2 \cdot \log_k n)$.
-Then the user would be able to fix a storage space $S \ge 4 \log_2 (n/b)$ and obtain the corresponding $k$ for a given $n$.
-\ule
-
-Finally, although such problem is interesting because targeted platforms could include FPGA, where efficient implementations exist \cite{swat_fpga}, several reasons made us considering this class of problem as a future work: \ul
-\item The most prominent problem in this category is Smith-Waterman, for which efficient implementation already exists\cite{swat_gpu},\cite{swat_mega}.
-Additionally, the authors are planning to write extensions to support some variants of this problem like Needleman-Wunsch.
-\item The implementation sensibly differs from the class of small problems, as the solving strategy is completely different from non-serial small problems, thereby requiring larger development time that would be out of the scope of this project.
-\item Finally, such an implementation would be only valuable for problems that are larger than the memory device, whereas smaller problems could perfectly use the existing implementation.
-\ule
-
-% Setup wavefront size (if needed) => just enlarge the matrix by 1 so that we go wavefront-to-wavefront. What's the maximal size of the wavefront ?
-% Can we avoid to store some matrices and put them in the wavefront ??
-% Split into blocks:
-% - Decide the shape of the blocks
-% - Decide the size of the blocks
-% - Decide of a strategy to store intermediate lines/columns: space/time tradeoff.
-% The delay of dependencies might also have an impact: if the matrix is too large to fit in the memory (device or main memory), it becomes necessary to maintain partial matrix content (all the intermediate elements) within the wavefront.
-Also the number of cost matrices might affect the performance, simply because maintaining them requires computations and memory accesses.
-
-% ------------------------------------------------------------------------------------------------
-\newpage
-\subsection{Memory layout}\label{mem_layout}
-A major bottleneck on massively parallel architecture with shared memory (like CUDA) is the global memory access.
-To address it, according to the manufacturer documentation \cite{nvidia}, it would be best if all threads access contiguous memory at the same time (coalesced memory access).
-This is justified by the memory hardware architecture, where additional latency (precharge rows and columns of the memory chip) is mandatory to access data at very different positions.
-Since all thread share the same global memory (and most of the time the same scheduler), accessing non-contiguous memory cumulates latencies before progress can be made.
-
-\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(16,2.5)
-	\put(1,0){\moveto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath\put(1.1,1.1){\linethickness{1.5pt}\vector(1,1){.75}}
-		\multiput(0,0)(0,.1){20}{\circle*{.03}}\multiput(0,0)(.1,0){20}{\circle*{.03}}}
-	\put(0,-.8){\makebox[4\unitlength]{Triangle}}
-
-	\put(7,0){\moveto(0,0)\lineto(2,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
-		\put(.5,1.5){\linethickness{1.5pt}\vector(1,-1){1}}}
-	\put(6,-.8){\makebox[4\unitlength]{Rectangle}}
-
-	\put(12,0){\moveto(2,0)\lineto(4,0)\lineto(2,2)\lineto(0,2)\closepath\strokepath
-		\put(1.6,0.6){\linethickness{1.5pt}\vector(1,1){.8}}
-		\multiput(0,0)(2,0){3}{\multiput(0,0)(0,.1){20}{\circle*{.03}}}
-		\multiput(0,0)(.1,0){20}{\circle*{.03}}\multiput(4,2)(-.1,0){20}{\circle*{.03}}}
-	\put(12,-.8){\makebox[4\unitlength]{Parallelogram}}
-\end{picture}\end{center}\caption{Matrix shapes, the arrow indicates the computation progress direction}\end{figure}
-
-Since progress is made along the diagonal of the matrices, a naive row/column addressing strategy would result in no coalesced accesses.
-Hence we address them by diagonal.
-Let $M_W$ the width of the matrix and $M_H$ its height.
-Coalesced addressing is most easy in the parallelogram matrix, as we could pretend that the parallelogram is simply a tilted rectangle  (with diagonal elements being contiguous in memory) as follows:
-\[(i,j) \to i + (j-i)*M_H\]
-
-Noticing that the triangle matrix shape is just half the parallelogram, we could use the same addressing scheme, but we would use twice more memory than necessary.
-Hence we need a more involved formula: the size of the triangle embedded in a square of side $k$ (incl.
-diagonal) is $\frac{k\cdot(k+1)}{2}$; knowing that fact, the index of the element $(i,j)$ of the triangle (starting with the main diagonal when $i=j$), we obtain the following formula (with $M_H=M_W=n$):
-\[(i,j) \to M-T+i \text{ with }\left\{\begin{array}{rcll}
-	M&=&\dfrac{(n\cdot (n+1)}{2} \vspace{.15cm} & \textit{total }\bigtriangleup\\
-	d&=&n+1+i-j \vspace{.15cm} & \textit{diagonal of }(i,j)\\
-	T &=&\dfrac{(d\cdot (d+1)}{2} & \bigtriangleup \textit{ of current diagonal}
-\end{array}\right.\]
-
-Finally, for the rectangular matrix, since the parallelogram indexing looks efficient, we want to reuse the same idea.
-However, embedding the rectangle within a parallelogram would have a very large overhead ($(M_H)^2$, by adding a triangle on each side of the rectangle).
-The solution consists into breaking the rectangle into multiple horizontal stripes of fixed height $B_H$, thereby dramatically reducing the size of the additional triangles.
-Finally, the stripes can be stitched together to form a single parallelogram continuing along the next stripe.
-Noticing that beyond a certain number of lines, coalescing access does not improve latency as memory is anyway not stored contiguously, we can make $B_H$ constant .
-
-\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(4,3.5)
-\put(0,-.5) {
-	\put(0,0){\color{lightgray}\moveto(4,0)\lineto(4.5,.5)\lineto(4,.5)\closepath\fillpath\moveto(0,4)\lineto(-.5,3.5)\lineto(0,3.5)\closepath\fillpath}
-	\put(1,3){\linethickness{1.5pt}\vector(1,-1){2}}
-	\multiput(0,.5)(0,.5){7}{\line(1,0){4}}
-	\moveto(0,0)\lineto(4,0)\lineto(4.5,.5)\lineto(4,.5)\lineto(4,4)\lineto(0,4)\lineto(-.5,3.5)\lineto(0,3.5)\closepath\strokepath
-}
-\end{picture}\end{center}\caption{Parallelogram memory representation of a rectangular matrix}\end{figure}
-
-It follows that the total memory required to store the matrix is:
-\[M_W \cdot \left\lceil\dfrac{M_H}{B_H}\right\rceil \cdot B_H  + (B_H)^2\]
-
-and the mapping of indices is given by:
-\[ (i,j) \to (B_H \cdot \underset{\rm diagonal}{(j + m)} + m + \underset{\text{stripe}}{\left\lfloor \dfrac{i}{B_H} \right\rfloor} \cdot \underset{\text{stripe size}}{M_W \cdot B_H}) \quad \text{with } m=i \mod B_H \]
-
-% ------------------------------------------------------------------------------------------------
-\subsection{LMS integration} \label{lms_use}
-At first glance, LMS seems the ideal candidate to transform Scala code into its C-like equivalent.
-However, the concern in this project is that the GPU code sensibly differers from the original CPU code because the two implementations serve different purposes: CPU version (Scala) is more general whereas the GPU version trades some functionalities for performance and suffer additional restrictions, in particular for memory management and alignment.
-
-Ad-hoc C translation seems more appropriate than using LMS for the following reasons: \ul
-\item LMS uses the type {\tt Rep[T]} to represent a lifted object of type {\tt T}.
-We tried to use LMS to generate both the Scala and CUDA program.
-This approach suffered from multiple complication from types, in particular it becomes very tedious to break a lifted tuple into a tuple of lifted values (that is convert {\tt Rep[(A,B,C)]} into {\tt (Rep[A],Rep[B],Rep[C])}), and this approach also limits the functionality of Scala parsers (as only a subset of the Scala language can be lifted into LMS).
-
-\item A second possibility would be to analyze the (user) Scala program and produce another Scala program (simplified) in the restricted subset that is compatible with LMS, hereby providing some ad-hoc transformations either during the conversion (user to simplified), or at a later stage during the optional optimization phase of LMS.
-This approach seems quite complicated and error prone as it introduces multiple sources of possible errors: during the ad-hoc conversion (user to simplified), during the optimization phase, and during the conversion to C code.
-Additionally, since dynamic programming require special scheduling to respect the dependencies between intermediate solutions, we would need at a later stage to combine ad-hoc scheduling code with the result of the LMS code generation.
-\item Since ADP normalizes the structure of dynamic programming, the concerns between grammar and algebra are well separated.
-Since the grammar has a limited number of parsers, it is possible to provide an ad-hoc implementation for each of them thereby allowing better integration.
-Given that the algebra can contain arbitrary used-defined functions, we avoid duplicating the efforts and rely on LMS to generate the corresponding C code.
-\ule
-For all the previous reasons it is more simple to provide an ad-hoc code generation for the grammar, whereas the algebra can be generated by LMS, hereby avoiding to duplicate code generation efforts while at the same time allowing us to integrate tightly the grammar parsers into the surrounding scheduling C code.
-
-From an engineering perspective although the features offered by LMS seem very desirable, we should refrain from adding unnecessary levels of complexity\footnote{KISS engineering principle: \url{http://en.wikipedia.org/wiki/KISS_principle}}.
-Also given the short time frame allowed to this project (with respect to the requirements), I considered that it was preferable to spend time in making progress in our project rather than helping improving and adding features to an existing project (but I would reconsider my position given different deadlines).
-Also during our attempts to move all the parsers on LMS, we also investigated in the Scala 2.10 macros (that operate on the Scala abstract syntax tree), which represent a possible alternative to LMS for generating  simple functions for a costing algebra.
-
-% ------------------------------------------------------------------------------------------------
-\subsection{Compilation stack}
-Since the generated code sensibly differs from the Scala version, due to the reasons previously discussed, we cannot reuse LMS, although we borrow some of its ideas.
-Having the full control on the compilation stack also provides us the following benefits:\ul
-\item Since CUDA target and runtime compilation/execution is not supported in LMS, but in Delite \cite{delite}: an additional framework that runs on top of LMS.
-Although conceptually very similar, implementing our own stack reduces the number of dependencies, hence possible sources of misconfiguration errors for the final user of our framework.
-Additionally, since we do not share much of the functionality of Delite, an ad-hoc stack helps keeping the project featherweight.
-\item LMS can generate code for monadic functions that operate on arrays.
-However dynamic programming problems might require multiple inputs (for multi-track grammars) and special scheduling (to respect dependencies in the matrix), hence we need to issue specific C/CUDA code to handle problems correctly.
-\item LMS works on array of primitive types, possibly array of structures broken into array of simple types.
-Since in DP problems, composite types represent a single logical element, and since we want to benefit from coalesced accesses (and possibly storing structures into efficient on-chip shared memory), we do not want to break structures.
-Also we want to offer support for tuples, which are a convenient way to write data containers (also we want to support case classes and composites types).
-\item Some information is only known at run-time (for instance input and matrix dimensions), hence we want to benefit from this knowledge as it helps computing the matrix indices more efficiently.
-Since such information could possibly be reused, we want to make the process as transparent as possible for the DSL user.
-\ule
-Finally, we reap most of the LMS benefits in the generation of user-defined functions, as they are completely independent of the rest of the generated program thus could be generated without additional processing.
-
-Although our compilation stack might look quite similar to that of Delite, we do not share any component but LMS for the user functions generation.
-The compilation and execution process works as follows:\ul
-\item Compilation: LMS generates the C code corresponding to the user-defined function and embeds them into the program bytecode (note that we present them separately for clarity).
-\item Runtime, for each parser (grammar+algebra): recurrences analysis is done in order to generate code (with placeholders for constants)
-\item Runtime, at every parse function call: the input size is known, hence replaced into the generic problem code, which is then processed by CUDA, C++ and Scala compilers.
-Then the JNI library resulting of the compilation is loaded and its corresponding Scala wrapper is invoked on the data to be processed.
-\ule
-
-\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(15,19)
-\put(0,17){\tbox{5}{1}{DSL libraries}}\put(6,17){\tbox{5}{1}{User program}}
-{\linethickness{1.5pt}\moveto(8,17)\lineto(8,16.25)\lineto(2.5,16.25)\strokepath\put(2.5,17){\vector(0,-1){2}}}\put(-1.5,16){\small Scala compiler}
-\put(8.5,17){\linethickness{1.5pt}\vector(0,-1){2}}\put(8.75,16){\small LMS}
-% -----
-\put(0,14){\tbox{5}{1}{JVM bytecode}}\put(6,14){\tbox{5}{1}{C functions}}
-\put(2.5,14){\vector(0,-1){2.75}}\put(8.5,14){\vector(0,-1){1.75}}
-\multiput(-2,13)(.15,0){130}{\circle*{.05}}\put(14,13.25){\small compilation}\put(14,12.5){\small run-time}
-% -----
-{\linethickness{1.5pt}\moveto(2.5,7)\lineto(2.5,11)\lineto(5,11)\strokepath\put(5,11){\vector(1,0){1}}}
-\put(2.75,9){\small recurrences analysis}\put(2.75,8.4){\small code generation}
-% -----
-\put(6,10){\tbox{5}{2}{C/CUDA/JNI code}}
-\put(8.5,10){\linethickness{1.5pt}\vector(0,-1){3}}\put(8.75,9){\small constants evaluation}\put(8.75,8.4){\small compilation}
-{\linethickness{1.5pt}\moveto(12,1.5)\lineto(12,8)\lineto(10,8)\strokepath\put(10,8){\vector(-1,0){1.5}}}\put(12.25,6.5){\rotatebox{-90}{\small input size}}
-% -----
-\put(0,2.5){\tbox{5}{4.5}{Scala parsers}}
-\put(6,2.5){\put(0,0){\tbox{5}{4.5}{}}
-	\put(.25,3.25){\tbox{4.5}{1}{CUDA kernels}}\put(.25,1.75){\tbox{4.5}{1}{JNI / host C}}\put(.25,.25){\tbox{4.5}{1}{Scala wrapper}}
-	\put(3.5,3.25){\vector(0,-1){.5}}\put(3.5,1.75){\vector(0,-1){.5}}\put(1.5,1.25){\vector(0,1){.5}}\put(1.5,2.75){\vector(0,1){.5}}}
-\put(13,1){\tbox[l0]{4}{1}{Input data}}
-\put(13,0){\tbox[l0]{4}{1}{Output data}}
-\put(1.5,1.5){\vector(0,1){1}}\put(7.5,1.5){\vector(0,1){1}}\moveto(1.5,2)\lineto(1.5,1.5)\lineto(13,1.5)\strokepath % input
-\put(9.5,.5){\line(0,1){2}}\moveto(3.5,2.5)\lineto(3.5,.5)\lineto(12,.5)\strokepath\put(12,.5){\vector(1,0){1}} % output
-
-\end{picture}\end{center}\caption{Compilation and execution scheme of a parser}\end{figure}
-
-%{\color{red}
-%\textbf{User-language}: define additional parameters for the recurrence\ul
-%\item Windowing (to convert non-serial into serial problems)
-%\item Input sizes, and alphabets (backtrack, input, cost)
-%\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
-%\item Recurrence functions, devices available
-%\item What to keep in memory (cost, backtrack or both).
-%\ule
-%$\Downarrow$ Conversion (using an existing technique)
-%
-%\textbf{Intermediate representation}
-%
-%$\Downarrow$ Optimizations\ul
-%\item Transform non-serial into serial \ul
-%	\item Use aggregation functions/transformations
-%	\item Use windowing from user (if no other technique succeed)
-%	\ule
-%\item Define the wavefront depth
-%\item Avoiding the cost matrix by moving it into the wavefront
-%\ule
-%
-%\textbf{Code specification}\ul
-%\item Kernel function (1-element function), inputs, outputs, wave front, dependencies, bit sizes
-%\item Device-level interface => setup the block sizes(w/h), input and memory sizes
-%\item Define the device-specific implementation of the block (CPU/FPGA/CUDA)
-%\item Define the co-processor memory aggregation function
-%\item Define the scheduling of the blocks and aggregation (software pipelining)
-%\item Define the data movement back and forth to disk
-%\ule
-%
-%$\Downarrow$ Generation\ul
-%\item Generate the kernel for specific device
-%\item Generate the scheduling and barriers
-%\ule
-%
-%\textbf{Binary program}
-%}
-%\ol
-%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
-%\item Let the user tune the window size if he wants to reduce non-serial to serial.
-%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
-%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
-%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
-%	\item Scheduler (CPU): interleave block computation and memory statistics
-%	\ule
-%\item Common description \ul
-%	\item Block kernel processor
-%	\item Full block
-%	\ule
-%\item Discussion: wavefront, design similarities, polyhedral theory
-%\ole
-
-%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
-%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
-%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
-%$\begin{array}{|C|C|C|C|C|C|} \hline
-%0 & 0 & 0 & 0 & 0 & 0 \mnl
-%0 &  &  &  &  & \mnl
-%0 & M_{23}  &  &  &  & \mnl
-%0 &  & \sum  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%0 &  &  &  &  & \mnl
-%\end{array}$
-%\end{document}
+\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Alphard/Alphard.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\usepackage{fullpage,epic,eepic,epsfig,makeidx,color,graphicx,moreverb,boxedminipage,version}
\sloppy

\excludeversion{tobedone}

\newcommand{\C}{\texttt{C}}
\newcommand{\gpl}{Gnu Public License}
\newcommand{\pip}{PiP}
\newcommand{\lpsolve}{LP-Solve}

\newcommand{\sitemmalfa}
{\texttt{{http://www.irisa.fr/cosi/ALPHA}{http://www.irisa.fr/api/ALPHA}}}
\newcommand{\emailmmalfa}{\texttt{alpha@irisa.fr}}

\newcommand{\Alpha}{{\sc Alpha}}
\newcommand{\AlpHard}{{\sc AlpHard}}
\newcommand{\MMA}{{\sc MMAlpha}}
\newcommand{\api}{{\sc api}}
\newcommand{\irisa}{ Irisa}
\newcommand{\vlsi}{{\sc vlsi}}
\newcommand{\vhdl}{{\sc vhdl}}
\newcommand{\alfa}{\Alpha}
\newcommand{\mmalfa}{\MMA}
\newcommand{\MMAlfa}{\MMA}
\newcommand{\mmalpha}{\MMA}
\newcommand{\mma}{{Mathematica}}
\newcommand{\polylib}{{\sc polylib}}
\newcommand{\Opt}[1]{{\rm\sl [} #1 {\rm\sl ]}}
\newcommand{\Group}[1]{{\rm\sl (} #1 {\rm\sl )}}
\newcommand{\Alt}{$\mid$}
\newcommand{\Program}{{\sl Program\ }}
\newcommand{\PDecl}{{\sl PDecl\ }}
\newcommand{\PointwiseDecl}{{\sl PointwiseDecl\ }}
\newcommand{\SystemDecl}{{\sl SystemDecl\ }}
\newcommand{\Name}{{\sl Name\ }}
\newcommand{\ScalarInDeclList}{{\sl ScalarInDeclList\ }}
\newcommand{\ScalarOutDecl}{{\sl ScalarOutDecl\ }}
\newcommand{\ScalarLocDeclList}{{\sl ScalarLocDeclList\ }}
\newcommand{\ScalarVarDecl}{{\sl ScalarVarDecl\ }}
\newcommand{\ScalarDeclList}{{\sl ScalarDeclList\ }}
\newcommand{\InputDeclList}{{\sl InputDeclList\ }}
\newcommand{\OutputDeclList}{{\sl OutputDeclList\ }}
\newcommand{\LocalDeclList}{{\sl LocalDeclList\ }}
\newcommand{\VarDeclList}{{\sl VarDeclList\ }}
\newcommand{\EquationBlock}{{\sl Equationblock\ }}
\newcommand{\ParamDecl}{{\sl ParamDecl\ }}
\newcommand{\VarDeclaration}{{\sl VarDeclaration\ }}
\newcommand{\Domain}{{\sl Domain\ }}
\newcommand{\IdentifierList}{{\sl IdentifierList\ }}
\newcommand{\Identifier}{{\sl Identifier\ }}
\newcommand{\ScalarType}{{\sl ScalarType\ }}
\newcommand{\IndexList}{{\sl IndexList\ }}
\newcommand{\ConstraintList}{{\sl ConstraintList\ }}
\newcommand{\AffineFunction}{{\sl AffineFunction\ }}
\newcommand{\Constraint}{{\sl Constraint\ }}
\newcommand{\IncreasingSeq}{{\sl IncreasingSeq\ }}
\newcommand{\DecreasingSeq}{{\sl DecreasingSeq\ }}
\newcommand{\EqualitySeq}{{\sl EqualitySeq\ }}
\newcommand{\IndexExpList}{{\sl IndexExpList\ }}
\newcommand{\EquationList}{{\sl EquationList\ }}
\newcommand{\Equation}{{\sl Equation\ }}
\newcommand{\ParamAssignation}{{\sl ParamAssignation\ }}
\newcommand{\InputList}{{\sl InputList\ }}
\newcommand{\ExtParamAssgn}{{\sl ExtParamAssgn\ }}
\newcommand{\ExtensionDomain}{{\sl ExtensionDomain\ }}
\newcommand{\ExpressionList}{{\sl ExpressionList\ }}
\newcommand{\Expression}{{\sl Expression\ }}
\newcommand{\Constant}{{\sl Constant\ }}
\newcommand{\IntegerConstant}{{\sl IntegerConstant\ }}
\newcommand{\RealConstant}{{\sl RealConstant\ }}
\newcommand{\BooleanConstant}{{\sl BooleanConstant\ }}
\newcommand{\BinaryOp}{{\sl BinaryOp\ }}
\newcommand{\CommutativeOp}{{\sl CommutativeOp\ }}
\newcommand{\RelativeOp}{{\sl RelativeOp\ }}
\newcommand{\UnaryOp}{{\sl UnaryOp\ }}
\newcommand{\IndexExpression}{{\sl IndexExpression\ }}
\newcommand{\IndexTerm}{{\sl IndexTerm\ }}
\newcommand{\Number}{{\sl Number\ }}
\newcommand{\Digit}{{\sl Digit\ }}
\newcommand{\Letter}{{\sl Letter\ }}
%\newcommand{\}{{\sl \ }}

\newcommand {\setn}   {\hbox{\it I\hskip -2pt N}}     % ensemble N
\newcommand {\setz}   {\hbox{\it Z\hskip -4pt Z}}     %ensemble Z
\newcommand {\setq}   {\hbox{\it Q\hskip -6pt {\it l}\hskip 4pt}}     % ensemble Q

\newcounter{Nex}

\newtheorem{ex}{Exemple}[section]
%\newenvironment{ex}
%{\vspace{0.5cm} \refstepcounter{Nex} {\bf Example ${\theNex}$:}
%\begin{quotation}}
%{\end{quotation}}

\makeindex

\begin{document}

\title{Getting started with \Alpha{}}
\author{Api, then Cosi, then R2D2 and Compsys\thanks{
Api, Cosi and R2D2 are the names of the 
research groups that successively hosted
research related on \alfa{} and \mmalpha{} at Irisa, Rennes, France. 
From 2001, Compsys in ENS Lyon also participated in 
the development of \mmalpha{}. Recently, Tanguy Risset moved to \textsc{Insa} de Lyon.}
\\Revision of \today{}}
\date{June 2004}
\maketitle

\begin{abstract}
This document is an introduction to the \alfa{} language
and to its use for regular program description, transformation, 
evaluation, and hardware implementation in the \mmalfa{} 
software. The \alfa{} syntax is presented by means of examples. 
Basic manipulations of \alfa{} programs are first shown. 
Then hardware generation and 
several advanced 
transformations -- pipelining, change of basis, substitution, 
normalization, and 
scheduling, -- are introduced and illustrated. 
Finally, the use of \alfa{} subsystems and the \AlpHard{}
hardware description language are presented.
\end{abstract}

\section{Introduction}
\label{intro}
This document\footnote{The source of this document is in:\\
\texttt{\$MMALPHA/doc/Quickstart/AlphaStart.tex}\\
and this file is in\\
\texttt{\$MMALPHA/doc/QuickStart/AlphaStart.pdf}\\
} should be read by all \alfa{} beginners.
It briefly presents the main features of the \alfa{} language and
the basic transformations of \alfa{} programs which are 
available in the \mmalfa{} environment.

\mmalfa{} is a free software, available under the 
\gpl{}. It can be downloaded from 
the site:

\sitemmalfa{}.

To be run, it requires
\mma{} version 3.0 or later.

If you have any problem while reading this document or while 
trying the
\mmalfa{} software, please send an e-mail to \emailmmalfa{}

\subsection*{What is \Alpha?}
\label{Whatis}
\nocite{Alpha97a,BQRR,Ris21,DupontQuRi95,DuRiRo97,QuRaRi96a,QuRaRi97a}
{\Alpha} is a functional data parallel 
language invented at 
{\irisa} in Rennes (France). 
The first definition of \alfa{} was proposed by Mauras~\cite{Mauras89} 
in 1989. The original motivation was to provide 
a language for expressing algorithms in an extended version of 
the formalism of recurrence equations
proposed by Karp, Miller and Winograd~\cite{KaMiWi67}. 
Other basic references are~\cite{Moldovan82,Quinton84c,Rajopadhye86}.
The goal of \alfa{} is to provide
a high-level tool for the synthesis of parallel regular {\vlsi} architectures.

Although {\Alpha} stands for the language itself, it is often also
associated with the environment in which it is currently developed:
{\bf {\MMA}}. {\MMA} is an interface based on the \mma{}
software from which one can manipulate {\Alpha} programs.

\mmalfa{} is still under development at \irisa{} and at ENS Lyon (France).

\subsection*{What is {\Alpha} for?}
\label{whatfor}
{\Alpha} is a research tool for 
various computer science fields such as functional language semantics,
parallelization, code generation, optimization, polyhedral theory,
{\vlsi} synthesis, systolic arrays, etc.

One important long term goal of \alfa{} is to promote the use of high-level 
functional languages for the synthesis of (parallel) {\vlsi}
architectures. From the short term, Alpha can be useful
for:
\begin{itemize}
\item Providing a correct recurrence equation specification for a particular 
algorithm (see section~\ref{language} and \ref{basic}).
\item Simulating such a specification (see section~\ref{simulation}).\index{Simulation}
\item Transforming and simplifying a recurrence equation specification (see 
sections~\ref{basic} and~\ref{advanced}).
\item Computing on convex polyhedra 
(see also~\texttt{{http://www.irisa.fr/polylib}}).
\index{polylib@\polylib}
\item Scheduling programs and detecting parallelism 
(see section~\ref{schedule}).
\end{itemize}

If you are ready to invest a little more time on {\MMA}, you will 
probably be able to:
\begin{itemize}
\item Generate {\vhdl} from this description.
\item Provide a design path from the high-level functional specification of
an algorithm to the layout description of a {\vlsi} algorithm
which implements it.
\end{itemize}

The remaining of this document is organized as follows. Section~\ref{howwork}
presents the \mmalfa{} interface; section~\ref{install} explains
the installation procedure; in section~\ref{language}, the \alfa{}
language is briefly presented, while in section~\ref{basic}, 
basic operations on \alfa{} programs are described; hardware
generation is presented in section~\ref{hardware}; in section~\ref{advanced},
other transformations of \alfa{} programs are explained, and 
in section~\ref{structured}, we explain how structured \alfa{} programs
can be structured and transformed. Finally, section~\ref{andnow}
concludes and gives some additional references for further 
exploration of \MMAlfa{}.

\section{How does \mmalfa{}{} work?}
\label{howwork}
{\mmalfa{}} is written in C and in \mma{}, but a user should only see
it through its \mma{} interface.

\mma{} provides an interpreted language with
high-level built-in functions for symbolic computations: {\mmalfa{}}
uses these facilities for transforming {\Alpha} programs. 
\mma{}\index{Mathematica@\mma{}}
embeds
also a general-purpose programming language.

The basic principle of the {\mmalfa{}} environment is the following one:
\mma{} sto\-res an internal representation of an {\Alpha} program
(called Abstract Syntax Tree or AST)\index{Abstract Syntax Tree}\index{AST}
and performs computations on this
internal representation via the user's commands or functions. These commands
can be for example: view an {\Alpha} program, check
its correctness, generate \C{} code to simulate it, generate
\vhdl{} code, etc. 
All these transformations are done on the AST which is stored
in a \mma{} variable named \texttt{\$result}.
\index{result@\texttt{\$result}}

Specific \C{} functions are used for two purposes: parse and unparse
\alfa{} programs or expressions, and perform computations on
polyhedra. All \C{} functions are called via \mma{}.
Most of them are available separately in the \polylib{}
\index{the polyhedral library@\polylib{}}
polyhedral library which constitutes the computationnal kernel of the
{\mmalfa{}} environment.

\section{Installing \mmalfa{}}
\label{install}
\index{Installation@Installing \mmalfa{}}
Before going on, you
should install the {\mmalfa{}} environment. 
The installation procedure is explained in
the \MMAlfa{} web site (\sitemmalfa{}).

The easiest way to use \mmalfa{} is to access it 
through its notebook interface. 
\index{notebook interface}
To do so, type \texttt{mathematica} under
\texttt{Unix}, or start \mma{}~3.0 in the Programs menu of 
Windows NT\footnote{Any \mma{} version later than 3.0 will work.}.
\footnote{One can also access 
the \mma{} kernel directly. Type \texttt{math} under
\texttt{Unix}, or start \mma{}~3.0~Kernel in the Programs menu of 
Windows NT. An alternative is to use the \mma{} 
kernel via \texttt{emacs}, but then the demonstration notebooks are
not accessible.}

Once \mmalfa{} is installed, open a new \mma{} notebook
and evaluate the command \texttt{start[]} in this notebook\footnote{Type the command, then press simultaneously the Shift and Enter keys.}.
\index{start@\texttt{start[]}}
This command opens a demonstration notebook which is part
of the \mmalfa{} distribution. 
Then open the cell named "Introduction Notebooks"\footnote{For 
the reader who is not yet familiar with \mma{}'s 
notebooks, notebooks are organized as a tree of cells. Each cell is
delimited by a brackets at its right. Selecting a cell (and all 
its subcells) is done by clicking once its bracket. 
To open a cell, double-click the bracket once the cell is selected.}, 
\index{Introduction notebooks}
and click on button "Getting started". This opens a notebook where all 
the examples of this document are shown.
\index{Getting-started notebook}

\section{The \alfa{} language}
\label{language}
In this section, we present the syntax of the \alfa{} language. 
\index{syntax of alpha@syntax of \alfa{}}
\subsection{{\Alpha} through examples}
\label{alphabyexamples}
We introduce here the basic
features of the language on the matrix-vector multiplication example
shown in Fig.~\ref{fig1}.

\begin{figure}[htbp]
\begin{verbatim}
system prodVect: {N | N>1}
  (a : {i,j|1 <= i,j <= N} of integer; 
   b : {i|1 <= i <= N} of integer)
returns (c : {i|1 <= i <= N} of  integer);
var 
  C :  {i,j|1 <= i <= N; 0<= j <=N} of  integer;
let
  C[i,j] = 
    case
      {|j=0} : 0[];
      {|j>=1} : C[i,j-1] + a[i,j] * b[j];
    esac;
  c[i]=C[i,N];
tel;   
\end{verbatim}
\caption{{\Alpha} program describing the matrix vector product}
\label{fig1}
\end{figure}

\subsection*{Systems, variables, domains and parameters}
An {\Alpha} program is a {\em system}. \alfa{} variables are generalized
arrays which can have any shape (not just rectangles).  The set of
indices of the array is called the {\em domain} of the variable.
\index{system}
\index{domain}
\index{variable}
Example~\ref{ex1} below shows the declaration of a variable {\tt a}
whose domain is the set of points $(i,j)$ in the triangle 
$\{0 \leq i \leq j;j \leq 10\}$.
\begin{ex}{~}
\begin{verbatim}
a : {i,j | 0<= i <= j; j <=10} of integer
\end{verbatim}
\label{ex1}
\end{ex} 
In the example of Fig.~\ref{fig1}, domains are indexed
by a parameter \texttt{N} which is declared, together with its
domain, right after the system name. The values of the parameters
can be constrained by any kind of 
affine constraints. In 
general, parameters allow generic descriptions of algorithms to 
be specified.

\subsection*{Input, output and local variables}
An \alfa{} system has input and output variables. 
Here inputs are \texttt{a} and \texttt{b}, and output
is \texttt{c}.
It may have local
variables, such as \texttt{C} here. Local variables are
declared after the keyword \texttt{var}.
\index{input variable}
\index{local variable}
\index{output variable}

\subsection*{Equations and expressions}
Each variable is defined by a unique equation which
usually has the form of a recurrence equation. 
\index{expression}
\index{equation}

The {\tt case}
construct allows one to define different values in different parts of
the domain.
\index{case expression}\index{case expression@\texttt{case}}

\begin{ex}{~}
\begin{verbatim}
a[i,j] = 
  case
    {| j = 0 } : 0[];
    {| j > 0 } : a[i,j-1]+1[];
  esac;
\end{verbatim}
\label{ex2} 
\end{ex}

The {\Alpha} expression of example~\ref{ex2} defines the values of
\texttt{a[i,0]} to be zero\footnote{Syntactic note: constants are zero
dimensional arrays hence the empty brackets in \texttt{0[]}.}
and recursively defines {\tt a} at all other points in its
domain. This equation defines a variable {\tt a} such that {\tt
a[i,j]=j}.

In Fig.~\ref{fig1}, variable \texttt{C} is also defined as
a recurrence equation, and its value at point $(i,j)$ 
is $\sum_{j=1}^{j=i} a_{i,j} \times b_{j}$.

\subsection{Array notation and standard notation}
The above equations use the so-called 
{\em array notation} of \alfa{}. 
\index{array notation}
\index{standard notation}
The
real syntax of {\Alpha} is sligthly less readable but more consistent
and logical from a semantic point of view. To illustrate this,
example~\ref{ex3} shows the same definition of {\tt a} in standard
notation.
\begin{ex}{~}
\begin{verbatim}
a = 
  case
    {i,j | j = 0 } : 0.(i,j -> );
    {i,j | j > 0 } : a.(i,j -> i,j-1) + 1.(i,j -> );
  esac;
\end{verbatim}
\label{ex3}
\end{ex}

\newcommand{\integer}{\texttt{integer}}
In such an expression, \texttt{a} denotes a variable, i.e., 
a function whose type is 
$$\integer{} \times \integer{} \rightarrow \integer{} \quad.$$
Expression \texttt{(i,j->i,j-1)} denotes
a dependence function\index{dependence function}, i.e., the affine function 
$$\integer{} \times \integer{} \rightarrow \integer{} \times \integer{}$$
which maps \texttt{(i,j)} to \texttt{(i,j-1)}. 
Expression \texttt{expr = a.(i,j->i,j-1)} denotes
the composition of \texttt{a} and the dependency function 
\texttt{(i,j -> i,j-1)}. 
In
other words, {\tt expr[i,j]} has the value {\tt a[i,j-1]} at each point
\texttt{(i,j)} such that \texttt{(i,j-1)} is in the domain of {\tt a}.

Similarly {\tt expr2 = 0.(i,j->)} means that {\tt expr2[i,j]} has
value \texttt{0} for all \texttt{(i,j)}. This notation describes
the extension (or broadcasting) of the constant \texttt{0} to 
all integral points \texttt{(i,j)} of the space.\index{constant}

An expression such as \texttt{\{i,j | j > 0 \} : a.(i,j->i,j-1) + 1.(i,j->)}
represents the restriction of expression\index{restriction}
\texttt{a.(i,j->i,j-1) + 1.(i,j->)} to the 
domain \texttt{\{i,j | j > 0\}} (functions of \alfa{} are
partial).

Finally, \texttt{expr = case exp1; exp2; esac} denotes 
a case definition, where \texttt{exp1} and \texttt{exp2} are
expressions on disjoint domains.\index{case expression}

In summary, any \alfa{} expression is either a variable, 
the composition of a variable and of a dependence function, 
a restriction, or a case statement\footnote{There exists another
construct, the \texttt{use} statement, which allows calls
to subsystems to be written. This statement is explained in 
section~\ref{structured}.}.

Note that the order of the equations as well as the order of the
expressions in a case statement is meaningless
as \alfa{} is declarative:
interchanging the two branches of the case in
example~\ref{ex2} would define exactly the same value for {\tt a}.

The
evaluation order is implicit and there are tools for finding schedules
for a given program. As {\Alpha} is a functional language, the only
constraint that any evaluation order must follow is that data
dependencies between instances of 
variables must be respected. In example~\ref{ex2},
obviously {\tt a[i,j]} must be computed after {\tt a[i,j-1]}.

\subsection{More on the syntax of \alfa{}}
Section~\ref{sectstruct} describes structured \alfa{} programs.
More details on the syntax of \alfa{} can be found in
appendix~\ref{alpha1}. 
Note that \alfa{} is case sensitive, and that 
underscores are allowed in identifiers. Avoid using 
underscores, as you might get into trouble when trying to 
generate \vhdl{} for example.

\section{Basic operations on {\Alpha} programs}
\label{basic}

This section presents the first commands that you should learn in
order to deal with {\Alpha} programs.

Write an {\Alpha}
program (such as the one of Fig.~\ref{fig1} for instance) using your
favourite text editor.

Let us say you called this file \texttt{prodVect.alpha}.
The commands described in this section allow you to
{\bf load} your program in \mma{}, {\bf view} the program in
\mma{} (array notation or standard notation), {\bf save} the
program in another file, perform a {\bf static analysis} and {\bf simulate}
the program.

All these examples are also available
in the \texttt{Getting started} notebook accessible by the 
\texttt{Master} notebook of \mmalfa{} (after starting
\mma{}, type the \mma{} command
{\tt start[]} in your notebook to  access the \texttt{Master} notebook of \mmalfa{}.)

If you are not familiar with \mma{}, the 
Introduction 
notebooks section of the
\texttt{Master} notebook
points towards a brief introduction to \mma{}.

\subsection{Loading and viewing an {\Alpha} program}
Once \mma{} has been started, commands can be send to 
the kernel through \texttt{Input} cells. 
\index{Directory@\texttt{Directory[]}}
The name of the working directory can be printed out
by typing:
\begin{verbatim}
Directory[]
\end{verbatim}

If you see that this directory is not the one where you have put 
{\tt prodVect.alpha}, change it by typing:
\begin{verbatim}
SetDirectory[ "pathname" ]
\end{verbatim}
\index{SetDirectory@\texttt{SetDirectory}}
where \texttt{pathname} is 
the directory that you want to become the current directory of 
\mmalpha{} (see the \mma{} Help or type \texttt{?Directory}).

You can now {\bf load} the {\Alpha} program into \mma{}
by typing:\\
\begin{verbatim}
load["prodVect.alpha"];
\end{verbatim}
\index{load@\texttt{load}}
\begin{small}
\begin{quotation}
Note that most often, it is useful to end 
\mmalfa{} commands with a semi-colon.
Indeed, \mmalfa{} commands are \mma{} functions, 
which return the AST of a transformed \alfa{} program.
If you forget the {\tt ;} symbol, 
\index{;@\texttt{;}}
\mma{} just displays the result of the function evaluation, 
which may sometimes take a few pages... 
\end{quotation}
\end{small}

As a side effect, the \texttt{load} command assigns the AST of
the parsed program to 
the global \mma{} variable {\tt \$result}\index{result@\texttt{\$result}}. 
In general, 
\texttt{\$result} contains the result of the most recent transformation.

\paragraph*{Troubleshooting}
Syntax errors in \alfa{} programs are sometimes difficult to understand. Here is 
a list of frequent errors.
\begin{itemize}
\item In domains, inequalities are separated by semicolons, not
colons. For example: \texttt{\{i | i<10 ; i> 20 \}}. Notice
that the domain \texttt{\{i,j | 1< i,j <20 \}}
represents the domain $\{i,j | 1 < i < 20 \wedge 
1 < j < 20 \}$, as the colon allows inequalities to 
be factorized.
\item Input variables are separated by semicolons.
\item All equations (even the last one) are ended by a semicolon.
\item All branches of a case (even the last one) are ended by a semicolon.
\item In the array notation, 
indexes of a right-hand side expression should appear in 
the right-hand side variable. For
example, \texttt{a[i,j] = b[k]} is wrong, as \texttt{k} is
not an index of \texttt{a}. Note that indexes represent 
positions, not absolute names. It is perfectly possible 
to use different index names in the declaration of a
variable, and in its definition, although this is a
bad practice...
\item Constants, such as \texttt{1[]} end up with brackets; boolean
constants are \texttt{true[]} and \texttt{false[]}.
\item Restrictions on branches of a case expression start with
\texttt{\{|}. 
\end{itemize}

\paragraph*{Bugs}
The \texttt{div} operator is available, but should be
replaced by a simple \texttt{/}. The \texttt{asave} function 
forgets to add a space after the \texttt{div} operator, so that
\texttt{a div 2[]} becomes \texttt{a div2[]}...

\subsection{Viewing {\Alpha} programs}
You can view the program that has been stored in {\tt \$result}\index{result@\texttt{\$result}} by 
evaluating:
\begin{verbatim}
ashow[]
\end{verbatim}
\index{ashow@\texttt{ashow}}
By default, \texttt{ashow} pretty prints the program contained 
in {\tt \$result}\index{result@\texttt{\$result}} in array notation, but more generally, \texttt{ashow[ var ]}
pretty prints the program contained in the \mma{} variable \texttt{var}.

To display a program in standard notation
use the
\texttt{show[]} command.
\index{show@\texttt{show}}
\subsection{On-line documentation and options}
\index{on-line documentation}
\index{options}
All \mma{} functions have an on-line documentation: 
\texttt{?ashow} gives the help on \texttt{ashow}. Commands
may also have options. Type \texttt{Options[ command ]} to list
the options of \texttt{command} together with their default value.
\index{on-line documentation}
\index{options}

The documentation of \MMAlfa{} is far from complete. 
Documentation notebooks are in directory 
\begin{verbatim}
$MMALPHA/doc/packages/
\end{verbatim}
and demos notebooks are available in 
\begin{verbatim}
$MMALPHA/demos/NOTEBOOKS/
\end{verbatim}
Documentation notebooks can be accessed
using the command \texttt{docLink[ "topic" ]}, and demos
can be open using \texttt{demoLink[ "demo" ]}: both commands
paste an active button in the current notebook. Try 
for example 
\begin{verbatim}
demoLink["Getting-started"]
\end{verbatim}

\subsection{Saving a program}
To save the {\Alpha} program in a file, use
the command {\tt asave} (or {\tt save} which saves in standard
notation). For instance: 
\begin{verbatim}
asave["myFile.alpha"]
\end{verbatim}
\index{asave@\texttt{asave}}
writes the program of Fig.~\ref{fig1} in file {\tt
myFile.alpha} in the current directory. 
This command is useful in order to save
the content of \texttt{\$result}\index{result@\texttt{\$result}} after some transformations.

\subsection{Analyzing an {\Alpha} program}
\label{analyzing}
Now that you have loaded an {\Alpha} program, you can start working on
it. Your first action should always be to check it for so-called {\em static
errors} by using the {\tt analyze} command:
\begin{verbatim}
analyze[]
\end{verbatim}
\index{analyze@\texttt{analyze}}
Information about potential errors in the {\Alpha} program
is printed out.

If the analysis is successful, the result is {\tt
True}. The static analyzer of \alfa{} does essentially two kinds
of verification: 
it checks the type of all expressions -- this is not a fantastic 
novelty -- but it also checks that variables are defined
in any point of their definition domain. This second kind of
verification is very powerful, and is much more original. 
\index{checking alpha programs@checking \alfa{} programs}

\paragraph*{Notes.}
\begin{itemize}
\item It is very important that your program passes successfully the 
\texttt{analyze} test. Indeed, some transformations assume that the program
is correct. 
\item Correcting the mistakes of an \alfa{} program may 
sometimes be difficult. Especially hard to fix are errors
involving values of the parameters. Do not hesitate to send 
a program to \emailmmalfa{} if you have some trouble: experts
will try to help you!
\item 
The \texttt{analyze} command allows structured systems to 
be checked by means of the \texttt{recurse} option. 
See \texttt{?recurse}.
\index{recurse@\texttt{recurse}}
\end{itemize}

\paragraph*{Troubleshooting}
If you use non conventional types, such as signed or 
unsigned short integer (denoted as \texttt{integer[S,5]} for
example), \texttt{analyze} is unable to check the type correctness
of expressions. Use the option \texttt{scalarTypeCheck -> False}
option to avoid error messages. 
\index{signed integers}
\index{unsigned integers}

\paragraph*{More information}
A notebook giving more details on the static analyzer can 
be accessed using \texttt{demoLink["Static"]}.
\index{demoLink@\texttt{demoLink}}

\subsection{Simulating a program}
\label{simulation}
The second step of your design flow 
should be a simulation. To do so, you
should first schedule the program, then generate a \C{} program
\footnote{Several \C{} code generators are available. As of 
the date of the current revision of this 
document, the simplest one is \texttt{cGen}, and we do not
present the other ones here.}.
\index{cGen@\texttt{cGen}}

First, schedule
the program by evaluating the command
\begin{verbatim}
schedule[]
\end{verbatim}
\index{schedule@\texttt{schedule}}
This command should give you a schedule of the program, 
unless there is a dependence cycle in the program, or the program
does not have a unidimensional schedule (see section~\ref{tssim} for
more information on the scheduler.)

If the scheduler is successful, a \C{} program can be generated using
\texttt{cGen}:\index{cGen@\texttt{cGen}}
\begin{verbatim}
cGen["prodVect.c", {"N"-> 10}]
\end{verbatim}
The \texttt{\{"N"-> 10\}}
argument indicates that the value of the parameter {\tt N} will be set
\index{parameter}
to 10 (the {\tt C} code generator is not parameter independent).
If there are more than one parameter, they are given in a list 
such as \texttt{\{"N"-> 10, "M" -> 32, ... \}}. 
By default, this program, once compiled, reads its input from the
standard input ({\tt stdin}) and prints its results on the 
standard output ({\tt stdout}). 
\index{standard input}
\index{standard output}

On \texttt{Unix}, \texttt{gcc -o prodVect prodVect.c} followed by
\texttt{./prodVect} will compile, then execute the simulated
program in the current directory. By default, this program 
will prompt you for the values of the input instances.

\paragraph*{Notes}
Options of \texttt{cGen} allow various forms of C programs 
to be generated. See the documentation notebook on 
code generation (\texttt{demoLink["cGen"]}).

\paragraph*{Troubleshooting}
\label{tssim}
As already mentioned, the scheduled may fail for several 
reasons.

If your program describes an infinite calculation\index{infinite calculation} (for
example, a filter), the scheduler will fail to find 
\index{filter}
a schedule, as by default, it tries to minimize the total
computation time. You can get a schedule by using the
command
\begin{verbatim}
schedule[ optimizationType -> Null ]
\end{verbatim}
\index{optimizationType@\texttt{optimizationType}}
Note that infinite calculations can also be represented
by \alfa{} programs where the iteration domain is bounded by
a parameter. This avoids the above problem.

Your program may not admit a unidimensional schedule. 
\index{unidimensional schedule}
To check this, call it with the appropriate option 
\begin{verbatim}
schedule[ multiDimensional -> True ];
\end{verbatim}
\index{multiDimensional@\texttt{multiDimensional}}
You may then get a result, where variables are ordered
with a multi-dimensional time. The \C{}-code generator
also works with multi-dimensional schedules. 
\index{multi-dimensional schedule}

If this is not successful, 
your program may contain a dependence cycle. 
\index{dependence cycle}
Checking this property is undecidable, and therefore, 
nothing can be done to help you. In general, this happens
because the program contains an error, and a careful check 
of the equations reveals it. (Removing some equations before
scheduling may help 
localizing the fault(s).)

\paragraph*{More on \C{} program generation}
See the \texttt{demoLink["CodeGen"]} notebook. 
\index{CodeGen@\texttt{CodeGen}}
\subsection{Scheduling}
\label{schedule}

\paragraph*{Principles}
The {\tt schedule} command looks for a schedule for an {\Alpha}
\index{schedule@\texttt{schedule}}
\index{Scheduling}
program.  The basic goal of the scheduler is to find a valid and good
evaluation order. Here, the term {\em good} depends on the
optimization criterion chosen: most often, it is the total evaluation
time of the program, but one may also consider other criteria.

The time is considered as a discrete single rate clock. The overall
idea of the scheduling process is to build a linear program
(LP) and to solve it with a software tool: 
this may be \pip{}\cite{FeautrierTa90}, or \lpsolve{}, or even the \mma{}
linear solver. 
\index{Pip@\pip{}}
\index{LPSolve@\lpsolve{}}
The \alfa{} scheduler provides several options to schedule 
a program. We consider here 
the simplest one (by default), called 
{\em  monodimensional affine-by-variable schedule}. 
\index{monodimensional}
\index{affine by variable}
This esoteric name means 
that the evaluation date $T_{\texttt{A}}(i,j)$ of a given variable
instance 
\texttt{A[i,j]}  
is given by an affine
function of the indices and parameters:\\
\centerline{ $T_{{\texttt{A}}}(i,j) =
\tau^i_{{\texttt{A}}} i + \tau^j_{{\texttt{A}}} j+ \tau^N_{{\texttt{A}}} N +
\alpha_{{\texttt{A}}}$} where $N$ is a 
parameter of the {\Alpha} program.
The coefficients $\tau^j_{{\texttt{A}}}$ of this function
are (in general) different for each variable in the system.

As already seen, 
the \mma{} command to schedule a program is:
\begin{verbatim}
schedule[]
\end{verbatim}
By
default, it schedules {\tt \$result}\index{result@\texttt{\$result}} and the resulting schedule is
placed in a global variable named {\tt \$schedule}.  
To visualize the current content of the \texttt{\$schedule} variable, 
use the command \texttt{showSchedResult[]}.
\index{showSchedResult@\texttt{showSchedResult}}
\index{schedule@\texttt{\$schedule}}
The
\texttt{schedule} function has many options. 
Type {\tt
Options[schedule]} for further information. Type {\tt
schedule[opt1->value1]} to change the default value of a particular
option {\tt opt1}.

\paragraph*{Refining the schedule}
Once a monodimensional schedule is found, the scheduler
may be used to refine the schedule, in order to obtain 
a "good" architecture. Two options can be used to do so.

The \texttt{addConstraints} option allows one to fix the
\index{addConstraints@\texttt{addConstraints}}
values of the coefficients of a particular timing-function, or
even to set the timing-function of a variable. Say for example, 
that you want the \texttt{a} variable to be available 
at time $\tau_{\texttt{a}}(i,j) = i + j$, and the \texttt{b}
variable to be available at time $\tau_{\texttt{b}}(i) = i$, 
then use the command
\begin{verbatim}
schedule[ addConstraints -> {"Ta[i,j,N]=i+j", "Tb[i,N]=i"} ]
\end{verbatim}
where \texttt{Ta[i,j,N]=i+j} sets the schedule of \texttt{a}
and \texttt{Tb[i,N]=i} sets the schedule of \texttt{b}. Notice
that the parameters -- here \texttt{N} -- must be included in the indexes of the 
timing-functions.

The \texttt{addConstraints} option allows also coefficients
to be set. Coefficient
$\tau^j_{{\texttt{A}}}$ of the timing-function of a 
variable \texttt{A} is named \texttt{TADj}, and constant 
$\alpha_{\texttt{A}}$ is named \texttt{CA}. 
To constrain the scheduler to set the first coefficient of the
timing-function of \texttt{a} and the constant of the 
same timing-function to be 1, evaluate: 
\begin{verbatim}
schedule[ addConstraints -> {"TaD1==1", "Ca==1"} ]
\end{verbatim}

The \texttt{durations} option allows one to fix the number
\index{durations@\texttt{durations}}
of cycles needed to evaluate a given equation of the 
\alfa{} program. With the current version of the scheduler, 
the \texttt{durations} option allows you to specify how
many time steps you allocate to the evaluation of one 
equation. Say for example that you would like to associate 
a duration of 7 to the evaluation of \texttt{C} and of
19 to the evaluation of \texttt{c}. Then you should
use
\begin{verbatim}
schedule[ durations -> {0, 0, 19, 7} ]
\end{verbatim}
The list of integer values corresponds to the input variables
first, then the output variables (in their order of appearance
in the header of the program), then the local variables (in their
order of declaration), \texttt{\{a, b, c, C\}} in 
our example. (Notice that assigning a duration to input variables
is meaningless.)

\paragraph*{How to schedule}
Here is a typical approach to schedule a program. 
\begin{itemize}
\item Use the plain \texttt{schedule[]} command first.
It will immediately reveal if the program has a unidimensional 
schedule, in which case, {\em la vie est belle}...
\item Remember that if your program describes an infinite
calculation, you should set the \texttt{optimizationType}
\index{optimizationType@\texttt{optimizationType}}
parameter to \texttt{Null}, or bound your iteration space
in the \alfa{} program.
\item If this does not provide a schedule, use the multi-dimensional
option to find a multi-dimensional schedule. But notice that 
\mmalpha{} cannot generate hardware for a multi-dimensional 
schedule... At least, you will be able to simulate the
\alfa{} program.
\item Once a schedule has been found, refine
it by adding constraints to the coefficients. 
\end{itemize}

\paragraph*{Troubleshooting}
The scheduler is probably the most important step of 
the design flow. As already seen, the scheduler is unable
to find out a schedule if the program does not admit a mono-dimensional 
schedule, or, if the program contains dependence cycles.

Currently, \mmalpha{} does not generate hardware for 
multi-dimensional scheduled programs\footnote{This is 
planned, but not yet realized...}.

The scheduler is quite robust, and not too slow. For 
very big programs, it is better to use structered 
programs, and to use the structured scheduler (see section~\ref{structured}).

Be careful if you use additional constraints or durations: you may specify
constraints that cannot be met by the scheduler...

\paragraph*{More information}
The schedule function is explained in more detail in
{the scheduler documentation}
given in file:
\begin{verbatim}
$MMALPHA/doc/Users/Scheduler_user_manual.pdf
\end{verbatim}

\subsection{Mapping the program to a parallel array}
\label{mapping}
Once a schedule (either mono or multi-dimensional) has been found, 
the program can be mapped to an architecture using the 
\texttt{appSched} function, for example:
\begin{verbatim}
appSched[]
\end{verbatim}
\index{appSched@\texttt{appSched}}
The plain form rewrites the \texttt{\$result}\index{result@\texttt{\$result}} program by applying 
space-time transformations to all variables in such a 
way that the first component of the index is the time, and 
the other components represent the coordinates of a processor
(also called, the allocation function).

\paragraph*{Notes}
The allocation is chosen automatically by \mmalpha{}. 
An option allows one to fix the allocation direction. See
the \texttt{projMatrix} option.

\section{Generating Hardware}
\label{hardware}
Once a system is scheduled and allocated, generating hardware
can be done in the following way.

First, transform the program into \alfa0{}, a 
\index{Alpha0@\alfa0{}}
subset of the language:
\begin{verbatim}
toAlpha0v2[];
\end{verbatim}
\index{toAlpha0v2@\texttt{toAlpha0v2}}
After execution of \texttt{toAlpha0v2}, \texttt{\$result}\index{result@\texttt{\$result}} contains one single
program, whose expressions can be interpreted as hardware components.

\begin{small}
\begin{quotation}
It is often a good idea to clean-up a little bit the program after
execution of \texttt{toAlpha0v2}. One can 
simplify the system by the command
\begin{verbatim}
simplifySystem[ alphaFormat -> Alpha0 ]
\end{verbatim}
\index{simplifySystem@\texttt{simplifySystem}}
\index{alphaFormat@\texttt{alphaFormat}}
\index{Alpha0@\texttt{Alpha0}}
Be careful not to forget the \texttt{alphaFormat} option, as otherwise, 
the simplification would destroy the case expressions which
represent multiplexers.

Warning: do not remove identity equations 
using 
\begin{verbatim}
removeIdEqus[];
\end{verbatim}
\index{removeEqus@\texttt{removeEqus}}
as it would remove the so-called {\em mirror equations} that
were added by \texttt{toAlpha0v2} for the outputs of the system.
\end{quotation}
\end{small}

Then rewrite it into \AlpHard{}, using
\index{AlpHard@\AlpHard{}}
the command:
\begin{verbatim}
alpha0ToAlphard[];
\end{verbatim}
\index{alpha0ToAlphard@\texttt{alpha0ToAlphard}}
After this command, the program becomes a set of structured
systems contained in \texttt{\$library}, and
\texttt{\$result}\index{result@\texttt{\$result}} contains the main subsystem
(see also Section~\ref{sectstruct} for more information on structured
systems). 
\index{Library@\texttt{\$library}}
Fix the value of the parameters:
\begin{verbatim}
fixParameter[ "N", 10 ];
\end{verbatim}
\index{fixParameter@\texttt{fixParameter}}

Finally, generate \vhdl{} code
\begin{verbatim}
a2v[]
\end{verbatim}
\index{a2v@\texttt{a2v}}
Function \texttt{a2v} generates in the current directory 
one \vhdl{} file for each subsystem
of \texttt{\$library}, except maybe the main subsystem which cannot be translated
into \vhdl{}.
\index{Vhdl@\vhdl{}}

To see the \vhdl{} code in the notebook, get as the current system
the subsystem that you want to see (\texttt{getSystem[ name ]}), 
then evaluate \texttt{showVhdl[]}.
\index{getSystem@\texttt{getSystem}}

\paragraph*{Troubleshooting}
Functions \texttt{toAlpha0v2} and \texttt{alpha0ToAlphard} require
the program in \texttt{\$result}\index{result@\texttt{\$result}} to meet very specific constraints, and fail 
if these constraints are not met.

\texttt{toAlpha0v2} requires that \texttt{\$result}\index{result@\texttt{\$result}} has been scheduled and 
allocated.

\texttt{alpha0ToAlphard} has to be executed after 
\texttt{toAlpha0v2} (and cannot be executed twice!). If
the outputs of the system are not mirror equations (of the
form \texttt{out = var} or \texttt{out = var.dep}) it will
fail. If all equations do not have the same dimension, both
\texttt{toAlpha0v2} and \texttt{alpha0ToAlphard} will fail.

\texttt{a2v} will almost always fail on the main calling
system which cannot be translated into \vhdl{}: you can 
just ignore this error message. An alternative is to remove
the calling subsystem of \texttt{\$library}, before calling \texttt{a2v}.

The most common problem with \texttt{a2v} is that the parameter
values are not fixed: then \texttt{a2v} fails to 
produce the controler and the modules. 
To see this, just display the current system using \texttt{ashow}.

To generate a \vhdl{} file for only one subsystem of the 
library, get this subsystem as the current system using 
\texttt{getSystem}, then run \texttt{a2v[ \$result ]}.

\paragraph*{Notes}
Recently, \texttt{toAlpha0v2} and \texttt{alpha0ToAlphard}
were extended in order to accept that some subsystems and/or
variables are kept untouched. For example
\begin{verbatim}
toAlpha0v2[ exceptions -> {sys, var} ]
\end{verbatim}
ignores all uses of subsystem \texttt{sys} as well as
the definition of \texttt{var}. 
In a similar fashion
\begin{verbatim}
alpha0ToAlphard[ exceptions -> {sys, var} ]
\end{verbatim}
will put in the interface system (i.e., the calling program)
all uses of subsystem \texttt{sys} as well as
the definition of \texttt{var}. This works only if the 
use is without extension, and if the inputs of the 
use are simple variables. 
\begin{tobedone}
TBD: 
Modify a2v not to translate $result, by default...
\end{tobedone}

\section{Other transformations of {\Alpha} programs}
\label{advanced}

In this section we briefly review a few more advanced manipulations of
{\Alpha}.

\subsection{Pipelining}
\index{Pipelining}
Pipelining is a transformation widely used in systolic synthesis.
It is also called {\em localization} or {\em uniformization}.
It consists basically of replacing a broadcasted value
by the pipeline of this value through all the computations that 
need it.

For instance, in the program of Fig.~\ref{fig1}, we see (last 
term in the second branch of the {\tt case} expression) that 
{\tt b[j]} is used for the computation of {\tt C[i,j]} 
for all $i$, $0\leq i \leq N$. This means that \texttt{b[j]}
will be broadcasted to all processors computing \texttt{C[i,j]}.

To introduce a new variable {\tt B1} which will pipeline the {\tt b[j]} 
value from the computation of {\tt C[j,0]} to 
{\tt C[j,1]}, ... , {\tt C[j,N]}, we
use the following command:
\begin{verbatim}
pipall["C","b.(i,j->j)","B1.(i,j->i+1,j)"];
\end{verbatim}
\index{pipeAll@\texttt{pipeAll}}
In this expression, the first argument is the variable whose
equation is to be modified, and the second argument is the
expression to be pipelined (standard notation is mandatory here)
and the last argument indicates the {\em direction} of the pipeline
\texttt{(i,j->i+1,j)}
as well as the name \texttt{B1} of the new variable introduced. 
After the
execution of this command, the program contained in {\tt \$result}\index{result@\texttt{\$result}} is
the one shown in Fig.~\ref{fig2}.
Note that the direction of pipeline gives the dependence vector of the
pipeline, and not the flow direction of the pipelined variable (see 
the definition of \texttt{B1} in Fig.~\ref{fig2}.)

\begin{figure}[ht]
\begin{verbatim}
system prodVect :{N | 2<=N}
  (a : {i,j | 1<=i<=N; 1<=j<=N} of boolean; 
   b : {i | 1<=i<=N} of boolean)
returns  (c : {i | 1<=i<=N} of boolean);
var
  B1 : {i,j | 1<=i<=N; 1<=j<=N; 2<=N} of boolean;
  C : {i,j | 1<=i<=N; 0<=j<=N} of boolean;
let
  B1[i,j] = 
      case
        {| i=1; 1<=j<=N; 2<=N} : b[j];
        {| 2<=i<=N; 1<=j<=N} : B1[i-1,j];
      esac;
  C[i,j] = 
      case
        {| j=0} : False[];
        {| 1<=j} : C[i,j-1] + a[i,j] * B1;
      esac;
  c[i] = C[i,N];
tel;
\end{verbatim}
\caption{{\Alpha} program of Fig.~\ref{fig1} after pipelining of {\tt b}
in the definition of \texttt{C}}
\label{fig2}
\end{figure}

Unlike previous transformations, pipelining changes the
{\Alpha} program, but the resulting program is equivalent to
the initial one. 
The modifications performed automatically by \texttt{pipeAll} are:
\begin{enumerate}
\item Determine the domain of {\tt B1} and add a
declaration for it.
\item Build the definition of {\tt B1} based on the dependency {\tt (i,j->i+1,j)} and the given initialization equation (here, {\tt b.(i,j->j)}).
\item Replace the original expression {\tt b.(i,j->j)} by {\tt B1}.
\end{enumerate}

\begin{tobedone}
\paragraph*{More information}
See the pipeline Notebook (\texttt{demoLink["Pipeline"]}).
TBD: Do the notebook...
\end{tobedone}

\paragraph*{Related functions}
\texttt{pipeline}, \texttt{pipeInfo}, \texttt{pipeVars}.

\paragraph*{Troubleshooting}
The pipeline transformations are difficult to use, as
they are not intuitive.

\subsection{Change of basis}
\label{cob}
The change of basis is another important transformation in systolic
array design. It allows variables to be re-indexed, and is often used
to map indices to time and space.

In the example of Fig.~\ref{fig2}, suppose that we wish to express
the computations in a new index basis {\tt i',j'} such that {\tt i'=i+j}, 
{\tt j'=j}. We can perform the following change of basis:
\begin{verbatim}
changeOfBasis["C.(i,j->i+j,j)"];
\end{verbatim}
\index{changeOfBasis@\texttt{changeOfBasis}}
\index{change of Basis}
This simply
indicates that the transformation is to be applied to variable {\tt C}
and that the new coordinates in term of the old ones are given by
the linear function \texttt{(i,j->i+1,j)}.
Note that a change of basis is meaningful only if
this linear function admits an integral left inverse: in this
example, its left inverse is obviously \texttt{(i,j->i-1,j)}. The
resulting program is shown in Fig.~\ref{fig3} (after its normalization).

\begin{figure}[ht]
\begin{verbatim}
system prodVect :{N | 2<=N}
                (a : {i,j | 1<=i<=N; 1<=j<=N} of boolean; 
                 b : {i | 1<=i<=N} of boolean)
       returns  (c : {i | 1<=i<=N} of boolean);
var
  B1 : {i,j | 1<=i<=N; 1<=j<=N; 2<=N} of boolean;
  C : {i,j | j+1<=i<=j+N; 0<=j<=N} of boolean;
let
  B1[i,j] = 
      case
        {| i=1; 1<=j<=N; 2<=N} : b[j];
        {| 2<=i<=N; 1<=j<=N} : B1[i-1,j];
      esac;
  C[i,j] = 
      case
        {| j=0} : False[];
        {| 1<=j} : C[i-1,j-1] + a[i-j,j] * B1[i-j,j];
      esac;
  c[i] = C[i+N,N];
tel;
\end{verbatim}
\caption{{\Alpha} program of Fig.~\ref{fig2} after the change of basis on 
 {\tt C}}
\label{fig3}
\end{figure}

\paragraph*{Notes}
The \texttt{changeOfBasis} transformation can be used to rename 
the indexes of a variable, for example:
\begin{verbatim}
changeOfBasis[ "C.(t,p->t,p)" ]
\end{verbatim}

It can also be used to place a local variable in a higher-dimensional
space. For example
\begin{verbatim}
changeOfBasis[ "C.(i,j->i,j,1)", {"i","j","k"} ]
\end{verbatim}
will add one dimension to the local \texttt{C} variable, and 
let its indexes become \texttt{i}, \texttt{j} and \texttt{k}.
\index{extended change of basis}

Finally, note that \texttt{changeOfBasis} returns an non 
normalized program.

\paragraph*{More information}
See the \texttt{ChangeOfBasis} notebook
\texttt{docLink["ChangeOfBasis"]}.

\subsection{Substitution}
Substitution allows a variable occurence 
to be replaced by its definition. For example:
\begin{verbatim}
substituteInDef[ "c", "C" ]
\end{verbatim}
\index{substitution}
\index{substituteInDef@\texttt{substituteInDef}}
substitutes the definition of \texttt{C} in equation defining
\texttt{c}.

\paragraph*{Notes.} The substitution does not normalize the 
program.

\paragraph*{Troubleshooting.} The substitution should normally
be restricted by the context domain of the expression, but this
is not done... Sometimes, the result is not correct.

\paragraph*{More information}
... available using \texttt{docLink[ "Substitution" ]}.

\subsection{Normalization}
\label{normalization}
\index{normalization}
The normalization transformation simplifies an {\Alpha} program into a
{particular} normal form called {\em case-restriction-dependency}.
This function is very useful when one performs several automatic
transformations that may render the program less and less
readable. The command is simply: 
\begin{verbatim}
normalize[];
\end{verbatim}
\index{normalize@\texttt{normalize}}
Another useful command is 
\begin{verbatim}
simplifySystem[];
\end{verbatim}
\index{simplification}
\index{simplifySystem@\texttt{simplifySystem}}
which normalizes and simplifies a program.

\paragraph*{More information}
See the \texttt{Normalization} notebook
(\texttt{docLink["Normalization"]}.)

\section{Structured  {\Alpha} programs}
\label{structured}
% Macros pour la syntaxe
\label{sectstruct}
\index{genericity}
\index{structured programming}
\index{program structures}
\index{structures of programs}
\alfa{} programs can be structured: this section explains 
how this can be done.

\subsection{Simple structures} 
Let us write an {\alfa} program for the addition of two integers
(or fixed-point numbers) expressed as bit vectors.  A binary
adder is classically described as a sequence of \emph{full adder}
operations with the propagation of a carry bit from one full adder to
the next one, as shown in Fig.\ref{adder}.
\index{adder}
\index{binary addition}
\index{full adder}

\begin{figure}[!ht]
  \centerline{
    \includegraphics{adder.pdf}
  }
\caption{Addition of two integers (coded as bit vectors), using \emph{full adders.}\label{adder}}
\end{figure}

The following \alfa{} system describes a \emph{full adder}:
\index{full adder}
\begin{verbatim}
      system FullAdder (A,B,Cin : boolean) 
               returns (X,Cout : boolean);
      let
        X = A xor B xor Cin;
        Cout = (A and B) or (A and Cin) or (B and Cin);
      tel;
\end{verbatim}

To build an adder using this program, we need to 
instanciate a collection of such systems,
as shown in Fig.\ref{adder}. The shape of this collection may
be expressed as the {\alfa} domain \texttt{\{~b~|~0<=b<W~\}} where
\texttt{W} is a size parameter giving the number of bits of the
adder.
\index{use statement}
\index{use statement@\texttt{use}}

The \textbf{use} construct of {\alfa} allows precisely that. The
following system describes in {\alfa} the adder given in
Fig.~\ref{adder}:
\index{adder}
\index{binary addition}
\begin{verbatim}
system Plus: {W|W>1} (A,B: {b| 0<=b<W} of boolean)              -- 1       
             returns (S : {b| 0<=b<=W} of boolean);             -- 2       
var                                                             -- 3       
  Cin, Cout, X : {b| 0<=b<W} of boolean;                        -- 4       
let                                                             -- 5       
  Cin[b] =                                                      -- 6       
    case                                                        -- 7       
      {| b=0} : 0[];                                            -- 8       
      {| b>0} : Cout[b-1];                                      -- 9       
    esac;                                                       -- 10      
  use {b| 0<=b<W} FullAdder[] (A,B,Cin) returns(X, Cout);       -- 11      
  S[b] =                                                        -- 12      
    case                                                        -- 13      
      {| b<W} : X;                                              -- 14      
      {| b=W} : Cout[W-1];                                      -- 15      
    esac;                                                       -- 16      
tel;                                                            -- 17      
\end{verbatim}
In this system, line 11 reads as follows: 
\index{use statement}
\index{use statement@\texttt{use}}
\begin{quote}
"Use (or instantiate)
a collection of
instances of the subsystem \texttt{FullAdder}. This collection has the
shape of the extension domain\index{extension domain} 
\verb!{ b | 0<=b<W }! and is thus indexed
by index \texttt{b}. Let the inputs of the \texttt{b}-th instance be
the variables \texttt{A}, \texttt{B} and \texttt{Cin} at point
\texttt{b}, and similarly let the outputs of this collection of
instances be the variables \texttt{X} and \texttt{Cout}."
\end{quote}
Lines~6-10 describe the carry propagation, and lines~12-16
define the output of this binary adder.

\index{dimension extension}
In other words, line~11 is a shortcut for the following equations,
which are those of the system \texttt{FullAdder} whith the dimension
of the variables extended from zero to one:
\begin{verbatim}
        X[b] = A[b] xor B[b] xor Cin[b];
        Cout[b] = (A[b] and B[b]) or (A[b] and Cin[b]) or (B[b] and Cin[b]);
\end{verbatim}

\paragraph{Note.} In general, the extension indexes are 
added {\em to the left} of the existing indexes. This
cannot be seen in this example, since the full adder subsystem
has no indexes.

\subsection{Handling structured programs}
A structured program is stored in {\mmalfa{}} as a {\mma{}} list of
systems called a \emph{library}. The default library is stored in the
global variable \verb!$library!.
\index{library}
\index{library@\texttt{\$library}}

A structured program may be written in one single file or several
distinct files.  In the former case the \texttt{load[]} function
returns a library composed of all the systems contained in the file,
and stores this library in \verb!$library!. %$ 
\index{load@\texttt{load}}

In addition, two functions, \texttt{putSystem[]} and
\texttt{getSystem[]}, may be used to get a system from a library
as the current system \texttt{\$result}\index{result@\texttt{\$result}}, and conversely
and to put back a modified system into a library. 
Typically a
system is extracted from the library as the {\em current}
system, modified by some program
transformation, and then put back in the library.
\index{putSystem@\texttt{putSystem[]}}
\index{getSystem@\texttt{getSystem[]}}

The commands to be used are \texttt{getSystem[]} and
\texttt{putSystem[]}.

\subsection{Program transformations associated with structures}

Most \mmalfa{} functions handle parameterized programs and \emph{use}
statements. There are, however, some major exceptions such as the
\texttt{cGen} translator\index{cGen@\texttt{cGen}} which generates code
only for {\em flat} {\Alpha} programs without subsystems. 
\mmalfa{}
provides functions to transform a structured program
into a flat equivalent one:
\begin{itemize}
\item \texttt{assignParameterValue[]}
\index{assignParameterValue@\texttt{assignParameterValue[]}}
gives a value to a size parameter, i.e. it refines a generic system
into a specialized one.
\item \texttt{inlineSubSystem[]}
\index{inlineSubSystem@\texttt{inlineSubSystem[]}}
expands a
\emph{use} statement, replacing it with the equations of the
corresponding subsystem, properly modified to take the dimension
extension into account.\index{flattening a structured program}
\item \texttt{inlineAll[]}
\index{inlining a subsystem}
\index{inlineAll@\texttt{inlineAll[]}}
recursively flattens a
structured {\alfa} program. \texttt{inlineAll[exceptions->\{sys\}]}
inlines all subsystems but those whose names appear in the 
exception list.
\end{itemize}

\paragraph*{More on subsystems}
For more information see 
{the subsystem documentation} in file
\begin{verbatim}
$MMALPHA/doc/Users/SubSystems.pdf
\end{verbatim}
If you are interested in scheduling structured subsystem, 
see an example in the \texttt{More} section of the 
\texttt{Master} notebook.

\paragraph*{Troubleshouting}
There is a nasty bug in \texttt{inlineSubSystem} and
\texttt{inlineAll}. If you try to inline, with extension, 
a scalar equation, this equation must be given in 
standard notation, not in array notation. In Example~\ref{bug},
the equation in the called system must not be
written \texttt{out[] = 2[] + in[]}.
\begin{ex}{~}
\begin{verbatim}
system called (in: integer) returns (out: integer);
let
  -- This works
  out = 2.(->) + in;
tel;
system caller : {N | N>=1}
 (v: {i| 0< i <= N} of integer) 
 returns (w: {i| 0< i <= N} of integer);
let
  use {i| 0<i<=N} called (v) returns (w);
tel;
\end{verbatim}
\label{bug}
\end{ex}
\section{And now?}
\label{andnow}
In this document, we have presented a few possibilities of 
\mmalfa{}. You should now know if you are interested in 
using the \mmalfa{} software.

If this is the case, you will find in the \alfa{} distribution 
some additional examples. See appendix~\ref{mmalphacontent}.

\bibliographystyle{alpha}
\bibliography{AlphaStart}

\newpage
\appendix
\section{Definition of \Alpha}
\label{alpha1}

\subsection{Meta Syntax}
\begin{tabbing}
xxxxxxx\= xxxxxxxxxxxxxxxxxx\= xxxxxxxxx\= xxxxxxx\= \kill
\>{\sl phrase* } \>===\>  zero or more repetitions of {\sl phrase}.\\
\> {\sl phrase1 {\sl \Alt} phrase2} \>===\> alternation, either {\sl
phrase1} or {\sl phrase2}.\\
\>  [\ldots] \>===\> optional phrase.\\
\> \Group{\ldots} \>===\> syntactic grouping.\\
\> {\bf \tt bold} \>===\> a terminal.\\
\> {\sl Italic} \>===\> a non-terminal.\\
\end{tabbing}

\subsection{Systems}
\Program{} stands for a library of \alfa{} programs. 
\PDecl{} (or \SystemDecl) is a single system. 
\Name{} is a system name. \ParamDecl{} is the declaration 
of a parameter domain. 
\InputDeclList{} is the list of input declarations. 
\OutputDeclList{} is the list of output declarations. 
\LocalDeclList{} is the list of local declarations. 
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\Program     \> :: \>\> \PDecl\ \PDecl *\\
\PDecl       \> :: \>\> \SystemDecl\\
\\\SystemDecl   \> :: \>\> system\ \Name \Opt{ : \ParamDecl }
                                    ( \InputDeclList )\\
\> \>\>\>  returns ( \OutputDeclList ) ;\\
\> \>\>\> \Opt{ var \LocalDeclList ;} \\
\> \>\>\> \EquationBlock ;\\
\\
\Name \>::\>\> \Identifier\\
\\
\ParamDecl  \> :: \>\> \Domain \\
\\
\InputDeclList \>::\>\> \VarDeclList\\
\OutputDeclList \>::\>\> \VarDeclList\\
\LocalDeclList \>::\>\> \VarDeclList\\
\end{tabbing}
}

\subsection{Declarations of variables}
\VarDeclList{} stands for a variable declaration. 
Notice that scalar types have been extended to 
special hardware types, such as 
\texttt{integer[S,10]} -- a signed
10 bit integer -- or
\texttt{integer[U,10]} -- an unsigned
10 bit integer. 
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\VarDeclList \>::\>\> \VarDeclaration *\\
\VarDeclaration \>::\>\> \IdentifierList\ : \Opt{\Domain of} \ScalarType ;\\
\ScalarType \> ::\>\> integer \Alt\ real \Alt\ boolean
\end{tabbing}
}

\subsection{Domains}
\Domain{} stands for a domain declaration. \IndexList{} is
a list of indexes, and \ConstraintList{} a list of constraints.

Notice that operations on domains such as intersection
(\texttt{\&}), union (\texttt{|}), complement (\texttt{~}) and
preimage by an affine function are allowed. The parser 
does the corresponding operation.

I am not sure that the \texttt{convex} syntax, nor the 
domain expressions fully work...

\ConstraintList{} is a list of constraints, each one being an
increasing constraint, a decreasing constraint, or an 
inequality.
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\Domain \>::\>\> \{ \IndexList \verb~|~ \ConstraintList \}\\
\>\>\Alt\> \Domain \verb~|~ \Domain\\
\>\>\Alt\> \Domain \verb~&~ \Domain\\
\>\>\Alt\> \Domain .\AffineFunction\\
\>\>\Alt\> \verb'~' \Domain\\
\>\>\Alt\> \Domain .convex\\
\>\>\Alt\> ( \Domain )\\
\\
\IndexList\>::\>\> \Opt{\IndexList ,} \Identifier\\
\\
\ConstraintList\>::\>\> \Opt{\ConstraintList ;} \Constraint\\
\Constraint \>::\>\> \IncreasingSeq \Alt\ \DecreasingSeq \Alt\ \EqualitySeq\\
\IncreasingSeq \>::\>\> \Group{ \IncreasingSeq \Alt\ \IndexExpList }
  \Group{ < \Alt\ <= } \IndexExpList\\
\DecreasingSeq \>::\>\> \Group{ \DecreasingSeq \Alt\ \IndexExpList }
  \Group{ > \Alt\ >= } \IndexExpList\\
\EqualitySeq\>::\>\> \Group{ \EqualitySeq \Alt\ \IndexExpList } = \IndexExpList
\end{tabbing}
}

\subsection{Equations}
\EquationBlock{} is the block of equation declarations. 
\Equation{} represents an equation, which can be either 
in array notation (when the lhs has the form \texttt{var[...]})
or in standard notation. An equation can also be a \texttt{use} statement.
\footnote{I am not sure of the syntax for the 
paramAssignation with the dot...}

{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\EquationBlock \>::\>\> let \EquationList tel \\
\EquationList  \>::\>\> \Opt{ \EquationList } \Equation\\
\Equation\>::\>\> \Identifier [ \IndexList ] = \Expression ;\\
\>\> \Alt\ \> \Identifier = \Expression ;\\
\>\> \Alt\ \> use \Opt{ \ExtensionDomain} \Identifier
                                    \Opt{.\ParamAssignation} \\
\>\>\>\>( \InputList )\\
\>\>\>\> returns ( \IdentifierList ) ;\\
\\
\ParamAssignation \>::\>\> \AffineFunction \\
\\
\InputList \>::\>\> \Opt{ \InputList , } \Expression\\
\\
\ExtensionDomain \>::\>\> \Domain \\
\end{tabbing}
}

\subsection{Expressions}
Expressions can be case expressions, if statements\footnote{\alfa{}
conditional statements are strict, that is to say, both branches
are evaluated, and moreover, the domain of the statement is 
the intersection of that of the condition and of the 
expressions.}, restrictions, affine dependencies, binary operations, 
unary operations, and reductions\footnote{Reductions are allowed, 
but few transformations are currently available.}
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\Expression\>::\>\> case \ExpressionList esac \\
\>\> \Alt\ \> if \Expression then \Expression else \Expression\\
\>\> \Alt\ \> \Domain :\Expression\\
\>\> \Alt\ \> \Expression .\AffineFunction\\
\>\> \Alt\ \> \Expression [ \IndexExpList ] \\
\>\> \Alt\ \> \Expression \BinaryOp \Expression\\
\>\> \Alt\ \> \BinaryOp ( \Expression, \Expression ) \\
\>\> \Alt\ \> \UnaryOp \Expression\\
\>\> \Alt\ \> reduce ( \CommutativeOp, \AffineFunction, \Expression ) \\
\>\> \Alt\ \> ( \Expression ) \\
\>\> \Alt\ \> \Identifier\\
\>\> \Alt\ \> \Constant\\
\\
\ExpressionList \>::\>\> \Opt{ \ExpressionList  } \Expression ;\\
\\
\BinaryOp \>::\>\> \CommutativeOp \Alt\ \RelativeOp \Alt\ - \Alt\ div \Alt\ mod\\
\CommutativeOp \>::\>\> + \Alt\ * \Alt\ / \Alt\ and \Alt\ or \Alt\ xor \Alt\ min \Alt\ max\\
\RelativeOp \>::\>\> = \Alt\ <> \Alt\ < \Alt\ <= \Alt\ > \Alt\ >=\\
\UnaryOp \>::\>\> - \Alt\ not \Alt\ sqrt\\
\\
\Constant \>::\>\> \IntegerConstant \Alt\ \RealConstant \Alt\ \BooleanConstant\\
\end{tabbing}
}
\subsection{Dependance Functions and Index Expressions}
\AffineFunction{} stands for affine functions. 
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\AffineFunction \>::\>\> ( \IndexList -> \IndexExpList )\\
\IndexExpList \>::\>\> \Opt{ \IndexExpList, } \IndexExpression\
          \Alt\ \IndexExpression\\
\IndexExpression \>::\>\> \IndexExpression \Group{+ \Alt\  - } \IndexTerm\
          \Alt\ [ - ] \IndexTerm\\
\IndexTerm \>::\>\> \IntegerConstant \Identifier \Alt\ \IntegerConstant \Alt\ \Identifier
\end{tabbing}
}

\subsection{Terminals}
{\tt
\begin{tabbing}
xxxxxxxxxxxxxxxx\= xxx\= xx\=  xx\= \kill
\IntegerConstant \>::\>\> \Opt{-} \Number\\
\RealConstant \>::\>\> \Opt{-} \Number .\Number\\
\BooleanConstant \>::\>\> true \Alt\ false \Alt True \Alt\ False\\
\Number \>::\>\> \Digit \Digit *\\
\Digit \>::\>\> 0 \Alt\ 1 \Alt\ldots\Alt\ 9\\
\Identifier \>::\>\> \Letter \Group{\Letter \Alt\ \Digit} *\\
\Letter \>::\>\> a \Alt\ldots\Alt\ z  \Alt\ A \Alt\ldots\Alt\ Z \Alt\ \verb'_'\\
\end{tabbing}
}
Note: avoid using underscores...

\section{Description of the internal format of ASTs}
\label{ast}
This section describes the format of Abstract Syntax Trees (AST) of
\alfa{} programs, as handled by \mma{}. In other words, 
the AST of a program is the \mma{} expression that \MMAlfa{} stores in 
variable \texttt{\$result} when the \texttt{load} command is 
executed. It can be displayed just by having \mma{} evaluate
the expression \texttt{\$result}.

In the following description, non terminals are
written inside angle brackets \verb\<>\.  
For readability, keywords are
written without the prefix \texttt{Alpha`} which is implicit. 
For example, the keyword \texttt{system} is actually represented
by the symbol \texttt{Alpha`system}. 
The documentation is presented in
five sections: Systems, Domains, Equations, Matrices and General.

\subsection{Systems}
\begin{verbatim}
   <library>       ::= {<system> , ... , <system>} 
   <system>        ::= system [ <system_id>,
                                <param_space>, 
                                <in_var>, 
                                <out_var>,
                                <local_var>, 
                                <equation_list> ]
   <param_space>::= <domain>
   <in_var>     ::= <declare_list>
   <out_var>    ::= <declare_list>
   <local_var>  ::= <declare_list>
   <declare>    ::= decl [ <id>, <data_type>, <domain> ]
   <data_type>  ::= integer | boolean | real | notype
\end{verbatim}

\subsection{Domains}
The domain specification closely follows the internal format of the domain
definition in the domain library. This was done to minimize the overhead of
domain storage and of making library calls.  In Mathematica, domains should
only be changed by making calls to the domain library.

\begin{verbatim}
   <domain>     ::= domain [ <dimension_number>,
                             <id_list>, 
                             <polyhedron_list> ] 
   <polyhedron> ::= pol [ <constraints_number>, 
                          <rays_number>,
                          <equations_number>, 
                          <lines_number>,
                          <constraint_list>, 
                          <ray_list> ] 
   <constraint> ::= { <const_type>, 
                      <number>, ... , <number> }
   <const_type> ::= 0 | 1
                    0 / 1 = constraint is equality / inequality
   <ray>        ::= { <ray_type>, 
                      <number>, ... , <number> }
   <ray_type>   ::= 0 | 1
                    0 / 1 = ray is line / ray
\end{verbatim}

\subsection{Equations}
\begin{verbatim}
   <equation>   ::= equation [ <id>, <exp> ]
                  | use [ <id>, 
                          <extension>, 
                          <param_assign>, 
                          <exp_list>,
                          <id_list> ]
   <extension>  ::= <domain>
   <param_assign> ::= <matrix>

   <exp>        ::= var[<id>]
                  | const[<number>] | const[<boolean>] | const[<real>]
                  | binop [ <bop>, <exp>, <exp> ]
                  | unop  [ <uop>, <exp> ]
                  | if [ <exp>, <exp>, <exp> ]
                  | affine [ <exp>, <matrix> ]
                  | restrict [ <domain>, <exp> ]
                  | case [ <exp_list> ]
                  | call [ <id>, <exp_list> ]
                  | reduce [ <casop>, <matrix>, <exp> ]
   <bop>        ::= add | sub | mul | div | idiv | mod | min | max
                  | eq | le | lt | gt | ge | ne | or | and | xor
   <unop>       ::= neg | not | sqrt 
   <casop>      ::= add | mul | and | or | xor | min | max
\end{verbatim}

\subsection{Matrices}
\begin{verbatim}
   <matrix>     ::= matrix [ <rows_number>,
                             <cols_number>, 
                             <id_list>, 
                             { { <number>, <number>, ... , <number> },
                               { <number>, <number>, ... , <number> },
                               ...
                               { <number>, <number>, ... , <number> } } ]
\end{verbatim}

\subsection{General specifications}
Numbers, ids, and lists, as used above, are defined generally
(with \verb\<*>\ representing any nonterminal ).

\begin{verbatim}
   <*_number>   ::= <number>
   <*_id>       ::= <id>
   <*_list>     ::= { <*>, <*>, ... , <*> }

   <number>     ::= [0-9][0-9]* | Infinity
   <real>       :=  <number>.<number>
   <boolean>    ::= True | False
   <id>         ::= "a name"
   <comment>    ::= (* blah blah blah *)
\end{verbatim}

\texttt{Reserved Keywords}

\begin{verbatim}
        Alpha`add       Alpha`affine    Alpha`and       Alpha`binop
        Alpha`boolean   Alpha`call      Alpha`case      Alpha`const
        Alpha`decl      Alpha`div       Alpha`domain    Alpha`eq
        Alpha`equation  Alpha`ge        Alpha`gt        Alpha`idiv
        Alpha`if        Alpha`integer   Alpha`le        Alpha`lt
        Alpha`matrix    Alpha`max       Alpha`min       Alpha`mod
        Alpha`mul       Alpha`ne        Alpha`neg       Alpha`not
        Alpha`notype    Alpha`or        Alpha`pol       Alpha`real
        Alpha`reduce    Alpha`restrict  Alpha`sqrt      Alpha`sub
        Alpha`system    Alpha`unknown   Alpha`unop      Alpha`use
        Alpha`var       Alpha`xor
\end{verbatim}

\section{A brief description of the \MMAlfa{} distribution}
\label{mmalphacontent}

\subsection{The \texttt{Master} notebook}
The master notebook is open by the \texttt{start[]} 
command. It contains 6 sections.
\begin{itemize}
\item The welcome section. 
\item The introduction notebooks: the 
\texttt{getting-started}\index{getting-started@The \texttt{Getting-started} notebook}
notebook, and the \texttt{mma-intro}\index{mma-intro@The \texttt{mma-intro} notebook}
notebook. 
\item The simple examples section. It contains pointers to the 
matrix-vector, the Fir filter, the Fifo, and the delay line 
demonstrations. 
\item The advances examples section. Here are presented a demonstration
of the Delayed Least-Mean Square filter, of the structured scheduler, 
and of the Samba architecture for DNA sequence alignment.
\footnote{Do the DLMS... Check the structured scheduler...}
\item More contains an access to the Domlib notebook, and some
suggestion about the organization of your own notebooks. It 
explained how you can set the variables \texttt{\$myNotebooks},
\texttt{\$myMasterNotebook}, then use the \texttt{myStart[]} 
and the \texttt{link}
commands to access directly your own working space. 
\item Finally, the Tests section gives access to a test notebook. 
\end{itemize}

\subsection{Documentation}
To explore the \MMAlfa{} distribution, you can follow the 
html files which are in each subdirectory of the distribution.
The structure of 
the distribution is as follows. The main directory is called \MMAlfa{}. It
contains another directory called Mathematica which is the 
\MMAlfa{} distribution properly speaking. In the \MMAlfa{} directory, 
the file \texttt{welcome.html} gives access to the documentation.

The Mathematica directory is organized as follows:
\begin{itemize}
\item \texttt{bin.cygwin} and \texttt{bin.solaris} 
directories contain the binary files for execution of \MMAlfa{} on 
the Windows NT and Solaris system respectively. 
\item \texttt{config} contains the configuration files.
\item \texttt{demos} contains the demonstrations notebooks.
\item \texttt{doc} contains the documentation notebooks.
\item \texttt{lib} contains the Mathematica packages which form 
\MMAlfa{}.
\item \texttt{sources} contains the source files of the \C{} programs
and of the latex documentation files. 
\item \texttt{tests} containes the test programs for \MMAlfa{}.
\end{itemize}

\printindex

\newpage
\tableofcontents

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Dataflow-Schedule/dataflow.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}
\usepackage{verbatim,version,epic,eepic}
\usepackage{makeidx}
\makeindex

\newcommand{\MMAlpha}{{\sc mmalpha}}
\newcommand{\mma}{{\sc Mathematica}}

\newcommand{\Alfa}{{\sc alpha}}
\newcommand{\alfa}{{\sc alpha}}
\newcommand{\myquotes}{"} 
\newcommand{\myitem}{\\}

\begin{document} 
\pagestyle{headings}
\thispagestyle{empty}
\title{About dataflow schedules}
\author{Patrice Quinton}
\date{\today}

\maketitle

\section{Introduction}
This short note describes some features that were added
to the \texttt{VertexSchedule.m} package during July 2004, 
in order to allow data-flow systems to be
scheduled. The theory of it is described in~\cite{quinton-asap2004}\footnote{Asap 2004 paper.} 
and it is briefly recalled here.

This note should be read while executing the examples in the directory 
\texttt{WCDMA-Periodic}. Currently, this directory is not part of the distribution 
of \mma{} but it soon will.

\section{Dataflow systems}
We call {\em elementary dataflow system} an \Alfa{} system 
where all symbols have a first index, say $i$, whose domain 
is $\{i | i\geq 0\}$. For example:
\begin{verbatim}
--
--   This system computes the complex multiplication
--   of two infinite complex flows of data
--
system complexMult 
  (aRe, bRe, aIm, bIm : {i|0<=i} of integer) -- input signal
returns  
  (sRe, sIm : {i|0<=i} of integer); -- output signal
let
  sRe[i] = aRe[i]*bRe[i] - aIm[i]*bIm[i];
  sIm[i] = aRe[i]*bIm[i] + aIm[i]*bRe[i];
tel;
\end{verbatim}
is a dataflow system that describes a simple complex multiplication of two 
infinite flows of data.

The schedule of such a system is easy to obtain, using for
example the following command
{\small
\begin{verbatim}
scd[optimizationType -> Null, 
       addConstraints -> {TaReD1 == TaImD1 == TbReD1 == 
       TbImD1 == 1, AaIm == 1}]
\end{verbatim}
}
Notice the options: the first one avoids trying to optimize
for duration, and the second one forces the value of the 
time component along first dimensions of variables to be 1.

The schedule a such a system can be saved in a file using 
the 
\begin{verbatim}
saveScheduleLibrary[ onlyMainSystem -> True]
\end{verbatim} 
command (see Section~\ref{saving}).
The suffix of the created
file is \texttt{scdlib}. 
The structure of this file is a list of schedules; a schedule is
a structure with \texttt{scheduleResult} head; the fields of this
structure
are the name of the system, a list corresponding to 
the parameters, and the list of variable schedules; the
schedule of a variable contains the variable name, the list of
its indexes (strings) and the schedule, in form 
\texttt{sched[ ... ]}.

Here is the 
schedule of \texttt{complexMult}, as it appears in the 
\texttt{WCDMA-Periodic} directory (this file is protected): 
{\footnotesize
\begin{verbatim}
{scheduleResult["complexMult", {}, {{"aIm", {"i"}, sched[{"$P"}, 0]}, 
   {"aRe", {"i"}, sched[{"$P"}, 0]}, {"bIm", {"i"}, sched[{"$P"}, 0]}, 
   {"bRe", {"i"}, sched[{"$P"}, 0]}}, {{"sIm", {"i"}, sched[{"$P"}, 1]}, 
   {"sRe", {"i"}, sched[{"$P"}, 1]}}]}
\end{verbatim}
}
In this example, there is a special feature: the schedule of the
variables contains a symbolic term \texttt{"\$P"}, which means 
that the schedule is given up to a symbolic {\em period} factor $\$P$.
This factor was set manually in the schedule file. It will allow a dataflow schedule to 
be computed later on.

\section{Up-sampling and down-sampling}
Two special systems are available: \texttt{oversampling} and
\texttt{undersampling}\footnote{The correct name should be up and
down sampling...}. The text of these systems is given in 
appendix \ref{oversampling} and \ref{undersampling}, respectively.
The first system takes the up-sampling by a $K$ factor of
its input $e$ and returns it in its output 
$s$ (the input is repeated $K$ times). 
The second one gives the down-sampling by a 
factor $K$ of its input $e$ and returns it in its output
$s$.

The schedule of the up-sampler, as given in the file 
\texttt{overSampling.scdlib} is:
\begin{verbatim}
{scheduleResult["overSampling", {"K"}, 
  {{"e", {"i", "K"}, sched[{"$P", 0}, 0]}}, 
  {{"s", {"j", "K"}, sched[{"$P"/"K", 0}, 0]}}]}
\end{verbatim}
It is actually a {\em symbolic scheduling}, 
given by the \mma{} expression
$\texttt{\$P}/\texttt{K}$.
This means  that if the input flow $e$ is run at period 
$\$P$, then the output $s$ is run at period $\$P/K$.

In a symmetric way, the schedule of the down sampler is 
(as found in file \texttt{underSampling.scdlib}: 
\begin{verbatim}
{scheduleResult["underSampling", {"K"}, 
  {{"e", {"i", "K"}, sched[{"$P"/"K", 0}, 0]}}, 
  {{"s", {"j", "K"}, sched[{"$P", 0}, 0]}}]}
\end{verbatim}

\section{Saving a Schedule}
\label{saving}
To obtain the schedule of a subsystem, the simplest way is 
to schedule it (using either \texttt{scd} or \texttt{schedule}), 
then to save it using the command:
\begin{verbatim}
saveScheduleLibrary[  onlyMainSystem -> True]
\end{verbatim}
This creates a file \texttt{name.scdlib} where \texttt{name}
is the current system name. 
The schedule can be edited manually and changed in order to 
contain the periodic factor \texttt{"\$P"}. It can be loaded in 
the \texttt{\$scheduleLibrary} variable using 
\begin{verbatim}
loadScheduleLibrary[ "systemName" ]
\end{verbatim}
The \texttt{saveScheduleLibrary} command has an 
option \texttt{onlyMainSystem}, the default option of which is
\texttt{False}.

\section{Scheduling a dataflow system}
To schedule a system which contains a set of subsystems, 
one uses the \texttt{scd} scheduler using the \texttt{periods} 
option
\begin{verbatim}
scd[ optimizationType -> Null, periods -> {p1, p2, ... }, ... ]
\end{verbatim}
where \texttt{periods} is mapped to the list of (integer) 
periods of the use statements of the calling system.

The order 
of these variables is that given in the dependence table and it
can be seen by printing the second part\footnote{The
structure of the dependence table is as follows:
\begin{itemize}
\item \texttt{dtable[ list of depend, list of dependuse ]},
\item a \texttt{depend} is a structure containing a domain, the
dependent var, the rhs var and a matrix that describes the dependence;
\item a \texttt{dependuse} is a structure containing 
the subsystem name, the list of input name, the list of output names, 
the rank of the use in the program, 
the domain of its parameters, and the matrix
of its parameters.
\end{itemize}
}
of this table using the \texttt{dep}
function:
\begin{verbatim}
ashow[ dep[][[2]] ]
\end{verbatim}
will print the second part of the dependence table.

For the moment, computing the periods is left to the user, 
but the method is not difficult (see~\cite{quinton-asap04}): it
amounts to solving a system of homogeneous equations of the form
$p2 = K p1$, where, for example, $p2$ is the period of a 
subsystem which is down sampled from the output of a down sampler
(and symmetrically for up sampling).

In the current model, we assume that there exists only on level
of hierarchy in the subsystems; we also assume that a given subsystem
has a uniform period. These hypothesis could be relaxed, by assuming
that a data-flow system has inputs and outputs with different 
dataflow periods.

\section{A First Example: the WCDMA Emitter}
\label{emitter}
In the example notebook, look at the WCDMA emitter example. 
\begin{verbatim}
load["WCDMAemitter.alpha"];
\end{verbatim}
loads the full program. Then
\begin{verbatim}
loadScheduleLibrary["ComplexMult"];
loadScheduleLibrary["OverSampling"];
loadScheduleLibrary["OVSF"];
loadScheduleLibrary["KASI"];
loadScheduleLibrary["Nyquist"];
loadScheduleLibrary["fir128_u"];
\end{verbatim}
loads in \texttt{\$scheduleLibrary} the schedules of the 
subsystems used in the emitter. The scheduler is called by the following command
\begin{verbatim}
scd[optimizationType -> Null, durations -> {0, 0, 1, 1, 1,
   1, 0, 0}, periods -> {16, 
    1024, 4, 4, 4, 4, 4, 4, 4, 1, 1, 1}, 
      addConstraints -> {TscD1 == TscMirrD1 == TsdD1, 
      TkascontrolCodeD2 == 
      TkascontrolCodeD3 == TkascontrolCodeD4 == 
      TkasdataCodeD2 == TkasdataCodeD3 == TkasdataCodeD4 == 
      TNyquistCodeD2 == 
      TNyquistCodeD3 == TNyquistCodeD4 ==
       TNyquistCodeD5 == TovsfControlCodeD2 == 
        TovsfControlCodeD3 == TovsfControlCodeD4 ==
         TovsfControlCodeD5 == TovsfDataCodeD2 == 
        TovsfDataCodeD3 == TovsfDataCodeD4 == 0, 
        TsccontrolD2 == TsccontrolD3 
        == TsccontrolD4 == TsccontrolD5 == TscdataD2 == 
        TscdataD3 == TscdataD4 == 
        TscdataD5 == 0, TscMirrD2 == TscMirrD3 == 
        TscMirrD5 == 0, TsdMirrD2 ==
         TsdMirrD3 == TsdMirrD5 == 0, TspcontrolD4 == 
         TspcontrolD2 == 
         TspcontrolD3 == 
        TspcontrolD5 == 0, TspdataD4 == 0}, 
        objFunction -> TscD1 + TscD2 + 
        TscD3 + TscD4 + TscD5 + TsdD1 + TsdD2 + 
        TsdD3 + TsdD4 + TsdD5]
\end{verbatim}
This gives the following schedule.
\begin{table}[htbp]
\begin{center}
\begin{tabular}{|l|c|}
\hline
\texttt{control[j]}&0\\
\texttt{controlMirr[j]}& $1024 j$\\
\texttt{data[i]}&0\\
\texttt{dataMirr[i]}&$i$\\
\texttt{kascontrolCode[j]}&$4 j$\\
\texttt{kasdataCode[i]}&$4 i$\\
\texttt{NyquistCode[j]}&$j+3$\\
\texttt{ovsfControlCode[j]}&$4j$\\
\texttt{sc[j]}&$j+2+KN$\\
\texttt{sccontrol[i]}&$4i+2$\\
\texttt{scdata[i]}&$4i+2$\\
\texttt{scMirr[j]}&$j+2+KN$\\
\texttt{sd[i]}&$2+i+KN$\\
\texttt{sdMirr[i]}&$2+i+KN$\\
\texttt{spcontrol[j]}&$4j$\\
\texttt{spdata[i]}&$4i$\\
\texttt{ssccontrol[i]}&$i+2$\\
\texttt{sscdata[i]}&$i+2$\\
\texttt{sscontrol[i]}&$4i$\\
\texttt{ssdata[i]}&$4i$\\
\hline
\end{tabular}
\end{center}
\caption{Schedules of the variables of the WCDMA emitter\label{tab:sched}}
\end{table}
Here are some explanations about the options.
\begin{itemize}
\item The \texttt{optimizationType} option allows a schedule to be found even when 
the domain of the variables is infinite (the default option tries to optimize the total
scheduling time, and this would fail). This is explained in the scheduler's manual.
\item The \texttt{durations} option allows one to assign different integral durations to 
each dependence. The order of the integer in this list is given by the order of
the dependences given by the \texttt{show[ dep[] ]} command.
\item The \texttt{periods} option is new. It allows one to provide an integral periodic factor for 
each one of the subsystems. The order of the periods corresponds to the order of the 
subsystems as given in the dependence table, and as shown\footnote{A recent modification done
on April 9, 2007.} by the \texttt{show[ dep[] ]}. To know more about how to find the 
periods, see Section~\ref{periods}. Period values are assigned to the $\$P$ 
parameter of each subsystem schedule, in such a way that this schedule is adapted
to the rate at which the system is able to run.
\item Constraints to the schedule are given in the \texttt{addConstraints} option. 
\item Finally, the objective function is given in the \texttt{objFunction} parameter.
\end{itemize}

Another, more complex, example is given in the notebook.

\section{How to Find a Schedule in Practice}
When solving this example, I did not found immediately the right 
schedule parameters. Here are some hints:
\begin{itemize}
\item Try to schedule a system in an incremental fashion: comment out some subsystems
until you find out a satisfying solution for some part of the system, then add progressively
new subsystems. Indeed, when the scheduler fails, it is very difficult to find out which one
of the constraints is not met and the reasons why it failed.
\item Setting the durations, the additional constraints, and the objective function may be postponed
until you find a solution to the whole systems with periods. Actually, the minimal 
parameters for the WCDMA emitter are:
\begin{verbatim}
scd[optimizationType -> Null, 
  periods -> {16, 1024, 4, 4, 4, 4, 4, 4, 4, 1, 1, 1}]
\end{verbatim}
This does not provide the optimal schedule
\end{itemize}
\section{How to Find Periods}
\label{periods}

\newpage
\appendix
\section{Up-sampling}
\label{oversampling}
\begin{verbatim}
--
--   This system oversamples an infinite integer 
--   input signal e with an over sampling factor K
--
system overSampling: {K|1<=K} 
 (e : {i|0<=i} of integer) -- input signal
returns  
 (s : {j|0<=j} of integer); -- output signal
var
 -- indexj[j] has value j
 -- jmodk[j] has value j mod K
 indexj, jmodk : {j|0<=j} of integer;
 -- The trick... Build this infinite array...
 E: {i,j|0<=i; 0<=j} of integer;
let
  -- Definition of indexj
  indexj[j] =
   case
     {|j=0}: 0[];
     {|j>0}: indexj[j-1]+1[];
   esac;
  -- Definition of jmodk
  jmodk[j] = 
  case
    {|j<K}: indexj[j];
    {|j>=K}: indexj[j-K];
  esac;
  -- Definition of E. 
  E[i,j] = 
   case
    {|j=0}: e[i];
    {|j>0}: if jmodk[j]=0[] then E[i+1,j-1] else E[i,j-1];
   esac;
  -- The result : take the first row of this infinite array
  s[j] = E[0,j];
tel;
\end{verbatim}

\newpage
\section{Down-sampling}

\label{undersampling}
\begin{verbatim}
system underSampling: {K|1<=K} 
 (e : {i|0<=i} of integer) -- input signal
returns  
 (s : {j|0<=j} of integer); -- output signal
var
 -- indexj[j] has value j
 -- jmodk[j] has value j mod K
 indexi : {i|0<=i} of integer;
 kindexj: {j|0<=j} of integer;
 kvalue: {k|0<=k<=K} of integer;
  -- The trick... Build this array...
 E: {i,j|0<=i; 0<=j} of integer;
let
  -- Definition of kvalue
  kvalue[k] =
    case
      {|k=0}: 0[];
      {|k>0}: kvalue[k-1]+1[];
    esac;
  -- Definition of indexi
  indexi[i] =
   case
     {|i=0}: 0[];
     {|i>0}: indexi[i-1]+1[];
   esac;
  -- Definition of kindexj
  -- This variable has value K times j
  kindexj[j] = 
  case
    {|j=0}: 0[];
    {|j>0}: kindexj[j-1]+kvalue[K];
  esac;
  -- Definition of E. 
  E[i,j] = 
   case
    {|j=0}: e[i];
    {|j>0}: if indexi[i]<kindexj[j] then E[i+1,j] else E[i+1,j-1];
   esac;
  -- The result : take the first row of this 
  s[j] = E[0,j];
tel;
\end{verbatim}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Developers/Alpha0v2/alpha0V2.tex
Added mmalpha (source and binary)

\documentclass[a4paper,11pt]{article}
%\usepackage{psfig}
\usepackage{epic}
\usepackage{eepic}

\newcommand{\mma}{{{\sc mma}lpha}}
\newcommand{\Alpha}{{{Alpha}}}
\newcommand{\AlphaZ}{{{\Alpha}0}}
\newcommand{\Alphard}{{Alphard}}
\newcommand{\vhdl}{{\sc vhdl} }

\newtheorem{definition}{Definition}[section]

\begin{document}
\title{Syntax and semantics of the {\AlphaZ} subset of Alpha (Version 2)}
\author{}
\date{\today}

 \maketitle

\section{Introduction}
{\AlphaZ} was introduced by C. Dezan in her Phd thesis (\cite{dezan}). It 
was conceived as the lowest level of {\Alpha} or equivalently as a subset of 
{\Alpha} with a structural interpretation. The main weakness of 
this subset came from the fact that {\Alpha} programs were not structured.
Since then, the structuring of 
{\Alpha} was provided by F. Dupont~\cite{DupontQuRi95} and the structured 
version of {\AlphaZ} was studied by P. Le Mo\"enner~\cite{Patricia96}.
{\Alphard} is now intended to be this subset of {\Alpha} with structural 
interpretation, from which we can translate {\Alpha} into {\vhdl} or other 
{\sc rtl} description languages. However, during the 
transformation path from {\Alpha} to {\Alphard}, the automatic structuring 
is a complicated process, hence it appear that the {\AlphaZ} format is still 
necessary as an intermediate format with all the structural information 
but in an unstructured form\footnote{il faudra unitliser deux mots different 
pour structural et structuring}. The current document describes the second 
version of the {\AlphaZ} format,

\section{Definition of {\AlphaZ}}
In this section we recall the format {\AlphaZ} 
introduced in~\cite{dezan}. 
Then we introduce the additionnal notions of version 2.
\subsection{Original {\AlphaZ}}
An {\Alpha} program is in {\AlphaZ} form if it meets the three following conditions:
\begin{enumerate}
\item there exits for all declared domains (except for the domains of 
inputs and outputs variables) an interpretation function for the indices
(see~\cite{dezan} for precise definition). Each index is either 
a temporal index or a spatial index.

\item each variable is a {\em signal} that has, for each spatial
interpretation, the temporal interpretation $\{t ~|~ t>=0\}$

\item The equations of the system define the outputs and local variables.
Each operator involved in the equations has a structural interpretation. 
These equations are of two types: data equations and control equations.
\begin{itemize} 
\item Data equations define the different signals of the program.
They are composed of the following operators:  
\begin{itemize} 
  \item pointwise operators represent the corresponding combinatorial 
operators. 
  \item restriction are used to restrict the spatial interpretation of 
indices.
\item  dependencies are temporal dependencies representing registers or
 spatial dependencies representing connections.
\item case expressions are spatial case allowing to gather several connections
\end{itemize}
\item control equations allow signals to be constructed from input variables
and the behavior to be controled. 
Such equations use the following constructions:
\begin{itemize}
\item input dependencies correspond to relations between formal 
input and variables of the programs indicating where and when 
input are sent.
\item output dependencies indicate where and when output results are valid.
\item case control (temporal case?)
 define control signals from boolean constant
(in a {\em finite automate} fashion). 
\end{itemize}
\end{itemize}
\end{enumerate}

\subsection{{\AlphaZ} version 2}
As a platform to {\Alphard}, we need to remove the second point of
the previous definition, namely the fact that all temporal domains are
of the form $\{t~|~ t>=\}$. The other basic principles remain the same
with slight differences. Spatial cases contain no dependencies,
connections are represented using a simple space dependency (no
case). Temporal cases are always nested in a spatial case.
Control initialization equations are called {\em control} equation,
the {\em Mirror} equations are the equations affecting inputs, output 
and control signal to local variables (hence, control signal 
become data signals). All these descriptions are summarized by the following 
definition:

\begin{enumerate}
\item There exits for all declared domains (except for the domains of 
inputs and outputs variables) an interpretation function for the indices
(see~\cite{dezan} for precise definition). Each index is either 
a temporal index or a spatial index.

\item The equations of the system define the output and local
variables.  Each operator involved in the equations has a structural
interpretation. These equations are of four types: data equations,
connection equations, control equations and mirror equations.
\begin{itemize} 
\item Data equations define the different signals of the program
which are necessarily local variables.
They are composed of the following operators:  
\begin{itemize} 
\item Pointwise operators represent the corresponding combinatorial 
operators. 
\item Restriction are used to restrict the spatial interpretation of 
indices.
\item Dependencies are temporal dependencies representing registers 
or identity dependencies representing connections between two signal inside one cell.
\item Cases are spatial case allowing several signals to be gathered.
\item The {\tt if} operator (with nested {\tt case} in each branches)
is interpreted as a multiplexer.
\end{itemize}
\item Connection equations are limited to the following form: a single
spatial dependency between two signals ({\tt A=B.(t,p->t,p-1)} for
instance).
\item Control equations allow control signals to be initialized. 
They define signals
which are only temporal (no spatial indices).
\item Mirror equations: equation between input variables of the system and 
local variable of the system (only for interface). equation limited to an 
affine function applied to a variable ({\tt A[t,p]=a[f(t,p)]} for input mirror
equation and {\tt b[i,j]=B[f(i,j)]} for output mirror equations)..
\end{itemize}
\end{enumerate}
\bibliography{/udd/risset/TEX/ref/newbib}
\bibliographystyle{unsrt}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Developers/Compiling/compiling.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\usepackage{fullpage,epic,eepic,epsfig,makeidx,color,moreverb,boxedminipage,version}

\sloppy

\excludeversion{tobedone}

\newcommand{\C}{\texttt{C}}
\newcommand{\gpl}{Gnu Public License}
\newcommand{\pip}{PiP}
\newcommand{\lpsolve}{LP-Solve}

\newcommand{\sitemmalfa}
{\texttt{\htmladdnormallink{http://www.irisa.fr/cosi/ALPHA}{http://www.irisa.fr/api/ALPHA}}}
\newcommand{\emailmmalfa}{\texttt{alpha@irisa.fr}}

\newcommand{\Alpha}{{\sc Alpha}}
\newcommand{\AlpHard}{{\sc AlpHard}}
\newcommand{\MMA}{{\sc MMAlpha}}
\newcommand{\irisa}{ Irisa}
\newcommand{\vlsi}{{\sc vlsi}}
\newcommand{\vhdl}{{\sc vhdl}}
\newcommand{\alfa}{\Alpha}
\newcommand{\mmalfa}{\MMA}
\newcommand{\MMAlfa}{\MMA}
\newcommand{\mmalpha}{\MMA}
\newcommand{\mma}{{Mathematica}}
\newcommand{\polylib}{{\sc polylib}}
\newcommand{\domlib}{{\sc domlib}}
\newcommand{\Opt}[1]{{\rm\sl [} #1 {\rm\sl ]}}
\newcommand{\Group}[1]{{\rm\sl (} #1 {\rm\sl )}}
\newcommand{\Alt}{$\mid$}

\newcommand {\setn}   {\hbox{\it I\hskip -2pt N}}     % ensemble N
\newcommand {\setz}   {\hbox{\it Z\hskip -4pt Z}}     %ensemble Z
\newcommand {\setq}   {\hbox{\it Q\hskip -6pt {\it l}\hskip 4pt}}     % ensemble Q

\newcommand{\ostype}{\texttt{\$OSTYPE}}
\newcommand{\envmmalpha}{\texttt{\$MMALPHA}}
\newcommand{\cygwin}{\texttt{cygwin}}
\newcommand{\cygwintrde}{\texttt{cygwin32}}
\newcommand{\linux}{\texttt{linux}}
\newcommand{\unix}{\texttt{unix}}
\newcommand{\darwin}{\texttt{darwin}}
\newcommand{\solaris}{\texttt{solaris}}
\newcommand{\sunqu}{\texttt{sun4}}
\newcommand{\sunos}{\texttt{sun OS}}

\newcommand{\varcc}{\texttt{CC}}
\newcommand{\varld}{\texttt{LD}}
\newcommand{\varpolylib}{\texttt{POLYLIB}}
\newcommand{\varcflags}{\texttt{CFLAGS}}
\newcommand{\varlflags}{\texttt{LFLAGS}}
\newcommand{\varbin}{\texttt{BIN}}
\newcommand{\varobj}{\texttt{OBJ}}
\newcommand{\vardepobj}{\texttt{DEPOBJ}}
\newcommand{\varmathprep}{\texttt{MATHPREP}}
\newcommand{\varmathprepcor}{\texttt{MATHPREPCOR}}
\newcommand{\varcppflags}{\texttt{CPPFLAGS}}
\newcommand{\varldflags}{\texttt{CLDFLAGS}}
\newcommand{\varloadlibes}{\texttt{LOADLIBS}}
\newcommand{\vardummy}{\texttt{DUMMY}}
\newcommand{\varname}{\texttt{NAME}}
\newcommand{\varobjs}{\texttt{OBJS}}
\newcommand{\varlibs}{\texttt{LIBS}}
\newcommand{\varpolydir}{\texttt{POLYDIR}}
\newcommand{\varmathinclude}{\texttt{MATHINCLUDE}}
\newcommand{\varlibdir}{\texttt{LIBDIR}}

%\newcommand{\domlib}{{Domlib}}
\newcommand{\zdomlib}{{ZDomlib}}
%\newcommand{\envmmalpha}{\texttt{\$MMALPHA}}
%\newcommand{\envostype}{\texttt{\$OSTYPE}}

\newcounter{Nex}

\newtheorem{ex}{Exemple}[section]

\makeindex

\newcommand{\ML}{{\sc Mathlink}}
\newcommand{\mprep}{{\sc mprep}}
%\newcommand{\mma}{{\sc Mathematica}}
\newcommand{\mkf}{{\texttt Makefile}}
\newcommand{\gcc}{\texttt{gcc}}
\newcommand{\doml}{\texttt{domlib.c}}
\newcommand{\tm}{\texttt{domlibtm.c}}
%\newcommand{\mmalfa}{\textsc{MMAlpha}}
\begin{document}

\title{Compiling \domlib{} and other C programs}
\author{Api, then Cosi, then R2D2 and Compsys\thanks{
Api, Cosi and R2D2 are the names of the 
research groups that successively hosted
research related on \alfa{} and \mmalpha{} at Irisa, Rennes, France. 
Since 2001, Compsys in ENS Lyon also participates in 
the development of \mmalpha{}}.\\Revision of \today{}}
\date{January 2009}
\maketitle

\begin{abstract}
This document explains how the C files for \domlib{}, \polylib{} etc.
are organized and compiled in the \mmalfa{} distribution.
\end{abstract}

%\title{Compiling C programs for \mmalfa{}}
%\author{Patrice Quinton}
%\date{September 2007}
%\maketitle

\section*{To do List}
\begin{description}
\item[2/1/2009:] seems that POLYLIB-LIB is useless. Check it...
\item[2008/10/04:] Cannot compile stuff on Linux, on Irisa machine, on \mma{} Version 4.0.1, since
I cannot run mprep... Why?
\item[2008/10/04:] Cannot compile Domlib properly for version \texttt{5.2}, but it works for version 6.0.
\item[X]  Clean the mess in this file
\item[?]  Add a DBG and a if in Makefile.darwin in domlib
\item[X]  Add varlist option to check the value of all variables
\end{description}

\section{Introduction}
\label{intro}
This document\footnote{The source of this document is in:\\
\texttt{\$MMALPHA/Mathematica/doc/sources/Compiling-Domlib/Compiling-Domlib.tex}\\
and this file should be in\\
\texttt{\$MMALPHA/Mathematica/doc/developer/Compiling-Domlib.pdf}\\
} is intended for \mmalfa{} developers who need to recompile or
modify the \domlib.

\section{About PolyLib}
Almost all C programs of \mma{} depend on the polyhedral library \polylib{}, that was developped initally
at IRISA and is now further developped and maintained in Strasburg. For convenience, 
the \mma{} distribution provides a version of \polylib{}. Before compiling
any program, you should make sure that it is properly installed.

To be more explicit, the \texttt{POLYLIB} variable of file \texttt{Makefile.config}
should contain the path to the directory where the \polylib{} is.

\section{Where are the executable files?}
They are located in \envmmalpha{}\verb\/bin.\\ostype{}, where \ostype{}
is the Operating System type. Currently, this type can be 
\cygwin{}, \cygwintrde{}, \linux{}.

\newcommand{\domlibexe}{\texttt{domlib}}
\newcommand{\pipexe}{\texttt{pip}}
\newcommand{\codegenexe}{\texttt{code\_gen}}
\newcommand{\readalphaexe}{\texttt{read\_alpha}}
\newcommand{\writealphaexe}{\texttt{write\_alpha}}
\newcommand{\writecexe}{\texttt{write\_c}}
\newcommand{\mathlink}{\texttt{mathlink}}

Typically, one finds the following executables:
\domlibexe{}, \pipexe{}, \codegenexe{}, \readalphaexe{}, 
\writealphaexe{} and \writecexe{}.

\footnote{In \texttt{bin.linux}, it seems that these executable files have
a different name... This information has to be checked.}

\section{Environment variables}
Two environment variables have to be set: 
\envmmalpha{}, \ostype{}.\footnote{Explain how to set these variables...}

\envmmalpha{} is the directory where \mmalfa{} is installed. It should have been defined 
for your installation to run (see the installation procedure).

\ostype{} is the type of your operating system, which should be set when starting 
a shell. (I do not know what it should be for \cygwin{}.)

\section{Where are the source files?}
The source files are in \envmmalpha{}\verb\/sources\. There is one directory for each 
executable.

\section{Compiling C files for MMAlpha}

\subsection{To Compile a New Version}
Compiling a new version is a somewhat tricky (and risky) operation. Tricky, since you
have to make sure that every variable, option etc. is set. Risky as you may erase some
stuff that works pretty well in an old version.

So first, make sure that you have a copy of the binaries and of the libraries which work...

Then, good luck!

First, a description of the various directories. These depend on several parameters:
your OS type (\darwin{}, \cygwin{}, \linux{}, \unix{}, \solaris{}, etc.) and the Mathematica version.
Your OS type should be set in an environment variable called OSTYPE, usually
defined by the system (this is true on Unix-like systems, not necessarily on 
Windows...).

\paragraph*{Compiling all programs}
The reference directory is \texttt{\$MMALPHA/sources}. It contains \texttt{Makefile} which
is used to compile whathever program you need. Before you compile run:
{\begin{verbatim}
make checkvars >>file
\end{verbatim}
This will check a couple of things and report the results in the \texttt{file} file. 
Examine \texttt{file} for possible errors.

To compile, run
{\begin{verbatim}
make all
\end{verbatim}

\paragraph*{Compiling one program}
To compile one particular program \texttt{pg}, run
{\begin{verbatim}
make pg
\end{verbatim}

You may even make several programs at once (\texttt{make pg1 pg2}).

\paragraph*{Cleaning}
To clean before compiling, run 
{\begin{verbatim}
make clean
\end{verbatim}
Seems however, not to clean everything.

\paragraph*{Checking variables}
To check the variables, run
{\begin{verbatim}
make checkvars
\end{verbatim}

\paragraph*{Testing stuff}
To test (no guarantee about what this tests), run
{\begin{verbatim}
make tests
\end{verbatim}

\subsection{Source directories}
The second important directory is \texttt?\$MMALPHA/sources/program} where your
program is (say for example Domlib). There, you will find source files for Domlib,
a make file (\texttt{Makefile}) that will be called by the main make file, and other 
special make files depending on your OS type: \texttt{Makefile.darwin} or
\texttt{Makefile.linux} etc...

\subsection{Objects directories}
There is also a directory \texttt{Obj.ostype} for each system type, which should be
created by \texttt{Makefile} (not sure it is), and where the objects file are put as well as 
the final binary.

Finally, the binaries are all copied in the directory \texttt{\$MMALPHA/bin.ostype}.

\section{Versions...}
The compilation process depends both on your operating system type and of your Mathematica version.
I try to provide as many versions as possible for \mma{} version 6, 5 and 4\footnote{I may soon give up for version 4, since
version 7 is already here... PQ. Jan 2009.}. However, prefered operating systems are Mac OSX and Linux.

Currently, the {\em mathematica version} is wired in the file \texttt{Makefile.config} of
the directory \texttt{\$MMALPHA/sources/MakeIncludes}. 
A variable called \texttt{MMAVERSION}
containts its number (for example, \texttt{6.0}, or \texttt{4.0.1}). The value of this variable
is very important. Indeed, the version number and the OS type define the suffixes of 
some make files. For example, if you want to compile for \mma{} version 6.0 on Linux,
then \texttt{Makefile.config} will call \texttt{Makefile.linux}, which in turn will call
\texttt{Makefile.linux.6.0} in the \texttt{MakeIncludes} directory. This allows some variables
to be set properly depending on the OS and the \mma{} version. To understand that it
is tricky, keep in mind that: 
\begin{itemize}
\item The location of the preparation program (\texttt{mprep}) and the compiling add-ons
of \mma{} necessary to prepare and compile \texttt{Domlib} depends strongly on the version.
\item After version 5, \texttt{Domlib} had to be changed, as the API of the 
\texttt{Mathlink} library were changed...
\item Some \texttt{lex} or \texttt{yacc} files had to be changed also for new versions of some OS...
\end{itemize}

Therefore, I suggest the following procedure.

\begin{enumerate}
\item Check that \envmmalpha{} and \ostype{} are properly set.
\item Look for your \mma{} version. To be sure, start \mma{} and evaluate \texttt{\$VersionNumber}.
\item In file \texttt{\envmmalpha{}/sources/MakeIncludes/Makefile.config}, check that the variable
\texttt{MMAVERSION} is set to your version number.
\item Check that there exist a file \texttt{Makefile.\ostype{}} in the directory
\texttt{\envmmalpha{}/sources/MakeIncludes}. If it does not exist, create it, on the
model of \texttt{Makefile.darwin} for example. You may have to modify the last line
of this file:
\begin{verbatim}
include ../MakeIncludes/Makefile.darwin.$(MMAVERSION)
\end{verbatim}
where you may replace \texttt{darwin} by the name of your operating system\footnote{I could have replaced
here the \texttt{darwin} name by variable \texttt{\ostype{}}, but... depending on your configuration, this
environment variable may be defined in various ways.}.
\item Check that there exist a file \texttt{Makefile.\ostype{}.MMAVERSION} in the directory
\texttt{\envmmalpha{}/sources/MakeIncludes}. If it does not exist, create it, for example, by
a modification of \texttt{Makefile.darwin.6.0}. In this file, you have to set the \texttt{MATHLINK}
variable in the second line (the remaining should be left unchanged.) This variable
says where the compiler additions of your \mma{} distribution are located. This depends on 
your configuration and on your \mma{} version. On my own machine, for version 6, they are 
{\small
\begin{verbatim}
MATHLINK = /Applications/Mathematica.app/SystemFiles/Links/MathLink/\
    DeveloperKit/CompilerAdditions
\end{verbatim}
}
and for version 5.2:
{\small
\begin{verbatim}
MATHLINK = /Applications/Mathematica\ 5.2.app/AddOns/MathLink/\
     DeveloperKit/MacOSX-x86/CompilerAdditions
\end{verbatim}
}
\end{enumerate}

If you have a problem, run a checkvars, and ... check the values of you variables.

\section{How things are}
The \texttt{sources} directory contains one directory per C software: \texttt{Code\_Gen}, 
\texttt{Domlib}, 
\texttt{Pip}, \texttt{Poly}, \texttt{Poly-darwin}, \texttt{Polylib}, \texttt{Pretty}, 
\texttt{Read\_Alpha}, \texttt{Write\_Alpha}.

The \texttt{sources} directory contains one directory per C software: \texttt{Code\_Gen}, 
\texttt{Domlib}, 
Pip, Poly, Poly-darwin, Polylib, Pretty, Read\_Alpha, Write\_Alpha.

\subsection{The \texttt{sources/Makefile}}

There is one Makefile in this directory. It contains a variable called DIR which contains
the list of programs to compile.
There is one \texttt{Makefile} in the source directory. It contains a variable called \texttt{DIR} which contains
programs to compile. For each program \texttt{pg} to compile, \texttt{Makefile}
calls 
\texttt{pg/Makefile}. For example, to compile \texttt{Domlib}, it goes in directory
\texttt{Domlib} and calls \texttt{make}.

\subsection{Where the binaries are}
All binaries produced by compilation are in file \texttt{../bin.OSTYPE}, 
where \texttt{OSTYPE} is defined as an
environment variable (this may cause a problem, as the \texttt{OSTYPE} is probably not
sufficient to decide which binary to use). Another flag, \texttt{MACHTYPE} is probably also needed...
at least for \texttt{MaCOS}.
\subsection{Where the libraries are}
The libraries produced by compilation are in file \texttt{../lib.OSTYPE}.

In each one of these directories, there is a makefile, which starts with an include
of another file \texttt{../MakeIncludes/Makefile.config}.

\subsection{The directory \texttt{sources/MakeIncludes}}
In this directory, there are a set of make files that are common to most of the 
programs to be compiled.
These files are :
\begin{itemize}
\item \texttt{Makefile.config}. A configuration file.
\item \texttt{Makefile.OSTYPE}, for all OS types. In this file, which are called by \texttt{Makefile.config},
there are configurations commands which depend on the OS type.
\item \texttt{Makefile.OSTYPE.MMAVERSION}. These files contain commands that are 
specific to a \mma{} version, for a particular OS type. They are called by 
\texttt{Makefile.OSTYPE}.
\item A file \texttt{Makefile.checkvars} which is called when the option \texttt{checkvars}
is set. This option does not work totally properly.
\end{itemize}

Thus we have the following chain of makes:
\begin{enumerate}
\item \texttt{Makefile} (in \texttt{sources})
\item Calls a make file \texttt{Makefile} located in some directory of DIR (say \texttt{read\_alpha}).
\begin{enumerate}
\item This \texttt{Makefile} includes \texttt{Makefile.config}, which is common to all programs, and
is in directory \texttt{sources/MakeIncludes}. 
\item \texttt{Makefile.config} includes itself a special configuration file customized for 
a system, named \texttt{Makefile.OSTYPE} and located in the
same directory, which may redefine some variables that were already defined in 
\texttt{Makefile.config}.
\item \texttt{Makefile.OSTYPE} includes another file, named
\begin{verbatim}
Makefile.OSTYPE.version
\end{verbatim}
located in the same directory, where 
\texttt{version} is the version number of Mathematica you are using (say 5.0, 5.2, 6.0). In this file, 
the set-up of variables related to Mathlink is done, as these may depend on the version 
of Mathematica.
\end{enumerate}
\item After including, \texttt{Makefile} includes also \texttt{Makefile.rules}
also located in \texttt{sources/MakeIncludes}, which contains
rules.  \texttt{Makefile.OSTYPE} in directory \texttt{read\_alpha}
where there are special rules for compiling this program on operating system \texttt{OSTYPE}. 
\end{enumerate}

\section{The configuration files, details}

\subsection{\texttt{Makefile.config}}
\label{config}
Defines \texttt{SHELL}, 
\texttt{MMAVERSION},
\texttt{LDLIB},
\texttt{CC},
\texttt{CFLAGS},
\texttt{CXX},
\texttt{YACC},
\texttt{LEX},
\texttt{LFLAGS},
\texttt{STRIP},
\texttt{MAKEFILES},
\texttt{BINDIR},
\texttt{LIBDIR},
\texttt{OBJDIR},
\texttt{NEEDBINDIR},
\texttt{NEEDLIBDIR},
\texttt{NEEDOBJDIR},
\texttt{SOURCES},
\texttt{POLYDIR},
\texttt{POLYINCLUDE},
\texttt{POLYOBJS},
\texttt{CODEGENDIR},
\texttt{CODEGENOBJS},
\texttt{PRETTYDIR},
\texttt{PRETTYOBJS},
\texttt{LPSOLVEDIR},
\texttt{OMEGALIB},
\texttt{OMEGABASIC}.

\subsection{The file Makefile.cygwin}
This is a configuration file where special variables are set. 
Defines SHELL, MMALPHA, MMAVERSION, SUFFIX, DEFINES, LD, PRECISIONFLAG, LDFLAGS, 
MATHLINK, OMEGADIR, MATHPREP, MATHPREPCOR, MATHINCLUDE, MATHLIB, 
MATHLOADLIB, MATHEXTRALIBS and MATHEXTRAOBJS.

All variables related to Mathlink are set in a file \texttt{Makefile.darwin.mmaversion}
where \texttt{mmaversion} is the number of the Mathematica version for which you
compile. There exists only a file for version 4.2 of Mathematica.

\paragraph*{Notes:} the Mathlink stuff is actually copied in the subdirectory
\$MMALPHA/sources/Mathlink.

\subsection{The file Makefile.linux}
Defines DEFINES, LD, PRECISIONFLAG, LDFLAGS, 
MATHLINK, OMEGADIR, MATHPREP, MATHPREPCOR, MATHINCLUDE, MATHLIB, 
MATHLOADLIB, MATHEXTRALIBS and MATHEXTRAOBJS. There is a file Makefile.linux-new. I
do not know the differences.

In the \texttt{linux} file, there is the definition of
\texttt{DEFINES}, 
\texttt{LD} (\texttt{gcc},
\texttt{PRECISIONFLAGS},
\texttt{LDFLAGS}. 
\texttt{OMEGADIR}.
Then this file calls in the same directory
another configutation file related to \mma{} (see \ref{mmafile}).

All variables related to Mathlink are set in a file \texttt{Makefile.linux.mmaversion}
where \texttt{mmaversion} is the number of the Mathematica version for which you
compile.

MATHLINK, OMEGADIR, MATHPREP, MATHPREPCOR, MATHINCLUDE, MATHLIB, 
MATHLOADLIB, MATHEXTRALIBS and MATHEXTRAOBJS.

\paragraph*{Notes:} the MMAVERSION variable is currently wired in the file
\texttt{makefile.linux} as 4.2, thus making a call to \texttt{makefile.linux.4.2}. 
There exists another such file, for version 3.0.1, referring to the location 
of Mathematica at Irisa, but I think that it is obsolete.

In the \texttt{cygwin} file, there are special definitions
(this file is not up-to-date).

In the \texttt{darwin} file, there is the definition of 
\texttt{DEFINES}, 
\texttt{LD} (\texttt{gcc},
\texttt{LDFLAGS}. Then this file calls in the same directory
another configutation file related to \mma{} (see \ref{mmafile}).

\subsection{The file Makefile.darwin}
Defines DEFINES, LD, PRECISIONFLAG, LDFLAGS, OMEGADIR.

All variables related to Mathlink are set in a file \texttt{Makefile.darwin.mmaversion}
where \texttt{mmaversion} is the number of the Mathematica version for which you
compile.

MATHLINK, OMEGADIR, MATHPREP, MATHPREPCOR, MATHINCLUDE, MATHLIB, 
MATHLOADLIB, MATHEXTRALIBS and MATHEXTRAOBJS.

\subsection{The Makefile.rules file}
This file contains general rules to compile programs. These rules are explained in the
file itself.

\subsubsection{List of variables, and where they are defined}

In Makefile.config, there is an inclusion of the definitions in MakeIncludes/Makefile.\$(OSTYPE)
which are peculiar to each OS type.

\begin{description}
\item[BINDIR]: in Makefile.config. \$(MMALPHA)/bin.\$(OSTYPE). Where the binaries will be
put. 
\item[CC]: in Makefile.config. I guess it is the compiler...
\item[CFLAGS]: in Makefile.config. -O3 -g -I\$(POLYINCLUDE) \$(PRECISIONFLAG). I guess that
these are the flags of the compiler.
\item[CODEGENDIR]: in Makefile.config. \$(SOURCES)/Code\_Gen. Where Code\_Gen is
\item[CODEGENOBJS]: in Makefile.config. \$(OBJDIR)/gen.o and nodeprocs.o. The objects of 
Code\_gen.
\item[CXX]: in Makefile.config g++
\item[DIR]: in Makefile. The list of directories where there are files to be compiled.
\item[LDLIB]: in Makefile.config. ld -r. It is the command for the linker.
\item[LEX]: in Makefile.config. flex
\item[LFLAGS]: in Makegile.config. Flags for flex
\item[LIBDIR]: in Makefile.config. \$MMALPHA/lib.\$(OSTYPE). Where the libraries are put.
\item[LPSOLVEDIR]: in Makefile.config. Where LP solve is, unused currently.
\item[OMEGALIB]: in Makefile.config. Where the OMEGA library is. unused currently.
\item[MMALPHA]: environment variable. May also be wired by a definition in Makefile.config
or in a system config file.
\item[MMAVERSION]: environment variable. May also be wired by a definition in Makefile.config
or in a Makefile.ostype.version file. Defines the version number of Mathematica. 
\item[MAKEFILES]: in Makefile.config. Contains the list of makefiles, i.e. Makefile itself (in the 
sources directory), Makefile.rules, Makefile.config, and Makefile.\$(OSTYPE) located in 
the directory MakeIncludes.
\item[NEEBINDIR]: in Makefile.config. \$(BINDIR)/.make. This variable is checked by a rule in
Makefile.rules. 
\item[NEEDLIBDIR]: in Makefile.config. \$(LIBDIR)/.make. Use?
\item[NEEDOBJDIR]: in Makefile.config. \$(OBJDIR)/.make. Use?
\item[OBJDIR]: in Makefile.config. Obj.\$(OSTYPE). Defines where the object files will be put.
This is relative to the directory of the application.
\item[OSTYPE]: environment variable
\item[POLYDIR]: in Makefile.config. \$(SOURCES)/Polylib. Where the source files of the polylib are.
\item[POLYOBJS]: in Makefile.config. The list of object files of the polylib. These files are all in 
\$(OBJDIR). I changed this line in July 2007, I do not know why... Maybe to add object files that were
needed...
\item[POLYXOBJ]: in Makefile.config. Another set of object files
\item[POLYMSG]: in Makefile.config. \$(OBJDIR)/errormsg.o Defines an object file where the 
error messages are defines for Polylib.
\item[PRETTYOBJS]
\item[PRETTYDIR]: in Makefile.config. \$(SOURCES)/Pretty. Where Pretty is.
\item[PRETTYOBJS]: in Makefile.config. $(PRETTYDIR)/$(OBJDIR)//itemprocs.o and writeitem.o. 
The objects of Pretty.
\item[SHELL]: in Makefile.config. /bin/sh
\item[SOURCES]: in Makefile.config. \$(MMALPHA)/sources. The source files of MMAlpha.
\item[STRIP]: in Makegile.config. strips. Meaning?
\item[YACC]: in Makefile.config. bison
\item[YFLAGS] in Makefile.config. Flags of Yacc
\end{description}

\subsubsection{ReadAlpha}

\begin{description}
\item First line includes Makefile.config
\item[CPPFLAGS] = -I\$(POLYDIR) -g (flags for cc)
\item[LOADLIBES] = empt
\item[DUMMY] = list of files to be destroyed when cleaning
\item[NAME] = the name of the program, read\_alpha
\item[OBJS] = the list of object files, \$(OBJDIR)/read\_alpha.o, node.o, \$(POLYOBJS)
\item[LIBS] = libraries to link again
\end{description}

Then we have a command of the form
\$(NAME) : \$(BINDIR)/\$(NAME)
	@\# bla

Then special rules or dependencies for .o files
Seems that the object files and the temporary binary are in Obj.darwin.

\subsubsection{WriteAlpha}

\begin{description}
\item First line includes Makefile.config
\item[CPPFLAGS] = -I\$(POLYDIR)
\item[LDFLAGS] = -L\$(LIBDIR) \# -mwindows
\item[LOADLIBES] = empty
\item[DUMMY] = list of files to be destroyed when cleaning
\item[NAME] = the name of the program, read\_alpha
\item[OBJS] = the list of object files, \$(OBJDIR)/yacc.o, nodeprocs.o, node2item.o, \$(POLYOBJS),
\$(POLYMSG) \$(PRETTYOBJS)
\item[LIBS] = libraries to link again, i.e, \$(LIBDIR)/libpoly.a, libpretty.a and libpolymsg.a
\end{description}

Then we have a command of the form
\$(NAME) : \$(BINDIR)/\$(NAME)
	@\# bla

Then special rules or dependencies for .o files
Seems that the object files and the temporary binary are in Obj.darwin.

\subsection{The files \texttt{sources/MakeIncludes/Makefiles.OSTYPE.Number}}
\label{mmafile}
There is one such file for each OS and each version of \mma{} which 
is supported. Currently: 
\begin{itemize}
\item For \texttt{darwin}: versions 5.2 and 6.0
\item For \texttt{linux}: versions 3.0.1 and 4.2. 
\end{itemize}
They define: 
\texttt{MATHLINK},
\texttt{MATHPREP},
\texttt{MATHINCLUDE},
\texttt{MATHLIB},
\texttt{MATHLOADLIB},
\texttt{MATHEXTRALIBS},
\texttt{MATHEXTRAOBJS}.

The \texttt{linux} version is set for IRISA and should be checked.

\paragraph*{Notes:} the MMAVERSION variable is currently wired in the file
\texttt{makefile.linux} as 4.2, thus making a call to \texttt{makefile.linux.4.2}. 
There exists another such file, for version 3.0.1, referring to the location 
of Mathematica at Irisa, but I think that it is obsolete.

\subsection{The \texttt{MakeIncludes/Makefile.rules} file}
\label{rules}
This file is called somewhere (I do not remember where) and sets common
dependence rules. There are rules for:
\begin{itemize}
\item 
\texttt{BINDIR},
\texttt{OBJDIR} (binaries, objects), 
\texttt{NEEDBINDIR} (to create BINDIR, create the directory etc.)
\texttt{NEEDLIBDIR},
\texttt{NEEDOBJDIR},
\texttt{LIBDIR} (for each library).
\end{itemize}

\section{Organization for \texttt{Domlib}}
The \texttt{Domlib/Makefile} calls another 
\texttt{Makefile.OSTYPE}, except for cygwin.

For each OS, there is a \texttt{Makefile.OSTYPE}. 
This Makefile calls first \texttt{MakeIncludes/Makefile.config} (see \ref{config}).

It calls then \texttt{MakeIncludes/Makefile.OSTYPE.MMAVERSION} (see \ref{mmafile})
(check that this file is not already called by the config file).
It defines
\texttt{POLYLIB-LIB},
\texttt{DBG},
\texttt{EXTRA\_CFLAGS}.
It contains dependence rules for 
\texttt{BINDIR},
\texttt{OBJDIR}. Seems that there is no call to Makefile.rules.

\paragraph*{Notes:} the Mathlink stuff is actually copied in the subdirectory
\$MMALPHA/sources/Mathlink.

This directory conaints a 
\texttt{MATHLINK} directory where all the stuff needed for the \mma{} interface
is put. The 
\texttt{Obj.OSTYPE} directory is where the object files will be put.

\section{\texttt{ReadAlpha}}

\begin{description}
\item First line includes Makefile.config
\item[\texttt{CPPFLAGS}] = -I\$(POLYDIR) -g (flags for cc)
\item[LOADLIBES] = empty
\item[DUMMY] = list of files to be destroyed when cleaning
\item[NAME] = the name of the program, read\_alpha
\item[OBJS] = the list of object files, \$(OBJDIR)/read\_alpha.o, node.o, \$(POLYOBJS)
\item[LIBS] = libraries to link again
\end{description}

These definitions are followed by special dependence rules, and 
also include the general rules in MakeIncludes/Makefile.rules. Finally, 
there is a file
\texttt{Makefile.checkvars} which contains the rules for 
checking variables. All variables are not checked.

\subsubsection{WriteAlpha}

\begin{description}
\item First line includes Makefile.config
\item[\texttt{CPPFLAGS}] = -I\$(POLYDIR)
\item[\texttt{LDFLAGS}] = -L\$(LIBDIR) \# -mwindows
\item[\texttt{LOADLIBES}] = empty
\item[\texttt{DUMMY}] = list of files to be destroyed when cleaning
\item[\texttt{NAME}] = the name of the program, read\_alpha
\item[\texttt{OBJS}] = the list of object files, \$(OBJDIR)/yacc.o, nodeprocs.o, node2item.o, \$(POLYOBJS),
\$(POLYMSG) \$(PRETTYOBJS)
\item[\texttt{LIBS}] = libraries to link again, i.e, \$(LIBDIR)/libpoly.a, libpretty.a and libpolymsg.a
\end{description}

These definitions are followed by special dependence rules, and 
also include the general rules in MakeIncludes/Makefile.rules. Finally, 
there is a file
\texttt{Makefile.checkvars} which contains the rules for 
checking variables. All variables are not checked.

\subsubsection{WriteC}

\begin{description}
\item First line includes Makefile.config
\item[\texttt{CPPFLAGS}] = -I\$(POLYDIR)
\item[\texttt{LDFLAGS}] = -L\$(LIBDIR) \# -mwindows
\item[\texttt{LOADLIBES}] = empty
\item[\texttt{DUMMY}] = list of files to be destroyed when cleaning
\item[\texttt{NAME}] = the name of the program, read\_alpha
\item[\texttt{OBJS}] = the list of object files, \$(OBJDIR)/yacc.o, nodeprocs.o, node2item.o, \$(POLYOBJS),
\$(POLYMSG) \$(PRETTYOBJS)
\item[\texttt{LIBS}] = libraries to link again, i.e, \$(LIBDIR)/libpoly.a, libpretty.a and libpolymsg.a
\end{description}

These definitions are followed by special dependence rules, and 
also include the general rules in MakeIncludes/Makefile.rules. Finally, 
there is a file
\texttt{Makefile.checkvars} which contains the rules for 
checking variables. All variables are not checked.

\section{List of variables, and where they are defined}

In Makefile.config, there is an inclusion of the definitions in MakeIncludes/Makefile.\$(OSTYPE)
which are peculiar to each OS type.

\begin{description}
\item[\texttt{BINDIR}]: in Makefile.config. \$(MMALPHA)/bin.\$(OSTYPE). Where the binaries will be
put. 
\item[\texttt{CC}]: in Makefile.config. I guess it is the compiler...
\item[\texttt{CFLAGS}]: in Makefile.config. -O3 -g -I\$(POLYINCLUDE) \$(PRECISIONFLAG). I guess that
these are the flags of the compiler.
\item[\texttt{CODEGENDIR}]: in Makefile.config. \$(SOURCES)/Code\_Gen. Where Code\_Gen is
\item[\texttt{CODEGENOBJS}]: in Makefile.config. \$(OBJDIR)/gen.o and nodeprocs.o. The objects of 
Code\_gen.
\item[\texttt{CXX}]: in Makefile.config g++
\item[\texttt{DEFINES}]: in Makefile.OSTYPE. 
\item[\texttt{DIR}]: in Makefile. The list of directories where there are files to be compiled. 
\item[\texttt{LDLIB}]: in Makefile.config. ld -r. It is the command for the linker.
\item[\texttt{LEX}]: in Makefile.config. flex
\item[\texttt{LFLAGS}]: in Makegile.config. Flags for flex
\item[\texttt{LDFLAGS}]: in Makegile.OSTYPE. Flags for the loader.
\item[\texttt{LIBDIR}]: in Makefile.config. \$MMALPHA/lib.\$(OSTYPE). Where the libraries are put.
\item[\texttt{LPSOLVEDIR}]: in Makefile.config. Where LP solve is, unused currently.
\item[\texttt{OMEGALIB}]: in Makefile.config. Where the OMEGA library is. unused currently.
\item[\texttt{MMALPHA}]: environment variable. May also be wired by a definition in Makefile.config
or in a system config file.
\item[\texttt{MMAVERSION}]: environment variable. May also be wired by a definition in Makefile.config
or in a Makefile.ostype.version file. Defines the version number of Mathematica. 
\item[\texttt{MAKEFILES}]: in Makefile.config. Contains the list of makefiles, i.e. Makefile itself (in the 
sources directory), Makefile.rules, Makefile.config, and Makefile.\$(OSTYPE) located in 
the directory MakeIncludes.
\item[\texttt{NEEBINDIR}]: in Makefile.config. \$(BINDIR)/.make. This variable is checked by a rule in
Makefile.rules. 
\item[\texttt{NEEDLIBDIR}]: in Makefile.config. \$(LIBDIR)/.make. Use?
\item[\texttt{NEEDOBJDIR}]: in Makefile.config. \$(OBJDIR)/.make. Use?
\item[\texttt{OBJDIR}]: in Makefile.config. Obj.\$(OSTYPE). Defines where the object files will be put.
This is relative to the directory of the application.
\item[\texttt{OMEGADIR}]: in Makefile.linux. Directory of the OMEGA software, not used
currently.
\item[\texttt{OSTYPE}]: environment variable
\item[\texttt{POLYDIR}]: in Makefile.config. \$(SOURCES)/Polylib. Where the source files of the polylib are.
\item[\texttt{POLYOBJS}]: in Makefile.config. The list of object files of the polylib. These files are all in 
\$(OBJDIR). I changed this line in July 2007, I do not know why... Maybe to add object files that were
needed...
\item[\texttt{POLYXOBJ}]: in Makefile.config. Another set of object files
\item[\texttt{POLYMSG}]: in Makefile.config. \$(OBJDIR)/errormsg.o Defines an object file where the 
error messages are defines for Polylib.
\item[\texttt{PRECISIONFLAGS}:] in Makefile.linux. Check this. 
\item[\texttt{PRETTYDIR}]: in Makefile.config. \$(SOURCES)/Pretty. Where Pretty is.
\item[\texttt{PRETTYOBJS}]: in Makefile.config. $(PRETTYDIR)/$(OBJDIR)//itemprocs.o and writeitem.o. 
The objects of Pretty.
\item[\texttt{SHELL}]: in Makefile.config. /bin/sh
\item[\texttt{SOURCES}]: in Makefile.config. \$(MMALPHA)/sources. The source files of MMAlpha.
\item[\texttt{STRIP}]: in Makegile.config. strips. Meaning?
\item[\texttt{YACC}]: in Makefile.config. bison
\item[\texttt{YFLAGS}]: in Makefile.config. Flags of Yacc
\item[\texttt{MATHLINK}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHPREP}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHINCLUDE}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHLIB}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHLOADLIB}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHEXTRALIBS}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{MATHEXTRAOBJS}]: in MakeIncludes/Makefile.OSTYPE.number.
\item[\texttt{EXTRA\_CFLAGS}]: in Domlib/Makefiles.darwin.
\item[\texttt{CPPFLAGS}]: in ReadAlpha, WriteAlpha, WriteC
\item[\texttt{LOADLIBES}]: in ReadAlpha, WriteAlpha, WriteC
\item[\texttt{DUMMY}]: in ReadAlpha, WriteAlpha, WriteC
\item[\texttt{NAME}]: in ReadAlpha, WriteAlpha, WriteC
\item[\texttt{OBJS}]: in ReadAlpha, WriteAlpha, WriteC
\item[\texttt{LIBS}]: in ReadAlpha, WriteAlpha, WriteC
\end{description}

\appendix

\section{Installing \polylib{}}
Look in MMALPHA/doc/Polylib/note.pdf for more details. I describe here the full 
procedure, but in the current distribution, \polylib{} version 5.22.3 is already
present in the distribution: you may then skip steps 1 and 2.

\begin{enumerate}
\item Get polylib from Strasburg site (http://icps.u-strasbg.fr/polylib/)
\item Copy it in \texttt{\$MMALPHA/sources}
\item Gunzip it and untar it (double click, or gunzip)
\item The name of this polylib should be something like \texttt{polylib-5.22.3}. Go in this
directory (in a shell).
\item Run the following commands:\label{trois}
\begin{verbatim}
./configure --prefix="$MMALPHA/sources/Polylib"
make
make install
\end{verbatim}
\item Run \texttt{make test} to check that everything is OK
\end{enumerate}

During step~\ref{trois}, the value of the environment variable
\texttt{\$MMALPHA} should be the path of your \mma{} installation.

When \polylib{} is installed, make sure that the value
of \texttt{POLYDIR} in file \texttt{Makefile.config} is that 
of the path to the \polylib{} (as the source files of \polylib{} are needed to
compile the C programs of \mma{}), and that

\section{Introduction}
This document explains how the \domlib{} library is installed, in this directory. It was
written while porting the \domlib{} on Mac OS X.

\section{To be checked}
Is your Polylib up-to-date? Look in sources, where you must find both Polylib and
Poly (where the files produced by the make command of Polylib are).

\section{Content of the directory}
The directory
contains the source, \texttt{domlib.c}, a make file, \texttt{Makefile}, the \ML{} template file 
\texttt{domlib.tm}. The make creates an object file \texttt{domlib.o}, a binary 
file \texttt{domlib}. On the fly, it translates the template file into a source file
\texttt{domlibtm.c}.

The file \mkf{} calls another make file, located in:
\begin{verbatim}
$MMALPHA/sources/MakeIncludes/Makefile.$OSTYPE
\end{verbatim}
In this make file, the location of various directories or programs are set. In particular, 
that of MLINKDIR.

To run properly, the Polylib must have been compiled previously. See documentation
on Polylib, but in a few words, just go in PolyLib and compile it by specifying
that the directory is Poly. This location is defined in Makefile.darwin, and it may be
changed. The commands to be done, once you have copied the gziped filed 
obtained from strasburg are:
\begin{verbatim}
./configure --prefix="$MMALPHA/sources/Poly"
make all
make install
\end{verbatim}

\section{The rational of the installation}
To be compiled, \domlib{} needs a \ML{} preprocessor called \mprep{}, and the
\polylib{} library. \ML{} compiler additions are also needed.

On this machine, \mma{} is located in\footnote{This depends on the current version of \mma{} that you have.}
\begin{verbatim}
/Applications/Mathematica 5.2.app/
\end{verbatim}
The variable \texttt{MLINKDIR} contains the location
\begin{verbatim}
/Applications/Mathematica 5.0.app/AddOns/Mathlink/DeveloperKit/MacOSX-x86
\end{verbatim}
On other systems, this location may change. If you do not have an Intel Processor, you
may have to switch \texttt{MLINKDIR} to either 
\begin{verbatim}
/Applications/Mathematica 5.0.app/AddOns/Mathlink/DeveloperKit/Darwin
\end{verbatim}
or
\begin{verbatim}
/Applications/Mathematica 5.0.app/AddOns/Mathlink/DeveloperKit/Darwin-PowerPC64
\end{verbatim}

In the \mkf{} , variables contain the system name, the place of the \polylib{} header files, 
the place of the \polylib{} library. The compiled additions are then located. 
Variable \texttt{POLYLIB\_INCLUDE} contains the directory where the include
files of the \polylib{} are. Similarly variable \texttt{POLYLIB\_LIB} 
contains the directory where the libraries of the \polylib{} are.

Variable \texttt{CADDSDIR} gives the place where the compiler additions of 
\mma{} are.

Variable \texttt{EXTRA\_CFLAGS} is set to 
\begin{verbatim}
-DLINEAR_VALUE_IS_LONGLONG -DUNIX
\end{verbatim}

Variable \texttt{INCDIR} is set to {CADDSDIR}.
Variable \texttt{INCPOLYDIR} is set to {POLYLIB\_INCLUDE}.
Same for \texttt{LIBDIR} and \texttt{LIBPOLYLIB}.

Variable \texttt{MPREP} is set to where \texttt{mprep} is.

The \texttt{domlibtm.c}
file is obtained by running \mprep{} (provided by \mma{}) on \texttt{domlib.tm}.

The \texttt{domlib} binary file is obtained by first running \gcc{} on 
both \doml{} and \tm{} then 
linking the object files with the appropriate \polylib{} library. This library is 
itself installed {\em separately}.\footnote{See companion \polylib{} documentation.} 
In the current version, we used the 64 bit version, but other versions are 
available.

The form of the gcc command is 
\begin{verbatim}
gcc -DLINEAR_VALUE_IS_LONGLONG -DUNIX domlibtm.o domlib.o 
  -L${LIBDIR} -L${CADDSDIR}
  -L${POLYLIB} -lpolylib64 -lML -o $@ -v
\end{verbatim}

\section{Running the \domlib{} from \mma{}}
To run this library from \mma{}, the following conditions must be satisfied.
\begin{enumerate}
\item The \domlib{} binary file must be located in a directory whose
address appears on the \texttt{\$Path} variable of \mma{}. 
\item To load the library, run 
\begin{verbatim}
Install[ "domlib" ]
\end{verbatim}
in \mma{}. The list of available functions in the library are obtained by evaluating
\begin{verbatim}
?Dom*
\end{verbatim}
\end{enumerate}
In fact, this library is usually run within the \mmalfa{} software, using the 
\texttt{Domlib} package. 
\section{Perspectives}
A better way to organize the library would be to make sure that it is portable
accross platforms. To do so, we plan to build a version using the \texttt{autoconf}
and \texttt{automake} software, and more generally, the \texttt{gnu} building tools.

\section{Additional remarks}
In this directory, the installation is done using a make file. To compile for
Mac OS X, the UNIX macro should be defined, and it is by means of the 
EXTRA\_CFLAGS option -DUNIX.

\section{Mathematica at Irisa}
At Irisa, Mathematica is located in directory
\begin{verbatim}
/soft/Mathematica
\end{verbatim}
Several versions are available. The most recent one is \verb\V4.0.2\. It is located
in 
\begin{verbatim}
/soft/Mathematica/V4.0.2
\end{verbatim}
Three version are available : for \texttt{Linux}, \texttt{SGI} and \texttt{Solaris}. 
Inside the directory, there are subdirectories \texttt{CompilerAdditions} where one can find
the \texttt{mprep} pre-processor.

The transit.irisa.fr machine is a Linux machine. I could not compile on this architecture, 
for unknown reasons. (mprep does not work).

The europe machine is a solaris machine. I was able to compile. I had to use in Makefile.solaris the 
same as in Linux. In particular, the MATHEXTRALIB has to be -lnsl -l socket
I had to add the \texttt{\$Path} definition, which is not standard (init file?).

\section{The Make file for the MMAlpha distribution}
In sources, one find Makefile. 
This file defines LDLIB as ld -r, CC as gcc, CFLAGS as -03 -g -I\$(POLYINCLUDE) \$(PRECISIONFLAG).
It defines some variables for Yacc and Lex.

It includes the makefile.system which is in the same directory MakeIncludes, here
Makefile.darwin.
In Makefile.darwin, LD is set to gcc, PRECISIONFLAGS, LDFLAGS, MATHLINK, MATHPREP,
MATHINCLUDE, MATHLIB, MATHLOADLIB (-lML -lm) and MATHEXTRALIBS is set
to empty.

It then defines MAKEFILES, 
BINDIR, LIBDIR, OBJDIR, NEEDBINDIR, NEEDLIBDIR, NEEDOBJDIR, 
SOURCES,

For all binary files in DIR, goes in directory, then runs make.

In the Domlib directory, one finds the Makefile.darwin file. It first calls the Makefile.config
in the directory MakeIncludes.

\section{A new version of \domlib{} (May 6, 2008)}
It appears that the MathLink functions were modified between \mma{} version 4
and version 5. I thus had to modify Domlib accordingly. The current version of 
Domlib, compiled with the tools fo version 6, is correct for version 5 and 6. I did
not try yet to recompile it for version 5, to see it there are some differences. Actually, 
I guess that there are such differences.

The Mathlink documentation of version 6.0 is quite useful. Actually, the main differences
is in the functions \texttt{MLGetFunction} and \texttt{MLCheckFunction}, which do not
support the same arguments... Too bad !

I added 3 functions to \domlib{}: \texttt{mirrorDomain}, \texttt{mirrorMatrix} 
and \texttt{mirrorZDomain}. These function may be useful to see if there is something
wrong.

The old version of domlib is kept in the directory \texttt{sources/Domlib-Version4}. The source
files are those which were compiled for Powermac on version 5.

\section{Compiling \domlib{}}
In the \zdomlib{} file, one finds the \texttt{domlib.c}
file, which is the source file of \domlib{}, the 
\texttt{domlib.tm} file which is the patterns of 
the \domlib{} functions for \mathlink{}, and finally, 
several make files.

I guess that compiling \domlib{} is done by simply typing
\texttt{make} in the \zdomlib{} directory. But the best approach
is to run the make file that is contained in \texttt{\$MMALPHA/sources}.

A \texttt{README} file gives some explanations regarding \domlib{}.
Beware that only the first few lines seem to be uptodate.

\subsubsection{The \texttt{Makefile} file}
The file \texttt{Makefile} contains a switch to make files that 
depend on the OS type (this one should be, I guess, in the 
environment variable \ostype{}). For example, for 
\cygwin{}, the corresponding make file is \texttt{Makefile.Visual}.

\subsubsection{The \texttt{Makefile.Visual} file}
The first thing that this make file does is to include the 
configuration make file, the name of which is \texttt{Makefile.config} and
which is located in the \envmmalpha{}\verb\MakeIncludes\ directory.

Therefore, you will have to look at the documentation of this
make file given in~\ref{doc-config-make} if you want to 
understand everyhing.

In addition, the configuration make file calls other make files 
situated in the same directory...

Coming back to \texttt{Makefile.Visual}, it defines 
the following variables:
\begin{itemize}
\item \varcc{}: the compiler, here \texttt{cl};
\item \varld{}: the linker, \texttt{link};
\item \varpolylib{}: where the \polylib{} is (it seems to me that
this variable is already set);
\item \varcflags{}: redefines the compilation flags;
\item \varlflags{}: refedines the link flags;
\item \varbin{}: gives the name of the binary file for domlib;
\item \varobj{}: where are the object codes. Here, they are in 
a directory \texttt{Obj.cygwin} located in the same directory.
\item \vardepobj{}: a list of object files which are going to be
generated by the compiling.
\end{itemize}
Then one finds some rules to create \domlib{}. The \texttt{all} rule
is for creating binary files. I do not understand totally the other 
rules...

One particular rule allows the \texttt{mlink.c} file to 
be created using \mmalpha{}'s tools from \texttt{domlib.tm}.

\subsubsection{The \texttt{Makefile.linux} file}
I am not sure that this file works properly. It defines 
\varcppflags{}, \varldflags{}, \varloadlibes{}, \vardummy{},
\varname{}, \varobjs{}, \varlibs{}. It uses \varpolydir{}, 
\varmathinclude{}, \varlibdir{}

\section{Conclusion}

\printindex

\newpage
\tableofcontents

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Developers/Writing-Tests/writing-a-test-file.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\begin{document}

\newcommand{\domlib}{{\sc domlib}}
\newcommand{\ML}{{\sc Mathlink}}
\newcommand{\mprep}{{\sc mprep}}
\newcommand{\polylib}{{\sc polylib}}
\newcommand{\mma}{{\sc Mathematica}}
\newcommand{\mkf}{{\texttt Makefile}}
\newcommand{\gcc}{\texttt{gcc}}
\newcommand{\doml}{\texttt{domlib.c}}
\newcommand{\tm}{\texttt{domlibtm.c}}
\newcommand{\mmAlpha}{\textsc{MMAlpha}}
\newcommand{\MMAlpha}{\textsc{MMAlpha}}
\newcommand{\MMA}{\textsc{Mathematica}}
\newcommand{\true}{\texttt{True}}
\newcommand{\false}{\texttt{False}}

\title{Writing a test file for \mmAlpha{}}
\author{Patrice Quinton}
\date{Version of \today}
\maketitle

\section{Introduction}
Testing programs is recommended by everybody... and done only
by a few people. \mmAlpha{} is no exception...

The purpose of this document is to explain what policy
is to be followed when one wants to develop a new package for
\mmAlpha{}.

A companion document, located in \texttt{\$MMALPHA/doc/Tests}
explains how to test the \mmAlpha{}.

In a few words, here are the recommendations:
\begin{enumerate}
\item All packages must be provided with a \MMA{} test file. 
Let \texttt{myPack.m} by a brand new \MMAlpha{} package (the
same policy applies to C program). Then, a \MMAlpha{} program, 
\texttt{TestmyPack.m} must be written, and placed in the 
directory \texttt{\$MMAlpha/tests} in a 
new directory \texttt{TestmyPack}. If you follow these
recommendations, your test program will be called when evaluating
the \MMAlpha{} expression \texttt{tests["myPack"]}. 
\item Your test program, \texttt{TextmyPack.m} must 
return a value \true{} if all tests
have been passed, \false{} otherwise. 
\item Test your test program! A good policy is to incorporate
all examples you have used when developing your code. 
Remember that your tests are going to be run among 
several thousands, so make sure that your tests are 
correct.
\end{enumerate}
In what follows, we present quickly facilities provided
for testing by \MMAlpha{}, and we give an example.

\section{Functions for testing}
Two functions are used: \texttt{tests}, and \texttt{testFunction}. These
functions were improved in the V2 version of \MMAlpha{}.

\subsection{The \texttt{tests} function}
This function is part of the \texttt{Alpha.m} package. 
Evaluating \texttt{tests[]} runs {\em all test files which
are in the test directory}. This may take a few hours... We
use it only to check a new version\footnote{This may change...}.

Evaluating \texttt{tests["myPack"]} runs the test program
for package \texttt{myPack}. You can therefore run tests 
for various \MMAlpha{} packages, for example, \texttt{tests["Alpha"]}
(try it!). 
Notice that \MMA{} error messages may happen during the test; this
does not mean that the test has failed. Indeed, some test programs
test that \MMAlpha{} detects abnormal situations.

\subsection{The \texttt{testFunction}}
A function, called, \texttt{testFunction}, is provided in 
the \texttt{Alpha.m} package, as a means to organize tests
in a systematic way. Using this function is recommended. Its operation
is as follows. 
\begin{verbatim}
testFunction[exp, result, testID]
\end{verbatim}
compares the evaluation of \texttt{exp} to \texttt{result}.
It returns 
\texttt{True} if they are equal, 
\texttt{False} and emits a message if they are different.
\texttt{testID} is a string which identifies the test.

This function can be used in many different ways. For example, 
here is a test of the \texttt{Alpha.m} package, which checks 
the \MMAlpha{} parser: 
\begin{verbatim}
testFunction[load["Test1.alpha"];$result,<<"RTest1","Alpha 1"]
\end{verbatim}
The evaluation of this expression loads the file \texttt{Test1.alpha}
which is supposed to be in the test directory, and compares the 
value of \texttt{\$result}, the \MMAlpha{} variable which contains
the current AST, with the content of the file \texttt{RTest1}, 
in the same directory. This test is identified as \texttt{"Alpha 1"}.

\section{An example}
Fig.~\ref{ex} gives an example of test file. 
\begin{figure}[htbp]
\begin{verbatim}
Module[{dir, res1, testresult},

  dir = Directory[];  (* Save current directory *)

  (* Go in test directory *)
  SetDirectory[Environment["MMALPHA"]<>"/tests/TestAlpha/"];
  Print["Test for Alpha.m"];

  (* Build a list of boolean values *)
  res1= 
  {
    testFunction[load["Test1.alpha"];$result,<<"RTest1","Alpha 1"],
    testFunction[load[22],Null,"Alpha 2"],
    testFunction[Check[show[],$Failed],Null,"Alpha 3"],
    testFunction[show[22],Null,"Alpha 4"],
    (* ... *)
    ReadList["MV1.c",Record,RecordSeparators->{}]===
      ReadList["MV1.c",Record,RecordSeparators->{}],True,"writeC 1"],
    DeleteFile["MV1.c"];True
  }

  (* And the list *)
  testResult = Apply[And,res1];

  (* Back to initial dir *)
  SetDirectory[dir];

  (* Diagnostic *)
  If[ testResult,
    Print["**** Test OK for Alpha.m "],
    Print["**** Something was wrong for Alpha.m"]];

  (* Do not forget to return the result *)
  testResult
] (* End module *)
\end{verbatim}
\caption{Example of test, adapted from the file 
\texttt{Alpha.m}}\label{ex}
\end{figure}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Distribution/Distribution.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}
\begin{document}

\newcommand{\mma}{\texttt{Mathematica}}
\newcommand{\mmalfa}{\texttt{MMAlpha}}

\title{Note about the Distribution of \mmalfa{}\footnote{Last update of this note on \today{}}}
\author{Patrice Quinton}
\date{January 2007}
\maketitle
\section{Introduction}
\label{introduction}
There is in the main directory a file named \texttt{CopyMMA.m}, which can be used to 
create a distribution of \mmalfa{}.

\section{How to run in}
To run this program, do the following:
\begin{enumerate}
\item Start \mma{} and start \mmalfa{}.
\item Make sure that you are in the \texttt{\$MMALPHA} directory.
\item If needed, edit the \texttt{CopyMMA.m} file in order to define
the source directory and the destination directory of the distribution.
\item Type \texttt{<<CopyMMA}.
\end{enumerate}
This should make it.

\section{Technical Information}
The \texttt{CopyMMA.m} file is a \mma{} program. A few comments are in order.
\begin{enumerate}
\item The source directory and the destination directory are {\em hardwired}. To change them,
change the value of variables \texttt{sourceMMA} and \texttt{destMMA} (around line 80 of the
program). 
\item The program uses two functions: \texttt{copyF}, which copies a file, and 
\texttt{copyDir}, which copies a full directory. These functions are straightforward. 
\item While building the distribution, a file named \texttt{content.txt} is created
and updated in the main directory of the destination. This file contains the description
of the files that were copied.
\item The content of the distribution is {\em hardwired}, i.e. the choice of directories
and files is in the \texttt{CopyMMA.m} program. This file is itself copied in the distribution, 
in such a way that you can check what was this program when the distribution was created.
\item If you open \texttt{CopyMMA.m}, you will see that directories are either wholly copied
-- with possibly deleting some useless files, -- or partially copied, file by file. Which method
was used depends on the number of files that had to be copied.
\end{enumerate}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Doc-MMA-Darwin/Darwin-note.tex
Added mmalpha (source and binary)

 \documentclass[11pt]{article}
\begin{document}

\newcommand{\mmalfa}{{\sc mmalpha}}
\newcommand{\dar}{{\sc darwin}}
\newcommand{\mma}{{\sc mathematica}}

\title{About the Darwin version of \mmalfa{}\\
Version of \today{}}
\author{Patrice Quinton}
\date{}
\maketitle
This short note presents the current status of the new \dar{} version of 
\mmalfa{}.

\section{How to run it}
\begin{enumerate}
\item Set the shell to \texttt{tcsh} (I do not know how to use \texttt{bash}). To do so, 
start the Terminal application, and look in the preference: you should set the
initial command to \texttt{/bin/tcsh}.
\item Create a \texttt{.cshrc} file in your home directory with the definition 
of the \texttt{MMAlpha} environment variable (where MMAlpha is), the command
to extend the search path. For example:
\begin{verbatim}
setenv MMALPHA /Users/quinton/mmalpha
alias mma '/Applications/Mathematica\ 5.2.app/Contents/MacOS/Mathematica'
set path=($MMALPHA/bin.$OSTYPE $path)
\end{verbatim}
First command defines the \texttt{MMAlpha} environment variable, the second one says where
Mathematica is, and the third one adds to the \texttt{path} variable the 
location of the binary files of \mma{}.
\item Create an \texttt{init.m} file in your home directory. An example of such a file is
in appendix~\ref{init}.
\item Add an alias with the location of Mathematica (see the \texttt{.cshrc} above.
\item Open a terminal window. This should run the \texttt{.cshrc} file 
\item Type \texttt{mma}.
\end{enumerate}

That should make it! Actually, it does not : the \texttt{init.m} file does not start...

\section{Compiling for \dar{}}

\section{Troubleshouting}
\texttt{init.m} does not start...

\appendix{}
\label{init}
\section{Typical \texttt{init.m} file}
There is no difference between this \texttt{init.m} file and 
another for Windows or Linux.
\begin{verbatim}
(* This is a simple init.m file *)
Module[ {env},
  Catch[
    env = Environment["MMALPHA"];
    If[ env === $Failed, 
          mmalpha::noenv = "The MMALPHA environment variable is not set";
          Throw[ Message[ mmalpha::noenv ] ]
    ];
    Get[ env<>"/config/init.m" ];
    Alpha`$myNotebooks = "/Users/quinton/myNotebooks";
    Alpha`$myMasterNotebook = "myMaster.nb";
    Print[ Alpha`$myNotebooks ];
  ]
]
\end{verbatim}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Install/How-to-install.tex
Added mmalpha (source and binary)

\section{How to install \MMAlfa{}}
%Initial version : Tanguy Risset
%Last Modification : Feb 2004 by Patrice Quinton

This section contains MMAlpha installation instructions for Unix users,
Windows NT or XP users, MacOsX and Linux users.\footnote{The source of this document
is in the \texttt{doc/Install} directory. It appears also as an appendix of the AlphaStart document.}

The \MMAlfa{} sofware can be downloaded at url \texttt{http://www.irisa.fr/cosi/ALPHA/}.
In case of problems, send a mail to \texttt{patrice.quinton@irisa.fr}~.

\subsection{Installing \MMAlfa{} on Unix-like systems} 
By Unix-like, we mean Unix, Mac OS X, or Linux. The following 
procedure has been checked for Unix and Mac OS X, not for Linux.
\begin{enumerate}
\item 
\label{stepOneMac}
This first action must be done only if 
\MMAlfa{} has not yet been installed on the server 
or the computer you
are using. If \MMAlfa{} is already installed, go to step~\ref{stepTwoMac}.

The current distribution, as obtained from the 
\MMAlfa{} web site, is a file named \texttt{mmalphaV2-1-0.zip}.
Double-click on this file to expand this directory.
You may also run the following command
\begin{verbatim}
unzip mmalphaV2-1-0.zip
\end{verbatim}
from a shell window.
%De-archive the complete distribution in some directory, say
%\texttt{/home/group/name/} (a directory called alpha is created)
%by executing the \texttt{install\_tar} script available at 
%\begin{verbatim}
%http://www.irisa.fr/cosi/ALPHA/DOWNLOAD
%\end{verbatim}
\item
\label{stepTwoMac}
Create a \texttt{MMALPHA} environment variable containing the
path of the directory where \mmalfa{} has been installed.
If you use a \texttt{csh} shell, you have to type:
\begin{verbatim}
setenv MMALPHA path-of-mmalpha
\end{verbatim}
and for a \texttt{bash} shell:
\begin{verbatim}
export MMALPHA=path-of-mmalpha
\end{verbatim}
These commands may alternatively be added to your \texttt{.cshrc}
or to your \texttt{.bashrc} file (or to your \texttt{.login} file) in order to be
executed automatically when starting a shell. 
\item
\label{stepThreeMac}
You have then to set 
the \texttt{PATH} environment variable to contain the directory
where the binary files of \mmalfa{} are. On Mac OS X, this directory
is \texttt{path-of-mmalpha/bin.darwin}. 
For a \texttt{csh} shell:
\begin{verbatim}
setenv PATH ${MMALPHA}/bin.darwin:${PATH}
\end{verbatim}
and for a \texttt{bash} shell:
\begin{verbatim}
export PATH=${MMALPHA}/bin.darwin:${PATH}
\end{verbatim}

Again, you may add this command to your \texttt{.cshrc} or
\texttt{.bashrc} file.
%\item 
%Execute the initialization script 
%that is now in \texttt{\$MMAlpha/init\_user.csh}
%(or in \texttt{\$MMAlpha/init\_user.bash} if you use the 
%\texttt{bash} shell) by the command:
%\begin{verbatim}
%source $MMALPHA/config/init_user.csh
%\end{verbatim}
%(or 
%\begin{verbatim}
%source $MMALPHA/config/init_user.csh
%\end{verbatim}
%for the \texttt{bash} shell.)
%The only files that should be edited by you are the
%files in directory \texttt{\$MMALPHA/config/}.
\item
\label{stepThree}
Copy  the  file \texttt{\$MMALPHA/config/init-for-you.m} under the 
name \texttt{init.m} to 
your \mma{} base directory\footnote{The \mma{} base directory
used to be the user home directory in \mma{} versions 
prior to version 5. Under \mma{}, the home directory
is given by evaluating the \texttt{\$HomeDirectory} \mma{} variable.
Since version 5 (see documentation of \mma{}), 
the \mma{} base directory is given by the
\texttt{\$UserBaseDirectory} variable. For example, in 
MacOS X, 
it is located in \texttt{\~/Library/Mathematica/Kernel} directory:
this is where you should put the \texttt{init.m} file.}
(or append it to your base \texttt{init.m} file if it already 
exists). 
For users unfamiliar with Mathematica, recall
that this \texttt{init.m} file is executed whenever Mathematica's kernel
is launched.  
\item
\label{stepFour}
It's ready! Run \mma{} by typing in a shell window the command:
\begin{verbatim}
mathematica
\end{verbatim}
and once a notebook is started, type and evaluate in it the command
\begin{verbatim}
start[]
\end{verbatim}
to launch the master notebook of \MMAlfa{}. From this master notebook, examples
and explanations are available.
\item To check that everything is OK, evaluate successively the following commands
\begin{verbatim}
test1[]
test2[]
test3[]
test4[]
\end{verbatim}
Each one of these command starts a set of tests and shoul return the value
\texttt{True}. If this is not the case, something in your installation is wrong. 
In case of problems, see Section~\ref{troubles}.

Notice that executing these
commands generates a lot of error messages, and even, unexpected messages
in the shell windows: as long as the final result of the test is \texttt{True}, this is
not a problem. 
\end{enumerate}

\paragraph*{Remarks}
\begin{itemize}
%\item The \texttt{PATH} location may change for the now
%\texttt{intel-pc} processors. I have not yet decided where it should be.

\item On MacOS X, 
\mma{} should be started from the Terminal application, as
otherwise, the \texttt{\$MMALPHA} environment variable 
cannot be set (at least, I do not know how to set this variable...). 
Moreover, after installing \mma{}, it is needed to add an alias
to start \mma{}. Usually, \mma{} is installed in directory \texttt{/Applications}, for
example : 
\begin{verbatim}
/Applications/Mathematica\ 5.1.app/Contents/MacOS/Mathematica
\end{verbatim}
So, you may add an alias in your \texttt{.cshrc} file. 
\end{itemize}

\subsection{Installing MMALPHA on Windows NT or Windows XP}
\paragraph*{Warning:} version V2 has not been tested on Windows NT, since I do not have
currently access to such a configuration. PQ, Jan. 2, 2009.
\begin{enumerate}
\item 
\label{stepOneWin}
This first action must be done only if 
\MMAlfa{} has not yet been installed in the server you
are using. If \MMAlfa{} is already installed, go to step~\ref{stepTwoWin}.
The current distribution is a file \texttt{mmalphaV2-1-0.zip} file. Double-click on this file to expand this directory.
\item
\label{stepTwoWin}
You have to set two environment variables: \texttt{MMALPHA} and 
\texttt{Path}. To do so:
\begin{enumerate}
\item
Open the system configuration panel (start \texttt{->} configuration panel)
\item Double click on the "System" icon.
\item Choose the "Advance" panel.
\item Click on "Environment variables".
\item Consider the user variable panel (top part).
\item Create a \texttt{MMALPHA} variable whose value is the path of the directory where
\MMAlfa{} is installed. Typically, the value of this variable should be 
\begin{verbatim}
C:\...\mmalphaV2-1-0
\end{verbatim}
To create such a variable, click on \texttt{New}, and fill 
the name and value. 
This step allows \mma{} to know
where \MMAlfa{} is located.
\item Append
\begin{verbatim}
;%MMALPHA%\bin.cygwin32
\end{verbatim}
to the user Path environment 
variable. To do so, select the \texttt{Path} variable, click \texttt{Modify}, 
place the cursor at the end of the string and type 
\begin{verbatim}
;%MMALPHA%\bin.cygwin32
\end{verbatim}
\footnote{Or possibly, cygwin.}This step allows the Domlib library
to be launched.
\end{enumerate}
\item		
\label{stepThree}
Copy  the  file \texttt{\$MMALPHA/init.m} to 
your \mma{} base directory\footnote{The \mma{} base directory
used to be the user home directory in \mma{} versions 
prior to version 5. Under \mma{}, the home directory
is given by evaluating the \texttt{\$HomeDirectory} \mma{} variable.
Since version 5 (see documentation of \mma{}), 
the \mma{} base directory is given by the
\texttt{\$UserBaseDirectory} variable. For example, in 
Windows,
it is probably located in \texttt{\~/Library/Mathematica/Kernel} directory:
this is where you should put the \texttt{init.m} file.}
(or append it to your base \texttt{init.m} file if it already 
exists). 
For users unfamiliar with Mathematica, recall
that this \texttt{init.m} file is executed whenever Mathematica's kernel
is launched.

\item
It's ready! Start \mma{}. The first evaluation in 
the initial notebook should start \MMAlfa{}. Normally, the
Messages window of \mma{} opens and contains a few
lines indicating that \mma{} was started successfully.
Evaluate start[] in any notebook to launch the master 
notebook of \mma{}.
From this master notebook, examples
and explanations are available.
\item To check that everything is OK, evaluate successively the following commands
\begin{verbatim}
test1[]
test2[]
test3[]
test4[]
\end{verbatim}
Each one of these command starts a set of tests and shoul return the value
\texttt{True}. If this is not the case, something in your installation is wrong. 
In case of problems, see Section~\ref{troubles}. 
Notice that executing these
commands generates a lot of error messages, and even, unexpected messages
in the shell windows: as long as the final result of the test is \texttt{True}, this is
not a problem. 
\end{enumerate}

\subsection{In Case of Problems}
\label{troubles}
Before reading this section, make sure that you have tried to evaluate
the \texttt{test1[]} through \texttt{test4[]} commands.

There is a \mma{} program called \texttt{diagnose.m}, located in 
\texttt{\$MMAlpha/config/diagnose.m}, that may help locate problems. 
To run it, start a fresh version of \MMAlfa{} (from the console window), 
then evaluate 
\begin{verbatim}
SetDirectory[ "/ your mmalpha dir/config"]
<<diagnose.m
\end{verbatim}
This will check various potential problem, report them in the \mma{} window, 
and write also a file \texttt{diagnosis.txt} in your home directory.

Here are a few difficulties you may encounter, and some ways to 
overcome them. Additional information is given in the test
documentation, in file
\begin{verbatim}
$MMALPHA/doc/Tests/testing-MMALPHA.pdf
\end{verbatim}

\subsection{The \texttt{MMALPHA} environment variable is not set}
To check this, evaluate
\begin{verbatim}
Environment[ "MMALPHA" ]
\end{verbatim}
in your notebook. If it does not answer the proper value, there
is a problem.

First, remember that you must start \mmalfa{} from a shell, otherwise, 
\mma{} does not inherit from your environment variables (this is true 
on Unix-like systems, not on Windows -like).

Second, it may be that your \texttt{init.m} file was not loaded: see next
section.

\subsubsection{The initialization file is not correctly installed}
Most of the problems come with the fact that
the \texttt{init.m} file may not be installed in the correct
directory. If this is the case, evaluating \texttt{test1[]} for
example will just return \texttt{test1[]} unevaluated, as
\mma{} was not able to load \mmalfa{}: indeed, loading
is done through the \texttt{init.m} file.

After launching \mma{}, make sure that the Kernel is also launched
by evaluating any \mma{} expression (for example, \texttt{2+2}). 
Before this evaluation is done, the \mma{}'s Kernel should 
evaluate the content of the \texttt{init.m} file, which results in 
a message in the \texttt{Messages} window of \mma{}. This
message ends with:
\begin{verbatim}
Alpha V2.0 Initialization
The Documentation can be found in ...
Current version in ...
Current directory is ...
If you use the notebook interface, you can open the master notebook: 
In[1]:= start[];
\end{verbatim}
If the Messages window is not opened, the \texttt{init.m} file was
not called and may not be in the proper directory. To check this, 
evaluate the expression \texttt{\$UserBaseDirectory} in \mma{}, 
and then check that the \texttt{init.m} file is in the 
\texttt{\$UserBaseDirectory/Kernel}.

\subsubsection{The Domlib does not start}
This happen if in the Messages window, the following 
message appears:
\begin{verbatim}
Warning: could not install domlib
\end{verbatim}

Also, if you run the \texttt{test1[]} command, 
you do have a \texttt{True} result.

Several reasons may be the cause of this problem.
\begin{enumerate}
\item Variable \texttt{\$PATH} may not contain the directory where
the Domlib is. The Domlib is in the directory \texttt{\$MMALPHA/bin.ostype},
where \texttt{ostype} is \texttt{darwin} for MacOS X, 
\texttt{linux} for Linux, and \texttt{cygwin32} for Windows. 
On Unix-like systems, check the value of this variable by
the command \texttt{echo \$PATH}. On Windows, check the value of
the \texttt{PATH} environment variable.
\item The binary file for the domlib may not be in the proper
directory. This means that \MMAlfa{} has not been installed 
properly, or altered. Again, check that the directory 
\texttt{\$MMALPHA/bin.ostype} contains \texttt{domlib}. If this is
not the case, unzip the \mmalfa{} distribution to get it (if you
do not find it in the distribution... just send me a mail !)
\item The \texttt{domlib} binary does not fit with your configuration. 
You then have to recompile it. I have not yet been able to configure
\mmalfa{} so that recompiling is easy, but I try. Please, refer
to the test documentation first, then to the documentation about
domlib.
\end{enumerate}

\subsubsection{Other problems}
You may encounter other problems, for example, some tests are corrects, 
some other are not. After the execution of the \texttt{test1[]} through
\texttt{test4[]} commands, you find a test report file in the 
\texttt{\$MMALPHA/tests} directory: send me this file, and I'll try
to find out what is wrong in order to help you.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Meta/bnfMeta.tex
Added mmalpha (source and binary)

<metarule> ::= <rule name> "::=" <abstract node> ":>" <semantics>
                       | <rule name> "::=" <switch_rule> ":>" <semantics>
<rule name> := <MMAsymbol>
<abstract node> ::= <MMAsymbol> "[" <list of arguments> "]"
<switch_rule> ::= <MMAsymbol> ":>" <list of switch arguments>
<semantics> ::= a call to a semantic function (*)
<argument> ::= <symbol> ":" <pattern>
                        | <MMAsymbol> ":" <rule name>
                        | <MMAsymbol> ":" "{" <rule name> "}"
<switch argument> ::= "_" <MMAsymbol> "->" <rule name>
						| <MMAsymbol> "->" "{" <rule name> "}"
<pattern> ::= any Mathematica pattern

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Meta/docMeta.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\usepackage{fullpage,epic,eepic,epsfig,makeidx,color,graphicx,moreverb,boxedminipage,version}
\sloppy

%\usepackage{fullpage,verbatim,vrbinput,vrbtabs,psfig,version,epic,eepic}
\usepackage{makeidx}
%\usepackage{french}
%\usepackage{mathematica}

\newcommand{\mmalpha}{{\sc mmalpha}}
\excludeversion{corrige}
\excludeversion{maversion}
\makeindex
\begin{document}

\title{The Meta Package}
\author{Patrice Quinton}
\date{Version of \today}
\maketitle
\abstract{This document presents the \texttt{Meta} package of \mmalpha{}. This
package allows one to define quickly translators for Mathematica expressions, in 
particular, \mmalpha{} Abstract Syntax Trees. Translators are specified by a set of rules
which are translated as a Mathematica package. When loaded, together with a 
separate semantic file, one gets a translator. The \texttt{meta} package is used in \mmalpha{}
to produce VHDL code. }

\section{Introduction}
The 
\texttt{meta}
 package allows one to define a set of meta rules for the translation of
 a Mathematica expression, and to translate these rules as a Mathematica
 package.
 The functions of this package can then be applied to an expression, in
 order to translate this expression into something useful.
 In the following, we explain how this meta translator operates, and we
 illustrate this by means of a few examples.
This package is used in various parts of \mmalpha{}.

Briefly speaking, creating a translator consists in:
\begin{enumerate}
\item writing a set of meta-rules in a \texttt{.meta} file,
\item writing a semantic file in a \texttt{.sem} file (not mandatory, but often convenient),
\item evaluating the \texttt{.meta} file with the \texttt{meta} function. This creates
a \texttt{.m} Mathematica package which contains the translator,
\item load the translator which then becomes available.
\end{enumerate}

The organization of this document is as follows. Section~\ref{define} presents
the syntax of meta-rules. In Section~\ref{using}, the use of the \texttt{meta} package is described. 
Options of the \texttt{meta} function are explained in Section~\ref{options}. Section~\ref{translation}
explains how meta-rules are translated. An example of translator is shown in 
Section~\ref{example}. The meta code and the semantic file of this translator are 
shown in Appendix~\ref{checkCellmeta} and Appendix~\ref{checkCellSem} respectively.

Directory \texttt{\$MMALPHA/doc/Packages/Meta} contains a notebook with 
examples. In particular, there is a nice translator of (a part of) Alpha to Mathematica.

You may access this notebook from the master notebook of 
\mmalpha{}\footnote{
Recall that the master notebook may be loaded by means of the command 
\texttt{on[]} from the Mathematica front-end.}
or by evaluating the expression
\begin{verbatim}
docLink["Meta"]
\end{verbatim}
in any notebook. This evaluation produces a button which opens the 
notebook.

\section{How to define meta rules}
\label{define}
The input file contains a list of meta-rules.
 Let 
\texttt{myTranslator.meta}
 be this file.
 A meta rule has the following syntax (in BNF form):
\verbatiminput{bnfMeta.tex}

A rule therefore describe a so-called 
{\em abstract node}, or a 
{\em switch rule}.

\subsection{Abstract nodes}
Abstract nodes represent expressions which correspond to an internal node
 of the abstract syntax tree.
 For example, the first level of an Alpha program has the form 
\texttt{system[n,p,id,od,ld,e]}, where 
\texttt{system} is the head, 
\texttt{n} is a string representing the name of the system, 
\texttt{p} is the parameter, 
\texttt{id} is the list of input variable declarations, 
\texttt{od} is the list of output variable declaration, 
\texttt{ld} the local declarations, and 
\texttt{e} the list of equations.
Such a node is described by the following meta-rule: 
\begin{verbatim}
SYSTEM ::= 
  system[ n:_String, p:DOMAIN, id:{DECLARATION}, od:{DECLARATION}, 
  ld:{DECLARATION}, e:{EQUATION} ] :> semantics...
\end{verbatim}
The left-hand-side of the rule, 
\texttt{system}, is the name of the rule.
 Following the 
\texttt{::=}, is the pattern which describes the form of the abstract node, followed,
 after the symbol 
\texttt{:>} by the semantics associated to this node.

The first argument of this pattern, 
\texttt{n:\_String}, has the form 
\texttt{<symbol>:<pattern>}.
 This pattern will be used to check that the corresponding terminal element
 in the AST is a string.
 If not, an error message will be issued.
 The 
\texttt{n} symbol names the argument which matches this position, and can be used
 in the semantics of the rule as explained later.

The second argument, 
\texttt{p:DOMAIN}, has the form 
\texttt{<symbol>:<rule name>}.
 It generates recursively a call to the meta-rule 
\texttt{DOMAIN.}
The subexpression which matches this second position is named 
\texttt{p}.
 The next argument, 
\texttt{id:{DECLARATION}}, is treated in a similar way, except that one expect a list of expressions
 matching the 
\texttt{DECLARATION} meta-rule.

\subsection{Switch rules}
\label{switchrules}
Switch rules allow alternate abstract rules to be called, depending on a
 pattern.
 For example, the following meta-rule is used to select the various forms
 of Alpha expressions:
\begin{verbatim}
EXPRESSION ::= 
  {_binop -> BINOP, 
   _var -> VAR, 
   _affine -> AFFINE, 
   _restrict -> RESTRICT 
  }
\end{verbatim}

The arguments of this rule are particular cases of Mathematica patterns,
 namely, 
\texttt{<typed\_blanks>}.
 What we call here a typed\_blank is a Mathematica expression of the form
 \texttt{\_type}, which matches any Mathematica expression whose head is 
\texttt{type}.

This rule can be read as follows: if the head of the current expression
 is 
\texttt{binop}, then apply meta-rule 
\texttt{BINOP}, if it is 
\texttt{var}, then apply meta-rule 
\texttt{VAR}, etc.

\subsection{Arguments}
\subsubsection{Arguments of abstract nodes}

The form of arguments depends on the type of subtree which is expected in
 the AST.
 If the subtree simply consists of a terminal element, a simple pattern
 is used, for example, 
\texttt{name:\_String}.

For non terminal subtrees, arguments of the form 
\texttt{<symbol>:<symbol>} or 
\texttt{<symbol>:{<symbol>}} are used.
 The latter form allows a list of expressions to be described.
 Finally, an argument of the form 
\texttt{<symbol>:{List}} generates a call to a particular function which expects a (Mathematica)
 list of objects, and returns this list unchanged.

\subsubsection{Switch arguments}

A switch argument has the form 
\texttt{\_Symbol -> <rule name>} or 
\texttt{\_Symbol -> {<rule name>}} .
 The 
\texttt{<rule name>} is the name of a rule which is selected by the head of the node, as specified
 by the pattern.
 As an example: 
\texttt{\_var -> varexpression} describes an alternative where the 
\texttt{varexpression} rule is called, when the head of the node is 
\texttt{var}.

\subsection{Semantic expressions}

The semantic part of a meta-rule is a Mathematica expression, involving
 symbols which name either arguments in the right-hand side part of the
 meta-rule, or results of the translation of these arguments by the semantic
 functions associated to the meta-rules.

Let us explain this on an example.
Consider the meta-rule:
\begin{verbatim}
EQUATION::= equation[c:_String, e:EXPRESSION] :> 
    semanticFunction[ equation, {c, tre} ] 
\end{verbatim}

The above rule describes the abstract node 
\texttt{EQUATION}.
 We expect this expression to contain a string, then an 
\texttt{EXPRESSION}, the latter being described by the 
\texttt{EXPRESSION} meta-rule.
 The string, when found, will be named 
\texttt{c}.
 The equation is named 
\texttt{e}.
 By convention, the result of the call to function 
\texttt{translateEXPRESSION} applied to 
\texttt{e} is named 
\texttt{tre} and can be used in the semantic part.
 The semantic expression, 
\texttt{semanticFunction[ equation, {c, tre} ]}, is a call to a function defined in a separate file, called the semantic
 file.
 There are absolutely no constraints in the way the semantic expression
 is defined.
 In this example, I choosed to name the functions I use 
\texttt{semanticFunction}, whose first parameter is the head of the expression, and the second one
 is the list formed by 
\texttt{c} and by the translation of the 
\texttt{EXPRESSION}.

\subsubsection*{Warning}
Although one can put any Mathematica expression as the semantic part of
 a meta-rule, calling a semantic function is a good way to avoid trouble.
 Indeed, when the meta translator reads the meta-rules, it uses the Mathematica
 interpreter which evaluates as many expressions as possible.
 Therefore, the result of the semantic expression might be very surprising.
 For example, lets assume that you write the semantic part
\texttt{c === "blabla"}
expecting the semantic of your rule to be 
\texttt{True} if 
\texttt{c} is equal to the string 
\texttt{"blabla"}, and 
\texttt{False} otherwise.
When the Mathematica parser reads this expression, it evaluates it, and
 the result of this evaluation is 
\texttt{False}, as the symbol 
\texttt{c} is different from the string 
\texttt{"blabla"}. The problem comes from the fact that we expected this
evaluation to happen later...
There are probably ways to avoid this kind of problem using complex Mathematica tricks (avoiding
Mathematica to evaluate an expression is sometimes hard...)

\section{Using Meta}
\label{using}
Assume you want to write a translator called 
\texttt{myTranslator}.
\begin{enumerate}
\item Write the meta-rules in a file named 
\texttt{myTranslator.meta}.
\item Write semantic functions (if needed) in a separate package named 
\texttt{myTranslator.sem} .
 This package should follow some rules which are described in appendix
\ref{checkCellSem}.
\item Load the \texttt{Meta.m} package (if it is not already loaded).
\item Evaluate the expression
\newline 
\texttt{meta[ "myTranslator", debug->True ]}
\item 
Load the file 
\texttt{myTranslator.m} which was created by 
\texttt{meta}.
\item To translate an expression \texttt{expr} of type 
\texttt{EXPRESSION}, evaluate
\newline 
\texttt{myTranslatorTranslateEXPRESSION[expr]}. 
Another better way is to introduce in the semantic package the function
 which you want to call to activate your translator on a tree. 
 See for example the semantic file given as an example in section \ref{checkCellSem}: 
 the semantic file contains the definition of a \texttt{checkCell} function.
\end{enumerate}

\section{Options of \texttt{meta}}
\label{options}
\begin{itemize}
\item \texttt{verbose}: if \texttt{True}, gives some information. Default value is 
\texttt{False}.
\item \texttt{debug}: if \texttt{True}, gives a lot of information. Default value is 
\texttt{False}.
\item \texttt{check}: if \texttt{True}, calls to functions are generated inside
a \texttt{Check} statement. This may help debugging the translator. Default value is 
\texttt{False}.
\item \texttt{directory}: gives the directory where the semantic file is to be
found and read. If \texttt{Null}, this directory is \texttt{\$MMALPHA/lib/Packages/Alpha}.
Default value is \texttt{Null}. 
\end{itemize}

\section{How meta-rules are translated}
\label{translation}
A meta rule named 
\texttt{RULE} is translated as a function 
\texttt{myTranslatorTranslateRULE}.
 This function is put in a package called 
\texttt{Alpha`myTranslator`}.
 In such a way, it is not possible to create conflicting names, provided
 the name of the translator is different from the names of the MMAlpha packages.

The package is equipped with a symbol called 
\texttt{myTranslatorDebug}, whose initial value is 
\texttt{False}.
 This symbol can be used as a debug flag by the semantic functions, in order
 to debug the translator.

The package 
\texttt{myTranslator.m} loads automatically the semantic file 
\texttt{myTranslator.sem} when it is itself loaded.

\section{An example}
\label{example}
Appendix \ref{checkCellmeta}
 shows the meta-rules for a program, named 
\texttt{checkCell}, which analyzes an Alpha 
program which can be interpreted as a 
\texttt{AlphHard} 
"cell", and reports errors if not.
A so-called cell is basically an Alpha program which describes a Register-Trans
fer-Level description of a piece of hardware (see
\cite{patricia}
 for details).

The meta-rules read an AST and check that this AST represents a cell.
 For example, such a program cannot contain use statements, etc.
 etc.

Appendix \ref{checkCellsem}
 presents the semantic file CheckCell.sem which is associated with the previous
 file.

\appendix

%\chapter{Appendix}

\section{Meta-rules for the checkCell translator\label{checkCellmeta}}

\verbatiminput{CheckCell.meta}

\section{Semantic file for the checkCell translator (part of)
\label{checkCellSem}}

\verbatiminput{CheckCell.sem}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/PeriodicSystems/PeriodicSystemGeneration.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}
\begin{document}

\newcommand{\mmalpha}{{\sc Mmalpha}}
\newcommand{\vhdl}{{\sc vhdl}}
\newcommand{\alfa}{{\sc alpha}}

\title{Periodic System Generation in \mmalpha{}}
\author{Anne-Marie Chana \and Patrice Quinton}
\date{Version of \today}
\maketitle

\tableofcontents

\section*{To do list}
\begin{itemize}
\item Separate the types in a separate component
\item Assign values to the coefficients of the filter
\item Infer le type de ovsfaddress etc... from that of the ROM
\item Before step 1, check that the system is well-formed
\item Step 2 assumes that there are no registers in the code, so check this...
\item During Step 3, we assume that the delay of a component is given by the schedule of
its first input, which is not necessarily the case... Check this and make an extension.
\end{itemize}

\section{Introduction}
This document presents the current state of development
of an extension of \mmalpha{} to generate multi-dimensional 
periodic systems. The theory underlying this work is described in~\cite{quinton-asap04}.

\section{The \texttt{synPeriodic} function}
This function is located in the \texttt{Synthesis.m} package. Its goal is to analyze a 
data-flow periodic systems and to find out the periods of each one of its components. 
It also generates \vhdl{} files for such a system.

Recall that such a system is a special case of \alfa{} system where all variables
are indexed with an infinite first dimension. This means that all variables of such
a system have a domain of the form \texttt{$\{ i, j, k ... | 0 \leq i , ... \}$}. The first
index, $i$, is considered as a time index.
It also uses subsystems, which have all 
the property that all their variables are also
indexed with a first infinite dimension.

Now the time dimension of each variable may be {\em stretched} or 
{\em contracted} depending on a fixed period. This means that for
some variables, the first index $i$ is in fact an affine function $a t + b$ of
an underlying, commun time clock, $t$. For another variable, this 
affine function might be of the form $c t + d$, where $c$ and $d$
are different from $a$ and $b$.

Some particular subsystems are oversamplers and undersamplers. They have
the property of changing the rate of the time dimension, i.e., the values of 
the periods $a$ or $c$ of the variables.

All other subsystems have the property of being {\em monochronous}: all
their variables have the same period. We expect these subsystems to represent
hardware components that may be run at various periods, depending on the 
way they are instanciated (we shall return to this point later on.)

In~\cite{quinton-asap04}, it is shown how one can compute the value of 
the periods of each individual signal or component of such a periodic
system, and from there, if a solution to this problem exists, to schedule in
detail the full system.

The \texttt{synPeriodic} function analyzes the graph of the \mmalpha{} data-flow system, 
and from its dependencies, it groups signals that have the same period (returned as 
list of list). Each one of the groups has un period, say $\lambda_i$, which is calculated
by this function. It then schedules the full system, and generates the \vhdl{} components
of it. We describe hereafter how this is done.

\section{Generation of \vhdl{} Programs}
\subsection{Generalities}
The generation of \vhdl{} programs is implemented as the 
\texttt{synPeriodic} \mmalpha{} function
of the \texttt{Synthesis.m} package.

The parameters of this function are not fully determined yet. 
\begin{verbatim}
synPeriodic[]
\end{verbatim}
generates a \vhdl{} file in the directory \texttt{\$MMALPHA/VHDL} and provides some
other results.

This function makes use of the \texttt{genVhdl} function, included in the 
\texttt{Vhdl2.m} package, and that we describe in appendix~\ref{genvhdl}.

Its operations are currently as follows:
\begin{description}
\item[Step 1:] Built the graph of periods, and solve the balance equations of this graph.
\item[Step 2:] Schedule the whole system, by injecting the value of the periods in the schedule
patterns of each component.
\item[Step 3:] Generate the delayed enable signals.
\item[Step 4:] Generate the enable signals.
\item[Step 5:] Generate the FSM that controls the periodic system.
\item[Step 6:] Generate the \vhdl{} code for the called systems.
\item[Step 7:] Generate the \vhdl{} code for the calling system.
\end{description}

We consider successively each one of these steps.

\subsection{How to Write a Periodic System}
Recall that all signals that are manipulated by periodic systems have a domain
whose first index is an infinite half-line. The other indexe can be anything 
(provided this matches the type of the subsystems).

A periodic system is made out of either:
\begin{itemize}
\item Connexions between periodic signals.
\begin{verbatim}
A = B;
\end{verbatim}
\item Combinatorial elements.
\begin{verbatim}
A = B + (C * D);
\end{verbatim}
\item Over-samplers.
\begin{verbatim}
use overSampling[6] (A) returns (B);
\end{verbatim}
The number inside brackets is the over-sampling factor. It is a fixed integer. 
Variable \texttt{A} is the input, and variable \texttt{B} is the output. 
\item Under-samplers.
\begin{verbatim}
use underSampling[6] (A) returns (B);
\end{verbatim}
Same remark as for the over-samplers.  
\item Calls to monochronous subsystems, among which, ROM and address generators.
\begin{verbatim}
use subsystem[ pm ] (A, B, etc.) returns (X, Y, etc.)
\end{verbatim}
\end{itemize}

\subsection{Step 1: Solve the Balance Equations}
Each variable and each component are assigned a period which is computed
by solving the so-called {\em balance equations}. The rules that are applied are
the following ones:
\begin{itemize}
\item Connected signals have the same period.
\item Inputs and outputs of a subsystem have the same period (subsystems are
assumed to be monochronous).
\item The period of the output of an over-sampler is equal to \texttt{x} times
that of its input, where \texttt{x} is the sampling factor.
\item The opposite is true for an under-sampler.
\end{itemize}
The algorithm consists in finding out the connected 
components of the dependence graph (whose elements have necessarily the same
period), and solve the balance equations. This returns one or several solutions.
Currently, only the case of one single solution is handled properly.

\subsection{Step 2: Schedule the System}
We assume that each subsystem has a parameterized schedule, which is put
in the schedule library. Building a parameterized schedule is explained in 
Appendix~\ref{paramschedule}.

Then, step 2 consists in scheduling the whole system.

\appendix
\section{The \texttt{genVhdl} Function}
\label{genvhdl}
The \texttt{genVhdl} functions allows one to generate component parts and 
architecture parts of several elements needed for the generation of 
periodic systems. Currently, this function allows the generation of:
\begin{itemize}
\item read-only memories,
\item finite-state machines,
\item periodic enable signals.
\end{itemize}

The principle of the generation is in all cases the same: it is made from \texttt{vhdl} pattern files
available in \texttt{\$MMALPHA/VHDL}. Place-holder are filled by the parameters given to 
\texttt{genVhdl}, depending on the type of \vhdl{} program to be generated. Place-holders
are character strings of the form \texttt{\$string\$}.

\section{Read-only Memories}
\label{gen-Rom}
Although this \vhdl{} element is not specific to periodic systems, we describe here the
use of \texttt{genVhdl} to generate a read-only memory.
\begin{verbatim}
 genVhdl["ROM" , "$wordLength$" -> "5",
             "$name$" -> "kasami", "$size$" -> "7",
             "$comment$" -> "Memory for Kasami coefficients",
             "$values$" -> "{1,1,0,0,1,8,2}"];
\end{verbatim}
The first parameter, \texttt{"ROM"}, identifies the type of block generated. Other parameters
are given as string replacement rules. The second parameter, gives the number of bits
of the words of the memory (it is mandatory). The third (mandatory) parameter is the name
given to the element. The fourth (mandatory) parameter is the number of words of 
the memory. A comment place-holder allows one to add a comment to 
the produced program. 
Finally, the last parameter is the values assigned to each word of the memory.
Its number should be equal to the size. There exist another place-holder in the 
ROM, namely, to define the address size, but it is computed automatically by the generator.

String rewriting rules can be given in any order.

This command, when executed, returns a list of 2 strings: the first one is the 
component corresponding to the ROM, and the second one is its architecture.

\begin{figure}[htbp]
\begin{verbatim}
COMPONENT kasami IS
  PORT 
  (
    address : IN  STD_LOGIC_VECTOR(3 DOWNTO 0);
    data : OUT STD_LOGIC_VECTOR(5 DOWNTO 0)
  );
END COMPONENT;
\end{verbatim}
\caption{Component of a ROM, as generated by the \texttt{genVhdl} 
command of~\ref{gen-Rom}.\label{figRom}}
\end{figure}

Figure~\ref{figRom} gives an example of memory generated.

\section{Periodic enable signals}
\label{gen-enable}
In periodic systems, each component is driven by a \texttt{clock-enable} signal
that controls its execution. The \texttt{genVhdl} function allows one to produce
a \vhdl{} component that generates a periodic signal that can be used to 
control such components. 
\begin{verbatim}
genVhdl[ "PeriodicEnable" ,  "$name$" -> "genEnable10",
     "$period$" -> "10",
     "$comment$" -> "Periodic enable generator with period 7"];
\end{verbatim}
The first component identifies the type of block produced. The second one, 
gives the name of the generated component. The third one gives the period
of the corresponding clock. Finally, the last one gives a comment.

Fig.~\ref{enable} shows an example of component for a 10 period enable.
Notice that this component has 3 inputs: the basic clock of the system, \texttt{clk}, 
a reset signal \texttt{rst}, and a {\em global enable} signal, \texttt{ceGen}. 
This signal, when false, has the effect of {\em freezing} the generation of 
the periodic enable during at least one \texttt{clk} clock cycle. 
\begin{figure}[htbp]
\begin{verbatim}
--
-- Component for a periodic enable of period 10
--

COMPONENT genEnable10 IS  
  PORT(
    clk : IN STD_LOGIC;
    rst : IN STD_LOGIC;
    ceGen : IN STD_LOGIC;
    periodicGe : OUT STD_LOGIC
  );
END COMPONENT;

\end{verbatim}
\caption{Component of a periodic enable signal generator, as produced by the
\texttt{genVhdl} command in~\ref{gen-enable}.\label{enable}}
\end{figure}

\section{Finite-state machine}
\label{gen-fsm}
The third element that the \texttt{genVhdl} function allows one to generate is 
a finite-state machine. Fig.~\ref{command-fsm} displays the form of the
\texttt{genVhdl} command to produce a finite-state machine. The first three 
parameters are position parameters: the first one identifies the type of block
generated, the second one gives the list (of strings) of the outputs of the 
fsm. The type of these outputs are assumed to be \texttt{STD\_LOGIC}. The 
third parameter describes the fsm. It is a list, the element of which are made
of an integer and a string. The integer gives the time when a given action
happens, and the string presents this action, in \texttt{vhdl} code. Usually, this
action involves setting values to the outputs.

The first element of this list should always be \texttt{\{ 0, "action" \}}, as
the initial state of the fsm is always named \texttt{state0}. 
\begin{figure}[t]
\begin{verbatim}
genVhdl[ "Fsm" ,
     {"fifo1_Out","fifo2_Out"},
     {
      {0,"BEGIN fifo1_Out <= '0'; fifo2_Out <= '1' END"},
      {4,"BEGIN fifo1_Out <= '1'; fifo2_Out <= '0' END"},
      {7,"BEGIN fifo1_Out <= '0'; fifo2_Out <= '1' END"},
      {22,"BEGIN fifo1_Out <= '1'; fifo2_Out <= '0' END"}
     },
     "$name$" -> "myFsm",
     "$comment$" -> "This is a Fsm"];
\end{verbatim}             
\caption{Command for generating a finite-state machine.\label{command-fsm}}
\end{figure}

Fig.~\ref{fsm} displays the component generated for a finite-state machine.
Notice that this fsm is controlled by a reset signal \texttt{rst} and a general
clock-enable signal \texttt{CE\_gen}. 
\begin{figure}[htbp]
\begin{verbatim}
--                                                                                 
-- This is a Fsm                                                                   
--                                                                                 
COMPONENT myFsm IS
  PORT(
    clk : IN STD_LOGIC;
    rst : IN STD_LOGIC;
    CE_gen : IN STD_LOGIC;
    fifo1_Out : OUT STD_LOGIC;
    fifo2_Out : OUT STD_LOGIC
  );
END myFsm;
\end{verbatim}
\caption{Component of a periodic enable signal generator, as produced by the
\texttt{genVhdl} command in Fig.~\ref{command-fsm}.\label{fsm}}
\end{figure}

\section{The FIR Filter}
The filter was generated using \mmalpha{}. It consists of a controller, a module
that instanciates three types of cells.

\subsection{The Module}
It has 6 parameters: the clock (\texttt{clk}), the clock-enable signal (\texttt{CE}), 
the reset signal (\texttt{Rst}), the coefficients (\texttt{wXMirr1In}), the x input (\texttt{xXMirr1In})
and the outputs (\texttt{Yout}).

Note: separate the package from the remaining of the filter.
Insert a return before the components...

The Module calls the controller, \texttt{ControlfirModule}. This controller has the
same signals (clock, clock-enable and reset) as the Module, and it returns a control
signal called \texttt{pipeCw7ctl1PXInit}.

It calls three types of cells. The simplest one is \texttt{cellfirModule4}, which 
does almost nothing: initialize the value of \texttt{Y} -- bounded to \texttt{Y4(0)} in the
call -- to 0, and transmits the
value of X (in the output signal \texttt{pipeCx5}, bounded to \texttt{pipeCx54(0)} in the
call.

The second cell is \texttt{cellfirModule1}. Besides the control signals (clock, clock-enable and
reset), it has four input signals: 
\begin{itemize}
\item the w coefficient (\texttt{wXMirr1}, binded to \texttt{wXMirr1(1)}),
\item the X input (\texttt{pipeCx5Reg3Xloc}), binded to \texttt{pipeCx5Reg3Xloc(1)}),
\item the Y input (\texttt{YReg5Xloc}) binded to \texttt{YReg5Xloc(1)}),
\item and
the pipe signal (\texttt{pipeCw7Xctl1PXInitXIn} binded to \texttt{pipeCw7Xctl1PXInitXIn(1)}).
\end{itemize}
It outputs similar signals:
\begin{itemize}
\item the w coefficient (\texttt{pipeCw7Xctl1P}, binded to \texttt{pipeCw7Xctl1P1(1)}),
\item the X output (\texttt{pipeCx5}), binded to \texttt{pipeCx51(1)}),
\item the pipe output signal (\texttt{pipeIOCw9} binded to \texttt{pipeIOCw91(1)}).
\item and the Y output (\texttt{Y}) binded to \texttt{Y1(1)}),
\end{itemize}

The second cell is \texttt{cellfirModule3}. Besides the control signals (clock, clock-enable and
reset), it has four input signals: 
\begin{itemize}
\item the w coefficient (\texttt{pipeCw7Xctl1PReg2Xloc}, binded to \texttt{pipeCw7Xctl1PReg2Xloc(p)}),
\item the X input (\texttt{pipeCx5Reg3Xloc}), binded to \texttt{pipeCx5Reg3Xloc(1)}),
\item the Y input (\texttt{YReg5Xloc}) binded to \texttt{YReg5Xloc(p)}),
\item and
the pipe signal (\texttt{pipeIOCw9Reg4Xloc} binded to \texttt{pipeIOCw9Reg4Xloc(p)}).
\end{itemize}
It outputs similar signals:
\begin{itemize}
\item the w coefficient (\texttt{pipeCw7Xctl1P}, binded to \texttt{pipeCw7Xctl1P1(1)}),
\item the X output (\texttt{pipeCx5}), binded to \texttt{pipeCx53(p)}),
\item and
the pipe output signal (\texttt{pipeIOCw9} binded to \texttt{pipeIOCw93(p)}), and
\item the Y output (\texttt{Y}) binded to \texttt{Y3(p)}).
\end{itemize}

\subsection{The Controller}
At reset time (reset is active at value 0), the controller initializes a counter to value $-1$, and enters state
\texttt{initState}. It remains in this state until the counter reaches the value
34. During the first 35 cycles, the output of the pipe signal is 0. It gets the 
value 1 during cycle 36 (when the counter reaches the value 34).

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/PeriodicSystems/component-descrition.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}

\usepackage{makeidx,fancyvrb,graphicx}

\begin{document}

\newcommand{\vhdl}{\textsc{vhdl}}
\title{Description of the \vhdl{} Library}
\author{}
\date{}
\maketitle

\newcommand{\rom}{{\texttt{ROM}}}
\section{The \rom{} Component}

\subsection{Component Description}
The \rom{} component is a combinatorial read-only
memory. It returns a data, as a signed number with
a specified length, specified by an integer address.

\subsection{Generation Process}
Use \texttt{genVhdl[ "ROM" ]} function in package \texttt{Vhdl2.m}.

\subsection{Files}
\begin{description}
\item[\texttt{ROM.vhdx}:] the architecture description of the ROM (see \ref{romarchi}).
\item[\texttt{ComponentOfROM.vhdx}:] the component description of the ROM (see \ref{romcomp}).
\end{description}

\newcommand{\sizeparam}{\texttt{\$size\$}}
\newcommand{\commentparam}{\texttt{\$comment\$}}
\newcommand{\wlmunparam}{\texttt{\$wordLengthMun\$}}
\newcommand{\nameparam}{\texttt{\$name\$}}
\newcommand{\valuesparam}{\texttt{\$values\$}}

\subsection{Parameters}
\begin{description}
\item[\sizeparam{}: ] the size of the read-only memory, in words.
\item[\commentparam{}: ] a comment.
\item[\valuesparam{}: ] the values associated to the ROM. This parameter is not
mandatory, and if not supplied, 0 values are assumed.
\item[\nameparam{}: ] the name assigned to the component.
\item[\wlmunparam{}: ] the word length (minus one) of the values in the ROM. 
\end{description}

\subsection{Example}

\begin{verbatim}
--
-- Component for a combinatoric ROM
-- ROM of size 32 and of word length 2
--

COMPONENT ROM3 IS
  PORT 
  (
    address : IN INTEGER RANGE 0 TO 31;
    data : OUT SIGNED (1 DOWNTO 0)
  );
END COMPONENT;
\end{verbatim}

\newcommand{\moduloaddress}{\texttt{moduloAddress}}
\section{The \moduloaddress{} Component}

\subsection{Component Description}
The \moduloaddress{} component is a component that generates...

\subsection{Generation Process}
Use \texttt{genVhdl[ "moduloaddress" ]} function in package \texttt{Vhdl2.m}.

\subsection{Files}
\begin{description}
\item[\texttt{ModuloAddress.vhdx}:] the architecture description of the modulo address generator (see \ref{romarchi}).
\item[\texttt{ComponentOfModuloAddress.vhdx}:] the component description of the modulo address generator (see \ref{romcomp}).
\end{description}

\newcommand{\periodparam}{\texttt{\$period\$}}
\newcommand{\periodmunparam}{\texttt{\$periodMun\$}}

\subsection{Parameters}
\begin{description}
\item[\periodparam{}: ] the period of the modulo generator.
\item[\commentparam{}: ] a comment.
\item[\periodmunparam{}: ] perion minus one. This is generated automatically by the
VHDL generator.
\item[\nameparam{}: ] the name assigned to the component.
\end{description}

\newpage
\appendix

\section{The ROM}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/ROM.vhdx}

\subsection{Component}
\label{romcomp}
\VerbatimInput{../../VHDL/ComponentOfROM.vhdx}

\newpage
\section{The Modulo Address Generator}
\subsection{Architecture}
\label{modumloaddressarchi}
\VerbatimInput{../../VHDL/ModuloAddress.vhdx}

\subsection{Component}
\label{modulocompcomp}
\VerbatimInput{../../VHDL/ComponentOfModuloAddress.vhdx}

\newpage
\section{The Registers}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/Registers.vhdx}

\subsection{Component}
\label{romcomp}
\VerbatimInput{../../VHDL/ComponentOfRegisters.vhdx}

\newpage
\section{The FSM component}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/Fsm.vhdx}

\subsection{Component}
\label{romcomp}
\VerbatimInput{../../VHDL/ComponentOfFsm.vhdx}

\subsection{Call}
\label{romarchi}
\VerbatimInput{../../VHDL/CallFsm.vhdx}

\newpage
\section{The Call reg component}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/CallReg.vhdx}

\section{The Call registers component}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/CallRegisters.vhdx}

\newpage
\section{The PeriodicEnable component}
\subsection{Architecture}
\label{romarchi}
\VerbatimInput{../../VHDL/PeriodicEnable.vhdx}

\subsection{Component}
\label{romcomp}
\VerbatimInput{../../VHDL/ComponentOfPeriodicEnable.vhdx}

\subsection{Call}
\label{romarchi}
\VerbatimInput{../../VHDL/CallPeriodicEnable.vhdx}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Pipeline/pipelineDoc.tex
Added mmalpha (source and binary)

\documentclass[a4paper,11pt]{article}
\usepackage{psfig}
\usepackage{epic}
\usepackage{eepic}

\title{Theoretical basis of the Pipeline and PipeControl package}
\author{Tanguy Risset \and Patrice Quinton\footnote{Last modification: 22 nov. 2009}}
\date{Version of \today}
\begin{document}

\maketitle
\section*{Introduction}
This document is intended to provide the precise semantics of 
the functions of the  {\tt Pipeline.m} package, together with 
their implementation details and examples of use.
These functions are: {\tt  pipeline, pipeall, pipeInput, pipeOutput,
pipeIO}.

\section{ToAlpha0v2}
\subsection{Space time separation}
See \texttt{ToAlpha0v2.m}.

Call to \texttt{spaceTimeDecomposition}, with time position and space position.
An option (exceptions) allows some variables to be ignored.

Compute the list of variables to consider : all variables, except output of 
use expressions. Then, call function \texttt{spaceTimeCase} for each
variable.

This function check that the variable is defined, and that the definition is
a case expression. If it is not the case, it returns the system unchanged.

Otherwise, it calls \texttt{makeSTCase}. This function will return a new
expression that will replace the old one.

Two cases to be considered: one or more than one branches.

If the expression has only one branch. We compute the domain
of the RHS expression (using \texttt{expDomain}). We compute the
intersection of this domain and of the LHS domain (actually, the
declaration domain of the variable), and we
signal if the expression domain is not included in the LHS domain. 
Then, we return this intersection domain.

If there are more than one branch, we proceed recursively. Let $d_0$
the domain of the first branch. Assume that the $n-1$ remaining
branches are in ST form, and let $d_i: exp_i$ be the
other branches.

\begin{itemize}
\Item compute the intersection $d_p$ of the space projection
of $d_i$ and the space projection of $d_i$\,;
\item if $d_p$ is empty, we add $d_i : exp_i$ to the list
of restrictions, and we keep $d_0$ unchanged for the next
round.
\item if not, we compute $d = d_0 \cap d_i$\,; and $dinter = d_i \ d$
\end{itemize}

\section{PipeControl}
The \texttt{pipeControl} function aims to optimize a 
control definition equation. Such an equation has the form
\begin{verbatim}
Vctrl[ t, p1, ... ] = 
case
  domProcs: 
	case
 	 domTime1: exp1;
 	 domTime2; exp2;
	esac;
esac;
\end{verbatim}
(It is called by the \texttt{pipeAllControl} function, which finds out all
variables that are supposed to be control equations. So, if called directly, 
the property is not necessarily checked, and the equation may not fit
this format.)

The idea is as follows. First, we try to find out a separating hyperplane
for both expressions. The idea would be to find out either a
processor domain, so that the value of the control signal is 
related to the processor position, or a time plane, so that 
the value of the control could be defined by a controller and
broadcasted to all processors.

Once such an hyperplane is found, it gives a direction
along which to propagate the \texttt{True} and \texttt{False}
values to the boundaries of the domain.

We first compute the direction vector of the hyperplane, and
then, we compute its null space vectors. We select these vectors
in such a way that their first component (related to time) is not
zero, otherwise, that would mean that we cannot pipeline, but only
delocalize the control.

\subsection{Checking}
The \mma{} function checks the following conditions:
\begin{enumerate}
\item The variable to pipeline exists (message \texttt{pipeControl::unknownvar}).
\item The dimension of the variable is less than or equal to 4.
\item The first dimension is the time.
\item 
\end{enumerate}

\end{verbatim}
\section{Pipeline}
\subsection{Explanation of the transformation}
The pipeline transformation is used to transform a program with 
a definition of a variable $var_1$ which contains an
expression $expr$ implying a non uniform dependence $f$ ($Ker(f) \neq
\emptyset$): $var_1[z] = F(expr[f(z)])$,  into an 
equivalent\footnote{equivalence to be proved} program which contains a uniform
dependence $d$\footnote{or several uniform dependencies} for the
definition of the same variable $var_1$.

The principle of the transformation is the following, 
the user gives : 
\begin{itemize}
\item the name of a new variable to be created 
 (pipeline variable: $pipeExpr$);
\item  the exact instance of 
the expression to be pipeline 
($expr[f(z)]$ in  $var_1$);
\item  and a pipeline vector
$d$. 
\end{itemize}
In the transformed program, 
the variable  $pipeExpr$ is defined on the whole domain $D$ where 
the expression $expr[f(z)]$ is used.
 $pipeExpr[x]$ is initialized to  $expr[f(x)]$ on the 
{\em border\footnote{at the points where we cannot retrieve one more pipeline vector without getting out of $D$} of $D$
 along $d$} (we call this border $B$).
 This value is  duplicated (we say {\em pipelined})
along $d$ on the rest of the domain $D$
($pipeExpr[z] = pipeExpr[z-d]$). Then, the corresponding part 
of the definition of $var_1$ is replaced by: 
$var_1[z] = pipeExpr[z]$ (see figure~\ref{fig0}).

This transformation cannot be done for every pipeline vector.
Note that  the transformation imply that the values
$pipeExpr[z+kd]$ $ \forall z \in B,\ k \in N$ will correspond to the
instance $expr[f(z)]$, thus we must have $expr[f(z)]=expr[f(z+kz)] \
\forall z \in B, \ k \in N$. This is verified if $d \in Ker(f)$. This 
imply that $f$ must be non invertible. Note that the pipeline vector 
may be a not primitive vector, but this may induce warning during 
analysis of the resulting program because it correspond to 
the assumption that the domain of use of the expression is not flat.

\subsection{Implementation}

\begin{figure}
\input{ex1.eepic}
\caption{Example of pipeline Transformation}
\label{fig0}
\end{figure}

The {\tt pipeline} function should be used by programmers only, 
the users should use {\tt pipeall}. In the parameter of 
the function, the name of the variable and the pipeline vector are catched in the 
same expression :{\tt "Name.function"} where the function is the translation
in the direction of the pipeline vector.

% The three forms possible for the pipeline function arguments are: 
%\begin{verbatim}
%pipeline[occur:List[_Integer..],pipeSpec_String]\n 
%pipeline[sys_Alpha`system,occur:List[_Integer..],pipeSpec_String]\n
%pipeline[sys:Alpha`system,occur:List[_Integer..],newVarName_String,
%  pipelineVector_Alpha`matrix]
%\end{verbatim}

At the moment, the domain of the new variable build is the
intersection of the context domain of the expression to pipeline
({\tt getContextDomain}) with the domain of the expression to
pipeline ({\tt expDomain}). If the pipeline vector is not in the
kernel of the dependence function, the transformation is
aborted.

\subsection{pipeall}

{\tt pipeall} is like a pipeline more convivial. If an expression is
present in several places in the definition of a variable, or even in several
definition of different variables, {\tt
pipeall} will pipeline all occurrences of this expression.  In that
case, the domain on which is defined the variable is the union of all
the domains which would result of each individual pipelines.
The main difference with pipeline are the way the arguments are asked 
(see {\tt ?pipeall}). Be careful, you cannot expressed the expression 
to be pipelined using the array notation, you must specify 
{\tt expr.(x->f(x))} instead of {\tt expr[f(x)]}.

\section{Pipeline of Inputs and Outputs}
the {\tt pipeline} transformation pipelines a dependance which was
  originally a broadcast (dependence non invertible).  We may need to
  pipeline an expression even if it is not broadcasted, this
  is particularly useful when the expression to pipeline must be
  input (resp. output) in an architecture, in which they must go
  through some cells before being used in a computation (array with a
  loading phase).  This can be done with the {\tt pipeIO} function, which 
  should be used through the functions {\tt pipeInput} and {\tt pipeOutput}.
  {\tt pipeIO} perform a {\em routing}. It takes some data at some 
  place of the iteration space and bring it into another place of the 
  iteration space.

\subsection{PipeInput}
\begin{figure}[ht]
\input{ex3.eepic}
\caption{Example of input pipe with {\tt pipeInput}}
\label{fig01}
\end{figure}

The user gives:
\begin{itemize}
\item the name of a new variable to be created 
 (pipeline variable: $pipeExpr$);
\item  the exact instance of 
the expression to be pipeline 
($expr[f(z)]$ in  $var_1[f(z)]$, note that $f$ may be non-singular);
\item  a pipeline vector
$d$;
\item and the half space  in which 
the pipeline is to be performed (bounded by an hyperplane $H$).
\end{itemize}

Pipelining an expression as
an input consists in the following transformation: we suppose the expression 
$expr[f(x)]$ is used 
an expression somewhere $x$  ($Var[x]=expr[f(x)]$ on domain $A$, 
see figure~\ref{fig01}).  After the transformation, 
the expression $expr$ is used somewhere else $y$ (along the bounding 
hyperplan $H$) and the value is propagated to location $x$ by the
pipeline variable ($Var[x]=pipeExpr[x]$ on the original domain $A$,
$pipeExpr[z]=pipeExpr[z-d]$ inside the pipeline Domain $D$,
 and $pipeExpr[y]=epxr[f[x]]$ on the border $B$).
This transformation is very close to the usual pipeline transformation

This transformation is illustrated on  figure~\ref{fig1}
and~\ref{fig2}. figure~\ref{fig1} represents the original program
and figure~\ref{fig2} represents the program after the execution of 
the command:  {\tt pipeInput["C", "b.(i,j->i)","B1.(i,j->i+1,j+1)",
" \{i,j | i >= 0 \} "]} and normalization.
 The example also corresponds to the illustration 
of figure~\ref{fig01}. $Var1$ is {\tt C}, $expr[f(x)]$ is {\tt b.(i,j->i)},
$d$ is {\tt (1,1)} (represented by {\tt (i,j->i+1,j+1)}),  $pipeVar$
is {\tt B1}, and $H$ is  {\tt \{i,j | i >= 0\}}.

\begin{figure}[ht]
\begin{verbatim}
system silly: {N | N>1}
                (a : {i,j|1 <= i,j <= N} of boolean; 
                 b : {i|1 <= i <= N} of boolean)
     returns (c : {i|1 <= i <= N} of boolean);
  var 
        C :  {i,j|1 <= i <= N; 0<= j <=N} of boolean;
  let
C[i,j] = case
       {|j=0} : b[i];
       {| j>=1} : C[i,j-1] + a[i,j];
         esac;
      c[i]=C[i,N];
  tel;   
\end{verbatim}
\caption{simple program before the use of pipeInput}
\label{fig1}
\end{figure}

\begin{figure}[ht]
\begin{verbatim}
......
var
  B1 : {i,j | (j+1,0)<=i<=j+N; j<=0; 2<=N} of integer;
  C : {i,j | 1<=i<=N; 0<=j<=N} of boolean;
let
  B1[i,j] = 
      case
        {| i=0; -N<=j<=-1; 2<=N} : b[i-j];
        {| 1<=i<=j+N; j<=0; 2<=N} : B1[i-1,j-1];
      esac;
  C[i,j] = 
      case
        {| 1<=i<=N; j=0; 2<=N} : B1;
        {| 1<=j} : C[i,j-1] + a[i,j];
      esac;
.......
\end{verbatim}
\caption{Program of figure~\ref{fig1}, after use of {\tt pipeInput}:
{\tt pipeInput["C", "b.(i,j->i)","B1.(i,j->i+1,j+1)",
" \{i,j | i >= 0 \} "]}}
\label{fig2}
\end{figure}

\subsection{PipeOutput}
\begin{figure}
\input{ex2.eepic}
\caption{Example of output pipe with {\tt pipeOutput}}
\label{fig02}
\end{figure}

The user gives (as for {\tt pipeInput}:
\begin{itemize}
\item the name of a new variable to be created 
 (pipeline variable: $pipeExpr$);
\item  the exact instance of 
the expression to be pipeline 
($expr[f(z)]$ in  $var_1[f(z)]$, note that $f$ may be non-singular);
\item  a pipeline vector
$d$;
\item and the half space  in which 
the pipeline is to be performed (bounded by an hyperplane $H$).
\end{itemize}

 Pipelining an expression as
an output consists in the following transformation:
 we suppose we use at some place $x$ 
an expression which  was produce at some place $y$
 ($Var[x]=expr[y]$ on domain $A$). In the transformed 
program, we use this expression in another variable  
at  place $y$ ($VarPipe[y] = expr[y]$) and the value 
in pipelined in $VarPipe$ until another place $f(x)$ where it is consumed 
by $Var$ ($Var[x] = VarPipe[f(x)]$ on domain $B$, see figure~\ref{fig02} for 
example of output pipe).

This transformation is illustrated on  figure~\ref{fig3}.
 figure~\ref{fig1} represents the original program
and figure~\ref{fig2} represents the program after the execution of 
the command:  {\tt pipeOutput["c", "C","C1.(i,j->i+1,j+1)",
"\{i,j| i <= N\}"]}. The transformation performed is illustrated on 
figure~\ref{fig02} where $Var$ is {\tt c}, $expr[f(x)]$ is {\tt C.(i,j->i,j)},
$d$ is {\tt (1,1)} (represented by {\tt (i,j->i+1,j+1)}),  $varPipe$
is {\tt C1} and $H$ is {\tt. \{i,j| i <= N\}}.

\begin{figure}
\begin{verbatim}
var

  C1 : {i,j | j-N+1<=i<=N; N<=j; 2<=N} of integer;
.........
  C1[i,j] = 
       case
         {| 1<=i<=N; j=N} : C;
          {| j-N+1<=i<=N; N+1<=j} :  C1[i-1,j-1];
       esac;
  c[i] = C1[N,-i+2N];
tel;
\end{verbatim}
\caption{program of figure~\ref{fig1}, after use of pipeOuput:
 {\tt pipeIO["c", "C","C1.(i,j->i+1,j+1)","\{i,j| i <= N\}"]}}
\label{fig3}
\end{figure}

\subsection{Implementation}
Both {\tt pipeOutput} and  {\tt pipeInput} are implemented
by the same function : {\tt pipeIO}, we will briefly described the 
implementation of this function here.
 The fact that the pipeline is an input pipe or an output pipe is
 determined by the scalar dot of the pipeline vector and 
the normal to the hyperplane $H$ bounding the half space.
If the pipeline vector goes {\em towards} $H$ then, this must be an output 
pipe, if it {\em comes from} $H$, this must be an input Pipe

Three domains are distinguished:
\begin{itemize}
 \item the domain of the original 
expression $expr[f(z)]$ (that we will call $domExpr$), 
which is the domain where
the pipeline is initialized in the case of an output pipe and the 
domain where the pipeline ends in the case of an input pipe 
(domain $A$ on figure~\ref{fig01} and~\ref{fig02}). 
\item the pipeline domain (that we will call $realDomPipe$) 
which is the domain  on which the value is pipelined 
(domain $D$ on figure~\ref{fig01} and~\ref{fig02}). 
\item the domains where the pipeline ends (that we will call $pipeEndDom$), 
which is in fact the domain where the pipeline is initialized in
 the case of an input pipe
(domain $B$ on figure~\ref{fig01} and~\ref{fig02}). 
\end{itemize}
$domExpr$ is computed as the intersection of the context domain of the
expression to pipeline ({\tt getContextDomain}) and the domain of the
expression itself ({\tt expDomain}). $realDomPipe$ is computed by
adding to $domExpr$ a ray (which is the pipeline vector in the case of
an output pipe and its opposite in the case of an input pipe) and then
by intersection the resulting domain with the half space bounded by 
 $H$. $pipeEndDom$ is computed by shifting $realPipeDomain$ by
the pipeline vector in case of input pipe (resp. its opposite in case
of output pipe) and retrieving the $realDomPipe$.

there remains the problem of founding, given on index point in
$pipeEndDom$, what is the corresponding point in $domExpr$ (resp. the
other way around in the case of input pipe). This is done by
observing the following fact. If we find a function of the indices
whose value is constant on the pipeline path and which is perfectly
determined by the pipeline path (unique for each pipeline path),
then giving this value will determine a unique antecedent point
$z_1$ in $exprDom$ and a unique antecedent point $z_2$ in
$pipeEndDom$. Hence, we will be able to use the {\tt
inverseInContext} function is order to find for a point in
$pipeEndDom$ the corresponding point in $exprDom$. the function to
build must have a square matrix thus we take a square $n\times n$
matrix of rank $n-1$ for which the kernel is generated by the
pipeline vector. And it works...  The use of {\tt inverseInContext}
impose that the domains of the expressions to pipeline must be flat
(dimension $n-1$).

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/PolyLib/note.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\begin{document}

\newcommand{\polylib}{{\sc PolyLib}}
\title{A note regarding the installation of \polylib{}}
\author{Patrice Quinton}
\date{October 23, 2007}
\maketitle
This short note contains a few remarks regarding the installation of \polylib{}.

%\section{Installing the 5.02 version}
%The version I run was 5.02, which I inherited from Tanguy Risset. The source should be
%on 
%\begin{verbatim}
%http://icps.u-strasbg.fr/~loechner/polylib
%\end{verbatim}
%but I did not check whether this version is the most current one.

%After untaring the tar file, I got a directory where the \texttt{doc/INSTALL} file
%contains some instructions, but this does not work.

%To obtain an executable, I did what Tanguy said:
%\begin{verbatim}
%./autogen.sh
%./configure --prefix="path"
%make
%make install
%\end{verbatim}

%The first command runs the \texttt{autogen.sh} file. However, for an unknown reason, 
%on the MacIntosh, the \texttt{libtoolize} tool is named \texttt{glibtoolize}... So the second
%line of the \texttt{autogen.sh} file has to be corrected.

\section{Installing the version of Strasburg}
I got the most recent available version of \polylib{} on site 
\begin{verbatim}
http://icps.u-strasbg.fr/~loechner/polylib
\end{verbatim}
and I tried to install it. In this version, the installation documentation is correct, as
\texttt{autoload.sh} does not need to be run prior to \texttt{configure}. By running
only 
\begin{verbatim}
./configure --prefix="$MMALPHA/sources/Poly"
make
make install
\end{verbatim}
The second command runs the configure tool that creates \texttt{makefiles} and prepares
libraries. The \texttt{--prefix} option specifies where the \polylib{} is to be created.

The third commands compiles everything, and the \texttt{make install} command 
creates the library. It adds binary files in the directory \texttt{path/bin}, libraries in directory \texttt{path/lib}, and 
finally, include files in directory \texttt{path/include}.

I got a runnable version, that I checked using \texttt{make longtest}. The 
\texttt{make test} command indicated in the install documentation does not work.

In this package, the html documentation is available, but the latex source files 
are not.

The documentation is also available.
\section{Tests}
Running \texttt{make check} runs the tests which are in directory \texttt{Tests}.

\section{Additional remarks}
Depending on whether you are on PPC or Intel processor, you have to recompile 
the Polylib. But this is not sufficient to generate a correct Domlib: you should recompile
Domlib, and make sure that {\em Read\_Alpha and other software are also recompiled}, 
as currently, this is how the lib.darwin directory is updated. If you do not do so, then
compiling Domlib may fail.

\section{Exploring web sites}
There are now several libraries for doing polyhedral computations. A survey of what these
libraries do (or cannot do) would be useful.

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/QuickStart/AlphaStart.tex
Added mmalpha (source and binary)

--- 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Alphard/Alphard.tex
@@
+%postcript file in $MMALPHA/doc/tutorial/QuickStart.ps
 \documentclass[12pt]{article}
@@
 \newcommand{\sitemmalfa}
-{\texttt{{http://www.irisa.fr/cosi/ALPHA}{http://www.irisa.fr/api/ALPHA}}}
+{\texttt{{http://www.irisa.fr/cosi/ALPHA}}}
 \newcommand{\emailmmalfa}{\texttt{alpha@irisa.fr}}
@@
 \title{Getting started with \Alpha{}}
-\author{Api, then Cosi, then R2D2 and Compsys\thanks{
-Api, Cosi and R2D2 are the names of the 
+\author{Api, then Cosi, then R2D2 and Compsys, then Cairn\thanks{
+Api, Cosi, R2D2 and Cairn are the names of the 
 research groups that successively hosted
@@
 \mmalfa{} is a free software, available under the 
-\gpl{}. It can be downloaded from 
-the site:
-
-\sitemmalfa{}.
-
+\gpl{}. 
+What is said here applies to Version 2.0 or \mmalfa{}, 
+which currently you may get by sending a mail to 
+\begin{verbatim}
+patrice.quinton@irisa.fr
+\end{verbatim}
 To be run, it requires
@@
 
-\mmalfa{} is still under development at \irisa{} and at ENS Lyon (France).
+\mmalfa{} is still under development at \irisa{} and at INSA in Lyon (France).
 
@@
 \item Providing a correct recurrence equation specification for a particular 
-algorithm (see section~\ref{language} and \ref{basic}).
-\item Simulating such a specification (see section~\ref{simulation}).\index{Simulation}
+algorithm (see Sections~\ref{language} and \ref{basic}).
+\item Simulating such a specification (see Section~\ref{simulation}).\index{Simulation}
 \item Transforming and simplifying a recurrence equation specification (see 
-sections~\ref{basic} and~\ref{advanced}).
+Sections~\ref{basic} and~\ref{advanced}).
 \item Computing on convex polyhedra 
@@
 \item Scheduling programs and detecting parallelism 
-(see section~\ref{schedule}).
+(see Section~\ref{schedule}).
 \end{itemize}
@@
 The remaining of this document is organized as follows. Section~\ref{howwork}
-presents the \mmalfa{} interface; section~\ref{install} explains
-the installation procedure; in section~\ref{language}, the \alfa{}
-language is briefly presented, while in section~\ref{basic}, 
+presents the \mmalfa{} interface; Section~\ref{install} explains
+the installation procedure; in Section~\ref{language}, the \alfa{}
+language is briefly presented, while in Section~\ref{basic}, 
 basic operations on \alfa{} programs are described; hardware
-generation is presented in section~\ref{hardware}; in section~\ref{advanced},
+generation is presented in Section~\ref{hardware}; in Section~\ref{advanced},
 other transformations of \alfa{} programs are explained, and 
-in section~\ref{structured}, we explain how structured \alfa{} programs
-can be structured and transformed. Finally, section~\ref{andnow}
+in Section~\ref{structured}, we explain how structured \alfa{} programs
+can be structured and transformed. Finally, Section~\ref{andnow}
 concludes and gives some additional references for further 
@@
 polyhedral library which constitutes the computationnal kernel of the
-{\mmalfa{}} environment.
+{\mmalfa{}} environment. \mma{} also uses the PiP software designed by Paul Feautrier
+(See \texttt{http://www.piplib.org/}).
 
@@
 should install the {\mmalfa{}} environment. 
-The installation procedure is explained in
+The installation procedure is explained 
+in Appendix~\ref{install} of this document.
+\MMAlfa{} can be downloaded from 
 the \MMAlfa{} web site (\sitemmalfa{}).
@@
 To do so, type \texttt{mathematica} under
-\texttt{Unix}, or start \mma{}~3.0 in the Programs menu of 
+\texttt{Unix}, or start \mma{} in the Programs menu of 
 Windows NT\footnote{Any \mma{} version later than 3.0 will work.}.
-\footnote{One can also access 
-the \mma{} kernel directly. Type \texttt{math} under
-\texttt{Unix}, or start \mma{}~3.0~Kernel in the Programs menu of 
-Windows NT. An alternative is to use the \mma{} 
-kernel via \texttt{emacs}, but then the demonstration notebooks are
-not accessible.}
+We assume in the following that you
+use \mma{} through its Notebook interface.
 
@@
 of the \mmalfa{} distribution.
-Then open the cell named "Introduction Notebooks"\footnote{For 
+
+Then open the synthesis notebook by clicking on the Synthesis
+button (in the Section called {\em New: synthesis notebooks}\footnote{For 
 the reader who is not yet familiar with \mma{}'s 
@@
 its subcells) is done by clicking once its bracket. 
-To open a cell, double-click the bracket once the cell is selected.}, 
+To open a cell, double-click the bracket once the cell is selected.}). In this 
+notebook, you will find several demonstrations examples: a FIR filter, 
+a distance calculation program, a matrix-vector multiplication, 
+a matrix-matrix multiplication, and a string alignment program called
+Samba. Currently, the documentation on the \texttt{syn} command is 
+available in 
+\begin{verbatim}
+$MMALPHA/doc/Synthesis/Synthesis.pdf
+\end{verbatim}
+
+Another possibility is to open the cell named "Introduction Notebooks", 
 \index{Introduction notebooks}
@@
 \subsection*{Systems, variables, domains and parameters}
-An {\Alpha} program is a {\em system}. \alfa{} variables are generalized
+An {\Alpha} program is a {\em system}. \alfa{} variables\footnote{In \Alpha{}, 
+identifiers are case sensitive. Undescores are accepted, as in 
+\texttt{A\_1} but {\em should be avoided}, as \mma{} identifiers
+cannot contain undescores and in some situations, this may result in 
+unexpected errors. Moreover, avoid using variables that are distinguished
+only by there case, as in \texttt{x} and \texttt{X}; since eventually,
+your program will be translated in \vhdl{} that is not case sensitive.} are generalized
 arrays which can have any shape (not just rectangles).  The set of
@@
 
+A frequent mistake: if you specify in the additional constraints a wrong coefficient, 
+the scheduler fails badly with a quite non specific message. For example, say you add
+a constraint such as:
+\begin{verbatim}
+TAD4 == 1
+\end{verbatim}
+although variable \texttt{A} has only 2 dimensions, then the scheduler fails, since it
+does not check the correctness of this constraint and therefore, calls \mma{}'s 
+optimizer with wrong arguments.
+
 \paragraph*{More information}
@@
 \paragraph*{Notes}
-The allocation is chosen automatically by \mmalpha{}. 
-An option allows one to fix the allocation direction. See
-the \texttt{projMatrix} option.
+When \texttt{appSched} is called in its simplest form, the allocation function
+is chosen automatically by \mmalpha{}. Sometimes, this allocation is not convenient.
+For example, it may happen that the processor dimension is mapped to 
+an infinite dimension of the calculation domain.
+
+Some options of \texttt{appSched} allow one to set the allocation direction. 
+The \texttt{projMatrix} option allows one to define the projection 
+matrix, that is to say, the linear function which defines the indices where
+the processor are mapped. For example:
+\begin{verbatim}
+appSched[ projMatrix -> {{1,0,0}, {0,0,1}} ]
+\end{verbatim}
+says that the processor indexes will be the first and the third one
+(assuming that there is no parameters.)
+
+The \texttt{projVector} option is another way of specifying the 
+allocation function:
+\begin{verbatim}
+appSched[ projVector -> {{0,1,0}} ]
+\end{verbatim}
+also projects on indexes 1 and 3.
+
+Both \texttt{projVector} and \texttt{projMatrix} 
+can be used only when all variables have the same dimension
+and are in the same {\em space}.
+
+If necessary, one can apply \texttt{appSched} variable by variable, 
+by using the \texttt{variables} option. For example:
+\begin{verbatim}
+appSched[ variables -> {"A", "b"}, projMatrix -> {{1,0,0}, {0,0,1}} ]
+\end{verbatim}
+applies the projection only on variables \texttt{A} and \texttt{b}.
+
+Another way of specifying the allocation function is to define
+what index has to be replaced by the schedule. For example:
+\begin{verbatim}
+appSched[ timeDimensions -> {2} ]
+\end{verbatim}
+says that the second dimension of the variables can be replaced by
+the schedule.
+
+This way of specifying the projection can be combined with the \texttt{variables} option:
+\begin{verbatim}
+appSched[ variables -> { {"A",{2}}, {"b",{1}}, {other,{3}} ]
+\end{verbatim}
+specifies that for variable \texttt{A}, the second dimension has to be replaced
+by the schedulig, for variable \texttt{b}, the first dimension, and for the other
+variables, the third dimension.
+
+If none of these possibilities works, the ultimate solution is to do the change of basis by hand. 
+Help may be obtained by running 
+\begin{verbatim}
+appSched[ verbose->True ]
+\end{verbatim}
+as this command will list the changes of bases that are applied.
+
+\paragraph*{Troubleshooting}
+Remember that no change of basis can be done on input or output variables\,! 
+Indeed, the change of basis preserves the semantics of an \alfa{} program, and 
+this is not compatible with a reindexing of inputs or outputs. If you really want to 
+reindex inputs or outputs, just add local variables where you copy the inputs or
+store the outputs.
 
@@
 
+\section{Limitations of \mma{}}
+This section contains a few hints to overcome limitations of \mma{} or
+to cope with known difficulties. Some of these hints are already mentioned in the
+document.
+
+\subsection{The \vhdl{} translater}
+Some situations may lead to unexpected errors when using the \texttt{a2v} 
+command. Some of them are easy to avoid.
+
+\subsubsection*{Parameters are forbidden!}
+First, remember that the main program should not contain parameters anymore. 
+The value of the parameters must have been set, before calling \texttt{a2v}, by the \texttt{fixParameter} 
+command. (Generating \vhdl{} for parameterized programs is very difficult, except for very simple
+cases.)
+
+\subsubsection*{Remove main program\footnote{This is not a joke \ldots}!}
+Second, notice that after the \texttt{alpha0ToAlphard} command, the \alfa{} program
+is separated into a controller, cells, and a module. All these elements are themselves
+called by a main program -- whose name is the name of the initial system -- that cannot always
+be translated into \vhdl{}. When calling \texttt{a2v}, the \vhdl{} translator may then fail on 
+this main program. This error has no consequence on the generation of \vhdl{} for the other
+subsystems. If you want to avoid it, just remove the main program from the library
+using the \texttt{removeSystem} command before calling \texttt{a2v}.
+
+\subsubsection*{Put a bound to your calculations!}
+\alfa{} allows unbounded programs to be written, filters for example. This happens for example, 
+by using variables with domains of the form
+\begin{verbatim}
+x : { k | 0 <= k } of integer;
+\end{verbatim}
+Currently, the \texttt{a2v} translator fails when generating a controller for such programs. 
+To overcome this, write instead
+\begin{verbatim}
+x : { k | 0 <= k <= B } of integer;
+\end{verbatim}
+where \texttt{B} is a parameter that will be set before calling \texttt{a2v}.
+
+\input{../Install/How-to-install}
+
 \printindex

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Alpha.tex
Added mmalpha (source and binary)

%
% File created on {2009, 4, 25, 18, 19, 40.078499} by makeDoc.
%
% Section header
\section{The Alpha` package } 
\label{label:Alpha`}
\input{incl-Alpha.tex}
\alphanote{Alpha}{Documentation revised on April 25, 2008}

\alphausage{Alpha}{\MMAlpha{} is a Mathematica-based implementation of the
ALPHA language.  The "Alpha`" package contains basic commands for
the \MMAlpha{} system.  The name "Alpha`" is also a root name for all
packages and symbols of \MMAlpha{}.}{Alpha.m}{Alpha`}
\index{Alpha}

\alphanote{Alpha}{In this package, there is a description of some structures (domain,
pol, zpol), of global variables (\$coneList, \$demoDirectory, \$depCone, 
\$library, \$masterNotebook, \$MMALPHA, \$myMasterNotebook, \$myNotebooks, 
\$program, \$result, \$rootDirectory, \$schedule, \$scheduleLibrary, 
\$testDirectory, \$tmpDirectory, \$showGraph, 
\$testResult, \$testDirectory, \$testIdentifier, \$testReportFile,
\$tmpDirectory, \$runTime,
\$vhdlPatternsDir), 
and finally of some functions. Among these, 
ashow, ashowLib, asave, asaveLib, getSystem, load, myStart, putSystem,
save, saveLib, show, showLib, showMat, start, and systemNames
are important for a user. Functions getPart, readAlpha, readDom, 
readExp, readDom, testFunction, tests, varTypes, writeC, astQ, 
fileName, openTemporary, getTemporary, packageLink, demoLink, docLink, 
link, exportAlphaFunctions, and setMMADir are useful for developers. 
Function writeTex, enTete, and wrap are unused or incomplete.}

\alphausage{domain}{domain[dim, idx, lpol] is the structure representing 
a domain. dim is the (integer) dimension of the domain, 
idx is the list of dimension names and lpol is a list of
polyhedra or of Z-polyhedra. See also ?pol or ?zpol.}{Alpha.m}{Alpha`}
\index{domain}

\alphanote{domain}{Domains of Alpha are either polyhedra, unions of polyhedra, Z-polyhedra, 
i.e., intersection of a polyhedron and of a lattice, or unions of 
Z-polyhedra. Note also that in domains of Z-Polyhedra, the 
index part is empty.}

\alphausage{pol}{pol[nbConstraints, nbRays, nbEq, nblines, cons, rays] is 
the representation of a polyhedron in Alpha.
nbConstraints (integer) is the number of
constraints (equalities and inequalities), 
nbRays (integer) is the number of rays
(including lines), nbEq (integer) is the number of equalities,
nblines (integer) is the number of lines, cons (matrix) is the
set of equalities and inequalities, and rays (matrix) is the 
set of rays and lines. Inequalities and rays are represented
by lists of integers with a leading 1, and equalities or lines
by lists of integers with a leading 0. See also ?domain and ?zpol.}{Alpha.m}{Alpha`}
\index{pol}

\alphausage{zpol}{zpol[m, lpol] is the representation of a Z-polyhedron in Alpha.
m is an Alpha matrix which represents the Z-lattice, and lpol
is a list of polyhedra. See also ?domain and ?pol.}{Alpha.m}{Alpha`}
\index{zpol}

\alphausage{\$coneList}{\$coneList contains the list of dependence cones of all 
the dependences of \$result. This variable is set by the 
initUniformization function, and it is a list of domains.}{Alpha.m}{Alpha`}
\index{\$coneList}

\alphausage{\$demoDirectory}{\$demoDirectory contains the path of the demo directory of \MMAlpha{}.}{Alpha.m}{Alpha`}
\index{\$demoDirectory}

\alphausage{\$depCone}{\$depCone is a glocal variable which contains the 
dependence cone of \$result. This variable is set by the
initUniformization function, and is in domain.}{Alpha.m}{Alpha`}
\index{\$depCone}

\alphausage{\$library}{\$library is the Mathematica symbol which holds the list of Alpha
programs (systems and subsystems) most recently loaded using
load. The system in \$result is also present in  \$library.}{Alpha.m}{Alpha`}
\index{\$library}

\alphausage{\$masterNotebook}{\$masterNotebook is the path of the master.nb notebook, which gives 
access to all demo notebooks on Alpha. To open it, use the command 
NotebookOpen[\$masterNotebook] or start[].}{Alpha.m}{Alpha`}
\index{\$masterNotebook}

\alphausage{\$MMALPHA}{\$MMALPHA contains the path of the \MMAlpha{} directory. This variable is 
set by an environment variable called MMALPHA.}{Alpha.m}{Alpha`}
\index{\$MMALPHA}

\alphausage{\$myMasterNotebook}{\$myMasterNotebook is the name of the notebook that you have placed
in the \$myNotebooks directory to access your own examples. To open
it, evaluate myStart[].}{Alpha.m}{Alpha`}
\index{\$myMasterNotebook}

\alphausage{\$myNotebooks}{\$myNotebooks contains the path of the directory containing your examples.
You can set this path in your init.m file. This file, placed in your home
directory, is loaded by Mathematica before any other Mathematica command.}{Alpha.m}{Alpha`}
\index{\$myNotebooks}

\alphausage{\$program}{\$program is a Mathematica symbol which keeps the abstract
syntax tree of the source of any transformation. It is initialized
by load, and changed when a (second) transformation is applied. The
most recent AST is kept in \$result (see ?\$result)}{Alpha.m}{Alpha`}
\index{\$program}

\alphausage{\$result}{\$result is the Mathematica symbol which holds the result of the
most recent operation (load or transformation). It is the default
source program for many transformations. When it is used by default by
a transformation, it is also modified by this transformation}{Alpha.m}{Alpha`}
\index{\$result}

\alphausage{\$rootDirectory}{\$rootDirectory contains the path of the \MMAlpha{} 
directory. It is set to \$MMALPHA}{Alpha.m}{Alpha`}
\index{\$rootDirectory}

\alphausage{\$schedule}{\$schedule is the Mathematica symbol that contains the schedule 
information of a system after the execution of the schedule or
the scd functions. The format of \$schedule is:

scheduleResult[name,

  List[\{var, varIndices, sched[tauVect, constCoef]\}],

  objFunction].

where name is the name of the system, var is a variable of sys,
varIndices is the list of indexes of var, tauVect is a list
of integer corresponding to the linear part of the schedule of
var, constCoef is the affine part, and objFunction is the 
objective function that was used to find the schedule.}{Alpha.m}{Alpha`}
\index{\$schedule}

\alphausage{\$scheduleLibrary}{\$scheduleLibrary is the list of the schedules that have been computed for 
various programs since the beginning of the current session.}{Alpha.m}{Alpha`}
\index{\$scheduleLibrary}

\alphausage{\$showGraph}{\$showGraph is a global variable that is set by the report function. Use is not
garanteed.}{Alpha.m}{Alpha`}
\index{\$showGraph}

\alphausage{\$testResult}{\$testResult is a global variable that is used to store the accumulated 
result value of a test.}{Alpha.m}{Alpha`}
\index{\$testResult}

\alphausage{\$testDirectory}{\$testDirectory is the path name of the directory that contains the 
test files for all packages. Currently set to \$MMALPHA/tests/}{Alpha.m}{Alpha`}
\index{\$testDirectory}

\alphausage{\$testIdentifier}{\$testIdentifier contains the name of the test that one wants to perform.
Default value is "".}{Alpha.m}{Alpha`}
\index{\$testIdentifier}

\alphausage{\$testReportFile}{\$testReportFile is the name of the reporting file for the tests. Its 
value is "" when no report file is open, otherwise, it contains the}{Alpha.m}{Alpha`}
\index{\$testReportFile}

\alphausage{\$tmpDirectory}{\$tmpDirectory is the pathname of the directory that will be used 
for all temporary files, i.e. /tmp/ on unix, C:/tmp/ on WindowsNT 
(strongly recommended).}{Alpha.m}{Alpha`}
\index{\$tmpDirectory}

\alphausage{\$runTime}{\$runTime is a global variable that stores the running time of 
the scheduler.}{Alpha.m}{Alpha`}
\index{\$runTime}

\alphausage{ashow}{ashow[sys] pretty prints in array notation the program contained in sys. Default value of sys is \$result. See also show.}{Alpha.m}{Alpha`}
\index{ashow}

\alphausage{ashowLib}{ashowLib[lib] prints in array notation all the systems of a library lib. 
Default value of lib is \$library. Warning, currently does not work 
on notebooks, use ashow[lib] instead. See also show, ashow.}{Alpha.m}{Alpha`}
\index{ashowLib}

\alphausage{asave}{asave[sys,filename] saves the array notation pretty printed version of
\$result in file filename. Default value of sys is \$result. See also
save, load, saveLib, asaveLib.}{Alpha.m}{Alpha`}
\index{asave}

\alphausage{asaveLib}{asavelib[lib,filename] saves in array notation all systems 
contained in library lib in file filename. Default value of 
lib is \$library. See also save, asave.}{Alpha.m}{Alpha`}
\index{asaveLib}

\alphausage{fshow}{obsolete form of show.}{Alpha.m}{Alpha`}
\index{fshow}

\alphausage{getPart}{getPart[exp,position] returns the subexpression of exp which is
designated by position.  The parameter position is a list of integers
following the convention of the Mathematica function Position. For
example, getPart[exp,\{2,3\}] identifies the third subtree of the second
subtree of exp. Warning: getPart does not accept in second parameter a
list of positions such as those returned by the Mathematica function
Position.}{Alpha.m}{Alpha`}
\index{getPart}

\alphausage{getSystem}{getSystem[sys,lib] extracts the system named sys from the library lib 
and selects it as the current system. Default value of lib is 
\$library. See also putSystem.}{Alpha.m}{Alpha`}
\index{getSystem}

\alphausage{load}{load[filename] parses the ALPHA program contained in file filename 
and returns the corresponding abstract syntax tree. As a side effect, 
symbols \$program and \$result are loaded with the parsed program.
See also save, asave, asaveLib.}{Alpha.m}{Alpha`}
\index{load}

\alphausage{myStart}{myStart[] opens the \$myMasterNotebook notebook placed in the
directory \$MMALPHA/myNotebooks.}{Alpha.m}{Alpha`}
\index{myStart}

\alphausage{on}{on[] is an obsolete form of start[]. It opens the Master 
Notebook.}{Alpha.m}{Alpha`}
\index{on}

\alphausage{putSystem}{putSystem[sys,lib] puts the system sys (default \$result) into library 
lib (default \$library). If sys already exists in lib, it is replaced, 
otherwise it is appended. See also getSystem.}{Alpha.m}{Alpha`}
\index{putSystem}

\alphausage{readAlpha}{readAlpha[filename] parses the ALPHA program in file filename
and returns its abstract syntax tree. Does not modify \$program 
nor \$result.}{Alpha.m}{Alpha`}
\index{readAlpha}

\alphausage{readDom}{readDom[dom] parses the ALPHA domain dom and returns its abstact syntax
tree. To parse a parametric domain such as \{i$|$i$<$N\}, it is
mandatory to supply the parameter domain or the parameter 
names as a second argument, for example:

readDom["\{i,j $|$ ... \}", \{"N"\}]

readDom["\{i,j $|$ ... \}",domain[1,\{"N"\}, ...] ] .}{Alpha.m}{Alpha`}
\index{readDom}

\alphausage{readExp}{readExp[exp] parses the ALPHA expression exp and returns its 
abstact syntax tree. readExp[exp,paramNames] or 
readExp[exp,paramDom] are used  to parse a parametric
expression. Example:

readExp["A.(i,j,N$ \rightarrow $N,N) + B.(i,N$ \rightarrow $i,N)"]

readExp["A.(i,j$ \rightarrow $N) + B.(i$ \rightarrow $i)", \{"N"\}]

readExp["A.(i,j$ \rightarrow $N) + B.(i$ \rightarrow $i)", domain[1,\{"N"\}, ...] ].}{Alpha.m}{Alpha`}
\index{readExp}

\alphausage{readMat}{readMat[dep] parses the ALPHA dependency dep and returns its matrix 
abstract syntax tree. To parse a parametric dependency such as (i,j$ \rightarrow $N),
it is mandatory to supply the parameter domain 
or the parameter names, e.g.:

readMat["(i,j,N$ \rightarrow $i)"]

readMat["(i,j$ \rightarrow $i)", \{"N"\}]

readMat["(i,j$ \rightarrow $i)",domain[1,\{"N"\}, ...] ] .}{Alpha.m}{Alpha`}
\index{readMat}

\alphausage{save}{save[sys,filename] saves the standard notation pretty printed
version of sys in file filename. Default value of sys is \$result.
See also load, asave, saveLib, asaveLib.}{Alpha.m}{Alpha`}
\index{save}

\alphausage{saveLib}{savelib[lib,filename] saves all systems contained in library lib
in file filename in standard notation.}{Alpha.m}{Alpha`}
\index{saveLib}

\alphausage{show}{show[var] pretty-prints the program, the domain, the matrix or 
the schedule contained in symbol var. Default value of var is 
\$result. show[var, p] pretty prints the domain or the matrix 
contained in symbol var taking parameter domain p into account. 
var and p should be abstract syntax trees.}{Alpha.m}{Alpha`}
\index{show}

\alphausage{showLib}{showLib[lib] prints all the systems of library lib in standard notation. 
Default value of lib is \$library. Warning, currently does not 
work in notebooks, use show[lib] instead}{Alpha.m}{Alpha`}
\index{showLib}

\alphausage{showMat}{showMat[m] pretty-prints the matrix m in Alpha format.}{Alpha.m}{Alpha`}
\index{showMat}

\alphausage{start}{start[] opens the Master Notebook.}{Alpha.m}{Alpha`}
\index{start}

\alphausage{systemNames}{systemNames[] returns the list of system names loaded in \$library.}{Alpha.m}{Alpha`}
\index{systemNames}

\alphausage{testFunction}{testFunction[expr,resOfExpr,message\_String] is used for building tests. It
evaluates expr and compare it the resOfExpr, return True if the two are 
equal, False otherwise and prints a standard message (using the information 
of message: usually a test number to identify the test). This function is
for developers.}{Alpha.m}{Alpha`}
\index{testFunction}

\alphausage{tests}{tests[PackageName] call the test procedure for the package 
PackageName. tests[] calls the test procedure 
for all \MMAlpha{} packages. Warning: a complete test takes a long time.}{Alpha.m}{Alpha`}
\index{tests}

\alphausage{varTypes}{varTypes[sys] lists the types of all variables of sys. Default value
of sys is \$result.}{Alpha.m}{Alpha`}
\index{varTypes}

\alphausage{writeC}{writeC[sys,f,opts] generates C code from the system sys in file f.  
Default value of sys is \$result and default value of f is "Alpha.c". 
opts are options that are sent to the C generator. 
Option "-p num1 num2 ..." sets Alpha parameters to value 
num1, num2, etc. The C code is correct only if all Alpha parameters 
are assigned values. The "-g" option provides a debug version, 
where all equation calls are printed out. The "-s" option is for 
the generation of C code to be interfaced with the Signal language. 
WARNING: writeC does not work properly for systems with unbounded domains.}{Alpha.m}{Alpha`}
\index{writeC}

\alphausage{writeTex}{writeTex[sys,f] generates the Latex form of the program
contained in symbol sys (default \$result) into file f 
(default "Alpha.tex") of the current directory. 
writeTex[...,"-a"] produces a program in array notation. 
Warning: writeTex overwrites an already existing output f file.}{Alpha.m}{Alpha`}
\index{writeTex}

\alphanote{writeTex}{This function is available only on Unix, and seldom used. Its 
output is not guaranteed.}

\alphausage{astQ}{astQ[exp] is True if expression exp is an AST, False otherwise.}{Alpha.m}{Alpha`}
\index{astQ}

\alphausage{fileName}{fileName[\{\_\_String\}] returns the path expression corresponding to the
file according to the operating system in use. Actually, this function 
is not necessary, since Mathematica converts all pathnames from the 
Unix form.}{Alpha.m}{Alpha`}
\index{fileName}

\alphausage{openTemporary}{openTemporary[] works as the Mathematica function OpenTemporary, but
opens the file in the \$tmpDirectory and return the corresponding 
stream. The command Close[openTemporary[] returns the name of 
the temporary file (warning: there is an inconsistency in this 
function due to Mathematica between the Unix version and the WindowsNT 
version: on Unix, the full name returned is ("/tmp/...") while 
on WindowsNT, only the name of the file is returned. Please use 
getTemporaryName[] to get the name of a temporary file instead.}{Alpha.m}{Alpha`}
\index{openTemporary}

\alphausage{getTemporaryName}{getTemporaryFile[] returns the full name of a new temporary file. 
It is equivalent to Close[OpenTemporary[]] on a unix plateform and 
it has the same behaviour on other plateforms.}{Alpha.m}{Alpha`}
\index{getTemporaryName}

\alphausage{demoLink}{demoLink[demo] creates an hyperlink button aiming at the Alpha 
demo notebook "demo.nb". This button may then be cut and pasted anywhere. 
Example:

demoLink["Fir"]

creates a button aiming at the demo notebook Fir.nb 
placed in directory \$MMALPHA/demos/NOTEBOOKS/Fir.

demoLink[d,demo] creates an hyperlink button aiming at notebook 
demo.nb in directory d of \$MMALPHA/demos/NOTEBOOKS.}{Alpha.m}{Alpha`}
\index{demoLink}

\alphausage{docLink}{docLink[file] creates an hyperlink button
aiming at the Alpha documentation notebook "file.nb". This button may then be
cut-and-pasted anywhere. docLink[d:\_String, file:\_String] creates an
hyperlink button aiming at notebook file.nb in directory d of
\$MMALPHA/doc/packages.}{Alpha.m}{Alpha`}
\index{docLink}

\alphausage{link}{link[file] creates an hyperlink button aiming at the Alpha
notebook "file.nb" in directory \$myNotebooks. This button may then
be cut-and-pasted anywhere. link[d, file] creates an
hyperlink button aiming at notebook "file.nb" in directory d of
\$myNotebooks.}{Alpha.m}{Alpha`}
\index{link}

\alphausage{exportAlphaFunctions}{exportAlphaFunctions[] writes DeclarePackage commands for all functions 
of the \MMAlpha{} standard packages. This function is automatically executed
when loading Alpha.m (see after the EndPackage instruction) whenever the
autoload.m does not exists, and creates this file. The autoload.m file 
is located in the Alpha directory. Each time you want to add a new package
to \MMAlpha{}, add an entry in the local variable "contexts" of Alpha.m, then 
remove the autoload.m file: a new one will be created.}{Alpha.m}{Alpha`}
\index{exportAlphaFunctions}

\alphausage{setMMADir}{setMMADir[List[\_\_\_String]] does the SetDirectory[] function work, but 
the path is specified as a list of strings and hence is valid on 
different plateforms (Windows, Unix).}{Alpha.m}{Alpha`}
\index{setMMADir}

\alphanote{setMMADir}{Actually this function is not
needed as Mathematica changes on the fly the path names depending on 
the platform.}

\alphausage{wrap}{wrap[exp,contextFile,result] evaluates the expression exp in the
context given in the file contextFile, and returns the result in 
the file result. This function is not currently finished.}{Alpha.m}{Alpha`}
\index{wrap}

\alphanote{wrap}{wrap is used for running examples in separate sessions
of Mathematica, for example, using a remote \MMAlpha{} installation. The
idea would be to control the execution, and to use only one licence
to run these examples. The price to pay would be the start time of 
Mathematica.}

\alphausage{enTete}{enTete[symb1\_] prints out the standard skeleton for
programming a new function named symb1 in \MMAlpha{} (for developper use
only).}{Alpha.m}{Alpha`}
\index{enTete}

\alphausage{mute}{mute is an option (Boolean). If True, function prints absolutely no information.}{Alpha.m}{Alpha`}
\index{mute}

\alphausage{test1}{test1[] executes part 1 of tests. See documentation in doc/Tests.}{Alpha.m}{Alpha`}
\index{test1}

\alphausage{test2}{test2[] executes part 2 of tests. See documentation in doc/Tests.}{Alpha.m}{Alpha`}
\index{test2}

\alphausage{test3}{test3[] executes part 3 of tests. See documentation in doc/Tests.}{Alpha.m}{Alpha`}
\index{test3}

\alphausage{test4}{test4[] executes part 4 of tests. See documentation in doc/Texts.}{Alpha.m}{Alpha`}
\index{test4}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Alphard.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 19.124067} by makeDoc.
%
% Section header
\section{The Alpha`Alphard` package } 
\label{label:Alpha`Alphard`}
\alphausage{alpha0ToAlphard}{alpha0ToAlphard[sys,controlSigList]  translate the Alpha0  system `sys' into an Alphard library, `controlSigList' is the list of names of the control signals.    alpha0ToAlphard[controlSigList] happens on   \$result and modify \$result and \$library}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{alpha0ToAlphard}

\alphausage{alpha0ToAlphardModule}{alpha0ToAlphardModule[sys,listRegion,controlSigList] translate the Alpha0   Module `sys' into an Alphard library. `listRegion' is the structural   information resulting from the function getArrayDomains[sys],  `controlSigList' is the list of names of the control signals. }{Alpha/Alphard.m}{Alpha`Alphard`}
\index{alpha0ToAlphardModule}

\alphausage{alphardFirstStep}{alphardFirstStep[] returns an integer which is the first logical steps of the  alphard architecture. The reset signal should be sent at this step.}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{alphardFirstStep}

\alphausage{alphardTimeLife}{alphardTimeLife[sys:\_Alpha`system] returns the time domain spanned by the hardware. This function should be called on the module of an Alphard program after the parameter have been assigned (your can call it before the assignement, but the result will be more difficult to interpret), alphardTimeLife[] applies on \$library (except the last program)  and take the union}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{alphardTimeLife}

\alphausage{getArrayDomains}{getArrayDomains[sys] computes all the spatial region with a  different behaviour determined by the system sys (on each  spatial region, cells are idential). The system sys must be in Alpha0 form. getArrayDomains[] applies to \$result.}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{getArrayDomains}

\alphausage{buildControler}{buildControler[sys,signalList] builds the subsystem which initializes all the control signals of `sys' indicated   by `signalList'. The system `sys' must be in Alpha0 form. buildControler[signalList] takes \$result as default   value for `sys'}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{buildControler}

\alphausage{isSpaceDepQ}{isSpaceDepQ[dep] checks whether the Alpha dependance `dep' is a space dependance  (in Alpha0). `dep' is an AST}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isSpaceDepQ}

\alphausage{isConnexionEqQ}{isConnexionEqQ[eq] checks whether the Alpha equation `eq' is a connexion equation (in Alpha0). `eq' is an AST}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isConnexionEqQ}

\alphausage{buildOneCell}{buildOneCell[sys,dom1,posList] return the Alphard cell (extracted form system `sys'), occuring on the  spatial region indicated by `dom1' and which  computes expressions indicated by the list of poisiton `posList' in AST `sys'. `dom' and `posList' can be obtained by the function  getArrayDomains[sys]. buildOneCell[sys,dom1,posList,controlSignals] do the same with the additionnal information of the list of control signals in `controlSignals'. }{Alpha/Alphard.m}{Alpha`Alphard`}
\index{buildOneCell}

\alphausage{isMirrorEqQ}{isMirrorEqQ[eq] checks whether the Alpha equation `eq' is a Mirror equation (in Alpha0). `eq' is an AST}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isMirrorEqQ}

\alphausage{buildInterface}{buildInterface[sys] returns two Alphard systems: the interface  and the module which correspond to system sys. The interface has  the same inputs and outputs as sys and simply calls the module  with scheduled input and output (Mirror variables for i/o).  buildInterface[] applies on \$result and modifies \$result and \$library.}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{buildInterface}

\alphausage{isModuleQ}{isModuleQ[sys] checks whether the Alpha system `sys' is a module or not}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isModuleQ}

\alphausage{removeSystem}{removeSystem[id\_String]  Remove the system named `id' from the library \$library  (changes \$program and \$result if system `id' is also \$result). }{Alpha/Alphard.m}{Alpha`Alphard`}
\index{removeSystem}

\alphausage{simplifyConnexions}{simplifyConnexions[sys] translate connexion syntax of connexion from Alpha0 mode to Alphard mode in system `sys', return the modified system. simplifyConnexions[] simplify connexion of \$result}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{simplifyConnexions}

\alphausage{normalizeIdDep}{normalizeIdDep[sys], normalize identity dependencies: replace in system `sys' all occurences of Alpha`var[a\_\_\_] which is not surrounded by an Alpha`affine by Alpha`affine[Alpha`var,identity] and return the modified system. normalizeIdDep[] apply the function to \$result and assigns the result to \$result. e.g A[t,p]=B is replaced by A[t,p]=B[t,p]. Warning, the transformation is not applied to the input arguments of Alpha`use }{Alpha/Alphard.m}{Alpha`Alphard`}
\index{normalizeIdDep}

\alphausage{normalizeIdDepInEq}{normalizeIdDep[sys,eq] normalize identity dependencies: replace in equation `eq' of system `sys' all occurences of Alpha`var[a\_\_\_] which is not surrounded by an Alpha`affine by Alpha`affine[Alpha`var,identity] and return the modified equation}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{normalizeIdDepInEq}

\alphausage{normalizeIdDepLib}{normalizeIdDepLib[] applies normalizeIdDep[] to all subsystems of the library, assigns it to \$library and  return the resulting Library (change also \$result)}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{normalizeIdDepLib}

\alphausage{isolateOutputList}{isolates outputs of a system in the library, so that an output is never used inside its generating system  (necessary for VHDL)}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isolateOutputList}

\alphausage{isolateOutput}{isolates outputs of a system in the library, so that an output is never used inside its generating system  (necessary for VHDL)}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{isolateOutput}

\alphausage{structureFrom}{structureFrom[indices\_List[\_\_String],listAllVars\_List[\_\_String]] build a subsystem from \$result which defines all the variables in `listAllVars', and for which the indices in `indices' will be the extension indices of the use set in \$result. All variables must give   the same domain when projecting then on the indices `indices'.  Can also be called in the forllowing form : structureFrom[indices], structureFrom[sys,indices], structureFrom[sys,indices,listStructVar],   structureFrom[sys,indices,listStructVar,newName]. Warning, not tested}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{structureFrom}

\alphausage{setOutputVar}{setOutputVar[sys1\_systel,var1\_String] returns sys1 in which var1 in moved from local vars to output vars (added as last output var). setOutputVar[var1] operates on \$result. WARNING this function changes the semantics of the program (the use calling this program are not valid anymore)}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{setOutputVar}

\alphausage{insertFunction}{inserFunction[var1\_String, func1\_String] insert the function func1 in the definition of var var1: if the definition of var1 was var1=expr1, the new definition is var1=func1(var1). This is mainly used for the introduction of functions like truncateLSB for bit level description of the operations.}{Alpha/Alphard.m}{Alpha`Alphard`}
\index{insertFunction}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/ChangeOfBasis.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 18.848102} by makeDoc.
%
% Section header
\section{The Alpha`ChangeOfBasis` package } 
\label{label:Alpha`ChangeOfBasis`}
\alphanote{ChangeOfBasis}{Documentation revised on August 3, 2004}

\alphausage{ChangeOfBasis}{The Alpha`Package` package contains the definition of the "change of basis" transformation (see changeOfBasis).}{Alpha/ChangeOfBasis.m}{Alpha`ChangeOfBasis`}
\index{ChangeOfBasis}

\alphausage{changeIndexes}{changeIndexes[ sys, var, rules ] replaces all indexes in the definition of var according to replacement rules. changeIndexes[ var, rules ] does the same on \$result. Remember that var names are strings. The declaration of var is  modified according to the rules. If var is the output of a use statement, all outputs of the use and the use itself are changed.}{Alpha/ChangeOfBasis.m}{Alpha`ChangeOfBasis`}
\index{changeIndexes}

\alphausage{changeOfBasis}{changeOfBasis[var.fn] returns a copy of \$result in which the variable var is reindexed by an unimodular affine function fn. The change of basis is specified as "B.(i,j,k $ \rightarrow $ i+1,j,k)" where B is the variable to be reindexed, and the change of basis matrix has the format of an Alpha dependence, with one difference: the index mapping function is understood as a mapping from initial to new position (and not the opposite). If it is square, this matrix should be unimodular. Non-square change of basis (also called generalized  change of basis) are allowed, in that case, the function should be called this way: changeOfBasis[var.fn, \{index, ...\}] applies a change of basis fn on variable var, except that the new indices of var are named according to the second argument.\\Example: \\changeOfBasis["B.(i,j $ \rightarrow $ 1,i,j)",\{"i1","j1","k1"\}]). \\changeOfBasis[sys,var.fn, \{index, ...\}]  applies a change of basis to system sys instead of \$result.  \\changeOfBasis["B.(i,j $ \rightarrow $ 1,i,j)", recurse$ \rightarrow $True]  recursively executes the change of basis on subsystem `subsys' called  with B as input or output. The recursive change of basis has  many restriction: as the semantics of the subsystem is modified,  we impose that there is only one occurence of subsys appearing  in the system. Moreover, the recursive change of basis can  modify only local indices (indices which are not extension  indices in the use of subsys) and must involve only local  indices and  parameter transmitted to the subsystem.  The  change of basis on the extension indices can be performed  with the function extDomainCOB[] (see ?extDomainCOB).}{Alpha/ChangeOfBasis.m}{Alpha`ChangeOfBasis`}
\index{changeOfBasis}

\alphausage{extDomainCOB}{extDomainCOB["subSysName.(j$ \rightarrow $g(j,N))"] applies the change of basis  to all the variables implied in the use of the subsystem subSysName  and change the extension domain accordingly. This change of basis  should be applied to a system where the following use appear:  \\use \{j$|$ ....\} subSysName[...] (...) returns (...) \\j are the extension indices, N are the parameter of the caller.  The change of basis must only depend on the extention indices and on the parameters as indicated here. The parameter assignement  function must be simple (i.e. each parameter of the subsystem is  assigned to a parameter of the caller, e.g. (N,M,P$ \rightarrow $P,N). This transformation should have no impact on the subsystems. }{Alpha/ChangeOfBasis.m}{Alpha`ChangeOfBasis`}
\index{extDomainCOB}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Control.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 19.976122} by makeDoc.
%
% Section header
\section{The Alpha`Control` package } 
\label{label:Alpha`Control`}
\alphausage{Control}{Package. Contains the definition of functions related to control signal generation: temporalCaseQ[], spatialCaseQ[], spaceTimeCase[], spaceTimeDecomposition[], needsMuxQ[], makeMuxControl[], makeAllMuxControl[], isControlEquQ[], controlVars[].}{Alpha/Control.m}{Alpha`Control`}
\index{Control}

\alphausage{makeOneMuxControl}{makeOneMuxControl[sys, tpos, spos, var, options] generates a multiplexer for the definition of var.}{Alpha/Control.m}{Alpha`Control`}
\index{makeOneMuxControl}

\alphausage{makeAllMuxControl}{makeAllMuxControl[sys, tpos, spos, options] generates multiplexers  and their control variables for all the variables in the system sys  needing it. The system sys must be in space/time form  (see spaceTimeDecomposition). The modified system is returned. makeAllMuxControl[tpos, spos, options], applies to \$result  and modifies it.}{Alpha/Control.m}{Alpha`Control`}
\index{makeAllMuxControl}

\alphausage{temporalCaseQ}{temporalCaseQ[sys, expr, post, poss] checks whether or  not the case expression expr of sys is a temporal case  (all alternatives are defined over the same spatial domain.)  post is the list of positions of temporal indices and poss is  the list of positions of spatial indices. WARNING: This function checks pure temporal case (no condition  involving space can change in the branches of the case)}{Alpha/Control.m}{Alpha`Control`}
\index{temporalCaseQ}

\alphausage{spatialCaseQ}{spatialCaseQ[sys, expr, post, poss] checks whether or not the case expression expr of sys is a spatial case (i.e., can be rewritten using conditions on spatial indices only).  post is the list of positions of temporal indices, and poss is the list of positions of spatial indices.}{Alpha/Control.m}{Alpha`Control`}
\index{spatialCaseQ}

\alphausage{spaceTimeCase}{spaceTimeCase[sys, varname, tpos, spos] unrolls the normalized  case-based definition of local variable varname to a double case  spatial then temporal. spaceTimeCase[varname, tpos, spos] applies to  \$result and modifies it.}{Alpha/Control.m}{Alpha`Control`}
\index{spaceTimeCase}

\alphausage{spaceTimeDecomposition}{spaceTimeDecomposition[sys,  tpos, spos] transforms all the local  variables of sys into their space-time case form. spaceTimeDecomposition[varname, tpos, spos] applies to \$result and modifies it.}{Alpha/Control.m}{Alpha`Control`}
\index{spaceTimeDecomposition}

\alphausage{needsMuxQ}{needsMuxQ[exp] returns True if expr of \$result needs a multiplexer,  False otherwise.}{Alpha/Control.m}{Alpha`Control`}
\index{needsMuxQ}

\alphausage{makeMuxControl}{makeMuxControl[var, ctrlVar, tpos, spos] builds the mux control variables for variable var. It is restricted to cases with two branches.  WARNING: DOES not check that the temporal cases are binary.}{Alpha/Control.m}{Alpha`Control`}
\index{makeMuxControl}

\alphausage{isControlEquQ}{isControlEquQ[equation] checks that equation is a controler equation.}{Alpha/Control.m}{Alpha`Control`}
\index{isControlEquQ}

\alphausage{controlVars}{controlVars[sys] returns the list of control variables in system sys. controlVars[] applies to \$result. A control variable is defined by a space/time case with two time branches equal to True and False.}{Alpha/Control.m}{Alpha`Control`}
\index{controlVars}

\alphausage{makeBinaryCases}{makeBinaryCases[ sys, var, opts ] replaces in the definition of  var, all case statements with more than 2 branches by binary  cases. makeBinaryCases[ var, opts ] does the same to \$result.}{Alpha/Control.m}{Alpha`Control`}
\index{makeBinaryCases}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Cut.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 19.090285} by makeDoc.
%
% Section header
\section{The Alpha`Cut` package } 
\label{label:Alpha`Cut`}
\alphanote{Cut}{Documentation revised on August 3, 2004.}

\alphausage{Cut}{Alpha`Cut` is the package which contains the cut, decompose  and merge transformations. }{Alpha/Cut.m}{Alpha`Cut`}
\index{Cut}

\alphausage{cut}{cut[var\_String, dom\_String, outvar1\_Sting, outvar2\_String] returns the program obtained by cutting the definition of var in program \$result into two definitions outvar1 and outvar2, where outvar1 is restricted to dom, and outvar2 is restricted to the complemtary space of dom. The result of cut is put in symbol \$result. cut[sys, var, dom, outvar1, outvar] does the same to program sys. The parameter dom may be specified either as a string, or as the internal Alpha form of a domain. \\Example: cut["A","\{i,j $|$ i$<$j \}","A1", "A2"].}{Alpha/Cut.m}{Alpha`Cut`}
\index{cut}

\alphausage{decompose}{decompose[expr,name] returns the program obtained by adding to  \$result a new equation "name = expr" and replacing the first  occurrence of expr by name. The program returned is assigned to \$result.  decompose[sys,expr,name] does the same to program sys, but does not modify \$result. expr and name are either strings or ast's. expr can also be specified using a Mathematica Position.}{Alpha/Cut.m}{Alpha`Cut`}
\index{decompose}

\alphausage{exprLocalEquivQ}{Predicate. Checks whether or not two expressions in an ALPHA programs are equivalent in any point of a given domain. Obsolete function.}{Alpha/Cut.m}{Alpha`Cut`}
\index{exprLocalEquivQ}

\alphausage{merge}{merge[var1, var2, newVar] replaces in \$result two local variable  definitions with a single one. The declaration domain of the new  variable is the union of the domains of the old ones. The definition  of the new variable is a case whose branches are definitions of  the old variables restricted to their respective domains. RHS occurrences of the old variables are replaced by the proper restrictions of the new variable.\\ merge[sys, var1, var2, newVar] replaces in system sys two local  variable definitions with a single one and returns the new system.}{Alpha/Cut.m}{Alpha`Cut`}
\index{merge}

\alphausage{mergeCaseBranches}{mergeCaseBranches[casePosition\_List, branchPosition\_List], merges (in \$result) the specified case branches if they contain identical expressions. mergeCaseBranches[sys, casePosition, branchPosition]  merges the  specified case branches of system sys if they contain  identical expressions and returns the new system. The domain of the  new branch may be non-convex.}{Alpha/Cut.m}{Alpha`Cut`}
\index{mergeCaseBranches}

\alphausage{mergeIdCaseBranches}{mergeIdCaseBranches[sys] tries to merge case branches that have  identical expressions (i.e. tries to inverse what splitCaseUnion did).  Warning, this function works only on normalized programs.}{Alpha/Cut.m}{Alpha`Cut`}
\index{mergeIdCaseBranches}

\alphausage{splitCaseUnion}{splitCaseUnion[sys] splits all the case branches that have a  non convex domain as restrict (union of convex) by several branches,  each corresponding to one convex domain. Warning, this function works only on normalized programs.}{Alpha/Cut.m}{Alpha`Cut`}
\index{splitCaseUnion}

\alphausage{unionMerge}{unionMerge[sys, firstVar, secondVar, resultingVar] is an extension of merge that handles overlapping definitions parameters. Returns the modified program. Warning obsolete function.}{Alpha/Cut.m}{Alpha`Cut`}
\index{unionMerge}

\alphanote{unionMerge}{unionMerge acts much like merge, the difference being that overlapping domains are accepted if both expressions are equivalent over the overlap. By equivalent we mean that the structures of the expressions are identical and that the images of the overlap domain through the dependence functions are equal for any dependence. To illustrate the latter, consider the expressions \\\\\\ \{i,j$|$i=j; 1$<$=j$<$=4\\\} : A.(i,j$ \rightarrow $i,j) \\\\\\and  \\\\\\\{i,j$|$i=j; 1$<$=j$<$=4\} : A.(i,j$ \rightarrow $j,j).\\\\\\ Although the dependence functions are not equal, these expressions are equivalent over the domain \{i,j$|$i=j; 1$<$=j$<$=4\}.}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Domlib.tex
Added mmalpha (source and binary)

%
% File created on {2009, 4, 25, 17, 47, 46.071200} by makeDoc.
%
% Section header
\section{The Alpha`Domlib` package } 
\label{label:Alpha`Domlib`}
\alphanote{Domlib}{Documentation revised on August 10, 2004}

\alphausage{DomLib}{Domlib is a library of domain functions. It contains the functions const2al,
const2mma, dom2mma, dom2al, domCompRays, DomAddRays, DomBasis, DomConstraints,
DomConvex, DomCost, DomDifference, DomEmpty, DomEmptyQ, DomEqualities,
DomEqualQ, DomExtend, DomImage, DomIntersection, DomIntEmptyQ, DomLeftHermite,
DomLTQ, DomMatrixSimplify, DomPreimage, DomProject, DomRays, DomRightHermite,
DomSimplify, DomSort, DomUnion, DomUniverse, DomUniverseQ, DomVertices,
DomVisual, DomZImage, DomZPreimage, hypercube, LatticeDifference,
LatticeImage, LatticeIntersection, LatticeHermite, LatticePreimage,
linearConstraintQ, linearExpQ, linHalfSpace, polToZpol, rays, vertices,
and zpolToPol.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomLib}

\alphanote{Domlib}{Domlib is based upon the PolyLib library, which is a set of public domain 
C programs first developed at Irisa. Domlib is interfaced to this library
using MathLink.}

\alphausage{domlib}{Math link. Link to the external program "domlib". The mechanism of 
MathLink is pretty complicated and needs to be clarified here.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{domlib}

\alphausage{const2al}{const2al[ind,c] translates in Alpha form the constraint c. 
ind is an index list (e.g. \{"i","j"\}) and c is a constraint in
form eq[i+2j,3] or ge[i+2j,3].}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{const2al}

\alphausage{const2mma}{const2mma[indexList,vectorList] converts a constraint, expressed as a list of 
index names and a list of vectors, into Mathematica form}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{const2mma}

\alphausage{dom2mma}{dom2mma[d] converts Alpha domain d into a pair \{constraints,index\} (e.g.
\{\{i,j\},\{i+j$>$=2,...\}\}) suitable for Mathematica.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{dom2mma}

\alphausage{dom2al}{dom2al[\{cst,ind\}] translates a domain given in MMA
form into its Alpha encoding. ind is a list of indexes, and cst is
a list of constraints of the form i+j $>$ 2. Warning: left-hand side is
the linear part, and right-hand side is an integer.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{dom2al}

\alphanote{dom2al}{In dom2al, the symbols must be strings, otherwise there is a problem.
This has to be checked.}

\alphausage{domCompRays}{domCompRays[dom] recomputes the rays of the Alpha domain dom and
returns an Alpha domain. domCompRays ignores the ray part of dom, and
recomputes it. This function allows one to modify a contraint in a domain,
and to update the domain accordingly.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{domCompRays}

\alphausage{domHalfSpaceQ}{domHalfSpaceQ[dom] is True if the Alpha domain dom is a half-space. dom 
is either a string or an ast.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{domHalfSpaceQ}

\alphausage{DomAddRays}{DomAddRays[dom,m] returns the domain dom augmented with rays from the matrix
m. rays are in Alpha format (i.e. n+2 components for n dimensions).

Example:

dom is \{i,j $|$ i $>$= 0\},

m=matrix[3, 4, \{\}, \{\{1, 0, 1, 0\}, \{1, 10, 10, 1\}\}],

DomAddRays[dom,m] returns \{i,j $|$ 0$<$=i$<$=(j,10)\}.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomAddRays}

\alphausage{DomBasis}{DomBasis[m] returns a row basis of the Alpha matrix m using Gauss 
Jordan-elimination.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomBasis}

\alphausage{DomConstraints}{DomConstraints[m] returns the minimum convex polyehdron defined by 
the constraint matrix m. DomConstraints[m1,m2] returns the Z-polyhedron 
obtained by image of the polyhedrom P by the invertible mapping m1, 
where P is the minimum convex Polyhedron satisfying the constraints 
given in m2, i.e., Z = m1(DomConstraints[m2]).}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomConstraints}

\alphausage{DomConvex}{DomConvex[d] returns the minimum convex polyhedron which encloses the
(polyhedral or Z) domain d. Warning (4/11/98), this function was bugged 
in Polylib, and the current implementation uses DomRays instead of 
DomConvex.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomConvex}

\alphausage{DomCost}{DomCost[d, c] returns the interval of values of the cost function c
evaluated over domain d = \{MinN, MinD, MinI, MaxN, MaxD, MaxI\}.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomCost}

\alphanote{DomCost}{DomCost is a function of the Domlib library, and what it does in 
not clear.}

\alphausage{DomDifference}{DomDifference[d1, d2] returns the domain difference of domain d1 less d2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomDifference}

\alphausage{DomEmpty}{DomEmpty[n] returns the empty domain of dimension n.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomEmpty}

\alphausage{DomEmptyQ}{DomEmptyQ[d] returns True if domain d is empty, False otherwise. The
test of emptyness is based uniquely on the rational polyhedron, not the 
integral polyhedron.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomEmptyQ}

\alphausage{DomEqualities}{DomEqualities[d] returns the matrix of equations (lineality space) of
domain d (does not handle union of convexes).}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomEqualities}

\alphausage{DomEqualQ}{DomEqualQ[d] returns True if domain d1 is equivalant to d2, False
otherwise.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomEqualQ}

\alphausage{DomExtend}{DomExtend[dom, idx] extends  dom to the indices in the index list idx. 
This list should contains the indices of dom.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomExtend}

\alphausage{DomImage}{DomImage[d,m] returns the image of the domain d under the transformation 
matrix m. This function always returns a polyhedral domain. 
If d is a Z-Domain, it returns the image of the rational polyhedral 
domain enclosing the Z-Domain.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomImage}

\alphausage{DomIntersection}{DomIntersection[d1, d2] returns the domain intersection of domains d1 and d2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomIntersection}

\alphausage{DomIntEmptyQ}{DomIntEmptyQ[dom1, dom2] returns True if dom1 contains no integral points
in the context of dom2, False otherwise. ??}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomIntEmptyQ}

\alphausage{DomLeftHermite}{DomLeftHermite[m] returns \{H, Q\} where m = HQ, Q unimodular, H hermite.
(Warning, bugged function please use hermiteL[]).}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomLeftHermite}

\alphausage{DomLTQ}{DomLTQ[dom1, dom2, idx, pdim] compares dom1 and dom2 at index position
idx (integer). pdim is the dimension of the parameter space. Returns 1 if
dom1$>$dom2, returns -1 if dom1$<$dom2, returns 0 if dom1$>$$<$dom2 (whatever it
means.)}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomLTQ}

\alphanote{DomLTQ}{DomLTQ does not check the type and number of parameters.}

\alphausage{DomMatrixSimplify}{DomMatrixSimplify[m1,m2] returns m1 simplified in the presence of m2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomMatrixSimplify}

\alphausage{DomPreimage}{DomPreimage[d,m] returns the preimage of the domain d under the transformation 
matrix m. This function always returns a polyhedral domain. For a
ZDomain, it returns the pre-image of the rational polyhedral domain 
enclosing the ZDomain.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomPreimage}

\alphausage{DomProject}{DomProject[dom, idx] projects dom onto the indices given 
in the index list idx. Reorders indices and changes the dimension
accordingly.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomProject}

\alphausage{DomRays}{DomRays[m] returns the minimum convex polyhedron defined by the rays 
given in the matrix m.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomRays}

\alphausage{DomRightHermite}{This function is bugged, use hermiteR instead.
DomRightHermite[m] returns the Hermite decomposition \{Q, H\} of the
Alpha matrix m, i.e. m = QH, Q unimodular, H Hermite.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomRightHermite}

\alphausage{DomSimplify}{DomSimplify[d1, d2] returns a domain equal to d1 simplified in 
the context of d2. In other words, we remove of the definition of
d1 all constraints which are implied by d2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomSimplify}

\alphausage{DomSort}{DomSort[ldom, idx, pdim, time:True$|$False] returns the 
topological ordered list of domains ldom. idx is the level to 
consider for sorting, pdim is the parameter space dimension. 
Returns a list of logical times (one per domain) if time is True, 
otherwise it returns a permutation of ldom. An application of this 
permutation to ldom returns a sorted list.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomSort}

\alphanote{DomSort}{This function seems to be fragile. The only place where it is used 
is in the INorm package.}

\alphausage{DomUnion}{DomUnion[d1, d2] returns the domain union of domains d1 and d2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomUnion}

\alphausage{DomUniverse}{DomUniverse[n]	returns the universe domain of dimension n.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomUniverse}

\alphausage{DomUniverseQ}{DomUniverseQ[d] returns True if domain d is the universe, False otherwise.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomUniverseQ}

\alphausage{DomVertices}{DomVertices[ldom, context] finds the parametrized vertices of a list 
of parametrized polyhedra ldom. context is the domain of 
parameters. It returns a list of pairs whose elements contain 
a parameter domain, and a list of parametrized vertices. 
DomVertices[pol, param] finds the parametrized vertices 
of a parametrized polyhedron pol.
DomVertices[pol], finds the vertices of the non-parametrized 
polyhedron pol. It returns \{\{e,l\}\} where e is the universe 0-domain and 
l is a list of non-parametrized vertices.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomVertices}

\alphanote{DomVertices}{This function seems to be fragile. It is used nowhere in \MMAlpha{}.}

\alphausage{DomVisual}{DomVisual[d1,d2] visualizes the domain d1 with the Opera tool. d2 is
the parameter domain. Warning if the domain d2 is ommited, it is
replaced by the parameter domain of \$result. WARNING: CURRENTLY NOT
AVAILABLE (the function does nothing).}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomVisual}

\alphausage{DomZImage}{DomZImage[dom, mat] finds the Z-image of the domain dom by the matrix mat.
This function may or may not return a Z-Domain depending on the 
transformation matrix.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomZImage}

\alphausage{DomZPreimage}{DomZPreimage[dom, mat] finds the Z-Preimage of the domain dom by the 
matrix mat. This function may or may not return a Z-Domain depending on 
the transformation matrix.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomZPreimage}

\alphausage{hypercube}{hypercube[n] creates a hypercubic domain (in Mathematica form) of dimension n, 
and size 10. dom2l[hypercube[n]] allows such a domain to 
be translated into Alpha form. Useful for testing purposes.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{hypercube}

\alphausage{LatticeDifference}{LatticeDifference[m1,m2] returns the difference of the lattices m1 and m2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{LatticeDifference}

\alphanote{LatticeDifference}{Seems to return a list of matrices.}

\alphausage{LatticeImage}{LatticeImage[m1,m2] returns the image of the lattice m1 by the function m2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{LatticeImage}

\alphausage{LatticeIntersection}{LatticeIntersection[m1,m2] returns the intersection of the lattices m1 and m2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{LatticeIntersection}

\alphausage{LatticeHermite}{LatticeHermite[m] returns the lattice m in Hermite Normal Form.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{LatticeHermite}

\alphausage{LatticePreimage}{LatticePreimage[m1,m2]	returns the preimage of the lattice m1 by the 
function m2.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{LatticePreimage}

\alphausage{linearConstraintQ}{linearConstraintQ[c,\{symbols\}] is True if c is a linear constraint
formed with symbols, False otherwise.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{linearConstraintQ}

\alphausage{linearExpQ}{linearExpQ[exp,\{symbols\}] is True if exp is a linear expression 
formed with symbols, False otherwise.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{linearExpQ}

\alphausage{linHalfSpace}{linHalfSpace[h] returns the Mathematica Matrix corresponding to the Alpha
half-space h.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{linHalfSpace}

\alphausage{polToZpol}{If the domain d is a union of polyhedra, polToZpol[d] returns the
equivalent Z-polyhedron, otherwise it returns d as is.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{polToZpol}

\alphausage{rays}{rays[d] returns the list of rays of the Alpha domain d. rays[\{const,index\}]
returns the vertices of d given in Mathematica form.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{rays}

\alphausage{vertices}{vertices[d] returns the list of vertices of the Alpha domain d. 
vertices[\{const,index\}] returns the vertices of d given in Mathematica form.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{vertices}

\alphausage{zpolToPol}{If the domain d is a Z-Domain, zpolToPol[d] returns the rational polyhedral
domain enclosing d, otherwise it returns d as is.}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{zpolToPol}

\alphausage{zpolIsPolQ}{zpolIsPolQ[d] is True if the Z-polyhedron d is actually a polyhedron
(i.e. has identity matrices as lattices).}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{zpolIsPolQ}

\alphausage{DomTrueRays}{DomTrueRays[ dom ] returns the list of true rays of dom, i.e. rays which
are not lines}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomTrueRays}

\alphausage{DomConstraintsOfDom}{DomConstrainsOfDom[ dom ] returns the constraints }{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomConstraintsOfDom}

\alphausage{DomLines}{DomLines[ dom ] returns the lines of dom}{Alpha/Domlib.m}{Alpha`Domlib`}
\index{DomLines}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/FarkasSchedule.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.758550} by makeDoc.
%
% Section header
\section{The Alpha`FarkasSchedule` package } 
\label{label:Alpha`FarkasSchedule`}
\alphausage{FarkasSchedule}{Package.  Schedule for  alpha program with Farkas method,   These function should be used through the schedule[] interface (Package Schedule.m)}{Alpha/FarkasSchedule.m}{Alpha`FarkasSchedule`}
\index{FarkasSchedule}

\alphausage{farkasSchedule}{(+) farkasSchedule[sys\_Alpha`system], finds the schedule of alpha system $<$sys$>$. farkasSchedule[] finds the schedule of \$result and puts the result in \$schedule.  farkasSchedule[sys\_Alpha`system, \{options\}] calls farkasSchedule with non default options. Options[farkasSchedule] provides the options of schedule, and ?option provides information on option.  ?\$schedule provides info on output formats of farkasSchedule).  By default, the   schedule is affine by variable The schedule computation may take a long time (2 minutes for 10 instructions). More information is available in the file \$MMALPHA/doc/user/Scheduler\_user\_manual.ps.}{Alpha/FarkasSchedule.m}{Alpha`FarkasSchedule`}
\index{farkasSchedule}

\alphausage{multiSched}{(+) multiSched[sys\_Alpha`system], finds a multiDimensionnal schedule alpha system $<$sys$>$ using the farkasSchedule scheduling function at each dimension. multiSched[] finds the schedule of \$result and puts the result in \$schedule. multiSched[sys\_Alpha`system, \{options\}] calls schedule with non default options. Options[multiSched] provides the options of schedule, and ?option provides information on option.  See also: applySchedule. }{Alpha/FarkasSchedule.m}{Alpha`FarkasSchedule`}
\index{multiSched}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/INorm.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 19.106120} by makeDoc.
%
% Section header
\section{The Alpha`INorm` package } 
\label{label:Alpha`INorm`}
\alphausage{iNorm}{iNorm[s:\_system, k:\_Integer, options] computes the  imperative form of system s. k is the dimension of time. As usual,  iNorm[k, options] means \$result = iNorm[\$result, k, options]. See  Options[iNorm] for options.}{Alpha/INorm.m}{Alpha`INorm`}
\index{iNorm}

\alphausage{rnf}{rnf[s:\_system] returns the R normal form of s.}{Alpha/INorm.m}{Alpha`INorm`}
\index{rnf}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Lexicographic.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 21.043002} by makeDoc.
%
% Section header
\section{The Alpha`Lexicographic` package } 
\label{label:Alpha`Lexicographic`}
\alphausage{domLexGreater}{domLexGreater[v:\_Matrix] returns a domain D such that for all z in D, z$>$v, where $>$ means lexicographically greater. V is a parametrized vertex. If v is a n*(p+1) matrix, D is a n+p domain, and the p last dimensions are parameter dimensions. Example: domLexGreater[\{\{1,0\},\{0,2\}\}] returns \{a,b,c $|$ c+1$<$=a\} $|$ \{a,b,c $|$ a=c; 3$<$=b\}, where c is the parameter index. domLexGreater[v:\_Vector] is a shortcut for 0-dimensional parameter space. (i.e. domLexGreater[\{0,0,1\}] means domLexGreater[\{\{0\},\{0\},\{1\}\}].}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{domLexGreater}

\alphausage{domLexLower}{domLexLower[v:\_Matrix] returns a domain D such that for all z in D, z$<$v, where $<$ means lexicographically lower. V is a parametrized vertex. If v is a n*(p+1) matrix, D is a n+p domain, and the p last dimensions are parameter dimensions. Example: domLexLower[\{\{1,0\},\{0,2\}\}] returns \{a,b,c $|$ a$<$=c-1\} $|$ \{a,b,c $|$ a=c; b$<$=1\}, where c is the parameter index. domLexLower[v:\_Vector] is a shortcut for 0-dimensional parameter space. (i.e. domLexLower[\{0,0,1\}] means domLexGreater[\{\{0\},\{0\},\{1\}\}].}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{domLexLower}

\alphausage{lexGreaterQ}{lexGreaterQ[a:\_Matrix, b:\_Matrix, p:\_domain] returns True  if a is lexicographicallygreater than b for any value of parameters in p. a and b  are parametrized vectors.}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{lexGreaterQ}

\alphausage{lexGreaterOrEqualQ}{lexGreaterOrEqualQ[a:\_Matrix, b:\_Matrix, p:\_domain] returns True if a is lexicographically greater or equal than b for any value of parameters in p. a and b are parametrized vectors.}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{lexGreaterOrEqualQ}

\alphausage{lexLowerQ}{lexLowerQ[a:\_Matrix, b:\_Matrix, p:\_domain] returns True  if a is lexicographically lower than b for any value of parameters in p. a and b  are parametrized vectors.}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{lexLowerQ}

\alphausage{lexLowerOrEqualQ}{lexLowerOrEqualQ[a:\_Matrix, b:\_Matrix, p:\_domain] returns True if a is lexicographically lower or equal than b for any value of parameters in p. a and b are parametrized vectors.}{Alpha/Lexicographic.m}{Alpha`Lexicographic`}
\index{lexLowerOrEqualQ}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/MakeDoc - copie.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 18.361709} by makeDoc.
%
% Section header
\section{The Alpha`MakeDoc` package } 
\label{label:Alpha`MakeDoc`}
\input{incl-MakeDoc.tex}
\alphausage{MakeDoc}{MakeDoc is a package which contains functions for the automatic creation of a reference Manual. MakeDoc contains functions doDoc and makeDoc.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{MakeDoc}

\alphanote{MakeDoc}{Functions of MakeDoc read all usage and note statements of a package, and for each one, they produce a Latex macro entry alphausage or alphanote. Before creating an entry, substitutions are performed on the text, so that special words (such as \$schedule) and special characters (such as \_) are transformed. The substitution list is in the private variable repRules in the file MakeDoc.m.}

\alphausage{doDoc}{doDoc[ package, texFile ] generates in file texFile the documentation LaTex file associated to a Mathematica file package. doDoc extracts the information contained in the usage and note descriptions placed at the beginning of file package, and produces an output Latex file. doDoc[ package ] automatically writes in a file which has the name  package where the .m suffix is replaced by .tex.  doDoc[ Package, functionList ] generates the documentation only for  the functions enumerated in functionList (as Strings). doDoc has  options fullLatex and callFile which allow a complete latex file,  resp. a calling Latex file to be produced. If there exists a file incl-textFile in the target directory, then a Latex input statement is added in the documentation package.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{doDoc}

\alphausage{fullLatex}{fullLatex is an option of doDoc and makeDoc. Default value is False. If True, the latex file contains the preamble and the \\end\{document\} statement.  Otherwise, it should be used in an input file}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{fullLatex}

\alphausage{targetDir}{option of doDoc. By default, it is "". Gives the name of a directory where the latex file and the calling latex file should be written.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{targetDir}

\alphausage{callFile}{callFile is an option of doDoc and makeDoc. Default valus is False. If True, a latex calling file is created.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{callFile}

\alphausage{sourceDir}{option of makeDoc. Default value is Null. If Null, the source package are those of \$MMALPHA. Their names are in the file \$MMALPHA/sources/ MaleDoc/List\_modules. Otherwise, these names are prefixed by the  value of the sourceDir option.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{sourceDir}

\alphausage{makeDoc}{This function does not work, currently. Use doDoc instead. makeDoc[] updates the Reference Manual of \MMAlpha{} by running  the file genReferenceManual.m in directory\\ \$MMALPHA/doc/ReferenceManual/\\ This produces new versions of the latex files describing the  packages. The Reference Manual can be produced using Latex on  referenceManual.tex. See also doDoc.}{Alpha/MakeDoc.m}{Alpha`MakeDoc`}
\index{makeDoc}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Matrix.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.885021} by makeDoc.
%
% Section header
\section{The Alpha`Matrix` package } 
\label{label:Alpha`Matrix`}
\alphanote{Matrix}{Documentation revised on March 11, 2008}

\alphausage{Matrix}{The Alpha`Matrix` package contains a few functions related to Alpha matrices. These functions are  alphaToMmaMatrix, composeAffines, convHull, convexize, convexizeAll, deleteColumn, deleteRow, determinant, dropColumns, dropParameters,  dropRows, emptyLinearPartQ, getLinearPart, getTranslationVector,  hermite, hermiteL, hermiteR, inverseMatrix, identityQ, idLinearPartQ, idMatrix, inverseInContext, nullSpaceVectors, mmaToAlphaMatrix,  nullLinearPartQ, simplifyAffines, solveDiophantine, smithNormalForm,  squareMatrixQ, suppressRowNum, translationMatrix, translationQ, unimodularQ,  composeAffines, inverseMatrix, unimodularQ, unimodularCompletion,  unimodCompl.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{Matrix}

\alphausage{addLinSpace}{Obsolete function, replaced by the generalized ChangeOfBasis.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{addLinSpace}

\alphausage{alphaToMmaMatrix}{alphaToMmaMatrix[mat] translates a linear function from Alpha format into Mathematica format (list of list). If the linear function mat was affine (non nul constant part) this constant part is lost.  \\ Example:\\ alphaToMmaMatrix[\\ matrix[3, 3, \{i, j\}, \{\{1, 2, 0\}, \{0, 3, 0\}, \{0, 0, 1\}\}]\\ ]\\ gives \{\{1, 2\}, \{0, 3\}\}.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{alphaToMmaMatrix}

\alphausage{canonicalProjection}{Obsolete. canonicalProjection[dom, pos] returns a projected domain obtained  by supressing the indices given by the integer list pos.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{canonicalProjection}

\alphanote{canonicalProjection}{OBSOLETE. Replaced by (generalized) ChangeOfBasis.}

\alphausage{composeAffines}{composeAffines[m1,m2] returns the composition of two affine matrices m1 and m2. These matrices are described in the Alpha format.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{composeAffines}

\alphausage{convHull}{convHull[ldom] returns the convex hull of a (possibly empty) list of  domains ldom.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{convHull}

\alphanote{convHull}{If the input list is empty, returns an empty list.}

\alphausage{convexize}{convexize[domain] returns the convex hull of domain if domain is convex, otherwise returns domain unmodified. Used to convexize a union  of domains (try to use DomConvex instead).}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{convexize}

\alphausage{convexizeAll}{convexizeAll[sys] tries to convexize all the non-convex domains of sys (default \$result) and returns the modified system.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{convexizeAll}

\alphausage{deleteColumn}{deleteColumn[m,k] removes the k-th column of the Alpha matrix m. It removes also index k of m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{deleteColumn}

\alphausage{deleteRow}{deleteRow[m,k] removes the k-th row of the Alpha matrix m. It does not touch the index list of m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{deleteRow}

\alphausage{determinant}{determinant[m] gives the determinant of a square Alpha matrix. The result can be integer or rationnal, the constant part is not taken into account.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{determinant}

\alphausage{dropColumns}{dropColumns[m,n] removes n columns of the Alpha matrix m, from front if n is positive, from end if n is negative. This function does not touch the  dimensions nor the indexes of the matrix.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{dropColumns}

\alphausage{dropParameters}{dropParameters[sys,m] removes the rows of Alpha matrix m corresponding to  parameters of system sys (default \$result). If there are too many  parameters, dropParameters returns m unchanged.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{dropParameters}

\alphausage{dropRows}{dropRows[m,n] removes n rows of the Alpha matrix m, from front if n is  positive, from the end otherwise. This function does not touch the  dimensions nor the indexes of the matrix.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{dropRows}

\alphausage{emptyLinearPartQ}{emptyLinearPartQ[m] is True if the linear part of matrix m is empty,  False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{emptyLinearPartQ}

\alphausage{getLinearPart}{getLinearPart[m] returns the linear part of an affine function m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{getLinearPart}

\alphausage{getTranslationVector}{getTranslationVector[m] returns the translation vector of a square  affine function m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{getTranslationVector}

\alphausage{hermite}{hermite[m] returns the left hermite decomposition \{h,u\} (as Alpha matrices) of m1 (such that m1=composeAffines[h,u]).\\ WARNING: Currently, this function only deals with linear function (the constant part is ignored).}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{hermite}

\alphausage{hermiteL}{hermiteL[m] performs the left Hermite decomposition of the Mathematica matrix m and returns mathematica matrices  \{H,Q\} where Q unimodular, H upper triangular and m = Dot[H,Q].}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{hermiteL}

\alphausage{hermiteR}{hermiteR[m] computes the right Hermite decomposition of the  Mathematica matrix m and returns Mathematica matrices \{H,Q\}, where Q is unimodular, H is lower triangular, and m=Dot[Q,H].}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{hermiteR}

\alphausage{inverseMatrix}{inverseMatrix[m] computes and returns the inverse of a square affine  matrix m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{inverseMatrix}

\alphausage{identityQ}{identityQ[m] is True if an Alpha affine function m is the identity,  False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{identityQ}

\alphausage{idLinearPartQ}{idLinearPartQ[m] is True if the linear part of the affine function given as Alpha matrix m is the identity, False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{idLinearPartQ}

\alphausage{idMatrix}{idMatrix[idx\_List,idy\_List] returns the transformation matrix for dependence (idx$ \rightarrow $idy), where idx and idy are lists of index names. It is assumed that names in idy are also in idx.  \\Example:\\ idMatrix[\{"i","j"\},\{"i","i","j"\}] = \\ matrix[4, 3, \{"i","j"\}, \{\{1,0, 0\}, \{1, 0, 0\}, \{0, 1, 0\}, \{0, 0, 1\}\}].}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{idMatrix}

\alphausage{inverseInContext}{inverseincontext[m,d] computes the inverse matrix of Alpha matrix m in  context with domain d, i.e., taking into account the lineality space  defined by d.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{inverseInContext}

\alphausage{isIdLinearPart}{obsolete form of idLinearPartQ.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{isIdLinearPart}

\alphausage{isNullLinearPart}{obsolete form of nullLinearPartQ.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{isNullLinearPart}

\alphausage{nullSpaceVectors}{nullSpaceVectors[mat] gives the list of vectors of the null space of Alpha matrix mat.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{nullSpaceVectors}

\alphausage{mmaToAlphaMatrix}{mmaToAlphaMatrix[m] returns the Alpha form of a Mathematica matrix m. mmaToAlphaMatrix[m,i] returns the Alpha form of matrix m with index (list of strings) i. mmaToAlphaMatrix[m,c:List[\_\_\_]] returns the Alpha form of a linear function Z $ \rightarrow $ mZ+c.  \\Example: \\mmaToAlphaMatrix[\{\{1, 2\},\{0, 3\}\},\{1,4\}]\\gives the Alpha matrix  which represents the affine function (i,j$ \rightarrow $i+2j+1,3j+4).   mmaToAlphaMatrix[m,c,i] returns the Alpha form of linear function Z $ \rightarrow $ mZ+c, with index (string list) i.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{mmaToAlphaMatrix}

\alphausage{nullLinearPartQ}{nullLinearPartQ[m] is True if the linear part of the affine function given as Alpha matrix m is null, False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{nullLinearPartQ}

\alphausage{solveDiophantine}{solveDiophantine[a,b] solves  the linear diophantine system  of equations aX = b where a is a MMA matrix, and b a MMA vector. The  solution has the form \{x1,n,M\} where x1 is a particular solution  of aX=b, n is the number of columns of M, and M is a matrix such  that X=Mt (t is a n-vector) is the general solution of aX=0. The  general solution of aX=b is x1+Mt. If the system has no integral solution, then x1 is \{\}. solveDiophantine[a] solves  the linear  diophantine system of equation  ax=0 where a is a MMA matrix.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{solveDiophantine}

\alphausage{smithNormalForm}{smithNormalForm[mat computes the Smith Normal Form of an Alpha matrix mat and returns \{u,s,v\} (Alpha matrices), where u,v are unimodular and s is diagonal such that mat = u.s.v.  smithNormalForm[m] computes the Smith Normal Form of a Mathematica matrix m.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{smithNormalForm}

\alphausage{squareMatrixQ}{squareMatrixQ[m] is True if m is a square Alpha matrix.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{squareMatrixQ}

\alphausage{subMatrices}{subMatrices[m1,m2] subtracts two matrices m1 and m2. This function is for internal use by the Pipeline function.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{subMatrices}

\alphausage{suppressRowNum}{suppressRowNum[mat,i] suppresses row i in (\MMAlpha{}) matrix mat.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{suppressRowNum}

\alphausage{translationMatrix}{translationMatrix[ind, vec] returns an Alpha translation matrix  corresponding to the function: (ind $ \rightarrow $ ind+vec), where ind is a list of indices and vec an integral vector.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{translationMatrix}

\alphausage{translationQ}{translationQ[m] is True if the full rank affine function m is a translation, False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{translationQ}

\alphausage{unimodularQ}{unimodularQ[m] is True if an \MMAlpha{} affine matrix m is square and unimodular, False otherwise.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{unimodularQ}

\alphausage{unimodularCompletion}{unimodularCompletion[vl] completes a vector list vl into a unimodular matrix which is returned. unimodularCompletion[mat] takes the vector  expressed as an Alpha matrix mat and completes it into a unimodular  Alpha matrix which is returned. If the original vector is of size n,  the resulting matrix is of size (n+1)*(n+1) and the first row of the  resulting matrix is the original vector with 0 appended as constant term. For example, \\unimodularCompletion[\{1,1,1\}] returns \\ matrix[4, 4, \{\}, \{\{1, 1, 1, 0\}, \{0, 0, 0, 1\}, \{0, 0, 1, 0\}, \{0, 1, 0, 0\}\}]\\ and  \\unimodularCompletion[matrix[4, 4, \{\}, \{\{1, 1, 1, 0\},  \{0, 0, 0, 1\}\}]  \\ returns \\ matrix[4, 4, \{\}, \{\{1, 1, 1, 0\}, \{0, 0, 0, 1\}, \{0, 0, 1, 0\},     \\\{0, 1, 0, 0\}\}].}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{unimodularCompletion}

\alphausage{unimodCompl}{unimodCompl[mat] returns an unimodular completion of matrix mat, or \$Failed if an error occurs. The result is an Alpha matrix. The indices of the result are those of mat, with some new arbitrary indices if necessary.  Exemple : completion of (i,j$ \rightarrow $2i) returns (i,j,k$ \rightarrow $2i+k,j,i). The optional parameter indicates the dimension of the parameter space (0 if none). \\unimodCompl[mat] is the mathematica version (no constant row and column). It returns a square unimodular \MMAlpha{} matrix. First lines of this matrix are the matrix mat (modulo an extension on the right).   \\ For exemple: \{\{2,0,0\},\{1,2,0\}\} will be completed as: \\ \{\{2,0,0,1,0\}, \{1,2,0,0,1\}, \{0,0,1,0,0\}, \{1,0,0,0,0\}, \{0,1,0,0,0\}\}. \\ unimodCompl[mat,nbPar] returns a MMA matrix or \$Failed if an error  occurs. mat is a (k*(n+p+1) MMA matrix, it represents a  k-dimensional affine function for a n-dim variable, with a p-dim parameter space. Thus, the last column is the constant column. nbPar is the parameter space dimension, i.e. p. The function computes n from p and mat. The result is a MMA matrix, which has at least (n+p+1) rows and exactly (n+p+1) columns. The first k rows are exactly mat. If the result isn't square, it means that some dimension will be added by the returned change of basis.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{unimodCompl}

\alphausage{simplifyAffines}{simplifyAffines[] simplifies all affine functions in system \$result.  \\simplifyAffines[sys] simplifies all affine functions on system sys.}{Alpha/Matrix.m}{Alpha`Matrix`}
\index{simplifyAffines}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Normalization.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 18.455255} by makeDoc.
%
% Section header
\section{The Alpha`Normalization` package } 
\label{label:Alpha`Normalization`}
\alphanote{Normalization}{Documentation revised on August 1, 2004}

\alphausage{Normalization}{Alpha`Normalization` is the package containing the definitions  of normalization rules along with basic  normalizing functions:  checkRestrictions, minRestrictInCtxt, normalize, normalize0,  normalizeDef, normalizeDef0, normalizeInCtxt, normalizeInCtxt0,  normalizeQ, normalize0Q, and simplifyInContext.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{Normalization}

\alphausage{checkRestrictions}{checkRestrictions[ast] checks for redundant restrictions in a normalized  system. Default ast is Alpha`\$result.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{checkRestrictions}

\alphanote{checkRestrictions}{This function is useful after normalizing a complete system, in order to remove redundant restrictions. It cannot be included in the set normalization rules because it needs contextual information wrt. the normalized expression (it requires the declaration of a variable).}

\alphausage{minRestrictInCtxt}{minRestrictInCtxt[ast,domain] simplifies all restrictions of a  program in the context of a specific domain. Returns the simplified  expression.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{minRestrictInCtxt}

\alphausage{normalize}{normalize[option] normalizes program \$result with the given option. normalize[sys,option] normalizes ALPHA expression sys with the given option. The available option is indexnorm which should be set to True if normalization of index expressions is desired. The default option is False.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalize}

\alphanote{normalize}{normalize[exp] computes the fixpoint of a set of normalization rules when applied to exp. The set of rules is contained in variable normalizationRules.}

\alphausage{normalize0}{normalize0[ast] normalizes an ALPHA expression ast wrt. Alpha0 rules  (allowing nested cases.) Default ast is Alpha`\$result.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalize0}

\alphanote{normalize0}{normalize[exp] computes the fixpoint of a set of normalization rules when applied to exp. The set of rules is contained in variable normalizationRules0.}

\alphausage{normalizeDef}{normalizeDef[symbol] normalizes the definition of ALPHA variable symbol in \$result. normalizeDef[sys,symbol] normalizes the definition of ALPHA variable symbol in program sys. Variable symbol is a string.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizeDef}

\alphanote{normalizeDef}{normalizeDef[exp, var] computes the fixpoint of a set of normalization rules when applied to the definition of var in exp. The set of rules is contained in variable normalizationRules.}

\alphausage{normalizeDef0}{normalizeDef0[symbol] normalizes the definition of ALPHA variable symbol in \$result wrt. the rules of Alpha0 (allowing nested cases). normalizeDef0[sys,symbol]normalizes the definition of ALPHA variable symbol in program sys. Variable symbol is a string.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizeDef0}

\alphanote{normalizeDef0}{normalizeDef[exp, var] computes the fixpoint of a set of normalization rules when applied to the definition of var in exp. The set of rules is contained in variable normalizationRules0.}

\alphausage{normalizationRules}{Set (list) of rewrite rules specifying the normalization of an ALPHA expression towards the CRD (Case-Restriction-Dependence)  normal form.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizationRules}

\alphanote{normalizationRules}{normalizationRules can be extended by appending/prepending new rules to the initial list. The rules are named and have the form  name = exp1 :$>$ exp2.}

\alphausage{normalizationRules0}{Set (list) of rewrite rules specifying the normalization of an ALPHA expression towards the Alpha0 (NestedCase-Restriction-Dependence) normal form.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizationRules0}

\alphanote{normalizationRules0}{normalizationRules0 can be extended by modifying the value of the rule list. The rules are named and have the form name = exp1 :$>$ exp2. }

\alphausage{normalizeInCtxt}{normalizeInCtxt[ast,domain] returns a normalized and reduced form of the restriction of ast to domain.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizeInCtxt}

\alphanote{normalizeInCtxt}{'normalizeInCtxt' should perhaps operate in a more intelligent way.}

\alphausage{normalizeInCtxt0}{Function. Normalizes an expression in the context of a specific domain.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizeInCtxt0}

\alphanote{normalizeInCtxt0}{normalizeInCtxt0 should perhaps operate in a more intelligent way.}

\alphausage{normalizeQ}{normalizeQ[ast] returns True if ast is normalized, False otherwise. Default value for ast is Alpha`\$result.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalizeQ}

\alphausage{normalize0Q}{normalize0Q[ast] returns True if ast is normalized, False otherwise. Default value for ast is Alpha`\$result.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{normalize0Q}

\alphausage{correctMat}{ in test }{Alpha/Normalization.m}{Alpha`Normalization`}
\index{correctMat}

\alphausage{correctAffineFunctions}{in test 2}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{correctAffineFunctions}

\alphausage{simplifyInContext}{simplifyInContext[] simplifies and reduces the Alpha system  \$result in the following ways: simplifies restriction domains  in context of declaration, simplifies transformations in context  of declaration. It operates on normalized or unnormalized programs. simplifyInContext[sys] simplifies and reduces the Alpha system  sys and returns the simplified system. Does not change \$result. simplifyInContext[sys, exp] simplifies expression exp in the  context of sys and returns the simplified expression.}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{simplifyInContext}

\alphausage{simplifySystem}{simplifySystem[] is just a shorthand for simplifyInContext[]; convexizeAll[];normalize[] (or normalize0 depending on the option).  simplifySystem[sys] return system sys modified without  modifying \$result}{Alpha/Normalization.m}{Alpha`Normalization`}
\index{simplifySystem}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Options.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 21.060492} by makeDoc.
%
% Section header
\section{The Alpha`Options` package } 
\label{label:Alpha`Options`}
\alphausage{debug}{option (Boolean). If True, debug mode. }{Alpha/Options.m}{Alpha`Options`}
\index{debug}

\alphausage{verbose}{option (Boolean). If True, print intermediate information.}{Alpha/Options.m}{Alpha`Options`}
\index{verbose}

\alphausage{recurse}{option (Boolean). If True, apply the function recursively to all the subsystem.}{Alpha/Options.m}{Alpha`Options`}
\index{recurse}

\alphausage{library}{option (List of Alpha`system). Defines the library which must be used by the function.}{Alpha/Options.m}{Alpha`Options`}
\index{library}

\alphausage{contextDomain}{option of addLocal. If True (default), the new variable is added with the contextual domain of the expression, with the form of addLocal with a position. Otherwise, the expression domain is  used.}{Alpha/Options.m}{Alpha`Options`}
\index{contextDomain}

\alphausage{invert}{option of serializeReduce. If True (default False),  the nullspace vector is inverted}{Alpha/Options.m}{Alpha`Options`}
\index{invert}

\alphausage{norm}{option of removeIdEqus. If True, normalization is done}{Alpha/Options.m}{Alpha`Options`}
\index{norm}

\alphausage{inputEquations}{option of removeIdEqus. If False, an equation of the forme X = in, where in is an input variable, is not removed.}{Alpha/Options.m}{Alpha`Options`}
\index{inputEquations}

\alphausage{allLibrary}{option of removeIdEqus (default False). If set to  True, all programs of library are treated.}{Alpha/Options.m}{Alpha`Options`}
\index{allLibrary}

\alphausage{resolutionSoft}{(+) resolutionSoft is an option of schedule. If resolutionSoft $ \rightarrow $ pip, the schedule is computed using Pip, through a file interface (implemented in Farkas resolution only).  None of the following are yet implemented: If resolutionSoft $ \rightarrow $ mma, the schedule is computed using the Mathematica function ConstrainedMin (implemented in Vertex resolution only).  resolutionSoft $ \rightarrow $ lpSolve the schedule is computed using the LpSolve library function ConstrainedMin (not implemented yet).}{Alpha/Options.m}{Alpha`Options`}
\index{resolutionSoft}

\alphausage{pip}{possible value (default) of the options resolutionSoft of schedule}{Alpha/Options.m}{Alpha`Options`}
\index{pip}

\alphausage{mma}{possible value  of the options resolutionSoft of schedule}{Alpha/Options.m}{Alpha`Options`}
\index{mma}

\alphausage{lpSolve}{possible value of the options resolutionSoft of schedule}{Alpha/Options.m}{Alpha`Options`}
\index{lpSolve}

\alphausage{schedMethod}{ Options of schedule (symbol), select the scheduling method to be used. schedMethod$ \rightarrow $farkas (default) stands for P. Feautrier's method (using PIP with file interface) implemented in the scheduleFarkas function. schedMethod$ \rightarrow $farkas2 stands for the most recent  implementation of this method (ModularSchedule) .schedMethod$ \rightarrow $Vertex stands for P. Quinton's vertex method (using MMA linear resolveur). WARNING: the setting of this options greatly influence the setting of the others options of schedule because they correspond to two completely different implementations}{Alpha/Options.m}{Alpha`Options`}
\index{schedMethod}

\alphausage{farkas}{value of options schedMethod of schedule, the schedule  will be computed with P. Feautrier's method (using PIP with file interface)}{Alpha/Options.m}{Alpha`Options`}
\index{farkas}

\alphausage{vertex}{value of options schedMethod of schedule, the schedule  will be computed with P. Quinton's vertex method (using MMA linear resolveur)}{Alpha/Options.m}{Alpha`Options`}
\index{vertex}

\alphausage{integerSolution}{(+) integerSolution is an option of schedule.  If integerSolution $ \rightarrow $ True (default for farkas resolution)  the schedule has integral coefficients. If integerSolution $ \rightarrow $ False, (default for Vertex resolution) the schedule may have rational coefficients.}{Alpha/Options.m}{Alpha`Options`}
\index{integerSolution}

\alphausage{bigParPos}{ bigParPos is an option of Pip related functions, Integer type. It indicates the position of the big Paramter. If set to negative integer, there is no big parameter}{Alpha/Options.m}{Alpha`Options`}
\index{bigParPos}

\alphausage{addConstraints}{option of schedule. The type is : List of  String or List of List of String (in case of MultiDim schedule),  default value : \{\}. Additional constraints for the LP.  add  constrains expressed in the strings to the current scheduling  process.  schedule[addConstraints$ \rightarrow $\{"TA[i,j,N]=i+2j-2","TBD2==2","TBD1+2TBD3$>$=1"\}] Impose that the schedule function of variable A is i+2j-2, Then we  have constraints on the coefficients of the timing function of the  B variable (the  second coefficient  must be 2, etc...). In case of a multi-dimensionnal scheduling, the  ith list is added to the constraints of the ith dimension}{Alpha/Options.m}{Alpha`Options`}
\index{addConstraints}

\alphausage{durations}{option of schedule (List) default \{\}.  select a mode for counting the duration of instructions.  durations $ \rightarrow $ \{\} : each equation is 1 (whatever complex is the computation) (default). durations $ \rightarrow $ ListOfInteger: The duration of each equation is given by the user the options "durationByEq" indicates whether these values stand for a duration by equation (default in Farkas resolution, not implemented in Vertex resolution) or a duration by dependence (default in Vertex resolution, not implemented in Farkas resolution). durations $ \rightarrow $ ListOfListOfInteger: same as previous one but for multidimensionnal scheduling (one list of integer by dimension).  }{Alpha/Options.m}{Alpha`Options`}
\index{durations}

\alphausage{durationByEq}{ option of schedule (boolean) indicates whether the duration values given in the "durations" option stand  for a duration by equation (default in Farkas resolution, not implemented in Vertex resolution) or a duration by dependence (default in Vertex resolution, not implemented in Farkas resolution). In the case of a duration by equation, the list must contain as many integer as there are VARIABLES (do not forget the inputs and output), the order is the order of declaration in the Alpha program. In the case of a duration by dependence, the   must contain as many integer as there are dependecies   the order is the order of the dependencies returned by the dep[] function }{Alpha/Options.m}{Alpha`Options`}
\index{durationByEq}

\alphausage{givenSchedVect}{options of Schedule, List of schedule given   for some variable, the syntaxe for each schedule is the one present   in \$schedule: \{\{a, \{i, j, N\}, sched[\{0, 0, 0\}, 0]\}\} (The list of   indice may be replaced by \{\}) }{Alpha/Options.m}{Alpha`Options`}
\index{givenSchedVect}

\alphausage{affineByVar}{Value of option scheduleType of schedule, the   resulting schedule will be affine by variable}{Alpha/Options.m}{Alpha`Options`}
\index{affineByVar}

\alphausage{sameLinearPart}{Value of option scheduleType of schedule, the   resulting schedule will be affine by variable with   the same linear part for all local variables}{Alpha/Options.m}{Alpha`Options`}
\index{sameLinearPart}

\alphausage{sameLinearPartExceptParam}{Value of option scheduleType of schedule, the   resulting schedule will be affine by variable with   the same linear part for all local variables}{Alpha/Options.m}{Alpha`Options`}
\index{sameLinearPartExceptParam}

\alphausage{scheduleType}{scheduleType is an option of schedule (symbol) which gives the type of the schedule searched.  scheduleType $ \rightarrow $ affineByVar (default) : affine by variable schedule.  scheduleType $ \rightarrow $ sameLinearPart: affine with constant linear part.  scheduleType $ \rightarrow $ sameLinearPartExceptParam: affine with constant linear part except for the parameters}{Alpha/Options.m}{Alpha`Options`}
\index{scheduleType}

\alphausage{optimizationType}{ options of schedule (symbol), indicates the objective fonction used by the schedule: minimizing time, delays on dependencies of nothing.  optimizationType$ \rightarrow $time tries to get aminimal time schedule (default).  optimizationType$ \rightarrow $delay tries to minimize the delays on the dependencies (not implemented currently) optimizationType$ \rightarrow $Null no objective function is provided (not implemented in Vertex resolution). In that case the coefficient of the timing function are minimized lexicographically (from output to input).}{Alpha/Options.m}{Alpha`Options`}
\index{optimizationType}

\alphausage{time}{possible value (default) of option optimizationType of schedule}{Alpha/Options.m}{Alpha`Options`}
\index{time}

\alphausage{delay}{possible value of option optimizationType of schedule}{Alpha/Options.m}{Alpha`Options`}
\index{delay}

\alphausage{multi}{possible value  of option optimizationType of schedule or of option structSchedType of buildSchedConstraints}{Alpha/Options.m}{Alpha`Options`}
\index{multi}

\alphausage{mono}{possible value  of option structSchedType of buildSchedConstraint}{Alpha/Options.m}{Alpha`Options`}
\index{mono}

\alphausage{scheduleDim}{Option of buildSched Constraint (integer or list of integer), indicates the dimension to which the constraints should be set}{Alpha/Options.m}{Alpha`Options`}
\index{scheduleDim}

\alphausage{structSchedType}{Option of buildSched Constraint (integer or list of integer), indicates whether the use will be pipelined (no additionnal dimension in the schedule) or considered as an instantaneous call (additionnal dimention in the schedule)}{Alpha/Options.m}{Alpha`Options`}
\index{structSchedType}

\alphausage{multiSchedDepth}{options of schedule (integer) indicating the depth of the  current schedule taking place in a multi dimensionnal scheduling process (this option should not be set by the user, it is for internal use only)}{Alpha/Options.m}{Alpha`Options`}
\index{multiSchedDepth}

\alphausage{onlyVar}{ Option of schedule which is a list of string (default value: all)  indicating the only variables that we wish to schedule. Warning, if  some variable are needed for the computation of the one indicated,  the function will try to find their execution dates in \$schedule  (only implemented in the Farkas method)}{Alpha/Options.m}{Alpha`Options`}
\index{onlyVar}

\alphausage{all}{value of various options (onlyVar, onlyDep,....)}{Alpha/Options.m}{Alpha`Options`}
\index{all}

\alphausage{onlyDep}{ Option of schedule which is a list of integer (default value: all) indicating the only dependences  that we wish to schedule (used for  multiSched[]) }{Alpha/Options.m}{Alpha`Options`}
\index{onlyDep}

\alphausage{objFunction}{objFunction is an option of schedule which gives the type of minimization done for minimizing what we try to minimize (i.e. time or delay or whatever). the technique for that is to build a function of the parameters of the system and to minimize the   coefficient of this function with respect to constraints which   ensure that this function is greater than what we want to minimize.  The technique used by default (objFunction$ \rightarrow $lexicographic)  in the Farkas resolution is to minimize   lexicographically the coefficient of this function in the order of   their declaration in the Alpha program. but one can chose to   minimize them in another order:   objFunction$ \rightarrow $lexicographic["N","P","M"] (not implemented anywhere) or to minimize a function of these coefficients example:   objFunction$ \rightarrow $2"N"+"P" (only implemented in the vertex        resolution) }{Alpha/Options.m}{Alpha`Options`}
\index{objFunction}

\alphausage{lexicographic}{possible value of the objFunction option of the   schedule function}{Alpha/Options.m}{Alpha`Options`}
\index{lexicographic}

\alphausage{checkSched}{(+) checkSched is an option of schedule which allows a schedule  to be checked for a given Alpha program. This option can be used only for affine by variable schedules. The form of this  option is checkSched $ \rightarrow $ list, where list has the same syntax as  \$schedule (see ?\$schedule for more information).\\ WARNING: this option is not implemented}{Alpha/Options.m}{Alpha`Options`}
\index{checkSched}

\alphausage{parameterConstraints}{parameterConstraints is an option of dep. If set, constraints on the parameters are included in the domains of the dependencies. Default is False}{Alpha/Options.m}{Alpha`Options`}
\index{parameterConstraints}

\alphausage{subSystems}{options of schedule (boolean) and dep indicating whether or not the function takes into account the calls to other   systems (default, false)}{Alpha/Options.m}{Alpha`Options`}
\index{subSystems}

\alphausage{subSystemSchedule}{option of schedule (list), indicates,  if the subSystems option is set to True, the schedules of the   different subsystems called in the system to schedule for performing   a structured scheduling}{Alpha/Options.m}{Alpha`Options`}
\index{subSystemSchedule}

\alphausage{multiDimensional}{ option of Schedule (boolean), indicates  if we perform a multi dimensional scheduling or mono dimensionnal   scheduming (default)}{Alpha/Options.m}{Alpha`Options`}
\index{multiDimensional}

\alphausage{outputForm}{options of schedule, gselect the output of the schedule which may be a schedule, a LP, a domain .... The default value is outputForm$ \rightarrow $scheduleResult i.e. the result isx a schedule (see ?\$schedule). other value: outputForm$ \rightarrow $lpSolve (Linear Programming problem (LP) formated for lp\_solve), outputForm$ \rightarrow $lpMMA (LP formated for MMA (to be implemented)), outputForm$ \rightarrow $domain (schedule polytope (Alpha`domain,Warning works for small programs))}{Alpha/Options.m}{Alpha`Options`}
\index{outputForm}

\alphausage{dataFlowConstantsNull}{dataFlowConstantsNull is an option of scd. Its default value is True, forcing the constant part of the first level of a data-flow schedule to be 0 for all variables.}{Alpha/Options.m}{Alpha`Options`}
\index{dataFlowConstantsNull}

\alphausage{dataFlowPeriod}{dataFlowPeriod is an option of scd. Its default value is 1. If set to an  integer, forces the period of the data-flow schedule to be this integer. If set to any non integral value, the data-flow period is choosen automatically,  and assumed to be $>$=1. Most often used in combination with dataFlowConstantsNull set to False.}{Alpha/Options.m}{Alpha`Options`}
\index{dataFlowPeriod}

\alphausage{dataFlowVariables}{dataFlowVariables is an option of scd, which is considered only for data-flow schedules. Its default value is $<$$<$all$>$$>$, meaning that all variables are considered for data-flow scheduling. If it is set to a list of strings, it gives the list of variables which must not be inserted in the dataflow variables. These variables will be considered only for scheduling in the next dimensions of the schedule.}{Alpha/Options.m}{Alpha`Options`}
\index{dataFlowVariables}

\alphausage{verticesPositives}{verticesPositives is an option of scd, scheduleConstraints, and  timeMinSchedConstraints. Its default value is False. When set, this option forces the scheduler to find timing-functions which are positive at the vertices of the domains of the variables. Usually, only constraints on rays of the domains are taken into account. This option is not used, as it most often results in non-integral (i.e. strictly rational) schedules.}{Alpha/Options.m}{Alpha`Options`}
\index{verticesPositives}

\alphausage{sortOrder}{sortOrder is an option of showSchedResult, which specifies the order in which the results are given. Default is noOrder. Other possibilities are lexicographic, or an integer which specifies the number of coefficient given backwards}{Alpha/Options.m}{Alpha`Options`}
\index{sortOrder}

\alphausage{noOrder}{noOrder is a value for the option sortOrder of showSchedResult}{Alpha/Options.m}{Alpha`Options`}
\index{noOrder}

\alphausage{eliminatesDoubles}{Options of dep[], if set to False  dep[] returns the usual dependence list. If set to True, the dependences   returned does not contain twice the same dependence}{Alpha/Options.m}{Alpha`Options`}
\index{eliminatesDoubles}

\alphausage{equalitySimpl}{ options of dep[], if set to True, try to   simplify the dep according to the context (default True)}{Alpha/Options.m}{Alpha`Options`}
\index{equalitySimpl}

\alphausage{scalarTypeCheck}{option to analyze[] (Boolean). If True, perform scalar type checking. For internal use mostly.}{Alpha/Options.m}{Alpha`Options`}
\index{scalarTypeCheck}

\alphausage{occurence}{option of inlineAll[] and inlineSubsystem[] (Integer).  When a given subsystem is used several times in the same caller system, the value of occurence selects the instance to be inlined.}{Alpha/Options.m}{Alpha`Options`}
\index{occurence}

\alphausage{rename}{option of inlineAll[] and inlineSubsystem[] (Boolean).  rename $ \rightarrow $ True  forces the renaming of all variables, whereas if rename $ \rightarrow $ False, variables are renamed only in case of conflict.}{Alpha/Options.m}{Alpha`Options`}
\index{rename}

\alphausage{renameCounter}{option of inlineAll[] and inlineSubsystem[] (Integer, default 1).  Sets the value to be appended to variable names in case when they are renamed to avoid name conflicts. }{Alpha/Options.m}{Alpha`Options`}
\index{renameCounter}

\alphausage{current}{option of inlineAll[] and inlineSubsystem[] (Boolean). If True, \$result and \$program are updated.}{Alpha/Options.m}{Alpha`Options`}
\index{current}

\alphausage{underscore}{option of inlineSubsystem and inlineAll (Boolean). When True (default), new identifier separator is \_, whereas it is X otherwise}{Alpha/Options.m}{Alpha`Options`}
\index{underscore}

\alphausage{indexnorm}{option of normalize[] (Boolean). indexnorm $ \rightarrow $ True   normalizes index expressions also. The default is False.}{Alpha/Options.m}{Alpha`Options`}
\index{indexnorm}

\alphausage{alphaFormat}{options of simplifySystem: alphaFormat$ \rightarrow $Alpha (default) simplify a system to the standard normalized Alpha form. alphaFormat$ \rightarrow $Alpha0 simplify to Alpha0 format, alphaFormat$ \rightarrow $AlpHard is not implemented}{Alpha/Options.m}{Alpha`Options`}
\index{alphaFormat}

\alphausage{Alpha0}{value of the alphaFormat options of simplifySystem}{Alpha/Options.m}{Alpha`Options`}
\index{Alpha0}

\alphausage{initZeroReg}{options of toAlpha0v2: initZeroReg$ \rightarrow $False  do not generate a control signal for the register that are initialized with a zero. These register will be initialized by the Rst signal}{Alpha/Options.m}{Alpha`Options`}
\index{initZeroReg}

\alphausage{verifyCone}{verifyCone is an option of uniformization which verifies whether the dependence cone is pointed if set to True (default value) or does not if it is False.}{Alpha/Options.m}{Alpha`Options`}
\index{verifyCone}

\alphausage{alignInp}{alignInp is an option of uniformization. If set to True the uniformize function will try to uniformize a dependence by aligning the input instead of routing the dependence. The default value is False.}{Alpha/Options.m}{Alpha`Options`}
\index{alignInp}

\alphausage{routeOnce}{routeOnce is an option of uniformization. If set to True the uniformize function  will perform routing for the first vector of the route. The default value is False.}{Alpha/Options.m}{Alpha`Options`}
\index{routeOnce}

\alphausage{tmpFile}{tmpFile is an option of uniformization. If set to True the uniformize function  will output a file in /tmp directory after each iteration of uniformization. The default value is False.}{Alpha/Options.m}{Alpha`Options`}
\index{tmpFile}

\alphausage{silent}{option of show (False). If silent is True, show does not print its result, but returns a string.}{Alpha/Options.m}{Alpha`Options`}
\index{silent}

\alphausage{allVariablesAllowed}{Option of changeOfBasis, default False. If True  the change of basis can be applied to any variable. Not yet operational...}{Alpha/Options.m}{Alpha`Options`}
\index{allVariablesAllowed}

\alphausage{showSquareDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showSquareDeps}

\alphausage{showNonSquareDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showNonSquareDeps}

\alphausage{showUniformDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showUniformDeps}

\alphausage{showNonUniformDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showNonUniformDeps}

\alphausage{showUniformizableDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showUniformizableDeps}

\alphausage{showNonUniformizableDeps}{option of uniformizeInfo}{Alpha/Options.m}{Alpha`Options`}
\index{showNonUniformizableDeps}

\alphausage{minimize}{option of simplex. Default is True}{Alpha/Options.m}{Alpha`Options`}
\index{minimize}

\alphausage{boundedDelay}{option of timeMinSchedConstraints. Currently not used.}{Alpha/Options.m}{Alpha`Options`}
\index{boundedDelay}

\alphausage{extraEdges}{Option of depGraph, allowing edges to be added}{Alpha/Options.m}{Alpha`Options`}
\index{extraEdges}

\alphausage{labelOffset}{Option of depGraph, allowing the offset of the position of a label  to be modified. Default is 0.025.}{Alpha/Options.m}{Alpha`Options`}
\index{labelOffset}

\alphausage{labelSize}{Option of depGraph, allowing the size of the labels to be changed. Default value is 0.1 .}{Alpha/Options.m}{Alpha`Options`}
\index{labelSize}

\alphausage{onlyIdDep}{Option of depGraph (default False), If set to True, build a graph where only  identity dependences are present. (require an aligned program).}{Alpha/Options.m}{Alpha`Options`}
\index{onlyIdDep}

\alphausage{cellType}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{cellType}

\alphausage{tempFile}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{tempFile}

\alphausage{vhdlLibrary}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{vhdlLibrary}

\alphausage{compass}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{compass}

\alphausage{compactCode}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{compactCode}

\alphausage{clockEnable}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{clockEnable}

\alphausage{skipLines}{Option of a2v}{Alpha/Options.m}{Alpha`Options`}
\index{skipLines}

\alphausage{stdLogic}{Option of a2v. If set, produces stdlogic types. Default value is True.}{Alpha/Options.m}{Alpha`Options`}
\index{stdLogic}

\alphausage{initialize}{Option of vhdlType. If set, produces initialized declarations}{Alpha/Options.m}{Alpha`Options`}
\index{initialize}

\alphausage{controler}{Option value for option cellType of alphaToVHDL}{Alpha/Options.m}{Alpha`Options`}
\index{controler}

\alphausage{cell}{Option value for option cellType of alphaToVHDL}{Alpha/Options.m}{Alpha`Options`}
\index{cell}

\alphausage{module}{Option value for option cellType of alphaToVHDL}{Alpha/Options.m}{Alpha`Options`}
\index{module}

\alphausage{rewrite}{rewite: option (boolean) for cGen. If True, output file is overwritten if it already exists.}{Alpha/Options.m}{Alpha`Options`}
\index{rewrite}

\alphausage{lyrtech}{Option of a2v. If True, generates strange code for the Lyrtech environment. For example, rst is not a std\_logic but a vector of std\_logic of size 1.}{Alpha/Options.m}{Alpha`Options`}
\index{lyrtech}

\alphausage{noPrint}{boolean option of cGen. If True, the generated C code does not provide  prompts before the inputs, nor before the outputs, which makes it easier to run a test from an input file}{Alpha/Options.m}{Alpha`Options`}
\index{noPrint}

\alphausage{alreadySchedule}{option of cGen (boolean, default false), if set to True, cGen does not  perform schedule neither change of Basis: WARNING not Implemented yet }{Alpha/Options.m}{Alpha`Options`}
\index{alreadySchedule}

\alphausage{stimuli}{option of cGen (boolean, default False), if set to True, cGen  generates one stimuli file by input and output variable}{Alpha/Options.m}{Alpha`Options`}
\index{stimuli}

\alphausage{interactive}{interactive is a (boolean) option of cGen. If True, cGen also generates  a function main which (i) reads system inputs on stdin (ii) evaluates the  system, (iii) prints the outputs on stdout}{Alpha/Options.m}{Alpha`Options`}
\index{interactive}

\alphausage{matlab}{matlab is a boolean option of cGen. If True, cGen also generates a  function mexFunction which is used for interface with matlab code}{Alpha/Options.m}{Alpha`Options`}
\index{matlab}

\alphausage{bitTrue}{bitTrue is a boolean option of cGen. If True, cGen includes the  bitOperators.h file in the C code and produces the compile script to  compile the code}{Alpha/Options.m}{Alpha`Options`}
\index{bitTrue}

\alphausage{debugC}{debugC: option (boolean) for VirtexGen. If True, Virtexgen generates  a function main which (i) reads system inputs on stdin (ii) evaluate  the system, (iii) prints the outputs on stdout (the output are not  significant in that case)}{Alpha/Options.m}{Alpha`Options`}
\index{debugC}

\alphausage{onlyLocalVars}{option used  when a function shouyld be applied sometimes nly on local variables }{Alpha/Options.m}{Alpha`Options`}
\index{onlyLocalVars}

\alphausage{projVector}{options of appSched to indicate what is the projection vector (appSched will try to find a unimodulary completion that is a projection along this vector) should be a Mathematica Vector (integer list) for instance: \{1,0,0\}(dimension is the number of indices plus parameter). Warning, this function does not try very hard to find a projection, the first one found is choosen. If it failed, please use the projMatrix options }{Alpha/Options.m}{Alpha`Options`}
\index{projVector}

\alphausage{projMatrix}{options of appSched to indicate what is the projection Matrix used for completing the schedule. should be a Mathematica Matrix (list of list of integer) for instance: \{\{1,0,0\},\{0,1,0\}\}. This matrix is a linear Matrix, i.e. it does not contains the affine part information (dimension is the number of indices plus parameter) }{Alpha/Options.m}{Alpha`Options`}
\index{projMatrix}

\alphausage{checkCase}{checkCase is an option of getContextDomain. If set, the function checks on the fly that the branches of the case expressions do not  overlap. False by default.}{Alpha/Options.m}{Alpha`Options`}
\index{checkCase}

\alphausage{mergeDomains}{mergeDomains is an option of simplifySpaceDom. If set (True), the domains are merged before the simplification is applied. Default value is False}{Alpha/Options.m}{Alpha`Options`}
\index{mergeDomains}

\alphausage{structured}{structured is an option of alpha0ToAlphard which specifies that a program may contain subsystems}{Alpha/Options.m}{Alpha`Options`}
\index{structured}

\alphausage{onlyModules}{onlyModules is an option of alpha2ToAlphard. If set (True), the program does not accept a system which is not a module. Default is False.}{Alpha/Options.m}{Alpha`Options`}
\index{onlyModules}

\alphausage{startTime}{startTime is an options of systemC generator for controleurs to indicate the  starting time of an Alphard program, i.e. the smallest value taken by the t index upon all the domains of the library, default Infinity}{Alpha/Options.m}{Alpha`Options`}
\index{startTime}

\alphausage{systemCFiles}{systemCFile is an options of systemC generator that indicates whether the codeis generated in one .h file (option value 1, default value) or in two files (.h and .cpp file, option value 2)}{Alpha/Options.m}{Alpha`Options`}
\index{systemCFiles}

\alphausage{busWidth}{options of socLibGen (default 32) set the bus Width used during the generation of the systemC  interface to Alphard programs}{Alpha/Options.m}{Alpha`Options`}
\index{busWidth}

\alphausage{inputOrOutput}{options of  socLibGen for internal use}{Alpha/Options.m}{Alpha`Options`}
\index{inputOrOutput}

\alphausage{exceptions}{option of several functions. Gives a list of variables or subsystems that has not to be considered for space time decomposition}{Alpha/Options.m}{Alpha`Options`}
\index{exceptions}

\alphausage{remIdDeps}{option of simplifySystem, false by default. If set, all identity dependences are removed. If it appears that this simplification is safe, It'll become a default option. PQ. July 2004.}{Alpha/Options.m}{Alpha`Options`}
\index{remIdDeps}

\alphausage{remIdEqus}{option of simplifySystem, false by default. If set, all identity equations are removed, unless the format is alpha0. If it appears  that this simplification is safe, It'll become a default option.  PQ. July 2004.}{Alpha/Options.m}{Alpha`Options`}
\index{remIdEqus}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/PipeControl.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.454932} by makeDoc.
%
% Section header
\section{The Alpha`PipeControl` package } 
\label{label:Alpha`PipeControl`}
\alphausage{eq2ge}{}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{eq2ge}

\alphausage{eq2le}{}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{eq2le}

\alphausage{PipeControl}{This package contains functions to transform case expressions for the pipeline of control variables.  These functions are: pipeControl, pipeAllControl, report, uniformizeMatrix, uniformizeDep, findSepHalfSpace, findPipeControl, domExplode.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{PipeControl}

\alphausage{pipeAllControl}{pipeAllControl[sys, options] generates pipeline control signals  for all the control variables present in sys (see pipeControl[])  and returns the modified system. pipeAllControl[options] applies  to \$result.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{pipeAllControl}

\alphausage{pipeControl}{pipeControl[var] generates a pipelined control signal from the definition of variable var.  pipeControl[sys, var] does the same to program contained in symbol sys. pipeControl checks that var is defined using a doubly nested case expression. The current version only covers the case when var is defined on a 2-dimensional domain, and we assume that the first dimension is the time. The new variable is named varPipe. pipeControl may not succeed for the following reasons: the pipeline direction is searched among the halfspaces found in the separating domains of the inside case expression and this heuristics may fail. In practice, we believe that it is reasonable...}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{pipeControl}

\alphausage{pipeInfo}{pipeInfo[occ:\{\_\_Integer\}] returns information about pipelining occurrence occ in \$result. If the occurrence can be piped, then pipeInfo returns a list formed by the occurence and the pipevector.  If not, it returns \$Failed. pipeInfo[] applies pipeInfo on all dependencies of  \$result. pipeInfo[ sys, occ ], or pipeInfo[ sys ] do the same to program sys. pipeInfo has an option execute $ \rightarrow $ True$|$False, When this option is set, pipeInfo executes the first pipe it has found}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{pipeInfo}

\alphausage{execute}{Option of pipeInfo. If True, pipeInfo tries to pipeline system, according to the informations it has found}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{execute}

\alphausage{pipeConstants}{Option of pipeInfo. If True, pipeInfo tries to pipeline constants}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{pipeConstants}

\alphausage{pipeVars}{pipeVars[] executes pipeInfo until it converges, a bounded number of times}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{pipeVars}

\alphausage{iterations}{Option of pipeVars. Gives the maximum number of times that pipeInfo can be called.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{iterations}

\alphausage{uniformizeMatrix}{uniformizeMatrix[ mat, contextDom ] tries to find out if mat can  be made uniform in the context defined by contextDom, and returns  the uniform matrix found.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{uniformizeMatrix}

\alphausage{constantizeMatrix}{constantizeMatrix[ mat, contextDom ] eliminates as many indexes  as possible in the context defined by contextDom, and returns the  matrix found.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{constantizeMatrix}

\alphausage{constantizeOccur}{constantizeOccur[occ] tries to transform into constants the dependency occ in program \$result, by eliminating as many redundant indexes as possible. The parameter occ specifies the position of the dependency, and can be obtained using the getOccurs function.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{constantizeOccur}

\alphausage{mkUniform}{mkUniform[occ] uniformizes the dependence occ in \$result. This function should be called only after getting the result of uniformizeOccur}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{mkUniform}

\alphausage{uniformizeOccur}{uniformizeOccur[occ] tries to uniformize the dependency occ in program \$result. The parameter occ specifies the position of the dependency, and can be obtained using the getOccurs function.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{uniformizeOccur}

\alphausage{reportDep}{reportDep[ repFile, occ, opts ] analyzes an dependency given by  its occurrence occ, with options opts, and writes the report in the output stream repFile. This function is called by report.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{reportDep}

\alphausage{report}{report[] analyzes all dependences of \$result and gives information on it in file current.report. report has options. report[name] analyzes a program in file name.alpha and puts the result in file name.report.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{report}

\alphausage{findSepHalfSpace}{findSepHalfSpace[exp] finds a list of separating hyperplanes for binary case expression exp. The method is heuristic in that the candidate hyperplanes are taken within the constraints defining the case expression. exp can be either a String or an Ast}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{findSepHalfSpace}

\alphausage{findPipeControl}{findPipeControl[x] finds if symbol x is defined using a doubly nested case expression, and returns the list of separating hyperplanes with time coefficient equal to 1, of the second level case expression. This function is used in the control signal generation.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{findPipeControl}

\alphausage{domExplode}{domExplode[dom] produces a list of halfspace from a domain definition.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{domExplode}

\alphausage{domExplode1}{domExplode1[dom] produces a list of halfspace from a domain definition.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{domExplode1}

\alphausage{route}{route[ occ, \{var, v\}, dom1, dom2 ] routes occurence occ in \$result, between dom1 and dom2, with routing vector v.}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{route}

\alphausage{delocalizeControl}{}{Alpha/PipeControl.m}{Alpha`PipeControl`}
\index{delocalizeControl}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Pipeline.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.427708} by makeDoc.
%
% Section header
\section{The Alpha`Pipeline` package } 
\label{label:Alpha`Pipeline`}
\alphausage{Pipeline}{Package. Contains the definition of the pipeline transformation. Containes the funcitons pipeline, pipeall, pipeIO}{Alpha/Pipeline.m}{Alpha`Pipeline`}
\index{Pipeline}

\alphausage{pipeline}{pipeline[position,pipespec] pipelines in system \$result expression given by position according to pipeline specification pipespec, and puts the result in \$result. pipeline[sys, position,pipespec] does the same to program contained in symbol sys. Parameter pipespec provides the name of the new variable, and the pipeline direction in the textual form "newname.translation" (example: "B.(i,j$ \rightarrow $i,j+1)" means that new pipeline variable is B and data movement is (0,1)). Parameter position gives the position of the expression in the program, using the conventions of the Mathematica function Position (see getPart).}{Alpha/Pipeline.m}{Alpha`Pipeline`}
\index{pipeline}

\alphausage{pipeall}{pipeall[var, exp, pipespec] pipelines in program \$result all occurrences of expression exp within equation whose left-hand side is var according to pipeline specification pipespec, and returns the result in \$result.  pipeall[sys, var, exp, pipespec] does the same to program contained in symbol sys.  pipeall[var, exp, pipespec, bd] pipelines exp of var definition  using pipespec, but extend the domain of pipeline to the boundary given by domain bd. Parameter var can be either a single variable,  a list of variables, or the empty list, in which case the expression is pipelined in all the program simultaneously. The parameters can be specified either in textual form or as  AST. For example,  pipeall["X","a.(i,j,k$ \rightarrow $i,j+1,k)","A.(i,j,k$ \rightarrow $i,j+1,k)"]  pipes all occurences of "a.(i,j,k$ \rightarrow $i,j+1,k)" in the definition of X,  using the pipeline specification "A.(i,j,k$ \rightarrow $i,j+1,k)". The pipeline specification contains the name to be used for the pipeline variable (here "A"), and the direction of pipeline given as a translation (here, vector (0,1,0)). Similarly,  pipeall["X","a.(i,j,k$ \rightarrow $i,j+1,k)","A.(i,j,k$ \rightarrow $i,j+1,k)",  "\{i,j,k$|$i$>$=0\}"]  pipes all occurences of "a.(i,j,k$ \rightarrow $i,j+1,k)" in the definition of X,  using the pipeline specification "A.(i,j,k$ \rightarrow $i,j+1,k)", withing  the boundary "\{i,j,k$|$i$>$=0\}"]. In case of error, pipeall returns Null, and leaves \$result unchanged.  For programming purposes, pipeall may  also be called with the form pipeall[sys,var,expression,pipespec],  where expression and pipeline are abstract syntax trees. Parameter expression can also be replaced by the occurrence number of the  expression. See also the function pipeInfo which gives hints about which pipeline commands to perform.}{Alpha/Pipeline.m}{Alpha`Pipeline`}
\index{pipeall}

\alphausage{pipeAll}{pipeall[var, exp, pipespec] pipelines in program \$result all occurrences of expression exp within equation whose left-hand side is var according to pipeline specification pipespec, and returns the result in \$result.  pipeall[sys, var, exp, pipespec] does the same to program contained in symbol sys.  pipeall[var, exp, pipespec, bd] pipelines exp of var definition  using pipespec, but extend the domain of pipeline to the boundary given by domain bd. Parameter var can be either a single variable,  a list of variables, or the empty list, in which case the expression is pipelined in all the program simultaneously. The parameters can be specified either in textual form or as  AST. For example,  pipeall["X","a.(i,j,k$ \rightarrow $i,j+1,k)","A.(i,j,k$ \rightarrow $i,j+1,k)"]  pipes all occurences of "a.(i,j,k$ \rightarrow $i,j+1,k)" in the definition of X,  using the pipeline specification "A.(i,j,k$ \rightarrow $i,j+1,k)". The pipeline specification contains the name to be used for the pipeline variable (here "A"), and the direction of pipeline given as a translation (here, vector (0,1,0)). Similarly,  pipeall["X","a.(i,j,k$ \rightarrow $i,j+1,k)","A.(i,j,k$ \rightarrow $i,j+1,k)",  "\{i,j,k$|$i$>$=0\}"]  pipes all occurences of "a.(i,j,k$ \rightarrow $i,j+1,k)" in the definition of X,  using the pipeline specification "A.(i,j,k$ \rightarrow $i,j+1,k)", withing  the boundary "\{i,j,k$|$i$>$=0\}"]. In case of error, pipeall returns Null, and leaves \$result unchanged.  For programming purposes, pipeall may  also be called with the form pipeall[sys,var,expression,pipespec],  where expression and pipeline are abstract syntax trees. Parameter expression can also be replaced by the occurrence number of the  expression. See also the function pipeInfo which gives hints about which pipeline commands to perform.}{Alpha/Pipeline.m}{Alpha`Pipeline`}
\index{pipeAll}

\alphausage{pipeIO}{pipeIO[var,exp,pipespec,boundary] pipelines in \$result I/O expression exp occuring in variable var to plane boundary according to pipeline specification pipespec. pipeIO[sys,var,exp,pipespec,boundary] does the same to program contained in symbol sys. Parameters var and exp are strings. Parameter pipespec is the pipeline specification string in the form "newVar.direction" (example: "X1.(t$ \rightarrow $t-1)").  Parameter boundary is given in textual form (example: " \\\{i,j $|$ 2i+3j+4$>$=0 \\\}").  For example, pipeIO["y","x","X.(t,n$ \rightarrow $t+1,n+1)"," \\\{t,n$|$n$>$=-1\\\}"] pipes all occurences of x in definition of y in \$result to plane " \\\{t,n $|$ n$>$=-1\\\}", with new name X, and pipeline direction (1,1).  If something wrong happens, pipeIO returns the system unmodified.  For programming purpose, exp, pipespec and boundary can be given as AST instead of textual form.  Also, the position of the expressions to pipeline may be given. See test file pipeIOT.m in \$MMALPHA/test.}{Alpha/Pipeline.m}{Alpha`Pipeline`}
\index{pipeIO}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Properties.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.514177} by makeDoc.
%
% Section header
\section{The Alpha`Properties` package } 
\label{label:Alpha`Properties`}
\alphanote{Properties}{Documentation checked on August 8, 2004}

\alphausage{Properties}{The Alpha`Properties` package contains the definition of various  property-checking functions. Should be used as the preferred place  for adding new property-checking predicates that might be of large  interest.}{Alpha/Properties.m}{Alpha`Properties`}
\index{Properties}

\alphanote{Properties}{This package should probably merged with other packages such as Tables, Substitution, etc.}

\alphausage{allDomEqualQ}{allDomEqualQ[domList] is True is all the domains in the list domList are equal.}{Alpha/Properties.m}{Alpha`Properties`}
\index{allDomEqualQ}

\alphausage{allDomDisjointQ}{allDomDisjointQ[domList] is True if all the domains in a list domList are two-by-two disjoint.}{Alpha/Properties.m}{Alpha`Properties`}
\index{allDomDisjointQ}

\alphausage{allDomUnion}{allDomUnion[domList] returns the domain which is the union of the domains in domList.}{Alpha/Properties.m}{Alpha`Properties`}
\index{allDomUnion}

\alphausage{uniformQ}{uniformQ[sys] is True if the Alpha program sys is uniform  (all dependences between local variables are translations),  False otherwise. The default value of sys is \$result.  uniformQ can also be applied to an Alpha matrix or to  an affine AST.}{Alpha/Properties.m}{Alpha`Properties`}
\index{uniformQ}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Reduction.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.522230} by makeDoc.
%
% Section header
\section{The Alpha`Reduction` package } 
\label{label:Alpha`Reduction`}
\alphanote{Reduction}{Documentation revised on August 8, 2004}

\alphausage{Reduction}{The Alpha`Reduction` package contains transformations to deal with  reduction operators.}{Alpha/Reduction.m}{Alpha`Reduction`}
\index{Reduction}

\alphanote{Reduction}{This package is not completed. In particular, it contains two functions splitReduction and splitReduce which do no work.}

\alphausage{serializeReduce}{serializeRecuce[pos] serializes the reduction at position pos in \$result. serializeReduce[pos, spec] serializes the reduction at position  pos in \$result using serialization specifier spec (spec is a string). Parameter pos may be either a position in the AST, or the name (string) of a variable whose definition has the form "pos = reduction". serializeReduce[sys, pos, spec] does the same to program sys. Parameter spec is a string containing the new variable name composed with the new serialized dependence (parameters need not be given  explicitly). For example: "Z.(i,k$ \rightarrow $i,k-1)" means that the serialized  variable will be "Z" and the serialize direction will be the vector  (0,-1). If no spec is given, the function guesses the direction of serialization, and the option (invert $ \rightarrow $ False$|$True) allows this direction to be changed. The guess is based on the null space of the reduction function, and it works only if this null space has exactly one vector.}{Alpha/Reduction.m}{Alpha`Reduction`}
\index{serializeReduce}

\alphausage{isolateReductions}{isolateReductions[] locates all the reductions in a system and isolates these reductions in new equations}{Alpha/Reduction.m}{Alpha`Reduction`}
\index{isolateReductions}

\alphausage{isolateOneReduction}{isolates one reduction}{Alpha/Reduction.m}{Alpha`Reduction`}
\index{isolateOneReduction}

\alphausage{splitReduction}{splitReduction[y] tries to rewrite the reduction at rhs of symbol y as a sequence of reduction. This function does not work currently.}{Alpha/Reduction.m}{Alpha`Reduction`}
\index{splitReduction}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Schedule.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.532490} by makeDoc.
%
% Section header
\section{The Alpha`Schedule` package } 
\label{label:Alpha`Schedule`}
\alphausage{Schedule}{Package.  Schedule for  alpha program The only function used is : schedule[]}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{Schedule}

\alphausage{structSched}{structSched[] tries to find a structured scheduling for \$result, using \$scheduleLibrary for schedule of subsystems. The default mode is linear, i.e. no dimension is added for  subsystems. By setting the option structSchedType to multi, the schedule of the subsystems will be in an additionnal dimension. Warning, in that case, you MUST set the depth of the schedule expected with the multSchedDepth options. Example: structSched[structSchedType$ \rightarrow $multi,multiSchedDepth$ \rightarrow $2] finds the schedule of \$result and puts}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{structSched}

\alphausage{schedule}{schedule[] finds the schedule of \$result and puts the result in \$schedule. schedule[sys\_Alpha`system], finds the schedule of alpha system `sys' and puts the result in \$schedule (affine by variable by default).  schedule[sys\_Alpha`system, options] calls schedule with non default options. Options[schedule] provides   the options of schedule, and ?opt1 provides information on option   opt1.  ?\$schedule provides info on the output format of the schedule.   The schedule computation may take a long time    (2 minutes for 20 instructions). More information is available in the file   \$MMALPHA/doc/user/<=Scheduler\_user\_manual.ps.}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{schedule}

\alphausage{checkOptions}{checkOption[sys\_Alpha`system,options\_\_\_Rule] check that the set of option of schedule is coherent, returned a new set of option in which some option have been changed by default to cope with incoherent options if possible}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{checkOptions}

\alphausage{benchSched}{benchSched[f] runs the scheduler on alpha file f, or on  list of alpha files f.}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{benchSched}

\alphausage{farkas}{option of benchSched}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{farkas}

\alphausage{vertex}{option of benchSched}{Alpha/Schedule.m}{Alpha`Schedule`}
\index{vertex}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/ScheduleTools.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.544242} by makeDoc.
%
% Section header
\section{The Alpha`ScheduleTools` package } 
\label{label:Alpha`ScheduleTools`}
\alphausage{applySchedule}{applySchedule[] applies the schedule \$schedule to \$result. The resulting system is the original one where all the change of basis in \$schedule have been performed and where the first index can interpreted as the time everywhere except for the input and output variables.  applySchedule[sys\_Alpha`system] applies the schedule \$schedule to system `sys'.  applySchedule[sys\_Alpha`system, sched\_Alpha`Schedule] applies the schedult `sched' to system `sys'. See also: schedule, \$schedule.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{applySchedule}

\alphausage{showSchedResult}{showSchedResult[sched1\_Alpha`ScheduleResult] pretty prints the schedule sched1, equivalent to show[sched1]}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{showSchedResult}

\alphausage{other}{other is an option of appSched which means for other variables}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{other}

\alphausage{renameIndices}{renameIndices[sys:Alpha`system,indiceList:\{\_\_\_String\}] rename the indices of all the local variables of the system `sys' according to the list given `indiceList' (including the output of `use'). renameIndices[sys:Alpha`system,indiceList:\{\_\_\_String\},var:(\_String$|$\{\_\_String\})] rename the change the name in only one variable `var' (or a list of given variable)}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{renameIndices}

\alphausage{appSched}{appSched[ sys, sch, options ] returns a scheduled system, according to  sch. appSched[ sch, options ] applies schedule sch to \$result.  appSched[ options ] applies schedule \$schedule to \$result. appSched is equivalent to ApplySchedule but works for  multidimensionnal scheduling. The projection direction is  chosen automatically by \MMAlpha{}, but if necessary, it can be chosen  using the projMatrix or the projVector options. See ?appSchedOptions for more informations.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{appSched}

\alphausage{appSchedOptions}{appSched chooses automatically the projection direction for variables,  and this sometimes may not be convenient. There are several ways to specify more precisely the schedule. Option projVector allows a  projection vector to be specified. For example, projVector $ \rightarrow $ \{1,0,0\} would say that the first component of all variables will be replaced by the time given by the schedule, in a 3-dimensional space. However, this option can be used only if all variables of the system are already  placed in the same space, which may not be the case. Similarly, one can specify the projection matrix, using the projMatrix option. For example, projMatrix $ \rightarrow $ \{\{1,0,0\},\{0,0,1\}\} will project all variables on components 1 and 3. Again, this may not be convenient if all variables are not in the same space. Another way of specifying the projection is to use the timeDimensions option. Specifying timeDimensions $ \rightarrow $ \{2\}  says that the component of all variables that will be replaced by  the time is the second dimension. This is equivalent to the option projVector $ \rightarrow $ \{0,1,0\}, but less dependant on the dimension of the  variables. Moreover, one can specify the time dimensions variable by variable, using the variables option. This option may take several forms. variables $ \rightarrow $ varlist, where varlist is a list of local variables, restricts appSched to be applied to only variables in varlist.  variables $ \rightarrow $ \{\{ var, timeDimensions \}, ... \} allows the value of option  timeDimensions to be attached to each individual variable. For example, variables $ \rightarrow $ \{\{ "A", \{2\} \}, \{ "B", \{1\} \}\} says that for variable  A, the time dimensions is 2, and for variable B, it is 1. One can also factorize this list as in variables $ \rightarrow $\{\{\{"A","C"\}, \{2\},  \{"B", \{1\}\}\}, and also use the keyword other to specify the  time dimensions for all remaining variables, as in  variables $ \rightarrow $ \{\{ "A", \{2\} \}, \{ "B", \{1\} \}, \{ other, \{3\} \}\}. Notice that timeDimensions is a list of dimensions: this is a provision for a multi-dimensional schedule, but this has not been tested yet. Also, remember that no garantee is given that these options provide a possible projection, as this depends on the existence of a unimodular transformation combining the schedule and the projection. In other words, appSched may fail sometimes. A final warning: projection vector or  matrices should include the parameters' dimensions.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{appSchedOptions}

\alphausage{timeDimensions}{timeDimensions is an option of appSched and appVarSched that gives the  list of dimensions that are to be assigned to time. This option  gives a hint to appSched. Say a variable has indexes i,j,k (including parameters), then timeDimensions$ \rightarrow $\{2\} means that index j is going to be replaced by the time and thus, indexes i and k will become  processors. This hint may not always work. See also ?appSchedOptions.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{timeDimensions}

\alphausage{variables}{option of appSched. It gives the list of variables for which the schedule must be applied. Default value is Null, and means all variables. See also ?appSchedOptions for the exact syntax of this option.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{variables}

\alphausage{appVarSched}{appVarSched[sys:\_system,var\_String,time\_Matrix,options] apply a single COB to a variable,  the schedule is to be given as a Mathematica matrix (List of List of integer). This function  is for internal use }{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{appVarSched}

\alphausage{convertSchedule}{convertSchedule[sch:\_scheduleResult]. Takes a schedule table and returns an equivalent time functions table. This format is used by lifetime, appSched, and other functions}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{convertSchedule}

\alphausage{identitySchedule}{identitySchedule[sys,dim] build the identity schedule of dimension `dim` dor system sys. The identity schedule simply consist of the lexicographic order  on the `dim`first indices (i.e. the program is already scheduled }{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{identitySchedule}

\alphausage{isScheduledQ}{isScheduleQ[sys,dim] check whether the sys is scheduled or not (with a schedule of dimension dim. isScheduledQ[sys] assumes a linear schedule}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{isScheduledQ}

\alphausage{testSched}{testSched[sys:\_system, sch:\_scheduleResult] tests the the schedule for the given system. When flow dependencies are not satisfied, it prints messages and returns False.  testSched[sys:\_system, sch:\_scheduleResult, var:\_String] test schedule for var only. Param sys may be a dependencies table.  Default: testSched[] test \$result with \$schedule.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{testSched}

\alphausage{equationOrderQ}{equationOrderQ[sys] checks if the equations are correctly ordered to generate code with a schedule with duration of 0 for some equation (i.e. graph of dependencies internal to te loop nest is topologicaly sorted; If the answer is Yes, the function returns True otherwise it attempts to find a permutation of the equations that is correct(returns a list of integer). This function does not take into account dependencies involving input or output variables WARNING, this function is temporarily disable because of an internal bug (TopologicalSort)}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{equationOrderQ}

\alphausage{reorderEquations}{ reorderEquations[sys,permutation] reorder the equation of system sys according to the permutation given (which only concern the local variables, see the function equationOrderQ). The output equation are put at the end of the program.}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{reorderEquations}

\alphausage{buildPseudoAlpha}{buildPseudoAlpha[sys\_] returns a system `sys2' in which all the subsystems called by `sys' have been replaced by equations on the output variables of the sub-system use. This function is used in order to be able schedule a structured system. buildPseudoAlpha[] is by default applied to \$result }{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{buildPseudoAlpha}

\alphausage{buildSchedConstraints}{buildSchedConstraints[sys\_,List[sched\_scheduleResult]] builds a list of constraints (strings) that correspond to the schedules given as second argument for all the subsystem called in `sys'. buildSchedConstraints[] calls the function buildSchedConstraints[\$result,\$scheduleLibrary]}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{buildSchedConstraints}

\alphausage{buildSchedConstraintForUse}{buildSchedConstraintForUse[sys\_,use1\_use,sched\_SchedulResult] build a list of constraints (strings) that correspond to the schedule given in sched of the subsystem used in `use1' which is part of `sys'}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{buildSchedConstraintForUse}

\alphausage{isReducibleQ}{isReducibleQ[sys] checks wether a structured Alpha program is `reducible' in the sens described in Irisa research report PI1140: the dependency between output and input of a subsystems have identity on the extension part}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{isReducibleQ}

\alphausage{addBufferVars}{addBufferVars[sys:\_Alpha`system] adds systematically a  local `buffer' variables  for each input and each output of each use of subsystems in `sys' (warning, if the input is not a simple variable, the function fails, this function is to be for internal use, please refer to simplifyUseInputs[])}{Alpha/ScheduleTools.m}{Alpha`ScheduleTools`}
\index{addBufferVars}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Schematics.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 21.172837} by makeDoc.
%
% Section header
\section{The Alpha`Schematics` package } 
\label{label:Alpha`Schematics`}
\alphausage{Schematics}{This package contains functions for drawing a raw schematics of an Alpha  programs}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{Schematics}

\alphausage{partShown}{partShown is an option of schematics. It defines the region of the plot that will be shown. Default is Automatic and corresponds to a region \{\{0,1\},\{0,1\}\}, which displays the full schematics. partShown $ \rightarrow $\{\{0.2,0.4\},\{0.3,1.2\}\} would display a region whose coordinates, relative to the schematics, are changed.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{partShown}

\alphausage{fSize}{fSize is an option of schematics, which changes the size of the font used to show symbols. Default value is 20 .}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{fSize}

\alphausage{sPositions}{sPositions is an option of schematics. Default value is square. sPositions $ \rightarrow $ Null stacks the operators.  sPositions $ \rightarrow $ \{columns $ \rightarrow $ nb\} gives the number of columns of the diagram.  sPositions may be a list of coordinates for the operators. In this case, make sure that the number of positions is the same as the number of operators.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{sPositions}

\alphausage{square}{square is an option value for option sPosition in  the schematics function. sPosition $ \rightarrow $ square asks for a square shaped schematics layout.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{square}

\alphausage{columns}{columns is an option value for option sPosition in  the schematics function. sPositions$ \rightarrow $\{columns$ \rightarrow $nb\} asks for a nb column schematics layout.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{columns}

\alphausage{yFactor}{yFactor is the form factor of boxes used by the schematics function. Default value is 0.2 .}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{yFactor}

\alphausage{offsetX}{offsetX is the X offset value for schematics. Default value is 0.5 .}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{offsetX}

\alphausage{offsetY}{offsetY is the Y offset value for schematics. Default value is 0.1 .}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{offsetY}

\alphausage{newName}{newName[ sys, v ] returns a new unique name built from variable v.  newName[ v ] does the same in \$result. This function cannot be used in any context. Use getNewName instead.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{newName}

\alphausage{skeleton}{skeleton[] gives the skeleton of \$result}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{skeleton}

\alphausage{schematics}{schematics[] draws a schematic diagram of the  program \$result. See Options[schematics] to get the options of  schematics.}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{schematics}

\alphausage{flattenEquation}{}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{flattenEquation}

\alphausage{flattenSkeleton}{ddddd}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{flattenSkeleton}

\alphausage{findAliases}{}{Alpha/Schematics.m}{Alpha`Schematics`}
\index{findAliases}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Semantics.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 18.751224} by makeDoc.
%
% Section header
\section{The Alpha`Semantics` package } 
\label{label:Alpha`Semantics`}
\alphanote{Semantics}{Documentation revised on August 10, 2004}

\alphausage{Semantics}{Alpha`Semantics` is the package which contains the functions for computing the type of Alpha expressions. These functions are changeType, expDimension, expDomain, expType, getContextDomain, matchTypes,  and replaceByEquivExpr.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{Semantics}

\alphausage{changeType}{changeType[sys,var,newType] changes the type of the variable var to newType in sys (default \$result). No typing compatibility check  is done: this is not a semantic preserving transformation. Mainly  used for changing integers to fixed bit width integers.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{changeType}

\alphausage{expDimension}{expDimension[sys,exp] computes the dimension of the expression exp the in  system contained in the system sys (default \$result). The expression may be given as a string, a position vector (following the convention of the Mathematica function Position) in the system, or as an AST. expDimension outputs error messages if the dimensions are not compatible, and returns dimension -1 in this case.  Remark: parameters are counted as additional dimensions.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{expDimension}

\alphausage{expDomain}{expDomain[sys,exp] computes and returns the domain of exp in program sys (default \$result) or \$Failed if the expression is badly formed. expDomain generates error and warning messages accordingly. The expression may be given as a string, a position vector (following the convention of the Mathematica function Position) in the system, or as an AST.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{expDomain}

\alphausage{expType}{expType[sys,exp] computes the type (integer, boolean, real, bottom) of an expression and checks its correctness using type--matching rules. Default value of sys is \$result. The expression may be given as a string, a position  vector (following the convention of the Mathematica function Position) in the system, or as an AST. expType returns bottom if the expression is not correctly typed.  In such a case, the offending tree and the colliding types are printed on the screen.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{expType}

\alphausage{getContextDomain}{getContextDomain[sys,pos] computes the context domain in which an expression at position pos in sys is used. Default value of sys is \$result. The context is the domain inherited by an expression from the constructs which surrounds it.  The position is counted from the root of the system following the Mathematica notion of position.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{getContextDomain}

\alphausage{matchTypes}{matchTypes[type1,type2] computes the type corresponding to the union of those of the parameters (integer, boolean, real) and returns the  resulting type or bottom if the types are incompatible.}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{matchTypes}

\alphausage{replaceByEquivExpr}{replaceByEquivExpr[sys,pos,expr] replace in sys the expression present at position pos by expr. Default value of sys is \$result. replaceByEquivExpr[sys,expr1,expr2] replace in sys all occurences of expr1 by expr2. expr1 and expr2 can be given in AST form of Alpha form.  Warning: the equivalence of the resulting program is NOT guaranteed. A test is made to check if both expressions are equivalent, but if this test fails, the replacement is done anyway. This function should be used for replacing an expression by an equivalent one when this cannot be done automatically (for instance, in presence of degenerated domains).}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{replaceByEquivExpr}

\alphausage{setBitWidth}{setBitWidth[var:\_String,value:\_Integer] change the scalar type of variable var1 in system sys if this variable has a specified bitwidth (e.g. integer[S,5]) and set it to the bitwidt value, if the variable does not have a specified bitwidth, it does nothing and issue a warning}{Alpha/Semantics.m}{Alpha`Semantics`}
\index{setBitWidth}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Static.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.855745} by makeDoc.
%
% Section header
\section{The Alpha`Static` package } 
\label{label:Alpha`Static`}
\alphanote{Static}{Documentation revised on August 10, 2004}

\alphausage{Static}{The Alpha`Static` package contains functions for the static analysis  of Alpha programs. These functions are: analyze, dep, and checkUseful.}{Alpha/Static.m}{Alpha`Static`}
\index{Static}

\alphausage{analyze}{analyze[sys] performs a static analysis of system sys (default \$result) and returns True if the analysis is successful (even with warnings), False otherwise. The analyze function has options verbose, recurse, library and scalarTypeCheck (see Options[analyze]). Option verbose (Boolean; default True) reports analysis progress.  Option recurse (Boolean; defaut False) recursively looks up for all the subsystems involved and analyzes them, too.  Option library (List; default \$library) gives the library to search for subsystems. Example of option usage: analyze[recurse$ \rightarrow $True, verbose$ \rightarrow $False]}{Alpha/Static.m}{Alpha`Static`}
\index{analyze}

\alphausage{dep}{dep[sys] generates a dependency table for sys (default \$result). The form of the table is:\\ dtable[\{\\   depend[ Domain, LHS\_var\_name, RHS\_var\_name, Matrix, RHS\_var\_domain],\\   depend[ Domain, LHS\_var\_name, RHS\_var\_name, Matrix, RHS\_var\_domain],\\   . . . \\ ].\\ The table may be pretty--printed using show[dep[]].}{Alpha/Static.m}{Alpha`Static`}
\index{dep}

\alphausage{dep1}{dep1[sys, \{occur:\{\_Integer..\}, occurs\_\_\_\}] generates dependency table entries for all occurrences in the list. Used by dep[] to generate the dependency table.}{Alpha/Static.m}{Alpha`Static`}
\index{dep1}

\alphausage{checkUseful}{checkUseful[sys, var] checks that the variable var is used in the system sys for all the points of its domain. Default value of sys is \$result. If var is omitted, the function is applied to all variables of sys.}{Alpha/Static.m}{Alpha`Static`}
\index{checkUseful}

\alphausage{checkDeclarations}{This function is a private function of the package.}{Alpha/Static.m}{Alpha`Static`}
\index{checkDeclarations}

\alphausage{buildLHSIdList}{This function is a private function of the package.}{Alpha/Static.m}{Alpha`Static`}
\index{buildLHSIdList}

\alphausage{buildRHSIdList}{This function is a private function of the package.}{Alpha/Static.m}{Alpha`Static`}
\index{buildRHSIdList}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/SubSystems.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.865821} by makeDoc.
%
% Section header
\section{The Alpha`SubSystems` package } 
\label{label:Alpha`SubSystems`}
\alphanote{SubSystems}{Documentation revised on August 10, 2004}

\alphanote{SubSystems}{Bugs: test[SubSystems] fails. inlineAll should test  that \$result contains a system.}

\alphausage{SubSystems}{The Alpha`SubSystems` package contains functions to operate on  Alpha subsystems. These functions are  assignParameterValue, assignParameterValueLib, fixParameterValue, inlineAll, inlineSubSystem, spread, removeIdEqus, simplifyUseInputs, substDom, subSystemUsedBy, and topoSort.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{SubSystems}

\alphausage{addParameterId}{function externalized for debugging purposes.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{addParameterId}

\alphausage{affExtHom}{function externalized for debugging purposes.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{affExtHom}

\alphausage{assignParameterValue}{assignParameterValue[param,v,sys] gives the value v to the parameter param in the Alpha system sys and returns the modified system. param is a string, and v an integer. Default value of sys is \$result.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{assignParameterValue}

\alphausage{assignParameterValueLib}{see fixParameter.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{assignParameterValueLib}

\alphausage{fixParameter}{fixParameter[param,v,systName] gives the value v to a parameter param in an Alpha system of name "systName" and in the call to "systName"  in \$library. It returns the new library (list of Alpha systems) and  modifies \$library (except if \$library is explicitely specified as the  4th parameter). If the system name is ommited, the function sets the  parameter in all the programs of \$library and in \$result.  WARNING: currently, this function supposes that the parameter "param" has the same value everywhere, you also have to ensure that \$result is the top calling system of the library (i.e. the module in an AlpHard program).}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{fixParameter}

\alphausage{inlineAll}{inlineAll[options] inlines all the subsystems of system \$result. Options are rename, renameCounter, verbose, caller, library, and current. (see Options[inLineAll] for default values).}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{inlineAll}

\alphausage{inliningRenameCounter}{Variable. Holds the counter value used to avoid name conflicts  while renaming the variables.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{inliningRenameCounter}

\alphausage{inlineSubsystem}{inlineSubsystem[name,options] searches the current system (see option : caller) for a use statement of the subsystem name, and replaces this use statement with its definition extracted from  \$library, with proper variable renaming and parameter instanciation. Returns the modified system if no error, the caller otherwise.  Side effect: if the "current" option is set, it sets \$program to the previous Alpha`\$result and sets \$result to the returned value. Options are occurence, rename, renameCounter, verbose, caller, library,  undescore, and current (default options in Options[inlineSubsystem]).}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{inlineSubsystem}

\alphausage{library}{library is an option of inlineAll and inlineSubsystem.   library $ \rightarrow $ list of systems (default \$library) specifies the  list of systems to search for a subsystem. When more than one  declaration of the same system appear in the library, the first  one is inlined.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{library}

\alphausage{occurence}{}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{occurence}

\alphausage{underscore}{option of inlineSubsystem and inlineAll (Boolean). When True (default), new identifier separator is \_, whereas it is X otherwise.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{underscore}

\alphausage{spread}{spread[ var, index ] replaces var in system \$result by a set of variables varI, where I is in the range of index of var. This  function is not completed and should not be used.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{spread}

\alphausage{removeIdEqus}{removeIdEqus[sys] removes in sys the equations of the form A=B as introduced for example by the inlineSubsystem and inlineAll  transformations.\\ Warning, this function does not normalize the resulting system, unless option norm $ \rightarrow $ True is used. removeIdEqus has options norm, allLibrary and inputEquations.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{removeIdEqus}

\alphausage{simplifyUseInputs}{simplifyUseInputs[sys] adds local buffer variables for each input which  is not already a simple variable (input to a use may be any expression). The new variable is defined on the 'real domain' of the expression: its  context domain intersected with its use domain.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{simplifyUseInputs}

\alphausage{substDom}{substDom[ dom, extDom, paramAssign, nbparams] computes the transformation of a domain dom in a subsystem inlining. For internal use only.}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{substDom}

\alphausage{subSystemUsedBy}{subSystemUsedBy[sys] returns the list of names of subsystems used by  system sys (default \$result).}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{subSystemUsedBy}

\alphausage{topoSort}{topoSort[lib] returns the list of system names of library  lib, sorted in the reverse hierachical order, i.e. if a system A  uses a system B and a system C, topoSort returns \{B, C, A\} or \{C, B, A\}}{Alpha/SubSystems.m}{Alpha`SubSystems`}
\index{topoSort}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Substitution.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 18.860811} by makeDoc.
%
% Section header
\section{The Alpha`Substitution` package } 
\label{label:Alpha`Substitution`}
\alphanote{Substitution}{Documentation revised on August 8, 2004}

\alphausage{Substitution}{Alpha`Substitution` is the package which contains the  functions for substituting Alpha variables.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{Substitution}

\alphausage{addLocal}{see addlocal in Substitution.m}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{addLocal}

\alphausage{addlocal}{addlocal[sys, var, exp] declares a new local variable var in system sys and defines it as exp on the domain of exp as calculated by expDomain. All instances of exp in sys are replaced with the variable var.\\ addlocal[sys, var, pos] adds a new local variable var defined on the  context domain of pos (as calculated by getContextDomain[]) and adds the definition of var as the expression at position pos. The expression defined by pos is replaced by var. Default value of  sys is \$result. WARNING: This function is kept for backward compatibility,  but it is not sufficiently specified, please use rather addLocalLHS or addLocalRHS functions}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{addlocal}

\alphausage{addLocalLHS}{addLocalRHS[var, exp] is similar to addlocal (see addlocal) but  its action is more clearly defined. From a set of equations such as \\A =...; X =...A...\\ addLocalLHS["B","A"] changes it into\\ B = A; A =...; X =...B....\\ i.e., B is added in the LHS of the equation using A.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{addLocalLHS}

\alphausage{addLocalRHS}{addLocalRHS[var, exp] is similar to addlocal (see addlocal) but  its action is more clearly defined. From a set of equations such as \\A =Y; X =...A...\\ addLocalRHS["B","A"] changes it into\\ A=B;B=Y;X=...B....,\\ i.e., B is added in the RHS of the equation using A, except if  A is an input, in which case it acts as addLocalLHS:\\ B=A; X=...B....}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{addLocalRHS}

\alphanote{addlocal}{addlocal is more or less the inverse transformation of substituteInDef.}

\alphausage{getNewName}{getNewName[sys,var] checks that the identifier "var" is not already  used in sys (default \$result) and returns it if not. If this  identifier already exists, it returns a modified version of "var"  by duplicating its last letter.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{getNewName}

\alphausage{getOccurs}{getOccurs[sys,p] finds out the positions of a pattern p in system sys,  and returns a Mathematica position specifier containing the list of  occurrences of p in \$result. The pattern p can be a string (e.g.  "A.(i,j$ \rightarrow $i+j)"), in which case it is parsed, or it can be an Alpha  AST. The result of getOccurs is a list of positions specifiers that are defined with respect to sys. A position specifier can then be  used in the function Part to access the element. For example, if the position is \{6,2,3\}, then Part[sys,6,2,3] gives the element. One can also use getPart[sys,\{6,2,3\}], which is easier to use directly.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{getOccurs}

\alphausage{getOccursInDef}{getOccursInDef[sys, var, p] lists the positions of occurrences of  a pattern p in the definition of variable lhs inside an ALPHA program sys (default \$result).  \\getOccursInDef[sys, var, p, rank] gives the position number rank in the result of getOccursInDef[sys, var, p]. rank specifies which occurrence to report (1 = first, 2 = second,  \{1,2\}= first and second, etc).}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{getOccursInDef}

\alphausage{replaceDefinition}{replaceDefinition[lhs, rhs] replaces the definition of a variable lhs  in an equation of program \$result with the Alpha expression rhs. replaceDefinition[sys, lhs, rhs] replaces the definition of variable  lhs in an ALPHA program sys with the Alpha expression rhs and return  the new system. lhs is a variable name (either symbol or string).  rhs is either an ast or a string.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{replaceDefinition}

\alphanote{replaceDefinition}{The function returns a copy of the original program in which the rhs of the equation defining the specified variable is replaced by an expression passed as parameter. The result of the substitution is not normalized. The meaning of the program may be changed by this transformation.}

\alphausage{substituteInDef}{substituteInDef[lhs, var] substitutes in \$result all occurrences of variable `var' in the RHS of the definition of variable `lhs' by the definition of `var', and returns the new system in \$result. substituteInDef[lhs,var,rank] substitutes occurrences `rank' of variable `var' in the RHS of the defition of variable `lhs' by the definition of `var'. The parameter rank specifies which occurrence to replace ( 1 = first, 2 = second, \{1,2\}= first and second, etc). substitute[sys,lhs,var] and substitute[sys,lhs,var,rank] do the same on program contained in symbol `sys'.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{substituteInDef}

\alphausage{occursInDefQ}{occursInDefQ[sys,var,p] returns True if the pattern p occurs in the of the definition of the variable var. Default value of sys is \$result. p can be either a string or an AST.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{occursInDefQ}

\alphausage{unusedVarQ}{unusedVarQ[sys,var] is True if var is used in the rhs of an equation of system sys, False otherwise. Default value of sys is \$result.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{unusedVarQ}

\alphausage{removeUnusedVar}{removeUnusedVar[sys,var] removes the definition of unused variable var in system sys. The default value of sys is \$result.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{removeUnusedVar}

\alphausage{removeAllUnusedVars}{removeAllUnusedVars[sys] removes the definitions of all unused local variables of system sys (default, \$result).}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{removeAllUnusedVars}

\alphausage{isOutputRegular}{isOutputRegular[sys,o] is True if output variable o has the form o = v where v is a simple variable. This predicate allows one to  detect non regular outputs}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{isOutputRegular}

\alphausage{areAllOutputsRegular}{areAllOutputsRegular[ sys ] is True if all outputs of sys have the form o = v, False otherwise. areAllOutputsRegular[] does the same to \$result.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{areAllOutputsRegular}

\alphausage{mkOutputRegular}{mkOutputRegular[sys,o] makes output variable o regular, if necessary. In other words, it makes sure that the definition of o has the form  o = v where v is a simple local variable. mkOutputRegular[o] does the same to \$result.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{mkOutputRegular}

\alphausage{mkAllOutputsRegular}{mkAllOutputsRegular[sys] makes all output variables of sys regular. mkAllOutputsRegular[] does the same on \$result.}{Alpha/Substitution.m}{Alpha`Substitution`}
\index{mkAllOutputsRegular}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Tables.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.925924} by makeDoc.
%
% Section header
\section{The Alpha`Tables` package } 
\label{label:Alpha`Tables`}
\alphanote{Matrix}{Documentation revised on March 11, 2008}

\alphausage{Tables}{The Alpha`Tables` package contains variable and function definitions relative to the retrieval of contextual information on nodes and  programs.  Functions are:   addAllParameterDomain, addParameterDomain, getDeclaration,  getDeclarationDomain, getDefinition, getDimension, getEquation, getIndexNames, getInputVars, getLocalVars, getOutputVars, getVariables, getSystemName, getSystemParameters, getSystemParameterDomain, symToString, lookUpFor, lookUpForPos, accessByPath, and undoModif.}{Alpha/Tables.m}{Alpha`Tables`}
\index{Tables}

\alphausage{addAllParameterDomain}{addAllParameterDomain[sys] adds the constraints present in the  parameter domain of system sys to each domain of the system.  addAllParameterDomain[] does the same to \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{addAllParameterDomain}

\alphausage{addParameterDomain}{addParameterDomain[ dom, paramdom ] adds to the constraints of dom  the constraints on the parameters defined in paramdom. For internal use  mostly.}{Alpha/Tables.m}{Alpha`Tables`}
\index{addParameterDomain}

\alphausage{getDeclaration}{getDeclaration[var] returns the complete declaration of variable var in system \$result or an empty list if the declaration does not exist. getDeclaration[sys, var] returns the declaration of var in system sys.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getDeclaration}

\alphausage{getDeclarationDomain}{getDeclarationDomain[v] returns the declaration domain of variable v  in system \$result. getDeclarationDomain[sys,v] returns the declaration domain  of variable v in system sys}{Alpha/Tables.m}{Alpha`Tables`}
\index{getDeclarationDomain}

\alphausage{getDefinition}{getDefinition[var] returns the RHS of the equation defining variable var or an empty list if the definition does not exist. If the variable is defined as output of a subsystem, returns an empty list and prints  an error message.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getDefinition}

\alphausage{getDimension}{getDimension[dom] returns the dimension of the Alpha domain dom.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getDimension}

\alphausage{getEquation}{getEquation[var] returns the equation defining a variable var or an  empty list if the definition does not exist. If the variable is the output of a subsystem, getEquation issues a warning and returns the use statement of this subsystem.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getEquation}

\alphausage{getIndexNames}{getIndexNames[sys, var] returns the list of index names from  a variable's declaration. getIndexNames[var] does the same to  \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getIndexNames}

\alphausage{getInputVars}{getInputVars[sys] returns the list of input variables used in system sys. getInputVars[] does the same to \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getInputVars}

\alphausage{getLocalVars}{getLocalVars[sys] returns the list of local variables used in system sys. getLocalVars[] does the same to \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getLocalVars}

\alphausage{getOutputVars}{getOutputVars[sys] returns the list of Output variables used in system sys. getOutputVars[] does the same to \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getOutputVars}

\alphausage{getVariables}{getVariables[sys] returns the list of variables used in system sys. getVariables[] does the same to \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getVariables}

\alphausage{getSystemName}{getSystemName[sys] return the name of system sys. getSystemName[] gives the name of \$result.}{Alpha/Tables.m}{Alpha`Tables`}
\index{getSystemName}

\alphausage{getSystemParameters}{getSystemParameters[sys] return the parameters of the system sys (default \$result)}{Alpha/Tables.m}{Alpha`Tables`}
\index{getSystemParameters}

\alphausage{getSystemParameterDomain}{getSystemParameterDomain[sys] return the parameters domain  of the system sys (default \$result)}{Alpha/Tables.m}{Alpha`Tables`}
\index{getSystemParameterDomain}

\alphausage{symToString}{symToString[symb\_] Converts a symbol to string. Leaves any other object simply evaluated. Returns the string containing its name according in current output format.}{Alpha/Tables.m}{Alpha`Tables`}
\index{symToString}

\alphausage{lookUpFor}{lookUpFor[sys, position:\{\_Integer...\}, operator\_Symbol] returns the   smallest tree whose root is the specified by operator and which   contains the specified position, or an empty list if operator not found.}{Alpha/Tables.m}{Alpha`Tables`}
\index{lookUpFor}

\alphausage{lookUpForPos}{lookUpForPos[sys, position:\{\_Integer...\}, operator\_Symbol] returns  the position of the nearest occurrence of a specific operator surrounding a specified position, or an empty list if operator not found.}{Alpha/Tables.m}{Alpha`Tables`}
\index{lookUpForPos}

\alphausage{accessByPath}{accessByPath[path:\{\_Integer..\}] extracts the part of  \$result specified by subtree position path.  accessByPath[tree\_, path:\{\_Integer..\}] extracts the part of a tree specified by its subtree position  specifier path}{Alpha/Tables.m}{Alpha`Tables`}
\index{accessByPath}

\alphanote{accessByPath}{Only one position can be supplied at a time, so a single tree is returned if the position is valid (no check is done.)}

\alphausage{undoModif}{undoModif[] called after a program transformation undoes its effects. Returns the program as before the last transformation.}{Alpha/Tables.m}{Alpha`Tables`}
\index{undoModif}

\alphanote{undoModif}{'undoModif' copies the contents of 'Alpha`\$program' into 'Alpha`\$result' so the result of the last transformation (see description of 'Alpha`\$result' and 'Alpha`\$program') is reset to the result of the previous one.}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/ToAlpha0v2.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.668673} by makeDoc.
%
% Section header
\section{The Alpha`ToAlpha0v2` package } 
\label{label:Alpha`ToAlpha0v2`}
\alphausage{ToAlpha0v2}{ToAlpha0v2: Package. Contains functions for the transformation of scheduled programs into Alpha0v2 form: decomposeSTdeps, makeInputMirrorEqus[], toAlpha0v2[]}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{ToAlpha0v2}

\alphausage{needSeparation}{debugUse}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{needSeparation}

\alphausage{toAlpha0v2}{toAlpha0v2[sys\_Alpha`system,options\_List] converts a scheduled program `sys' (see schedule[] and applySchedule[]) to Alpha0v2 form. The program `sys' is assumed to be such that all the local variables have the same dimension (see uniformizeDims[]) and all the dependencies are uniform. This function calls in sequence spaceTimeDecomposition[] makeAllMuxControl[], pipeAllControl[] and decomposeSTdeps[]. See these transformations for further information. WARNING: if there is more than one space index then pipeAllControl[] doesn't work yet. If sys is not present then the transformations apply to \$result and modify it. The option list is optional. }{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{toAlpha0v2}

\alphausage{steps}{steps is an option of toAlpha0v2. It gives a list of numbers, each one corresponding to a step.}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{steps}

\alphausage{makeSimpleExpr}{makeSimpleExpr[sys] renames all the subexpressions which are not simple until the program as no more composed expression (in the sense of the analysis performed by needSeparation). Currently, the expression simplified are only the one involving multiplexer and operators}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{makeSimpleExpr}

\alphausage{decomposeSTdeps}{decomposeSTdeps[sys] splits, in sys, each dependency involving space and time into two dependencies, one on space, the other on time. decomposeSTdeps[] applies to \$result and modifies it.}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{decomposeSTdeps}

\alphausage{makeInputMirrorEqus}{makeInputMirrorEqus[sys\_Alpha`system,options\_List] adds to system `sys' mirror equations fot the inputs. This low-level transformation is needed before the translator to AlpHard is invoked. `sys' is optional, defaults to \$result, and the option list is optional, too (see Options[makeInputMirrorEqus]).}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{makeInputMirrorEqus}

\alphausage{reuseCommonExpr}{reuseCommonExpr[sys\_Alpha`system] attemps to gather expressions used several time by adding local variables for these expressions in system `sys'.  reuseCommonExpr[sys\_Alpha`system,var\_String,expr\_] specifically add   variable `var' for expr `expr' (which s in ast form). Currently only   RHS of equation are scanned up for multiple uses}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{reuseCommonExpr}

\alphausage{integerToBooleanSyst}{integerToBooleanSyst[$<$syst\_Alpha`system$>$] Function, change all * in and and all + in or. Returns the resulting  AST}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{integerToBooleanSyst}

\alphausage{booleanToIntegerSyst}{booleanToInteger[$<$syst\_Alpha`system$>$] Function, change all and in * and all or in +. Returns the resulting  AST}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{booleanToIntegerSyst}

\alphausage{correctIdEqs}{debug}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{correctIdEqs}

\alphausage{splitMax}{splitMax[sys]returns the system sys in which  the Max4 has been replaced by calls to Max operators, splitMax[] applies this to \$result}{Alpha/ToAlpha0v2.m}{Alpha`ToAlpha0v2`}
\index{splitMax}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/VertexSchedule.tex
Added mmalpha (source and binary)

%
% File created on {2009, 4, 25, 18, 19, 40.128578} by makeDoc.
%
% Section header
\section{The Alpha`VertexSchedule` package } 
\label{label:Alpha`VertexSchedule`}
\input{incl-VertexSchedule.tex}
\alphanote{VertexSchedule}{Documentation revised on August 1st, 2004.}

\alphausage{VertexSchedule}{The Alpha`VertexSchedule` package is one of the two scheduling packages
of \MMAlpha{}. Functions of the Alpha`Schedule` package should be preferred,
but in some cases, VertexSchedule offers some interesting functionalities.
Symbols defined in this package are \$dependencyConstraints, 
\$optimalityConstraints, \$scheduleDepTable, \$scheduleLibrary, 
\$scheduleMDSol. Functions are checkSchedOptions, constOf, depComponents,
depCycles, depGraph, DomDimension, DomSingletonQ, DomTrueDimension,
eliminateVar, encodeSchedule, equations, ishow, isIndentityOnDim,
getLinPart, getUseSchedule, loadScheduleLibrary, matrixTransPart, 
polDomainQ, timeMinSchedConstraints, saveScheduleLibrary, scd, slackOf,
sortEquations, zpolDomainQ
}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{VertexSchedule}

\alphausage{\$dependencyConstraints}{\$dependencyConstraints contains the  dependency constraints used for
solving the scheduling. It is a list whose entries correspond directly
to the entries of the deptable. The last element is the set of
dependency variables. This global variable is set when calling the
scheduleConstraints  function}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{\$dependencyConstraints}

\alphausage{\$optimalityConstraints}{\$optimalityConstraints contains the  optimality (total time, etc)
constraints used for solving the scheduling. It is a list, whose
entries are pairs \{variables, constraints\}. This global variable is set
by the timeMinSchedConstraints function}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{\$optimalityConstraints}

\alphausage{\$scheduleDepTable}{when set, \$scheduleDepTable containts the dependency table of \$result. It is
computed by a call to scd.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{\$scheduleDepTable}

\alphausage{\$scheduleLibrary}{\$scheduleLibrary containts the schedules found by the scheduler when
called on the elements of the current library. These schedules are used
when the scheduler is called with the structured option}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{\$scheduleLibrary}

\alphausage{\$scheduleMDSol}{\$scheduleMDSol is an array which gathers information during a call to 
the scd function. \$scheduleMDSol[i] contains a list \{o,c,v,sol\}
consisting  of the objective function, constraints, variables, and
solution to the linear problem solved when dealing with dimension i of
the schedule. For a unidimensional schedule, only level 1 is set. For a
dataflow schedule, level 1 corresponds to  the dataflow leve, and other
levels to the multidimensional schedule of the  inside indexes.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{\$scheduleMDSol}

\alphausage{adaptUses}{adaptUses[] modifies all use statements, by adding extra input 
equations}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{adaptUses}

\alphausage{listOfAdaptedSignals}{listOfAdaptedSignals is an option of adaptUses, False by default. If
True, adaptUses returns a list formed of the equivalence between the
adapted signals and the old signals}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{listOfAdaptedSignals}

\alphausage{addSchedule}{addSchedule is an option of loadScheduleLibrary. Default value is 
True. If True, the schedule contained in file sys.scdlib is added
to \$scheduleLibrary, where it replaces a previous schedule for the
current system.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{addSchedule}

\alphausage{affineDepConsts}{affineDepConsts[ dp, v, dur, taus, alphas, opts ] computes the constraints
due to dependency do, on variable v, with duration dur, coefficients taus and
alpha, and options opts.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{affineDepConsts}

\alphausage{checkSchedOptions}{checkSchedOptions[options] checks the options against the 
options of scd}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{checkSchedOptions}

\alphausage{constOf}{constOf[symbol] gives the value of symbol}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{constOf}

\alphausage{depComponents}{depComponents[] computes the strongly connected components of 
the dependence graph}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{depComponents}

\alphausage{durationsNonZero}{durationsNonZero is an option of scd. It gives the rank in the dependence
table of the dependences the duration of which is non zero. Those durations
are assumed to be 1. A negative rank is counted from backward from the end
of the dependence table.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{durationsNonZero}

\alphausage{depCycles}{depCycles[] computes the elementary cycles in the dependence graph. 
depCycles[ selfDep$ \rightarrow $False ] computes the elementary cycles, 
without the self dependences}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{depCycles}

\alphausage{depGraph}{depGraph[] computes and displays the dependence graph of \$result in
the format of the Combinatorica package. This function assumes that
the dependence graph has already been computed. See also Options[
depGraph ]. Inputs are green, outputs are blue and selfvertices are
red.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{depGraph}

\alphausage{depGraphViz}{depGraphViz[] computes and displays the dependence graph of \$result in
the format of the GraphViz package. This function assumes that
the dependence graph has already been computed. See also Options[
depGraph ]. Inputs are green, outputs are blue and selfvertices are
red. Option displaySchedule allows the schedule to be displayed.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{depGraphViz}

\alphausage{displaySchedule}{displaySchedule is an option of depGraphViz that adds the schedule
of all variables}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{displaySchedule}

\alphausage{showViz}{showViz[] draws a dot file for the alpha program}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{showViz}

\alphausage{variables}{variables is an option of several functions, for example depGraph or
appSched. Default value is all, meaning all variables of a system}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{variables}

\alphausage{extraEdges}{Option of depGraph. Seems to be unused...}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{extraEdges}

\alphausage{variables}{Option of depGraph. Default value is all. Allows one to speficy the list
of variables to which the graph is restricted.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{variables}

\alphausage{factor}{Option of depGraph. Default value is 1. Gives a size factor for arrow heads
and points of the dependence graph.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{factor}

\alphausage{labelSize}{Option of depGraph. Default is 0.1. Size of the labels.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{labelSize}

\alphausage{labelFactor}{Option of depGraph. Default is 1.2. Gives the distances of the labels 
from the center of the graph.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{labelFactor}

\alphausage{DomDimension}{DomDimension[d] gives the dimension of domain d, including the 
parameter dimension (see DomTrueDimension to get the dimension 
with de parameter dimension)}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{DomDimension}

\alphausage{DomSingletonQ}{DomSingletonQ[sys, d] is True if the true dimension of d is 0 and 
if d is the universal domain. DomSingletonQ[d] is equivalent to 
DomSingletonQ[\$result,d]}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{DomSingletonQ}

\alphausage{DomTrueDimension}{DomTrueDimension[sys,d] gives the dimension of domain d less the 
dimension of the parameter space of sys. DomTrueDimension[d] is 
equivalent to DomDimension[\$result,d]}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{DomTrueDimension}

\alphausage{eliminateVar}{eliminateVar[v] removes variable v, after checking that the variable
is useless}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{eliminateVar}

\alphausage{encodeSchedule}{encodeSchedule[sys, params, sol, opts] returns the encoding of the
schedule sol of system sys, with parameters params, using options opts.
Among the options of encodeSchedule, one option allows one to choose
between the Farkas format and the Vertex format}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{encodeSchedule}

\alphausage{equations}{equations is an option value of the option variables of depGraph. If 
True, the variables of are ordered in the order given by the equations, 
with input first.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{equations}

\alphausage{ishow}{ishow[] pretty-prints \$result, without calling the external C 
program. It supports some features of the imperative form (such as
aliases) which ashow does not support.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{ishow}

\alphausage{isIdentityOnDim}{isIdentityOnDim[mat,\{1,3\}] (for instance) 
checks that matrix mat1 is identity on dimension 1 and 3}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{isIdentityOnDim}

\alphausage{getLinPart}{getLinPart[m] computes the mma form of the linear part of the Alpha
matrix m, considered as an affine function. Note: use getLinearPart
instead.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{getLinPart}

\alphausage{getUseSchedule}{getUseSchedule[ x ] obtains the schedule of the use entry of a dependency table x,
by looking in the global variable \$scheduleLibrary.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{getUseSchedule}

\alphausage{loadScheduleLibrary}{loadScheduleLibrary[sys] loads in global variable \$scheduleLibrary 
the content of the file sys.scdlib}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{loadScheduleLibrary}

\alphausage{matrixTransPart}{matrixTransPart[m] computes the mma form of the translation part of
the Alpha matrix m, considered as an affine function.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{matrixTransPart}

\alphausage{parameterRules}{parameterRules is an option of loadScheduleLibrary. Default value
is \{\}. It contains replacement rules for the parameters of the 
schedule which is loaded. For example, if option parameterRules $ \rightarrow $ \{"K"$ \rightarrow $2\} 
is set, then parameter "K" will be replaced by value 2 while reading
the schedule.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{parameterRules}

\alphausage{periodicFactor}{periodicFactor is an option of getUseSchedule. Its default value is 1.The first parameter of the schedule of a subsystem is multiplied byfactor}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{periodicFactor}

\alphausage{saveScheduleLibrary}{saveScheduleLibrary[sys] saves the content of global variable \$scheduleLibrary 
in the file sys.scdlib. saveScheduleLibrary[] saves the content of \$scheduleLibrary in
the file name.scdlib, where name is the name of the system currently contained
in \$result. To save only the schedule of the system in \$result, use 
saveScheduleLibrary[ onlyMainSystem $ \rightarrow $ True ].}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{saveScheduleLibrary}

\alphausage{onlyMainSystem}{onlyMainSystem is an option of saveScheduleLibrary. If True, only 
the schedule of the system in \$result is saved, otherwise (default)
all schedules in \$scheduleLibrary are saved.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{onlyMainSystem}

\alphausage{timeMinSchedConstraints}{Timeminschedconstraints[] computes the constraints needed to minimize
the total execution time of
\$result. timeMinSchedConstraints[var\_String] computes only the
dependence constraints for variable var. Options of
timeMinSchedConstraints are sameLinearPart (default False),
addConstraints (\{\}) and verbose (False).}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{timeMinSchedConstraints}

\alphausage{saveSchedule}{saveSchedule is an option of scd. If True, \$schedule is saved in 
file sys.scd, where sys is the name of the scheduled system.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{saveSchedule}

\alphausage{scd}{***scd[] finds out the schedule that minimizes the total computation time
for a bounded Alpha program. Fails if the system is not bounded. scd can 
also be used to find a multidimensional schedule, or a dataflow schedule.
An option allows a structured schedule to be found.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{scd}

\alphausage{periods}{periods is a parameter of scd. It is the list of periods associated with
the subsystems, whenever these subsystems contain the parameters "\$P".}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{periods}

\alphausage{onlyUseDep}{onlyUseDep is an option of scd that specifies the use dependencies that are
to be kept. Default value is all, otherwise, it is a list of integers.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{onlyUseDep}

\alphausage{simplex}{simplex[o,c,v] finds the minimum of function o on contraints c (list) 
for list of variables v. Variables are not assumed to be linear}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{simplex}

\alphausage{slackOf}{slackOf[listOfConstraints] gives the difference between the lhs and 
the rhs in the constraints, using the solution contained in \$scheduleMDSol[1].
This function allows somehow to explore a given schedule.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{slackOf}

\alphausage{slg}{slg[g] shows graph g. slg[g,Directed] shows g as a directed graph.
This function is a slight modification of ShowLabeledGraph.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{slg}

\alphausage{pal}{}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{pal}

\alphausage{sortEquations}{sortEquations[sys:\_system] returns a the same system where the equation 
have been reordered in such a way that all computation executed 
simultaneously (with duration 0) can be executed in the textual order,
it currently assumes that the time is one dimensionnal and the first indice.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{sortEquations}

\alphausage{zpolDomainQ}{zpolDomainQ[d] is true if d is a domain with z-polyhedra}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{zpolDomainQ}

\alphausage{matrix2mma}{matrix2mma[m] computes the mma form of the Alpha matrix m.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{matrix2mma}

\alphausage{eqsDomain}{}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{eqsDomain}

\alphausage{statScheduleConstraints}{statScheduleConstraints[] does as scheduleConstraints, but provides 
a few statistics}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{statScheduleConstraints}

\alphausage{summaryScheduleConstraints}{summaryScheduleConstraints[] computes the constraints needed to minimize the 
total time, and projects these constraints on the Input and Output variables.}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{summaryScheduleConstraints}

\alphausage{variablesOf}{variablesOf[ exp ] gives the list of variables in the 
list of expression exp}{Alpha/VertexSchedule.m}{Alpha`VertexSchedule`}
\index{variablesOf}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Vhdl2.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.779597} by makeDoc.
%
% Section header
\section{The Alpha`Vhdl2` package } 
\label{label:Alpha`Vhdl2`}
\alphausage{Vhdl2}{Alpha`Vhdl2` contains the Vhdl generator. The main function is a2v.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{Vhdl2}

\alphausage{\$vhdlCurrent}{vhdlCurrent is a global variable owned by Vhdl, which contains the  current program}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{\$vhdlCurrent}

\alphausage{\$vhdlOutputFile}{}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{\$vhdlOutputFile}

\alphausage{a2v}{a2v[] translates in Vhdl all the files contained in \$library.  a2v[lib] translates in Vhdl all the files contained in library lib. a2v[sys] translates in Vhdl the system sys. a2v has many  options (see Options[a2v]). a2v[elem,tinit] runs a2v on library or system elem, and fixes the initial value of the time to the integer value tinit.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{a2v}

\alphausage{bitWidth}{bitWidth[sys,var] gives the bit width of variable var. Default  value of sys is \$result.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{bitWidth}

\alphausage{bitWidthOfExpr}{bitWidth[sys,expr] gives the bit width of an expression. Default of sys is \$result.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{bitWidthOfExpr}

\alphausage{getVhdlType}{getVhdlType[sys,var] return the vhdl type of the element of array var}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{getVhdlType}

\alphausage{showVhdl}{showVhdl[] prints the Vhdl generated for the system in \$result.  showVhdl[s] show the content of the file s.vhd.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{showVhdl}

\alphausage{stim}{stim[] post processes all the simuli files (replace blanks by zeros).  These files are generated from a C code in hexadecimal and there is no  format in C to write zeros in for the first digits.}{Alpha/Vhdl2.m}{Alpha`Vhdl2`}
\index{stim}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/VhdlTestBench.tex
Added mmalpha (source and binary)

%
% File created on {2007, 11, 15, 17, 10, 20.994002} by makeDoc.
%
% Section header
\section{The Alpha`VhdlTestBench` package } 
\label{label:Alpha`VhdlTestBench`}
\alphausage{VhdlTestBanch}{Package,contains definition of funtion for generating test benches for VHDL generated from Alphard}{Alpha/VhdlTestBench.m}{Alpha`VhdlTestBench`}
\index{VhdlTestBanch}

\alphausage{vhdlTestBenchGen}{vhdlTestBenchGen[sys] generates a VHDL test bench for  the VHDL component generated from sys with the a2v command. sys mustbe a valid Alphard system, it works if sys is a cell, a module or a controller (not an interface), as for a2v, the parameters must be assigned. vhdlTestBenchGen[] applies on \$result}{Alpha/VhdlTestBench.m}{Alpha`VhdlTestBench`}
\index{vhdlTestBenchGen}

\alphausage{mkVhdlLoop}{test}{Alpha/VhdlTestBench.m}{Alpha`VhdlTestBench`}
\index{mkVhdlLoop}

\alphausage{mkVhdlLowerBound}{test}{Alpha/VhdlTestBench.m}{Alpha`VhdlTestBench`}
\index{mkVhdlLowerBound}

\alphausage{mkVhdlUpperBound}{test}{Alpha/VhdlTestBench.m}{Alpha`VhdlTestBench`}
\index{mkVhdlUpperBound}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Visual.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.838607} by makeDoc.
%
% Section header
\section{The Alpha`Visual` package } 
\label{label:Alpha`Visual`}
\alphanote{Visual}{Documentation revised on August 10, 2004}

\alphausage{Visual}{The Alpha`Visual` packages contains a few functions to visualize 1D or 2D domains. The main function is showDomain. Function getBoundingBox is also used in some other packages, as well as function bbDomain,  which is poorly written.}{Alpha/Visual.m}{Alpha`Visual`}
\index{Visual}

\alphausage{bbDomain}{bbDomain[d] returns a list representing the bounding box of domain d. This has the form \{\{xmin,xmax\},\{ymin,ymax\},\{xneg,xpos\}, \{yneg,ypos\}\} where \{xmin,xmax\},\{ymin,ymax\} is the bounding box of the vertices, \{xneg,xpos\}, \{yneg,ypos\} indicate that the domain is infinite in one of these directions. If xneg and xpos are 0, no rays. If xneg is 1, negative ray, if yneg is 1, positive ray.  Same for yneg and ypos}{Alpha/Visual.m}{Alpha`Visual`}
\index{bbDomain}

\alphausage{getBoundingBox}{getBoundingBox[dom] gives the bounds of the smallest rectangle containing dom. This function is temporary, in particular it does not handle domains that are union of polyhedra (in fact it work only on convex polyhedra. It should be reimplemented using DomProject).}{Alpha/Visual.m}{Alpha`Visual`}
\index{getBoundingBox}

\alphausage{showDomain}{showDomain[d] displays a plot of any 1D or 2D domain or any list of 1D or 2D domains. showDomain[d, name] displays d with the title name specified.}{Alpha/Visual.m}{Alpha`Visual`}
\index{showDomain}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Visual3D.tex
Added mmalpha (source and binary)

%
% File created on {2008, 3, 11, 16, 16, 25.845024} by makeDoc.
%
% Section header
\section{The Alpha`Visual3D` package } 
\label{label:Alpha`Visual3D`}
\alphanote{Visual3D}{Documentation revised on August 10, 2004}

\alphausage{Visual3D}{The Alpha`Visual3D` package contains a few functions to visualize and animate 3D domains. The main function is vshow, other functions  are facets, and listPlanes. Some functions are exported only for debugging purposes.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{Visual3D}

\alphausage{facets}{facets[dom] computes the list of polygons corresponding to the domain dom.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{facets}

\alphausage{listPlanes}{listPlanes[l,p] returns the pair \{l,lp\} where lp is the lists of planes in p which contain point l.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{listPlanes}

\alphausage{maxR}{maxR is an option of vshow. It allows the viewpoint of the final picture to  be set. See stepR for more details.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{maxR}

\alphausage{minR}{minR is an option of vshow. It allows the r parameter of the viewpoint  to be changed. By default, minR is 1. See also stepR and maxR, and ViewPoint of Mathematica.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{minR}

\alphausage{orderPolygon}{only here for debugging purposes.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{orderPolygon}

\alphausage{stepR}{stepR is an option of vshow. It allows the step of the r parameter of the  viewpoint to be changed, when one wants the domains to be animated.  By default, stepR is 1. Combined with minR and maxR, it allows a sequence of pictures with viewpoint between minR and maxR, by steps stepR to be  drawn.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{stepR}

\alphausage{threeDDomainQ}{only here for debugging purposes.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{threeDDomainQ}

\alphausage{twoDDomainQ}{only here for debugging purposes.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{twoDDomainQ}

\alphausage{vp1}{vp1 is an option of vshow, allowing the first parameter of ViewPoint to be changed. Default value is 3.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{vp1}

\alphausage{vp3}{vp3 is an option of vshow, allowing the third parameter of ViewPoint to  be changed. Default value is 1.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{vp3}

\alphausage{vshow}{vshow[d] shows a 2D or 3D graphic picture of the 3D domain d.  vshow[var] shows the domain of variable var in \$result (see Options[vshow] for more details).}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{vshow}

\alphausage{units}{only here for debugging purposes.}{Alpha/Visual3D.m}{Alpha`Visual3D`}
\index{units}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/Zpol.tex
Added mmalpha (source and binary)

%
% File created on {2009, 4, 25, 17, 47, 46.104429} by makeDoc.
%
% Section header
\section{The Alpha`Zpol` package } 
\label{label:Alpha`Zpol`}
\alphanote{Zpol}{Documentation revised on August 10, 2004}

\alphausage{Zpol}{The Alpha`Zpol` package contains additional functions for computing
with Z-polyhedra. These functions are expressLatticeWithMat, 
expressZpolWithMat, getMatOfZpol, and getPolOfZpol.}{Alpha/Zpol.m}{Alpha`Zpol`}
\index{Zpol}

\alphanote{Zpol}{The internal representation of a Zpol is:
domain[dim\_Integer, \{\_\_\_String\},
\{\_\_\_zpol[m\_matrix,\{\_\_\_pol\}]\}] (which is m(p)).}

\alphausage{expressLatticeWithMat}{expressLatticeWithMat[sys,m] returns a system where all Z-polyhedra 
using lattices corresponding to the matrix m are rewritten so as 
to appear with matrix m as lattice. Default value of sys is \$result.
expressLatticeWithMat[sys,m,pos] modifies only the 
Z-domain at position pos of the Alpha abstract syntax tree.}{Alpha/Zpol.m}{Alpha`Zpol`}
\index{expressLatticeWithMat}

\alphausage{expressZpolWithMat}{expressZpolWithMat[z,M] attempts to change the representation    
of the Z-polyhedron z. Assuming z = N(P), we expect a result of the 
form w = M(P') where M is the matrix given as second argument, hence 
it returns M(P') where P'=M\^\{-1\} N(P). If M and N do not represent 
the same lattice, it gives a warning and return z unchanged. 
If z is a union of Z-polyhedra, it fails.}{Alpha/Zpol.m}{Alpha`Zpol`}
\index{expressZpolWithMat}

\alphausage{getMatOfZpol}{getMatOfZpol[z] returns the matrix used for identifying the 
lattice of a Z-polyhedron z. If z=M(P), it returns M. If Z is already a 
polyhedron, it returns the identity matrix. If z is a union of 
Z-polyhedra, it fails.}{Alpha/Zpol.m}{Alpha`Zpol`}
\index{getMatOfZpol}

\alphausage{getPolOfZpol}{getPolOfZpol[z] returns the polyhedron used as source for generating 
a Z-polyhedron z. If z=M(P), it returns P. If z is already 
a polyhedron, it returns z. If z is a union of Z-polyhedra, it fails.}{Alpha/Zpol.m}{Alpha`Zpol`}
\index{getPolOfZpol}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/incl-Alpha.tex
Added mmalpha (source and binary)

% 
\subsubsection{Introduction}
The \texttt{Alpha.m} package is the main package of Mathematica.
It is located in the directory \texttt{\$MMALPHA/lib/Packages}.
It contains the definition of the main symbols and data structures
of \MMAlpha{}.

A special word about the \texttt{exportAlphaFunctions} function.
This function takes as a parameter the autoloadFile. This file
is located in 
\begin{verbatim}
$MMAlpha/lib/Packages/Alpha/autoload.m
\end{verbatim}
%$
and is created automatically by the \texttt{exportAlphaFunctions}.

At the end of the \texttt{Alpha.m}, a few instructions are executed
when loading this package. If the \texttt{autoloadFile} does not
exists, \texttt{exportAlphaFunctions} is called and creates it.

The function \texttt{exportAlphaFunctions} uses a local variable
\texttt{contexts} of \texttt{Alpha.m} which contains the list of 
the packages that need to be loaded when using \MMAlpha{}. This variable
has to be changed when a new package is added. For each package
of \texttt{contexts}, \texttt{exportAlphaFunctions} creates an 
entry in the autoload file that declares the corresponding package
using the \texttt{DeclarePackage} function of Mathematica. In this
way, all packages are declared, and when a symbol of the package
is evaluated for the first time, 
the corresponding package is automatically loaded.

The remaining of this chapter describes the functions of the 
Alpha package.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/incl-MakeDoc.tex
Added mmalpha (source and binary)

\subsection{Note regarding the documentation of \MMAlfa{}}
\label{doc-makedoc}
The \MMAlfa{} software is (partially) documented in various
ways.\footnote{This chapter has been
proof-read by Patrice Quinton on January 8, 2007.}
\begin{itemize}
\item A \texttt{html} documentation is included in the distribution.
It can be accessed from the \texttt{welcome.html} file in the 
\texttt{\$MMalpha{}} directory. 
\item A {\em getting started} document is available in form of
a \texttt{pdf}
file. It is accessible from the initial \texttt{html} welcome
page and is located in:
\begin{verbatim} 
$MMAlpha/doc/QuickStart/AlphaStart.pdf
\end{verbatim} 
This document is the most up-to-date tutorial for \MMAlpha{}. It does not
describe all features of \MMAlpha{}, but normally, allows a user to
discover the main features of the software.
\item A reference manual is being developped in:
\begin{verbatim} 
$MMAlpha/doc/ReferenceManual
\end{verbatim}
and it is described hereafter.
\item Other documentation files exist, but they are not guaranteed to 
be up-to-date.
\end{itemize}

The reference manual of \MMALPHA{} is obtained through the use 
of commands in the \texttt{Makedoc} package. 
It is produced by executing the program called \texttt{genReferenceManual.m}
which is situated in the same directory.

This program contains a set of calls to the function \texttt{doDoc}
of the \texttt{Makedoc} package (see \texttt{?doDoc}). Each one of
these calls reads a package -- for example, \texttt{Alpha/Makedoc.m} --
whose name is relative to directory
\begin{verbatim} 
$MMALPHA/lib/Packages
\end{verbatim}
and translates it into a \LaTeX{} file -- for example, \texttt{Makedoc.tex}, --
in target directory \texttt{\$MMALPHA/doc/ReferenceManual}.

The reference manual itself is edited manually. It is in the 
file:
\begin{verbatim}
referenceManual.tex
\end{verbatim}
of the same directory. It contains
basically input statements for the documentation files created by 
\texttt{doDoc}. This function (which is documented hereafter), has
several options: one can either generate a standalone \LaTeX{} file, or
an input file (this helps for editing the documentation), one can 
specify a list of functions that are only to be documented, one can 
specify where the file has to be put, etc.

The reference manual is based on the usage and note statements which are
contained in a given package. In there exists a file named 
\texttt{incl-textFile} in the target directory, then an include 
statement for this file is added in the generated \LaTeX{} file.
For example, the documentation that you are reading is in the 
file \texttt{incl-Makedoc.text} file, in directory:
\begin{verbatim}
$MMALPHA/doc/referenceManual
\end{verbatim}
Therefore, when the 
\texttt{Makedoc.m} file is processed by the corresponding 
statement in the \texttt{genReferenceManual.m} program, i.e.:
\begin{small}
\begin{verbatim}
doDoc["Alpha/MakeDoc.m", "MakeDoc.tex", 
  targetDir -> Environment["MMALPHA"] <> "/doc/ReferenceManual"];
\end{verbatim}
\end{small}
an input statement is included in the \texttt{Makedoc.tex} file 
which is generated.

The \texttt{doDoc} function tries to modify the content of the
usage and note statements so that special symbols are correctly
output in \LaTeX{}. These transformations are described in the
variable \texttt{repRules} of the package, and may be modified. 
Index statements are generated for each usage statement.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/ReferenceManual/referenceManual.tex
Added mmalpha (source and binary)

\documentclass[11pt]{report}
\usepackage{verbatim,version,epic,eepic}
\usepackage{makeidx}
\makeindex

\newcommand{\MMAlpha}{{\sc mmalpha}}
\newcommand{\MMALPHA}{{\sc mmalpha}}

\newcommand{\MMAlfa}{{\sc mmalpha}}
\newcommand{\Alfa}{{\sc alpha}}
\newcommand{\mma}{{\sc Mathematica}}
\newcommand{\alfa}{{\sc alpha}}
\newcommand{\myquotes}{"} 
\newcommand{\myitem}{\\}

\newcommand{\alphausage} [4]
{\subsection{#1} \noindent
{\rm #2}\\{\bf Defined in file:} #3\\{\bf Package:} #4}

\newcommand{\alphanote} [2]
{~\\\noindent{\footnotesize #2}}

\newcommand{\packageusage} [3]
{\subsection{#1} 
{\rm #2}}

\begin{document} 
\pagestyle{headings}
\thispagestyle{empty}
\title{MMAlpha Version 2.0 -- Reference Manual\thanks{Last update of
\today{}}} 
\author{Api/Cosi/R2D2/Compsys/Cairn}

\date{March 2008}

\maketitle

\tableofcontents

\newpage 
\chapter{Introduction}
This document is the reference Manual of \MMAlpha{}. It concerns
Version 2.0 of \MMAlpha{}, but is still under construction.
It has been
(almost) automatically produced from the usage and note statements
of the \MMAlpha{} packages. In Chapter~\ref{basicfunctions}, we present
the main package of \MMAlpha{}. 
Chapter~\ref{domainfunctions} concerns
functions on domains. In Chapter~\ref{staticfunctions}, functions 
related to the static analysis of \alfa{} programs are described.
etc.

This manual is part of the general documentation of \MMAlpha{}.
This documentation can be accessed through the html hierarchy 
that follows the \MMAlpha{} hierarchy. The origin of this 
hierarchy is
\begin{verbatim}
$MMALPHA/doc/index.html
\end{verbatim}
%$
\section*{Some conventions}
Many functions of \MMAlpha{} operate on the Abstract Syntax Tree (AST) of
an \alfa{} program and return the new transformed AST. For example, 
evaluating \texttt{normalize[sys]} returns a normalized version of 
the \texttt{sys}, which may be any \mma{} expression whose evaluation
is the AST of an \Alfa{} program. The result of this transformation may
be stored in another variable, e.g.:
\begin{verbatim}
x = normalize[sys];
\end{verbatim}
and even manipulated by a \mma{} function.

However, \MMAlpha{} provides another, implicit way of manipulating
an \Alfa{} program. When an Alfa{} program is parsed and loaded
using the \texttt{load} function (see \texttt{?load}), 
it becomes the {\em current \alfa{} 
program} and is stored in a variable 
named \$result. By default, functions operate on \$result and 
modify \$result, in addition to returning the transformed program.
For example, 
\begin{verbatim}
normalize[];
\end{verbatim}
puts in \$result the normalized version of \$result, and 
returns this value. (This is the reason of the \texttt{;} ending
this command, otherwise, \mma{} would print out the AST on the notebook.)
Before modifying \$result, this variable is stored in another variable
named \$program, so that the value of the current AST before 
the normalization is kept. The \texttt{undo[]} command allows 
\$result to be restored to its previous value.

\section{Content of the Reference Manual}
The reference manual is structured in the following way.
\begin{enumerate}
\item The mail \LaTeX{} file, \texttt{referenceManual.tex}, is located in 
directory:
\begin{verbatim}
../doc/RefefenceManual
\end{verbatim}
It contains an introduction (that you are currently reading) and calls to 
the various documentation packages, which are produced using commands
of the \texttt{Makedoc} package (see~\ref{doc-makedoc} for details.
\item The reference manual contains 
the following
Chapters.
\begin{itemize}
\item Chapter~\ref{basicfunctions} presents the functions contained in the \texttt{Alpha} 
package. 
\item Chapter~\ref{domainfunctions} presents the functions contained in the \texttt{Domlib},
\texttt{Zpol},
\texttt{Visual}, and \texttt{Visual3D}
packages. 
\item Chapter~\ref{staticfunctions} presents the functions contained in the \texttt{Static},
and \texttt{Semantics}
packages. 
\item Chapter~\ref{subsystemfunctions} presents the functions contained in the \texttt{SubSystems},
package. 
\item Chapter~\ref{manipfunctions} presents the functions contained in the \texttt{Matrix},
and \texttt{Tables}
packages. 
\item Chapter~\ref{elementaryfunctions} presents the functions contained in the \texttt{ChangeOfBasis},
\texttt{Cut},
\texttt{Normalization}, 
\texttt{Reduction}, 
\texttt{Substitution}, 
and \texttt{Substitution}
packages. {\em This chapter has not been proof-read.}
\item Chapter~\ref{scheduling} presents the functions contained in the \texttt{ScheduleTools}, 
\texttt{VertexSchedule}, 
and \texttt{FarkasSchedule}
packages. {\em This chapter has not been proof-read.}
\item Chapter~\ref{elementaryfunctions} presents the functions contained in the \texttt{Pipeline},
\texttt{Control},
and \texttt{PipeControl}
packages. {\em This chapter has not been proof-read.}
\item Chapter~\ref{uniformization} presents the functions contained in the \texttt{Pipeline},
\texttt{Control},
and \texttt{PipeControl}
packages. 
\item Chapter~\ref{backend} presents the functions contained in the \texttt{ToAlpha0v2},
\texttt{Alphard},
\texttt{Vhdl2},
and \texttt{VhdlTestBench}
packages. 
\item Chapter~\ref{utilities} presents the functions contained in the \texttt{MakeDoc},
package. 
\end{itemize}
\item The status of the documentation contained in each package 
is described in a note at the beginning of the package: when the documentation has been
revised, it appears in a note following the subsection of the package. A revision
means that I checked that the content of the usage and note statements of the
package appear correctly on the reference manual, but does not mean that the
documentation itself is correct.
\end{enumerate}
\newpage
\section*{Acknowledgments (Patrice Quinton)}
This version of \MMAlpha{} was developed over the years by the contribution of
many people. With the risk of forgetting some contributors (sorry for this), 
here is a list of these contributors:
\begin{itemize}
\item Christophe Mauras invented the \Alfa{} language in his PhD thesis
en 1989, and the actual language is almost its initial version.
\item The late Herv\'e Le Verge designed with Christophe the first 
version of \Alfa{} (called \Alfa{} du Centaur, as it was designed 
using the Centaur system of Inria.) Herv\'e also designed the first
version of the Polylib which is the engine of \MMAlfa{}.
\item Doran Wilde designed the first version of \MMAlfa{}, i.e., 
the initial \mma{} version of our tool. It started in 1989. 
\item Zbignew Chamski participated to the development of 
this first version. 
\item Tanguy Risset was, and still is, the main architect and 
contributor of \MMAlfa{}. 
\item Florent Dupont de Dinechin implemented a significant part of 
\MMAlfa{} (subsystems, semantic analysis, and part of the back end 
process).
\item Patricia Le Moenner implemented the first version of the 
Vhdl translator.
\item Anne-Claire Guillou contributed to the current Vhdl translater, 
and designed several circuits.
\item Sanjay Rajopadhye was a deep inspirator of \MMAlfa{}.
\item Fabien Quiller\'e implemented mainly the C code generator.
\item Several indian students contributed to the development of
\MMAlfa{} during internships.
\end{itemize}

I was often asked why \MMAlfa{} was developed using \mma{}. The 
choice was made in 1989, mainly in order to take advantage of the 
available formal calculations embedded in this tool. There are
disadvantages to this choice, one being that people have
to buy this software in order to use \MMAlfa{}. I believe that this
is not the main point, however, since \mma{} is available in many
universities, and companies interested in our software could 
buy it easily. Probably more important is that \mma{} does not fit in 
the current {\em culture} of developers.

But from the point of view of a prototype developper, \mma{} is 
a very nice tool. Our software is available on {\em all platforms}
thanks to the portability of \mma{}\footnote{By the way, the 
part of \MMAlfa{} that creates most portability problems is 
the Domlib library that is written in C.}. I have never experienced 
any problem of running out of space, and time constraints have 
never been a significant issue. Moreover, every 2 years or so, 
\MMAlfa{} accelerates significantly when I change my laptop\footnote{The only
occasion when this was not true is when I recently changed my Apple MacBook 
into a MacBook Pro, where I was disappointed to see the performance
degrade. But it may be only because I did not recompile Domlib...}.

\chapter{Basic Functions}
\label{basicfunctions}

\section{Introduction}
Basic functions are mainly in the Alpha package.
\input{Alpha}

\chapter{Functions on domains}
\label{domainfunctions}
\input{Domlib}%Proof-read
\input{Zpol}%Proof-read
\input{Visual}%Proof-read
\input{Visual3D}%Proof-read

\chapter{Static Analysis}
\label{staticfunctions}
\input{Static}
\input{Semantics}

\chapter{Subsystems}
This package contains functions to handle Alpha subsystems.
\label{subsystemfunctions}
\input{SubSystems}

\chapter{Manipulating Alpha Expressions}
\label{manipfunctions}
The Matrix and Tables packages contain some functions to manipulate
Alpha expressions and programs. Other such functions are scattered
in other packages, as they have been written sometimes on the fly
by the authors of these packages. A developper has to read this
chapter carefully in order not to reinvent the wheel.
\input{Matrix}
\input{Tables}

\chapter{Elementary Program Transformations}
\label{elementaryfunctions}
\input{ChangeOfBasis} %Proof-read
\input{Cut} %Proof-read
\input{Normalization} %Proof-read
\input{Reduction} %Proof-read
\input{Substitution}

\chapter{Scheduling}
\label{scheduling}
\input{ScheduleTools}
\input{Schedule}
\input{VertexSchedule}
\input{FarkasSchedule}

\chapter{Uniformization}
\label{uniformization}
\input{Pipeline}
\input{Control}
\input{PipeControl}

\chapter{Back end process}
\label{backend}
\input{ToAlpha0v2}
\input{Alphard}
\input{Vhdl2}
\input{VhdlTestBench}

\chapter{Utilities}
\label{utilities}
\input{MakeDoc}%Proof-read

\chapter{Add-ons}
\input{BinExpansion}
\input{BitOperators}
\input{Decomposition}
\input{DesignManagement}
\input{INorm}
\input{INormalize}
\input{Lexicographic}
\input{LinearAlg}
\input{Matrix}
\input{Meta}
\input{ModularSchedule}
\input{Omega}
\input{Options}
\input{Properties}
\input{Reduction}
\input{Schematics}
\input{Semantics}
\input{Static}
\input{SubSystems}
\input{SystemCTestBench}
\input{SystemCTools}
\input{Tables}
\input{Transformation}
\input{Uniformization}
\input{UniformizationTools}
\input{Visual}
\input{Visual3D}
\input{Zpol}

\printindex

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Schedule/Scheduler_dev_man.tex
Added mmalpha (source and binary)

\documentstyle[fullpage]{article}

\newcommand{\aalpha}{{\sc Alpha}}
\newcommand{\ammalpha}{{\sc MmAlpha}}
\newcommand{\err}{\top}
\newcommand{\indef}{\bot}
\newtheorem{theorem}{theorem}

\title{reference manual for the Alpha {\bf Scheduler}}
\author{Tanguy Risset}
\date{\today}

\begin{document}
%\bibliographystyle{plain}

\maketitle

This documentation provides help for the user of the {\aalpha} 
scheduler developer. It is available in the file 
{\tt \$MMALPHA/doc/developper/Scheduler\_dev\_man.ps}

\section{Introduction}

 The basic goal of the scheduler 
is to find a valid execution order that is {\em good} with respect to 
a particular criterion. The theoretical basis for the scheduling process 
is inherited from the systolic array synthesis researches and from the 
automatic parallization researches. Our  implementation makes use of the 
procedure defined in~\cite{Feautrier92aa}. Rouglhy speaking, 
the time is considered as a discrete  
single clock. the idea of the scheduling process is to 
gather all the constraints that must verify the schedule in a 
 linear programming problem (LP) and to solve it with a particular 
software.

This document is not intended to be a comprehensive explanation of the
implementation, it is an introduction to the overall architecture of
the package {\tt NewSchedule.m} which is quite big (about 3000 lignes).
The developper should carefully read this before getting into the
code.  The source code of {\tt NewSchedule.m} is commented in such a way
that the present document document should be enough to understand
everything.

\section{theoretical basis}
\label{theoric}
The schedule is based on the method proposed in~\cite{Feautrier92aa}
and uses the farkas lemma. We give her the
form of this lemma that we use for ensuring constraints
\begin{theorem}[Affine form of Farkas lemma]
\label{affineFeautrier}
Let ${\cal D}$ be a polyhedron defined by the inequalities $Ax + b \geq 0 $,
Then the affine form $\phi$ is non-negative on ${\cal D}$ if and only
if there exists $\lambda_0,\Lambda \geq 0$ such that $\phi=\Lambda
\left(\begin{array}{cc} A & b \end{array}\right) +\lambda_0$
\end{theorem}
{\bf Proof:} See~\cite{Schrijver86} .

It is easy to see how we will make use of this lemma. To ensure a linear 
constraint (which can always be written as the fact that a linear function is positive on a domain), we will try to find $\Lambda,\lambda_0$ such that 
the above equality is verified (we transformed a $\forall x \in ...$ in 
a $\exists y\geq 0 s.t....$).  The method proposed 
in~\cite{Feautrier92aa} distinguish
three type of constraints that a  scheduling function $T$ must ensure. 
\begin{itemize}
\item {\bf Positivity constraints}: for each variable $A$ of the program, for any 
point $(x,N)$ in the domain of $A$, we must have $T_A(x,N) \geq 0$.
\item {\bf Dependency constraints}: for any 
dependence $A[x,N] = F(\ldots,B[I(x,N)], \ldots)$, for each point $(x,N)$ 
where this dependence occurs, we must have: 
$T_A(x,N) - T_B(I(x,N)) - \delta \geq 0 $ 
(here $\delta$ is a delai associated to 
the computation of $A$.
\item {\bf Objective function constraints}: These constraints depend on the type of 
objective function choosen. If we chose to minimiz
e the overall execution time, 
we introduced a function $f(N)$ depending only upon the parameters of
 the program, which will be greater than any time scheduled (i.e. for each variable $A$,  for any 
point $(x,N)$ in the domains of $A$, we must have 
$f(N) - T_A(x,N) \geq 0$).
\end{itemize}
As index functions are affine functions
 in \aalpha, all these constraints are of 
the required form: ensure that an affine function is positive on a particular 
domain, thus we can use  the Farkas lemma. 
For each  constraints, we will use the formulae of 
theorem~\ref{affineFeautrier} and 
 obtain a system of equalities that will be solve with mppip.
\begin{itemize}
\item Positivity constraints: if a variable $A$ is defined on the domain
$D_A=\{x ~|~ Cx + b \geq 0 \}$ ($x$ has $n$ components and $C$ has 
$k$ rows), the scheduling function 
$T_A(x) = Tx+ t$ is positive on $D_A$ if and only if the system: \\
\centerline{
$\left\{\begin{array}{l} (T\ t) - \Lambda.(C\ b) - \lambda_0 = 0   \\
\lambda_i \geq 0, 0 \leq i \leq k\end{array}\right.$
}
has a solution (where $\Lambda$ is a $k$-vectors whose $i^{th}$ component 
is $\lambda_i$). Hence we get $n+1$ equation: 
\begin{itemize}
\item for each dimension $i$, if $C_{\cdot i}$ is the $i^{th}$ 
column of C:  $\tau_i   - \Lambda.C_{\cdot i} = 0$
\item $t -  \Lambda.b - \lambda_0 = 0$
\end{itemize}

\item dependency constraints: $A[x]$ depends upon $B[I(x)]$ where 
$I(x) = Dx+d$ on the domain $D_{AB} = \{x ~|~ Cx + b \geq 0 \}$
($x$ has $n$ components and $C$ has $k$ rows) the scheduling functions 
$T_A(x) = T_Ax+ t_A$ and  $T_B(x) = T_Bx+ t_B$ respect the dependency 
if an only if the system:\\
\centerline{
$\left\{\begin{array}{l} (T_A\ t_A) - (T_B \ t_B).(D \ d) -  \delta_{BA} - M.(C\ b) - \mu_0 = 0   \\
\mu_i \geq 0, 0 \leq i \leq k\end{array}\right.$
}
has a solution (where $M$ is a $k$-vectors whose $i^{th}$ component 
is $\mu_i$, $\delta_{BA}$ is the delai necessary between the production of $B$ and the production of $A$). Hence we get $n+1$ equation: 
\begin{itemize}
\item  for each dimension $i$,  $\tau_{A,i}   - (T_B \ t_B).D_{\cdot i} - 
M.C_{\cdot i} = 0$
\item $t_A -  (T_B \ t_B).d -\delta_{BA}- M.b - \mu_0 = 0$
\end{itemize}

\item  Objective function constraints: 
if a variable $A$ is defined on the domain
$D_A$ with $D_A=\{(x,N) ~|~ A\left(\begin{array}{c} x\\ N \end{array}\right) + b \geq 0 \}$ ($x$ has $n$ components, $N$ is the parameter vectors has $m$ component and $A$ has 
$k$ rows), the scheduling function 
$T_A(x) = T_A\left(\begin{array}{c} x\\ N \end{array}\right) + t_A$ 
lower than the objective function $\Phi.N + \phi_0$ on  $D_A$ 
if and only if the system: \\
\centerline{$\left\{\begin{array}{l} (0 \ \Phi \ \phi_0) - (T_A\ t_A) - R.(A\ b) - \rho_0 = 0   \\
\rho_i \geq 0, 0 \leq i \leq k\end{array}\right.$
}
has a solution (where $R$ is a $k$-vectors whose $i^{th}$ component 
is $\rho_i$).  Hence we get $n+m+1$ equation: 
\begin{itemize}
\item for each $i, \ 1 \leq i \leq n$:  $-\tau_{A,i}   - R.A_{\cdot i} = 0$
\item for each $i, \  n+1 \leq i \leq n+m$: $\phi_{i-n} -  \tau_{A,i}   - R.A_{\cdot i} = 0$
\item $\phi_0 - t_A - R.b - \rho_0 = 0$
\end{itemize}
\end{itemize}

\section{Algorithm of the schedule function}

The basic structure of the fonction schedule is the following:
\begin{enumerate}
\item Perform various analysis (get the list of variable, domains for each 
 variables,  dependencies, duration for each instruction \ldots)
\item Build the list of constraints of the linear program (three types 
of constraints: constraints ensuring positivity for timing function,
constraints  ensuring dependancies, constraints for the objective function).
\item Write the LP in a unix file in {\tt Pip}. Solve the LP with 
the software  {\tt mppip}, and get the result.
\item Print the result and get the final output form.
\end{enumerate}

\section{Step 1: various analysis}
\begin{itemize}
\item {\tt addAllParameterDomain}. The syntax of parameter in Alpha allow 
some context constraints that are implicit in all domains of the \aalpha 
program.  
If we want to apply Farkas lemma, we need all the constraints of the domain, 
hence, we add {\em by hand} these constraints to all domains of the program.

\item {\tt domains of variables}. We have choosen to distinguish three
types of variables: input variables, local variables and Output
variables. For each variable, the domain is convexise, i.e. we take
the convex hull of the real domain of the variable (this allow to
simplify the resulting LP, but imply a less powerfull scheduler).

\item {\tt dependencies}. The dependence analysis is first done 
by the {\tt dep[] } function. Then,  the list of dependancies are 
classified according to the type of dependancy (6 type of dependency, in this 
order: Local variable 
to Output variable, Input to Output, Output to Local,  Input to Local,
Local to Local, Output to Output).

\item {\tt makeNumInstr}. This function numbers all variable. The number 
of a variable is its order of declaration in the Alpha program. 
This function build a list of couple: {\tt  \{name\_String,number\_Integer\}}
 that will be used 
in the rest of the program to make the correspondance between
 the name of a variable and its number.

\item {\tt makeRefline}. At this point of the program, we are able to
know exactly how many variables will have the linear program generated
(how many $\lambda$'s, $\rho$'s, etc...). As we will see below, we
will first build a constraint matrix in which each row has the same
structure.  This structure is indicated by the {\tt refline}
constructed here.  {\tt refline} contains 8 sublist.  The first one
indicates how many coefficient there are in the objective function
$\Phi$. The second one indicates, for each variable $A$, the dimension
of the domain of $A$ (hence the number of $\tau_A$'s). The third one
indicates how many variables will be dedicated to the constant part of
the scheduling functions (the $\alpha$'s). The fourth indicates, for
each dependancy, how many farkas coefficient ($mu$) will be introduced
by the equation needed to ensure this dependancy. The fifth indicates,
for each variable, how many farkas coefficient ($\lambda$)
 will be necessary to 
ensure the positivity condition for this variable, The sixth indicates, 
for each variable, how many farkas coefficient ($\rho$)
 will be necessary for the objective function  constraints. the $7^{th}$ and 
$8^{th}$ elements are set to {\tt \{1\}} and will be explained later.

As we have classified the variables in three categories and
the dependancies in 6 categories, elements $2,3,5,6$ of refline are
divised in three sublist and element $4$ of refline is divised in 6
sublist.  The precise structure of {\tt refline } is clearly commented
in the {\tt NewSchedule.m } package.  For the moment, we just  need to 
know that this variable will be used in the rest of the program for 
getting information on the program (for instance: how many input variable, 
how many $\lambda$'s related to variable $A$ etc.).

\end{itemize}

\section{Step 2: building the list of constraint}
the list of constraints is built by the function {\tt makeTableEq}.
As we have explained in section~\ref{theoric}, there are three types
of constraints. The positivity constraints are set for the Input,
Output and Local variables, but we impose the objective function
constraint only on the output variables. the dependency constraints
are built for the six type of dependencies. The constraints are the
elements of the {\tt structTableEq}. Each constraint is a list which
have the following structure: \\
\begin{verbatim}
{ PhiCoeffs_IntList,
{ TauLocal_IntListList,TauOutput_IntListList,TauInput_IntListList},
{ AlphaLocal_IntList,AlphaOutput_IntList,AlphaInput_IntList},
{ LambdaLocToOut_IntListList,  LambdaInToOut_IntListList,
  LambdaOutToLoc_IntListList,  LambdaInToLoc_IntListList,
  LambdaLocToLoc_IntListList,  LambdaOutToOut_IntListList},
{ MuLoc_IntListList,  MuOut_IntListList,  MuIn_IntListList },
{ RhoLoc_IntListList,  RhoOut_IntListList,  RhoIn_IntListList },
{ CstCoeff_Int },
{ EqOrIneq_Int} }
\end{verbatim}
In the above structure, {\tt \_IntList} indicates that the element considered
is a list of integer (and {\tt \_IntListList}, a list of list of integer). 
In a constraints {\tt c1} of this type, 
{\tt PhiCoeffs} is the list of factors of the 
coefficients of the objective function $\Phi$ in {\tt c1}.
 One element of {\tt TauLocal} correspond to a local variable (say $A$) 
and gives the factor (in c1)  of the components 
of the scheduling vectors of $A$. 
One element of {\tt LambdaLocToOut} corresponds to a particular 
dependency from a local variable to
 an output variable and gives the coefficients (in c1)
 of the  farkas coefficient $\lambda$ introduced for this dependency,
 and so on ....

For instance, if $A$ is the first local 
variable of the Alpha program which has only one parameter 
 (the domain of $A$ has dimension 3),
 and we want to impose the constraint: 
$\phi_1 - 2\tau_{A,1} = 0 $, this constraint we be stored in one row as:
\begin{verbatim}
{ {1,0},{{{-2,0,0},{0.... (* structured list of  0 in the remaining *) ,{0}}
\end{verbatim}
and the real constraint can be obtained by 
$expr1 = (1\ 0).\left(\begin{array}{c} \phi_1 \\ \phi_0 \end{array}\right)+
 (-2\ 0 \ 0).\left(\begin{array}{c} \tau_{A,1} \\ 
\tau_{A,2} \\ \tau_{A,3}  \end{array}\right)+ \ldots  = 0 $.

In {\tt NewSchedule.m}  such structured  constraints are always assigned to 
variables which are called {\tt line0} or {\tt line1} (like the lines of the 
constraint matrix). Usually, {\tt line0} consists of a structured constraints
corresponding to 0=0 (build from the informations contained in {\tt refline}),
 then it is assigned to {\tt line1}, and
 the  components of {\tt line1} are iteratively modified to obtain a 
valid constraint as indicated in section~\ref{theoric}.
 The $7^th$  element of  line1 is a list of one single integer
which is the constant that appear in the constraint. The $8^th$  
 element of  {\tt line1} is a list of one single integer which indicates if the constraints is and equality (value 0) or an inequality (value 1: $expr1 \geq 0$)

\section{Step 3: Solving the LP}

{\tt fichierPip}  writes the constraints of 
{\tt structTableEq} matrix in the mppip format in a file, 
then calls {\tt mpppip} to solve the LP and read the solution.

{\tt fichierPip} first calls {\tt matPip} which flatten the output of
each line of {\tt structTablEq}. At this point, the order of the coefficients 
of the $\tau$'s is inverted in order to minimize first the coefficients 
of the parameters. {\tt matPip} outputs the matrix {\tt matrice} 
 Then, {\tt writefile} writes the matrix in a file (this part is
quite long for big files). During this phase we add a single parameter
to the LP which is specific to {\tt Pip } behaviour. As we want to allow some 
variables to be negative (the $\tau$'s for instance), we have to add a 
{\em big parameter} $G$ and replace each occurrence of a $\tau$ by $\tau-G$. 
This manipulation is explained in~\cite{Feautrier88a}, This should be transparent 
and the coefficient of the big parameter in the solution is not important.

After the matrix is written, we perform some unix manipulations on the
file to obtain the good mppip input format. Then, we run 
mppip~\cite{Feautrier92aa}  and
the result is read by {\tt readPipResult},

\section{Step 4: final output form}
{\tt reorderSchedResult} place back the coefficients of the parameters
at the end of the scheduling vectors.
 {\tt getFinalOutputForm} pretty prints the result to the screen and 
set the output format to the following form: \\
{\tt <schedResult>=Alpha`ScheduleResult[scheduleType\_Integer,<sched3List>]\\
<sched3>=$\begin{array}{l} \mbox{\{ nameVar\_String,}\\
          \mbox{ indices\_List,}\\
           \mbox{ Alpha`sched[tauVector\_List,constCoef\_Integer] \}}
\end{array}$
}

 \bibliographystyle{alpha99}
\bibliography{/udd/risset/TEX/ref/newbib}
\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Schedule/Scheduler_user_manual.tex
Added mmalpha (source and binary)

\documentclass[11pt,fullpage]{article}

\usepackage{makeidx}
\newcommand{\aalpha}{{\sc Alpha}}
\newcommand{\ammalpha}{{\sc MmAlpha}}
\newcommand{\err}{\top}
\newcommand{\indef}{\bot}
\newcommand{\mma}{{\sc Mathematica}}
\makeindex

\title{User manual of  the Alpha {\bf Scheduler}\\
Version of \today{}}
\author{Tanguy Risset and Patrice Quinton}
\date{\today}

\begin{document}
%\bibliographystyle{plain}

\maketitle
\tableofcontents
\newpage
This documentation\footnote{It improves and replaces the previous 
scheduler manual that was  called \texttt{docSched.tex}. Unfortunately, this documentation 
in far from being complete.} provides some information regarding the {\aalpha} 
schedulers. It is available in the file:\\
{\tt \$MMALPHA/doc/Schedule/Scheduler\_user\_manual.pdf}

Section~\ref{Introduction} describes the principles of the Farkas and
the vertex schedulers. Section~\ref{sec2} presents the main utilization modes
of the \texttt{schedule} function, 
while section~\ref{sec3} describes the options of this function. 
In section~\ref{Result}, the output format of 
the scheduler is detailed. 
Section~\ref{settings} describes briefly the \texttt{Schedule.m}
package. A short example is shown in section~\ref{Examples}. 
Troubles are described in section~\ref{troubles}.

\section{Introduction}
\label{Introduction}
The scheduler is a package of {\ammalpha} that attempts 
to find a schedule for a given {\aalpha} program. An {\aalpha} 
program~\cite{Wilde93,Mauras89} has no sequential ordering: 
any sequential or parallel ordering of the computations is semantically 
valid provided it respects the data dependencies of the program.

Scheduling an {\aalpha} program
consists in giving a computation date for each variable of the
program. The time is considered as a discrete
single clock. The basic goal of the scheduler is to find a valid execution order
that is {\em good} with respect to a particular criterion. The
theoretical basis for the scheduling process is inherited from 
research on
systolic array synthesis and on automatic
parallelization. Currently two techniques are implemented for
the computation of a scheduling function, we will call them the {\em
Farkas method}~\cite{Feautrier92aa} (method used by default), and the
{\em vertex method}~\cite{MQRS90}.

In order to compute a schedule, the scheduler gathers all the constraints that
the schedule must meet and builds a linear program (LP) which is then
solved with a particular software.  
The two implemented methods
proceed differently but both give a {\em
mono-dimensional affine by variable} schedule. They can also provide
a {\em multi-dimensional affine by variable} schedule by changing the
options of the schedule (see section~\ref{sec2}).

A mono-dimensional affine by variable schedule assigns to each \aalpha{} equation of
the form
{\tt
A[i,j] = \ldots }, an execution date $T_A(i,j)$  which 
is given by an affine function of the indices
($i,j$) and of the system parameters ($N$):
$$T_{A}(i,j) = \tau^i_{A}
i + \tau^j_{A} j+ \tau^N_{A} N + \alpha_{A} \quad .$$ 
The vector $(\tau^i_{A} ,
\tau^j_{A},\tau^N_{A})$ is called the {\em scheduling vector}.  For
instance: $T_A(i,j) = 2i + j +3$ is a mono-dimensional affine schedule
for variable $A$; Its corresponding scheduling vector is $(2,1)$.

One important subset of mono-dimensional affine schedules are {\em  affine with
constant linear part} schedules, where the value of the scheduling vector
is  the same for each local variable of the program;
In this case, all variables must have
the same number of indices.

Multidimensional schedules are used for programs that do not admit
a linear execution time (for instance, if the execution time is
$N^2$). In this case, the schedule function is a vector of linear
functions and the order that is defined by the schedule 
is the lexicographic order on the components
(see~\cite{Feautrier92bb} for details).

\section{The \ammalpha{} \texttt{schedule} function }
\label{sec2} 
The  function to be called  in order to schedule a program is
 {\tt schedule[]}. Its effects is to schedule the {\aalpha} program contained in 
 the 
 {\tt \$result} \ammalpha{} variable,  and to put the result in 
 the global variable {\tt \$schedule} (see section~\ref{sec5} for the structure of the
resulting schedule). Options allow several parameters to be changed (see section~\ref{sec3}).

The possible forms of the use of the function {\tt schedule} are:
\begin{itemize}
\item {\tt schedule[]}: finds an affine by variable schedule for {\tt
\$result} which minimizes the global execution time of the system and assigns this schedule to
{\tt \$schedule}.
\item {\tt schedule[sys]}:
finds an affine by variable schedule for the {\aalpha} system {\tt sys}; the schedule minimizes 
the global execution time. The resulting schedule is assigned to
{\tt \$schedule}
\item {\tt
schedule[option\_1->value\_1,\ldots,option\_n->value\_n]}:
finds a schedule for {\tt \$result} which respects the chosen options and
assign it to {\tt \$schedule}
\item {\tt
schedule[sys,option\_1->value\_1,\ldots,option\_n->value\_n]}\\
find a schedule for the {\aalpha} system {\tt sys} which respects the chosen options and
assigns it to {\tt \$schedule}.
\end{itemize}

\section{Options of the \texttt{schedule} function}
\label{sec3}
Options have default values indicated hereafter. To change these
values, put one of the corresponding rules as a parameter to the {\tt
schedule} function. 
The Farkas method and the vertex method have
completely different implementations, hence their options are sometimes
different. Choosing between the Farkas method
and the vertex method is done with the {\tt schedMethod} option. 
\subsection{Main options}
\subsubsection*{schedMethod }
\label{schedMethod}
This option\footnote{As of \today{}, this option does not work. To use the 
vertex method, run the \texttt{scd[]} command.} allows one to switch from the Farkas method to the vertex
method. Its type is symbolic. Changing the value of this option greatly
influences the other options (for instance, it changes default values
of some options: {\tt durationByEq},...), be sure to check the others
options you use. The possible values are:
\begin{itemize}
\item {\tt schedMethod -> farkas} (default) selects the Farkas method.
\item {\tt schedMethod -> vertex} selects the vertex method.   
\end{itemize}
One of the main differences between the two methods is that the Farkas
method uses the \texttt{Pip} software to solve the LP while the vertex method
uses the \mma{} linear solver.

\subsubsection*{scheduleType}
\label{schedType}\index{scheduleType}
This option gives the type of schedule that \texttt{schedule} looks for.
Its type is symbolic, and its possible values are:
\begin{itemize}
\item {\tt scheduleType -> affineByVar} (default): affine by variable schedule.
\item {\tt scheduleType -> sameLinearPart}:   affine by variable scheduling with constant linear part. This option is often used for systolic designs. 
\item {\tt scheduleType -> sameLinearPartExceptParam}:  affine by variable schedule with constant linear part except for the parameters.
\end{itemize}
The default option allows one to get any kind of affine schedule. Option \texttt{sameLinearPart}
constraints the schedule to have all a common linear part. As a consequence, all variables
must have the same dimension (not checked).

\subsubsection*{multidimensional}
\label{multiDimensional}\index{multidimensional}
By default, the scheduler looks for a mono-dimensional schedule. If the
\texttt{multidimensional} is set, the scheduler also looks for multi-dimensional 
schedules. 
This option is boolean, hence
its possible values are:
\begin{itemize}
\item {\tt multidimensional -> False} (default): the schedule will be mono-dimensional.
\item {\tt multiDimensional -> True}: the schedule will be
multidimensional. If you set this options, be aware that it greatly
influences the value of other options: {\tt addConstraints}, {\tt durations} 
must be of type list of list (one list by schedule dimension, this means that you have to know the number of dimensions of the resulting schedule). Options {\tt optimizationTime} is automatically set to multi.
\end{itemize}

\subsubsection*{optimizationType}
\label{optimizationType}\index{optimizationType}
This option sets the objective function chosen. Its type is symbolic, and
 its possible values are:
\begin{itemize}
\item {\tt optimizationType -> time} (default): the total latency is
minimized (In the Farkas method, this minimization always correspond 
to the lexicographic minimization of the coefficients of the global
execution time which is an affine function of the parameters. In the
vertex method the way this minimization is performed can be changed by
the option {\tt objFunction}).
\item {\tt optimizationType -> Null }: no objective function (the
coefficients of the scheduling vectors are minimized in a
lexicographic order)
\item {\tt optimizationType -> delay}: tries to minimize the delays on
the dependencies (not implemented currently).
\item {\tt optimizationType -> multi} is for internal use, when a
multidimensional scheduling is computed, just be aware that the three above values for this option are not compatible with 
multi-dimensional scheduling.
\end{itemize}

%\subsubsection*{ratOrInt}
%This option indicates whether the resolution of the LP by mppip is
% done in integer mode (resulting scheudling vectors with integral
% coefficients) or in rationnal mode (resulting scheduling vectors may contains
% rationnal coefficients).  Its type is integer, the possible values
% are:
%\begin{itemize}
%\item {\tt \{ratOrInt -> 0\} } : rationnal solution, in this case, the
%scheduling function obtained by taking the floor of the scheduling
%functions is valid (\cite{Feautrier92aa}).
%\item {\tt \{ratOrInt ->  1\} } : (default) integer solution
%\end{itemize}

\subsubsection*{addConstraints}
\label{addConstraints}\index{addConstraints}
This option allows one to add some constraints to the generated LP.
Adding constraints is very important if you want to control the
resulting schedule.  The type of this option is a list of strings, each one of which
represent a constraint. The type of constraints authorized are
affine constraints on scheduling vectors.
Constraints can have two forms: one can force the value of 
a scheduling
vector or one can force coefficients of schedule to meet some constraints:
\begin{itemize}
\item Forcing a  variable $A[i,j]$ to be scheduled at time $i+2j+2$ can be done 
by the adding {\tt "TA[i,j]=i+2j+2"} to the constraints list.
\item For more precise constraints,  one can directly access each component 
of the schedule functions of each variable. For instance {\tt TAD2}
represents coefficient $\tau_A^j$ of the schedule of $A$ and {\tt CA} represents
the constant coefficient $\alpha_A$). With these names, one can add linear constraints 
on these coefficients using operators {\tt ==} or {\tt >=}. For instance,\\
{\tt \{"TAD1 == 1","TAD2 == 2", "CA >= 2" \} }\\is the same constraint
as above except that the constant is allowed to be greater than two.
\end{itemize}  
These options can also be used during the search of a multi-dimensional schedule.
In that case, its value is a list of lists of string, each
list of strings corresponding to constraints imposed on a dimension of the schedule.
Example of
use for mono-dimensional scheduling:
\begin{verbatim}
schedule[addConstraints-> {"TA[i,j,N]=i+2j-2",
    "T1D2==2","T1D1+2T1D3>=1"}]}
\end{verbatim}

\subsection*{durations}
\label{durations}\index{durations}
This option 
allows the user to specify precisely the execution time for each equation
or for each dependence. By default, each equation is assumed to need exactly
one cycle to be evaluated. Sometimes, one want to be more precise, 
especially when designing a circuit. For example, an equation of 
the form \texttt{A = B} imposes that \texttt{A} be scheduled one cycle
after \texttt{B}.

The type of this option is a
list of integers. It may be affected by the value of the {\tt durationByEq}
option.

For the Farkas method, 
the possible values of this option are:
\begin{itemize}
\item {\tt durations -> \{\}} (default): each equation 
%(if durationByEq is True) 
lasts one cycle.
%(resp. each dependence is 1 if durationByEq is False),
% whatever complex is the computation (default value).
\item {\tt duration -> \{0,0,1,1,0,...0,3,1\}} (list of integers):
\begin{itemize}
\item If the \texttt{durationByEq} option is set to \texttt{True} (default in the Farkas
method): integer number $i$ indicates the duration of the equation
defining variable number $i$ of the Alpha program. The 
variables are numbered in their order of declaration in the 
\aalpha{} system:  input variables,
then output variables, and finally local variables.
\item If the \texttt{durationByEq} option is set to \texttt{False} (default in the vertex method),
 integer number $i$ indicating the duration of dependence number $i$.  The 
 dependencies are numbered in the order given by the \texttt{dep} function: to
 get the list of dependences run the function \texttt{show[ dep[] ]}.
\end{itemize}
\end{itemize}
\paragraph*{Note:} as of \today, I think that the option \texttt{durationByEq -> True} 
is only valid for the Farkas method and the option \texttt{durationByEq -> False} is only valid
for the vertex method.

\subsubsection*{outputForm}
\label{outputForm}\index{outputForm}
This options allows a non standard output to be obtained. The
standard schedule output is described in section~\ref{sec5}. 
One can
also obtain as a result the linear program to be solved in various formats, or the
schedule polytope, i.e. the polytope which contains all the valid
solutions in the \aalpha{} format. The possible values of \texttt{outputForm} are:
\begin{itemize}
\item {\tt outputForm -> scheduleResult} (default): standard schedule
output form (Alpha`ScheduleResult structure, see section~\ref{sec5}).
\item {\tt outputForm -> lpSolve}: outputs in the format of the
lp\_solve software,
the linear programming
problem to solve in order to find the schedule.
\item {\tt outputForm -> lpMMA}: outputs the linear programming problem
to solve in order top find the schedule in the format of the linear
solver of Mathematica. Warning; as of \today{}, this option is not
implemented.
\item {\tt outputForm -> domain}: outputs the schedule polytope,
i.e. the \aalpha{} domain which is composed of all the constraints of the LP
to solve. WARNING, this works only for SMALL programs (~3 instructions), otherwise the domain is too big to be handled by polylib.
\end{itemize}

\subsubsection*{debug}
\index{debug}\label{debug}
Prints additional information and does not destroy the temporary files build
for the interface with PIP. The type of this option is boolean, possible value are:
\begin{itemize}
\item  {\tt  debug -> False} (default):   debug mode not set.
\item  {\tt  debug -> True }: debug mode set.
\end{itemize}

\subsubsection*{verbose}
\index{verbose}\label{verbose}
 If set, the scheduler returns additional information, otherwise, 
 it returns only the result. 
 This option is boolean, its possible values are:
\begin{itemize}
\item  {\tt  verbose -> True} (default): normal printing.
\item  {\tt  verbose -> False}: nothing is printed out.
\end{itemize}

\subsection{Advanced options}
These options are here for an advanced use of the schedule function.
\subsubsection*{resolutionSoft}
This function indicates which software will be used to solve the LP. Its type is symbolic, the possible values are: 
\begin{itemize}
\item {\tt resolutionSoft -> pip}: uses P. Feautrier's PIP software (only
available for the Farkas method).
\item {\tt resolutionSoft -> mma}: uses the \texttt{ConstrainedMin} function of
\mma{} (only implemented in the vertex method).
\item {\tt resolutionSoft -> lpSolve}: uses the {\tt lp\_solve} software(not implemented yet).
\end{itemize}

\subsubsection*{objFunction}
\index{objFunction}
This option is used to indicate how the minimization of the objective function 
(which is usually a function of the parameters) is performed. 
Its type is symbolic, the possible values are: 
\begin{itemize}
\item {\tt objFunction -> lexicographic}: minimizes   lexicographically 
the coefficient  of this function in the order of
the declaration of the corresponding parameters in the Alpha program (default in Fakas implementation).
\item {\tt objFunction -> lexicographic["N","P","M"]}: (not implemented anywhere) 
\item {\tt objFunction -> 2"N" + "P"} minimize 2 time the coefficient of parameter "N" plus one time the coefficient of parameter "P" (only implemented in the vertex resolution).
\end{itemize}

\subsubsection*{onlyVar}
\index{onlyVar}
Indicates which variables to schedule (useful if you have a very long program and which to schedule only part of it. Its type is a list of strings, the possible values are: 
\begin{itemize}
\item {\tt onlyVar -> all} (default value): schedules all the variables.
\item {\tt onlyVar -> \{"a","B","c"\}} (list of strings): schedules only the specified variables,   Warning, if  some variables are
   needed for the computation of the variables listed in the option,  the function will try to
   find their execution dates in \$schedule (This feature is only implemented in the Farkas method).
\end{itemize}

\subsubsection*{onlyDep}
\index{onlyDep}
Indicates which dependences to schedule (used for multidimensional scheduling).
Its type is a list of
integers and its possible values are:
\begin{itemize}
\item {\tt onlyDep -> all} (default value): schedules all the dependencies.
\item {\tt onlyDep -> \{1,4,5\}} (list of integers): schedules only the specified dependencies, the number correspond to their position in the list of dependencies which is returned by  the dep[] function (only implemented in the Farkas method).
\end{itemize}

\subsubsection*{subSystems}
\index{subSystems}
Indicates
whether or not the schedule takes into account calls to other subsystems.
\begin{itemize}
\item {\tt subSystems -> False}: (default value).
\item {\tt  subSystems -> True}.
\end{itemize}
In the Farkas method, the default value is not to consider subsystems, and if there
are some calls, the method fails. Therefore, to schedule a structured system, one
must run:
\begin{verbatim}
schedule[ subSystem -> True ]
\end{verbatim}
or equivalently:
\begin{verbatim}
structSched[]
\end{verbatim}

\subsubsection*{subSystemsSchedule}
\index{subSystemsSchedule}
Indicates where are the schedules of the subsystems used in the 
system we schedule. Its type is a list of schedules (see the format in section~\ref{sec5}.
\begin{itemize}
\item {\tt subSystemsSchedule -> \$scheduleLibrary} (default value).
\item {\tt  subSystemsSchedule -> List[schedule..]}.
\end{itemize}
Note that the scheduler automatically appends to \texttt{\$scheduleLibrary} 
the schedules of all new systems that it schedules.

\section{Format of \texttt{schedule}'s results}
\label{Result}
\label{sec5}
The result given by the function {\tt schedule} has a special form.
Two new head names are
introduced: {\tt Alpha`scheduleResult} and {\tt Alpha`sched}.

The
outermost structure is  a structure starting with the head {\tt
Alpha`ScheduleResult}, where the first argument is the name of the system
(string)  and where the
second argument is the schedule itself.

The Last argument of {\tt Alpha`scheduleResult} is for internal use.

The syntax of this structure is
described here.\\
\begin{verbatim}
<schedResult> = 
  Alpha`ScheduleResult[ 
    scheduleType_Integer, <sched3List>, objFunc_]

<sched3List> = { nameVar_String, indices_List, 
  Alpha`sched[ tauVector_List, constCoef_Integer ] }
\end{verbatim}
Example: 
\begin{verbatim}
scheduleResult[ "test1",
   {{"a", {"i", "k"}, sched[ {2, 2}, -4]}, 
    {"b", {"j"}, sched[ {0}, 0]}, {"c", {"i"}, sched[{0}, 15 ]}, 
    {"A", {"i", "k"}, sched[ {2, 2}, -3 ]}, 
    {"C", {"i", "k"}, sched[ {2, 2}, -2 ]}}, {15} ]
\end{verbatim}

Other examples are given in section~\ref{Examples}.

\section{Technical settings}
\label{settings}
The packages is called {\tt Schedule.m}. 
It uses packages \\{\tt FarkasSchedule},
{\tt VertexSchedule} and {\tt scheduleTools.m}.

The Farkas implementation
uses the {\tt pip} software (version D.1~\cite{Feautrier88b}) which must be present in the 
binary directory\footnote{Explain where...}.

The communication between \mma{} and {\tt pip} is done by files. These
files are written in the directory indicated by the \mma{} variable \texttt{\$tmpDirectory}.  
Currently, all methods should work on Solaris, Windows and MacOS X plateforms.

\section{Examples}
\label{Examples} 
Consider the program of Fig.~\ref{figureEx}. In the following, we
give the result of the scheduler with different options.
\begin{figure}[htbp]
\begin{verbatim}
system matmult : {M |M>1} 
  (a,b : {i,j | 1<=i,j<=M} of real)
returns
  (c : {i,j | 1<=i,j<=M } of real);
var
  C : {i,j,k | 1<=i,j<=M; 0<=k<=M} of real; 
let
  c[i,j] = C[i,j,M];
  C[i,j,k] = case
        {|k=0} : 0[];
        {|1<=k<=M} : C[i,j,k-1]+a[i,k]*b[k,j];
  esac;
tel;
\end{verbatim}
\caption{Matrix multiplication}
\label{figureEx}
\end{figure}

\subsubsection*{Default use}
If you type the following command (after loading the program of Fig.~\ref{figureEx}):\\
{\tt schedule[]}\\
the output written on the screen session should look like:
\begin{small}
\begin{verbatim}
In[65]:=
schedule[]
Checking options...
Dependence analysis...
Building LP...
LP: 92 variables, 80 Constraints
Writing file for PIP....
Solving the LP...
Total execution Time: 1+M
T_a{{i,j,M} = 0
T_b{i,j,M} = 0
T_c{i,j,M} = 1+M
T_C{i,j,k,M} = k

Out[65]=
scheduleResult[matmult,{{a,{i,j,M},sched[{0,0,
  0},0]},{b,{i,j,M},sched[{0,0,0},0]},{c,{i,j,M},sched[{0,0,1},1]},{
      C,{i,j,k,M},sched[{0,0,1,0},0]}},{1,1}]
\end{verbatim}
\end{small}
The first lines indicates which computations are performed. 
% the lines starting by {\tt Version} are output by
%Pip. 
Then the result is printed on the screen. Here, for instance
the schedule is affine by variable. {\tt c[i,j,k]} is computed at time
{\tt 1+M}. The result (after {\tt Out[65]}) is the corresponding
\mma{} structure assigned to {\tt \$schedule}.

\subsubsection*{Adding a constraint}
Suppose that we want to impose that variable {\tt C[i,j,k]} be
schedule at time $2i+j+k+7$ and that $\tau_a^i=\tau_b^i=\tau_c^i$. 
We have to add the two constraints: {\tt "TC[i,j,k]=2i+j+k+7"},
{\tt "TaD1==TbD2"} and {\tt "TaD1==TcD2"}, hence the command is: 
\begin{verbatim}
In[15]:= schedule[addConstraints->{"TA[i,j,k]=2i+j+k+7",
  "TaD1==TbD2", "TaD1==TcD1}]
\end{verbatim}
The result is 
\begin{verbatim}
T_a{i, j, N} = i
T_b{i, j, N} = i
T_c{i, j, N} = 8 + i + j + 2 M
T_C{i, j, k, N} = 7 + 2i + j +  k
\end{verbatim}
Notice that in this schedule, \texttt{c} and \texttt{C} do not
have the same $\tau^i$ coefficient. We may, if needed, fix this by adding the
constraint \texttt{TcD1==TCD1}.

\section{Troubles}
\label{troubles}
\subsection*{No schedule is found}
If no schedule is found, a message tells the user
and there may be several reasons. 
\begin{itemize}
\item No schedule exists (no way of solving this problem). You should be aware that
finding a schedule for a system of recurrence equations is in the general case undecidable.
Finding out an affine schedule is decidable, but an absence of affine schedule does
not guarantee that there does not exist a schedule. Most often, however, the scheduler
fails because there is an error in the \aalpha{} program which contains a dependence
cycle: the only solution is to check the program (use the \texttt{analyze} command, then 
check your program "by hand"). 
\item No schedule of the chosen type exists: try a multi-dimensional schedule, or
try another type of schedule.
%In this case, you can try
% the old scheduler which was more powerful (ask a \aalpha
% developer).
\item There exists a schedule but the time is not bounded. 
In this case try with the option {\tt objFunction} set to 1): this prevents the scheduler
to try optimizing the total time of the algorithm. 
\item  The program is not semantically correct: try the 
\texttt{analyze} function.
\end{itemize}

\subsection*{An awful error message}
In general, when something goes wrong during the scheduling, the error is captured correctly.
Sometimes,
the error may come with the following message:
\begin{verbatim}
General::aofil: /tmp/mat.tmp already open as /tmp/mat.tmp.
OpenWrite::noopen: Cannot open /tmp/mat.tmp.
General::stream: \$Failed is not a string, InputStream[ ], 
or OutputStream[ ].
\end{verbatim}
This may come from the fact that you have interrupted the previous
execution of the scheduler: \mma{} tries then to open a file which was
not closed. You can close these files (here {\tt /tmp/mat.tmp}) by
typing: 
\begin{verbatim}
Close["/tmp/mat.tmp"]
\end{verbatim}
In general it is not recommended to interrupt the evaluation of the schedule function.
\bibliographystyle{alpha}
\bibliography{biblio}
\printindex

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Synthesis/Synthesis.tex
Added mmalpha (source and binary)

\documentclass[11pt,fullpage]{article}

\usepackage{makeidx,fancyvrb,graphicx}

\newcommand{\alfa}{{\sc Alpha}}
\newcommand{\ammalpha}{{\sc MmAlpha}}
\newcommand{\mmalpha}{{\sc MmAlpha}}
\newcommand{\err}{\top}
\newcommand{\indef}{\bot}
\newcommand{\mma}{{\sc Mathematica}}
\newcommand{\syn}{\texttt{syn}}
\newcommand{\vhdl}{{\sc vhdl}}
\makeindex

\title{The Synthesis Program of \ammalpha{}\\
Version of \today{}}
\author{Patrice Quinton}
\date{Version of July 2008\footnote{There have been modifications since this note was written.}}

\begin{document}
%\bibliographystyle{plain}

\maketitle
%\tableofcontents
%\newpage
\begin{abstract}
This documentation presents the \texttt{syn} command of \ammalpha{}. This command,
which stands for {\em synthesize}, allows an \alfa{} program to be translated into \vhdl{},
whenever possible, by running in a systematic way the commands necessary to transform
the program into \vhdl{}. 
\end{abstract}

\section{Introduction}
\label{Introduction}
The \syn{} command allows a \vhdl{} program to be synthesized directly from
an \alfa{} file. In this document, we explain how this command can be used and
how it is implemented.
To better follow this documentation, you may lauch \mmalpha{} and type 
the command: \texttt{start[]}. A notebook will open, and in this notebook, go in the section
{\em New: synthesis notebooks} where you will find the examples presented
here.

\section{Basic usage}
\begin{verbatim}
syn[ "file.alpha" ]
\end{verbatim}
in its simplest form.

However, there are several options that allow the basic method to be changed, if
needed.

All functions are called with a special parameter \texttt{mute}: by default, they
are silent.

During the execution of one of the transformation steps, if an error message is triggered, then
\syn{} aborts. It then writes the \alfa{} program in the state that it had when the error 
occured.

A directory is created where various files produced during the 
synthesis are stored. A trace file, named
\texttt{trace.txt}, contains the list of \mmalpha{} commands that have been executed
until something wrong happened.

Finally, if everything goes well, a report file is written in this directory.

Example:
\begin{verbatim}
syn["fir.alpha"]
\end{verbatim}
synthesizes the program contained in file \texttt{file.alpha}, 
and shown in Appendix~\ref{firprogram}.

This basic command stops just before generating \vhdl{}, as the values of the 
\texttt{K} and \texttt{M} parameters are not set. To obtain \vhdl{}, run 
\begin{verbatim}
syn["fir.alpha", parameterRules -> {"N" -> 20, "M" -> 100}]
\end{verbatim}

The synthesis is almost silent, and takes a few seconds. If it succeeds, a 
congratulation message is issued\footnote{You'll understand why when you
start synthesizing you own program!}.

You get more information by adding the \texttt{verbose} option:
\begin{verbatim}
syn["fir.alpha", parameterRules -> {"N" -> 20, "M" -> 100}, 
  verbose->True]
\end{verbatim}

The \texttt{syn} program has a few options that can be displayed by
evaluating \texttt{Options[ syn ]}.

\subsection{The \texttt{SYN} directory}
The execution creates a directory named \texttt{"firSYN"} which contains: 
\begin{itemize}
\item \texttt{fir.report}: a file giving an analysis of the program after the 
\texttt{Alpha0} code generation (corresponding to the file 
\texttt{firAlpha0.alpha"}. 
\item \texttt{fir.scd}: the value of the schedule that was produced and used.
\item \texttt{firAlpha0.alpha}: the program, after \texttt{Alpha0} generation.
\item \texttt{firAlphard.alpha}: the program, after \texttt{Alphard} generation.
\item \texttt{firParametersFixed.alpha}: the program, after parameters are fixed. This is
the last transformation step before \vhdl{} generation.
\item \texttt{firPiped.alpha}: the program, after piping variables.
\item \texttt{firScheduled.alpha}: the program, after scheduling and placement.
\item \texttt{trace.txt}: the list of \mmalpha{} commands executed.
\item \texttt{VHDL}: a directory that contains the \vhdl{} programs.
\end{itemize}

If the directory contains a file including \texttt{Wrong} in its name, then one of
the steps went wrong (or maybe, this was produced during a previous execution 
of \texttt{syn}, as the directory is not cleaned.)

\section{The \texttt{VHDL} Directory}
The \texttt{VHDL} directory contains the following files, if the synthesis was
successful: 
\begin{itemize}
\item \texttt{ControlfirModule.component}: the component description of the 
controller part.
\item \texttt{ControlfirModule.vhd}: the \vhdl{} description of the controller part.
\item \texttt{cellfirModule1.component}: the component description of the first cell.
\item \texttt{cellfirModule1.vhd}: the \vhdl{} description of the first cell.
\item \texttt{cellfirModule2.component}: the component description of the second cell.
\item \texttt{cellfirModule2.vhd}: the \vhdl{} description of the first cell.
\item \texttt{cellfirModule4.component}: the component description of cell number 4.
\item \texttt{cellfirModule4.vhd}: the \vhdl{} description of cell number 4.
\item \texttt{definition.vhd}: a useless file.
\item \texttt{fir.component}: the component description of the fir.
\item \texttt{fir.vhd}: the \vhdl{} description of the fir filter.
\item \texttt{firModule.component}: the component description of the 
module part.
\item \texttt{firModule.vhd}: the \vhdl{} description of the module part.
\end{itemize}
The generation of \vhdl{} follows the structure shown in Fig.~\ref{firvhdl}. The main 
file, \texttt{fir.vhd}, may not be usable for reasons that are difficult to explain here...
But \vhdl{} file called \texttt{firModule.vhd} should be synthesizable. It calls a 
controller and instanciates three types of basic cells. 
\begin{figure}[ht]
\centerline{
\includegraphics[width=\textwidth]{firvhdl.pdf}
}
\caption{Structure of \vhdl{} generated for the FIR filter\label{firvhdl}}
\end{figure}

\section{What \syn{} does}
\begin{enumerate}
\item Loads the specified file. If this step is successful, the {\em synthesis directory} is created. Its
name is \texttt{xxxSyn} where \texttt{xxx} is the name of the local directory.
\item Inline program, if needed. In other words, if the loaded program contains several
systems, then the last one that was loaded is inlined. If this step is successful, then
only this last system is kept in the library, and it is saved under the name \texttt{yyyInlined.alpha}
in the synthesis directory, where \texttt{yyy} is the name of this system.
\item Checks the program, by doing a static analysis.
\item Schedule the program. Without options, \texttt{schedule[]} is applied. Options can 
be provided (see Section~\ref{options}) to either call \texttt{scd[]} and give 
hints to the scheduler. Found schedule is saved in the synthesis directory and if no 
schedule is found, the file is saved under the name \texttt{yyyWrongScheduled.alpha}.
\item Applies the schedule. Result is saved in file \texttt{yyyScheduled.alpha}.
\item Pipes variables using \texttt{pipeVars} command. 
Result is saved in file \texttt{yyyPiped.alpha} (or \texttt{yyyWrongPiped.alpha} if an
error occured).
\item Generates 
\end{enumerate}

\section{Options}
\label{options}
\begin{description}
\item[\texttt{optionsOfScheduler}:] contains options that are passed to the scheduler
\item[\texttt{debug}:] debug option
\item[\texttt{verbose}:] 
\item[\texttt{schedMethod}:] allows the scheduler type to be specified. Values are 
\texttt{farkas} (by default) or otherwise (any other value), the vertex method.
\item[\texttt{parameterRules}:] specifies an association list of parameter values.
\end{description}

\section{Report}
\label{report}
\begin{enumerate}
\item Writes \texttt{Equation}
\item Gives the declaration of the lhs
\item Prints out the equation
\item Prints out the type of the lhs
\item Signals if equation is an output equation (the lhs is an output variable), or a local
equation (the lhs is local)
\item Signals if we have an input equation, i.e., an equation of the form \texttt{lhs = input}\footnote{This should be extended to the case where there is an domain, and possibly an dependency.}.
\item Prints out the indexes.
\item Signals if equation is scheduled. It could be a scalar, a variable with one dimension of
time, possibly with several dimensions of space.

\end{enumerate}

\appendix
\newpage
\VerbatimInput{fir.alpha}
%\newpage
\section{\vhdl{} Model Generated for the FIR Module}
This is the file \texttt{firModule.vhd}. 
\VerbatimInput{firModule.vhd}
\section{\vhdl{} Model Generated for the FIR cell 2}
This is the file \texttt{cellfirModule2.vhd}. 
\VerbatimInput{cellfirModule2.vhd}

%\bibliographystyle{alpha}
%\bibliography{biblio}
\printindex

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Technical-Manual/Technical-Manual.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}
\begin{document}

\newcommand{\mma}{\texttt{Mathematica}}
\newcommand{\mmalfa}{\texttt{MMAlpha}}

\title{Technical manual of \mmalfa{}\footnote{Version of \today{}}}
\author{Patrice Quinton}
\date{May 2005}
\maketitle
\section{Introduction}
\label{introduction}
This document presents the organisation of the LipForge distribution of 
\mmalfa{}. This new archive is being built by Tanguy Risset and Patrice Quinton.

\newcommand{\bincyg}{\texttt{bin.cygwin}}
\newcommand{\bindarw}{\texttt{bin.darwin}}
\newcommand{\binlinux}{\texttt{bin.linux}}
\newcommand{\config}{\texttt{config}}
\newcommand{\demos}{\texttt{demos}}
\newcommand{\sources}{\texttt{sources}}
\newcommand{\tests}{\texttt{tests}}
\newcommand{\doc}{\texttt{doc}}
\newcommand{\lib}{\texttt{lib}}
\newcommand{\libcyg}{\texttt{lib.cygwin}}
\newcommand{\libdarw}{\texttt{lib.darwin}}

\newcommand{\readme}{\texttt{readme}}
\newcommand{\cgen}{\texttt{Code\_gen}}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\bin}{\texttt{bin}}

\section{Organization}
\label{organization}
The distribution contains the following directories:
\begin{enumerate}
\item \bincyg{}, \bindarw{}, \binlinux{}, etc. are the directories where the binary 
files for various systems are put. These binary files are the result of 
compiling source files.
\item \config{} contains various files needed to configure the installation.
\item \demos{} contains notebooks for various demonstrations of \mmalfa{}.
\item \doc{} contains the documentation.
\item \lib{} contains the mma{} packages.
\item \libcyg{}, \libdarw{}, etc. contain some library files obtained as
byproducts of compiling source files.
\item \sources{} contains the source files.
\item \tests{} contains test files.
\end{enumerate}
In addition, the main directory contains a \readme{} file, and several files
for the \texttt{html} documentation.

The main directory also contains a file \texttt{CopyMMA.m}. This is a 
\mma{} program that can be used to create a distribution. The documentation 
of this program is in directory \texttt{doc/Distribution}.

\section{The \sources{} directory}
\label{sources}
\subsection{Content}
The \sources{} directory contains one directory for each 
\texttt{C} program used by \mmalfa{}. As any other directory, it contains
a \readme{} file, \texttt{html} files.

\newcommand{\domlib}{\texttt{Domlib}}
\newcommand{\makeinclude}{\texttt{Makeinclude}}
\newcommand{\mathlink}{\texttt{Mathlink}}
\newcommand{\pip}{\texttt{Pip}}
\newcommand{\polylib}{\texttt{Polylib}}
\newcommand{\pretty}{\texttt{Pretty}}
\newcommand{\readalfa}{\texttt{Read\_Alpha}}
\newcommand{\writealfa}{\texttt{Write\_Alpha}}
The directories are:
\begin{enumerate}
\item \cgen{},
\item \domlib{},
\item \makeinclude{},
\item \mathlink{} (obsolete),
\item \pip{},
\item \polylib{},
\item \pretty{}
\item \readalfa{},
\item \writealfa{}.
\end{enumerate}

It contains also a sed directory, which is obsolete. The \mathlink{} directory is
also obsolete.

It contains a make file, named \makefile{}.

\subsection{Compiling source files}
It is possible to recompile the source files. The 
method is as follows.

\newcommand{\envostype}{\texttt{OSTYPE}}
\newcommand{\envmmalpha}{\texttt{MMALPHA}}
\newcommand{\vardir}{\texttt{DIR}}

In the \sources{} directory, type
\begin{verbatim}
make all
\end{verbatim}
The  \makefile{} is then executed. If the \envostype{} variable is set, 
the corresponding binary files are produced and put in the 
corresponding \bin{} directory. In case of problem, check the
value of \envostype{} and of \envmmalpha{}.

\subsection{How it works}
The \makefile{} in \sources{} contains a few variable definitions and 
a few rules. The \vardir{} variable defines the list of directories
where files should be compiled. 
The \texttt{all} rule, defines what should be done for all element
of \vardir{}: namely, go in the corresponding directory, and run make. 
The \texttt{all} rule depends on any file contained in the bin or lib
directories.

The \texttt{clean} rule defines what should be done to clean directories.
Again, it consists in going in the corresponding directory, and running
\begin{verbatim}
make -f Makefile clean
\end{verbatim}

Finally, two rules help create the lib and bin directories, if they
are missing.

Once a directory is entered, say \pretty{}, this directory contains
a make file named \makefile{}, some source files, some Object
directories, a \readme{} file and \texttt{html} files.

The organization of this \makefile{} depends on the considered 
source directory.

\subsubsection{The \pretty{} directory}
For the \pretty{} directory, it is very simple, and is explained in 
the corresponding \makefile{}. This directory contains 
some source files which are needed for \readalfa{} and 
\writealfa{}.

\subsubsection{The \cgen{} directory}
For the \cgen{} directory, the structure of the make file
is much more complicated.

First, the \makefile{} calls the file \texttt{Makefile.config} which is
situated in the \makeinclude{} directory, where some variables 
will be set. This make file also calls system dependent make file, 
say \texttt{Makefile.darwin} for the \texttt{darwin} system, where
some system dependent definitions are set.

\newcommand{\varcppflags}{\texttt{CPPFLAGS}}
\newcommand{\varldflags}{\texttt{LDFLAGS}}
\newcommand{\vardummy}{\texttt{DUMMY}}
\newcommand{\varloadlibes}{\texttt{LOADLIBES}}
\newcommand{\varname}{\texttt{NAME}}
\newcommand{\varobjs}{\texttt{OBJS}}
\newcommand{\varlibs}{\texttt{LIBS}}
\newcommand{\rules}{\texttt{rules}}
\newcommand{\varcc}{\texttt{CC}}
\newcommand{\varpolyinclude}{\texttt{POLYINCLUDE}}
\newcommand{\varbindir}{\texttt{BINDIR}}
\newcommand{\varobjdir}{\texttt{OBJDIR}}
\newcommand{\varextraflags}{\texttt{EXTRA\_FLAGS}}

Then, back to the local make file, some other variables are defined:
\begin{itemize}
\item \varcppflags{}: include paths for the compiler,
\item \varldflags{}: library paths for the loader (where libraries should be looked for),
\item \varloadlibes{}: libraries for the loader,
\item \vardummy{}: list of files to be removed when cleaning
\item \varname{}: the name of the binary file to be created
\item \varobjs{}: the list of all object files
\item \varlibs{}: the list of libraries needed. This variable is used for completing 
make dependencies.
\end{itemize}

After these definition, one additional rule (which I do not understand).

And finally, call to a generic \makefile{}.\rules{} (which is barely understandable...).

\subsubsection{The \readalfa{} and \writealfa{} directories}
These directories are defined in much the same way as the \cgen{} 
directory.

\subsubsection{The \domlib{} directory}
\label{domlib}
The \domlib{} directory is different, after a deep modification that was
recently done, in order to become less dependent on the \polylib{}
which is distributed within \mmalfa{}.
\begin{itemize}
\item First, the main \makefile{} file contains a switch depending on the
system type. If the OS is Cygwin, the Visual C++ environment is called, 
otherwise, an appropriate make file is called. 
\item I only define here what a make file for a Unix-like system looks like.
It contains the definition of a few variables.
\begin{itemize}
\item \varcc{}: which compiler to use (could be removed)
\item \varpolyinclude{}: the directory where the include files for the \polylib{}
are. In the current version, this directory is where the \polylib{} installation 
puts these files. (A companion document explains how to install the \polylib{}, 
from the Web distribution. In the installation procedure, one must 
run a command
\begin{verbatim}
./configure.in --prefix="Path"
make
make install
\end{verbatim}
where \texttt{Path} is the directory where the \texttt{include} and \texttt{lib}
directories of \polylib{} are put. Say for example, you install 
\polylib{} with 
\begin{verbatim}
./configure.in --prefix="$MMALPHA/sources/Poly/$OSTYPE"
make
make install
\end{verbatim}
then (provided \envmmalpha{} and \envostype{} are properly set), 
you will find in \sources{}\texttt{/Poly/}\envostype{} the include
and lib files.
\item \varbindir{}: where the binary files will be put, enventually,
\item \varobjdir{}: where the object files will be put, during compilation,
\item \varextraflags{}: additional flags for the compiler. Uses some
variables that are set by \texttt{MakeIncludes/Makefile.\$OSTYPE},
\end{itemize}

\end{itemize}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Test-Bench/Test-Bench.tex
Added mmalpha (source and binary)

\documentclass[11pt]{article}
\usepackage{fullpage,makeidx,fancyvrb,graphicx}
\usepackage{moreverb}

\begin{document}

\newcommand{\mmalfa}{{\sc MMAlpha}}
\newcommand{\mma}{{\sc mathematica}}
\newcommand{\alfa}{{\sc Alpha}}
\newcommand{\vhdl}{{\sc vhdl}}
\newcommand{\alfazero}{{\sc alpha0}}
\title{The \mmalfa{} Test Bench\thanks{Version printed on \today}}
\author{Patrice Quinton}
\date{February 24, 2010}
\maketitle
\begin{abstract}
We describe a program to generate test bench automatically for 
\mmalfa{} programs. We also describe how to generate stimuli files, and
how to generate a bit true simulator. 
\end{abstract}

\section{Introduction}
In Section~\ref{tb}, we present how to generate a test bench.
In Section~\ref{stimuli}, we present how to generate stimuli files. 
In Section~\ref{bittrue}, we explain how to simulate a program using 
a bit true simulation.

\section{Test Bench Generation}
\label{tb}
The \texttt{VhdlTestBench} package contains a function that allows a \vhdl{} test bench
to be automatically generated for an \alfa{} program. This program must be in \textsc{Alpha0}
format, such that obtained after a successful utilization of the \texttt{a2v} command. This
means that the current program (contained in the library variable, \texttt{\$library}), has been translated
into \vhdl{}.

The function is called \texttt{vhdlTestBenchGen}.
The test bench is produced in a file called \texttt{fileTB.vhd}, where \texttt{file} is
the name of the system currently contained in the \texttt{sys} parameter of the
\texttt{vhdlTestBenchGen} function\footnote{See the usage of this function.}.

The test bench file is written in the current directory of \textsc{Mathematica}. If such
a file already exists, it is copied in a temporary file named
\texttt{file\_.TB.vhd} (a previous version of such a file is deleted, if any exists).

The test-bench that is produced must be used with stimuli files for each
of the input variables of the \alfa{} program; it will produce one output file for
each output variable of the \alfa{} program. How to produce these stimuli files
is explained in Section~\ref{stimuli} of this documentation.

The \texttt{vhdlTestBenchGen} function allows test benches to be generated for each one of the 
elements of the library. Recall that, after generating an \alfazero{}
program, all modules produced are contained in an \mmalfa{} 
variable called \texttt{\$library}. The current program is in
variable \texttt{\$result}, and it is usually the main program. Say this 
program is \texttt{fir} (corresponding to an initial \texttt{fir.alpha} \alfa{}
program), then executing
\begin{verbatim}
vhdlTestBenchGen[ ]
\end{verbatim}
will produced a test-bench named \texttt{firTB.vhd}. It is essential that 
the content of \texttt{\$library} be consistent with the program for which the
test bench is generated.

The \texttt{vhdlTestBenchGen} function checks various possible errors:
\begin{itemize}
\item The given program must be recognized as a cell, a module or a controller
by \mmalfa{} (using the \texttt{checkCell},  \texttt{checkModule}, or
 \texttt{checkController} functions.
 \item The name of the given program must appear in the library contained 
 in \texttt{\$library}.
 \item 
\end{itemize}

More can be found on this topic

\section{Technicalities}
\label{technicalities}
\begin{itemize}
\item
There is a provision for generating a \vhdl{} file with the 1987 \vhdl{}
syntax (see the \texttt{s87} flag, directly in the code). 
\item
The clock rate is 20\,ns, and the clock initialization time is 30\,ns. 
\item
If there is already a \texttt{component} file for the program contained
in \texttt{\$result} (say for example, \texttt{fir.component}) in the directory, 
this component is used in the test bench, otherwise, a component is built
on the fly (using the \texttt{buildComponent} function). Therefore, if there 
exists an inconsistency between the program and its component, an 
error may happen. 
\end{itemize}

\section{Meaning of the signals}
Here is a description of some of the signals in the test bench produced. 
\begin{itemize}
\item \texttt{clk\_rate}: 
\item \texttt{clk\_init}: 
\item \texttt{sig\_Name}: a signal associated to \alfa{} variable \texttt{Name};
\item \texttt{comp}: the name of the instanciated component;
\item \texttt{ce}: clock enable signal;
\item \texttt{rst\_0}: initial value of the reset signal;
\item \texttt{temp\_buffer}: 
\item \texttt{temp\_buffer\_out}: 
\item \texttt{endstim}: a boolean variable that indicates the end of the stimulus file
\item \texttt{good}: a boolean variable used to check the reading of a stimulus file
\item \texttt{temp\_buffer\_Name}: a temporary buffer for \alfa{} variable \texttt{Name};
\item \texttt{stim\_file\_Name}: the declaration of the stimulus file for variable \texttt{Name}:
\item \texttt{stim\_line\_Name}: the declaration of the file line for variable \texttt{Name}:
\item \texttt{timecounter}: time counter for the design;
\end{itemize}

\section{Producing Stimuli}
\label{stimuli}
As already said, there exists a stimuli file for each input variable. 
If \texttt{inVar} is the name of an input variable of the \alfa{} program, then its
stimuli file is named \texttt{stimInVar.txt}. This file contains one value for each instant
of time for which a value is required by the program.

Recall that, for any \alfa{} variable, its {\em life time} is given by projecting its domain 
on the \texttt{t} index. Say this life interval is $10 \leq \texttt{t} \leq 20$, then
values will be required only for time instants between 10 and 20, therefore, the 
stimuli file contains 11 values.

Producing automatically stimuli files can be done either manually, or by using 
the \texttt{cGen} program, as described here.

To better explain this program, assume we want to produce stimuli files for
the following program \texttt{fir}:

\VerbatimInput{fir.alpha}

We load this file, and we schedule it:
\begin{verbatim}
load[ "file.alpha" ];
scd[]
\end{verbatim}
Then, we run the following program

\begin{verbatim}
cGen[ "file.c", { "K" -> 3, "N" -> 10 }, stimuli -> True  ]
\end{verbatim}

This produces in file \texttt{file.c} a \texttt{C} program that simulates the 
operations of the \alfa{} program. Compiling this program :
\begin{verbatim}
gcc file.c -o fir
\end{verbatim}
produces a binary file \texttt{fir} that we can run 
\begin{verbatim}
./fr
\end{verbatim}
The effect of this execution is to ask the values of the input variables and to return, 
eventually, the values of the output variables. Here is what you see at the 
console: 
\begin{verbatim}
x[0]?0
x[1]?1
x[2]?2
x[3]?3
x[4]?4
x[5]?5
x[6]?6
x[7]?7
x[8]?8
x[9]?9
x[10]?10
w[1]?1
w[2]?0
w[3]?0
y[3]=2
y[4]=3
y[5]=4
y[6]=5
y[7]=6
y[8]=7
y[9]=8
y[10]=9
\end{verbatim}
The input chosen are \texttt{x[i] = i}, and \texttt{w[0] = 1}, 
\texttt{w[1] = w[2] = 0}, which gives the result shown.

As a side effect, this evaluation produces also 3 files: 
input files \texttt{stim\_x.txt} and \texttt{stim\_w.txt} where input values for
x and w are recorded, and an output file \texttt{stim\_y.txt}.
Here is the \texttt{stim\_x.txt} file: 
\VerbatimInput{stim_x.txt}
Here is the \texttt{stim\_w.txt} file: 
\VerbatimInput{stim_w.txt}
Here is the \texttt{stim\_y.txt} file: 
\VerbatimInput{stim_y.txt}

The \texttt{cGen} function has several options. The \texttt{noPrint} option, when set to
\texttt{True}, does not prints out the name of the inputs and outputs values (for example,
\texttt{x[1]=00000001}, but only the result. The resulting files may then be used to
be run with the test bench.

This example is presented in the \texttt{Fir} notebook in \texttt{mmalpha/demos/NOTEBOOKS/Fir}.
\section{Bit True Simulation}
\label{bittrue}
Documentation to be done.
\newcommand{\stim}{\texttt{alpHardStim}}
\section{Generation of test files}

\label{stim}
The \texttt{VhdlTestBench} package contains now a new function, called \stim{}, which
allows stimuli files to be generated for the \vhdl{} test bench. For each input \texttt{in}, 
it produces an input file called \texttt{alpHardStimin.txt}, and similarly for each 
output. This file contains, in the order given by the time variable, the command to
be executed to generate...

\section{The \texttt{alpha2mma} Package}
\label{translator}
There exists a prototype program that allows an \alfa{} program to be 
translated into \mma{}. This package is called \texttt{alpha2mma.m}. It is currently
available in the directory
\begin{verbatim}
$MMALPHA/doc/Packages/Meta
\end{verbatim}
as it is an example of using the \texttt{Meta} package.

The \texttt{alpha2mma.m} program is itself produced by the use of
the \texttt{meta} command, applied to a description of the \alfa{} syntax
given in the meta file \texttt{alpha2mma.meta}.
See in the notebook \texttt{meta.nb} located in the same directory how to
call the translator.

Once the \texttt{alpha2mma.m} file has been produced, one loads it by
the command 
\begin{verbatim}
<<alpha2mma.m
\end{verbatim}

Then to translate a program, say \texttt{test1.alpha} into a \mma{} program, one
does:
\begin{verbatim}
load[ "test1.alpha" ];
alpha2mma[]
\end{verbatim}
(there exist a \texttt{debug} option, as usual).

This produces in the current directory a file called \texttt{test1.simul}, which contains
definitions for the input, local and output variables.

Then, one loads this file:
\begin{verbatim}
<<test1.simul
\end{verbatim}
and one calls the simul function :
\begin{verbatim}
simul["test1"]
\end{verbatim}

This function has the only effect of entering the context
\texttt{Alpha`Simul`test1`} in which all definitions of the variables
are available, as functions.

To be more explicit, assume that the \texttt{test1.alpha} file looks
as follows:
\begin{verbatim}
system test1   (a :  integer; b :  integer)
returns (c :  integer);
let
  c = a + b;
tel;
\end{verbatim}

This program is translated into the following \mma{} program:
\begin{verbatim}
Begin["Alpha`Simul`test1`"];
Clear[ Alpha`Simul`test1`a, Alpha`Simul`test1`b, Alpha`Simul`test1`c];

    (* Definition of c *)
    Alpha`Simul`test1`c[f:(_Integer|_Symbol)...] := Plus[ a[f], b[f] ];
    Alpha`Simul`test1`c[ ___ ] := Message[Alpha`Simul`test1`c::params ];
End[];
\end{verbatim}

As you can see, this program enters the special context
\texttt{Alpha`Simul`test1`}, then clears the variables \texttt{a},
\texttt{b} and \texttt{c} in this context, and defines a \texttt{c} function.
The form of this function is quite special. It says basically that 
\texttt{c[ f ] := a[ f ] + b[ f ]}, where \texttt{f} is any sequence of integers or
symbols. This allows the function to be used for any sequence
of indices, a property that will be useful later on when one wants to
evaluate a structured \alfa{} program.

When one runs \texttt{simul[ "test1" ]}, one enters the (unique) simulation context
of \texttt{test1}, where the definition of \texttt{c} are available. In this context, 
if we just evaluate \texttt{c[]}, we get the result \texttt{a[] + b[]}; in the
same manned, we can evaluate any \texttt{c[ i1, i2, ..., ik ]} expression, 
where the \texttt{ik} are either integers or symbols of \mma{}, to 
get the result 
\texttt{a[ i1, i2, ... , ik ] + b[  i1, i2, ... , ik]}.

Consider now a more involved example, where one wants to 
build a binary adder in \alfa{}. The program is given in Fig.~\ref{figadder}.
\begin{figure}[htbp]
\begin{verbatim}
system FullAdder (A :  boolean;
                  B :  boolean;
                  Cin :  boolean)
       returns   (X :  boolean;
                  Cout :  boolean);
let
  X[] = A xor B xor Cin;
  Cout[] = A and B or A and Cin or B and Cin;
tel;

system Plus :  {W | 2<=W}
               (A : {b | 0<=b<=W} of boolean;
                B : {b | 0<=b<=W} of boolean)
       returns (S : {b | 0<=b<=W} of boolean);
var
  Cin : {b | 0<=b<=W-1} of boolean;
  Cout : {b | 0<=b<=W-1} of boolean;
  X : {b | 0<=b<=W-1} of boolean;
let
  use {b | 0<=b<=W-1} FullAdder[] (A, B, Cin) returns  (X, Cout) ;
  Cin[b] =
      case
        { | b=0} : False[];
        { | 1<=b} : Cout[b-1];
      esac;
  S[b] =
      case
        { | b<=W-1} : X[b];
        { | b=W} : Cout[W-1];
      esac;
tel;
\end{verbatim}
\caption{The binary Plus program\label{figadder}}
\end{figure}

It contains two parts: a full adder (\texttt{FullAdder} program), and 
a \texttt{Plus} program, that allows one to build a binary adder of 
\texttt{W} bits by instanciating \texttt{W} times the full adder.

Translating this program into \mma{} involves (currently) two steps. First, 
one translate the \texttt{FullAdder} program. To this end, assuming
that the program is in the file \texttt{fulladder.alpha}, one loads this program, 
one select the \texttt{FullAdder} subsystem, and one translates it :

\begin{verbatim}
load[ "fulladder.alpha" ];
getSystem[ "FullAdder" ];
alpha2mma[];
\end{verbatim}

One does the same for the main program, \texttt{Plus}
\begin{verbatim}
getSystem[ "Plus" ];
alpha2mma[];
\end{verbatim}

Then, one loads both programs :
\begin{verbatim}
<<"Plus.simul"
<<"FullAdder.simul"
\end{verbatim}
which creates definitions for the variables

\newpage
\appendix
\section{The Test Bench for the Fir Filter}
\label{tbfile}
\VerbatimInput{fir_TB.vhd}

\newpage
\section{The Stimuli Generator for the Fir Filter}
\label{stimfile}
This program was generated with the default value of the \texttt{noPrint} option.
\VerbatimInput{fir.c}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tests/testing-MMALPHA.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\begin{document}

\newcommand{\domlib}{{\sc domlib}}
\newcommand{\ML}{{\sc Mathlink}}
\newcommand{\mprep}{{\sc mprep}}
\newcommand{\polylib}{{\sc polylib}}
\newcommand{\mma}{{\sc Mathematica}}
\newcommand{\mkf}{{\texttt Makefile}}
\newcommand{\gcc}{\texttt{gcc}}
\newcommand{\doml}{\texttt{domlib.c}}
\newcommand{\tm}{\texttt{domlibtm.c}}
\newcommand{\mmAlpha}{\textsc{MMAlpha}}
\newcommand{\MMAlpha}{\textsc{MMAlpha}}
\newcommand{\MMA}{\textsc{Mathematica}}
\newcommand{\true}{\texttt{True}}
\newcommand{\false}{\texttt{False}}

\title{Testing \mmAlpha{}}
\author{Patrice Quinton}
\date{Version of \today}
\maketitle

\begin{abstract}
This document gives information on testing \mmAlpha{}. Some other information 
is also available in \texttt{doc/Install}. Information regarding compiling \texttt{domlib}
is given in \texttt{doc/Developers/Compiling}.
\end{abstract}
\section{Right after installing \mmAlpha{}}
\mmAlpha{} provides testing facilities which are explained in this document. 
After installation, run \texttt{test1[]}. This checks various things, in particular, that 
the \texttt{domlib} is properly installed. 
The report of this test is written in file \texttt{\$MMALPHA/tests/testReportFile.txt}. 
If this test does not return the value \texttt{True}, please send a mail together
with this file to 
\begin{verbatim}
patrice.quinton@irisa.fr
\end{verbatim}

\paragraph{Warning:} Notice that executing \texttt{test1[]} erases the previous version of the test 
report file.

\paragraph{Do it yourself:} You may also fix by yourself some common errors: report to Section~\label{troubleshooting}.

\paragraph{Tested configurations}
The current distribution runs properly on a MacBook Pro, whith \mma{} version 5 and 6. 
Version 7.0 of \mma{} is under test.

\section{More tests}
Other tests are done using the commands 
\texttt{test2[]}, 
\texttt{test3[]}, 
and \texttt{test4[]}. All these tests append information in the 
file \texttt{\$MMALPHA/tests/testReportFile.txt}.

If one of these tests fails (i.e., it does not return the value \texttt{True}),
please send a mail together
with the test report file to 
\begin{verbatim}
patrice.quinton@irisa.fr
\end{verbatim}

If all these tests are successful, your installation should be correct.

\section{Exploring tests}
Commands \texttt{test1}, \texttt{test2}, etc. make use of another test program, 
which allows each individual package to be tested. For example, 
\begin{verbatim}
test[ "Domlib" ]
\end{verbatim}
runs a test for the domlib and returns \texttt{True} if the test was successful. This 
command writes information in the test report file. If one test fails, the name
of this test is written in the test report file. Say test \texttt{"domlib 23"} fails, then 
you can replay it by evaluating : 
\begin{verbatim}
test[ "Domlib" , "domlib 23"]
\end{verbatim}
The nature of this test will then be displayed. 
Notice however, that running an individual test may not always provide the
same result as when it is run in the context of a full package test. The reason is that
some of the package test were written before this facility was offered. I am trying\footnote{As of December 2008.}
to fix this, but test programs are numerous and long...

Test files are in directory \texttt{\$MMALPHA/Tests}. Commands \texttt{test1} to
\texttt{test4} are part of \texttt{\$MMALPHA/lib/Packages/Alpha.m}.

\section{Common Problems at Installation and How to Fix Them}
(See also \texttt{doc/Install}).

Check this in order. 
\begin{enumerate}
\item Is environment variable \texttt{\$MMALPHA} set? If this is not the case, \mma{} does not
start \MMAlpha{}. 
To check this, start \MMAlpha{} and evaluate
\begin{verbatim}
Environment[ "MMALPHA" ]
\end{verbatim}
If this variable was not set, then the result is \texttt{\$Failed}, otherwise, it should be the path
of the directory where you installed \MMAlpha{}. To fix this problem, report to the installation 
manual. Remember on \texttt{MacOSX} that you must start \mma{} from a shell window, and not
directly by double-clicking on the \mma{} icon: in this case, \mma{} does not import the 
environment variables.
\item Does the environment variable \texttt{\$PATH} contain the directory where the binaries
of \MMAlpha{} are ? If this is not the case, \texttt{domlib} cannot be loaded. 
Check this by evaluating
\begin{verbatim}
$Path
\end{verbatim}
and checking that there appears a path of the form \texttt"{... /mmalpha/bin.ostype"} where 
\texttt{ostype} depends on your operating system. For \texttt{MacOSX}, is should be
\texttt{bin.darwin}, for \texttt{Linux}, it should be \texttt{bin.linux}, etc. To fix this, report to
the installation manual and set the environment variable. You can get the value of your
OS type by evaluating \texttt{Environment[ "OSTYPE"]} in \mma{}. 
\item Does the directory of the binaries exist? To check this, get the value of environment 
variable \texttt{\$OSTYPE}, and check that there exist a directory \texttt{bin.ostype} in 
your \MMAlpha{} repository. If such a directory does not exist, the problem has to be
reported to the author of this note... 
\item Even if the \texttt{bin.ostype} directory exists, it may not contain the proper 
binary files: \texttt{domlib}, etc. 
\end{enumerate}

\end{document}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/appendixdemos.tex
Added mmalpha (source and binary)

\chapter{The demos of {\mmalfa}}
\label{demos}

{\sc Version of \today{}}

This chapter presents a few demos of
{\mmalfa}.\index{demos,demonstrations} 
There two types of demos: some demos
have a notebook interface, and the others
should be executed in kernel mode. There are also various
documentations available with the \alfa{} archive.

\section{Notebook demos}
All these demonstrations can be accessed directly by 
hyperlinks of the 
so-called {\em master} notebook, which is started by
evaluating the command \texttt{start[]}.

\begin{description}
\item [Introduction.] Introduction to {\Alpha}

\item [Domlib.] Elementary computations on domains with Domlib.

\item [Fir: Design of a FIR filter.]  The FIR demo shows the typical
steps for the synthesis of a linear architecture for a Finite Impulse
Response filter.  The demo comprises the following steps: loading the
initial program, pipelining, scheduling, assign parameter values,
control signal generation, {\alphaz} generation, {\alphard} generation.

\item [matvect: architecture for matrix vector
multiplication.]

The MATVECT demo shows the typical steps for the
synthesis of a linear architecture for a Matrix Vector Multiplication.
The demo comprises the following steps: loading the initial program,
pipelining, scheduling, assign parameter values, control signal
generation,  {\alphaz} generation, {\alphard} generation.

\item[Tutorial\_intro: ]
very introductory tutorial. Corresponds to document~\cite{first-tutorial}.

\item[mma-intro: ]
quick introduction to \mma{}. Explains how Alpha ASTs are represented
in \mma{}. Warning: the abstract syntax presentation may not be as reliable as
the appendix of this manual (see chapter~\ref{appendixalpha}).
\item[Kalman: ]
this example describes an application of \mmalfa{} to a Kalman
filter. This application, quite complex, shows how to use 
structured scheduling. Reference~\cite{parelec} describes the 
background of this application. 
\item[Kalman-Sqrt: ]
this example describes an application of \mmalfa{} to a Kalman
filter, using a so-called square root covariance method. 
This application, quite complex, shows how to use 
structured scheduling. Reference~\cite{cce} describes the 
background of this application.

\item[Neural-Network: ]
this notebook is not ready yet, but will be soon.
\item[Fuzzy-Logic: ]
an application of fuzzy logic to channel equalization. 
\end{description}

Another serie of notebook allows one to become familiar with
different packages of \mmalfa{}: 
\begin{description}
\item[Alpha: ] the Alpha package.
\item[Control: ] the Control package.
\item[Matrix: ] the Matrix package.
\item[ChangeOfBasis: ] the ChangeOfBasis package.
\item[Normalization: ] the Normalization package.
\item[Cut: ] the Cut package.
\item[Decomposition: ] the Decomposition package.
\item[Static: ] the Static package.
\item[PipeControl: ] the PipeControl package. 
\item[Vhdl: ] the Vhdl package.
\item[CheckAlpHard: ] the AlphHard package.
\item[Meta: ] the Meta package.
\item[Schedule: ] the various schedule packages.
\item[dataFlowSchedule: ]

\item[matlib: ]

\item[Visual: ] the Visual package.

\end{description}

\section{kernel demos}
These demos are avaiblable at the following address:
\begin{verbatim}
/udd/alpha/alpha\_beta/Mathematica/demos/demos.html
\end{verbatim}

\begin{description}
 \item [INIT: An initiation to {\mmalfa}.] The INIT demo shows a few
  basic functions of MMAlpha. The demo comprises the following steps:
  loading and showing a program, normalizing, showing domains etc...
  \index{INIT demo}

 \item [FIR: Design of a FIR filter.]  The FIR demo shows the typical
steps for the synthesis of a linear architecture for a Finite Impulse
Response filter.  The demo comprises the following steps: loading the
initial program, pipelining, scheduling, assign parameter values,
control signal generation, {\alphaz} generation, {\alphard} generation.\index{FIR demo, Finite impulse
response filter}

 \item [MATVECT: architecture for matrix vector
multiplication.] 
The MATVECT demo shows the typical steps for the
synthesis of a linear architecture for a Matrix Vector Multiplication.
The demo comprises the following steps: loading the initial program,
pipelining, scheduling, assign parameter values, control signal
generation,  {\alphaz} generation, {\alphard} generation.\index{MATVECT demo,
Matrix vector architecture}

% \item [VERIF: Verification of the Matrix vector architecture.]
%The VERIF demo shows the verification of a linear architecture for a
%Matrix Vector Multiplication. The demo shows the use of subsystems,
%the inlining of subsystems, the use of libraries, C code generation
%and execution.\index{VERIF demo}

 \item [SVD: analysis of a Singular Value Decomposition
algorithm.]  The SVD demo shows the analysis of a larger Alpha
program. It shows the use of subsystems, the nlining subsystems, the
CNF code generation, the C code generation and execution.

% \item [SVDUPDATE: another demo on Singular Value Decomposition.]
%The SVD demo shows the analysis of another version of the SVD algorithm.

 \item [BINMULT: Synthesis of a multiplier.]  This demonstration
shows the synthesis of a bit-serial multiplier, from an initial
specification structured into several parameterized subsystems.  The
demo script contains the following steps: loading and displaying a
multi-system program, conversion into C code for simulation,
pipelining of the diffusions in the initial program, flattenning all
the systems for synthesis, scheduling before space/time
transformation, space/time transformation, pipelining of the inputs
(so that all the data enters the first processor in bit-serial
manner), simulation of the resulting array.

\item [ESTIM: synthesis of motion estimation] a two dimensionnal example, with 
simulation, on a real image, derivation of {\alphaz} code (warning, long demo).
\end{description}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/appendixtechset.tex
Added mmalpha (source and binary)

\chapter{Installing \mmalfa{}}
\label{appendixtechset}
\label{install}

\newcommand{\mmalpha}{{MMalpha}}
\input{install1.tex}

\section{Getting started}
You may use either {\mma} under {\tt emacs}, or directly on the Unix 
system. 
\begin{itemize}
\item under {\tt emacs}, type {\tt  esc-X Mathematica}. 
\item type math (use the {\mma} textual interface).
\item type mathematica (use the {\mma} textual interface).
\end{itemize}
%Do not use a Notebook, as the output of the pretty printer would not
%be available.
When the {\mma} session is started, you can type {\tt goDemo["INIT"]}
for the initiation demonstration.

When the {\mma} session is started, you can type {\tt on[]} or
\texttt{start[]} for the initiation demonstration.

\section{{\mmalpha} file hierarchy}
\includegraphics[width=12cm]{figures/fileHierarchy}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapalpha0.tex
Added mmalpha (source and binary)

\chapter{The Alpha0 format}
\label{chapalfha0}

{\sc Version of \today}

In this chapter, we describe the {\Alpha0} format which is basically a non-structured version of the {\AlpHard} language. We also 
 describe the methodology to go from an {\Alpha} program to an {\Alpha0} program and then how to translate it into {\AlpHard}.

\subsection{Definition of {\AlphaZ}}
{\AlphaZ} was introduced by C. Dezan in her Phd thesis
(\cite{dezan}). It was conceived as the lowest level of {\Alpha} or
equivalently as a subset of {\Alpha} which can describe circuits at the 
register transfer level. the main weakness of this subset came from the fact
that {\Alpha} programs were not structured.  Since, the structuring of
{\Alpha} was provided by F. Dupont~\cite{DupontQuRi95} and the
structured version of {\AlphaZ} was studied by P. Le
Mo\"enner~\cite{Patricia96}.  {\AlpHard} (described in
chapter~\ref{chapalphard}) is now intended to be this subset of
{\Alpha} with structural interpretation, from which we can translate
{\Alpha} into {\vhdl} or other {\sc rtl} description
languages. However, during the transformation path from {\Alpha} to
{\AlpHard}, the automatic structuring is a complicated process, hence
it appeared that the {\AlphaZ} format is still necessary as an
intermediate format with all the register transfer level informations
but in an unstructured form. We describe hereafter, the second version
 the {\AlphaZ} format (Alpha0v2) which is slightly different from the version described in~\cite{dezan}

% As a platform to {\AlpHard}, we need to remove the second point of
% the previous definition, namely the fact that all temporal domains are
% of the form $\{t~|~ t>=\}$. The other basic principles remains the same
% with slight differences. spatial cases contain no dependencies,
% connections are represented using a simple space dependency (no
% case). Temporal cases are always nested in a spatial case.
% Control initialization  equations are called {\em control} equation,
% the {\em Mirror} equations are the equations affecting  input, output 
% and control signal to local variables (hence, control signal 
% become data signals). All these descriptions are summarized by the following 
% definition:

An {\Alpha} program is in {\AlphaZ} form if it contains no {\tt use} statement
and if it meets the following conditions:

\begin{enumerate}
\item there exits, for all declared domains (except for the domains of 
inputs and outputs variables) an interpretation function for the indices
(see~\cite{dezan} for precise definition of {\em interpretation function}).
 Basically, each index is either 
a temporal index (indicating living dates of the signal) or a spatial index
(indicating in which processor the signal lives).

\item The equations of the system define the output and local
 variables.  Each operator involved in the equations has a structural
 interpretation. these equations are of four types: data equations,
 connection equations, control equations and mirror equations.
\begin{itemize} 
\item data equations define the different signals of the program
which are necessarily local variables.
They are composed of the following operators:  
\begin{itemize} 
  \item pointwise operators represent the corresponding combinatorial 
operators. 
  \item restriction are used to restrict the spatial interpretation of 
indices.
\item  dependencies are temporal dependencies representing registers 
or identity dependencies representing connections between two signal inside one cell.
\item case are spatial case allowing to gather several signals.
\item the {\tt if} operator (with nested {\tt case} in each branches)
is interpreted as a multiplexer.
\end{itemize}
\item connection equations are limited to the following form: a single
spatial dependency between two signals ({\tt A=B.(t,p->t,p-1)} for
instance).
\item control equations allow initialize control signals. they define signals
which are only temporal (no spatial indices).
\item Mirror equations: equation between input variables of the system and 
local variable of the system (only for interface). equation limited to an 
affine function applied to a variable ({\tt A[t,p]=a[f(t,p)]} for input mirror
equation and {\tt b[i,j]=B[f(i,j)]} for output mirror equations)..
\end{itemize}
\end{enumerate}

Examples of {\AlphaZ} programs are present in the directory 
{\tt \$MMALPHA/examples/Alpha0}, a small example is presented hereafter
 (figure~\ref{fig04}).

\section{From {\Alpha} to {\AlphaZ}}
The first attempt to go from High level specification in {\Alpha}
to {\sc rtl} specification in {\Alpha0} has been done be Sie and 
Wilde~\cite{SieWilde94}. This methodology has been 
extended and we discribe it briefly here, more informations
can be obtained by executing the Notebooks demos: {\tt Fir} and {Matvect}.

Consider the very simple program of figure~\ref{fig01}

{\small
\begin{figure}[h]
\begin{verbatim}
system NtimeNot :{N | 2<=N}
                (a : {i | 1<=i<=N} of boolean)
       returns  (b : {i | 1<=i<=N} of boolean);
var
  Acc : {i,j | 1<=i<=N; 1<=j<=N} of boolean;
let
  Acc[i,j] = 
      case
        {| j=1} : a[i];
        {| 2<=j} : not Acc[i,j-1];
      esac;
  b[i] = Acc[i,N];
tel;
\end{verbatim}
\caption{An {\Alpha} program computing {\tt N} times {\tt Not} 
on an array {\tt a}}
\label{fig01}
\end{figure}
}

Translating the program in {\AlphaZ} consists in the following steps: 
\begin{enumerate}
 \item {\bf Uniformize the program}. Here the program is already
uniform, in general this procedure may be complicated {\mmalpha}
provides tools for the automatic or designer-guided uniformization
(see the documentation on the chapter~\ref{chapuniformize}).
\item {\bf Schedule the program}. It consists in finding an execution
date (and a place also) for each computation. Usually, once
uniformized, the program should be scheduled with a unique linear part
(see the chapter~\ref{chapscheduling} for detail on scheduling). The
function {\tt applySchedule} choose an allocation function. In our example, 
you can execute  the following commands:
\begin{verbatim}
schedule[]
applySchedule[]
renameIndices[{"t","p"}]
\end{verbatim}
We obtain a linear array of processors,
the resulting program is shown in figure~\ref{fig02}
{\small
\begin{figure}[h]
\begin{center}
\begin{verbatim}
system NtimeNot :{N | 2<=N}
                (a : {i | 1<=i<=N} of boolean)
       returns  (b : {i | 1<=i<=N} of boolean);
var
  Acc : {t,p | 1<=t<=N; 1<=p<=N} of boolean;
let
  Acc[t,p] = 
      case
        {| t=1} : a[p];
        {| 2<=t} : not Acc[t-1,p];
      esac;
  b[i] = Acc[N,i];
tel;
\end{verbatim}
\caption{Program of figure~\ref{fig01} after applying schedule}
\label{fig02}
\end{center}
\end{figure}
}

\item {\bf Pipeline Inputs and Output}. You can see in figure~\ref{fig02}
that the  {\tt a} is input in every processors. If we want only the first processor to input data from the host, we could have pipelined the input with the following command:\\
\verb/pipeIO["Acc","a.(t,p->p)","aIn.(t,p->t+1,p+1)","{t,p| p>=1}"]/. We can do the same treatment for the output. Here we chose not to pipeline the I/O for sake of simplicity.

\item {\bf Go Down to Alpha0}. When the previous steps have been correctly 
executed, this step should be automatic with {\tt toAlpha0v2[]} command. In 
Our case, if we apply this function to the program of figure~\ref{fig02},
the printing on the screen are the following:
\begin{verbatim}
In[__]:= toAlpha0v2[];
Time index: {1}  space indices: {2}
Calling spaceTimeDecomposition[];
Calling makeAllMuxControl[];
  Equation of Acc...
   is in ST form
    Adding multiplexer.
  Equation of b...
Calling pipeAllControl[];
  Pipelining control for: Acc_ctl1
     From dimension 2 To dimension 1
       Warning, no pipe vector was found for control signal Acc_ctl1
       --> assuming broadcasted signal
     Control generated in cell: {p | 1<=p<=N; 2<=N}
Calling decomposeSTdeps[];
  In equation of Acc, adding a local variable: Acc_reg1
 Decomposing the space/time dependencies
Calling makeInputMirrorEqus[];
  Adding mirror equation for input a
Out[__]:=
\end{verbatim}

The important treatment here is the generation and pipeline of a
control signal. Indeed, the condition {\tt t=1} of the program of
figure~\ref{fig02} has to be realized in hardware.  Hence a control
signal was generated and a multiplexer added in each processors. Then
{\mmalpha} tries to pipeline the control signal (it often happens in
systolic architectures that the control can be pipelined). In our
particular example, all processors start their computation
simultaneously, hence the control signal is not pipelined but
broadcasted. Other treatments are just syntactic rewriting. The result
{\AlphaZ} program is show on figure~\ref{fig03}. note the precise information
that appears on the life time of each signal (in this simple program,
each signal lives from step 1 to step {\tt N}, but in general it is much 
more complicated) and on the input date of each data (mirror equation 
contains the information of {\em how to use} the hardware -- i.e. interface 
specification --).
\end{enumerate}
{\small
\begin{figure}[h]
\begin{verbatim}
system NtimeNot :{N | 2<=N}
                (a : {i | 1<=i<=N} of boolean)
       returns  (b : {i | 1<=i<=N} of boolean);
var
  a_mirr1 : {t,p | t=1; 1<=p<=N} of boolean;
  Acc_reg1 : {t,p | 2<=t<=N; 1<=p<=N} of boolean;
  Acc_ctl1_In : {t,p | 1<=t<=N; 1<=p<=N} of boolean;
  Acc : {t,p | 1<=t<=N; 1<=p<=N} of boolean;
  Acc_ctl1 : {t | 1<=t<=N} of boolean;
let
  a_mirr1[t,p] = a[t+p-1];
  Acc_reg1[t,p] = Acc[t-1,p];
  Acc_ctl1_In[t,p] = Acc_ctl1[t];
  Acc_ctl1[t] = 
      case
        case
          {| t=1} : True[];
          {| 2<=t} : False[];
        esac;
      esac;
  Acc[t,p] = 
      case
        {| 1<=t} : if (Acc_ctl1_In) then 
               case
                 {| t=1} : a_mirr1;
                 {| 2<=t} : False[];
               esac else 
               case
                 {| t=1} : False[];
                 {| 2<=t} : not Acc_reg1;
               esac;
      esac;
  b[i] = Acc[N,i];
tel;
\end{verbatim}
\caption{Program {\AlphaZ} derived from the program of
figure~\ref{fig01}.  This {\AlphaZ} program describes a linear array
of {\tt N} processors.  The equation defining {\tt Acc} represent a
multiplexer controlled by {\tt Acc\_ctl1} selecting {\tt a\_mirr1} or
{\tt Acc\_reg1}. The equation defining {\tt Acc\_ctl1\_In} is a control
equation. The equations defining {\tt a\_mirr1} and {\tt b} are mirror
equations. The equation defining {\tt Acc\_reg1} is interpreted as a
register. The equation defining {\tt Acc\_ctl1\_In} is interpreted as 
a connection equation: broadcast of the control  signal {\tt Acc\_ctl1}
to all processors.}
\label{fig03}
\end{figure}
}
\section{From {\AlphaZ} to {\AlpHard}}
The translation to {\AlpHard} is automatic. the command to use is {\tt
alpha0toAlphard[]}. {\mmalpha} automatically detects all the spatial
regions on which the computations are identical (the whole spatial
region in our case: {\tt $\{ p \| 1 \leq p \leq N\}$} and structures
the program with one cell (i.e. one subsystem) for each region, one
controller for initialization of control signal, one module for the
complete circuit and one interface to keep the same I/O as the
original program of figure~\ref{fig01}. The interface system obtained in our 
example is shown on figure~\ref{fig04}.
{\small
\begin{figure}[h]
\begin{verbatim}
system NtimeNot :{N | 2<=N}
                (a : {i | 1<=i<=N} of boolean)
       returns  (b : {i | 1<=i<=N} of boolean);
var
  a_mirr1 : {t,p | t=1; 1<=p<=N} of boolean;
  Acc : {t,p | 1<=t<=N; 1<=p<=N} of boolean;
let
  a_mirr1[t,p] = a[t+p-1];
  b[i] = Acc[N,i];
  use  NtimeNotModule[N] (a_mirr1) returns  (Acc) ;
tel;
\end{verbatim}
\caption{interface of the {\AlpHard} Program obtain from the program of figure~\ref{fig03} by the {\tt alpha0ToAlphard[]} command}
\label{fig04}
\end{figure}
}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapalphaarray.tex
Added mmalpha (source and binary)

\chapter{Arrays of synchronous operators}
\label{chapalphaarray}
In this chapter, we show how arrays of hardware operators can be modelled in
{\alfa}\index{arrays}. In chapter~\ref{chapalphascal}, the basic
operators of the language were described. As we shall see here, a 
simple extension of these operators allows two-dimensional -- and
higher-dimensional -- variables and expressions to be used.

\section{Describing array of elements}
Figure~\ref{adder-array} shows an array of adders. 
\section{An array of adders}
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
%\colorbox{white}{
\centerline{\includegraphics{figures/exemple_9}}
%}
\caption{An array of adders}\label{adder-array}
\end{figure}

The corresponding {\alfa} code is shown on figure~\ref{adder-array-alpha}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder-array.alpha}
\caption{An array of adders in {\alfa}}\label{adder-array-alpha}
\end{figure}
Blabla
\section{Change of basis}
Any {\alfa} program can be seen as a black box which takes input
variables, and produces output variables. Therefore, intermediate
variables are unimportant, and, in particular, one can reindex them
using one-to-one index mappings. Obviously, 
\begin{verbatim}
A[i] = B[i];
B[i] = C[i];
\end{verbatim}
is identical to 
\begin{verbatim}
A[i] = B[i-1];
B[i] = C[i+1];
\end{verbatim}
 The change of basis operation does just this reindexing. In the
context of {\alfa}, the reindexing functions that we consider need to
be restricted to affine {\em unimodular functions},\index{affine
functions}\index{unimodular functions} that is to say, affine mappings
which admit an integral inverse.

Consider variable {\tt Z} in program~\ref{adder-array-alpha}. Consider
the reindexing function $B$ from $\Z^2$ to itself, defined by $B(t,p)
= (t+p,p)$. One can show that $B$ is unimodular. Let us denote
$B$ using the {\alfa} syntax, as {\tt (t,p -> t+p,p)}. This function 
is one-to-one, and its inverse is $B^{-1}$, denoted {\tt (t,p -> t-p,p)}.

To apply the change of basis defined by $B$ on variable {\tt Z}, 
we need to do the following operations:
\begin{enumerate}
\item Replace the domain of definition of {\tt Z} by its image under
the $B$ mapping. As the domain of {\tt Z} is 
{\tt \{t,p|t>=1;4>=p;p>=0\}}, its image by $B$ is 
{\tt \{t,p | p+1<=t; 0<=p<=4\}}.
\item Replace any occurrence of {\tt Z} in a right-hand side of 
an equation by {\tt Z.(t,p -> t+p,p)}.
\item Replace the entire equation {\tt Z = expr} defining {\tt Z}
by {\tt Z = expr.(t,p -> t-p,p)}.
\end{enumerate}

The new program is shown in figure~\ref{adder-array-bn-alpha}. We can normalize
it, which leads to program~\ref{adder-array-cb-alpha}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder-array-bn.alpha}
\caption{Program adderarray, after change of basis {\tt (t,p->t+p,p)}
performed on {\tt Z}}\label{adder-array-bn-alpha}
\end{figure}

This new program can be interpreted as the architecture shown in 
figure~\ref{adder-array-cb-tex}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder-array-cb.alpha}
\caption{Program adderarray, after change of basis 
and normalization}\label{adder-array-cb-alpha}
\end{figure}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
%\colorbox{white}{
\centerline{\includegraphics{figures/exemple_10}}
%}
\caption{An array of adders}\label{adder-array-cb-tex}
\end{figure}

% Reduction
%\centerline{\input{/udd/quinton/documents/lncs_574/seminaire_litp/figures/exemple_8}}
%\input{/udd/quinton/documents/lncs_574/seminaire_litp/exemples_alpha/exemple_8}
%
%\transparent{0cm}{{Convolution}}{0cm}
%{\small
%\input{/udd/quinton/cours/cours_nouveau/convgen}
%}
%\transparent{0cm}{{Convolution apr\`{e}s s\'{e}rialisation}}{0cm}
%{\small
%\input{/udd/quinton/cours/cours_nouveau/conv}
%}
%\transparent{0cm}{{Convolution, apr\`{e}s changement de base}}{0cm}
%{\small
%\input{/udd/quinton/cours/cours_nouveau/conv1}
%}
%\begin{slide}{}
%\titre{Conception typique (Mathematica Alpha)}
%{\small
%\begin{enumerate}
%\item Analyse syntaxique, v\'{e}rification
%\item Placement des variables (manuel)
%\item Uniformisation (automatique)
%\item Ordonnancement (automatique)
%\item Changement  de base (automatique)
%\item Separation espace et temps
%\item Remplacement des {\bf case} par des signaux de contr\^{o}le (automatique)
%\item Generation circuit (MadMacs ou VHDL) 
%\end{enumerate}
%}
%\end{slide}

\subsection*{Processing this example in {\mmalfa}}
Load the array of adders:
\begin{verbatim}
load["adder-array.alpha"];   (* loads the alpha file *)
ashow[];                     (* shows its content *)
\end{verbatim}
Apply a change of basis:
\begin{verbatim}
changeOfBasis["B.(t,p->t+p,p)"];  (* apply a change of basis *)
ashow[];                     (* shows the result *)
\end{verbatim}
Notice that the result of a change of basis is automatically normalized.

\section{Conclusion}
In this chapter, we have seen how arrays of elements can be 
described using {\alfa}. In order to present hardware examples, we
have limited ourself to programs where the indexes can be
interpreted as the time and the space (i.e., {\tt t} and {\tt p}).
Actually, {\alfa} programs are not restricted to such indexes. 
{\em Au contraire}, hardware-like {\alfa} programs represent the 
ultimate goal of a synthesis process which starts from a 
behavioural specification. They are in fact a proper subset of all 
{\alfa} programs.

In Chapter~\ref{chapfunctional}, we shall consider {\alfa} from the
data parallel perspective.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapalphard.tex
Added mmalpha (source and binary)

\chapter{The {\alfhard} language}
\label{chapalfhard}

{\sc Version of \today}

In this chapter, we describe the {\alfhard} language. We also 
 describe the {\vhdl} generator.

\section{\AlpHard}
\label{alphard}
The main goal of \alfa{} is to allow someone to produce 
a hardware implementation for an \alfa{} specification. To this
end, a subset of \alfa{}, called \AlpHard{}, is defined. 
We quickly introduce this language.

\subsection{Basic concepts}

{\AlpHard} is a subset of the {\Alpha} language that enables a
structural definition of a regular architecture (systolic arrays,
etc.) to be given.

{\AlpHard} is intended to meet two important goals.  First, {\AlpHard}
programs are obtained by automatic transformation of {\Alpha}
programs, and hence, \AlpHard{} it must be a coherent subset of
{\Alpha}. Second, the architectural description given by {\AlpHard}
must:
 \begin{itemize}
 \item Provide {\em structuring} because complex design process must be
   hierarchical.
 \item Provide {\em genericity} in order to allow component reuse.
 \item Allow {\em regularity} to be described, in order to reuse hardware
   descriptions and simplify the design process.
\end{itemize}

Thus, {\AlpHard} is hierarchical.  At the lowest level of description
  one has {\em cells} consisting of combinational circuits, registers,
  multiplexors, etc.  A cell may also contain other cells.  At the
  same level in the hierarchy, we may also have {\em controllers}
  responsible for initialization\footnote{In systolic arrays, the
  control signals are themselves distributed in a regular fashion, and
  their data-path can also be described in terms of cells.}.  At this
  point we have a description of a piece of circuit that has no
  ``spatial dimension'': it represents a single processing element
  which may later be instantiated at multiple spatial locations (like
  the full adder presented in section~\ref{sectstruct}).

The next level of the hierarchy is the {\em module} which allows the
user to specify how different cells are assembled regularly in one or
more spatial dimensions. This is achieved by {\em instantiating}
previously declared cells. Typically, controllers are instantiated
only once with no spatial replication. The separation of temporal and
spatial aspects is also reflected by the fact that the equations
describing the behavior of cells have only one local index variable
(time), and in the equations for modules, the {\em dependencies} have
only spatial indices, indicating pure (delay-less) interconnections
(thus all registers are viewed as part of the cells).

\subsection{A small example}

These ideas are illustrated in figure~\ref{figstupid}
and~\ref{figstupid2}, which show a three-cell module, where each cell
is a simple register-inverter. The {\AlpHard} code corresponding to
the {\em cell} definition is in figure~\ref{figstupid2}. 
The 
system {\tt RegInvCell} has two parameters (lines 1) corresponding
to its start time and its duration.  Lines 3-5 are the declaration of
input and output signals of the cell, the {\em domain} between the
braces represents the lifetime of signals {\tt A1} and {\tt B1}.  The
equation defining {\tt B1} (line 7) consists of an
inverter combined with a delay {\tt A1[t-1]} which 
represents a register.  Notice that the only index appearing in the
equations is the time, {\tt t}.
{\small
\begin{figure}[h]
\begin{verbatim}
system                                                          -- 1  
RegInvCell:{Tinit,Duration| Tinit,Duration >= 0}                -- 2  
           (A1 : {t | Tinit<=t<=Tinit+Duration} of boolean)     -- 3  
    returns                                                     -- 4  
           (B1 : {t | Tinit+1<=t<=Tinit+Duration+1} of boolean);-- 5  
let                                                             -- 6  
        B1[t] = not A1[t-1];                                    -- 7 
tel;                                                            -- 8

\end{verbatim}

\begin{verbatim}
system
RegInvModule:{Tinit,Duration,Size| Tinit,Duration, Size > 0}     -- 9  
	(a: {t|  Tinit<=t<=Tinit+Duration} of boolean)           -- 10
   returns                                                       -- 11  
	(b: {t|  Tinit+Size<=t<=Tinit+Duration+Size} of boolean);-- 12
var                                                             
A : {t,p | Tinit+p-1<=t<=Tinit+Duration+p-1; 1<= p <= Size}  of boolean;  
B : {t,p |  Tinit+p<=t<=Tinit+Duration+p; 1<= p <= Size}  of boolean;
let
use {p | 1<= p <= Size} RegInvCell[Tinit+p-1,Duration]           -- 17
                        (A) returns (B);                         -- 18
A[t,p] = case                                                    -- 19   
      {| p=1}: a[t];                                             -- 20   
      {| p>1}: B[t,p-1];                                         -- 21   
      esac;                                                      -- 22   
b[t]=B[t,Size];                                                  -- 23
tel;                                       
\end{verbatim}
\caption{An {\AlpHard} program describing a simple cell, ({\tt
    RegInvCell}), and instantiation of {\tt Size} copies of it in a module
({\tt regInvModule}).
  Note how {\tt p} is used to specify the value {\tt Tinit} for each instance
  (the {\tt p}-th cell starts {\tt p-1} cycles after the first one).}
\label{figstupid}
\end{figure}
}

The {\tt RegInvModule} system is a {\em module} which uses the cell
 {\tt RegInvCell} .  Multiple copies are instantiated by means of the
construct,
\verb.use {p|1<=p<=Size}..  Such a multiple instantiation also means that the
variables {\tt A} and {\tt B} will have two indices, {\tt t}
(inherited from the {\tt RegInvCell} definition) and {\tt p} (from the
multiple use in line 17).  We also see that the value of the {\tt Tinit}
parameter is instantiated as a function of {\tt p} so that the cells start at
different instants of time. Lines 19-22 specify the connections between the
input/output of the cells in the module.  Notice that the constraints in
the {\tt case} involve only spatial indices (if it had involved temporal 
indices --i.e. {\tt t}--, we would need some additional control hardware not described here).  Also note that the time index,
{\tt t}, is the same on both sides of the equation (no register outside the cells).

\begin{figure}[ht]
$$\includegraphics{figures/stupid1}$$
\caption{A simple architecture consisting of three identical cells}
\label{figstupid2}
\end{figure}

For more information on {\AlpHard} see~\cite{Patricia96}
\cite{Lemoenne} and there are 
example of Alphard programs in the the directory 
{\tt \$MMALPHA/examples/Alphard}

To have information on
the {\vhdl} generator, see the notebooks: 
\begin{itemize}
\item matvect demonstration,
\item Fir demonstration,
\item Vhdl.
\end{itemize}
They are all available as hyperlinks in the \texttt{master.nb}
that you can start using the command \texttt{on[]} or
\texttt{start[]}.

\section{generating {\vhdl} from {\AlpHard}}
The {\AlpHard} format was conceived by P. Le Moënner together
with the translator to {\vhdl}. The targeted {\vhdl} is 
{\em synthetizable \vhdl}, it has been tested on commercial tools like
 Compass, Synopsis,\ldots. Currently, the {\vhdl} translator allows
to generate {\vhdl} code from {\AlpHard} programs which represent
 linear systolic arrays. The reason why we cannot generate code for 
2-dimensional arrays is that the {\em synthesizable  \vhdl 
subset}~\cite{VHDLSynth} does not allow nested {\tt generate} instructions\footnote{The {\vhdl}  {\tt generate} instruction is used for expressing the 
repetitive use of a cell in an array}.

\subsection{Setting up the translation}
the only operation  that have to be done before translating 
{\AlpHard} code is to affect the parameters value in the module and the controler
(we cannot
described arrays with parametric size in {\vhdl}). As the current {\AlpHard}
program is a library, this process may be a little complicated (you have 
to ensure that the parameter instantiations between the caller and the 
called program is respected). Currently this can be done with
 {\tt assignParameterValue} function. this function assign the parameter value 
in one system ({\tt \$result}), if you want to recursively assign the parameter 
value to the systems called by {\tt \$result}, you can use the
 {\tt fixParameter} function which assigns a value to a given parameter for 
all systems in {\tt \$library}\footnote{You have to ensure that this parameter 
has indeed exactly the same value in all systems of the library, this is not the case for {\tt Tinit} in our example}. In the example of figure~\ref{figstupid}, 
you can use the following commands:\\
\begin{verbatim}
 assignParameterValue["Size",10];
assignParameterValue["Tinit",0];
assignParameterValue["Duration",100]
\end{verbatim}
The resulting program is shown in figure~\ref{figstupid3}. The command
{\tt alphardToVHDL[]} will generate one file per system of the library.
Here, one file for the cell (named
{\tt regInvCell.vhd}) and one file for the module (named
{\tt regInvModule.vhd}). The {\vhdl} code for the
cell {\tt RegInvCell}  and the module {\tt RegInvModule}  are
 shown in figure~\ref{figVH1} and~\ref{figVHD2}

{\small
\begin{figure}[h]
\begin{verbatim}
system RegInvCell :{Tinit,Duration | 0<=Tinit; 0<=Duration}
                  (A1 : {t | Tinit<=t<=Tinit+Duration} of boolean)
       returns    (B1 : {t | Tinit+1<=t<=Tinit+Duration+1} of boolean);
var
  A2 : {t | Tinit+1<=t<=Tinit+Duration+1; 0<=Tinit} of boolean;
let
  A2[t] = A1[t-1];
  B1[t] = not A2;
tel;
\end{verbatim}
\begin{verbatim}
system RegInvModule (a : {t | 0<=t<=100} of boolean)
       returns      (b : {t | 10<=t<=110} of boolean);
var
  A : {t,p | p-1<=t<=p+99; 1<=p<=10} of boolean;
  B : {t,p | p<=t<=p+100; 1<=p<=10} of boolean;
let
  use {p | 1<=p<=10} RegInvCell[p-1,100] (A) returns  (B) ;
  A[t,p] = 
      case
        {| p=1} : a[t];
        {| 2<=p} : B[t,p-1];
      esac;
  b[t] = B[t,10];
tel;
\end{verbatim}
\caption{The  {\AlpHard} program of figure~\ref{figstupid} with particular 
value for the parameters of the module. Note that the cell is still parameterized.}
\label{figstupid3}
\end{figure}
}

{\small
\begin{figure}[h]
\begin{verbatim}
-- VHDL Model Created for "system RegInvCell" 
 --  20/5/1999 11:27:36

library IEEE;
use IEEE.std_logic_1164.all;
library COMPASS_LIB;
 use COMPASS_LIB.STDCOMP.all;
library COMPASS_LIB; 
 use COMPASS_LIB.COMPASS.all;

entity RegInvCell is
      Port ( Ck : In std_logic;
      A1 : In std_logic;
      B1 : Out std_logic );
end RegInvCell;

architecture Behavioral of RegInvCell is

  signal A2 : std_logic;

begin

  process(ck)
  begin
    if (ck='1' AND ck'event) then
        A2 <= A1;
    end if;
  end process;

  B1 <= ( not A2);

end Behavioral;
\end{verbatim}
\caption{{\vhdl} code generated from  the {\tt RegInvCell} cell of figure~\ref{figstupid3}}
\label{figVHD1}
\end{figure}
}

{\small
\begin{figure}[h]
\begin{verbatim}
todo
\end{verbatim}
\caption{{\vhdl} code generated from  the {\tt RegInvModule} cell of 
figure~\ref{figstupid3}}
\label{figVHD2}
\end{figure}
}

\subsection{Description of the generated \vhdl}
{\tt todo:} describe the translation of operators, registers, multiplexer,
controler, cells, modules.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapalphascal.tex
Added mmalpha (source and binary)

\chapter{Modelling synchronous architectures}
\label{chapalphascal}

\begin{tobedone}
\begin{enumerate}
\item Redraw figures (after reinstalling X11 and xfig)
\end{enumerate}
\end{tobedone}

In this chapter, we present the {\alfa} language through the 
description of very simple -- and seemingly artificial -- examples
of synchronous circuits\index{synchronous circuits}. Our goal is to show the basic constructors
of the language: their syntax, their meaning, and how they can be
used in practice. 
% 
%\newpage
\section{Pointwise operators\index{pointwise operators}}
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
%\colorbox{white}{
\centerline{\includegraphics{figures/exemple_1}}
\caption{An adder\label{adder1}}
\end{figure}
Consider the architecture\index{architecture} shown in figure~\ref{adder1}: there is a 
simple adder\index{adder}, 
which takes to input flows, respectively {\tt x} and 
{\tt y}, and returns {\tt S}.

This architecture is described by the {\alfa} program of figure \ref{adder-alpha}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder.alpha}
\caption{An adder in {\alfa}}\label{adder-alpha}
\end{figure}
A program is called a {\em system}\index{system}. 
Line 1 is the head of the system, 
with name {\tt exemple1}. This system has two input arguments
named
{\tt x} and {\tt y}. These arguments, called {\em variables}\index{input
variables}, are
defined on the set {\tt \{t | 1 <= t\}}. Line 3 defines a local variable\index{local variable}
{\tt S} (for the moment, this variable is useless, but we shall
soon find a use for it). 
Between the {\tt let} and {\tt tel} keywords of lines 4 and
7, we have two equations\index{equation}: 
line 5 says that {\tt S} is the pointwise sum of 
{\tt x} and {\tt y}. In other words, $\forall t, S_t = x_t + y_t$. 
Line 6 says that {\tt z} and {\tt S} are identical.
\subsection*{Processing this example in {\mmalfa}}
Here are the Mathematica expressions that allow one to try
the previous example under MMAlpha.
%A more complete introduction to {\mmalfa} is given in 
%appendix~\ref{mma-introduction}.
\begin{verbatim}
load["adder.alpha"]    (* loads the alpha file *)
ashow[];               (* shows its content *)
\end{verbatim}

%
%\newpage
\section{Delays as dependence fonctions}
We now modify this example, by adding a delay\index{delay} on the output, as
shown in figure \ref{adderDelayed}. 
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exemple_2}}
\caption{Adder delayed}\label{adderDelayed}
\end{figure}
The corresponding {\alfa} program is shown in figure \ref{adderDelayed-alpha}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adderDelayed.alpha}
\caption{Adder plus delay}\label{adderDelayed-alpha}
\end{figure}
In this example, the definition domains\index{domain} of {\tt z} and {\tt S} have
changed: indeed, the output of the adder\index{adder (delayed)} is defined only one
cycle after the input. Thus, {\tt z} and {\tt S} are declared on 
domain {\tt \{t|2<=t\}}. Instruction 4 now defines {\tt S} as being the
sum of variables {\tt x+y}, delayed by 1.

%
%\newpage
\section{Retiming\index{retiming}}
{\alfa} provides tools for retiming synchronous architectures. 
In the example of figure~\ref{adderDelayed}, one can shift the 
delay from the output of the adder to its inputs. 
This corresponds to replacing line 4 in program~\ref{adderDelayed-alpha}
by \verb\S[t] = x[t-1] + y[t-1];\. The corresponding architecture is shown 
in figure~\ref{adderRetimed}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exemple_3}}
\caption{Retiming the adder}\label{adderRetimed}
\end{figure}
\subsection*{Retiming in {\mmalfa}}
Do 
\begin{verbatim}
load["adderDelayed.alpha"]; (* load the alpha program *)
ashow[];                      (* show it *)
normalize[];                  (* simplify it *)
ashow[];                      (* show the result *)
asave["file.alpha"];          (* save program *)
\end{verbatim}

%
%\newpage
\section{Case and restrictions}
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exemple_5}}
\caption{Adding a multiplexer}\label{adderMultiplexer}
\end{figure}

Two other constructs of the language are the case\index{case} and the
restriction.  Consider the example of figure~\ref{adderMultiplexer},
which shows a new delayed and retimed version of the adder, where the
output is taken through a multiplexer\index{multiplexer}. In other
words, depending on the value of index {\tt t}, {\tt S} is either 0 or
the sum {\tt x[t-1] + y[t-1]}. This is reflected in the definition of
{\tt S} in figure~\ref{addermultiplexer-alpha} (lines 2--5.) This
definition uses a case expression. Each branch\index{branch (of a case
expression)} of the case is an {\alfa} expression. The first one is
the constant\index{constant} $0$, denoted as {\tt 0[]} in {\alfa}. It is
considered only when {\tt t} is equal to $1$, which is reflected in the
condition\index{condition} {\tt \{| t = 1\}}. The second branch
corresponds to the expression {\tt x[t] + y[t]}, taken when the
condition {\tt \{|t>1\}} is true. Actually, we call an expression such
as {\tt \{|t>1\}} a {\em restriction}.\index{restriction}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adderMultiplexer.alpha}
\caption{Adder and multiplexer}\label{addermultiplexer-alpha}
\end{figure}
The definition of {\tt z} also uses a restriction. The definition of 
line 6 says that {\tt z} is defined as {\tt S}, but only for time 
instants {\tt t} such that {\tt \{|t>=5\}}. 
\subsection*{Exercices in {\mma}}
Load and parse program adderMultiplexer.alpha.
\begin{verbatim}
load[adderMultiplexer.alpha];
ashow[adderMultiplexer];
\end{verbatim}

%
%\newpage
\section{Array and full form of {\alfa}}
\label{arrayform}

In all the above examples, we have used the so called {\em array
form}\index{array form} of {\alfa}. The full form\index{full form} of
the language is actually a little bit different, and a little bit more
difficult to read initially. It is however needed in order to
understand some of the properties of the language.

Let us consider again the definition of {\tt S} in
program~\ref{addermultiplexer-alpha}, line 2 -- 5:
\begin{verbatim}
 S[t] = case
     {|t=1}: 0[];
     {|t>1}: x[t-1] + y[t-1];
     esac;
\end{verbatim}
The full form of this expression is in fact: 
\begin{verbatim}
 S = case
      {t|t=1}: 0.(t->);
      {t|t>1}: x.(t->t-1) + y.(t->t-1);
  esac;
\end{verbatim}
To explain this, we need to go more deeply in the 
structure of the language. In {\alfa}, any expression is a function from
a set of points -- called the {\em domain} of the expression --, 
to some value set depending on the type of the expression. 
Expressions are either 
\begin{itemize}
\item variables -- e.g. {\tt S},\index{variables} 
\item scalar constants\index{constants}
\item pointwise operations on expressions -- e.g. {\tt x + y}\index{pointwise 
operators}
\item application of dependence functions to expressions, -- e.g. 
{\tt exp.f} where {\tt f} is a dependence function explained later,\index{dependence function}
\item restriction of expressions, -- e.g. {\tt dom: exp}, where 
{\tt dom} is a domain description\index{restriction},
\item case expressions -- e.g. {\tt case e1 ; e2; esac}, where {\tt e1}
and {\tt e2} are expressions.\index{case expression}
\end{itemize}

Let us examine expressions in more details. 
\subsection{Domains}
\index{domain}

Domains of {\alfa} are restricted to {\em polyhedra}, (more precisely,
unions of finitely may polyhedra).  A polyhedron is a set of integral
points whose coordinate satisfy linear inequalities.  For example:
{\tt \{t | 1 <= t <= 10\}} denotes such a set. So also does {\tt \{i,j
| i>=10; 0 <= j <= i\}} for example. {\texttt \{i | 0 < i < 10\} ||
\{i | 15 <= i\}} is a union of polyhedra.  \index{polyhedron}

\subsection{Variables}
\index{variable}
Variables are functions from a domain to a set of values. For example, 
in program~\ref{addermultiplexer-alpha}, {\tt S} is a function from 
the domain {\tt \{t|1<=t\}} to the set of integers. Thus, {\tt S}
represents a collection of indexed values $S_t$, where {\it t} belongs
to the domain of {\tt S}.

{\alfa} is a strongly-typed language, where variables are declared,
together with their domain. This allows many compile time checks to be
performed\index{type-checking} (see chapter~\ref{chapanalyze}).

Input variables are defined in the head of the system
definition. Output variables are defined after the key-word {\tt
result}. Local variables are defined after the key-word {\texttt var}
and the keyword {\tt let}.

\subsection{Constants}
In {\alfa}, constants are considered as functions from the
singleton set $Z^0$ to a type. For example, $1$ is the function 
from $Z^0$ to the integer value $1$. 
\subsection{Pointwise operations}
\index{pointwise operations}

Pointwise operations generalize classical operations like +, -, * on
integers, or {\tt and}, {\tt not}, {\tt or} on booleans. So, {\tt x +
y} represents the pointwise operation on values $x_t$ and $y_t$ for
all $t$. There is however, an important point to notice: if {\tt op}
is a pointwise operator and {\tt e1} and {\tt e2} are expressions,
then the domain of the expression {\tt e1 op e2} is the {intersection}
of the domains of {\tt e1} and {\tt e2}.\index{domain of a pointwise
expression}

\subsection{Dependence functions}
\index{dependence function}

Dependence functions are affine functions on indices.
They are denoted as
\begin{center}
\texttt{(idx, idx, ... -> iexp, iexp, ... )}
\end{center}
where \texttt{idx} is an index name and \texttt{iexp} is an affine
function of indices.  For example: {\tt (i,j -> i+2, j+3i-4)\}}
denotes such a function.  So does also the function {\tt f=(t ->
t-1)}. Therefore, {\tt x.(t -> t-1)} is an {\alfa} expression, whose
meaning is as follows.  Remember that {\tt x} is a function from the
domain {\tt \{t | 1<=t\}} to the set of integers. The composition of
functions {\tt x o f} is therefore defined, and this is just what {\tt
x.(t -> t-1)} represents. It is easy to see that this function is
defined on the domain {\tt \{t | t>=2\}}.  More generally, the domain
of any expression {\tt e.f} is ${\tt f}^{-1}(\mbox{dom}({\tt e}))$,
that is to say, the pre-image of the domain of {\tt e} by function
{\tt f}.

Note that a function {\tt (->t)} denotes a mapping from the set $Z^0$
to $Z$. Similarly, {\tt (t ->)} denotes a mapping from the set $Z$ to
$Z^0$. Such functions are used to extend scalar constants to higher
dimensional domains.\index{Extension of constants}

\subsection{Restriction expressions}

\index{restriction} As said before, restrictions allows one to
restrict the definition of an expression to a subdomain. To be
consistent with the other language constructs, restrictions are
defined as polyhedra, and therefore are denoted just as domains. For
example, {\tt \{t | t>=5\}: S} is a restriction of the expression
\texttt{S} to domain \texttt{\{t | t >= 5\}}. The domain of a
restricted expression \texttt{dom: e} is the intersection of the
domain of the expression and the domain defined by the restriction,
i.e., $\texttt{dom} \cap \texttt{e}$.\index{domain of a restriction}

\subsection{Case expressions}

\index{case expression} As already seen, case expressions allow one to
define a new function case by case. A case expression is defined on
the union of the domains of its branch. Let {\tt case exp1; exp2 esac}
be a case expression, with branches {\tt exp1} and {\tt exp2}. Let
{\tt d1} and {\tt d2} be respectively the domains of {\tt exp1} and
{\tt exp2}. Assume that {\tt d1} and {\tt d2} have the same
dimensionality and are disjoint (these properties are checked during
static analysis). Assume also that {\tt exp1} and {\tt exp2} have the
same type, say {\tt T}.  Then {\tt case exp1; exp2 esac} is a function
from ${\tt d1} \cup {\tt d2}$ to {\tt T}, whose value is that of {\tt
exp1} on {\tt d1}, and that of {\tt exp2} on domain {\tt d2}. For 
a detailed desciption of the domain computations and its use for the
static analysis, see section~\ref{static-analysis}.

\subsection*{Back to our example}
We are now ready to consider again the example corresponding to figure~\ref{addermultiplexer-alpha}. The definition of {\tt S} is:
\begin{verbatim}
 S = case
      {t|t=1}: 0.(t->);
      {t|t>1}: x.(t->t-1) + y.(t->t-1);
  esac;
\end{verbatim}
Let us analyze each branch. {\tt 0} is a function from $Z^0$ to 
the integer value $0$. Composing this function with the function 
\texttt{(t ->} extends the constant {\tt 0} to the set $Z$. In other 
words, {\tt 0.(t->)} denotes the function from $Z$ to integers,
whose value is {\tt 0} everywhere. 
{\tt \{t | t=1\}: 0.(t->)} denotes the restriction of this function 
to the single point {\tt t=1} of $Z$.

We can analyze the second branch in a similar fashion. We would conclude
that it represents a function from the subset {\tt \{t | t>1\}} of $Z$, 
whose value at point {\tt t} is $x_{t-1} + y_{t-1}$.

Finally, the total case expression defines {\tt S} as a function from
{\tt \{t | t >=1\}} to the integer set, whose value is {\tt 0} 
when {\tt t = 1}, $x_{t-1}+ y_{t-1}$ otherwise.

\subsection{Array form}
\index{array form}

The array form is just a convenient way of presenting {\alfa} programs, 
in a more ``readable'' form. The rule is simple to explain on the 
definition of {\tt S}:
\begin{itemize}
\item Dependence functions of the form {\tt .(indices -> exprs)} are rewritten
as {\tt [exprs]}, 
\item Restrictions of the form {\tt \{indices | inequalities\}} are rewritten
{\tt \{| inequalities\}},
\item Left-hand side variables are appended with the indices from the
right hand side.
\end{itemize}
Thus, the definition of {\tt S} in array form is
\begin{verbatim}
 S[t] = case
      {|t=1}: 0[];
      {|t>1}: x[t-1] + y[t-1];
  esac;
\end{verbatim}
This notation can be used safely for any normalized expression, as 
we shall see in section~\ref{substitution}.
\subsection*{Exercices in {\mmalfa}}
Load a program, and see its full form using the {\tt show[]}
function, or its array form using {\tt ashow[]}.
%\newpage
\section{Substitution and normalization}
\label{substitution}
One of the properties of {\alfa} is called the referential transparency\,:
\index{referential transparency}
it means that the meaning of an expression is independent of the context
where it appears. 
For this reason, any symbol can be safely replaced by its definition. 
For example, by replacing the {\tt S} symbol in the definition of 
{\tt z} in program~\ref{addermultiplexer-alpha}, we obtain the 
program shown in figure~\ref{addermultiplexer-alpha}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adderSubstituted.alpha}
\caption{Program of figure~\ref{addermultiplexer-alpha} after substitution of {\tt S} in 
definition of {\tt z}}\label{addersubstituted-alpha}
\end{figure}

Let us consider the new definition of {\tt z}:
\begin{verbatim}
z = {t | 5<=t} : 
         case
           {t | t=1} : 0.(t->);
           {t | 2<=t} : x.(t->t-1) + y.(t->t-1);
         esac.(t->t);
\end{verbatim}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adderNormalized.alpha}
\caption{Normalization of program~\ref{addersubstituted-alpha} (in
array form)}
\label{addernormalized-alpha}
\end{figure}

It is rather intuitive that the first restriction {\tt \{t | 5<=t\}}
can be distributed in the branches of the case expression. Combining
(i.e., intersecting) this restriction with the {\tt \{t | t=1\}} leads
obviously to an empty set, and consequently, to an empty
branch. Similarly, combining {\tt \{t | 5<=t\}} with {\tt \{t | 2<=t\}}
leads to {\tt \{t | 5<=t\}}. In summary, we can imagine that the full
definition of {\tt z} can be simplified to something like {\tt z =
x.(t->t-1) + y.(t->t-1)}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exemple_7}}
\caption{Architecture after substitution and normalization}
\label{addernormalized}
\end{figure}
These intuitive properties are in fact true, and such a simplification
is done in a systematic way by means of the normalization
transformation: one can prove that any expression can be simplified
into a normalized one, with at most one level of outermost case, one
level of restrictions, and one level of
dependences.\index{normalization}\index{simplification of expressions}

After normalization, the program becomes that of
figure~\ref{addernormalized-alpha}.

We can interpret this program as representing the synchronous
architecture of figure~\ref{addernormalized}.
We can see -- and this is of course straightforward -- that the 
multiplexer is indeed useless, since we are interested in the 
value of {\tt z} only when {\tt t>1}. This normalization 
property is one of the most important one of {\alfa}.

\subsection*{Exercices in {\mma}}
\subsubsection*{Exercice 1}
Write an {\alfa} program to model the 
following
architecture:
%of figure~\ref{exo1}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exo1}}
\caption{Exercice 1}
\label{exo1}
\end{figure}
\subsubsection*{Exercice 2}
Write an {\alfa} program to model 
the circuit of figure~\ref{exo2}.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/exo2}}
\caption{Exercice 2}\label{exo2}
\end{figure}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapanalyze.tex
Added mmalpha (source and binary)

\chapter{Static analysis of {\alfa} programs\label{static}\label{chapanalyze}}

\section{Introduction}

This chapter describes the use of the static analysis tool
\texttt{analyze[]}.\index{analyze[]}  
It is divided in two parts: the first part deals with the
static analysis of a single system, and the second describes the
analysis of a structured program consisting of several systems. Any
beginner in {\alfa} should read the first part, while the
second part is left to more experienced users.

\subsection{What is static analysis?\index{static analysis}}

It is impossible to ensure that an {\alfa} system computes the
expected result (to start with, the termination of such computation is
well known to be indecidable). However there are a few necessary
conditions for a system to be valid which may be verified
\emph{statically}, that is

\begin{itemize}

\item independently of any set of input values that may be fed to the
system, and
\item in a manner that is valid for any of these set of values.

\end{itemize}

\subsection{What does the static analyzer do?}

The {\alfa} static analyzer basically verifies the following rule:

\emph{For each point of the domain of a variable, there is
one and only one computation defining the value of the variable at
this point.}

The static analyzer checks that this rule is ensured and outputs error
messages giving the points where a variable is over- or under-defined.

\subsection{When should the static analyzer be used?}

This tool is very useful while writing and debugging {\alfa} code, and
should be invoked systematically. Besides,

{\large
\begin{center}
\emph{{\alfa} program transformations are only guaranteed to work}

\emph{on programs that pass the static analysis without error messages.}

\end{center}
}

This is in particularly true of the {\alfa} to C translator \index{writeC}
\texttt{writeC[]}: the C code generated by this command is likely to
cause run-time errors if the initial {\alfa} program doesn't pass the
static analysis. Therefore the static analyzer should be called before any
simulation of the {\alfa} program.

\subsection{An example\index{gaussian elimination}}

We will use in this chapter the following two-system example program
which transforms a matrix into triangular form as first step of a
Gaussian elimination\footnote{This program is neither optimal nor
complete, it was written for the purpose of demonstrating the
\textsc{analyze[]} function. Writing a complete Gaussian elimination
is left as an exercise to the reader.}.

The first system takes a square matrix and zeroes all the elements
below the \texttt{K}-th diagonal of this matrix (one slice of
Fig.\ref{gauss-slices}):

\begin{verbatim}
 system ZeroColumn: {N,K| 1<=K<N} (A: {i,j| 1<=i,j<=N} of real)
                          returns (Ar: {i,j| 1<=i,j<=N} of real);
 let
   Ar[i,j] = case
             {| i<=K}      : A[i,j];
             {| i>K; j<=K} : 0[];
             {| i>K; j>K}  : A[i,j] - A[K,j]*A[i,K]/A[K,K];
             esac;
 tel; 
\end{verbatim}

The second system uses N instances of the first to compute the
triangular matrix (see Fig.ref{gauss-slices}):

\begin{verbatim}
 system Gauss: {N | N>1} (A: {i,j | 1<=i,j<=N} of real)
                 returns (T: {i,j | 1<=i,j<=N} of real);
 var Ak : {i,j,k| 1<=i,j<=N; 1<=k<=N} of real;
     Ak1: {i,j,k| 1<=i,j<=N; 1<=k<N} of real;
 let
   use {k| 1<=k<N} ZeroColumn[N,k] (Ak) returns (Ak1);
   Ak[i,j,k] = case
               {| k=1} : A[i,j];
               {| k>1} : Ak1[i,j,k-1];
               esac;
   T[i,j] = Ak[i,j,N];
 tel;
\end{verbatim}

\begin{figure}
\centerline{
	\includegraphics[width=8cm]{figures/gauss}
        }
\label{gauss-slices}
\caption{Triangularization in two systems}
\end{figure}

What kind of information does the static analyzer give? Suppose a typo
has replaced an \texttt{j} index with a \texttt{i} index, leading to
the following \texttt{ZeroColumn} program:

\begin{verbatim}
 system ZeroColumn: {N,K| 1<=K<N} (A: {i,j| 1<=i,j<=N} of real)
                          returns (Ar: {i,j| 1<=i,j<=N} of real);
 let
   Ar[i,j] = case
             {| i<=K}      : A[i,j];
             {| i>K; i<=K} : 0[];     -- The typo is hidden here
             {| i>K; j>K}  : A[i,j] - A[K,j]*A[i,K]/A[K,K];
             esac;
 tel;
\end{verbatim}

This error can't be detected by a parser program -- this program is
syntactically correct. However, invoking the static analysis will
yield the following messages:\index{analyze[]}

\begin{verbatim}
In[5]:= analyze[];

WARNING: This expression has an empty domain :
{i,j | i=0; j=0; N=0; K=0; 1=0} : 0.(i,j->)

ERROR: Variable Ar not defined over the domain :
{i,j | K+1<=i<=N; 1<=j<=K}

*** Analysis failed ***
\end{verbatim}

A warning and an error tell us that there is an error in the equation
defining \texttt{Ar}. Several \emph{error domains} may help us pinpoint
precisely where the error is. The first warning message is enough
to spot that the error is in the \texttt{case} subexpression
containing the \texttt{0[]}. The \texttt{1=0} equation in the domain
indicates that this domain is empty.\index{empty domain}

As this example shows, it may take a certain amount of experience
to fully understand the error messages: the error domains
are not always in the most readable form. However the indications
given usually allow to spot the problem precisely: in our example the
combination of both previous messages points exactly to the cause of
the error.

The remainder of this paper describes the static analysis tool in more
details.

\section{Static analysis of an {\alfa} system}
\label{static-analysis}

\subsection{The domain of an expression\label{expr_check}}

\index{expression domain}\index{domain of an expression} In {\alfa}
every variable is declared with a polyhedral domain defining the set
where it is expected to contain a value.\index{declaration
domain}\index{domain of a variable}

When an expression is built using such variables, this expression
inherits the domain of these variables: for example if \texttt{A} and
\texttt{B} are two expressions defined over some square domain
\texttt{\{i,j|0<i,j<10\}}, then their
sum \texttt{A+B} is defined everywhere both \texttt{A} and \texttt{B}
are defined, that is on the same square domain.

Now if the domains of \texttt{A} and \texttt{B} are different, then
the sum is still defined everywhere both \texttt{A} and \texttt{B} are
defined, that is on the intersection of the domains of \texttt{A} and
\texttt{B}.

It is possible to carry these ideas further, and thus to define the
domain of any {\alfa} expression, knowing the domains of the
subexpressions (see the discussion of section~\ref{arrayform},
page~\pageref{arrayform}). The rules to apply, given below, are actually part of
the definition of the semantics\index{semantics of {\alfa}} of the
language. They rely only on operators preserving the set of {\alfa}
domains and thus may be computed automatically: this is what the
static analyzer does.

 \begin{center}
  \begin{tabular}{rc}
   {Constants} & \fbox{$Dom(c) = \mathbf{Z}^0$} \vspace{1ex}\\ 
   {Variables} & \fbox{$Dom(V) \textrm{ is declared in the header}$}\vspace{1ex}\\ 
   {Unary operators} & \fbox{$Dom(-e) = Dom(e)$}\vspace{1ex}\\ 
   {Binary operators} & \fbox{$Dom(e_1+e_2) = Dom(e_1) \cap Dom(e_2)$}\\
   & {\small The sum of two variables is defined where both variables are defined}\vspace{1ex}\\ 
   {Ternary operators} & \fbox{$Dom(\texttt{if}\ e_1\ \texttt{then}\ e_2\ \texttt{else}\ e_3) = 
            \bigcap_{i=1}^3 Dom(e_i)$}\\
   & {\small The \texttt{if}\ \texttt{then}\ \texttt{else}\ is considered as a ternary operator}\\
   & {\small and is defined where the condition and both alternatives are defined}\vspace{1ex}\\ 
   {Restriction} & \fbox{$Dom(D : e) = D\cap Dom(e)$}\\
   {\small $D$ is a domain} 
   & {\small This operator restricts an expression to D}
   \vspace{1ex}\\
   {Affine dependency} & \fbox{$Dom(e[f]) = f^{-1}(Dom(e))$}\\
   {\small $f$ is an affine function}
    & {\small The value of $e[f]$ at point \textbf{z} is the value of $e$}\\
   {\small of the indices of the LHS}
    & {\small at point $f(\mathbf{z})$, hence the domain of $e[f]$}
  \vspace{1ex}\\
   {\texttt{case} operator} 
   & \fbox{$Dom(\texttt{case} e_1;..;e_n;\texttt{esac}) = \bigcup_{i=1}^n Dom(e_i)$}\\
   & {\small The \texttt{case} operator allows the piecewise definition of an expression}\\
   & {\small by several subexpressions $e_i$ with disjoint domains.}
  \end{tabular}
 \end{center}

The function \texttt{expDomain[]} \index{expDomain[]} may be used to
compute the domain of any {\alfa} expression, following these rules.

Now during the computation of the domain of an expression, the
\texttt{case} operator \index{case} introduces the possibility of
having more than one subexpression define the value of the same point
of the domain.  Therefore the analysis tool has to check that the
intersections of the domains of the case subexpressions are empty. It
computes this intersection, and if it not empty it issues an error
message as in the following example:
\index{overlap of case statements}\index{analysis of the case}\index{case validity}

\textbf{Example:} in the equation defining \texttt{Ar}, if the
second line of the \texttt{case} was wrongly written:
\begin{verbatim}
      {| i>=K; j<=K}: 0[];
\end{verbatim}
The analysis tool will output the following message:
\begin{verbatim}
 ERROR: in case statement: ...,
        domains of subexpressions overlap on:
        {i,j | i=K; 1<=j<=K}
\end{verbatim}

There are other useful informations which the static analyzer
provides. For example it is useful to detect an empty expression
i.e. an expression with an empty domain: such an expression is 
in the best case pointless (in a case statement), and may be a source of
errors. \index{expression has an empty domain}\index{empty domain}
To avoid cascaded error messages, only the
deepest empty subexpression is reported to the user.

\textbf{Example:}
 Still in the same equation, a mistake in the first \texttt{case} subexpression:
\begin{verbatim}
      {| i<=0}     : A[i,j];
\end{verbatim}
will cause the following messages:
\begin{verbatim}
 WARNING: This expression has an empty domain: 
          {| i<=0} : A[i,j]
 ERROR:   Variable Ar not defined over the domain: 
          {i,j| 1<=i<=K; 1<=j<=N}
\end{verbatim}

\subsection{Equation analysis\label{equ_check}}

\index{equation analysis}\index{analysis of an equation} Now we
describe how the static analyzer works: it considers each equation
$V[i,j\ldots] = e$, where $e$ is an {\alfa} expression.  To ensure
that there is at least one computation defining the value of $V$ for
each point $(i,j,\ldots)$ of its domain, the analysis tool computes
$D' = Dom({V}) \setminus Dom({e})$, where $\setminus$ denotes the set
difference, $Dom(V)$ is the domain of the variable $V$ (declared in
the header of the system), and $Dom(e)$ is the domain of the
expression as defined above.\index{declaration domain}\index{domain of
a variable}

If $D'$ is non empty, an error is issued, stating that $V$ is not
defined over $D'$. This error domain will be useful to the user to
spot the problem. \index{variable not defined}

\textbf{Example:} In the equation of the system \texttt{ZeroColumn} 
 defining \texttt{Ar}, 
suppose the first line of the \texttt{case} statement was mistyped:
\begin{verbatim}
      {| i<K}     : A[i,j];    -- instead of {| i<=K} ...
\end{verbatim} 
(notice the \verb~<K~ instead of \verb~<=K~). The analysis tool will output
 the following message:
\begin{verbatim}
 ERROR: Variable Ar not defined over the domain:
        {i,j| i=K; 1<=j<=N}
\end{verbatim}

\subsection{Parameter related analysis\label{param-check}}

\index{parameter analysis}\index{analysis of a parameterized system}
If the system considered is parameterized, the static analysis process
may be usefully refined by taking the parameters into
account. Consider again the \texttt{Gauss} system:

\begin{verbatim}
 system Gauss: {N | N>1} (A: {i,j | 1<=i,j<=N} of real)
                 returns (T: {i,j | 1<=i,j<=N} of real);
 var Ak : {i,j,k| 1<=i,j<=N; 1<=k<=N} of real;
     Ak1: {i,j,k| 1<=i,j<=N; 1<=k<N} of real;
 let
   use {k| 1<=k<N} ZeroColumn[N,k] (Ak) returns (Ak1);
   Ak[i,j,k] = case
               {| k=1} : A[i,j];
               {| k>1} : Ak1[i,j,k-1];
               esac;
   T[i,j] = Ak[i,j,N];
 tel;
\end{verbatim}

Suppose there was no restriction on the parameter \texttt{N} of the
system \texttt{Gauss}. Its header would be: \mbox{\verb~ system Gauss:
\{N |\} ~}.  Now obviously for negative values of \texttt{N}, all the
variables of this system have an empty domain, \index{expression has
an empty domain}\index{empty domain} which should be pointed to the
user as a possible source of errors.  The static analyzer performs
such parameter-related checks.

\textbf{Example:}
 We may restrict the parameter \texttt{N} of the system \texttt{Gauss} to
 be positive:
\begin{verbatim}
 system Gauss: {N | N>0}  
\end{verbatim}
One may check that the system is still valid, even for \texttt{N=0}.
 However the analysis will issue the following message:
\begin{verbatim}
 WARNING: for parameters {N| N=1}, the expression Ak1 has an empty domain
\end{verbatim}

It is good programming practice to ensure that the system is valid for
all the values of its parameter domain. It becomes mandatory if the
program is composed of several systems, as shown in the following
section.

\section{Analysis of structured programs}
\index{analysis of structured programs}

\subsection{Analysis of \texttt{use}\ statements\label{useanalysis}}
\index{analysis of the use}\index{use statement analysis}\index{validity of the use}

The analysis of a \texttt{use} statement is deduced from its
\index{substitution semantics}\index{semantics of the use}
\emph{substitution semantics}: in short, a program containing a
\texttt{use}\ statement is (by definition of the use) equivalent to
one where this statement has been replaced with the body of the
subsystem (properly modified to take into account the extra dimensions
and the affine parameter assignment) and additional equations to
perform the I/O passing: input equations relate the actual inputs and
the formal ones~:
\begin{verbatim}
      SubSystemInputVariable = ActualInputExpression ;
\end{verbatim} 
and output equations relate the actual outputs and the formal ones~:
\begin{verbatim}
        ActualOutputVariable = SubSystemOutputVariable ;
\end{verbatim}

\paragraph{Global checks}
The first validity conditions are that the subsystem has been declared
somewhere in the program, that the correct number of actual
inputs/outputs are given, and that their respective dimensions match
the formal ones.  The tool also checks, using the same techniques as
previously, that the extension domain is non-empty for all the values
of the parameters.

\index{parameter assignment}\index{assignment of the
parameters}\index{parameters}\index{validity of the parameter
assignment} 
Then we consider the values given to the parameters of
the subsystem by the caller. The parameters of the subsystem are
an affine function of the caller parameter and,
possibly, the extension indices.  The analyzer checks that, for all
the possible values of the caller parameters, and for all the points
in the extension domain, the values assigned to the subsystem
parameters fall within the range permitted, i.e. within the parameter
domain of the subsystem.  Otherwise an error message is issued,
showing a domain which is the set of points where parameters are given
but not expected.

\textbf{Example:}
In the \texttt{Gauss} system, the following \texttt{use}\ statement:
\begin{verbatim}
    use {k| 0<=k<=N} ZeroColumn[N,k] (Ak) returns (Ak1);
\end{verbatim}
will cause the following message:
\begin{verbatim}
 ERROR : in statement ``use ... ZeroColumn...'',
         parameter values in {N,K| K=0, N>=1} not allowed.
\end{verbatim}

Obviously, the more restricted the parameter domain of the subsystem,
the more acurate the checks performed here.

\paragraph{Input/Output checks}

\index{subsystem input/output}
The substitution semantics implies that the verifications to be
performed on the I/Os of a subsystem \texttt{use}\ may be deduced from
those of the equations described in \ref{equ_check}. Error messages
are given accordingly.

\subsection{Global analysis of a library of systems}

\index{analysis of a structured program}
The validity of a \texttt{use}\ statement is then implied by the
validity of the subsystem on its parameter domain, the condition that
all the parameter values assigned by a \texttt{use} are allowed, and
the validity of the virtual I/O passing equations.

We may thus describe the general down-top verification method for a
complete program. Such a program is an acyclic graph of {\alfa}
systems using each other (system \texttt{use} can not be mutually
recursive). Systems without a \texttt{use}\ statement in their
equations are called {\em leaves}. A system using a subsystem is
called a {\em parent} of this subsystem.
\begin{itemize}
\item First, the leaves are analyzed, and their parameter domain is
restricted as much as possible. No warning message should remain.  For
example, for the system \texttt{ZeroColumn}, we have to restrict
\texttt{K} and \texttt{N} at least to the domain given in the
correct version of this sytem (it is possible to constraint the
parameters more than what the analysis suggests. For example we could
put a bound on \texttt{N} depending on the application aimed at).

\item Then the parents of the leaves are analyzed. If they are written
to use a leaf with illegal values of its parameters, the tool
will spot it and the programmer will be invited either to correct the
error, or to restrict more the caller parameters. Meanwhile, the other
equations of the caller are also analyzed, with the same effect.

\item This process is repeated on the parents of the parents, and so
on until the whole program passes the static analysis.
\end{itemize}

Note that one of the options to \texttt{analyze[]} decides whether the
analysis is performed recursively on all the subsystems of the system
currently being analyzed, or only to this system.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapcodegen.tex
Added mmalpha (source and binary)

\chapter{Code generation and simulation of {\alfa} programs}
\label{chapcodegen}
In this chapter, we explain how to generate C code from an
{\alfa} program. As seen in Chapter~\ref{chapfunctional}, 
{\alfa} programs are functional, and therefore, do not convey any
execution ordering. However, the validation of {\alfa} programs, 
or even, the generation of code for a general-purpose or
dedicated processor, requires the possibility to simulate the 
evaluation of {\alfa} expressions.

This can be done using the {\tt writeC} {\mmalfa} command, as seen here.

\section*{An example}
Consider the following {\alfa} program:
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder-WriteC.alpha}
\caption{A very simple example\label{adderwritec}
}
\end{figure}
To simulate this program, we first load this program, 
then generate a C program in file ``adder.c'', using 
\begin{verbatim}
load["adder-WriteC.alpha"];  (* load file *)
writeC["adder.c"];
\end{verbatim}
Then we can compile this program: 
\begin{verbatim}
cc -o adder adder.c
\end{verbatim}
The execution of this program is shown here:
\begin{verbatim}
adder
Input x =2
x = 2
Input y =3
y = 3
z = 5
\end{verbatim}
The C code prompts for the value of the input variables {\tt x} and
{\tt y}.
\section{Program with parameters}
Programs with parameters cannot be simulated without giving 
a value to the parameters. This is obtained through an option of
{\tt writeC}. 
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/adder-WriteC1.alpha}
\caption{A very simple example\label{adderwritec1}
}
\end{figure}
\begin{verbatim}
load["adder-WriteC1.alpha"];  (* load program *)
writeC["adder1.c","-p 3"];
cc -o adder1 adder1.c
adder1Input x[1] =2
x[1]= 2
Input x[2] =3
x[2]= 3
Input x[3] =4
x[3]= 4
Input y[1] =-2
y[1]= -2
Input y[2] =-3
y[2]= -3
Input y[3] =-4
y[3]= -4
z[1]= 0
z[2]= 0
z[3]= 0
\end{verbatim}
\subsection{Additional examples}
\begin{verbatim}
SetDirectory["/home/frodon/d01/api/quinton/alpha/sobel"];
load["sobel.a"];ashow[];
analyze[];
writeC["-p 5 35 3"];
schedule[{objFunction -> 1,addConstraints -> {"Tpixel[i,N,M,p]=i"},duration -> 2}]
\end{verbatim}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapderiving.tex
Added mmalpha (source and binary)

\chapter{Deriving \index{systolic}systolic architectures using {\alfa}}
\label{chapderiving}

{\sc Version of \today}

In this chapter, we describe a methodology for deriving
systolic architectures using {\mmalfa}. This \index{design
methodology}methodology is illustrated by means of an {\alfa}
program for the matrix-vector multiplication. A companion
notebook called \texttt{matvect.nb} allows the reader to
execute interactively this demonstration. This notebook is
situated in directory:\\
\texttt{\$MMALPHA/demos/NOTEBOOKS/Matvect}\\
A link to this notebook is available in the 
master notebook (see~\ref{master}).

\section{Introduction}

The methodology presented here is inherited from the research of the
``systolic community''~\cite{Meg1,KungHT,QuintonRo89} and from its
adaptation to {\alfa}~\cite{SieWilde94,Dinechin96c,Patricia96}.

The successive steps of the derivation are the following ones:
\begin{enumerate}
\item High level specification in {\alfa}.
\item Uniformization.
\item Scheduling and mapping.
\item Control signal generation.
\item \label{step5} Generation of an {\alphard} specification. 	\\
~\\These steps may be followed by three additional ones: 
\item Translation to {\vhdl}.
\item Logic synthesis.
\item Layout synthesis.
\end{enumerate}

The five first transformations deal with producing a so-called
\alphard{} program.  {\alphard} is a subset of the {\alfa} language
used for architectural description at the register transfer level.
The description of {\alphard} format can be found
in~\cite{Patricia96}, and is presented in chapter~\ref{chapvhdl}.  In
the present chapter, we only describe how to perform the first four above
steps using {\mmalfa}. The result will be a single {\alfa}
program which can be interpreted as a systolic array for executing the
original {\alfa} specification. The main difference between this
format -- called {\alphaz} -- and {\alphard} is their structuring: an
{\alphard} program is structured into several sub-systems while an
{\alphaz} program contains only one {\alfa} system.

\section{Example of the matrix-vector product\index{matrix-vector product}}
We start by an \alfa{} description of the algorithm, 
as shown in figure~\ref{figder1}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MatVect.alpha}
\caption{{\alfa} program for the matrix-vector}
\label{figder1}
\end{figure}

The dependence graph of this program 
is shown in the left hand side of figure~\ref{figder2}, 
for $N=M=4$. In this
figure, it can be seen that each component of the {\tt b} vectors is
broadcasted to several computations. For instance, $b_1$ is used in
the computation of $ a_{1,1}, a_{1,2}, a_{1,3}$ and $a_{1,4}$. The
first step of the synthesis consists of ``pipelining'' these values
in order to obtain a uniform dependence graph such as 
that shown in the right hand side of
figure~\ref{figder2}. This step is often called {\em uniformization}
or {\em localization}
in the litterature.

\subsection{Uniformization\index{Uniformization}}

Methodologies for automatic uniformization were
proposed by several researchers
(see~\cite{QuintonVa88} for example). However, performing the
uniformization process completely automatically is difficult as there
are many possibilities for uniformizing a given program. Hence, 
this transformation requires some help from the user.

Before proceeding, we first perform a pre-processing 
transformation, which consists of adding a local 
variable \texttt{A} to replace the occurrence of
\texttt{a} in the definition of \texttt{C}. The purpose of
this modification is to anticipate a change of basis 
that will be done later on, and which cannot be applied on
an input variable without altering the semantics of
the program. The \mmalfa{} function which does this
is:\\
\texttt{addLocal["A = a"];}\\
%We first perform a single pre-processing step which consists of adding all
%the constraints present in the parameter domain declaration (present
%at the top of the system) inside every domain appearing in the
%program. Then we add, for each input variable, a local variable
%(remember that we cannot perform transformations involving inputs and
%output variables). Next, we have to \index{pipeline}pipeline the
%broadcasted expressions.

We are now ready to uniformize the occurence \texttt{b[i]}
in the definition of \texttt{C}. 
The pipeline vector will be the direction of
the $\vec{\imath}$ axis in our example, as shown in 
figure~\ref{figder2}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/tr1}}
\caption{Dependence graph for the program of figure~\ref{figder1}, before 
and after uniformization.}
\label{figder2}
\end{figure}

In our example, the {\mmalfa} function for uniformizing is: 
\begin{verbatim}
pipeall["C","b.(i,j->j)","B1.(i,j->i+1,j)"];
\end{verbatim}
The parameters given to 
{\tt pipeall} can be read as follows: 
\begin{quote}
In the definition of {\tt C},
pipeline the expression
{\tt  b.(i,j->j)}. The pipeline vector is
$(1,0)$, the variable which will pipeline the value will be called
{\tt B1}.
\end{quote}
Two remarks: 
\begin{itemize}
\item The array notation {\tt b[i]} cannot be used
here, as it would be ambiguous. 
\item The pipeline vector is indicated as a translation in
{\alfa} form.  
\end{itemize}
The result of applying these commands is shown in
figure~\ref{figder3} where only the equation of the resulting {\alfa}
program are shown.  Note that the domains of the branches
of \texttt{B1} definition have been computed by {\tt
pipeall}. Note also that the dependence graph of this program
-- represented in the right hand side of figure~\ref{figder2} -- is
now uniform. For more information on the pipelining process refer
to the
documentation~\cite{docpipeline}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\begin{verbatim}
B1[i,j] = 
      case
        {| i=1; 1<=j<=M; 2<=N; 2<=M} : b[j];
        {| 2<=i<=N; 1<=j<=M; 2<=M} : B1[i-1,j];
      esac;
A[i,j] = a;
C[i,j] = 
      case
        {| j=0} : 0[];
        {| 1<=j} : C[i,j-1] + A[i,j] * B1;
      esac;
\end{verbatim}
\caption{Definitions of \texttt{B}, \texttt{A}, and \texttt{C}
of program of figure~\ref{figder1} after uniformization}
\label{figder3}
\end{figure}

\subsection{\index{Scheduling}Scheduling and \index{mapping}mapping}
\label{schedAlpha}

A schedule of this program is found by the {\tt schedule} function.  
(see chapter~\ref{chapscheduling} for more details.) 
Here we call this function with the option
{\tt scheduleType} in order to find out
a schedule where 
all variables share the same linear part: this
preserve the uniformity of the program. 
The {\mmalfa} command which does this is: 
\begin{verbatim}
schedule[scheduleType->sameLinearPart];
\end{verbatim}
The result yields $T_C[i,j] = i+j$ and is illustrated in
figure~\ref{figder4}. The mapping must be given by the
user. Following the standard systolic methodology, the 
\index{allocation}allocation
function $a$ is a projection of the $n$-dimensional index space onto a
$n-1$-dimensional space. Here, we assume that we want to  project on the
$\vec{\jmath}$ axis ({\tt (i,j -> j)} (see figure~\ref{figder4}). This results
in a \index{linear array}linear array of $N$ processors.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/tr3}}
\caption{Scheduling and mapping of the uniform program}
\label{figder4}
\end{figure}

\subsection{Applying a spatio-temporal reindexing}
We now have all the spatio-temporal information -- timing, and spatial
mapping -- for obtaining a systolic array.  We can make this
information explicit in the {\alfa} program by performing a change of basis: 
in this way, in the new domains, the first index will represent the
execution date and the second index will represent the processor
number.

%In our case, this corresponds to the \index{change of
%basis}change of basis by the function \texttt{f}: {\tt (i,j->i+j,j)}. 
%The following sequence of \mma{} expressions does
%the job: 
%\begin{verbatim}
%f = readMat["(i,j,N,M->i+j,i,N,M)"];
%id2 = readMat["(t,p,N,M->t,p,N,M)"];
%g = composeAffines[f,id2]
%changeOfBasis["C",g]
%changeOfBasis["A",g]
%changeOfBasis["B1",g]
%\end{verbatim}
%Recall that \texttt{id = expression} denotes an assignment in \mma{}.
%The first line assigns to symbol $f$ the expression obtained by
%parsing the
%\alfa{} matrix expression \texttt{"(i,j,N,M->i+j,i,N,M)"}
%using the {\tt readMat} function. 
%(Notice that this expression contains
%the parameters \texttt{N} and 
%\texttt{M} of the {\tt MatVect} program.)
%The {\tt id2}
%symbol is assigned a function to rename the indices 
%($(t,p)$ instead of $(i,j)$) for readability. 
%Then the change of basis 
%is performed on all local variables using the
%\texttt{changeOfBasis} functio.

The function which does this transformation is:\\
\begin{verbatim}
applySchedule[];
renameIndices[{"t","p"}];
\end{verbatim}
\index{\texttt{applySchedule}}
(The command \texttt{rename} allows one to change the 
names of the indices.)
The resulting program is shown in 
figure~\ref{figder5}.  This program is scheduled and mapped: for
example, {\tt C[t,p]} will be computed at step {\tt t} on processor
{\tt p}.

Note that \texttt{applySchedule} chooses the 
mapping of the program. It is possible to modify this
mapping by applying another change of basis.\footnote{There is an 
incoherence here between the notebook and this chapter...}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MVsched.alpha}
\caption{{\alfa} program after change of basis}
\label{figder5}
\end{figure}

\subsection{Generation of an {\alphaz} program}
\label{alphaZ}
The generation of {\alphaz} code consists of \index{Control signal}control
signal generation, and simplification of the code. 
The control signal
generation is necessary when the computation of a variable 
changes over execution time. For instance,
consider variable {\tt C} of the program of
figure~\ref{figder5}. 
When evaluating {\tt C[t,p]}, a processor {\tt p}
must know whether {\tt t=p} of {\tt t>p}, and this can only be done
with the help of a control signal. In addition, this control signal
should be inialized by a controller.

The situation is different for variable \texttt{B1}.
There are two different definitions
of {\tt B1[t,p]}
depending on whether {\tt p=1} or {\tt p>1}: 
{\tt b[t-p]} and {\tt B1[t-1,p-1]}.
But these conditions
do not involve the time {\tt t}. As a consequence, 
no control signal needs to be generated, as these
conditions will be directly reflected in the structure of 
cell number 1, and cells number \texttt{p}, \texttt{p>1}.

\begin{figure}[htbp]
\begin{verbatim}
C[t,p] = 
    case
      {| p<=t<=p+M; 1<=p<=N; 2<=N; 2<=M} : 
         case
           {| t=p; 1<=p<=N; 2<=N; 2<=M} : (if (loadC) then 0[] else True[]);
           {| p+1<=t<=p+M; 1<=p<=N; 2<=N; 2<=M} : 
              (if (not loadC) then C[t-1,p] + A[t,p] * B1[t,p] else False[]);
         esac;
    esac;
loadC[t,p] = 
    case
      {| p<=t<=p+M; 1<=p<=N; 2<=N; 2<=M} : 
         case
           {| t=p; 1<=p<=N; 2<=N; 2<=M} : True[];
           {| p+1<=t<=p+M; 1<=p<=N; 2<=N; 2<=M} : False[];
         esac;
    esac;
\end{verbatim}
\caption{Definition of {\tt C} after space-time case separation, and 
control signal generation.}
\label{figder6}
\end{figure}

Control signal generation is performed by a single call to the \mma{}
function {\tt toAlpha0v2}:
\begin{verbatim}
toAlpha0v2[];
\end{verbatim}
The \alphaz{} program is shown in figure~\ref{figder7}. It can directly 
be interpreted as the array of figure~\ref{figder8}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MVfinal.alpha}
\caption{ {\alphaz} program before translation in {\alphard}, this program 
represent the array of figure~\ref{figder8}}
\label{figder7}
\end{figure}

%Before generating control
%signals, we perform a preliminary transformation called 
%{\em space-time case
%separation}, which consists of replacing the usual one level case with
%a two level case, the first level involving conditions involving
%only processor numbers.  Figure~\ref{figder6} shows how 
%the definition of {\tt C} has been modified
%after the space-time case separation. Note
%that the first case level contains only one branch as there are no
%condition involving only processors indices.
%
%In the program of figure~\ref{figder5}, only {\tt C} needs a
%control signal. The variable needing a control signal 
%can be automatically detected (after space-time case separation)
%with the command {\tt needsMuxQ}. For instance,  
%{\tt needsMuxQ[getDefinition["C"]]} returns {\tt True}.
%The control signal generation 
%is made automatically with the function {\tt makeMuxControl}, for
%these actions, the commands are: 
%
%\begin{verbatim}
%spaceTimeCase["C", {1}, {2}];
%spaceTimeCase["A", {1}, {2}];
%spaceTimeCase["B1", {1}, {2}]; 
%makeMuxControl["C", "loadC", {1}, {2}];
%\end{verbatim}
%
%
%At this point, current implementation state imposes to fix the values
%of the parameters.  After the signal has been generated, it is
%necessary to pipeline it to the border of the array, this should be
%done with the {\tt pipeControl} command (or with the {\tt pipeline}
%command if you want a very specific pipeline). The last two command
%call two low-level transformations: split dependencies. Hence the
%{\mmalfa} commands are:
%
%\begin{verbatim}
%pipeControl["loadC",{1,0}];
%assignParameterValue["N",4];
%assignParameterValue["M",4];
%EquationSpatTemp[]
%makeSimpleExpr[]
%\end{verbatim}

\subsection{Obtaining an {\alphard} specification}

{\alphard} is a subset of {\Alpha} which represents hardware in a
structured manner. {\alphard} will be described precisely in
section~\ref{alphard}.  From the {\alphaz} specification, {\alphard}
code is obtained automatically by the command {\tt alpha0ToAlphard}

\begin{verbatim}
alpha0ToAlphard[controlVars[]];
\end{verbatim}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\centerline{\includegraphics{figures/tr6}}
\caption{\index{array}Array obtained after high level synthesis (here for $N=3$)}
\label{figder8}
\end{figure}

\section{Appendix: list of commands}
\begin{verbatim}
addLocal["A = a"];
pipeall["C","b.(i,j->j)","B1.(i,j->i+1,j)"];
schedule[scheduleType->sameLinearPart];
applySchedule[];
renameIndices[{"t","p"}];
toAlpha0v2[];
alpha0ToAlphard[controlVars[]];
\end{verbatim}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapfunctional.tex
Added mmalpha (source and binary)

\chapter{The Alpha Language and System -- Another point of view}
\label{chapfunctional}
\section{Introduction}

\begin{tobedone}
The last figure is garbage. The psfig inside has been commented out. 
See reference document in old pdf turorial.
\end{tobedone}
Alpha is a functional language, based on recurrence
equations\index{recurrence equation}.  All variables are type-declared
at the beginning of a program, and represent multidimensional arrays,
whose shapes can be arbitrary polyhedra.  For example, the declaration
\verb!A: {i,j| 0<j<i<=N} of real!, specifies a (strictly) lower
triangular, real matrix\index{lower triangular linear system}.  To
introduce the main features of the language, consider the problem of
solving a system of linear inequalities, $Ax=b$, where $A$ is a lower
triangular, $n\times n$ matrix with unit diagonal.  It is well known
that this can be solved using forward substitution\index{forward
substitution}, as given by the following formula.  \[\mbox{for } i =
1\ldots n,~~x_i = b_i - \sum_{j=1}^{i-1} A_{i,j} x_j\]

\begin{figure}[tb]
%\htmlimage{scale=\myscale}
\setlength{\unitlength}{10mm}
\begin{center}
%\begin{picture}(16,2.5)
%\thicklines
%\put(-0.1,-0.7){\framebox(16.2,3.7){}}
%\begin{boxedminipage}[b]{16cm}
%{\footnotesize
\begin{verbatim}
system ForwardSubstitution : { N | N>1 }       -- comments are like this
                 ( A : { i,j | 0<j<i<=N } of real;    -- a 2D input variable
                   B : { i   | 0<i<=N } of real)      -- a 1D input variable
         returns ( X : { i   | 0<i<=N } of real );    -- a 1D output variable
  let
    X[i] = B[i] - reduce(+, (i,j -> i), A * X.(i,j->j))[i]
  tel;
\end{verbatim}
%\end{boxedminipage}
%\end{picture}
\end{center}
%\vspace*{-0.8cm}
\caption{Alpha program for the forward substitution algorithm}
\label{fig-fs}
\end{figure}

  The corresponding Alpha program (Fig.~\ref{fig-fs}) is identical, except for
syntactic sugar.  The first line names the system and declares that it has a
positive integer parameter, {\tt N}.  The next three lines are declarations of
the input and output variables of the system.  Note how each variable is
defined over a certain range of indices, specified by a set of linear
inequalities.  This defines a polyhedron, called its {\em domain}.  For
example, the domain of {\tt A} is triangular, while {\tt B} and {\tt X} are
one dimensional vectors.  Variables declared before the ``{\tt returns}''
keyword are input variables, and those after it are outputs.  Although our
example does not illustrate it, a system may also have local variables, which
are declared after the system header, using the keyword {\tt var}.  The body
of the program is a number of equations delineated by the {\tt let} and {\tt
tel} keywords.  Here, we have a single equation corresponding to the above
formula, which specifies that we compute {\tt X[i]} by performing a reduction
(any associative and commutative binary operator may be used; here we have
{\tt +}).  The {\tt (i,j->i)} specifies a projection of a two dimensional
index space to one dimension.  Intuitively, it states that within the body of
the {\tt reduce}, we have two index variables, {\tt i} and {\tt j}, but the
latter is not visible outside the scope of the {\tt reduce}.  The body of the
{\tt reduce} is the expression, {\tt A * X.(i,j->j)}, where {\tt (i,j->j)}, is
a {\em dependency}. It denotes the fact that to compute the body at index
point, {\tt [i,j]}, we need the value of {\tt X} at index point {\tt j} (the
dependency of {\tt A} is not explicitly written---it is the identity).
Dependencies are an important aspect of Alpha.  They describe ``indexing
functions'' and have the syntax,
 \verb! (idx, idx, ...  -> i-expr, i-expr, ...)!, where each {\tt idx} is an
index name, and {\tt i-expr} is an {\em affine} expression of the system
parameters and the {\tt idx}'s.  The Alpha system uses this syntax for
specifying a multidimensional affine function in many different contexts.

\section{Polyhedra and affine functions}
  Alpha imposes the restriction that all variable domains are polyhedral, and
all dependencies are {\em affine} functions.  Although this may constrain our
expressive power, it is a conscious choice that was made for many reasons.  A
large number of linear algebra and DSP and image processing algorithms fall
within these limitations.  It is also known 
%\cite{feautrier91}, 
that any
nested loop program (in a conventional language) that satisfies certain
constraints (loop bounds are affine functions of the parameters and outer
indices, and the body contains only assignments to array variables which are
accessed using only affine functions of the loop indices) can be modeled by a
formalism that is similar to an Alpha program.  Such assumptions are made in
many techniques used in parallelizing compilers.  By concentrating explicitly
on this subset of programs, we can make use of powerful static analysis and
optimization tools based on polyhedra and linear programming (implemented in
our {\bf polyhedral library}.
%\cite{wilde93-poly}).  
Finally, it is precisely
these restrictions that interact coherently and form the foundation of our
transformational system.  They also ensure that all Alpha programs can be put
into a ``normal'' form, and in fact, all examples that we use here have been
so normalized.

\section{Transformations in Alpha}

  We now introduce some of the Alpha transformations used in converting our
program into a single assignment C program.  
\subsection{Serialization}
The first one is used to {\bf
serialize} the reduction operation, and requires two parameters: the
``direction'' in which we accumulate the partial sums, and the name of the
temporary variable.  For our example, if we choose to accumulate in the {\em
increasing} direction of {\tt j} (specified as ``{\tt (i,j->i,j+1)}'') and
name the new variable, {\tt f}, we obtain the Alpha program shown in
Fig.~\ref{fig-serial}.  We also see two new syntactic constructs, the {\em
case} (which has the usual meaning), and the {\em restrict}, which has the
syntax,
 \verb!<polyhedron>: <expr>!, and denotes the expression, {\tt <expr>} but
restricted to the subset of index points in {\tt <polyhedron>}.

  Observe that the domain of {\tt f} is \verb!{i,j| 0<=j<i; 2<=i<=N}! (a
``nearly triangular'' trapezium defined by the points {\tt [2,0]} {\tt [2,1]}
{\tt [N,N-1]} and {\tt [N,0]}).  Also note that the equation for {\tt X} now
has the expression {\tt f[i,i-1]} replacing the reduction, and the new
equation for {\tt f} consists of an initialization at the boundary, and an
accumulation using the {\em body of the original reduction}.  All this is
automatically determined by the system, using the polyhedral library.

\begin{figure}[tb]
%\htmlimage{scale=\myscale}
\setlength{\unitlength}{10mm}
\begin{center}
\begin{picture}(16,5.7)
\thicklines
\put(-0.1,-0.7){\framebox(16.2,6.7){}}
\put(11,0.5){\framebox{The domain of {\tt f}.}}
\put(11,1.3){\begin{picture}(6,3)
              \put(0,0){\includegraphics{figures/domain-f.pdf}}
              \put(-0.3,3.2){$i$}
              \put(0.5,3.6){$j$}
              \put(-1,2.7){\small\tt [2,0]}
              \put(1,2.7){\small\tt [2,1]}
              \put(-1,0){\small\tt [N,0]}
              \put(3.4,0){\small\tt [N,N-1]}
              \end{picture}}
\begin{boxedminipage}[b]{16cm}
\hspace*{2cm}$\vdots$\small
\begin{verbatim}
var
  f : {i,j | (2,j+1)<=i<=N; 0<=j} of real;
let
  f[i,j] = case
        {| j=0} : 0[];
        {| 1<=j} : f[i,j-1] + A * X[j];
           esac;
  X[i] = case
        {| 2<=i} : B[i] - f[i,i-1];
        {| i=1} : B[i];
         esac;
tel;
\end{verbatim}
\end{boxedminipage}
\end{picture}
\end{center}
\vspace*{-0.8cm}
\caption{The forward substitution program after serialization}
\label{fig-serial}
\end{figure}

\subsection{Change of basis}
  Perhaps the most important transformation in the Alpha system is the {\bf
change of basis}.  The intuition behind it is as follows.  Since an Alpha
variable can be viewed as a multidimensional array defined over a polyhedral
domain, we should be able to ``move'' (or otherwise change the ``shape'' of)
its domain and construct an equivalent program.  For example, suppose we want
to rotate the domain of {\tt f} in Fig.~\ref{fig-serial} counterclockwise
by $45^\circ$, i.e., we desire to map the point {\tt [1,0]} to {\tt [1,1]},
and {\tt [1,1]} to {\tt [0,1]} (this makes a vertical line diagonal, and a
diagonal one horizontal).  It is easy to verify that the mapping
 \verb!(i,j -> i-j,i)! achieves this; the resulting program and transformed
domain are shown in Fig.~\ref{fig-transf} (the changes are highlighted).  In
general, when variable {\tt V} is transformed, the system must determine: (i)
its new domain, (ii) the new case structure of its equation, (iii) the new
dependencies for the uses of {\em all} variables in the equation for {\tt V},
and (iv) the new dependencies for {\em uses} of {\tt V} (in all other
equations).  All of this is done automatically using the polyhedral library,
and relies on the fact that the language allows a coherent interaction among
the domains, the dependencies and the (normal form) structure of the programs.

\begin{figure}[tb]
%\htmlimage{scale=\myscale}
\setlength{\unitlength}{10mm}
\begin{center}
\begin{picture}(16,5)
\thicklines
\put(-0.1,-0.7){\framebox(16.2,6.3){}}
%\put(3.4,4.5){\oval(4.7,0.5)}
%\put(2.5,3){\oval(2.2,0.9)}
%\put(6.9,2.8){\oval(5.9,0.5)}
%\put(5.5,1.55){\oval(1.4,0.5)}
\put(11,1){\begin{picture}(6,3)
             \put(0,0){\includegraphics{figures/rot-f}}
             \put(-0.3,3){$i$}
             \put(0.5,3.6){$j$}
             \put(0.3,2.3){\small\tt [2,2]}
             \put(0.8,3.5){\small\tt [1,2]}
             \put(3,3.5){\small\tt [0,N]}
             \put(2.5,-0.3){\small\tt [N,N]}
              \end{picture}}
\begin{minipage}[b]{16cm}
\hspace*{2cm}$\vdots$\small
\begin{verbatim}
  f : {i,j | 1<=i<=j; 2<=j<=N} of real;
let
  f[i,j] = case
        {| i=j}    : 0[];
        {| i<=j-1} : f[i+1,j] + A[j,-i+j] * X[-i+j];
           esac;
  X[i] = case
        {| 2<=i} : B[i] - f[1,i];
        {| i=1} : B[i] - 0[];
         esac;
tel;
\end{verbatim}
\end{minipage}
\end{picture}
\end{center}
\vspace*{-0.8cm}
\caption{The transformed program and the new domain of {\tt f}}
\label{fig-transf}
\end{figure}

The syntax for the affine mapping in a change of basis
transformation is the same as for dependencies, but there is a subtle
difference.  We should be able to transform the variable in any
(affine) manner, but we must always ensure that the {\em number} of
index points remains invariant.  For this, the affine change of basis
must admit an integral inverse (i.e., be {\em unimodular}).  We have
also developed a {\em generalized change of basis}, which allows
``non-square'' mappings (non-unimodular transformations are encoded as
a special case of this).

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapscheduling.tex
Added mmalpha (source and binary)

\newcommand{\aalpha}{{\alfa}}
\newcommand{\lp}{{\sc lp}}
\chapter{\index{Scheduling}Scheduling {\alfa} programs}
\label{chapscheduling}

{\sc Version of \today}

This chapter explains how to use the scheduler for finding valid
execution order of computations of {\alfa} programs.

\section{Introduction}
An {\aalpha} program does not convey any sequential ordering: an
execution order is semantically valid provided it respects the data
dependencies of the program.  Consider for instance the program of
figure~\ref{figSched1} which represents the computation of the product
of a $M\times N$ matrix \texttt{a} by a $M$ vector \texttt{b}. To execute
such a program on a sequential machine, we must chose a
computation
order which respects data dependencies. Such an
order is shown in figure~\ref{figSched2}-(a). Another possible
order would be the demand-driven execution scheme used in the
simulation, as shown in chapter~\ref{chapcodegen}\,: to evaluate a
variable, evaluate recursively the expressions that this variable
depends on.  Parallel executions are also possible. The program of
figure~\ref{figSched2}-(b) shows a possible parallel order for
computing the matrix-vector multiplication.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MatVect.alpha}
\caption{{\aalpha} program for the matrix-vector product, 
no execution order is specified.}
\label{figSched1}
\end{figure}

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\begin{minipage}{8cm}
\begin{verbatim}
For i=1 to N
   C[i,0]=0
EndFor
For i=1 to N
   For j=1 to M
       C[i,j]=C[i,j-1]+
                a[i,j] * b[j]
   EndFor
EndFor
For i=1 to N
   c[i]=C[i,M]
EndFor

           (a) 
\end{verbatim}
\end{minipage}
\begin{minipage}{8cm}
\begin{verbatim}
ForAll i=1 to N
   C[i,0]=0
EndFor
For j=1 to M
   ForAll i=1 to N
       C[i,j]=C[i,j-1]+
                 a[i,j] * b[j]
   EndForAll
EndFor
ForAll i=1 to N
   c[i]=C[i,M]
EndForAll

           (b)
\end{verbatim}
\end{minipage}
\caption{
Two possible order of execution for the program of figure~\ref{figSched1}. The 
left one is parallel. (\texttt{ForAll} represents a parallel loop).}
\label{figSched2}
\end{figure}

The basic goal of the scheduler is to find out valid execution orders, 
called {\em schedules} in what follows. 
A secondary goal is to find out {\em good} schedules. 
However, there is no best schedule: the quality of
a schedule depends on a particular criterion.

In our model, the time is considered as a discrete single clock and we
look for parallel ordering of the computations.  The theoretical basis
for the scheduling process is inherited from the research on systolic
array and automatic parallelization. The scheduler implements two
different procedures: one, called the Farkas method, is defined
in~\cite{Feautrier92aa}; the other one, called the vertex method, is
presented in~\cite{citationdespaa}.

The scheduler of \alfa{} is an analysis tool: it does not modify the
current program in \texttt{\$result}.  The result is the description of a
possible parallel schedule. For instance, for the program of
figure~\ref{figSched1}, if we assume that the inputs are available at
time 0 and that each assignement takes 1 clock tick, we
will obtain the information that, for all \texttt{i}, value {\tt
C[i,j]} can be computed at step \texttt{j}.  Thus, for all \texttt{i}, the
values \texttt{c[i]} can be computed at step \texttt{M+1}. This schedule
can be summarized by:
\begin{equation}
\label{Schedeqn1}
\begin{array}{l} T_a[i,j] = 0 \\ T_b[i] = 0 \\ 
T_C[i,j] = j \\ T_c[i] = M+1 \end{array} 
\end{equation}
This particular schedule corresponds exactly
to the parallel execution order of the program of figure~\ref{figSched2}-(b).

In the present chapter, we are interested in {\em affine by variable}
schedules, i.e schedules in which, for each variable, the execution
date is an affine function of the indices (and possibly of the
structural parameters). The idea of the scheduling process is to
gather all the constraints that the schedule must verify in a linear
programming problem ({\sc lp}) and to solve this \lp{} using a {\sc lp}-solver.

\section{Basic Farkas scheduler}

The scheduling process is quite complex, since many parameters
can influence the result. In this section, 
we present the basics of the scheduler.

\subsection{How to use the \texttt{schedule} function}

The function to be called in order to schedule a program is {\tt
schedule}. Its effects is to find out a schedule for the {\aalpha} program
contained in \texttt{\$result}, and to put the schedule in the
global \mma{} variable \texttt{\$schedule}. Options allow one to modify
parameters, hence the possible forms of a call to the function {\tt
schedule} listed here: 
\begin{itemize}
\item \texttt{schedule[]}\\ finds an affine-by-variable schedule for {\tt
\$result} and assigns it to
\texttt{\$schedule} (this schedule attempts to  minimize
 the global execution time).
\item \texttt{schedule[sys]}\\ 
finds an affine-by-variable schedule for the {\aalpha} system \texttt{sys}
 and assigns it to
\texttt{\$schedule}
\item {\tt
schedule[option\_1->value\_1,\ldots,option\_n->value\_n]}\\
finds a schedule for \texttt{\$result} which respects the chosen options and
assigns it to \texttt{\$schedule}
\item {\tt
schedule[sys,option\_1->value\_1,\ldots,option\_n->value\_n]}\\
finds a schedule for the {\aalpha} system \texttt{sys} which
 respects the chosen options and
assigns it to \texttt{\$schedule}
\end{itemize}

For example, calling the scheduler on the matrix-vector multiplication
of figure~\ref{figSched1} prints out the following informations:
\begin{verbatim}
In[30]:= schedule[];
Dependence analysis...
Duration : 1 for each equation 
Building LP...
LP: 82 Constraints
Writing file for PIP....
Solving the LP...
Version C.3 MultiPrecision (mpip, rev. 1.3.0)
cross : 407646, alloc : 1
Max cross-product result: 4 (1 digits, base 10)
Max numerator term: 8 (2 digits, base 10)
Max denominator: 2 (1 digits, base 10)
n 1 u 130''' s 11'''
T_a{i, j, N, M} = 0
T_b{i, N, M} = 0
T_c{i, N, M} = 1 + M
T_C{i, j, N, M} = j

Out[30]= scheduleResult[1, {{a, {i, j, N, M}, sched[{0, 0, 0, 0}, 0]}, 
   {b, {i, N, M}, sched[{0, 0, 0}, 0]}, {c, {i, N, M}, sched[{0, 0, 1}, 1]}, 
   {C, {i, j, N, M}, sched[{0, 1, 0, 0}, 0]}}]
\end{verbatim}

The first lines indicate which computations are performed.  The Farkas
scheduler first built a \lp{}, then writes this \lp{} in a file,
then calls a \lp{}-solver. Writing this file takes some time, thus
informations are displayed for the user not to become anxious.

The lines starting by \texttt{Version} are output by the \lp{}-solver
\pip{}. The result of calling Pip is then shown. Here, for
instance the schedule says that \texttt{B[i,j,k]} is computed at time
\texttt{i}. The result (after \texttt{Out[30]}) is the corresponding
\mma{} expression representing \texttt{\$schedule}.

For such a little program, only a few seconds are necessary to find
out a schedule. But it may take quite a long time for larger {\aalpha}
programs.

\subsection{Format of the output of \texttt{schedule}}

The output of the \texttt{schedule} function has a special form. 
It is enclosed in a structure whose head is
\texttt{Alpha`ScheduleResult}. The first argument of this
structure is the type of 
the schedule (integer, coded as the option scheduleType, 
see section~\ref{schedType}) and its 
second argument is the schedule itself. 
The syntax of this structure is:\\
\texttt{
<schedResult> ::= Alpha`scheduleResult[scheduleType\_Integer,\{sched...\}]\\
$\begin{array}{l} \mbox{ <sched>} ::= \\ ~\\  ~ \end{array}\begin{array}{l} \mbox{\{ nameVar\_String,}\\
          \mbox{ indices\_List,}\\
           \mbox{ Alpha`sched[tauVector\_List,constCoef\_Integer] \}}
\end{array}$
}\\
The effect of the \texttt{schedule} function is to assign this form 
to te global variable \texttt{\$schedule}.

Another example is given in section~\ref{ExampleSched}. 
A schedule can be displayed using the \texttt{show} command:
\begin{verbatim}
In[31]:=show[$schedule]

T_a{i, j, N, M} = 0
T_b{i, N, M} = 0
T_c{i, N, M} = 1 + M
T_C{i, j, N, M} = j

Out[31]:=
\end{verbatim}

\subsection{Using the result of the scheduler}

We may use the result of the scheduler as an information on the program, 
but we can also tranform the {\aalpha} program such that this information
become explicit in the program.  For instance, on the matrix-vector example, 
if we perform the \index{change of basis}change of basis
\texttt{(i,j -> j,i)} on variable \texttt{C}
and if we rename (for instance) 
the first index \texttt{t} and the second one \texttt{j},
we obtain the program of figure~\ref{figSched3}.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MatVectSched.alpha}
\caption{Scheduled version of  program of figure~\ref{figSched1}}
\label{figSched3}
\end{figure}

This program is functionnaly equivalent to the one of figure~\ref{figSched1}
but it shows exactly at which clock tick each  value of the local variable
\texttt{C} is computed: \texttt{C[t,i]} is computed at step \texttt{t}. We call
a program which has this property a \index{scheduled program} {\em
scheduled} {\aalpha} program.  Of course, in a scheduled {\aalpha}
program, the first index of the local variables defines a partial
order which is compatible with data dependencies.

The scheduled program of figure~\ref{figSched3} has been obtained by
evaluating the expression
\index{applySchedule}\texttt{applySchedule[]}.  This command applies a
change of basis to each local variable of the program contained in
\texttt{\$result} in such a way that the first index in the new basis
represent the time. The change of basis is not performed on the inputs
and outputs, as otherwise the new program would not be equivalent to the
previous one. Another example of the use of \texttt{applySchedule[]}
is shown in section~\ref{ExampleSched}.

\section{Advanced scheduling}
In this section we introduce some of the parameters of the 
\texttt{schedule} command which allow more precise results to be
obtained.  A detailed description of this function can be found in
the documentation 
file \texttt{\$MMALPHA/doc/user/Scheduler\_user\_manual.ps}.

\subsection{Options of schedule}
The \texttt{schedule} function has many options. These options
have default values indicated hereafter. To change these
values, put one of the corresponding replacement
rules as a parameter to the
\texttt{schedule} function (see the example in section~\ref{ExampleSched}).

\paragraph{Option \texttt{scheduleType}}
\label{schedType}
This option gives the type of schedule.
Its possible values are:
\begin{itemize}
\item \texttt{scheduleType -> affineByVar} (default) 
affine by variable scheduling;
\item \texttt{scheduleType -> sameLinearPart} affine by variable scheduling with constant linear part;
\item \texttt{scheduleType -> ?}  affine by variable scheduling with constant linear part except for the parameters.
\end{itemize}

\paragraph{objFunction }
This option gives the objective function chosen. Its type is integer,
the possible values are:
\begin{itemize}
\item \texttt{objFunction -> 0 } (default) the total latency is minimized. 
\item \texttt{objFunction -> 1 } no objective function (the
coefficients of the scheduling vectors are minimized in a
lexicographic order).

It is mandatory not to minimize the total execution time when this 
time is not bounded (otherwise, the schedule will fail).
\end{itemize}

\paragraph{ratOrInt}
The resolution of the linear programming problem generated is done 
by the software {\sc mp}Pip which is an infinite precision version
 of the sofware {\sc pip}~\cite{Feautrier88a}. The resolution can be done with
integer programming or rational linear programming.
  This option indicates whether the resolution of the {\sc lp} by {\sc mp}Pip
 is
 done in integer mode (resulting scheduling vectors with integral
 coefficients) or in rationnal mode (resulting scheduling vectors may contains
 rationnal coefficients).  Its type is integer, the possible values
 are:
\begin{itemize}
\item \texttt{ratOrInt -> 0 } : rationnal solution.
\item \texttt{ratOrInt ->  1 } : (default) integer solution.
\end{itemize}

\paragraph{addConstraints}
This option allows to add some constraints to the {\sc lp} generated.
This option is very usefull to guide the scheduling process.
 Its type is a list of string, each string representing a constraint. The
 constraints authorized are affine constraints on scheduling
vectors.
There are two types of  constraint added, one can force a scheduling
vector value or simply set linear constraints on its components. 
\begin{itemize}
\item forcing a  variable $A$ to be schedule at time i+2j+2 can be done 
with the constraints: \texttt{"TA[i,j]=i+2j+2" }. 
\item for more precise constraint,  one can directly access to each components 
of the schedule functions of each variable. For instance \texttt{TAD2}
 will represent the variable coefficient of the second indice
 in the schedule of variable \texttt{A} (and \texttt{CA} will represent 
the constant part in the schedule of variable \texttt{A}).
 With these names, one can set linear constraints 
on these coefficients using operators \texttt{==} or \texttt{>=}. For instance,
\texttt{\{"TAD1 == 1","TAD2 == 2", "CA >= 2" \} } is the same constraint
as above except thar the constant is allowed to be greater than two.
\end{itemize}  
Example of use:\\
\texttt{schedule[addConstraints->\{"TA[i,j,N]=i+2j-2",
"TBD2==2","TBD1+2TCD3>=1"\}]}

\paragraph{duration} indicates how to count the duration of each computation. The possible value for the duration option are: 
\begin{itemize}
\item \texttt{duration -> \{\} }: (default) each equation takes 1 step
	 (whatever complex the computation is).
\item \texttt{duration -> \{Integer...\} }:

The duration of each equation is given by the user. The list must
contain as many integer as there are variables (do not forget the
input variables). For instance on the program of figure~\ref{figSched1}, the
command \texttt{schedule[]} is equivalent to the command:
\texttt{schedule[duration->\{1,1,1,1\}]}.
\end{itemize}

\section{Another example}
\label{ExampleSched} 
Consider the program of figure~\ref{figSched4}, which represent a
uniform program for the \index{multiplication of matrices}
multiplication of matrices. in this section, we give the result of the
schedule with different options.
\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MMunif.alpha}
\caption{Uniform matrix matrix product}
\label{figSched4}
\end{figure}

\paragraph{Default use}
If you type the following command (after having loaded the program of figure~\ref{figSched4}):\\
\texttt{schedule[]}\\
the result should be: 
\begin{verbatim}
T_a{i, j, N} = 0
T_b{i, j, N} = 0
T_c{i, j, N} = 1 + 2 N
T_B{i, j, k, N} = i
T_A{i, j, k, N} = j
T_C{i, j, k, N} = k + N

Out[14]= scheduleResult[1, {{a, {i, j, N}, sched[{0, 0, 0}, 0]}, 
   {b, {i, j, N}, sched[{0, 0, 0}, 0]}, {c, {i, j, N}, sched[{0, 0, 2}, 1]}, 
   {B, {i, j, k, N}, sched[{1, 0, 0, 0}, 0]}, 
   {A, {i, j, k, N}, sched[{0, 1, 0, 0}, 0]}, 
   {C, {i, j, k, N}, sched[{0, 0, 1, 1}, 0]}}]
\end{verbatim}
The result (after \texttt{Out[14]}) is the
Mathematica structure assigned to \texttt{\$schedule}.

\paragraph{Adding a constraint}
Suppose that the inputs \texttt{a[i,j]} and \texttt{b[i,j]} arrive
respectively at time \texttt{i+j} and \texttt{i+j+3}. Moreover, we
want a schedule of unique linear part (because this allow a locally
connected array to be obtained from a uniform program).  We have to
add the two constraints: \texttt{"Ta[i,j]=i+j"} and
\texttt{"Tb[i,j]=i+j+3"}. Also, we need to set the option scheduleType
to 2, hence the command is:
\begin{verbatim}
In[15]:= schedule[addConstraints->{"Ta[i,j]=i+j","Tb[i,j]=i+j+3"},
                   scheduleType->2]
\end{verbatim}
The result is 
\begin{verbatim}
T_a{i, j, N} = i + j
T_b{i, j, N} = 3 + i + j
T_c{i, j, N} = 5 + i + j + N
T_B{i, j, k, N} = 3 + i + j + k
T_A{i, j, k, N} = i + j + k
T_C{i, j, k, N} = 4 + i + j + k
\end{verbatim}

Note that the unique linear part option is applied on local variables
only (indeed, it cannot be enforced on inputs and outputs which do not have
the same number of indices.) After scheduling, one can obtain a
scheduled program by typing the command: \texttt{applySchedule[]}. The
result is the program of figure~\ref{figSched5}.

Warning: the \texttt{applySchedule} function finds the change of basis 
by unimodular completion of the scheduling vector. The unimodular 
completion is not always possible (for instance, if the scheduling 
vector is null), in that case, applySchedule will perform 
a generalized change of basis and add a new dimension to the domain of 
the corresponding variable. A warning is printed out during this operation.

\begin{figure}[htbp]
%\htmlimage{scale=\myscale}
\verbatiminput{alpha_progs/MMunifSched.alpha}
\caption{Scheduled  matrix matrix product}
\label{figSched5}
\end{figure}

\subsection{What if no schedule can be found?}
Due to the restriction to linear schedule, there may 
happen that the schedule function only answers: 
\texttt{No schedule was found, sorry...}
There maybe several reasons for that:
\begin{itemize}
\item The program is not semantically correct, try \texttt{analyze[]}.
\item There exists a schedule but the time is not bounded. 
In this case try with the option \texttt{objFunction} set to 1.
\item No affine one dimensional schedule exists. You can try to find
a multidimensional schedule with the \texttt{multiSched}.
\item No schedule exists (there is no way of solving this problem).
\end{itemize}

\section{To come soon}
\begin{itemize}
\item Multi-dimensional schedule.
\item Structured scheduling.
\item Data-flow scheduling. 
\end{itemize}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapstart.tex
Added mmalpha (source and binary)

\chapter{A guided tour of \alfa{}}
\label{chapstart}

{\sc Version of \today}

This chapter briefly presents the main features of the \alfa{} 
language and the basic operations of the \mmalfa{} environment.
Section~\ref{chapstart:examples} presents several examples of 
\alfa{} programs. In section~\ref{chapstart:basic}, we introduce
and illustrate a few basic manipulations of \alfa{} programs, 
whereas section~\ref{chapstart:advanced} presents more advanced
transformations. Structured programs are shown in~\ref{chapstart:sectstruct}.

\section{Examples of {\Alpha} programs}
\label{chapstart:examples}
For a complete description of the {\Alpha} language, refer to
appendix~\ref{alpha1} and~\ref{alpha2} or
to~\cite{wilde-tech94,Mauras89,DupontQuRi95}. We introduce the basic
features of the langage on the exemple of the matrix-vector
multiplication.

An {\Alpha} program is a {\em system}. \alfa{} variables are generalized
arrays which can have any shape (not just rectangles).  The set of
indices of the array is called the {\em domain} of the variable.
Example~\ref{ex1} below shows the declaration of a variable {\tt a}
whose domain is  the set of points $(i,j)$ in the triangle:  $0 \leq i \leq j$;
$j \leq 10$.

\begin{ex}
~\\
\texttt{a : \{i,j~|~ 0<= i <= j; j <=10 \} }
\label{ex1}
\end{ex}

An \alfa{} system has input and output variables, it may have local
variables and also size parameters which allows parameterized programs
to be defined.  Variables are defined by a single equation (which
usually has the form of a recurrence equation). The {\tt case}
construct allows one to define different values in different parts of
the domain.

\begin{ex}
\texttt{
\\
a[i,j] = \\
case\\
\ \ \ \ \{| j = 0 \} : 0[];\\
\ \ \ \   \{| j > 0 \} : a[i,j-1]+1[];\\
         esac;
}
\label{ex2} 
\end{ex}

The {\Alpha} expression of example~\ref{ex2} defines the values of
{\tt a[i,0]} to be zero\footnote{Syntactic note: constants are zero
dimensionnal arrays hence the empty brackets in \texttt{0[]}. This
notation is the source of many syntax errors, and we plan to modify
it...} and recursively defines {\tt a} at all the other points in its
domain. This equation defines a variable {\tt a} such that {\tt
a[i,j]=j}. The above equation is printed in {\em array notation}.  The
real syntax of {\Alpha} is sligthly less readable but more consistent
and logical from a semantic point of view.  To illustrate this,
example~\ref{ex3} shows the same definition of {\tt a} in standard
notation.

\begin{ex}
{\tt 
~\\
a = \\
case\\
\ \ \ \ \       \{i,j | j = 0 \} : 0.(i,j->);\\
\ \ \ \ \      \{i,j | j > 0 \} : a.(i,j->i,j-1)+1.(i,j->);\\
    esac;\\
}
\label{ex3}
\end{ex}

Expression {\tt expr = a.(i,j->i,j-1)} should be read as {\em {\tt
a} composed with the dependency function} \texttt{f(i,j)=(i,j-1)}. In
other words, {\tt expr[i,j]} has value {\tt a[i,j-1]} at each point
\texttt{(i,j)} such that \texttt{(i,j-1)} is in the domain of {\tt a}. 
Similarly {\tt expr2 = 0.(i,j->)} means that {\tt expr2[i,j]} has
value \texttt{0} for all \texttt{(i,j)}.

Note also that there is no sequential in the different
computations: interchanging the two branches of the case in
example~\ref{ex2} would define exactly the same value for {\tt a}:
\alfa{} is therefore a declarative language.
The
evaluation order is implicit and there are tools for finding schedules
for a given program. As {\Alpha} is a functional language, the only
constraint that an evaluation order must follow is the data
dependencies between variables. In example~\ref{ex2},
obviously {\tt a[i,j]} must be computed after {\tt a[i,j-1]}.

More details on the langage can be found in
appendix~\ref{alpha1}. Figure~\ref{fig1} shows an \alfa{} program to
compute the multiplication of a matrix of size $n\times n$ and a vector
of size $n$.
\begin{figure}[htbp]
\begin{verbatim}
system prodVect: {N | N>1}
               (a : {i,j|1 <= i,j <= N} of integer; 
                b : {i|1 <= i <= N} of integer)
       returns (c : {i|1 <= i <= N} of  integer);
  var 
	C :  {i,j|1 <= i <= N; 0<= j <=N} of  integer;
  let
    C[i,j] = case
              {|j=0} : 0[];
              {|j>=1} : C[i,j-1] + a[i,j] * b[j];
             esac;
    c[i]=C[i,N];
  tel;   
\end{verbatim}
\caption{{\Alpha} program describing the matrix vector multiplication}
\label{fig1}
\end{figure}

\section{Basic manipulation of {\Alpha} programs}
\label{chapstart:basic}

This section presents the first commands that you should learn in
order to deal with {\Alpha} programs. Before reading this section, you
should have set the different environment variables necessary to use
the {\mmalfa{}} environment. This procedure is described in
appendix~\ref{install}.

There are several ways of using \mmalfa{}:
\begin{enumerate}
\item Using the notebook interface. Type \texttt{mathematica} under
\texttt{Unix}, or start \mma{}~3.0 in the Programs menu of 
Windows NT.
\item Using the \mma{} kernel directly. Type \texttt{math} under
\texttt{Unix}, or start \mma{}~3.0~Kernel in the Programs menu of 
Windows NT.
\item Using the \mma{} kernel via \texttt{emacs} (\texttt{Unix}
only).
\end{enumerate}

The second method is not recommended. The first one is easy, but
some people prefer using the third one. In the following, we shall
assume that we interact with the kernel (using the second
or the third method), but all commands we describe can also be
used in a notebook.

Once {\mmalfa{}} is installed, start \mma{} and write an {\Alpha}
program (such as the one of figure~\ref{fig1} for instance) using your
favorite text editor. Say you called this file {\tt
prodVect.alpha}. The commands described in this section allow you to
{\bf load} your program into \mma{}, {\bf view} the program in
\mma{} (array notation or standard notation), {\bf save} the
program in another file, perform {\bf static analysis} and {\bf schedule}
the program. By the way, all these examples are alse available
in the \texttt{Getting-started} notebook accessible by the 
\texttt{Master} notebook of \mmalfa{}.

\subsection{Loading and viewing an {\Alpha} program}
When \mma{} is loaded, a few welcome messages about {\Alpha} are
printed out. 
Then one gets the usual \mma{} prompt:\\ {\tt In[1]:=
}\\ 
The name of the working directory can be printed out
by typing:\\
{\tt In[1]:=Directory[] }\\
If you see that this directory is not the one where you have put 
{\tt prodVect.alpha}, change it by typing:\\
{\tt In[2]:= SetDirectory[" the directory you want "] }\\

You can now {\bf load} the {\Alpha} program into \mma{}
by typing:\\
{\tt In[3]:= load["prodVect.alpha"]; }\\
Note that most often, \mmalfa{} commands should end with a semi-colon.
The reason is the following one: \mmalfa{} commands are \mma{} functions, 
which most often return a transformed \alfa{} program, expressed
as its AST. 
If you forget the {\tt ;} symbol,
\mma{} just prints out the result of the function evaluation, 
which sometimes may take a few pages... 
The side effect of \texttt{load} is to assign the AST  to 
the global \mma{} variable {\tt \$result}.

You can view the program that has been stored in {\tt \$result} by typing:\\
{\tt In[4]:= show[] } \\
By default, \texttt{show} pretty prints the program contained 
in {\tt \$result}, but more generally, \\\texttt{show[ var ]}\\would
pretty print the program contained in \mma{} variable \texttt{var}.

By the way, all \mma{} functions have an on-line documentation: 
\texttt{?show} gives the help on \texttt{show}. Commands
may have options. Type \texttt{Options[ command ]} to list
the options of \texttt{command} together with their default value.

You may have noticed that the program printed on the screen looks
different from the one of figure~\ref{fig1}: this program is in
standard notation. If you want to print it in array notation (which is
much more readable) you should evaluate: \\ {\tt In[6] := ashow[]}\\

To save the {\Alpha} program in another file, use
the command {\tt save} (or {\tt asave} which writes in array
notation).  For instance: \\ {\tt In[7]:= asave["myFile.alpha"] }\\
will write program of figure~\ref{fig1} in file {\tt
myFile.alpha}.
This command is needed if one wants to save
the content of \texttt{\$result} after some transformations.

\subsection{Analyzing and simulating an {\Alpha} program}

Now that you have loaded an {\Alpha} program, you can start working on
it. Your first action should be to check it for so-called {\em static
errors} by using the command {\tt analyze}:\\ {\tt In[8]:=
analyze[]}\\ Information about possible errors of the {\Alpha} program
are printed out. If the analysis is successful, the result is {\tt
True}. The static analyzer of \alfa{} does essentially two verifications: 
it checks the type of expressions -- this is not a fantastic 
novelty, -- but it also checks that variables have a definition 
in any point of their domain definition. This second verification
is very powerful, and is much more original.

Another interesting analysis tool is the {\tt scheduler}. It
finds (whenever possible) a linear schedule for your {\Alpha} program that
minimizes its execution time. The use of the scheduler is detailed in
subsection~\ref{schedule}.

Suppose now that you want to evaluate the {\Alpha} program you just
loaded. There is no real compiler for {\Alpha} but we can generate
{\tt C} code that evaluates \alfa{} in a demand driven way. 
The command for generating {\tt C} code is {\tt writeC}:\\
{\tt In[9]:= writeC["prodVect.c","-p 10"] }\\ The {\tt "-p 10"}
argument indicates that the value of the parameter {\tt N} will be set
to 10 (the {\tt C} code is not parameter independent).
By
default, this program, once compiled, reads its input from the
standard input ({\tt stdin}) and prints on the standard output ({\tt
stdout}).

\section{Advanced manipulation of {\Alpha} programs}
\label{chapstart:advanced}

In this section we briefly review a few more advanced manipulations of
{\Alpha}. Additional information is in the {\Alpha} tutorial and the
{\Alpha} reference manual (these documents are part of the \mmalfa{}
distribution).

\subsection{Pipelining}
Pipelining is a transformation widely used in systolic synthesis.
It is also called {\em localization} or {\em uniformization}.
It consists basically of replacing a broadcast 
by the pipeline of this value through all the computations that 
need it.

For instance, in the program of figure~\ref{fig1}, we see (last 
term in the second branch of the {\tt case}) that 
{\tt b[j]} is used for the computation of {\tt C[i,j]} 
for all $i$, $0\leq i \leq N$.

To introduce a new variable {\tt B1} which will pipeline the {\tt b[j]} 
value from the computation of {\tt C[j,0]} to 
{\tt C[j,1]}, ... , {\tt C[j,N]}, we
use the following command:\\
{\tt In[10]:= pipall["C","b.(i,j->j)","B1.(i,j->i+1,j)"]; }\\
In this expression, the first argument is the variable whose
equation is to be modified, the second argument is the
expression to be pipelined (standard notation is mandatory here)
and the last argument indicates the {\em direction} of the pipeline
as well as the name of the new variable introduced. 
After the
execution of this command, the program contained in {\tt \$result} is
the one shown in figure~\ref{fig2}.

\begin{figure}[ht]
\begin{verbatim}
system prodVect :{N | 2<=N}
                (a : {i,j | 1<=i<=N; 1<=j<=N} of boolean; 
                 b : {i | 1<=i<=N} of boolean)
       returns  (c : {i | 1<=i<=N} of boolean);
var
  B1 : {i,j | 1<=i<=N; 1<=j<=N; 2<=N} of boolean;
  C : {i,j | 1<=i<=N; 0<=j<=N} of boolean;
let
  B1[i,j] = 
      case
        {| i=1; 1<=j<=N; 2<=N} : b[j];
        {| 2<=i<=N; 1<=j<=N} : B1[i-1,j];
      esac;
  C[i,j] = 
      case
        {| j=0} : False[];
        {| 1<=j} : C[i,j-1] + a[i,j] * B1;
      esac;
  c[i] = C[i,N];
tel;
\end{verbatim}
\caption{{\Alpha} program of figure~\ref{fig1} after pipelining of {\tt b}
in the definition of \texttt{C}}
\label{fig2}
\end{figure}

Unlike previous transformations, pipelining changes the
{\Alpha} program, but the resulting program is equivalent to
the initial one. 
The modifications performed automatically by \texttt{pipeAll} are:
\begin{enumerate}
\item Determine the domain of {\tt B1} and add a
declaration for it.
\item Build the definition of {\tt B1} based on the dependency ({\tt (i,j->i+1,j)}) and the initialisation given ({\tt b.(i,j->j)}).
\item Replace the original expression ({\tt b.(i,j->j)}) by {\tt B1}.
\end{enumerate}

\subsection{Change of basis}
The change of basis is another important transformation in systolic
array design. It allows 
variables to be re-indexed, and is often used to map indices to time and space.

In the example of figure~\ref{fig2}, suppose that we wish to express
the computations in a new index basis {\tt i',j'} such that {\tt i'=i+j}, 
{\tt j'=j}. We can perform the following change of basis:
\\ {\tt In[11]:= changeOfBasis["C.(i,j->i+j,j)"]; } \\ This simply
indicates that the transformation is to be applied to variable {\tt C}
and that the new coordinates in term of the old ones are given by
the linear function \texttt{(i,j->i+1,j)}.
Note that a change of basis is meaningful only if
this linear function admits an integral left inverse: in this
example, its left inverse is obviously \texttt{(i,j->i-1,j)}. The
resulting program is shown in figure~\ref{fig3}.

\begin{figure}[ht]
\begin{verbatim}
system prodVect :{N | 2<=N}
                (a : {i,j | 1<=i<=N; 1<=j<=N} of boolean; 
                 b : {i | 1<=i<=N} of boolean)
       returns  (c : {i | 1<=i<=N} of boolean);
var
  B1 : {i,j | 1<=i<=N; 1<=j<=N; 2<=N} of boolean;
  C : {i,j | j+1<=i<=j+N; 0<=j<=N} of boolean;
let
  B1[i,j] = 
      case
        {| i=1; 1<=j<=N; 2<=N} : b[j];
        {| 2<=i<=N; 1<=j<=N} : B1[i-1,j];
      esac;
  C[i,j] = 
      case
        {| j=0} : False[];
        {| 1<=j} : C[i-1,j-1] + a[i-j,j] * B1[i-j,j];
      esac;
  c[i] = C[i+N,N];
tel;
\end{verbatim}
\caption{{\Alpha} program of figure~\ref{fig2} after the change of basis on 
 {\tt C}}
\label{fig3}
\end{figure}

%\subsection{Substitution}
%{\tt To be written soon...}
\subsection{Normalization}
\label{normalization}
The normalization transformation simplifies an {\Alpha} program into a
{particular} normal form called {\em case-restriction-dependency}.
This function is very useful when one performs several automatic
transformations that may render the program less and less
readable. The command is simply: \\
{\tt In[12]:= normalize[];}\\This transformation is illustrated in the
\alfa{} tutorial.

\subsection{Scheduling}
\label{schedule}

The {\tt schedule} command looks for a schedule for an {\Alpha}
program.  The basic goal of the scheduler is to find a valid and good
evaluation order. Here, the term {\em good} depends on the
optimization criterion choosen: most often, it is the total evaluation
time of the program, but one may also consider other criteria.

The time is considered as a discrete single rate clock. The overall
idea of the scheduling process is to build a linear programming
problem (LP) and to solve it with a software tool: 
this may be \pip{}\cite{pip}, or \lpsolve{}\cite{lpsolve}, or even the \mma{}
linear solver.

The \alfa{} scheduler provides several options to schedule 
a program. We consider here 
the simplest one (by default), called 
{\em  monodimensional affine-by-variable schedule}. 
This esoteric name means 
that the evaluation date $T_A(i,j)$ of a given computation
\texttt{A[i,j]}  
is given by an affine
function of the indices and parameters:\\
\centerline{ $T_{A}(i,j) =
\tau^i_{A} i + \tau^j_{A} j+ \tau^N_{A} N +
\alpha_{A}$} where $N$ is a 
parameter of the {\Alpha} program.
The coefficients of this function
are (in general) different for each variable in the system

The command to schedule a program is :\\ 
\texttt{schedule[]}\\By default, it schedules {\tt \$result}
and the resulting schedule is
placed in a global variable named {\tt \$schedule}.

The \texttt{schedule} function has many options 
(type {\tt Options[schedule]} for further information.)
Some uses of the function {\tt schedule} are:
\begin{itemize}
\item {\tt schedule[]}\\ find an affine by variable schedule  for {\tt
\$result} which minimizes the global execution time and assign it to
{\tt \$schedule}.
\item {\tt schedule[sys]}\\ find an affine by variable schedule for
the {\Alpha} system {\tt sys} which minimizes the global execution
time and assign it to {\tt \$schedule}.
\item {\tt
schedule[\{option1->value1,\ldots,optionn->valuen\}]}\\
find a schedule for {\tt \$result} which respects the chosen options and
assign it to {\tt \$schedule}.
\item {\tt
schedule[sys,\{option1->value1,\ldots,optionn->valuen\}]}\\
find a schedule for the {\Alpha} system {\tt sys} which respects the chosen options and
assign it to {\tt \$schedule}.
\end{itemize}
The schedule function is explained in more detail in
the scheduler documentation
given in file:
\begin{verbatim}
$MMALPHA/doc/user/docSched.dvi
\end{verbatim}

\section{Structured  {\Alpha} programs}
% Macros pour la syntaxe
\label{chapstart:sectstruct}

\index{genericity}\index{structured programming}\index{program structures}\index{structures of programs}
\alfa{} programs can be structured: this section explains 
how this can be done.

\subsection{Simple structures} 
Let us write an {\alfa} program for the addition of two integers
(or fixed-point numbers) expressed as bit vectors.  A binary
adder is classically described as a sequence of \emph{full adder}
operations with the propagation of a carry bit from one full adder to
the next one, as shown in Fig.\ref{binaryadder}.
\index{adder}\index{binary addition}

\begin{figure}[!ht]
  \centerline{
    \includegraphics{adder.pdf}
  }
\caption{Addition of two integers (coded as bit vectors), using \emph{full adders}\label{binaryadder}}
\end{figure}

The following \alfa{} system describes a \emph{full adder}:
\index{full adder}
\begin{verbatim}
      system FullAdder (A,B,Cin : boolean) 
               returns (X,Cout : boolean);
      let
        X = A xor B xor Cin;
        Cout = (A and B) or (A and Cin) or (B and Cin);
      tel;
\end{verbatim}

To build an adder using this program, we need to 
instanciate a collection of such system,
as shown in Fig.\ref{binaryadder}. The shape of this collection may
be expressed as the {\alfa} domain \texttt{\{b|0<=b<W\}} where
\texttt{W} is a size parameter giving the number of bits of the
adder.
\index{use statement}

The \textbf{use} construct of {\alfa} allows precisely that: the
following system describes in {\alfa} the adder given in
figure~\ref{binaryadder}:
\index{adder}\index{binary addition}

\begin{verbatim}
system Plus: {W|W>1} (A,B: {b| 0<=b<W} of boolean)              -- 1       
             returns (S : {b| 0<=b<=W} of boolean);             -- 2       
var                                                             -- 3       
  Cin, Cout, X : {b| 0<=b<W} of boolean;                        -- 4       
let                                                             -- 5       
  Cin[b] =                                                      -- 6       
    case                                                        -- 7       
      {| b=0} : 0[];                                            -- 8       
      {| b>0} : Cout[b-1];                                      -- 9       
    esac;                                                       -- 10      
  use {b| 0<=b<W} FullAdder[] (A,B,Cin) returns(X, Cout);       -- 11      
  S[b] =                                                        -- 12      
    case                                                        -- 13      
      {| b<W} : X;                                              -- 14      
      {| b=W} : Cout[W-1];                                      -- 15      
    esac;                                                       -- 16      
tel;                                                            -- 17      
\end{verbatim}

\index{use statement}
In this system, line 11 reads as follows: 
\begin{quote}
``Use (or instantiate)
a collection of
instances of the subsystem \texttt{FullAdder}. This collection has the
shape of the extension domain\index{extension domain} \verb!{b| 0<=b<W}! and is thus indexed
by index \texttt{b}. Let the inputs of the \texttt{b}-th instance be
the variables \texttt{A}, \texttt{B} and \texttt{Cin} at point
\texttt{b}, and similarly let the outputs of this collection of
instances be the variables \texttt{X} and \texttt{Cout}.''
\end{quote}
Lines~6-10 describe the carry propagation, and lines~12-16
define the output of this binary adder.

\index{dimension extension}
In other words, line~11 is a shortcut for the following equations,
which are those of the system \texttt{FullAdder} whith the dimension
of the variables extended from zero to one:

\begin{verbatim}
        X[b] = A[b] xor B[b] xor Cin[b];
        Cout[b] = (A[b] and B[b]) or (A[b] and Cin[b]) or (B[b] and Cin[b]);
\end{verbatim}

\subsection{Syntax of the \emph{use} construct}

\index{use syntax}
The \textbf{use}\ construct appears at the syntactic level of an
equation, since it is basically a shortcut for a set of
equations. Here is the general syntax of an equation/use (see
appendix~\ref{alpha1} for the meta syntax):
{\tt
\begin{tabbing}
xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\=  \kill
\textsl{Equation} ::=\\
\>\> \textsl{Identifier} = \textsl{Expression} ;\\
\> \Alt\ \> use \Opt{ \textsl{ExtensionDomain}} \textsl{Identifier}\\
\>\>\>\>\>\> \Opt{[\textsl{ParamAssignment} ]} \\
\>\>\>\>\>\>( \textsl{ExpressionList} )\\
\>\>\> returns\>\>\> ( \textsl{IdentifierList} ) ;
\end{tabbing}
}

\index{parameter assignation}\index{assignation of the
parameters}\index{parameters}
In this syntax we see that there is an optional \emph{parameter
assignment} which is discussed in the following.
In the previous addition the subsystem \texttt{FullAdder} has
no parameters, and the parameter assignment is therefore empty.

\subsection{Manipulating structured programs}

A structured program is stored in {\mmalfa{}} as a {\mma{}} list of
systems called a \emph{library}. The default library is stored in the
global variable \verb!$library!.\index{$library}

A structured program may be written in one single file or several
distinct files.  In the former case the \texttt{load[]} function
returns a library composed of all the systems contained in the file,
and stores this library in \verb!$library!. %$

If the program is stored in several files, it is the responsability of
the user to build a proper library, i.e. a {\mmalfa{}} list of all the
systems needed by the hierarchical structure of the program. For this
purpose, the user will typically use {\mmalfa{}} list manipulating
functions such as \texttt{Join[], Append[]}, etc.\index{structured
programs}

In addition, two functions, \texttt{putSystem[]} and
\texttt{getSystem[]}, may be used to extract a system from a library
and to put back a modified system in a library. Typically a
system is extracted from the library as the {\em current}
system, modified by some program
transformation, and then put back in the library.\index{putSystem[]}\index{\getSystem[]}

\subsection{Program transformations associated with structures}

Most \mmalfa{} functions handle parameterized programs and \emph{use}
statements. There are, however, some major exceptions such as the
\texttt{writeC[]} translator \index{writeC[]} which generates code
only for flat {\Alpha} programs without subsystems. 
\mmalfa{}
provides functions to transform a structured program
into a ``flat'' equivalent one~:

\begin{itemize}
\item \texttt{assignParameterValue[]} \index{assignParameterValue[]}
gives a value to a size parameter, i.e. it refines a generic system
into a specialized one.
\item \texttt{inlineSubSystem[]} \index{inlineSubSystem[]} expands a
\emph{use} statement, replacing it with the equations of the
corresponding subsystem, properly modified to take the dimension
extension into account.\index{flattening a structured program}
\item \texttt{inlineAll[]} \index{inlineAll[]} recursively flattens a
structured {\alfa} program.
\end{itemize}
For more information see 
the subsystem documentation
\begin{verbatim}
$MMALPHA/doc/user/SubSystems.dvi
\end{verbatim}

\section*{Appendix: The polyhedral library}
\label{polylib}
See~\cite{Wilde93}(\texttt{http://www.irisa.fr/EXTERNE/bibli/pi/pi785.html}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/chapsubsystems.tex
Added mmalpha (source and binary)

% Macros pour la syntaxe

\chapter{Structured programming in {\alfa}}
\label{chapssubsystems}

\index{genericity}\index{structured programming}\index{program structures}\index{structures of programs}

\section{Introduction}

This chapter presents  generic and structured
programming in {\alfa}.

It first introduces the notion of \emph{size parameters} of an {\alfa}
system: a size parameter is a special index, often noted in capital
letters, used to give generic values to the size of the
problem. For example, when talking of a NxN matrix product, ``N'' is a
size parameter. In an {\alfa} system, such a parameter will appear in
the domains and affine functions as an index which is global to all
the domains of the system. The reader may already have noticed such
size parameters in some of the previous examples, as their use is
quite intuitive. This chapter will present them in greater detail.

The \emph{program structures} in {\alfa} are then introduced: so far
all the algorithms were simple enough to be described as a single
{\alfa} system, but one may want to structure a complex problem into
several smaller and simpler {\alfa} systems. The structure mechanisms
presented here allow this.

\section{Parameters\label{parameters}}

In most applications, it is useful to deal with \emph{parameterized}
algorithms. The polyhedral SARE model allows for a simple but powerful
parameter scheme which is demonstrated in the following program, a
simple matrix-vector product.

\begin{verbatim}
system matvect :{N | N>1}                               -- 1  
               (M : {i,j | 1<=i,j<=N} of real;          -- 2  
                V : {j | 1<=j<=N} of real)              -- 3  
       returns (R : {i | 1<=i<=N} of real);             -- 4  
var                                                     -- 5  
  C : {i,j | 1<=i,j<=N} of real;                        -- 6  
let                                                     -- 7  
  C = case                                              -- 8  
        {i,j | j=0} : 0.(i,j->);                        -- 9  
        {i,j | j>0} : C.(i,j->i,j-1) + M * V.(i,j->j);  -- 10 
      esac;                                             -- 11 
  R = C.(i->i,N);                                       -- 12 
tel;                                                    -- 13 
\end{verbatim}

In this system, \texttt{N} is a \emph{size parameter}\index{size
parameters}\index{parameters} which is declared in the header of the
system using  a \emph{parameter domain},\index{parameter
domain}\index{domain of the parameters} here \texttt{\{N |N>1\}}. This
parameter domain may be any {\alfa} domain, in particular there is no
limit on the number of parameters, and this domain may express any
affine inequation between the parameters. For example, \texttt{\{M,N|
M<2N\}} is a valid parameter domain.

Such a parameter may then appear anywhere in the system where indices
are allowed, that is, in domains (see lines~2-4 of
the previous program) and in affine functions (see line~12).

This parameterization retains all the properties of the language, for
parameters are actually special indices which are shared by all the
variables and subexpressions of a system. For example the
parameterized domain of the variable \texttt{M}, defined line~2 as
\texttt{\{i,j| 1<=i,j<=N\}}, actually denotes the domain
\texttt{\{i,j,N| 1<=i,j<=N\}}. Similarly, the parameterized affine
function \texttt{(i->i,N)} of line~12 actually represents the closed
function \texttt{(i,N->i,N,N)}. Thus a parameterized system is nothing
but a usual SARE where all the objects share the parameter
indices. Actually the system manipulated by {\mmalfa} is the
following:

\begin{verbatim}
system matvect                                                    -- 1  
               (M : {i,j,N | 1<=i,j<=N} of real;                  -- 2  
                V : {j,N | 1<=j<=N} of real)                      -- 3  
       returns (R : {i,N | 1<=i<=N} of real);                     -- 4  
var                                                               -- 5  
  C : {i,j,N | 1<=i,j<=N} of real;                                -- 6  
let                                                               -- 7  
  C = case                                                        -- 8  
        {i,j,N | j=0} : 0.(i,j,N->);                              -- 9  
        {i,j,N | j>0} : C.(i,j,N->i,j-1,N) + M * V.(i,j,N->j,N);  -- 10 
      esac;                                                       -- 11 
  R = C.(i,N->i,N,N);                                             -- 12 
tel;                                                              -- 13 
\end{verbatim}

\section{Structures\label{structures}}

\index{structured programming}\index{program structures}\index{structures of programs}
The issue of structuring a complex algorithm into a hierarchy of SAREs
is more complex than it seems. Obviously, it is partly addressed by
the decomposition of the problem into equations: it is always possible
to break an equation into two simpler equations, using an auxilliary
variable to hold a subexpression of the initial expression. The
reverse operation, replacing a variable appearing in an expression
with its definition, is also always possible. The {\mmalfa}
environment provides assistance for both these operations, thus ensuring
that the resulting system is equivalent to the initial one\footnote{These
operations are similar to $\beta$-conversion in the
lambda-calculus.}.\index{beta conversion}

Structuring using equations, however, basically remains first-order
structuring (in the usual functional meaning), and is thus
limited. These limits appear more clearly on the following example:
suppose we want to write in {\alfa} an algorithm for an
image processing application, which involves time-varying pixel matrices.
Such matrices are represented in {\alfa} as three-dimensional data
arrays (two matrix dimensions, and one -- possibly unbounded -- time
dimension) as represented on Fig.\ref{CollectionLineaire}.
\index{time-varying matrices}
\begin{figure}[!t]
  \centerline{
    \includegraphics{figures/matcoll}
  }
\caption{Linear collection of bidimensional matrices
\label{CollectionLineaire}}
\end{figure}

Now let us try to re-use the equations of the matrix-vector product to
operate on these time-varying matrices. Obviously it is not possible
in a straightforward manner, for the inputs don't have the proper
dimension. For example we will need to rewrite the whole of the
equation defining \texttt{C} to add one dimension to the domains and
the affine functions, yielding the following equation:

\begin{verbatim}
  C = case                                                       -- 8  
        {i,j,t | j=0}: 0.(i,j,t->);                              -- 9  
        {i,j,t | j>0}: C.(i,j,t->i,j-1,t) + M * V.(i,j,t->j,t);  -- 10 
      esac;                                                      -- 11 
\end{verbatim}

Structuring with variable will never be adequate when such a
\emph{dimension extension} \index{dimension extension}is needed, which
is the usual case.  The remainder of this chapter presents the {\alfa}
program structures addressing this problem.

\section{Simple structures}

Let us try and write an {\alfa} SARE for the addition of two integers
(or fixed-point numbers) written in binary notation.  Such a binary
addition is classically described as a sequence of \emph{full adder}
operations with the propagation of a carry bit from one full adder to
the following, as described by Fig.\ref{adder}.
\index{adder}\index{binary addition}

\begin{figure}[!ht]
  \centerline{
    \includegraphics{figures/adder}
  }
\caption{Addition of integers coded in boolean, using \emph{full adders}\label{adder}}
\end{figure}

The following system is a \emph{full adder} function written in {\alfa}. 
\index{full adder}

\begin{verbatim}
      system FullAdder (A,B,Cin : boolean) 
               returns (X,Cout : boolean);
      let
        X = A xor B xor Cin;
        Cout = (A and B) or (A and Cin) or (B and Cin);
      tel;
\end{verbatim}

To build an adder system with this program, we need to be able to
express that we have a linear collection of instances of this system,
as shown in Fig.\ref{adder}. The shape of this linear collection may
be expressed as an {\alfa} domain, say \texttt{\{b|0<=b<W\}} where
\texttt{W} is a size parameter giving the number of bits.
\index{use statement}

The structure construct \textbf{use} in {\alfa} allows precisely that:
the following system describes in {\alfa} the adder given Fig.\ref{adder}: 
\index{adder}\index{binary addition}

\begin{verbatim}
     system Plus: {W|W>1} (A,B: {b| 0<=b<W} of boolean)         -- 1  
                  returns (S : {b| 0<=b<=W} of boolean);        -- 2  
     var                                                        -- 3  
       Cin, Cout, X : {b| 0<=b<W} of boolean;                   -- 4  
     let                                                        -- 5  
       use {b| 0<=b<W} FullAdder[] (A,B,Cin) returns(X, Cout);  -- 6  
       Cin[b] =                                                 -- 7  
         case                                                   -- 8  
           {| b=0} : 0[];                                       -- 9  
           {| b>0} : Cout[b-1];                                 -- 10 
         esac;                                                  -- 11 
       S[b] =                                                   -- 12 
         case                                                   -- 13 
           {| b<W} : X;                                         -- 14 
           {| b=W} : Cout[W-1];                                 -- 15 
         esac;                                                  -- 16 
     tel;                                                       -- 17 
\end{verbatim}

\index{use statement}
In this system, the line 6 reads as follows: ``Use a collection of
instances of the subsystem \texttt{FullAdder}. This collection has the
shape of the extension domain\index{extension domain} \verb!{b|0<=b<W}! 
and is thus indexed by index \texttt{b}. Let the inputs of
the \texttt{b}-th instance be the variables \texttt{A}, \texttt{B} and
\texttt{Cin} at point
\texttt{b}, and similarly let the outputs of this collection of
instances be the variables \texttt{X} and \texttt{Cout}.''
(The lines~7-11 describe the carry propagation, and lines~12-16
define the output of this binary adder.)

\index{dimension extension}
In other words, line~11 is a shortcut for the following equations,
which are those of the system \texttt{FullAdder} whith the dimension
of the variables extended from zero to one:

\begin{verbatim}
        X[b] = A[b] xor B[b] xor Cin[b];
        Cout[b] = (A[b] and B[b]) or (A[b] and Cin[b]) or (B[b] and Cin[b]);
\end{verbatim}

\section{Syntax of the \emph{use} construct}

\index{use syntax}
The \textbf{use}\ construct appears at the syntactic
level of an equation, since it is basically a shortcut for a set of
equations. Here is the general syntax of an equation/use:
{\tt
\begin{tabbing}
xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\= xxx\=  \kill
\textsl{Equation} ::\\
\>\> \textsl{Identifier} = \textsl{Expression} ;\\
\> \Alt\ \> use \Opt{ \textsl{ExtensionDomain}} \textsl{Identifier}\\
\>\>\>\>\>\> \Opt{[\textsl{ParamAssignment} ]} \\
\>\>\>\>\>\>( \textsl{ExpressionList} )\\
\>\>\> returns\>\>\> ( \textsl{IdentifierList} ) ;
\end{tabbing}
}

\index{parameter assignment}\index{assignment of the
parameters}\index{parameters}
In this syntax we see that there is an optional \emph{parameter
assignment} which is discussed in the following.
In the previous addition, the subsystem \texttt{FullAdder} has
no parameters, and the parameter assignment is therefore empty.

\section{Binary multiplication in {\alfa}}

When performed by hand, a multiplication is basically a collection
of additions, as shown by Fig.\ref{intmul}. \index{binary multiplication}

\begin{figure}[!ht]
\begin{center}
    \includegraphics{figures/intmul}
  \caption{Product of two fixed point reals in binary representation\label{intmul} }
\end{center}
\end{figure}

The following program is the {\alfa}
incarnation of Fig.\ref{intmul}. Line~8 performs the binary product of
all the bits of the first operand by each bit of the second. Line~10
describes a linear collection of additions, indexed by \texttt{m}
which is the row index in Fig.\ref{intmul}. Lines~12-17 link the
result of one additions to the input of the following.

\begin{verbatim}
    system Times: {W|W>2} (A,B: {b| 0<=b<W} of boolean)   -- 1  
                  returns (X  : {b| 0<=b<W} of boolean);  -- 2  
    var                                                   -- 3  
      P : {b,m| 0<=b,m<W } of boolean;                    -- 4  
      Si : {b,m| 0<=b<W; 0<m<W } of boolean;              -- 5  
      So : {b,m| 0<=b<=W; 0<m<W } of boolean;             -- 6  
    let                                                   -- 7  
      P[b,m] = A[b] and B[m];                             -- 8  
      use {m| 0<m<W} Plus[W] (Si,P) returns (So);         -- 9   
                                                          -- 10 
      Si[b,m] =                                           -- 12 
                                                          -- 11
        case                                              -- 13 
          {| m=1} : P[b,m-1];                             -- 14 
          {| m>1} : So[b+1,m-1];                          -- 15 
        esac;                                             -- 16 
      X[b] = So[b+1,W-1];                                 -- 17 
    tel;                                                  -- 18 
\end{verbatim}

Here the parameter assignment \texttt{W} just equates the bit size
parameter of the subsystem \texttt{Plus} and that of the
multiplication. In the general case, however, this parameter
assignment may be any affine function, which proves very
powerful. Figure~\ref{accumdraw} describes the problem of accumulating
binary numbers without overflow: the sum of two $N$-bits numbers may
be a $N+1$ bits number, therefore we need to use additions of
increasing sizes to avoid loss of bits. This is expressed in {\alfa} as:
\begin{verbatim}
         use {i| 1<=i<N} Plus[W+i-1] (A1,A2) returns (Acc);
\end{verbatim}
\index{parameter assignment}\index{assignment of the
parameters}\index{parameters}
\begin{figure}[!ht]
\centerline{
   \includegraphics{figures/Accum}
    }
\caption{Binary numbers accumulation. \label{accumdraw} }
\end{figure}

Another example where the natural structuring of an algorithm
makes use of a parameter assignment depending on the extension indices
is the Gaussian elimination given at the beginning of
chapter~\ref{static}. The reader is invited to try and understand this
program.

\index{extension domain}
Note that in all our examples, the extension domain is
monodimensional, but this is not a rule: in the general case the
extension domain may be any arbitrary {\alfa} domain. For example, in
an image processing application, a 2-dimensional extension domain may
be used to apply some function to all the points of an image.

\section{Manipulating structured programs}

A structured program is stored in {\mmalfa} as a {\mma} list of
systems called a \emph{library}. The default library is stored in the
global variable \verb!$library!.\index{$library}

A structured program may be written in one single file or several
distinct files.  In the latter case the \texttt{load[]} function
returns a library composed of all the systems contained in the file,
and stores this library in \verb!$library!. %$

If the program is stored in several files, it is the responsibility of
the user to build a proper library, i.e. a {\mma} list of all the
systems needed by the hierarchical structure of the program. For this
purpose, the user will typically use {\mma} list manipulating
functions such as \texttt{Join[], Append[]}...\index{structured
programs}

In addition, two functions, \texttt{putSystem[]} and
\texttt{getSystem[]}, may be used to extract a system from a library
and to put back a modified system in a library. Typically one of the
system is extracted from the library, modified by some program
transformation, and then put back in the library.\index{putSystem[]}\index{\getSystem[]}

\section{Program transformations associated with structures}

Most {\mmalfa} functions handle parameterized programs and \emph{use}
statements. There are, however, some major exceptions such as the
\texttt{writeC[]} translator. \index{writeC[]}In this case, {\mmalfa}
provides program transformations transforming a structured program
into a simpler equivalent one~:

\begin{itemize}

\item \texttt{assignParameterValue[]} \index{assignParameterValue[]}
gives a value to a size parameter, i.e. it refines a generic system
into a specialized one.

\item \texttt{inlineSubSystem[]} \index{inlineSubSystem[]} expands a
\emph{use} statement, replacing it with the equations of the
corresponding subsystem, properly modified to take the dimension
extension into account.\index{flattenning a structured program}

\item \texttt{inlineAll[]} \index{inlineAll[]} recursively flattens a
structured {\alfa} program.

\end{itemize}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/demos.tex
Added mmalpha (source and binary)

--- 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/appendixdemos.tex
@@
 \label{demos}
+This chapter presents a few demos of {\mmalfa}.\index{demos,demonstrations}
+There two types of demos, the ones that are to be seen in a bloc note interface and the ones that should be executed in kernel mode. There are also various documentations available with the archive.
 
-{\sc Version of \today{}}
-
-This chapter presents a few demos of
-{\mmalfa}.\index{demos,demonstrations} 
-There two types of demos: some demos
-have a notebook interface, and the others
-should be executed in kernel mode. There are also various
-documentations available with the \alfa{} archive.
-
-\section{Notebook demos}
-All these demonstrations can be accessed directly by 
-hyperlinks of the 
-so-called {\em master} notebook, which is started by
-evaluating the command \texttt{start[]}.
+\section{Bloc notes demos}
 
@@
  \item [Introduction.] Introduction to {\Alpha}
-
 \item [Domlib.] Elementary computations on domains with Domlib.
 
-\item [Fir: Design of a FIR filter.]  The FIR demo shows the typical
+ \item [FIR: Design of a FIR filter.]  The FIR demo shows the typical
 steps for the synthesis of a linear architecture for a Finite Impulse
@@
 
-\item [matvect: architecture for matrix vector
+ \item [MATVECT: architecture for matrix vector
 multiplication.] 
-
 The MATVECT demo shows the typical steps for the
@@
 
-\item[Tutorial\_intro: ]
-very introductory tutorial. Corresponds to document~\cite{first-tutorial}.
-
-\item[mma-intro: ]
-quick introduction to \mma{}. Explains how Alpha ASTs are represented
-in \mma{}. Warning: the abstract syntax presentation may not be as reliable as
-the appendix of this manual (see chapter~\ref{appendixalpha}).
-\item[Kalman: ]
-this example describes an application of \mmalfa{} to a Kalman
-filter. This application, quite complex, shows how to use 
-structured scheduling. Reference~\cite{parelec} describes the 
-background of this application. 
-\item[Kalman-Sqrt: ]
-this example describes an application of \mmalfa{} to a Kalman
-filter, using a so-called square root covariance method. 
-This application, quite complex, shows how to use 
-structured scheduling. Reference~\cite{cce} describes the 
-background of this application.
-
-\item[Neural-Network: ]
-this notebook is not ready yet, but will be soon.
-\item[Fuzzy-Logic: ]
-an application of fuzzy logic to channel equalization. 
-\end{description}
-
-Another serie of notebook allows one to become familiar with
-different packages of \mmalfa{}: 
-\begin{description}
-\item[Alpha: ] the Alpha package.
-\item[Control: ] the Control package.
-\item[Matrix: ] the Matrix package.
-\item[ChangeOfBasis: ] the ChangeOfBasis package.
-\item[Normalization: ] the Normalization package.
-\item[Cut: ] the Cut package.
-\item[Decomposition: ] the Decomposition package.
-\item[Static: ] the Static package.
-\item[PipeControl: ] the PipeControl package. 
-\item[Vhdl: ] the Vhdl package.
-\item[CheckAlpHard: ] the AlphHard package.
-\item[Meta: ] the Meta package.
-\item[Schedule: ] the various schedule packages.
-\item[dataFlowSchedule: ]
-
-\item[matlib: ]
-
-\item[Visual: ] the Visual package.
-
 \end{description}
@@
 These demos are avaiblable at the following address:
-\begin{verbatim}
-/udd/alpha/alpha\_beta/Mathematica/demos/demos.html
-\end{verbatim}
+\htmladdnormallink{file:/udd/alpha/alpha\_beta/Mathematica/demos/demos.html}
+{file:/udd/alpha/alpha\_beta/Mathematica/demos/demos.html}
 

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/install.tex
Added mmalpha (source and binary)

--- 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/appendixtechset.tex
@@
 \chapter{Installing \mmalfa{}}
+
 \label{appendixtechset}
+
 \label{install}
@@
 \newcommand{\mmalpha}{{MMalpha}}
+
 \input{install1.tex}
@@
 \section{Getting started}
+
 You may use either {\mma} under {\tt emacs}, or directly on the Unix
+
 system.
+
 \begin{itemize}
+
 \item under {\tt emacs}, type {\tt  esc-X Mathematica}.
+
 \item type math (use the {\mma} textual interface).
+
 \item type mathematica (use the {\mma} textual interface).
+
 \end{itemize}
+
 %Do not use a Notebook, as the output of the pretty printer would not
+
 %be available.
+
 When the {\mma} session is started, you can type {\tt goDemo["INIT"]}
+
 for the initiation demonstration.
@@
 When the {\mma} session is started, you can type {\tt on[]} or
+
 \texttt{start[]} for the initiation demonstration.
-
-\section{{\mmalpha} file hierarchy}
-\includegraphics[width=12cm]{figures/fileHierarchy}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/introduction.tex
Added mmalpha (source and binary)

\chapter*{Foreword}
\mmalfa{} is a free software, available under the 
\gpl{}. It can be downloaded from 
the site:\\
\sitemmalfa{}\\
If you have any problem 
in understanding this document or trying the \mmalfa{} software, please 
send an e-mail to the following address:
{\tt alpha@irisa.fr}.

Although we cannot guarantee a strong support, we shall try
our best to help solving your problem.

Many people contributed to \mmalfa{}. 
\begin{itemize}
\item Pierrick Gachet designed an early version of \alfa{}.
\item Christophe Mauras designed the first real version of \alfa{}, and
implemented the ancestor of \mmalfa{}, called {\sc Alpha du Centaur}.
\item Hervé Le Verge continued the work of Christophe Mauras, and 
worked on reduction operators. He also implemented the first version
of the polyhedral library.
\item Doran Wilde designed the first version of the parser, of the
pretty-printers, of \texttt{WriteC}, and of many early packages
of \mmalfa{}. He developped a new, more efficient version of the 
polyhedral library. 
\item Zbignew Chamsky contributed to the design of the first
\mmalfa{} packages. 
\item Florent Dupont de Dinechin implemented the structuration of 
\alfa{}, and deeply improved the \mmalfa{} environment. 
\item Tanguy Risset is the conductor of the \mmalfa{} 
entreprise. He designed the scheduler, and many other packages.
\item Patrice Quinton was the inspirer of \alfa{}, and implemented
several minor packages of \alfa{}. He also designed numerous
notebook. 
\item Sanjay Rajopadhye insuffled many ideas and designed the 
reduction expansion package. 
\item Patricia Le Moenner implemented the Vhdl generator.
\item Fabien Quilleré is currently implementing a new C generator.
\end{itemize}

We also thank the EEC, who sponsored the development of \alfa{}
through several grants. The late C$^3$ research cooperation 
network also helped initial research on \alfa{}.

\chapter{Introduction}

\label{chap1}
In this document, we introduce the basics of {\alfa} and we explain
how to use {\mmalfa}.

%What is Alpha
\section{What is \Alpha?}
{\alfa} is a language for the specification and the derivation of
systolic algorithms and architectures, and more generally, for the
specification and design of parallel algorithms and architectures based on
the formalism called the {\em polyhedral model}.

\label{Whatis}
\nocite{Alpha97a,Alpha97b}

% History
{\Alpha} is a functional data parallel language invented in the {\api}
research group in the {\irisa} laboratory in Rennes. The project was
headed by Patrice Quinton and the first definition of the language was
proposed by Mauras in~\cite{Mauras89} in 1989. The original motivation
was to provide a language for expressing algorithms in an extended
version of the formalism of recurrence equations proposed by Karp,
Miller and Winograd~\cite{KaMiWi67}. The goal of this language was to
provide a high level tool for the synthesis of parallel regular
{\vlsi} architectures.

Although {\Alpha} stands for the language itself, it is often also
associated with the environment in which it is currently developed:
{\bf {\mmalfa{}}}. {\mmalfa{}} is an interface based on the {\mma{}}
software from which one can manipulate {\Alpha} programs.

%Motivations of {\alfa}

\section{What is {\Alpha} for?}
\label{whatfor}
{\Alpha} is currently a research tool which provides research problems
in different computer science areas fields like: functional language semantics,
parallelization, code generation, optimization, polyhedral theory,
{\vlsi} synthesis, systolic arrays, etc.

One important long term goal is to promote the use of high level
functional languages for the synthesis of parallel {\vlsi}
architectures.

%What can we do with it?

\section{What can you expect from \alfa{}?}
From the short term point of view, \alfa{} can be
useful for:
\begin{itemize}
\item Providing a correct recurrence equation specification for a particular 
algorithm.

\item Simulating such a specification.

\item Transforming and simplifying a recurrence equation specification.

\item Computing on convex polyhedra.

\item Scheduling and detecting parallelism.

\end{itemize}

If you are ready to invest a little more time on {\mmalfa{}}, you will 
probably be able to 
\begin{itemize}
\item Describe a systolic architecture with {\AlpHard} (see
section~\ref{alphard}).
\item Generate {\vhdl} from this description.
\item Provide a path from high level functional specification of
an algorithm to the layout description of a {\vlsi} algorithm
realizing it.

\end{itemize}

If you becomed hooked to \alfa{}, you will be able to take advantage
of a {\em completely open framework} in which you will be 
able to design your own tools, or interface your own programs: 
the source code of all functions described in this manual is
available.

% Principles

\section{How does \mmalfa{}{} work?}

{\mmalfa{}} is written in C (for a small part) and in
\mma{} (for a larger part).

The user should only see the \mma{} interface.
\mma{} provides an interpreted language with high level built-in
function for symbolic computations:  {\mmalfa{}} uses these facilities for
transforming {\Alpha} programs. \mma{} has also a nice programming
language in which one can do general computation as in any other
programming language.

The basic principle of the {\mmalfa{}} environment is the following one:
\mma{} stores an internal representation of an {\Alpha} program
(called Abstract Syntax Tree or AST) and perform computations on this
internal representation via user's commands. These commands can be for
example: viewing the {\Alpha} program, checking its correctness,
adding a variable, generating C code to simulate it, generate a
systolic array, generate \vhdl{} code, etc. All the computations
happen on the internal AST which is stored in the \mma{} environment.

Specific modules in C are used for two purposes: various parsers and
unparsers and computations on polyhedra. All the C functions are
accessed via \mma{}, but they can also be called from a C program:
this is particulary useful for the polyhedral library.  The polyhedral
library, written by H. Le Verge and D. Wilde (see
appendix~\ref{polylib}) is the computationnal kernel of the
{\mmalfa{}} environment. Its use provides many transformations that
cannot be done easily by hand.

%Describe the document

\section{What is in this document}
The organization of this document is as follows. 
Chapter~\ref{chapstart} is a brief overview of the 
\mmalfa{} environment.

In chapters~\ref{chapalphascal} and \ref{chapalphaarray}, we introduce
{\alfa} following an architecture-oriented approach. In other words,
we describe {\alfa} programs whose interpretation is always an
architecture.

Chapter~\ref{chapderiving} illustrates the
derivation of systolic architectures from a high-level description.

In chapter~\ref{chapfunctional}, we introduce {\alfa} as a functional 
data parallel programming language. This point of view is complementary
to that of the first two chapters, in that it show how to use {\alfa}
as an algorithmic language.

Chapter~\ref{chapssubsystems} deals with structured and parameterised
{\alfa}, which gives the full power of the language, both for the 
description of hardware and the specification of algorithms.

Chapters~\ref{chapanalyze} and~\ref{chapcodegen} are more ``practical''
and tool-oriented.  In chapter~\ref{chapanalyze}, we describe the
static analyzer of {\mmalfa}\,: it is a useful tool to check some
properties of an {\alfa} program.  Chapter~\ref{chapcodegen} explaines
how to generate code and simulate {\alfa} programs.

Chapter~\ref{chapscheduling} is devoted to scheduling {\alfa}
programs. Scheduling is one of the essential steps of the derivation
of architectures as well as compiling {\alfa} to sequential and
parallel machines.

Chapter~\ref{chapalfhard} (still under heavy work...) is
intended to explain how \AlpHard{} is defined
and how to 
generate \vhdl{} code frome \alfa{}.

%A roadmap for the tutorial.

\section{How to read this tutorial?}
Depending on your interest in \mmalfa{}, you will be interested
in particular chapters of this tutorial.

First, read chapter~\ref{chapstart}: it does not hurt... and 
does not take too much time. By the way, there is a separate, 
slightly extended version of this chapter available in document
~\cite{gettingstarted}.

Chapter~\ref{chapderiving} will be worth reading, if you are
interested in systolic array design.

You may skip chapters~\ref{chapalphascal} and \ref{chapalphaarray}
if you are not immediately interested in deeply understanding 
the mechanisms of \alfa{}.

If you are interested in writing a "real" \alfa{} program, 
you will have to read chapters~\ref{chapssubsystems}, 
\ref{chapanalyze}, \ref{chapcodegen}, and~\ref{chapscheduling}.

Finally, if you want to produce "real hardware", read chapter
~\ref{chapalfhard}.

\section{Other documents}
This tutorial is not the only information available on 
\alfa{} and \mmalfa{}. There are papers, for the academic 
oriented people. There are also many demonstration notebooks. 
To access them, start \mma{}, type the command \texttt{start[]}, 
and enjoy\footnote{or dislike...}.

\section{What we plan in the (near?) future}
The current tutorial is not yet complete. Several aspects of 
{\alfa} and {\mmalfa} (already implemented) need to be described
in detail. Among them, the generation of {\sc Vhdl}, the subset
of {\alfa} called {\alfhard}, and the functions in {\mmalfa} 
which allow one to manipulate convex domains and {\alfa} 
programs. Knowing how long it takes to write and check a 
documentation, we have chosen to release a \mmalfa{} version 
without a complete documentation, for those who are interested
in working with us.

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/rest-chap1.tex
Added mmalpha (source and binary)

\chapter{Modelling synchronous architectures}
\label{chap2}
In this chapter, we present the {\alfa} language through the 
description of very simple -- and somehow artificial -- examples
of synchronous circuits. Our goal is to show the basic constructors
of the language: their syntax, their meaning, and how they can be
used in practice. 
% 
\section{Pointwise operators\index{pointwise operators}}
\begin{figure}[htbp]
\htmlimage{scale=1}
\colorbox{red}{
\centerline{\input{/udd/quinton/documents/lncs_574/seminaire_litp/figures/exemple_1}}}
\caption{An adder}\label{adder}
\end{figure}
Consider the architecture\index{architecture} shown in figure~\ref{adder}: there is a 
simple adder, which takes to input flows, respectively {\tt x} and 
{\tt y}, and returns {\tt S}.

This architecture is described by the {\alfa} program of figure \ref{adder-alpha}.
\begin{figure}[htbp]
\htmlimage{scale=1}
%\input{adder.alpha}
\begin{verbatim}
system exemple1 (x,y: {t|1<=t} of integer)     -- 1
returns (z: {t|1<=t} of integer);              -- 2
var S: {t|1<=t} of integer;                    -- 3
let                                            -- 4
  S = x + y;                                   -- 5
  z = S;                                       -- 6
tel;                                           -- 7
\end{verbatim}
\caption{An adder in {\alfa}}\label{adder-alpha}
\end{figure}
This program is called a {\em system}. Line 1 is the head of the system, 
with name {\tt exemple1}. This system has two input arguments named
{\tt x} and {\tt y}. These arguments, called {\em variables}, are
defined on the set {\tt \{t|1<=t\}}. Line 3 defines a local variable
{\tt S}. Between the {\tt let} and {\tt tel} keywords of lines 4 and
7, we have two equations: line 5 says that {\tt S} is the pointwise sum of 
{\tt x} and {\tt y}. In other words, $\forall t, S_t = x_t + y_t$. 
Line 6 says that {\tt z} and {\tt S} are identical.
\subsection*{Processing this example in {\mmalfa}}
A few words on {\mma}. A more complete introduction is given in 
appendix~\ref{mma-introduction}
\begin{verbatim}
load["adder.alpha"]    (* loads the alpha file *)
ashow[];               (* shows its content *)
\end{verbatim}

%
\section{Substitution}
\begin{figure}[htbp]
\htmlimage{scale=1}
%\input{adderMultiplexerSubs.alpha}
\begin{verbatim}
system adderMultiplexerSubs (x,y : {t|1<=t} of integer)
       returns (z : {t|5<=t} of integer);
  var
    S : {t|1<=t} of integer;
  let
    S = case
        {t|t=1} : 0.(t->);
        {t|t>1} : x.(t->t-1) + y.(t->t-1);
        esac;
    z = {t|t>=5} : case
                   {t|t=1} : 0.(t->);
                   {t|t>1} : x.(t->t-1) + y.(t->t-1);
                   esac;
  tel;
\end{verbatim}
\caption{Substitution}\label{adderMultiplexerSubs-alpha}
\end{figure}
\subsection*{Exercices in {\mma}}

%
\section{Normalization}
\begin{figure}[htbp]
\htmlimage{scale=1}
\centerline{\input{/udd/quinton/documents/lncs_574/seminaire_litp/figures/exemple_7}}
\caption{Normalizing}\label{adderMultiplexerNorm}
\end{figure}
The corresponding program is shown in figure \ref{adderMultiplexerNorm-alpha}.
\begin{figure}[htbp]
\htmlimage{scale=1}
%\input{adderMultiplexerNorm.alpha}
\begin{verbatim}
system adderMultiplexerNorm (x,y : {t|t>=1} of integer)
       returns (z : {t|t>=5} of integer);
  var
    S : {t|t>=1} of integer;
  let
    S = case
        {t|1=t} : 0.(t->);
        {t|t>1} : x.(t->t-1) + y.(t->t-1);
        esac;
    z = {t|t>=5} : x.(t->t-1) + y.(t->t-1);
  tel;
\end{verbatim}
\caption{Normalization}\label{adderMultiplexerNorm-alpha}
\end{figure}
\subsection*{Exercices in {\mma}}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Tutorial/vhdl.tex
Added mmalpha (source and binary)

\section{Translating to \vhdl{}}

\alfa{} programs cannot be translated directly to

\vhdl{} (that would be to easy!).

Only a subset of \alfa{}, called \alphard{}, can be

translated.

The \texttt{alphaToVHDL} command, or the \texttt{av} command which is

similar, take an \alphard{} program, and translate it into \vhdl{}.

Normally, \alphard{} programs should be produced using the synthesis

process, which ends up using the \texttt{AlphaToHard} function:

if this is the case, the synthesized program should be correct by

construction, and the \texttt{alphaToVHDL} function provides a

correct \vhdl{} model too. However, it is possible to

write directly \alphard{} programs, and one can check such

programs, using various \texttt{check} functions (see~\ref{check}).

However, it is interesting to look in detail into the form of the

\alphard{} language. This it what we do here, by examining how various

pieces of hardware are represented in \alphard{}, and translated in

\vhdl{}.

*****************

\texttt{alphaToVHDL[]} generates \vhdl{} files from the library

\texttt{$library}. Initial

activity time of the corresponding architecture is 0.

\texttt{alphaToVHDL[tinit]}

generates \vhdl{} files from \texttt{$library}

with initial activity time \texttt{tinit}.

\texttt{alphaToVHDL[lib, tinit]} generates \vdhl{}

files from the library (list of

systems) or the single system \texttt{lib}.

Actually, it is equivalent to use the

\texttt{av} command, which is shorter.

Note the following. First, both commands translate all programs loaded in

\texttt{$library}. This means that, if you modify the current program in

\texttt{$result}, \texttt{$library}

will be changed only if you put

\texttt{$result} back in \texttt{$library}

(see the \texttt{putSystem} command.)

Second, the \vhdl{} files which are generated have the name

\texttt{pg.vhd}

where

\texttt{pg}

is the actual system name of the loaded program (not to be confused with

the name of the file containing it.)

\subsection{Options}

\texttt{alphaToVHDL} and \texttt{av} have options.

\begin{itemize}

\item \texttt{debug->True}

provide a trace of execution of the translater. It may be helpful to

detect which part of the program was not accepted by the translater.

\item \texttt{verbose->True}

gives a little bit more information.

\item \texttt{tempFile -> False}

puts the \vhdl{} files in the /tmp/ directory\footnote{I do not

know where this temporary directory is in Win NT.}

\end{itemize}

\section{How \alphard{} program are written}

Combinatorial logic and register

The following program is a cell containing an adder. As all cells, it is

parameterized by the itinial time, named

\texttt{Tinit}

and the duration of its use, named

\texttt{Duration}.

The domain of these parameters is not important for combinatorial logic,

but uncorrect values may cause the

\texttt{analyze}

command to fail on such a program.

Here is the program:

\begin{verbatim}

system addreg: {Tinit, Duration | Tinit>0; Duration>1}

(x,y : {t|Tinit<= t<= Tinit+Duration-1} of integer)

returns (z: {t|Tinit+1<= t<= Tinit+Duration-1} of integer);

var X, Y : {t|Tinit+1<= t<= Tinit+Duration-1} of integer;

let

	X[t] = x[t-1];

	Y[t] = y[t-1];

	z[t] = X + Y;

tel;

\end{verbatim}

We now call the translator:

\begin{verbatim}

av[verbose -> True, tempFile -> True]

\end{verbatim}

and here is the generated \vhdl{} code:

\begin{verbatim}

-- VHDL Model Created for "system addreg"

 --  27/4/1999 18:11:54

library IEEE;

use IEEE.std_logic_1164.all;

library COMPASS_LIB;

 use COMPASS_LIB.STDCOMP.all;

library COMPASS_LIB;

 use COMPASS_LIB.COMPASS.all;

entity addreg is

      Port ( Ck : In std_logic;

      x : In Integer range 0 to 255;

      y : In Integer range 0 to 255;

      z : Out Integer range 0 to 255 );

end addreg;

architecture Behavioral of addreg is

  signal X : Integer range 0 to 255;

  signal Y : Integer range 0 to 255;

begin

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        X <= x;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        Y <= y;

    end if;

  end process;

  z <= (X + Y);

end Behavioral;

\end{verbatim}

\section{Cell 1 of the FIR}

This program is a cell of the FIR. It consists of combinational logic

and registers.

\begin{verbatim}

-- Example for VHDL translator -- September 1998

-- This example is a cell extracted from the FIR module

system cellFirModule3 :{p,N,M | 2<=p<=N; N<=M-1}

  (Y_reg7loc : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of integer;

   xPipe_reg5loc : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of integer;

   HPipeES_reg4loc : {t | 2p-N-2<=t<=p-1; 2<=p<=N; N<=M-1} of integer;

   xPipeES_reg3loc : {t | 2p-N-1<=t<=p; 2<=p<=N; N<=M-1} of integer;

   xPipe_ctl2P_reg2loc : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of boolean;

   HPipe_ctl1P_reg1loc : {t | p-1<=t<=p-N+M-1; 2<=p<=N; N<=M-1} of boolean

  )

returns

  (Y : {t | p+2<=t<=p-N+M+2; 2<=p<=N; N<=M-1} of integer;

   xPipe : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of integer;

   HPipeES : {t | 2p-N<=t<=p; 2<=p; N<=M-1} of integer;

   xPipeES : {t | 2p-N+1<=t<=p+1; 2<=p; N<=M-1} of integer;

   xPipe_ctl2P : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of boolean;

   HPipe_ctl1P : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of boolean

  );

var

  HPipe_ctl1Ploc6 : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of boolean;

  xPipe_ctl2Ploc5 : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of boolean;

  xPipeESloc4 : {t | 2p-N+1<=t<=p+1; 2<=p; N<=M-1} of integer;

  HPipeESloc3 : {t | 2p-N<=t<=p; 2<=p; N<=M-1} of integer;

  xPipeloc2 : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of integer;

  HPipe : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of integer;

  HPipe_ctl1 : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of boolean;

  HPipe_ctl1P_reg1 : {t | p<=t<=p-N+M; 2<=p<=N; N<=M-1} of boolean;

  HPipeES_reg4 : {t | 2p-N<=t<=p; 2<=p; N<=M-1} of integer;

  HPipe_reg6 : {t | p+1<=t<=p-N+M; 2<=p<=N} of integer;

  HPipe_reg8 : {t | p+2<=t<=p-N+M+2; 2<=p<=N; N<=M-1} of integer;

  xPipe_ctl2 : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of boolean;

  xPipe_ctl2P_reg2 : {t | p+1<=t<=p-N+M+1; 2<=p<=N; N<=M-1} of boolean;

  xPipeES_reg3 : {t | 2p-N+1<=t<=p+1; 2<=p; N<=M-1} of integer;

  xPipe_reg5 : {t | p+2<=t<=p-N+M+1; 2<=p<=N} of integer;

  xPipe_reg9 : {t | p+2<=t<=p-N+M+2; 2<=p<=N; N<=M-1} of integer;

  Y_reg7 : {t | p+2<=t<=p-N+M+2; 2<=p<=N; N<=M-1} of integer;

let

  HPipe_ctl1P[t] = HPipe_ctl1Ploc6[t];

  xPipe_ctl2P[t] = xPipe_ctl2Ploc5[t];

  xPipeES[t] = xPipeESloc4[t];

  HPipeES[t] = HPipeESloc3[t];

  xPipe[t] = xPipeloc2[t];

  Y_reg7[t] = Y_reg7loc[t-1];

  xPipe_reg5[t] = xPipe_reg5loc[t-2];

  HPipeES_reg4[t] = HPipeES_reg4loc[t-2];

  xPipeES_reg3[t] = xPipeES_reg3loc[t-2];

  xPipe_ctl2P_reg2[t] = xPipe_ctl2P_reg2loc[t-1];

  HPipe_ctl1P_reg1[t] = HPipe_ctl1P_reg1loc[t-1];

  xPipe_reg9[t] = xPipeloc2[t-1];

  HPipe_reg8[t] = HPipe[t-2];

  HPipe_reg6[t] = HPipe[t-1];

  xPipeESloc4[t] = xPipeES_reg3[t];

  HPipeESloc3[t] = HPipeES_reg4[t];

  xPipe_ctl2Ploc5[t] = xPipe_ctl2P_reg2[t];

  xPipe_ctl2[t] = xPipe_ctl2Ploc5[t];

  xPipeloc2[t] =

    case

      {| t=p+1; 2<=p<=N; N<=M-1} :

         if (xPipe_ctl2[t]) then xPipeESloc4[t] else 0[];

      {| p+2<=t<=p-N+M+1; 2<=p<=N} :

         if (xPipe_ctl2[t]) then 0[] else xPipe_reg5[t];

    esac;

  HPipe_ctl1Ploc6[t] = HPipe_ctl1P_reg1[t];

  HPipe_ctl1[t] = HPipe_ctl1Ploc6[t];

  HPipe[t] =

    case

      {| t=p; 2<=p<=N; N<=M-1} :

         if (HPipe_ctl1[t]) then HPipeESloc3[t] else 0[];

      {| p+1<=t<=p-N+M; 2<=p<=N} :

         if (HPipe_ctl1[t]) then 0[] else HPipe_reg6[t];

      esac;

  Y[t] = Y_reg7[t] + HPipe_reg8[t] * xPipe_reg9[t];

tel;

\end{verbatim}

And here is the generated \vhdl{} program:

\begin{verbatim}

-- VHDL Model Created for "system cellFirModule3"

 --  27/4/1999 19:14:50

library IEEE;

use IEEE.std_logic_1164.all;

library COMPASS_LIB;

 use COMPASS_LIB.STDCOMP.all;

library COMPASS_LIB;

 use COMPASS_LIB.COMPASS.all;

entity cellFirModule3 is

      Port ( Ck : In std_logic;

      Y_reg7loc : In Integer range 0 to 255;

      xPipe_reg5loc : In Integer range 0 to 255;

      HPipeES_reg4loc : In Integer range 0 to 255;

      xPipeES_reg3loc : In Integer range 0 to 255;

      xPipe_ctl2P_reg2loc : In std_logic;

      HPipe_ctl1P_reg1loc : In std_logic;

      Y : Out Integer range 0 to 255;

      xPipe : Out Integer range 0 to 255;

      HPipeES : Out Integer range 0 to 255;

      xPipeES : Out Integer range 0 to 255;

      xPipe_ctl2P : Out std_logic;

      HPipe_ctl1P : Out std_logic );

end cellFirModule3;

architecture Behavioral of cellFirModule3 is

  signal HPipe_ctl1Ploc6 : std_logic;

  signal xPipe_ctl2Ploc5 : std_logic;

  signal xPipeESloc4 : Integer range 0 to 255;

  signal HPipeESloc3 : Integer range 0 to 255;

  signal xPipeloc2 : Integer range 0 to 255;

  signal HPipe : Integer range 0 to 255;

  signal HPipe_ctl1 : std_logic;

  signal HPipe_ctl1P_reg1 : std_logic;

  signal HPipeES_reg4 : Integer range 0 to 255;

  signal HPipe_reg6 : Integer range 0 to 255;

  signal HPipe_reg8 : Integer range 0 to 255;

  signal xPipe_ctl2 : std_logic;

  signal xPipe_ctl2P_reg2 : std_logic;

  signal xPipeES_reg3 : Integer range 0 to 255;

  signal xPipe_reg5 : Integer range 0 to 255;

  signal xPipe_reg9 : Integer range 0 to 255;

  signal Y_reg7 : Integer range 0 to 255;

  signal xPipe_reg5_S1 : Integer range 0 to 255;

  signal HPipeES_reg4_S1 : Integer range 0 to 255;

  signal xPipeES_reg3_S1 : Integer range 0 to 255;

  signal HPipe_reg8_S1 : Integer range 0 to 255;

begin

  HPipe_ctl1P <= HPipe_ctl1Ploc6;

  xPipe_ctl2P <= xPipe_ctl2Ploc5;

  xPipeES <= xPipeESloc4;

  HPipeES <= HPipeESloc3;

  xPipe <= xPipeloc2;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        Y_reg7 <= Y_reg7loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipe_reg5_S1 <= xPipe_reg5loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipe_reg5 <= xPipe_reg5_S1;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipeES_reg4_S1 <= HPipeES_reg4loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipeES_reg4 <= HPipeES_reg4_S1;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipeES_reg3_S1 <= xPipeES_reg3loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipeES_reg3 <= xPipeES_reg3_S1;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipe_ctl2P_reg2 <= xPipe_ctl2P_reg2loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipe_ctl1P_reg1 <= HPipe_ctl1P_reg1loc;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        xPipe_reg9 <= xPipeloc2;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipe_reg8_S1 <= HPipe;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipe_reg8 <= HPipe_reg8_S1;

    end if;

  end process;

  process(ck)

  begin

    if (ck='1' AND ck'event) then

        HPipe_reg6 <= HPipe;

    end if;

  end process;

  xPipeESloc4 <= xPipeES_reg3;

  HPipeESloc3 <= HPipeES_reg4;

  xPipe_ctl2Ploc5 <= xPipe_ctl2P_reg2;

  xPipe_ctl2 <= xPipe_ctl2Ploc5;

  xPipeloc2 <=

    xPipeESloc4 when xPipe_ctl2 = '1'  else

    0 when xPipe_ctl2 = '1'  else xPipe_reg5;

  HPipe_ctl1Ploc6 <= HPipe_ctl1P_reg1;

  HPipe_ctl1 <= HPipe_ctl1Ploc6;

  HPipe <=

    HPipeESloc3 when HPipe_ctl1 = '1'  else

    0 when HPipe_ctl1 = '1'  else HPipe_reg6;

  Y <= (Y_reg7 + (HPipe_reg8 * xPipe_reg9));

end Behavioral;

\end{verbatim}

\section{Control cell of the FIR}

Here is the controller for the FIR program.

\begin{verbatim}

system controlFirModule :{N,M | 2<=N<=M-1}

  ( )

returns

  (HPipe_ctl1P_Init : {t | t=1; 2<=N<=M-1} |

                      {t | 2<=t<=-N+M+1; 2<=N} of boolean;

   xPipe_ctl2P_Init : {t | t=2; 2<=N<=M-1} |

                      {t | 3<=t<=-N+M+2; 2<=N} of boolean);

let

  HPipe_ctl1P_Init[t] =

      case

        {| t=1; 2<=N<=M-1} : True[];

        {| 2<=t<=-N+M+1; 2<=N} : False[];

      esac;

  xPipe_ctl2P_Init[t] =

      case

        {| t=2; 2<=N<=M-1} : True[];

        {| 3<=t<=-N+M+2; 2<=N} : False[];

      esac;

tel;

\end{verbatim}

This program is parameterized, and therefore, a \vhdl{}

program cannot be generated without assigning values to

the parameters. This is done by means of the following

commands:

\begin{verbatim}

load["controlFirModule.alpha"];

assignParameterValue["N", 5];

assignParameterValue["M", 10];

av[tempFile -> True]

\end{verbatim}

The \vhdl{} program for this controller is presented

in appendix~\ref{vdhl:appendix}.

\section*{Appendix: \vhdl{} code of the Fir controller}

\label{vhdl:appendix}

\begin{verbatim}

-- VHDL Model Created for "system controlFirModule"

 --  27/4/1999 18:52:17

library IEEE;

use IEEE.std_logic_1164.all;

entity controlFirModule is

      Port ( Ck : In std_logic;

      Rst : In std_logic;

      HPipe_ctl1P_Init : Out std_logic;

      xPipe_ctl2P_Init : Out std_logic );

end controlFirModule;

architecture state_machine of controlFirModule is

  signal cpt: Integer;

  type HPipe_ctl1P_InitStates is (E0,E0bis,E1,E2);

  signal currentHPipe_ctl1P_InitState, nextHPipe_ctl1P_InitState :HPipe_ctl1P_InitStates;

  type xPipe_ctl2P_InitStates is (E0,E0bis,E1,E2);

  signal currentxPipe_ctl2P_InitState, nextxPipe_ctl2P_InitState :xPipe_ctl2P_InitStates;

BEGIN

reset_sm : PROCESS

 begin

-- compass stateMachine adj currentHPipe_ctl1P_InitState

-- compass stateMachine adj currentxPipe_ctl2P_InitState

  WAIT UNTIL (Ck = '1' AND Ck'event);

  IF Rst ='1' THEN

   cpt <= -1;

   currentHPipe_ctl1P_InitState <= E0;

   currentxPipe_ctl2P_InitState <= E0;

  ELSE

   cpt <= cpt + 1;

   currentHPipe_ctl1P_InitState <= nextHPipe_ctl1P_InitState;

   currentxPipe_ctl2P_InitState <= nextxPipe_ctl2P_InitState;

  END IF;

 END PROCESS;

evolution_HPipe_ctl1P_Init : PROCESS(cpt, currentHPipe_ctl1P_InitState)

 begin

 CASE currentHPipe_ctl1P_InitState IS

  WHEN E0 => IF (cpt < 0) THEN nextHPipe_ctl1P_InitState <= E0;

    ELSIF (cpt = 0) THEN nextHPipe_ctl1P_InitState <= E1; END IF;

  WHEN E1 =>  IF (cpt = 1) THEN nextHPipe_ctl1P_InitState <= E2; END IF;

  WHEN E2 => IF ((cpt >= 2) AND (cpt < 6 )) THEN nextHPipe_ctl1P_InitState <= E2;  END IF;

   IF (cpt = 6) THEN nextHPipe_ctl1P_InitState <= E0bis; END IF;-- remise a zero de la SM

  WHEN OTHERS => -- erreurs et hors service

   nextHPipe_ctl1P_InitState <= E0bis ;

 END CASE;

END PROCESS;

output_HPipe_ctl1P_Init : PROCESS(currentHPipe_ctl1P_InitState)

 begin

 CASE currentHPipe_ctl1P_InitState IS

  WHEN E1=>HPipe_ctl1P_Init <= '1';

  WHEN E2=>HPipe_ctl1P_Init <= '0';

  WHEN OTHERS =>

   HPipe_ctl1P_Init <= 'X';

 END CASE;

END PROCESS;

evolution_xPipe_ctl2P_Init : PROCESS(cpt, currentxPipe_ctl2P_InitState)

 begin

 CASE currentxPipe_ctl2P_InitState IS

  WHEN E0 => IF (cpt < 1) THEN nextxPipe_ctl2P_InitState <= E0;

    ELSIF (cpt = 1) THEN nextxPipe_ctl2P_InitState <= E1; END IF;

  WHEN E1 =>  IF (cpt = 2) THEN nextxPipe_ctl2P_InitState <= E2; END IF;

  WHEN E2 => IF ((cpt >= 3) AND (cpt < 7 )) THEN nextxPipe_ctl2P_InitState <= E2;  END IF;

   IF (cpt = 7) THEN nextxPipe_ctl2P_InitState <= E0bis; END IF;-- remise a zero de la SM

  WHEN OTHERS => -- erreurs et hors service

   nextxPipe_ctl2P_InitState <= E0bis ;

 END CASE;

END PROCESS;

output_xPipe_ctl2P_Init : PROCESS(currentxPipe_ctl2P_InitState)

 begin

 CASE currentxPipe_ctl2P_InitState IS

  WHEN E1=>xPipe_ctl2P_Init <= '1';

  WHEN E2=>xPipe_ctl2P_Init <= '0';

  WHEN OTHERS =>

   xPipe_ctl2P_Init <= 'X';

 END CASE;

END PROCESS;

END state_machine;

\end{verbatim}

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/maintenance/Status.tex
Added mmalpha (source and binary)

\documentclass[12pt]{article}
\begin{document}

\newcommand{\mmalfa}{{\sc MMAlpha}}
\newcommand{\mma}{{\sc Mathematica}}
\title{Current status of \mmalfa\\
Last change: \today}
\author{Patrice Quinton}
\date{}
\maketitle

\section{Introduction}
This document is where I put things I have to remember about the current status of 
\mmalfa{}. The document is available directly in the main directory of the
distribution.

\section{A day to day memo}
A * before a date means that the correction was committed. A \texttt{?} means that I
am not sure that it was committed. A \texttt{-} means that it has to be done. A \texttt{!} 
says that I should commit the correction.

\begin{description}
\item[18/1/2009:] 
\item[14/1/2009:] corrections à faire.
\begin{itemize}
\item tailles des données: 2 bits en entrée, idem pour les rom, résultats sur 4 bits à la sortie des mult, 
6 bits pour ovsf, une ligne pour compléter à 16 bits
\item regarder si on met des signed ou des std vectors
\item emetteur : ds le composant moduloaddress, erreur sur le type de port (std logic vectors)
\item taille des compteurts de periodic est à revoir (periodic enable4)
pour les stdlogic vector, les valeurs son entre ""
\item enlever size dans la liste de sensibilité de periodicenable
\item trop de ports dans les rom? est-ce qu'on fait un registre
\item dans la rom, changer la limite de la decl de addr
\item enlever size dans moduloaddress de la liste de sensibilité
\item remplacer values
\end{itemize}
\item[24/05/2010:] creating a new version MMALPHA-2-1-0. To be checked:
\begin{itemize}
\item In demos, check that there are not identifiers with identical names. Change analyze accordingly.
\item Check the init.m file in the config file.
\end{itemize}
\item[10/1/2009:] change fixParameter, so that the parameters are changed, depending on their
use... Has to be modified. Actually, it would be better to generate Vhdl using rules
for the parameters, and to generate on the fly as many elements than there are 
different calls.
\item[9/1/2009:] Made a few modifications in the Vertex Scheduler in order to be adapted
to the new version where coefficients are MMA variables. Made a correction to a bug in 
Simplex: if a variable of a constraint does not appear in the list of the variables of Minimize, then
the MMA function never stops.
\item[*6/1/2009:] New version released and installed. Waiting for an answer of Arpit. 
\item[4/1/2009:] Modification of the Web site. New version almost ready. Cleaning of some 
notebooks before installation. Created a test \texttt{TestSynthesis} and added it to the
distribution.
\item[2/1/2009:] Trying to check a new version. Tested it on Linux (machine sabre at IRISA). Succeeded
in compliling Polylib, but not domlib, as the MMA version is too old. Has to recover a previous 
version of the Domlib to make the test. 
\item[1/1/2009:] I made again changes in the tests, adding the defiition of a package
for each one of these, otherwise, global symbols are created when reading the file.
Has to change the documentation accordingly. In \texttt{TestMatrix.m}, there are
some tests (by inclusion of files) which were left as they were, and have to be 
included.
\item[31/12/2008:] I made major modifications and committed almost everything (except 
allocate.m and PipeControl-new.m which are not read. See 19/10/2008). I prepared a new version 
\texttt{mmalphaV2-0-1}. New tests were built (test1, 2, 3 and 4), but they are to 
be changed in order to introduce symbols in packages, and not in 
Global. I also made a correction to \texttt{dom2mma} in order to create the 
symbols that are the translation in \mma{} of those of the domain, in a work space
called \texttt{Alpha`Work`}. This had an effect on many packages. I also had to 
replace \texttt{Cut.m} by \texttt{CutMMA.m}, as there is a conflict with a function
in the \texttt{Combinatorica} package.
\item[*12/12/2008:] modification to Options.m (vhdlDir option usage was made more
precise).
\item[*12/12/2008:] modifications done in Vhdl2.m. Added an option vhdlPatterns to 
a2v, when true, Rom and a bunch of components are not generated. Instead, 
patterns are inserted in the file, and they are filled afterwards. Modifications were done
also to autoload.m, to VhdlCell.sem, and Synthesis.m.
\item[*4/12/2008:] started to implement the vhdl generator for periodic systems. In Alpha.m, 
there is a special version of load, which reads templates (suffixes are vhdx). In Vhdl2.m, 
created a function genVhdl that allows vhdl files to be created from templates. The templates
are currently in \$MMALPHA/VHDL. There are functions for ROM, for Fsm and for Periodic
enable signals. A documentation has to be provided. A function synPeriodic was included
in the \texttt{Synthesis.m} package. 
\item[*30/11/2008:] controlled a few traces in Alphard.m and ToAlpha0v2.m. Not commited yet.
\item[*30/11/2008:] made a small modification in vhdlCell.sem and vhdlModule.m, as there
was an extra message when the mute option was set. Not committed yet.
\item[*30/11/2008:] In Control.m, there was a lot of uncontrolled traces, set when I wanted
to make modifications to MakeSTCase. I controlled the traces. 
\item[*22/10/2008:] remove the definitions and work in the input of the files. Add a 
return.
\item[*21/10/2008:] tried the palindrome. Made modifications to ToAlpha0v2, but not sure
that it is necessary. 
\item[19/10/2008:] I made modifications to PipeControl in order to add the automatic
detection of pipeIO in pipeIO. I also rewrote part of the generation of pipeall commands, in 
pipeInfo. I added a few new functions to get information on polyhedra, regarding space
and time matter. The modification seems to work (on all 1D examples). Still a couple of 
extra prints to remove. I did not yet tested the 2D cases (matrix multiplication). I did not 
commit all this stuff. These modifs are currently in PipeControl-new.m and the previous
stuff is in PipeControl.m. Has to be checked.

\item[08/10/2008:] it would be good to separate the type declarations (package des types) 
in order to get the testbench separately.
\item[*07/10/2008:] there was a mistake in the controller generator, as the initial time
was shifted by the mintime, but not all instants of time. I made the correction (now, things start at 
mintime, until truetime, then we change to false time). Has to commit stuff. Added the
printing of the fsm in a2v when verbose is true. Only package to commit : vhdlCont.sem.
\item[*04/10/2008:] Tried to recompile \texttt{Write\_Alpha} on my laptop, and did not succeed, 
as the gcc loader has something wrong... I thought that it 
was probably related to the new version of MacOS. Was right, 
updated XCode, and it worked, except for read\_Alpha. Actually, there was 3 problems. First, 
there are some warnings in the compilation of yacc.y in some directories, because functions
strcat, strcopy and strlen seem to be used in a wrong way, but I believe that it does nothing more
than giving a warning message. There was a syntax error in the yacc.y file for Write\_C : a
semicolon missing. Seems that previously, there was no such problem... There was another
problem in yacc.y for read\_Alpha. It issues an error message "? bad char: (octal 0)" in a
systematic way. This message comes from a fprintf at the very end of the file. I just removed
the error message, and everything works fine. Why??? no idea.
\item[*04/10/2008:] Made a modification in AlpHard.m, to the function alpha0ToAlphard. Now, 
at the end, all systems are cleaned (option simplify $\rightarrow$ True added) before saving the library.
Note that there is an error in the synthesis of the fir, as the allocation in incorrect if one takes
the simplest schedule. One has to make sure that the first coefficient of the schedule
is 1. Also, possible to synthesize the fir without pipeing variables. Should add a noPipe 
option to the function syn, and see what happens. 
\item[16/09/2008:] started a new package, called Allocate.m, not yet automatically
downloaded. A corresponding example is in myNotebooks/Allocate. There is a function
allocTable, that creates an allocation table, with the name of the system,
its inputs, outputs, its processor space, its operators and their timing space. 
It is displayed with showAT. Its purpuse is to prepare something to merge
differente cells into a single one. 
\item[*13/09/2008:] added a new option to function scd. Option durationsNonZero gives the positions
of the dependences in the dep table which are non zero (duration assumed to be one). 
\item[*6/09/2008:] made a new version, seems to be running properly on Linux. 
Modified the Web site. On my laptop, the new Web site is in directory Copy-MMALPHA-Site/ALPHA.
It is accessible through 
\begin{verbatim}
http://webdav.irisa.fr/www.irisa.fr/htdocs/cosi
\end{verbatim}. 
I have to upgrade this... In the new version, called mmalphaV2-0-1.zip, I have
to modify the diagnose.m file, as it does not work properly for Mathematica version 4.0.
Also, add the information that unzip works... 
\item[3/09/2008:] tried to get a running Linux version. I zipped mmalpha, made a 
copy of it in my directory at irisa. Then I did a make checkvars in sources. It did not
work. I had to replace the makefile.linux in Domlib by a copy of the makefile.darwin. 
I also had to create a Makefile.linux.4.2 in MakeIncludes. Then make checkvars
worked. Then I made make all. Did not work. Seems that I have to provide an old
version of Domlib... because of the incompatibility of the Mathlink library. 
I made a copy of Domlib, and now, I put the old version of domlib.c. But I have
to compile PolyLib... Actually, everything works with the bin.linux version... seems so.
I forgot to set the \$path variable.
\item[*28/07/2008:] corrected the inlining in synthesis. In fact, it is needed to remove identical
equations, normalize, then simplify the system. In the normalization, there is something
strange: the parameter domain is not projected over the domains... Moreover, in a case
statement, if a domain is not empty, then the branch is not removed... 
\item[*26/07/2008:] now using eclipse to manage mmalpha... Wrote a first documentation 
for synthesis. With eclipse, I had a few problems. 
\item[*26/07/2008:] cleaned and committed some demos (Fir, Samba, etc.)
\item[*23/07/2008:] installed in my directory software MinGW, with examples. Contains
\texttt{gcc} to compile for Windows and Linux architectures. Not tried yet...
\item[*7/05/2008:] continued to add the mute option in AlphaToAlpha0v2, Alphard. 
In Subsystem, this is more difficult... since no options in Fixparameters...
\item[*5/05/2008:] succeeded in compiling Domlib. See Domlib documentation for
explanations... Should commit it soon. 
\item[*2/05/2008:] tried to recompile \texttt{writeC}. First, the use of this function is not clear, 
as if no file is given, we have the same input as if a parameter string was given... Second, 
there are missing libraries in the Write\_C program, and I added them. Third, the use of 
strcat seems not to be accepted by gcc. I replaced it by strcopy, and afterwards, I got
a linkedit error. Actually, I had to put the new libraries in yacc.y (which produces yacc.c), 
and make sure it was produced in one single text command, otherwise too many arguments
in Vsep. 
\item[*25/04/2008:] added \texttt{mute} option to \texttt{schedule.m}, \texttt{farkasSchedule.m},
\texttt{analyze}, etc. Committed.
\item[*21/04/2008:] added a \texttt{mute} option to \texttt{Alpha.m} package. \texttt{Alpha.m}, 
\texttt{Synthesis.m}, \texttt{Options.m} and \texttt{Vhdl2.m} were committed. 
Also the demo
directory for \texttt{Synthesis}. Committed.
\item[*20/04/2008:] created a new demo notebooks called \texttt{Synthesis}. Demo
of \texttt{Samba} does not work. Add a parameter for
\texttt{a2v} with the directory where we create \texttt{vhdl} files. Also check all options of
\texttt{a2v} in \texttt{Options.m} and develop them.
\item[19/04/2008:] created a Synthesis.m package, containing a \texttt[syn] function. Documentation
has to be done. Also commit \texttt{Alpha.m} and \texttt{autoload.m}. Next improvements : 
modify \texttt{syn} to handle structured designs, first by simple inlining, then using
real structured design. Add the possibility of designing the vhdl and the schedules, by
reading complex IPs. This could be done by modifying load in such a way that it
reads files with a different suffix using MMA, and then find out the structure of the 
program. 
\item[*18/04/2008:] Committed everything in lib, except \texttt{Normalization.m}. Corrected
a bug in Domlib (in DomEmptyQ). I removed the call to solveEqualities in noSolutions, 
and replaced it by a call to \texttt{solveDiophantine} which works. Should add a note in 
the documentation to explain that \texttt{DomEmptyQ} does not find really if a domain is
empty. Notice also that one can use the function Reduce to check if a domain is empty.
\texttt{Reduce[{i>1,i<2},{i,j},Integer]}. 
\item[*1/04/2008:] Dans Substitution.m, removed a \texttt{;;}. This created pbs in version 6.0. I.
\item[*1/04/2008:] Change in Alpha.m to include Visual3D. I think that it was committed.
\item[*1/04/2008:] Tried to clean up tests. Also, in packages, whenever there are accents, 
there is a problem in version 6.0. Guess it was committed.
\item[*1/04/2008:] In Domlib, I changed Global`scdc into scdc, which may be a problem?
\item[*1/04/2008:] TestsDomlib.m was modified and committed.
\item[?8/01/2008:] Dans pipeVars[], s'il y a un pb lors de l'application d'un pipeline, le
programme boucle...Voir myNotebook firlspg. 
\item[5/01/2008:] The test for the Static package does not work. Also, check the Static doc package.
Check also the doc package of StructuredScheduler. There is an error in the test. 
In analyze, there is a pb when the output of a system is produced by a subsystem (see
in doc package for StructuredScheduler). The test for Substitution is wrong. 
Le package de Uniformization ne marche pas... Il n'était pas installé, il faut tout revoir. 
\item[*31/12/2007:] I checked the Schedule package. I made corrections to scd so that it
can work on multidimensional cases... There is a bug in schematics, that you can 
see in the matmult example (after change of basis). See the nice example of test bench 
generation in matvect.
\item[29/12/2007:] the Shedule doc package has to be checked... Was not commited yet. 
\item[*29/12/2007:] the PipeControl doc package has been copied. Some options of report
are interesting and have to be documented. Committed.
\item[29/12/2007:] the Meta documentation (in doc/Meta and in doc/Packages/Meta) was
checked, updated and committed. The alpha2mma translator still works, but has to 
be completely debugged... The lyx file was translated and may now be forgotten... 
(I did not remember this lyx file, and I therefore was not able to find out this
documentation which I wrote a while ago...)
\item[*26/12/2007:] checking the repository. Committed various modifications in doc. 
Adding the Doc/Package directory. In the CheckAlpHard dir, there are some .simul files
which result, I think, from an attempt to write a simulator of Alpha in MMA. But I do not
know where the corresponding meta files are. In the Meta package doc, there is a 
lyx file which gives a precious documentation on the Meta package... Has to be translated...
\item[15/11/2007:] added functions isOutputRegular, areAllOutputsRegular, mkOutputRegular
and mkAllOutputsRegular to Substitution.m. Also changed getNewName and newName (in
schematics.m). Goal is to detect (in analyze) that outputs are not of the form o = v
where v is a variable and provide a function to change this. Has to commit Analyze, 
Static.m and Substitution.m. Also added MakeDoc.m in packages, and has to 
commit this. There are strange bugs, but documentation is OK. Has to commit TestsSubstitione.
There are bugs in tests of Static.
Commit change to VertexSchedule.m
\item[4/11/2007:] trying the Neural Network of Sanjay in myNotebooks, NN-Sanjay
BackPropNN (main.nb). Everything goes well until the ToAlpha0V2, where I added
some traces. Could not solve the problem. See notebook. 
\item[23/10/2007:] trying to install mmalpha at CSU. Did not work, because Polylib
was not adapted to PPC Mac... Still a problem. I recompiled Polylib (notice that the documentation
regarding Polylib is not up-to-date). Then, I had the problem that the Mathematica version
was wrong. Then, I had to recompile Domlib together with the other programs to get
a new library for Polylib... Eventually, I succeded. Morality : first, get the
result of compilation... Second, try to adapt the compiler to the PPC case... 
The right way to do this would be to generate different bin: bin.macosXppc and bin.macosXintel...
\item[7/9/2007 -- Pip:] found the bug, it was in FarkasSchedule, the calls to 
sed. A corrected version of this package is in the desktop.
\item[29/8/2007 -- Pip:] There are problems regarding Pip on Windows, and I cannot 
find out what these problems are. Actually, there are also some problems with Domlib.
In the old version of Anne-Marie, Pip works, but tests["Domlib"] leads to a bug...
In the current version of Anne-Marie, Pip does not work: it returns an error message
saying that there is no solution (the same version works on the Mac). If I run the old
version with the new directory of Pip and Domlib, I also get an error in Pip, but
this time, the error is different as there is a call to DomMatrixSimplify which is
wrong... So there are some interferences between pip and Domlib. Actually, 
DomMatrixSimplify is used in Domlib, so I do not know where FarkasSchedule.m
calls it. Also, if
I run the new mmalpha with the old binaries, it does not work either... So I
put some Print statements in the FarkasSchedule.m of my Mac version and I 
will try to explore this with Anne-Marie... 
\item[24/7/2007 -- Domlib:] To try to correct the error in Polylib, I worked on 
Domlib in order to be able to recompile it. See in doc-install-domlib my notes
regarding Xcode etc. and all I could not fix.

In the makefiles of Domlib, I made a few changes in order to recompile it. In 
particular, I set the place of Mathlink to the MacOSX-x86 version.

Polylib works correctly, but not domlib. readDom works also nicely. So the 
problem seems to come from Mathlink, or maybe, domlib.c is not uptodate 
anymore... 
\item[20/7/2007 -- Polylib:] I again tried to look at Polylib and recompiling all stuff on 
Intel machine.

First, I downloaded polylib from strasburg (see the note in doc/PolyLib). 
This was successful.

Actually, it seems that to put a new version of PolyLib in MMAlpha, 
one just download it, then install it following the procedure described in the documentation
(see again the note), and then one recompile domlib etc.

I did this in mmalpha-2 (after changing MMALPHA variable).

Then, I made a copy of mmalpha into mmalpha-2. I changed the setenv of MMALPHA (otherwise
the makefile in the sources dir may not work).

I edited the makefile in the sources dir. I changed the DIR variable to only Pretty. 
Then, I made make clean, which removes Obj.dir in Pretty, and I made a make all.
There are some errors, but it produces in Obj.darwin in Pretty a pretty binary file.

Then I tried the same for read\_Alpha. This does not work. It compiles properly read\_alpha.c, 
node.c, matrix.c, and vector.c but not polyhedron.c.

Then, I replaced the Polylib directory in the sources by the new Polylib, and... it worked !

Then, I tried the same for Write\_Alpha, and it did not work...

In the meantime, when putting in DIR of makefile Pretty + read\_ALpha + Write\_ALpha, 
I got a loop...

Then I tried the same only for writeAlpha, and it worked ! Seems that make clean does not
work properly.

I tried again clean with all files in Dir and it worked...

I tried the same for Write\_C and gen.o is missing... I tried to add gen.o from Codegen in 
Obj.darwin of Write\_C, but it did not work.

OK. I went back to Domlib. I had to modify several things. First, make sure that the new Polylib
was recompiled, and installed in directory MMALPHA/sources/Poly. (By the way, in mmalpha, 
I made a copy of bin.darwin renamed bin.darwin-ppc, because I know that this one works
on any mac. I also made a copy of Poly into Poly-darwin, just in case. With the indroduction
of intel mac, all this stuff has to made cleaner...

In case of backup, there is always the solution of the dest-mmalpha-o, and also
the lip forge repository.

For Domlib to compile, changes have to be made in Makefile.darwin of Domlib (where is Polylib), 
Makefile of MakeIncludes (where the mprep is ).

\item[20/7/2007 -- Domlib.m:] there is a bug (I guess) in \texttt{DomSimplify}. The
example is shown here:
\begin{verbatim}
d1 = readDom[  "{ i | i<=0}" ];
d2 = readDom[  "{ i | 0<=i}" ];
domSimplify[ d1, d2 ]
\end{verbatim}
will return domain $\{ i | 1>=0 \}$, i.e. the full space. 
\begin{verbatim}
d1 = readDom[  "{ i | i=0}" ];
d2 = readDom[  "{ i | 0<=i}" ];
domSimplify[ d1, d2 ]
\end{verbatim}
will return domain $\{ i | i=0 \}$, which is correct. When looking in Polylib, I discover
that there is another version of

\item[*15/4/2007 -- All:] checked committements of lib.cygwin, lib.darwin, bin.cygwin, and
doc, until Schedule.
\item[*11/4/2007 -- All:] committed the config directory
\item[*10/4/2007 -- All:] committed all modifications to the distribution in lib. Added \texttt{Visual3D}
and \texttt{Schematics} in the distribution. 
\item[*9/4/2007 -- All:] checked that bin. are OK. 
\item[*8/4/2007 -- Alpha.m:] corrected to print use dependencies. This was not possible in the
MacOSX version.
\item[*7/4/2007 -- VertexSchedule.m:] Edition of some usages in VertexSchedule was done (parameterValues
for example).
\item[*6/4/2007 -- VertexSchedule:] was modified in order to comply with old version
of Mathematica.
\item[*11/12/2006 -- VertexSchedule:] has been modified for the new version of 
MMAlpha. Added options. A new function, changeIndexes, was added to 
changeOfBasis.
\item[*15/9/2006 -- appSched:] has been modified in order to change easily.
\item[9/9/2006: ] The 2D filter works, almost. There is a problem with appSched. In this example, 
when it is applied, it results in an infinite array. The solution is then to apply changes of bases
individually. I also put a check in Vhdl2 and vhdlModule.sem when a use has an 
infinite extension domain. This problem has been fixed. 
\item[5/9/2006:] Tried to work out a 2D filter given by Anne-Marie. See myNotebooks/filter2D. 
The problem was related to structured schedule. Remember that using schedule (Farkas 
scheduler) one needs to use structSched (or schedule[ subSystems->True ]. This is explained
in the schedule documentation. 
\item[25/5/2006:] Received a notebook of Anne-Marie (see myNotebooks/Chana) where she
tries to obtain vhdl code from a simple program. This does not work for several reasons, and
it could be possible to fix these problems. In particular, creating Vhdl code could be possible
by extension of the method to 0 dimensional variables. 
\item[*25/3/2006 -- PipeInfo:] Tanguy sent me a notebook, and PipeInfo does not work. Actually, 
there is a new version corrected in PipeControl-new.m. Removed in it the 
latex command and also, checked that no pipeline is done on a union of a domain.
Has to be commited.
\item[*1/3/2006 -- Schedule, FarkasSchedule, ScheduleTools:] I made a few modifications to \texttt{Schedule.m}, 
\texttt{FarkasSchedule.m}, and maybe also to \texttt{VertexSchedule.m} and
\texttt{ScheduleTools.m}. The modification was to unify the way schedules are
added to \texttt{\$scheduleLibrary}. I added to \texttt{Schedule.m} an internal
function \texttt{storeSchedule} of \texttt{VertexSchedule.m} which adds a
schedule to \texttt{\$scheduleLibrary} only by keeping the input and output
schedules. Thus, one may now schedule a subsystem using both \texttt{scd} or
\texttt{schedule} and still be able to use one or the other scheduler for the 
structured scheduline. 
\item[*15/2/2005 -- Schematics: ] while trying to use \texttt{schematics}, I noticed that the \texttt{Schematic.m}
package is not in the current distribution. OK, committed.
\item[2/12/2005 -- Scheduler] corrections done to the scheduler documentation in \verb!/doc/Schedule!
Added in the doc notebooks the Schedule directory. Not commited yet.
\item[1/12/2005:] created an example called SignAlpha.nb, for Anne-Marie
\item[30/11/2005 -- StructuredScheduled:] added StructuredScheduler in the demos. Is not totally checked yet, and
was not committed.
\item[30/11/2005 -- Demos: ] added Fifo in the demos, but it does not work completely... Not committed yet.
\end{description}

\section{The to do list}
\begin{description}
\item[10/1/2009:] make the corrections seen by the installation of Anne-Marie
\item[9/1/2009:] check that the scheduler is adapted to the new version with string additionnal constraints
and update documentation
\item[*4/1/2009:] update web site
\item[4/1/2009:] check that testSynthesis work, create and test new distribution
\item[2/1/2009:] remove documentation DOMLIB-Install
\item[2/1/2009:] there is a bus error in readAlpha in test4[]
\item[2/1/2009:] look in FarkasSchedule.m to see what happens with \$Runtime aso
\item[* 2/1/2009:] update installation documentation.
\item[2/1/2009:] filter more carefully the arguments of ashow and show. When called with 22 for example,
ashow tries show before falling, which causes a message to be issued in the shell
\item[2/1/2009:] repair \texttt{showLib} (prints out the library directly in the shell)
\item[* 2/1/2009:] clean \texttt{TestAlpha.m} from parasite shell error messages.
\item[* 2/1/2009:] remove ugly messages issued in the shell window when trying to remove files from 
PiP. Probably have to redirect stderr to another file?
\item poursuivre l'analyse de top\_fir pour enlever la fifo et aussi regarder l'affaire de
l'initialisation
\item comprendre pourquoi YOutType n'apparaît nulle part
\item demander à Steven si le fonctionnement des mémoires est correcte
\item peut-on mettre un type en généric. Voir tutorial...
\item If you apply \texttt{toAlpha0v2} to a program that has not been appscheded, it 
fails...
\item There is a message in the Write\_Alpha program, "Could not find the declaration..." that should be removed, as it pollutes the main window...
\item In the config directory, clean red files of the LipForge repository
\item In the lib directory, look at the files which are not in the distribution and add those
which are useful. They are shown in red in the \texttt{Description.xls} file in 
the \texttt{maintenance} directory.
\item By the way, I noticed that \texttt{structSched} does not
work for multi-dimensional schedules, and that would be great to have it. 
\item Add a variable \$pipeInfo that contains the result of the PipeInfo command.
\item Commit \texttt{Schedule.m} etc. (see above).
\item Add \texttt{Schematics.m} to the distribution.
\item Commit new document packages
\item Complete the description of appendices in this document using the README
file in this directory
\item Check demo FIFO
\item Add bibliography to scheduler documentation and add documentation for the
scd command
\item Check the doc package directory for the vertex scheduler
\item Create a doc package directory for the farkas scheduler
\item Extend the vhdl generator to the case of non parallel programs (see myNotebooks/Chana)
\end{description}

\section{The done list}
\begin{description}
\item[2/1/2009:] clean \texttt{TestSubstitution.m} which creates parasite symbols.
\item[-2/1/2009:] Remove \texttt{\$NewName} in Alpha.m. Done.
\item[-1/1/2009:] SystemProgramming is missing in CopyMMA. Some tests are missing in CopyMMA. Could
not understand why CopyDirectory does not work uniquely for TestSchedule ! Done anyway.
\end{description}

\appendix
\section{How to use LipForge}
This version of \mmalfa{} comes from the 
LipForge repository. To use if, the following operations are needed:
\begin{itemize}
\item Open a terminal window.
\item Export the environment variable \verb\CVS\\_RSH\ using
the \texttt{bash} command:
\begin{verbatim}
export CVS_RSH=ssh
\end{verbatim}
or the \texttt{csh} command:
\begin{verbatim}
setenv CVS_RSH ssh
\end{verbatim}
\item Each time you want to apply a \texttt{cvs} command, say 
\texttt{ccc}, use
\begin{verbatim}
cvs -z3 -d:ext:quinton@lipforge.ens-lyon.fr:/local/chroot\
/cvsroot/irisa ccc
\end{verbatim}
Alternatively, save the first part of the command (before \texttt{ccc}) as
an alias
\begin{verbatim}
alias cvs='cvs -z3 -d:ext:quinton@lipforge.ens-lyon.fr:/local\
/chroot/cvsroot/irisa'
\end{verbatim}
and then use the command
\begin{verbatim}
cvs ccc
\end{verbatim}
\end{itemize}
\newcommand{\macosx}{{\sc MacOSX}}

\section{How to connect to my files at Irisa}
In the Finder, select a server. 
\begin{verbatim}
smb://nas1a.irisa.fr/quinton
\end{verbatim}

\section{How to run \mmalfa{} on another server}
On a Linux server
\begin{verbatim}
ssh sabre.irisa.fr
\end{verbatim}

\section{How to access to the web server of MMAlpha}
In the Finder, select a server. 
\begin{verbatim}
http://webdav.irisa/www.irisa.fr/htdocs/cosi
\end{verbatim}
On my laptop, the old site is in
\begin{verbatim}
Copy-MMALPHA-Site
\end{verbatim}

\section{How to run MMA on Mac OS X}
For unknown reasons, it is not possible to start \mmalfa{} directly
from the normal interface of \macosx{}. The solution:
\begin{enumerate}
\item Open a terminal window.
\item Make sure that an environment variable \texttt{MMALPHA} is 
set to the directory where the \mma{} distribution is placed. For example
\begin{verbatim}
export MMALPHA=/Users/quinton/MMAlpha
\end{verbatim}
or under \texttt{csh}:
\begin{verbatim}
setenv MMALPHA /Users/quinton/MMAlpha
\end{verbatim}
\item Then run \mma{}. The \mma{} software is placed 
in directory 
\begin{verbatim}
/Applications/Mathematica 5.1.app/Contents/MacOS
\end{verbatim}
(this may differ depending on the version number.)
\item Normally, \mma{} when started should call a
\texttt{init.m} file, but for unknown reasons, this is not so...
\end{enumerate}
Additional comments:
\begin{itemize}
\item Add a command to \texttt{.alias} in order to be able
to call \mma{} directly, such as for example:
\begin{verbatim}
alias mma /Applications/Mathematica 5.1.app/Contents/MacOS/Mathematica
\end{verbatim}
\item Add a command in the .cshrc file, in order to 
\end{itemize}
\begin{verbatim}
setenv MMALPHA /Users/quinton/mmalpha
echo 'This is my MMA environment: ' ${MMALPHA}
#   For Mac OS X
alias mma '/Applications/Mathematica\ 5.1.app/Contents/MacOS/Mathematica'
alias cvsmma 'cvs -z3 -d:ext:quinton@lipforge.ens-lyon.fr:/local/chroot/cvsroot/irisa'
set path=($MMALPHA/bin.$OSTYPE $path)
set path=(/usr/local/bin $path)
\end{verbatim}
\end{document}

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.gecos.mmalpha/Examples/MMAlpha/UniformizationTools.m

@@ -1,0 +1,213 @@
+(* file: $MMALPHA/lib/Package/UniformizationTools.m
+   AUTHOR :Manjunathaiah. M  
+             University of Reading, U.K.
+           and  Tanguy Risset
+             IRISA, France
+	     
+	     The support provided by EPSRC grant "SARACEN"
+	     (No. GR/K95413) and franco British collaboration 
+	     grant "Remit" (No. PN 97.093) is acknowledged
+	     in the developpement of this package
+
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+
+BeginPackage["Alpha`UniformizationTools`", {"Alpha`Domlib`",
+					    "Alpha`",
+					    "Alpha`Options`",
+					    "Alpha`Matrix`",
+					    "Alpha`Tables`",
+					    "Alpha`Zpol`"}]
+
+
+
+vecInBasisQ::usage="vecInBasisQ[vec_List,basis1_basis]
+check if vector  `vec' is in the space indicated by `basis1'. `basis1Xs'
+is in the format of the result of the findHermiteBasis[] function,
+i.e. an integer `k' and a basis, vecInBasisQ checks whether the vector
+is in the space spanned by the `k' first vectors of the basis. "
+
+getVertices::usage= "getVertices[dom1_Alpha`domain] returns the list
+of vertices of the domain `dom1' as a list of vectors
+ (6 indices-> 6 components)"
+
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.gecos.mmalpha/src/examples/MMAlphaCommandExample.java

@@ -1,0 +1,20 @@
+package examples;
+
+import java.io.IOException;
+
+import com.wolfram.jlink.MathLinkException;
+
+import fr.irisa.cairn.gecos.bindings.mathematica.MathematicaASTPrinter;
+import fr.irisa.cairn.gecos.bindings.mathematica.MathematicaKernel;
+import fr.irisa.cairn.model.mathematica.Node;
+
+public class MMAlphaCommandExample {
+	
+	public static void main(String[] args) throws MathLinkException, IOException {
+		MathematicaKernel kernel = new MathematicaKernel();
+		Node res = kernel.execute("load[\"/home/sderrien/workspace-helios/fr.irisa.cairn.gecos.mmalpha/Examples/MV1.alpha\"]");
+		kernel.saveResultAsXML("Examples/MV1_alpha.mathematica");
+		System.out.println(MathematicaASTPrinter.getPrinter().printTree(res));
+	}
+
+}

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext.editor/src/fr/irisa/cairn/model/mathematica/presentation/MathematicaEditor.java

@@ -1,0 +1,1816 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ * $Id$
+ */
+package fr.irisa.cairn.model.mathematica.presentation;
+
+
+import java.io.IOException;
+import java.io.InputStream;
+
+import java.util.ArrayList;
+import java.util.Collection;
+import java.util.Collections;
+import java.util.EventObject;
+import java.util.HashMap;
+import java.util.Iterator;
+import java.util.LinkedHashMap;
+import java.util.List;
+import java.util.Map;
+
+import org.eclipse.core.resources.IFile;
+import org.eclipse.core.resources.IMarker;
+import org.eclipse.core.resources.IResource;
+import org.eclipse.core.resources.IResourceChangeEvent;
+import org.eclipse.core.resources.IResourceChangeListener;
+import org.eclipse.core.resources.IResourceDelta;
+import org.eclipse.core.resources.IResourceDeltaVisitor;
+import org.eclipse.core.resources.ResourcesPlugin;
+
+import org.eclipse.core.runtime.CoreException;
+import org.eclipse.core.runtime.IPath;
+import org.eclipse.core.runtime.IProgressMonitor;
+import org.eclipse.core.runtime.NullProgressMonitor;
+
+import org.eclipse.jface.action.IMenuListener;
+import org.eclipse.jface.action.IMenuManager;
+import org.eclipse.jface.action.IStatusLineManager;
+import org.eclipse.jface.action.IToolBarManager;
+import org.eclipse.jface.action.MenuManager;
+import org.eclipse.jface.action.Separator;
+
+import org.eclipse.jface.dialogs.MessageDialog;
+import org.eclipse.jface.dialogs.ProgressMonitorDialog;
+
+import org.eclipse.jface.viewers.ColumnWeightData;
+import org.eclipse.jface.viewers.ISelection;
+import org.eclipse.jface.viewers.ISelectionChangedListener;
+import org.eclipse.jface.viewers.ISelectionProvider;
+import org.eclipse.jface.viewers.IStructuredSelection;
+import org.eclipse.jface.viewers.ListViewer;
+import org.eclipse.jface.viewers.SelectionChangedEvent;
+import org.eclipse.jface.viewers.StructuredSelection;
+import org.eclipse.jface.viewers.StructuredViewer;
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/ASTLeaf.java

@@ -1,0 +1,21 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica;
+
+
+/**
+ * <!-- begin-user-doc -->
+ * A representation of the model object '<em><b>AST Leaf</b></em>'.
+ * <!-- end-user-doc -->
+ *
+ *
+ * @see fr.irisa.cairn.model.mathematica.MathematicaPackage#getASTLeaf()
+ * @model
+ * @generated
+ */
+public interface ASTLeaf extends Node
+{
+} // ASTLeaf

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/IntLeaf.java

@@ -1,0 +1,80 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica;
+
+
+/**
+ * <!-- begin-user-doc -->
+ * A representation of the model object '<em><b>Int Leaf</b></em>'.
+ * <!-- end-user-doc -->
+ *
+ * <p>
+ * The following features are supported:
+ * <ul>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.IntLeaf#getSigned <em>Signed</em>}</li>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.IntLeaf#getValue <em>Value</em>}</li>
+ * </ul>
+ * </p>
+ *
+ * @see fr.irisa.cairn.model.mathematica.MathematicaPackage#getIntLeaf()
+ * @model
+ * @generated
+ */
+public interface IntLeaf extends ASTLeaf
+{
+  /**
+   * Returns the value of the '<em><b>Signed</b></em>' attribute.
+   * <!-- begin-user-doc -->
+   * <p>
+   * If the meaning of the '<em>Signed</em>' attribute isn't clear,
+   * there really should be more of a description here...
+   * </p>
+   * <!-- end-user-doc -->
+   * @return the value of the '<em>Signed</em>' attribute.
+   * @see #setSigned(String)
+   * @see fr.irisa.cairn.model.mathematica.MathematicaPackage#getIntLeaf_Signed()
+   * @model
+   * @generated
+   */
+  String getSigned();
+
+  /**
+   * Sets the value of the '{@link fr.irisa.cairn.model.mathematica.IntLeaf#getSigned <em>Signed</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @param value the new value of the '<em>Signed</em>' attribute.
+   * @see #getSigned()
+   * @generated
+   */
+  void setSigned(String value);
+
+  /**
+   * Returns the value of the '<em><b>Value</b></em>' attribute.
+   * <!-- begin-user-doc -->
+   * <p>
+   * If the meaning of the '<em>Value</em>' attribute isn't clear,
+   * there really should be more of a description here...
+   * </p>
+   * <!-- end-user-doc -->
+   * @return the value of the '<em>Value</em>' attribute.
+   * @see #setValue(int)
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/impl/ASTNodeImpl.java

@@ -1,0 +1,242 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica.impl;
+
+import fr.irisa.cairn.model.mathematica.ASTNode;
+import fr.irisa.cairn.model.mathematica.MathematicaPackage;
+import fr.irisa.cairn.model.mathematica.Node;
+
+import java.util.Collection;
+
+import org.eclipse.emf.common.notify.Notification;
+import org.eclipse.emf.common.notify.NotificationChain;
+
+import org.eclipse.emf.common.util.EList;
+
+import org.eclipse.emf.ecore.EClass;
+import org.eclipse.emf.ecore.InternalEObject;
+
+import org.eclipse.emf.ecore.impl.ENotificationImpl;
+
+import org.eclipse.emf.ecore.util.EObjectContainmentEList;
+import org.eclipse.emf.ecore.util.InternalEList;
+
+/**
+ * <!-- begin-user-doc -->
+ * An implementation of the model object '<em><b>AST Node</b></em>'.
+ * <!-- end-user-doc -->
+ * <p>
+ * The following features are implemented:
+ * <ul>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.ASTNodeImpl#getName <em>Name</em>}</li>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.ASTNodeImpl#getChildren <em>Children</em>}</li>
+ * </ul>
+ * </p>
+ *
+ * @generated
+ */
+public class ASTNodeImpl extends NodeImpl implements ASTNode
+{
+  /**
+   * The default value of the '{@link #getName() <em>Name</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getName()
+   * @generated
+   * @ordered
+   */
+  protected static final String NAME_EDEFAULT = null;
+
+  /**
+   * The cached value of the '{@link #getName() <em>Name</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getName()
+   * @generated
+   * @ordered
+   */
+  protected String name = NAME_EDEFAULT;
+
+  /**
+   * The cached value of the '{@link #getChildren() <em>Children</em>}' containment reference list.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getChildren()
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/impl/BuiltInNodeImpl.java

@@ -1,0 +1,242 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica.impl;
+
+import fr.irisa.cairn.model.mathematica.BuiltInNode;
+import fr.irisa.cairn.model.mathematica.MathematicaPackage;
+import fr.irisa.cairn.model.mathematica.Node;
+
+import java.util.Collection;
+
+import org.eclipse.emf.common.notify.Notification;
+import org.eclipse.emf.common.notify.NotificationChain;
+
+import org.eclipse.emf.common.util.EList;
+
+import org.eclipse.emf.ecore.EClass;
+import org.eclipse.emf.ecore.InternalEObject;
+
+import org.eclipse.emf.ecore.impl.ENotificationImpl;
+
+import org.eclipse.emf.ecore.util.EObjectContainmentEList;
+import org.eclipse.emf.ecore.util.InternalEList;
+
+/**
+ * <!-- begin-user-doc -->
+ * An implementation of the model object '<em><b>Built In Node</b></em>'.
+ * <!-- end-user-doc -->
+ * <p>
+ * The following features are implemented:
+ * <ul>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.BuiltInNodeImpl#getKeyword <em>Keyword</em>}</li>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.BuiltInNodeImpl#getChildren <em>Children</em>}</li>
+ * </ul>
+ * </p>
+ *
+ * @generated
+ */
+public class BuiltInNodeImpl extends NodeImpl implements BuiltInNode
+{
+  /**
+   * The default value of the '{@link #getKeyword() <em>Keyword</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getKeyword()
+   * @generated
+   * @ordered
+   */
+  protected static final String KEYWORD_EDEFAULT = null;
+
+  /**
+   * The cached value of the '{@link #getKeyword() <em>Keyword</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getKeyword()
+   * @generated
+   * @ordered
+   */
+  protected String keyword = KEYWORD_EDEFAULT;
+
+  /**
+   * The cached value of the '{@link #getChildren() <em>Children</em>}' containment reference list.
+   * <!-- begin-user-doc -->
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/impl/FloatLeafImpl.java

@@ -1,0 +1,291 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica.impl;
+
+import fr.irisa.cairn.model.mathematica.FloatLeaf;
+import fr.irisa.cairn.model.mathematica.MathematicaPackage;
+
+import org.eclipse.emf.common.notify.Notification;
+
+import org.eclipse.emf.ecore.EClass;
+
+import org.eclipse.emf.ecore.impl.ENotificationImpl;
+
+/**
+ * <!-- begin-user-doc -->
+ * An implementation of the model object '<em><b>Float Leaf</b></em>'.
+ * <!-- end-user-doc -->
+ * <p>
+ * The following features are implemented:
+ * <ul>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.FloatLeafImpl#getSigned <em>Signed</em>}</li>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.FloatLeafImpl#getA <em>A</em>}</li>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.FloatLeafImpl#getB <em>B</em>}</li>
+ * </ul>
+ * </p>
+ *
+ * @generated
+ */
+public class FloatLeafImpl extends ASTLeafImpl implements FloatLeaf
+{
+  /**
+   * The default value of the '{@link #getSigned() <em>Signed</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getSigned()
+   * @generated
+   * @ordered
+   */
+  protected static final String SIGNED_EDEFAULT = null;
+
+  /**
+   * The cached value of the '{@link #getSigned() <em>Signed</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getSigned()
+   * @generated
+   * @ordered
+   */
+  protected String signed = SIGNED_EDEFAULT;
+
+  /**
+   * The default value of the '{@link #getA() <em>A</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getA()
+   * @generated
+   * @ordered
+   */
+  protected static final int A_EDEFAULT = 0;
+
+  /**
+   * The cached value of the '{@link #getA() <em>A</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getA()
[…]

### 2013-01-31 resources/mmalpha/gecos-mmalpha/fr.irisa.cairn.model.mathematica.xtext/src-gen/fr/irisa/cairn/model/mathematica/impl/SymbolLeafImpl.java

@@ -1,0 +1,179 @@
+/**
+ * <copyright>
+ * </copyright>
+ *
+ */
+package fr.irisa.cairn.model.mathematica.impl;
+
+import fr.irisa.cairn.model.mathematica.MathematicaPackage;
+import fr.irisa.cairn.model.mathematica.SymbolLeaf;
+
+import org.eclipse.emf.common.notify.Notification;
+
+import org.eclipse.emf.ecore.EClass;
+
+import org.eclipse.emf.ecore.impl.ENotificationImpl;
+
+/**
+ * <!-- begin-user-doc -->
+ * An implementation of the model object '<em><b>Symbol Leaf</b></em>'.
+ * <!-- end-user-doc -->
+ * <p>
+ * The following features are implemented:
+ * <ul>
+ *   <li>{@link fr.irisa.cairn.model.mathematica.impl.SymbolLeafImpl#getName <em>Name</em>}</li>
+ * </ul>
+ * </p>
+ *
+ * @generated
+ */
+public class SymbolLeafImpl extends ASTLeafImpl implements SymbolLeaf
+{
+  /**
+   * The default value of the '{@link #getName() <em>Name</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getName()
+   * @generated
+   * @ordered
+   */
+  protected static final String NAME_EDEFAULT = null;
+
+  /**
+   * The cached value of the '{@link #getName() <em>Name</em>}' attribute.
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @see #getName()
+   * @generated
+   * @ordered
+   */
+  protected String name = NAME_EDEFAULT;
+
+  /**
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @generated
+   */
+  protected SymbolLeafImpl()
+  {
+    super();
+  }
+
+  /**
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @generated
+   */
+  @Override
+  protected EClass eStaticClass()
+  {
+    return MathematicaPackage.Literals.SYMBOL_LEAF;
+  }
+
+  /**
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
+   * @generated
+   */
+  public String getName()
+  {
+    return name;
+  }
+
+  /**
+   * <!-- begin-user-doc -->
+   * <!-- end-user-doc -->
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/sources/Read_Alpha/node.c

@@ -1,0 +1,1548 @@
+ /*  file: $MMALPHA/sources/Read_Alpha/nodes.c
+   AUTHOR : Doran Wilde
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   */
+/* node.c */
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include "nodetypes.h"
+#include <polylib/polylib.h>
+#include "node.h"
+extern	FILE *yyout;
+extern	void yyerror(char *);
+extern	void yyerror2(char *, char *);
+extern	void yyerror3(char *, char *, char *);
+extern	void yywarning(char *);
+extern	void yywarning2(char *, char *);
+extern	void yywarning3(char *, char *, char *);
+extern  int  do_let;
+
+/*------------------------------------------------------------------*/
+/* internal memory management system                                */
+/*------------------------------------------------------------------*/
+static node	*allocated=(node *)0,
+		*unallocated=(node *)0,
+		*idallocated=(node *)0;
+
+/* add node n to allocation list a */
+static	void 	add_to_alist(ap, n)
+node **ap, *n;
+{   node *a=*ap;
+    if (!a) { *ap = n; return; }
+    n->nx = a;		/* add n to front of list */
+    a->pv->nx = n;	/* end of old list point to new head */
+    n->pv = a->pv;	/* new head points to end of list */
+    a->pv = n;		/* old head points to new head */
+}
+
+/* remove n from allocation list a */
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/sources/Write_C/Test/test2.c

@@ -1,0 +1,79 @@
+/* system foobug2 */
+
+/* C-Code generated by Alpha Code Generator version 1.2 */
+
+#include <math.h>
+typedef struct { int value;
+		 int computed; } intvar;
+typedef struct { int value;
+		 int computed; } boolvar;
+typedef struct { double value;
+		 int computed; } realvar;
+double atof();
+int    atoi();
+
+#define min(x,y) ((x)<(y)?(x):(y))
+#define max(x,y) ((x)>(y)?(x):(y))
+#define INFINITY 0x7fffffff
+#define SHR(x,y) ((x)/(double)(1<<(y)))
+#define SHL(x,y) ((x)*(double)(1<<(y)))
+#define EXP(x,y) pow((x),(y))
+#define TRUNCATE(x) ((int)(x))
+#define CEILING(x) ((int)(ceil(x)))
+#define FLOOR(x) ((int)(floor(x)))
+#define ROUND(x) ((int)(rint(x)))
+#define FLOAT(x) ((double)(x))
+
+/* parameters not fixed at compile time */
+
+/* no input variables */
+
+/* output variables */
+intvar __a[4];
+#define a(i)	__a[(i)-1]
+
+/* local variables */
+intvar __A[4];
+#define A(i)	__A[(i)-1]
+
+static int compute_a();
+
+static int compute_A();
+
+/* --let equations */
+static int compute_A(i)
+int i;
+{
+  intvar *tmp;
+/* printf("Computing: A[%d]\n" ,i); */
+  tmp = &(A(i));
+  if (!tmp->computed) {
+          tmp->value = ( -i+1>=0 || i-3>=0 )  ? ( 0 ) : 
+         ( i-2==0 )  ? ( 1 ) : 
+         ( printf("? case error\n"), exit(-1));
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+static int compute_a(i)
+int i;
+{
+  intvar *tmp;
+/* printf("Computing: a[%d]\n" ,i); */
+  tmp = &(a(i));
+  if (!tmp->computed) {
+          tmp->value = compute_A(i);
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+int main()
+{
+   int i ;
+   
+   /* --inputs */
+   /* --outputs */
+   for (i=1; i<=4; i++)
+   {  printf("a[%d]= %d\n", i, compute_a(i) );
+   }
+}

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/tests/TestCodegen/C/simple1.c

@@ -1,0 +1,162 @@
+/* Generated: 1/10/2008 at 19:29:29 */
+/* Code generated by MMAlpha code generator version 0.2.6 (02/02/2001 11:35) FQ */
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include <assert.h>
+ 
+ 
+ 
+
+#define min(a,b) ((a) > (b) ? (b) : (a))
+#define max(a,b) ((a) > (b) ? (a) : (b))
+#define power(a, i) ((a)^(i))
+static int rfloor (int a, int b) {
+  assert (b>0);
+  return ((a<0) ? ((a+1)/b)-1 : a/b);
+}
+static int rceil (int a, int b) {
+  assert (b>0);
+  return ((a>0) ?  ((a-1)/b)+1 : a/b);
+}
+
+void SimpleEx(float* _a, float* _b, float* _c) {
+  /* Aliases for all variables */
+#define a(n) _a[(-1 + (n))]
+#define b(n, i) _b[(-1 + 3*(i) + (n))]
+#define c(n, i) _c[(-1 + 3*(i) + (n))]
+#define A(t, p) _A[(5*(p) + -7 + (t))]
+#define Amirr(t, p1, p2) _Amirr[(-1 + (p2))]
+#define B_mirr(t, p1, p2) _B_mirr[(5*(p2) + (p1))]
+#define B_reg(t, p1, p2) _B_reg[(-5 + 5*(p2) + (p1))]
+#define C(t, p) _C[(5*(p) + -8 + (t))]
+#define outc(t, p) c(p, (-4 + t))
+   
+  /* Allocate memory for local variables */
+  float * _A = (float *) malloc(sizeof(float)*(35));
+  float * _Amirr = (float *) malloc(sizeof(float)*(3));
+  float * _B_mirr = (float *) malloc(sizeof(float)*(35));
+  float * _B_reg = (float *) malloc(sizeof(float)*(35));
+  float * _C = (float *) malloc(sizeof(float)*(35));
+  
+   
+  /* Loop variables */
+  int i;
+  int n;
+  int p;
+  int p1;
+  int p2;
+  int t;
+   
+  /* A few loops */
+  t = 1;
+  for (p1 = 0; p1 <= 4; ++p1) {
+    for (p2 = p1; p2 <= (2 + p1); ++p2) {
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/tests/TestCodegen/testLud/check.c

@@ -1,0 +1,55 @@
+#include<stdio.h>
+#include<stdlib.h>
+#include<string.h>
+
+/*FILE *fopen(const char *filename, const char *mode);*/
+
+
+
+int main(int argc,char * argv[])
+{
+  FILE  *outNow,*outPrev;
+  float temp1,temp2;
+  char *inpstr;
+  int n,i;
+  
+ 
+
+  system("rm temp");
+  
+  inpstr = "lud < inLud >temp"; 
+
+  system(inpstr);
+
+ 
+  outNow  = fopen("temp","r");
+  outPrev = fopen("outLud","r");    /*open the file written and previous output file */
+ 
+ 
+
+  n=0;
+  while( fscanf(outNow,"%f",&temp1) != EOF |  fscanf(outPrev,"%f",&temp2) != EOF )
+    {  
+      if(temp1 != temp2)
+	{
+	
+	  n=1;
+	}
+    }
+
+  fclose(outNow);
+  fclose(outPrev);
+ 
+  if(n==1)
+    { 
+      fprintf(stderr,"Lud was not correct");
+    }else{
+      fprintf(stderr,"Lud was correct");
+  
+    }
+  
+  
+}
+
+
+

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/tests/TestInstall/TestInstall.m

@@ -1,0 +1,118 @@
+BeginPackage["Alpha`TestInstall`",
+  {"Alpha`","Alpha`Domlib`","Alpha`Options`","Alpha`Static`"}];
+
+Begin["`Private`"];
+
+$testResult = True; 
+
+Module[ {res,file,globalres,path},
+
+globalres =
+{
+  testFunction[
+    res = True;
+    Check[
+      Close[ $testReportFile ]; 
+      $testReportFile = OpenWrite[ Environment["MMALPHA"]<>
+       "/tests/TestingReport.txt"],
+      res = False
+    ];
+
+    If[ res, 
+      WriteString[ $testReportFile, "Test report file was cleaned" ];
+      WriteString[ $testReportFile, "\nDate: ", DateString[] ];
+      WriteString[ $testReportFile, "\nEntering installation test" ],
+      Print[ "Could not open file $testReportFile" ]
+    ];
+    res,
+    True,
+    "Install 1"
+  ]
+,
+  testFunction[
+    res = MatchQ[Environment["MMALPHA"],_String];
+    If[ res, 
+      WriteString[ $testReportFile, "\nMMALPHA is set to: ", 
+        Environment["MMALPHA"]],
+      WriteString[ $testReportFile, "\nMMALPHA improperly set" ];
+      Throw[ res = False ]
+    ];
+    res,
+    True,
+    "Install 2"
+  ]
+,
+  testFunction[
+    res = Cases[ Links[], LinkObject["domlib",__] ];
+    Print[ res ];
+    Which[ 
+      res === {},
+      WriteString[ $testReportFile, "\ndomlib not open"]; res = False,
+      Length[ res ] > 1,
+      WriteString[ $testReportFile, "\nseveral domlib are open" ]; 
+        res = False,
+      Length[ res ] === 1,
+        WriteString[ $testReportFile, "\ndomlib open in link: ", res[[1]] ]; 
+        res = True
+    ];
+    res,
+    True,
+    "Install 3"
+  ]
+,
+  testFunction[
+    res = True;
+    Check[
+      path = Environment["PATH"],
+      res = False
+    ];
+
+    If[ res, 
+      WriteString[ $testReportFile, "\nPATH environment set" ],
+      Print[ "\nCould not get Path environment variable" ]
+    ];
+    res,
+    True,
+    "Install 4"
+  ]
+,
+  testFunction[
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/tests/TestMatrix/TestMatrix.m

@@ -1,0 +1,366 @@
+BeginPackage["Alpha`TestMatrix`",
+  {"Alpha`","Alpha`Domlib`","Alpha`Tables`","Alpha`Matrix`"}];
+
+Begin["`Private`"];
+
+Module[{dir,t1,t2,res},
+
+$testResult = True; 
+
+res=
+{
+  (* Test of translationMatrix *)
+
+  testFunction[translationMatrix[{"i", "j"}, {1, 2}], 
+    matrix[3, 3, {"i", "j"}, {{1, 0, 1}, {0, 1, 2}, {0, 0, 1}}],"Matrix 1"]
+,
+  testFunction[translationMatrix[{"i", "j"}, {1, 2, 3}], $Failed,"Matrix 2"]
+, 
+  testFunction[translationMatrix[2], $Failed, "Matrix 3"]
+,
+  (* Test of composeAffines *)
+  testFunction[
+    composeAffines[readMat["(i,j->j,i)"], readMat["(i,j->j+1,i+3)"]], 
+    matrix[3, 3, {"i", "j"}, {{1, 0, 3}, {0, 1, 1}, {0, 0, 1}}], 
+    "Matrix 4"]
+,
+  testFunction[composeAffines[2], $Failed, "Matrix 5"]
+,
+  testFunction[
+    composeAffines[readMat["(i,j->j,i)"], readMat["(i,j->j,i,i)"]], 
+    $Failed, "Matrix 6"]
+, 
+  (* Test of unimodularQ *)
+  testFunction[unimodularQ[readMat["(i,j->j,i)"]], True, "Matrix 7"]
+,
+  testFunction[unimodularQ[readMat["(i,j->j,2i)"]], False, 
+    "Matrix 8"]
+,
+  testFunction[unimodularQ[readMat["(i,j->j)"]], False, "Matrix 9"]
+,
+  testFunction[unimodularQ[22], $Failed, "Matrix 10"]
+,
+  (* Test of inverseMatrix *)
+  testFunction[inverseMatrix[readMat["(i,j->i+1,j+1)"]], 
+    matrix[3, 
+      3, {"i", "j"}, {{1, 0, -1}, {0, 1, -1}, {0, 0, 1}}], 
+    "Matrix 11"]
+,
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-0-2/tests/TestSchedule/TestSchedule.m

@@ -1,0 +1,381 @@
+BeginPackage["Alpha`TestSchedule`",
+  {"Alpha`","Alpha`Domlib`","Alpha`Schedule`","Alpha`ScheduleTools`",
+   "Alpha`VertexSchedule`","Alpha`FarkasSchedule`", "Alpha`Options`",
+   "Alpha`Tables`"}];
+
+Begin["`Private`"];
+
+(* Written by Tanguy Risset, updated by P. Quinton, Dec. 2008 *)
+Module[ {dir, res, scd1, scd2, ast, testResult},
+
+  $testResult = True; 
+
+  res = 
+    {
+      testFunction[
+        load["Test1.alpha"]; scd1 = schedule[];
+        testSched[ $result, $schedule ], True, "Schedule 1"
+      ]
+    ,
+      testFunction[
+        load["Test2.alpha"];
+        schedule[]; testSched[ $result,$schedule ],
+        True, "Schedule 2.1"
+      ]
+    ,
+      testFunction[
+        load["Test1.alpha"];
+        scd[]; testSched[ $result,$schedule ],
+        True, "Schedule 2.2"
+      ]
+    ,
+      testFunction[
+        load["Test3.alpha"];
+        schedule[]; testSched[ $result,$schedule ],
+        True, "Schedule 3.1"
+      ]
+    ,
+      testFunction[
+        load["Test3.alpha"];
+        scd[]; testSched[ $result,$schedule ],
+        True, "Schedule 3.2"
+      ]
+    ,
+      testFunction[
+        load["Test4.alpha"];
+        schedule[ scheduleType -> sameLinearPart]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule 4.1"
+      ]
+    ,
+      testFunction[
+        load["Test4.alpha"];
+        scd[ scheduleType -> sameLinearPart]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule 4.2"
+      ]
+    ,
+      testFunction[
+        load["TestAffine2.alpha"];
+        schedule[]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule affine 1.1"
+      ]
+    ,
+      testFunction[
+        load["TestUnion.alpha"];
+        schedule[ ratOrInt -> 0 ]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule affine 2.1"
+      ]
+    ,
+      testFunction[
+        load["TestUnion.alpha"];
+        scd[ ]; 
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/demos/NOTEBOOKS/Fir/fir.c

@@ -1,0 +1,109 @@
+/* Generated: 25/4/2010 at 16:2:2 */
+/* Code generated by MMAlpha code generator version 0.2.6 (02/02/2001 11:35) FQ */
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include <assert.h>
+ 
+ 
+ 
+
+#define min(a,b) ((a) > (b) ? (b) : (a))
+#define max(a,b) ((a) > (b) ? (a) : (b))
+#define power(a, i) ((a)^(i))
+static int rfloor (int a, int b) {
+  assert (b>0);
+  return ((a<0) ? ((a+1)/b)-1 : a/b);
+}
+static int rceil (int a, int b) {
+  assert (b>0);
+  return ((a>0) ?  ((a-1)/b)+1 : a/b);
+}
+
+void fir(int* _x, int* _w, int* _y) {
+  /* Aliases for all variables */
+#define x(i) _x[(i)]
+#define w(k) _w[(-1 + (k))]
+#define y(i) _y[(-3 + (i))]
+#define YLOC(t, p) _YLOC[(-12 + 4*(p) + (t))]
+#define outy(t, p) y(p)
+   
+  /* Allocate memory for local variables */
+  int * _YLOC = (int *) malloc(sizeof(int)*(32));
+  
+   
+  /* Loop variables */
+  int i;
+  int p;
+  int t;
+   
+  /* A few loops */
+  t = 0;
+  for (p = 3; p <= 10; ++p) {
+    YLOC(t, p) = 0;
+  }
+  for (t = 1; t <= 3; ++t) {
+    for (p = 3; p <= 10; ++p) {
+      YLOC(t, p) = (YLOC((-1 + t), p) + (w(t) * x((p + -t))));
+    }
+  }
+  t = 4;
+  for (p = 3; p <= 10; ++p) {
+    outy(t, p) = YLOC(3, p);
+  }
+   
+  /* Clean up local variables' memory */
+  /* Commented out because it was crashing at run time*/
+  /*
+  free(_YLOC);
+  
+  */
+  /* And finally undef aliases */
+#undef x
+#undef w
+#undef y
+#undef YLOC
+#undef outy
+}
+ 
+int main (void) {
+#define x(i) _x[(i)]
+#define w(k) _w[(-1 + (k))]
+#define y(i) _y[(-3 + (i))]
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Developers/Writing-a-package/MyPackage.m

@@ -1,0 +1,255 @@
+(* 
+  The following is an example of MMAlpha package. Its structure is 
+  roughly as follows: 
+   - Gnu license
+   - Package declaration and usage statements for exported objects.
+   - Private part of the package.
+
+  Read if carefully.
+
+  The way this package is written is not always met by all MMAlpha 
+  packages, simply because it resulted from a pretty long experience
+  of using MMA (some of these experiences were pretty bad). 
+  Do not complain about this : MMAlpha was written by many people over
+  about 15 years, and Tanguy and myself make our best to clean code
+  whenever we find some time. But time is a very expensive resource...
+
+  Patrice Quinton
+*)
+
+(* file: $MMALPHA/lib/Package/MyPackage.m
+   AUTHOR : The Alpha Contributors 
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   etc...
+
+*)
+BeginPackage["Alpha`MyPackage`", {"Alpha`Domlib`","Alpha`",
+                                      "Alpha`Options`",
+				      "Alpha`Matrix`",
+                                      "Alpha`Substitution`",
+                                      "Alpha`Normalization`",
+				      "Alpha`Tables`"}];
+
+(* 
+  The BeginPackage statement has two arguments. The first one defines
+  the name of the package, and the second one is a list of all packages
+  whose functions are needed by functions of myPackage. 
+
+  The name of a package is a string of the form Alpha`MyPackage` .
+  All packages (except the root Alpha` are placed in directory 
+  $MMALPHA/lib/Packages/Alpha in a file named MyPackage.m .
+
+  Package start with an uppercase letter. Functions of MMAlpha all 
+  start with a lowercase letter to avoid potential name conflicts 
+  with Mathematica functions. 
+*)
+
+(*
+  The commented lines below are automatically updated by CVS. 
+  See the CVS usage.
+*)
+
+(* $Id: MyPackage.m,v 1.1 2006/05/28 16:22:24 quinton Exp $
+
+   changelog:
+
+   07/15/1993, D. Wilde:	full text specification of COBs for batch use.
+   11/29/1993, Z. Chamski:	fixed normalization problem preventing
+				 identity changes of basis from changing index
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Packages/CheckAlphard/CheckCell.m

@@ -1,0 +1,392 @@
+BeginPackage["Alpha`CheckCell`",{"Alpha`",
+"Alpha`Domlib`",
+"Alpha`Tables`",
+"Alpha`Matrix`", 
+"Alpha`Options`", 
+"Alpha`Static`"}];
+CheckCellDebug::usage = "Debug Switch";
+CheckCellDebug = False;
+CheckCellTranslatelist::usage = "";
+CheckCellTranslateSYSTEMDECLARATION::usage = "";
+CheckCellTranslateDECLARATION::usage = "";
+CheckCellTranslateDOMAINPARAM::usage = "";
+CheckCellTranslateDOMAIN::usage = "";
+CheckCellTranslatePOLYHEDRON::usage = "";
+CheckCellTranslateCONSTRAINT::usage = "";
+CheckCellTranslateGENERATOR::usage = "";
+CheckCellTranslateEQUATION::usage = "";
+CheckCellTranslateASSIGNMENT::usage = "";
+CheckCellTranslateELEMENTS::usage = "";
+CheckCellTranslateAFFEXP::usage = "";
+CheckCellTranslateMATRIX::usage = "";
+CheckCellTranslateMATRIXNUM::usage = "";
+CheckCellTranslateBINEXP::usage = "";
+CheckCellTranslateUNEXP::usage = "";
+CheckCellTranslateVAREXPORCONST::usage = "";
+CheckCellTranslateVAREXPE::usage = "";
+CheckCellTranslateCONSTEXPE::usage = "";
+CheckCellTranslateVAREXP::usage = "";
+CheckCellTranslateCONSTEXP::usage = "";
+CheckCellTranslateSUBEXPRESSION::usage = "";
+CheckCellTranslateCASEMUX::usage = "";
+CheckCellTranslateRESTEXP::usage = "";
+CheckCellTranslateIFEXP::usage = "";
+CheckCellTranslateCONTROLEXPRESSION::usage = "";
+CheckCellTranslateIFMUX::usage = "";
+CheckCellTranslateSUBMUX2::usage = "";
+CheckCellTranslateCASEXP::usage = "";
+CheckCellTranslateRESTEXPMUX2::usage = "";
+Begin[ "`Private`" ];
+With[{f=$rootDirectory<>"/"<>fileName[{"lib","Packages","Alpha","CheckCell.sem"}]},
+  Check[ Get[ f ],
+       Print[ "Error while trying reading file CheckCell "]]
+];
+Clear[ CheckCellTranslateList ];
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Packages/CodeGen/verySimple.c

@@ -1,0 +1,85 @@
+/* Generated: 27/12/2002 at 18:31:31 */
+/* Code generated by MMAlpha code generator version 0.2.6 (02/02/2001 11:35) FQ */
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include <assert.h>
+ 
+ 
+
+#define min(a,b) ((a) > (b) ? (b) : (a))
+#define max(a,b) ((a) > (b) ? (a) : (b))
+#define power(a, i) ((a)^(i))
+static int rfloor (int a, int b) {
+  assert (b>0);
+  return ((a<0) ? ((a+1)/b)-1 : a/b);
+}
+static int rceil (int a, int b) {
+  assert (b>0);
+  return ((a>0) ?  ((a-1)/b)+1 : a/b);
+}
+
+void SimpleEx(float* _a, float* _c, int N) {
+  /* aliases for all variables */
+#define a(n) _a[(-1 + (n))]
+#define c(n) _c[(-1 + (n))]
+#define A(t, p) _A[(-1 + (p))]
+#define outc(t, p) c(p)
+   
+  /* allocate memory for local variables */
+  float * _A = (float *) malloc(sizeof(float)*(N));
+  
+   
+  /* loops variables */
+  int n;
+  int p;
+  int t;
+   
+  /* a few loops */
+  t = 1;
+  for (p = 1; p <= N; ++p) {
+    A(t, p) = (a(p) * a(p));
+  }
+  t = 2;
+  for (p = 1; p <= N; ++p) {
+    outc(t, p) = A(1, p);
+  }
+   
+  /* clean up local variables' memory */
+  /* commented out because it was crashing at run time*/
+  /*
+  free(_A);
+  
+  */
+  /* and finally undef aliases */
+#undef a
+#undef c
+#undef A
+#undef outc
+}
+ 
+int main (void) {
+#define a(n) _a[(-1 + (n))]
+#define c(n) _c[(-1 + (n))]
+  float * _a = (float *) malloc(sizeof(float)*(N));
+  float * _c = (float *) malloc(sizeof(float)*(N));
+  int n;
+   
+  for (n = 1; n <= N; ++n) {
+    fprintf(stdout, "a[%i]?", n);
+    fscanf(stdin, "%f", &(a(n)));
+     
+  }
+  SimpleEx(_a, _c);
+  for (n = 1; n <= N; ++n) {
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/doc/Packages/Meta/test.m

@@ -1,0 +1,7 @@
+BeginPackage["monPackage`"];
+
+x = 2;
+Get["test2.m"];
+y
+
+EndPackage[];

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/Control.m

@@ -1,0 +1,1480 @@
+(* file: $MMALPHA/lib/Package/Alpha.m
+   AUTHOR : Patricia Le Moenner
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+
+BeginPackage["Alpha`Control`", {"Alpha`Domlib`",
+				"Alpha`",
+				"Alpha`Options`",
+				"Alpha`Matrix`",
+				"Alpha`Properties`",
+				"Alpha`PipeControl`",
+				"Alpha`Tables`",
+				"Alpha`Normalization`",
+				"Alpha`Substitution`",
+				"Alpha`Semantics`",
+				"Alpha`Schedule`"}]
+
+(* $Id: Control.m,v 1.4 2010/04/10 19:52:14 quinton Exp $
+ 
+ changelog: none
+ *)
+
+Control::usage = 
+"Package. Contains the definition of functions related to control
+signal generation: temporalCaseQ[], spatialCaseQ[], spaceTimeCase[],
+spaceTimeDecomposition[], needsMuxQ[], makeMuxControl[],
+makeAllMuxControl[], isControlEquQ[], controlVars[]."
+
+Control::latex = 
+"The \\texttt{Control} package contains functions related to 
+the generation of control signal. These functions are\,:\n 
+\begin{itemize}\n
+\\item \\texttt{controlVars}.\n
+\\item \\texttt{isControlEquQ},\n
+\\item \\texttt{makeMuxControl},\n
+\\item \\texttt{makeAllMuxControl},\n
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/Lexicographic.m

@@ -1,0 +1,370 @@
+(* file: $MMALPHA/lib/Package/Alpha.m
+   AUTHOR : The Alpha Contributors 
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+ 
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+ 
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+(*
+        $Author: trisset $
+        $Date: 2005/03/11 16:40:17 $
+        $Source: /local/chroot/cvsroot/irisa/mmalpha/lib/Packages/Alpha/Lexicographic.m,v $
+        $Revision: 1.1 $
+        $Log: Lexicographic.m,v $
+        Revision 1.1  2005/03/11 16:40:17  trisset
+        added all remaining packages to V2
+
+        Revision 1.2  1999/04/26 12:30:04  quillere
+        removed Alpha`Tools` need
+
+        Revision 1.1  1999/04/26 09:19:51  quillere
+        added temporary AppSched packages -- will have dispatch it
+
+*)
+
+BeginPackage["Alpha`Lexicographic`"];
+
+Unprotect[domLexGreater, domLexLower,
+	  lexLowerQ, lexGreaterQ,
+	  lexLowerOrEqualQ, lexGreaterOrEqualQ];
+
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/PipeControl-saved.m

@@ -1,0 +1,3192 @@
+(* file: $MMALPHA/lib/Package/Alpha.m
+   AUTHOR : Patricia Le Moenner, Patrice Quinton
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+BeginPackage["Alpha`PipeControl`",{"Alpha`Domlib`",
+				   "Alpha`",
+				   "Alpha`Options`",
+				   "Alpha`Matrix`",
+				   "Alpha`Properties`",
+				   "Alpha`Tables`",
+				   "Alpha`Semantics`",
+				   "Alpha`Substitution`",
+				   "Alpha`Normalization`",
+				   "Alpha`Pipeline`",
+				   "Alpha`Control`",
+				   "Alpha`Alphard`",
+(*
+                                   "Alpha`Uniformization`", 
+*)
+				     "Alpha`CodeGen`Domains`"
+			     }];
+
+(* changelog:
+    28/3/95 P. Quinton
+*)
+(* Standard head for CVS
+
+	$Author: quinton $
+	$Date: 2009/05/22 10:24:37 $
+	$Source: /local/chroot/cvsroot/irisa/mmalpha/lib/Packages/Alpha/PipeControl.m,v $
+	$Revision: 1.11 $
+	$Log: PipeControl.m,v $
+	Revision 1.11  2009/05/22 10:24:37  quinton
+	May 2009
+	
+	Revision 1.10  2008/12/29 17:37:02  quinton
+	Minor corrections
+	
+	Revision 1.9  2008/11/30 11:35:07  quinton
+	Commented a Print
+	
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/SubSystems.m

@@ -1,0 +1,1777 @@
+(* file: $MMALPHA/lib/Package/Alpha.m
+   AUTHOR :  Florent dupont de Dinechin
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+
+BeginPackage["Alpha`SubSystems`", {"Alpha`Domlib`",
+				   "Alpha`",
+				   "Alpha`Options`",
+				   "Alpha`Matrix`",
+				   "Alpha`Tables`",
+				   "Alpha`Semantics`",
+				   "Alpha`Substitution`",
+				   "Alpha`Normalization`",
+				   "Alpha`CutMMA`",
+				   "Alpha`ChangeOfBasis`",
+                                   "Alpha`Visual`"}  ]
+
+(* $Id: SubSystems.m,v 1.4 2009/05/22 10:24:36 quinton Exp $
+
+   changelog:
+   02/08/1995, F. de Dinechin: added expandSystem[]
+   02/14/1995, F. de Dinechin: expandSystem[] works with m_use
+   03/27/1995, F. de Dinechin: m_use removed, file and lib functions moved
+			to Alpha.m, assignParameterValue added
+   04/28/1995, F. de Dinechin: inlineSubsystem[] added (options). Added
+                          proper index renaming.
+   05/09/1995, P. Quinton: small usage modifs...
+   02/27/1995, F. de Dinechin: replaced Print[] by Message[] in case of
+   an error, so that callers (e.g. analyze) may detect the error by Check[];
+  NOTE: .
+*)
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/VhdlTestBench.m

@@ -1,0 +1,1644 @@
+BeginPackage["Alpha`vhdlTestBench`",
+{"Alpha`",
+ "Alpha`Domlib`",
+ "Alpha`Tables`",
+ "Alpha`Matrix`", 
+ "Alpha`Options`", 
+ "Alpha`Vhdl2`", 
+ "Alpha`Visual`",
+		       (* "Alpha`VertexSchedule`", *)
+ "Alpha`Substitution`",
+ "Alpha`Static`",
+ "Alpha`Semantics`",
+ "Alpha`CheckModule`", 
+ "Alpha`CheckCell`", 
+ "Alpha`CheckController`", 
+ "Alpha`VertexSchedule`", 
+ "Alpha`Alphard`"
+}
+];
+
+(* Standard head for CVS
+
+	$Author: quinton $
+	$Date: 2008/12/29 17:55:02 $
+	$Source: /local/chroot/cvsroot/irisa/mmalpha/lib/Packages/Alpha/VhdlTestBench.m,v $
+        $Revision: 1.7 $
+	$Log: VhdlTestBench.m,v $
+	Revision 1.7  2008/12/29 17:55:02  quinton
+	Minor corrections.
+	
+	Revision 1.6  2005/07/04 08:26:17  ebecheto
+	little modif of a typing error
+	
+	Revision 1.5  2005/05/20 13:18:49  trisset
+	correct VhdlTestBench.m
+	
+	Revision 1.4  2005/05/13 08:21:10  ebecheto
+	Ed modif for truncate_*SB procedure
+	
+	Revision 1.3  2005/03/31 13:43:24  trisset
+	update VhdlTestBench for 2D array (started)
+	
+	Revision 1.2  2005/03/14 16:18:59  trisset
+	added VhdlTestBench
+	
+	Revision 1.1  2005/03/14 14:39:11  trisset
+	added VhdlTestBench
+	
+	Revision 1.2  2004/04/26 08:34:14  risset
+	added system File before patrice visit
+	
+	Revision 1.1  2003/12/12 13:13:24  risset
+	 add vhdlTestBench
+	
+
+*)
+(*
+
+ Generation of test bench files from Alphard 
+ 
+*)
+
+VhdlTestBanch::usage = 
+"Package,contains definition of funtion for generating test benches for VHDL generated from Alphard"
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/Visual3D.m

@@ -1,0 +1,821 @@
+BeginPackage["Alpha`Visual3D`",
+  {"Alpha`",
+   "Alpha`Tables`",
+   "Alpha`Semantics`",
+   "Alpha`Domlib`"
+      (*                               "Graphics`Colors`" *)
+  }];
+
+(* This package contains a few functions to show 3D polyhedra.
+   Function planes is used to compute the faces of a polyhedron 
+   given by its vertices. This function is not useful if MMAlpha
+   is used. It returns a pair composed of the sets of vertices 
+   of each face, and the planes delimiting the faces. The second
+   main function of this package is myPolyhedron, which takes a
+   list of planes, and a list of faces - as produced by planes, -
+   and produces a list of convex polygons, which can be displayed
+   by MMA without additional triangularization *)
+
+Visual3D::note = "Documentation revised on August 10, 2004";
+
+Visual3D::usage =
+"The Alpha`Visual3D` package contains a few functions to visualize
+and animate 3D domains. The main function is vshow, other functions 
+are facets, and listPlanes. Some functions are exported only for debugging
+purposes.";
+
+facets::usage = 
+"facets[dom] computes the list of polygons corresponding to the domain dom."; 
+
+listPlanes::usage = 
+"listPlanes[l,p] returns the pair {l,lp} where lp is the lists of planes in p which contain point l.";
+
+maxR::usage = 
+"maxR is an option of vshow. It allows the viewpoint of the final picture to 
+be set. See stepR for more details.";
+
+minR::usage = 
+"minR is an option of vshow. It allows the r parameter of the viewpoint 
+to be changed. By default, minR is 1. See also stepR and maxR,
+and ViewPoint of Mathematica.";
+
+orderPolygon::usage = "only here for debugging purposes.";
+
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Alpha/Zpol.m

@@ -1,0 +1,261 @@
+(* file: $MMALPHA/lib/Package/Alpha.m
+   AUTHOR :  Tanguy Risset
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA.   
+*)
+BeginPackage["Alpha`Zpol`", {"Alpha`Domlib`",  
+			     "Alpha`",
+                               "Alpha`Options`",
+                               "Alpha`Matrix`",
+			     "Alpha`Tables`"
+			     }]
+
+(* $Id: Zpol.m,v 1.1 2005/03/11 16:40:17 trisset Exp $  *)
+(* Standard head for CVS
+
+	$Author: trisset $
+	$Date: 2005/03/11 16:40:17 $
+	$Source: /local/chroot/cvsroot/irisa/mmalpha/lib/Packages/Alpha/Zpol.m,v $
+	$Revision: 1.1 $
+	$Log: Zpol.m,v $
+	Revision 1.1  2005/03/11 16:40:17  trisset
+	added all remaining packages to V2
+	
+	Revision 1.4  2004/09/16 13:45:23  quinton
+	Updated documentation
+	
+	Revision 1.3  2004/08/02 13:13:26  quinton
+	Documentation updated for reference manual
+	
+	Revision 1.2  2004/07/08 14:06:51  risset
+	commit Zpol changes
+	
+	Revision 1.1  2001/05/07 14:26:26  risset
+	restaur Zpol.m
+	
+	Revision 1.16  2001/05/07 13:39:44  risset
+	test for control M
+	
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/lib/Packages/Fail.m

@@ -1,0 +1,22 @@
+BeginPackage["Fail`"];
+
+Attributes[verbFail] = {};
+Attributes[chkFail] = {};
+ClearAll[chkFail];
+ClearAll[verbFail];
+
+chkFail::usage = "chkFail[a] throws an exception if a is $Failed, else
+returns a";
+verbFail::usage = "verbFail[mess, expr] prints the message if expr is
+$Failed, and returns exp";
+
+Begin["`Private`"];
+
+chkFail[a:_] := If[a===$Failed||a===Null,Throw[$Failed],a];
+verbFail[mess:_, expr:_] := If[expr===$Failed,Message[mess];$Failed,expr];
+
+Attributes[verbFail] = {Protected, ReadProtected, HoldFirst};
+Attributes[check] = {Protected, ReadProtected};
+
+End[];
+EndPackage[];

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/sources/Pretty/readitem.c

@@ -1,0 +1,262 @@
+/*  file: $MMALPHA/sources/Pretty/readitem.c
+   AUTHOR : Doran Wilde
+   CONTACT : http://www.irisa.fr/api/ALPHA
+   COPYRIGHT  (C) INRIA
+   
+   This library is free software; you can redistribute it and/or
+   modify it under the terms of the GNU Library General Public
+   License as published by the Free Software Foundation; either
+   version 2 of the License, or (at your option) any later version.
+   (see file : $MMALPHA/LICENSING).
+
+   This library is distributed in the hope that it will be useful,
+   but WITHOUT ANY WARRANTY; without even the implied warranty of
+   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+   Library General Public License for more details.
+
+   You should have received a copy of the GNU Library General Public
+   License along with this library(see file : $MMALPHA/LICENSING);
+   if not, write to the Free Software Foundation, Inc., 59 Temple
+   Place - Suite 330, Boston, MA  02111-1307, USA. */  
+/*
+/*
+ * This program outputs a typeset piece of text
+ */
+
+#include <stdio.h>
+#include <stdlib.h>
+#include "item.h"
+#include "itemprocs.h"
+#include "writeitem.h"
+
+typedef
+struct token_type_ {
+        union {
+                struct {
+                        int             value;
+                }               number;
+                struct {
+                        char            string[128];
+                }               text;
+        }               the;
+}               token_type;
+
+static token_type      token;
+static item            *head;
+
+#include "lex.c"
+FILE *ref=0;
+
+static void
+open_paren_test(identifier)
+   char            identifier[];
+{
+   if (yylex() != '(') {
+      printf("An open parenthesis was expected by %s .", identifier);
+      exit(1);
+   }
+}
+
+static void
+close_paren_test(identifier)
+   char            identifier[];
+{
+   if (yylex() != ')') {
+      printf("A close parenthesis was expected by %s .", identifier);
+      exit(1);
+   }
+}
+
+static void
+comma_test(identifier)
+   char            identifier[];
+{
+   if (yylex() != ',') {
+      printf("A comma was expected by %s .", identifier);
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/sources/Write_Alpha/lex.c

@@ -1,0 +1,2220 @@
+
+#line 3 "<stdout>"
+
+#define  YY_INT_ALIGNED short int
+
+/* A lexical scanner generated by flex */
+
+#define FLEX_SCANNER
+#define YY_FLEX_MAJOR_VERSION 2
+#define YY_FLEX_MINOR_VERSION 5
+#define YY_FLEX_SUBMINOR_VERSION 33
+#if YY_FLEX_SUBMINOR_VERSION > 0
+#define FLEX_BETA
+#endif
+
+/* First, we deal with  platform-specific or compiler-specific issues. */
+
+/* begin standard C headers. */
+#include <stdio.h>
+#include <string.h>
+#include <errno.h>
+#include <stdlib.h>
+
+/* end standard C headers. */
+
+/* flex integer type definitions */
+
+#ifndef FLEXINT_H
+#define FLEXINT_H
+
+/* C99 systems have <inttypes.h>. Non-C99 systems may or may not. */
+
+#if __STDC_VERSION__ >= 199901L
+
+/* C99 says to define __STDC_LIMIT_MACROS before including stdint.h,
+ * if you want the limit (max/min) macros for int types. 
+ */
+#ifndef __STDC_LIMIT_MACROS
+#define __STDC_LIMIT_MACROS 1
+#endif
+
+#include <inttypes.h>
+typedef int8_t flex_int8_t;
+typedef uint8_t flex_uint8_t;
+typedef int16_t flex_int16_t;
+typedef uint16_t flex_uint16_t;
+typedef int32_t flex_int32_t;
+typedef uint32_t flex_uint32_t;
+#else
+typedef signed char flex_int8_t;
+typedef short int flex_int16_t;
+typedef int flex_int32_t;
+typedef unsigned char flex_uint8_t; 
+typedef unsigned short int flex_uint16_t;
+typedef unsigned int flex_uint32_t;
+#endif /* ! C99 */
+
+/* Limits of integral types. */
+#ifndef INT8_MIN
+#define INT8_MIN               (-128)
+#endif
+#ifndef INT16_MIN
+#define INT16_MIN              (-32767-1)
+#endif
+#ifndef INT32_MIN
+#define INT32_MIN              (-2147483647-1)
+#endif
+#ifndef INT8_MAX
+#define INT8_MAX               (127)
+#endif
+#ifndef INT16_MAX
+#define INT16_MAX              (32767)
+#endif
+#ifndef INT32_MAX
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/sources/Write_Alpha/nodeprocs.h

@@ -1,0 +1,28 @@
+/*
+ *   This file was automatically generated by version 1.7 of cextract.
+ *   Manual editing not recommended.
+ *
+ *   Created: Mon Jul 12 10:43:22 1999
+ */
+#ifndef _nodeprocs_
+#define _nodeprocs_
+#if __STDC__
+
+extern node *add_to_list ( node *n, node *e );
+extern void free_node ( node *n );
+extern node *new_id ( char *s );
+extern node *new_list ( node *e );
+extern node *new_node ( nodetype t );
+extern void remove_from_list ( node *n, node *e );
+
+#else /* __STDC__ */
+
+extern node *add_to_list (/* node *n, node *e */);
+extern void free_node (/* node *n */);
+extern node *new_id (/* char *s */);
+extern node *new_list (/* node *e */);
+extern node *new_node (/* nodetype t */);
+extern void remove_from_list (/* node *n, node *e */);
+
+#endif /* __STDC__ */
+#endif /* _nodeprocs_ */

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/sources/Write_Alpha/writeitem.h

@@ -1,0 +1,19 @@
+/*
+ *   This file was automatically generated by version 1.7 of cextract.
+ *   Manual editing not recommended.
+ *
+ *   Created: Thu Apr 24 10:21:53 1997
+ */
+#ifndef _writeitem_
+#define _writeitem_
+#if __STDC__
+
+extern void print_item ( item *x );
+extern void print_item_list ( item *x );
+
+#else /* __STDC__ */
+
+extern void print_item_list (/* item *x */);
+
+#endif /* __STDC__ */
+#endif /* _writeitem_ */

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/sources/Write_C/Test/pdscal.c

@@ -1,0 +1,99 @@
+/* system pdscal */
+
+/* C-Code generated by Alpha Code Generator version 1.2 */
+
+#include <math.h>
+typedef struct { int value;
+		 int computed; } intvar;
+typedef struct { int value;
+		 int computed; } boolvar;
+typedef struct { double value;
+		 int computed; } realvar;
+double atof();
+int    atoi();
+
+#define min(x,y) ((x)<(y)?(x):(y))
+#define max(x,y) ((x)>(y)?(x):(y))
+#define INFINITY 0x7fffffff
+
+/* parameters */
+
+/* input variables */
+char __s[32];
+realvar _V[8];
+#define V(i,p)	_V[(i)+4*(p)-5]
+realvar _pivot;
+#define pivot()	_pivot
+
+/* output variables */
+realvar _Vdiv[8];
+#define Vdiv(i,p)	_Vdiv[(i)+4*(p)-5]
+
+/* no local variables */
+
+static double compute_V(i,p)
+int i,p;
+{
+  realvar *tmp;
+/* printf("Computing: V[%d,%d]\n" ,i,p); */
+  tmp = &(V(i,p));
+  if (!tmp->computed) {
+          printf("Input V[%d,%d] =" ,i,p);
+     gets(__s);
+     tmp->value = atof(__s);
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+static double compute_pivot()
+{
+  realvar *tmp;
+/* printf("Computing: pivot[]\n"); */
+  tmp = &(pivot());
+  if (!tmp->computed) {
+          printf("Input pivot =" );
+     gets(__s);
+     tmp->value = atof(__s);
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+static double compute_Vdiv();
+
+/* --let equations */
+static double compute_Vdiv(i,p)
+int i,p;
+{
+  realvar *tmp;
+/* printf("Computing: Vdiv[%d,%d]\n" ,i,p); */
+  tmp = &(Vdiv(i,p));
+  if (!tmp->computed) {
+          tmp->value = ( -p+1>=0 )  ? ( compute_V(i,p) ) : 
+         ( p-2==0 )  ? ( ( i-3>=0 && -i+4>=0 )  ? 
+                     ( compute_V(i,p) / compute_pivot() ) : 
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestCodegen/C/simple1.c

@@ -1,0 +1,162 @@
+/* Generated: 23/4/2010 at 14:30:30 */
+/* Code generated by MMAlpha code generator version 0.2.6 (02/02/2001 11:35) FQ */
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+#include <assert.h>
+ 
+ 
+ 
+
+#define min(a,b) ((a) > (b) ? (b) : (a))
+#define max(a,b) ((a) > (b) ? (a) : (b))
+#define power(a, i) ((a)^(i))
+static int rfloor (int a, int b) {
+  assert (b>0);
+  return ((a<0) ? ((a+1)/b)-1 : a/b);
+}
+static int rceil (int a, int b) {
+  assert (b>0);
+  return ((a>0) ?  ((a-1)/b)+1 : a/b);
+}
+
+void SimpleEx(float* _a, float* _b, float* _c) {
+  /* Aliases for all variables */
+#define a(n) _a[(-1 + (n))]
+#define b(n, i) _b[(-1 + 3*(i) + (n))]
+#define c(n, i) _c[(-1 + 3*(i) + (n))]
+#define A(t, p) _A[(5*(p) + -7 + (t))]
+#define Amirr(t, p1, p2) _Amirr[(-1 + (p2))]
+#define B_mirr(t, p1, p2) _B_mirr[(5*(p2) + (p1))]
+#define B_reg(t, p1, p2) _B_reg[(-5 + 5*(p2) + (p1))]
+#define C(t, p) _C[(5*(p) + -8 + (t))]
+#define outc(t, p) c(p, (-4 + t))
+   
+  /* Allocate memory for local variables */
+  float * _A = (float *) malloc(sizeof(float)*(35));
+  float * _Amirr = (float *) malloc(sizeof(float)*(3));
+  float * _B_mirr = (float *) malloc(sizeof(float)*(35));
+  float * _B_reg = (float *) malloc(sizeof(float)*(35));
+  float * _C = (float *) malloc(sizeof(float)*(35));
+  
+   
+  /* Loop variables */
+  int i;
+  int n;
+  int p;
+  int p1;
+  int p2;
+  int t;
+   
+  /* A few loops */
+  t = 1;
+  for (p1 = 0; p1 <= 4; ++p1) {
+    for (p2 = p1; p2 <= (2 + p1); ++p2) {
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestDomlib/TestDomlib-Short.m

@@ -1,0 +1,39 @@
+(* 
+
+$Id
+$Author
+$Date
+$Source
+$Revision
+$Log
+
+*)
+(* Updated by P. Quinton. 2/01/2003 and August 1st, 2004*)
+Module[
+  { i, j, dir, listResDomlib, listResZDomlib, res, d1, m1, m2},
+  Print["Test for Domlib"];
+  dir = Directory[]; 
+  SetDirectory[Environment["MMALPHA"]<>"/tests/TestDomlib"];
+
+  listResDomlib=
+  {
+    Print["Test of DomIntersection"];
+    d1=readDom["{i,j|0<=i;j>=0}"];d2=readDom["{i,j|j>=0}"];
+    testFunction[DomIntersection[d1,d2], 
+		domain[2,{"i","j"},{pol[3,3,0,0,
+					{{1,1,0,0},{1,0,1,0},{1,0,0,1}},
+					{{1,1,0,0},{1,0,1,0},{1,0,0,1}}]}], 
+
+		"Domlib 14"]
+  };
+
+  res = Apply[ And, Join[ listResDomlib, listResZDomlib ] ];
+
+  If[ res,
+    Print["No problem occured during test of Domlib"],
+    Print["WARNING:  problems occured during test of Domlib"]
+  ];
+
+  SetDirectory[dir];
+  res
+]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestNormalization/TestNormalization.m

@@ -1,0 +1,114 @@
+BeginPackage["Alpha`TestNormalization`",
+  {"Alpha`","Alpha`Domlib`","Alpha`Tables`", "Alpha`Substitution`","Alpha`Normalization`"}];
+
+Begin["`Private`"];
+
+(* Written by P. Quinton, 4/1/98 *)
+(* Updated and checked on dec. 27, 2002 *)
+
+Module[ {dir, res, allTest, a, b, ast2, ast3, ast4, ast5}, 
+
+$testResult = True; 
+
+res=
+  {
+  testFunction[
+    load["Test1.alpha"];
+    addLocal["U","Z"];substituteInDef["Z","U"];
+    substituteInDef["Z","U"];
+    normalize[];
+    $result,<<RTest11,"Normalize 1"
+],
+  testFunction[
+    $result = <<RTest11;
+    substituteInDef["U","U"];normalize[];
+    (* $result>>RTest12; *)
+    $result,
+    <<RTest12,"Normalize 2"
+],
+  testFunction[
+    $result = <<RTest12;
+    normalize["Z"];
+    (* $result>>RTest13; *)
+    $result,<<RTest13,"Normalize 3"
+],
+  testFunction[
+    load["Test2.alpha"];
+    normalizeInCtxt[getDefinition["Z"],readDom["{i|i>=0}"]],
+    <<RTest2,"Normalize 4"
+  ],
+  testFunction[
+    load["Test3.alpha"];
+    simplifyInContext[];
+    (* $result>>RTest3; *)
+    $result,<<RTest3,"Normalize 5"
+  ],
+  testFunction[
+    load["tn1.alpha"];
+    simplifyInContext[],<<AST1,"Normalize 6"
+  ],
+  testFunction[
+    $result = <<AST1;
+    simplifyInContext[],<<AST1,"Normalize 7"
+  ],
+  testFunction[
+    load["tn1.alpha"];
+    ast2=substituteInDef["A","B"];
+    ast2,<<AST2,"Normalize 8"
+  ],
+  testFunction[
+    $result = <<AST2;
+    ast3=substituteInDef["A","A"];
+    ast3,<<AST3,"Normalize 9"]
+  ,
+  testFunction[
+    $result = <<AST3;
+    ast4=substituteInDef["A","A"];
+    ast4,<<AST4,"Normalize 10"]
+  ,
+  Print["Normalizing... (takes a little while)..."];
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestPipeControl/TestPipeControl.m

@@ -1,0 +1,100 @@
+(* ::Package:: *)
+
+BeginPackage["Alpha`TestPipeControl`",
+  {"Alpha`","Alpha`Domlib`","Alpha`PipeControl`","Alpha`Pipeline`",
+  "Alpha`Static`"}];
+
+Begin["`Private`"]
+
+Module[
+  {res,aaa,bbb,ccc,p1,p2,i,ii,ji,kk,j,k,finalRes},
+
+res = 
+{
+  testFunction[
+    const2al[{ii,ji,kk},
+    Alpha`ge[2ii+kk,2]],{1,2,0,1,-2},
+    "PipeControl 1"]
+,
+  testFunction[
+    const2mma[{i,j,k},{0, 2, 0, 1, -2}],
+    List[List[i, j, k],Equal[Plus[Times[2, i], k],2]],
+    "PipeControl 2"
+  ]
+,
+  testFunction[
+    const2mma[{"i","j","k"},{0, 2, 0, 1, -2}],
+    List[List[Alpha`Work`i, Alpha`Work`j, Alpha`Work`k],
+      Equal[Plus[Times[2, Alpha`Work`i], Alpha`Work`k],2]],
+    "PipeControl 3"
+  ]
+,
+  testFunction[dom2mma["{i,j,k|i=2;j>=k+2i}"],
+    {{Alpha`Work`i==2,Alpha`Work`j-Alpha`Work`k>=4},{Alpha`Work`i,Alpha`Work`j,
+    Alpha`Work`k}},
+    "PipeControl 4"]
+,
+  testFunction[dom2al[dom2mma["{i,j,k|i=2;j>=k+2i}"]],
+    readDom["{i, j, k|i = 2;j - k>= 4}"],"PipeControl 5"]
+,
+  testFunction[
+    aaa = readDom["{i,j,k|i=2;j>=k+2i}"];
+    ccc = domExplode[aaa];
+    ccc,{readDom["{i, j, k|i = 2}"], 
+    readDom["{i, j, k|(j - k) >= 4}"]},"PipeControl 6"]
+,
+  testFunction[domHalfSpaceQ[readDom["{i,j|i+j>=2}"]],True,
+    "PipeControl 7"]
+,
+  testFunction[domHalfSpaceQ[readDom["{i,j|i+j>=2;i=3}"]],False,
+  "PipeControl 8"]
+,
+  testFunction[domHalfSpaceQ[v],False,"PipeControl 9"]
+,
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestPipeline/TestPipelineMerged.m

@@ -1,0 +1,376 @@
+<<<<<<< TestPipeline.m
+dir = Directory[]
+ SetDirectory[Environment["MMALPHA"]<>"/tests/TestPipeline"]; 
+
+res = 
+{
+load["MV1.alpha"];
+Print["testing pipeall"];
+p1 = pipeall["C", "b.(i,j->j)", "BB.(i,j->i+1,j)"];
+If [p1===<<MV61.ast,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p2 = pipeall["C", "b.(i,j->j)", "BB.(i,j->i+1,j)", "{i,j| i>=-1}"];
+If [p2===<<MV62.ast,
+    Print["OK"];True,Print["Problem"];False]; True,
+load["MV1.alpha"];
+p3 = pipeall[{"C"}, "b.(i,j->j)", "BB.(i,j->i+1,j)"];
+If [p1===p3,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p4 = pipeall[{"C"}, "b.(i,j->j)", "BB.(i,j->i+1,j)", "{i,j| i>=-1}"];
+If [p4===p2,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p5 = pipeall[$result,"C", "b.(i,j->j)", "BB.(i,j->i+1,j)"];
+If [p1===p5 && p5=!=$result ,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p6 = pipeall[$result,"C", "b.(i,j->j)", "BB.(i,j->i+1,j)", "{i,j| i>=-1}"];
+If [p6===p2 || p6 =!= $result,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p5 = pipeall[$result,{"C"}, "b.(i,j->j)", "BB.(i,j->i+1,j)"];
+If [p1===p5 && p5=!=$result ,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
+p6 = pipeall[$result,{"C"}, "b.(i,j->j)", "BB.(i,j->i+1,j)", "{i,j| i>=-1}"];
+If [p6===p2,
+    Print["OK"];True,Print["Problem"];False],
+load["MV1.alpha"];
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestSchedule/TestSchedule.m

@@ -1,0 +1,388 @@
+BeginPackage["Alpha`TestSchedule`",
+  {"Alpha`","Alpha`Domlib`","Alpha`Schedule`","Alpha`ScheduleTools`",
+   "Alpha`VertexSchedule`","Alpha`FarkasSchedule`", "Alpha`Options`",
+   "Alpha`Tables`"}];
+
+Begin["`Private`"];
+
+(* Written by Tanguy Risset, updated by P. Quinton, Dec. 2008 *)
+Module[ {dir, res, scd1, scd2, ast, testResult},
+
+  $testResult = True; 
+
+  res = 
+    {
+      testFunction[
+        load["Test1.alpha"]; scd1 = schedule[];
+        testSched[ $result, $schedule ], True, "Schedule 1"
+      ]
+    ,
+      testFunction[
+        load["Test2.alpha"];
+        schedule[]; testSched[ $result,$schedule ],
+        True, "Schedule 2.1"
+      ]
+    ,
+      testFunction[
+        load["Test1.alpha"];
+        scd[]; testSched[ $result,$schedule ],
+        True, "Schedule 2.2"
+      ]
+    ,
+      testFunction[
+        load["Test3.alpha"];
+        schedule[]; testSched[ $result,$schedule ],
+        True, "Schedule 3.1"
+      ]
+    ,
+      testFunction[
+        load["Test3.alpha"];
+        scd[]; testSched[ $result,$schedule ],
+        True, "Schedule 3.2"
+      ]
+    ,
+      testFunction[
+        load["Test4.alpha"];
+        schedule[ scheduleType -> sameLinearPart]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule 4.1"
+      ]
+    ,
+      testFunction[
+        load["Test4.alpha"];
+        scd[ scheduleType -> sameLinearPart]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule 4.2"
+      ]
+    ,
+      testFunction[
+        load["TestAffine2.alpha"];
+        schedule[]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule affine 1.1"
+      ]
+    ,
+      testFunction[
+        load["TestUnion.alpha"];
+        schedule[ ratOrInt -> 0 ]; 
+        testSched[ $result,$schedule ],
+        True, "Schedule affine 2.1"
+      ]
+    ,
+      testFunction[
+        load["TestUnion.alpha"];
+        scd[ ]; 
[…]

### 2013-01-31 resources/mmalpha/mmalphaV2-1-0_binary/tests/TestWrite_C/test2.c

@@ -1,0 +1,58 @@
+/* system foobug2 */
+
+typedef struct { int value;
+		 int computed; } intvar;
+typedef struct { int value;
+		 int computed; } boolvar;
+typedef struct { double value;
+		 int computed; } realvar;
+double atof();
+int    atoi();
+
+/* --parameters */
+
+/* --input variables */
+
+/* --output variables */
+
+intvar a[4];
+int compute_a();
+
+/* --local variables */
+
+intvar A[4];
+int compute_A();
+
+/* --let equations */
+
+int compute_A(i)
+int i;
+{
+  intvar *tmp;
+  tmp = &(A[i-1]);
+  if (!tmp->computed) {
+     tmp->value = ( -i+1>=0 || i-3>=0 )  ? ( 0 ) : 
+         ( i-2==0 )  ? ( 1 ) : 
+         ( printf("? case statement error"), exit(-1));
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+int compute_a(i)
+int i;
+{
+  intvar *tmp;
+  tmp = &(a[i-1]);
+  if (!tmp->computed) {
+     tmp->value = compute_A(i);
+     tmp->computed = 1;}
+  return tmp->value;
+}
+
+int main()
+{
+   { int i ;
+   for (i=1; i<=4; i++) {
+   printf("a[%d]= %d\n", i, compute_a(i) );
+   }}
+}

### 2013-11-05 docs/report/conclusion.tex
Nits.

@@
 We have demonstrated that our implementation, DynaProg, is able to deal with simple (Smith-Waterman) and complex (Zuker and RNAfold) dynamic programming problems on sequences that are expressed by a grammar and an algebra.
-With this formalism, their expression is simplified and exploiting the dynamic programming results becomes possible thanks to systematic backtrack encoding that allows traces to be exchanged between algebrae.
+With this formalism, their expression is simplified and exploiting the dynamic programming results becomes possible thanks to systematic backtrack encoding that allows traces to be exchanged between algebras.
 This implementation is publicly available\footnote{\url{https://github.com/manojo/lamp-dp-mt/}} and can immediately be used in Scala projects.

### 2013-11-05 docs/report/design.tex
Nits.

@@
 
-Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebrae).
+Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebras).
 
@@
 LMS only supports a subset of Scala.
-Embedding both algebrae and grammars into the LMS framework would reduce the expressivity of algebrae.
+Embedding both algebras and grammars into the LMS framework would reduce the expressivity of algebras.
 Taking the example of matrix multiplication, the user would like to not only solve the dynamic problem on GPU, but also leverage the graphic card, the disk or the network to compute the matrices products.
@@
 For the generation of algebra functions, since CUDA types are restricted, they are representable in LMS.
-Hence we can leverage LMS for CUDA-compatible algebrae whereas regular Scala can be used for complex algebrae being executed on CPU exclusively.
+Hence we can leverage LMS for CUDA-compatible algebras whereas regular Scala can be used for complex algebras being executed on CPU exclusively.
 

### 2013-11-05 docs/report/implementation.tex
Nits.

@@
 \subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
-First define a signature that can fit both algebrae, then specify for each algebra the related functions.
+First define a signature that can fit both algebras, then specify for each algebra the related functions.
 In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general).
@@
 
-Finally, we describe the associated grammar and the programs that mixes the algebrae and the grammar.
+Finally, we describe the associated grammar and the programs that mixes the algebras and the grammar.
 Note that we need one instance of each pair of grammar and algebra.
@@
 
-Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebrae easily.
+Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebras easily.
 The signature needs to inherit either from {\tt Signature} or {\tt RNASignature}, would the RNA folding energies be needed.

### 2013-11-05 docs/report/introduction.tex
Nits.

@@
 
-A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, which allows the user to combine one grammar with different algebrae or vice versa.
+A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, which allows the user to combine one grammar with different algebras or vice versa.
 Because recurrence relations are expressed by a parsing grammar, ADP makes the candidate structure explicit and hides tabulation indices, thereby preventing potential errors.
@@
 
-Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebrae} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
+Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebras} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature.
 The concrete program instance combines the algebra with the grammar.
@@
 The research on ad-hoc implementation has focused on three kind of problems:\ul
-\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
+\item General problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
 \item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
